### 2019-04-08 biotime.Rmd
Scripts for prelim analysis of trait timeseries.

---
title: "BioTime"
author: "Kari Norman"
date: "4/1/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(here)
library(tidyverse)
library(lazyeval)
library(rdataretriever)

```

Import occurence data from Biotime
```{r}
#rdataretriever::install("biotime", connection = "csv", data_dir = here("data"))

data <- read_csv(here("data", "biotime_metadata.csv")) %>% filter(taxa %in% c("Birds", "Mammals")) %>%
  select(study_id, taxa) %>%
  left_join(read_csv(here("data", "biotime_query.csv")), by = "study_id")

```

Get mammal and trait data
```{r}
bird_trait <- read_csv(here("data/elton_traits/elton_traits_BirdFuncDat.csv"))
mamm_trait <- read.csv(here("data/elton_traits/elton_traits_MammFuncDat.csv"))
```

Start with trying to merge bird data via taxadb
```{r}
library(taxadb)

get_match_counts <- function(data, species_col){
  species_col <- lazyeval::as_name(species_col)
  data %>% 
    select(!!species_col) %>% 
    drop_na() %>%
    distinct() %>%
    mutate(
      gbif = get_ids(!!species_col, "gbif"),
      col = get_ids(!!species_col, "col" ),
      itis = get_ids(!!species_col, "itis"),
      ncbi = get_ids(!!species_col, "ncbi"),
      wd = get_ids(!!species_col, "wd"  ),
      iucn = get_ids(!!species_col, "iucn"),
      ott = get_ids(!!species_col, "ott" ) 
    ) %>%
    #select(!!call("-", species_col)) %>%
    purrr::map_dbl(function(x) sum(!is.na(x)))
}

##Counts show that OTT is the best authority for both data sets
get_match_counts(data %>% filter(taxa == "Birds"), "genus_species")
get_match_counts(bird_trait, "scientific")

```

Merge functional traits on OTT ids
```{r}
#Getting matches by scientific name 
bird_data <- data %>% 
  filter(taxa == "Birds") %>%
  select(genus_species) %>%
  drop_na() %>%
  distinct() %>%
  mutate(id = get_ids(genus_species, "ott")) %>%
  left_join(data, by = "genus_species")

bird_common <- bird_data %>%
  filter(is.na(id)) %>%
  select(genus_species) %>%
  distinct() %>%
  by_rank("Aves", "class", "ott")

sci_names <- by_name(data$genus_species, "ott")

comm_names <- by_common(data$genus, "ott")
```

### 2019-04-11 biotime.Rmd
merge data.

@@
 
-Start with trying to merge bird data via taxadb
+Check which authority would result in the most matches 
 ```{r}
@@
 get_match_counts(data %>% filter(taxa == "Birds"), "genus_species")
+get_match_counts(data, "genus_species")
 get_match_counts(bird_trait, "scientific")
-
+get_match_counts(mamm_trait, "scientific")
 ```
 
-Merge functional traits on OTT ids
+OTT is the best authority but doesn't have common names, so we use the next best ITIS
 ```{r}
 #Getting matches by scientific name 
-bird_data <- data %>% 
-  filter(taxa == "Birds") %>%
+biotime <- data %>% 
+  #filter(taxa == "Birds") %>%
   select(genus_species) %>%
@@
   distinct() %>%
-  mutate(id = get_ids(genus_species, "ott")) %>%
+  mutate(id = get_ids(genus_species, "itis")) %>%
   left_join(data, by = "genus_species")
 
-bird_common <- bird_data %>%
-  filter(is.na(id)) %>%
-  select(genus_species) %>%
-  distinct() %>%
-  by_rank("Aves", "class", "ott")
+#Getting matches by sci name and common name 
+biotime_ids <- by_name(unique(data$genus_species), "itis") %>%
+  bind_rows(by_common(unique(data$genus_species), "itis")) %>%
+  drop_na(acceptedNameUsageID)
 
-sci_names <- by_name(data$genus_species, "ott")
+#There is not a unique taxonID for each species, different pop's have different ID's for some marine mammals
+#Names with more than one match that need to be resolved manually
+#biotime_ids %>% group_by(sort) %>% filter(n()>1) %>% View()
 
-comm_names <- by_common(data$genus, "ott")
+biotime_data <- biotime_ids %>%
+  group_by(sort) %>%
+  filter(n() < 2) %>%
+  ungroup(sort) %>%
+  select(taxonID,input) %>%
+  left_join(data, by = c("input" = "genus_species"))
 ```

### 2019-04-16 biotime.Rmd
Merge biotime and trait data, check coverage.

@@
 
-OTT is the best authority but doesn't have common names, so we use the next best ITIS
+OTT is the best authority but doesn't have common names, so we use the next best ITIS. We'll start with BioTime data:
+
+##Biotime
+
 ```{r}
@@
   bind_rows(by_common(unique(data$genus_species), "itis")) %>%
+  filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
   drop_na(acceptedNameUsageID)
+```
 
-#There is not a unique taxonID for each species, different pop's have different ID's for some marine mammals
-#Names with more than one match that need to be resolved manually
-#biotime_ids %>% group_by(sort) %>% filter(n()>1) %>% View()
+Some names match to multiple acceptedUsage ID's, so we have to manually choose which one we want. In this case it's just the Blue-winged warbler
+```{r}
+unres <- biotime_ids %>%
+  distinct(acceptedNameUsageID, input, sort) %>%
+  group_by(sort) %>% 
+  filter(n()>1) %>% 
+  ungroup()
+
+unres
+```
+
+Add the unresolved ID to all the other ids and join with biotime data
+```{r}
+biotime_ids <- unres %>%
+left_join(biotime_ids) %>% #join observations back with all the columns
+  filter(taxonomicStatus == "accepted") %>% #want the accepted ID's
+  bind_rows(biotime_ids %>% filter(!sort %in% unres$sort)) #join to resolved ID's, excluding the observations we resolved manually
 
 biotime_data <- biotime_ids %>%
-  group_by(sort) %>%
-  filter(n() < 2) %>%
-  ungroup(sort) %>%
-  select(taxonID,input) %>%
-  left_join(data, by = c("input" = "genus_species"))
+  select(id = acceptedNameUsageID,input) %>%
+  distinct() %>%
+  right_join(data, by = c("input" = "genus_species")) %>%
+  rename(genus_species = input)
 ```
+
+##Elton Traits
+
+```{r}
+bird_ids <- by_name(unique(bird_trait$scientific), "itis") %>%
+  bind_rows(by_common(unique(bird_trait$scientific), "itis")) %>%
+  #filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
+  drop_na(acceptedNameUsageID)
+```
+
+check if all bird biotime data w/id's has an elton match. There are 54 species without a match
+```{r}
+biotime_data %>%
+  filter(taxa == "Birds") %>%
+  select(id) %>%
+  distinct() %>%
+  filter(!id %in% bird_ids$acceptedNameUsageID) %>%
+  dim()
+```
+
+and 605 entries in the species column without an ID
+```{r}
+biotime_data %>%
+  filter(taxa == "Birds") %>%
+  select(id, genus_species) %>%
+  distinct() %>%
+  filter(is.na(id)) %>%
+  dim()
+```

### 2019-04-16 biotime.Rmd
Get duplicate id's via function. Get complete trait data with id's.

@@
 ```{r}
-unres <- biotime_ids %>%
+
+get_dupe_ids <- function(data){
+  biotime_ids %>%
     distinct(acceptedNameUsageID, input, sort) %>%
@@
     ungroup()
+}
 
+unres <- get_dupe_ids(biotime_ids)
 unres 
@@
 biotime_ids <- unres %>%
-left_join(biotime_ids) %>% #join observations back with all the columns
-  filter(taxonomicStatus == "accepted") %>% #want the accepted ID's
+left_join(biotime_ids) %>% #join ID's back with all the columns
+  filter(taxonomicStatus == "accepted") %>% #want the accepted ID from the two Warbler options
   bind_rows(biotime_ids %>% filter(!sort %in% unres$sort)) #join to resolved ID's, excluding the observations we resolved manually
@@
 ##Elton Traits
-
+Get id's for the Elton Bird trait data
 ```{r}
@@
   drop_na(acceptedNameUsageID)
+```
+
+Are there any unresolved ids? Yes - just one, which is a synonym to two different accepted id's
+```{r}
+unres_trait <- get_dupe_ids(bird_ids) %>%
+  left_join(bird_ids)
+
+unres_trait
+
+#resolve to the desired ID (in this case the one with complete hierarchy data)
+bird_ids <- unres_trait %>% 
+  filter(!is.na(kingdom)) %>%
+  bind_rows(bird_ids %>% filter(!sort %in% unres_trait$sort))
+
+bird_trait_data <- bird_ids  %>%
+  select(acceptedNameUsageID, input) %>%
+  distinct() %>%
+  right_join(bird_trait, by = c("input" = "scientific")) %>%
+  rename(scientific = input)
 ```
@@
   distinct() %>%
-  filter(!id %in% bird_ids$acceptedNameUsageID) %>%
+  filter(!id %in% bird_trait_ids$acceptedNameUsageID) %>%
   dim()
@@
 
-and 605 entries in the species column without an ID
+and 605 species without an ID
 ```{r}
@@
 ```
+
+Are there any studies for which we have trait data for all species?
+```{r}
+bird_data <- biotime_data %>%
+  filter(taxa == "Birds") %>%
+  left_join(bird_ids)
+```

### 2019-04-16 biotime.Rmd
Join biotime and trait data.

@@
 bird_trait_data <- bird_ids  %>%
-  select(acceptedNameUsageID, input) %>%
+  select(id = acceptedNameUsageID, input) %>%
   distinct() %>%
@@
   filter(taxa == "Birds") %>%
-  left_join(bird_ids)
+  left_join(bird_trait_data) 
 ```

### 2019-04-17 biotime.Rmd
Don't join on NAs.
Exclude old taxonomy columns from trait data, preserve just original sci name and common name columns.

--- 2019-04-16 biotime.Rmd
@@
 biotime_data <- biotime_ids %>%
-  select(id = acceptedNameUsageID,input) %>%
+  select(id = acceptedNameUsageID,input, scientificName) %>%
   distinct() %>%
-  right_join(data, by = c("input" = "genus_species")) %>%
-  rename(genus_species = input)
+  right_join(data %>% select(-species, -genus), by = c("input" = "genus_species")) %>%
+  rename(originalSciName = input)
 ```
@@
 
+#only want one column for old name and one column for new name, these are the old ID columns to remove
+col_ex <- c("specid", "passnonpass", "iocorder", "blfamilylatin", "blfamilyenglish", "blfamsequid", "taxo")
+
 bird_trait_data <- bird_ids  %>%
-  select(acceptedNameUsageID, input) %>%
+  select(id = acceptedNameUsageID, input) %>%
   distinct() %>%
-  right_join(bird_trait, by = c("input" = "scientific")) %>%
-  rename(scientific = input)
+  right_join(bird_trait %>% select(.dots = -col_ex), by = c("input" = "scientific")) %>%
+  rename(originalSciName = input)
 ```
@@
 ```{r}
+#join trait and biotime data
 bird_data <- biotime_data %>%
-  filter(taxa == "Birds") %>%
-  left_join(bird_ids)
+  #filter(taxa == "Birds") %>%
+  rowid_to_column("row") %>%
+  left_join(bird_trait_data %>% 
+              select(), 
+            na_matches = "never")
+
+coverage <- bird_data %>%
+  group_by(study_id) %>%
+  summarise(perc = sum(!is.na(id))/n()) %>%
+  arrange(desc(perc))
+
+head(coverage)
 ```

### 2019-04-17 biotime.Rmd
Search for common names with the appropriate column, give preference to species that match on the scientific column. Modify the unresolved function to gropu by the appropriate id (when do different columns are used in by_, sort will not be the appropriate unique grouping column.)

@@
 
-get_dupe_ids <- function(data){
-  biotime_ids %>%
-    distinct(acceptedNameUsageID, input, sort) %>%
-    group_by(sort) %>% 
+get_dupe_ids <- function(data, orig_id){
+  orig_id <- lazyeval::as_name(orig_id)
+
+  data %>%
+    distinct(acceptedNameUsageID, input, !!orig_id) %>%
+    group_by(!!orig_id) %>% 
     filter(n()>1) %>% 
@@
 
-unres <- get_dupe_ids(biotime_ids)
+unres <- get_dupe_ids(biotime_ids, "sort")
 unres 
@@
 ```{r}
-bird_ids <- by_name(unique(bird_trait$scientific), "itis") %>%
-  bind_rows(by_common(unique(bird_trait$scientific), "itis")) %>%
+sci_names <-  by_name(unique(bird_trait$scientific), "itis") %>%
+  drop_na(acceptedNameUsageID) %>%
+  #get original specid so we can figure out how many don't have matches
+  left_join(bird_trait %>% select(specid, scientific), by = c("input" = "scientific")) %>%
+  mutate(match_type = "scientific")
+
+com_names <- bird_trait %>% 
+  filter(!specid %in% sci_names$specid) %>%
+  pull(english) %>%
+  unique() %>%
+  by_common("itis") %>%
+  #mutate(match_type = "common")
   #filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
-  drop_na(acceptedNameUsageID)
+  drop_na(acceptedNameUsageID) %>%
+  left_join(bird_trait %>% select(specid, english), by = c("input" = "english"))%>%
+  mutate(match_type = "common")
+
+#there are both sci and common names, only want common name 
+bird_ids <- bind_rows(sci_names, com_names)
 ```
@@
 ```{r}
-unres_trait <- get_dupe_ids(bird_ids) %>%
+unres_trait <- get_dupe_ids(bird_ids, "specid") %>%
   left_join(bird_ids)

### 2019-04-18 biotime.Rmd
Join trait id's back to trait data, taking into account whether the id was matched on a scientific or common name.

@@
   right_join(data %>% select(-species, -genus), by = c("input" = "genus_species")) %>%
-  rename(originalSciName = input)
+  rename(matchingName = input)
 ```
@@
 #only want one column for old name and one column for new name, these are the old ID columns to remove
-col_ex <- c("specid", "passnonpass", "iocorder", "blfamilylatin", "blfamilyenglish", "blfamsequid", "taxo")
+col_ex <- c("passnonpass", "iocorder", "blfamilylatin", "blfamilyenglish", "blfamsequid", "taxo")
 
-bird_trait_data <- bird_ids  %>%
-  select(id = acceptedNameUsageID, input) %>%
+#join on scientific and common
+sci_matches <- bird_ids  %>%
+  filter(match_type == "scientific") %>%
+  select(id = acceptedNameUsageID, input, scientificName) %>%
   distinct() %>%
-  right_join(bird_trait %>% select(.dots = -col_ex), by = c("input" = "scientific")) %>%
-  rename(originalSciName = input)
+  right_join(bird_trait %>% select(.dots = -col_ex), by = c("input" = "scientific"), na_matches = "never") %>%
+  drop_na(id) %>%
+  distinct() %>%
+  select(-english)
+
+comm_matches <- bird_ids  %>%
+  filter(match_type == "common") %>%
+  select(id = acceptedNameUsageID, input, scientificName) %>%
+  distinct() %>%
+  right_join(bird_trait %>% filter(!specid %in% sci_matches$specid) %>% select(.dots = -col_ex), by = c("input" = "english"), na_matches = "never") %>%
+  distinct() %>%
+  rename(matchingName = input) %>%
+  select(-scientific)
+
+#there are more rows after ID data is joined because for species that matched on common names there may be more than one scientificName 
+bird_trait_data <- bind_rows(sci_matches, comm_matches)
 ```
@@
   distinct() %>%
-  filter(!id %in% bird_trait_ids$acceptedNameUsageID) %>%
+  filter(!id %in% bird_trait_data$id) %>%
   dim()

### 2019-04-18 biotime.Rmd
Propogate changed column names to later code.

@@
 
-check if all bird biotime data w/id's has an elton match. There are 54 species without a match
+check if all bird biotime data w/id's has an elton match. There are 38 species without a match
 ```{r}
@@
   filter(taxa == "Birds") %>%
-  select(id, genus_species) %>%
+  select(id, matchingName) %>%
   distinct() %>%

### 2019-04-24 biotime.Rmd
Make id matching for trait data a function.

@@
 Get id's for the Elton Bird trait data
+
+```{r}
+get_trait_ids <- function(data, provider, id, common = TRUE){
+  #id <- lazyeval::as_name(id)
+  sci_match <-  by_name(unique(data$scientific), provider) %>%
+    drop_na(acceptedNameUsageID) %>%
+    #get original specid so we can figure out ho w many don't have matches
+    left_join(data %>% select_(id, "scientific"), by = c("input" = "scientific")) %>%
+    mutate(match_type = "scientific")
+
+  if(common){
+    com_names <- data %>% 
+      #the id argument should be inplace of specid, but it doesn't work for some reason
+      filter(!specid %in% sci_match$specid) %>%
+      pull(english) %>%
+      unique() %>%
+      by_common("itis") %>%
+      #mutate(match_type = "common")
+      #filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
+      drop_na(acceptedNameUsageID) %>%
+      left_join(bird_trait %>% select_(id, "english"), by = c("input" = "english"))%>%
+      mutate(match_type = "common")
+
+    #there are both sci and common names, only want common name 
+    return(bind_rows(sci_match, com_names))
+  }else{ return(sci_match)}
+}
+
+test <- get_trait_ids(bird_trait, "itis", "specid")
+test_mamm <- get_trait_ids(mamm_trait, "itis", "msw3_id", FALSE)
+
+```
+
 ```{r}
@@
 
-and 605 species without an ID
+and 605 bird species without an ID, 657 total species
 ```{r}
 biotime_data %>%
-  filter(taxa == "Birds") %>%
+  #filter(taxa == "Birds") %>%
   select(id, matchingName) %>%
@@
   #filter(taxa == "Birds") %>%
-  rowid_to_column("row") %>%
-  left_join(bird_trait_data %>% 
-              select(), 
+  left_join(bird_trait_data, by = "id",
             na_matches = "never")
@@
   group_by(study_id) %>%
-  summarise(perc = sum(!is.na(id))/n()) %>%
+  summarise(perc = sum(!is.na(specid))/n()) %>%
   arrange(desc(perc))

### 2019-04-24 biotime.Rmd
Find unresolved id's for mammal trait data and join back to original data.

@@
 
-test <- get_trait_ids(bird_trait, "itis", "specid")
-test_mamm <- get_trait_ids(mamm_trait, "itis", "msw3_id", FALSE)
+bird_ids <- get_trait_ids(bird_trait, "itis", "specid")
+mammal_ids <- get_trait_ids(mamm_trait, "itis", "msw3_id", FALSE)
 
@@
 
-```{r}
-sci_names <-  by_name(unique(bird_trait$scientific), "itis") %>%
-  drop_na(acceptedNameUsageID) %>%
-  #get original specid so we can figure out how many don't have matches
-  left_join(bird_trait %>% select(specid, scientific), by = c("input" = "scientific")) %>%
-  mutate(match_type = "scientific")
-
-com_names <- bird_trait %>% 
-  filter(!specid %in% sci_names$specid) %>%
-  pull(english) %>%
-  unique() %>%
-  by_common("itis") %>%
-  #mutate(match_type = "common")
-  #filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
-  drop_na(acceptedNameUsageID) %>%
-  left_join(bird_trait %>% select(specid, english), by = c("input" = "english"))%>%
-  mutate(match_type = "common")
-
-#there are both sci and common names, only want common name 
-bird_ids <- bind_rows(sci_names, com_names)
-```
-
-Are there any unresolved ids? Yes - just one, which is a synonym to two different accepted id's
+Are there any unresolved ids for birds? Yes - just one, which is a synonym to two different accepted id's
 ```{r}
@@
   bind_rows(bird_ids %>% filter(!sort %in% unres_trait$sort))
+```
 
+What about for mammals? Yes - also a synonym to two different accepted id's. Further research shows that the Callospermophilus saturatus uses the subgenus, so we'll keep the name with the genus 
+```{r}
+unres_mamm_trait <- get_dupe_ids(mammal_ids, "msw3_id") %>% 
+  left_join(mammal_ids)
+
+unres_mamm_trait
+
+mammal_ids <- unres_mamm_trait %>% 
+  filter(acceptedNameUsageID == "ITIS:632452") %>%
+  bind_rows(mammal_ids %>% filter(!sort %in% unres_mamm_trait$sort))
+```
+
+Join bird ids back to original trait data (to get data that didn't have an id match)
+```{r}
 #only want one column for old name and one column for new name, these are the old ID columns to remove
@@
 bird_trait_data <- bind_rows(sci_matches, comm_matches)
+```
+
+Do the same for mammal trait data
+```{r}
+mamm_trait_data <- mammal_ids %>%
+  select(id = acceptedNameUsageID, input, scientificName) %>%
+  distinct() %>%
+  right_join(mamm_trait %>% select(-mswfamilylatin), by = c("input" = "scientific"), na_matches = "never") %>%
+  drop_na(id) %>%
+  distinct() 
 ```

### 2019-04-25 biotime.Rmd
Tried to use synonyms() to check for id's with other providers, but it doesn't return the input column so you can't join back to original data.

@@
   bind_rows(by_common(unique(data$genus_species), "itis")) %>%
-  filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
+  #filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
   drop_na(acceptedNameUsageID)
@@
   rename(matchingName = input)
+```
+
+Do any unmatched species have matches with other providers?
+```{r}
+get_match_counts(biotime_data %>% filter(is.na(id)), "matchingName")
+
+#match all the un-ID'd names to other providers, check if synonyms given by those providers match known ITIS species 
+syn_res <- bind_rows(synonyms(biotime_data %>% filter(is.na(id)) %>% pull(matchingName) %>% unique(), "ott"),
+                  synonyms(biotime_data %>% filter(is.na(id)) %>% pull(matchingName) %>% unique(), "gbif"),
+                  synonyms(biotime_data %>% filter(is.na(id)) %>% pull(matchingName) %>% unique(), "col"),
+                  synonyms(biotime_data %>% filter(is.na(id)) %>% pull(matchingName) %>% unique(), "ncbi"),
+                  synonyms(biotime_data %>% filter(is.na(id)) %>% pull(matchingName) %>% unique(), "wd"),
+                  synonyms(biotime_data %>% filter(is.na(id)) %>% pull(matchingName) %>% unique(), "iucn")
+                  ) %>%
+  pull(synonym) %>% 
+  na.omit() %>% 
+  unique() %>%
+  by_name("itis") %>%
+  select(id = acceptedNameUsageID, input, scientificName) %>% 
+  drop_na(id) %>% 
+  left_join(data %>% select(-species, -genus), by = c("input" = "genus_species")) %>%
+  rename(matchingName = input)
+
+#Below add's more rows, should just replace them 
+biotime_data <- biotime_data %>% filter(!matchingName %in% syn_res$matchingName) %>%
+  bind_rows(syn_res)
+
+#These two should be equal, why aren't they??
+dim(syn_res)  
+biotime_data %>% filter(matchingName %in% syn_res$matchingName) %>% dim()
+biotime_data %>% filter(is.na(id)) %>% dim()
 ```
@@
 
-Join bird ids back to original trait data (to get data that didn't have an id match)
+Join bird ids back to original trait data
 ```{r}
@@
   select(id = acceptedNameUsageID, input, scientificName) %>%
-  distinct() %>%
   right_join(mamm_trait %>% select(-mswfamilylatin), by = c("input" = "scientific"), na_matches = "never") %>%
@@
 
-check if all bird biotime data w/id's has an elton match. There are 38 species without a match
+#Checking Coverage of ID's and trait data -> biotime
+
+check if all bird & mammal biotime data w/id's has an elton match. There are 258 species without a match (38 are birds).
 ```{r}
 biotime_data %>%
-  filter(taxa == "Birds") %>%
+  #filter(taxa == "Birds") %>%
   select(id) %>%

### 2019-05-03 data-raw/biotime.Rmd
Sort out package structure, move raw data to installed files.

--- 2019-04-24 biotime.Rmd
@@
 
-data <- read_csv(here("data", "biotime_metadata.csv")) %>% filter(taxa %in% c("Birds", "Mammals")) %>%
+data <- read_csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS")) %>% 
+  filter(taxa %in% c("Birds", "Mammals")) %>%
   select(study_id, taxa) %>%
-  left_join(read_csv(here("data", "biotime_query.csv")), by = "study_id")
-
+  left_join(read_csv(system.file("extdata", "biotime/biotime_query.csv", package = "biodivTS")), by = "study_id")
 ```
@@
 ```{r}
-bird_trait <- read_csv(here("data/elton_traits/elton_traits_BirdFuncDat.csv"))
-mamm_trait <- read.csv(here("data/elton_traits/elton_traits_MammFuncDat.csv"))
+bird_trait <- read_csv(system.file("extdata", "elton_traits/elton_traits_BirdFuncDat.csv", package = "biodivTS"))
+mamm_trait <- read.csv(system.file("extdata", "elton_traits/elton_traits_MammFuncDat.csv", package = "biodivTS"))
 ```

### 2019-05-03 data-raw/amphibio.Rmd
Move helpful functions out of biotime script so they can be used across processing scripts.

---
title: "AmphibBio"
author: "Kari Norman"
date: "5/3/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)

source(here::here("data-raw", "helper_functions.R"))
```

```{r}
#original data from https://figshare.com/articles/Oliveira_et_al_AmphiBIO_v1/4644424

amph <- read_csv(system.file("extdata", "amphibio/AmphiBIO_v1.csv", package = "biodivTS")) %>%
  unite(scientificName, Genus, Species)

amph_ids <- by_name(amph$Species, "itis") %>%
  filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
  drop_na(acceptedNameUsageID)
```

### 2019-05-03 data-raw/biotime.Rmd
Move helpful functions out of biotime script so they can be used across processing scripts.

@@
 library(rdataretriever)
+library(taxadb)
 
+source(here::here("data-raw", "helper_functions.R"))
 ```
@@
 ```{r}
-library(taxadb)
-
-get_match_counts <- function(data, species_col){
-  species_col <- lazyeval::as_name(species_col)
-  data %>% 
-    select(!!species_col) %>% 
-    drop_na() %>%
-    distinct() %>%
-    mutate(
-      gbif = get_ids(!!species_col, "gbif"),
-      col = get_ids(!!species_col, "col" ),
-      itis = get_ids(!!species_col, "itis"),
-      ncbi = get_ids(!!species_col, "ncbi"),
-      wd = get_ids(!!species_col, "wd"  ),
-      iucn = get_ids(!!species_col, "iucn"),
-      ott = get_ids(!!species_col, "ott" ) 
-    ) %>%
-    #select(!!call("-", species_col)) %>%
-    purrr::map_dbl(function(x) sum(!is.na(x)))
-}
-
 ##Counts show that OTT is the best authority for both data sets
@@
 ```{r}
-
-get_dupe_ids <- function(data, orig_id){
-  orig_id <- lazyeval::as_name(orig_id)
-
-  data %>%
-    distinct(acceptedNameUsageID, input, !!orig_id) %>%
-    group_by(!!orig_id) %>% 
-    filter(n()>1) %>% 
-    ungroup()
-}
-
 unres <- get_dupe_ids(biotime_ids, "sort")

### 2019-05-13 data-raw/amphibio.Rmd
Match amphibian traits to ID's

@@
 
-amph_ids <- by_name(amph$Species, "itis") %>%
-  filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
-  drop_na(acceptedNameUsageID)
+amph_ids <- amph %>%
+  mutate(id = get_ids(Species, "itis")) %>%
+  filter(!is.na(id))
+
+# Using by_ doesn't give any additional ids
+# amph_ids <- by_name(amph$Species, "itis") %>%
+#   filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
+#   drop_na(acceptedNameUsageID)
+
 ```

### 2019-05-15 biotime.Rmd
clean up synonym resolutions, fix the join.

@@
 
+unmatched <- biotime_data %>% filter(is.na(id)) %>% pull(matchingName) %>% unique()
+providers <- c("ott", "gbif", "col", "ncbi", "wd", "iucn")
+
 #match all the un-ID'd names to other providers, check if synonyms given by those providers match known ITIS species 
-syn_res <- bind_rows(synonyms(biotime_data %>% filter(is.na(id)) %>% pull(matchingName) %>% unique(), "ott"),
-                  synonyms(biotime_data %>% filter(is.na(id)) %>% pull(matchingName) %>% unique(), "gbif"),
-                  synonyms(biotime_data %>% filter(is.na(id)) %>% pull(matchingName) %>% unique(), "col"),
-                  synonyms(biotime_data %>% filter(is.na(id)) %>% pull(matchingName) %>% unique(), "ncbi"),
-                  synonyms(biotime_data %>% filter(is.na(id)) %>% pull(matchingName) %>% unique(), "wd"),
-                  synonyms(biotime_data %>% filter(is.na(id)) %>% pull(matchingName) %>% unique(), "iucn")
-                  ) %>%
-  pull(synonym) %>% 
-  na.omit() %>% 
-  unique() %>%
-  by_name("itis") %>%
+alt_names <- map_df(providers, function(name) synonyms(unmatched, name)) %>%
+  drop_na(acceptedNameUsageID)
+
+syn_res <- bind_rows(by_name(na.omit(unique(alt_names$synonym)),"itis"),
+                     by_name(na.omit(unique(alt_names$acceptedNameUsage)),"itis")) %>%
   select(id = acceptedNameUsageID, input, scientificName) %>% 
   drop_na(id) %>% 
-  left_join(data %>% select(-species, -genus), by = c("input" = "genus_species")) %>%
+  distinct() %>%
+  inner_join(data %>% select(-species, -genus), by = c("input" = "genus_species"), na_matches =  "never") %>%
   rename(matchingName = input)
@@
 
-#These two should be equal, why aren't they??
-dim(syn_res)  
-biotime_data %>% filter(matchingName %in% syn_res$matchingName) %>% dim()
-biotime_data %>% filter(is.na(id)) %>% dim()
+# #These two should be equal, why aren't they??
+# dim(syn_res)  
+# biotime_data %>% filter(matchingName %in% syn_res$matchingName) %>% dim()
+# biotime_data %>% filter(is.na(id)) %>% dim()
 ```

### 2019-05-16 biotime.Rmd
Clean up checking alt matches back to itis, preserve original input column to join back to data.

@@
 alt_names <- map_df(providers, function(name) synonyms(unmatched, name)) %>%
-  drop_na(acceptedNameUsageID)
+  drop_na(acceptedNameUsageID) %>%
+  #new matches can be in either the synonym or acceptedNameUsage column, so let's combine them
+  gather(type, altProviderName, synonym, acceptedNameUsage) %>%
+  select(altProviderName, input) %>% 
+  distinct()
 
-syn_res <- bind_rows(by_name(na.omit(unique(alt_names$synonym)),"itis"),
-                     by_name(na.omit(unique(alt_names$acceptedNameUsage)),"itis")) %>%
+#match accepted names from 
+syn_res <- by_name(na.omit(unique(alt_names$altProviderName)),"itis") %>%
+                       rename(altProviderName = input) %>%
+                       left_join(alt_names) %>%
   select(id = acceptedNameUsageID, input, scientificName) %>% 
@@
   distinct() %>%
-  inner_join(data %>% select(-species, -genus), by = c("input" = "genus_species"), na_matches =  "never") %>%
+  left_join(data %>% select(-species, -genus), by = c("input" = "genus_species"), na_matches =  "never") %>%
   rename(matchingName = input)

### 2019-05-20 biotime.Rmd
Deal with duplicate matches from the alternate provider matching.

@@
 ```{r}
-get_match_counts(biotime_data %>% filter(is.na(id)), "matchingName")
+#get_match_counts(biotime_data %>% filter(is.na(id)), "matchingName")
 
@@
 syn_res <- by_name(na.omit(unique(alt_names$altProviderName)),"itis") %>%
+  drop_na(acceptedNameUsageID) %>% 
   rename(altProviderName = input) %>%
-                       left_join(alt_names) %>%
+  left_join(alt_names, na_matches = "never")
+
+#some names have multiple matches (same ID but multiple scinames), have to pick one
+## in this case all the duplicates have an accepted and a synonym, but there are some single matches that are just synonyms
+## so we have to find duplicate ID's and then filter
+alt_dupes <- syn_res %>% 
+  group_by(acceptedNameUsageID, input) %>%
+  filter(n() > 1)
+
+#filter duplicates for accepted, and replace old entries 
+syn_res <- alt_dupes %>% 
+  filter(taxonomicStatus == "accepted") %>% 
+  bind_rows(syn_res %>% 
+              filter(!acceptedNameUsageID %in% alt_dupes$acceptedNameUsageID)) %>%
+#then finish clean up 
   select(id = acceptedNameUsageID, input, scientificName) %>% 
-  drop_na(id) %>% 
   distinct() %>%

### 2019-05-20 biotime.Rmd
Actually get rid of all the duplicates! Only want one id per input name.

--- 2019-05-16 biotime.Rmd
@@
 syn_res <- by_name(na.omit(unique(alt_names$altProviderName)),"itis") %>%
+  drop_na(acceptedNameUsageID) %>% 
   rename(altProviderName = input) %>%
-                       left_join(alt_names) %>%
+  left_join(alt_names, na_matches = "never")
+
+#some names have multiple matches (same ID but multiple scinames), have to pick one
+## in this case all the duplicates have an accepted and a synonym, but there are some single matches that are just synonyms
+## so we have to find duplicate ID's and then filter
+alt_dupes <- syn_res %>% 
+  group_by(input) %>%
+  filter(n() > 1)
+
+#filter duplicates for accepted, and replace old entries 
+syn_res <- alt_dupes %>% 
+  filter(taxonRank == "species", taxonomicStatus == "accepted") %>% 
+  bind_rows(syn_res %>% 
+              filter(!acceptedNameUsageID %in% alt_dupes$acceptedNameUsageID)) %>%
+#then finish clean up 
   select(id = acceptedNameUsageID, input, scientificName) %>% 
-  drop_na(id) %>% 
   distinct() %>%
   left_join(data %>% select(-species, -genus), by = c("input" = "genus_species"), na_matches =  "never") %>%
-  rename(matchingName = input)
+  rename(matchingName = input) %>%
+  ungroup()
 
@@
 
-# #These two should be equal, why aren't they??
-# dim(syn_res)  
-# biotime_data %>% filter(matchingName %in% syn_res$matchingName) %>% dim()
-# biotime_data %>% filter(is.na(id)) %>% dim()
 ```

### 2019-05-20 data-raw/biotime.Rmd

@@
   rename(matchingName = input)
+```
+
+Do any unmatched species have matches with other providers?
+```{r}
+#get_match_counts(biotime_data %>% filter(is.na(id)), "matchingName")
+
+unmatched <- biotime_data %>% filter(is.na(id)) %>% pull(matchingName) %>% unique()
+providers <- c("ott", "gbif", "col", "ncbi", "wd", "iucn")
+
+#match all the un-ID'd names to other providers, check if synonyms given by those providers match known ITIS species 
+alt_names <- map_df(providers, function(name) synonyms(unmatched, name)) %>%
+  drop_na(acceptedNameUsageID) %>%
+  #new matches can be in either the synonym or acceptedNameUsage column, so let's combine them
+  gather(type, altProviderName, synonym, acceptedNameUsage) %>%
+  select(altProviderName, input) %>% 
+  distinct()
+
+#match accepted names from 
+syn_res <- by_name(na.omit(unique(alt_names$altProviderName)),"itis") %>%
+  drop_na(acceptedNameUsageID) %>% 
+  rename(altProviderName = input) %>%
+  left_join(alt_names, na_matches = "never")
+
+#some names have multiple matches (same ID but multiple scinames), have to pick one
+## in this case all the duplicates have an accepted and a synonym, but there are some single matches that are just synonyms
+## so we have to find duplicate ID's and then filter
+alt_dupes <- syn_res %>% 
+  group_by(input) %>%
+  filter(n() > 1)
+
+#filter duplicates for accepted, and replace old entries 
+syn_res <- alt_dupes %>% 
+  filter(taxonRank == "species", taxonomicStatus == "accepted") %>% 
+  bind_rows(syn_res %>% 
+              filter(!acceptedNameUsageID %in% alt_dupes$acceptedNameUsageID)) %>%
+#then finish clean up 
+  select(id = acceptedNameUsageID, input, scientificName) %>% 
+  distinct() %>%
+  left_join(data %>% select(-species, -genus), by = c("input" = "genus_species"), na_matches =  "never") %>%
+  rename(matchingName = input) %>%
+  ungroup()
+
+#Below add's more rows, should just replace them 
+biotime_data <- biotime_data %>% filter(!matchingName %in% syn_res$matchingName) %>%
+  bind_rows(syn_res)
+
 ```

### 2019-05-20 analysis/coverage_check.Rmd
Split biotime and elton id labeling into two separate scripts, move coverage checking into analysis folder. Save all the data products in appropriate places.

--- 2019-05-03 data-raw/biotime.Rmd
@@
 ---
-title: "BioTime"
+title: "Check Coverage"
 author: "Kari Norman"
-date: "4/1/2019"
+date: "5/20/2019"
 output: html_document
@@
 ```{r}
-library(here)
-library(tidyverse)
-library(lazyeval)
-library(rdataretriever)
-library(taxadb)
-
-source(here::here("data-raw", "helper_functions.R"))
+load(here::here("data", "biotime_data.rda"))
+load(here::here("data", "elton_mamm.rda"))
+load(here::here("data", "elton_bird.rda"))
 ```
 
-Import occurence data from Biotime
-```{r}
-#rdataretriever::install("biotime", connection = "csv", data_dir = here("data"))
+#Checking Coverage of ID's and trait data -> biotime
 
-data <- read_csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS")) %>% 
-  filter(taxa %in% c("Birds", "Mammals")) %>%
-  select(study_id, taxa) %>%
-  left_join(read_csv(system.file("extdata", "biotime/biotime_query.csv", package = "biodivTS")), by = "study_id")
-```
-
-Get mammal and trait data
-```{r}
-bird_trait <- read_csv(system.file("extdata", "elton_traits/elton_traits_BirdFuncDat.csv", package = "biodivTS"))
-mamm_trait <- read.csv(system.file("extdata", "elton_traits/elton_traits_MammFuncDat.csv", package = "biodivTS"))
-```
-
-Check which authority would result in the most matches 
-```{r}
-##Counts show that OTT is the best authority for both data sets
-get_match_counts(data %>% filter(taxa == "Birds"), "genus_species")
-get_match_counts(data, "genus_species")
-get_match_counts(bird_trait, "scientific")
-get_match_counts(mamm_trait, "scientific")
-```
-
-OTT is the best authority but doesn't have common names, so we use the next best ITIS. We'll start with BioTime data:
-
-##Biotime
-
-```{r}
-#Getting matches by scientific name 
-biotime <- data %>% 
-  #filter(taxa == "Birds") %>%
-  select(genus_species) %>%
-  drop_na() %>%
-  distinct() %>%
-  mutate(id = get_ids(genus_species, "itis")) %>%
-  left_join(data, by = "genus_species")
-
-#Getting matches by sci name and common name 
-biotime_ids <- by_name(unique(data$genus_species), "itis") %>%
-  bind_rows(by_common(unique(data$genus_species), "itis")) %>%
-  filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
-  drop_na(acceptedNameUsageID)
-```
-
-Some names match to multiple acceptedUsage ID's, so we have to manually choose which one we want. In this case it's just the Blue-winged warbler
-```{r}
-unres <- get_dupe_ids(biotime_ids, "sort")
-unres 
-```
-
-Add the unresolved ID to all the other ids and join with biotime data
-```{r}
-biotime_ids <- unres %>%
-left_join(biotime_ids) %>% #join ID's back with all the columns
-  filter(taxonomicStatus == "accepted") %>% #want the accepted ID from the two Warbler options
-  bind_rows(biotime_ids %>% filter(!sort %in% unres$sort)) #join to resolved ID's, excluding the observations we resolved manually
-
-biotime_data <- biotime_ids %>%
-  select(id = acceptedNameUsageID,input, scientificName) %>%
-  distinct() %>%
-  right_join(data %>% select(-species, -genus), by = c("input" = "genus_species")) %>%
-  rename(matchingName = input)
-```
-
-##Elton Traits
-Get id's for the Elton Bird trait data
-
-```{r}
-get_trait_ids <- function(data, provider, id, common = TRUE){
-  #id <- lazyeval::as_name(id)
-  sci_match <-  by_name(unique(data$scientific), provider) %>%
-    drop_na(acceptedNameUsageID) %>%
-    #get original specid so we can figure out ho w many don't have matches
-    left_join(data %>% select_(id, "scientific"), by = c("input" = "scientific")) %>%
-    mutate(match_type = "scientific")
-
-  if(common){
-    com_names <- data %>% 
-      #the id argument should be inplace of specid, but it doesn't work for some reason
-      filter(!specid %in% sci_match$specid) %>%
-      pull(english) %>%
-      unique() %>%
-      by_common("itis") %>%
-      #mutate(match_type = "common")
-      #filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
-      drop_na(acceptedNameUsageID) %>%
-      left_join(bird_trait %>% select_(id, "english"), by = c("input" = "english"))%>%
-      mutate(match_type = "common")
-
-    #there are both sci and common names, only want common name 
-    return(bind_rows(sci_match, com_names))
-  }else{ return(sci_match)}
-}
-
-bird_ids <- get_trait_ids(bird_trait, "itis", "specid")
-mammal_ids <- get_trait_ids(mamm_trait, "itis", "msw3_id", FALSE)
-
-```
-
-Are there any unresolved ids for birds? Yes - just one, which is a synonym to two different accepted id's
-```{r}
-unres_trait <- get_dupe_ids(bird_ids, "specid") %>%
-  left_join(bird_ids)
-
-unres_trait
-
-#resolve to the desired ID (in this case the one with complete hierarchy data)
-bird_ids <- unres_trait %>% 
-  filter(!is.na(kingdom)) %>%
-  bind_rows(bird_ids %>% filter(!sort %in% unres_trait$sort))
-```
-
-What about for mammals? Yes - also a synonym to two different accepted id's. Further research shows that the Callospermophilus saturatus uses the subgenus, so we'll keep the name with the genus 
-```{r}
-unres_mamm_trait <- get_dupe_ids(mammal_ids, "msw3_id") %>% 
-  left_join(mammal_ids)
-
-unres_mamm_trait
-
-mammal_ids <- unres_mamm_trait %>% 
-  filter(acceptedNameUsageID == "ITIS:632452") %>%
-  bind_rows(mammal_ids %>% filter(!sort %in% unres_mamm_trait$sort))
-```
-
-Join bird ids back to original trait data (to get data that didn't have an id match)
-```{r}
-#only want one column for old name and one column for new name, these are the old ID columns to remove
-col_ex <- c("passnonpass", "iocorder", "blfamilylatin", "blfamilyenglish", "blfamsequid", "taxo")
-
-#join on scientific and common
-sci_matches <- bird_ids  %>%
-  filter(match_type == "scientific") %>%
-  select(id = acceptedNameUsageID, input, scientificName) %>%
-  distinct() %>%
-  right_join(bird_trait %>% select(.dots = -col_ex), by = c("input" = "scientific"), na_matches = "never") %>%
-  drop_na(id) %>%
-  distinct() %>%
-  select(-english)
-
-comm_matches <- bird_ids  %>%
-  filter(match_type == "common") %>%
-  select(id = acceptedNameUsageID, input, scientificName) %>%
-  distinct() %>%
-  right_join(bird_trait %>% filter(!specid %in% sci_matches$specid) %>% select(.dots = -col_ex), by = c("input" = "english"), na_matches = "never") %>%
-  distinct() %>%
-  rename(matchingName = input) %>%
-  select(-scientific)
-
-#there are more rows after ID data is joined because for species that matched on common names there may be more than one scientificName 
-bird_trait_data <- bind_rows(sci_matches, comm_matches)
-```
-
-Do the same for mammal trait data
-```{r}
-mamm_trait_data <- mammal_ids %>%
-  select(id = acceptedNameUsageID, input, scientificName) %>%
-  distinct() %>%
-  right_join(mamm_trait %>% select(-mswfamilylatin), by = c("input" = "scientific"), na_matches = "never") %>%
-  drop_na(id) %>%
-  distinct() 
-```
-
-check if all bird biotime data w/id's has an elton match. There are 38 species without a match
+check if all bird & mammal biotime data w/id's has an elton match. There are 258 species without a match (38 are birds).
 ```{r}
 biotime_data %>%
-  filter(taxa == "Birds") %>%
+  #filter(taxa == "Birds") %>%
   select(id) %>%
   distinct() %>%
-  filter(!id %in% bird_trait_data$id) %>%
+  filter(!id %in% elton_bird$id) %>%
   dim()
@@
   #filter(taxa == "Birds") %>%
-  left_join(bird_trait_data, by = "id",
+  left_join(elton_bird, by = "id",
             na_matches = "never")

### 2019-05-20 data-raw/biotime.Rmd
Split biotime and elton id labeling into two separate scripts, move coverage checking into analysis folder. Save all the data products in appropriate places.

@@
 
-Get mammal and trait data
-```{r}
-bird_trait <- read_csv(system.file("extdata", "elton_traits/elton_traits_BirdFuncDat.csv", package = "biodivTS"))
-mamm_trait <- read.csv(system.file("extdata", "elton_traits/elton_traits_MammFuncDat.csv", package = "biodivTS"))
-```
-
 Check which authority would result in the most matches 
@@
 get_match_counts(data, "genus_species")
-get_match_counts(bird_trait, "scientific")
-get_match_counts(mamm_trait, "scientific")
 ```
@@
 
-##Elton Traits
-Get id's for the Elton Bird trait data
-
 ```{r}
-get_trait_ids <- function(data, provider, id, common = TRUE){
-  #id <- lazyeval::as_name(id)
-  sci_match <-  by_name(unique(data$scientific), provider) %>%
-    drop_na(acceptedNameUsageID) %>%
-    #get original specid so we can figure out ho w many don't have matches
-    left_join(data %>% select_(id, "scientific"), by = c("input" = "scientific")) %>%
-    mutate(match_type = "scientific")
-
-  if(common){
-    com_names <- data %>% 
-      #the id argument should be inplace of specid, but it doesn't work for some reason
-      filter(!specid %in% sci_match$specid) %>%
-      pull(english) %>%
-      unique() %>%
-      by_common("itis") %>%
-      #mutate(match_type = "common")
-      #filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
-      drop_na(acceptedNameUsageID) %>%
-      left_join(bird_trait %>% select_(id, "english"), by = c("input" = "english"))%>%
-      mutate(match_type = "common")
-
-    #there are both sci and common names, only want common name 
-    return(bind_rows(sci_match, com_names))
-  }else{ return(sci_match)}
-}
-
-bird_ids <- get_trait_ids(bird_trait, "itis", "specid")
-mammal_ids <- get_trait_ids(mamm_trait, "itis", "msw3_id", FALSE)
-
+usethis::use_data(biotime_data)
 ```
-
-Are there any unresolved ids for birds? Yes - just one, which is a synonym to two different accepted id's
-```{r}
-unres_trait <- get_dupe_ids(bird_ids, "specid") %>%
-  left_join(bird_ids)
-
-unres_trait
-
-#resolve to the desired ID (in this case the one with complete hierarchy data)
-bird_ids <- unres_trait %>% 
-  filter(!is.na(kingdom)) %>%
-  bind_rows(bird_ids %>% filter(!sort %in% unres_trait$sort))
-```
-
-What about for mammals? Yes - also a synonym to two different accepted id's. Further research shows that the Callospermophilus saturatus uses the subgenus, so we'll keep the name with the genus 
-```{r}
-unres_mamm_trait <- get_dupe_ids(mammal_ids, "msw3_id") %>% 
-  left_join(mammal_ids)
-
-unres_mamm_trait
-
-mammal_ids <- unres_mamm_trait %>% 
-  filter(acceptedNameUsageID == "ITIS:632452") %>%
-  bind_rows(mammal_ids %>% filter(!sort %in% unres_mamm_trait$sort))
-```
-
-Join bird ids back to original trait data
-```{r}
-#only want one column for old name and one column for new name, these are the old ID columns to remove
-col_ex <- c("passnonpass", "iocorder", "blfamilylatin", "blfamilyenglish", "blfamsequid", "taxo")
-
-#join on scientific and common
-sci_matches <- bird_ids  %>%
-  filter(match_type == "scientific") %>%
-  select(id = acceptedNameUsageID, input, scientificName) %>%
-  distinct() %>%
-  right_join(bird_trait %>% select(.dots = -col_ex), by = c("input" = "scientific"), na_matches = "never") %>%
-  drop_na(id) %>%
-  distinct() %>%
-  select(-english)
-
-comm_matches <- bird_ids  %>%
-  filter(match_type == "common") %>%
-  select(id = acceptedNameUsageID, input, scientificName) %>%
-  distinct() %>%
-  right_join(bird_trait %>% filter(!specid %in% sci_matches$specid) %>% select(.dots = -col_ex), by = c("input" = "english"), na_matches = "never") %>%
-  distinct() %>%
-  rename(matchingName = input) %>%
-  select(-scientific)
-
-#there are more rows after ID data is joined because for species that matched on common names there may be more than one scientificName 
-bird_trait_data <- bind_rows(sci_matches, comm_matches)
-```
-
-Do the same for mammal trait data
-```{r}
-mamm_trait_data <- mammal_ids %>%
-  select(id = acceptedNameUsageID, input, scientificName) %>%
-  right_join(mamm_trait %>% select(-mswfamilylatin), by = c("input" = "scientific"), na_matches = "never") %>%
-  drop_na(id) %>%
-  distinct() 
-```
-
-#Checking Coverage of ID's and trait data -> biotime
-
-check if all bird & mammal biotime data w/id's has an elton match. There are 258 species without a match (38 are birds).
-```{r}
-biotime_data %>%
-  #filter(taxa == "Birds") %>%
-  select(id) %>%
-  distinct() %>%
-  filter(!id %in% bird_trait_data$id) %>%
-  dim()
-```
-
-and 605 bird species without an ID, 657 total species
-```{r}
-biotime_data %>%
-  #filter(taxa == "Birds") %>%
-  select(id, matchingName) %>%
-  distinct() %>%
-  filter(is.na(id)) %>%
-  dim()
-```
-
-Are there any studies for which we have trait data for all species?
-```{r}
-#join trait and biotime data
-bird_data <- biotime_data %>%
-  #filter(taxa == "Birds") %>%
-  left_join(bird_trait_data, by = "id",
-            na_matches = "never")
-
-coverage <- bird_data %>%
-  group_by(study_id) %>%
-  summarise(perc = sum(!is.na(specid))/n()) %>%
-  arrange(desc(perc))
-
-head(coverage)
-```

### 2019-05-20 data-raw/elton.Rmd
Split biotime and elton id labeling into two separate scripts, move coverage checking into analysis folder. Save all the data products in appropriate places.

--- 2019-05-16 biotime.Rmd
@@
 ---
-title: "BioTime"
+title: "elton trait"
 author: "Kari Norman"
-date: "4/1/2019"
+date: "5/20/2019"
 output: html_document
@@
 library(rdataretriever)
+library(taxadb)
 
-```
-
-Import occurence data from Biotime
-```{r}
-#rdataretriever::install("biotime", connection = "csv", data_dir = here("data"))
-
-data <- read_csv(here("data", "biotime_metadata.csv")) %>% filter(taxa %in% c("Birds", "Mammals")) %>%
-  select(study_id, taxa) %>%
-  left_join(read_csv(here("data", "biotime_query.csv")), by = "study_id")
-
+source(here::here("data-raw", "helper_functions.R"))
 ```
@@
 ```{r}
-library(taxadb)
-
-get_match_counts <- function(data, species_col){
-  species_col <- lazyeval::as_name(species_col)
-  data %>% 
-    select(!!species_col) %>% 
-    drop_na() %>%
-    distinct() %>%
-    mutate(
-      gbif = get_ids(!!species_col, "gbif"),
-      col = get_ids(!!species_col, "col" ),
-      itis = get_ids(!!species_col, "itis"),
-      ncbi = get_ids(!!species_col, "ncbi"),
-      wd = get_ids(!!species_col, "wd"  ),
-      iucn = get_ids(!!species_col, "iucn"),
-      ott = get_ids(!!species_col, "ott" ) 
-    ) %>%
-    #select(!!call("-", species_col)) %>%
-    purrr::map_dbl(function(x) sum(!is.na(x)))
-}
-
 ##Counts show that OTT is the best authority for both data sets
-get_match_counts(data %>% filter(taxa == "Birds"), "genus_species")
-get_match_counts(data, "genus_species")
 get_match_counts(bird_trait, "scientific")
@@
 
-OTT is the best authority but doesn't have common names, so we use the next best ITIS. We'll start with BioTime data:
-
-##Biotime
-
-```{r}
-#Getting matches by scientific name 
-biotime <- data %>% 
-  #filter(taxa == "Birds") %>%
-  select(genus_species) %>%
-  drop_na() %>%
-  distinct() %>%
-  mutate(id = get_ids(genus_species, "itis")) %>%
-  left_join(data, by = "genus_species")
-
-#Getting matches by sci name and common name 
-biotime_ids <- by_name(unique(data$genus_species), "itis") %>%
-  bind_rows(by_common(unique(data$genus_species), "itis")) %>%
-  #filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
-  drop_na(acceptedNameUsageID)
-```
-
-Some names match to multiple acceptedUsage ID's, so we have to manually choose which one we want. In this case it's just the Blue-winged warbler
-```{r}
-
-get_dupe_ids <- function(data, orig_id){
-  orig_id <- lazyeval::as_name(orig_id)
-
-  data %>%
-    distinct(acceptedNameUsageID, input, !!orig_id) %>%
-    group_by(!!orig_id) %>% 
-    filter(n()>1) %>% 
-    ungroup()
-}
-
-unres <- get_dupe_ids(biotime_ids, "sort")
-unres 
-```
-
-Add the unresolved ID to all the other ids and join with biotime data
-```{r}
-biotime_ids <- unres %>%
-left_join(biotime_ids) %>% #join ID's back with all the columns
-  filter(taxonomicStatus == "accepted") %>% #want the accepted ID from the two Warbler options
-  bind_rows(biotime_ids %>% filter(!sort %in% unres$sort)) #join to resolved ID's, excluding the observations we resolved manually
-
-biotime_data <- biotime_ids %>%
-  select(id = acceptedNameUsageID,input, scientificName) %>%
-  distinct() %>%
-  right_join(data %>% select(-species, -genus), by = c("input" = "genus_species")) %>%
-  rename(matchingName = input)
-```
-
-Do any unmatched species have matches with other providers?
-```{r}
-get_match_counts(biotime_data %>% filter(is.na(id)), "matchingName")
-
-unmatched <- biotime_data %>% filter(is.na(id)) %>% pull(matchingName) %>% unique()
-providers <- c("ott", "gbif", "col", "ncbi", "wd", "iucn")
-
-#match all the un-ID'd names to other providers, check if synonyms given by those providers match known ITIS species 
-alt_names <- map_df(providers, function(name) synonyms(unmatched, name)) %>%
-  drop_na(acceptedNameUsageID) %>%
-  #new matches can be in either the synonym or acceptedNameUsage column, so let's combine them
-  gather(type, altProviderName, synonym, acceptedNameUsage) %>%
-  select(altProviderName, input) %>% 
-  distinct()
-
-#match accepted names from 
-syn_res <- by_name(na.omit(unique(alt_names$altProviderName)),"itis") %>%
-                       rename(altProviderName = input) %>%
-                       left_join(alt_names) %>%
-  select(id = acceptedNameUsageID, input, scientificName) %>% 
-  drop_na(id) %>% 
-  distinct() %>%
-  left_join(data %>% select(-species, -genus), by = c("input" = "genus_species"), na_matches =  "never") %>%
-  rename(matchingName = input)
-
-#Below add's more rows, should just replace them 
-biotime_data <- biotime_data %>% filter(!matchingName %in% syn_res$matchingName) %>%
-  bind_rows(syn_res)
-
-# #These two should be equal, why aren't they??
-# dim(syn_res)  
-# biotime_data %>% filter(matchingName %in% syn_res$matchingName) %>% dim()
-# biotime_data %>% filter(is.na(id)) %>% dim()
-```
-
-##Elton Traits
-Get id's for the Elton Bird trait data
-
+Get id's for the Elton trait data
 ```{r}
@@
 #there are more rows after ID data is joined because for species that matched on common names there may be more than one scientificName 
-bird_trait_data <- bind_rows(sci_matches, comm_matches)
+elton_bird <- bind_rows(sci_matches, comm_matches)
 ```
@@
 ```{r}
-mamm_trait_data <- mammal_ids %>%
+elton_mamm <- mammal_ids %>%
   select(id = acceptedNameUsageID, input, scientificName) %>%
@@
 
-#Checking Coverage of ID's and trait data -> biotime
-
-check if all bird & mammal biotime data w/id's has an elton match. There are 258 species without a match (38 are birds).
 ```{r}
-biotime_data %>%
-  #filter(taxa == "Birds") %>%
-  select(id) %>%
-  distinct() %>%
-  filter(!id %in% bird_trait_data$id) %>%
-  dim()
+usethis::use_data(elton_bird, elton_mamm)
 ```
-
-and 605 bird species without an ID, 657 total species
-```{r}
-biotime_data %>%
-  #filter(taxa == "Birds") %>%
-  select(id, matchingName) %>%
-  distinct() %>%
-  filter(is.na(id)) %>%
-  dim()
-```
-
-Are there any studies for which we have trait data for all species?
-```{r}
-#join trait and biotime data
-bird_data <- biotime_data %>%
-  #filter(taxa == "Birds") %>%
-  left_join(bird_trait_data, by = "id",
-            na_matches = "never")
-
-coverage <- bird_data %>%
-  group_by(study_id) %>%
-  summarise(perc = sum(!is.na(specid))/n()) %>%
-  arrange(desc(perc))
-
-head(coverage)
-```

### 2019-06-06 data-raw/biotime.Rmd
Resolve all of the duplicate id's

@@
   left_join(biotime_ids) %>% #join ID's back with all the columns
-  filter(taxonomicStatus == "accepted") %>% #want the accepted ID from the two Warbler options
+  filter(taxonomicStatus == "accepted", taxonRank == "species") %>% #want the accepted ID from the two Warbler options
   bind_rows(biotime_ids %>% filter(!sort %in% unres$sort)) #join to resolved ID's, excluding the observations we resolved manually

### 2019-07-01 analysis/coverage_check.Rmd
Update species counts.

@@
 ```{r}
+library(tidyverse)
+```
+
+```{r}
 load(here::here("data", "biotime_data.rda"))
@@
 
-check if all bird & mammal biotime data w/id's has an elton match. There are 258 species without a match (38 are birds).
+check if all bird & mammal biotime data w/id's has an elton match. There are 81 species without a match (67 are birds).
+```{r}
+no_trait <- biotime_data %>%
+  #filter(taxa == "Birds") %>%
+  select(id, taxa) %>%
+  drop_na(id) %>%
+  distinct() %>%
+  filter(!id %in% elton_bird$id, !id %in% elton_mamm$id)
+
+dim(no_trait)
+```
+
+and 552 species without an ID, some of which are species codes
 ```{r}
@@
   #filter(taxa == "Birds") %>%
-  select(id) %>%
-  distinct() %>%
-  filter(!id %in% elton_bird$id) %>%
-  dim()
-```
-
-and 605 bird species without an ID, 657 total species
-```{r}
-biotime_data %>%
-  #filter(taxa == "Birds") %>%
-  select(id, matchingName) %>%
+  select(id, matchingName, taxa) %>%
   distinct() %>%
@@
 bird_data <- biotime_data %>%
-  #filter(taxa == "Birds") %>%
+  filter(taxa == "Birds") %>%
   left_join(elton_bird, by = "id",

### 2019-07-01 data-raw/biotime.Rmd
modify to include all taxa in biotime data

--- 2019-05-20 data-raw/biotime.Rmd
@@
 data <- read_csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS")) %>% 
-  filter(taxa %in% c("Birds", "Mammals")) %>%
   select(study_id, taxa) %>%
@@
 ```{r}
-##Counts show that OTT is the best authority for both data sets
+##Counts show that OTT is the best authority 
 get_match_counts(data %>% filter(taxa == "Birds"), "genus_species")
@@
 ```{r}
-#Getting matches by scientific name 
-biotime <- data %>% 
-  #filter(taxa == "Birds") %>%
-  select(genus_species) %>%
-  drop_na() %>%
-  distinct() %>%
-  mutate(id = get_ids(genus_species, "itis")) %>%
-  left_join(data, by = "genus_species")
+# #Getting matches by scientific name 
+# biotime <- data %>% 
+#   select(genus_species) %>%
+#   drop_na() %>%
+#   distinct() %>%
+#   mutate(id = get_ids(genus_species, "itis")) %>%
+#   left_join(data, by = "genus_species")
 
@@
   bind_rows(by_common(unique(data$genus_species), "itis")) %>%
-  #filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
+  filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
   drop_na(acceptedNameUsageID)
@@
 
-Some names match to multiple acceptedUsage ID's, so we have to manually choose which one we want. In this case it's just the Blue-winged warbler
+Some names match to multiple acceptedUsage ID's, so we have to manually choose which one we want.
 ```{r}
@@
   left_join(biotime_ids) %>% #join ID's back with all the columns
-  filter(taxonomicStatus == "accepted") %>% #want the accepted ID from the two Warbler options
+  filter(acceptedNameUsageID %in% c(
+    "ITIS:159807", #Heterodontus zebra, all the studies including it are using sci names, so we should match to the sci name match not common name match
+    "ITIS:175125" #The Redhead is coming from bird studies, so choose the bird match, 
+    #the rest can't be match due to too little information
+  )) %>%
   bind_rows(biotime_ids %>% filter(!sort %in% unres$sort)) #join to resolved ID's, excluding the observations we resolved manually
@@
   select(altProviderName, input) %>% 
-  distinct()
+  distinct() %>%
+  drop_na(altProviderName)
 
@@
   rename(altProviderName = input) %>%
-  left_join(alt_names, na_matches = "never")
+  left_join(alt_names, na_matches = "never", by = "altProviderName")
 
@@
 alt_dupes <- syn_res %>%
+  select(-taxonID) %>%
+  filter(taxonRank == "species") %>%
   group_by(input) %>%
@@
 
-#filter duplicates for accepted, and replace old entries 
-syn_res <- alt_dupes %>% 
-  filter(taxonRank == "species", taxonomicStatus == "accepted") %>% 
-  bind_rows(syn_res %>% 
-              filter(!acceptedNameUsageID %in% alt_dupes$acceptedNameUsageID)) %>%
-#then finish clean up 
+# #filter duplicates for accepted, and replace old entries 
+# syn_res <- alt_dupes %>% 
+#   filter(acceptedNameUsageID %in% c(
+#     ITIS:331277, 
+#   )) %>% 
+#   bind_rows(syn_res %>% 
+#               filter(!acceptedNameUsageID %in% alt_dupes$acceptedNameUsageID)) %>%
+# #then finish clean up 
+#   select(id = acceptedNameUsageID, input, scientificName) %>% 
+#   distinct() %>%
+#   left_join(data %>% select(-species, -genus), by = c("input" = "genus_species"), na_matches =  "never") %>%
+#   rename(matchingName = input) %>%
+#   ungroup()
+
+#Just get rid of those with duplicates, deal with them later
+add_syn_res <- syn_res %>% 
+  filter(!input %in% alt_dupes$input) %>% #excluding duplicate matched species
   select(id = acceptedNameUsageID, input, scientificName) %>%
@@
 
-#Below add's more rows, should just replace them 
-biotime_data <- biotime_data %>% filter(!matchingName %in% syn_res$matchingName) %>%
-  bind_rows(syn_res)
+biotime_data <- biotime_data %>% 
+  filter(!matchingName %in% add_syn_res$matchingName) %>%
+  distinct() %>%
+  bind_rows(add_syn_res)
 

### 2019-07-02 data-raw/biotime.Rmd
Filter biotime data not original data.

@@
   select(-taxonID) %>%
-  filter(taxonRank == "species") %>%
+  #filter(taxonRank == "species") %>%
   group_by(input) %>%
@@
   filter(!input %in% alt_dupes$input) %>% #excluding duplicate matched species
-  select(id = acceptedNameUsageID, input, scientificName) %>%
+  select(id = acceptedNameUsageID, matchingName = input, scientificName) %>%
   distinct() %>%
-  left_join(data %>% select(-species, -genus), by = c("input" = "genus_species"), na_matches =  "never") %>%
-  rename(matchingName = input) %>%
-  ungroup()
+  left_join(biotime_data%>% select(-c(id, scientificName)), by = "matchingName", na_matches =  "never") %>%
+  ungroup() %>%
+  distinct()
 
@@
   filter(!matchingName %in% add_syn_res$matchingName) %>%
-  distinct() %>%
   bind_rows(add_syn_res)

### 2019-07-02 analysis/coverage_check.Rmd
Check coverage for mammals and get id counts for all species (not just birds and mammals).

@@
 
-check if all bird & mammal biotime data w/id's has an elton match. There are 81 species without a match (67 are birds).
+check if all bird & mammal biotime data w/id's has an elton match. There are 63 species without a match (54 are birds).
 ```{r}
 no_trait <- biotime_data %>%
-  #filter(taxa == "Birds") %>%
-  select(id, taxa) %>%
+  filter(taxa %in% c("Birds", "Mammals")) %>%
+  select(id, matchingName, taxa) %>%
   drop_na(id) %>%
@@
 
-and 552 species without an ID, some of which are species codes
+and 20384 species without an ID, some of which are species codes
 ```{r}
@@
   select(id, matchingName, taxa) %>%
+  filter(is.na(id)) %>%
   distinct() %>%
-  filter(is.na(id)) %>%
   dim()
@@
 
-coverage <- bird_data %>%
+mamm_data <- biotime_data %>%
+  filter(taxa == "Mammals") %>%
+  left_join(elton_mamm, by = "id",
+            na_matches = "never")
+
+bird_coverage <- bird_data %>%
   group_by(study_id) %>%
@@
 
-head(coverage)
+mamm_coverage <- mamm_data %>%
+  group_by(study_id) %>%
+  summarise(perc = sum(!is.na(msw3_id))/n()) %>%
+  arrange(desc(perc))
+
+head(bird_coverage)
+head(mamm_coverage)
 ```

### 2019-07-02 data-raw/elton.Rmd
Include all trait data, even species without id's. Standardize origin sci name column name.

@@
   right_join(bird_trait %>% select(.dots = -col_ex), by = c("input" = "scientific"), na_matches = "never") %>%
-  drop_na(id) %>%
+  #drop_na(id) %>%
   distinct() %>%
@@
 #there are more rows after ID data is joined because for species that matched on common names there may be more than one scientificName 
-elton_bird <- bind_rows(sci_matches, comm_matches)
+elton_bird <- bind_rows(sci_matches, comm_matches) %>% rename(sourceName = input)
 ```
@@
   right_join(mamm_trait %>% select(-mswfamilylatin), by = c("input" = "scientific"), na_matches = "never") %>%
-  drop_na(id) %>%
-  distinct() 
+  distinct() %>%
+  rename(sourceName = input)
 ```

### 2019-07-02 data-raw/amphibio.Rmd
Don't need new sci name column, it's already called species.
Standardize name of the original sci name column.

@@
 
-amph <- read_csv(system.file("extdata", "amphibio/AmphiBIO_v1.csv", package = "biodivTS")) %>%
-  unite(scientificName, Genus, Species)
+amph <- read_csv(system.file("extdata", "amphibio/AmphiBIO_v1.csv", package = "biodivTS"))
 
@@
   mutate(id = get_ids(Species, "itis")) %>%
-  filter(!is.na(id))
+  filter(!is.na(id)) %>%
+  mutate(scientificName = get_names(id, "itis")) %>%
+  select(-c("Order", "Family", "Genus")) %>%
+  rename(sourceName = Species)
 

### 2019-07-02 data-raw/biotime.Rmd
Standardize name of the original sci name column.

@@
   right_join(data %>% select(-species, -genus), by = c("input" = "genus_species")) %>%
-  rename(matchingName = input)
+  rename(sourceName = input)
 ```
@@
 ```{r}
-#get_match_counts(biotime_data %>% filter(is.na(id)), "matchingName")
+#get_match_counts(biotime_data %>% filter(is.na(id)), "sourceName")
 
-unmatched <- biotime_data %>% filter(is.na(id)) %>% pull(matchingName) %>% unique()
+unmatched <- biotime_data %>% filter(is.na(id)) %>% pull(sourceName) %>% unique()
 providers <- c("ott", "gbif", "col", "ncbi", "wd", "iucn")
@@
 #   left_join(data %>% select(-species, -genus), by = c("input" = "genus_species"), na_matches =  "never") %>%
-#   rename(matchingName = input) %>%
+#   rename(sourceName = input) %>%
 #   ungroup()
@@
   filter(!input %in% alt_dupes$input) %>% #excluding duplicate matched species
-  select(id = acceptedNameUsageID, matchingName = input, scientificName) %>%
+  select(id = acceptedNameUsageID, sourceName = input, scientificName) %>%
   distinct() %>%
-  left_join(biotime_data%>% select(-c(id, scientificName)), by = "matchingName", na_matches =  "never") %>%
+  left_join(biotime_data%>% select(-c(id, scientificName)), by = "sourceName", na_matches =  "never") %>%
   ungroup() %>%
@@
 biotime_data <- biotime_data %>% 
-  filter(!matchingName %in% add_syn_res$matchingName) %>%
+  filter(!sourceName %in% add_syn_res$sourceName) %>%
   bind_rows(add_syn_res)

### 2019-07-02 data-raw/amphibio.Rmd
Keep all data, even species without id's. Match dataframe name to rda file name.

--- 2019-05-03 data-raw/amphibio.Rmd
@@
 
-amph <- read_csv(system.file("extdata", "amphibio/AmphiBIO_v1.csv", package = "biodivTS")) %>%
-  unite(scientificName, Genus, Species)
+amph <- read_csv(system.file("extdata", "amphibio/AmphiBIO_v1.csv", package = "biodivTS"))
 
-amph_ids <- by_name(amph$Species, "itis") %>%
-  filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
-  drop_na(acceptedNameUsageID)
+amphibio <- amph %>%
+  mutate(id = get_ids(Species, "itis")) %>%
+  filter(!is.na(id)) %>%
+  mutate(scientificName = get_names(id, "itis")) %>%
+  select(id, scientificName, Species) %>%
+  right_join(select(amph, -id), by = "Species") %>%
+  rename(sourceName = Species) %>%
+  select(-c("Order", "Family", "Genus"))
+
+# Using by_ doesn't give any additional ids
+# amphibio <- by_name(amph$Species, "itis") %>%
+#   filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
+#   drop_na(acceptedNameUsageID)
+
+usethis::use_data(amphibio)
 ```

### 2019-07-03 data-raw/biotime.Rmd
Make function for chacking names missing ID's against other providers to get synonyms that might have ids.

@@
 ```{r}
-#get_match_counts(biotime_data %>% filter(is.na(id)), "sourceName")
 
-unmatched <- biotime_data %>% filter(is.na(id)) %>% pull(sourceName) %>% unique()
-providers <- c("ott", "gbif", "col", "ncbi", "wd", "iucn")
-
-#match all the un-ID'd names to other providers, check if synonyms given by those providers match known ITIS species 
-alt_names <- map_df(providers, function(name) synonyms(unmatched, name)) %>%
-  drop_na(acceptedNameUsageID) %>%
-  #new matches can be in either the synonym or acceptedNameUsage column, so let's combine them
-  gather(type, altProviderName, synonym, acceptedNameUsage) %>%
-  select(altProviderName, input) %>% 
-  distinct() %>%
-  drop_na(altProviderName)
-
-#match accepted names from 
-syn_res <- by_name(na.omit(unique(alt_names$altProviderName)),"itis") %>%
-  drop_na(acceptedNameUsageID) %>% 
-  rename(altProviderName = input) %>%
-  left_join(alt_names, na_matches = "never", by = "altProviderName")
+#Check for potential matches from synonyms found in the other providers
+syn_res <- match_providers(biotime_data, "itis")
 

### 2019-07-03 data-raw/helper_functions.R

@@ -28,3 +28,25 @@ get_dupe_ids <- function(data, orig_id){
     filter(n()>1) %>%
     ungroup()
 }
+
+#Check if synonyms from other providers have matches in the original database
+match_providers <- function(data, original_provider) {
+  unmatched <- data %>% filter(is.na(id)) %>% pull(sourceName) %>% unique()
+  providers <- c("itis", "ncbi", "col", "gbif", "fb", "wd", "ott", "iucn")
+
+  #match all the un-ID'd names to other providers, check if synonyms given by those providers match known ITIS species
+  alt_names <-map_df(providers[providers != original_provider], function(name) synonyms(unmatched, name)) %>%
+    drop_na(acceptedNameUsageID) %>%
+    #new matches can be in either the synonym or acceptedNameUsage column, so let's combine them
+    gather(type, altProviderName, synonym, acceptedNameUsage) %>%
+    select(altProviderName, input) %>%
+    distinct() %>%
+    drop_na(altProviderName)
+
+  #match accepted names from
+  syn_res <- by_name(na.omit(unique(alt_names$altProviderName)), original_provider) %>%
+    drop_na(acceptedNameUsageID) %>%
+    rename(altProviderName = input) %>%
+    left_join(alt_names, na_matches = "never", by = "altProviderName")
+
+}

### 2019-07-03 data-raw/amphibio.Rmd
Resolve missing ids against other providers.

@@
 
+Get id's for amphibio database from itis
 ```{r}
@@
 #   drop_na(acceptedNameUsageID)
+```
 
+Check for matches from other providers
+```{r}
+syn_res <- match_providers(amphibio, "itis")
+
+#names that resolve to more than one id
+alt_dupes <- syn_res %>%
+  select(acceptedNameUsageID, input) %>%
+  distinct() %>%
+  group_by(input) %>%
+  filter(n() > 1)
+
+add_syn_res <- syn_res %>% 
+  filter(!input %in% alt_dupes$input) %>% #excluding duplicate matched species
+  select(id = acceptedNameUsageID, sourceName = input, scientificName) %>%
+  distinct() %>%
+  group_by(sourceName) %>%
+  top_n(1, scientificName) %>%
+  left_join(amphibio %>% select(-c(id, scientificName)), by = "sourceName", na_matches =  "never") %>%
+  ungroup() %>%
+  distinct()
+
+amphibio <- amphibio %>% 
+  filter(!sourceName %in% add_syn_res$sourceName) %>%
+  bind_rows(add_syn_res)
+```
+
+```{r}
 usethis::use_data(amphibio)

### 2019-07-03 data-raw/biotime.Rmd
Deal with names that only have one id but multiple potential accepted sci names, randomly pick one.

@@
 alt_dupes <- syn_res %>%
-  select(-taxonID) %>%
-  #filter(taxonRank == "species") %>%
+  select(acceptedNameUsageID, input) %>%
+  distinct() %>%
   group_by(input) %>%
@@
   distinct() %>%
+  group_by(sourceName) %>%
+  top_n(1, scientificName) %>% #some have multiple sci names for same id, just pick on 
   left_join(biotime_data%>% select(-c(id, scientificName)), by = "sourceName", na_matches =  "never") %>%

### 2019-07-03 data-raw/elton.Rmd
Check alternative providers for unresolved names.

@@
 
+Check against alternative providers, first for mammals
+```{r}
+syn_res <- match_providers(elton_mamm, "itis")
+
+#names that resolve to more than one id
+alt_dupes <- syn_res %>%
+  select(acceptedNameUsageID, input) %>%
+  distinct() %>%
+  group_by(input) %>%
+  filter(n() > 1)
+
+add_syn_res <- syn_res %>% 
+  filter(!input %in% alt_dupes$input) %>% #excluding duplicate matched species
+  select(id = acceptedNameUsageID, sourceName = input, scientificName) %>%
+  distinct() %>%
+  group_by(sourceName) %>%
+  top_n(1, scientificName) %>%
+  left_join(elton_mamm %>% select(-c(id, scientificName)), by = "sourceName", na_matches =  "never") %>%
+  ungroup() %>%
+  distinct()
+
+elton_mamm <- elton_mamm %>% 
+  filter(!sourceName %in% add_syn_res$sourceName) %>%
+  bind_rows(add_syn_res)
+```
+
+then birds
+```{r}
+syn_res <- match_providers(elton_bird, "itis")
+
+#names that resolve to more than one id
+alt_dupes <- syn_res %>%
+  select(acceptedNameUsageID, input) %>%
+  distinct() %>%
+  group_by(input) %>%
+  filter(n() > 1)
+
+add_syn_res <- syn_res %>% 
+  filter(!input %in% alt_dupes$input) %>% #excluding duplicate matched species
+  select(id = acceptedNameUsageID, sourceName = input, scientificName) %>%
+  distinct() %>%
+  group_by(sourceName) %>%
+  top_n(1, scientificName) %>%
+  left_join(elton_bird %>% select(-c(id, scientificName)), by = "sourceName", na_matches =  "never") %>%
+  ungroup() %>%
+  distinct()
+
+elton_bird <- elton_bird %>% 
+  filter(!sourceName %in% add_syn_res$sourceName) %>%
+  bind_rows(add_syn_res)
+```
+
 ```{r}

### 2019-07-03 data-raw/helper_functions.R
Also check common names against alternate providers.

@@ -30,7 +30,7 @@ get_dupe_ids <- function(data, orig_id){
 }
 
 #Check if synonyms from other providers have matches in the original database
-match_providers <- function(data, original_provider) {
+match_providers <- function(data, original_provider, common = FALSE) {
   unmatched <- data %>% filter(is.na(id)) %>% pull(sourceName) %>% unique()
   providers <- c("itis", "ncbi", "col", "gbif", "fb", "wd", "ott", "iucn")
 
@@ -48,5 +48,22 @@ match_providers <- function(data, original_provider) {
     drop_na(acceptedNameUsageID) %>%
     rename(altProviderName = input) %>%
     left_join(alt_names, na_matches = "never", by = "altProviderName")
+
+  if (common == TRUE){
+    common_providers <- c("col", "fb", "gbif", "itis", "iucn", "ncbi", "slb")
+    unmatched <- unmatched[unmatched %in% syn_res$input]
+
+    common_match <- map_df(common_providers[common_providers != original_provider],
+           function(name) by_common(unmatched, name)) %>%
+      drop_na(acceptedNameUsageID)
+
+    sci_match <- by_name(unique(common_match$scientificName), "itis") %>%
+      rename(altProviderName = input) %>%
+      drop_na(acceptedNameUsageID) %>%
+      left_join(common_match %>% select(scientificName, input),
+                by = c("altProviderName" = "scientificName"))
+
+    return(bind_rows(syn_res, sci_match))
+  }
 
 }

### 2019-07-03 analysis/coverage_check.Rmd
Clean up coverage function. Also get coverage of amphibian studies.

@@
 load(here::here("data", "elton_bird.rda"))
+load(here::here("data", "amphibio.rda"))
 ```
@@
 
-and 20384 species without an ID, some of which are species codes
+and 19869 species without an ID, some of which are species codes
 ```{r}
@@
   #filter(taxa == "Birds") %>%
-  select(id, matchingName, taxa) %>%
+  select(id, sourceName, taxa) %>%
   filter(is.na(id)) %>%
@@
 ```{r}
-#join trait and biotime data
-bird_data <- biotime_data %>%
-  filter(taxa == "Birds") %>%
-  left_join(elton_bird, by = "id",
+get_coverage <- function(taxa, trait_data){
+  taxa_data <- biotime_data %>%
+    select(-sourceName, -scientificName) %>%
+  filter(taxa == taxa) %>%
+  left_join(trait_data, by = "id",
             na_matches = "never")
 
-mamm_data <- biotime_data %>%
-  filter(taxa == "Mammals") %>%
-  left_join(elton_mamm, by = "id",
-            na_matches = "never")
+  coverage <- taxa_data %>%
+  group_by(study_id) %>%
+  summarise(perc = sum(!is.na(sourceName))/n()) %>%
+  arrange(desc(perc))
+}
 
-bird_coverage <- bird_data %>%
-  group_by(study_id) %>%
-  summarise(perc = sum(!is.na(specid))/n()) %>%
-  arrange(desc(perc))
-
-mamm_coverage <- mamm_data %>%
-  group_by(study_id) %>%
-  summarise(perc = sum(!is.na(msw3_id))/n()) %>%
-  arrange(desc(perc))
-
-head(bird_coverage)
-head(mamm_coverage)
+get_coverage("Birds", elton_bird) %>% View()
+get_coverage("Mammals", elton_mamm) %>% View()
+get_coverage("Amphibians", amphibio) %>% View()
 ```

### 2019-07-05 analysis/coverage_check.Rmd
Modify so we can get both coverage and merged data from map calls.

@@
 ```{r}
-get_coverage <- function(taxa, trait_data){
-  taxa_data <- biotime_data %>%
+merge_data <- function(taxa, trait_data){
+  biotime_data %>%
     select(-sourceName, -scientificName) %>%
@@
             na_matches = "never")
+}
 
+get_coverage <- function(taxa_data){
   coverage <- taxa_data %>%
@@
 
-get_coverage("Birds", elton_bird) %>% View()
-get_coverage("Mammals", elton_mamm) %>% View()
-get_coverage("Amphibians", amphibio) %>% View()
+get_coverage(merge_data("Birds", elton_bird)) %>% View
+get_coverage(merge_data("Mammals", elton_mamm)) %>% View()
+get_coverage(merge_data("Amphibians", amphibio)) %>% View()
 ```
+
+Get data w/traits for all studies that have >85% trait coverage
+```{r}
+#set coverage threshold for analysis dataset
+perc <- 0.85
+
+traits <- list(elton_bird, elton_mamm, amphibio)
+taxa <- list("Birds", "Mammals", "Amphibians")
+
+biotime_traits_list <- map2(taxa, traits, ~merge_data(taxa = .x, trait_data = .y))
+coverage <- map_dfr(biotime_traits_list, get_coverage) %>%
+  filter(perc > perc)
+
+biotime_traits <- biotime_traits_list %>%
+  bind_rows() %>%
+  filter(study_id %in% coverage$study_id)
+```

### 2019-07-05 data-raw/elton.Rmd
Rename was adding an extra matchingName column.

@@
   distinct() %>%
-  rename(matchingName = input) %>%
   select(-scientific)

### 2019-07-05 data-raw/helper_functions.R

@@ -64,6 +64,5 @@ match_providers <- function(data, original_provider, common = FALSE) {
                 by = c("altProviderName" = "scientificName"))
 
     return(bind_rows(syn_res, sci_match))
-  }
-
+  } else {return(syn_res)}
 }

### 2019-07-05 analysis/coverage_check.Rmd
Excluded data without trait matches to get complete dataset.

@@
   filter(taxa %in% c("Birds", "Mammals")) %>%
-  select(id, matchingName, taxa) %>%
+  select(id, sourceName, taxa) %>%
   drop_na(id) %>%
@@
 #set coverage threshold for analysis dataset
-perc <- 0.85
+percent <- 0.85
 
@@
 coverage <- map_dfr(biotime_traits_list, get_coverage) %>%
-  filter(perc > perc)
+  filter(perc > percent)
 
-biotime_traits <- biotime_traits_list %>%
-  bind_rows() %>%
+biotime_traits <- map_dfr(biotime_traits_list, ~drop_na(., scientificName)) %>%
+  #bind_rows() %>%
   filter(study_id %in% coverage$study_id)

### 2019-07-18 analysis/merge_data.Rmd
Rename
Calculate FD metrics for each year and time point.

--- 2019-07-05 analysis/coverage_check.Rmd
@@
 library(tidyverse)
+library(FD)
 ```
@@
 biotime_traits <- map_dfr(biotime_traits_list, ~drop_na(., scientificName)) %>%
-  #bind_rows() %>%
   filter(study_id %in% coverage$study_id)
 ```
+
+```{r}
+usethis::use_data(biotime_traits)
+```
+
+Get FD metrics for each study 
+```{r}
+load(here::here("data", "biotime_traits.rda"))
+
+get_fd <- function(data){
+  species_mat <- data %>%
+    select(year, id) %>% 
+    arrange(year, id) %>%
+    mutate(present = 1) %>%
+    spread(id, value = present, fill = 0) %>%
+    select(-year)
+
+  years <- data %>% 
+    select(year) %>%
+    distinct() %>%
+    arrange(year)
+
+  trait_mat <- data %>%
+    select(id, diet_inv, diet_vend, diet_vect, diet_vfish, diet_vunk, diet_scav, diet_fruit, 
+           diet_nect, diet_seed, diet_planto, diet_certainty, forstrat_watbelowsurf,
+           forstrat_wataroundsurf, forstrat_ground, forstrat_understory, forstrat_midhigh,
+           forstrat_canopy, forstrat_aerial, pelagicspecialist, forstrat_speclevel, nocturnal, 
+           bodymass_value, bodymass_speclevel, forstrat_value, activity_nocturnal, activity_crepuscular, 
+           activity_diurnal, Fos, Ter, Aqu, Arb, Leaves, Flowers, Seeds, Fruits, Arthro, Vert, 
+           Diu, Noc, Crepu, Wet_warm, Wet_cold, Dry_warm, Dry_cold, Body_mass_g, Age_at_maturity_min_y, 
+           Age_at_maturity_max_y, Body_size_mm, Size_at_maturity_min_mm, Size_at_maturity_max_mm, 
+           Longevity_max_y , Litter_size_min_n, Litter_size_max_n, Reproductive_output_y, 
+           Offspring_size_min_mm, Offspring_size_max_mm, Dir, Lar, Viv, OBS 
+    ) %>%
+    distinct() %>%
+    select_if(~ length(unique(.)) > 1) %>% #remove columns that have the same value for all columns
+    mutate_at(vars(-id), list(~scales::rescale(.,to=c(0, 1)))) %>% #rescale variables
+    arrange(id) %>%
+    column_to_rownames("id")
+
+  as.tibble(dbFD(trait_mat, species_mat, w.abun = FALSE))[,1:8] %>%
+    cbind(years, .)
+}
+
+get_fd_safe <- possibly(get_fd, NA)
+
+test_data <- biotime_traits %>% 
+  filter(study_id %in% c(39, 46, 47, 51)) %>%
+  group_by(study_id) %>%
+  nest() %>%
+  mutate(metrics = map(data, ~get_fd_safe(.x))) %>%
+  filter(!is.na(metrics)) %>%
+  select(-data) %>%
+  unnest()
+```

### 2019-08-06 analysis/merge_data.Rmd
Add joggle argument to qhull calculation to fix precision problems.

@@
 get_fd <- function(data){
+  #print(max(data$record_id))
   species_mat <- data %>%
@@
 
-  as.tibble(dbFD(trait_mat, species_mat, w.abun = FALSE))[,1:8] %>%
+  as.tibble(dbFD_joggle(trait_mat, species_mat, w.abun = FALSE))[,1:8] %>%
     cbind(years, .)
@@
 test_data <- biotime_traits %>% 
-  filter(study_id %in% c(39, 46, 47, 51)) %>%
-  group_by(study_id) %>%
+  group_by(study_id, plot) %>%
   nest() %>%

### 2019-08-06 R/dbFD_joggle.R

@@ -1,0 +1,716 @@
+dbFD_joggle <- function (x, a, w, w.abun = TRUE, stand.x = TRUE, ord = c("podani",
+                                                          "metric"), asym.bin = NULL, corr = c("sqrt", "cailliez",
+                                                                                               "lingoes", "none"), calc.FRic = TRUE, m = "max", stand.FRic = FALSE,
+          scale.RaoQ = FALSE, calc.FGR = FALSE, clust.type = "ward",
+          km.inf.gr = 2, km.sup.gr = nrow(x) - 1, km.iter = 100, km.crit = c("calinski",
+                                                                             "ssi"), calc.CWM = TRUE, CWM.type = c("dom", "all"),
+          calc.FDiv = TRUE, dist.bin = 2, print.pco = FALSE, messages = TRUE)
+{
+  tol <- .Machine$double.eps
+  corr <- match.arg(corr)
+  ord <- match.arg(ord)
+  CWM.type <- match.arg(CWM.type)
+  km.crit <- match.arg(km.crit)
+  if (!is.logical(messages))
+    stop("'messages' must be TRUE or FALSE.", "\n")
+  if (!is.logical(stand.FRic))
+    stop("'stand.FRic' must be TRUE or FALSE.", "\n")
+  if (!is.logical(stand.x))
+    stop("'stand.x' must be TRUE or FALSE.", "\n")
+  if (!is.logical(w.abun))
+    stop("'w.abun' must be TRUE or FALSE.", "\n")
+  if (!is.logical(calc.FRic))
+    stop("'calc.FRic' must be TRUE or FALSE.", "\n")
+  if (!is.logical(calc.FDiv))
+    stop("'calc.FDiv' must be TRUE or FALSE.", "\n")
+  if (!is.logical(calc.FGR))
+    stop("'calc.FGR' musts be TRUE or FALSE.", "\n")
+  if (!is.logical(calc.CWM))
+    stop("'calc.CWM' must be TRUE or FALSE.", "\n")
+  if (!is.logical(scale.RaoQ))
+    stop("'scale.RaoQ' must be TRUE or FALSE.", "\n")
+  if (!is.logical(print.pco))
+    stop("'print.pco' must be TRUE or FALSE.", "\n")
+  if (is.matrix(x) | is.data.frame(x)) {
+    is.dist.x <- FALSE
+    s.x <- dim(x)[1]
+    t.x <- dim(x)[2]
+    if (is.null(row.names(x)))
+      stop("'x' must have row names.", "\n")
[…]

### 2019-08-07 data-raw/amphibio.Rmd
Handle multiple trait database matches to the same ITIS id.

@@
 
+Check for multiple matches to the same ID
+```{r}
+#get ids that have more than one entry
+dupe_ids <- amphibio %>% 
+  select(id, sourceName) %>% 
+  drop_na(id) %>%
+  distinct() %>% 
+  group_by(id) %>% 
+  filter(n() > 1) %>%
+  pull(id)
+
+#get data associated with ids
+dupe_data <- amphibio %>% 
+  filter(id %in% dupe_ids)
+
+#resolve so that traits are averaged for species that need to be combined
+resolved <- dupe_data %>%
+  select(-scientificName, -sourceName) %>%
+  group_by(id) %>%
+  summarise_all(~ if (all(is.na(.))) {NA} else {mean(., na.rm = TRUE)}) %>%
+  mutate(scientificName = get_names(id))
+
+#remove unresolved data and replace with new resolved
+amphibio <-  amphibio %>% 
+  anti_join(dupe_data) %>% 
+  bind_rows(resolved)
+
+```
+
 ```{r}

### 2019-08-07 data-raw/elton.Rmd
Standardize trait names, remove unecessary columns.

@@
 
+Grab only the trait columns we're interested in, standardize column names
+```{r}
+elton_bird <- elton_bird %>% 
+  select(-specid, -ends_with("source"), -ends_with("enteredby"), -ends_with("certainty"), -ends_with("speclevel"), -ends_with("comment"))
+
+elton_mamm <- elton_mamm %>%
+  select(-msw3_id, -forstrat_value, -ends_with("source"), -ends_with("certainty"), -ends_with("speclevel"), -ends_with("comment"))
+```
+
 ```{r}

### 2019-08-07 data-raw/helper_functions.R
Make function for resolving multiple trait entries to one id, resolve for elton database as well.

@@ -66,3 +66,31 @@ match_providers <- function(data, original_provider, common = FALSE) {
     return(bind_rows(syn_res, sci_match))
   } else {return(syn_res)}
 }
+
+#Identify ids that have matched to multiple entries (for trait databases), and average within ID so there is only one entry
+undupe_ids <- function(data) {
+  #get ids that have more than one entry
+  dupe_ids <- data %>%
+    select(id, sourceName) %>%
+    drop_na(id) %>%
+    distinct() %>%
+    group_by(id) %>%
+    filter(n() > 1) %>%
+    pull(id)
+
+  #get data associated with ids
+  dupe_data <- data %>%
+    filter(id %in% dupe_ids)
+
+  #resolve so that traits are averaged for species that need to be combined
+  resolved <- dupe_data %>%
+    select(-scientificName,-sourceName) %>%
+    group_by(id) %>%
+    summarise_all( ~ if (all(is.na(.))) {NA} else {mean(., na.rm = TRUE)}) %>%
+    mutate(scientificName = get_names(id))
+
+  #remove unresolved data and replace with new resolved
+  resolved_data <- data %>%
+    anti_join(dupe_data) %>%
+    bind_rows(resolved)
+}

### 2019-08-07 analysis/merge_data.Rmd
Standardize trait column names, simplify trait matrix column selection.

--- 2019-07-05 analysis/coverage_check.Rmd
@@
 biotime_traits <- map_dfr(biotime_traits_list, ~drop_na(., scientificName)) %>%
-  #bind_rows() %>%
   filter(study_id %in% coverage$study_id)
 ```
+
+```{r}
+usethis::use_data(biotime_traits)
+```
+
+Get FD metrics for each study 
+```{r}
+load(here::here("data", "biotime_traits.rda"))
+
+get_fd <- function(data){
+  #print(max(data$record_id))
+  species_mat <- data %>%
+    select(year, id) %>% 
+    arrange(year, id) %>%
+    mutate(present = 1) %>%
+    spread(id, value = present, fill = 0) %>%
+    select(-year)
+
+  years <- data %>% 
+    select(year) %>%
+    distinct() %>%
+    arrange(year)
+
+  trait_mat <- data %>%
+    select(-c(id, taxa, record_id, day, month, year, sample_desc, id_species, 
+              latitude, longitude, sum_abundance, sum_biomass, sourceName, scientificName)
+    ) %>%
+    distinct() %>%
+    select_if(~ length(unique(.)) > 1) %>% #remove columns that have the same value for all columns
+    mutate_at(vars(-id), list(~scales::rescale(.,to=c(0, 1)))) %>% #rescale variables
+    arrange(id) %>%
+    column_to_rownames("id")
+
+  as.tibble(dbFD_joggle(trait_mat, species_mat, w.abun = FALSE))[,1:8] %>%
+    cbind(years, .)
+}
+
+get_fd_safe <- possibly(get_fd, NA)
+
+test_data <- biotime_traits %>% 
+  group_by(study_id, plot) %>%
+  nest() %>%
+  mutate(metrics = map(data, ~get_fd_safe(.x))) %>%
+  filter(!is.na(metrics)) %>%
+  select(-data) %>%
+  unnest()
+```

### 2019-08-08 analysis/merge_data.Rmd
debug fd calculation with joggle

@@
   trait_mat <- data %>%
-    select(-c(id, taxa, record_id, day, month, year, sample_desc, id_species, 
+    select(-c(taxa, record_id, day, month, year, sample_desc, id_species, 
               latitude, longitude, sum_abundance, sum_biomass, sourceName, scientificName)
@@
     select_if(~ length(unique(.)) > 1) %>% #remove columns that have the same value for all columns
-    mutate_at(vars(-id), list(~scales::rescale(.,to=c(0, 1)))) %>% #rescale variables
+    mutate_at(vars(-id, -diet_5cat), list(~scales::rescale(.,to=c(0, 1)))) %>% #rescale variables
     arrange(id) %>%
@@
 
-get_fd_safe <- possibly(get_fd, NA)
+#get_fd_safe <- possibly(get_fd, NA)
 
@@
   nest() %>%
-  mutate(metrics = map(data, ~get_fd_safe(.x))) %>%
-  filter(!is.na(metrics)) %>%
+  mutate(metrics = map(data, ~get_fd(.x))) %>%
   select(-data) %>%

### 2019-08-08 R/dbFD_joggle.R

@@ -467,7 +467,7 @@ dbFD_joggle <- function (x, a, w, w.abun = TRUE, stand.x = TRUE, ord = c("podani
                   "\n")
         }
         if (stand.FRic) {
-          hull.all <- convhulln(traits.FRic, "FA", "QJ")
+          hull.all <- convhulln(traits.FRic, c("FA", "QJ"))
           FRic.all <- hull.all$vol
         }
       }
@@ -625,7 +625,7 @@ dbFD_joggle <- function (x, a, w, w.abun = TRUE, stand.x = TRUE, ord = c("podani
           if (!warning)
             thresh <- 3
           if (nb.sp[i] >= thresh) {
-            convhull <- convhulln(tr.FRic, "FA", "QJ")
+            convhull <- convhulln(tr.FRic, c("FA", "QJ"))
             if (!stand.FRic)
               FRic[i] <- convhull$vol
             if (stand.FRic)
@@ -669,7 +669,7 @@ dbFD_joggle <- function (x, a, w, w.abun = TRUE, stand.x = TRUE, ord = c("podani
     if (calc.FDiv & calc.FRic) {
       if (any(x.class2 == "numeric") & dim(tr.FRic)[2] >
           1 & nb.sp[i] >= 3) {
-        vert0 <- convhulln(tr.FRic, "Fx TO 'vert.txt'", "QJ")
+        vert0 <- convhulln(tr.FRic, c("Fx TO 'vert.txt'", "QJ"))
         vert1 <- scan("vert.txt", quiet = T)
         vert2 <- vert1 + 1
         vertices <- vert2[-1]

### 2019-08-08 analysis/merge_data.Rmd
debug

--- 2019-08-07 analysis/merge_data.Rmd
@@
 get_fd <- function(data){
-  #print(max(data$record_id))
+  print(max(data$record_id))
   species_mat <- data %>%
     select(year, id) %>% 
+    distinct() %>%
     arrange(year, id) %>%
@@
     select_if(~ length(unique(.)) > 1) %>% #remove columns that have the same value for all columns
-    mutate_at(vars(-id), list(~scales::rescale(.,to=c(0, 1)))) %>% #rescale variables
+    mutate_at(vars(-id, -ends_with("5cat")), list(~scales::rescale(.,to=c(0, 1)))) %>% #rescale variables
     arrange(id) %>%
@@
   mutate(metrics = map(data, ~get_fd_safe(.x))) %>%
-  filter(!is.na(metrics)) %>%
   select(-data) %>%

### 2019-08-14 analysis/merge_data.Rmd
Switch to group_map, write out fd data for each site/plot

@@
 
-get_fd <- function(data){
+get_fd <- function(data, study, plot){
   print(max(data$record_id))
@@
 
+  dir.create(paste0(here::here("data"), "/fd_output"), showWarnings = FALSE)
+
   as.tibble(dbFD_joggle(trait_mat, species_mat, w.abun = FALSE))[,1:8] %>%
-    cbind(years, .)
+    cbind(years, .) %>%
+    mutate(study_id = study, plot = plot) %>%
+    write_tsv(paste0(here::here("data"), "/fd_output/", "s", study, "_p", plot, "_fd.tsv"))
 }
@@
 
-test_data <- biotime_traits %>% 
+fd_data <- biotime_traits %>% 
+  filter(study_id == 56) %>%
+  group_by(study_id, year, plot) %>%
+  filter(n_distinct(id) > 4) %>%
+  ungroup() %>%
   group_by(study_id, plot) %>%
-  nest() %>%
-  mutate(metrics = map(data, ~get_fd_safe(.x))) %>%
-  select(-data) %>%
-  unnest()
+  group_map(~get_fd(.x, study = .y$study_id, plot = .y$plot))
 ```

### 2019-08-26 analysis/merge_data.Rmd

@@
 fd_data <- biotime_traits %>% 
-  filter(study_id == 56) %>%
+  filter(!is.na(-plot)) %>%
   group_by(study_id, year, plot) %>%
@@
   group_by(study_id, plot) %>%
-  group_map(~get_fd(.x, study = .y$study_id, plot = .y$plot))
+  group_map(~get_fd_safe(.x, study = .y$study_id, plot = .y$plot))
 ```
+
+Read data back in:
+```{r}
+fd_data <- dir(here::here("data", "fd_output"), "*.tsv") %>%
+  paste0(here::here("data", "fd_output"), "/", .) %>%
+  map_dfr(read_tsv)
+```

### 2019-09-03 data-raw/web_scrape.Rmd
Web scraping fishtraits database with rselenium.

---
title: "fishtraits"
author: "Kari Norman"
date: "8/28/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(RSelenium)
```

```{r}
clickObject <- function(selector){
  object <- rd$findElement(using = 'css selector', selector)
  object$clickElement()
}
```
connect to Selenium server and navigate to the page
```{r}
driver <- rsDriver(browser = c("chrome"), chromever = "76.0.3809.126")
rd <- driver[["client"]]
rd$open()

rd$navigate("http://www.fishtraits.info/search/taxo/")
```

see if we can get the list of all possible search items for the FID dropdown (second is dependent on the first input)
```{r}
#click on drop down element
clickObject('.ui-combobox .ui-icon-triangle-1-s')

#get text of the list
fid_list <- rd$findElement('css selector', '#ui-id-1')
fid_text <- unlist(strsplit(unlist(fid_list$getElementText()), "\n"))
```

Choose FID and Genus
```{r}
#select and search the FID dropdown
fid <- rd$findElement(using = 'css selector', "#fid-combo")
fid$clearElement()
#pass to search box
fid$sendKeysToElement(list(fid_text[1]))
#click matching one
clickObject("#ui-id-1")

#get list of Genus's that are now possible
clickObject(".ui-combobox2 .ui-icon-triangle-1-s")
genus_list <- rd$findElement('css selector', '#ui-id-2')
genus_text <- unlist(strsplit(unlist(genus_list$getElementText()), "\n"))

#select and search the Genus dropdown
genus <- rd$findElement(using = 'css selector', "#genus-combo")
genus$clearElement()
genus$sendKeysToElement(list(genus_text[1]))
#click matching one
clickObject('#ui-id-2')
```

Pick the columns we want
```{r}
#make sure the appropriate column boxes are checked
columns <- rd$findElement(using = 'css selector', '#ui-accordion-ac1-header-1')
columns$clickElement()

#selectors for all boxes that need to be checked
```

Submit search and download
```{r}
#select search box
clickObject("#searchSubmit")
#and download
clickObject("#downloadbtn")
```

```{r}
driver$server$stop()
```

### 2019-09-03 data-raw/web_scrape.Rmd
loop across FID's and genus's

@@
 ```{r}
+
+for (i in 1:length(fid_text)) {
   #select and search the FID dropdown
@@
   #pass to search box
-fid$sendKeysToElement(list(fid_text[1]))
+  fid$sendKeysToElement(list(fid_text[i]))
   #click matching one
@@
 
+  for (j in 1:length(genus_text)) {
     #select and search the Genus dropdown
@@
     genus$clearElement()
-genus$sendKeysToElement(list(genus_text[1]))
+    genus$sendKeysToElement(list(genus_text[j]))
     #click matching one
     clickObject('#ui-id-2')
-```
 
-Pick the columns we want
-```{r}
     #make sure the appropriate column boxes are checked
@@
     #selectors for all boxes that need to be checked
+
+    #select search box
+    clickObject("#searchSubmit")
+    #and download
+    clickObject("#downloadbtn")
+  }
+}
+```
+
+Pick the columns we want
+```{r}
+
 ```
@@
 ```{r}
-#select search box
-clickObject("#searchSubmit")
-#and download
-clickObject("#downloadbtn")
+
 ```

### 2019-09-03 data-raw/web_scrape.Rmd
Add intentional time lags so dropdown menus have time to catch up with click commands.

@@
 
-for (i in 1:length(fid_text)) {
+for (i in 6:length(fid_text)) {
   #select and search the FID dropdown
@@
   #pass to search box
+  print(fid_text[i])
   fid$sendKeysToElement(list(fid_text[i]))
   #click matching one
+  Sys.sleep(0.15)
   clickObject("#ui-id-1")
@@
   clickObject(".ui-combobox2 .ui-icon-triangle-1-s")
+  Sys.sleep(0.15)
   genus_list <- rd$findElement('css selector', '#ui-id-2')
   genus_text <- unlist(strsplit(unlist(genus_list$getElementText()), "\n"))
+  print(genus_text)
 
@@
     genus$sendKeysToElement(list(genus_text[j]))
+    Sys.sleep(0.15)
     #click matching one
@@
     #make sure the appropriate column boxes are checked
-    columns <- rd$findElement(using = 'css selector', '#ui-accordion-ac1-header-1')
-    columns$clickElement()
-
+    # columns <- rd$findElement(using = 'css selector', '#ui-accordion-ac1-header-1')
+    # columns$clickElement()
+    # 
     #selectors for all boxes that need to be checked

### 2019-09-03 data-raw/web_scrape.Rmd
clean up

@@
 
-Choose FID and Genus
+Choose FID and Genus.
+All the boxes under the "columns" header unfortunately need to be checked manually because css selectors couldn't be identified for each individual box.
+Only have to do it once though!
 ```{r}
 
-for (i in 6:length(fid_text)) {
+for (i in 1:length(fid_text)) {
   #select and search the FID dropdown
@@
   #click matching one
-  Sys.sleep(0.15)
+  Sys.sleep(0.25)
   clickObject("#ui-id-1")
@@
   clickObject(".ui-combobox2 .ui-icon-triangle-1-s")
-  Sys.sleep(0.15)
+  Sys.sleep(0.25)
   genus_list <- rd$findElement('css selector', '#ui-id-2')
@@
 
-    #make sure the appropriate column boxes are checked
-    # columns <- rd$findElement(using = 'css selector', '#ui-accordion-ac1-header-1')
-    # columns$clickElement()
-    # 
-    #selectors for all boxes that need to be checked
-
     #select search box
@@
 
-Pick the columns we want
-```{r}
-
-```
-
-Submit search and download
-```{r}
-
-```
-
 ```{r}

### 2019-09-06 data-raw/fishtraits.Rmd
Fishtraits cleaning

---
title: "fishtraits"
author: "Kari Norman"
date: "9/4/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(taxadb)
```

Read in files
```{r}
fishtraits <- dir(here::here("inst", "extdata", "fishtraits"), "*.csv") %>%
  paste0(here::here("inst", "extdata", "fishtraits"), "/", .) %>%
  map_dfr(read_csv) %>% 
  filter(id != 810) %>% #one species has a duplicate entry with no data for some reason
  select(scientificName = scientificname, itistsn, commonName = commonname, altcommonName = othernames, 
                      diet_nonfeed = nonfeed, diet_benthic = benthic, diet_surwcol = surwcol, diet_algphyto = algphyto, diet_macvascu = macvascu, 
                      diet_detritus = detritus, diet_inv = invlvfsh, diet_fshcrcrb = fshcrcrb, diet_blood = blood, diet_eggs = eggs, diet_other = other, 
                      max_length = maxtl, matuage, longevity, fecundity, serial, season, starts_with("a_"), starts_with("b_"), starts_with("c1_"),
                      euryhaline, mintemp, maxtemp, muck, claysilt, sand, gravel, cobble, boulder, bedrock, vegetat, debrdetr, lwd,
                      pelagic, preflot, preflen, largeriv, smallriv, creek, sprgsubt, lacustrine, potanadr, lowland, upland, montane,
                      slowcurr, modcurr, fastcurr)

#find species without a TRUE habitat type 
missing_hab <- fishtraits %>% 
  #select(starts_with("a_"), starts_with("b_")) %>%
  filter_at(vars(starts_with("a_"), starts_with("b_"), starts_with("c1_")),  ~ !. %in% TRUE) %>%
  pull(itistsn) %>%
  unique()

fishtraits_clean <- fishtraits %>%
  gather(habitat, value, starts_with("a_"), starts_with("b_"), starts_with("c1_")) %>%
  mutate(habitat = case_when(
    itistsn %in% missing_hab ~ NA_character_,
    TRUE ~ habitat
    )
  ) %>% 
  filter(value == TRUE | is.na(habitat), itistsn != 0) %>%
  select(-value) %>%
  distinct()

```

### 2019-09-09 data-raw/fishtraits.Rmd
Get id's for fishtraits

@@
 ```{r}
-fishtraits <- dir(here::here("inst", "extdata", "fishtraits"), "*.csv") %>%
+fishtraits_data <- dir(here::here("inst", "extdata", "fishtraits"), "*.csv") %>%
   paste0(here::here("inst", "extdata", "fishtraits"), "/", .) %>%
@@
   filter(id != 810) %>% #one species has a duplicate entry with no data for some reason
-  select(scientificName = scientificname, itistsn, commonName = commonname, altcommonName = othernames, 
+  select(scientificName = scientificname, itistsn,
                       diet_nonfeed = nonfeed, diet_benthic = benthic, diet_surwcol = surwcol, diet_algphyto = algphyto, diet_macvascu = macvascu, 
@@
 #find species without a TRUE habitat type 
-missing_hab <- fishtraits %>% 
+missing_hab <- fishtraits_data %>% 
   #select(starts_with("a_"), starts_with("b_")) %>%
@@
 
-fishtraits_clean <- fishtraits %>%
+fishtraits_clean <- fishtraits_data %>%
   gather(habitat, value, starts_with("a_"), starts_with("b_"), starts_with("c1_")) %>%
@@
   select(-value) %>%
-  distinct()
+  distinct() %>%
+  mutate_at(vars(-c(scientificName, itistsn, max_length, matuage, longevity, fecundity, 
+                    season, mintemp, maxtemp, habitat)), ~ as.integer(.x))
 
 ```
+
+Get id's! They seemingly already have ITIS id's, but let's double check that they're up to date with what we have through taxadb
+```{r}
+fishtraits <- fishtraits_clean %>%
+  mutate(itistsn = paste0("ITIS:", itistsn), 
+         id = get_ids(scientificName))
+
+#There are no instances where the database id doesn't match the id from taxadb
+fishtraits %>% filter(itistsn != id) %>% dim()
+
+#so we'll just keep that ID column
+fishtraits <- fishtraits %>%
+  mutate(id = itistsn) %>%
+  select(-itistsn)
+```
+
+```{r}
+usethis::use_data(fishtraits)
+```

### 2019-09-10 data-raw/marinetraits.Rmd
Formating marine traits data - not high quality

---
title: "marine traits"
author: "Kari Norman"
date: "9/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
mt <- list.files(system.file("extdata", "marinetraits", package = "biodivTS")) %>%
  system.file("extdata", "marinetraits", ., package = "biodivTS") %>%
  map_dfr(., read.csv) %>%
  select(AphiaID, scientificName = ScientificName, taxonRank, taxonomicStatus, measurementID, 
         measurementType, measurementValue, measurementUnit) %>%
  filter(taxonRank == "Species") %>% 
  distinct()

#create ID column
id_num <- 0
obsID <- vector("list", length = length(mt$measurementType))

for (i in 1:length(mt$measurementType)){
  if (isTRUE(grepl(">", mt$measurementType[i]))){
    obsID[i] <- id_num
  } else{
    id_num <- id_num + 1
    obsID[i] <- id_num
  }
}

mt <- cbind(mt, obsID = unlist(obsID)) %>% 
  spread(measurementType, measurementValue) %>%
  select(-c("Body size > Stage", "Feedingtype > Host", "Feedingtype > Stage", 
            "Body size > Locality (MRGID)", "Body size > Dimension", "Body size > Gender"))

coalesce_all_columns <- function(df) {
  return(coalesce(!!! as.list(df)))
}

mt_clean <- mt %>%
  group_by(obsID) %>%
  summarise_all(coalesce_all_columns) %>% 
  group_by(scientificName) %>%
  summarise_all(coalesce_all_columns)
```

### 2019-09-11 analysis/merge_data.Rmd
Code for picking up FD calculations where they leave off after crashing.

--- 2019-08-14 analysis/merge_data.Rmd
@@
 library(FD)
+devtools::load_all()
 ```
@@
 
-fd_data <- biotime_traits %>% 
-  filter(study_id == 56) %>%
+fd <- biotime_traits %>% 
+  filter(!is.na(plot)) %>%
   group_by(study_id, year, plot) %>%
@@
   group_by(study_id, plot) %>%
-  group_map(~get_fd(.x, study = .y$study_id, plot = .y$plot))
+  group_map(~get_fd_safe(.x, study = .y$study_id, plot = .y$plot))
 ```
+
+Read data back in:
+```{r}
+files <-  dir(here::here("data", "fd_output"), "*.tsv")
+
+fd_data <- files %>%
+  paste0(here::here("data", "fd_output"), "/", .) %>%
+  map_dfr(read_tsv)
+
+fd <- biotime_traits %>% 
+  filter(!is.na(plot), !study_id %in% c(195, 360)) %>%
+  anti_join(fd_data) %>%
+  group_by(study_id, year, plot) %>%
+  filter(n_distinct(id) > 4) %>%
+  ungroup() %>%
+  group_by(study_id, plot) %>%
+  group_map(~get_fd_safe(.x, study = .y$study_id, plot = .y$plot))
+
+#4610981
+#7611121
+
+```

### 2019-09-12 data-raw/bientraits.Rmd
First commit bien traits

---
title: "BIEN trait"
author: "Kari Norman"
date: "9/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(BIEN)
library(taxadb)

source(here::here("data-raw", "helper_functions.R"))
```

Let's get them traits
```{r}
trait_types <- na.omit(BIEN_trait_list())

BIEN_data <- map_dfr(trait_types$trait_name, BIEN_trait_trait)

bien <- BIEN_data %>%
  select(scientificName = scrubbed_species_binomial, trait_name, trait_value, unit, method, latitude, longitude, elevation_m, obsID = id) %>%
  spread()
```

Get id's for just the unique 
```{r}
bien_ids <- bien %>% 
  select(scientifcName) %>%
  distinct() %>%
  mutate(id = get_ids(scientificName))
```

```{r}
usethis::use_data(bien)
```

### 2019-09-19 data-raw/fishbase.Rmd
Get traits using rfishbase

---
title: "Fishbase"
author: "Kari Norman"
date: "9/19/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rfishbase)

```

Potential tables of interest: diet, ecology, ecosystem, fecundity, fooditems, maturity, popchar, popgrowth, reproduc, spawning, speed

other: morphdat, oxygen

```{r}
docs(table = "diet")
```

### 2019-09-19 data-raw/elton.Rmd
Make function for resolving multiple trait entries to one id, resolve for elton database as well.
Adjustments with new itis table.

@@
 bird_trait <- read_csv(system.file("extdata", "elton_traits/elton_traits_BirdFuncDat.csv", package = "biodivTS"))
-mamm_trait <- read.csv(system.file("extdata", "elton_traits/elton_traits_MammFuncDat.csv", package = "biodivTS"))
+mamm_trait <- read.csv(system.file("extdata", "elton_traits/elton_traits_MammFuncDat.csv", package = "biodivTS")) %>%
+  rename(nocturnal = activity_nocturnal, crepuscular = activity_crepuscular, diurnal = activity_diurnal)
 ```
@@
 
-Are there any unresolved ids for birds? Yes - just one, which is a synonym to two different accepted id's
+Are there any unresolved ids for birds? Yes - just one, which matched to a species and a subspecies
 ```{r}
@@
 bird_ids <- unres_trait %>% 
-  filter(!is.na(kingdom)) %>%
+  filter(taxonRank == "species") %>%
   bind_rows(bird_ids %>% filter(!sort %in% unres_trait$sort))
@@
 
-What about for mammals? Yes - also a synonym to two different accepted id's. Further research shows that the Callospermophilus saturatus uses the subgenus, so we'll keep the name with the genus 
+What about for mammals? Nope!
 ```{r}
@@
 unres_mamm_trait
-
-mammal_ids <- unres_mamm_trait %>% 
-  filter(acceptedNameUsageID == "ITIS:632452") %>%
-  bind_rows(mammal_ids %>% filter(!sort %in% unres_mamm_trait$sort))
 ```
@@
 
+Check for multiple matches to the same ID
+```{r}
+elton_bird <- undupe_ids(elton_bird)
+
+elton_mamm <- undupe_ids(elton_mamm)
+```
+
 ```{r}

### 2019-09-20 analysis/merge_data.Rmd
Update with correct itis table

--- 2019-08-26 analysis/merge_data.Rmd
@@
 
-check if all bird & mammal biotime data w/id's has an elton match. There are 63 species without a match (54 are birds).
+check if all bird & mammal biotime data w/id's has an elton match. There are 73 species without a match (63 are birds).
 ```{r}
@@
 
-and 19869 species without an ID, some of which are species codes
+and 20136 species without an ID, some of which are species codes
 ```{r}

### 2019-10-03 data-raw/amphibio.Rmd
Make function for resolving multiple trait entries to one id, resolve for elton database as well.
Standardize trait column names, simplify trait matrix column selection.

--- 2019-07-03 data-raw/amphibio.Rmd
@@
 library(tidyverse)
+library(taxadb)
 
@@
 
-amph <- read_csv(system.file("extdata", "amphibio/AmphiBIO_v1.csv", package = "biodivTS"))
+amph <- read_csv(system.file("extdata", "amphibio/AmphiBIO_v1.csv", package = "biodivTS")) %>%
+  rename_all(tolower) %>%
+  rename(diet_leaves = leaves, diet_flowers = flowers, diet_seed = seeds, diet_fruit = fruits, diet_inv = arthro, diet_vert = vert,
+         diurnal = diu, nocturnal = noc, crepuscular = crepu, bodymass_value = body_mass_g) %>%
+  mutate(diet_seed = as.numeric(diet_seed))
 
 amphibio <- amph %>%
-  mutate(id = get_ids(Species, "itis")) %>%
+  mutate(id = get_ids(species, "itis")) %>%
   filter(!is.na(id)) %>%
   mutate(scientificName = get_names(id, "itis")) %>%
-  select(id, scientificName, Species) %>%
-  right_join(select(amph, -id), by = "Species") %>%
-  rename(sourceName = Species) %>%
-  select(-c("Order", "Family", "Genus"))
+  select(id, scientificName, species) %>%
+  right_join(select(amph, -id), by = "species") %>%
+  rename(sourceName = species) %>%
+  select(-c("order", "family", "genus"))
 
@@
 
+Select relevant trait columns
+```{r}
+amphibio <- amphibio %>%
+  select(-obs)
+```
+
+Check for multiple matches to the same ID
+```{r}
+amphibio <- undupe_ids(amphibio)
+```
+
 ```{r}

### 2019-10-03 data-raw/biotime.Rmd
Also check common names against alternate providers.

@@
   left_join(read_csv(system.file("extdata", "biotime/biotime_query.csv", package = "biodivTS")), by = "study_id")
+```
+
+Many of the studies with only one plot have NA for the plot number, but we want to distinguish between NA's in studies that have multiple plots and unlabled studies.
+```{r}
+#find studies that only have one plot labeled NA
+sing_study <- data %>% 
+  select(study_id, plot) %>%
+  distinct() %>%
+  group_by(study_id) %>%
+  count() %>%
+  filter(n == 1) %>%
+  pull(study_id)
+
+#replace plot NA's that are really studies with only one plot with a plot label 
+data <- data %>% 
+  mutate(plot = case_when(
+    study_id %in% sing_study & is.na(plot) ~ 1,
+    TRUE ~ plot
+  ))
 ```
@@
 #Check for potential matches from synonyms found in the other providers
-syn_res <- match_providers(biotime_data, "itis")
+syn_res <- match_providers(biotime_data, "itis", common = TRUE)
 

### 2019-10-07 analysis/merge_data.Rmd
Create standard FD and study tables

--- 2019-09-11 analysis/merge_data.Rmd
@@
 
-check if all bird & mammal biotime data w/id's has an elton match. There are 63 species without a match (54 are birds).
+check if all bird & mammal biotime data w/id's has an elton match. There are 68 species without a match (63 are birds).
 ```{r}
@@
 biotime_traits_list <- map2(taxa, traits, ~merge_data(taxa = .x, trait_data = .y))
+
+#this gives a coverage for each trait dataset, so potentially a lot of coverage with one dataset and no coverage on all the rest
+#need to get rid of zeros when there is coverage from another dataset so we only have one entry per study
 coverage <- map_dfr(biotime_traits_list, get_coverage) %>% 
-  filter(perc > percent)
+  filter(perc > 0)%>%
+  right_join(biotime_data %>% select(study_id) %>% distinct()) %>%
+  mutate(perc = replace_na(perc, 0))
+
+inc_coverage <- coverage %>% filter(perc > percent)
 
 biotime_traits <- map_dfr(biotime_traits_list, ~drop_na(., scientificName)) %>%
-  filter(study_id %in% coverage$study_id)
+  filter(study_id %in% inc_coverage$study_id)
 ```
@@
   ungroup() %>%
+  filter(study_id == 516) %>%
   group_by(study_id, plot) %>%
-  group_map(~get_fd_safe(.x, study = .y$study_id, plot = .y$plot))
+  group_map(~get_fd_safe(.x, study = .y$study_id, plot = .y$plot)) %>%
+  bind_rows()
 ```
@@
 ```{r}
-files <-  dir(here::here("data", "fd_output"), "*.tsv")
-
-fd_data <- files %>%
-  paste0(here::here("data", "fd_output"), "/", .) %>%
-  map_dfr(read_tsv)
-
-fd <- biotime_traits %>% 
-  filter(!is.na(plot), !study_id %in% c(195, 360)) %>%
-  anti_join(fd_data) %>%
-  group_by(study_id, year, plot) %>%
-  filter(n_distinct(id) > 4) %>%
-  ungroup() %>%
-  group_by(study_id, plot) %>%
-  group_map(~get_fd_safe(.x, study = .y$study_id, plot = .y$plot))
-
-#4610981
-#7611121
+# files <-  dir(here::here("data", "fd_output"), "*.tsv")
+# 
+# fd_data <- files %>%
+#   paste0(here::here("data", "fd_output"), "/", .) %>%
+#   map_dfr(read_tsv)
+# 
+# fd <- biotime_traits %>% 
+#   filter(!is.na(plot), !study_id %in% c(195, 360)) %>%
+#   anti_join(fd_data) %>%
+#   group_by(study_id, year, plot) %>%
+#   filter(n_distinct(id) > 4) %>%
+#   ungroup() %>%
+#   group_by(study_id, plot) %>%
+#   group_map(~get_fd_safe(.x, study = .y$study_id, plot = .y$plot))
+# 
+# #4610981
+# #7611121
 
 ```
+
+Format tables
+```{r}
+fd_table <- fd %>% 
+  select(year, study_id, plot, richness = nbsp, everything(), -sing.sp) %>%
+  gather(metric, value, contains("F"), RaoQ) %>%
+  View()
+
+study_table <- biotime_data %>%
+  group_by(study_id) %>%
+  mutate(min_year = min(year), max_year = max(year)) %>% 
+  #select(study_id, plot, taxa, latitude, longitude, min_year, max_year) %>%
+  select(study_id, taxa, min_year, max_year) %>%
+  distinct() %>%
+  left_join(coverage, by = "study_id")
+```

### 2019-10-07 analysis/merge_data.Rmd
Revamp how coverage is measured

@@
 ```{r}
-merge_data <- function(taxa, trait_data){
+merge_data <- function(trait_data){
   biotime_data %>%
     select(-sourceName, -scientificName) %>%
-  filter(taxa == taxa) %>%
   left_join(trait_data, by = "id",
-            na_matches = "never")
+            na_matches = "never") %>%
+    drop_na(scientificName)
 }
-
-get_coverage <- function(taxa_data){
-  coverage <- taxa_data %>%
-  group_by(study_id) %>%
-  summarise(perc = sum(!is.na(sourceName))/n()) %>%
-  arrange(desc(perc))
-}
-
-get_coverage(merge_data("Birds", elton_bird)) %>% View
-get_coverage(merge_data("Mammals", elton_mamm)) %>% View()
-get_coverage(merge_data("Amphibians", amphibio)) %>% View()
 ```
@@
 traits <- list(elton_bird, elton_mamm, amphibio)
-taxa <- list("Birds", "Mammals", "Amphibians")
 
-biotime_traits_list <- map2(taxa, traits, ~merge_data(taxa = .x, trait_data = .y))
+comp_biotime_traits <- map_dfr(traits, merge_data) %>%
+  right_join(biotime_data %>% select(-sourceName, -scientificName))
+
+#always check that one species didn't match to multiple datasets, below should return and empty dataframe
+#comp_biotime_traits %>% group_by(record_id) %>% filter(n() > 1) %>% dim()
 
@@
 #need to get rid of zeros when there is coverage from another dataset so we only have one entry per study
-coverage <- map_dfr(biotime_traits_list, get_coverage) %>% 
-  filter(perc > 0)%>%
-  right_join(biotime_data %>% select(study_id) %>% distinct()) %>%
-  mutate(perc = replace_na(perc, 0))
+coverage <- comp_biotime_traits %>%
+  group_by(study_id) %>%
+  summarise(perc = sum(!is.na(sourceName))/n()) %>%
+  arrange(desc(perc))
 
@@
 
-biotime_traits <- map_dfr(biotime_traits_list, ~drop_na(., scientificName)) %>%
+biotime_traits <- comp_biotime_traits %>%
   filter(study_id %in% inc_coverage$study_id)

### 2019-10-08 analysis/merge_data.Rmd
Drop species without trait data from biotime_data.

@@
 biotime_traits <- comp_biotime_traits %>%
-  filter(study_id %in% inc_coverage$study_id)
+  filter(study_id %in% inc_coverage$study_id, !is.na(sourceName))
 ```
@@
 usethis::use_data(biotime_traits)
+```
+
+Create table of info for each study
+```{r}
+study_table <- biotime_data %>%
+  group_by(study_id) %>%
+  mutate(min_year = min(year), max_year = max(year)) %>% 
+  #select(study_id, plot, taxa, latitude, longitude, min_year, max_year) %>%
+  select(study_id, taxa, min_year, max_year) %>%
+  distinct() %>%
+  left_join(coverage %>% select(study_id, coverage = perc), by = "study_id")
+
+usethis::use_data(study_table)
 ```
@@
   View()
-
-study_table <- biotime_data %>%
-  group_by(study_id) %>%
-  mutate(min_year = min(year), max_year = max(year)) %>% 
-  #select(study_id, plot, taxa, latitude, longitude, min_year, max_year) %>%
-  select(study_id, taxa, min_year, max_year) %>%
-  distinct() %>%
-  left_join(coverage, by = "study_id")
 ```

### 2019-10-08 analysis/merge_data.Rmd
Add the number of traits input to the FD output

@@
     cbind(years, .) %>%
-    mutate(study_id = study, plot = plot) %>%
+    mutate(study_id = study, plot = plot, n_traits = ncol(trait_mat)) %>%
     write_tsv(paste0(here::here("data"), "/fd_output/", "s", study, "_p", plot, "_fd.tsv"))

### 2019-10-22 analysis/merge_data.Rmd
Adapt to a mapping framework

@@
 
-fd <- biotime_traits %>% 
+data <- biotime_traits %>% 
   filter(!is.na(plot)) %>%
@@
   ungroup() %>%
-  filter(study_id == 516) %>%
-  group_by(study_id, plot) %>%
-  group_map(~get_fd_safe(.x, study = .y$study_id, plot = .y$plot)) %>%
-  bind_rows()
+  filter(study_id == 516)
+
+combs <- cross2(unique(data$study_id), unique(data$plot))
+
+fd <- map_df(combs, ~ get_fd_safe(data = data, study = .x[1], plot = .x[2]))  
 ```

### 2019-10-22 analysis/merge_data.Rmd
Merge conflict

@@
 
-get_fd <- function(data, study, plot){
+get_fd <- function(data, study, plot_num){
   print(max(data$record_id))
+
+  data <- data %>%
+    filter(study_id == study, plot == plot_num)
+
   species_mat <- data %>%
@@
 
-  as.tibble(dbFD_joggle(trait_mat, species_mat, w.abun = FALSE))[,1:8] %>%
+  as_tibble(dbFD_joggle(trait_mat, species_mat, w.abun = FALSE))[,1:8] %>%
     cbind(years, .) %>%
-    mutate(study_id = study, plot = plot, n_traits = ncol(trait_mat)) %>%
-    write_tsv(paste0(here::here("data"), "/fd_output/", "s", study, "_p", plot, "_fd.tsv"))
+    mutate(study_id = unlist(study), plot = unlist(plot_num)) %>%
+    write_tsv(paste0(here::here("data"), "/fd_output/", "s", study, "_p", plot_num, "_fd.tsv"))
 }
@@
 data <- biotime_traits %>% 
-  filter(!is.na(plot)) %>%
+  filter(!is.na(plot), !is.na(scientificName)) %>% #remove unlabeled plots and observations without traits
   group_by(study_id, year, plot) %>%
@@
 
-fd <- map_df(combs, ~ get_fd_safe(data = data, study = .x[1], plot = .x[2]))  
+fd <- map_dfr(combs, ~ get_fd_safe(data = data, study = .x[1], plot = .x[2]))  
 ```
@@
 ```{r}
-# files <-  dir(here::here("data", "fd_output"), "*.tsv")
-# 
-# fd_data <- files %>%
-#   paste0(here::here("data", "fd_output"), "/", .) %>%
-#   map_dfr(read_tsv)
-# 
-# fd <- biotime_traits %>% 
-#   filter(!is.na(plot), !study_id %in% c(195, 360)) %>%
-#   anti_join(fd_data) %>%
-#   group_by(study_id, year, plot) %>%
-#   filter(n_distinct(id) > 4) %>%
-#   ungroup() %>%
-#   group_by(study_id, plot) %>%
-#   group_map(~get_fd_safe(.x, study = .y$study_id, plot = .y$plot))
-# 
-# #4610981
-# #7611121
+files <-  dir(here::here("data", "fd_output"), "*.tsv")
+
+fd_data <- files %>%
+  paste0(here::here("data", "fd_output"), "/", .) %>%
+  map_dfr(read_tsv)
+
+fd <- biotime_traits %>%
+  filter(!is.na(plot), !study_id %in% c(195, 360)) %>%
+  anti_join(fd_data) %>%
+  group_by(study_id, year, plot) %>%
+  filter(n_distinct(id) > 4) %>%
+  ungroup() %>%
+  group_by(study_id, plot) %>%
+  group_map(~get_fd_safe(.x, study = .y$study_id, plot = .y$plot))
+
+#4610981
+#7611121
 

### 2019-10-22 analysis/merge_data.Rmd
add parallelization

@@
 ```{r}
+library(furrr)
 load(here::here("data", "biotime_traits.rda"))
@@
 
-fd <- map_dfr(combs, ~ get_fd_safe(data = data, study = .x[1], plot = .x[2]))  
+plan(multiprocess)
+fd <- future_map_dfr(combs, ~ get_fd_safe(data = data, study = .x[1], plot = .x[2]))  
 ```

### 2019-10-22 analysis/merge_data.Rmd
Create temporary directories to avoid race condition for the convexhull vertice writeout

@@
 
-  dir.create(paste0(here::here("data"), "/fd_output"), showWarnings = FALSE)
+  fs::dir_create(paste0(here::here("data"), "/fd_output"))
+
+  #the convexhull c code creates an output file of vertices, in order to not have multiple threads writing
+  #to the same file we have to create individual working directories
+  tmp_dir <- paste0(here::here("data", "fd_output", "tmp"), "/tmp_", study, plot_num)
+  fs::dir_create(tmp_dir)
+  setwd(tmp_dir)
 
@@
     write_tsv(paste0(here::here("data"), "/fd_output/", "s", study, "_p", plot_num, "_fd.tsv"))
+
+  setwd(here::here())
 }

### 2019-10-22 analysis/merge_data.Rmd
Return the correct object from FD calculation function.

@@
 
-  as_tibble(dbFD_joggle(trait_mat, species_mat, w.abun = FALSE))[,1:8] %>%
+  fd_out <- as_tibble(dbFD_joggle(trait_mat, species_mat, w.abun = FALSE))[,1:8] %>%
     cbind(years, .) %>%
@@
   setwd(here::here())
+  return(fd_out)
 }
@@
 ```{r}
-files <-  dir(here::here("data", "fd_output"), "*.tsv")
-
-fd_data <- files %>%
-  paste0(here::here("data", "fd_output"), "/", .) %>%
-  map_dfr(read_tsv)
-
-fd <- biotime_traits %>%
-  filter(!is.na(plot), !study_id %in% c(195, 360)) %>%
-  anti_join(fd_data) %>%
-  group_by(study_id, year, plot) %>%
-  filter(n_distinct(id) > 4) %>%
-  ungroup() %>%
-  group_by(study_id, plot) %>%
-  group_map(~get_fd_safe(.x, study = .y$study_id, plot = .y$plot))
-
-#4610981
-#7611121
+# files <-  dir(here::here("data", "fd_output"), "*.tsv")
+# 
+# fd_data <- files %>%
+#   paste0(here::here("data", "fd_output"), "/", .) %>%
+#   map_dfr(read_tsv)
+# 
+# fd <- biotime_traits %>%
+#   filter(!is.na(plot), !study_id %in% c(195, 360)) %>%
+#   anti_join(fd_data) %>%
+#   group_by(study_id, year, plot) %>%
+#   filter(n_distinct(id) > 4) %>%
+#   ungroup() %>%
+#   group_by(study_id, plot) %>%
+#   group_map(~get_fd_safe(.x, study = .y$study_id, plot = .y$plot))
+# 
+# #4610981
+# #7611121
 

### 2019-10-22 analysis/merge_data.Rmd
Include number of traits in the data/metric table.

@@
     cbind(years, .) %>%
-    mutate(study_id = unlist(study), plot = unlist(plot_num)) %>%
+    mutate(study_id = unlist(study), plot = unlist(plot_num), trait_count = dim(trait_mat)[2]) %>%
     write_tsv(paste0(here::here("data"), "/fd_output/", "s", study, "_p", plot_num, "_fd.tsv"))

### 2019-10-25 analysis/merge_data.Rmd
Get FD for all data

@@
   ungroup() %>%
-  filter(study_id == 516)
+  filter(!study_id %in% c(195, 360), #exclude because they break
+         !study_id %in% c(217, 327)) #exclude to get the file size smaller
 
@@
 plan(multiprocess)
-fd <- future_map_dfr(combs, ~ get_fd_safe(data = data, study = .x[1], plot = .x[2]))  
+system.time(fd <- future_map_dfr(combs, ~ get_fd_safe(data = data, study = .x[1], plot = .x[2])))
 ```

### 2019-11-13 analysis/merge_data.Rmd
modify future_map

--- 2019-10-22 analysis/merge_data.Rmd
@@
 Get FD metrics for each study
+
+FD function:
 ```{r}
@@
 get_fd <- function(data, study, plot_num){
-  print(max(data$record_id))
+  require(FD)
 
@@
     filter(study_id == study, plot == plot_num)
+
+  print(max(data$record_id))
 
@@
 get_fd_safe <- possibly(get_fd, NA)
+```
 
+First pass at calculating FD metrics
+```{r}
 data <- biotime_traits %>% 
@@
   ungroup() %>%
-  filter(study_id == 516)
+  filter(!study_id %in% c(195, 360), #exclude because they break
+         !study_id %in% c(217, 327)) #exclude to get the file size smaller
 
-combs <- cross2(unique(data$study_id), unique(data$plot))
+combs <- data %>%
+  select(study_id, plot) %>%
+  distinct()
 
 plan(multiprocess)
-fd <- future_map_dfr(combs, ~ get_fd_safe(data = data, study = .x[1], plot = .x[2]))  
+system.time(fd <- future_pmap_dfr(combs, ~ get_fd(data = data, study = .x, plot = .y)))
 ```
 
-Read data back in:
+Read data back in, excluding sites/plots as they become a problem:
 ```{r}
-# files <-  dir(here::here("data", "fd_output"), "*.tsv")
-# 
-# fd_data <- files %>%
-#   paste0(here::here("data", "fd_output"), "/", .) %>%
-#   map_dfr(read_tsv)
-# 
-# fd <- biotime_traits %>%
-#   filter(!is.na(plot), !study_id %in% c(195, 360)) %>%
-#   anti_join(fd_data) %>%
-#   group_by(study_id, year, plot) %>%
-#   filter(n_distinct(id) > 4) %>%
-#   ungroup() %>%
-#   group_by(study_id, plot) %>%
-#   group_map(~get_fd_safe(.x, study = .y$study_id, plot = .y$plot))
-# 
-# #4610981
-# #7611121
+files <-  dir(here::here("data", "fd_output"), "*.tsv")
 
+fd_data <- files %>%
+  paste0(here::here("data", "fd_output"), "/", .) %>%
+  map_dfr(read_tsv)
+
+remain <- biotime_traits %>%
+  filter(!is.na(plot), !is.na(scientificName)) %>% #remove unlabeled plots and observations without traits
+  filter(!is.na(plot), !study_id %in% c(195, 360),
+         !study_id %in% c(217, 327), #exclude to get the file size smaller
+         !study_id %in% c(59, 166, 337), #NA's in the distance matrix
+         study_id != 169,  #Zero distance error
+         !study_id %in% c(372,374), #Qhull error code 4
+         study_id != 333) %>% #species distance matrix not Euclidean after 'sqrt' correction
+  filter(study_id != 56 | plot != 24) %>% #NA's in the distance matrix
+  anti_join(fd_data, by = c("year", "study_id", "plot")) %>%
+  group_by(study_id, year, plot) %>%
+  filter(n_distinct(id) > 4) %>%
+  ungroup()
+
+combs <- remain %>%
+  select(study_id, plot) %>%
+  distinct()
+
+plan(multiprocess)
+system.time(fd <- future_pmap_dfr(combs, ~ get_fd(data = remain, study = .x, plot = .y)))
 ```
@@
 ```{r}
-fd_table <- fd %>% 
+fd_table <- fd_data %>% 
   select(year, study_id, plot, richness = nbsp, everything(), -sing.sp) %>%
-  gather(metric, value, contains("F"), RaoQ) %>%
-  View()
+  gather(metric, value, contains("F"), RaoQ)
+
+usethis::use_data(fd_table)
 ```

### 2019-11-13 data-raw/fishbase.Rmd
explore column names
little stuff

@@
 ```{r}
-docs(table = "diet")
+tables <- c("diet", "ecology", "ecosystem", "fecundity", "fooditems", "maturity", "popchar", "popgrowth", "reproduc", "spawning", "speed", "morphdat", "oxygen")
+columns <- purrr::map(tables, ~docs(table = .x)$column_name) %>% unlist()
 ```
+
+download all the tables
+```{r}
+#diet <- diet()
+
+eco <- ecology() %>% 
+  select(-c(autoctr, StockCode, EcologyRefNo, DietRemark, FoodRemark, AddRems, OutsideHost, OHRemarks, InsideHost),
+                            -ends_with("Ref"), -starts_with("Associations"))
+
+ecosys <- ecosystem() %>%
+  select(EcosystemType, Salinity, Climate, Polar, Boreal, Temperate)
+
+fec <- fecundity()
+
+food <- fooditems()
+
+mat <- maturity()
+
+```

### 2019-11-13 R/dbFD_joggle.R

@@ -118,11 +118,11 @@ dbFD_joggle <- function (x, a, w, w.abun = TRUE, stand.x = TRUE, ord = c("podani
           x.dist <- dist(x.s)
         }
         else {
-          x.dist <- gowdis(x, w = w, ord = ord, asym.bin = asym.bin)
+          x.dist <- FD::gowdis(x, w = w, ord = ord, asym.bin = asym.bin)
         }
       }
       else {
-        x.dist <- gowdis(x, w = w, ord = ord, asym.bin = asym.bin)
+        x.dist <- FD::gowdis(x, w = w, ord = ord, asym.bin = asym.bin)
       }
     }
     if (t.x == 1) {
@@ -225,7 +225,7 @@ dbFD_joggle <- function (x, a, w, w.abun = TRUE, stand.x = TRUE, ord = c("podani
     else x <- x
     x <- data.frame(x)
     dimnames(x) <- list(x.rn, "Trait")
-    x.dist <- gowdis(x, w = w, ord = ord, asym.bin = asym.bin)
+    x.dist <- FD::gowdis(x, w = w, ord = ord, asym.bin = asym.bin)
   }
   if (is.factor(x) & !is.ordered(x)) {
     if (any(is.na(x))) {

### 2019-11-21 analysis/merge_data.Rmd
First attempt at correction via a simple glm.

@@
 ```
+
+#Correct FD for species richness
+
+First for one study, 311
+```{r}
+model_data <- fd_table %>% filter(metric == "FRic", study_id == 311)
+fit <- glm(value ~ year + richness, data = model_data)
+
+pred_vars <- model_data %>% 
+  select(richness, year) %>% 
+  distinct() %>%
+  mutate(fric_pred = predict(fit, ., type = "response"))
+
+plot_data <- left_join(model_data, pred_vars) %>%
+  mutate(fric_adj = value - fric_pred,
+         fric_adj = ifelse(fric_adj > 0, fric_adj, 0))
+
+plot_data %>%
+  ggplot() +
+  geom_line(aes(x = year, y = value), color = "blue") +
+  geom_line(aes(x = year, y = fric_pred), color = "red") +
+  geom_line(aes(x = year, y = fric_adj), color = "black") +
+  scale_x_continuous(breaks = scales::pretty_breaks()) +
+  theme(legend.position = "none") +
+  ylab("Fric")
+```

### 2019-11-21 analysis/merge_data.Rmd
First attempt at correction via a simple glm.

@@
   biotime_data %>%
-    select(-sourceName, -scientificName) %>%
-  left_join(trait_data, by = "id",
-            na_matches = "never") %>%
-    drop_na(scientificName)
+    select(-scientificName) %>%
+    rename(biotimeName = sourceName) %>%
+  left_join(trait_data %>% rename(traitName = sourceName), by = "id",
+            na_matches = "never") #%>%
+   # drop_na(scientificName)
 }
@@
 Get data w/traits for all studies that have >85% trait coverage
+
+`id` is `NA` if there was no `id` match
+`biotimeName` or `traitName` is `NA` if that ID is the result of an average of multliple species matches
+`scientificName` is `NA` when there was no trait data for a given ID
 ```{r}
@@
 coverage <- comp_biotime_traits %>%
+  select(study_id, scientificName) %>%
+  distinct() %>%
   group_by(study_id) %>%
-  summarise(perc = sum(!is.na(sourceName))/n()) %>%
+  summarise(perc = sum(!is.na(scientificName))/n()) %>%
   arrange(desc(perc))
@@
 biotime_traits <- comp_biotime_traits %>%
-  filter(study_id %in% inc_coverage$study_id, !is.na(sourceName))
+  filter(study_id %in% inc_coverage$study_id, !is.na(scientificName)) #get studies with appropriate coverage and drop observations that we don't have trait data for
 ```

### 2020-03-05 analysis/visualization.Rmd
First commit of data visualizations.

---
title: "Visualization"
author: "Kari Norman"
date: "9/18/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

```{r}
load(here::here("data", "biotime_data.rda"))
```

```{r}
usa <- map_data("usa")

plt <- ggplot() +
  geom_polygon(data = usa, aes(x=long, y = lat, group = group), fill = NA, color = "black") +
  coord_fixed(1.3)

plt +
  geom_point(data = biotime_data %>% filter(taxa == "Fish"), aes(x = longitude, y = latitude), color = "black", size = .25)

```

#Visualizing FD trends
```{r}
load(here::here("data", "fd_table.rda"))

fric_plot <- fd_table %>%
  filter(metric == "FRic") %>%
  ggplot(aes(
    x = year,
    y = value,
    colour = as.factor(plot)
  )) +
  geom_line() +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  facet_wrap( ~ as.factor(study_id), scales = "free") +
  theme(legend.position = "none") +
  ylab("Fric")

ggsave("fric_trends.png", fric_plot)
```

Reproduce Dornelas figure
```{r}
#get climate classification
metadata <- read_csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS"))

fd_table %>%
  filter(metric == "FRic") %>%
  left_join(metadata %>% select(study_id, climate)) %>%
  unite(id, study_id, plot) %>%
  ggplot(aes(x = year, y = log(value))) +
  geom_smooth(aes(color = climate, group = id), method = "glm", se = FALSE) +
  geom_point(aes(color = climate, group = id)) +
  geom_smooth(color = "black", method = "glm", se = FALSE) +
  theme(legend.position = "none") 
```

### 2020-03-05 analysis/merge_data.Rmd
Make glm correction a function.

--- 2019-11-21 analysis/merge_data.Rmd
@@
 ```{r}
-model_data <- fd_table %>% filter(metric == "FRic", study_id == 311)
-fit <- glm(value ~ year + richness, data = model_data)
+
+glm_correction <- function(study, plot_num){
+  model_data <- fd_table %>% filter(metric == "FRic", study_id == study, plot == plot_num)
+  fit <- glm(value ~ richness, data = model_data)
 
   pred_vars <- model_data %>% 
-  select(richness, year) %>% 
+    select(richness) %>% 
     distinct() %>%
@@
            fric_adj = ifelse(fric_adj > 0, fric_adj, 0))
+
+  return(plot_data)
+}
+
+plot_data <- glm_correction(311, 1)
 

### 2020-06-04 analysis/ts_null_model.Rmd
filter plot_id

---
title: "null_model"
author: "Kari Norman"
date: "6/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(biodivTS)
```

Create null models that would be transferable to any timeseries by treating the species pool as all species observed across the timeseries.

Simulation functions
```{r}
get_plot_sample <- function(data, filter_plot, n){
  plot_data <- data %>% filter(plot_id == filter_plot)

  rich_levels <- plot_data %>%
    select(year, id) %>%
    distinct() %>%
    count(year) %>%
    pull(n) %>%
    unique()

  species_pool <- plot_data %>%
    pull(id) %>%
    na.omit() %>%
    unique() #get column of unique scientific names

  if (length(species_pool) < max(rich_levels)) browser()

  get_samp_species <- function(richness){
    samp_species <- sample(species_pool, richness)
    df <- cbind(richness, samp_species)
    return(df)
  }

  site_samples <- data.frame()
  for (j in 1:n){
    iter_samples <- data.frame()
    for(i in rich_levels){
    iter_samples <- rbind(iter_samples, get_samp_species(i))
    }
    site_samples <- rbind(site_samples, iter_samples %>% mutate(n = j))
  }
  site_samples %>%
    mutate(plot_id = filter_plot, value = 1) %>%
    pivot_wider(names_from = samp_species, values_from = value)
}

#drop trait columns that are the same value for every observation, and therefore providing no additional info
clean_trait_matrix <- function(trait_matrix){
  uniq_vals <- sapply(trait_matrix, function(x){length(unique(x))}) #get number of unique values of each trait
  col_names <- names(which(uniq_vals == 1)) #find columns where there is only one value for all observations
  return(select(trait_matrix, -col_names))
}

get_trait_matrix <- function(data, species_list, trait_list){ 
  traits <- data %>%
    filter(id %in% species_list) %>%
    dplyr::select(id, all_of(trait_list)) %>%
    distinct() %>%
    arrange(id) %>%
    column_to_rownames(var = "id")
}
```

Get dataframe of samples for each plot and corresponding species and trait matrices
```{r}
load(here::here("data", "biotime_traits.rda"))

biotime_traits <- biotime_traits %>%
  mutate(plot_id = paste0(study_id, plot)) %>%
  filter(!is.na(id))

n = 100

sample_occurrence <- map_dfr(unique(biotime_traits$plot_id), get_plot_sample, data = biotime_traits, n = n) %>%
  replace(is.na(.), 0)

#colnames(biotime_traits)[17:64]

get_plot_FD <- function(data, trait_range, occ_df, filter_plot) {
  sample_species_mat <- occ_df %>%
    filter(plot_id == filter_plot) %>%
    select(-c(richness, plot_id, n)) %>% #want only species x site info
    select_if(!colSums(.) %in% c(0, NA)) %>% #remove species that don't occur in the plot
    select(sort(tidyselect::peek_vars())) #sort columns to be in alphabetical order, required for dbFDI()

  traits <- colnames(data)[trait_range]
  sample_trait_mat <-
    clean_trait_matrix(get_trait_matrix(data = data, colnames(sample_species_mat), traits))

  sample_fd <- as.data.frame(dbFD_joggle(x = sample_trait_mat, a = sample_species_mat, w.abun = FALSE))

  df <- occ_df %>%
    filter(plot_id == filter_plot) %>%
    select(richness, plot_id, n) %>%
    bind_cols(sample_fd) %>%
    mutate(merge_test = case_when(richness == nbsp ~ TRUE,
                                  TRUE ~ FALSE))

  if (sum(df$merge_test) != dim(df)[1])
    warning("richness and region columns may not have been appropriately joined with FD data")

  return(df %>% select(-merge_test))
}

test <- get_plot_FD(biotime_traits, 17:64, sample_occurrence, 5617)
```

BAT
```{r}
library(BAT)

get_plot_BAT <- function(data, trait_range, occ_df, filter_plot) {
  sample_species_mat <- occ_df %>%
    filter(plot_id == filter_plot) %>%
    select(-c(richness, plot_id, n)) %>% #want only species x site info
    select_if(!colSums(.) %in% c(0, NA)) %>% #remove species that don't occur in the plot
    select(sort(tidyselect::peek_vars())) #sort columns to be in alphabetical order, required for dbFDI()

  traits <- colnames(data)[trait_range]
  sample_trait_mat <-
    clean_trait_matrix(get_trait_matrix(data = data, colnames(sample_species_mat), traits))

  #get distance matrix
  dist_traits <-  gowdis(sample_trait_mat)
  trait.pco <- dudi.pco(quasieuclid(dist_traits), scannf = FALSE, full = TRUE)

}

test <- get_plot_BAT(biotime_traits, 17:64, sample_occurrence, 5617)

ower.mat_spider <- gower.dist(traits)
euc.pco <- pco(gower.mat_spider,k=3)

barplot(euc.pco$eig)
(euc.pco$eig[1]+euc.pco$eig[2]+euc.pco$eig[3])/sum(euc.pco$eig[euc.pco$eig > 0]) #0.955 variance explained by first 3 axes

traits <- euc.pco$points
row.names(traits) <- row.names(traits2)

```

### 2020-06-04 analysis/ts_null_model.Rmd
first commit

@@
 
+```{r}
 ```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo = TRUE)
-```
-
-```{r}
-library(tidyverse)
-library(biodivTS)
 ```
@@
 ```{r}
-get_plot_sample <- function(data, filter_plot, n){
-  plot_data <- data %>% filter(plot_id == filter_plot)
+load(here::here("data", "biotime_traits.rda"))
 
-  rich_levels <- plot_data %>%
+get_plot_sample <- function(data, study, plot_id, n){
+  rich_levels <- data %>%
+    filter(study_id == study, plot == plot_id) %>%
     select(year, id) %>%
@@
 
-  species_pool <- plot_data %>%
+  species_pool <- data %>% 
+    filter(study_id == study, plot == plot_id) %>%
     pull(id) %>%
-    na.omit() %>%
     unique() #get column of unique scientific names
-
-  if (length(species_pool) < max(rich_levels)) browser()
 
@@
   site_samples %>%
-    mutate(plot_id = filter_plot, value = 1) %>%
-    pivot_wider(names_from = samp_species, values_from = value)
+    mutate(study_id = study, plot = plot_id, value = 1) %>%
+    spread(samp_species, value)
+ # return(site_samples %>% mutate(study_id = study, plot = plot_id))
 }
@@
 
-get_trait_matrix <- function(data, species_list, trait_list){ 
-  traits <- data %>%
-    filter(id %in% species_list) %>%
-    dplyr::select(id, all_of(trait_list)) %>%
-    distinct() %>%
-    arrange(id) %>%
-    column_to_rownames(var = "id")
+get_sample_FD <- function(data, n){
+  #get random sample for each region and richness level
+  plots <- data %>%
+    select(site_id, plot) %>%
+    distinct()
+
+  sample_occurrence <- pmap_dfr(plots[1:10,], get_plot_sample)
+
+  sample_species_mat <- select(sample_occurrence, -richness, -site_id) %>% #want only species x site info
+    select(sort(current_vars())) #sort columns to be in alphabetical order, required for dbFDI()
+  sample_trait_mat <- clean_trait_matrix(get_trait_matrix(colnames(sample_species_mat)))
+
+  sample_fd <- as.data.frame(dbFD_joggle(x = sample_trait_mat, a = sample_species_mat, w.abun = FALSE))
+
+  df <- sample_occurrence %>%
+    select(region, richness) %>%
+    bind_cols(sample_fd) %>%
+    mutate(merge_test = case_when(
+      richness == nbsp ~ TRUE,
+      TRUE ~ FALSE
+    ))
+
+  if(sum(df$merge_test) != dim(df)[1]) warning("richness and region columns may not have been appropriately joined with FD data")
+
+  path <- paste0(here(), "/data/fd_site_samples/", x, "_tmp.tsv.bz2")
+  write_tsv(df, path)
+  #pb_upload(path)
 }
@@
 
-Get dataframe of samples for each plot and corresponding species and trait matrices
 ```{r}
-load(here::here("data", "biotime_traits.rda"))
+n <- 100 #sample size
 
-biotime_traits <- biotime_traits %>%
-  mutate(plot_id = paste0(study_id, plot)) %>%
-  filter(!is.na(id))
-
-n = 100
-
-sample_occurrence <- map_dfr(unique(biotime_traits$plot_id), get_plot_sample, data = biotime_traits, n = n) %>%
-  replace(is.na(.), 0)
-
-#colnames(biotime_traits)[17:64]
-
-get_plot_FD <- function(data, trait_range, occ_df, filter_plot) {
-  sample_species_mat <- occ_df %>%
-    filter(plot_id == filter_plot) %>%
-    select(-c(richness, plot_id, n)) %>% #want only species x site info
-    select_if(!colSums(.) %in% c(0, NA)) %>% #remove species that don't occur in the plot
-    select(sort(tidyselect::peek_vars())) #sort columns to be in alphabetical order, required for dbFDI()
-
-  traits <- colnames(data)[trait_range]
-  sample_trait_mat <-
-    clean_trait_matrix(get_trait_matrix(data = data, colnames(sample_species_mat), traits))
-
-  sample_fd <- as.data.frame(dbFD_joggle(x = sample_trait_mat, a = sample_species_mat, w.abun = FALSE))
-
-  df <- occ_df %>%
-    filter(plot_id == filter_plot) %>%
-    select(richness, plot_id, n) %>%
-    bind_cols(sample_fd) %>%
-    mutate(merge_test = case_when(richness == nbsp ~ TRUE,
-                                  TRUE ~ FALSE))
-
-  if (sum(df$merge_test) != dim(df)[1])
-    warning("richness and region columns may not have been appropriately joined with FD data")
-
-  return(df %>% select(-merge_test))
+if(!dir.exists(here::here("data", "fd_site_samples"))){
+  dir.create(paste0(here::here(), "/data/fd_site_samples"))
+  system.time({out <- future_map(1:n, get_sample_FD)})
 }
 
-test <- get_plot_FD(biotime_traits, 17:64, sample_occurrence, 5617)
+sim <- purrr::map_dfr(fs::dir_ls(path = here::here("data", "fd_samples"), glob="*.tsv.bz2"), readr::read_tsv, .id = "sample")
+
 ```
 
-BAT
-```{r}
-library(BAT)
-
-get_plot_BAT <- function(data, trait_range, occ_df, filter_plot) {
-  sample_species_mat <- occ_df %>%
-    filter(plot_id == filter_plot) %>%
-    select(-c(richness, plot_id, n)) %>% #want only species x site info
-    select_if(!colSums(.) %in% c(0, NA)) %>% #remove species that don't occur in the plot
-    select(sort(tidyselect::peek_vars())) #sort columns to be in alphabetical order, required for dbFDI()
-
-  traits <- colnames(data)[trait_range]
-  sample_trait_mat <-
-    clean_trait_matrix(get_trait_matrix(data = data, colnames(sample_species_mat), traits))
-
-  #get distance matrix
-  dist_traits <-  gowdis(sample_trait_mat)
-  trait.pco <- dudi.pco(quasieuclid(dist_traits), scannf = FALSE, full = TRUE)
-
-}
-
-test <- get_plot_BAT(biotime_traits, 17:64, sample_occurrence, 5617)
-
-ower.mat_spider <- gower.dist(traits)
-euc.pco <- pco(gower.mat_spider,k=3)
-
-barplot(euc.pco$eig)
-(euc.pco$eig[1]+euc.pco$eig[2]+euc.pco$eig[3])/sum(euc.pco$eig[euc.pco$eig > 0]) #0.955 variance explained by first 3 axes
-
-traits <- euc.pco$points
-row.names(traits) <- row.names(traits2)
-
 ```

### 2020-06-09 analysis/ts_null_model.Rmd
Function to get sample occurrences for all studies and plots.

@@
 knitr::opts_chunk$set(echo = TRUE)
+```
+
+```{r}
+library(tidyverse)
 ```
@@
     pull(id) %>%
+    na.omit() %>%
     unique() #get column of unique scientific names
@@
     mutate(study_id = study, plot = plot_id, value = 1) %>%
-    spread(samp_species, value)
- # return(site_samples %>% mutate(study_id = study, plot = plot_id))
+    pivot_wider(names_from = samp_species, values_from = value, values_fill = list(value = 0))
 }
@@
 
-get_sample_FD <- function(data, n){
-  #get random sample for each region and richness level
-  plots <- data %>%
-    select(site_id, plot) %>%
-    distinct()
-
-  sample_occurrence <- pmap_dfr(plots[1:10,], get_plot_sample)
-
-  sample_species_mat <- select(sample_occurrence, -richness, -site_id) %>% #want only species x site info
-    select(sort(current_vars())) #sort columns to be in alphabetical order, required for dbFDI()
-  sample_trait_mat <- clean_trait_matrix(get_trait_matrix(colnames(sample_species_mat)))
-
-  sample_fd <- as.data.frame(dbFD_joggle(x = sample_trait_mat, a = sample_species_mat, w.abun = FALSE))
-
-  df <- sample_occurrence %>%
-    select(region, richness) %>%
-    bind_cols(sample_fd) %>%
-    mutate(merge_test = case_when(
-      richness == nbsp ~ TRUE,
-      TRUE ~ FALSE
-    ))
-
-  if(sum(df$merge_test) != dim(df)[1]) warning("richness and region columns may not have been appropriately joined with FD data")
-
-  path <- paste0(here(), "/data/fd_site_samples/", x, "_tmp.tsv.bz2")
-  write_tsv(df, path)
-  #pb_upload(path)
+get_trait_matrix <- function(species_list, trait_list){ 
+  traits <- biotime_traits %>%
+    filter(id %in% species_list) %>%
+    dplyr::select(id, all_of(trait_list)) %>%
+    distinct() %>%
+    arrange(id) %>%
+    column_to_rownames(var = "id")
 }
@@
 
+Get dataframe of samples for each plot and corresponding species and trait matrices
 ```{r}
-n <- 100 #sample size
+n = 100
 
-if(!dir.exists(here::here("data", "fd_site_samples"))){
-  dir.create(paste0(here::here(), "/data/fd_site_samples"))
-  system.time({out <- future_map(1:n, get_sample_FD)})
-}
+plots <- biotime_traits %>%
+    select(study = study_id, plot_id = plot) %>%
+    distinct()
 
-sim <- purrr::map_dfr(fs::dir_ls(path = here::here("data", "fd_samples"), glob="*.tsv.bz2"), readr::read_tsv, .id = "sample")
+sample_occurrence <- pmap_dfr(plots[1:10,], get_plot_sample, data = biotime_traits, n = n) %>%
+  replace(is.na(.), 0)
+
+sample_species_mat <- select(sample_occurrence, -c(richness, study_id, plot, n)) %>% #want only species x site info
+  select(sort(tidyselect::peek_vars())) #sort columns to be in alphabetical order, required for dbFDI()
+
+traits <- colnames(biotime_traits)[17:64]
+sample_trait_mat <- clean_trait_matrix(get_trait_matrix(colnames(sample_species_mat), traits))
+```
+
+Explore different functional diversity approaches, starting with original FD application
+
+```{r}
+library(FD)
+
+#get diversity metrics from FD package
+sample_fd <- as.data.frame(dbFD(x = sample_trait_mat, a = sample_species_mat, w.abun = FALSE))
+
+ #checks for merging output with richness/study info
+df <- sample_occurrence %>%
+  select(richness, study_id, plot, n) %>%
+  bind_cols(sample_fd) %>%
+  mutate(merge_test = case_when(richness == nbsp ~ TRUE,
+                                TRUE ~ FALSE))
+
+if (sum(df$merge_test) != dim(df)[1])
+  warning("richness and region columns may not have been appropriately joined with FD data")
+```
+
+BAT
+```{r}
+library(BAT)
 
 ```
-
-```

### 2020-06-17 analysis/merge_data.Rmd
add BAT function for calculating metrics

@@
 
-First pass at calculating FD metrics
+First pass at calculating FD metrics from FD package:
 ```{r}
@@
 
+Let's also do the functions from BAT package
+```{r}
+#function for getting PCA axes with at least 95% coverage
+axes95 <- function(mat){
+  axes <- pco(mat)
+  eigen <- axes$eig
+
+  threshold <- sum(eigen[eigen > 0]) * 0.95
+
+  sums <- cumsum(eigen)
+  num_axes <- Position(function(x) x > threshold, sums)
+
+  return(num_axes)
+}
+
+get_BAT_fd <- function(data, study, plot_num){
+  require(BAT)
+
+  data <- data %>%
+    na.omit(id) %>%
+    filter(study_id == study, plot == plot_num)
+
+  print(max(data$record_id))
+
+  species_mat <- data %>%
+    select(year, id) %>% 
+    distinct() %>%
+    arrange(year, id) %>%
+    mutate(present = 1) %>%
+    spread(id, value = present, fill = 0) %>%
+    select(-year)
+
+  years <- data %>% 
+    select(year) %>%
+    distinct() %>%
+    arrange(year)
+
+  trait_mat <- data %>%
+    select(-c(taxa, record_id, day, month, year, sample_desc, id_species, 
+              latitude, longitude, sum_abundance, sum_biomass, sourceName, scientificName)
+    ) %>%
+    distinct() %>%
+    select_if(~ length(unique(.)) > 1) %>% #remove columns that have the same value for all columns
+    mutate_at(vars(-id, -ends_with("5cat")), list(~scales::rescale(.,to=c(0, 1)))) %>% #rescale variables
+    arrange(id) %>%
+    column_to_rownames("id")
+
+  #get PCA trait matrix
+  trait_dist <- gower.dist(trait_mat)
+  trait_pco <- pco(trait_dist, k = axes95(trait_dist))
+
+  return(rownames(trait_pco$points) <- rownames)
+  # fs::dir_create(paste0(here::here("data"), "/fd_output"))
+  # 
+  # #the convexhull c code creates an output file of vertices, in order to not have multiple threads writing
+  # #to the same file we have to create individual working directories
+  # tmp_dir <- paste0(here::here("data", "fd_output", "tmp"), "/tmp_", study, plot_num)
+  # fs::dir_create(tmp_dir)
+  # setwd(tmp_dir)
+  # 
+  # fd_out <- as_tibble(dbFD_joggle(trait_mat, species_mat, w.abun = FALSE))[,1:8] %>%
+  #   cbind(years, .) %>%
+  #   mutate(study_id = unlist(study), plot = unlist(plot_num), trait_count = dim(trait_mat)[2]) %>%
+  #   write_tsv(paste0(here::here("data"), "/fd_output/", "s", study, "_p", plot_num, "_fd.tsv"))
+  # 
+  # setwd(here::here())
+  # return(fd_out)
+}
+
+```
+
 Format tables

### 2020-06-17 analysis/merge_data.Rmd
add BAT function for calculating metrics

@@
   biotime_data %>%
-    select(-scientificName) %>%
-    rename(biotimeName = sourceName) %>%
-  left_join(trait_data %>% rename(traitName = sourceName), by = "id",
-            na_matches = "never") #%>%
-   # drop_na(scientificName)
+    select(-sourceName, -scientificName) %>%
+  left_join(trait_data, by = "id",
+            na_matches = "never") %>%
+    drop_na(scientificName)
 }
@@
 Get data w/traits for all studies that have >85% trait coverage
-
-`id` is `NA` if there was no `id` match
-`biotimeName` or `traitName` is `NA` if that ID is the result of an average of multliple species matches
-`scientificName` is `NA` when there was no trait data for a given ID
 ```{r}
@@
 coverage <- comp_biotime_traits %>%
-  select(study_id, scientificName) %>%
-  distinct() %>%
   group_by(study_id) %>%
-  summarise(perc = sum(!is.na(scientificName))/n()) %>%
+  summarise(perc = sum(!is.na(sourceName))/n()) %>%
   arrange(desc(perc))
@@
 biotime_traits <- comp_biotime_traits %>%
-  filter(study_id %in% inc_coverage$study_id, !is.na(scientificName)) #get studies with appropriate coverage and drop observations that we don't have trait data for
+  filter(study_id %in% inc_coverage$study_id, !is.na(sourceName))
 ```

### 2020-06-17 analysis/merge_data.Rmd
add BAT function for calculating metrics

--- 2020-03-05 analysis/merge_data.Rmd
@@
 biotime_traits <- comp_biotime_traits %>%
-  filter(study_id %in% inc_coverage$study_id, !is.na(scientificName)) #get studies with appropriate coverage and drop observations that we don't have trait data for
+  #get studies with appropriate coverage and drop observations that we don't have trait data for
+  filter(study_id %in% inc_coverage$study_id, !is.na(scientificName)) %>% 
+  mutate(plot_id = paste0(study_id, plot))
 ```
@@
 
-get_fd <- function(data, study, plot_num){
+get_fd <- function(data, filter_plot){
   require(FD)
@@
   data <- data %>%
-    filter(study_id == study, plot == plot_num)
+    filter(plot_id == filter_plot)
 
-  print(max(data$record_id))
+  print(filter_plot)
 
@@
     select(-c(taxa, record_id, day, month, year, sample_desc, id_species, 
-              latitude, longitude, sum_abundance, sum_biomass, sourceName, scientificName)
+              latitude, longitude, sum_abundance, sum_biomass, traitName, 
+              biotimeName, scientificName)
     ) %>%
@@
   #to the same file we have to create individual working directories
-  tmp_dir <- paste0(here::here("data", "fd_output", "tmp"), "/tmp_", study, plot_num)
+  tmp_dir <- paste0(here::here("data", "fd_output", "tmp"), "/tmp_", filter_plot)
   fs::dir_create(tmp_dir)
@@
     cbind(years, .) %>%
-    mutate(study_id = unlist(study), plot = unlist(plot_num), trait_count = dim(trait_mat)[2]) %>%
-    write_tsv(paste0(here::here("data"), "/fd_output/", "s", study, "_p", plot_num, "_fd.tsv"))
+    mutate(plot_id = unlist(filter_plot), trait_count = dim(trait_mat)[2]) %>%
+    write_tsv(paste0(here::here("data"), "/fd_output/", filter_plot, "_fd.tsv"))
 
@@
 
-First pass at calculating FD metrics
+First pass at calculating FD metrics from FD package:
 ```{r}
 data <- biotime_traits %>% 
-  filter(!is.na(plot), !is.na(scientificName)) %>% #remove unlabeled plots and observations without traits
-  group_by(study_id, year, plot) %>%
+  filter(!is.na(plot)) %>% #remove unlabeled plots 
+  group_by(plot_id, year) %>%
   filter(n_distinct(id) > 4) %>%
-  ungroup() %>%
-  filter(!study_id %in% c(195, 360), #exclude because they break
-         !study_id %in% c(217, 327)) #exclude to get the file size smaller
+  ungroup()# %>%
+  #filter(!study_id %in% c(195, 360), #exclude because they break
+         #!study_id %in% c(217, 327)) #exclude to get the file size smaller
 
-combs <- data %>%
-  select(study_id, plot) %>%
-  distinct()
+plots <- data %>%
+  pull(plot_id) %>%
+  unique()
 
 plan(multiprocess)
-system.time(fd <- future_pmap_dfr(combs, ~ get_fd(data = data, study = .x, plot = .y)))
+system.time(fd <- future_map_dfr(plots[1:10], get_fd, data = data))
 ```
@@
 
+Let's also do the functions from BAT package
+```{r}
+#function for getting PCA axes with at least 95% coverage
+axes95 <- function(mat){
+  axes <- pco(mat)
+  eigen <- axes$eig
+
+  threshold <- sum(eigen[eigen > 0]) * 0.95
+
+  sums <- cumsum(eigen)
+  num_axes <- Position(function(x) x > threshold, sums)
+
+  return(num_axes)
+}
+
+get_BAT_fd <- function(data, filter_plot){
+  require(BAT)
+  require(StatMatch)
+  require(labdsv)
+
+  data <- data %>% filter(plot_id == filter_plot)
+
+  print(max(data$record_id))
+
+  species_mat <- data %>%
+    select(year, id) %>% 
+    distinct() %>%
+    arrange(year, id) %>%
+    mutate(present = 1) %>%
+    spread(id, value = present, fill = 0) %>%
+    column_to_rownames("year")
+
+  trait_mat <- data %>%
+    select(-c(taxa, record_id, day, month, year, sample_desc, id_species, 
+              latitude, longitude, sum_abundance, sum_biomass, biotimeName, 
+              traitName, scientificName)
+    ) %>%
+    distinct() %>%
+    select_if(~ length(unique(.)) > 1) %>% #remove columns that have the same value for all columns
+    mutate_at(vars(-id, -ends_with("5cat")), list(~scales::rescale(.,to=c(0, 1)))) %>% #rescale variables
+    arrange(id) %>%
+    column_to_rownames("id")
+
+  #get PCA trait matrix
+  trait_dist <- gower.dist(trait_mat)
+  trait_pco <- pco(trait_dist, k = axes95(trait_dist))
+
+  traits <- trait_pco$points
+  rownames(traits) <- rownames(trait_mat)
+
+  #getFD metrics
+  kernelFD <- kernel.alpha(comm=as.matrix(species_mat),trait=traits,abund=FALSE,method="gaussian",return.hv=TRUE)
+  hv <- kernelFD[[2]]
+
+  disp <- kernel.dispersion(comm=hv,frac=0.1,func='dissimilarity') 
+  div <- kernel.dispersion(comm=hv,frac=0.6, func='divergence')
+  even <- kernel.evenness(comm=hv)
+
+  results <- data.frame(richness = as.numeric(kernelFD[[1]]), disp, div, even) %>% 
+    mutate(plot_id = filter_plot)
+
+  return(results)
+}
+
+get_BAT_fd(biotime_traits, 376300153)
+```
+
 Format tables
@@
 ```{r}
-
 glm_correction <- function(study, plot_num){

### 2020-06-19 data-raw/amphibio.Rmd
update helper functions and fix amphbio processing

--- 2019-07-03 data-raw/amphibio.Rmd
@@
 
-amph <- read_csv(system.file("extdata", "amphibio/AmphiBIO_v1.csv", package = "biodivTS"))
+amph <- read_csv(system.file("extdata", "amphibio/AmphiBIO_v1.csv", package = "biodivTS")) %>%
+  select(-id) %>%
+  rename_all(tolower) %>%
+  rename(diet_leaves = leaves, diet_flowers = flowers, diet_seed = seeds, diet_fruit = fruits, diet_inv = arthro, diet_vert = vert,
+         diurnal = diu, nocturnal = noc, crepuscular = crepu, bodymass_value = body_mass_g) %>%
+  mutate(diet_seed = as.numeric(diet_seed))
 
 amphibio <- amph %>%
-  mutate(id = get_ids(Species, "itis")) %>%
+  mutate(id = get_ids(species, "itis")) %>%
   filter(!is.na(id)) %>%
   mutate(scientificName = get_names(id, "itis")) %>%
-  select(id, scientificName, Species) %>%
-  right_join(select(amph, -id), by = "Species") %>%
-  rename(sourceName = Species) %>%
-  select(-c("Order", "Family", "Genus"))
+  select(id, scientificName, species) %>%
+  right_join(amph, by = "species") %>%
+  rename(sourceName = species) %>%
+  select(-c("order", "family", "genus"))
 
-# Using by_ doesn't give any additional ids
-# amphibio <- by_name(amph$Species, "itis") %>%
-#   filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
-#   drop_na(acceptedNameUsageID)
+#two entries need to be resolved by hand
+add_ids <- amphibio %>%
+  filter(is.na(id)) %>%
+  pull(sourceName) %>%
+  filter_name(., "itis") %>%
+  filter(!is.na(acceptedNameUsageID)) %>%
+  select(scientificName, id = acceptedNameUsageID) %>%
+  distinct() %>%
+  left_join(amph, by = c("scientificName" = "species")) %>%
+  select(-c("order", "family", "genus")) %>%
+  mutate(sourceName = scientificName)
+
+amphibio <- amphibio %>% 
+  filter(!sourceName %in% add_ids$sourceName) %>%
+  bind_rows(add_ids)
 ```
@@
   filter(!input %in% alt_dupes$input) %>% #excluding duplicate matched species
-  select(id = acceptedNameUsageID, sourceName = input, scientificName) %>%
+  select(id = acceptedNameUsageID, sourceName = input) %>%
   distinct() %>%
-  group_by(sourceName) %>%
-  top_n(1, scientificName) %>%
+  mutate(scientificName = get_names(id, "itis")) %>%
   left_join(amphibio %>% select(-c(id, scientificName)), by = "sourceName", na_matches =  "never") %>%
-  ungroup() %>%
   distinct()

### 2020-06-19 data-raw/helper_functions.R

@@ -34,9 +34,11 @@ match_providers <- function(data, original_provider, common = FALSE) {
   unmatched <- data %>% filter(is.na(id)) %>% pull(sourceName) %>% unique()
   providers <- c("itis", "ncbi", "col", "gbif", "fb", "wd", "ott", "iucn")
 
+  synonyms <-
+
   #match all the un-ID'd names to other providers, check if synonyms given by those providers match known ITIS species
-  alt_names <-map_df(providers[providers != original_provider], function(name) taxadb::synonyms(unmatched, name)) %>%
-    drop_na(acceptedNameUsageID) %>%
+  alt_names <-map_df(providers[providers != original_provider], function(name) synonyms(unmatched, name)) %>%
+    drop_na(synonym) %>%
     #new matches can be in either the synonym or acceptedNameUsage column, so let's combine them
     gather(type, altProviderName, synonym, acceptedNameUsage) %>%
     select(altProviderName, input) %>%
@@ -44,7 +46,7 @@ match_providers <- function(data, original_provider, common = FALSE) {
     drop_na(altProviderName)
 
   #match accepted names from
-  syn_res <- by_name(na.omit(unique(alt_names$altProviderName)), original_provider) %>%
+  syn_res <- filter_name(na.omit(unique(alt_names$altProviderName)), original_provider) %>%
     drop_na(acceptedNameUsageID) %>%
     rename(altProviderName = input) %>%
     left_join(alt_names, na_matches = "never", by = "altProviderName")
@@ -54,10 +56,10 @@ match_providers <- function(data, original_provider, common = FALSE) {
     unmatched <- unmatched[unmatched %in% syn_res$input]
 
     common_match <- map_df(common_providers[common_providers != original_provider],
-           function(name) by_common(unmatched, name)) %>%
+           function(name) filter_common(unmatched, name)) %>%
       drop_na(acceptedNameUsageID)
 
-    sci_match <- by_name(unique(common_match$scientificName), "itis") %>%
+    sci_match <- filter_name(unique(common_match$scientificName), "itis") %>%
       rename(altProviderName = input) %>%
       drop_na(acceptedNameUsageID) %>%
       left_join(common_match %>% select(scientificName, input),
@@ -93,4 +95,101 @@ undupe_ids <- function(data) {
   resolved_data <- data %>%
[…]

### 2020-06-22 analysis/merge_data.Rmd
clarify the NA scenarios for different species identification variables, make sure I'm excluding the appropriate one.

--- 2020-06-17 analysis/merge_data.Rmd
@@
 
-First pass at calculating FD metrics from FD package:
+First pass at calculating FD metrics
 ```{r}
@@
 
-Let's also do the functions from BAT package
-```{r}
-#function for getting PCA axes with at least 95% coverage
-axes95 <- function(mat){
-  axes <- pco(mat)
-  eigen <- axes$eig
-
-  threshold <- sum(eigen[eigen > 0]) * 0.95
-
-  sums <- cumsum(eigen)
-  num_axes <- Position(function(x) x > threshold, sums)
-
-  return(num_axes)
-}
-
-get_BAT_fd <- function(data, study, plot_num){
-  require(BAT)
-
-  data <- data %>%
-    na.omit(id) %>%
-    filter(study_id == study, plot == plot_num)
-
-  print(max(data$record_id))
-
-  species_mat <- data %>%
-    select(year, id) %>% 
-    distinct() %>%
-    arrange(year, id) %>%
-    mutate(present = 1) %>%
-    spread(id, value = present, fill = 0) %>%
-    select(-year)
-
-  years <- data %>% 
-    select(year) %>%
-    distinct() %>%
-    arrange(year)
-
-  trait_mat <- data %>%
-    select(-c(taxa, record_id, day, month, year, sample_desc, id_species, 
-              latitude, longitude, sum_abundance, sum_biomass, sourceName, scientificName)
-    ) %>%
-    distinct() %>%
-    select_if(~ length(unique(.)) > 1) %>% #remove columns that have the same value for all columns
-    mutate_at(vars(-id, -ends_with("5cat")), list(~scales::rescale(.,to=c(0, 1)))) %>% #rescale variables
-    arrange(id) %>%
-    column_to_rownames("id")
-
-  #get PCA trait matrix
-  trait_dist <- gower.dist(trait_mat)
-  trait_pco <- pco(trait_dist, k = axes95(trait_dist))
-
-  return(rownames(trait_pco$points) <- rownames)
-  # fs::dir_create(paste0(here::here("data"), "/fd_output"))
-  # 
-  # #the convexhull c code creates an output file of vertices, in order to not have multiple threads writing
-  # #to the same file we have to create individual working directories
-  # tmp_dir <- paste0(here::here("data", "fd_output", "tmp"), "/tmp_", study, plot_num)
-  # fs::dir_create(tmp_dir)
-  # setwd(tmp_dir)
-  # 
-  # fd_out <- as_tibble(dbFD_joggle(trait_mat, species_mat, w.abun = FALSE))[,1:8] %>%
-  #   cbind(years, .) %>%
-  #   mutate(study_id = unlist(study), plot = unlist(plot_num), trait_count = dim(trait_mat)[2]) %>%
-  #   write_tsv(paste0(here::here("data"), "/fd_output/", "s", study, "_p", plot_num, "_fd.tsv"))
-  # 
-  # setwd(here::here())
-  # return(fd_out)
-}
-
-```
-
 Format tables
@@
 ```
-
-#Correct FD for species richness
-
-First for one study, 311
-```{r}
-
-glm_correction <- function(study, plot_num){
-  model_data <- fd_table %>% filter(metric == "FRic", study_id == study, plot == plot_num)
-  fit <- glm(value ~ richness, data = model_data)
-
-  pred_vars <- model_data %>% 
-    select(richness) %>% 
-    distinct() %>%
-    mutate(fric_pred = predict(fit, ., type = "response"))
-
-  plot_data <- left_join(model_data, pred_vars) %>%
-    mutate(fric_adj = value - fric_pred,
-           fric_adj = ifelse(fric_adj > 0, fric_adj, 0))
-
-  return(plot_data)
-}
-
-plot_data <- glm_correction(311, 1)
-
-plot_data %>%
-  ggplot() +
-  geom_line(aes(x = year, y = value), color = "blue") +
-  geom_line(aes(x = year, y = fric_pred), color = "red") +
-  geom_line(aes(x = year, y = fric_adj), color = "black") +
-  scale_x_continuous(breaks = scales::pretty_breaks()) +
-  theme(legend.position = "none") +
-  ylab("Fric")
-```

### 2020-06-23 analysis/merge_data.Rmd
Calculate BAT metrics in function. Update functions to map over plot_id variable.

--- 2020-06-17 analysis/merge_data.Rmd
@@
 ```{r}
+
 glm_correction <- function(study, plot_num){

### 2020-06-24 analysis/calc_fd.Rmd
Move FD calculations to their own script.

--- 2020-06-23 analysis/merge_data.Rmd
@@
 ---
-title: "Check Coverage"
-author: "Kari Norman"
-date: "5/20/2019"
+title: "Calculate FD"
 output: html_document
@@
 library(tidyverse)
+library(furrr)
 library(FD)
-devtools::load_all()
-```
-
-```{r}
-load(here::here("data", "biotime_data.rda"))
-load(here::here("data", "elton_mamm.rda"))
-load(here::here("data", "elton_bird.rda"))
-load(here::here("data", "amphibio.rda"))
-```
-
-#Checking Coverage of ID's and trait data -> biotime
-
-check if all bird & mammal biotime data w/id's has an elton match. There are 68 species without a match (63 are birds).
-```{r}
-no_trait <- biotime_data %>%
-  filter(taxa %in% c("Birds", "Mammals")) %>%
-  select(id, sourceName, taxa) %>%
-  drop_na(id) %>%
-  distinct() %>%
-  filter(!id %in% elton_bird$id, !id %in% elton_mamm$id)
-
-dim(no_trait)
-```
-
-and 19869 species without an ID, some of which are species codes
-```{r}
-biotime_data %>%
-  #filter(taxa == "Birds") %>%
-  select(id, sourceName, taxa) %>%
-  filter(is.na(id)) %>%
-  distinct() %>%
-  dim()
-```
-
-Are there any studies for which we have trait data for all species?
-```{r}
-merge_data <- function(trait_data){
-  biotime_data %>%
-    select(-scientificName) %>%
-    rename(biotimeName = sourceName) %>%
-  left_join(trait_data %>% rename(traitName = sourceName), by = "id",
-            na_matches = "never") #%>%
-   # drop_na(scientificName)
-}
-```
-
-Get data w/traits for all studies that have >85% trait coverage
-
-`id` is `NA` if there was no `id` match
-`biotimeName` or `traitName` is `NA` if that ID is the result of an average of multliple species matches
-`scientificName` is `NA` when there was no trait data for a given ID
-```{r}
-#set coverage threshold for analysis dataset
-percent <- 0.85
-
-traits <- list(elton_bird, elton_mamm, amphibio)
-
-comp_biotime_traits <- map_dfr(traits, merge_data) %>%
-  right_join(biotime_data %>% select(-sourceName, -scientificName))
-
-#always check that one species didn't match to multiple datasets, below should return and empty dataframe
-#comp_biotime_traits %>% group_by(record_id) %>% filter(n() > 1) %>% dim()
-
-#this gives a coverage for each trait dataset, so potentially a lot of coverage with one dataset and no coverage on all the rest
-#need to get rid of zeros when there is coverage from another dataset so we only have one entry per study
-coverage <- comp_biotime_traits %>%
-  select(study_id, scientificName) %>%
-  distinct() %>%
-  group_by(study_id) %>%
-  summarise(perc = sum(!is.na(scientificName))/n()) %>%
-  arrange(desc(perc))
-
-inc_coverage <- coverage %>% filter(perc > percent)
-
-biotime_traits <- comp_biotime_traits %>%
-  #get studies with appropriate coverage and drop observations that we don't have trait data for
-  filter(study_id %in% inc_coverage$study_id, !is.na(scientificName)) %>% 
-  mutate(plot_id = paste0(study_id, plot))
-```
-
-```{r}
-usethis::use_data(biotime_traits)
-```
-
-Create table of info for each study
-```{r}
-study_table <- biotime_data %>%
-  group_by(study_id) %>%
-  mutate(min_year = min(year), max_year = max(year)) %>% 
-  #select(study_id, plot, taxa, latitude, longitude, min_year, max_year) %>%
-  select(study_id, taxa, min_year, max_year) %>%
-  distinct() %>%
-  left_join(coverage %>% select(study_id, coverage = perc), by = "study_id")
-
-usethis::use_data(study_table)
 ```
@@
 ```{r}
-library(furrr)
 load(here::here("data", "biotime_traits.rda"))
@@
 
-plan(multiprocess)
-system.time(fd <- future_map_dfr(plots[1:10], get_fd, data = data))
+library(furrr)
+future::plan(multiprocess)
+system.time(fd <- future_map_dfr(plots, get_fd_safe, data = data))
 ```
@@
 
-combs <- remain %>%
-  select(study_id, plot) %>%
-  distinct()
+plots <- remain %>%
+  pull(plot_id) %>%
+  unique()
 
-plan(multiprocess)
-system.time(fd <- future_pmap_dfr(combs, ~ get_fd(data = remain, study = .x, plot = .y)))
+library(furrr)
+future::plan(multiprocess)
+system.time(fd <- future_map_dfr(plots, get_fd_safe, data = data))
 ```

### 2020-06-24 analysis/merge_data.Rmd
Move FD calculations to their own script.

--- 2020-06-22 analysis/merge_data.Rmd
@@
 ```
-
-Get FD metrics for each study
-
-FD function:
-```{r}
-library(furrr)
-load(here::here("data", "biotime_traits.rda"))
-
-get_fd <- function(data, study, plot_num){
-  require(FD)
-
-  data <- data %>%
-    filter(study_id == study, plot == plot_num)
-
-  print(max(data$record_id))
-
-  species_mat <- data %>%
-    select(year, id) %>% 
-    distinct() %>%
-    arrange(year, id) %>%
-    mutate(present = 1) %>%
-    spread(id, value = present, fill = 0) %>%
-    select(-year)
-
-  years <- data %>% 
-    select(year) %>%
-    distinct() %>%
-    arrange(year)
-
-  trait_mat <- data %>%
-    select(-c(taxa, record_id, day, month, year, sample_desc, id_species, 
-              latitude, longitude, sum_abundance, sum_biomass, sourceName, scientificName)
-    ) %>%
-    distinct() %>%
-    select_if(~ length(unique(.)) > 1) %>% #remove columns that have the same value for all columns
-    mutate_at(vars(-id, -ends_with("5cat")), list(~scales::rescale(.,to=c(0, 1)))) %>% #rescale variables
-    arrange(id) %>%
-    column_to_rownames("id")
-
-  fs::dir_create(paste0(here::here("data"), "/fd_output"))
-
-  #the convexhull c code creates an output file of vertices, in order to not have multiple threads writing
-  #to the same file we have to create individual working directories
-  tmp_dir <- paste0(here::here("data", "fd_output", "tmp"), "/tmp_", study, plot_num)
-  fs::dir_create(tmp_dir)
-  setwd(tmp_dir)
-
-  fd_out <- as_tibble(dbFD_joggle(trait_mat, species_mat, w.abun = FALSE))[,1:8] %>%
-    cbind(years, .) %>%
-    mutate(study_id = unlist(study), plot = unlist(plot_num), trait_count = dim(trait_mat)[2]) %>%
-    write_tsv(paste0(here::here("data"), "/fd_output/", "s", study, "_p", plot_num, "_fd.tsv"))
-
-  setwd(here::here())
-  return(fd_out)
-}
-
-get_fd_safe <- possibly(get_fd, NA)
-```
-
-First pass at calculating FD metrics
-```{r}
-data <- biotime_traits %>% 
-  filter(!is.na(plot), !is.na(scientificName)) %>% #remove unlabeled plots and observations without traits
-  group_by(study_id, year, plot) %>%
-  filter(n_distinct(id) > 4) %>%
-  ungroup() %>%
-  filter(!study_id %in% c(195, 360), #exclude because they break
-         !study_id %in% c(217, 327)) #exclude to get the file size smaller
-
-combs <- data %>%
-  select(study_id, plot) %>%
-  distinct()
-
-plan(multiprocess)
-system.time(fd <- future_pmap_dfr(combs, ~ get_fd(data = data, study = .x, plot = .y)))
-```
-
-Read data back in, excluding sites/plots as they become a problem:
-```{r}
-files <-  dir(here::here("data", "fd_output"), "*.tsv")
-
-fd_data <- files %>%
-  paste0(here::here("data", "fd_output"), "/", .) %>%
-  map_dfr(read_tsv)
-
-remain <- biotime_traits %>%
-  filter(!is.na(plot), !is.na(scientificName)) %>% #remove unlabeled plots and observations without traits
-  filter(!is.na(plot), !study_id %in% c(195, 360),
-         !study_id %in% c(217, 327), #exclude to get the file size smaller
-         !study_id %in% c(59, 166, 337), #NA's in the distance matrix
-         study_id != 169,  #Zero distance error
-         !study_id %in% c(372,374), #Qhull error code 4
-         study_id != 333) %>% #species distance matrix not Euclidean after 'sqrt' correction
-  filter(study_id != 56 | plot != 24) %>% #NA's in the distance matrix
-  anti_join(fd_data, by = c("year", "study_id", "plot")) %>%
-  group_by(study_id, year, plot) %>%
-  filter(n_distinct(id) > 4) %>%
-  ungroup()
-
-combs <- remain %>%
-  select(study_id, plot) %>%
-  distinct()
-
-plan(multiprocess)
-system.time(fd <- future_pmap_dfr(combs, ~ get_fd(data = remain, study = .x, plot = .y)))
-```
-
-Format tables
-```{r}
-fd_table <- fd_data %>% 
-  select(year, study_id, plot, richness = nbsp, everything(), -sing.sp) %>%
-  gather(metric, value, contains("F"), RaoQ)
-
-usethis::use_data(fd_table)
-```

### 2020-06-25 analysis/calc_fd.Rmd
Calculate FD in chunks to reduce parallelization overhead

@@
 
-get_fd <- function(data, filter_plot){
+get_fd <- function(obs_data, filter_plot){
   require(FD)
 
-  data <- data %>%
+  data <- obs_data %>%
     filter(plot_id == filter_plot)
@@
   setwd(here::here())
+  print(filter_plot)
   return(fd_out)
@@
   filter(n_distinct(id) > 4) %>%
+  group_by(plot_id) %>%
+  filter(n_distinct(year) > 4) %>% 
   ungroup()# %>%
@@
 system.time(fd <- future_map_dfr(plots, get_fd_safe, data = data))
+```
+
+The dataset is too big to copy to each parallel thread, so we split the dataset into chunks 
+```{r}
+#number of chunks to split original dataset into 
+num_groups = 4
+
+plot_sizes <- data %>%
+  count(plot_id) %>%
+  arrange(desc(n)) %>%
+  mutate(group = rep_len(1:num_groups, dim(.)[1]))
+
+get_parallel_FD <- function(data, group_num){
+  group_plots <- plot_sizes %>% filter(group == group_num)
+  group_data <- data %>% filter(plot_id %in% group_plots$plot_id)
+
+  future::plan(multiprocess)
+  fd <- future_map_dfr(group_plots$plot_id, get_fd_safe, data = group_data)
+
+  print(group_num)
+  return(FD)
+}
+
+fd <- map_dfr(1:num_groups, get_parallel_FD, data = data)
 ```

### 2020-07-01 R/pins_initialize.R
first commit

@@ -1,0 +1,13 @@
+##Pins test
+library(tidyverse)
+library(pins)
+
+#check github options to make sure files are added to a release and not the repo
+getOption("pins.github.release") == 0
+#if above expression is FALSE, change .Rprofile option to (usethis::edit_r_profile):
+#options(pins.github.release = 0)
+
+board_register_github(repo = "karinorman/biodivTS", branch = "master")
+
+#to pin:
+#pin(object_name, board = "github")

### 2020-07-02 analysis/calc_fd.Rmd
work around the potential empty dataframes

@@
   future::plan(multiprocess)
-  fd <- future_map_dfr(group_plots$plot_id, get_fd_safe, data = group_data)
+  fd <- future_map(group_plots$plot_id, get_fd_safe, data = group_data) %>%
+    rbind()
 
@@
 
-fd <- map_dfr(1:num_groups, get_parallel_FD, data = data)
+fd <- map(1:num_groups, get_parallel_FD, data = data)
 ```
@@
 
-remain <- biotime_traits %>%
-  filter(!is.na(plot), !is.na(scientificName)) %>% #remove unlabeled plots and observations without traits
-  filter(!is.na(plot), !study_id %in% c(195, 360),
-         !study_id %in% c(217, 327), #exclude to get the file size smaller
-         !study_id %in% c(59, 166, 337), #NA's in the distance matrix
-         study_id != 169,  #Zero distance error
-         !study_id %in% c(372,374), #Qhull error code 4
-         study_id != 333) %>% #species distance matrix not Euclidean after 'sqrt' correction
-  filter(study_id != 56 | plot != 24) %>% #NA's in the distance matrix
-  anti_join(fd_data, by = c("year", "study_id", "plot")) %>%
-  group_by(study_id, year, plot) %>%
-  filter(n_distinct(id) > 4) %>%
-  ungroup()
+# remain <- data %>%
+#   filter(!is.na(plot)) %>% #remove unlabeled plots 
+#   filter(!is.na(plot), !study_id %in% c(195, 360),
+#          !study_id %in% c(217, 327), #exclude to get the file size smaller
+#          !study_id %in% c(59, 166, 337), #NA's in the distance matrix
+#          study_id != 169,  #Zero distance error
+#          !study_id %in% c(372,374), #Qhull error code 4
+#          study_id != 333) %>% #species distance matrix not Euclidean after 'sqrt' correction
+#   filter(study_id != 56 | plot != 24) %>% #NA's in the distance matrix
+#   anti_join(fd_data, by = c("year", "study_id", "plot")) %>%
+#   group_by(study_id, year, plot) %>%
+#   filter(n_distinct(id) > 4) %>%
+#   ungroup()
 
-plots <- remain %>%
-  pull(plot_id) %>%
-  unique()
+plots_left <- plots[!plots %in% fd_data$plot_id]
 
@@
 future::plan(multiprocess)
-system.time(fd <- future_map_dfr(plots, get_fd_safe, data = data))
+system.time(fd <- future_map(plots_left, get_fd_safe, data = data))
 ```

### 2020-07-03 analysis/ts_null_model.Rmd
refactor get_plot_FD function
save fd data for null model as csv, only include plots that meet the criteria for inclusion.

--- 2020-06-04 analysis/ts_null_model.Rmd
@@
 library(biodivTS)
+library(FD)
 ```
@@
 clean_trait_matrix <- function(trait_matrix){
-  uniq_vals <- sapply(trait_matrix, function(x){length(unique(x))}) #get number of unique values of each trait
-  col_names <- names(which(uniq_vals == 1)) #find columns where there is only one value for all observations
-  return(select(trait_matrix, -col_names))
+  # uniq_vals <- sapply(trait_matrix, function(x){length(unique(x))}) #get number of unique values of each trait
+  # col_names <- names(which(uniq_vals == 1)) #find columns where there is only one value for all observations
+  # return(select(trait_matrix, -col_names))
+  trait_matrix %>%
+  select_if(~ length(unique(.)) > 1) %>% #remove columns that have the same value for all columns
+  mutate_at(vars(-id, -ends_with("5cat")), list(~scales::rescale(.,to=c(0, 1)))) %>% #rescale variables
+  arrange(id) %>%
+  column_to_rownames(var = "id")
 }
@@
     dplyr::select(id, all_of(trait_list)) %>%
-    distinct() %>%
-    arrange(id) %>%
-    column_to_rownames(var = "id")
+    distinct()
 }
@@
 
-biotime_traits <- biotime_traits %>%
-  mutate(plot_id = paste0(study_id, plot)) %>%
-  filter(!is.na(id))
-
 n = 100
 
-sample_occurrence <- map_dfr(unique(biotime_traits$plot_id), get_plot_sample, data = biotime_traits, n = n) %>%
+#get the plots that meet our criteria for calculating FD
+sample_plots <- biotime_traits %>%
+  filter(!is.na(plot)) %>% #remove unlabeled plots 
+  group_by(plot_id, year) %>%
+  filter(n_distinct(id) > 4) %>%
+  group_by(plot_id) %>%
+  filter(n_distinct(year) > 4) %>% 
+  ungroup()
+
+sample_occurrence <- map_dfr(unique(sample_plots$plot_id), get_plot_sample, data = sample_plots, n = n) %>%
   replace(is.na(.), 0)
@@
 get_plot_FD <- function(data, trait_range, occ_df, filter_plot) {
-  sample_species_mat <- occ_df %>%
-    filter(plot_id == filter_plot) %>%
+  plot_samples <- occ_df %>%
+    filter(plot_id == filter_plot)
+
+  sample_species_mat <- plot_samples %>%
     select(-c(richness, plot_id, n)) %>% #want only species x site info
@@
 
-  sample_fd <- as.data.frame(dbFD_joggle(x = sample_trait_mat, a = sample_species_mat, w.abun = FALSE))
+  fs::dir_create(paste0(here::here("data"), "/fd_output"))
+
+  #the convexhull c code creates an output file of vertices, in order to not have multiple threads writing
+  #to the same file we have to create individual working directories
+  tmp_dir <- paste0(here::here("data", "fd_output", "tmp"), "/tmp_", filter_plot)
+  fs::dir_create(tmp_dir)
+  setwd(tmp_dir)
+
+  print(filter_plot)
+  sample_fd <- as.data.frame(dbFD_joggle(x = sample_trait_mat, a = sample_species_mat, w.abun = FALSE)) %>%
+    cbind(plot_samples %>% select(richness, plot_id, n))
+
+  save_dir <- paste0(here::here("data", "fd_output", "null"))
+  fs::dir_create(save_dir)
+  write_tsv(sample_fd, paste0(save_dir, "/null", filter_plot, ".csv"))
 
@@
 
-test <- get_plot_FD(biotime_traits, 17:64, sample_occurrence, 5617)
+#test <- get_plot_FD(sample_plots, 18:65, sample_occurrence, 391)
+null_FD <- map_dfr(unique(sample_plots$plot_id)[1:3], get_plot_FD, 
+                   data = sample_plots, trait_range = 18:65, occ_df = sample_occurrence)
 ```

### 2020-07-03 analysis/calc_fd.Rmd
save fd data for null model as csv, only include plots that meet the criteria for inclusion.

--- 2020-06-25 analysis/calc_fd.Rmd
@@
     cbind(years, .) %>%
-    mutate(plot_id = unlist(filter_plot), trait_count = dim(trait_mat)[2]) %>%
-    write_tsv(paste0(here::here("data"), "/fd_output/", filter_plot, "_fd.tsv"))
+    mutate(plot_id = unlist(filter_plot), trait_count = dim(trait_mat)[2])
+    write_tsv(fd_out, paste0(here::here("data"), "/fd_output/", filter_plot, "_fd.tsv"))
 

### 2020-07-21 analysis/ts_null_model.Rmd
save fd data for null model as csv, only include plots that meet the criteria for inclusion.
Null model in parallel.

@@
 
+Calculate null models in parallel
+```{r}
+#number of chunks to split original dataset into 
+num_groups = 4
+
+plot_sizes <- sample_plots %>%
+  count(plot_id) %>%
+  arrange(desc(n)) %>%
+  mutate(group = rep_len(1:num_groups, dim(.)[1]))
+
+get_parallel_FD <- function(data, group_num){
+  group_plots <- plot_sizes %>% filter(group == group_num)
+  group_data <- data %>% filter(plot_id %in% group_plots$plot_id)
+
+  future::plan(multiprocess)
+  fd <- map_dfr(group_plots$plot_id[1:3], get_plot_FD, 
+                   data = group_data, trait_range = 18:65, occ_df = sample_occurrence)
+    #future_map(group_plots$plot_id, get_fd_safe, data = group_data)
+
+  print(group_num)
+  return(fd)
+}
+
+fd <- map_dfr(1:num_groups, get_parallel_FD, data = data)
+```
+
 BAT

### 2020-07-21 analysis/calc_fd.Rmd
Null model in parallel.

--- 2020-07-02 analysis/calc_fd.Rmd
@@
   print(group_num)
-  return(FD)
+  return(fd)
 }

### 2020-07-24 analysis/calc_fd.Rmd

@@
   future::plan(multiprocess)
-  fd <- future_map(group_plots$plot_id, get_fd_safe, data = group_data) %>%
+  fd <- future_map(group_plots$plot_id, get_fd_safe, obs_data = group_data) %>%
     rbind()
@@
 
-fd <- map(1:num_groups, get_parallel_FD, data = data)
+fd <- map(1:num_groups, get_parallel_FD, obs_data = data)
 ```
@@
 future::plan(multiprocess)
-system.time(fd <- future_map(plots_left, get_fd_safe, data = data))
+system.time(fd <- future_map(plots_left, get_fd_safe, obs_data = data))
 ```

### 2020-07-25 analysis/pins_initialize.R
Add github_pat

@@ -7,6 +7,10 @@ getOption("pins.github.release") == 0
 #if above expression is FALSE, change .Rprofile option to (usethis::edit_r_profile):
 #options(pins.github.release = 0)
 
+#make sure the github PAT is set
+#usethis::edit_r_environ()
+#GITHUB_PAT="1e03178a12e2c53fb5f640cb30f4473349be3b30"
+
 board_register_github(repo = "karinorman/biodivTS", branch = "master")
 
 #to pin:

### 2020-07-25 analysis/calc_fd.Rmd
Clean up reading back in data and creating fd_table.

--- 2020-07-03 analysis/calc_fd.Rmd
@@
   print(filter_plot)
-  return(fd_out)
+  #return(fd_out)
 }
@@
   filter(n_distinct(year) > 4) %>% 
-  ungroup()# %>%
-  #filter(!study_id %in% c(195, 360), #exclude because they break
-         #!study_id %in% c(217, 327)) #exclude to get the file size smaller
+  ungroup() #%>%
+  #filter(#!study_id %in% c(195, 360), #exclude because they break
+         #!study_id %in% c(217, 327, 195)) #exclude to get the file size smaller
 
@@
 
-library(furrr)
-future::plan(multiprocess)
-system.time(fd <- future_map_dfr(plots, get_fd_safe, data = data))
+#system.time(fd <- future_map(plots[1:3], get_fd_safe, obs_data = data))
 ```
@@
 
-get_parallel_FD <- function(data, group_num){
+get_parallel_FD <- function(df, group_num){
   group_plots <- plot_sizes %>% filter(group == group_num)
-  group_data <- data %>% filter(plot_id %in% group_plots$plot_id)
+  group_data <- df %>% filter(plot_id %in% group_plots$plot_id)
 
-  future::plan(multiprocess)
-  fd <- future_map_dfr(group_plots$plot_id, get_fd_safe, data = group_data)
+  future::plan("multiprocess", workers = 4)
+  fd <- future_map(group_plots$plot_id, get_fd_safe, obs_data = group_data) %>%
+    rbind()
 
   print(group_num)
-  return(FD)
+  return(fd)
 }
 
-fd <- map_dfr(1:num_groups, get_parallel_FD, data = data)
+#fd <- map(1:num_groups, get_parallel_FD, df = data)
 ```
@@
 
-remain <- biotime_traits %>%
-  filter(!is.na(plot), !is.na(scientificName)) %>% #remove unlabeled plots and observations without traits
-  filter(!is.na(plot), !study_id %in% c(195, 360),
-         !study_id %in% c(217, 327), #exclude to get the file size smaller
-         !study_id %in% c(59, 166, 337), #NA's in the distance matrix
-         study_id != 169,  #Zero distance error
-         !study_id %in% c(372,374), #Qhull error code 4
-         study_id != 333) %>% #species distance matrix not Euclidean after 'sqrt' correction
-  filter(study_id != 56 | plot != 24) %>% #NA's in the distance matrix
-  anti_join(fd_data, by = c("year", "study_id", "plot")) %>%
-  group_by(study_id, year, plot) %>%
-  filter(n_distinct(id) > 4) %>%
-  ungroup()
+remain <- data %>%
+  filter(!plot_id %in% fd_data$plot_id, !plot_id %in% c(1661, 196456, 3721, 3601))
 
-plots <- remain %>%
-  pull(plot_id) %>%
-  unique()
+plot_sizes <- remain %>%
+  count(plot_id) %>%
+  arrange(desc(n)) %>%
+  mutate(group = rep_len(1:num_groups, dim(.)[1]))
 
-library(furrr)
-future::plan(multiprocess)
-system.time(fd <- future_map_dfr(plots, get_fd_safe, data = data))
+#fd <- map(1:num_groups, get_parallel_FD, df = remain)
+
+system.time(fd <- map(plot_sizes$plot_id, get_fd_safe, obs_data = data))
 ```
@@
 fd_table <- fd_data %>% 
-  select(year, study_id, plot, richness = nbsp, everything(), -sing.sp) %>%
-  gather(metric, value, contains("F"), RaoQ)
+  select(year, plot_id, richness = nbsp, everything(), -sing.sp) %>%
+  gather(metric, value, contains("F"), RaoQ) %>%
+  left_join(biotime_traits %>% 
+              select(plot_id, study_id, plot) %>% 
+              mutate(plot_id = as.numeric(plot_id)) %>%
+                       distinct())
 

### 2020-07-29 analysis/calc_fd.Rmd
Write out attempted files

@@
 
-  print(filter_plot)
+  fs::dir_create(paste0(here::here("data"), "/fd_attempt"))
+  write_tsv(data.frame(), paste0(here::here("data"), "/fd_attempt/", filter_plot, ".tsv"))
 
@@
 
+attempted <- dir(here::here("data", "fd_attempt"), "*.tsv") %>%
+  str_replace(".tsv", "") %>%
+  as.numeric()
+
 remain <- data %>%
-  filter(!plot_id %in% fd_data$plot_id, !plot_id %in% c(1661, 196456, 3721, 3601))
+  filter(!plot_id %in% fd_data$plot_id, 
+         !plot_id %in% attempted,
+         !plot_id %in% c(1661, 196456, 3721, 3601, 2171))#1952507
 

### 2020-07-29 analysis/pins_initialize.R
Remove the PAT

@@ -9,7 +9,6 @@ getOption("pins.github.release") == 0
 
 #make sure the github PAT is set
 #usethis::edit_r_environ()
-#GITHUB_PAT="1e03178a12e2c53fb5f640cb30f4473349be3b30"
 
 board_register_github(repo = "karinorman/biodivTS", branch = "master")
 

### 2020-08-19 analysis/visualization.Rmd
Update plots

@@
 ```{r}
-load(here::here("data", "fd_table.rda"))
+#load(here::here("data", "fd_table.rda"))
+fd_table <- pins::pin_get("fd_table", board = "github")
 
@@
   scale_x_continuous(breaks = scales::pretty_breaks()) +
-  facet_wrap( ~ as.factor(study_id), scales = "free") +
+  facet_wrap( ~ as.factor(plot_id), scales = "free") +
   theme(legend.position = "none") +
@@
 
-fd_table %>%
+fric_trend <- fd_table %>%
   filter(metric == "FRic") %>%
   left_join(metadata %>% select(study_id, climate)) %>%
-  unite(id, study_id, plot) %>%
   ggplot(aes(x = year, y = log(value))) +
-  geom_smooth(aes(color = climate, group = id), method = "glm", se = FALSE) +
-  geom_point(aes(color = climate, group = id)) +
+  # geom_smooth(aes(color = climate, group = plot_id), method = "glm", se = FALSE) +
+  # geom_point(aes(color = climate, group = plot_id)) +
+  geom_smooth(aes(group = plot_id), color = "grey", method = "glm", se = FALSE) +
+  geom_point(aes(group = plot_id) ,color = "grey") +
   geom_smooth(color = "black", method = "glm", se = FALSE) +
   theme(legend.position = "none") 
+ggsave("fric_trends.png", fric_plot)
 ```
+
+metric plotting function
+```{r}
+library(lme4)
+
+plot_metric <- function(metric_name, ylabel){
+  data <- fd_table %>%
+    filter(metric == metric_name) %>%
+    mutate(plot_id = as.factor(plot_id), logvalue = log(value))
+
+  mod <- lmer(logvalue ~ year + (1|plot_id), data = data)
+  coefs <- summary(mod)$coefficients
+
+  print(metric_name) 
+  confint(mod)
+
+  pred <- data.frame(year = min(data$year):max(data$year)) %>%
+    mutate(pred = coefs[1,1] + year * coefs[2,1])
+
+  print()
+  trend_plot <- data %>%
+    left_join(pred, by = "year") %>%
+    #left_join(metadata %>% select(study_id, climate)) %>%
+    ggplot(aes(x = year, y = logvalue)) +
+    # geom_smooth(aes(color = climate, group = plot_id), method = "glm", se = FALSE) +
+    # geom_point(aes(color = climate, group = plot_id)) +
+    geom_point(aes(group = plot_id) ,color = "grey") +
+    geom_smooth(aes(group = plot_id), color = "darkgrey", method = "glm", se = FALSE) +
+    #geom_smooth(color = "black", method = "glm", se = FALSE) +
+    geom_line(mapping = aes(y = pred)) +
+    theme(legend.position = "none") +
+    theme_classic() +
+    ylab(ylabel)
+
+  ggsave(paste0(metric_name, "_trend.png"), trend_plot)
+  return(trend_plot)
+}
+
+plot_metric("FRic", "log(FD)")
+
+plot_map <- tibble(metric_name = c("FRic", "FEve", "FDiv", "FDis"), 
+           ylabel = c("log(FD)", "log(Evenness)", "log(Divergence)", "log(Dispersion)"))
+pmap(plot_map, plot_metric)
+```
+
+get model info
+```{r}
+get_metric_conint <- function(metric_name){
+ data <- fd_table %>%
+    filter(metric == metric_name) %>%
+    mutate(plot_id = as.factor(plot_id), logvalue = log(value))
+
+  mod <- lmer(logvalue ~ year + (1|plot_id), data = data)
+  coefs <- summary(mod)$coefficients
+
+  confint(mod)
+  # as_tibble(confint(mod)) %>%
+  #   mutate(metric = metric_name)
+}
+
+map(c("FRic", "FEve", "FDiv", "FDis"), get_metric_conint)
+```
+
+```{r}
+library(lme4)
+
+fric_data <- fd_table %>%
+  filter(metric == "FRic") %>%
+  mutate(plot_id = as.factor(plot_id), logvalue = log(value))
+
+fric_mod <- lmer(logvalue ~ year + (1|plot_id), data = fric_data)
+coefs <- summary(fric_mod)$coefficients
+
+pred <- data.frame(year = min(fric_data$year):max(fric_data$year)) %>%
+                     mutate(pred = coefs[1,1] + year * coefs[2,1])
+
+fric_trend <- fric_data %>%
+  left_join(pred, by = "year") %>%
+  #left_join(metadata %>% select(study_id, climate)) %>%
+  ggplot(aes(x = year, y = logvalue)) +
+  # geom_smooth(aes(color = climate, group = plot_id), method = "glm", se = FALSE) +
+  # geom_point(aes(color = climate, group = plot_id)) +
+  geom_smooth(aes(group = plot_id), color = "grey", method = "glm", se = FALSE) +
+  geom_point(aes(group = plot_id) ,color = "grey") +
+  #geom_smooth(color = "black", method = "glm", se = FALSE) +
+  geom_line(mapping = aes(y = pred)) +
+  theme(legend.position = "none") +
+  ylab("log(FD)")
+ggsave("fric_trend.png", fric_trend)
+```
+
+```{r}
+feve_data <- fd_table %>%
+  filter(metric == "FEve") %>%
+  mutate(plot_id = as.factor(plot_id), logvalue = log(value))
+
+feve_mod <- glm(logvalue ~ year + plot_id, data = feve_data)
+
+#df3 = cbind(df3, pred = predict(fm1))
+
+feve_trend <- feve_data %>%
+  #cbind(pred = predict(fric_mod)) %>%
+  #left_join(metadata %>% select(study_id, climate)) %>%
+  ggplot(aes(x = year, y = logvalue)) +
+  # geom_smooth(aes(color = climate, group = plot_id), method = "glm", se = FALSE) +
+  # geom_point(aes(color = climate, group = plot_id)) +
+  geom_smooth(aes(group = plot_id), color = "grey", method = "glm", se = FALSE) +
+  geom_point(aes(group = plot_id) ,color = "grey") +
+  geom_smooth(color = "black", method = "glm", se = FALSE) +
+  #geom_line(mapping = aes(y = pred)) +
+  theme(legend.position = "none") +
+  ylab("log(Evenness)")
+ggsave("feve_trend.png", feve_trend)
+```
+
+```{r}
+fdis_data <- fd_table %>%
+  filter(metric == "FDis") %>%
+  mutate(plot_id = as.factor(plot_id), logvalue = log(value))
+
+fdis_mod <- glm(logvalue ~ year + plot_id, data = fdis_data)
+
+#df3 = cbind(df3, pred = predict(fm1))
+
+fdis_trend <- fdis_data %>%
+  #cbind(pred = predict(fric_mod)) %>%
+  #left_join(metadata %>% select(study_id, climate)) %>%
+  ggplot(aes(x = year, y = logvalue)) +
+  # geom_smooth(aes(color = climate, group = plot_id), method = "glm", se = FALSE) +
+  # geom_point(aes(color = climate, group = plot_id)) +
+  geom_smooth(aes(group = plot_id), color = "grey", method = "glm", se = FALSE) +
+  geom_point(aes(group = plot_id) ,color = "grey") +
+  geom_smooth(color = "black", method = "glm", se = FALSE) +
+  #geom_line(mapping = aes(y = pred)) +
+  theme(legend.position = "none") +
+  ylab("log(Dispersion)")
+ggsave("fdis_trend.png", fdis_trend)
+```
+
+```{r}
+fdiv_data <- fd_table %>%
+  filter(metric == "FDiv") %>%
+  mutate(plot_id = as.factor(plot_id), logvalue = log(value))
+
+fdiv_mod <- glm(logvalue ~ year + plot_id, data = fdiv_data)
+
+#df3 = cbind(df3, pred = predict(fm1))
+
+fdiv_trend <- fdiv_data %>%
+  #cbind(pred = predict(fric_mod)) %>%
+  #left_join(metadata %>% select(study_id, climate)) %>%
+  ggplot(aes(x = year, y = logvalue)) +
+  # geom_smooth(aes(color = climate, group = plot_id), method = "glm", se = FALSE) +
+  # geom_point(aes(color = climate, group = plot_id)) +
+  geom_smooth(aes(group = plot_id), color = "grey", method = "glm", se = FALSE) +
+  geom_point(aes(group = plot_id) ,color = "grey") +
+  geom_smooth(color = "black", method = "glm", se = FALSE) +
+  #geom_line(mapping = aes(y = pred)) +
+  theme(legend.position = "none") +
+  ylab("log(Dispersion)")
+ggsave("fdis_trend.png", fdiv_trend)
+```

### 2020-08-24 analysis/calc_fd.Rmd
Incorporate abundance into FD calculations and scale using the z score

@@
 library(FD)
+library(biodivTS)
 ```
@@
 ```{r}
-load(here::here("data", "biotime_traits.rda"))
+biotime_traits <- pins::pin_get("biotime-traits", board = "github")
 
@@
   species_mat <- data %>%
-    select(year, id) %>% 
+    select(year, id, sum_abundance) %>% 
     distinct() %>%
     arrange(year, id) %>%
-    mutate(present = 1) %>%
-    spread(id, value = present, fill = 0) %>%
+    spread(id, value = sum_abundance, fill = 0) %>%
     select(-year)
@@
     distinct() %>%
-    select_if(~ length(unique(.)) > 1) %>% #remove columns that have the same value for all columns
-    mutate_at(vars(-id, -ends_with("5cat")), list(~scales::rescale(.,to=c(0, 1)))) %>% #rescale variables
+    select_if(~ length(unique(na.omit(.))) > 1) #remove columns that have the same value for all columns
+
+  #get binary variables so they can be excluded from rescaling 
+  bin_vars <- map(trait_mat, ~ all(na.omit(.) %in% 0:1))
+
+  trait_mat <- trait_mat %>%
+    mutate_at(vars(-c(id, names(bin_vars[bin_vars == TRUE])), -ends_with("5cat")), 
+              list(~as.numeric(scale(.)))) %>% #rescale variables, not binary or categorical
     arrange(id) %>%

### 2020-08-31 analysis/calc_fd.Rmd
Cap number of PCA axes and use Cailliez rather than square root correction.

@@
 library(biodivTS)
+
+pins::board_register_github(repo = "karinorman/biodivTS", branch = "master")
 ```
@@
 
-get_fd <- function(obs_data, filter_plot){
+get_fd <- function(obs_data, filter_plot, ...){
   require(FD)
@@
   species_mat <- data %>%
-    select(year, id, sum_abundance) %>% 
-    distinct() %>%
-    arrange(year, id) %>%
-    spread(id, value = sum_abundance, fill = 0) %>%
+    group_by(year, id) %>%
+    summarise(abun = sum(sum_abundance)) %>%
+    ungroup() %>%
+    arrange(year) %>%
+    arrange(id) %>%
+    pivot_wider(names_from = id, values_from = abun, values_fill = 0) %>%
     select(-year)
@@
 
-  fd_out <- as_tibble(dbFD_joggle(trait_mat, species_mat, w.abun = FALSE))[,1:8] %>%
+  fd_out <- as_tibble(dbFD_joggle(trait_mat, species_mat, w.abun = TRUE, m = "min", ...))[,1:8] %>%
     cbind(years, .) %>%
@@
          #!study_id %in% c(217, 327, 195)) #exclude to get the file size smaller
-
-plots <- data %>%
-  pull(plot_id) %>%
-  unique()
-
-#system.time(fd <- future_map(plots[1:3], get_fd_safe, obs_data = data))
 ```
@@
 
-attempted <- dir(here::here("data", "fd_attempt"), "*.tsv") %>%
-  str_replace(".tsv", "") %>%
-  as.numeric()
-
 remain <- data %>%
-  filter(!plot_id %in% fd_data$plot_id, 
-         !plot_id %in% attempted,
-         !plot_id %in% c(1661, 196456, 3721, 3601, 2171))#1952507
+  filter(!plot_id %in% fd_data$plot_id)
 
 plot_sizes <- remain %>%
-  count(plot_id) %>%
-  arrange(desc(n)) %>%
-  mutate(group = rep_len(1:num_groups, dim(.)[1]))
+  count(plot_id)
 
-#fd <- map(1:num_groups, get_parallel_FD, df = remain)
-
-system.time(fd <- map(plot_sizes$plot_id, get_fd_safe, obs_data = data))
+#get_fd_error <- safely(get_fd)
+#system.time(fd <- map(plot_sizes$plot_id, get_fd_error, obs_data = data, corr = "cailliez"))
+system.time(fd <- map(plot_sizes$plot_id, get_fd_safe, obs_data = data, corr = "cailliez"))
 ```
@@
 usethis::use_data(fd_table)
+pins::pin(fd_table, board = "github")
 ```

### 2020-09-02 analysis/database_access.Rmd
First commit, accessing Biotime database

---
title: "Biotime Database Access"
author: "Kari Norman"
date: "8/31/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(DBI)
library(dbplyr)
library(tidyverse)

con <- dbConnect(
  drv = RMariaDB::MariaDB(), 
  dbname = "biotime",
  username = "root",
  password = "Ka1172353!", 
  host = "localhost"
)

biotime_sql <- tbl(con, "allrawdata") %>% 
  collect() %>% 
  rename_all(.funs = tolower) %>%
  select(-id_all_raw_data)
```

```{r}
dupes <- biotime_sql %>% group_by(study_id, plot, id_species, year, month, day) %>% filter(n() >1)

year_biotime <- biotime_sql %>% 
  group_by(id_species, plot, year, study_id) %>% 
  summarise(year_abun = sum(na.omit(abundance)), 
            year_biomass = sum(na.omit(biomass)))

#should be zero to be appropriately aggregated
#year_biotime %>% group_by(study_id, plot, year, id_species) %>% filter(n()>1) %>% View()

usethis::use_data(year_biotime)
pins::pin(year_biotime, board = "github")
```

### 2020-09-02 analysis/merge_data.Rmd
Update biotime_traits for new biotime data.

--- 2020-06-24 analysis/merge_data.Rmd
@@
 
-check if all bird & mammal biotime data w/id's has an elton match. There are 68 species without a match (63 are birds).
+check if all bird & mammal biotime data w/id's has an elton match. There are 24 species without a match.
 ```{r}
@@
 
-and 19869 species without an ID, some of which are species codes
+and 25268 species without an ID, some of which are species codes
 ```{r}

### 2020-09-03 analysis/calc_fd.Rmd
Calculate FD not in parallel

@@
          #!study_id %in% c(217, 327, 195)) #exclude to get the file size smaller
+
+plots <- data %>%
+  pull(plot_id) %>%
+  unique()
+
+system.time(fd <- map(plots, get_fd_safe, obs_data = data))
 ```

### 2020-09-03 analysis/database_access.Rmd

@@
   summarise(year_abun = sum(na.omit(abundance)), 
-            year_biomass = sum(na.omit(biomass)))
+            year_biomass = sum(na.omit(biomass))) %>%
+  left_join(tbl(con, "datasets") %>% select(taxa = TAXA, study_id = STUDY_ID) %>% collect()) %>%
+  left_join(tbl(con, "species") %>% select(id_species = ID_SPECIES, genus_species = GENUS_SPECIES) %>% collect())
 

### 2020-09-03 data-raw/biotime.Rmd

@@
 library(lazyeval)
-library(rdataretriever)
 library(taxadb)
@@
 #rdataretriever::install("biotime", connection = "csv", data_dir = here("data"))
+# 
+# data <- read_csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS")) %>%
+#   select(study_id, taxa) %>%
+#   left_join(read_csv(system.file("extdata", "biotime/biotime_query.csv", package = "biodivTS")), by = "study_id")
 
-data <- read_csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS")) %>% 
-  select(study_id, taxa) %>%
-  left_join(read_csv(system.file("extdata", "biotime/biotime_query.csv", package = "biodivTS")), by = "study_id")
+data <- pins::pin_get("year-biotime", board = "github")
 ```
@@
   distinct() %>%
-  group_by(study_id) %>%
-  count() %>%
+  count(study_id) %>%
   filter(n == 1) %>%
@@
   mutate(plot = case_when(
-    study_id %in% sing_study & is.na(plot) ~ 1,
+    study_id %in% sing_study & is.na(plot) ~ "A",
     TRUE ~ plot
@@
 #Getting matches by sci name and common name 
-biotime_ids <- by_name(unique(data$genus_species), "itis") %>%
-  bind_rows(by_common(unique(data$genus_species), "itis")) %>%
+biotime_ids <- filter_name(unique(data$genus_species), "itis") %>%
+  bind_rows(filter_common(unique(data$genus_species), "itis")) %>%
   filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
@@
   filter(acceptedNameUsageID %in% c(
+    "ITIS:1026896", #accepted on the ITIS website
+    "ITIS:172921",
+    "ITIS:683209",
+    "ITIS:527684",
+    "ITIS:782604",
+    "ITIS:28107",
+    "ITIS:508923",
     "ITIS:159807", #Heterodontus zebra, all the studies including it are using sci names, so we should match to the sci name match not common name match
@@
   distinct() %>%
-  right_join(data %>% select(-species, -genus), by = c("input" = "genus_species")) %>%
+  right_join(data, by = c("input" = "genus_species")) %>%
   rename(sourceName = input)

### 2020-09-04 analysis/calc_fd.Rmd
Modify for new biotime format

@@
   species_mat <- data %>%
-    group_by(year, id) %>%
-    summarise(abun = sum(sum_abundance)) %>%
-    ungroup() %>%
+    select(year, id, year_abun) %>%
     arrange(year) %>%
     arrange(id) %>%
-    pivot_wider(names_from = id, values_from = abun, values_fill = 0) %>%
+    pivot_wider(names_from = id, values_from = year_abun, values_fill = 0) %>%
     select(-year)
@@
   trait_mat <- data %>%
-    select(-c(taxa, record_id, day, month, year, sample_desc, id_species, 
-              latitude, longitude, sum_abundance, sum_biomass, traitName, 
-              biotimeName, scientificName)
+    select(-c(biotimeName, taxa, id_species, plot, year, year_abun, 
+              year_biomass, traitName, scientificName, plot_id, study_id)
     ) %>%
@@
   bin_vars <- map(trait_mat, ~ all(na.omit(.) %in% 0:1))
+  bin_vars <- names(bin_vars[bin_vars == TRUE])
 
   trait_mat <- trait_mat %>%
-    mutate_at(vars(-c(id, names(bin_vars[bin_vars == TRUE])), -ends_with("5cat")), 
+    mutate_at(vars(-c(id, bin_vars), -ends_with("5cat")), 
               list(~as.numeric(scale(.)))) %>% #rescale variables, not binary or categorical
+    mutate_at(vars(bin_vars), as.character(.)) %>%
     arrange(id) %>%
@@
 
-system.time(fd <- map(plots, get_fd_safe, obs_data = data))
+system.time(fd <- map(plots, get_fd_safe, obs_data = data, m = "min", corr = "cailliez"))
 ```

### 2020-09-04 data-raw/biotime.Rmd
Label all plots not just NA's

@@
   mutate(plot = case_when(
-    study_id %in% sing_study & is.na(plot) ~ "A",
+    study_id %in% sing_study & plot %in% c(NA, "", "#REF!") ~ "A",
     TRUE ~ plot

### 2020-09-07 analysis/merge_data.Rmd
Calculate FD not in parallel
mod for biotime query data

@@
 
-check if all bird & mammal biotime data w/id's has an elton match. There are 24 species without a match.
+check if all bird & mammal biotime data w/id's has an elton match. There are 18 species without a match.
 ```{r}
@@
   group_by(study_id) %>%
-  summarise(perc = sum(!is.na(scientificName))/n()) %>%
+  summarise(perc = sum(!is.na(id))/n()) %>%
   arrange(desc(perc))
@@
 usethis::use_data(biotime_traits)
+pins::pin(biotime_traits, board = "github")
 ```
@@
   group_by(study_id) %>%
-  mutate(min_year = min(year), max_year = max(year)) %>% 
+  mutate(min_year = min(year), max_year = max(year), n_year = n_distinct(year)) %>% 
   #select(study_id, plot, taxa, latitude, longitude, min_year, max_year) %>%

### 2020-09-07 data-raw/biotime.Rmd
mod for biotime query data

@@
 
+Some ID's matched to more than one species in biotime, so we need to consolidate those entries so there is only one per ID (by removing the sourceName) 
+```{r}
+#find the entries for species that match to the same ID as another species
+dupe_id_entry <- biotime_data %>% 
+  group_by(id) %>% 
+  filter(n_distinct(sourceName) > 1)
+
+biotime_data <- setdiff(biotime_data, dupe_id_entry) %>%
+  bind_rows(dupe_id_entry %>% 
+              select(-sourceName, id_species) %>% 
+              group_by(id, scientificName, plot, year, study_id, taxa) %>% 
+              summarize(year_abun = sum(year_abun), year_biomass = sum(year_biomass)))
+```
+
 Do any unmatched species have matches with other providers?
+
+NOT EXECUTED FOR CURRENT ITERATION (9/4/20)
 ```{r}
-
-#Check for potential matches from synonyms found in the other providers
-syn_res <- match_providers(biotime_data, "itis", common = TRUE)
-
-#some names have multiple matches (same ID but multiple scinames), have to pick one
-## in this case all the duplicates have an accepted and a synonym, but there are some single matches that are just synonyms
-## so we have to find duplicate ID's and then filter
-alt_dupes <- syn_res %>%
-  select(acceptedNameUsageID, input) %>%
-  distinct() %>%
-  group_by(input) %>%
-  filter(n() > 1)
-
-# #filter duplicates for accepted, and replace old entries 
-# syn_res <- alt_dupes %>% 
-#   filter(acceptedNameUsageID %in% c(
-#     ITIS:331277, 
-#   )) %>% 
-#   bind_rows(syn_res %>% 
-#               filter(!acceptedNameUsageID %in% alt_dupes$acceptedNameUsageID)) %>%
-# #then finish clean up 
-#   select(id = acceptedNameUsageID, input, scientificName) %>% 
+# 
+# #Check for potential matches from synonyms found in the other providers
+# syn_res <- match_providers(biotime_data, "itis", common = TRUE)
+# 
+# #some names have multiple matches (same ID but multiple scinames), have to pick one
+# ## in this case all the duplicates have an accepted and a synonym, but there are some single matches that are just synonyms
+# ## so we have to find duplicate ID's and then filter
+# alt_dupes <- syn_res %>%
+#   select(acceptedNameUsageID, input) %>%
 #   distinct() %>%
-#   left_join(data %>% select(-species, -genus), by = c("input" = "genus_species"), na_matches =  "never") %>%
-#   rename(sourceName = input) %>%
-#   ungroup()
-
-#Just get rid of those with duplicates, deal with them later
-add_syn_res <- syn_res %>% 
-  filter(!input %in% alt_dupes$input) %>% #excluding duplicate matched species
-  select(id = acceptedNameUsageID, sourceName = input, scientificName) %>%
-  distinct() %>%
-  group_by(sourceName) %>%
-  top_n(1, scientificName) %>% #some have multiple sci names for same id, just pick on 
-  left_join(biotime_data%>% select(-c(id, scientificName)), by = "sourceName", na_matches =  "never") %>%
-  ungroup() %>%
-  distinct()
-
-biotime_data <- biotime_data %>% 
-  filter(!sourceName %in% add_syn_res$sourceName) %>%
-  bind_rows(add_syn_res)
+#   group_by(input) %>%
+#   filter(n() > 1)
+# 
+# # #filter duplicates for accepted, and replace old entries 
+# # syn_res <- alt_dupes %>% 
+# #   filter(acceptedNameUsageID %in% c(
+# #     ITIS:331277, 
+# #   )) %>% 
+# #   bind_rows(syn_res %>% 
+# #               filter(!acceptedNameUsageID %in% alt_dupes$acceptedNameUsageID)) %>%
+# # #then finish clean up 
+# #   select(id = acceptedNameUsageID, input, scientificName) %>% 
+# #   distinct() %>%
+# #   left_join(data %>% select(-species, -genus), by = c("input" = "genus_species"), na_matches =  "never") %>%
+# #   rename(sourceName = input) %>%
+# #   ungroup()
+# 
+# #Just get rid of those with duplicates, deal with them later
+# add_syn_res <- syn_res %>% 
+#   filter(!input %in% alt_dupes$input) %>% #excluding duplicate matched species
+#   select(id = acceptedNameUsageID, sourceName = input, scientificName) %>%
+#   distinct() %>%
+#   group_by(sourceName) %>%
+#   top_n(1, scientificName) %>% #some have multiple sci names for same id, just pick on 
+#   left_join(biotime_data%>% select(-c(id, scientificName)), by = "sourceName", na_matches =  "never") %>%
+#   ungroup() %>%
+#   distinct()
+# 
+# biotime_data <- biotime_data %>% 
+#   filter(!sourceName %in% add_syn_res$sourceName) %>%
+#   bind_rows(add_syn_res)
 

### 2020-09-07 analysis/visualization.Rmd

--- 2020-03-05 analysis/visualization.Rmd
@@
 
-fd_table %>%
+fric_trend <- fd_table %>%
   filter(metric == "FRic") %>%
   left_join(metadata %>% select(study_id, climate)) %>%
-  unite(id, study_id, plot) %>%
   ggplot(aes(x = year, y = log(value))) +
-  geom_smooth(aes(color = climate, group = id), method = "glm", se = FALSE) +
-  geom_point(aes(color = climate, group = id)) +
+  # geom_smooth(aes(color = climate, group = plot_id), method = "glm", se = FALSE) +
+  # geom_point(aes(color = climate, group = plot_id)) +
+  geom_smooth(aes(group = plot_id), color = "grey", method = "glm", se = FALSE) +
+  geom_point(aes(group = plot_id) ,color = "grey") +
   geom_smooth(color = "black", method = "glm", se = FALSE) +
   theme(legend.position = "none") 
+ggsave("fric_trends.png", fric_plot)
 ```
+
+metric plotting function
+```{r}
+library(lme4)
+
+plot_metric <- function(metric_name, ylabel){
+  data <- fd_table %>%
+    filter(metric == metric_name) %>%
+    mutate(plot_id = as.factor(plot_id), logvalue = log(value))
+
+  mod <- lmer(logvalue ~ year + (1|plot_id), data = data)
+  coefs <- summary(mod)$coefficients
+
+  print(metric_name) 
+  print(summary(mod))
+  confint(mod)
+
+  pred <- data.frame(year = min(data$year):max(data$year)) %>%
+    mutate(pred = coefs[1,1] + year * coefs[2,1])
+
+  trend_plot <- data %>%
+    left_join(pred, by = "year") %>%
+    #left_join(metadata %>% select(study_id, climate)) %>%
+    ggplot(aes(x = year, y = logvalue)) +
+    # geom_smooth(aes(color = climate, group = plot_id), method = "glm", se = FALSE) +
+    # geom_point(aes(color = climate, group = plot_id)) +
+    geom_point(aes(group = plot_id) ,color = "grey") +
+    geom_smooth(aes(group = plot_id), color = "darkgrey", method = "glm", se = FALSE) +
+    #geom_smooth(color = "black", method = "glm", se = FALSE) +
+    geom_line(mapping = aes(y = pred)) +
+    theme(legend.position = "none") +
+    theme_classic() +
+    ylab(ylabel)
+
+  ggsave(paste0(metric_name, "_trend.png"), trend_plot)
+  return(trend_plot)
+}
+
+plot_metric("FRic", "log(FD)")
+
+plot_map <- tibble(metric_name = c("FRic", "FEve", "FDiv", "FDis"), 
+           ylabel = c("log(FD)", "log(Evenness)", "log(Divergence)", "log(Dispersion)"))
+pmap(plot_map, plot_metric)
+```
+
+get model info
+```{r}
+get_metric_conint <- function(metric_name){
+ data <- fd_table %>%
+    filter(metric == metric_name) %>%
+    mutate(plot_id = as.factor(plot_id), logvalue = log(value))
+
+  mod <- lmer(logvalue ~ year + (1|plot_id), data = data)
+  coefs <- summary(mod)$coefficients
+
+  confint(mod)
+  # as_tibble(confint(mod)) %>%
+  #   mutate(metric = metric_name)
+}
+
+map(c("FRic", "FEve", "FDiv", "FDis"), get_metric_conint)
+```

### 2020-09-07 analysis/merge_data.Rmd
Get rid of duplicates

@@
 comp_biotime_traits <- map_dfr(traits, merge_data) %>%
-  right_join(biotime_data %>% select(-sourceName, -scientificName))
+  right_join(biotime_data %>% select(-sourceName, -scientificName)) %>%
+  distinct()
 
@@
 usethis::use_data(study_table)
+pins::pin(study_table, board = "github")
 ```

### 2020-09-08 data-raw/biotime.Rmd
Switch to base R read csv to avoid parsing failures

@@
 # 
-# data <- read_csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS")) %>%
+# data <- read.csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS")) %>%
 #   select(study_id, taxa) %>%
-#   left_join(read_csv(system.file("extdata", "biotime/biotime_query.csv", package = "biodivTS")), by = "study_id")
+#   left_join(read.csv(system.file("extdata", "biotime/biotime_query.csv", package = "biodivTS")), by = "study_id")
 

### 2020-09-08 data-raw/biotime.Rmd
Get rid of duplicates from synonym scientific names matching to the same ID.

@@
 biotime_data <- biotime_ids %>%
-  select(id = acceptedNameUsageID,input, scientificName) %>%
+  select(id = acceptedNameUsageID,input) %>%
   distinct() %>%
   right_join(data, by = c("input" = "genus_species")) %>%
-  rename(sourceName = input)
+  rename(sourceName = input) %>%
+  mutate(scientificName = taxadb::get_names(id))
 ```

### 2020-09-08 analysis/merge_data.Rmd
Add number of plots to each study.

@@
   group_by(study_id) %>%
-  mutate(min_year = min(year), max_year = max(year), n_year = n_distinct(year)) %>% 
-  #select(study_id, plot, taxa, latitude, longitude, min_year, max_year) %>%
-  select(study_id, taxa, min_year, max_year) %>%
+  mutate(min_year = min(year), max_year = max(year), n_year = n_distinct(year), plots = n_distinct(plot)) %>% 
+  select(study_id, taxa, min_year, max_year, plots) %>%
   distinct() %>%

### 2020-09-09 analysis/ts_null_model.Rmd
Null model in parallel.
Modify get_plot_sample and get_plot_FD for new data structure.

@@
 library(FD)
+
+pins::board_register_github(repo = "karinorman/biodivTS", branch = "master")
 ```
@@
 get_plot_sample <- function(data, filter_plot, n){
-  plot_data <- data %>% filter(plot_id == filter_plot)
-
-  rich_levels <- plot_data %>%
-    select(year, id) %>%
-    distinct() %>%
-    count(year) %>%
-    pull(n) %>%
-    unique()
+  print(filter_plot)
+  plot_data <- data %>% filter(plot_id == filter_plot) %>%
+    select(year, id, year_abun)
 
@@
 
-  if (length(species_pool) < max(rich_levels)) browser()
+  get_samp_species <- function(data){
+    samp_species <- sample(species_pool, n_distinct(data$id))
+    samp_abun <- sample(data$year_abun)
 
-  get_samp_species <- function(richness){
-    samp_species <- sample(species_pool, richness)
-    df <- cbind(richness, samp_species)
+    df <- as_tibble(cbind(samp_species, samp_abun)) %>% mutate(year = unique(data$year))
     return(df)
@@
     iter_samples <- data.frame()
-    for(i in rich_levels){
-    iter_samples <- rbind(iter_samples, get_samp_species(i))
+    for(i in unique(plot_data$year)){
+      year_data <- plot_data %>% filter(year == i)
+      iter_samples <- rbind(iter_samples, get_samp_species(year_data))
     }
@@
   site_samples %>%
-    mutate(plot_id = filter_plot, value = 1) %>%
-    pivot_wider(names_from = samp_species, values_from = value)
+    mutate(plot_id = filter_plot) %>%
+    pivot_wider(names_from = samp_species, values_from = samp_abun)
 }
@@
 
-sample_occurrence <- map_dfr(unique(sample_plots$plot_id), get_plot_sample, data = sample_plots, n = n) %>%
-  replace(is.na(.), 0)
+# sample_occurrence <- map(unique(sample_plots$plot_id), get_plot_sample, data = sample_plots, n = n) %>%
+#   bind_rows() %>%
+#   replace(is.na(.), 0)
+# pins::pin(sample_occurrence, board = "github")
+sample_occurrence <- pins::pin_get("sample_occurrence", board = "github")
 
-#colnames(biotime_traits)[17:64]
-
-get_plot_FD <- function(data, trait_range, occ_df, filter_plot) {
+get_plot_FD <- function(data, occ_df, filter_plot, ...) {
   plot_samples <- occ_df %>%
@@
   sample_species_mat <- plot_samples %>%
-    select(-c(richness, plot_id, n)) %>% #want only species x site info
-    select_if(!colSums(.) %in% c(0, NA)) %>% #remove species that don't occur in the plot
+    select(-c(year, plot_id, n)) %>% #want only species x site info
+    select_if(!colSums(na.omit(.)) %in% c(0, NA)) %>% #remove species that don't occur in the plot
     select(sort(tidyselect::peek_vars())) #sort columns to be in alphabetical order, required for dbFDI()
 
-  traits <- colnames(data)[trait_range]
-  sample_trait_mat <-
-    clean_trait_matrix(get_trait_matrix(data = data, colnames(sample_species_mat), traits))
+  trait_mat <- data %>%
+    filter(id %in% colnames(sample_species_mat)) %>%
+    select(-c(biotimeName, taxa, id_species, plot, year, year_abun, 
+              year_biomass, traitName, scientificName, plot_id, study_id)
+    ) %>%
+    distinct() %>%
+    select_if(~ length(unique(na.omit(.))) > 1)  #remove columns that have the same value for all columns
 
-  fs::dir_create(paste0(here::here("data"), "/fd_output"))
+  #get binary variables so they can be excluded from rescaling 
+  bin_vars <- map(trait_mat, ~ all(na.omit(.) %in% 0:1))
+
+  trait_mat <- trait_mat %>%
+    mutate_at(vars(-c(id, names(bin_vars[bin_vars == TRUE])), -ends_with("5cat")), 
+              list(~as.numeric(scale(.)))) %>% #rescale variables, not binary or categorical
+    arrange(id) %>%
+    column_to_rownames("id")
+
+  save_dir <- here::here("data", "fd_null_output")
+  fs::dir_create(save_dir)
 
@@
   #to the same file we have to create individual working directories
-  tmp_dir <- paste0(here::here("data", "fd_output", "tmp"), "/tmp_", filter_plot)
+  tmp_dir <- paste0(save_dir, "/tmp", "/tmp_", filter_plot)
   fs::dir_create(tmp_dir)
@@
   print(filter_plot)
-  sample_fd <- as.data.frame(dbFD_joggle(x = sample_trait_mat, a = sample_species_mat, w.abun = FALSE)) %>%
-    cbind(plot_samples %>% select(richness, plot_id, n))
+  sample_fd <- as.data.frame(dbFD_joggle(x = trait_mat, a = sample_species_mat, w.abun = TRUE, ...)) %>%
+    cbind(plot_samples %>% select(year, plot_id, n))
 
-  save_dir <- paste0(here::here("data", "fd_output", "null"))
-  fs::dir_create(save_dir)
+  setwd(here::here())
   write_tsv(sample_fd, paste0(save_dir, "/null", filter_plot, ".csv"))
 
-  df <- occ_df %>%
-    filter(plot_id == filter_plot) %>%
-    select(richness, plot_id, n) %>%
-    bind_cols(sample_fd) %>%
-    mutate(merge_test = case_when(richness == nbsp ~ TRUE,
-                                  TRUE ~ FALSE))
+  # df <- occ_df %>%
+  #   filter(plot_id == filter_plot) %>%
+  #   select(richness, plot_id, n) %>%
+  #   bind_cols(sample_fd) %>%
+  #   mutate(merge_test = case_when(richness == nbsp ~ TRUE,
+  #                                 TRUE ~ FALSE))
+  # 
+  # if (sum(df$merge_test) != dim(df)[1])
+  #   warning("richness and region columns may not have been appropriately joined with FD data")
 
-  if (sum(df$merge_test) != dim(df)[1])
-    warning("richness and region columns may not have been appropriately joined with FD data")
-
-  return(df %>% select(-merge_test))
+  return(sample_fd)
 }
 
-#test <- get_plot_FD(sample_plots, 18:65, sample_occurrence, 391)
 null_FD <- map_dfr(unique(sample_plots$plot_id)[1:3], get_plot_FD, 
-                   data = sample_plots, trait_range = 18:65, occ_df = sample_occurrence)
+                   data = sample_plots, occ_df = sample_occurrence,
+                   m = "min", corr = "cailliez")
 ```
@@
 row.names(traits) <- row.names(traits2)
-
 ```

### 2020-09-09 analysis/ts_null_model.Rmd
Modify to calculate in parallel

@@
 }
-
-#drop trait columns that are the same value for every observation, and therefore providing no additional info
-clean_trait_matrix <- function(trait_matrix){
-  # uniq_vals <- sapply(trait_matrix, function(x){length(unique(x))}) #get number of unique values of each trait
-  # col_names <- names(which(uniq_vals == 1)) #find columns where there is only one value for all observations
-  # return(select(trait_matrix, -col_names))
-  trait_matrix %>%
-  select_if(~ length(unique(.)) > 1) %>% #remove columns that have the same value for all columns
-  mutate_at(vars(-id, -ends_with("5cat")), list(~scales::rescale(.,to=c(0, 1)))) %>% #rescale variables
-  arrange(id) %>%
-  column_to_rownames(var = "id")
-}
-
-get_trait_matrix <- function(data, species_list, trait_list){ 
-  traits <- data %>%
-    filter(id %in% species_list) %>%
-    dplyr::select(id, all_of(trait_list)) %>%
-    distinct()
-}
 ```
@@
 
+trait_data <- biotime_traits %>% 
+  select(-c(biotimeName, taxa, id_species, plot, year, year_abun, 
+              year_biomass, traitName, scientificName, plot_id, study_id)) %>%
+  distinct()
+
 # sample_occurrence <- map(unique(sample_plots$plot_id), get_plot_sample, data = sample_plots, n = n) %>%
@@
 
-get_plot_FD <- function(data, occ_df, filter_plot, ...) {
+get_plot_FD <- function(traits, occ_df, filter_plot, ...) {
   plot_samples <- occ_df %>%
@@
 
-  trait_mat <- data %>%
+  trait_mat <- traits %>%
     filter(id %in% colnames(sample_species_mat)) %>%
-    select(-c(biotimeName, taxa, id_species, plot, year, year_abun, 
-              year_biomass, traitName, scientificName, plot_id, study_id)
-    ) %>%
-    distinct() %>%
     select_if(~ length(unique(na.omit(.))) > 1)  #remove columns that have the same value for all columns
@@
 null_FD <- map_dfr(unique(sample_plots$plot_id)[1:3], get_plot_FD, 
-                   data = sample_plots, occ_df = sample_occurrence,
+                   traits = trait_data, occ_df = sample_occurrence,
                    m = "min", corr = "cailliez")
@@
 
-get_parallel_FD <- function(data, group_num){
+get_parallel_FD <- function(group_num){
+  browser()
   group_plots <- plot_sizes %>% filter(group == group_num)
-  group_data <- data %>% filter(plot_id %in% group_plots$plot_id)
+  group_data <- sample_occurrence %>% filter(plot_id %in% group_plots$plot_id)
 
@@
   fd <- map_dfr(group_plots$plot_id[1:3], get_plot_FD, 
-                   data = group_data, trait_range = 18:65, occ_df = sample_occurrence)
+                   traits = group_trait_data, occ_df = group_data)
     #future_map(group_plots$plot_id, get_fd_safe, data = group_data)
@@
 
-fd <- map_dfr(1:num_groups, get_parallel_FD, data = data)
+fd <- map_dfr(1:num_groups, get_parallel_FD)
 ```

### 2020-09-11 analysis/calc_fd.Rmd
Calculate FD based on the weight column, read in plots as characters

--- 2020-08-31 analysis/calc_fd.Rmd
@@
 
+  # #check which column the abundaces weights are stored in (abundance or biomass)
+  # if (sum(data$year_abun) == 0) {weight <- "year_biomass"} else {weight <- "year_abun"}
+
   species_mat <- data %>%
-    group_by(year, id) %>%
-    summarise(abun = sum(sum_abundance)) %>%
-    ungroup() %>%
+    select(year, id, weight) %>%
     arrange(year) %>%
     arrange(id) %>%
-    pivot_wider(names_from = id, values_from = abun, values_fill = 0) %>%
+    pivot_wider(names_from = id, values_from = weight, values_fill = 0) %>%
     select(-year)
@@
   trait_mat <- data %>%
-    select(-c(taxa, record_id, day, month, year, sample_desc, id_species, 
-              latitude, longitude, sum_abundance, sum_biomass, traitName, 
-              biotimeName, scientificName)
+    select(-c(biotimeName, taxa, id_species, plot, year, weight, 
+              traitName, scientificName, plot_id, study_id)
     ) %>%
@@
 
-  fd_out <- as_tibble(dbFD_joggle(trait_mat, species_mat, w.abun = TRUE, m = "min", ...))[,1:8] %>%
+  fd_out <- as_tibble(dbFD_joggle(trait_mat, species_mat, w.abun = TRUE, ...))[,1:8] %>%
     cbind(years, .) %>%
@@
   filter(n_distinct(year) > 4) %>% 
+  filter(plot_id != "169A") %>% #soemthing strange going on with this one, a duplicate species obs in 2006
   ungroup() #%>%
@@
          #!study_id %in% c(217, 327, 195)) #exclude to get the file size smaller
+
+plots <- data %>%
+  pull(plot_id) %>%
+  unique()
+
+system.time(fd <- map(plots, get_fd_safe, obs_data = data, m = "min", corr = "cailliez"))
 ```
@@
   paste0(here::here("data", "fd_output"), "/", .) %>%
-  map_dfr(read_tsv)
+  map_dfr(~read_tsv(.x, col_types = "dddddddddcd")) #%>% mutate(plot_id = as.character(plot_id)))
 
@@
 
-#get_fd_error <- safely(get_fd)
-#system.time(fd <- map(plot_sizes$plot_id, get_fd_error, obs_data = data, corr = "cailliez"))
-system.time(fd <- map(plot_sizes$plot_id, get_fd_safe, obs_data = data, corr = "cailliez"))
+get_fd_error <- safely(get_fd)
+system.time(fd <- map(plot_sizes$plot_id, get_fd_error, obs_data = data, m = "min", corr = "cailliez"))
+system.time(fd <- map(plot_sizes$plot_id, get_fd_safe, obs_data = data,  m = "min", corr = "cailliez"))
 ```
@@
               select(plot_id, study_id, plot) %>%
-              mutate(plot_id = as.numeric(plot_id)) %>%
                        distinct())

### 2020-09-11 analysis/ts_null_model.Rmd
Remove old columns

@@
 
-trait_data <- biotime_traits %>% 
-  select(-c(biotimeName, taxa, id_species, plot, year, year_abun, 
-              year_biomass, traitName, scientificName, plot_id, study_id)) %>%
+trait_data <- sample_plots %>% 
+  select(-c(biotimeName, taxa, id_species, plot, year, weight, 
+            traitName, scientificName, plot_id, study_id)) %>%
   distinct()

### 2020-09-11 analysis/ts_null_model.Rmd
Use weight column

@@
 library(FD)
+library(furrr)
 
@@
   plot_data <- data %>% filter(plot_id == filter_plot) %>%
-    select(year, id, year_abun)
+    select(year, id, weight)
 
@@
     samp_species <- sample(species_pool, n_distinct(data$id))
-    samp_abun <- sample(data$year_abun)
+    samp_abun <- sample(data$weight)
 
@@
 
-# sample_occurrence <- map(unique(sample_plots$plot_id), get_plot_sample, data = sample_plots, n = n) %>%
-#   bind_rows() %>%
-#   replace(is.na(.), 0)
-# pins::pin(sample_occurrence, board = "github")
+sample_occurrence <- future_map(unique(sample_plots$plot_id), get_plot_sample, data = sample_plots, n = n) %>%
+  bind_rows() %>%
+  replace(is.na(.), 0)
+pins::pin(sample_occurrence, board = "github")
 sample_occurrence <- pins::pin_get("sample_occurrence", board = "github")

### 2020-09-14 analysis/ts_null_model.Rmd
Save out samples incrementally

@@
   }
-  site_samples %>%
+  plot_samples <- site_samples %>%
     mutate(plot_id = filter_plot) %>%
     pivot_wider(names_from = samp_species, values_from = samp_abun)
+
+  save_dir <- here::here("data", "null_samples")
+  fs::dir_create(save_dir)
+  write.csv(plot_samples, paste0(save_dir, "/", filter_plot, "_samps.csv"))
+
+  return(plot_samples)
 }
@@
   distinct()
+```
 
+Get sample dataframe
+```{r}
+future::plan(multiprocess)
 sample_occurrence <- future_map(unique(sample_plots$plot_id), get_plot_sample, data = sample_plots, n = n) %>%
@@
   replace(is.na(.), 0)
+
+files <-  dir(here::here("data", "null_samples"), "*.csv")
+sample_occurrence <- files %>%
+  paste0(here::here("data", "null_samples"), "/", .) %>%
+  map_dfr(~read_csv(.x) %>% mutate(plot_id = as.character(plot_id)) %>% replace(is.na(.), 0))
+
 pins::pin(sample_occurrence, board = "github")
 sample_occurrence <- pins::pin_get("sample_occurrence", board = "github")
+```
 
+```{r}
 get_plot_FD <- function(traits, occ_df, filter_plot, ...) {

### 2020-09-14 analysis/ts_null_model.Rmd
parallel FD using new settings

--- 2020-09-11 analysis/ts_null_model.Rmd
@@
 get_parallel_FD <- function(group_num){
-  browser()
   group_plots <- plot_sizes %>% filter(group == group_num)
   group_data <- sample_occurrence %>% filter(plot_id %in% group_plots$plot_id)
+  group_trait_data <- trait_data %>% filter(id %in% colnames(group_data)[-(1:3)])
 
@@
   fd <- map_dfr(group_plots$plot_id[1:3], get_plot_FD,
-                   traits = group_trait_data, occ_df = group_data)
+                traits = group_trait_data, occ_df = group_data,
+                m = "min", corr = "cailliez")
     #future_map(group_plots$plot_id, get_fd_safe, data = group_data)

### 2020-09-15 analysis/BAT_explore.Rmd
Calculate Null models by reading in individual plot files.

--- 2020-09-14 analysis/ts_null_model.Rmd
@@
 ---
-title: "null_model"
-author: "Kari Norman"
-date: "6/2/2020"
+title: "BAT"
 output: html_document
@@
 library(tidyverse)
-library(biodivTS)
-library(FD)
-library(furrr)
-
-pins::board_register_github(repo = "karinorman/biodivTS", branch = "master")
-```
-
-Create null models that would be transferable to any timeseries by treating the species pool as all species observed across the timeseries.
-
-Simulation functions
-```{r}
-get_plot_sample <- function(data, filter_plot, n){
-  print(filter_plot)
-  plot_data <- data %>% filter(plot_id == filter_plot) %>%
-    select(year, id, weight)
-
-  species_pool <- plot_data %>%
-    pull(id) %>%
-    na.omit() %>%
-    unique() #get column of unique scientific names
-
-  get_samp_species <- function(data){
-    samp_species <- sample(species_pool, n_distinct(data$id))
-    samp_abun <- sample(data$weight)
-
-    df <- as_tibble(cbind(samp_species, samp_abun)) %>% mutate(year = unique(data$year))
-    return(df)
-  }
-
-  site_samples <- data.frame()
-  for (j in 1:n){
-    iter_samples <- data.frame()
-    for(i in unique(plot_data$year)){
-      year_data <- plot_data %>% filter(year == i)
-      iter_samples <- rbind(iter_samples, get_samp_species(year_data))
-    }
-    site_samples <- rbind(site_samples, iter_samples %>% mutate(n = j))
-  }
-  site_samples %>%
-    mutate(plot_id = filter_plot) %>%
-    pivot_wider(names_from = samp_species, values_from = samp_abun)
-}
-```
-
-Get dataframe of samples for each plot and corresponding species and trait matrices
-```{r}
-biotime_traits <- pins::pin_get("biotime-traits", board = "github")
-
-n = 100
-
-#get the plots that meet our criteria for calculating FD
-sample_plots <- biotime_traits %>%
-  filter(!is.na(plot)) %>% #remove unlabeled plots 
-  group_by(plot_id, year) %>%
-  filter(n_distinct(id) > 4) %>%
-  group_by(plot_id) %>%
-  filter(n_distinct(year) > 4) %>%
-  ungroup()
-
-trait_data <- sample_plots %>% 
-  select(-c(biotimeName, taxa, id_species, plot, year, weight, 
-            traitName, scientificName, plot_id, study_id)) %>%
-  distinct()
-
-sample_occurrence <- future_map(unique(sample_plots$plot_id), get_plot_sample, data = sample_plots, n = n) %>%
-  bind_rows() %>%
-  replace(is.na(.), 0)
-pins::pin(sample_occurrence, board = "github")
-sample_occurrence <- pins::pin_get("sample_occurrence", board = "github")
-
-get_plot_FD <- function(traits, occ_df, filter_plot, ...) {
-  plot_samples <- occ_df %>%
-    filter(plot_id == filter_plot)
-
-  sample_species_mat <- plot_samples %>%
-    select(-c(year, plot_id, n)) %>% #want only species x site info
-    select_if(!colSums(na.omit(.)) %in% c(0, NA)) %>% #remove species that don't occur in the plot
-    select(sort(tidyselect::peek_vars())) #sort columns to be in alphabetical order, required for dbFDI()
-
-  trait_mat <- traits %>%
-    filter(id %in% colnames(sample_species_mat)) %>%
-    select_if(~ length(unique(na.omit(.))) > 1)  #remove columns that have the same value for all columns
-
-  #get binary variables so they can be excluded from rescaling 
-  bin_vars <- map(trait_mat, ~ all(na.omit(.) %in% 0:1))
-
-  trait_mat <- trait_mat %>%
-    mutate_at(vars(-c(id, names(bin_vars[bin_vars == TRUE])), -ends_with("5cat")), 
-              list(~as.numeric(scale(.)))) %>% #rescale variables, not binary or categorical
-    arrange(id) %>%
-    column_to_rownames("id")
-
-  save_dir <- here::here("data", "fd_null_output")
-  fs::dir_create(save_dir)
-
-  #the convexhull c code creates an output file of vertices, in order to not have multiple threads writing
-  #to the same file we have to create individual working directories
-  tmp_dir <- paste0(save_dir, "/tmp", "/tmp_", filter_plot)
-  fs::dir_create(tmp_dir)
-  setwd(tmp_dir)
-
-  print(filter_plot)
-  sample_fd <- as.data.frame(dbFD_joggle(x = trait_mat, a = sample_species_mat, w.abun = TRUE, ...)) %>%
-    cbind(plot_samples %>% select(year, plot_id, n))
-
-  setwd(here::here())
-  write_tsv(sample_fd, paste0(save_dir, "/null", filter_plot, ".csv"))
-
-  # df <- occ_df %>%
-  #   filter(plot_id == filter_plot) %>%
-  #   select(richness, plot_id, n) %>%
-  #   bind_cols(sample_fd) %>%
-  #   mutate(merge_test = case_when(richness == nbsp ~ TRUE,
-  #                                 TRUE ~ FALSE))
-  # 
-  # if (sum(df$merge_test) != dim(df)[1])
-  #   warning("richness and region columns may not have been appropriately joined with FD data")
-
-  return(sample_fd)
-}
-
-null_FD <- map_dfr(unique(sample_plots$plot_id)[1:3], get_plot_FD, 
-                   traits = trait_data, occ_df = sample_occurrence,
-                   m = "min", corr = "cailliez")
-```
-
-Calculate null models in parallel
-```{r}
-#number of chunks to split original dataset into 
-num_groups = 4
-
-plot_sizes <- sample_plots %>%
-  count(plot_id) %>%
-  arrange(desc(n)) %>%
-  mutate(group = rep_len(1:num_groups, dim(.)[1]))
-
-get_parallel_FD <- function(group_num){
-  group_plots <- plot_sizes %>% filter(group == group_num)
-  group_data <- sample_occurrence %>% filter(plot_id %in% group_plots$plot_id)
-  group_trait_data <- trait_data %>% filter(id %in% colnames(group_data)[-(1:3)])
-
-  future::plan(multiprocess)
-  fd <- map_dfr(group_plots$plot_id[1:3], get_plot_FD,
-                traits = group_trait_data, occ_df = group_data,
-                m = "min", corr = "cailliez")
-    #future_map(group_plots$plot_id, get_fd_safe, data = group_data)
-
-  print(group_num)
-  return(fd)
-}
-
-fd <- map_dfr(1:num_groups, get_parallel_FD)
 ```

### 2020-09-15 analysis/ts_null_model.Rmd
Calculate Null models by reading in individual plot files.

--- 2020-09-14 analysis/ts_null_model.Rmd
@@
   distinct()
+
+rm(biotime_traits)
 ```
@@
   replace(is.na(.), 0)
-
-files <-  dir(here::here("data", "null_samples"), "*.csv")
-sample_occurrence <- files %>%
-  paste0(here::here("data", "null_samples"), "/", .) %>%
-  map_dfr(~read_csv(.x) %>% mutate(plot_id = as.character(plot_id)) %>% replace(is.na(.), 0))
-
-pins::pin(sample_occurrence, board = "github")
-sample_occurrence <- pins::pin_get("sample_occurrence", board = "github")
 ```
@@
 ```{r}
-get_plot_FD <- function(traits, occ_df, filter_plot, ...) {
-  plot_samples <- occ_df %>%
-    filter(plot_id == filter_plot)
+get_nullplot_FD <- function(file_path, traits, ...) {
+
+  plot_samples <- read_csv(file_path) %>% mutate(plot_id = as.character(plot_id)) %>% replace(is.na(.), 0)
 
   sample_species_mat <- plot_samples %>%
-    select(-c(year, plot_id, n)) %>% #want only species x site info
-    select_if(!colSums(na.omit(.)) %in% c(0, NA)) %>% #remove species that don't occur in the plot
+    select(-c(X1, year, plot_id, n)) %>% #want only species x site info
     select(sort(tidyselect::peek_vars())) #sort columns to be in alphabetical order, required for dbFDI()
@@
   #to the same file we have to create individual working directories
-  tmp_dir <- paste0(save_dir, "/tmp", "/tmp_", filter_plot)
+  tmp_dir <- paste0(save_dir, "/tmp", "/tmp_", unique(plot_samples$plot_id))
   fs::dir_create(tmp_dir)
@@
 
-  print(filter_plot)
+  print(unique(plot_samples$plot_id))
   sample_fd <- as.data.frame(dbFD_joggle(x = trait_mat, a = sample_species_mat, w.abun = TRUE, ...)) %>%
@@
   setwd(here::here())
-  write_tsv(sample_fd, paste0(save_dir, "/null", filter_plot, ".csv"))
-
-  # df <- occ_df %>%
-  #   filter(plot_id == filter_plot) %>%
-  #   select(richness, plot_id, n) %>%
-  #   bind_cols(sample_fd) %>%
-  #   mutate(merge_test = case_when(richness == nbsp ~ TRUE,
-  #                                 TRUE ~ FALSE))
-  # 
-  # if (sum(df$merge_test) != dim(df)[1])
-  #   warning("richness and region columns may not have been appropriately joined with FD data")
+  write_tsv(sample_fd, paste0(save_dir, "/null", unique(plot_samples$plot_id), ".csv"))
 
@@
 
-null_FD <- map_dfr(unique(sample_plots$plot_id)[1:3], get_plot_FD, 
-                   traits = trait_data, occ_df = sample_occurrence,
+files <-  dir(here::here("data", "null_samples"), "*.csv") %>%
+  paste0(here::here("data", "null_samples"), "/", .)
+
+plan(multiprocess)
+null_FD <- future_map_dfr(files, get_nullplot_FD, traits = trait_data,
                    m = "min", corr = "cailliez")
 ```
-
-Calculate null models in parallel
-```{r}
-#number of chunks to split original dataset into 
-num_groups = 4
-
-plot_sizes <- sample_plots %>%
-  count(plot_id) %>%
-  arrange(desc(n)) %>%
-  mutate(group = rep_len(1:num_groups, dim(.)[1]))
-
-get_parallel_FD <- function(group_num){
-  browser()
-  group_plots <- plot_sizes %>% filter(group == group_num)
-  group_data <- sample_occurrence %>% filter(plot_id %in% group_plots$plot_id)
-
-  future::plan(multiprocess)
-  fd <- map_dfr(group_plots$plot_id[1:3], get_plot_FD, 
-                   traits = group_trait_data, occ_df = group_data)
-    #future_map(group_plots$plot_id, get_fd_safe, data = group_data)
-
-  print(group_num)
-  return(fd)
-}
-
-fd <- map_dfr(1:num_groups, get_parallel_FD)
-```
-
-BAT
-```{r}
-library(BAT)
-
-get_plot_BAT <- function(data, trait_range, occ_df, filter_plot) {
-  sample_species_mat <- occ_df %>%
-    filter(plot_id == filter_plot) %>%
-    select(-c(richness, plot_id, n)) %>% #want only species x site info
-    select_if(!colSums(.) %in% c(0, NA)) %>% #remove species that don't occur in the plot
-    select(sort(tidyselect::peek_vars())) #sort columns to be in alphabetical order, required for dbFDI()
-
-  traits <- colnames(data)[trait_range]
-  sample_trait_mat <-
-    clean_trait_matrix(get_trait_matrix(data = data, colnames(sample_species_mat), traits))
-
-  #get distance matrix
-  dist_traits <-  gowdis(sample_trait_mat)
-  trait.pco <- dudi.pco(quasieuclid(dist_traits), scannf = FALSE, full = TRUE)
-
-}
-
-test <- get_plot_BAT(biotime_traits, 17:64, sample_occurrence, 5617)
-
-ower.mat_spider <- gower.dist(traits)
-euc.pco <- pco(gower.mat_spider,k=3)
-
-barplot(euc.pco$eig)
-(euc.pco$eig[1]+euc.pco$eig[2]+euc.pco$eig[3])/sum(euc.pco$eig[euc.pco$eig > 0]) #0.955 variance explained by first 3 axes
-
-traits <- euc.pco$points
-row.names(traits) <- row.names(traits2)
-```

### 2020-09-15 analysis/ts_null_model.Rmd
read in null samples and process into table

@@
 ```
+
+Create Null table
+```{r}
+path <- here::here("data", "fd_null_output")
+null_files <-  dir(path, "*.csv") %>%
+  paste0(path, "/", .)
+
+process_nulls <- function(file_path){
+  data <- read_tsv(file_path) %>%
+    mutate(plot_id = as.character(plot_id)) %>%
+    select(year, plot_id, richness = nbsp, everything(), -sing.sp, -starts_with("CWM"), -n) %>%
+    group_by(year)
+
+  means <- data %>%
+    select(-plot_id, -richness) %>%
+    summarise(across(everything(), mean)) %>%
+    pivot_longer(cols = -year, names_to = "metric", values_to = "mean")
+
+  sds <- data %>%
+    select(-plot_id, -richness) %>%
+    summarise(across(everything(), sd)) %>%
+    pivot_longer(cols = -year, names_to = "metric", values_to = "sd")
+
+  stats <- left_join(means, sds, by = c("year", "metric")) %>%
+    left_join(data %>% select(year, plot_id, richness) %>% distinct(), by = "year") %>%
+    mutate(se = sd/sqrt(n),
+           lowerCI = mean - qt(0.975, n - 1) * se, 
+           upperCI = mean + qt(0.975, n - 1) * se)  %>%
+    select(year, plot_id, everything())
+
+  return(stats)
+}
+
+plan(multiprocess)
+null_table <- future_map_dfr(null_files, process_nulls)
+
+usethis::use_data(null_table)
+pins::pin(null_table, board = "github")
+```

### 2020-09-15 analysis/calc_fd.Rmd
Add BAT stuff previously in calc_fd.Rmd.

@@
 
-Let's also do the functions from BAT package
-```{r}
-#function for getting PCA axes with at least 95% coverage
-axes95 <- function(mat){
-  axes <- pco(mat)
-  eigen <- axes$eig
-
-  threshold <- sum(eigen[eigen > 0]) * 0.95
-
-  sums <- cumsum(eigen)
-  num_axes <- Position(function(x) x > threshold, sums)
-
-  return(num_axes)
-}
-
-get_BAT_fd <- function(data, filter_plot){
-  require(BAT)
-  require(StatMatch)
-  require(labdsv)
-
-  data <- data %>% filter(plot_id == filter_plot)
-
-  print(max(data$record_id))
-
-  species_mat <- data %>%
-    select(year, id) %>% 
-    distinct() %>%
-    arrange(year, id) %>%
-    mutate(present = 1) %>%
-    spread(id, value = present, fill = 0) %>%
-    column_to_rownames("year")
-
-  trait_mat <- data %>%
-    select(-c(taxa, record_id, day, month, year, sample_desc, id_species, 
-              latitude, longitude, sum_abundance, sum_biomass, biotimeName, 
-              traitName, scientificName)
-    ) %>%
-    distinct() %>%
-    select_if(~ length(unique(.)) > 1) %>% #remove columns that have the same value for all columns
-    mutate_at(vars(-id, -ends_with("5cat")), list(~scales::rescale(.,to=c(0, 1)))) %>% #rescale variables
-    arrange(id) %>%
-    column_to_rownames("id")
-
-  #get PCA trait matrix
-  trait_dist <- gower.dist(trait_mat)
-  trait_pco <- pco(trait_dist, k = axes95(trait_dist))
-
-  traits <- trait_pco$points
-  rownames(traits) <- rownames(trait_mat)
-
-  #getFD metrics
-  kernelFD <- kernel.alpha(comm=as.matrix(species_mat),trait=traits,abund=FALSE,method="gaussian",return.hv=TRUE)
-  hv <- kernelFD[[2]]
-
-  disp <- kernel.dispersion(comm=hv,frac=0.1,func='dissimilarity') 
-  div <- kernel.dispersion(comm=hv,frac=0.6, func='divergence')
-  even <- kernel.evenness(comm=hv)
-
-  results <- data.frame(richness = as.numeric(kernelFD[[1]]), disp, div, even) %>% 
-    mutate(plot_id = filter_plot)
-
-  return(results)
-}
-
-get_BAT_fd(biotime_traits, 376300153)
-```
-
 Format tables

### 2020-09-18 analysis/calc_fd.Rmd
Add null values to fd_table and calculate SES, make sure plot names are being read correctly and years are appropriately labeled to null output.

@@
     select(year, id, weight) %>%
-    arrange(year) %>%
     arrange(id) %>%
-    pivot_wider(names_from = id, values_from = weight, values_fill = 0) %>%
-    select(-year)
+    pivot_wider(names_from = id, values_from = weight, values_fill = 0)
 
-  years <- data %>% 
-    select(year) %>%
-    distinct() %>%
-    arrange(year)
+  years <- select(species_mat, year)
 
@@
 
-  fd_out <- as_tibble(dbFD_joggle(trait_mat, species_mat, w.abun = TRUE, ...))[,1:8] %>%
+  fd_out <- as_tibble(dbFD_joggle(trait_mat, select(species_mat, -year), w.abun = TRUE, ...))[,1:8] %>%
     cbind(years, .) %>%
@@
   future::plan("multiprocess", workers = 4)
-  fd <- future_map(group_plots$plot_id, get_fd_safe, obs_data = group_data) %>%
+  fd <- future_map(group_plots$plot_id, get_fd_safe, obs_data = group_data, 
+                   m = "min", corr = "cailliez") %>%
     rbind()
@@
 
-#fd <- map(1:num_groups, get_parallel_FD, df = data)
+fd <- map(1:num_groups, get_parallel_FD, df = data)
 ```
@@
 
-Format tables
+Format FD table with null values
 ```{r}
+null_table <- pins::pin_get("null-table", board = "github")
+
 fd_table <- fd_data %>% 
@@
               select(plot_id, study_id, plot) %>%
-                       distinct())
+                       distinct()) %>%
+  left_join(null_table %>%
+              select(plot_id, year, richness, metric, mean, sd), 
+            by = c("year", "plot_id", "richness", "metric")) %>%
+  mutate(SES = (value - mean)/sd)
 

### 2020-09-18 analysis/ts_null_model.Rmd
Add null values to fd_table and calculate SES, make sure plot names are being read correctly and years are appropriately labeled to null output.

@@
 
-  plot_samples <- read_csv(file_path) %>% mutate(plot_id = as.character(plot_id)) %>% replace(is.na(.), 0)
+  plot_samples <- read_csv(file_path, col_types = cols(plot_id = "c")) %>% mutate(plot_id = as.character(plot_id)) %>% replace(is.na(.), 0)
 
@@
   setwd(here::here())
-  write_tsv(sample_fd, paste0(save_dir, "/null", unique(plot_samples$plot_id), ".csv"))
+  write_tsv(sample_fd, paste0(save_dir, "/null", unique(plot_samples$plot_id), ".tsv"))
 
@@
 path <- here::here("data", "fd_null_output")
-null_files <-  dir(path, "*.csv") %>%
+null_files <-  dir(path, "*.tsv") %>%
   paste0(path, "/", .)
 
-process_nulls <- function(file_path){
-  data <- read_tsv(file_path) %>%
-    mutate(plot_id = as.character(plot_id)) %>%
+process_nulls <- function(file_path, n){
+  data <- read_tsv(file_path, col_types = cols(plot_id = "c")) %>%
     select(year, plot_id, richness = nbsp, everything(), -sing.sp, -starts_with("CWM"), -n) %>%
@@
 plan(multiprocess)
-null_table <- future_map_dfr(null_files, process_nulls)
+null_table <- future_map_dfr(null_files, process_nulls, n = n)
 

### 2020-09-22 paper/manuscript.Rmd
First commit

---
title: "manuscript"
author: "Kari Norman"
date: "8/24/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###Methods
- matched all species to an ID using taxadb, if more than one species name matched to the same ID they were considered the same species and observations of those species were consolidated into one
- only include time series with at least 4 species and at least 4 years
- drop traits that were the same for all observed species
- all non-categorical or binary traits were z-score scaled to give each trait equal weight (@lepsQuantifyingInterpretingFunctional2006, @schleuterUserGuideFunctional2010)
- raw abundance counts within a year for each plot
- needed euclidian distance matrix - used calliez correction because it was a sufficient correction for all plots even though it led to a marginal decrease in "quality" for the corrected hull
- weight by abundance if available and biomass if not
- null model stuff via @swensonBiogeographyFilteringWoody2012
- for years where all species in the species pool were observed the null model sd is zero for functional richness, since FRic does not incorporate abundance changes. 
- drop species with trait match

### 2020-09-23 analysis/calc_fd.Rmd
Eliminate plots that don't meet criteria at the get go when creating the master dataset (biotime_traits)

@@
 ```{r}
-biotime_traits <- pins::pin_get("biotime-traits", board = "github")
-
 get_fd <- function(obs_data, filter_plot, ...){
@@
 ```{r}
-data <- biotime_traits %>% 
-  filter(!is.na(plot)) %>% #remove unlabeled plots 
-  group_by(plot_id, year) %>%
-  filter(n_distinct(id) > 4) %>%
-  group_by(plot_id) %>%
-  filter(n_distinct(year) > 4) %>% 
-  filter(plot_id != "169A") %>% #soemthing strange going on with this one, a duplicate species obs in 2006
-  ungroup() #%>%
-  #filter(#!study_id %in% c(195, 360), #exclude because they break
-         #!study_id %in% c(217, 327, 195)) #exclude to get the file size smaller
+data <- pins::pin_get("biotime-traits", board = "github")
 

### 2020-09-23 analysis/ts_null_model.Rmd
Eliminate plots that don't meet criteria at the get go when creating the master dataset (biotime_traits)

@@
 
-#get the plots that meet our criteria for calculating FD
-sample_plots <- biotime_traits %>%
-  filter(!is.na(plot)) %>% #remove unlabeled plots 
-  group_by(plot_id, year) %>%
-  filter(n_distinct(id) > 4) %>%
-  group_by(plot_id) %>%
-  filter(n_distinct(year) > 4) %>%
-  ungroup()
-
-trait_data <- sample_plots %>% 
+trait_data <- biotime_traits %>% 
   select(-c(biotimeName, taxa, id_species, plot, year, weight, 
@@
 future::plan(multiprocess)
-sample_occurrence <- future_map(unique(sample_plots$plot_id), get_plot_sample, data = sample_plots, n = n) %>%
+sample_occurrence <- future_map(unique(biotime_traits$plot_id), get_plot_sample, data = biotime_traits, n = n) %>%
   bind_rows() %>%

### 2020-10-12 analysis/ts_null_model.Rmd
write out PCA traits

@@
   print(unique(plot_samples$plot_id))
-  sample_fd <- as.data.frame(dbFD_joggle(x = trait_mat, a = sample_species_mat, w.abun = TRUE, ...)) %>%
+  sample_fd <- as.data.frame(dbFD_mine(x = trait_mat, a = sample_species_mat, w.abun = TRUE, ...)$fd_met) %>%
     cbind(plot_samples %>% select(year, plot_id, n))

### 2020-10-12 analysis/visualization.Rmd

@@
 #load(here::here("data", "fd_table.rda"))
-fd_table <- pins::pin_get("fd_table", board = "github")
+fd_table <- pins::pin_get("fd-table", board = "github")
 
@@
     filter(metric == metric_name) %>%
-    mutate(plot_id = as.factor(plot_id), logvalue = log(value))
+    mutate(plot_id = as.factor(plot_id), logSES = log(SES))
 
-  mod <- lmer(logvalue ~ year + (1|plot_id), data = data)
+  mod <- lmer(logSES ~ year + (1|plot_id), data = data)
   coefs <- summary(mod)$coefficients
@@
     #left_join(metadata %>% select(study_id, climate)) %>%
-    ggplot(aes(x = year, y = logvalue)) +
+    ggplot(aes(x = year, y = logSES)) +
     # geom_smooth(aes(color = climate, group = plot_id), method = "glm", se = FALSE) +

### 2020-10-12 analysis/calc_fd.Rmd

@@
 library(biodivTS)
+
+source(here::here("R", "dbFD_mine.R"))
 
@@
 
-  fd_out <- as_tibble(dbFD_joggle(trait_mat, select(species_mat, -year), w.abun = TRUE, ...))[,1:8] %>%
+  fd_data <- dbFD_mine(trait_mat, select(species_mat, -year), w.abun = TRUE, ...)
+  fd_out <- as_tibble(fd_data$fd_met)[,1:8] %>%
     cbind(years, .) %>%
     mutate(plot_id = unlist(filter_plot), trait_count = dim(trait_mat)[2])
+
     write_tsv(fd_out, paste0(here::here("data"), "/fd_output/", filter_plot, "_fd.tsv"))
+    write_tsv(fd_data$pca_traits, paste0(here::here("data"), "/fd_output/", filter_plot, "_traits.tsv"))
 
@@
 ```{r}
-files <-  dir(here::here("data", "fd_output"), "*.tsv")
+files <-  dir(here::here("data", "fd_output"), "*_fd.tsv")
 

### 2020-10-12 analysis/merge_data.Rmd
Create one weight column that is either abundance or biomass

@@
 biotime_traits <- comp_biotime_traits %>%
-  #get studies with appropriate coverage and drop observations that we don't have trait data for
-  filter(study_id %in% inc_coverage$study_id, !is.na(scientificName)) %>% 
-  mutate(plot_id = paste0(study_id, plot))
+  #get studies with appropriate coverage and drop observations that we don't have trait data for, and unlabeled non-id'd species
+  filter(study_id %in% inc_coverage$study_id, !is.na(scientificName), !is.na(plot)) %>% 
+  mutate(plot_id = paste0(study_id, plot)) %>%
+  group_by(plot_id, year) %>%
+  filter(n_distinct(id) > 4) %>%
+  group_by(plot_id) %>%
+  filter(n_distinct(year) > 4) %>%
+  ungroup()
+```
+
+Create one weight column 
+```{r}
+biotime_traits <- biotime_traits %>%
+  mutate(weight = case_when(
+    year_abun != 0 ~ year_abun,
+    TRUE ~ year_biomass
+  )) %>%
+  select(-c(year_abun, year_biomass))
 ```

### 2020-10-12 paper/manuscript.Rmd
Flesh out outline

@@
 ---
-title: "manuscript"
-author: "Kari Norman"
-date: "8/24/2020"
-output: html_document
+title: "Functional trait data tell new story about biodiversity change"
+author: 
+  - name: "Kari E A Norman"
+    affiliation: ucb
+  - name: "Carl Boettiger"
+    affiliation: ucb, ropensci
+address:
+  - code: ucb
+    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
+  - code: ropensci
+    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA" 
+abstract: |
+
+journal: "Ecography"
+date: "`r Sys.Date()`"
+bibliography: refs.bib
+layout: 3p
+header-includes:
+   - \usepackage{lineno}
+   - \linenumbers
+output: 
+  rticles::elsevier_article:
+    includes:
+      in_header: preamble.tex
+editor_options: 
+  markdown: 
+    wrap: 72
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
+### Introduction
 
-###Methods
-- matched all species to an ID using taxadb, if more than one species name matched to the same ID they were considered the same species and observations of those species were consolidated into one
-- only include time series with at least 4 species and at least 4 years
+Ecological communities are experiencing unprecedented change as a result
+of anthropogenic pressures such as climate change, land use change, and
+invasive species.
+
+#### Paragraph 1 - communities are changing, we need to know how and what that means for ecosystem properties
+
+-   communities are changing as a result of a whole bunch of different
+    pressures
+
+-   a few well documented phenomena: accelerating extinction rates, what
+    else?
+
+-   it's essential for conservation and ecology to understand what these
+    changes mean for ecosystem properties like stability, function, etc
+
+#### Paragraph 2 - species diversity is the go to - it's not working
+
+-   traditionally the most common method of capturing change in
+    communities is biodiversity
+
+-   despite other well documented changes, species diversity doesn't
+    seem to capture local community-level changes
+    ([@dornelasAssemblageTimeSeries2014],
+    [@suppSpecieslevelCommunitylevelResponses2014],
+    [@vellendGlobalMetaanalysisReveals2013]), though there is evidence
+    of a significant increase in turnover
+
+-   We need to look beyond species and their abundances to understand
+    the kinds of changes happening in communities
+
+#### Paragraph 3 - functional diversity is the answer!
+
+-   Functional diversity provides a potentially powerful tool for
+    examining biodiversity changes because it captures characteristics
+    beyond species identity.
+
+-   Powerful functional lens ([@mcgillRebuildingCommunityEcology2006])
+
+-   Functional diversity can tells us a story independent of species
+    richness, as huge functional changes can occur while species
+    richness (or even diversity!) stays the same. This especially gives
+    insight into the turnover we know is occurring
+
+-   We examine 100's of mammal, bird, and amphibian timeseries from the
+    Biotime database paired with functional trait data to assess overall
+    trends in functional change and identify where functional diversity
+    tells us a story beyond species richness
+
+### Methods
+
+#### *Data*
+
+Biodiversity timeseries were obtained from the Biotime database, a
+global repository of assemblage timeseries with 386 individual studies
+from multiple taxa.
+
+-   matched all species to an ID using taxadb, if more than one species
+    name matched to the same ID they were considered the same species
+    and observations of those species were consolidated into one - only
+    include time series with at least 4 species and at least 4 years
+
 -   drop traits that were the same for all observed species
-- all non-categorical or binary traits were z-score scaled to give each trait equal weight (@lepsQuantifyingInterpretingFunctional2006, @schleuterUserGuideFunctional2010)
+
+-   all non-categorical or binary traits were z-score scaled to give
+    each trait equal weight
+    ([@lepsQuantifyingInterpretingFunctional2006],
+    [@schleuterUserGuideFunctional2010])
+
 -   raw abundance counts within a year for each plot
-- needed euclidian distance matrix - used calliez correction because it was a sufficient correction for all plots even though it led to a marginal decrease in "quality" for the corrected hull
+
+-   needed euclidian distance matrix - used calliez correction because
+    it was a sufficient correction for all plots even though it led to a
+    marginal decrease in "quality" for the corrected hull
+
 -   weight by abundance if available and biomass if not
-- null model stuff via @swensonBiogeographyFilteringWoody2012
-- for years where all species in the species pool were observed the null model sd is zero for functional richness, since FRic does not incorporate abundance changes. 
-- drop species with trait match
+
+-   null model stuff via [@swensonBiogeographyFilteringWoody2012] - for
+    years where all species in the species pool were observed the null
+    model sd is zero for functional richness, since FRic does not
+    incorporate abundance changes. - drop species with trait match

### 2020-10-15 data-raw/biotime.Rmd
Move biotime database access to data-raw script.

@@
 library(taxadb)
+library(DBI)
+library(dbplyr)
 
@@
 
-Import occurence data from Biotime
+Import occurrence data from Biotime
+
+Access database
 ```{r}
-#rdataretriever::install("biotime", connection = "csv", data_dir = here("data"))
-# 
-# data <- read.csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS")) %>%
-#   select(study_id, taxa) %>%
-#   left_join(read.csv(system.file("extdata", "biotime/biotime_query.csv", package = "biodivTS")), by = "study_id")
+#personal computer
+con <- dbConnect(
+  drv = RMariaDB::MariaDB(), 
+  dbname = "biotime",
+  username = "root",
+  password = "Ka1172353!", 
+  host = "localhost"
+)
 
-data <- pins::pin_get("year-biotime", board = "github")
+#thelio
+con <- DBI::dbConnect(RMariaDB::MariaDB(), 
+                      host="mariadb", 
+                      user="root", 
+                      password="password", 
+                      dbname="biotime")
+
+biotime_sql <- tbl(con, "allrawdata") %>% 
+  collect() %>% 
+  rename_all(.funs = tolower) %>%
+  select(-id_all_raw_data)
+```
+```{r}
+data <- biotime_sql %>% 
+  group_by(id_species, plot, year, study_id) %>% 
+  summarise(year_abun = sum(na.omit(abundance)), 
+            year_biomass = sum(na.omit(biomass))) %>%
+  left_join(tbl(con, "datasets") %>% select(taxa = TAXA, study_id = STUDY_ID) %>% collect()) %>%
+  left_join(tbl(con, "species") %>% select(id_species = ID_SPECIES, genus_species = GENUS_SPECIES) %>% collect()) %>%
+  ungroup()
+
+#should be zero to be appropriately aggregated
+#data %>% group_by(study_id, plot, year, id_species) %>% filter(n()>1) %>% View()
 ```
 
-Many of the studies with only one plot have NA for the plot number, but we want to distinguish between NA's in studies that have multiple plots and unlabled studies.
+Replace NA plots with a plot label (these are studies with no plot subdivision)
 ```{r}
-#find studies that only have one plot labeled NA
-sing_study <- data %>% 
-  select(study_id, plot) %>%
-  distinct() %>%
-  count(study_id) %>%
-  filter(n == 1) %>%
-  pull(study_id)
-
 #replace plot NA's that are really studies with only one plot with a plot label 
@@
   mutate(plot = case_when(
-    study_id %in% sing_study & plot %in% c(NA, "", "#REF!") ~ "A",
+    is.na(plot) ~ "A",
     TRUE ~ plot
@@
 ```{r}
-usethis::use_data(biotime_data)
+pins::pin(biotime_data, board = "github")
 ```

### 2020-10-15 analysis/merge_data.Rmd
Fix trait merge and coverage calculation.

@@
 ```{r}
-load(here::here("data", "biotime_data.rda"))
+biotime_data <- pins::pin_get("biotime-data", board = "github")
+#load(here::here("data", "biotime_data.rda"))
 load(here::here("data", "elton_mamm.rda"))
@@
 load(here::here("data", "amphibio.rda"))
+
+biotime_data <- biotime_data %>%
+  ungroup() %>%
+  mutate(obsID = row_number())
 ```
@@
   left_join(trait_data %>% rename(traitName = sourceName), by = "id",
-            na_matches = "never") #%>%
-   # drop_na(scientificName)
+            na_matches = "never") %>%
+   drop_na(scientificName)
 }
@@
 
-comp_biotime_traits <- map_dfr(traits, merge_data) %>%
-  right_join(biotime_data %>% select(-sourceName, -scientificName)) %>%
-  distinct()
+trait_merge <- map_dfr(traits, merge_data)
+
+comp_biotime_traits <- trait_merge %>%
+  bind_rows(biotime_data %>% 
+              filter(!obsID %in% trait_merge$obsID) %>%
+              select(-scientificName) %>%
+              rename(biotimeName = sourceName))# %>%
+  #distinct()
 
@@
 coverage <- comp_biotime_traits %>%
-  select(study_id, scientificName) %>%
+  select(study_id, id, biotimeName, traitName) %>%
   distinct() %>%
   group_by(study_id) %>%
-  summarise(perc = sum(!is.na(id))/n()) %>%
+  summarise(perc = sum(!is.na(traitName))/n()) %>%
   arrange(desc(perc))
@@
 ```{r}
-study_table <- biotime_data %>%
-  group_by(study_id) %>%
-  mutate(min_year = min(year), max_year = max(year), n_year = n_distinct(year), plots = n_distinct(plot)) %>% 
-  select(study_id, taxa, min_year, max_year, plots) %>%
-  distinct() %>%
+metadata <- pins::pin_get("metadata", board = "github")
+
+study_table <- biotime_traits %>%
+  select(study_id, plot) %>%
+  count(study_id, name = "n_plots") %>% 
+  left_join(metadata %>% select(study_id, realm, climate, habitat, biome_map, taxa, 
+                                organisms, start_year, end_year), by = "study_id") %>%
   left_join(coverage %>% select(study_id, coverage = perc), by = "study_id")

### 2020-10-15 paper/manuscript.Rmd

@@
 global repository of assemblage timeseries with 386 individual studies
-from multiple taxa.
+from multiple taxa. Timeseries were paired with their functional trait
+data based on the predominant taxa in the timeseries. Due to limitations
+in trait data and processing feasability, we included only mammal, bird,
+and amphibian timeseries. Trait data were obtained from Elton Trait
+Database (mammals and birds; [@wilman2014]) and Amphibio (amphibians,
+[@oliveira2017]).
 
--   matched all species to an ID using taxadb, if more than one species
-    name matched to the same ID they were considered the same species
-    and observations of those species were consolidated into one - only
-    include time series with at least 4 species and at least 4 years
+-   something here describing where the data in those databases comes
+    from, that they're species-level means, etc**;**
 
--   drop traits that were the same for all observed species
+-   also more description of how biotime is structured (studies, some
+    with plots, etc)
 
--   all non-categorical or binary traits were z-score scaled to give
-    each trait equal weight
-    ([@lepsQuantifyingInterpretingFunctional2006],
-    [@schleuterUserGuideFunctional2010])
+In order to ensure taxonomic consistency across datasets, biotime
+species were paired with trait data based on their species identifier
+from the Integrated Taxonomic Information System database (citation),
+obtained through the taxadb R package
+([@normanTaxadbHighperformanceLocal2020]). If more than one species in
+the assemblage data resolved to the same identifier, observations were
+considered the same species. For trait data, traits for all species of
+the same identifier were averaged. Only studies with at least 85% trait
+coverage were included. In order to have a sufficient number of
+observation years to observe a trend and a sufficient number of species
+to calculate functional diversity metrics, studies with fewer than 4
+years and years with fewer than 4 species observed were also excluded.
+
+Functional diversity metrics were calculated for total yearly abundance.
+For many studies there were a different number of samples within each
+year. To adjust for this inconsistency in sampling effort we used
+sample-based rarefaction by bootstrap resampling within year for each
+timeseries based on the smallest number of samples in a year for that
+timeseries.
 
 -   raw abundance counts within a year for each plot
+
+#### *Functional diversity metrics*
+
+We calculated yearly metrics of functional diversity and turnover for
+each plot within each study. All available trait data for each study was
+included in the trait matrix with the exception of traits that were the
+same value for all observed species in the study. All continuous traits
+were z-score scaled to give each trait equal weight in the trait space
+([@lepsQuantifyingInterpretingFunctional2006],
+[@schleuterUserGuideFunctional2010]).
+
+We used the function dbFD from the FD package (citation) to calculated
+PCA-scaled trait matrices for each species in a timeseries and
+functional diversity (FD), functional evenness (FEve), functional
+diveregence (FDiv), and functional dispersion (FDis) based on those
+trait matrices. The number of trait axes was limited to the maximum
+number of traits that fulfills the criteria $s >= s^t$, where $s$ is the
+number of species and \$t\$ is the number of traits. This limitation
+allows for a sufficient number of axes to capture the trait space while
+maintaining computational feasibility for metric calculations. Metrics
+incorporated weighting based on either species abundance or biomass
+based on the methods for each study. The PCA-scaled trait matrices were
+then used as input to the BAT package....
 
@@
 
--   weight by abundance if available and biomass if not
-
 -   null model stuff via [@swensonBiogeographyFilteringWoody2012] - for
@@
     incorporate abundance changes. - drop species with trait match
+
+#### *Analysis*

### 2020-10-16 02_rarefy_griddedData_clusterVersion.R
add rarefying and data processing scripts from https://doi.org/10.5281/zenodo.1473860.

@@ -1,0 +1,401 @@
+##======================================================================
+##	23 Feb 2017: code modified to run on iDiv eve cluster
+##	returns all rarefied metrics for 1 resamples; to be combined and means
+##	calculated in separate script
+##======================================================================
+# Primary Code Authors: Shane Blowes and Sarah Supp
+# Email: sablowes@gmail.com, sarah@weecology.org
+##======================================================================
+rm(list=ls())
+
+##	load packages
+library(dplyr)
+library(purrr)
+library(tibble)
+library(tidyr)
+library(lazyeval)
+library(vegan)
+library(betapart)
+
+##==========================================
+##	Get the gridded data locally
+load('data/BioTIME_grid_filtered.Rdata')
+
+##	use the job and task id as a counter for resampling and to set the seed
+##	of the random number generator
+uniq_id <- Sys.getenv('UNIQ_ID')
+seed <- uniq_id
+#seed <- substr(seed, 1, 4)
+set.seed(seed)
+
+#=================================FUNCTION TO RAREFY DATA============================
+
+rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resamples=100, trimsamples=FALSE){
+
+  #	CALCULATE RAREFIED METRICS for each study for all years
+  #	restrict calculations to where there is abundance>0 AND
+  #	following removal of NAs and 0's there are still more than 2 years
+
+  # Check if the data is count abundance: if yes, calculate all rarefied metrics
+  # Check if the data is presence or biomass: if yes, calculate only S and Jaccards
+  # If is.na(ABUNDANCE_TYPE), then should calculate on the Biomass column
+  # grid: input dataset
+  # type: count, presence, or biomass
+  # resamples: the number of bootstrap resampling events desired (default is 100)
+  # trimsamples: TRUE means that years with < 1/2 the average number of samples should be removed to avoid excessive information loss. Default is FALSE.
+  # calculate the number of sampling events per year, and find the minimum
+  # resample the data to rarefy the diversity metrics
+  # output a new dataframe
+
[…]

### 2020-10-16 data-raw/01_Study_to_Grid.R
add rarefying and data processing scripts from https://doi.org/10.5281/zenodo.1473860.

@@ -1,0 +1,622 @@
+##======================================================================
+##	Before rarefaction we classify studies as having samples
+##	from one or more locations in space: Single Location (SL) and
+##	many location (ML). We use this classification, and the mean size of
+##  SL studies to grid all other ML studies to a similar extent, using dggrid.
+##  New cell centres are estimated based on rarefyID (study_cell).
+##  Gridded studies are then further filtered based on coverage, omitting years
+##  below the threshold for abundance data, and entire "rarefyID"s below the
+##  threshold for presence and biomass data.
+
+# Primary Code Authors: Shane Blowes and Sarah Supp
+# Email: sablowes@gmail.com, sarah@weecology.org
+# DISCLAIMER: code is under development
+# 07/03/2017
+
+# Input is the BioTIME database and metadata
+
+# Coverage:
+# Abundance-based coverage is calculated as C-hat corrected (Eqn 4a, Chao & Jost, Ecology)
+# Incidence-based coverage is calculated as C-hat corrected, using years as samples
+#   (Eqn Chat_sample(T) from Table 2, Chao et al. 2014)
+# Note that the resolution of coverage (and omission) differs for abundance data vs. presence and biomass data
+
+# Output is a gridded version of the BioTIME database, with a new column for rarefyID (the "studyID" to use henceforth)
+#  that has been filtered for adequate abundance- or incidence-based coverage.
+#   BioTIME_grid_filtered.Rdata (data.frame to be used for rarefaction in rarefy_griddedData_allmetrics.R)
+##======================================================================
+
+rm(list=ls())
+##
+library(tidyverse)
+library(dggridR)
+library(scales)
+library(vegan)
+library(iNEXT)
+
+##	get the data and metadata
+##	Get the raw data locally
+bt <- read.csv(system.file("extdata", "biotime/biotime_query.csv", package = "biodivTS")) %>%
+  rename_with(toupper)
+##	Get the meta data locally
[…]

### 2020-10-17 analysis/merge_data.Rmd
Create biotime table that includes all data that fits trait matching criteria. Create reference trait table.

@@
 
-#this gives a coverage for each trait dataset, so potentially a lot of coverage with one dataset and no coverage on all the rest
-#need to get rid of zeros when there is coverage from another dataset so we only have one entry per study
 coverage <- comp_biotime_traits %>%
@@
   #get studies with appropriate coverage and drop observations that we don't have trait data for, and unlabeled non-id'd species
-  filter(study_id %in% inc_coverage$study_id, !is.na(scientificName), !is.na(plot)) %>% 
-  mutate(plot_id = paste0(study_id, plot)) %>%
-  group_by(plot_id, year) %>%
+  filter(study_id %in% inc_coverage$study_id, !is.na(scientificName)) %>% 
+  group_by(rarefyid, year) %>%
   filter(n_distinct(id) > 4) %>%
-  group_by(plot_id) %>%
-  filter(n_distinct(year) > 4) %>%
   ungroup()
-```
-
-Create one weight column 
-```{r}
-biotime_traits <- biotime_traits %>%
-  mutate(weight = case_when(
-    year_abun != 0 ~ year_abun,
-    TRUE ~ year_biomass
-  )) %>%
-  select(-c(year_abun, year_biomass))
 ```
@@
 pins::pin(biotime_traits, board = "github")
+```
+
+Get biotime data filtered by trait coverage
+```{r}
+bt_cols <- colnames(biotime_data)[!colnames(biotime_data) %in% 
+                                         c("sourceName", "obsID", "scientificName")]
+bt_traitfiltered <- biotime_traits %>% 
+  select(all_of(bt_cols)) %>%
+  rename(species = id)
+
+pins::pin(bt_traitfiltered, board = "github")
+```
+
+Create a reference table of trait data
+```{r}
+trait_cols <- c(colnames(amphibio), colnames(elton_bird), colnames(elton_mamm))
+trait_cols <- trait_cols[!trait_cols %in% c("sourceName", "scientificName")]
+
+trait_ref <- biotime_traits %>%
+  select(all_of(trait_cols)) %>%
+  rename(species = id) %>%
+  distinct()
+
+pins::pin(trait_ref, board = "github")
 ```

### 2020-10-17 data-raw/biotime.Rmd
Create biotime table that includes all data that fits trait matching criteria. Create reference trait table.

@@
 
+Load data from outside processing script
+```{r}
+load(here::here('data/BioTIME_grid_filtered.Rdata'))
+data <- bt_grid_filtered %>%
+  rename_with(tolower) %>%
+  mutate(species = str_replace(species, "_", " "))
+
+rm(bt_grid_filtered)
+```
+
 Check which authority would result in the most matches 
@@
 ```{r}
-# #Getting matches by scientific name 
-# biotime <- data %>% 
-#   select(genus_species) %>%
-#   drop_na() %>%
-#   distinct() %>%
-#   mutate(id = get_ids(genus_species, "itis")) %>%
-#   left_join(data, by = "genus_species")
-
 #Getting matches by sci name and common name 
-biotime_ids <- filter_name(unique(data$genus_species), "itis") %>%
-  bind_rows(filter_common(unique(data$genus_species), "itis")) %>%
+biotime_ids <- filter_name(unique(data$species), "itis") %>%
+  bind_rows(filter_common(unique(data$species), "itis")) %>%
   filter(taxonRank == "species") %>% #only want ID's to species, since that's the level of the trait data
@@
     "ITIS:782604",
-    "ITIS:28107",
-    "ITIS:508923",
-    "ITIS:159807", #Heterodontus zebra, all the studies including it are using sci names, so we should match to the sci name match not common name match
-    "ITIS:175125" #The Redhead is coming from bird studies, so choose the bird match, 
-    #the rest can't be match due to too little information
+    "ITIS:159807"
   )) %>%
@@
   distinct() %>%
-  right_join(data, by = c("input" = "genus_species")) %>%
+  right_join(data, by = c("input" = "species")) %>%
   rename(sourceName = input) %>%
-  mutate(scientificName = taxadb::get_names(id))
+  mutate(scientificName = taxadb::get_names(id)) %>%
+  mutate(row_num = row_number())
 ```
@@
 
-biotime_data <- setdiff(biotime_data, dupe_id_entry) %>%
+biotime_data <- biotime_data %>%
+  filter(!row_num %in% dupe_id_entry$row_num) %>% #setdiff(biotime_data, dupe_id_entry) %>%
   bind_rows(dupe_id_entry %>% 
-              select(-sourceName, id_species) %>% 
-              group_by(id, scientificName, plot, year, study_id, taxa) %>% 
-              summarize(year_abun = sum(year_abun), year_biomass = sum(year_biomass)))
+              select(-sourceName) %>% 
+              group_by(across(c(-row_num, -abundance, -biomass))) %>%
+              summarize(abundance = sum(abundance), biomass = sum(biomass))) %>%
+  select(-row_num)
 ```
@@
 # #   distinct() %>%
-# #   left_join(data %>% select(-species, -genus), by = c("input" = "genus_species"), na_matches =  "never") %>%
+# #   left_join(data %>% select(-species, -genus), by = c("input" = "species"), na_matches =  "never") %>%
 # #   rename(sourceName = input) %>%

### 2020-10-17 data-raw/01_Study_to_Grid.R

@@ -36,10 +36,10 @@ library(iNEXT)
 
 ##	get the data and metadata
 ##	Get the raw data locally
-bt <- read.csv(system.file("extdata", "biotime/biotime_query.csv", package = "biodivTS")) %>%
+bt <- read.csv("data/biotime_query.csv") %>%
   rename_with(toupper)
 ##	Get the meta data locally
-meta <- read.csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS")) %>%
+meta <- read.csv("data/biotime_metadata.csv") %>%
   rename_with(toupper)
 
 ##	join abundance records with the metadata

### 2020-10-19 R/dbFD_mine.R
Modify rarefaction function to calculate FD metrics inline.

@@ -1,0 +1,716 @@
+dbFD_mine <- function (x, a, w, w.abun = TRUE, stand.x = TRUE, ord = c("podani",
+                                                          "metric"), asym.bin = NULL, corr = c("sqrt", "cailliez",
+                                                                                               "lingoes", "none"), calc.FRic = TRUE, m = "max", stand.FRic = FALSE,
+          scale.RaoQ = FALSE, calc.FGR = FALSE, clust.type = "ward",
+          km.inf.gr = 2, km.sup.gr = nrow(x) - 1, km.iter = 100, km.crit = c("calinski",
+                                                                             "ssi"), calc.CWM = TRUE, CWM.type = c("dom", "all"),
+          calc.FDiv = TRUE, dist.bin = 2, print.pco = FALSE, messages = TRUE)
+{
+  tol <- .Machine$double.eps
+  corr <- match.arg(corr)
+  ord <- match.arg(ord)
+  CWM.type <- match.arg(CWM.type)
+  km.crit <- match.arg(km.crit)
+  if (!is.logical(messages))
+    stop("'messages' must be TRUE or FALSE.", "\n")
+  if (!is.logical(stand.FRic))
+    stop("'stand.FRic' must be TRUE or FALSE.", "\n")
+  if (!is.logical(stand.x))
+    stop("'stand.x' must be TRUE or FALSE.", "\n")
+  if (!is.logical(w.abun))
+    stop("'w.abun' must be TRUE or FALSE.", "\n")
+  if (!is.logical(calc.FRic))
+    stop("'calc.FRic' must be TRUE or FALSE.", "\n")
+  if (!is.logical(calc.FDiv))
+    stop("'calc.FDiv' must be TRUE or FALSE.", "\n")
+  if (!is.logical(calc.FGR))
+    stop("'calc.FGR' musts be TRUE or FALSE.", "\n")
+  if (!is.logical(calc.CWM))
+    stop("'calc.CWM' must be TRUE or FALSE.", "\n")
+  if (!is.logical(scale.RaoQ))
+    stop("'scale.RaoQ' must be TRUE or FALSE.", "\n")
+  if (!is.logical(print.pco))
+    stop("'print.pco' must be TRUE or FALSE.", "\n")
+  if (is.matrix(x) | is.data.frame(x)) {
+    is.dist.x <- FALSE
+    s.x <- dim(x)[1]
+    t.x <- dim(x)[2]
+    if (is.null(row.names(x)))
[…]

### 2020-10-19 analysis/02_rarefy_griddedData_clusterVersion.R

@@ -19,18 +19,22 @@ library(betapart)
 
 ##==========================================
 ##	Get the gridded data locally
-load('data/BioTIME_grid_filtered.Rdata')
+#load('data/BioTIME_grid_filtered.Rdata')
+bt_grid_filtered <- pins::pin_get("bt-traitfiltered", board = "github") %>%
+  rename_with(toupper) %>%
+  rename(Species = SPECIES, StudyMethod = STUDYMETHOD, ObsEventID = OBSEVENTID,
+         rarefyID = RAREFYID, cell = CELL, Abundance = ABUNDANCE, Biomass = BIOMASS)
 
 ##	use the job and task id as a counter for resampling and to set the seed
 ##	of the random number generator
-uniq_id <- Sys.getenv('UNIQ_ID')
+uniq_id <- runif(1, min = 1, max = 2000)
 seed <- uniq_id
 #seed <- substr(seed, 1, 4)
 set.seed(seed)
 
 #=================================FUNCTION TO RAREFY DATA============================
 
-rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resamples=100, trimsamples=FALSE){
+rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resamples=100, trimsamples=FALSE, ...){
 
   #	CALCULATE RAREFIED METRICS for each study for all years
   #	restrict calculations to where there is abundance>0 AND
@@ -131,7 +135,7 @@ rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resam
     # remove 0's or NA's (to catch any places where abundance wasn't actually recorded or the species is indicated as absent)
     filter_(.dots=zero_NA_filter) %>%
     # depending on type: nest(Species, Abundance) OR nest(Species, Biomass)
-    nest_(key_col="data", nest_cols=c("Species", field)) %>%
+    nest(data=c("Species", field)) %>%
     # reduce to studies that have more than two time points for a given cell
     group_by(rarefyID) %>%
     #keeps all study_cells with 2 or more years of data
@@ -180,21 +184,31 @@ rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resam
 
       # create community matrix of rarefied sample
[…]

### 2020-10-19 R/get_FD.R
Modify rarefaction function to calculate FD metrics inline.

@@ -1,0 +1,15 @@
+get_FD <- function(species_mat, trait_mat, year_list, data_id, samp_id, ...){
+  #the convexhull c code creates an output file of vertices, in order to not have multiple threads writing
+  #to the same file we have to create individual working directories
+  tmp_dir <- paste0(here::here("tmp"), "/tmp_", data_id, "_", samp_id)
+  fs::dir_create(tmp_dir)
+  setwd(tmp_dir)
+
+  fd_data <- dbFD_mine(trait_mat, species_mat, w.abun = TRUE, ...)
+  fd_out <- as_tibble(fd_data$fd_met)[,1:8] %>%
+    cbind(year = year_list, .)
+
+  setwd(here::here())
+
+  return(list(fd_met = fd_out, pca_traits = fd_data$pca_traits))
+}

### 2020-10-19 R/get_traitMat.R

@@ -1,0 +1,15 @@
+get_traitMat <- function(species_list, trait_data){
+
+  trait_mat <- trait_data %>%
+    filter(Species %in% species_list) %>%
+    select_if(~ length(unique(na.omit(.))) > 1)  #remove columns that have the same value for all columns
+
+  #get binary variables so they can be excluded from rescaling
+  bin_vars <- map(trait_mat, ~ all(na.omit(.) %in% 0:1))
+
+  trait_mat <- trait_mat %>%
+    mutate_at(vars(-c(Species, names(bin_vars[bin_vars == TRUE])), -ends_with("5cat")),
+              list(~as.numeric(scale(.)))) %>% #rescale variables, not binary or categorical
+    arrange(Species) %>%
+    column_to_rownames("Species")
+}

### 2020-10-20 analysis/03_collate_rarefied_resamps_median.R
Add functional diversity metrics to calculation and aggregation scripts.

@@ -1,0 +1,102 @@
+##============================================================
+##	script to combine rarefied resamples and calculate mean
+##	for analysis
+library(dplyr)
+library(tibble)
+##============================================================
+##	set the pattern to load the files to be compiled
+path <- here::here("data", "rarefied_metrics")
+filelist <-  dir(path, "*.rda") %>%
+  paste0(path, "/", .)
+nSimul <-  length(filelist)
+
+##	initialise data.frames to store all rarefied resamples
+rarefy_abund_all <- data_frame()
+#rarefy_biomass_all <- data_frame()
+rarefy_pres_all <- data_frame()
+
+for (iSimul in 1:nSimul) {
+	##	load a rarefied sample
+	load(filelist[iSimul])
+	print(iSimul)
+	##	add a column describing measurement type
+	rarefy_abund <- rarefy_abund %>% mutate(measurement_type = 'abundance')
+	#rarefy_biomass <- rarefy_biomass %>% mutate(measurement_type = 'biomass')
+	rarefy_pres <- rarefy_pres %>% mutate(measurement_type = 'presence')
+
+	##	collate rarefied samples
+	rarefy_abund_all <- bind_rows(rarefy_abund_all, rarefy_abund)
+	#rarefy_biomass_all <- bind_rows(rarefy_biomass_all, rarefy_biomass)
+	rarefy_pres_all <- bind_rows(rarefy_pres_all, rarefy_pres)
+}
+
+##	put them all together
+rarefied_metrics <- bind_rows(rarefy_abund_all, #rarefy_biomass_all,
+                              rarefy_pres_all)
+
+##	pull out new metadata
+new_meta <- rarefied_metrics %>%
+	distinct(rarefyID, SamplePool, SampleN, num_years, duration, startYear, endYear)
+
+
+##	calculate the medians for all the metrics
+rarefied_medians <- ungroup(rarefied_metrics) %>%
+  group_by(rarefyID, YEAR, cell) %>%
+  dplyr::summarise(
+    N = median(N),
+    N_int = round(median(N)),
+    S = median(S),	# as above?
+    S_int = round(median(S)),
+    PIE = median(PIE),
+    ENSPIE = median(ENSPIE),
+    ENSPIE_int = round(median(ENSPIE)),
+    pielou = median(pielou),
+    Hill1 = median(Hill1),
+    #MSA = median(MSA),
[…]

### 2020-10-20 R/get_FD.R
Add functional diversity metrics to calculation and aggregation scripts.

@@ -5,7 +5,7 @@ get_FD <- function(species_mat, trait_mat, year_list, data_id, samp_id, ...){
   fs::dir_create(tmp_dir)
   setwd(tmp_dir)
 
-  fd_data <- dbFD_mine(trait_mat, species_mat, w.abun = TRUE, ...)
+  fd_data <- dbFD_mine(trait_mat, species_mat, ...)
   fd_out <- as_tibble(fd_data$fd_met)[,1:8] %>%
     cbind(year = year_list, .)
 

### 2020-10-20 analysis/02_rarefy_griddedData_clusterVersion.R

@@ -16,6 +16,7 @@ library(tidyr)
 library(lazyeval)
 library(vegan)
 library(betapart)
+library(furrr)
 
 ##==========================================
 ##	Get the gridded data locally
@@ -149,7 +150,7 @@ rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resam
   for(i in 1:resamples){
 
     ## loop to do rarefaction for each study
-    for(j in 1:length(unique(bt_grid_nest$rarefyID))){
+    for(j in 1:length(unique(bt_grid_nest$rarefyID)[1:5])){
       print(paste('rarefaction', i, 'out of', resamples, 'for study_cell', j, '(', unique(bt_grid_nest$rarefyID)[j], ')',  'in', length(unique(bt_grid_nest$rarefyID))))
 
       ##	get the jth study_cell
@@ -231,8 +232,9 @@ rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resam
 
         # calculated functional diversity
         FD_mets <- get_FD(species_mat = rare_comm, trait_mat = traits, year_list = years,
-                          data_id = rare_samp$rarefyID, samp_id = uniq_id, m = "min", corr = "cailliez")
-        J_func_components <- funcbeta.pair(x = rare_comm_binary, traits = FD_mets$pca_traits, index.family='jaccard')	# distance
+                          data_id = rare_samp$rarefyID, samp_id = uniq_id,
+                          w.abun = TRUE, m = "min", corr = "cailliez")
+        J_func_components <- functional.beta.pair(x = rare_comm_binary, traits = FD_mets$pca_traits, index.family='jaccard')	# distance
         Jbeta_func <- as.matrix(J_func_components$funct.beta.jac)
         Jtu_func <- as.matrix(J_func_components$funct.beta.jtu)
         Jne_func <- as.matrix(J_func_components$funct.beta.jne)
@@ -332,7 +334,15 @@ rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resam
         Jtu <- as.matrix(J_components$beta.jtu)
         Jne <- as.matrix(J_components$beta.jne)
 
-        # How baseline is calculated.
+        # calculated functional diversity
[…]

### 2020-11-19 analysis/02_rarefy_griddedData_clusterVersion.R
Calculate FD with appropriate columns for all scenarios, read in FD with rarefyID. Calculate medians by dropping NA's.

@@ -375,6 +375,11 @@ bt_grid_pres <- bt_grid_filtered %>%
 rarefy_abund <- rarefy_diversity(grid=bt_grid_abund, trait_data = trait_ref, type="count", trait_axes = "min", parallel = TRUE, resamples=200, core_num = 44)
 rarefy_pres <- rarefy_diversity(grid=bt_grid_pres, trait_data = trait_ref, type="presence", trait_axes = "min", parallel = TRUE, resamples=200, core_num = 20)
 
+loadRData <- function(file_name){
+  load(file_name)
+  get(ls()[ls() != "file_name"])
+}
+
 calc_FD_mets <- function(file_path, traits){
   rare_comm_save <- loadRData(file_path)
 
@@ -466,7 +471,9 @@ calc_FD_mets <- function(file_path, traits){
 
     } else{
       biochange_metrics <- FD_mets$fd_met %>%
-        select(-c(nbsp, sing.sp))
+        select(-c(nbsp, sing.sp)) %>%
+        mutate(cell = unique(rare_comm_save$cell), rarefy_resamp = unique(rare_comm_save$rarefy_resamp)) %>%
+        rename(YEAR = year)
     }
   }else{ biochange_metrics <- tibble() }
 
@@ -475,6 +482,35 @@ calc_FD_mets <- function(file_path, traits){
   save(biochange_metrics, file=paste0(path, "/", unique(rare_comm_save$type),
                                       "_cell", unique(rare_comm_save$rarefyID),
                                       "_sample", unique(rare_comm_save$rarefy_resamp), ".rda"))
+
+}
+
+get_FD_errors <- function(file_path, traits){
+  rare_comm_save <- loadRData(file_path)
+
+  years <- rare_comm_save$YEAR
+
+  species_mat <- rare_comm_save %>%
+    column_to_rownames("YEAR") %>%
+    dplyr::select(-c(rarefyID, cell, type, rarefy_resamp))
+
+  # get trait matrix for the species in the sample
+  traits <- get_traitMat(colnames(species_mat), trait_data = traits)
+
+  # create FD function that returns an empty dataframe instead of erroring if FD can't be calculated
+  get_FD_safe <- safely(get_FD, otherwise = data_frame())
+
+  # calculated functional diversity
+  if(unique(rare_comm_save$type) == "count"){
+    FD_mets <- get_FD_safe(species_mat = species_mat, trait_mat = traits, year_list = years,
[…]

### 2020-11-19 analysis/02_rarefy_griddedData_clusterVersion.R
Calculate FD with appropriate columns for all scenarios, read in FD with rarefyID. Calculate medians by dropping NA's.

@@ -373,6 +373,11 @@ bt_grid_pres <- bt_grid_filtered %>%
 rarefy_abund <- rarefy_diversity(grid=bt_grid_abund, trait_data = trait_ref, type="count", trait_axes = "min", parallel = TRUE, resamples=200, core_num = 44)
 rarefy_pres <- rarefy_diversity(grid=bt_grid_pres, trait_data = trait_ref, type="presence", trait_axes = "min", parallel = TRUE, resamples=200, core_num = 20)
 
+loadRData <- function(file_name){
+  load(file_name)
+  get(ls()[ls() != "file_name"])
+}
+
 calc_FD_mets <- function(file_path, traits){
   rare_comm_save <- loadRData(file_path)
 
@@ -464,7 +469,9 @@ calc_FD_mets <- function(file_path, traits){
 
     } else{
       biochange_metrics <- FD_mets$fd_met %>%
-        select(-c(nbsp, sing.sp))
+        select(-c(nbsp, sing.sp)) %>%
+        mutate(cell = unique(rare_comm_save$cell), rarefy_resamp = unique(rare_comm_save$rarefy_resamp)) %>%
+        rename(YEAR = year)
     }
   }else{ biochange_metrics <- tibble() }
 
@@ -473,6 +480,35 @@ calc_FD_mets <- function(file_path, traits){
   save(biochange_metrics, file=paste0(path, "/", unique(rare_comm_save$type),
                                       "_cell", unique(rare_comm_save$rarefyID),
                                       "_sample", unique(rare_comm_save$rarefy_resamp), ".rda"))
+
+}
+
+get_FD_errors <- function(file_path, traits){
+  rare_comm_save <- loadRData(file_path)
+
+  years <- rare_comm_save$YEAR
+
+  species_mat <- rare_comm_save %>%
+    column_to_rownames("YEAR") %>%
+    dplyr::select(-c(rarefyID, cell, type, rarefy_resamp))
+
+  # get trait matrix for the species in the sample
+  traits <- get_traitMat(colnames(species_mat), trait_data = traits)
+
+  # create FD function that returns an empty dataframe instead of erroring if FD can't be calculated
+  get_FD_safe <- safely(get_FD, otherwise = data_frame())
+
+  # calculated functional diversity
+  if(unique(rare_comm_save$type) == "count"){
+    FD_mets <- get_FD_safe(species_mat = species_mat, trait_mat = traits, year_list = years,
[…]

### 2020-11-19 analysis/03_collate_rarefied_resamps_median.R

@@ -17,7 +17,18 @@ path <- here::here("data", "rarefied_metrics", "fd")
 filelist <-  dir(path, "*.rda") %>%
   paste0(path, "/", .)
 
-fd_metrics <- purrr::map_dfr(filelist, function(x) {load(x)
+fd_metrics <- purrr::map_dfr(filelist, function(x) {
+  load(x)
+  split_path <- str_split(x, "/") %>%
+    unlist(.) %>%
+    last() %>%
+    str_split(., "_") %>%
+    unlist()
+
+  rarefyID <- paste0(str_replace(split_path[2], "cell", ""), "_", split_path[3])
+
+  biochange_metrics <- biochange_metrics %>%
+    mutate(rarefyID = rarefyID)
   return(biochange_metrics)
 })
 
@@ -28,13 +39,14 @@ rarefied_metrics <- full_join(species_metrics, fd_metrics)
 new_meta <- rarefied_metrics %>%
 	distinct(rarefyID, SamplePool, SampleN, num_years, duration, startYear, endYear)
 
-
-##	calculate the medians for all the metrics
+##	calculate the medians for all the metrics for studies that were rarefied
 rarefied_medians <- ungroup(rarefied_metrics) %>%
-  select(-c(SamplePool, SampleN, num_years, duration, startYear, endYear, rarefy_resamp)) %>%
-  group_by(rarefyID, YEAR, cell) %>%
-  dplyr::summarise(across(.cols = everything(), median)) %>%
-  ungroup()
+  filter(rarefied == TRUE) %>%
+  select(-c(SamplePool, SampleN, num_years, duration, startYear, endYear, rarefy_resamp, CWM)) %>%
+  group_by(rarefyID, YEAR, cell, rarefied, type) %>%
+  dplyr::summarise(across(.cols = everything(), ~median(.x, na.rm = TRUE))) %>%
+  ungroup() %>%
+  bind_rows(ungroup(rarefied_metrics) %>% filter(rarefied == FALSE))
 
 rarefied_ints <- ungroup(rarefied_metrics) %>%
   select(-c(SamplePool, SampleN, num_years, duration, startYear, endYear, rarefy_resamp)) %>%
@@ -45,8 +57,8 @@ rarefied_ints <- ungroup(rarefied_metrics) %>%
   ungroup()
 
 ##	recombine with new metadata
-rarefied_medians <- inner_join(new_meta, rarefied_medians, by='rarefyID') %>%
[…]

### 2020-11-19 analysis/03_collate_rarefied_resamps_median.R
Deal with duplicate columns for join.

@@ -40,13 +40,15 @@ new_meta <- rarefied_metrics %>%
 	distinct(rarefyID, SamplePool, SampleN, num_years, duration, startYear, endYear)
 
 ##	calculate the medians for all the metrics for studies that were rarefied
-rarefied_medians <- ungroup(rarefied_metrics) %>%
+metric_cols <- ungroup(rarefied_metrics) %>%
+  select(-c(SamplePool, SampleN, num_years, duration, startYear, endYear, rarefy_resamp, CWM))
+
+rarefied_medians <- metric_cols %>%
   filter(rarefied == TRUE) %>%
-  select(-c(SamplePool, SampleN, num_years, duration, startYear, endYear, rarefy_resamp, CWM)) %>%
   group_by(rarefyID, YEAR, cell, rarefied, type) %>%
   dplyr::summarise(across(.cols = everything(), ~median(.x, na.rm = TRUE))) %>%
   ungroup() %>%
-  bind_rows(ungroup(rarefied_metrics) %>% filter(rarefied == FALSE))
+  bind_rows(metric_cols %>% filter(rarefied == FALSE))
 
 rarefied_ints <- ungroup(rarefied_metrics) %>%
   select(-c(SamplePool, SampleN, num_years, duration, startYear, endYear, rarefy_resamp)) %>%
@@ -62,3 +64,4 @@ rarefied_metrics <- inner_join(new_meta, rarefied_medians, by='rarefyID') %>%
 
 ##	save
 usethis::use_data(rarefied_metrics)
+pins::pin(rarefied_metrics, board = "github")

### 2020-11-19 analysis/04_meta_join.R
first commit

@@ -1,0 +1,116 @@
+##--------------Script to be run after rarefaction: join rarefied_medians with:
+## 1) biome meta data (these are the new Biomes from Faye)
+## 2) old meta data (needed for taxa meta data)
+## 3) do some tidying up: classify data types, remove time series with only 1 year,
+##                        add covariates - centred year, logN, logENSPIE, logS,
+##                        add comparison to first year to dissimilarity metrics, get spatial information
+##                        for individual cells,
+
+##--------------join new biome metadata with the rarefied_medians for modelling-----------------
+
+##	libraries
+library(tidyverse)
+pins::board_register_github(repo = "karinorman/biodivTS", branch = "master")
+
+# rarefied metrics ready for modelling
+rarefied_metrics <- pins::pin_get("rarefied-metrics", board = "github")
+meta <- pins::pin_get("metadata", board = "github")
+
+# load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
+# # new meta data
+# biomes <- read_csv('~/Dropbox/BiogeoBioTIME/rarefIDBiomeEco.csv')
+#
+# ##	filter the new biomes meta data to rarefyIDs that we have for modelling
+# biomes <- biomes %>% filter(rarefyID %in% rarefied_medians$rarefyID)
+#
+# # join
+# rarefied_medians <- inner_join(biomes %>% select(rarefyID, Biome, Ecoregion, X9, eco3, eco4, ECOREGION, Distance, Manual), rarefied_medians, by='rarefyID')
+#
+# # join with 'old' meta data too
+# meta <- read_csv('~/Dropbox/BioTIMELatest/bioTIMEmetadataJune.csv')
+# # meta <- read_csv('/home/sarah/Dropbox/BioTIMELatest/bioTIMEmetadataFeb.csv')
+# #str(meta)
+
+##	merge with 'old' metadata with the rarefied metrics
+rarefied_medians <- rarefied_metrics %>%
+  separate(rarefyID, c("STUDY_ID", "cell2"), "_", remove=FALSE, convert=TRUE) %>%
+  dplyr::select(-cell2)
+rarefied_medians <- inner_join(rarefied_medians, meta, by= c('STUDY_ID' = "study_id"))
+rarefied_medians <- as_tibble(rarefied_medians)
+
+##================================BEFORE MODEL FITTING======================================
[…]

### 2020-11-20 analysis/03_collate_rarefied_resamps_median.R
Cleanup columns in final dataframe, remove samples with only one species.

@@ -33,7 +33,8 @@ fd_metrics <- purrr::map_dfr(filelist, function(x) {
 })
 
 ##	put them all together
-rarefied_metrics <- full_join(species_metrics, fd_metrics)
+rarefied_metrics <- full_join(species_metrics, fd_metrics) %>%
+  filter(S != 1) # remove samples that had one species as they result in strange metric values
 
 ##	pull out new metadata
 new_meta <- rarefied_metrics %>%

### 2020-11-20 analysis/04_meta_join.R
Cleanup columns in final dataframe, remove samples with only one species.

@@ -14,7 +14,8 @@ pins::board_register_github(repo = "karinorman/biodivTS", branch = "master")
 
 # rarefied metrics ready for modelling
 rarefied_metrics <- pins::pin_get("rarefied-metrics", board = "github")
-meta <- pins::pin_get("metadata", board = "github")
+meta <- pins::pin_get("metadata", board = "github") %>%
+  select(study_id, realm, climate, habitat, biome_map, taxa, organisms, cent_lat, cent_long, abundance_type)
 
 # load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
 # # new meta data
@@ -76,10 +77,10 @@ rarefied_medians <- rarefied_medians %>%
     taxa_mod = ifelse((taxa=='Marine invertebrates' | taxa=='Terrestrial invertebrates' | taxa=='Freshwater invertebrates'),
                       'Invertebrates',
                       ifelse((taxa=='Marine plants' | taxa=='Terrestrial plants' | taxa=='Freshwater plants'),
-                             'Plant', taxa)))#,
+                             'Plant', taxa)),
     # simplified climate/latitudinal bands
-    # climate_mod = ifelse(abs(rarefyID_y) > 60, 'Polar',
-    #                      ifelse(abs(rarefyID_y) < 23.5, 'Tropical', 'Temperate')))
+     climate_mod = ifelse(abs(cent_lat) > 60, 'Polar',
+                          ifelse(abs(cent_lat) < 23.5, 'Tropical', 'Temperate')))
 
 ##	add turnover comparisons of first year to itself: similarity==1, dissimiliarity==0
 ##	also, add indicator variable for complete turnover (similarity==0)
@@ -111,6 +112,5 @@ rarefied_medians <- rarefied_medians %>%
     Jtu_01_func = ifelse(Jtu_base_func==Jbeta_base, 1, 0),		# 100% contribution of turnover (species replacement)
     Jne_01_func = ifelse(Jne_base_func==Jbeta_base, 1, 0)) %>%	# 100% contribution of nestedness ()
   ungroup()
-
 
 pins::pin(rarefied_medians, board = "github")

### 2020-11-20 analysis/visualization.Rmd
Modify column names to match new dataframe.

@@
 ```{r}
-load(here::here("data", "biotime_data.rda"))
+metrics <- pins::pin_get("rarefied-medians", board = "github") %>%
+  pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "cYEAR", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
+  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID"))) 
 ```
@@
 plot_metric <- function(metric_name, ylabel){
-  data <- fd_table %>%
-    filter(metric == metric_name) %>%
-    mutate(plot_id = as.factor(plot_id), logSES = log(SES))
+  data <- metrics %>%
+    filter(metric == metric_name, !is.na(value), !is.infinite(value)) %>%
+    mutate(logvalue = log(value))
 
-  mod <- lmer(logSES ~ year + (1|plot_id), data = data)
+  mod <- lmer(logvalue ~ year + (1|rarefyID), data = data)
   coefs <- summary(mod)$coefficients
@@
     #left_join(metadata %>% select(study_id, climate)) %>%
-    ggplot(aes(x = year, y = logSES)) +
-    # geom_smooth(aes(color = climate, group = plot_id), method = "glm", se = FALSE) +
-    # geom_point(aes(color = climate, group = plot_id)) +
-    geom_point(aes(group = plot_id) ,color = "grey") +
-    geom_smooth(aes(group = plot_id), color = "darkgrey", method = "glm", se = FALSE) +
+    ggplot(aes(x = year, y = logvalue)) +
+    # geom_smooth(aes(color = climate, group = rarefyID), method = "glm", se = FALSE) +
+    # geom_point(aes(color = climate, group = rarefyID)) +
+    geom_point(aes(group = rarefyID) ,color = "grey") +
+    geom_smooth(aes(group = rarefyID), color = "darkgrey", method = "glm", se = FALSE) +
     #geom_smooth(color = "black", method = "glm", se = FALSE) +
@@
 get_metric_conint <- function(metric_name){
- data <- fd_table %>%
+ data <- metrics %>%
     filter(metric == metric_name) %>%
-    mutate(plot_id = as.factor(plot_id), logSES = log(SES))
+    mutate(rarefyID = as.factor(rarefyID), logValue = log(value))
 
-  mod <- lmer(logSES ~ year + (1|plot_id), data = data)
+  mod <- lmer(logValue ~ year + (1|rarefyID), data = data)
   coefs <- summary(mod)$coefficients

### 2020-12-01 analysis/bt/code/00_HLM_Jtu.R
First commit

@@ -1,0 +1,50 @@
+##======================================================================
+##  Hierarchical models: fit to either all studies (S, Jaccard), or only those with
+##	count data (N, PIE, etc)
+##======================================================================
+rm(list=ls())
+
+##	
+#withr::with_libpaths(new = '/home/blowes/R/x86_64-pc-linux-gnu-library/3.3/', library(brms))
+library(brms)
+library(dplyr)
+library(tidyr)
+##======================================================================
+##	alternatively load averaged metrics
+load('/gpfs1/data/sChange/BioTime/data/rarefied_medians.Rdata')
+
+##======================================================================		
+##	count cells within realm/climate/taxa hierarchy
+cell_count <- rarefied_medians %>%
+        group_by(Biome, taxa_mod) %>%
+        dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
+        ungroup()
+
+##	rejoin
+rarefied_medians <- left_join(cell_count, rarefied_medians, 
+	by=c('Biome','taxa_mod'))
+
+# need to set first year to zero dissimilarity for consecutive beta metrics
+#rarefied_medians <- rarefied_medians %>%
+#        mutate(Jtu_next = ifelse(YEAR==startYear, 0, Jtu_next))
+
+##	filter to ecoregions with >3 cells
+rarefied_medians <- rarefied_medians %>%
+        filter(n_cells > 3 & BROAD_TYPE == 'count')
+
+##	set some weakly regularising priors...
+hier_prior <- c(set_prior(prior = 'normal(0,1)', class='b', coef='cYEAR'), 	# global slope
+	set_prior(prior = 'normal(0,2)', class='Intercept', coef=''), 		# global intercept
+	set_prior(prior = 'cauchy(0,2)', class='sd'),							# group-level intercepts and slopes
+	set_prior(prior = 'lkj(2)', class='cor'))
+
+Jtu_norm_BTSRfyID <- brm(bf(Jtu_base ~ cYEAR + (cYEAR|Biome/taxa_mod/STUDY_ID/rarefyID), 
+	family = brmsfamily('gaussian')),	
+	data= rarefied_medians,
+	prior=hier_prior,
+	inits = '0',
+	iter = 2000,
+	cores = 4,
+	chains = 4)
+
+save(Jtu_norm_BTSRfyID, file=Sys.getenv('OFILE'))

### 2020-12-01 analysis/bt/code/02c_bt_taxa_to_singleLocations.R

@@ -1,0 +1,355 @@
+# code to get geography for plotting study estimates: get one or more points for biome-taxa-study combinations
+source('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/bt/code/02b_studyLevel_coef_wrangle.R')
+world <- map_data('world')
+
+# function to find the convex hull of a polygon
+find_hull <- function(df) df[chull(df$rarefyID_x, df$rarefyID_y), ]
+# function to find centre of hull
+hull_centre <- function(rarefyID_x, rarefyID_y){
+  hc_x = geosphere::geomean(cbind(x = rarefyID_x, y = rarefyID_y))[1]
+  hc_y = geosphere::geomean(cbind(x = rarefyID_x, y = rarefyID_y))[2]
+  return(cbind.data.frame(hc_x, hc_y))
+}  
+
+##  get unique coords for each biome, taxa combination
+studies_ <- BT_study_estimate %>% 
+  distinct(STUDY_ID)
+
+coords <- rarefied_medians %>% 
+  distinct(REALM, Biome, taxa_mod, STUDY_ID, rarefyID, rarefyID_x, rarefyID_y) %>% 
+  # filter to the studies for which we have study-level slope estimates
+  filter(STUDY_ID %in% studies_$STUDY_ID)
+
+
+##--------marine realm wrangle------------------
+marine_hulls <- plyr::ddply(filter(coords, REALM=='Marine'),
+                     plyr::.(REALM, Biome, taxa_mod, STUDY_ID), find_hull) %>%
+  as_tibble()
+
+
+marine_nest <- marine_hulls %>%
+  group_by(REALM, Biome, taxa_mod, STUDY_ID) %>%
+  filter(n_distinct(rarefyID) > 1) %>%
+  nest(rarefyID_x, rarefyID_y) %>% 
+  mutate(hull_centre = purrr::map(data, ~ hull_centre(.x$rarefyID_x, .x$rarefyID_y)))
+  
+
+mar_hull_cent <- marine_nest %>% unnest(hull_centre)
+
+# we want a point for each unique combination of biome/taxa/study: ok
+mar_hull_cent %>% 
+  group_by(Biome, taxa_mod) %>% 
+  summarise(n.study = n_distinct(STUDY_ID), # 
+            points = n_distinct(hc_x, hc_y)) %>% 
+  filter(points!=n.study)
+
+# and we need points for the studies that only had one location 
[…]

### 2020-12-01 analysis/bt/code/02d_bt_biome_map_wrangle.R

@@ -1,0 +1,240 @@
+# read terrestrial shape file
+
+# code modified 080219: map results from new models (S, Jtu, Jne)
+
+# load new results and raw data
+load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/bt/model_fits_output/BT_model_coef_ranef.Rdata')
+load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
+
+
+# load biomes---------------------
+library(tidyverse)
+world <- map_data('world')
+
+
+library(sf)
+# library(tidyverse) 
+ 
+mar_shp <- sf::st_read('~/Dropbox/BiogeoBioTIME/biomes/marineInUSe.shp', quiet = T) %>% select(-lat, -long)
+terr_shp <- sf::st_read('~/Dropbox/BiogeoBioTIME/biomes/terrInUSe.shp', quiet = T)  %>% select(-lat, -long)
+fresh_shp <- sf::st_read('~/Dropbox/BiogeoBioTIME/biomes/freshinUse.shp', quiet = T)  %>% select(-lat, -long)
+
+# create spatialdataframes (sp package) for use with geom_polygon
+marine_spdf <- as(mar_shp, 'Spatial')
+terr_spdf <- as(terr_shp, 'Spatial')
+fresh_spdf <- as(fresh_shp, 'Spatial')
+
+
+#-------process marine shp file: add slope ranef and flatten for ggplot---------
+mar_biomes <- rarefied_medians %>% 
+  filter(REALM=='Marine') %>%
+  distinct(Biome) %>%
+  mutate(Biome = gsub("_", " ", Biome))
+
+# filter spatialdataframe to these biomes only
+marine_spdf <- marine_spdf[which(marine_spdf$Biome %in% mar_biomes$Biome),]
+
+# put the slope ranef in after fixing names
+marine_biome_re <- bind_cols(BT_biome_ranef %>% 
+                               filter(model=='S_pois_new') %>% 
+                               mutate(S_biome_re = Slope_ranef) %>%
+                               select(S_biome_re, Biome),
+                             BT_biome_ranef %>% 
+                               filter(model=='Jtu_new_norm') %>% 
+                               mutate(Jtu_biome_re = Slope_ranef) %>%
+                               select(Jtu_biome_re),
+                             BT_biome_ranef %>% 
+                               filter(model=='Jne_new_norm') %>%
+                               mutate(Jne_biome_re = Slope_ranef) %>%
[…]

### 2020-12-01 analysis/bt/code/03a_bt_plot_global_estimates.R

@@ -1,0 +1,133 @@
+# update code for new models
+library(brms)
+library(tidyverse)
+
+setwd('~/Dropbox/1current/BioTime/local_code/hierarchical/6results/bt/figs/')
+
+# load model fits to extract posterior samples
+load('~/Desktop/revision_models/Jtu_norm_BTSRfyID_count.Rdata')
+load('~/Desktop/revision_models/S_pois_BTSRfyID_expPrior.Rdata')
+load('~/Desktop/revision_models/Jne_norm_BTSRfyID_count-4950633.Rdata')
+
+# want coefficient plot of the posterior distribution of the global slope
+jne_post <- posterior_samples(Jne_norm_BTSRfyID, pars = 'b_cYEAR', subset = 2000:3000)
+
+jne_summary <- jne_post %>%
+  summarise(
+    median = median(b_cYEAR),
+    lower50 = quantile(b_cYEAR, probs = 0.25),
+    upper50 = quantile(b_cYEAR, probs = 0.75),
+    lower90 = quantile(b_cYEAR, probs = 0.05),
+    upper90 = quantile(b_cYEAR, probs = 0.95))
+
+jtu_post <- posterior_samples(Jtu_norm_BTSRfyID, pars = 'b_cYEAR', subset = 2000:3000)
+
+jtu_summary <- jtu_post %>%
+  summarise(
+    median = median(b_cYEAR),
+    lower50 = quantile(b_cYEAR, probs = 0.25),
+    upper50 = quantile(b_cYEAR, probs = 0.75),
+    lower90 = quantile(b_cYEAR, probs = 0.05),
+    upper90 = quantile(b_cYEAR, probs = 0.95))
+
+s_post <- posterior_samples(S_pois_BTSRfyID_countData, pars = 'b_cYEAR', subset = 2000:3000)
+
+s_summary <- s_post %>%
+  summarise(
+    median = median(b_cYEAR),
+    lower50 = quantile(b_cYEAR, probs = 0.25),
+    upper50 = quantile(b_cYEAR, probs = 0.75),
+    lower90 = quantile(b_cYEAR, probs = 0.05),
+    upper90 = quantile(b_cYEAR, probs = 0.95),
+    lower95 = quantile(b_cYEAR, probs = 0.025),
+    upper95 = quantile(b_cYEAR, probs = 0.975))
+
+# tidy up (the model objects are big)
[…]

### 2020-12-01 analysis/bt/code/03c_bt_alpha_beta_plots.R

@@ -1,0 +1,285 @@
+# code to examine relationships between alpha and beta-diversity change at the study level (new model)
+
+#--Taxa-level plots------
+study_corr <- study_corr %>%
+  mutate(realm2 = ifelse(REALM=='Marine', REALM, 'Terrestrial and freshwater'),
+         realm_label = ifelse(REALM=='A.   Marine', REALM, 'B.   Terrestrial and freshwater'))
+
+# rename group of taxa labelled 'All' to 'Multiple'
+study_corr$taxa_mod2 <- factor(study_corr$taxa_mod, 
+                              levels = c("Fish", "Benthos", "Birds", "Invertebrates", "Plant", "Amphibians",
+                                         "All", "Marine invertebrates/plants", "Mammals"),
+                              labels = c("Fish", "Benthos", "Birds", "Invertebrates", "Plant", "Amphibians",
+                                         "Multiple taxa", "Marine invertebrates/plants", "Mammals"))
+
+# set shapes for plotting groups of taxa
+shapes_mod2 = c('Invertebrates' = 0, 'Fish' = 1, 'Benthos' = 2, 'Birds' = 15, 'Mammals' = 17,
+                'Plant' = 5, 'Multiple taxa' = 6, 'Marine invertebrates/plants' = 7, 'Amphibians' = 8)
+
+# shapes for realm
+shapes_realm <- c('Marine' = 15, 'Terrestrial' = 16, 'Freshwater' = 17)
+
+# colours for each taxa
+taxa_col = c('Multiple taxa' = '#e6f598',
+             'Amphibians' = '#fee08b',
+             'Benthos' = '#5e4fa2',
+             'Birds' = '#f46d43',
+             'Fish' = '#3288bd',
+             'Invertebrates' = '#abdda4',
+             'Mammals' = '#9e0142',
+             'Marine invertebrates/plants' = '#66c2a5',
+             'Plant' = '#fdae61')
+
+# colours for conceptual model relating changes in alpha & beta diversity
+# blue green (colour blind friendly - from colorbrewer2.org) l.blue, d.blue, l.green, d.green
+# quad_col = c('c1' = '#a6cee3', 'c2' = '#1f78b4', 'c3' = '#b2df8a', 'c4' = '#33a02c')
[…]

### 2020-12-01 analysis/rlt/code/03_rlt_posterior_density_plots.R

@@ -1,0 +1,531 @@
+# plot posterior densities of the rct study level estimates (at the realm-climate level)
+rm(list=ls())
+# setwd('~/Desktop/revision_code/')
+# rm(list=ls())
+library(brms) 
+library(tidyverse)
+library(ggridges)
+#------posterior samples from biome and biome-taxa level----------
+load('~/Desktop/revision_models/S_pois_RCTSRfyID_count-4950831.Rdata')
+load('~/Desktop/revision_models/Jtu_norm_RCTSRfyID_count.Rdata')
+
+# first get posterior draws for overall trend
+rct_global_posterior <- tibble(
+  S_global = posterior_samples(S_pois_RCTSRfyID,
+                               pars = 'b_cYEAR',
+                               exact_match = TRUE,
+                               subset = floor(runif(n = 1000,
+                                                    min = 1, max = 2000))) %>% unlist() %>% as.numeric(),
+  Jtu_global = posterior_samples(Jtu_norm_RCTSRfyID,
+                                 pars = 'b_cYEAR',
+                                 exact_match = TRUE,
+                                 subset = floor(runif(n = 1000,
+                                                      min = 1, max = 2000))) %>% unlist() %>% as.numeric())
+
+rct_levels <- S_pois_RCTSRfyID$data %>% 
+  as_tibble() %>% 
+  distinct(rlm_clm_txa) %>% 
+  mutate(level = rlm_clm_txa) %>% 
+  nest(level)
+
+study_levels <- S_pois_RCTSRfyID$data %>% 
+  as_tibble() %>% 
+  distinct(`rlm_clm_txa:STUDY_ID`) %>% 
+  mutate(level = `rlm_clm_txa:STUDY_ID`) %>% 
+  nest(level)
+
+rct_sample_posterior <- rct_levels %>%
+  mutate(S_posteriorSamp = purrr::map(data, ~posterior_samples(S_pois_RCTSRfyID, 
+                                                               pars = paste('r_rlm_clm_txa[', as.character(.x$level), ',cYEAR]', sep=''),
+                                                               exact = TRUE,
+                                                               subset = floor(runif(n = 1000,
+                                                                                    min = 1, max = 2000))) %>% unlist() %>% as.numeric()),
+         Jtu_posterior = purrr::map(data, ~posterior_samples(Jtu_norm_RCTSRfyID,
+                                                             pars = paste('r_rlm_clm_txa[', as.character(.x$level), ',cYEAR]', sep =''),
+                                                             exact = TRUE,
[…]

### 2020-12-01 analysis/rrt/04_model_coef_compare.R

@@ -1,0 +1,730 @@
+# code to compare results of alternative models
+
+# high-level comparison - bt versus rrt: 
+# compare the biome/taxon level to the realm-region-taxon levels
+
+
+# low level comparisons: compare slope estimates for studies and cells that the different models share
+# biome-taxon versus realm-climate-taxon and bt versus realm-region-taxon
+
+rm(list=ls())
+
+library(tidyverse)
+
+# biome-taxa model
+load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/bt/model_fits_output/BT_model_coef_ranef.Rdata')
+# realm-climate-taxa model
+load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/rct/model_fits_output/rlm_clm_txa_modelCoefs.Rdata')
+# realm-region-taxa model
+load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/rrt/model_coef_ranef/rlm_reg_txa_coef_ranef.Rdata')
+
+#--------rarefied_medians.Rdata for meta data not in model object
+load('~/Dropbox/BiogeoBioTIME/rarefied_medians_continents.Rdata')
+rarefied_medians_region <- rarefied_medians
+load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
+
+rarefied_medians <- rarefied_medians %>%
+  group_by(Biome, taxa_mod) %>%
+  mutate(n_cells = n_distinct(rarefyID)) %>%
+  ungroup() %>%
+  filter(n_cells > 3)
+
+realm_climate_biome <- rarefied_medians %>%
+  distinct(REALM, climate_mod, Biome) %>%
+  ungroup()
+
+# want unique climate (latitudinal bands for each biome)
+# this is to compare like with like when comparing bt with rct:
+# both rct estimates and bimoe estimates span multiple latitudinal bands, 
+# here I constrain biomes and to later make comparisons with rct estimate
+# from the same latitude
+realm_climate_biome <- realm_climate_biome %>%
+  # filter(Biome %in% biome_levels_model$Biome) %>%
+  mutate(climate_mod = ifelse(Biome=='Arctic', 'Polar', 
+                              ifelse(Biome=='Boreal_Forests_Taiga', 'Temperate',
+                                     ifelse(Biome=='Cold_Temperate_Northeast_Pacific', 'Temperate',
+                                            ifelse(Biome=='Continental_High_Antarctic', 'Polar',
[…]

### 2020-12-01 analysis/bt/code/00_HLM_Jne.R
First commit

@@ -1,0 +1,50 @@
+##======================================================================
+##  Hierarchical models: fit to either all studies (S, Jaccard), or only those with
+##	count data (N, PIE, etc)
+##======================================================================
+rm(list=ls())
+
+##	
+#withr::with_libpaths(new = '/home/blowes/R/x86_64-pc-linux-gnu-library/3.3/', library(brms))
+library(brms)
+library(dplyr)
+library(tidyr)
+##======================================================================
+##	alternatively load averaged metrics
+load('/gpfs1/data/sChange/BioTime/data/rarefied_medians.Rdata')
+
+##======================================================================		
+##	count cells within realm/climate/taxa hierarchy
+cell_count <- rarefied_medians %>%
+        group_by(Biome, taxa_mod) %>%
+        dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
+        ungroup()
+
+##	rejoin
+rarefied_medians <- left_join(cell_count, rarefied_medians, 
+	by=c('Biome','taxa_mod'))
+
+# need to set first year to zero dissimilarity for consecutive beta metrics
+#rarefied_medians <- rarefied_medians %>%
+#        mutate(Jne_next = ifelse(YEAR==startYear, 0, Jne_next))
+
+##	filter to ecoregions with >3 cells
+rarefied_medians <- rarefied_medians %>%
+        filter(n_cells > 3)
+
+##	set some weakly regularising priors...
+hier_prior <- c(set_prior(prior = 'normal(0,1)', class='b', coef='cYEAR'), 	# global slope
+	set_prior(prior = 'normal(0,2)', class='Intercept', coef=''), 		# global intercept
+	set_prior(prior = 'cauchy(0,2)', class='sd'),							# group-level intercepts and slopes
+	set_prior(prior = 'lkj(2)', class='cor'))
+
+Jne_norm_BTSRfyID <- brm(bf(Jne_base ~ cYEAR + (cYEAR|Biome/taxa_mod/STUDY_ID/rarefyID), 
+	family = brmsfamily('gaussian')),	
+	data= rarefied_medians %>% filter(BROAD_TYPE=='count'),
+	prior=hier_prior,
+	inits = '0',
+	iter = 2000,
+	cores = 4,
+	chains = 4)
+
+save(Jne_norm_BTSRfyID, file=Sys.getenv('OFILE'))

### 2020-12-01 analysis/bt/code/00a_nuts_diagnostics.R

@@ -1,0 +1,92 @@
+# code to check nuts (no u-turn sampler) diagnostics for all models
+# code to run remotely and save results to file
+rm(list=ls())
+
+library(brms)
+library(rstan)
+library(dplyr)
+library(tibble)
+
+# source nuts diagnostics file (from Michael Betancourt)
+source('/gpfs1/data/sChange/BioTime/hierarchical/final_results/stan_utility.R')
+
+# #------------load rlm_clm_txa models------------
+# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/results/Jne_norm_RCTSRfyID-4950830.Rdata')
+# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jne_norm_RCTSRfyID_slice1.Rdata')
+# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jne_norm_RCTSR_slice2.Rdata')
+# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jne_norm_RCTSRfyID_slice3.Rdata')
+# 
+# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/results/Jtu_norm_RCTSRfyID_count-4950828.Rdata')
+# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jtu_norm_RCTSRfyID_slice1-4951060.Rdata')
+# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jtu_norm_RCTSRfyID_slice2.Rdata')
+# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jtu_norm_RCTSRfyID_slice3.Rdata')
+#  
+# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/results/S_pois_RCTRfyID_count-4436001.Rdata')
+# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/S_pois_RCTSR_slice1.Rdata')
+# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/S_pois_RCTSRfyID_slice2-4950841.Rdata')
[…]

### 2020-12-01 analysis/bt/code/03b_bt_biome_map_plot.R

@@ -1,0 +1,323 @@
+# code to plot biome maps
+
+# run code to plot posterior distribution of global estimates for inset
+source('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/bt/code/03a_bt_plot_global_estimates.R')
+# run code to wrangle the data for mapping
+source('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/bt/code/02d_bt_biome_map_wrangle.R')
+
+#----------plot Jtu biome random effects and taxa estimates------------------
+# label for the legend
+Jtu_leg_label <- c(Jtu_re_quantile[1:4], 0, Jtu_re_quantile[5:6]) %>% signif(1) %>% as.numeric() %>% format(scientific = T)
+# alt legend label with the numbers representing biome estimates (not random effects)
+Jtu_estimate_label <- as.numeric(signif(c(Jtu_re_quantile[1:4], 0, Jtu_re_quantile[5:6]) + 
+                                          BT_global_estimates %>% filter(model=='Jtu_new_norm' & time_period=='ALL') %>% 
+                                          filter(term=='Slope') %>% .$Estimate, 2) )
+
+combined_label <- paste0(paste(Jtu_leg_label, Jtu_estimate_label, sep = ' ('),')') 
+
+mar_biome_plot_Jtu <-  ggplot() +
+  geom_polygon(data = filter(alldf, Realm2 == 'Marine'),
+               aes(x = long, y = lat, group = group, linetype = NA, fill = Jtu_re), 
+               alpha = 0.85) +#colour = 'black', 
+  geom_polygon(data=world, aes(long, lat, group = group), colour=NA, fill='#7f7f7f', size=0) +
+  geom_point(data = filter(coords2, REALM == 'Marine'),
+             aes(x = rarefyID_x, y = rarefyID_y, colour = Jtu_re), size = 0.5, stroke = 1.2) +
+  scale_fill_manual(name = '', values = c('1' = '#b2182b', '2' = '#d6604d', '3' = '#f4a582',
+                                          '4' = '#fddbc7', '5' = '#4393c3', '6' = '#2166ac'), 
+                    breaks = c('1', '2', '3', '4', '5', '6'),
+                    labels = c(combined_label[-1]), drop = FALSE) +
+  scale_colour_manual(guide = FALSE, values = c('1' = '#b2182b', '2' = '#d6604d', '3' = '#f4a582',
[…]

### 2020-12-01 analysis/bt/code/06_bt_rct_new_models_coefCompare.R

@@ -1,0 +1,213 @@
+# want to compare the slope estimates for studies and rarefyID's from
+# the bt versus the rct models
+
+rm(list=ls())
+
+library(tidyverse)
+
+load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/bt/model_fits_output/BT_model_coef_ranef.Rdata')
+load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/rct/model_fits_output/rlm_clm_txa_modelCoefs.Rdata')
+
+#--------rarefied_medians.Rdata for meta data not in model object
+load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
+
+rarefied_medians <- rarefied_medians %>%
+  group_by(Biome, taxa_mod) %>%
+  mutate(n_cells = n_distinct(rarefyID)) %>%
+  ungroup() %>%
+  filter(n_cells > 3)
+
+realm_climate_biome <- rarefied_medians %>%
+  distinct(REALM, climate_mod, Biome) %>%
+  ungroup()
+
+# want unique climate (latitudinal bands for each biome)
+# this is to compare like with like: some of the rct study estimates span multiple latitudinal bands
+realm_climate_biome <- realm_climate_biome %>%
+  # filter(Biome %in% biome_levels_model$Biome) %>%
+  mutate(climate_mod = ifelse(Biome=='Arctic', 'Polar', 
+                              ifelse(Biome=='Boreal_Forests_Taiga', 'Temperate',
+                                     ifelse(Biome=='Cold_Temperate_Northeast_Pacific', 'Temperate',
+                                            ifelse(Biome=='Continental_High_Antarctic', 'Polar',
+                                                   ifelse(Biome=='Hawaii', 'Tropical', 
+                                                          ifelse(Biome=='Northern_European_Seas', 'Temperate',
+                                                                 ifelse(Biome=='Subantarctic_Islands', 'Polar',
+                                                                        ifelse(Biome=='Tropical_and_Subtropical_Grasslands_Savannas_and_Shrublands', 'Temperate',
+                                                                               ifelse(Biome=='Tropical_and_Subtropical_Moist_Broadleaf_Forests', 'Tropical', 
[…]

### 2020-12-01 analysis/rlt/code/00a_nuts_diagnostics.R

@@ -1,0 +1,56 @@
+# code to check nuts (no u-turn sampler) diagnostics for all models
+# code to run remotely and save results to file
+rm(list=ls())
+
+library(brms)
+library(rstan)
+library(dplyr)
+library(tibble)
+
+# source nuts diagnostics file (from Michael Betancourt)
+source('/gpfs1/data/sChange/BioTime/hierarchical/final_results/stan_utility.R')
+
+#------------load rlm_clm_txa models------------
+load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/results/Jne_norm_RCTSRfyID-4950830.Rdata')
+load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jne_norm_RCTSRfyID_slice1.Rdata')
+load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jne_norm_RCTSR_slice2.Rdata')
+load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jne_norm_RCTSRfyID_slice3.Rdata')
+
+load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/results/Jtu_norm_RCTSRfyID_count-4950828.Rdata')
+load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jtu_norm_RCTSRfyID_slice1-4951060.Rdata')
+load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jtu_norm_RCTSRfyID_slice2.Rdata')
+load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jtu_norm_RCTSRfyID_slice3.Rdata')
+
+load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/results/S_pois_RCTRfyID_count-4436001.Rdata')
+load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/S_pois_RCTSR_slice1.Rdata')
+load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/S_pois_RCTSRfyID_slice2-4950841.Rdata')
[…]

### 2020-12-01 analysis/rlt/code/05_rlt_sensitivity_plots.R

@@ -1,0 +1,251 @@
+# code modified 120219: new models
+##------data and libraries-----------------------
+rm(list=ls())
+
+library(brms)
+library(tidyverse)
+
+load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/rct/model_fits_output/rlm_clm_txa_modelCoefs.Rdata')
+
+##---------add the number of time series used to estimate each slope-------------
+##----the data models were fit too
+load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
+
+##	create new covariate that is concatenation of realm_climate_taxa
+rarefied_medians <- rarefied_medians %>%
+  unite(rlm_clm_txa, REALM, climate_mod, taxa_mod, remove=FALSE)
+
+## get duration, num_years and other metadata for each rarefyID
+cell_meta <- rarefied_medians %>%
+  distinct(REALM, climate_mod, taxa_mod, rarefyID, rarefyID_x, rarefyID_y, num_years, duration, SamplePool, SampleN, startYear, endYear)
+
+initialS <- rarefied_medians %>% distinct(rarefyID, .keep_all=TRUE) %>%
+  filter(YEAR == startYear) %>%
+  mutate(initialS = S) %>%
+  select(rarefyID, initialS)
+
+cell_meta <- inner_join(cell_meta, initialS, by='rarefyID')
+
+
+## need to split level into rct and rarefyID
+rarefyID_estimate <- bind_rows(
+  # need to do this differently for the new model
+  rlm_clm_txa_rarefyID_estimate %>%
+  filter(model=='Jtu_new_norm' | model=='Jne_new_norm' | model=='S_pois_new') %>% 
+  separate(level, into = c('Realm', 'Climate', 'Taxa', 'STUDY_ID', 'b5' , 'b6'), sep = '_', remove = FALSE) %>%
+  unite(rarefyID, b5, b6))
+
+# check if we have metadata for all the cells in the model
+rarefyID_estimate %>% 
+  filter(!rarefyID %in% cell_meta$rarefyID)
+
+## combine with rarefyID estimates for plotting
+rarefyID_estimate <- right_join(cell_meta, rarefyID_estimate, by = 'rarefyID') %>%
+  mutate(change = ifelse(lower_slope > 0, 'up',
+                         ifelse(upper_slope < 0, 'down', 'neutral'))) %>%
+  unite(RealmClimate, c(Realm,Climate), remove = FALSE) 
+
[…]

### 2020-12-03 analysis/03_collate_rarefied_resamps_median.R
Remove duplicates on import of metrics.

@@ -11,7 +11,7 @@ filelist <-  dir(path, "*.rda") %>%
 
 species_metrics <- purrr::map_dfr(filelist, function(x) {load(x)
   return(rarefied_metrics)
-	})
+	}) %>% distinct()
 
 path <- here::here("data", "rarefied_metrics", "fd")
 filelist <-  dir(path, "*.rda") %>%
@@ -30,7 +30,7 @@ fd_metrics <- purrr::map_dfr(filelist, function(x) {
   biochange_metrics <- biochange_metrics %>%
     mutate(rarefyID = rarefyID)
   return(biochange_metrics)
-})
+}) %>% distinct()
 
 ##	put them all together
 rarefied_metrics <- full_join(species_metrics, fd_metrics) %>%

### 2020-12-03 analysis/model_metrics.Rmd
First commit

---
title: "Models"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(moderndive)
library(lme4)
library(broom.mixed)
```

Get data and put it in long form 
```{r}
metrics <- pins::pin_get("rarefied-medians", board = "github") %>%
  select(-cYEAR) %>%
  pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID"))) 
```
####Data exploration

How many years are missing FD metrics? Only a handful! Let's exclude them so we can get an accurate view of the number of years for which we have data for each timeseries                                                                                                     
```{r}
no_fd_years <- metrics %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis"), is.na(value))

dim(no_fd_years)

metrics <- setdiff(metrics, no_fd_years)
```

Visualize number of years for each timeseries
```{r}
hist_numYears <- metrics %>%
  select(year, rarefyID) %>% 
  distinct() %>% 
  count(rarefyID, name = "num_years") %>% 
  ggplot() +
  geom_histogram(aes(x = num_years),
                 binwidth = 2) +
  scale_x_continuous(name = 'Number of years sampled',
                     breaks = c(2,4,8,16,32,48,64,96)) +
  #scale_y_continuous(breaks = c(0,2500,5000,7500,10000,12500, 15000, 17500, 20000,22500)) +
  labs(tag = 'B',
       y = 'Number of cells') +
  theme_bw() +
  theme(panel.grid.minor.x = element_blank())
```

Function to fit GLM to overall metric trend
```{r}
metric_mean_fit <- metrics %>%
  filter(!is.na(value)) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year + (1|rarefyID), data = .x)))

metric_mean_coef <- metric_mean_fit %>%
  mutate(coef = map(model, tidy)) %>% 
  select(-data, -model) %>% 
  unnest(cols = c(coef))
```

```{r}
indiv_ts_fit <- metrics %>% 
  filter(!is.na(value)) %>%
  group_by_at(setdiff(names(metrics), c("year", "value"))) %>%
  filter(n_distinct(year) > 1) %>% #filter out rarefyID's that don't have multiple years after dropping Na's
  group_nest() %>%
  mutate(model = map(data, ~ lm(value ~ year, data = .x)))

indiv_ts_coef <- indiv_ts_fit %>%
  mutate(coef = map(model, tidy)) %>% 
  select(-data, -model) %>% 
  unnest(cols = c(coef))
```

### 2021-01-14 analysis/model_metrics.Rmd
Update models

@@
 library(tidyverse)
-library(moderndive)
+#library(moderndive)
 library(lme4)
@@
 ```{r}
+pins::board_register_github(repo = "karinorman/biodivTS", branch = "master")
+
 metrics <- pins::pin_get("rarefied-medians", board = "github") %>%
@@
   pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
-  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID"))) 
+  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
+  mutate(logvalue = log(value + abs(floor(min(na.omit(value)))))) # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
 ```
@@
 
-Function to fit GLM to overall metric trend
+Fit GLM to overall metric trend
 ```{r}
+# model for overall trend (random effect w/varying intercept)
+## should be GLS according to Dornelas supplement, but glmer isn't happy
 metric_mean_fit <- metrics %>%
@@
   group_nest(metric) %>%
-  mutate(model = map(data, ~lmer(value ~ year + (1|rarefyID), data = .x)))
+  mutate(model = map(data, ~lmer(logvalue ~ year + (1|study_id) + (1|rarefyID), data = .x)))
 
@@
   mutate(coef = map(model, tidy)) %>% 
+  #mutate(confint = map(model, confint)) %>%
   select(-data, -model) %>% 
   unnest(cols = c(coef))
+
+# model for overall trend (random effect w/varying intercept and slope)
+metric_slope_fit <- metrics %>%
+  filter(!is.na(value)) %>%
+  group_nest(metric) %>%
+  mutate(model = map(data, ~lmer(logvalue ~ year + (1 + year | rarefyID), data = .x)))
+
+metric_slope_coef <- metric_slope_fit %>%
+  mutate(coef = map(model, tidy)) %>% 
+  #mutate(confint = map(model, confint)) %>%
+  select(-data, -model) %>% 
+  unnest(cols = c(coef))
+
+# metric_pred_fit <- metrics %>%
+#   filter(!is.na(value)) %>%
+#   group_nest(metric) %>%
+#   mutate(model = map(data, ~lmer(logvalue ~ year + climate + taxa + (1|study_id/rarefyID), data = .x)))
+# 
+# metric_mean_coef_base <- metric_mean_fit_base %>%
+#   mutate(coef = map(model, tidy)) %>% 
+#   mutate(confint = map(model, confint)) %>%
+#   select(-data, -model) %>% 
+#   unnest(cols = c(coef))
+# 
+# metric_mean_coef <- metric_mean_fit %>%
+#   mutate(coef = map(model, tidy)) %>% 
+#   #mutate(confint = map(model, confint)) %>%
+#   select(-data, -model) %>% 
+#   unnest(cols = c(coef))
 ```

### 2021-01-15 data-raw/elton.Rmd
Clean up data storage

@@
 ```{r}
+#save locally
 usethis::use_data(elton_bird, elton_mamm)
+
+#save to cache
+pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
+pins::pin(elton_bird, board = "github")
+pins::pin(elton_mamm, board = "github")
 ```

### 2021-01-15 analysis/01_filter_by_traits.Rmd
Update file saving to outside cache.

--- 2020-10-17 analysis/merge_data.Rmd
@@
 ```{r}
+pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
+
 biotime_data <- pins::pin_get("biotime-data", board = "github")
-#load(here::here("data", "biotime_data.rda"))
-load(here::here("data", "elton_mamm.rda"))
-load(here::here("data", "elton_bird.rda"))
-load(here::here("data", "amphibio.rda"))
+elton_mamm <- pins::pin_get("elton-mamm", board = "github")
+elton_bird <- pins::pin_get("elton-bird", board = "github")
+amphibio <- pins::pin_get("amphibio", board = "github")
 
@@
 
+#save locally and in cache
+usethis::use_data(bt_traitfiltered)
 pins::pin(bt_traitfiltered, board = "github")
@@
   select(all_of(trait_cols)) %>%
-  rename(species = id) %>%
+  rename(Species = id) %>%
   distinct()
 
+#save locally and in cache
+usethis::use_data(trait_ref)
 pins::pin(trait_ref, board = "github")
@@
 
+#save locally and in cache
 usethis::use_data(study_table)

### 2021-01-15 data-raw/02_biotime_ids.Rmd
Update file saving to outside cache.

--- 2020-10-17 data-raw/biotime.Rmd
@@
 
-Import occurrence data from Biotime
+## Get Biotime Occurrence
+
+### Option for Getting data from SQL query (not grid filtered)
 
@@
 ```{r}
-#personal computer
-con <- dbConnect(
-  drv = RMariaDB::MariaDB(), 
-  dbname = "biotime",
-  username = "root",
-  password = "Ka1172353!", 
-  host = "localhost"
-)
-
-#thelio
-con <- DBI::dbConnect(RMariaDB::MariaDB(), 
-                      host="mariadb", 
-                      user="root", 
-                      password="password", 
-                      dbname="biotime")
-
-biotime_sql <- tbl(con, "allrawdata") %>% 
-  collect() %>% 
-  rename_all(.funs = tolower) %>%
-  select(-id_all_raw_data)
-```
-```{r}
-data <- biotime_sql %>% 
-  group_by(id_species, plot, year, study_id) %>% 
-  summarise(year_abun = sum(na.omit(abundance)), 
-            year_biomass = sum(na.omit(biomass))) %>%
-  left_join(tbl(con, "datasets") %>% select(taxa = TAXA, study_id = STUDY_ID) %>% collect()) %>%
-  left_join(tbl(con, "species") %>% select(id_species = ID_SPECIES, genus_species = GENUS_SPECIES) %>% collect()) %>%
-  ungroup()
-
-#should be zero to be appropriately aggregated
-#data %>% group_by(study_id, plot, year, id_species) %>% filter(n()>1) %>% View()
+# #personal computer
+# con <- dbConnect(
+#   drv = RMariaDB::MariaDB(), 
+#   dbname = "biotime",
+#   username = "root",
+#   password = "Ka1172353!", 
+#   host = "localhost"
+# )
+# 
+# #thelio
+# con <- DBI::dbConnect(RMariaDB::MariaDB(), 
+#                       host="mariadb", 
+#                       user="root", 
+#                       password="password", 
+#                       dbname="biotime")
+# 
+# biotime_sql <- tbl(con, "allrawdata") %>% 
+#   collect() %>% 
+#   rename_all(.funs = tolower) %>%
+#   select(-id_all_raw_data)
+# ```
+# ```{r}
+# data <- biotime_sql %>% 
+#   group_by(id_species, plot, year, study_id) %>% 
+#   summarise(year_abun = sum(na.omit(abundance)), 
+#             year_biomass = sum(na.omit(biomass))) %>%
+#   left_join(tbl(con, "datasets") %>% select(taxa = TAXA, study_id = STUDY_ID) %>% collect()) %>%
+#   left_join(tbl(con, "species") %>% select(id_species = ID_SPECIES, genus_species = GENUS_SPECIES) %>% collect()) %>%
+#   ungroup()
+# 
+# #should be zero to be appropriately aggregated
+# #data %>% group_by(study_id, plot, year, id_species) %>% filter(n()>1) %>% View()
 ```
@@
 ```{r}
-#replace plot NA's that are really studies with only one plot with a plot label 
-data <- data %>% 
-  mutate(plot = case_when(
-    is.na(plot) ~ "A",
-    TRUE ~ plot
-  ))
+# #replace plot NA's that are really studies with only one plot with a plot label 
+# data <- data %>% 
+#   mutate(plot = case_when(
+#     is.na(plot) ~ "A",
+#     TRUE ~ plot
+#   ))
 ```
 
-Load data from outside processing script
+### Or, from the grid processing script
 ```{r}
+#file created by 01_biotime_grid.R
 load(here::here('data/BioTIME_grid_filtered.Rdata'))
@@
 
+### Now get id's
 Check which authority would result in the most matches 
@@
 
-OTT is the best authority but doesn't have common names, so we use the next best ITIS. We'll start with BioTime data:
-
-##Biotime
+OTT is the best authority but doesn't have common names, so we use the next best ITIS
 
@@
 ```{r}
+#save locally
+usethis::use_data(biotime_data)
+
+#save to cache
+pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
 pins::pin(biotime_data, board = "github")

### 2021-01-15 data-raw/01_biotime_grid.R

@@ -26,8 +26,6 @@
 #   BioTIME_grid_filtered.Rdata (data.frame to be used for rarefaction in rarefy_griddedData_allmetrics.R)
 ##======================================================================
 
-rm(list=ls())
-##
 library(tidyverse)
 library(dggridR)
 library(scales)
@@ -37,9 +35,10 @@ library(iNEXT)
 ##	get the data and metadata
 ##	Get the raw data locally
 bt <- read.csv("data/biotime_query.csv") %>%
+  read_csv(system.file("extdata", "biotime/BioTIMEQuery02_04_2018 2.csv", package = "biodivTS")) %>%
   rename_with(toupper)
 ##	Get the meta data locally
-meta <- read.csv("data/biotime_metadata.csv") %>%
+meta <- read.csv("biotime/BioTIMEQuery02_04_2018 2.csv") %>%
   rename_with(toupper)
 
 ##	join abundance records with the metadata
@@ -419,204 +418,210 @@ bt_pabm_filtered <- bt_grid %>%
 #   Note: count data was filtered at rarefyID + YEAR, and presence and biomass data was filtered at rarefyID
 bt_grid_filtered <- rbind(bt_count_filtered, bt_pabm_filtered)
 
+#save locally
+usethis::use_data(dgg)
+usethis::use_data(bt_grid_filtered)
 
-save(dgg, bt_grid_filtered, file='data/BioTIME_grid_filtered.Rdata')
+#save to cache
+pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
+pins::pin(dgg, board = "github")
+pins::pin(bt_grid_filtered, board = "github")
 
-
-##==================================== VISUALIZE DATA ===========================================
-##	want to examine coverage and numbers of samples per cell through time
-##	calculate diversity metrics within cells for each study through time
-
-##	Calculate biodiversity metrics for grid cells (count data only)
-bt_gridMetrics_count <- bt_grid_collate %>%
-	# count data first...
-	filter(ABUNDANCE_TYPE!='Presence/Absence' & ABUNDANCE_TYPE!='<NA>') %>%
[…]

### 2021-01-15 analysis/04_meta_join.R
Modify file saving to cache

@@ -10,29 +10,14 @@
 
 ##	libraries
 library(tidyverse)
-pins::board_register_github(repo = "karinorman/biodivTS", branch = "master")
+pins::board_register_github(repo = "karinorman/biodivTS_data", branch = "master")
 
-# rarefied metrics ready for modelling
+# rarefied metrics ready for modeling
 rarefied_metrics <- pins::pin_get("rarefied-metrics", board = "github")
-meta <- pins::pin_get("metadata", board = "github") %>%
+meta <- pins::pin_get("meta", board = "github") %>%
   select(study_id, realm, climate, habitat, biome_map, taxa, organisms, cent_lat, cent_long, abundance_type)
 
-# load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
-# # new meta data
-# biomes <- read_csv('~/Dropbox/BiogeoBioTIME/rarefIDBiomeEco.csv')
-#
-# ##	filter the new biomes meta data to rarefyIDs that we have for modelling
-# biomes <- biomes %>% filter(rarefyID %in% rarefied_medians$rarefyID)
-#
-# # join
-# rarefied_medians <- inner_join(biomes %>% select(rarefyID, Biome, Ecoregion, X9, eco3, eco4, ECOREGION, Distance, Manual), rarefied_medians, by='rarefyID')
-#
-# # join with 'old' meta data too
-# meta <- read_csv('~/Dropbox/BioTIMELatest/bioTIMEmetadataJune.csv')
-# # meta <- read_csv('/home/sarah/Dropbox/BioTIMELatest/bioTIMEmetadataFeb.csv')
-# #str(meta)
-
-##	merge with 'old' metadata with the rarefied metrics
+##	merge metadata with the rarefied metrics
 rarefied_medians <- rarefied_metrics %>%
   separate(rarefyID, c("STUDY_ID", "cell2"), "_", remove=FALSE, convert=TRUE) %>%
   dplyr::select(-cell2)

### 2021-01-15 data-raw/01_biotime_grid.R

@@ -40,6 +40,9 @@ bt <- read.csv("data/biotime_query.csv") %>%
 ##	Get the meta data locally
 meta <- read.csv("biotime/BioTIMEQuery02_04_2018 2.csv") %>%
   rename_with(toupper)
+# pin to use later
+pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
+pins::pin(meta, board = "github")
 
 ##	join abundance records with the metadata
 bt <- inner_join(meta, bt, by='STUDY_ID')
@@ -423,7 +426,6 @@ usethis::use_data(dgg)
 usethis::use_data(bt_grid_filtered)
 
 #save to cache
-pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
 pins::pin(dgg, board = "github")
 pins::pin(bt_grid_filtered, board = "github")
 

### 2021-01-15 analysis/02_rarefy_griddedData_clusterVersion.R
Modify file saving to cache

@@ -21,10 +21,9 @@ library(FD)
 devtools::load_all()
 
 ##==========================================
-pins::board_register_github(repo = "karinorman/biodivTS", branch = "master")
+pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
 
 ##	Get the gridded data locally
-#load('data/BioTIME_grid_filtered.Rdata')
 bt_grid_filtered <- pins::pin_get("bt-traitfiltered", board = "github") %>%
   rename_with(toupper) %>%
   rename(Species = SPECIES, StudyMethod = STUDYMETHOD, ObsEventID = OBSEVENTID,

### 2021-01-15 analysis/03_collate_rarefied_resamps_median.R

@@ -65,4 +65,6 @@ rarefied_metrics <- inner_join(new_meta, rarefied_medians, by='rarefyID') %>%
 
 ##	save
 usethis::use_data(rarefied_metrics)
+
+pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
 pins::pin(rarefied_metrics, board = "github")

### 2021-01-15 test.R
Don't need that any more

@@ -1,97 +1,0 @@
-
-  # calculating between year similarities (NOT DISTANCE!) with Jaccard, Morisita-Horn, Chao and Pearson correlations
-  Pearsoncor <- cor(t(log(rare_comm+1)), method='pearson')
-  Jacsim <- as.matrix(1-vegdist(rare_comm, method='jaccard', binary=TRUE))
-  Hornsim <- as.matrix(1-vegdist(rare_comm, method='horn'))
-  Chaosim <- as.matrix(1-vegdist(rare_comm, method='chao'))
-  Gainssim <- as.matrix(designdist(rare_comm, method = "B-J", terms = "binary",  abcd = FALSE, alphagamma = FALSE, "gains"))
-  Lossessim <- as.matrix(designdist(rare_comm, method = "A-J", terms = "binary",  abcd = FALSE, alphagamma = FALSE, "losses"))
-  # two steps for Jaccard components (so as calculation is done only once)
-  J_components <- beta.pair(rare_comm_binary, index.family='jaccard')	# distance
-  Jbeta <- as.matrix(J_components$beta.jac)
-  Jtu <- as.matrix(J_components$beta.jtu)
-  Jne <- as.matrix(J_components$beta.jne)
-  n <- length(years)
-
-  # calculated functional diversity
-  FD_mets <- get_FD(species_mat = rare_comm, trait_mat = traits, year_list = years,
-                    data_id = rare_samp$rarefyID, samp_id = uniq_id,
-                    w.abun = TRUE, m = "min", corr = "cailliez")
-  J_func_components <- functional.beta.pair(x = rare_comm_binary, traits = FD_mets$pca_traits, index.family='jaccard')	# distance
-  Jbeta_func <- as.matrix(J_func_components$funct.beta.jac)
-  Jtu_func <- as.matrix(J_func_components$funct.beta.jtu)
-  Jne_func <- as.matrix(J_func_components$funct.beta.jne)
-
-  # How baseline is calculated.
-  # NB: we do not compare first year with itself here (to be added before model fitting later)
-  simbaseline[counter2:(counter2+n-2),] <- cbind(
[…]

### 2021-01-15 analysis/other_Blowes_analysis/07_sim_repeat_sampling.R
Reorganize analysis folder

@@ -1,0 +1,168 @@
+# source code to do turnover calculations
+rm(list=ls())
+
+library(tidyverse)
+source('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/06_rarefysamplesturnoverbnh.R')
+
+set.seed(403)
+
+# simulation parameters
+n_sim <- 1000
+# length of the time series for each simulation
+# time_series_length <- floor(runif(n_sim, min = 2, max = 50))
+# what is the empirical distribution of time series length in our analyses?
+load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
+cell_count <- rarefied_medians %>%
+  group_by(Biome, taxa_mod) %>%
+  dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
+  ungroup() 
+
+##	rejoin
+rarefied_medians <- left_join(cell_count, rarefied_medians, by=c('Biome', 'taxa_mod'))
+
+# get parameters of poisson lognormal for duration
+duration_poilog <- poilog::poilogMLE(rarefied_medians %>% 
+                    group_by(rarefyID) %>%
+                    # distinct(num_years) %>%
+                    distinct(duration) %>% .$duration)
+
+##	filter to count data and biome/taxa combinations with >3 cells
+# these are the data that the model was fit to
+rarefied_medians <- rarefied_medians %>%
+  filter(BROAD_TYPE=='count' & n_cells > 3)
+# duration
+inset <- rarefied_medians %>% 
+  group_by(rarefyID) %>%
+  # distinct(num_years) %>%
+  distinct(duration) %>%
+  # summarise(mean(log(duration)),
+            # sd(log(duration)))
+ ggplot() +
+  geom_density(aes(x = duration, fill = 'empirical')) +
+  # squint (or assume something) and it could be lognormal 
+  geom_density(data = data.frame(duration_sim = rlnorm(1000,2.3,0.65)),#poilog::rpoilog(n_distinct(rarefied_medians %>%
+                                                                             # distinct(duration) %>% 
+                                                                             # .$duration),
+                                                                # mu = duration_poilog$par[1],
+                                                                # sig = duration_poilog$par[2])),#sads::rsad(#S = 50000, 
+                                                           # frac = 1,
[…]

### 2021-01-15 analysis/02_rarefy_griddedData_clusterVersion.R
rebase

@@ -165,7 +165,7 @@ rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resam
 
       ##	get the jth study_cell
       study <- bt_grid_nest %>%
-        filter(rarefyID==unique(bt_grid_nest$rarefyID)[j])
+        filter(rarefyID==filter_id)
 
       # get minimum sample size for rarefaction
       min_samp <- study %>% distinct(min_samp) %>% .$min_samp
@@ -394,11 +394,10 @@ rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resam
     dir.create(dir)
     save(rarefied_metrics, file=paste0(dir, "/resample_", type, i, ".rda"))
 
-    rarefied_metrics
   }	# rarefaction loop
 
-  parallel::stopCluster(cl)
-  return(rarefied_metrics)
+  rarefied_metrics <- as_tibble(rarefied_metrics)
+  rarefied_metrics <- inner_join(new_meta, rarefied_metrics, by = "rarefyID")
 } # END function
 
 ##================================== Calculate mean rarefied diversity for each data type=======================================
@@ -421,7 +420,6 @@ calc_FD_mets <- function(file_path){
   betapart_safe <- possibly(functional.beta.pair, otherwise = data.frame())
 
   # calculated functional diversity
-  if(rare_comm_save$)
   FD_mets <- get_FD_safe(species_mat = rare_comm, trait_mat = traits, year_list = years,
                          data_id = rare_samp$rarefyID, samp_id = i,
                          w.abun = TRUE, m = trait_axes, corr = "cailliez")
@@ -445,4 +443,5 @@ calc_FD_mets <- function(file_path){
   }
 
 }
+
 

### 2021-01-15 analysis/02_rarefy_griddedData_clusterVersion.R
Rebase

@@ -155,13 +155,14 @@ rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resam
   }
 
   ##	rarefy rarefy_resamps times
-  rarefied_metrics <- foreach(i = 1:resamples, .combine=rbind) %dopar% {
-  #rarefied_metrics <- for(i in 1:resamples){
+  #rarefied_metrics <- foreach(i = 1:resamples, .combine=rbind) %dopar% {
+  rarefied_metrics <- for(i in 1:resamples){
 
     ## loop to do rarefaction for each study
     for(j in 1:length(unique(bt_grid_nest$rarefyID))){
     #for(j in 1:3){
-      print(paste('rarefaction', i, 'out of', resamples, 'for study_cell', j, '(', unique(bt_grid_nest$rarefyID)[j], ')',  'in', length(unique(bt_grid_nest$rarefyID))))
+      filter_id <- unique(bt_grid_nest$rarefyID)[j]
+      print(paste('rarefaction', i, 'out of', resamples, 'for study_cell', j, '(', filter_id, ')',  'in', length(unique(bt_grid_nest$rarefyID))))
 
       ##	get the jth study_cell
       study <- bt_grid_nest %>%
@@ -198,6 +199,13 @@ rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resam
         arrange(Species) %>% #order species so the trait matrix can in the same order
         spread(Species, Abundance, fill=0)
 
+      #save out sample species matrix
+      rare_comm_save <- rare_comm %>% mutate(type = type)
+      path <- here::here("data", "rarefied_samples")
+      dir.create(path)
+      save(rare_comm_save, file=paste0(path, "/", type, "_cell", unique(study$rarefyID), "_sample", i, ".rda"))
+
+
       #get vector of years in the same order as the community matrix to tack back on to FD dataframe later
       years <- rare_comm$YEAR
 
@@ -207,9 +215,6 @@ rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resam
 
       # betapart requires presence/absence matrix for Jaccard calculations of turnover/nestedness
       rare_comm_binary <- with(rare_comm, ifelse(rare_comm > 0, 1, 0))
-
-      # get trait matrix for the species in the sample
-      #traits <- get_traitMat(rare_samp$Species, ...)
 
[…]

### 2021-01-15 analysis/02_rarefy_griddedData_clusterVersion.R
Rebase

@@ -76,18 +76,18 @@ rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resam
     # calculate how many observations per year per study
     dplyr::summarise(nsamples = n_distinct(ObsEventID))
 
-    # Calculate the minimum number of samples per cell
-    min_samp <- ungroup(nsamples) %>% group_by(rarefyID) %>%
-      mutate(min_samp = min(nsamples), max_samp = max(nsamples)) %>%
-      #check for studies that have consistent sampling and don't need to be rarefied
-      mutate(rarefy = if_else(min_samp != max_samp, TRUE, FALSE)) %>%
-      # retain only the rows with the minimum sample size for a given cell
-      select(-max_samp, -nsamples, -YEAR) %>%
-      distinct()
+  # Calculate the minimum number of samples per cell
+  min_samp <- ungroup(nsamples) %>% group_by(rarefyID) %>%
+    mutate(min_samp = min(nsamples), max_samp = max(nsamples)) %>%
+    #check for studies that have consistent sampling and don't need to be rarefied
+    mutate(rarefy = if_else(min_samp != max_samp, TRUE, FALSE)) %>%
+    # retain only the rows with the minimum sample size for a given cell
+    select(-max_samp, -nsamples, -YEAR) %>%
+    distinct()
 
-    #	Add the min_samp to the data and tidy a little
-    grid <- inner_join(grid, min_samp)
-    rm(min_samp)
+  #	Add the min_samp to the data and tidy a little
+  grid <- inner_join(grid, min_samp)
+  rm(min_samp)
 
   # Re-calculate metadata
   new_meta <- ungroup(grid) %>%
@@ -106,7 +106,8 @@ rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resam
       # duration of time series, start and end points
       duration = max(YEAR) - min(YEAR) + 1,
       startYear = min(YEAR),
-      endYear = max(YEAR))
+      endYear = max(YEAR),
+      rarefied = unique(rarefy))
 
   #	Create dataframe where unique observations (i.e., the data of an
   #	ObsEventID's [individual species abundances])
@@ -123,42 +124,51 @@ rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resam
     filter(n_distinct(YEAR)>=2) %>%
     ungroup()
 
[…]

### 2021-01-15 analysis/model_metrics.Rmd
Update plots

@@
 ```{r}
-pins::board_register_github(repo = "karinorman/biodivTS", branch = "master")
+pins::board_register_github(repo = "karinorman/biodivTS_data", branch = "master")
 
@@
 # model for overall trend (random effect w/varying intercept and slope)
-metric_slope_fit <- metrics %>%
-  filter(!is.na(value)) %>%
-  group_nest(metric) %>%
-  mutate(model = map(data, ~lmer(logvalue ~ year + (1 + year | rarefyID), data = .x)))
-
-metric_slope_coef <- metric_slope_fit %>%
-  mutate(coef = map(model, tidy)) %>% 
-  #mutate(confint = map(model, confint)) %>%
-  select(-data, -model) %>% 
-  unnest(cols = c(coef))
-
-# metric_pred_fit <- metrics %>%
+# metric_slope_fit <- metrics %>%
 #   filter(!is.na(value)) %>%
 #   group_nest(metric) %>%
-#   mutate(model = map(data, ~lmer(logvalue ~ year + climate + taxa + (1|study_id/rarefyID), data = .x)))
+#   mutate(model = map(data, ~lmer(logvalue ~ year + (1 + year | rarefyID), data = .x)))
 # 
-# metric_mean_coef_base <- metric_mean_fit_base %>%
-#   mutate(coef = map(model, tidy)) %>% 
-#   mutate(confint = map(model, confint)) %>%
-#   select(-data, -model) %>% 
-#   unnest(cols = c(coef))
-# 
-# metric_mean_coef <- metric_mean_fit %>%
+# metric_slope_coef <- metric_slope_fit %>%
 #   mutate(coef = map(model, tidy)) %>% 
@@
 ```
+
+Plot of S and diversity metrics
+```{r}
+indiv_ts_coef %>%
+  filter(term == "year", metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
+  select(metric, estimate) %>% 
+  ggplot() +
+  geom_histogram(aes(x = estimate)) + #,
+                 #binwidth = 2) +
+  scale_x_continuous(name = 'slope') +
+  #scale_y_continuous(breaks = c(0,2500,5000,7500,10000,12500, 15000, 17500, 20000,22500)) +
+  labs(tag = 'B',
+       y = 'Number of cells') +
+  theme_bw() +
+  theme(panel.grid.minor.x = element_blank()) +
+  facet_wrap(vars(metric), scales = "free")
+```

### 2021-01-15 analysis/visualization.Rmd
Data import from pins.

@@
 ```{r}
+pins::board_register_github(repo = "karinorman/biodivTS_data", branch = "master")
+
 metrics <- pins::pin_get("rarefied-medians", board = "github") %>%

### 2021-01-25 analysis/05_model_metrics.Rmd
Get p-values for models, plot dornelas figures

--- 2021-01-15 analysis/model_metrics.Rmd
@@
   filter(!is.na(value)) %>%
-  group_nest(metric) %>%
-  mutate(model = map(data, ~lmer(logvalue ~ year + (1|study_id) + (1|rarefyID), data = .x)))
+  mutate(minyear = min(year), maxyear = max(year)) %>%
+  group_nest(metric, minyear, maxyear) %>%
+  #mutate(model = map(data, ~lmer(logvalue ~ year + (1|study_id) + (1|rarefyID), data = .x)),
+  mutate(model = map(data, ~lmer(logvalue ~ year + (1|rarefyID), data = .x)),
+         prediction = map(model, ~predict(.x), newdata = as.data.frame(year = minyear:maxyear)))
 
@@
   unnest(cols = c(coef))
+
+mean_tests <- metric_mean_fit %>%
+  #filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
+   mutate(p_value = map(model, ~drop1(.x, test = "Chisq") %>% select(`Pr(Chi)`) %>% slice(2) %>% pull())) %>% 
+  select(metric, p_value) %>%
+  left_join(metric_mean_coef %>% 
+              filter(term == "year") %>% 
+              select(metric, estimate))
 
@@
   filter(!is.na(value)) %>%
-  group_by_at(setdiff(names(metrics), c("year", "value"))) %>%
+  group_by_at(setdiff(names(metrics), c("year", "value", "logvalue"))) %>%
   filter(n_distinct(year) > 1) %>% #filter out rarefyID's that don't have multiple years after dropping Na's
   group_nest() %>%
-  mutate(model = map(data, ~ lm(value ~ year, data = .x)))
+  mutate(model = map(data, ~ lm(logvalue ~ year, data = .x)))
 
@@
 ```
+Reproduce Dornelas figure - all data points & regressions w/overall mean
+```{r}
+#get climate classification
+metadata <- read_csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS"))
+
+fric_trend <- metrics %>%
+  filter(metric == "FRic") %>%
+  ggplot(aes(x = year, y = logvalue)) +
+  # geom_smooth(aes(color = climate, group = plot_id), method = "glm", se = FALSE) +
+  # geom_point(aes(color = climate, group = plot_id)) +
+  geom_smooth(aes(group = rarefyID), color = "grey", method = "glm", se = FALSE) +
+  geom_point(aes(group = rarefyID) ,color = "grey") +
+  geom_smooth(color = "black", method = "glm", se = FALSE) +
+  theme(legend.position = "none") 
+ggsave("fric_trends.png", fric_plot)
+```
+
+```{r}
+library(lme4)
+
+plot_metric <- function(metric_name, ylabel){
+
+  data <- metrics %>%
+    filter(metric == metric_name, !is.na(value), !is.infinite(value))
+
+  model_coef <- metric_mean_coef %>%
+    filter(metric == metric_name) %>%
+    select(term, estimate)
+
+  intercept <- model_coef %>% filter(term == "(Intercept)") %>% pull(estimate)
+  slope <- model_coef %>% filter(term == "year") %>% pull(estimate)
+
+  pred <- data.frame(year = min(data$year):max(data$year)) %>%
+    mutate(pred = intercept + year * slope)
+
+  trend_plot <- data %>%
+    left_join(pred, by = "year") %>%
+    #left_join(metadata %>% select(study_id, climate)) %>%
+    ggplot(aes(x = year, y = logvalue)) +
+    # geom_smooth(aes(color = climate, group = rarefyID), method = "glm", se = FALSE) +
+    # geom_point(aes(color = climate, group = rarefyID)) +
+    geom_point(aes(group = rarefyID) ,color = "grey", size = 0.30) +
+    geom_smooth(aes(color = climate_mod, group = rarefyID), method = "glm", se = FALSE, size = 0.5) +
+    #geom_smooth(color = "black", method = "glm", se = FALSE) +
+    geom_line(mapping = aes(y = pred)) +
+    theme(legend.position = "none") +
+    theme_classic() +
+    ylab(ylabel) +
+    labs(color = "Climate")
+
+  ggsave(paste0(metric_name, "_trend.png"), trend_plot)
+  return(trend_plot)
+}
+
+plot_metric("FRic", "log(FD)")
+
+plot_map <- tibble(metric_name = c("S", "FRic", "FEve", "FDiv", "FDis"), 
+           ylabel = c("log(S)", "log(FD)", "log(Evenness)", "log(Divergence)", "log(Dispersion)"))
+pmap(plot_map, plot_metric)
+```
+
+```{r}
+
+```

### 2021-02-01 analysis/05_model_metrics.Rmd
Fit models based on numer of plots within study.

@@
   rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
-  mutate(logvalue = log(value + abs(floor(min(na.omit(value)))))) # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
+  mutate(logvalue = log(value + abs(floor(min(na.omit(value)))))) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
+  filter(!is.na(value))
 ```
@@
 metric_mean_fit <- metrics %>%
-  filter(!is.na(value)) %>%
   mutate(minyear = min(year), maxyear = max(year)) %>%
   group_nest(metric, minyear, maxyear) %>%
-  #mutate(model = map(data, ~lmer(logvalue ~ year + (1|study_id) + (1|rarefyID), data = .x)),
-  mutate(model = map(data, ~lmer(logvalue ~ year + (1|rarefyID), data = .x)),
-         prediction = map(model, ~predict(.x), newdata = as.data.frame(year = minyear:maxyear)))
+  mutate(model = map(data, ~lmer(logvalue ~ year + (1|study_id) + (1|rarefyID), 
+                                 data = .x, control = lmerControl(optimizer ="Nelder_Mead"))))
 
@@
               select(metric, estimate))
+```
 
-# model for overall trend (random effect w/varying intercept and slope)
-# metric_slope_fit <- metrics %>%
-#   filter(!is.na(value)) %>%
-#   group_nest(metric) %>%
-#   mutate(model = map(data, ~lmer(logvalue ~ year + (1 + year | rarefyID), data = .x)))
-# 
-# metric_slope_coef <- metric_slope_fit %>%
-#   mutate(coef = map(model, tidy)) %>% 
-#   #mutate(confint = map(model, confint)) %>%
-#   select(-data, -model) %>% 
-#   unnest(cols = c(coef))
+Fit model to individual studies - if there is more than one timeseries in a study, account for that
+```{r}
+sing_study <- metrics %>%
+  group_by(study_id) %>%
+  filter(n_distinct(rarefyID) == 1) %>% 
+  pull(study_id) %>% unique()
+
+# indiv_data <- metrics %>%
+#   mutate(rarefyID = as.factor(rarefyID)) %>%
+#   group_by_at(setdiff(names(metrics), c("year", "value", "logvalue", "rarefyID", "cell", "samplen"))) %>% 
+#   #filter(n_distinct(year) > 1) %>% #filter out rarefyID's that don't have multiple years after dropping Na's
+#   group_nest()
+
+indiv_data <- metrics %>%
+  mutate(rarefyID = as.factor(rarefyID)) %>%
+  group_by(study_id) %>%
+  group_nest()
+
+indiv_ts_sing_fit <- indiv_data %>%
+  filter(study_id %in% sing_study) %>%
+  mutate(model = map(data, ~ lm(logvalue ~ year, data = .x))) %>%
+  mutate(model_type = "lm")
+
+indiv_ts_fit <- indiv_data %>%
+  filter(!study_id %in% sing_study) %>%
+  mutate(model = map(data, ~lmer(logvalue ~ (1 + year|rarefyID), 
+                                 data = .x, control = lmerControl(optimizer ="Nelder_Mead")))) %>%
+  mutate(model_type = "mixed",
+         p_value = map(model, ~drop1(.x, test = "Chisq") %>% select(`Pr(Chi)`) %>% slice(2) %>% pull()))
+
+indiv_ts_coef <- bind_rows(indiv_ts_fit, indiv_ts_sing_fit) %>%
+  mutate(coef = map(model, tidy)) %>% 
+  select(-data, -model) %>% 
+  unnest(cols = c(coef))
 ```
@@
 ```{r}
-indiv_ts_fit <- metrics %>% 
-  filter(!is.na(value)) %>%
-  group_by_at(setdiff(names(metrics), c("year", "value", "logvalue"))) %>%
-  filter(n_distinct(year) > 1) %>% #filter out rarefyID's that don't have multiple years after dropping Na's
-  group_nest() %>%
-  mutate(model = map(data, ~ lm(logvalue ~ year, data = .x)))
+fit_mod <- function(data){
+  print(unique(data$rarefyID)) %>%
+  lmer(logvalue ~ (1 + year|rarefyID), 
+                                 data = data, control = lmerControl(optimizer ="Nelder_Mead"))
+}
 
-indiv_ts_coef <- indiv_ts_fit %>%
-  mutate(coef = map(model, tidy)) %>% 
-  select(-data, -model) %>% 
-  unnest(cols = c(coef))
+indiv_data %>%
+  filter(!study_id %in% sing_study) %>%
+  mutate(model = map(data, ~lmer(logvalue ~ ( 1+ year|rarefyID), 
+                                 data = .x, control = lmerControl(optimizer ="Nelder_Mead"))))
+
+fit_mod <- function(filter_id){
+  filter_data <- metrics %>% filter(study_id == filter_id)
+
+  print(filter_id)
+  lmer(logvalue ~ (1+ year|rarefyID), 
+                                 data = filter_data)
+}
+
+mult_studies <- metrics %>% filter(!study_id %in% sing_study) %>% pull(study_id) %>% unique()
+map(mult_studies, fit_mod)
+
 ```

### 2021-02-05 analysis/05_model_metrics.Rmd
Fitting with and without random slope

@@
 Fit GLM to overall metric trend
+
+Convergence problems!! See here:
+https://joshua-nugent.github.io/allFit/
+https://rstudio-pubs-static.s3.amazonaws.com/33653_57fc7b8e5d484c909b615d8633c01d51.html
+https://biologyforfun.wordpress.com/2018/04/09/help-i-have-convergence-warnings/
+
+Maybe need something other than a gaussian distribution?
+
+Model checks:
+https://debruine.github.io/posts/normality/
+performance::check_model()
+
 ```{r}
 # model for overall trend (random effect w/varying intercept)
-## should be GLS according to Dornelas supplement, but glmer isn't happy
 metric_mean_fit <- metrics %>%
-  mutate(minyear = min(year), maxyear = max(year)) %>%
-  group_nest(metric, minyear, maxyear) %>%
-  mutate(model = map(data, ~lmer(logvalue ~ year + (1|study_id) + (1|rarefyID), 
-                                 data = .x, control = lmerControl(optimizer ="Nelder_Mead"))))
+  filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
+  group_nest(metric) %>%
+  mutate(model = map(data, ~lmer(logvalue ~ year + (1|study_id/rarefyID), 
+                                 data = .x, control = lmerControl(optimizer ="Nelder_Mead"))),
+           null_model = map(data, ~lmer(logvalue ~ (1|study_id/rarefyID), 
+                                 data = .x, control = lmerControl(optimizer ="Nelder_Mead"))),
+           test = map2(model, null_model, anova),
+           test_res = map(test, tidy),
+           coef = map(model, tidy)
+           ) %>% 
+  unnest(cols = c(test_res)) %>%
+  select(metric, model, coef, p.value) %>%
+  filter(!is.na(p.value)) %>%
+  unnest(cols = c(coef)) %>%
+  filter(term == "year") %>%
+  select(-c(effect, group, term, statistic))
 
-metric_mean_coef <- metric_mean_fit %>%
-  mutate(coef = map(model, tidy)) %>% 
-  #mutate(confint = map(model, confint)) %>%
-  select(-data, -model) %>% 
-  unnest(cols = c(coef))
-
-mean_tests <- metric_mean_fit %>%
-  #filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
-   mutate(p_value = map(model, ~drop1(.x, test = "Chisq") %>% select(`Pr(Chi)`) %>% slice(2) %>% pull())) %>% 
-  select(metric, p_value) %>%
-  left_join(metric_mean_coef %>% 
+metric_mean_fit_slope <- metrics %>%
+  filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
+  group_nest(metric) %>%
+  mutate(model = map(data, ~lmer(logvalue ~ (year|study_id/rarefyID), 
+                                 data = .x, #control = lmerControl(optimizer ="Nelder_Mead"),
+                                 REML = FALSE)),
+           null_model = map(data, ~lmer(logvalue ~ (1|study_id/rarefyID), 
+                                 data = .x, #control = lmerControl(optimizer ="Nelder_Mead"),
+                                 REML = FALSE)),
+           test = map2(model, null_model, anova),
+           test_res = map(test, tidy),
+           coef = map(model, tidy)
+           ) %>% 
+  unnest(cols = c(test_res)) %>%
+  select(metric, model, coef, p.value) %>%
+  filter(!is.na(p.value)) %>%
+  unnest(cols = c(coef)) %>%
   filter(term == "year") %>%
-              select(metric, estimate))
+  select(-c(effect, group, term, statistic))
 ```

### 2021-02-05 paper/manuscript.Rmd
Methods notes

@@
 global repository of assemblage timeseries with 386 individual studies
-from multiple taxa. Timeseries were paired with their functional trait
-data based on the predominant taxa in the timeseries. Due to limitations
-in trait data and processing feasability, we included only mammal, bird,
-and amphibian timeseries. Trait data were obtained from Elton Trait
-Database (mammals and birds; [@wilman2014]) and Amphibio (amphibians,
-[@oliveira2017]).
+from multiple taxa. Following best practices for the dataset
+[@blowes2019] to account for differences in scale between studies, we
+aggregated based on a grid system:
+
+*annotation of* `01_biotime_grid.R`
+
+-   label single location (SL) or multiple location (ML), ML with an
+    extent less than ((SL extent mean + SL extent SD) are assigned to be
+    SL
+
+    -   SL coords are the central lat and long, ML coords are the center
+        of convex hull of locations within a rarefyID (grid cell)
+
+-   exclude studies with one year of data
+
+-   create global grid with resolution based on the cutoff study area
+    size (SL extent mean + SL extent SD)
+
+-   collate observations for the same study within cells/years
+
+-   remove years (for count) or rarefyID's (for abundance and biomass)
+    that do not meet coverage requirement of
+
+Timeseries were paired with their functional trait data based on the
+predominant taxa in the timeseries. Due to limitations in trait data and
+processing feasability, we included only mammal, bird, and amphibian
+timeseries. Trait data were obtained from Elton Trait Database (mammals
+and birds; [@wilman2014]) and Amphibio (amphibians, [@oliveira2017]).
 

### 2021-02-11 paper/manuscript.Rmd
Update intro

@@
 
+Change is a fundamental process shaping ecological communities across
+the globe. Natural background rates of change are also now punctated by
+shifts as a result of anthropogenic pressures such as climate change,
+land use change, and invasive species. One of the most widely accepted
+phenomena characterizing the world's ecological communities is that of
+change.
+
 Ecological communities are experiencing unprecedented change as a result
 of anthropogenic pressures such as climate change, land use change, and
-invasive species.
+invasive species. Impacts of these pressures are well documented in an
+accelerating global extinction rate (citation), shifts in the nature of
+well studied systems (natural history sense, e.g. forests (FIA), coral
+bleaching?), increasing fragmentation etc (how do we know things are
+bad??). Despite all this, a developing body of work indicates that one
+of the most ubiquitous metrics of an local-scale ecosystem, the number
+of species, does not show a clear global decline. How do we reconcile
+documented change in communities with this finding? We assess/make the
+case for/present a powerful alternative approach to understanding
+biodiveristy change through a functional lens by assess functional
+change through time for mammal, bird, and amphibian communities across
+the globe.
+
+The number of species (species richness) is the defualt approach for
+measuring community structure, especially in studies comparing trends
+across many assemblages of different types.
 

### 2021-02-11 paper/manuscript.Rmd
Flesh out intro

--- 2021-02-05 paper/manuscript.Rmd
@@
     includes:
-      in_header: preamble.tex
+     # in_header: preamble.tex
 editor_options: 
@@
 of anthropogenic pressures such as climate change, land use change, and
-invasive species.
+invasive species. Impacts of these pressures are well documented on a
+global scale with an accelerating global extinction rate (citation) and
+fundamental changes in some of the most well-known systems (e.g. coral
+bleaching. At the local scale however the most common metric of
+biodiversity, species richness, tells a different story. Recent
+large-scale assessments of trends in biodiversity over time have shown a
+decoupling of community composition from richness [@brose2016], evidence
+of community-level regulation of richness [@gotelli2017], and no net
+change in local biodiversity despite widespread environmental change
+[@brown2001; @dornelas2014; @vellend2013;, @vellend2017] despite ongoing
+species turnover. While communities are clearly changing, our most
+common species-based approaches do not fully capture the nature of that
+change.
+
+Functional diversity offers a potentially powerful alternative for
+detecting and describing community change by providing a mechanistic
+link between the organisms in a system and the processes they perform
+[@mcgill2006]. By summarizing functional traits rather than species,
+functional diversity metrics capture the disproportionate impact of
+losses or gains of functionally unique species. Functional diversity
+metrics are particularly suited for understanding local-scale
+biodiversity patterns as they can detect shifts in community structure
+independent of species change. The rich history of connecting functional
+diversity to ecosystem function and resilience and community assembly
+provide further context in which to interpret changes in functional
+diversity.
+<!-- turn that last sentence into its own paragraph? e.g. what functional diversity tells us about community structure and function -->
+
+Assessments of broad-scale temporal change in functional diversity have
+previously been limited by a lack of functional trait data. The majority
+of work has therefore focused largely on system-specific studies with
+traits collected in situ. However, ongoing efforts to assemble
+functional traits for a variety of taxa have made synthesis of existing
+community assemblage data with functional traits possible for the first
+time [@trindade-santos2020; @jarzyna2016; @barnagaud2017], providing
+initial looks at how functional diversity changes on a broad-scale. We
+approach local biodiversity change through this lens, providing the
+first cross-taxa and cross-system assessment of functional diversity
+change through time.
 

### 2021-02-24 paper/manuscript.Rmd
Revamp methods, refine intro.

--- 2021-02-05 paper/manuscript.Rmd
@@
    - \linenumbers
-output: 
-  rticles::elsevier_article:
-    includes:
-      in_header: preamble.tex
+output: word_document
+  #rticles::elsevier_article:
+   # includes:
+     # in_header: preamble.tex
 editor_options: 
@@
 of anthropogenic pressures such as climate change, land use change, and
-invasive species.
+invasive species. Impacts of these pressures are well documented on a
+global scale with an accelerating global extinction rate (citation) and
+fundamental changes in some of the most well-studied systems (e.g. coral
+bleaching). At the local scale however, species diversity tells a
+different story. Recent synthesis of local trends in biodiversity over
+time have shown a decoupling of community composition from richness
+[@brose2016], evidence of community-level regulation of richness
+[@gotelli2017], and no net change in local biodiversity despite
+widespread environmental change and ongoing species turnover
+[@brown2001; @dornelas2014; @vellend2013; , @vellend2017]. While
+communities are clearly changing, our most common species-based
+approaches do not fully capture the nature of that change.
 
-#### Paragraph 1 - communities are changing, we need to know how and what that means for ecosystem properties
+Functional diversity offers a potentially powerful alternative for
+detecting and describing community change by providing a mechanistic
+link between the organisms in a system and the processes they perform
+[@mcgill2006]. By summarizing functional traits rather than species,
+functional diversity metrics capture the disproportionate impact of
+losses or gains of functionally unique species. Functional diversity
+metrics are particularly suited for understanding local-scale
+biodiversity patterns as they can detect shifts in community structure
+independent of species change.
 
--   communities are changing as a result of a whole bunch of different
-    pressures
+<!-- Functional diversity also sheds light on critical ecological processes. -->
 
--   a few well documented phenomena: accelerating extinction rates, what
-    else?
+Functional diversity also sheds further light on critical ecosystem and
+community processes. For example, there is increasing evidence that
+functional diversity is a better predictor of ecosystem function than
+species-based metrics [@cadotte2011; @gagic2015], and that different
+facets of functional diversity also play a role in maintaining ecosystem
+stability [\@craven2018; @morin2014]. Further, biogeographic assessments
+of functional diversity give foundational insight into how community
+assembly plays out across systems at a standardized scale [@violle2014;
+@weiher2011].
 
--   it's essential for conservation and ecology to understand what these
-    changes mean for ecosystem properties like stability, function, etc
+<!-- What can we say about how global change is impacting:stability, resilience, function, assembly, ecosystem processes (and ecosystem services) - need a functional grounding for all those concepts.
+Link specific patterns in FD metrics to impacts on those concepts.
+-->
 
-#### Paragraph 2 - species diversity is the go to - it's not working
+Assessments of broad-scale temporal change in functional diversity have
+previously been limited by a lack of functional trait data. The majority
+of work has therefore focused largely on system-specific studies with
+traits collected *in situ*. Ongoing efforts to assemble functional
+traits for a variety of taxa have made synthesis of existing community
+assemblage data and functional traits possible for the first time,
+providing initial looks at how functional diversity changes on a
+broad-scale [@trindade-santos2020; @jarzyna2016; @barnagaud2017].
+<!-- so what do these studies say? does that go here? --> However to
+date there has been no comprehensive assessment of functional change
+through time on a broad geographic and taxonomic extent.\
 
--   traditionally the most common method of capturing change in
-    communities is biodiversity
-
--   despite other well documented changes, species diversity doesn't
-    seem to capture local community-level changes
-    ([@dornelasAssemblageTimeSeries2014],
-    [@suppSpecieslevelCommunitylevelResponses2014],
-    [@vellendGlobalMetaanalysisReveals2013]), though there is evidence
-    of a significant increase in turnover
-
--   We need to look beyond species and their abundances to understand
-    the kinds of changes happening in communities
-
-#### Paragraph 3 - functional diversity is the answer!
-
--   Functional diversity provides a potentially powerful tool for
-    examining biodiversity changes because it captures characteristics
-    beyond species identity.
-
--   Powerful functional lens ([@mcgillRebuildingCommunityEcology2006])
-
--   Functional diversity can tells us a story independent of species
-    richness, as huge functional changes can occur while species
-    richness (or even diversity!) stays the same. This especially gives
-    insight into the turnover we know is occurring
-
--   We examine 100's of mammal, bird, and amphibian timeseries from the
-    Biotime database paired with functional trait data to assess overall
-    trends in functional change and identify where functional diversity
-    tells us a story beyond species richness
+Here we perform the first cross-taxa and cross-system assessment of
+functional diversity change through time. We assess 100's of mammal,
+bird, and amphibian functional diversity timeseries to determine how
+functional diversity change compares to species change. We also identify
+3 distinct types of functional change common across communities and
+their implications for ecological processes.
 
@@
 
-Biodiversity timeseries were obtained from the Biotime database, a
-global repository of assemblage timeseries with 386 individual studies
-from multiple taxa. Following best practices for the dataset
-[@blowes2019] to account for differences in scale between studies, we
-aggregated based on a grid system:
+We obtained mammal, bird, and amphibian timeseries from the Biotime
+database, a global repository of high quality assemblage timeseries. All
+studies included in the database follow consistent sampling protocols
+and represent full assemblages rather than population-level data
+[@dornelas2018]. Following best practices for the database [@blowes2019]
+and to maintain a consistently defined local scale, studies with
+multiple sample locations were split into separate time series based a
+global grid wherein all samples within a single grid cell were
+considered a single time series. The size of the cell was determined
+based on the sample extent of the single location studies (see
+[@dornelas2018] for details on how sample extents were defined), with
+the area of each cell set to one standard deviation away from the mean
+of the single extent locations.
 
-*annotation of* `01_biotime_grid.R`
-
--   label single location (SL) or multiple location (ML), ML with an
-    extent less than ((SL extent mean + SL extent SD) are assigned to be
-    SL
-
-    -   SL coords are the central lat and long, ML coords are the center
-        of convex hull of locations within a rarefyID (grid cell)
-
--   exclude studies with one year of data
-
--   create global grid with resolution based on the cutoff study area
-    size (SL extent mean + SL extent SD)
-
--   collate observations for the same study within cells/years
-
--   remove years (for count) or rarefyID's (for abundance and biomass)
-    that do not meet coverage requirement of
-
-Timeseries were paired with their functional trait data based on the
-predominant taxa in the timeseries. Due to limitations in trait data and
-processing feasability, we included only mammal, bird, and amphibian
-timeseries. Trait data were obtained from Elton Trait Database (mammals
-and birds; [@wilman2014]) and Amphibio (amphibians, [@oliveira2017]).
-
--   something here describing where the data in those databases comes
-    from, that they're species-level means, etc**;**
-
--   also more description of how biotime is structured (studies, some
-    with plots, etc)
+We gathered trait data from the Elton Trait Database (mammals and birds;
+[@wilman2014]) and Amphibio (amphibians, [@oliveira2017]). These
+databases include species-level means for traits that best define
+species' function in the community including body size, diet, and
+behavioral characteristics <!-- what's in the amphibian database? -->.
+For the full list of traits included in the analysis see the
+supplemental materials.
 
@@
 from the Integrated Taxonomic Information System database (citation),
-obtained through the taxadb R package
-([@normanTaxadbHighperformanceLocal2020]). If more than one species in
-the assemblage data resolved to the same identifier, observations were
-considered the same species. For trait data, traits for all species of
-the same identifier were averaged. Only studies with at least 85% trait
-coverage were included. In order to have a sufficient number of
-observation years to observe a trend and a sufficient number of species
-to calculate functional diversity metrics, studies with fewer than 4
-years and years with fewer than 4 species observed were also excluded.
+obtained through the taxadb R package ([@norman2020]). If more than one
+species in the assemblage data resolved to the same identifier,
+observations were considered the same species. For trait data, traits
+for all species of the same identifier were averaged. Only studies with
+at least 85% trait coverage were included. In order to have a sufficient
+number of species to calculate functional diversity metrics years with
+fewer than 4 species observed were excluded.
 
-Functional diversity metrics were calculated for total yearly abundance.
-For many studies there were a different number of samples within each
-year. To adjust for this inconsistency in sampling effort we used
-sample-based rarefaction by bootstrap resampling within year for each
-timeseries based on the smallest number of samples in a year for that
-timeseries.
-
--   raw abundance counts within a year for each plot
+Many studies had a variable number of samples within years. To adjust
+for this sampling effort inconsistency we used sample-based rarefaction
+by bootstrap resampling within years for each timeseries based on the
+smallest number of samples in a year for that timeseries.
 
@@
 We calculated yearly metrics of functional diversity and turnover for
-each plot within each study. All available trait data for each study was
-included in the trait matrix with the exception of traits that were the
-same value for all observed species in the study. All continuous traits
-were z-score scaled to give each trait equal weight in the trait space
-([@lepsQuantifyingInterpretingFunctional2006],
-[@schleuterUserGuideFunctional2010]).
+each timeseries within each study. All available trait data for each
+study was included in the trait matrix with the exception of traits that
+were the same value for all observed species in the study. All
+continuous traits were z-score scaled to give each trait equal weight in
+the trait space ([@leps2006; @schleuter2010]). We also calculated basic
+species richness, *S.*
 
@@
 incorporated weighting based on either species abundance or biomass
-based on the methods for each study. The PCA-scaled trait matrices were
-then used as input to the BAT package....
-
--   needed euclidian distance matrix - used calliez correction because
-    it was a sufficient correction for all plots even though it led to a
-    marginal decrease in "quality" for the corrected hull
-
--   null model stuff via [@swensonBiogeographyFilteringWoody2012] - for
-    years where all species in the species pool were observed the null
-    model sd is zero for functional richness, since FRic does not
-    incorporate abundance changes. - drop species with trait match
+based on the methods for each study.
 
 #### *Analysis*
+
+We modeled each diversity metric using a Generalized Linear Mixed
+Effects model with year as the predictor and mixed effects of timeseries
+nested within study. <!-- more to come! -->

### 2021-03-09 analysis/05_model_metrics.Rmd
Debugging problem models, add fixed effect for year and scale variables.

--- 2021-02-01 analysis/05_model_metrics.Rmd
@@
   rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
-  mutate(logvalue = log(value + abs(floor(min(na.omit(value)))))) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
+  mutate(logvalue = log(value + abs(floor(min(na.omit(value))))),
+         year_scaled = scale(year)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
   filter(!is.na(value)) 
@@
 
+Visualize distributions of response
+```{r}
+metrics %>%
+  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
+  ggplot() +
+  geom_histogram(aes(x = logvalue),
+                 bins = 60) +
+  facet_wrap(~metric, scales = "free") +
+  labs(title = "Distribution of Log values")
+
+metrics %>%
+  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
+  ggplot() +
+  geom_histogram(aes(x = value),
+                 bins = 60) +
+  facet_wrap(~metric, scales = "free") +
+  labs(title = "Distribution of raw values")
+```
+
 Fit GLM to overall metric trend
+
+Convergence problems!! See here:
+https://joshua-nugent.github.io/allFit/
+https://rstudio-pubs-static.s3.amazonaws.com/33653_57fc7b8e5d484c909b615d8633c01d51.html
+https://biologyforfun.wordpress.com/2018/04/09/help-i-have-convergence-warnings/
+
+Maybe need something other than a gaussian distribution?
+
+Model checks:
+https://debruine.github.io/posts/normality/
+performance::check_model()
+
 ```{r}
 # model for overall trend (random effect w/varying intercept)
-## should be GLS according to Dornelas supplement, but glmer isn't happy
 metric_mean_fit <- metrics %>%
-  mutate(minyear = min(year), maxyear = max(year)) %>%
-  group_nest(metric, minyear, maxyear) %>%
-  mutate(model = map(data, ~lmer(logvalue ~ year + (1|study_id) + (1|rarefyID), 
-                                 data = .x, control = lmerControl(optimizer ="Nelder_Mead"))))
+  filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S_int")) %>%
+  group_nest(metric) %>%
+  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (1|study_id/rarefyID), data = .x,
+                  control = lmerControl(optCtrl = list(maxit = 1e9,
+                                                    maxfun = 1e9,
+                                                    xtol_abs = 1e-11,
+                                                    ftol_abs = 1e-11,
+                                                    maxeval = 1e9)))),
+         null_model = map(data, ~lmer(logvalue ~ (1|study_id/rarefyID), data = .x,
+                  control = lmerControl(optCtrl = list(maxit = 1e9,
+                                                    maxfun = 1e9,
+                                                    xtol_abs = 1e-11,
+                                                    ftol_abs = 1e-11,
+                                                    maxeval = 1e9)))),
 
-metric_mean_coef <- metric_mean_fit %>%
-  mutate(coef = map(model, tidy)) %>% 
-  #mutate(confint = map(model, confint)) %>%
-  select(-data, -model) %>% 
-  unnest(cols = c(coef))
+           test = map2(model, null_model, anova),
+           test_res = map(test, tidy),
+           coef = map(model, tidy)
+           ) %>% 
+  unnest(cols = c(test_res)) %>%
+  select(metric, model, coef, p.value) %>%
+  filter(!is.na(p.value)) %>%
+  unnest(cols = c(coef)) %>%
+  filter(term == "year_scaled") %>%
+  select(-c(effect, group, term, statistic))
 
-mean_tests <- metric_mean_fit %>%
-  #filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
-   mutate(p_value = map(model, ~drop1(.x, test = "Chisq") %>% select(`Pr(Chi)`) %>% slice(2) %>% pull())) %>% 
-  select(metric, p_value) %>%
-  left_join(metric_mean_coef %>% 
-              filter(term == "year") %>% 
-              select(metric, estimate))
+metric_mean_fit_slope <- metrics %>%
+  filter(metric %in%  c("FRic", "FEve", "FDiv", "S_int")) %>%
+  group_nest(metric) %>%
+  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
+                                 data = .x)),
+           null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
+                                 data = .x)),
+           test = map2(model, null_model, anova),
+           test_res = map(test, tidy),
+           coef = map(model, tidy)
+           ) %>% 
+  unnest(cols = c(test_res)) %>%
+  select(metric, model, coef, p.value) %>%
+  filter(!is.na(p.value)) %>%
+  unnest(cols = c(coef)) %>%
+  filter(term == "year_scaled") %>%
+  select(-c(effect, group, term, statistic))
 ```
@@
 ```
+Fit a single model to play around with 
+```{r}
+eve <- metrics %>% 
+  filter(metric == "FEve")
 
-```{r}
+feve_fit <- glmer(value ~ year + (1|study_id/rarefyID), data = eve, family = Gamma,
+                  control = glmerControl(optCtrl = list(maxit = 1e9,
+                                                    maxfun = 1e9,
+                                                    xtol_abs = 1e-11,
+                                                    ftol_abs = 1e-11,
+                                                    maxeval = 1e9)))
+
+#find best optimizer
+diff_optims <- allFit(feve_fit, maxfun = 1e5)
+
+optimx_options <- c("L-BFGS-B", "nlminb", "nlm", "bobyqa", "nmkb", "hjkb")
+
+for(i in 1:length(optimx_options)){
+  model_flex <- glmer(value ~ year + (1|study_id/rarefyID), data = eve, family = Gamma,
+                     control = glmerControl(optimizer = "optimx",
+                                           optCtrl = list(method = optimx_options[i],
+                                                                   maxit = 1e9)))
+  print(model_flex@optinfo$conv$lme4$messages)
+}
+
+algoptions <- c("NLOPT_LN_PRAXIS", "NLOPT_GN_CRS2_LM",
+"NLOPT_LN_COBYLA", "NLOPT_LN_NEWUOA",
+"NLOPT_LN_NEWUOA_BOUND", "NLOPT_LN_NELDERMEAD",
+"NLOPT_LN_SBPLX", "NLOPT_LN_BOBYQA")
+
+for(i in 1:length(algoptions)){
+  model_flex <- glmer(value ~ year + (1|study_id/rarefyID), data = eve, family = Gamma,
+                     control = glmerControl(optimizer = "nloptwrap",
+                                           optCtrl = list(algorithm = algoptions[i],
+                                                          maxeval = 1e7,
+                                                          xtol_abs = 1e-9,
+                                                          ftol_abs = 1e-9)))
+  if(is.null(model_flex@optinfo$conv$lme4$messages)){
+    print(paste0("One of the nloptwrap options, ", algoptions[i],", worked!"))
+    print(summary(model_flex))
+    break
+  }
+}
+
 fit_mod <- function(data){

### 2021-03-09 analysis/problem_model.Rmd
Debugging problem models, add fixed effect for year and scale variables.

---
title: "Models"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(lme4)
library(broom.mixed)
```

Get data and put it in long form 
```{r}
pins::board_register_github(repo = "karinorman/biodivTS_data", branch = "master")

metrics <- pins::pin_get("rarefied-medians", board = "github") %>%
  select(-cYEAR) %>%
  pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
  mutate(logvalue = log(value + abs(floor(min(na.omit(value)))))) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
  filter(!is.na(value))
```

####Data exploration

Visualize distributions of response
```{r}
metrics %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S_int")) %>%
  ggplot() +
  geom_histogram(aes(x = logvalue),
                 bins = 60) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Distribution of Log values")

metrics %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S_int")) %>%
  ggplot() +
  geom_histogram(aes(x = value),
                 bins = 60) +
  facet_wrap(~metric, scales = "free") +
  labs(title = "Distribution of raw values")
```

Fit a single model to play around with 
```{r}
test_data <- metrics %>% 
  filter(metric == "FRic") %>%
  mutate(study_id = as.factor(study_id),
         year_center = scale(year),
         value_scale = scale(value))

#lognormal
ln_fit <- lmer(value_scale ~ year_center + (year_center|study_id/rarefyID), data = test_data,
               control = lmerControl(optCtrl = list(maxit = 1e9,
                                                    maxfun = 1e9,
                                                    xtol_abs = 1e-11,
                                                    ftol_abs = 1e-11,
                                                    maxeval = 1e9)))

gamma_fit <- glmer(value ~ year_center + (year_center|study_id/rarefyID), data = test_data, family = Gamma)

poiss_fit <- glmer(value ~ year_center + (year_center|study_id/rarefyID), data = test_data, family = poisson(link = "log"))

nb_fit <- glmer.nb(value ~ year_center + (year_center|study_id/rarefyID), data = test_data)
```

```{r}
options(mc.cores = parallel::detectCores())
stan_fit <- stan_lmer(value_scale ~ year_center + (year_center|study_id/rarefyID), data = test_data, iter = 500)

launch_shinystan(stan_fit,ppd = FALSE)
```

Look for best optimizer using `allFit()`, doesn't look like any of them work
```{r}
#find best optimizer
#lognormal fit
ln_optims <- allFit(ln_fit, maxfun = 1e5)

is.OK <- sapply(ln_optims, is, "merMod")
ln_optims.OK <- ln_optims[is.OK]
lapply(ln_optims.OK,function(x) x@optinfo$conv$lme4$messages)

```

Let's try all the options for both optimix and nloptwrap, first for the generalized option w/Gamma distribution
```{r}
optimx_options <- c("L-BFGS-B", "nlminb", "nlm", "bobyqa", "nmkb", "hjkb")

for(i in 1:length(optimx_options)){
  model_flex <- lmer(logvalue ~ year_center + (year_center|study_id/rarefyID), data = test_data,
                     control = lmerControl(optimizer = "optimx",
                                           optCtrl = list(method = optimx_options[i],
                                                                   maxit = 1e9)))
  if(is.null(model_flex@optinfo$conv$lme4$messages)){
    print(paste0("One of the optimx options, ", optimx_options[i],", worked!"))
    print(summary(model_flex))
    break
  }
}

algoptions <- c("NLOPT_LN_PRAXIS", "NLOPT_GN_CRS2_LM",
"NLOPT_LN_COBYLA", "NLOPT_LN_NEWUOA",
"NLOPT_LN_NEWUOA_BOUND", "NLOPT_LN_NELDERMEAD",
"NLOPT_LN_SBPLX", "NLOPT_LN_BOBYQA")

for(i in 1:length(algoptions)){
  model_flex <- lmer(logvalue ~ year_center + (year_center|study_id/rarefyID), data = test_data,
                     control = lmerControl(optimizer = "nloptwrap",
                                           optCtrl = list(algorithm = algoptions[i],
                                                          maxeval = 1e7,
                                                          xtol_abs = 1e-9,
                                                          ftol_abs = 1e-9)))
  if(is.null(model_flex@optinfo$conv$lme4$messages)){
    print(paste0("One of the nloptwrap options, ", algoptions[i],", worked!"))
    print(summary(model_flex))
    break
  }
}
```

```{r}

optimx_options <- c("L-BFGS-B", "nlminb", "nlm", "bobyqa", "nmkb", "hjkb")

for(i in 1:length(optimx_options)){
  model_flex <- glmer(value ~ (year|study_id/rarefyID), data = test_data, family = Gamma,
                     control = glmerControl(optimizer = "optimx",
                                           optCtrl = list(method = optimx_options[i],
                                                                   maxit = 1e9)))
    if(is.null(model_flex@optinfo$conv$lme4$messages)){
    print(paste0("One of the optimx options, ", optimx_options[i],", worked!"))
    print(summary(model_flex))
    break
  }
}

#nloptwrap
algoptions <- c("NLOPT_LN_PRAXIS", "NLOPT_GN_CRS2_LM",
"NLOPT_LN_COBYLA", "NLOPT_LN_NEWUOA",
"NLOPT_LN_NEWUOA_BOUND", "NLOPT_LN_NELDERMEAD",
"NLOPT_LN_SBPLX", "NLOPT_LN_BOBYQA")

for(i in 1:length(algoptions)){
  model_flex <- glmer(value ~ (year|study_id/rarefyID), data = test_data, family = Gamma,
                     control = glmerControl(optimizer = "nloptwrap",
                                           optCtrl = list(algorithm = algoptions[i],
                                                          maxeval = 1e7,
                                                          xtol_abs = 1e-9,
                                                          ftol_abs = 1e-9)))
  if(is.null(model_flex@optinfo$conv$lme4$messages)){
    print(paste0("One of the nloptwrap options, ", algoptions[i],", worked!"))
    print(summary(model_flex))
    break
  }
}

```

### 2021-03-09 analysis/visualization.Rmd
Debugging problem models, add fixed effect for year and scale variables.

@@
 ```{r}
-usa <- map_data("usa")
+world <- map_data("world")
 
 plt <- ggplot() +
-  geom_polygon(data = usa, aes(x=long, y = lat, group = group), fill = NA, color = "black") +
-  coord_fixed(1.3)
+  geom_polygon(data = world, aes(x=long, y = lat, group = group), fill = NA, color = "black", size = .25) +
+  coord_fixed(1.3) +
+  theme_minimal()
 
 plt +
-  geom_point(data = biotime_data %>% filter(taxa == "Fish"), aes(x = longitude, y = latitude), color = "black", size = .25)
+  geom_point(data = metrics %>% distinct(cent_lat, cent_long), aes(x = cent_long, y = cent_lat), color = "#cc5968", size = .75)
+
+ggsave("study_map.png")
 

### 2021-03-09 paper/manuscript.Rmd
Add null model section.

@@
 of anthropogenic pressures such as climate change, land use change, and
-invasive species. Impacts of these pressures are well documented on a
-global scale with an accelerating global extinction rate (citation) and
-fundamental changes in some of the most well-studied systems (e.g. coral
+invasive species. Impacts of these pressures are well documented at a
+global scale by an accelerating global extinction rate [@barnosky2011],
+and fundamental changes in some of the well-studied systems (e.g. coral
 bleaching). At the local scale however, species diversity tells a
 different story. Recent synthesis of local trends in biodiversity over
-time have shown a decoupling of community composition from richness
-[@brose2016], evidence of community-level regulation of richness
-[@gotelli2017], and no net change in local biodiversity despite
-widespread environmental change and ongoing species turnover
-[@brown2001; @dornelas2014; @vellend2013; , @vellend2017]. While
-communities are clearly changing, our most common species-based
-approaches do not fully capture the nature of that change.
+time have found no net change in local biodiversity despite ongoing
+species turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017]
+and evidence of a decoupling of community composition from richness
+[@brose2016; [@gotelli2017]. While communities are clearly changing, our
+most common species-based approaches do not fully capture the nature of
+that change.
 
@@
 losses or gains of functionally unique species. Functional diversity
-metrics are particularly suited for understanding local-scale
+metrics are particularly suited for understanding community-level
 biodiversity patterns as they can detect shifts in community structure
@@
 
-<!-- Functional diversity also sheds light on critical ecological processes. -->
-
-Functional diversity also sheds further light on critical ecosystem and
-community processes. For example, there is increasing evidence that
-functional diversity is a better predictor of ecosystem function than
-species-based metrics [@cadotte2011; @gagic2015], and that different
-facets of functional diversity also play a role in maintaining ecosystem
-stability [\@craven2018; @morin2014]. Further, biogeographic assessments
-of functional diversity give foundational insight into how community
+Functional diversity sheds further light on critical ecosystem and
+community processes potentially impacted by anthropogenic change. For
+example, there is increasing evidence that functional diversity is a
+better predictor of ecosystem function than species-based metrics
+[@cadotte2011; @gagic2015], and that different facets of functional
+diversity also play a role in maintaining ecosystem stability
+[\@craven2018; @morin2014]. Further, biogeographic assessments of
+functional diversity give foundational insight into how community
 assembly plays out across systems at a standardized scale [@violle2014;
@@
 providing initial looks at how functional diversity changes on a
-broad-scale [@trindade-santos2020; @jarzyna2016; @barnagaud2017].
-<!-- so what do these studies say? does that go here? --> However to
-date there has been no comprehensive assessment of functional change
+broad-scale for specific taxa [@trindade-santos2020; @jarzyna2016;
+@barnagaud2017].
+<!-- so what do these studies say? does that go here? --> However, to
+date there has been no cross-taxa assessment of functional change
 through time on a broad geographic and taxonomic extent.\
 
-Here we perform the first cross-taxa and cross-system assessment of
+Here we perform the first multitaxa, multisystem assessment of
 functional diversity change through time. We assess 100's of mammal,
-bird, and amphibian functional diversity timeseries to determine how
-functional diversity change compares to species change. We also identify
-3 distinct types of functional change common across communities and
-their implications for ecological processes.
+bird, and amphibian functional diversity timeseries to 1) determine how
+functional diversity change compares to species change and 2) identify 3
+distinct types of functional change common across communities.
 
@@
 We obtained mammal, bird, and amphibian timeseries from the Biotime
-database, a global repository of high quality assemblage timeseries. All
-studies included in the database follow consistent sampling protocols
-and represent full assemblages rather than population-level data
-[@dornelas2018]. Following best practices for the database [@blowes2019]
-and to maintain a consistently defined local scale, studies with
-multiple sample locations were split into separate time series based a
-global grid wherein all samples within a single grid cell were
-considered a single time series. The size of the cell was determined
-based on the sample extent of the single location studies (see
-[@dornelas2018] for details on how sample extents were defined), with
-the area of each cell set to one standard deviation away from the mean
-of the single extent locations.
+database, a global repository of high quality community assemblage
+timeseries. All studies included in the database follow consistent
+sampling protocols and represent full assemblages rather than
+population-level data [@dornelas2018]. Following best practices for the
+database [@blowes2019] studies with multiple sample locations were split
+into individual timeseries following a standardized scale. A global grid
+was defined with a cell size determined based on the sample extent of
+studies with only a single location (see [@dornelas2018] for details on
+how sample extents were defined), with the area of each cell set to one
+standard deviation away from the mean of the single extent locations.
+All samples from a study within a single cell were considered to be a
+single timeseries, and species abundances were combined for all samples.
 
@@
 
-In order to ensure taxonomic consistency across datasets, biotime
+In order to ensure taxonomic consistency across datasets, Biotime
 species were paired with trait data based on their species identifier
 from the Integrated Taxonomic Information System database (citation),
-obtained through the taxadb R package ([@norman2020]). If more than one
+obtained through the taxadb R package [@norman2020]. If more than one
 species in the assemblage data resolved to the same identifier,
@@
 Many studies had a variable number of samples within years. To adjust
-for this sampling effort inconsistency we used sample-based rarefaction
-by bootstrap resampling within years for each timeseries based on the
-smallest number of samples in a year for that timeseries.
+for this inconsistency in sampling effort we used sample-based
+rarefaction by bootstrap resampling within years for each timeseries
+based on the smallest number of samples in a year for that timeseries.
 
@@
 continuous traits were z-score scaled to give each trait equal weight in
-the trait space ([@leps2006; @schleuter2010]). We also calculated basic
+the trait space [@leps2006; @schleuter2010]. We also calculated basic
 species richness, *S.*
@@
 We used the function dbFD from the FD package (citation) to calculated
-PCA-scaled trait matrices for each species in a timeseries and
 functional diversity (FD), functional evenness (FEve), functional
-diveregence (FDiv), and functional dispersion (FDis) based on those
-trait matrices. The number of trait axes was limited to the maximum
-number of traits that fulfills the criteria $s >= s^t$, where $s$ is the
-number of species and \$t\$ is the number of traits. This limitation
-allows for a sufficient number of axes to capture the trait space while
-maintaining computational feasibility for metric calculations. Metrics
-incorporated weighting based on either species abundance or biomass
-based on the methods for each study.
+diveregence (FDiv), and functional dispersion (FDis). The number of
+trait axes was limited to the maximum number of traits that fulfills the
+criteria $s >= s^t$, where $s$ is the number of species and \$t\$ is the
+number of traits. This restriction allows for a sufficient number of
+axes to capture the trait space while maintaining computational
+feasibility for metric calculations. Metrics incorporated weighting
+based on either species abundance or biomass based on the methods for
+each study.
+
+***Null Models***
+
+In order to control for changes in functional diversity metrics due to
+species richness, we calculated the standardized effect size (SES) for
+each metric from null estimates [@swenson2012]. For each richness level
+observed in a timeseries, our null model randomly selected the same
+number of species from the species pool, which included all species
+observed over the course of the timeseries. We then randomly assigned
+abundances from the original sample to each species, and all functional
+diversity metrics were calculated. This process was repeated 500 times
+to get a estimate and standard deviation of the null expectation for the
+metric for that timeseries at that richness level. We used these values
+to calculated SES using the following formula:
+$SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
 
@@
 
-We modeled each diversity metric using a Generalized Linear Mixed
-Effects model with year as the predictor and mixed effects of timeseries
-nested within study. <!-- more to come! -->
+We modeled each diversity metric using a Linear Mixed Effects model with
+a random slope and intercept for year and a nested random effect of
+timeseries nested within study.

### 2021-03-10 data-raw/01_biotime_grid.R
Get rid of abundance coverage filtering.

@@ -233,198 +233,206 @@ bt_grid <- bt_grid %>% unite(col=rarefyID, STUDY_ID, cell, sep="_", remove=FALSE
 # add column for ObsEventID (STUDY_ID + PLOT + YEAR = discrete sampling events within a year)
 bt_grid <- bt_grid %>% unite(col=ObsEventID, rarefyID, PLOT, YEAR, sep="_", remove=FALSE)
 
-#save locally
-usethis::use_data(bt_grid)
-
-#save to cache
-pins::pin(bt_grid, board = "github")
-
-
-##==================================== EXPLORE AND FILTER ON COVERAGE ========================================
-##	first collate taxa within cells, to calculate coverage for count data (see below for incidence)
 bt_grid_collate <- bt_grid %>%
-  group_by(CLIMATE, REALM, TAXA, StudyMethod, ABUNDANCE_TYPE, rarefyID, STUDY_ID, YEAR, cell, Species) %>%
+  group_by(CLIMATE, REALM, TAXA, StudyMethod, ABUNDANCE_TYPE, BIOMASS_TYPE, rarefyID, STUDY_ID, YEAR, cell, Species) %>%
   summarise(
     Abundance = sum(as.numeric(Abundance), na.rm=TRUE),
     Biomass = sum(as.numeric(Biomass), na.rm=TRUE)) %>%
-  ungroup()
-
-abund_coverage <- bt_grid_collate %>%
-  # get only rows representing count data (abundance)
-  filter(ABUNDANCE_TYPE!='Presence/Absence' & ABUNDANCE_TYPE!='<NA>') %>%
-  # remove zeroes and NAs(some studies are designated count, but record Biomass estimates)
-  filter(Abundance > 0 & !is.na(Abundance)) %>%
-  group_by(CLIMATE, REALM, TAXA, StudyMethod, ABUNDANCE_TYPE, rarefyID, STUDY_ID, YEAR, cell) %>%
-  summarise(
-    # how many singletons
-    singletons = sum(Abundance==1),
-    # how many doubletons
-    doubletons = sum(Abundance==2),
-    # how many individuals in total sample
-    N = sum(Abundance),
-    # eqn 4a in Chao & Jost 2012 Ecology (==eqn 12 in Chao et al 2014 Ecol Monogr)
-    Chat = 1 - (singletons/N) * (((N-1)*singletons)/((N-1)*singletons + 2*doubletons)),
-    # with fix from Chao et al 2014 Subroutine for iNEXT code (appendix)
[…]

### 2021-03-10 data-raw/02_biotime_ids.Rmd
Get from pins

@@
 #file created by 01_biotime_grid.R
-load(here::here('data/BioTIME_grid_filtered.Rdata'))
+#load(here::here('data/BioTIME_grid_filtered.Rdata'))
+bt_grid_filtered <- pins::pin_get("bt-grid-filtered", board = "github")
+
 data <- bt_grid_filtered %>%

### 2021-03-10 paper/manuscript.Rmd
Fix citation.

@@
 diversity also play a role in maintaining ecosystem stability
-[\@craven2018; @morin2014]. Further, biogeographic assessments of
+[@craven2018; @morin2014]. Further, biogeographic assessments of
 functional diversity give foundational insight into how community
@@
 
-***Null Models***
+#### ***Null Models***
 

### 2021-03-10 data-raw/02_biotime_ids.Rmd
update pin import

--- 2021-01-15 data-raw/02_biotime_ids.Rmd
@@
 #file created by 01_biotime_grid.R
-load(here::here('data/BioTIME_grid_filtered.Rdata'))
-data <- bt_grid_filtered %>%
+#load(here::here('data/BioTIME_grid_filtered.Rdata'))
+bt_grid_col <- pins::pin_get("bt-grid-collate", board = "github_data")
+
+data <- bt_grid_col %>%
   rename_with(tolower) %>%
@@
 
-rm(bt_grid_filtered)
+rm(bt_grid_col)
 ```

### 2021-03-11 data-raw/02_biotime_ids.Rmd
bt_grid instead of collated

--- 2021-01-15 data-raw/02_biotime_ids.Rmd
@@
 #file created by 01_biotime_grid.R
-load(here::here('data/BioTIME_grid_filtered.Rdata'))
-data <- bt_grid_filtered %>%
+#load(here::here('data/BioTIME_grid_filtered.Rdata'))
+bt_grid <- pins::pin_get("bt-grid", board = "github_data")
+
+data <- bt_grid %>%
   rename_with(tolower) %>%
@@
 
-rm(bt_grid_filtered)
+rm(bt_grid)
 ```

### 2021-03-11 data-raw/02_biotime_ids.Rmd
Resolve ID's appropriately

--- 2021-03-10 data-raw/02_biotime_ids.Rmd
@@
 #file created by 01_biotime_grid.R
-#load(here::here('data/BioTIME_grid_filtered.Rdata'))
-bt_grid_col <- pins::pin_get("bt-grid-collate", board = "github_data")
+pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
+bt_grid_col <- pins::pin_get("bt-grid-collate", board = "github")
 
@@
     "ITIS:172921", 
-    "ITIS:683209",
+    "ITIS:508923",
     "ITIS:527684", 
     "ITIS:782604", 
-    "ITIS:159807"
+    "ITIS:159807", 
+    "ITIS:28034"
   )) %>%
@@
 #save to cache
-pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
 pins::pin(biotime_data, board = "github")

### 2021-03-11 data-raw/02_biotime_ids.Rmd

@@
 #file created by 01_biotime_grid.R
-pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
-bt_grid_col <- pins::pin_get("bt-grid-collate", board = "github")
+bt_grid <- pins::pin_get("bt-grid", board = "github_data")
 
-data <- bt_grid_col %>%
+data <- bt_grid %>%
   rename_with(tolower) %>%
@@
 
-rm(bt_grid_col)
+rm(bt_grid)
 ```

### 2021-03-11 analysis/01_filter_by_traits.Rmd
allow 75% trait coverage

@@
 #set coverage threshold for analysis dataset
-percent <- 0.85
+percent <- 0.75
 
@@
   ungroup()
-```
-
-```{r}
-usethis::use_data(biotime_traits)
-pins::pin(biotime_traits, board = "github")
 ```

### 2021-03-11 data-raw/02_biotime_ids.Rmd
allow 75% trait coverage

@@
 #file created by 01_biotime_grid.R
-bt_grid <- pins::pin_get("bt-grid", board = "github_data")
+pins::board_register_github(repo = "karinorman/biodivTS_data", branch = "master", name = "github")
+bt_grid <- pins::pin_get("bt-grid", board = "github")
 
@@
     "ITIS:159807", 
-    "ITIS:28034"
+    "ITIS:28034",
+    "ITIS:71868",
+    "ITIS:683103"
   )) %>%

### 2021-03-12 analysis/05_model_metrics.Rmd
Clean up modeling and visualization

@@
 metrics <- pins::pin_get("rarefied-medians", board = "github") %>%
-  select(-cYEAR) %>%
   pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
@@
   select(-c(effect, group, term, statistic))
+
+model_comp <- metric_mean_fit %>%
+  select(metric, no_slope_estimate = estimate, no_slope_pval = p.value) %>%
+  left_join(metric_mean_fit_slope %>% 
+  select(metric, slope_estimate = estimate, slope_pval = p.value ))
 ```
 
-Fit model to individual studies - if there is more than one timeseries in a study, account for that
-```{r}
-sing_study <- metrics %>%
-  group_by(study_id) %>%
-  filter(n_distinct(rarefyID) == 1) %>% 
-  pull(study_id) %>% unique()
-
-# indiv_data <- metrics %>%
-#   mutate(rarefyID = as.factor(rarefyID)) %>%
-#   group_by_at(setdiff(names(metrics), c("year", "value", "logvalue", "rarefyID", "cell", "samplen"))) %>% 
-#   #filter(n_distinct(year) > 1) %>% #filter out rarefyID's that don't have multiple years after dropping Na's
-#   group_nest()
-
-indiv_data <- metrics %>%
-  mutate(rarefyID = as.factor(rarefyID)) %>%
-  group_by(study_id) %>%
-  group_nest()
-
-indiv_ts_sing_fit <- indiv_data %>%
-  filter(study_id %in% sing_study) %>%
-  mutate(model = map(data, ~ lm(logvalue ~ year, data = .x))) %>%
-  mutate(model_type = "lm")
-
-indiv_ts_fit <- indiv_data %>%
-  filter(!study_id %in% sing_study) %>%
-  mutate(model = map(data, ~lmer(logvalue ~ (1 + year|rarefyID), 
-                                 data = .x, control = lmerControl(optimizer ="Nelder_Mead")))) %>%
-  mutate(model_type = "mixed",
-         p_value = map(model, ~drop1(.x, test = "Chisq") %>% select(`Pr(Chi)`) %>% slice(2) %>% pull()))
-
-indiv_ts_coef <- bind_rows(indiv_ts_fit, indiv_ts_sing_fit) %>%
-  mutate(coef = map(model, tidy)) %>% 
-  select(-data, -model) %>% 
-  unnest(cols = c(coef))
-```
-Fit a single model to play around with 
-```{r}
-eve <- metrics %>% 
-  filter(metric == "FEve")
-
-feve_fit <- glmer(value ~ year + (1|study_id/rarefyID), data = eve, family = Gamma,
-                  control = glmerControl(optCtrl = list(maxit = 1e9,
-                                                    maxfun = 1e9,
-                                                    xtol_abs = 1e-11,
-                                                    ftol_abs = 1e-11,
-                                                    maxeval = 1e9)))
-
-#find best optimizer
-diff_optims <- allFit(feve_fit, maxfun = 1e5)
-
-optimx_options <- c("L-BFGS-B", "nlminb", "nlm", "bobyqa", "nmkb", "hjkb")
-
-for(i in 1:length(optimx_options)){
-  model_flex <- glmer(value ~ year + (1|study_id/rarefyID), data = eve, family = Gamma,
-                     control = glmerControl(optimizer = "optimx",
-                                           optCtrl = list(method = optimx_options[i],
-                                                                   maxit = 1e9)))
-  print(model_flex@optinfo$conv$lme4$messages)
-}
-
-algoptions <- c("NLOPT_LN_PRAXIS", "NLOPT_GN_CRS2_LM",
-"NLOPT_LN_COBYLA", "NLOPT_LN_NEWUOA",
-"NLOPT_LN_NEWUOA_BOUND", "NLOPT_LN_NELDERMEAD",
-"NLOPT_LN_SBPLX", "NLOPT_LN_BOBYQA")
-
-for(i in 1:length(algoptions)){
-  model_flex <- glmer(value ~ year + (1|study_id/rarefyID), data = eve, family = Gamma,
-                     control = glmerControl(optimizer = "nloptwrap",
-                                           optCtrl = list(algorithm = algoptions[i],
-                                                          maxeval = 1e7,
-                                                          xtol_abs = 1e-9,
-                                                          ftol_abs = 1e-9)))
-  if(is.null(model_flex@optinfo$conv$lme4$messages)){
-    print(paste0("One of the nloptwrap options, ", algoptions[i],", worked!"))
-    print(summary(model_flex))
-    break
-  }
-}
-
-fit_mod <- function(data){
-  print(unique(data$rarefyID)) %>%
-  lmer(logvalue ~ (1 + year|rarefyID), 
-                                 data = data, control = lmerControl(optimizer ="Nelder_Mead"))
-}
-
-indiv_data %>%
-  filter(!study_id %in% sing_study) %>%
-  mutate(model = map(data, ~lmer(logvalue ~ ( 1+ year|rarefyID), 
-                                 data = .x, control = lmerControl(optimizer ="Nelder_Mead"))))
-
-fit_mod <- function(filter_id){
-  filter_data <- metrics %>% filter(study_id == filter_id)
-
-  print(filter_id)
-  lmer(logvalue ~ (1+ year|rarefyID), 
-                                 data = filter_data)
-}
-
-mult_studies <- metrics %>% filter(!study_id %in% sing_study) %>% pull(study_id) %>% unique()
-map(mult_studies, fit_mod)
-
-```
-
-Plot of S and diversity metrics
-```{r}
-indiv_ts_coef %>%
-  filter(term == "year", metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
-  select(metric, estimate) %>% 
-  ggplot() +
-  geom_histogram(aes(x = estimate)) + #,
-                 #binwidth = 2) +
-  scale_x_continuous(name = 'slope') +
-  #scale_y_continuous(breaks = c(0,2500,5000,7500,10000,12500, 15000, 17500, 20000,22500)) +
-  labs(tag = 'B',
-       y = 'Number of cells') +
-  theme_bw() +
-  theme(panel.grid.minor.x = element_blank()) +
-  facet_wrap(vars(metric), scales = "free")
-```
 Reproduce Dornelas figure - all data points & regressions w/overall mean
@@
 
-plot_metric <- function(metric_name, ylabel){
+plot_metric <- function(metric_name, ylabel, data, model_coef){
 
-  data <- metrics %>%
+  data <- data %>%
     filter(metric == metric_name, !is.na(value), !is.infinite(value))
 
-  model_coef <- metric_mean_coef %>%
+  model_coef <- model_coef %>%
     filter(metric == metric_name) %>%
@@
   intercept <- model_coef %>% filter(term == "(Intercept)") %>% pull(estimate)
-  slope <- model_coef %>% filter(term == "year") %>% pull(estimate)
+  slope <- model_coef %>% filter(term == "year_scaled") %>% pull(estimate)
 
-  pred <- data.frame(year = min(data$year):max(data$year)) %>%
-    mutate(pred = intercept + year * slope)
+  pred <- data %>%
+    select(year_scaled) %>%
+    distinct() %>%
+    mutate(pred = intercept + year_scaled * slope)
 
   trend_plot <- data %>%
-    left_join(pred, by = "year") %>%
+    left_join(pred, by = "year_scaled") %>%
     #left_join(metadata %>% select(study_id, climate)) %>%
-    ggplot(aes(x = year, y = logvalue)) +
+    ggplot(aes(x = year_scaled, y = logvalue)) +
     # geom_smooth(aes(color = climate, group = rarefyID), method = "glm", se = FALSE) +
@@
 
+model_coef <- metric_mean_fit_slope %>%
+  select(metric, model) %>%
+   mutate(coef = map(model, tidy)) %>% 
+  select(-model) %>% 
+  unnest(cols = c(coef))
+
 plot_metric("FRic", "log(FD)")
 
-plot_map <- tibble(metric_name = c("S", "FRic", "FEve", "FDiv", "FDis"), 
-           ylabel = c("log(S)", "log(FD)", "log(Evenness)", "log(Divergence)", "log(Dispersion)"))
-pmap(plot_map, plot_metric)
+plot_map <- tibble(metric_name = c("S_int", "FRic", "FEve", "FDiv", "FDis"), 
+           ylabel = c("log(S_int)", "log(FD)", "log(Evenness)", "log(Divergence)", "log(Dispersion)"))
+pmap(plot_map, plot_metric, data = metrics, model_coef = model_coef)
 ```

### 2021-03-12 analysis/04_meta_join.R

@@ -42,11 +42,11 @@ rarefied_medians <- rarefied_medians %>% filter(!(rarefyID %in% one_year_only$ra
 
 ##	centre year so intercept is mean year of time series
 ##  add log-transformed responses for modelling
-rarefied_medians <- rarefied_medians %>%
-  mutate(cYEAR = YEAR - mean(YEAR),
-         logS = log(S),
-         logN = log(N),
-         logENSPIE = log(ENSPIE))
+# rarefied_medians <- rarefied_medians %>%
+#   mutate(cYEAR = YEAR - mean(YEAR),
+#          logS = log(S),
+#          logN = log(N),
+#          logENSPIE = log(ENSPIE))
 
 ##	want coordinates of rarefyIDs (cells)
 #load('~/Dropbox/BiogeoBioTIME/rarefyID_cell_centre_011017.Rdata')

### 2021-03-16 analysis/rarefy_null_models.Rmd
first commit

---
title: "Rarefied Null Models"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(FD)
library(furrr)

pins::board_register_github(repo = "karinorman/biodivTS_data", branch = "master")
```

Get dataframe of filenames and corresponding rarefyID's
```{r}
path <- here::here("data", "rarefied_samples")
files <- list.files(path) %>%
  paste0(path, "/", .)

samples <- purrr::map_dfr(files, function(x) {load(x)
  return(rare_comm_save)
	}) %>% distinct()
```

Function to get all rarefied samples for an ID
```{r}
# get_rarefy_samps <- function(id, path = here::here("data", "rarefied_samples")){
#   print(id)
#   samp_data <- list.files(path = path, pattern = paste0( "*.", id, ".*")) %>% 
#     paste0(path, "/", .) %>%
#      purrr::map_dfr(., function(x) {load(x)
#        return(rare_comm_save)
#        }) 
#   return(samp_data)
# }
```
Get species pools for each timeseries
```{r}
bt_grid_filtered <- pins::pin_get("bt-traitfiltered", board = "github")
trait_ref <- pins::pin_get("trait-ref", board = "github")

species_pools <- bt_grid_filtered %>%
  select(rarefyid, species) %>% 
  distinct()

rm(bt_grid_filtered)
```

Simulation functions
```{r}
get_plot_sample <- function(file, n){
  #print(filter_id)
  load(file)
  comm <- rare_comm_save %>%
    pivot_longer(cols = starts_with("ITIS"), names_to = "species", values_to = "weight") %>%
    filter(weight != 0) %>%
    rename_with(tolower)

  #check that there are at least two species for all years, otherwise stop sampling
  species_count <- comm %>% select(year, species) %>% count(year)
  if (min(species_count$n) < 2){
    save_dir <- here::here("data", "no_null_sample")
    fs::dir_create(save_dir)
    write.csv(rare_comm_save, paste0(save_dir, "/", filter_rarefyid, "_samp", samp_rarefy_resamp, ".csv"))
  }
  filter_rarefyid <- unique(comm$rarefyid) %>% unlist()
  samp_rarefy_resamp <- unique(comm$rarefy_resamp) %>% unlist()

  # plot_data <- data %>% filter(rarefyID == filter_id) %>%
  #   select(year, id, weight)

  species_pool <- species_pools %>%
    filter(rarefyid == filter_rarefyid) %>%
    pull(species) %>%
    na.omit() %>%
    unique() #get column of unique scientific names

  get_samp_species <- function(data){
    samp_species <- sample(species_pool, n_distinct(data$species))
    samp_abun <- sample(data$weight)

    df <- as_tibble(cbind(samp_species, samp_abun)) %>% mutate(year = unique(data$year))
    return(df)
  }

  site_samples <- data.frame()
  for (j in 1:n){
    iter_samples <- data.frame()
    for(i in unique(comm$year)){
      year_data <- comm %>% filter(year == i)
      iter_samples <- rbind(iter_samples, get_samp_species(year_data))
    }
    site_samples <- rbind(site_samples, iter_samples %>% mutate(n = j))
  }
  plot_samples <- site_samples %>%
    mutate(rarefyid = filter_rarefyid, rarefy_resamp = samp_rarefy_resamp) %>%
    pivot_wider(names_from = samp_species, values_from = samp_abun)

  save_dir <- here::here("data", "null_samples")
  fs::dir_create(save_dir)
  write.csv(plot_samples, paste0(save_dir, "/", filter_rarefyid, "_samp", samp_rarefy_resamp, ".csv"))

  return(plot_samples)
}
```

Get sample dataframe
```{r}
future::plan(multiprocess)
sample_occurrence <- future_map(files, get_plot_sample, n = 100) %>%
  bind_rows() %>%
  replace(is.na(.), 0)
```
Figure out which samples didn't get null samples
```{r}
null_files <- list.files(here::here("data", "null_samples")) %>%
  paste0(here::here("data", "null_samples"), "/", .)

null_samples <- map_dfr(null_files, ~read_csv(.x) %>% select(rarefyid, rarefy_resamp) %>% distinct()) %>%
  mutate(rarefy_samp_file = paste0(path, "/count_cell", rarefyid, "_sample", rarefy_resamp, ".rda"))

missing_files <- setdiff(files, null_samples$rarefy_samp_file)

future::plan(multiprocess)
sample_occurrence <- future_map(missing_files, get_plot_sample, n = 100) %>%
  bind_rows() %>%
  replace(is.na(.), 0)
```

Get richness levels for a rarefyID
```{r}
get_rarefy_richness <- function(id){
  get_rarefy_samps(id) %>%
    pivot_longer(starts_with("ITIS:"), names_to = "speciesID", values_to = "abundance") %>% 
    select(YEAR, rarefy_resamp, speciesID) %>%
    group_by(rarefy_resamp, YEAR) %>% 
    summarise(richness = n_distinct(speciesID)) %>% 
    distinct()
}

rich_levels <- map_dfr(ids, get_rarefy_richness)
```

### 2021-03-24 analysis/ts_null_model.Rmd
Clean up and clarifications

@@
 library(tidyverse)
-library(biodivTS)
+#library(biodivTS)
 library(FD)
@@
 
-pins::board_register_github(repo = "karinorman/biodivTS", branch = "master")
+pins::board_register_github(repo = "karinorman/biodivTS_data", branch = "master")
 ```
@@
 ```{r}
-biotime_traits <- pins::pin_get("biotime-traits", board = "github")
+biotime_traits <- pins::pin_get("bt-traitfiltered", board = "github")
 

### 2021-03-24 analysis/02_rarefy_griddedData_clusterVersion.R

@@ -514,6 +514,6 @@ path <- here::here("data", "rarefied_samples")
 files <- dir(path, "*.rda") %>%
   paste0(path, "/", .)
 
-plan("multiprocess", workers = 4)
+plan("multiprocess", workers = 30)
 furrr::future_map(files, calc_FD_mets, traits = trait_ref)
 

### 2021-03-24 analysis/01_filter_by_traits.Rmd

@@
                                          c("sourceName", "obsID", "scientificName")]
-bt_traitfiltered <- biotime_traits %>% 
+biotime_traits <- biotime_traits %>% 
   select(all_of(bt_cols)) %>%
   rename(species = id)
+```
 
+Split timeseries that have more than one class of species (split up birds and mammals)
+```{r}
+#get the class of each species
+tax_info <- taxadb::filter_id(unique(biotime_traits$species), "itis")
+biotime_traits <- biotime_traits %>% 
+  left_join(tax_info %>% select(input, class), by = c("species" = "input"))
+
+#find all the studies with more than one taxa type
+mix_study <- biotime_traits %>% 
+  select(species, class, taxa, study_id) %>% 
+  distinct() %>% 
+  count(study_id, class, taxa) %>% 
+  group_by(study_id) %>%
+  filter(n_distinct(class, na.rm = TRUE) > 1) %>% 
+  pull(study_id) %>%
+  unique()
+
+#split into two different timeseries based on whether they're bird or mammal observations,
+#remove years with too few species
+mix_split <- biotime_traits %>%
+  filter(study_id %in% mix_study) %>%
+  mutate(rarefyid = if_else(class == "Aves", paste0(rarefyid, "_bird"), paste0(rarefyid, "_mamm"))) %>%
+  group_by(rarefyid, year) %>% 
+  filter(n_distinct(species) > 4)
+
+#update that studies that needed to be split with new rarefyid breakdowns
+bt_traitfiltered <- biotime_traits %>%
+  filter(!study_id %in% mix_study) %>%
+  bind_rows(mix_split) %>%
+  select(-class)
+```
+
+```{r}
 #save locally and in cache

### 2021-03-24 analysis/01_filter_by_traits.Rmd
Clean up

@@
   filter(study_id %in% mix_study) %>%
-  mutate(rarefyid = if_else(class == "Aves", paste0(rarefyid, "_bird"), paste0(rarefyid, "_mamm"))) %>%
+  mutate(rarefyid = if_else(class == "Aves", paste0(rarefyid, "_bird"), paste0(rarefyid, "_mamm")),
+         taxa = if_else(class == "Aves", "Birds", paste0(rarefyid, "Mammals"))) %>%
   group_by(rarefyid, year) %>% 

### 2021-03-24 analysis/05_model_metrics.Rmd
Clean up

@@
 
+Kitchen sink model
+```{r}
+all_model <- metrics %>%
+  filter(metric %in%  c("FRic", "FEve", "FDiv", "S_int")) %>%
+  group_nest(metric) %>%
+  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + taxa + realm + (year_scaled|study_id/rarefyID), 
+                                 data = .x)))#,
+  #          null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
+  #                                data = .x)),
+  #          test = map2(model, null_model, anova),
+  #          test_res = map(test, tidy),
+  #          coef = map(model, tidy)
+  #          ) %>% 
+  # unnest(cols = c(test_res)) %>%
+  # select(metric, model, coef, p.value) %>%
+  # filter(!is.na(p.value)) %>%
+  # unnest(cols = c(coef)) %>%
+  # filter(term == "year_scaled") %>%
+  # select(-c(effect, group, term, statistic))
+
+```
+
 Reproduce Dornelas figure - all data points & regressions w/overall mean

### 2021-03-24 analysis/ts_null_model.Rmd
Clean up

--- 2020-10-12 analysis/ts_null_model.Rmd
@@
 future::plan(multiprocess)
-sample_occurrence <- future_map(unique(biotime_traits$plot_id), get_plot_sample, data = biotime_traits, n = n) %>%
+sample_occurrence <- future_map(unique(biotime_traits$rarefyid), get_plot_sample, data = biotime_traits, n = n) %>%
   bind_rows() %>%

### 2021-03-24 analysis/rarefy_null_models.Rmd

@@
 
-samples <- purrr::map_dfr(files, function(x) {load(x)
-  return(rare_comm_save)
-	}) %>% distinct()
+# samples <- purrr::map_dfr(files, function(x) {load(x)
+#   return(rare_comm_save)
+# 	}) %>% distinct()
 ```
@@
 
+  filter_rarefyid <- unique(comm$rarefyid) %>% unlist()
+  samp_rarefy_resamp <- unique(comm$rarefy_resamp) %>% unlist()
+  print(filter_rarefyid)
+
   #check that there are at least two species for all years, otherwise stop sampling
@@
     write.csv(rare_comm_save, paste0(save_dir, "/", filter_rarefyid, "_samp", samp_rarefy_resamp, ".csv"))
+    return()
   }
-  filter_rarefyid <- unique(comm$rarefyid) %>% unlist()
-  samp_rarefy_resamp <- unique(comm$rarefy_resamp) %>% unlist()
 
@@
 ```{r}
-future::plan(multiprocess)
-sample_occurrence <- future_map(files, get_plot_sample, n = 100) %>%
-  bind_rows() %>%
-  replace(is.na(.), 0)
+future::plan(multiprocess, workers = 30)
+sample_occurrence <- future_map(files, get_plot_sample, n = 500, 
+                                .options = furrr_options(seed = TRUE)) #%>%
+  # bind_rows() %>%
+  # replace(is.na(.), 0)
 ```
@@
 
-null_samples <- map_dfr(null_files, ~read_csv(.x) %>% select(rarefyid, rarefy_resamp) %>% distinct()) %>%
-  mutate(rarefy_samp_file = paste0(path, "/count_cell", rarefyid, "_sample", rarefy_resamp, ".rda"))
+nulls <- as.data.frame(null_files) %>% 
+  mutate(store_nums = str_extract_all(null_files, "[0-9]+")) %>%
+  unnest_wider(store_nums) %>%
+  unite(rarefyid, `...1`, `...2`, sep = "_") %>% 
+  rename(rarefy_resamp = `...3`) %>%
+  mutate(rarefy_files = paste0(path, "/count_cell", rarefyid, "_sample", rarefy_resamp, ".rda"))
 
-missing_files <- setdiff(files, null_samples$rarefy_samp_file)
+# null_samples <- map_dfr(null_files, ~read_csv(.x) %>% select(rarefyid, rarefy_resamp) %>% distinct()) %>%
+#   mutate(rarefy_samp_file = paste0(path, "/count_cell", rarefyid, "_sample", rarefy_resamp, ".rda"))
+
+missing_files <- setdiff(files, nulls$rarefy_files)
 
 future::plan(multiprocess)
-sample_occurrence <- future_map(missing_files, get_plot_sample, n = 100) %>%
+sample_occurrence <- future_map(missing_files, get_plot_sample, n = 500,
+                                .options = furrr_options(seed = TRUE)) %>%
   bind_rows() %>%
@@
 
-Get richness levels for a rarefyID
 ```{r}
-get_rarefy_richness <- function(id){
-  get_rarefy_samps(id) %>%
-    pivot_longer(starts_with("ITIS:"), names_to = "speciesID", values_to = "abundance") %>% 
-    select(YEAR, rarefy_resamp, speciesID) %>%
-    group_by(rarefy_resamp, YEAR) %>% 
-    summarise(richness = n_distinct(speciesID)) %>% 
-    distinct()
+get_nullplot_FD <- function(file_path, traits, ...) {
+  library(FD)
+
+  plot_samples <- read_csv(file_path) %>% 
+    replace(is.na(.), 0)
+
+  sample_species_mat <- plot_samples %>%
+    select(starts_with("ITIS")) %>% #want only species x site info
+    select(sort(tidyselect::peek_vars())) #sort columns to be in alphabetical order, required for dbFDI()
+
+  print(head(sample_species_mat))
+  trait_mat <- get_traitMat(colnames(sample_species_mat), trait_data = traits)
+
+  save_dir <- here::here("data", "fd_null_output")
+  fs::dir_create(save_dir)
+
+  get_FD_safe <- possibly(get_FD, otherwise = data_frame())
+
+  max_val <- max(unlist(map(sample_species_mat, max)))
+  if(max_val > 1){
+    FD_mets <- get_FD_safe(species_mat = sample_species_mat, trait_mat = trait_mat, year_list = plot_samples$year,
+                           data_id = unique(plot_samples$rarefyid), samp_id = unique(plot_samples$rarefy_resamp),
+                           w.abun = TRUE, m = "min", corr = "cailliez")
+  }else{
+    FD_mets <- get_FD_safe(species_mat = sample_species_mat, trait_mat = traits, year_list = plot_samples$year,
+                           data_id = unique(plot_samples$rarefyid), samp_id = unique(plot_samples$rarefy_resamp),
+                           w.abun = FALSE, m = "min", corr = "cailliez")
   }
 
-rich_levels <- map_dfr(ids, get_rarefy_richness)
+  sample_fd <- FD_mets$fd_met %>% cbind(plot_samples %>% select(rarefyid, rarefy_resamp, n))
+
+  setwd(here::here())
+  write_tsv(sample_fd, paste0(save_dir, "/null_rarefy", unique(plot_samples$rarefyid), "samp",  unique(plot_samples$rarefy_resamp), ".tsv"))
+
+  return(sample_fd)
+}
+
+null_files <-  dir(here::here("data", "null_samples"), "*.csv") %>%
+  paste0(here::here("data", "null_samples"), "/", .)
+
+#in order to get this to run in parallel, must devtools::install() for the biodivTS package
+plan(multiprocess, workers = 30)
+null_FD <- future_map_dfr(null_files, ~get_nullplot_FD(.x, traits = trait_ref),
+                           .options = furrr_options(seed = TRUE))
 ```
+
+Create Null table
+```{r}
+path <- here::here("data", "fd_null_output")
+null_files <-  dir(path, "*.tsv") %>%
+  paste0(path, "/", .)
+
+process_nulls <- function(file_path, n){
+  data <- read_tsv(file_path, col_types = cols(plot_id = "c")) %>%
+    select(year, plot_id, richness = nbsp, everything(), -sing.sp, -starts_with("CWM"), -n) %>%
+    group_by(year)
+
+  means <- data %>%
+    select(-plot_id, -richness) %>%
+    summarise(across(everything(), mean)) %>%
+    pivot_longer(cols = -year, names_to = "metric", values_to = "mean")
+
+  sds <- data %>%
+    select(-plot_id, -richness) %>%
+    summarise(across(everything(), sd)) %>%
+    pivot_longer(cols = -year, names_to = "metric", values_to = "sd")
+
+  stats <- left_join(means, sds, by = c("year", "metric")) %>%
+    left_join(data %>% select(year, plot_id, richness) %>% distinct(), by = "year") %>%
+    mutate(se = sd/sqrt(n),
+           lowerCI = mean - qt(0.975, n - 1) * se, 
+           upperCI = mean + qt(0.975, n - 1) * se)  %>%
+    select(year, plot_id, everything())
+
+  return(stats)
+}
+
+plan(multiprocess)
+null_table <- future_map_dfr(null_files, process_nulls, n = n)
+
+usethis::use_data(null_table)
+pins::pin(null_table, board = "github")
+```

### 2021-03-24 analysis/rarefy_null_models.Rmd
Update for new file structure.

@@
 process_nulls <- function(file_path, n){
-  data <- read_tsv(file_path, col_types = cols(plot_id = "c")) %>%
-    select(year, plot_id, richness = nbsp, everything(), -sing.sp, -starts_with("CWM"), -n) %>%
+  data <- read_tsv(file_path) %>%
+    rename(richness = nbsp) %>%
+    select(-sing.sp) %>%
     group_by(year)
@@
   means <- data %>%
-    select(-plot_id, -richness) %>%
+    select(-rarefyid, -richness) %>%
     summarise(across(everything(), mean)) %>%
@@
   sds <- data %>%
-    select(-plot_id, -richness) %>%
+    select(-rarefyid, -richness) %>%
     summarise(across(everything(), sd)) %>%
@@
   stats <- left_join(means, sds, by = c("year", "metric")) %>%
-    left_join(data %>% select(year, plot_id, richness) %>% distinct(), by = "year") %>%
+    left_join(data %>% select(year, rarefyid, rarefy_resamp, richness) %>% distinct(), by = "year") %>%
     mutate(se = sd/sqrt(n),
@@
            upperCI = mean + qt(0.975, n - 1) * se)  %>%
-    select(year, plot_id, everything())
+    select(year, rarefyid, everything())
 
@@
 plan(multiprocess)
-null_table <- future_map_dfr(null_files, process_nulls, n = n)
+null_table <- future_map_dfr(null_files, process_nulls, n = 500)
 

### 2021-03-24 analysis/03_collate_rarefied_resamps_median.R

@@ -3,6 +3,7 @@
 ##	for analysis
 library(dplyr)
 library(tibble)
+library(stringr)
 ##============================================================
 ##	set the pattern to load the files to be compiled
 path <- here::here("data", "rarefied_metrics")
@@ -17,18 +18,23 @@ path <- here::here("data", "rarefied_metrics", "fd")
 filelist <-  dir(path, "*.rda") %>%
   paste0(path, "/", .)
 
+# fd_metrics <- purrr::map_dfr(filelist, function(x) {
+#   load(x)
+#   split_path <- str_split(x, "/") %>%
+#     unlist(.) %>%
+#     last() %>%
+#     str_split(., "_") %>%
+#     unlist()
+#
+#   rarefyID <- paste0(str_replace(split_path[2], "cell", ""), "_", split_path[3])
+#
+#   biochange_metrics <- biochange_metrics %>%
+#     mutate(rarefyID = rarefyID)
+#   return(biochange_metrics)
+# }) %>% distinct()
+
 fd_metrics <- purrr::map_dfr(filelist, function(x) {
   load(x)
-  split_path <- str_split(x, "/") %>%
-    unlist(.) %>%
-    last() %>%
-    str_split(., "_") %>%
-    unlist()
-
-  rarefyID <- paste0(str_replace(split_path[2], "cell", ""), "_", split_path[3])
-
-  biochange_metrics <- biochange_metrics %>%
-    mutate(rarefyID = rarefyID)
   return(biochange_metrics)
 }) %>% distinct()
 

### 2021-03-25 analysis/03_collate_rarefied_resamps_median.R
Remove old function

@@ -18,21 +18,6 @@ path <- here::here("data", "rarefied_metrics", "fd")
 filelist <-  dir(path, "*.rda") %>%
   paste0(path, "/", .)
 
-# fd_metrics <- purrr::map_dfr(filelist, function(x) {
-#   load(x)
-#   split_path <- str_split(x, "/") %>%
-#     unlist(.) %>%
-#     last() %>%
-#     str_split(., "_") %>%
-#     unlist()
-#
-#   rarefyID <- paste0(str_replace(split_path[2], "cell", ""), "_", split_path[3])
-#
-#   biochange_metrics <- biochange_metrics %>%
-#     mutate(rarefyID = rarefyID)
-#   return(biochange_metrics)
-# }) %>% distinct()
-
 fd_metrics <- purrr::map_dfr(filelist, function(x) {
   load(x)
   return(biochange_metrics)

### 2021-03-25 analysis/invasive_data.Rmd
Commit

---
title: "GBIF invasive species"
author: "Kari Norman"
date: "3/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rgbif)
library(geonames)
```

Get GBIF data
```{r}
#get all the datasets from the Invasive Species Specialist Group
datasets <- dataset_search(publishingOrg = "cdef28b1-db4e-4c58-aa71-3c5238c2d0b5", limit = 300)
datasets_list <- datasets$data

occ <- occ_download(dataset = "b351a324-77c4-41c9-a909-f30f77268bc4")
```

Figure out which country each study is located in 
```{r}
pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
options(geonamesUsername="getafreeusernamefromgeonames.org")
options(gbif_u)

obs_data <- pins::pin_get("bt-traitfiltered", board = "github") %>%
  select(study_id, species) %>%
  distinct()

meta <- pins::pin_get("meta", board = "github") %>%
  select(study_id, cent_lat, cent_long) %>%
  filter(study_id %in% obs_data$study_id)

get_GNcountry <- possibly(GNcountryCode, otherwise = NA_character_)

get_country <- function(study_id, cent_lat, cent_long){
  con_data <- get_GNcountry(lat = cent_lat, lng = cent_long, lang = "en", radius = 50)
  name <- ifelse(length(con_data) == 1, con_data, con_data$countryName)

  return(tibble(study_id = study_id, country = name))
}

country_data <- pmap_dfr(meta, get_country) %>%
  #hand resolve study_ids in the ocean
  mutate(country = case_when(
  study_id == 108 ~ "Namibia", 
  study_id %in% c(166, 169) ~ "United States", 
  study_id == 172 ~ "Morocco", 
  TRUE ~ country
))

obs_data <- obs_data %>%
  left_join(country_data)
```

Get datasets relevant for observed studies
```{r}
countries <- unique(country_data$country)
datasets_list %>% filter(str_detect(datasetTitle, countries)) %>% View()

```

### 2021-03-26 analysis/03_collate_rarefied_resamps_median.R
Update collation to incorporate null models

@@ -5,7 +5,8 @@ library(dplyr)
 library(tibble)
 library(stringr)
 ##============================================================
-##	set the pattern to load the files to be compiled
+
+##	load species based metrics
 path <- here::here("data", "rarefied_metrics")
 filelist <-  dir(path, "*.rda") %>%
   paste0(path, "/", .)
@@ -14,18 +15,29 @@ species_metrics <- purrr::map_dfr(filelist, function(x) {load(x)
   return(rarefied_metrics)
 	}) %>% distinct()
 
+## load fd based metrics
 path <- here::here("data", "rarefied_metrics", "fd")
 filelist <-  dir(path, "*.rda") %>%
   paste0(path, "/", .)
 
 fd_metrics <- purrr::map_dfr(filelist, function(x) {
-  load(x)
-  return(biochange_metrics)
-}) %>% distinct()
+    load(x)
+    rarefyID <- str_match(x,"_cell\\s*(.*?)\\s*_sample")[,2]
 
+    biochange_metrics <- biochange_metrics %>%
+      mutate(rarefyID = rarefyID)
+    return(biochange_metrics)
+  }) %>% distinct()
+
+## get null models
+null_table <- pins::pin_get("null-table", board = "github")
+
+### Merge rarefied species and fd metrics and null models
 ##	put them all together
-rarefied_metrics <- full_join(species_metrics, fd_metrics) %>%
+rarefied_metrics <- full_join(species_metrics, fd_metrics,
+                              c("YEAR", "cell", "rarefy_resamp", "rarefyID")) %>%
   filter(S != 1) # remove samples that had one species as they result in strange metric values
+
 
 ##	pull out new metadata
 new_meta <- rarefied_metrics %>%
@@ -33,26 +45,41 @@ new_meta <- rarefied_metrics %>%
 
 ##	calculate the medians for all the metrics for studies that were rarefied
 metric_cols <- ungroup(rarefied_metrics) %>%
-  select(-c(SamplePool, SampleN, num_years, duration, startYear, endYear, rarefy_resamp, CWM))
+  select(-c(SamplePool, SampleN, num_years, duration, startYear, endYear, CWM, type)) %>%
+  pivot_longer(-c(rarefyID, rarefied, YEAR, cell, rarefy_resamp), names_to = "metric") %>%
[…]

### 2021-04-06 analysis/05_model_metrics.Rmd
incorporate null model samples

--- 2021-03-12 analysis/05_model_metrics.Rmd
@@
 
-metrics <- pins::pin_get("rarefied-medians", board = "github") %>%
-  pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
+metrics <- pins::pin_get("rarefied-metrics", board = "github") %>%
+  # pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
   rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
-  mutate(logvalue = log(value + abs(floor(min(na.omit(value))))),
-         year_scaled = scale(year)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
-  filter(!is.na(value)) 
+  filter(!is.na(value)) %>%
+  mutate(logvalue = log(value + 1),
+         year_scaled = scale(year)) # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
+
 ```
@@
 metrics %>%
-  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
+  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "SES_FDis", "SES_FDiv", "SES_FEve", "SES_FRic")) %>%
   ggplot() +
@@
 ```{r}
+#find rarefyID's that have inf value for some metrics and exclude
+cont_inf <- metrics %>% filter(is.infinite(value), !metric %in% c("qual.FRic",  "SES_qual.FRic")) %>% pull(rarefyID)
+
+#get rarefyID's that don't have enough null samples
+missing_null <- bind_rows(metrics %>% 
+  filter(n_missing_nulls == 1, commplete_null_samps == FALSE),
+  metrics %>% filter(n_missing_nulls > 100)) %>%
+  pull(rarefyID) %>%
+  unique()
+
+model_data <- metrics %>% 
+  filter(!rarefyID %in% cont_inf, !rarefyID %in% missing_null) %>%
+  mutate(study_id = substr(rarefyID, 1, 3))
+
 # model for overall trend (random effect w/varying intercept)
-metric_mean_fit <- metrics %>%
-  filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S_int")) %>%
+# metric_mean_fit <- model_data %>%
+#   filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
+#   group_nest(metric) %>%
+#   mutate(model = map(data, ~lmer(log(value + 1) ~ year_scaled + (1|study_id/rarefyID), data = .x, REML = FALSE,
+#                   control = lmerControl(optCtrl = list(maxit = 1e9,
+#                                                     maxfun = 1e9,
+#                                                     xtol_abs = 1e-11,
+#                                                     ftol_abs = 1e-11,
+#                                                     maxeval = 1e9)))),
+#          null_model = map(data, ~lmer(logvalue ~ (1|study_id/rarefyID), data = .x, REML = FALSE,
+#                   control = lmerControl(optCtrl = list(maxit = 1e9,
+#                                                     maxfun = 1e9,
+#                                                     xtol_abs = 1e-11,
+#                                                     ftol_abs = 1e-11,
+#                                                     maxeval = 1e9)))),
+#            
+#            test = map2(model, null_model, anova),
+#            test_res = map(test, tidy),
+#            coef = map(model, tidy)
+#            ) %>% 
+#   unnest(cols = c(test_res)) %>%
+#   select(metric, model, coef, p.value) %>%
+#   filter(!is.na(p.value)) %>%
+#   unnest(cols = c(coef)) %>%
+#   filter(term == "year_scaled") %>%
+#   select(-c(effect, group, term, statistic))
+
+SES_rand_slope <- model_data %>%
+  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis")) %>%
   group_nest(metric) %>%
-  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (1|study_id/rarefyID), data = .x,
-                  control = lmerControl(optCtrl = list(maxit = 1e9,
-                                                    maxfun = 1e9,
-                                                    xtol_abs = 1e-11,
-                                                    ftol_abs = 1e-11,
-                                                    maxeval = 1e9)))),
-         null_model = map(data, ~lmer(logvalue ~ (1|study_id/rarefyID), data = .x,
-                  control = lmerControl(optCtrl = list(maxit = 1e9,
-                                                    maxfun = 1e9,
-                                                    xtol_abs = 1e-11,
-                                                    ftol_abs = 1e-11,
-                                                    maxeval = 1e9)))),
-
-           test = map2(model, null_model, anova),
-           test_res = map(test, tidy),
-           coef = map(model, tidy)
-           ) %>% 
-  unnest(cols = c(test_res)) %>%
-  select(metric, model, coef, p.value) %>%
-  filter(!is.na(p.value)) %>%
-  unnest(cols = c(coef)) %>%
-  filter(term == "year_scaled") %>%
-  select(-c(effect, group, term, statistic))
-
-metric_mean_fit_slope <- metrics %>%
-  filter(metric %in%  c("FRic", "FEve", "FDiv", "S_int")) %>%
-  group_nest(metric) %>%
-  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
+  mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                  data = .x)),
-           null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
+           null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
                                  data = .x)),
@@
 
-model_comp <- metric_mean_fit %>%
-  select(metric, no_slope_estimate = estimate, no_slope_pval = p.value) %>%
-  left_join(metric_mean_fit_slope %>% 
-  select(metric, slope_estimate = estimate, slope_pval = p.value ))
+metric_rand_slope <- model_data %>%
+  filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
+  group_nest(metric) %>%
+  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
+                                 data = .x, REML = FALSE)),
+           null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
+                                 data = .x, REML = FALSE)),
+           test = map2(model, null_model, anova),
+           test_res = map(test, tidy),
+           coef = map(model, tidy)
+           ) %>% 
+  unnest(cols = c(test_res)) %>%
+  select(metric, model, coef, p.value) %>%
+  filter(!is.na(p.value)) %>%
+  unnest(cols = c(coef)) %>%
+  filter(term == "year_scaled") %>%
+  select(-c(effect, group, term, statistic))
+
+# model_comp <- metric_mean_fit %>%
+#   select(metric, no_slope_estimate = estimate, no_slope_pval = p.value) %>%
+#   left_join(metric_mean_fit_slope %>% 
+#   select(metric, slope_estimate = estimate, slope_pval = p.value ))
+
+metric_model_table <- bind_rows(select(SES_rand_slope, -model), select(metric_rand_slope, -model))
+```
+
+Kitchen sink model
+```{r}
+all_model <- model_data %>%
+  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis", "S")) %>%
+  group_nest(metric) %>%
+  mutate(model = map(data, ~lmer(value ~ year_scaled + taxa + realm + (year_scaled|study_id/rarefyID), 
+                                 data = .x)))#,
+  #          null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
+  #                                data = .x)),
+  #          test = map2(model, null_model, anova),
+  #          test_res = map(test, tidy),
+  #          coef = map(model, tidy)
+  #          ) %>% 
+  # unnest(cols = c(test_res)) %>%
+  # select(metric, model, coef, p.value) %>%
+  # filter(!is.na(p.value)) %>%
+  # unnest(cols = c(coef)) %>%
+  # filter(term == "year_scaled") %>%
+  # select(-c(effect, group, term, statistic))
+
 ```
@@
 
+Get study level slop estimates
 ```{r}
+get_blup <- function(metric, model){
+  coef(model)$study_id %>%
+    rownames_to_column("study_id") %>%
+    mutate(metric = metric)
+}
 
+blup <- pmap_dfr(bind_rows(select(SES_rand_slope, metric, model), select(metric_rand_slope,  metric, model)), get_blup) %>%
+  select(-"(Intercept)") %>%
+  pivot_wider(names_from = "metric", values_from = "year_scaled")
+
+#pairwise plots using GGally
+ggpairs(blup %>% 
+          select(-c("study_id", "FDis", "FDiv", "FEve", "FRic")))
+
+#pairwise plots using GGPlot
+gatherpairs <- function(data, ..., 
+                        xkey = '.xkey', xvalue = '.xvalue',
+                        ykey = '.ykey', yvalue = '.yvalue',
+                        na.rm = FALSE, convert = FALSE, factor_key = FALSE) {
+  vars <- quos(...)
+  xkey <- enquo(xkey)
+  xvalue <- enquo(xvalue)
+  ykey <- enquo(ykey)
+  yvalue <- enquo(yvalue)
+
+  data %>% {
+    cbind(gather(., key = !!xkey, value = !!xvalue, !!!vars,
+                 na.rm = na.rm, convert = convert, factor_key = factor_key),
+          select(., !!!vars)) 
+  } %>% gather(., key = !!ykey, value = !!yvalue, !!!vars,
+               na.rm = na.rm, convert = convert, factor_key = factor_key)
+}
+
+blup %>% 
+  gatherpairs("SES_FDis", "SES_FDiv", "SES_FEve", "SES_FRic", "S") %>% {
+  ggplot(., aes(x = .xvalue, y = .yvalue)) +
+      geom_point() + 
+      #geom_smooth(method = 'lm') +
+      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
+      scale_color_brewer(type = 'qual')
+  }
+
+blup %>%
+  select(-c("FDis", "FDiv", "FEve", "FRic")) %>%
+  pivot_longer(-one_of("study_id", "S"), names_to = "metric") %>%
+  ggplot(aes(x = S, y = value)) +
+  geom_point() +
+  facet_wrap(~metric) +
+  xlim(-1.3, 1.3) +
+  ylim(-0.6, 0.6) +
+    theme_classic() +
+    geom_hline(yintercept = 0) +
+    geom_vline(xintercept = 0) + 
+  ylab("Metric Slope") +
+  xlab("Richness Slope")
+
+blup %>%
+  select(-c("FDis", "FDiv", "FEve", "FRic")) %>%
+  pivot_longer(-one_of("study_id", "SES_FRic"), names_to = "metric") %>%
+  ggplot(aes(x = SES_FRic, y = value)) +
+  geom_point() +
+  facet_wrap(~metric) +
+  xlim(-.25, .25) +
+  ylim(-0.6, 0.6) +
+    theme_classic() +
+    geom_hline(yintercept = 0) +
+    geom_vline(xintercept = 0) + 
+  ylab("Metric Slope") +
+  xlab("Functional Richness Slope")
 ```

### 2021-04-06 analysis/03_collate_rarefied_resamps_median.R

@@ -48,7 +48,22 @@ metric_cols <- ungroup(rarefied_metrics) %>%
   select(-c(SamplePool, SampleN, num_years, duration, startYear, endYear, CWM, type)) %>%
   pivot_longer(-c(rarefyID, rarefied, YEAR, cell, rarefy_resamp), names_to = "metric") %>%
   left_join(null_table %>% select(-c(richness, se, lowerCI, upperCI)),
-            by = c("YEAR" = "year", "rarefyID" = "rarefyid", "rarefy_resamp" = "rarefy_resamp", "metric" = "metric")) %>%
+            by = c("YEAR" = "year", "rarefyID" = "rarefyid", "rarefy_resamp" = "rarefy_resamp", "metric" = "metric"))
+
+# get a dataframe of rarefyID's and rarefication samples that we don't have null_model for, and a count of the number of resamples that are missing
+no_null <- metric_cols %>%
+  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis"), is.na(mean)) %>%
+  select(rarefyID, YEAR, rarefy_resamp) %>%
+  distinct()
+
+missing_nulls <- null_table %>%
+  right_join(no_null, by = c("rarefyid" = "rarefyID", "year" = "YEAR", "rarefy_resamp")) %>%
+  group_by(rarefyid) %>%
+  summarise(n_missing_nulls = n_distinct(rarefy_resamp)) %>%
+  mutate(commplete_null_samps = "FALSE")
+
+# calculate SES and incorporate into metric columns
+metric_cols <- metric_cols %>%
   mutate(SES = (value - mean)/sd) %>%
   select(-c(mean, sd)) %>%
   #the next three lines go back to wide format for both SES and unadjusted values, then to long format incorporating SES values
@@ -78,7 +93,8 @@ rarefied_medians <- metric_cols %>%
 #   ungroup()
 
 ##	recombine with new metadata
-rarefied_metrics <- inner_join(new_meta, rarefied_medians, by='rarefyID') #%>%
+rarefied_metrics <- inner_join(new_meta, rarefied_medians, by='rarefyID') %>%
+  left_join(missing_nulls, by = c("rarefyID" = "rarefyid"))
   #inner_join(rarefied_ints)
 
 ##	save

### 2021-04-06 analysis/invasive_data.Rmd
get invasive data

@@
 pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
-options(geonamesUsername="getafreeusernamefromgeonames.org")
-options(gbif_u)
+options(geonamesUsername="karinorman")
 
@@
 
+Read in Invasive Data
+```{r}
+invasive_files <- list.dirs(here::here("data", "invasive"))
+
+taxon <- read_tsv(paste0(invasive_files[2], "/", "taxon.txt"))
+profile <- read_tsv(paste0(invasive_files[2], "/", "speciesprofile.txt"))
+dist <- read_tsv(paste0(invasive_files[2], "/", "distribution.txt")) %>%
+  select(id, countryCode)
+
+data <- left_join(taxon, profile) %>% left_join(dist)
+
+```
+
 Get datasets relevant for observed studies

### 2021-04-08 analysis/03_rarefy_null_models.Rmd
Reordering of scripts to add null model script

--- 2021-03-24 analysis/rarefy_null_models.Rmd
@@
 ```
+
 Figure out which samples didn't get null samples
@@
 
+  plot_samples %>% select(rarefyid, rarefy_resamp) %>% distinct() %>% print()
+
   sample_species_mat <- plot_samples %>%
@@
 
-  print(head(sample_species_mat))
+  #print(head(sample_species_mat))
   trait_mat <- get_traitMat(colnames(sample_species_mat), trait_data = traits)
@@
 
+Figure out which nulls still need to be calculated
+```{r}
+null_files <- dir(here::here("data", "null_samples"), "*.csv") 
+null_files <- as.data.frame(null_files) %>%
+  mutate(rarefyid = str_match(null_files, "(.*?)\\s*_samp")[,2],
+         rarefy_samp = str_match(null_files, "samp\\s*(.*?)\\s*.csv")[,2])
+
+fd_files <- dir(here::here("data", "fd_null_output"), "*.tsv")
+fd_files <- as.data.frame(fd_files) %>%
+  mutate(rarefyid = str_match(fd_files, "null_rarefy\\s*(.*?)\\s*samp")[,2],
+         rarefy_samp = str_match(fd_files, "samp\\s*(.*?)\\s*.tsv")[,2])
+
+missing_files <- null_files %>%
+  left_join(fd_files) %>% 
+  filter(is.na(fd_files)) %>%
+  pull(null_files) %>%
+  paste0(here::here("data", "null_samples/"), .)
+
+get_nullplot_FD_safe <- possibly(get_nullplot_FD, otherwise = data_frame())
+plan(multiprocess, workers = 120)
+null_FD <- future_map_dfr(missing_files, ~get_nullplot_FD_safe(.x, traits = trait_ref),
+                           .options = furrr_options(seed = TRUE))
+```
+
 Create Null table
@@
     rename(richness = nbsp) %>%
-    select(-sing.sp) %>%
+    select(-sing.sp, -n) %>%
     group_by(year)
@@
   means <- data %>%
-    select(-rarefyid, -richness) %>%
-    summarise(across(everything(), mean)) %>%
+    select(-rarefyid, -richness, -rarefy_resamp) %>%
+    summarise(across(everything(), ~mean(.x, na.rm = TRUE))) %>%
     pivot_longer(cols = -year, names_to = "metric", values_to = "mean")
@@
   sds <- data %>%
-    select(-rarefyid, -richness) %>%
-    summarise(across(everything(), sd)) %>%
+    select(-rarefyid, -richness, -rarefy_resamp) %>%
+    summarise(across(everything(), ~sd(.x, na.rm = TRUE))) %>%
     pivot_longer(cols = -year, names_to = "metric", values_to = "sd")
@@
            upperCI = mean + qt(0.975, n - 1) * se)  %>%
-    select(year, rarefyid, everything())
+    select(year, rarefyid, rarefy_resamp, everything())
 
@@
 
-plan(multiprocess)
+plan(multiprocess, workers = 120)
 null_table <- future_map_dfr(null_files, process_nulls, n = 500)

### 2021-04-08 analysis/05_model_metrics.Rmd

@@
 
+#get metadata
+meta <- pins::pin_get("meta", board = "github") %>%
+  select(study_id, realm, climate, habitat, protected_area, biome_map)
+
+meta <- pins::pin_get("bt-traitfiltered", board = "github") %>%
+  select(study_id, rarefyid, taxa) %>% distinct() %>%
+  left_join(meta)
+
+#get metric data
 metrics <- pins::pin_get("rarefied-metrics", board = "github") %>%
@@
   filter(!is.na(value)) %>%
-  mutate(logvalue = log(value + 1),
-         year_scaled = scale(year)) # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
+  mutate(logvalue = log(value + 1)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
+  left_join(meta, c("rarefyID" = "rarefyid"))
+```
 
-```
 ####Data exploration
@@
 ```{r}
-no_fd_years <- metrics %>%
-  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis"), is.na(value))
-
-dim(no_fd_years)
-
-metrics <- setdiff(metrics, no_fd_years)
+# no_fd_years <- metrics %>%
+#   filter(metric %in% c("FRic", "FEve", "FDiv", "FDis"), is.na(value))
+# 
+# dim(no_fd_years)
+# 
+# metrics <- setdiff(metrics, no_fd_years)
 ```
@@
   filter(!rarefyID %in% cont_inf, !rarefyID %in% missing_null) %>%
-  mutate(study_id = substr(rarefyID, 1, 3))
+  mutate(study_id =  str_extract(rarefyID, "[^_]+")) %>%
+  mutate(year_scaled = scale(year), 
+         scale_center = attributes(year_scaled)$`scaled:center`, 
+         scale = attributes(year_scaled)$`scaled:scale`)
 
@@
 SES_rand_slope <- model_data %>%
-  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis")) %>%
+  #filter(protected_area == "FALSE") %>%
+  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis")) %>%
+  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
   group_nest(metric) %>%
@@
 metric_rand_slope <- model_data %>%
-  filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
+  #filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
+  filter(metric %in%  c("FRic", "FEve", "FDiv", "S")) %>%
   group_nest(metric) %>%
@@
 all_model <- model_data %>%
-  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis", "S")) %>%
+  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis")) %>%
   group_nest(metric) %>%
   mutate(model = map(data, ~lmer(value ~ year_scaled + taxa + realm + (year_scaled|study_id/rarefyID), 
-                                 data = .x)))#,
-  #          null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
-  #                                data = .x)),
-  #          test = map2(model, null_model, anova),
-  #          test_res = map(test, tidy),
-  #          coef = map(model, tidy)
-  #          ) %>% 
-  # unnest(cols = c(test_res)) %>%
-  # select(metric, model, coef, p.value) %>%
-  # filter(!is.na(p.value)) %>%
-  # unnest(cols = c(coef)) %>%
-  # filter(term == "year_scaled") %>%
-  # select(-c(effect, group, term, statistic))
+                                 data = .x)),
+           null_model = map(data, ~lmer(value ~ taxa + realm + (year_scaled|study_id/rarefyID),
+                                 data = .x)),
+           test = map2(model, null_model, anova),
+           test_res = map(test, tidy),
+           coef = map(model, tidy)
+           ) %>%
+  unnest(cols = c(test_res)) %>%
+  select(metric, model, coef, p.value) %>%
+  filter(!is.na(p.value)) %>%
+  unnest(cols = c(coef)) %>%
+  filter(term == "year_scaled") %>%
+  select(-c(effect, group, term, statistic))
 
@@
 ```
+
+Clustering 
+```{r}
+scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)
+
+blup_mat <- blup %>%  
+  #these studies have NA's for some reason, remove them 
+  filter(!study_id %in% c(445, 446, 348, 382)) %>%
+  select(study_id, starts_with("SES"), S) %>%
+  column_to_rownames("study_id") %>%
+  mutate(across(.fns = scale2)) %>%
+  as.matrix()
+
+clmethods <- c("hierarchical","kmeans","pam")
+intern <- clValid(blup_mat, nClust = 2:6, clMethods = clmethods, validation = "internal")
+stab <- clValid(blup_mat, nClust = 2:6, clMethods = clmethods, validation = "stability")
+
+summary(intern)
+optimalScores(stab)
+
+#get the hierarchical clustering object
+clust <- intern@clusterObjs$hierarchical
+cat5 <- cutree(clust, k = 5)
+cat3 <- cutree(clust, k = 3)
+
+blup_clust <- blup %>% 
+  left_join(as.data.frame(cat3) %>%
+  rownames_to_column("study_id")) %>%
+  left_join(as.data.frame(cat5) %>%
+  rownames_to_column("study_id"))
+```

### 2021-04-08 analysis/04_collate_rarefied_resamps_median.R
cleanup final columns

@@ -36,7 +36,8 @@ null_table <- pins::pin_get("null-table", board = "github")
 ##	put them all together
 rarefied_metrics <- full_join(species_metrics, fd_metrics,
                               c("YEAR", "cell", "rarefy_resamp", "rarefyID")) %>%
-  filter(S != 1) # remove samples that had one species as they result in strange metric values
+  filter(S != 1) %>% # remove samples that had one species as they result in strange metric values
+  select(-qual.FRic)
 
 
 ##	pull out new metadata
@@ -53,11 +54,11 @@ metric_cols <- ungroup(rarefied_metrics) %>%
 # get a dataframe of rarefyID's and rarefication samples that we don't have null_model for, and a count of the number of resamples that are missing
 no_null <- metric_cols %>%
   filter(metric %in% c("FRic", "FEve", "FDiv", "FDis"), is.na(mean)) %>%
-  select(rarefyID, YEAR, rarefy_resamp) %>%
+  select(rarefyID, rarefy_resamp) %>%
   distinct()
 
 missing_nulls <- null_table %>%
-  right_join(no_null, by = c("rarefyid" = "rarefyID", "year" = "YEAR", "rarefy_resamp")) %>%
+  right_join(no_null, by = c("rarefyid" = "rarefyID","rarefy_resamp")) %>%
   group_by(rarefyid) %>%
   summarise(n_missing_nulls = n_distinct(rarefy_resamp)) %>%
   mutate(commplete_null_samps = "FALSE")

### 2021-04-08 analysis/05_model_metrics.Rmd
Scale year by metric

@@
   filter(!rarefyID %in% cont_inf, !rarefyID %in% missing_null) %>%
+  group_by(metric) %>%
   mutate(study_id =  str_extract(rarefyID, "[^_]+")) %>%
@@
          scale_center = attributes(year_scaled)$`scaled:center`, 
-         scale = attributes(year_scaled)$`scaled:scale`)
+         scale = attributes(year_scaled)$`scaled:scale`) %>%
+  ungroup()
 
@@
   select(-c(effect, group, term, statistic))
-
 ```

### 2021-04-09 analysis/invasive_data.Rmd
Function to read in invasive species data

@@
 library(geonames)
+library(countrycode)
+library(taxadb)
 ```
@@
 get_country <- function(study_id, cent_lat, cent_long){
-  con_data <- get_GNcountry(lat = cent_lat, lng = cent_long, lang = "en", radius = 50)
+  #con_data <- get_GNcountry(lat = cent_lat, lng = cent_long, lang = "en", radius = 50)
+  con_data <- get_GNcountry(lat = cent_lat, lng = cent_long, radius = 50)
   name <- ifelse(length(con_data) == 1, con_data, con_data$countryName)
@@
   study_id == 108 ~ "Namibia", 
-  study_id %in% c(166, 169) ~ "United States", 
+  study_id %in% c(56, 166, 169) ~ "United States", 
   study_id == 172 ~ "Morocco", 
   TRUE ~ country
+)) %>%
+  mutate(code = countryname(country, destination = "genc2c")) %>%
+  mutate(code = case_when(
+    code == "CA" ~ "US", 
+    code == "PR" ~ "DO",
+    TRUE ~ code
   ))
@@
 
-taxon <- read_tsv(paste0(invasive_files[2], "/", "taxon.txt"))
-profile <- read_tsv(paste0(invasive_files[2], "/", "speciesprofile.txt"))
-dist <- read_tsv(paste0(invasive_files[2], "/", "distribution.txt")) %>%
+invasive_data <- map_dfr(invasive_files[2:21], function(x){
+  taxon <- read_tsv(paste0(x, "/", "taxon.txt"))
+  profile <- read_tsv(paste0(x, "/", "speciesprofile.txt"))
+  dist <- read_tsv(paste0(x, "/", "distribution.txt")) %>%
     select(id, countryCode)
@@
   data <- left_join(taxon, profile) %>% left_join(dist)
+  return(data)
+})
+```
 
-```
+Substituting years:
+- orig: Canada, data: United States
+- orig: Puerto Rico, data: Dominican Republic
 

### 2021-04-09 analysis/problem_model.Rmd
Clean up data to match model_metrics

@@
 
-metrics <- pins::pin_get("rarefied-medians", board = "github") %>%
-  select(-cYEAR) %>%
-  pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
+metrics <- pins::pin_get("rarefied-metrics", board = "github") %>%
+  # pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
   rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
-  mutate(logvalue = log(value + abs(floor(min(na.omit(value)))))) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
-  filter(!is.na(value))
+  filter(!is.na(value)) %>%
+  mutate(logvalue = log(value + 1),
+         year_scaled = scale(year)) # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
+
+cont_inf <- metrics %>% filter(is.infinite(value), !metric %in% c("qual.FRic",  "SES_qual.FRic")) %>% pull(rarefyID)
+
+#get rarefyID's that don't have enough null samples
+missing_null <- bind_rows(metrics %>% 
+  filter(n_missing_nulls == 1, commplete_null_samps == FALSE),
+  metrics %>% filter(n_missing_nulls > 100)) %>%
+  pull(rarefyID) %>%
+  unique()
+
+model_data <- metrics %>% 
+  filter(!rarefyID %in% cont_inf, !rarefyID %in% missing_null) %>%
+  mutate(study_id = substr(rarefyID, 1, 3))
 ```
@@
 ```{r}
-metrics %>%
+model_data %>%
   filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S_int")) %>%
@@
 
-metrics %>%
+model_data %>%
   filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S_int")) %>%
@@
 ```{r}
-test_data <- metrics %>% 
-  filter(metric == "FRic") %>%
-  mutate(study_id = as.factor(study_id),
-         year_center = scale(year),
-         value_scale = scale(value))
+test_data <- model_data %>% 
+  filter(metric == "FEve") %>%
+  mutate(study_id = as.factor(study_id))
 
 #lognormal
-ln_fit <- lmer(value_scale ~ year_center + (year_center|study_id/rarefyID), data = test_data,
+ln_fit <- lmer(logvalue ~ year_center + (year_center|study_id/rarefyID), data = test_data,
                control = lmerControl(optCtrl = list(maxit = 1e9,
@@
 ```
+
+```{r}
+test_data <- model_data %>% 
+  filter(metric == "SES_FEve") %>%
+  mutate(study_id = as.factor(study_id))
+
+all_model <- lmer(value ~ year_scaled + taxa + realm + (year_scaled|study_id/rarefyID), 
+                                 data = test_data)
+```

### 2021-04-13 analysis/func_distinctiveness.Rmd
First commit

---
title: "Functional Rarity"
output: html_document
---
```{r}
library(funrar)
library(biodivTS)
```

Calculate functional distinctiveness for each species based on the species pool (defined by all species observed over the course of the timeseries)

Get data
```{r}
pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")

##	Get the gridded data locally
bt_data <- pins::pin_get("bt-traitfiltered", board = "github") %>%
  rename_with(tolower)

trait_ref <- pins::pin_get("trait-ref", board = "github")

bt_abund <- bt_data %>%
  filter(abundance_type %in% c("Count", "Density", "MeanCount"))

bt_pres <- bt_data %>%
  filter(abundance_type == "Presence/Absence")
```

```{r}
get_ts_distinct <- function(data, trait_data, ts_id){
  species_data <- data %>% 
    filter(rarefyid == ts_id) %>%
    select(species, abundance, year) %>%
    pivot_wider(names_from = "species", values_from = "abundance", values_fill = 0) %>%
    column_to_rownames("year")

  rel_species <- make_relative(as.matrix(species))
  # traits <- trait_data %>% 
  #   filter(Species %in% unique(species$species)) %>% 
  #   select_if(~length(unique(na.omit(.))) > 1) %>%
  #   column_to_rownames("Species")
  trait_dist <- get_traitMat(colnames(species_data), trait_data)
  #bin_vars <- map(trait_mat, ~all(na.omit(.) %in% 0:1))

  trait_dist <- trait_dist %>%
    mutate_if(is.character, as.factor) %>%
    compute_dist_matrix(., metric = "gower")

  distinctive <- as.data.frame(distinctiveness(pres_matrix = rel_species, dist_matrix = trait_dist)) %>%
    rownames_to_column("year") %>%
    mutate(rarefyid = ts_id)

  return(distinctive)
}

get_ts_distinct(bt_data, trait_ref, "39_407478")

path <- here::here("data", "rarefied_samples")
files <- dir(path, "*.rda") %>%
  paste0(path, "/", .)

plan("multiprocess", workers = 30)
furrr::future_map(files, get_ts_distinct, traits = trait_ref)
```

### 2021-04-14 analysis/05_model_metrics.Rmd
Clean up plotting, update model output processing to reflect lmerTest output

@@
 library(broom.mixed)
+library(lmerTest)
 ```
@@
 metrics %>%
-  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
+  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_hind")) %>%
   ggplot() +
@@
 metrics %>%
-  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "SES_FDis", "SES_FDiv", "SES_FEve", "SES_FRic")) %>%
+  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "SES_FDis", "SES_FDiv", "SES_FEve", "SES_FRic", "Jaccard_hind")) %>%
   ggplot() +
@@
   #filter(protected_area == "FALSE") %>%
-  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis")) %>%
-  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
+  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis", )) %>%
+  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "Jaccard_next")) %>%
   group_nest(metric) %>%
@@
                                  data = .x)),
-           null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
-                                 data = .x)),
-           test = map2(model, null_model, anova),
-           test_res = map(test, tidy),
+           # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
+           #                       data = .x)),
+           # test = map2(model, null_model, anova),
+           # test_res = map(test, tidy),
            coef = map(model, tidy)
            ) %>% 
-  unnest(cols = c(test_res)) %>%
-  select(metric, model, coef, p.value) %>%
-  filter(!is.na(p.value)) %>%
+  select(metric, model, coef) %>%
   unnest(cols = c(coef)) %>%
@@
                                  data = .x, REML = FALSE)),
-           null_model = map(data, ~lmer(logvalue ~ (year_scaled|study_id/rarefyID), 
-                                 data = .x, REML = FALSE)),
-           test = map2(model, null_model, anova),
-           test_res = map(test, tidy),
            coef = map(model, tidy)
            ) %>% 
-  unnest(cols = c(test_res)) %>%
-  select(metric, model, coef, p.value) %>%
-  filter(!is.na(p.value)) %>%
+  select(metric, model, coef) %>%
   unnest(cols = c(coef)) %>%
@@
 ```
+Models with categorical study characteristics as random effects
 
+Taxa:
+```{r}
+taxa_slope_model <- model_data %>%
+  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
+  group_nest(metric) %>%
+  mutate(model = map(data, ~lmer(value ~ year_scaled + taxa + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
+                                 data = .x, REML = TRUE)),
+           coef = map(model, tidy)) %>%
+  select(metric, coef) %>%
+  unnest(cols = c(coef)) %>%
+  #filter(term == "year_scaled") %>%
+  filter(term %in% c("year_scaled", "taxaBirds", "taxaMammals")) %>%
+  select(-c(effect, group, statistic))
+
+taxa_slope_model_S <- model_data %>% 
+  filter(metric == "S") %>%
+  lmer(logvalue ~ year_scaled + taxa + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
+                                 data = ., REML = TRUE)
+```
+Realm:
+```{r}
+realm_slope_model <- model_data %>%
+  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
+  group_nest(metric) %>%
+  mutate(model = map(data, ~lmer(value ~ year_scaled + realm + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
+                                 data = .x, REML = TRUE)),
+           coef = map(model, tidy)) %>%
+  select(metric, coef) %>%
+  unnest(cols = c(coef)) %>%
+  #filter(term == "year_scaled") %>%
+  filter(effect == "fixed") %>%
+  select(-c(effect, group, statistic))
+
+realm_slope_model_S <- model_data %>% 
+  filter(metric == "S") %>%
+  lmer(logvalue ~ year_scaled + realm + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
+                                 data = ., REML = TRUE)
+```
+Climate:
+```{r}
+climate_slope_model <- model_data %>%
+  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
+  group_nest(metric) %>%
+  mutate(model = map(data, ~lmer(value ~ year_scaled + climate + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
+                                 data = .x, REML = TRUE)),
+           coef = map(model, tidy)) %>%
+  select(metric, coef) %>%
+  unnest(cols = c(coef)) %>%
+  #filter(term == "year_scaled") %>%
+  filter(effect == "fixed") %>%
+  select(-c(effect, group, statistic))
+
+climate_slope_model_S <- model_data %>% 
+  filter(metric == "S") %>%
+  lmer(logvalue ~ year_scaled + climate + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
+                                 data = ., REML = TRUE)
+```
+
+Get study level slop estimates
+```{r}
+get_blup <- function(metric, model){
+  coef(model)$study_id %>%
+    rownames_to_column("study_id") %>%
+    mutate(metric = metric)
+}
+
+blup <- pmap_dfr(bind_rows(select(SES_rand_slope, metric, model), select(metric_rand_slope,  metric, model)), get_blup) %>%
+  rename(intercept = "(Intercept)")
+
+blup_slope <- blup %>%
+  pivot_wider(names_from = "metric", values_from = "year_scaled")
+
+#pairwise plots using GGally
+ggpairs(blup_slope %>% 
+          select(-c("study_id", "FDiv", "FEve", "FRic")))
+
+#pairwise plots using GGPlot
+gatherpairs <- function(data, ..., 
+                        xkey = '.xkey', xvalue = '.xvalue',
+                        ykey = '.ykey', yvalue = '.yvalue',
+                        na.rm = FALSE, convert = FALSE, factor_key = FALSE) {
+  vars <- quos(...)
+  xkey <- enquo(xkey)
+  xvalue <- enquo(xvalue)
+  ykey <- enquo(ykey)
+  yvalue <- enquo(yvalue)
+
+  data %>% {
+    cbind(gather(., key = !!xkey, value = !!xvalue, !!!vars,
+                 na.rm = na.rm, convert = convert, factor_key = factor_key),
+          select(., !!!vars)) 
+  } %>% gather(., key = !!ykey, value = !!yvalue, !!!vars,
+               na.rm = na.rm, convert = convert, factor_key = factor_key)
+}
+
+blup_slope %>% 
+  gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S") %>% {
+  ggplot(., aes(x = .xvalue, y = .yvalue)) +
+      geom_point() + 
+      #geom_smooth(method = 'lm') +
+      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
+      scale_color_brewer(type = 'qual')
+  }
+
+blup_slope %>%
+  select(-c("FDiv", "FEve", "FRic")) %>%
+  pivot_longer(-one_of("study_id", "S"), names_to = "metric") %>%
+  ggplot(aes(x = S, y = value)) +
+  geom_point() +
+  facet_wrap(~metric) +
+  xlim(-1.3, 1.3) +
+  ylim(-0.6, 0.6) +
+    theme_classic() +
+    geom_hline(yintercept = 0) +
+    geom_vline(xintercept = 0) + 
+  ylab("Metric Slope") +
+  xlab("Richness Slope")
+
+blup_slope %>%
+  select(-c("FDiv", "FEve", "FRic")) %>%
+  pivot_longer(-one_of("study_id", "SES_FRic"), names_to = "metric") %>%
+  ggplot(aes(x = SES_FRic, y = value)) +
+  geom_point() +
+  facet_wrap(~metric) +
+  xlim(-.25, .25) +
+  ylim(-0.6, 0.6) +
+    theme_classic() +
+    geom_hline(yintercept = 0) +
+    geom_vline(xintercept = 0) + 
+  ylab("Metric Slope") +
+  xlab("Functional Richness Slope")
+```
+
+Clustering 
+```{r}
+library(clValid)
+scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)
+
+blup_mat <- blup_slope %>%  
+  #these studies have NA's for some reason, remove them 
+  filter(!study_id %in% c(445, 446, 348, 382, 312)) %>%
+  select(study_id, starts_with("SES"), S) %>%
+  column_to_rownames("study_id") %>%
+  mutate(across(.fns = scale2)) %>%
+  as.matrix()
+
+clmethods <- c("hierarchical","kmeans","pam")
+intern <- clValid(blup_mat, nClust = 2:6, clMethods = clmethods, validation = "internal")
+stab <- clValid(blup_mat, nClust = 2:6, clMethods = clmethods, validation = "stability")
+
+summary(intern)
+optimalScores(stab)
+
+#get the hierarchical clustering object
+clust <- intern@clusterObjs$hierarchical
+cat5 <- cutree(clust, k = 5)
+cat3 <- cutree(clust, k = 3)
+
+blup_clust <- blup_slope %>% 
+  left_join(as.data.frame(cat3) %>%
+  rownames_to_column("study_id")) %>%
+  left_join(as.data.frame(cat5) %>%
+  rownames_to_column("study_id"))
+```
+```{r}
+blup_clust %>% 
+  gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S") %>% {
+  ggplot(., aes(x = .xvalue, y = .yvalue, color = as.factor(cat5))) +
+      geom_point() + 
+      #geom_smooth(method = 'lm') +
+      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
+      scale_color_brewer(type = 'qual')
+  }
+```
+
+##Figures
 Reproduce Dornelas figure - all data points & regressions w/overall mean
@@
 
-plot_metric <- function(metric_name, ylabel, data, model_coef){
+plot_metric <- function(metric_name, ylabel, yval, data, model_coef){
 
@@
     #left_join(metadata %>% select(study_id, climate)) %>%
-    ggplot(aes(x = year_scaled, y = logvalue)) +
+    ggplot(aes(x = year, y = !!sym(yval))) +
     # geom_smooth(aes(color = climate, group = rarefyID), method = "glm", se = FALSE) +
     # geom_point(aes(color = climate, group = rarefyID)) +
-    geom_point(aes(group = rarefyID) ,color = "grey", size = 0.30) +
-    geom_smooth(aes(color = climate_mod, group = rarefyID), method = "glm", se = FALSE, size = 0.5) +
+    geom_point(aes(group = rarefyID) ,color = "grey", size = 0.50) +
+    geom_smooth(aes(color = climate, group = study_id), method = "glm", se = FALSE, size = 0.7) +
     #geom_smooth(color = "black", method = "glm", se = FALSE) +
@@
     ylab(ylabel) +
-    labs(color = "Climate")
+    labs(color = "Climate") +
+    scale_fill_jcolors(palette = "pal7")
 
-  ggsave(paste0(metric_name, "_trend.png"), trend_plot)
+  ggsave(paste0(here::here("figures"), "/",metric_name, "_trend.png"), trend_plot)
   return(trend_plot)
@@
 
-model_coef <- metric_mean_fit_slope %>%
+model_coef <- bind_rows(SES_rand_slope, metric_rand_slope) %>%
   select(metric, model) %>%
@@
 
-plot_map <- tibble(metric_name = c("S_int", "FRic", "FEve", "FDiv", "FDis"), 
-           ylabel = c("log(S_int)", "log(FD)", "log(Evenness)", "log(Divergence)", "log(Dispersion)"))
-pmap(plot_map, plot_metric, data = metrics, model_coef = model_coef)
+plot_map <- tibble(metric_name = c("S", "SES_FRic", "Jaccard_base"), 
+           ylabel = c("log(S)", "Functional Richness", "Jaccard"),
+           yval = c("logvalue", "value", "value"))
+m_plots <- pmap(plot_map, plot_metric, data = model_data, model_coef = model_coef)
+
+plot_join <- plot_grid(plotlist=m_plots, labels = c("A", "B", "C"), nrow = 1)
 ```
-
-Get study level slop estimates
-```{r}
-get_blup <- function(metric, model){
-  coef(model)$study_id %>%
-    rownames_to_column("study_id") %>%
-    mutate(metric = metric)
-}
-
-blup <- pmap_dfr(bind_rows(select(SES_rand_slope, metric, model), select(metric_rand_slope,  metric, model)), get_blup) %>%
-  select(-"(Intercept)") %>%
-  pivot_wider(names_from = "metric", values_from = "year_scaled")
-
-#pairwise plots using GGally
-ggpairs(blup %>% 
-          select(-c("study_id", "FDis", "FDiv", "FEve", "FRic")))
-
-#pairwise plots using GGPlot
-gatherpairs <- function(data, ..., 
-                        xkey = '.xkey', xvalue = '.xvalue',
-                        ykey = '.ykey', yvalue = '.yvalue',
-                        na.rm = FALSE, convert = FALSE, factor_key = FALSE) {
-  vars <- quos(...)
-  xkey <- enquo(xkey)
-  xvalue <- enquo(xvalue)
-  ykey <- enquo(ykey)
-  yvalue <- enquo(yvalue)
-
-  data %>% {
-    cbind(gather(., key = !!xkey, value = !!xvalue, !!!vars,
-                 na.rm = na.rm, convert = convert, factor_key = factor_key),
-          select(., !!!vars)) 
-  } %>% gather(., key = !!ykey, value = !!yvalue, !!!vars,
-               na.rm = na.rm, convert = convert, factor_key = factor_key)
-}
-
-blup %>% 
-  gatherpairs("SES_FDis", "SES_FDiv", "SES_FEve", "SES_FRic", "S") %>% {
-  ggplot(., aes(x = .xvalue, y = .yvalue)) +
-      geom_point() + 
-      #geom_smooth(method = 'lm') +
-      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
-      scale_color_brewer(type = 'qual')
-  }
-
-blup %>%
-  select(-c("FDis", "FDiv", "FEve", "FRic")) %>%
-  pivot_longer(-one_of("study_id", "S"), names_to = "metric") %>%
-  ggplot(aes(x = S, y = value)) +
-  geom_point() +
-  facet_wrap(~metric) +
-  xlim(-1.3, 1.3) +
-  ylim(-0.6, 0.6) +
-    theme_classic() +
-    geom_hline(yintercept = 0) +
-    geom_vline(xintercept = 0) + 
-  ylab("Metric Slope") +
-  xlab("Richness Slope")
-
-blup %>%
-  select(-c("FDis", "FDiv", "FEve", "FRic")) %>%
-  pivot_longer(-one_of("study_id", "SES_FRic"), names_to = "metric") %>%
-  ggplot(aes(x = SES_FRic, y = value)) +
-  geom_point() +
-  facet_wrap(~metric) +
-  xlim(-.25, .25) +
-  ylim(-0.6, 0.6) +
-    theme_classic() +
-    geom_hline(yintercept = 0) +
-    geom_vline(xintercept = 0) + 
-  ylab("Metric Slope") +
-  xlab("Functional Richness Slope")
-```
-
-Clustering 
-```{r}
-scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)
-
-blup_mat <- blup %>%  
-  #these studies have NA's for some reason, remove them 
-  filter(!study_id %in% c(445, 446, 348, 382)) %>%
-  select(study_id, starts_with("SES"), S) %>%
-  column_to_rownames("study_id") %>%
-  mutate(across(.fns = scale2)) %>%
-  as.matrix()
-
-clmethods <- c("hierarchical","kmeans","pam")
-intern <- clValid(blup_mat, nClust = 2:6, clMethods = clmethods, validation = "internal")
-stab <- clValid(blup_mat, nClust = 2:6, clMethods = clmethods, validation = "stability")
-
-summary(intern)
-optimalScores(stab)
-
-#get the hierarchical clustering object
-clust <- intern@clusterObjs$hierarchical
-cat5 <- cutree(clust, k = 5)
-cat3 <- cutree(clust, k = 3)
-
-blup_clust <- blup %>% 
-  left_join(as.data.frame(cat3) %>%
-  rownames_to_column("study_id")) %>%
-  left_join(as.data.frame(cat5) %>%
-  rownames_to_column("study_id"))
-```

### 2021-04-15 paper/manuscript.Rmd
Clean up methods, add discussion

@@
 global scale by an accelerating global extinction rate [@barnosky2011],
-and fundamental changes in some of the well-studied systems (e.g. coral
-bleaching). At the local scale however, species diversity tells a
+and fundamental changes in some of the most well-studied systems (e.g.
+coral bleaching). At the local scale however, species diversity tells a
 different story. Recent synthesis of local trends in biodiversity over
@@
 and evidence of a decoupling of community composition from richness
-[@brose2016; [@gotelli2017]. While communities are clearly changing, our
-most common species-based approaches do not fully capture the nature of
-that change.
+[[@brose2016]; [@gotelli2017]. Potential hypothesis for the balance of
+etc..<!-- potential indication of biological homogenization due to shifting ranges and/or invasive species-->
+While communities are clearly changing, our most common species-based
+approaches do not fully capture the nature of that change.
 
@@
 losses or gains of functionally unique species. Functional diversity
-metrics are particularly suited for understanding community-level
-biodiversity patterns as they can detect shifts in community structure
-independent of species change.
+metrics are therefore particularly well suited for addressing change in
+communities with constant species richness.
 
-Functional diversity sheds further light on critical ecosystem and
-community processes potentially impacted by anthropogenic change. For
-example, there is increasing evidence that functional diversity is a
-better predictor of ecosystem function than species-based metrics
-[@cadotte2011; @gagic2015], and that different facets of functional
-diversity also play a role in maintaining ecosystem stability
+Functional diversity also has the potential to shed light on critical
+ecosystem and community processes potentially impacted by anthropogenic
+change. For example, there is increasing evidence that functional
+diversity is a better predictor of ecosystem function than species-based
+metrics [@cadotte2011; @gagic2015], and that different facets of
+functional diversity also play a role in maintaining ecosystem stability
 [@craven2018; @morin2014]. Further, biogeographic assessments of
-functional diversity give foundational insight into how community
-assembly plays out across systems at a standardized scale [@violle2014;
-@weiher2011].
+functional diversity may give foundational insight into how community
+assembly plays out across systems [@violle2014; @weiher2011].
 
@@
 -->
+
+<!-- paragraph about what we already know about global functional changes -->
 
@@
 providing initial looks at how functional diversity changes on a
-broad-scale for specific taxa [@trindade-santos2020; @jarzyna2016;
-@barnagaud2017].
+broad-scale for specific taxa (e.g. fish[@trindade-santos2020], birds
+[@jarzyna2016; @barnagaud2017]).
 <!-- so what do these studies say? does that go here? --> However, to
 date there has been no cross-taxa assessment of functional change
-through time on a broad geographic and taxonomic extent.\
+through time across a broad geographic and taxonomic extent.
 
 Here we perform the first multitaxa, multisystem assessment of
-functional diversity change through time. We assess 100's of mammal,
+functional diversity change through time. We assess thousands of mammal,
 bird, and amphibian functional diversity timeseries to 1) determine how
-functional diversity change compares to species change and 2) identify 3
-distinct types of functional change common across communities.
+functional diversity change compares to species change and 2) identify
+three distinct types of functional change common across communities.
+While our data represent the largest collection of functional diversity
+timeseries to date, they are not a representative sample of systems
+experiencing biodiversity change. We do not therefore seek to classify
+the global trend in functional diversity change, but rather assess the
+utility of a functional lens in understanding biodiversity change.
 
-### Methods
+### Material and Methods
 
@@
 timeseries. All studies included in the database follow consistent
-sampling protocols and represent full assemblages rather than
-population-level data [@dornelas2018]. Following best practices for the
+sampling protocols and represent full assemblages rather than population
+of single species. [@dornelas2018]. Following best practices for the
 database [@blowes2019] studies with multiple sample locations were split
@@
 supplemental materials.
+<!-- limitations of species level trait data? -->
 
@@
 
-#### *Functional diversity metrics*
+Our final dataset included 2439 timeseries from 54 studies in 21
+countries and 15 biomes. Only four studies (consisting of 11 timeseries)
+come from Amphibian studies due to the limited availability of amphibian
+timeseries and low species richness values in those timeseries. For a
+full breakdown of studies and their characteristics see the supplement.
 
-We calculated yearly metrics of functional diversity and turnover for
-each timeseries within each study. All available trait data for each
-study was included in the trait matrix with the exception of traits that
-were the same value for all observed species in the study. All
-continuous traits were z-score scaled to give each trait equal weight in
-the trait space [@leps2006; @schleuter2010]. We also calculated basic
-species richness, *S.*
+#### *Diversity Metrics*
 
-We used the function dbFD from the FD package (citation) to calculated
-functional diversity (FD), functional evenness (FEve), functional
-diveregence (FDiv), and functional dispersion (FDis). The number of
-trait axes was limited to the maximum number of traits that fulfills the
-criteria $s >= s^t$, where $s$ is the number of species and \$t\$ is the
-number of traits. This restriction allows for a sufficient number of
-axes to capture the trait space while maintaining computational
-feasibility for metric calculations. Metrics incorporated weighting
-based on either species abundance or biomass based on the methods for
-each study.
+We calculated yearly metrics of functional and species diversity for
+each timeseries. Species-based metrics include species richness (*S*)
+and Jaccard similarity (*J*) as a measure of turnover. Jaccard
+similarity was calculated relative to the first observed year for a
+timeseries. A negative trend would therefore indicate increasing
+turnover.
 
-#### ***Null Models***
+Functional diversity metrics were calculated using the dbFD function
+from the FD R package. Here we report functional richness (*FRic*),
+functional evenness (FEve), and functional diveregence (FDiv) which
+together describe three complementary characteristics of the functional
+space [@mason2005]. FRic assesses the volume of the trait space occupied
+by species in the community. FEve describes how species are distributed
+across the trait space and how abundance is distributed across species.
+FDiv measures the degree to which species and their abundances maximize
+differences in the functional space. Higher values of FDiv therefore
+correspond to communities where many highly abundant species are on the
+edges of the trait space.
 
-In order to control for changes in functional diversity metrics due to
-species richness, we calculated the standardized effect size (SES) for
-each metric from null estimates [@swenson2012]. For each richness level
-observed in a timeseries, our null model randomly selected the same
-number of species from the species pool, which included all species
-observed over the course of the timeseries. We then randomly assigned
-abundances from the original sample to each species, and all functional
-diversity metrics were calculated. This process was repeated 500 times
-to get a estimate and standard deviation of the null expectation for the
-metric for that timeseries at that richness level. We used these values
-to calculated SES using the following formula:
-$SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
+All available trait data for each study were included in functional
+diversity calculations with the exception of traits that were the same
+value for all observed species in the study. All continuous traits were
+z-score scaled to give each trait equal weight in the trait space
+[@leps2006; @schleuter2010]. The number of trait axes was limited to the
+maximum number of traits that fulfills the criteria $s >= s^t$, where
+$s$ is the number of species and \$t\$ is the number of traits. This
+restriction allows for a sufficient number of axes to capture the trait
+space while maintaining computational feasibility. Metrics incorporated
+weighting based on species abundance where available.
+
+#### *Null Models*
+
+We corrected for bias in functional diversity metrics due to changes in
+species richness by calculating the standardized effect size (SES) for
+each metric from null estimates [@swenson2012]. Null model corrections
+allow us to assess the degree to which the observed functional diversity
+metric deviates from the value expected by chance in a randomly
+assembled community. Null estimates were calculated for each rarefied
+sample by randomly sampling species from the species pool for each year
+and randomly assigning observed abundances to species. Species pools
+included all species observed for a timeseries. This process was
+repeated 500 times to get an estimate and standard deviation of the null
+expectation for the metric for that timeseries. We used these values to
+calculate SES using the following formula:
+$SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$. Unless otherwise
+specified, functional diversity results are reported for standardized
+values.
 
@@
 
-We modeled each diversity metric using a Linear Mixed Effects model with
-a random slope and intercept for year and a nested random effect of
-timeseries nested within study.
+We estimated overall trends for each diversity metric using a Linear
+Mixed Effects model with a random slope and intercept for each
+timeseries nested within study. Differences in trends by taxa, biome,
+and realm were also modeled individually as covariates in models of the
+same structure.
+
+We assessed the type of changes happening in studies across metrics by
+performing a cluster analysis of the study-level trends for each metric
+using hierarchical clustering. Studies were clustered based on the
+overall temporal slope for each metric. We included all standardized
+functional diversity metrics and species richness to represent
+independent facets of biodiversity change.
+<!-- why do some studies not have blup estimates for FRic?-->
+
+### Results
+
+We found no significant temporal trend in species richness or functional
+diversity metrics for either observed or standardized metrics. The lack
+of trend also held when data was broken down by taxa, biome and realm.
+We did find a significant decrease in Jaccard similarity, indicating
+increasing turnover through time.
+
+At the study level, the majority of studies (78%) experienced a
+significant temporal change in species richness and all functional
+diversity metrics. Only four studies experienced no significant trend in
+species richness clustering analysis showed that all but a few sites
+were part of a cluster characterized by minimal deviance from a zero
+temporal trend estimate for all metrics (Table 1). We identified four
+other clusters containing six studies that were characterized by outlier
+trends in at least one of the metrics and near-zero trends in the rest.
+The first contained two studies with the most strongly decreasing
+evenness. The second, two studies with the most strongly increasing
+functional divergence and increasing functional richness. The third
+contained one study with strongly decreasing functional divergence and
+richness, and the fourth strongly decreasing species richness with
+increasing functional richness.
+
+### Discussion
+
+Our study represents the first broad-scale multi-taxa assessment of
+functional change through time, giving a first look at overall trends in
+functional diversity and the range of changes happening in mammal, bird,
+and amphibian communities.
+
+We found no systematic temporal change in any measured dimension of
+functional diversity or species richness, even within specific realms,
+biomes, or taxonomic groups. Species-based results are consistent with
+other broad scale analyses of biodiversity change through time
+[@dornelas2014] that found no overall trend in species richness despite
+increasing turnover. While nearly all of the studies included in our
+data experienced a significant increase or decrease in species richness,
+they were observed in equal numbers resulting in an overall trend not
+significantly different from zero. Surprisingly, functional diversity
+metrics also did not capture any overall trends, indicating that similar
+to species richness, communities are experience a wide range of
+functional shifts that cannot be explained by their taxa, realm, or
+climate.
+
+While species-level metrics are the only functional data available to
+perform an analysis of this scale, they are unable to detect functional
+shifts in communities with little to no turnover as they do not capture
+intraspecific shifts.
+
+Data limitations:
+
+-   species-level traits
+
+-   not a lot of changes in these systems already?
+
+#### Add to supplement
+
+-   breakdown of number of studies/timeseries for each taxa/realm etc

### 2021-04-16 paper/manuscript.Rmd
Add corelation analysis, build discussion

@@
 database [@blowes2019] studies with multiple sample locations were split
-into individual timeseries following a standardized scale. A global grid
-was defined with a cell size determined based on the sample extent of
+into individual timeseries following a standardized scale. Scale was set
+by a global grid with cell size determined based on the sample extent of
 studies with only a single location (see [@dornelas2018] for details on
@@
 for all species of the same identifier were averaged. Only studies with
-at least 85% trait coverage were included. In order to have a sufficient
+at least 75% trait coverage were included. In order to have a sufficient
 number of species to calculate functional diversity metrics years with
@@
 
-Many studies had a variable number of samples within years. To adjust
+Many studies had a variable number of samples within years. To account
 for this inconsistency in sampling effort we used sample-based
@@
 come from Amphibian studies due to the limited availability of amphibian
-timeseries and low species richness values in those timeseries. For a
-full breakdown of studies and their characteristics see the supplement.
+timeseries and low species richness values for assemblages. For a full
+breakdown of studies and their characteristics see the supplement.
 
@@
 specified, functional diversity results are reported for standardized
-values.
+values for the rest of the paper.
 
 #### *Analysis*
+
+*Overall Trends*
 
@@
 
-We assessed the type of changes happening in studies across metrics by
-performing a cluster analysis of the study-level trends for each metric
-using hierarchical clustering. Studies were clustered based on the
-overall temporal slope for each metric. We included all standardized
-functional diversity metrics and species richness to represent
-independent facets of biodiversity change.
+*Study-level Trends*
+
+We obtained study-level estimates of temporal change from the Best
+Linear Unbiased Predictors (BLUP's) for each overall trend model and
+assessed them in two ways. First, we looked at pairwise relationships
+between all metrics to assess what kinds of community shifts are
+commonly found together. We obtained correlations and bootstrapped
+p-values using *corr.test* and *corCI* functions in the *psych* R
+package (citation). Second, we assessed the relationship between all
+metrics simultaneously by performing a cluster analysis which grouped
+each study based on their similarities across all metrics. We used a
+hierarchical clustering algorithm through the function *hclust* from the
+R package *stats.* We included all standardized functional diversity
+metrics, species richness, and Jaccard similarity in the analysis. We
+report a qualitative interpretation of the groups from the cluster
+analysis.
+
 <!-- why do some studies not have blup estimates for FRic?-->
@@
 We found no significant temporal trend in species richness or functional
-diversity metrics for either observed or standardized metrics. The lack
-of trend also held when data was broken down by taxa, biome and realm.
-We did find a significant decrease in Jaccard similarity, indicating
-increasing turnover through time.
+diversity metrics for either observed or standardized metrics. There
+continued to be no trend for studies broken down by taxa, biome or
+realm. We did find a significant overall decrease in Jaccard similarity,
+indicating increasing turnover through time.
 
@@
 significant temporal change in species richness and all functional
-diversity metrics. Only four studies experienced no significant trend in
-species richness clustering analysis showed that all but a few sites
-were part of a cluster characterized by minimal deviance from a zero
-temporal trend estimate for all metrics (Table 1). We identified four
-other clusters containing six studies that were characterized by outlier
-trends in at least one of the metrics and near-zero trends in the rest.
-The first contained two studies with the most strongly decreasing
-evenness. The second, two studies with the most strongly increasing
-functional divergence and increasing functional richness. The third
-contained one study with strongly decreasing functional divergence and
-richness, and the fourth strongly decreasing species richness with
-increasing functional richness.
+diversity metrics. 92% of sites experienced a significant trend for
+species richness, 88% for functional functional richness, 96% for
+functional divergence, 88% for functional evenness, and 35% for Jaccard
+similarity. Only four studies experienced no significant trend in
+species richness and were also characterized by significantly decreasing
+functional evenness. There was only one significant pairwise correlation
+between the metrics, a positive relationship between Functional
+Divergence and Functional Evenness, indicating that studies experiencing
+an increase in Divergence frequently experience and increase in
+Evenness.
+
+Cluster analysis grouped all but a few studies into a cluster
+characterized by non-siginficant or very small significant slopes for
+all metrics. We identified four other clusters containing six studies
+that were characterized by outlier trends in at least one of the metrics
+and near-zero trends in the rest. The first contained two studies with
+the most strongly decreasing evenness. The second, two studies with the
+most strongly increasing functional divergence and increasing functional
+richness. The third contained one study with strongly decreasing
+functional divergence and richness, and the fourth strongly decreasing
+species richness with increasing functional richness.
 
@@
 
-Our study represents the first broad-scale multi-taxa assessment of
+Our study represents the largest broad-scale multi-taxa assessment of
 functional change through time, giving a first look at overall trends in
 functional diversity and the range of changes happening in mammal, bird,
-and amphibian communities.
+and amphibian communities. We found no systematic temporal change in any
+measured dimension of functional diversity or species richness, even
+within specific realms, biomes, or taxonomic groups. Species-based
+results are consistent with other broad scale analyses of biodiversity
+change through time [@dornelas2014] that found no overall trend in
+species richness despite increasing turnover. While nearly all of the
+studies included in our data experienced a significant increase or
+decrease in species richness, they were observed in equal numbers
+resulting in an overall trend not significantly different from zero.
+Surprisingly, there was also no general trend in functional diversity
+metrics indicating that, similar to species richness, communities are
+experience a wide range of functional shifts. Non significant results
+for study characteristics confirm that functional changes cannot be
+explained by the taxa, realm, or climate characteristics of a given
+study.
 
-We found no systematic temporal change in any measured dimension of
-functional diversity or species richness, even within specific realms,
-biomes, or taxonomic groups. Species-based results are consistent with
-other broad scale analyses of biodiversity change through time
-[@dornelas2014] that found no overall trend in species richness despite
-increasing turnover. While nearly all of the studies included in our
-data experienced a significant increase or decrease in species richness,
-they were observed in equal numbers resulting in an overall trend not
-significantly different from zero. Surprisingly, functional diversity
-metrics also did not capture any overall trends, indicating that similar
-to species richness, communities are experience a wide range of
-functional shifts that cannot be explained by their taxa, realm, or
-climate.
+Individual studies in our dataset demonstrate the variety of species and
+functional shifts that can happen concurrently in a community. Notably,
+many studies did not experience increasing turnover (65%) despite having
+significant change in functional dimensions.
+
+-   very few sites experienced increasing turnover, but the ones that
+    did were significant!
+
+-   what does it me for turnover not to be increasing but the function
+    of communities to be changing
+
+-   Biological homogenization is one of the hypothesized meaningful
+    forces
 

### 2021-04-19 paper/manuscript.Rmd
Discussion, list of papers to go back to

@@
 ---
-title: "Functional trait data tell new story about biodiversity change"
+title: "Functional timeseries reveal no systemic shifts for vertebrate communities"
 author: 
@@
    - \linenumbers
-output: word_document
-  #rticles::elsevier_article:
-   # includes:
-    #  in_header: preamble.tex
+output: #word_document
+  rticles::elsevier_article:
+  includes:
+   in_header: preamble.tex
 editor_options: 
@@
 and evidence of a decoupling of community composition from richness
-[[@brose2016]; [@gotelli2017]. Potential hypothesis for the balance of
+[@brose2016; @gotelli2017] . Potential hypothesis for the balance of
 etc..<!-- potential indication of biological homogenization due to shifting ranges and/or invasive species-->
@@
 
-Functional diversity also has the potential to shed light on critical
-ecosystem and community processes potentially impacted by anthropogenic
-change. For example, there is increasing evidence that functional
-diversity is a better predictor of ecosystem function than species-based
-metrics [@cadotte2011; @gagic2015], and that different facets of
-functional diversity also play a role in maintaining ecosystem stability
-[@craven2018; @morin2014]. Further, biogeographic assessments of
-functional diversity may give foundational insight into how community
-assembly plays out across systems [@violle2014; @weiher2011].
+Loss of functional diversity is also linked to impacts on critical
+ecosystem and community processes. For example, there is increasing
+evidence that functional diversity is a better predictor of ecosystem
+function than species-based metrics [@cadotte2011; @gagic2015], and that
+different facets of functional diversity play essential roles in
+maintaining ecosystem stability [@craven2018; @morin2014]. Further,
+maintenance of functional diversity is key for maximizing adaptive
+potential of communities under climate change.
+
+<!-- Functional loss is often implicit in work linking species loss to impacts on ecosystem processes despite a wide range of documented relationships between functional and species change [@biswas2011].
+Examining functional and species change side by side across a variety of communities -->
 
@@
 <!-- so what do these studies say? does that go here? --> However, to
-date there has been no cross-taxa assessment of functional change
-through time across a broad geographic and taxonomic extent.
+date there has been no cross-taxa assessment of temporal functional
+change for a broad geographic and taxonomic extent.
 
-Here we perform the first multitaxa, multisystem assessment of
+Here we perform the first multi-taxa, multi-system assessment of
 functional diversity change through time. We assess thousands of mammal,
 bird, and amphibian functional diversity timeseries to 1) determine how
-functional diversity change compares to species change and 2) identify
-three distinct types of functional change common across communities.
-While our data represent the largest collection of functional diversity
-timeseries to date, they are not a representative sample of systems
-experiencing biodiversity change. We do not therefore seek to classify
-the global trend in functional diversity change, but rather assess the
-utility of a functional lens in understanding biodiversity change.
+functional diversity change compares to species change and 2) assess the
+types of functional shifts common across communities.
+<!-- While our data represent the largest collection of functional diversity timeseries to date, they are not a representative sample of systems experiencing biodiversity change.
+We do not therefore seek to classify the global trend in functional diversity change, but rather assess the utility of a functional lens in understanding biodiversity change.-->
 
@@
 
-Functional diversity metrics were calculated using the dbFD function
-from the FD R package. Here we report functional richness (*FRic*),
-functional evenness (FEve), and functional diveregence (FDiv) which
+Functional diversity metrics were calculated using the *dbFD* function
+from the *FD* R package. Here we report functional richness (*FRic*),
+functional evenness (*FEve*), and functional diveregence (*FDiv*) which
 together describe three complementary characteristics of the functional
-space [@mason2005]. FRic assesses the volume of the trait space occupied
-by species in the community. FEve describes how species are distributed
-across the trait space and how abundance is distributed across species.
-FDiv measures the degree to which species and their abundances maximize
-differences in the functional space. Higher values of FDiv therefore
-correspond to communities where many highly abundant species are on the
-edges of the trait space.
+space [@mason2005]. *FRic* assesses the volume of the trait space
+occupied by species in the community. *FEve* describes how species are
+distributed across the trait space and how abundance is distributed
+across species. *FDiv* measures the degree to which species and their
+abundances maximize differences in the functional space. Higher values
+of *FDiv* therefore correspond to communities where many highly abundant
+species are on the edges of the trait space.
 
@@
 specified, functional diversity results are reported for standardized
-values for the rest of the paper.
+values for the remainder of the paper.
 
@@
 timeseries nested within study. Differences in trends by taxa, biome,
-and realm were also modeled individually as covariates in models of the
+and realm were each modeled individually as covariates in models of the
 same structure.
@@
 We found no significant temporal trend in species richness or functional
-diversity metrics for either observed or standardized metrics. There
-continued to be no trend for studies broken down by taxa, biome or
-realm. We did find a significant overall decrease in Jaccard similarity,
-indicating increasing turnover through time.
+diversity metrics for either observed or standardized metrics. There was
+also no trend for studies broken down by taxa, biome or realm. We did
+find a significant overall decrease in Jaccard similarity, indicating
+increasing turnover through time.
 
@@
 Divergence and Functional Evenness, indicating that studies experiencing
-an increase in Divergence frequently experience and increase in
-Evenness.
+an increase in Divergence frequently experience an increase in Evenness.
 
 Cluster analysis grouped all but a few studies into a cluster
-characterized by non-siginficant or very small significant slopes for
+characterized by non-significant or very small significant slopes for
 all metrics. We identified four other clusters containing six studies
@@
 functional diversity and the range of changes happening in mammal, bird,
-and amphibian communities. We found no systematic temporal change in any
-measured dimension of functional diversity or species richness, even
-within specific realms, biomes, or taxonomic groups. Species-based
-results are consistent with other broad scale analyses of biodiversity
-change through time [@dornelas2014] that found no overall trend in
-species richness despite increasing turnover. While nearly all of the
-studies included in our data experienced a significant increase or
+and amphibian communities. Consistent with previous species-based
+syntheses, we found no overall trend in species diversity accompanied by
+increasing turnover through time [@dornelas2018]. While nearly all of
+the studies included in our data experienced a significant increase or
 decrease in species richness, they were observed in equal numbers
 resulting in an overall trend not significantly different from zero.
-Surprisingly, there was also no general trend in functional diversity
-metrics indicating that, similar to species richness, communities are
-experience a wide range of functional shifts. Non significant results
-for study characteristics confirm that functional changes cannot be
-explained by the taxa, realm, or climate characteristics of a given
-study.
+
+Surprisingly, we did not detect an overall trend in any of the observed
+or corrected functional diversity metrics. Similar to species richness,
+communities are experiencing a wide range of functional shifts that
+balance each other out. While our dataset likely under-samples areas
+with most direct human impact, anthropogenic change is hypothesized to
+impact even the most remote systems. Conservatively, our findings
+therefore indicate that in unperturbed mammal, bird, and amphibian
+communities there is no evidence of systemic functional richness loss or
+functional change. A lack of trend for data broken down by realm, biome,
+and taxonomic group gives further evidence that directional functional
+change is absent from all systems observed in our dataset.
 
 Individual studies in our dataset demonstrate the variety of species and
-functional shifts that can happen concurrently in a community. Notably,
-many studies did not experience increasing turnover (65%) despite having
-significant change in functional dimensions.
+functional shifts that can happen concurrently in a community. With only
+one significant pairwise relationship between Functional Divergence and
+Functional Evenness, all other possible combinations of species and
+functional change were observed in at least one study. Notably, there is
+no relationship between species richness change and change in functional
+richness, meaning that even for studies that may be experiencing species
+losses there is not necessarily an accompanying loss of functional
+richness and in some cases there may even be an increase. These
+communities are gaining new species that are functional distinct enough
+to make up for functional losses due to species losses, a pattern that
+would be further magnified if the species lost are not functionally
+distinct members of the assemblage.
 
--   very few sites experienced increasing turnover, but the ones that
-    did were significant!
+The overall trend of increasing turnover is consistent with work
+indicating higher than expected background rates of turnover in
+contemporary communities [@gotelli2017; @magurran2018]. However, many
+studies did not experience increasing turnover (65%), suggesting that it
+is not universal. Even for those studies without a turnover trend,
+significant functional shifts occurred. While some communities may be
+changing more rapidly than usual, that turnover is not necessarily an
+indication of specific kind of functional change.
 
--   what does it me for turnover not to be increasing but the function
-    of communities to be changing
+Our results have important implications for how we approach work linking
+species loss to ecosystem processes, which often implicitly assume
+concurrent functional loss. There is no evidence of directional
+functional loss or even functional shifts accompany species loss in
+these natural communities. Building on a rich body of work linking
+species diversity changes to changes in ecosystem function, resilience,
+and stability (citations), it will be essential to tease apart effects
+due to species change from those due to functional shifts if we want to
+infer impacts in natural communities.
+
+<!-- The absence of systematic loss of functional richness or functional divergence suggests that widespread biological homogenization is not occurring in sampled communities.
+Biological homogenization occurs when endemic or rare species are lost and replaced with generalist or non-native species, resulting in a loss of -->
+
+As the first broad scale analysis of temporal functional diversity
+change we make x broad conclusions. First,
+
+Above all our study shows that patterns of functional change can not be
+inferred from
+
+-   how are these findings consistent with expectations for hypothesized
+    shifts in functional change?
 
@@
     forces
+
+-   other unmeasured taxa may be harder hit? different patterns for fish
+    assemblages
+
+-   contextualize functional shifts in other temporal studies
 
@@
 
+Papers to look at:
+
+Magurran et al 2018
+
+Magurran et al 2019 (temporal beta diversity)
+
+McGill 2015
+
+Crossley 2020 - not sure what to do with it
+
+Liang et al 2019 - functional homogenization
+
+Finderup Nielsen et al 2019 - biotic homogenization
+
+Dornelas et al 2019
+
+MacLean et al 2018
+
+Hilldebrand 2018
+
+**Ideas to tie in:**
+
+-   species richness loss linked to impacts on processes, in doing so
+    often inferring functional loss, but the two do not go hand in hand
+    in natural communities
+
+-   no evidence of biological homogenization
+
+-   how is this consistent with theory
+
 #### Add to supplement

### 2021-04-21 paper/manuscript.Rmd
Writing group feedback

@@
 coral bleaching). At the local scale however, species diversity tells a
-different story. Recent synthesis of local trends in biodiversity over
-time have found no net change in local biodiversity despite ongoing
-species turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017]
-and evidence of a decoupling of community composition from richness
-[@brose2016; @gotelli2017] . Potential hypothesis for the balance of
-etc..<!-- potential indication of biological homogenization due to shifting ranges and/or invasive species-->
+different story. Recent syntheses of local trends in biodiversity over
+time have found no net change in local species diversity despite ongoing
+turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and
+evidence of a significant shifts in community composition underlying
+consistent species richness [@brose2016; @gotelli2017] .
+<!-- potential indication of biological homogenization due to shifting ranges and/or invasive species-->
 While communities are clearly changing, our most common species-based
@@
 link between the organisms in a system and the processes they perform
-[@mcgill2006]. By summarizing functional traits rather than species,
-functional diversity metrics capture the disproportionate impact of
-losses or gains of functionally unique species. Functional diversity
-metrics are therefore particularly well suited for addressing change in
-communities with constant species richness.
+[@mcgill2006]. By describing the functional trait space rather than
+species, functional diversity metrics capture the disproportionate
+impact of losses or gains of functionally unique species. Functional
+diversity metrics are therefore particularly well-suited for addressing
+change in communities with constant species richness.
 
@@
 maintenance of functional diversity is key for maximizing adaptive
-potential of communities under climate change.
+potential of communities under climate change (citation).
 
@@
 assemblage data and functional traits possible for the first time,
-providing initial looks at how functional diversity changes on a
-broad-scale for specific taxa (e.g. fish[@trindade-santos2020], birds
+providing initial insights into the ways functional diversity changes on
+a broad scale for specific taxa (e.g. fish[@trindade-santos2020], birds
 [@jarzyna2016; @barnagaud2017]).
@@
 bird, and amphibian functional diversity timeseries to 1) determine how
-functional diversity change compares to species change and 2) assess the
-types of functional shifts common across communities.
+functional shifts compares to species change and 2) assess the types of
+functional shifts that are most common across communities.
 <!-- While our data represent the largest collection of functional diversity timeseries to date, they are not a representative sample of systems experiencing biodiversity change.
@@
 sampling protocols and represent full assemblages rather than population
-of single species. [@dornelas2018]. Following best practices for the
-database [@blowes2019] studies with multiple sample locations were split
-into individual timeseries following a standardized scale. Scale was set
-by a global grid with cell size determined based on the sample extent of
-studies with only a single location (see [@dornelas2018] for details on
-how sample extents were defined), with the area of each cell set to one
-standard deviation away from the mean of the single extent locations.
-All samples from a study within a single cell were considered to be a
-single timeseries, and species abundances were combined for all samples.
+of single species [@dornelas2018]. Following best practices for the
+database [@blowes2019], studies with multiple sample locations were
+split into individual timeseries following a standardized spatial scale.
+Scale was set by a global grid with cell size determined based on the
+sample extent of studies with only a single location (see
+[@dornelas2018] for details on how sample extents were defined), with
+the area of each cell set to one standard deviation away from the mean
+of the single extent locations. All samples from a study within a single
+cell were considered to be a single timeseries, and species abundances
+were combined for all samples.
 
@@
 species' function in the community including body size, diet, and
-behavioral characteristics <!-- what's in the amphibian database? -->.
+behavioral characteristics<!-- what's in the amphibian database? -->.
 For the full list of traits included in the analysis see the
@@
 at least 75% trait coverage were included. In order to have a sufficient
-number of species to calculate functional diversity metrics years with
+number of species to calculate functional diversity metrics, years with
 fewer than 4 species observed were excluded.
@@
 
-Our final dataset included 2439 timeseries from 54 studies in 21
-countries and 15 biomes. Only four studies (consisting of 11 timeseries)
-come from Amphibian studies due to the limited availability of amphibian
-timeseries and low species richness values for assemblages. For a full
-breakdown of studies and their characteristics see the supplement.
+Our final dataset included 2,439 timeseries from 54 studies in 21
+countries and 15 biomes <!-- list traits -->. Only four studies
+(consisting of 11 timeseries) came from Amphibian studies due to the
+limited availability of amphibian timeseries and low species richness
+values for assemblages. For a full breakdown of studies and their
+characteristics, see the supplement.
 
@@
 similarity was calculated relative to the first observed year for a
-timeseries. A negative trend would therefore indicate increasing
+timeseries. A negative trend in *J* would therefore indicate increasing
 turnover.
@@
 Mixed Effects model with a random slope and intercept for each
-timeseries nested within study. Differences in trends by taxa, biome,
-and realm were each modeled individually as covariates in models of the
-same structure.
+timeseries nested within the study. Differences in trends by taxa,
+biome, and realm were each modeled individually as covariates in models
+of the same structure.
 
@@
 We obtained study-level estimates of temporal change from the Best
-Linear Unbiased Predictors (BLUP's) for each overall trend model and
+Linear Unbiased Predictors (BLUPs) for each overall trend model and
 assessed them in two ways. First, we looked at pairwise relationships
@@
 functional divergence, 88% for functional evenness, and 35% for Jaccard
-similarity. Only four studies experienced no significant trend in
-species richness and were also characterized by significantly decreasing
-functional evenness. There was only one significant pairwise correlation
-between the metrics, a positive relationship between Functional
-Divergence and Functional Evenness, indicating that studies experiencing
-an increase in Divergence frequently experience an increase in Evenness.
+similarity. Only 4 (cite studies) out of 54 studies experienced no
+significant trend in species richness and were also characterized by
+significantly decreasing functional evenness. There was only one
+significant pairwise correlation between the metrics, a positive
+relationship between Functional Divergence and Functional Evenness,
+indicating that communities experiencing an increase in Divergence
+frequently experience an increase in Evenness.
 
@@
 all metrics. We identified four other clusters containing six studies
-that were characterized by outlier trends in at least one of the metrics
-and near-zero trends in the rest. The first contained two studies with
-the most strongly decreasing evenness. The second, two studies with the
-most strongly increasing functional divergence and increasing functional
-richness. The third contained one study with strongly decreasing
-functional divergence and richness, and the fourth strongly decreasing
-species richness with increasing functional richness.
+(list studies) that were characterized by outlier trends in at least one
+of the metrics and near-zero trends in the rest. The first contained two
+studies with the most strongly decreasing evenness. The second, two
+studies with the most strongly increasing functional divergence and
+increasing functional richness. The third contained one study with
+strongly decreasing functional divergence and richness, and the fourth
+strongly decreasing species richness with increasing functional
+richness.
 
@@
 the studies included in our data experienced a significant increase or
-decrease in species richness, they were observed in equal numbers
+decrease in species richness, trends were observed in equal numbers
 resulting in an overall trend not significantly different from zero.
@@
 richness and in some cases there may even be an increase. These
-communities are gaining new species that are functional distinct enough
-to make up for functional losses due to species losses, a pattern that
-would be further magnified if the species lost are not functionally
-distinct members of the assemblage.
+communities are gaining new species that are functionally distinct
+enough to make up for functional losses due to species losses. This
+pattern would be further magnified if the species lost are not
+functionally distinct members of the assemblage.
 
@@
 functional loss or even functional shifts accompany species loss in
-these natural communities. Building on a rich body of work linking
-species diversity changes to changes in ecosystem function, resilience,
-and stability (citations), it will be essential to tease apart effects
-due to species change from those due to functional shifts if we want to
-infer impacts in natural communities.
+these natural communities.
+<!-- clarify that this is about relationship between concurrent patterns in a single community -->
+Building on a rich body of work linking species diversity changes to
+changes in ecosystem function, resilience, and stability (citations), it
+will be essential to tease apart effects due to species change from
+those due to functional shifts if we want to infer impacts in natural
+communities.
 

### 2021-05-07 analysis/05_model_metrics.Rmd
Incorporate Human Footprint data

@@
 library(lmerTest)
+library(cowplot)
 ```
@@
 all_model <- model_data %>%
-  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis")) %>%
+  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv", "S", "Jaccard_base")) %>%
   group_nest(metric) %>%
@@
                                  data = .x)),
-           null_model = map(data, ~lmer(value ~ taxa + realm + (year_scaled|study_id/rarefyID),
-                                 data = .x)),
-           test = map2(model, null_model, anova),
-           test_res = map(test, tidy),
            coef = map(model, tidy)
            ) %>%
-  unnest(cols = c(test_res)) %>%
-  select(metric, model, coef, p.value) %>%
-  filter(!is.na(p.value)) %>%
+  select(metric, model, coef) %>%
   unnest(cols = c(coef)) %>%
@@
 
-Get study level slop estimates
+Get study level slope estimates
 ```{r}
-get_blup <- function(metric, model){
-  coef(model)$study_id %>%
-    rownames_to_column("study_id") %>%
-    mutate(metric = metric)
+get_study_slopes <- function(metric, model){
+  year_scaled_coef <- tidy(model) %>%
+    filter(term == "year_scaled")
+
+  year_scaled_est <- pull(year_scaled_coef, estimate)
+
+  year_scaled_var <- year_scaled_coef %>% 
+    mutate(var = std.error*std.error) %>%
+    pull(var)
+
+  study_ests <- broom.mixed::tidy(model, effects="ran_vals") %>%
+    filter(group == "study_id", term == "year_scaled") %>% 
+    select(-c(effect, group, term)) %>%
+    rename(study_id = level, cond.std.error = std.error, cond.estimate = estimate) %>%
+    mutate(estimate = cond.estimate + year_scaled_est,
+      cond.var = cond.std.error*cond.std.error, 
+           var = cond.var + year_scaled_var,
+           std.error = sqrt(var),
+           upr.ci = estimate + (2*std.error),
+           lwr.ci = estimate - (2*std.error),
+           sig = case_when(
+             lwr.ci < 0 & upr.ci > 0 ~ TRUE,
+             TRUE ~ FALSE
+           ),
+           metric = metric)
 }
 
-blup <- pmap_dfr(bind_rows(select(SES_rand_slope, metric, model), select(metric_rand_slope,  metric, model)), get_blup) %>%
-  rename(intercept = "(Intercept)")
+study_slopes <- pmap_dfr(bind_rows(select(SES_rand_slope, metric, model), select(metric_rand_slope,  metric, model)), get_study_slopes)
 
-blup_slope <- blup %>%
-  pivot_wider(names_from = "metric", values_from = "year_scaled")
+slopes_wide <- study_slopes %>%
+  select(study_id, metric, estimate) %>%
+  pivot_wider(names_from = "metric", values_from = "estimate")
+```
 
+Summaries of study slopes
+```{r}
+#First - how many studys had a significant trend for each metric? 
+study_slopes %>% 
+  filter(sig == TRUE) %>% 
+  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
+  group_by(metric) %>% 
+  summarise(n_study_id = n_distinct(study_id)) %>%
+  mutate(percent = n_study_id/54)
+
+#And how many were significant for all but Jaccard?
+study_slopes %>% 
+  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "S")) %>% 
+  group_by(study_id) %>% filter(sum(sig) == 4) %>% pull(study_id) %>% 
+  n_distinct()
+
+#Breakdown of relationship between functional richness and species richness
+s_sig <- study_slopes %>% filter(metric == "S", sig == TRUE) %>% pull(study_id) %>% unique()
+s_sig_not_fric <- study_slopes %>% 
+  filter(study_id %in% s_sig, metric == "FRic", sig == FALSE) %>% 
+  pull(study_id) %>%
+  unique()
+
+both_sig <- study_slopes %>% 
+  filter(study_id %in% s_sig, metric == "FRic", sig == TRUE) %>%
+  pull(study_id)
+
+both_pos
+fric_pos
+s_pos_fric_neg <- study_slopes %>%
+  filter(study_id %in% both_sig) %>% 
+  group_by(study_id) %>%
+  filter(sum(sign(estimate)) == 0)
+
+```
+Pairwise correlation analysis
+```{r}
+slope_corr <- corr.test(slopes_wide %>% ungroup() %>% select(starts_with("SES"), "S", "Jaccard_base"))
+
+slope_corr_bootstrap <- corCi(slopes_wide %>% ungroup() %>% select(starts_with("SES"), "S", "Jaccard_base"))
+```
+
+Plots of study-level slopes
+```{r}
 #pairwise plots using GGally
-ggpairs(blup_slope %>% 
+ggpairs(slopes_wide %>% 
           select(-c("study_id", "FDiv", "FEve", "FRic")))
@@
 
-blup_slope %>% 
+slopes_wide %>% 
   gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S") %>% {
@@
 
-blup_slope %>%
+slopes_wide %>%
   select(-c("FDiv", "FEve", "FRic")) %>%
@@
 
-blup_mat <- blup_slope %>%  
-  #these studies have NA's for some reason, remove them 
-  filter(!study_id %in% c(445, 446, 348, 382, 312)) %>%
-  select(study_id, starts_with("SES"), S) %>%
+scale_slopes <- slopes_wide %>%  
+  drop_na() %>%
+  select(study_id, starts_with("SES"), S, Jaccard_base) %>%
   column_to_rownames("study_id") %>%
@@
 clmethods <- c("hierarchical","kmeans","pam")
-intern <- clValid(blup_mat, nClust = 2:6, clMethods = clmethods, validation = "internal")
-stab <- clValid(blup_mat, nClust = 2:6, clMethods = clmethods, validation = "stability")
+intern <- clValid(scale_slopes, nClust = 2:6, clMethods = clmethods, validation = "internal")
+stab <- clValid(scale_slopes, nClust = 2:6, clMethods = clmethods, validation = "stability")
 
@@
 
-blup_clust <- blup_slope %>% 
+slopes_clust <- slopes_wide %>% 
+  ungroup() %>%
   left_join(as.data.frame(cat3) %>%
@@
 ```
+
 ```{r}
-blup_clust %>% 
-  gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S") %>% {
+slopes_clust %>% 
+  gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S", "Jaccard_base") %>% {
   ggplot(., aes(x = .xvalue, y = .yvalue, color = as.factor(cat5))) +
@@
     labs(color = "Climate") +
-    scale_fill_jcolors(palette = "pal7")
+    scale_color_manual(values = c("#32a251", 
+                "#ff7f0f", "#39737c", "#b85a0d", "#ffd94a"))
 
@@
 
-plot_metric("FRic", "log(FD)")
-
 plot_map <- tibble(metric_name = c("S", "SES_FRic", "Jaccard_base"), 
@@
 
-plot_join <- plot_grid(plotlist=m_plots, labels = c("A", "B", "C"), nrow = 1)
+plot_join <- plot_grid(plotlist=m_plots, labels = c("A", "B", "C"), nrow = 3)
+ggsave2("3met_long.jpeg")
 ```

### 2021-05-13 analysis/05_model_metrics.Rmd
Reorganize

@@
   left_join(meta, c("rarefyID" = "rarefyid"))
+
+## Now the data that actually makes it into the model ##
+
+#find rarefyID's that have inf value for some metrics and exclude
+cont_inf <- metrics %>% filter(is.infinite(value), !metric %in% c("qual.FRic",  "SES_qual.FRic")) %>% pull(rarefyID)
+
+#get rarefyID's that don't have enough null samples
+missing_null <- bind_rows(metrics %>% 
+  filter(n_missing_nulls == 1, commplete_null_samps == FALSE),
+  metrics %>% filter(n_missing_nulls > 100)) %>%
+  pull(rarefyID) %>%
+  unique()
+
+model_data <- metrics %>% 
+  filter(!rarefyID %in% cont_inf, !rarefyID %in% missing_null) %>%
+  group_by(metric) %>%
+  mutate(study_id =  str_extract(rarefyID, "[^_]+")) %>%
+  mutate(year_scaled = scale(year), 
+         scale_center = attributes(year_scaled)$`scaled:center`, 
+         scale = attributes(year_scaled)$`scaled:scale`) %>%
+  ungroup()
+
+pins::pin(model_data, board = "github")
+```
+Add human footprint data
+```{r}
+hfp_data <- pins::pin_get("hfp-data", board = "github")
+
+model_data_hfp <- model_data %>% 
+  #dplyr::select(rarefyID, year) %>% 
+  #distinct() %>%
+  mutate(hfp_year = case_when(
+    year < 2001 ~ 1993,
+    TRUE ~ 2009
+  ), 
+  hfp_rarefyID = str_replace_all(rarefyID, pattern = "_bird", ""), 
+  hfp_rarefyID = str_replace_all(hfp_rarefyID, pattern = "_mamm", "")) %>%
+  left_join(hfp_data %>% dplyr::select(-STUDY_ID) %>% mutate(year = as.numeric(year)), by = c("hfp_year" = "year", "hfp_rarefyID" = "rarefyID")) %>%
+  dplyr::select(-hfp_year)
+
 ```
@@
 ####Data exploration
-
-How many years are missing FD metrics?
-Only a handful!
-Let's exclude them so we can get an accurate view of the number of years for which we have data for each timeseries
-```{r}
-# no_fd_years <- metrics %>%
-#   filter(metric %in% c("FRic", "FEve", "FDiv", "FDis"), is.na(value))
-# 
-# dim(no_fd_years)
-# 
-# metrics <- setdiff(metrics, no_fd_years)
-```
 
@@
 
-Fit GLM to overall metric trend
+####Fit GLM to overall metric trend
 
@@
 ```{r}
-#find rarefyID's that have inf value for some metrics and exclude
-cont_inf <- metrics %>% filter(is.infinite(value), !metric %in% c("qual.FRic",  "SES_qual.FRic")) %>% pull(rarefyID)
-
-#get rarefyID's that don't have enough null samples
-missing_null <- bind_rows(metrics %>% 
-  filter(n_missing_nulls == 1, commplete_null_samps == FALSE),
-  metrics %>% filter(n_missing_nulls > 100)) %>%
-  pull(rarefyID) %>%
-  unique()
-
-model_data <- metrics %>% 
-  filter(!rarefyID %in% cont_inf, !rarefyID %in% missing_null) %>%
-  group_by(metric) %>%
-  mutate(study_id =  str_extract(rarefyID, "[^_]+")) %>%
-  mutate(year_scaled = scale(year), 
-         scale_center = attributes(year_scaled)$`scaled:center`, 
-         scale = attributes(year_scaled)$`scaled:scale`) %>%
-  ungroup()
-
 # model for overall trend (random effect w/varying intercept)

### 2021-05-13 analysis/visualization.Rmd
Reorganize

--- 2020-11-20 analysis/visualization.Rmd
@@
 ---
-title: "Visualization"
-author: "Kari Norman"
-date: "9/18/2019"
+title: "worldmap"
 output: html_document
@@
 
-```{r}
-library(tidyverse)
+## R Markdown
+
+This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.
+
+When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
+You can embed an R code chunk like this:
+
+```{r cars}
+library(ggmap)
+library(ggplot2)
+library(dplyr)
+library(ggthemes)
+library(mapproj)
+library(stringr)
 ```
@@
 ```{r}
-metrics <- pins::pin_get("rarefied-medians", board = "github") %>%
-  pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "cYEAR", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
-  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID"))) 
+pins::board_register_github(repo = "karinorman/biodivTS_data", branch = "master")
+country_data <- pins::pin_get("country-data-bind", board = "github") %>%
+  mutate(country = str_remove(country, "Commonwealth of The "),
+         country = str_remove(country, "Kingdom of "),
+         country = str_remove(country, "Republic of "),
+         country = str_remove(country, "Kingdom of "),
+         country = str_remove(country, "Federative Republic of "),
+         country = str_remove(country, "Federative "),) %>%
+  mutate(country = case_when(
+    country == "United States" ~ "USA",
+    country == "Russian Federation" ~ "Russia",
+    country == "United Great Britain and Northern Ireland" ~ "UK",
+    country == "Czechia" ~ "Czech Republic",
+    TRUE ~ country
+  )) %>% 
+  count(country, code)
+
+mapdata <- map_data("world") %>%
+  left_join(country_data, by = c("region" = "country"))
+
+mapdata %>%
+  ggplot(aes(long, lat)) +
+  geom_polygon(mapping = aes(group = group, fill = n)) +
+  ggthemes::theme_map() +
+  #scale_fill_continuous(color = "#32A251")
+  scale_fill_gradient(low = "white", high = "#39737c")
+
+ggsave("study_map.jpeg")
 ```
@@
 ```{r}
-usa <- map_data("usa")
+#get rarefyid centers
+included_ids <- pins::pin_get("model-data", board = "github") %>%
+  select(rarefyID, taxa, climate, realm) %>%
+  distinct() %>% 
+  mutate(rarefyID = str_replace_all(rarefyID, pattern = "_bird", ""), 
+  rarefyID = str_replace_all(rarefyID, pattern = "_mamm", ""))
 
-plt <- ggplot() +
-  geom_polygon(data = usa, aes(x=long, y = lat, group = group), fill = NA, color = "black") +
-  coord_fixed(1.3)
+study_data <- pins::pin_get("rarefyID-cell-centre", board = "github") %>%
+  inner_join(included_ids)
 
-plt +
-  geom_point(data = biotime_data %>% filter(taxa == "Fish"), aes(x = longitude, y = latitude), color = "black", size = .25)
-
+map_data("world") %>%
+  ggplot(aes(long, lat, group = group)) +
+  geom_polygon(fill = "white", colour = "grey") + 
+  coord_map(xlim=c(-180,180), ylim=c(-57, 90)) +
+  geom_point(data = study_data, aes(rarefyID_x, rarefyID_y, color = taxa), alpha = 0.5, inherit.aes = FALSE) +
+  ggthemes::theme_map()
 ```
-
-#Visualizing FD trends
-```{r}
-#load(here::here("data", "fd_table.rda"))
-fd_table <- pins::pin_get("fd-table", board = "github")
-
-fric_plot <- fd_table %>%
-  filter(metric == "FRic") %>%
-  ggplot(aes(
-    x = year,
-    y = value,
-    colour = as.factor(plot)
-  )) +
-  geom_line() +
-  scale_x_continuous(breaks = scales::pretty_breaks()) +
-  facet_wrap( ~ as.factor(plot_id), scales = "free") +
-  theme(legend.position = "none") +
-  ylab("Fric")
-
-ggsave("fric_trends.png", fric_plot)
-```
-
-Reproduce Dornelas figure
-```{r}
-#get climate classification
-metadata <- read_csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS"))
-
-fric_trend <- fd_table %>%
-  filter(metric == "FRic") %>%
-  left_join(metadata %>% select(study_id, climate)) %>%
-  ggplot(aes(x = year, y = log(value))) +
-  # geom_smooth(aes(color = climate, group = plot_id), method = "glm", se = FALSE) +
-  # geom_point(aes(color = climate, group = plot_id)) +
-  geom_smooth(aes(group = plot_id), color = "grey", method = "glm", se = FALSE) +
-  geom_point(aes(group = plot_id) ,color = "grey") +
-  geom_smooth(color = "black", method = "glm", se = FALSE) +
-  theme(legend.position = "none") 
-ggsave("fric_trends.png", fric_plot)
-```
-
-metric plotting function
-```{r}
-library(lme4)
-
-plot_metric <- function(metric_name, ylabel){
-  data <- metrics %>%
-    filter(metric == metric_name, !is.na(value), !is.infinite(value)) %>%
-    mutate(logvalue = log(value))
-
-  mod <- lmer(logvalue ~ year + (1|rarefyID), data = data)
-  coefs <- summary(mod)$coefficients
-
-  print(metric_name) 
-  print(summary(mod))
-  confint(mod)
-
-  pred <- data.frame(year = min(data$year):max(data$year)) %>%
-    mutate(pred = coefs[1,1] + year * coefs[2,1])
-
-  trend_plot <- data %>%
-    left_join(pred, by = "year") %>%
-    #left_join(metadata %>% select(study_id, climate)) %>%
-    ggplot(aes(x = year, y = logvalue)) +
-    # geom_smooth(aes(color = climate, group = rarefyID), method = "glm", se = FALSE) +
-    # geom_point(aes(color = climate, group = rarefyID)) +
-    geom_point(aes(group = rarefyID) ,color = "grey") +
-    geom_smooth(aes(group = rarefyID), color = "darkgrey", method = "glm", se = FALSE) +
-    #geom_smooth(color = "black", method = "glm", se = FALSE) +
-    geom_line(mapping = aes(y = pred)) +
-    theme(legend.position = "none") +
-    theme_classic() +
-    ylab(ylabel)
-
-  ggsave(paste0(metric_name, "_trend.png"), trend_plot)
-  return(trend_plot)
-}
-
-plot_metric("FRic", "log(FD)")
-
-plot_map <- tibble(metric_name = c("FRic", "FEve", "FDiv", "FDis"), 
-           ylabel = c("log(FD)", "log(Evenness)", "log(Divergence)", "log(Dispersion)"))
-pmap(plot_map, plot_metric)
-```
-
-get model info
-```{r}
-get_metric_conint <- function(metric_name){
- data <- metrics %>%
-    filter(metric == metric_name) %>%
-    mutate(rarefyID = as.factor(rarefyID), logValue = log(value))
-
-  mod <- lmer(logValue ~ year + (1|rarefyID), data = data)
-  coefs <- summary(mod)$coefficients
-
-  confint(mod)
-  # as_tibble(confint(mod)) %>%
-  #   mutate(metric = metric_name)
-}
-
-map(c("FRic", "FEve", "FDiv", "FDis"), get_metric_conint)
-```

### 2021-05-13 data-raw/01_biotime_grid.R
Clean up commits

@@ -219,7 +219,10 @@ ML_coords <- ungroup(bt) %>%
 rarefyID_cell_centre <- bind_rows(rarefyID_cell_centre, SL_coords, ML_coords)
 ##	not sure why but I have multiple entries for each rarefyID!
 rarefyID_cell_centre <- rarefyID_cell_centre %>% distinct(STUDY_ID, rarefyID, cell_extent, rarefyID_x, rarefyID_y)
-#save(rarefyID_cell_centre, file='rarefyID_cell_centre_011017.Rdata')
+
+#save to cache
+pins::pin(rarefyID_cell_centre, board = "github")
+
 ##======================================================================
 ##	reduce to data required for rarefying, and rename a couple of columns
 ##	NB: we will rarefy to the smallest number of ObsEventID's within studies

### 2021-05-31 analysis/04_collate_rarefied_resamps_median.R
observations with a null model with no variation (sd = 0) assigned to zero

@@ -68,7 +68,7 @@ metric_cols <- metric_cols %>%
   mutate(SES = (value - mean)/sd) %>%
   #for years where species richness = richness of the species pool, FRic sd = 0, so use the original value with no SES correction
   mutate(SES = case_when(
-    sd == 0 ~ value,
+    sd == 0 ~ 0,
     TRUE ~ SES)) %>%
   select(-c(mean, sd)) %>%
   #the next three lines go back to wide format for both SES and unadjusted values, then to long format incorporating SES values

### 2021-05-31 analysis/05_model_metrics.Rmd
Restructuring

@@
 ```
-
+#### Read in data, get final dataset for models
 Get data and put it in long form 
@@
   rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
-  filter(!is.na(value)) %>%
   mutate(logvalue = log(value + 1)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
@@
 ## Now the data that actually makes it into the model ##
-
-#find rarefyID's that have inf value for some metrics and exclude
-cont_inf <- metrics %>% filter(is.infinite(value), !metric %in% c("qual.FRic",  "SES_qual.FRic")) %>% pull(rarefyID)
 
@@
 model_data <- metrics %>% 
-  filter(!rarefyID %in% cont_inf, !rarefyID %in% missing_null) %>%
+  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "Jaccard_next",
+                       "FRic", "FEve", "FDiv", "S"),
+         !rarefyID %in% missing_null) %>%
   group_by(metric) %>%
@@
          scale = attributes(year_scaled)$`scaled:scale`) %>%
-  ungroup()
+  ungroup() %>%
+  #this study has many low species richness observations, so FD metrics weren't calculated, just drop it
+  filter(study_id != 348)
 
@@
 
-####Fit GLM to overall metric trend
+####Fit GLMs to overall patterns and data broken down by groups
 
@@
 https://biologyforfun.wordpress.com/2018/04/09/help-i-have-convergence-warnings/
-
-Maybe need something other than a gaussian distribution?
 
@@
            # test_res = map(test, tidy),
+           coef = map(model, tidy),
+           ) %>% 
+  select(metric, model, coef) %>%
+  unnest(cols = c(coef)) %>%
+  filter(term == "year_scaled") %>%
+  select(-c(effect, group, term, statistic))
+
+#check that they all used the same number of observations
+map(SES_rand_slope$model, nobs)
+
+metric_rand_slope <- model_data %>%
+  #filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
+  filter(metric %in%  c("S")) %>%
+  #it wants its own scaling to converge for some reason
+  mutate(year_scaled = scale(year)) %>%
+  group_nest(metric) %>%
+  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
+                                 data = .x, REML = TRUE)),
            coef = map(model, tidy)
@@
 
-metric_rand_slope <- model_data %>%
-  #filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
-  filter(metric %in%  c("FRic", "FEve", "FDiv", "S")) %>%
-  group_nest(metric) %>%
-  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
-                                 data = .x, REML = FALSE)),
-           coef = map(model, tidy)
-           ) %>% 
-  select(metric, model, coef) %>%
-  unnest(cols = c(coef)) %>%
-  filter(term == "year_scaled") %>%
-  select(-c(effect, group, term, statistic))
+#check that they all used the same number of observations
+map(metric_rand_slope$model, nobs)
 
@@
 
+#### Process Model Output
+
 Get study level slope estimates
@@
 ```{r}
-#First - how many studys had a significant trend for each metric? 
+#First - how many studies had a significant trend for each metric? 
 study_slopes %>% 
@@
 ```
+
+#### Analizing slopes
 Pairwise correlation analysis
@@
 
-Plots of study-level slopes
+Clustering 
+```{r}
+library(clValid)
+scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)
+
+scale_slopes <- slopes_wide %>%  
+  drop_na() %>%
+  select(study_id, starts_with("SES"), S, Jaccard_base) %>%
+  column_to_rownames("study_id") %>%
+  mutate(across(.fns = scale2)) %>%
+  as.matrix()
+
+clmethods <- c("hierarchical","kmeans","pam")
+intern <- clValid(scale_slopes, nClust = 2:6, clMethods = clmethods, validation = "internal")
+stab <- clValid(scale_slopes, nClust = 2:6, clMethods = clmethods, validation = "stability")
+
+summary(intern)
+optimalScores(stab)
+
+#get the hierarchical clustering object
+clust <- intern@clusterObjs$hierarchical
+cat5 <- cutree(clust, k = 5)
+cat3 <- cutree(clust, k = 3)
+
+slopes_clust <- slopes_wide %>% 
+  ungroup() %>%
+  left_join(as.data.frame(cat3) %>%
+  rownames_to_column("study_id")) %>%
+  left_join(as.data.frame(cat5) %>%
+  rownames_to_column("study_id"))
+```
+
+```{r}
+slopes_clust %>% 
+  gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S", "Jaccard_base") %>% {
+  ggplot(., aes(x = .xvalue, y = .yvalue, color = as.factor(cat5))) +
+      geom_point() + 
+      #geom_smooth(method = 'lm') +
+      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
+      scale_color_brewer(type = 'qual')
+  }
+```
+
+####Figures
+
+Pairwise plots of study-level slopes
 ```{r}
@@
 
-Clustering 
 ```{r}
-library(clValid)
-scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)
-
-scale_slopes <- slopes_wide %>%  
-  drop_na() %>%
-  select(study_id, starts_with("SES"), S, Jaccard_base) %>%
-  column_to_rownames("study_id") %>%
-  mutate(across(.fns = scale2)) %>%
-  as.matrix()
-
-clmethods <- c("hierarchical","kmeans","pam")
-intern <- clValid(scale_slopes, nClust = 2:6, clMethods = clmethods, validation = "internal")
-stab <- clValid(scale_slopes, nClust = 2:6, clMethods = clmethods, validation = "stability")
-
-summary(intern)
-optimalScores(stab)
-
-#get the hierarchical clustering object
-clust <- intern@clusterObjs$hierarchical
-cat5 <- cutree(clust, k = 5)
-cat3 <- cutree(clust, k = 3)
-
-slopes_clust <- slopes_wide %>% 
-  ungroup() %>%
-  left_join(as.data.frame(cat3) %>%
-  rownames_to_column("study_id")) %>%
-  left_join(as.data.frame(cat5) %>%
-  rownames_to_column("study_id"))
-```
-
-```{r}
-slopes_clust %>% 
-  gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S", "Jaccard_base") %>% {
-  ggplot(., aes(x = .xvalue, y = .yvalue, color = as.factor(cat5))) +
-      geom_point() + 
-      #geom_smooth(method = 'lm') +
-      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
-      scale_color_brewer(type = 'qual')
-  }
-```
-
-##Figures
-Reproduce Dornelas figure - all data points & regressions w/overall mean
-```{r}
-#get climate classification
-metadata <- read_csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS"))
-
-fric_trend <- metrics %>%
-  filter(metric == "FRic") %>%
-  ggplot(aes(x = year, y = logvalue)) +
-  # geom_smooth(aes(color = climate, group = plot_id), method = "glm", se = FALSE) +
-  # geom_point(aes(color = climate, group = plot_id)) +
-  geom_smooth(aes(group = rarefyID), color = "grey", method = "glm", se = FALSE) +
-  geom_point(aes(group = rarefyID) ,color = "grey") +
-  geom_smooth(color = "black", method = "glm", se = FALSE) +
-  theme(legend.position = "none") 
-ggsave("fric_trends.png", fric_plot)
-```
-
-```{r}
-library(lme4)
-
 plot_metric <- function(metric_name, ylabel, yval, data, model_coef){

### 2021-06-03 analysis/05_model_metrics.Rmd
Fix scaling so all models converge, all plots in one figure with a legend.

@@
 ```{r}
-# model for overall trend (random effect w/varying intercept)
-# metric_mean_fit <- model_data %>%
-#   filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
-#   group_nest(metric) %>%
-#   mutate(model = map(data, ~lmer(log(value + 1) ~ year_scaled + (1|study_id/rarefyID), data = .x, REML = FALSE,
-#                   control = lmerControl(optCtrl = list(maxit = 1e9,
-#                                                     maxfun = 1e9,
-#                                                     xtol_abs = 1e-11,
-#                                                     ftol_abs = 1e-11,
-#                                                     maxeval = 1e9)))),
-#          null_model = map(data, ~lmer(logvalue ~ (1|study_id/rarefyID), data = .x, REML = FALSE,
-#                   control = lmerControl(optCtrl = list(maxit = 1e9,
-#                                                     maxfun = 1e9,
-#                                                     xtol_abs = 1e-11,
-#                                                     ftol_abs = 1e-11,
-#                                                     maxeval = 1e9)))),
-#            
-#            test = map2(model, null_model, anova),
-#            test_res = map(test, tidy),
-#            coef = map(model, tidy)
-#            ) %>% 
-#   unnest(cols = c(test_res)) %>%
-#   select(metric, model, coef, p.value) %>%
-#   filter(!is.na(p.value)) %>%
-#   unnest(cols = c(coef)) %>%
-#   filter(term == "year_scaled") %>%
-#   select(-c(effect, group, term, statistic))
-
 SES_rand_slope <- model_data %>%
-  #filter(protected_area == "FALSE") %>%
-  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis", )) %>%
-  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "Jaccard_next")) %>%
+  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
   group_nest(metric) %>%
-  mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
+  mutate(#year_center = map(data, ~scale(.x$year)),
+         model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                  data = .x)),
@@
 
-metric_rand_slope <- model_data %>%
-  #filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
+#these two metrics are cranky and want their own year scaling to converge
+
+#Jaccard Base
+jacc_base_mod <-  model_data %>%
+  filter(metric == ("Jaccard_base")) %>%
+  mutate(year_scaled = scale(year)) %>%
+  group_nest(metric) %>%
+  mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
+                                 data = .x)),
+           coef = map(model, tidy),
+           ) %>% 
+  select(metric, model, coef) %>%
+  unnest(cols = c(coef)) %>%
+  filter(term == "year_scaled") %>%
+  select(-c(effect, group, term, statistic))
+
+# S model
+s_mod <- model_data %>%
   filter(metric %in%  c("S")) %>%
@@
 
-#check that they all used the same number of observations
-map(metric_rand_slope$model, nobs)
-
-# model_comp <- metric_mean_fit %>%
-#   select(metric, no_slope_estimate = estimate, no_slope_pval = p.value) %>%
-#   left_join(metric_mean_fit_slope %>% 
-#   select(metric, slope_estimate = estimate, slope_pval = p.value ))
-
-metric_model_table <- bind_rows(select(SES_rand_slope, -model), select(metric_rand_slope, -model))
+metric_model_table <- bind_rows(SES_rand_slope, s_mod, jacc_base_mod)
 ```
@@
 taxa_slope_model_S <- model_data %>% 
+  mutate(year_scaled = scale(year)) %>% 
   filter(metric == "S") %>%
@@
 realm_slope_model_S <- model_data %>% 
+  mutate(year_scaled = scale(year)) %>% 
   filter(metric == "S") %>%
@@
 climate_slope_model_S <- model_data %>% 
+  mutate(year_scaled = scale(year)) %>% 
   filter(metric == "S") %>%
@@
 
-study_slopes <- pmap_dfr(bind_rows(select(SES_rand_slope, metric, model), select(metric_rand_slope,  metric, model)), get_study_slopes)
+study_slopes <- pmap_dfr(select(metric_model_table,  metric, model), get_study_slopes)
 
@@
   summarise(n_study_id = n_distinct(study_id)) %>%
-  mutate(percent = n_study_id/54)
+  mutate(percent = n_study_id/53)
 
@@
 ```{r}
-slope_corr <- corr.test(slopes_wide %>% ungroup() %>% select(starts_with("SES"), "S", "Jaccard_base"))
+library(psych)
+#slope_corr <- corr.test(slopes_wide %>% ungroup() %>% select(starts_with("SES"), "S", "Jaccard_base"))
 
@@
 ```{r}
-plot_metric <- function(metric_name, ylabel, yval, data, model_coef){
+library(cowplot)
+library(RColorBrewer)
+
+plot_metric <- function(metric_name, ylabel, yval, xlabel, data, model_coef, color_list){
+
+  if(is.na(xlabel)) xlabel <- NULL
 
@@
     geom_line(mapping = aes(y = pred)) +
-    theme(legend.position = "none") +
     theme_classic() +
+    theme(axis.title = element_text(size=7), legend.position = "none") +
     ylab(ylabel) +
+    xlab(xlabel) +
     labs(color = "Climate") +
-    scale_color_manual(values = c("#32a251", 
-                "#ff7f0f", "#39737c", "#b85a0d", "#ffd94a"))
+    scale_x_continuous(breaks = scales::pretty_breaks(n = 6)) +
+    scale_colour_manual(values = color_list, drop = FALSE)
+
+    #scale_color_manual(values = c("#32a251", 
+               # "#ff7f0f", "#39737c", "#b85a0d", "#ffd94a"))
 
@@
 
-model_coef <- bind_rows(SES_rand_slope, metric_rand_slope) %>%
+model_coef <- metric_model_table %>%
   select(metric, model) %>%
@@
 
-plot_map <- tibble(metric_name = c("S", "SES_FRic", "Jaccard_base"), 
-           ylabel = c("log(S)", "Functional Richness", "Jaccard"),
-           yval = c("logvalue", "value", "value"))
-m_plots <- pmap(plot_map, plot_metric, data = model_data, model_coef = model_coef)
+pal <- brewer.pal(n = 5, name = "Dark2")
+group_colors <- c("Global" = pal[1], "Polar/Temperate" = pal[2], "Temperate/Tropical" = pal[3], "Tropical" = pal[4], "Temperate" = pal[5], "Overall Mean" = "#000000")
 
-plot_join <- plot_grid(plotlist=m_plots, labels = c("A", "B", "C"), nrow = 3)
-ggsave2("3met_long.jpeg")
+plot_map <- tibble(metric_name = c("S", "Jaccard_base", "SES_FRic", "SES_FDiv", "SES_FEve"), 
+           ylabel = c("log(S)", "Jaccard", "Functional Richness SES", "Functional Divergence SES", "Functional Evenness SES"),
+           #ylabel = c("log(S)", "Jaccard", "FRic SES", "FDiv SES", "FEve SES"),
+           yval = c("logvalue", "value", "value", "value", "value"),
+           xlabel = c(NA, NA, NA, "year", "year"))
+m_plots <- pmap(plot_map, plot_metric, data = model_data, model_coef = model_coef, color_list = group_colors)
+
+plot_join <- plot_grid(plotlist=m_plots, labels = "AUTO", nrow = 3, align = 'vh',
+                       hjust = -1, axis = "l", scale = 0.9)
+
+leg_plot <- model_data %>%
+  bind_rows(data.frame(climate = c("Overall Mean", "Overall Mean"), year = c(2019, 2020), value = c(1, 1))) %>%
+  ggplot(aes(x = year, y = value)) +
+  geom_point(color = "grey", size = 0.50) +
+  geom_smooth(aes(color = climate, group = study_id), method = "glm", se = FALSE, size = 0.7) +
+  theme_classic() +
+  theme(axis.title = element_text(size=7)) +
+  scale_colour_manual(values = group_colors, drop = FALSE)
+
+######
+group.colors <- c(A = "#333BFF", B = "#CC6600", C ="#9633FF", D = "#E2FF33", E = "#E3DB71")
+
+simplePlot <- function(DT, tit) 
+  ggplot(DT ,aes(x=Name, y=Value, fill=Group)) +
+  geom_bar(stat="identity") + xlab("") + ggtitle(tit) +
+  #Specify colours
+  scale_fill_manual(values=group.colors)
+#####
+
+# extract the legend from one of the plots
+legend <- get_legend(leg_plot)
+
+# add the legend to the row we made earlier. Give it one-third of 
+# the width of one plot (via rel_widths).
+plot_join + draw_grob(legend,  2/3.3, -0.3, .3/3.3, 1, scale = 0.9)
+
+ggsave2(here::here("figures", "3met_long.jpeg"))
 ```

### 2021-06-04 analysis/visualization.Rmd
map by taxa

@@
 
-ggsave("study_map.jpeg")
+#ggsave("study_map.jpeg")
 ```
@@
   ggthemes::theme_map()
+
+ggsave(here::here("figures/study_map.jpeg"))
 ```

### 2021-06-18 paper/manuscript.Rmd
Add header
Update figures, revamp discussion

@@
 
-journal: "Ecography"
+journal: "Ecology Letters"
 date: "`r Sys.Date()`"
@@
   markdown: 
-    wrap: 72
+    wrap: sentence
 ---
+
+```{r message=FALSE, include = FALSE}
+library(kableExtra)
+library(magrittr)
+library(tidyverse)
+#library(printr)
+library(rticles)
+library(knitr)
+
+knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
+
+# no sci notation integers pleases
+options(scipen=999)
+```
 
@@
 
-Ecological communities are experiencing unprecedented change as a result
-of anthropogenic pressures such as climate change, land use change, and
-invasive species. Impacts of these pressures are well documented at a
-global scale by an accelerating global extinction rate [@barnosky2011],
-and fundamental changes in some of the most well-studied systems (e.g.
-coral bleaching). At the local scale however, species diversity tells a
-different story. Recent syntheses of local trends in biodiversity over
-time have found no net change in local species diversity despite ongoing
-turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and
-evidence of a significant shifts in community composition underlying
-consistent species richness [@brose2016; @gotelli2017] .
-<!-- potential indication of biological homogenization due to shifting ranges and/or invasive species-->
-While communities are clearly changing, our most common species-based
-approaches do not fully capture the nature of that change.
+Ecological communities are experiencing unprecedented change as a result of anthropogenic pressures such as climate change, land use change, and invasive species.
+Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011], and fundamental changes in some of the most well-studied systems (e.g.
+coral bleaching).
+At the local scale however, species diversity tells a different story.
+Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of a significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017] .
+<!-- potential indication of biological homogenization due to shifting ranges and/or invasive species--> While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.
 
-Functional diversity offers a potentially powerful alternative for
-detecting and describing community change by providing a mechanistic
-link between the organisms in a system and the processes they perform
-[@mcgill2006]. By describing the functional trait space rather than
-species, functional diversity metrics capture the disproportionate
-impact of losses or gains of functionally unique species. Functional
-diversity metrics are therefore particularly well-suited for addressing
-change in communities with constant species richness.
+Functional diversity offers a potentially powerful alternative for detecting and describing community change by providing a mechanistic link between the organisms in a system and the processes they perform [@mcgill2006].
+By describing the functional trait space rather than species, functional diversity metrics capture the disproportionate impact of losses or gains of functionally unique species.
+Functional diversity metrics are therefore particularly well-suited for assessing the implications of community shifts in the full range of species change scenarios.
 
-Loss of functional diversity is also linked to impacts on critical
-ecosystem and community processes. For example, there is increasing
-evidence that functional diversity is a better predictor of ecosystem
-function than species-based metrics [@cadotte2011; @gagic2015], and that
-different facets of functional diversity play essential roles in
-maintaining ecosystem stability [@craven2018; @morin2014]. Further,
-maintenance of functional diversity is key for maximizing adaptive
-potential of communities under climate change (citation).
+Trends in functional composition also have important implications for how we assesses potential changes in ecosystem stability, function, and resilience.
+Functional loss is often implicit in work linking biodiversity loss to impacts on ecosystem processes despite a wide range of documented relationships between functional and species change in natural communities [@biswas2011].
+<!-- For example, hypotheses underpining the biodiversity ecosystem function relationship: insert function/stability theory here.
+--> Where functional diversity is explicitly considered, there is increasing evidence it is a better predictor of ecosystem function than species-based metrics [@cadotte2011; @gagic2015], and that different facets of functional diversity play essential roles in maintaining ecosystem stability [@craven2018; @morin2014].
+Tracking the relationship between species and functional changes across a wide variety of communities therefore provides an opportunity to assess not only the prevalence of concurrent species and functional loss, but also the evidence underpinning hypothesis between biodiversity loss and ecosystem processes.
 
-<!-- Functional loss is often implicit in work linking species loss to impacts on ecosystem processes despite a wide range of documented relationships between functional and species change [@biswas2011].
-Examining functional and species change side by side across a variety of communities -->
+<!-- paragraph on maintenence of functional structure through time -->
 
-<!-- What can we say about how global change is impacting:stability, resilience, function, assembly, ecosystem processes (and ecosystem services) - need a functional grounding for all those concepts.
-Link specific patterns in FD metrics to impacts on those concepts.
--->
+Assessments of broad-scale temporal change in functional diversity have previously been limited by a lack of functional trait data.
+The majority of work has therefore focused largely on system-specific studies with traits collected *in situ*.
+Ongoing efforts to assemble functional traits for a variety of taxa have made synthesis of existing community assemblage data and functional traits possible for the first time, providing initial insights into the ways functional diversity changes on a broad scale for specific taxa [e.g.
+fish @trindade-santos2020, birds @jarzyna2016; @barnagaud2017].
+<!-- so what do these studies say? does that go here? --> However, to date there has been no cross-taxa assessment of temporal functional change for a broad geographic and taxonomic extent.
 
-<!-- paragraph about what we already know about global functional changes -->
-
-Assessments of broad-scale temporal change in functional diversity have
-previously been limited by a lack of functional trait data. The majority
-of work has therefore focused largely on system-specific studies with
-traits collected *in situ*. Ongoing efforts to assemble functional
-traits for a variety of taxa have made synthesis of existing community
-assemblage data and functional traits possible for the first time,
-providing initial insights into the ways functional diversity changes on
-a broad scale for specific taxa (e.g. fish[@trindade-santos2020], birds
-[@jarzyna2016; @barnagaud2017]).
-<!-- so what do these studies say? does that go here? --> However, to
-date there has been no cross-taxa assessment of temporal functional
-change for a broad geographic and taxonomic extent.
-
-Here we perform the first multi-taxa, multi-system assessment of
-functional diversity change through time. We assess thousands of mammal,
-bird, and amphibian functional diversity timeseries to 1) determine how
-functional shifts compares to species change and 2) assess the types of
-functional shifts that are most common across communities.
+Here we perform the first multi-taxa, multi-system assessment of functional diversity change through time.
+We assess thousands of mammal, bird, and amphibian functional diversity timeseries to 1) determine how functional shifts compares to species change and 2) assess the types of functional shifts that are most common across communities.
 <!-- While our data represent the largest collection of functional diversity timeseries to date, they are not a representative sample of systems experiencing biodiversity change.
@@
 
-We obtained mammal, bird, and amphibian timeseries from the Biotime
-database, a global repository of high quality community assemblage
-timeseries. All studies included in the database follow consistent
-sampling protocols and represent full assemblages rather than population
-of single species [@dornelas2018]. Following best practices for the
-database [@blowes2019], studies with multiple sample locations were
-split into individual timeseries following a standardized spatial scale.
-Scale was set by a global grid with cell size determined based on the
-sample extent of studies with only a single location (see
-[@dornelas2018] for details on how sample extents were defined), with
-the area of each cell set to one standard deviation away from the mean
-of the single extent locations. All samples from a study within a single
-cell were considered to be a single timeseries, and species abundances
-were combined for all samples.
+We obtained mammal, bird, and amphibian timeseries from the Biotime database, a global repository of high quality community assemblage timeseries.
+All studies included in the database follow consistent sampling protocols and represent full assemblages rather than populations of single species [@dornelas2018].
+Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual timeseries following a standardized spatial scale.
+Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018] for details on how sample extents were defined), with the area of each cell set to one standard deviation away from the mean of the single extent locations.
+All samples from a study within a single cell were considered to be a single timeseries, and species abundances were combined for all samples.
 
-We gathered trait data from the Elton Trait Database (mammals and birds;
-[@wilman2014]) and Amphibio (amphibians, [@oliveira2017]). These
-databases include species-level means for traits that best define
-species' function in the community including body size, diet, and
-behavioral characteristics<!-- what's in the amphibian database? -->.
-For the full list of traits included in the analysis see the
-supplemental materials.
+```{r echo = FALSE, results = "asis"}
+tibble(Taxa = c("Mammals", "Birds", "Amphibians"),
+       Timeseries_Count = c(48, 2380, 11),
+       Species_Count = c(184, 700, 184),
+       Trait_Source = c("Elton Traits", "Elton Traits", "Amphibio"),
+       Traits = c("body mass, diet, active diel period", "body mass, diet, nocturnality, \nforest foraging strata, pelagic specialist", "habitat, diet, active diel period, \nactivity seasonality, body mass, body length, \nmin maturation size, max maturation size, \nmin offspring size, max offspring size, \nreproductive output, breeding strategy")
+) %>%
+  mutate_all(linebreak, align = "l") %>% 
+  knitr::kable(format = "latex", col.names = c("Taxa", "Number of \nTimeseries", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
+               caption = "Summary of the data in the final trait database.") %>%
+   kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
+   #kableExtra::column_spec(2, width = "5cm") #%>%
+  # kableExtra::column_spec(4, width = "10cm")
+```
+
+We gathered trait data from the Elton Trait Database [mammals and birds, @wilman2014] and Amphibio [amphibians, @oliveira2017].
+These databases include species-level means for traits that best define species' function in the community including body size, diet, and behavioral characteristics.
+For the full list of traits included in the analysis for each taxa see Table 1.
 <!-- limitations of species level trait data? -->
 
-In order to ensure taxonomic consistency across datasets, Biotime
-species were paired with trait data based on their species identifier
-from the Integrated Taxonomic Information System database (citation),
-obtained through the taxadb R package [@norman2020]. If more than one
-species in the assemblage data resolved to the same identifier,
-observations were considered the same species. For trait data, traits
-for all species of the same identifier were averaged. Only studies with
-at least 75% trait coverage were included. In order to have a sufficient
-number of species to calculate functional diversity metrics, years with
-fewer than 4 species observed were excluded.
+In order to ensure taxonomic consistency across datasets, Biotime species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (citation), obtained through the taxadb R package [@norman2020].
+If more than one species in the assemblage data resolved to the same identifier, observations were considered the same species.
+For trait data, traits for all species of the same identifier were averaged.
+Only studies with at least 75% trait coverage were included and observations for species with no trait data were excluded.
+In order to have a sufficient number of species to calculate functional diversity metrics, years with fewer than 5 species observed were also excluded.
 
-Many studies had a variable number of samples within years. To account
-for this inconsistency in sampling effort we used sample-based
-rarefaction by bootstrap resampling within years for each timeseries
-based on the smallest number of samples in a year for that timeseries.
+Many studies had a variable number of samples within years.
+To account for this inconsistency in sampling effort we used sample-based rarefaction by bootstrap resampling within years for each timeseries based on the smallest number of samples in a year for that timeseries.
 
-Our final dataset included 2,439 timeseries from 54 studies in 21
-countries and 15 biomes <!-- list traits -->. Only four studies
-(consisting of 11 timeseries) came from Amphibian studies due to the
-limited availability of amphibian timeseries and low species richness
-values for assemblages. For a full breakdown of studies and their
-characteristics, see the supplement.
+Our final dataset included 2,443 timeseries from 53 studies in 21 countries and 15 biomes and 13 different traits (Fig 1).
+Only four studies (consisting of 11 timeseries) came from Amphibian studies due to the limited availability of amphibian timeseries and low species richness values for assemblages (Table 1).
+For a full breakdown of studies and their characteristics, see the supplement.
+
+```{r taxa-map, fig.cap="Map of timeseries locations with points color corresponding to taxa.", echo=FALSE}
+knitr::include_graphics(here::here("figures/study_map.jpeg"))
+```
 
@@
 
-We calculated yearly metrics of functional and species diversity for
-each timeseries. Species-based metrics include species richness (*S*)
-and Jaccard similarity (*J*) as a measure of turnover. Jaccard
-similarity was calculated relative to the first observed year for a
-timeseries. A negative trend in *J* would therefore indicate increasing
-turnover.
+We calculated yearly metrics of functional and species diversity for each timeseries.
+Species-based metrics include species richness (*S*) and Jaccard similarity (*J*) as a measure of turnover.
+Jaccard similarity was calculated relative to the first observed year for a timeseries.
+A negative trend in *J* would therefore indicate increasing turnover.
 
-Functional diversity metrics were calculated using the *dbFD* function
-from the *FD* R package. Here we report functional richness (*FRic*),
-functional evenness (*FEve*), and functional diveregence (*FDiv*) which
-together describe three complementary characteristics of the functional
-space [@mason2005]. *FRic* assesses the volume of the trait space
-occupied by species in the community. *FEve* describes how species are
-distributed across the trait space and how abundance is distributed
-across species. *FDiv* measures the degree to which species and their
-abundances maximize differences in the functional space. Higher values
-of *FDiv* therefore correspond to communities where many highly abundant
-species are on the edges of the trait space.
+Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package.
+Here we report functional richness (*FRic*), functional evenness (*FEve*), and functional diveregence (*FDiv*) which together describe three complementary characteristics of the functional space [@mason2005].
+*FRic* assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
+*FEve* describes how species are distributed across the trait space and how abundance is distributed across species.
+Higher values of *FEve* indicate more even spacing of species in the trait space and individuals across species.
+*FDiv* measures the degree to which species and their abundances maximize differences in the functional space.
+Higher values of *FDiv* therefore correspond to communities where many highly abundant species are on the edges of the trait space.
+<!-- use these metrics because they're common, not because of their results -->
 
-All available trait data for each study were included in functional
-diversity calculations with the exception of traits that were the same
-value for all observed species in the study. All continuous traits were
-z-score scaled to give each trait equal weight in the trait space
-[@leps2006; @schleuter2010]. The number of trait axes was limited to the
-maximum number of traits that fulfills the criteria $s >= s^t$, where
-$s$ is the number of species and \$t\$ is the number of traits. This
-restriction allows for a sufficient number of axes to capture the trait
-space while maintaining computational feasibility. Metrics incorporated
-weighting based on species abundance where available.
+All available trait data for each study were included in functional diversity calculations with the exception of traits that were the same value for all observed species in the study.
+All continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
+The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= s^t$, where $s$ is the number of species and \$t\$ is the number of traits.
+This restriction allows for a sufficient number of axes to capture the trait space while maintaining computational feasibility.
+Metrics incorporated weighting based on species abundance where available.
 
@@
 
-We corrected for bias in functional diversity metrics due to changes in
-species richness by calculating the standardized effect size (SES) for
-each metric from null estimates [@swenson2012]. Null model corrections
-allow us to assess the degree to which the observed functional diversity
-metric deviates from the value expected by chance in a randomly
-assembled community. Null estimates were calculated for each rarefied
-sample by randomly sampling species from the species pool for each year
-and randomly assigning observed abundances to species. Species pools
-included all species observed for a timeseries. This process was
-repeated 500 times to get an estimate and standard deviation of the null
-expectation for the metric for that timeseries. We used these values to
-calculate SES using the following formula:
-$SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$. Unless otherwise
-specified, functional diversity results are reported for standardized
-values for the remainder of the paper.
+We corrected for bias in functional diversity metrics due to changes in species richness by calculating the standardized effect size (SES) for each metric from null estimates [@swenson2012].
+Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
+Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species.
+Species pools included all species observed for a timeseries.
+This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for that timeseries.
+We used these values to calculate SES using the following formula: $SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
+Unless otherwise specified, functional diversity results are reported for standardized values for the remainder of the paper.
 
@@
 
-We estimated overall trends for each diversity metric using a Linear
-Mixed Effects model with a random slope and intercept for each
-timeseries nested within the study. Differences in trends by taxa,
-biome, and realm were each modeled individually as covariates in models
-of the same structure.
+We estimated overall trends for each diversity metric using a Linear Mixed Effects model with a random slope and intercept for each timeseries nested within the study.
+Differences in trends by taxa, biome, and realm were each modeled individually as covariates in models of the same structure.
 
@@
 
-We obtained study-level estimates of temporal change from the Best
-Linear Unbiased Predictors (BLUPs) for each overall trend model and
-assessed them in two ways. First, we looked at pairwise relationships
-between all metrics to assess what kinds of community shifts are
-commonly found together. We obtained correlations and bootstrapped
-p-values using *corr.test* and *corCI* functions in the *psych* R
-package (citation). Second, we assessed the relationship between all
-metrics simultaneously by performing a cluster analysis which grouped
-each study based on their similarities across all metrics. We used a
-hierarchical clustering algorithm through the function *hclust* from the
-R package *stats.* We included all standardized functional diversity
-metrics, species richness, and Jaccard similarity in the analysis. We
-report a qualitative interpretation of the groups from the cluster
-analysis.
+We obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model and assessed them in two ways.
+First, we looked at pairwise relationships between all metrics to assess what kinds of community shifts are commonly found together.
+We obtained correlations and bootstrapped p-values using *corr.test* and *corCI* functions in the *psych* R package (citation).
+Second, we assessed the relationship between all metrics simultaneously by performing a cluster analysis which grouped each study based on their similarities across all metrics.
+We used a hierarchical clustering algorithm through the function *hclust* from the R package *stats.* We included all standardized functional diversity metrics, species richness, and Jaccard similarity in the analysis.
+We report a qualitative interpretation of the groups from the cluster analysis.
 
-<!-- why do some studies not have blup estimates for FRic?-->
+```{r timeseries-plot, fig.cap="Plots of timeseries-level trends with line color corresponding to climatic region, with data points in grey and the overall metric mean in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES", echo=FALSE}
+knitr::include_graphics(here::here("figures/3met_long.jpeg"))
+```
 
@@
 
-We found no significant temporal trend in species richness or functional
-diversity metrics for either observed or standardized metrics. There was
-also no trend for studies broken down by taxa, biome or realm. We did
-find a significant overall decrease in Jaccard similarity, indicating
-increasing turnover through time.
+<!-- Despite significant trends at the study level for the majority of studies,-->
 
-At the study level, the majority of studies (78%) experienced a
-significant temporal change in species richness and all functional
-diversity metrics. 92% of sites experienced a significant trend for
-species richness, 88% for functional functional richness, 96% for
-functional divergence, 88% for functional evenness, and 35% for Jaccard
-similarity. Only 4 (cite studies) out of 54 studies experienced no
-significant trend in species richness and were also characterized by
-significantly decreasing functional evenness. There was only one
-significant pairwise correlation between the metrics, a positive
-relationship between Functional Divergence and Functional Evenness,
-indicating that communities experiencing an increase in Divergence
-frequently experience an increase in Evenness.
+We found no significant overall trend in species richness or functional diversity metrics (Fig 2).
+There was also no temporal trend for studies broken down by taxa, biome or realm.
+We did find a significant overall decrease in Jaccard similarity, indicating increasing turnover through time.
+Non-significant overall trends indicate that an equal number of studies experienced increasing and decreasing trends in those metrics.
 
-Cluster analysis grouped all but a few studies into a cluster
-characterized by non-significant or very small significant slopes for
-all metrics. We identified four other clusters containing six studies
-(list studies) that were characterized by outlier trends in at least one
-of the metrics and near-zero trends in the rest. The first contained two
-studies with the most strongly decreasing evenness. The second, two
-studies with the most strongly increasing functional divergence and
-increasing functional richness. The third contained one study with
-strongly decreasing functional divergence and richness, and the fourth
-strongly decreasing species richness with increasing functional
-richness.
+At the study level, the majority of studies (86%) experienced a significant temporal change in species richness and all functional diversity metrics.
+92% of studies experienced a significant trend for species richness, 88% for functional richness, 98% for functional divergence, 96% for functional evenness, and 30% for Jaccard similarity.
+Only 4 (cite studies) out of 53 studies experienced no significant trend in species richness and were also characterized by significantly decreasing functional evenness.
+There was only one significant pairwise correlation between the metrics, a positive relationship between Functional Divergence and Functional Evenness, indicating that communities experiencing an increase in Divergence frequently experience an increase in Evenness.
+
+Cluster analysis grouped all but a few studies into a cluster characterized by non-significant or very small significant slopes for all metrics.
+We identified four other clusters containing six studies (list studies) that were characterized by outlier trends in at least one of the metrics and near-zero trends in the rest.
+The first contained two studies with the most strongly decreasing evenness.
+The second, two studies with the most strongly increasing functional divergence and increasing functional richness.
+The third contained one study with strongly decreasing functional divergence and richness, and the fourth strongly decreasing species richness with increasing functional richness.
 
@@
 
-Our study represents the largest broad-scale multi-taxa assessment of
-functional change through time, giving a first look at overall trends in
-functional diversity and the range of changes happening in mammal, bird,
-and amphibian communities. Consistent with previous species-based
-syntheses, we found no overall trend in species diversity accompanied by
-increasing turnover through time [@dornelas2018]. While nearly all of
-the studies included in our data experienced a significant increase or
-decrease in species richness, trends were observed in equal numbers
-resulting in an overall trend not significantly different from zero.
+<!-- Our study represents the largest broad-scale multi-taxa assessment of functional change through time, giving a first look at overall trends in functional diversity and the range of changes happening in mammal, bird, and amphibian communities.
+Consistent with previous species-based syntheses, we found no overall trend in species diversity accompanied by increasing turnover through time [@dornelas2018].
+While nearly all of the studies included in our data experienced a significant increase or decrease in species richness, trends were observed in equal numbers resulting in an overall trend not significantly different from zero.
+-->
 
-Surprisingly, we did not detect an overall trend in any of the observed
-or corrected functional diversity metrics. Similar to species richness,
-communities are experiencing a wide range of functional shifts that
-balance each other out. While our dataset likely under-samples areas
-with most direct human impact, anthropogenic change is hypothesized to
-impact even the most remote systems. Conservatively, our findings
-therefore indicate that in unperturbed mammal, bird, and amphibian
-communities there is no evidence of systemic functional richness loss or
-functional change. A lack of trend for data broken down by realm, biome,
-and taxonomic group gives further evidence that directional functional
-change is absent from all systems observed in our dataset.
+<!-- Surprisingly, we did not detect an overall trend in any of the functional diversity metrics.
+Similar to species richness, communities are experiencing a wide range of functional shifts that balance each other out.
+While our dataset likely under-samples areas with most direct human impact, anthropogenic change is hypothesized to impact even the most remote systems.
+Conservatively, our findings therefore indicate that in unperturbed mammal, bird, and amphibian communities there is no evidence of systemic functional richness loss or functional change.
+A lack of trend for data broken down by realm, biome, and taxonomic group gives further evidence that directional functional change is absent from all systems observed in our dataset.
+-->
 
-Individual studies in our dataset demonstrate the variety of species and
-functional shifts that can happen concurrently in a community. With only
-one significant pairwise relationship between Functional Divergence and
-Functional Evenness, all other possible combinations of species and
-functional change were observed in at least one study. Notably, there is
-no relationship between species richness change and change in functional
-richness, meaning that even for studies that may be experiencing species
-losses there is not necessarily an accompanying loss of functional
-richness and in some cases there may even be an increase. These
-communities are gaining new species that are functionally distinct
-enough to make up for functional losses due to species losses. This
-pattern would be further magnified if the species lost are not
-functionally distinct members of the assemblage.
+Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at overall trends in functional diversity and the range of changes happening in mammal, bird, and amphibian communities.
+Surprisingly, we did not detect an overall trend in any of the calculated functional diversity metrics.
+While nearly all of the studies included in our data experienced a significant increase or decrease in at least one of the metrics, trends were observed in equal numbers resulting in an overall trend that was not significantly different from zero.
+As with previous species-based syntheses, we also found no overall trend in species diversity accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics may be consistent with similar well-documented species derived trends.
+We find no evidence of systemic functional richness loss or functional change.
+A lack of trend for data broken down by realm, biome, and taxonomic group gives further evidence that directional functional change is absent from all systems observed in our dataset.
 
-The overall trend of increasing turnover is consistent with work
-indicating higher than expected background rates of turnover in
-contemporary communities [@gotelli2017; @magurran2018]. However, many
-studies did not experience increasing turnover (65%), suggesting that it
-is not universal. Even for those studies without a turnover trend,
-significant functional shifts occurred. While some communities may be
-changing more rapidly than usual, that turnover is not necessarily an
-indication of specific kind of functional change.
+Individual studies in our dataset demonstrate the variety of species and functional shifts that can happen concurrently in a community.
+With only one significant pairwise relationship between study-level functional divergence slopes and functional evenness slopes, all other possible combinations of species and functional change were observed in at least one study.
+Notably, there is no relationship between species richness change and change in functional richness, meaning that communities experiencing species losses do not necessarily experience accompanying functional richness, and in some cases may even experience an increase.
+<!-- need to put a disclaimer about what "increasing" means here - it's increased richness relative to the expectation for a community of that size --> These communities gain species that expand the boundaries of the trait space while losing interior species.
 
-Our results have important implications for how we approach work linking
-species loss to ecosystem processes, which often implicitly assume
-concurrent functional loss. There is no evidence of directional
-functional loss or even functional shifts accompany species loss in
-these natural communities.
-<!-- clarify that this is about relationship between concurrent patterns in a single community -->
-Building on a rich body of work linking species diversity changes to
-changes in ecosystem function, resilience, and stability (citations), it
-will be essential to tease apart effects due to species change from
-those due to functional shifts if we want to infer impacts in natural
-communities.
+The overall trend of increasing turnover is consistent with work indicating higher than expected background rates of turnover in contemporary communities [@gotelli2017; @magurran2018].
+However, many studies did not experience increasing turnover (65%), suggesting that it is not universal.
+Even for those studies without a turnover trend, significant functional shifts occurred.
+While some communities may be changing more rapidly than usual, that turnover is not necessarily an indication of specific kind of functional change.
+
+*Methodological Limitations*
+
+This study represents best practices in biodiversity synthesis and functional diversity analysis, however methodological limitations may still be obscuring potential systematic shifts in functional composition.
+First, as previously acknowledged, the Biotime database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
+While both of these challenges are pervasive in Ecology generally and could have potentially insidious implications for synthesis work, we can only use the data we have.
+We do however emphasize that our study included many timeseries that experienced a significant species loss (\~ 52%), indicating that this is not simply a synthesis of pristine systems.
+
+Second, despite using the most comprehensive trait databases for these taxa, we were still limited to species-level means of the traits deemed important by database creators.
+The importance of intraspecific variation is well documented [@desroches2018; @violle2012], however individual-level traits are rarely collected along side monitoring data, especially for the longest running efforts.
+Further investment in individual level trait data for long-term monitoring will be critical for assessing finer scale shifts that likely occur more commonly in communities and potentially preceded bigger changes.
+
+Third, while we use here the most common metrics for describing functional diversity there are a number of possible metrics for assessing a functional space, with new ones frequently proposed.
+There may be a significant shift in a characteristic of the functional space that was poorly captured by our chosen metrics.
+<!-- Further exploration of complementarity of traits... -->
+
+*Implications for Ecological Process*
+
+Our results highlight a gap in both empirical and theoretical work linking shifts in community composition to functional change.
+While a large body of work assesses study-level insights into temporal functional responses (citations), a macroecological perspective illustrates the urgent need for future work exploring the factors important in determining functional change under different species change scenarios.
+Due to the many previously discussed barriers to directly assessing functional shifts, a framework for predicting functional change from community characteristics would be a substantial step forward in understanding biodiversity change.
+
+Much of the work linking species change (or loss) to ecological process either implicitly or explicitly assumes functional loss follows from species loss in natural communities.
+For example ....
+This assumption is clearly refuted by our results.
+
+Our results have important implications for how we approach work linking species loss to ecosystem processes, which often implicitly assume concurrent functional loss.
+There is no evidence of directional functional loss or even functional shifts accompany species loss in these natural communities.
+<!-- clarify that this is about relationship between concurrent patterns in a single community --> Building on a rich body of work linking species diversity changes to changes in ecosystem function, resilience, and stability (citations), it will be essential to tease apart effects due to species change from those due to functional shifts if we want to infer impacts in natural communities.
+--\>
+
+<!-- call for theory development, punt back to community ecology - we don't know what this means! -->
+
+Our results emphasize the importance of directly assessing functional change rather than assuming functional loss accompanies species loss.
+
+-   getting functional data is costly, so we'd like to be able to say something about udnerlying functional composition without it
+
+-   list of studies where they make functional inference without only species level data
+
+Also illustrates the lack of theoretical development surrounding the relationship between functional and species change, and when you would expect different kinds of relationships.
+
+-   also theory around the implications of functional shifts for ecological processes
+
+**Notes**
 
@@
 
-As the first broad scale analysis of temporal functional diversity
-change we make x broad conclusions. First,
+As the first broad scale analysis of temporal functional diversity change we make x broad conclusions.
+First,
 
-Above all our study shows that patterns of functional change can not be
-inferred from
+Above all our study shows that patterns of functional change can not be inferred from
 
--   how are these findings consistent with expectations for hypothesized
-    shifts in functional change?
+-   how are these findings consistent with expectations for hypothesized shifts in functional change?
 
--   Biological homogenization is one of the hypothesized meaningful
-    forces
+-   Biological homogenization is one of the hypothesized meaningful forces
 
--   other unmeasured taxa may be harder hit? different patterns for fish
-    assemblages
+-   other unmeasured taxa may be harder hit?
+    different patterns for fish assemblages
 
@@
 
-While species-level metrics are the only functional data available to
-perform an analysis of this scale, they are unable to detect functional
-shifts in communities with little to no turnover as they do not capture
-intraspecific shifts.
+While species-level metrics are the only functional data available to perform an analysis of this scale, they are unable to detect functional shifts in communities with little to no turnover as they do not capture intraspecific shifts.
 
@@
 
--   species richness loss linked to impacts on processes, in doing so
-    often inferring functional loss, but the two do not go hand in hand
-    in natural communities
+-   species richness loss linked to impacts on processes, in doing so often inferring functional loss, but the two do not go hand in hand in natural communities
 
@@
 
+-   land use history and disturbance showing different relationships between species and functional diversity (Cadotte 2011)
+
 #### Add to supplement

### 2021-06-21 paper/manuscript.Rmd
Add Carl comments

@@
 ---
-title: "Functional timeseries reveal no systemic shifts for vertebrate communities"
+title: "Functional timeseries reveal no systemic shifts in vertebrate communities"
 author: 
@@
     affiliation: ucb, ropensci
+  - name: "Perry de Valpine"
+    affiliation: ucb
 address:
@@
 
-*Methodological Limitations*
+[@osullivan2021]
+
+<!-- Link turnover paragraph to background turnover rate, maintenence of species diversity (i.e.
+turnover is happening!
+what does that mean for speices div?
+maintained.
+for functional diversity?
+maintained.) -->
+
+*Implications for Future Work*
+
+Here's the prediction of the field, here is the data of the field, given that the data doesn't match the prediction it's on the field to resolve these issues.
+
+Why do we find no significant trend in the global data despite extensive predictions to the contrary from both theory and local studies?
+First, there are a variety of data limitations to consider:
 
 This study represents best practices in biodiversity synthesis and functional diversity analysis, however methodological limitations may still be obscuring potential systematic shifts in functional composition.
-First, as previously acknowledged, the Biotime database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
+First, the Biotime database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
 While both of these challenges are pervasive in Ecology generally and could have potentially insidious implications for synthesis work, we can only use the data we have.
 We do however emphasize that our study included many timeseries that experienced a significant species loss (\~ 52%), indicating that this is not simply a synthesis of pristine systems.
+
+You'd imagine the pattern would still be visible even with North American birds...
 
@@
 
-*Implications for Ecological Process*
+-   one opening sentence that links back to the null result
+
+Even with the right data and metrics, you may not see a result!
+Due to the wide variety of mechanisms at play we amy not expect to see a global trend even when the local trends are significant.
 
@@
 --\>
+
+We can not understand all functional loss in terms of species loss
+
+-   lean in to four points in the abstraction
+
+    -   raises questions for future data collection and theoretical development
+
+-   field that's full of idle speculation
+
+-   find papers that make claims we're "testing"
 
@@
 -   breakdown of number of studies/timeseries for each taxa/realm etc
+
+They are limitations of the field not our limitations
+
+-   acknowledge those things without taking the blame
+
+-   lean into it as a conclusion
+
+-   we found a null result and its your fault
+
+-   fixing x, y, z could resolve the null result, but there are good reasons it might not
+
+-   talk about balance of local trends as the fourth reason
+
+There are reasons we haven't got trait data, hoped there would be an easy way out, but these results show that there isn't
+
+Is there a null result for some deep theoretical reason or
+
+we could've been lucky that we would see the pattern with crappy data, but it isn't
+
+-   the thing we thought would be there isn't there, so what can we do to solve it
+
+clever way to do temporal coverage - maps by decade?
+
+-   number of data points through time (bar chart colored by taxa)
+
+figure 1.
+illustrating functional diversity metrics
+
+figure 2.
+the processing workflow

### 2021-06-30 analysis/visualization.Rmd
Add historgrams of the duration of timeseries.

@@
 ```{r}
+#library(forcats)
+
 #get rarefyid centers
 included_ids <- pins::pin_get("model-data", board = "github") %>%
-  select(rarefyID, taxa, climate, realm) %>%
+  select(rarefyID, taxa, climate, realm, duration) %>%
   distinct() %>% 
@@
 study_data <- pins::pin_get("rarefyID-cell-centre", board = "github") %>%
-  inner_join(included_ids)
+  inner_join(included_ids) #%>%
+  #mutate(taxa = as.factor(taxa), taxa = fct_relevel(taxa, "Birds"))
 
-map_data("world") %>%
+study_map <- map_data("world") %>%
   ggplot(aes(long, lat, group = group)) +
   geom_polygon(fill = "white", colour = "grey") + 
-  coord_map(xlim=c(-180,180), ylim=c(-57, 90)) +
-  geom_point(data = study_data, aes(rarefyID_x, rarefyID_y, color = taxa), alpha = 0.5, inherit.aes = FALSE) +
-  ggthemes::theme_map()
+  coord_map(xlim=c(-180,180), ylim=c(-57, 90), projection = "mercator") +
+  geom_point(data = study_data, aes(rarefyID_x, rarefyID_y, color = taxa), alpha = 0.3, inherit.aes = FALSE) +
+  ggthemes::theme_map() +
+  theme(legend.title  = element_blank())
 
@@
 ```
+
+```{r}
+duration_hist <- study_data %>% 
+  #filter(taxa == "Birds") %>%
+  ggplot(aes(duration)) + 
+  geom_histogram(aes(fill = taxa)) + 
+  theme_classic() + 
+  facet_wrap(~taxa, scales = "free") +
+  theme(strip.background = element_blank(), strip.text = element_blank(),
+        legend.position = "none")
+
+plot_grid(study_map, duration_hist, labels = "AUTO", nrow = 2, align = 'vh', hjust = -1, axis = "l", scale = c(1,0.9), rel_heights = c(2, 1))
+
+ggsave2(here::here("figures/study_map_hist.jpeg"))
+```

### 2021-06-30 analysis/05_model_metrics.Rmd
Models of study-level slope estimates. Minor changes to figures.

@@
 metric_model_table <- bind_rows(SES_rand_slope, s_mod, jacc_base_mod)
+
+#save coefficients and models
+usethis::use_data(metric_model_table)
+pins::pin(metric_model_table, board = "github")
 ```
@@
 ```{r}
-taxa_slope_model <- model_data %>%
-  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
+# taxa_slope_model <- model_data %>%
+#   filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
+#   group_nest(metric) %>%
+#   mutate(model = map(data, ~lmer(value ~ year_scaled + taxa + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
+#                                  data = .x, REML = TRUE)),
+#            coef = map(model, tidy)) %>%
+#   select(metric, coef) %>%
+#   unnest(cols = c(coef)) %>%
+#   #filter(term == "year_scaled") %>%
+#   filter(term %in% c("year_scaled", "taxaBirds", "taxaMammals")) %>%
+#   select(-c(effect, group, statistic))
+# 
+# taxa_slope_model_S <- model_data %>% 
+#   mutate(year_scaled = scale(year)) %>% 
+#   filter(metric == "S") %>%
+#   lmer(logvalue ~ year_scaled + taxa + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
+#                                  data = ., REML = TRUE)
+
+SES_rand_slope <- model_data %>%
+  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
   group_nest(metric) %>%
-  mutate(model = map(data, ~lmer(value ~ year_scaled + taxa + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
-                                 data = .x, REML = TRUE)),
-           coef = map(model, tidy)) %>%
-  select(metric, coef) %>%
+  mutate(#year_center = map(data, ~scale(.x$year)),
+         model = map(data, ~lmer(value ~ taxa + (1|study_id) + (|study_id:rarefyID), 
+                                 data = .x)),
+           # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
+           #                       data = .x)),
+           # test = map2(model, null_model, anova),
+           # test_res = map(test, tidy),
+           coef = map(model, tidy),
+           ) %>% 
+  select(metric, model, coef) %>%
   unnest(cols = c(coef)) %>%
-  #filter(term == "year_scaled") %>%
-  filter(term %in% c("year_scaled", "taxaBirds", "taxaMammals")) %>%
-  select(-c(effect, group, statistic))
-
-taxa_slope_model_S <- model_data %>% 
-  mutate(year_scaled = scale(year)) %>% 
-  filter(metric == "S") %>%
-  lmer(logvalue ~ year_scaled + taxa + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
-                                 data = ., REML = TRUE)
+  filter(term == "year_scaled") %>%
+  select(-c(effect, group, term, statistic))
 ```
@@
 #### Analizing slopes
+Is slope a function of timeseries characteristics?
+```{r}
+study_slope_data <- study_slopes %>%
+  select(study_id, metric, estimate) %>%
+  left_join(meta %>% 
+              select(-rarefyid) %>% 
+              distinct() %>%
+              mutate(study_id = as.character(study_id))) %>%
+  left_join(model_data %>% select(study_id, startyear, duration) %>% distinct())
+
+#taxa model
+taxa_slope <- study_slope_data %>%
+  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
+  group_nest(metric) %>%
+  mutate(#year_center = map(data, ~scale(.x$year)),
+         #model = map(data, ~lmer(estimate ~ taxa + (1|taxa), 
+         model = map(data, ~glm(estimate ~ taxa, 
+                                 data = .x)),
+           coef = map(model, tidy),
+           ) %>% 
+  select(metric, coef) %>%
+  unnest(cols = c(coef)) %>%
+  mutate(covariate = "taxa")
+
+climate_slope <- study_slope_data %>%
+  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
+  group_nest(metric) %>%
+  mutate(#year_center = map(data, ~scale(.x$year)),
+         #model = map(data, ~lmer(estimate ~ taxa + (1|taxa), 
+         model = map(data, ~glm(estimate ~ climate, 
+                                 data = .x)),
+           coef = map(model, tidy),
+           ) %>% 
+  select(metric, coef) %>%
+  unnest(cols = c(coef)) %>%
+  mutate(covariate = "climate")
+
+realm_slope <- study_slope_data %>%
+  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
+  group_nest(metric) %>%
+  mutate(#year_center = map(data, ~scale(.x$year)),
+         #model = map(data, ~lmer(estimate ~ taxa + (1|taxa), 
+         model = map(data, ~glm(estimate ~ realm, 
+                                 data = .x)),
+           coef = map(model, tidy),
+           ) %>% 
+  select(metric, coef) %>%
+  unnest(cols = c(coef)) %>%
+  mutate(covariate = "realm")
+
+duration_slope <- study_slope_data %>%
+  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
+  group_nest(metric) %>%
+  mutate(#year_center = map(data, ~scale(.x$year)),
+         #model = map(data, ~lmer(estimate ~ taxa + (1|taxa), 
+         model = map(data, ~glm(estimate ~ duration, 
+                                 data = .x)),
+           coef = map(model, tidy),
+           ) %>% 
+  select(metric, coef) %>%
+  unnest(cols = c(coef)) %>%
+  mutate(covariate = "duration")
+
+startyear_slope <- study_slope_data %>%
+  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
+  group_nest(metric) %>%
+  mutate(#year_center = map(data, ~scale(.x$year)),
+         #model = map(data, ~lmer(estimate ~ taxa + (1|taxa), 
+         model = map(data, ~glm(estimate ~ startyear, 
+                                 data = .x)),
+           coef = map(model, tidy),
+           ) %>% 
+  select(metric, coef) %>%
+  unnest(cols = c(coef)) %>%
+  mutate(covariate = "startyear")
+
+#All model estimates in one dataframe
+slope_models <- bind_rows(taxa_slope, realm_slope, climate_slope, duration_slope, startyear_slope)
+
+# Which predictor variables were significant?
+slope_models %>% filter(term != "(Intercept)", p.value < 0.05) %>% View()
+```
+
 Pairwise correlation analysis
@@
     # geom_point(aes(color = climate, group = rarefyID)) +
-    geom_point(aes(group = rarefyID) ,color = "grey", size = 0.50) +
+    geom_point(aes(group = rarefyID) ,color = "grey", size = 0.25) +
     geom_smooth(aes(color = climate, group = study_id), method = "glm", se = FALSE, size = 0.7) +
@@
 
-######
-group.colors <- c(A = "#333BFF", B = "#CC6600", C ="#9633FF", D = "#E2FF33", E = "#E3DB71")
-
-simplePlot <- function(DT, tit) 
-  ggplot(DT ,aes(x=Name, y=Value, fill=Group)) +
-  geom_bar(stat="identity") + xlab("") + ggtitle(tit) +
-  #Specify colours
-  scale_fill_manual(values=group.colors)
-#####
-
 # extract the legend from one of the plots
@@
 # the width of one plot (via rel_widths).
-plot_join + draw_grob(legend,  2/3.3, -0.3, .3/3.3, 1, scale = 0.9)
+#plot_join + draw_grob(legend,  2/3.3, -0.3, .3/3.3, 1, scale = 0.6)
+plot_join + draw_grob(legend,  2/3.3, -0.3,.3/3.3, 1, scale = 0.01)
 

### 2021-06-30 analysis/05_model_metrics.Rmd
Get rid of old model attempts to include study attributes.

@@
 pins::pin(metric_model_table, board = "github")
-```
-
-Kitchen sink model
-```{r}
-all_model <- model_data %>%
-  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv", "S", "Jaccard_base")) %>%
-  group_nest(metric) %>%
-  mutate(model = map(data, ~lmer(value ~ year_scaled + taxa + realm + (year_scaled|study_id/rarefyID), 
-                                 data = .x)),
-           coef = map(model, tidy)
-           ) %>%
-  select(metric, model, coef) %>%
-  unnest(cols = c(coef)) %>%
-  filter(term == "year_scaled") %>%
-  select(-c(effect, group, term, statistic))
-```
-Models with categorical study characteristics as random effects
-
-Taxa:
-```{r}
-# taxa_slope_model <- model_data %>%
-#   filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
-#   group_nest(metric) %>%
-#   mutate(model = map(data, ~lmer(value ~ year_scaled + taxa + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
-#                                  data = .x, REML = TRUE)),
-#            coef = map(model, tidy)) %>%
-#   select(metric, coef) %>%
-#   unnest(cols = c(coef)) %>%
-#   #filter(term == "year_scaled") %>%
-#   filter(term %in% c("year_scaled", "taxaBirds", "taxaMammals")) %>%
-#   select(-c(effect, group, statistic))
-# 
-# taxa_slope_model_S <- model_data %>% 
-#   mutate(year_scaled = scale(year)) %>% 
-#   filter(metric == "S") %>%
-#   lmer(logvalue ~ year_scaled + taxa + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
-#                                  data = ., REML = TRUE)
-
-SES_rand_slope <- model_data %>%
-  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
-  group_nest(metric) %>%
-  mutate(#year_center = map(data, ~scale(.x$year)),
-         model = map(data, ~lmer(value ~ taxa + (1|study_id) + (|study_id:rarefyID), 
-                                 data = .x)),
-           # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
-           #                       data = .x)),
-           # test = map2(model, null_model, anova),
-           # test_res = map(test, tidy),
-           coef = map(model, tidy),
-           ) %>% 
-  select(metric, model, coef) %>%
-  unnest(cols = c(coef)) %>%
-  filter(term == "year_scaled") %>%
-  select(-c(effect, group, term, statistic))
-```
-Realm:
-```{r}
-realm_slope_model <- model_data %>%
-  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
-  group_nest(metric) %>%
-  mutate(model = map(data, ~lmer(value ~ year_scaled + realm + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
-                                 data = .x, REML = TRUE)),
-           coef = map(model, tidy)) %>%
-  select(metric, coef) %>%
-  unnest(cols = c(coef)) %>%
-  #filter(term == "year_scaled") %>%
-  filter(effect == "fixed") %>%
-  select(-c(effect, group, statistic))
-
-realm_slope_model_S <- model_data %>% 
-  mutate(year_scaled = scale(year)) %>% 
-  filter(metric == "S") %>%
-  lmer(logvalue ~ year_scaled + realm + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
-                                 data = ., REML = TRUE)
-```
-Climate:
-```{r}
-climate_slope_model <- model_data %>%
-  filter(metric %in%   c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
-  group_nest(metric) %>%
-  mutate(model = map(data, ~lmer(value ~ year_scaled + climate + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
-                                 data = .x, REML = TRUE)),
-           coef = map(model, tidy)) %>%
-  select(metric, coef) %>%
-  unnest(cols = c(coef)) %>%
-  #filter(term == "year_scaled") %>%
-  filter(effect == "fixed") %>%
-  select(-c(effect, group, statistic))
-
-climate_slope_model_S <- model_data %>% 
-  mutate(year_scaled = scale(year)) %>% 
-  filter(metric == "S") %>%
-  lmer(logvalue ~ year_scaled + climate + (year_scaled|study_id/rarefyID), # + (year_scaled|taxa), 
-                                 data = ., REML = TRUE)
 ```

### 2021-06-30 analysis/visualization.Rmd
Minor label changes

@@
   facet_wrap(~taxa, scales = "free") +
-  theme(strip.background = element_blank(), strip.text = element_blank(),
-        legend.position = "none")
+  xlab("Timeseries Duration") +
+  theme(strip.background = element_blank(), 
+        strip.text = element_blank(),
+        legend.position = "none", 
+        axis.title = element_text(size=8))
 

### 2021-07-21 paper/manuscript.Rmd
small comment

@@
 
-Trends in functional composition also have important implications for how we assesses potential changes in ecosystem stability, function, and resilience.
+Beyond simply characterizing changes in community structure, trends in functional composition also have important implications for ecosystem stability, function, and resilience.
+There is increasing evidence functional diversity is a better predictor of ecosystem function than species-based metrics [@cadotte2011; @gagic2015], and that different facets of functional diversity play essential roles in maintaining ecosystem stability [@craven2018; @morin2014].
+Indeed, almost all hypothesized mechanisms underpinning the relationship between species diversity and ecosystem function are trait-dependent [@hillebrand2009].
+Determining functional trends therefore gives a more complete picture of potential trends in critical ecosystem processes.
+
+While functional loss is frequently cited as one of the most pressing concerns of the anthropocene (CITATIONS), observed and projected functional diversity trends are less certain.
+Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020].
+And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@matuoka2020].
+Overall trends are therefore necessary...
+Clearly functional loss is not inevitable in scenarios of perturbation or even species loss [@daz2001].
+
+<!-- trends in functional composition also have important implications for how we assesses potential changes in ecosystem stability, function, and resilience.
 Functional loss is often implicit in work linking biodiversity loss to impacts on ecosystem processes despite a wide range of documented relationships between functional and species change in natural communities [@biswas2011].
-<!-- For example, hypotheses underpining the biodiversity ecosystem function relationship: insert function/stability theory here.
---> Where functional diversity is explicitly considered, there is increasing evidence it is a better predictor of ecosystem function than species-based metrics [@cadotte2011; @gagic2015], and that different facets of functional diversity play essential roles in maintaining ecosystem stability [@craven2018; @morin2014].
+Where functional diversity is explicitly considered, there is increasing evidence it is a better predictor of ecosystem function than species-based metrics [@cadotte2011; @gagic2015], and that different facets of functional diversity play essential roles in maintaining ecosystem stability [@craven2018; @morin2014].
 Tracking the relationship between species and functional changes across a wide variety of communities therefore provides an opportunity to assess not only the prevalence of concurrent species and functional loss, but also the evidence underpinning hypothesis between biodiversity loss and ecosystem processes.
+-->
 
@@
 Unless otherwise specified, functional diversity results are reported for standardized values for the remainder of the paper.
+
+<!-- description of edge case where community = species pool -->
 

### 2021-07-26 analysis/conceptual_fig.Rmd
First commit

---
title: "Conceptual FD Figure"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(dplyr)
```

Plot a normal curve
```{r}
ggplot(data.frame(x = c(-5, 5)), aes(x)) + 
  stat_function(fun = dnorm, args = list(mean = 0, sd = .9), col='red', xlim = c(-3, 3)) +
  stat_function(fun = dnorm, args = list(mean = 1, sd = .5), col='blue', xlim = c(-2, 4)) +
  xlim(c(-5, 5)) +
  ylim(c(0,1)) +
  theme_classic() +
  theme(axis.text.x=element_blank(),
        #axis.ticks.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

Richness plots 
```{r}
get_xlim <- function(center, edge){
  return(list(center - edge, center + edge))
}

low_rich <- 
  ggplot(data.frame(x = c(-5, 5)), aes(x)) + 
  stat_function(fun = dnorm, args = list(mean = -3, sd = .9), xlim = c(-5, -1)) +
  stat_function(fun = dnorm, args = list(mean = 3, sd = .9), xlim = c(1, 5)) +
  xlim(c(-5, 5)) +
  #ylim(c(0,1)) +
  scale_y_continuous(expand = expansion(mult = c(0, .1)), limits = c(.05, 1)) +
  theme_classic() +
  ylab("S") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_text(margin = margin(t = 0, r = 50, b = 0, l = 0)),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  ggtitle("Low")

low_rich

high_rich <- 
  ggplot(data.frame(x = c(-5, 5)), aes(x)) + 
  #stat_function(fun = dnorm, args = list(mean = -3, sd = .9), xlim = c(-5, -1)) +
  #stat_function(fun = dnorm, args = list(mean = 3, sd = .9), xlim = c(1, 5)) +
  pmap(tibble(mean = c(-3, -2, -1.5, 0, 1.25, 3), 
  sd = rep(0.9, 6)) %>%
    rowwise %>%
    mutate(xlim = list(get_xlim(mean, 2))),
  function(mean, sd, xlim) stat_function(fun = dnorm, args = list(mean = mean, sd = sd), xlim = unlist(xlim))
  )  +
  xlim(c(-5, 5)) +
  #ylim(c(0,1)) +
  scale_y_continuous(expand = expansion(mult = c(0, .1)), limits = c(.05, 1)) +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  ggtitle("High")
```

Functional Richness
```{r}
low_fric <- 
  ggplot(data.frame(x = c(-5, 5)), aes(x)) + 
  #stat_function(fun = dnorm, args = list(mean = -3, sd = .9), xlim = c(-5, -1)) +
  #stat_function(fun = dnorm, args = list(mean = 3, sd = .9), xlim = c(1, 5)) +
  pmap(tibble(mean = c(-.5, 0,.25), 
  sd = rep(0.9, 3)) %>%
    rowwise %>%
    mutate(xlim = list(get_xlim(mean, 2))),
  function(mean, sd, xlim) stat_function(fun = dnorm, args = list(mean = mean, sd = sd), xlim = unlist(xlim))
  )  +
  xlim(c(-5, 5)) +
  #ylim(c(0,1)) +
  scale_y_continuous(expand = expansion(mult = c(0, .1)), limits = c(.05, 1)) +
  theme_classic() +
  ylab("FRic") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_text(margin = margin(t = 0, r = 50, b = 0, l = 0)),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())

low_fric

high_fric <- 
  ggplot(data.frame(x = c(-5, 5)), aes(x)) + 
  #stat_function(fun = dnorm, args = list(mean = -3, sd = .9), xlim = c(-5, -1)) +
  #stat_function(fun = dnorm, args = list(mean = 3, sd = .9), xlim = c(1, 5)) +
  pmap(tibble(mean = c(-3, -1.5, 1, 3), 
  sd = rep(0.9, 4)) %>%
    rowwise %>%
    mutate(xlim = list(get_xlim(mean, 2))),
  function(mean, sd, xlim) stat_function(fun = dnorm, args = list(mean = mean, sd = sd), xlim = unlist(xlim))
  )  +
  xlim(c(-5, 5)) +
  #ylim(c(0,1)) +
  scale_y_continuous(expand = expansion(mult = c(0, .1)), limits = c(.05, 1)) +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())

high_fric
```

Evenness 
```{r}
low_eve <- 
  ggplot(data.frame(x = c(-5, 5)), aes(x)) + 
  #stat_function(fun = dnorm, args = list(mean = -3, sd = .9), xlim = c(-5, -1)) +
  #stat_function(fun = dnorm, args = list(mean = 3, sd = .9), xlim = c(1, 5)) +
  pmap(tibble(mean = c(-3, -2, -1.5, 0, 1.25, 3), 
  sd = c(1.25, 0.5, 1.25, 1.25, 0.65, 1.25),
  plot_range = c(2.5, 2, 2.5, 2.5, 2, 2.5)) %>%
    rowwise %>%
    mutate(xlim = list(get_xlim(mean, plot_range))) %>%
    select(-plot_range),
  function(mean, sd, xlim) stat_function(fun = dnorm, args = list(mean = mean, sd = sd), xlim = unlist(xlim))
  )  +
  xlim(c(-5.25, 5.25)) +
  #ylim(c(0,1)) +
  scale_y_continuous(expand = expansion(mult = c(0, .1)), limits = c(.05, 1)) +
  theme_classic() +
  ylab("FEve") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_text(margin = margin(t = 0, r = 50, b = 0, l = 0)),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())

low_eve

high_eve <- ggplot(data.frame(x = c(-5, 5)), aes(x)) + 
  #stat_function(fun = dnorm, args = list(mean = -3, sd = .9), xlim = c(-5, -1)) +
  #stat_function(fun = dnorm, args = list(mean = 3, sd = .9), xlim = c(1, 5)) +
  pmap(tibble(mean = c(-3, -2, -1.5, 0, 1.25, 3), 
  sd = rep(0.9, 6)) %>%
    rowwise %>%
    mutate(xlim = list(get_xlim(mean, 2))),
  function(mean, sd, xlim) stat_function(fun = dnorm, args = list(mean = mean, sd = sd), xlim = unlist(xlim))
  )  +
  xlim(c(-5, 5)) +
  #ylim(c(0,1)) +
  scale_y_continuous(expand = expansion(mult = c(0, .1)), limits = c(.05, 1)) +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

Divergence 
```{r}
low_div <-  
ggplot(data.frame(x = c(-5, 5)), aes(x)) + 
  #stat_function(fun = dnorm, args = list(mean = -3, sd = .9), xlim = c(-5, -1)) +
  #stat_function(fun = dnorm, args = list(mean = 3, sd = .9), xlim = c(1, 5)) +
  pmap(tibble(mean = c(-3, -2, -1.5, 0, 1.25, 3), 
  sd = c(0.9, 0.9, 0.55, 0.5, 0.9, 0.9)) %>%
    rowwise %>%
    mutate(xlim = list(get_xlim(mean, 2))),
  function(mean, sd, xlim) stat_function(fun = dnorm, args = list(mean = mean, sd = sd), xlim = unlist(xlim))
  )  +
  xlim(c(-5, 5)) +
  #ylim(c(0,1)) +
  scale_y_continuous(expand = expansion(mult = c(0, .1)), limits = c(.05, 1)) +
  theme_classic() +
  ylab("FDiv") +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_text(margin = margin(t = 0, r = 50, b = 0, l = 0)),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())

low_div

-.5, 0,.25
high_div <- 
ggplot(data.frame(x = c(-5, 5)), aes(x)) + 
  #stat_function(fun = dnorm, args = list(mean = -3, sd = .9), xlim = c(-5, -1)) +
  #stat_function(fun = dnorm, args = list(mean = 3, sd = .9), xlim = c(1, 5)) +
  pmap(tibble(mean = c(-3, -.5, -0.15, 0,.25, 3), 
  sd = c(0.55, 0.9, 0.9, 0.9, 0.9, 0.5)) %>%
    rowwise %>%
    mutate(xlim = list(get_xlim(mean, 2))),
  function(mean, sd, xlim) stat_function(fun = dnorm, args = list(mean = mean, sd = sd), xlim = unlist(xlim))
  )  +
  xlim(c(-5, 5)) +
  #ylim(c(0,1)) +
  scale_y_continuous(expand = expansion(mult = c(0, .1)), limits = c(.05, 1)) +
  theme_classic() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())

high_div
```

```{r}
library(cowplot)
library(grid)
library(gridExtra)

plot_join <- plot_grid(low_rich, high_rich, low_fric, high_fric, 
                       low_eve, high_eve, low_div, high_div, 
                       nrow = 4, align = 'vh',
                       hjust = -1, axis = "l", scale = 0.9, 
                       label_x = "Trait Axis", label_y = "Abundance") #+
  #theme(plot.margin = margin(0, 0, 0, 30))

plot_join

# add_x <- ggdraw(add_sub(plot_join, "Trait Axis", 
#                y=2, x=0.5, hjust = 0.5, vjust=4.5))
# add_y <- add_sub(add_x, "Abundance",  
#                y=2, x=0.03, hjust = 0, vjust=0.5, angle = 90)
# ggdraw(add_y)
# 
# ggdraw(plot_join) + draw_label("Trait Axis", size = 14, hjust = 0.5, vjust = 0)

y.grob <- textGrob("Abundance", rot=90)

x.grob <- textGrob("Trait Axis")

grid.arrange(arrangeGrob(plot_join, left = y.grob, bottom = x.grob))

ggsave(here::here("figures/conceptual_fig.jpeg"))

 grid.arrange(arrangeGrob(p,p,p,p,top=textGrob("Sample Title One"), ncol=1), 
 arrangeGrob(p,p,p,p,top=textGrob("Sample Title Two"), ncol=1),   
 ncol = 2)
```
patchwork
```{r}
library(patchwork)
low_rich + high_rich + low_fric + high_fric + 
                       low_eve + high_eve + low_div + high_div + plot_layout(ncol = 2) %>%
  add_global_label(Ylab="Abundance", Ygap=0.05)
```

### 2021-07-26 analysis/conceptual_fig.Rmd
Create new column for y axis labels, save figure

@@
         axis.ticks.x=element_blank(),
-        axis.title.y=element_text(margin = margin(t = 0, r = 50, b = 0, l = 0)),
+        axis.title.y=element_text(angle = 0, vjust = 0.5),
         axis.text.y=element_blank(),
@@
 
-low_rich
+low_rich_fin <- low_rich + ylab(NULL)
 
@@
 
-low_fric
+low_fric_fin <- low_fric + ylab(NULL)
 
@@
 
-low_eve
+low_eve_fin <- low_eve + ylab(NULL)
 
@@
 
-low_div
+low_div_fin <- low_div + ylab(NULL)
 
--.5, 0,.25
 high_div <- 
@@
 
-plot_join <- plot_grid(low_rich, high_rich, low_fric, high_fric, 
-                       low_eve, high_eve, low_div, high_div, 
+plot_join <- plot_grid(get_plot_component(low_rich, "ylab-l"), low_rich_fin, high_rich, 
+                       get_plot_component(low_fric, "ylab-l"), low_fric_fin, high_fric, 
+                       get_plot_component(low_eve, "ylab-l"), low_eve_fin, high_eve, 
+                       get_plot_component(low_div, "ylab-l"), low_div_fin, high_div, 
                        nrow = 4, align = 'vh',
                        hjust = -1, axis = "l", scale = 0.9, 
-                       label_x = "Trait Axis", label_y = "Abundance") #+
-  #theme(plot.margin = margin(0, 0, 0, 30))
+                       label_x = "Trait Axis", label_y = "Abundance")
 
@@
 
-# add_x <- ggdraw(add_sub(plot_join, "Trait Axis", 
-#                y=2, x=0.5, hjust = 0.5, vjust=4.5))
-# add_y <- add_sub(add_x, "Abundance",  
-#                y=2, x=0.03, hjust = 0, vjust=0.5, angle = 90)
-# ggdraw(add_y)
-# 
-# ggdraw(plot_join) + draw_label("Trait Axis", size = 14, hjust = 0.5, vjust = 0)
+y.grob <- textGrob("Abundance",  
+            gp = gpar(fontsize = 14),
+            vjust = 15, rot=90)
 
-y.grob <- textGrob("Abundance", rot=90)
+x.grob <- textGrob("Trait Axis", hjust = -1.25)
 
-x.grob <- textGrob("Trait Axis")
+final <- grid.arrange(arrangeGrob(plot_join, left = y.grob, bottom = x.grob))
 
-grid.arrange(arrangeGrob(plot_join, left = y.grob, bottom = x.grob))
+ggsave(here::here("figures/conceptual_fig.jpeg"), final, height = 6, width = 7)
 
-ggsave(here::here("figures/conceptual_fig.jpeg"))
-
- grid.arrange(arrangeGrob(p,p,p,p,top=textGrob("Sample Title One"), ncol=1), 
- arrangeGrob(p,p,p,p,top=textGrob("Sample Title Two"), ncol=1),   
- ncol = 2)
 ```
@@
 ```{r}
-library(patchwork)
-low_rich + high_rich + low_fric + high_fric + 
-                       low_eve + high_eve + low_div + high_div + plot_layout(ncol = 2) %>%
-  add_global_label(Ylab="Abundance", Ygap=0.05)
+# library(patchwork)
+# low_rich + high_rich + low_fric + high_fric + 
+#                        low_eve + high_eve + low_div + high_div + plot_layout(ncol = 2) %>%
+#   add_global_label(Ylab="Abundance", Ygap=0.05)
 ```

### 2021-07-26 paper/manuscript.Rmd
Rewrite intro, add conceptual figure

@@
     affiliation: ucb
+  - name: "Perry de Valpine"
+    affiliation: ucb
   - name: "Carl Boettiger"
     affiliation: ucb, ropensci
-  - name: "Perry de Valpine"
-    affiliation: ucb
 address:
@@
 
-knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE)
+knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE, out.width = "\\textwidth")
 
@@
 By describing the functional trait space rather than species, functional diversity metrics capture the disproportionate impact of losses or gains of functionally unique species.
-Functional diversity metrics are therefore particularly well-suited for assessing the implications of community shifts in the full range of species change scenarios.
+Functional diversity metrics are therefore uniquely suited for assessing the implications of community shifts in the full range of species change scenarios.
 
@@
 
-While functional loss is frequently cited as one of the most pressing concerns of the anthropocene (CITATIONS), observed and projected functional diversity trends are less certain.
+While functional loss is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], it is not necessarily inevitable even in scenarios of species loss [@daz2001].
 Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020].
-And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@matuoka2020].
-Overall trends are therefore necessary...
-Clearly functional loss is not inevitable in scenarios of perturbation or even species loss [@daz2001].
-
-<!-- trends in functional composition also have important implications for how we assesses potential changes in ecosystem stability, function, and resilience.
-Functional loss is often implicit in work linking biodiversity loss to impacts on ecosystem processes despite a wide range of documented relationships between functional and species change in natural communities [@biswas2011].
-Where functional diversity is explicitly considered, there is increasing evidence it is a better predictor of ecosystem function than species-based metrics [@cadotte2011; @gagic2015], and that different facets of functional diversity play essential roles in maintaining ecosystem stability [@craven2018; @morin2014].
-Tracking the relationship between species and functional changes across a wide variety of communities therefore provides an opportunity to assess not only the prevalence of concurrent species and functional loss, but also the evidence underpinning hypothesis between biodiversity loss and ecosystem processes.
--->
-
-<!-- paragraph on maintenence of functional structure through time -->
+And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @matuoka2020].
+It is therefore critical to establish whether or not functional loss is the general rule for communities, and at what rate functional loss is occurring.
 
@@
 ```{r taxa-map, fig.cap="Map of timeseries locations with points color corresponding to taxa.", echo=FALSE}
-knitr::include_graphics(here::here("figures/study_map.jpeg"))
+knitr::include_graphics(here::here("figures/study_map_hist.jpeg"))
 ```
@@
 <!-- use these metrics because they're common, not because of their results -->
+
+```{r metric-fi, fig.cap="Example distributions of species and abundance for a single trait axis for high and low values of our measured metrics.", echo=FALSE, strip.white=TRUE}
+knitr::include_graphics(here::here("figures/conceptual_fig.jpeg"))
+```
 
@@
 We report a qualitative interpretation of the groups from the cluster analysis.
+
+*Study-level Trends*
+
+We assessed the impact of study-level characteristics on slope estimates...
 
@@
 As with previous species-based syntheses, we also found no overall trend in species diversity accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics may be consistent with similar well-documented species derived trends.
-We find no evidence of systemic functional richness loss or functional change.
+We found no evidence of systemic functional richness loss or functional change.
 A lack of trend for data broken down by realm, biome, and taxonomic group gives further evidence that directional functional change is absent from all systems observed in our dataset.

### 2021-07-26 analysis/05_model_metrics.Rmd
clean up

@@
 taxa_slope <- study_slope_data %>%
+  mutate(taxa = as.factor(taxa)) %>%
   #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
   group_nest(metric) %>%
-  mutate(#year_center = map(data, ~scale(.x$year)),
-         #model = map(data, ~lmer(estimate ~ taxa + (1|taxa), 
-         model = map(data, ~glm(estimate ~ taxa, 
+  mutate(model = map(data, ~glm(estimate ~ taxa, 
                                  data = .x)),

### 2021-07-29 analysis/05_model_metrics.Rmd
Mapping approach instead of tidymodels

@@
 ```{r}
-hfp_data <- pins::pin_get("hfp-data", board = "github")
-
-model_data_hfp <- model_data %>% 
-  #dplyr::select(rarefyID, year) %>% 
-  #distinct() %>%
-  mutate(hfp_year = case_when(
-    year < 2001 ~ 1993,
-    TRUE ~ 2009
-  ), 
-  hfp_rarefyID = str_replace_all(rarefyID, pattern = "_bird", ""), 
-  hfp_rarefyID = str_replace_all(hfp_rarefyID, pattern = "_mamm", "")) %>%
-  left_join(hfp_data %>% dplyr::select(-STUDY_ID) %>% mutate(year = as.numeric(year)), by = c("hfp_year" = "year", "hfp_rarefyID" = "rarefyID")) %>%
-  dplyr::select(-hfp_year)
+# hfp_data <- pins::pin_get("hfp-data", board = "github")
+# 
+# model_data_hfp <- model_data %>% 
+#   #dplyr::select(rarefyID, year) %>% 
+#   #distinct() %>%
+#   mutate(hfp_year = case_when(
+#     year < 2001 ~ 1993,
+#     TRUE ~ 2009
+#   ), 
+#   hfp_rarefyID = str_replace_all(rarefyID, pattern = "_bird", ""), 
+#   hfp_rarefyID = str_replace_all(hfp_rarefyID, pattern = "_mamm", "")) %>%
+#   left_join(hfp_data %>% dplyr::select(-STUDY_ID) %>% mutate(year = as.numeric(year)), by = c("hfp_year" = "year", "hfp_rarefyID" = "rarefyID")) %>%
+#   dplyr::select(-hfp_year)
 
@@
 
+Function approach to model
 ```{r}
-SES_rand_slope <- model_data %>%
-  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
-  group_nest(metric) %>%
-  mutate(#year_center = map(data, ~scale(.x$year)),
-         model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
-                                 data = .x)),
-           # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
-           #                       data = .x)),
-           # test = map2(model, null_model, anova),
-           # test_res = map(test, tidy),
-           coef = map(model, tidy),
-           ) %>% 
-  select(metric, model, coef) %>%
-  unnest(cols = c(coef)) %>%
-  filter(term == "year_scaled") %>%
-  select(-c(effect, group, term, statistic))
+get_model <- function(metric_name, formula, model_id, data, ...){
+  print(c(metric_name, model_id))
+  qlmer <- quietly(lmer)
 
-#check that they all used the same number of observations
-map(SES_rand_slope$model, nobs)
+  fit <- data %>%
+    filter(metric == metric_name) %>%
+    mutate(year_scaled = scale(year)) %>%
+    qlmer(formula = formula, data = ., ...)
 
-#these two metrics are cranky and want their own year scaling to converge
+  tidy(fit$result) %>%
+   mutate(metric = metric_name, model = list(fit), model_id = model_id, warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
+}
 
-#Jaccard Base
-jacc_base_mod <-  model_data %>%
-  filter(metric == ("Jaccard_base")) %>%
-  mutate(year_scaled = scale(year)) %>%
-  group_nest(metric) %>%
-  mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
-                                 data = .x)),
-           coef = map(model, tidy),
-           ) %>% 
-  select(metric, model, coef) %>%
-  unnest(cols = c(coef)) %>%
-  filter(term == "year_scaled") %>%
-  select(-c(effect, group, term, statistic))
+get_model(data = model_data, metric_name = "SES_FRic", "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)", model_id = "base")
+```
 
-# S model
-s_mod <- model_data %>%
-  filter(metric %in%  c("S")) %>%
-  #it wants its own scaling to converge for some reason
-  mutate(year_scaled = scale(year)) %>%
-  group_nest(metric) %>%
-  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
-                                 data = .x, REML = TRUE)),
-           coef = map(model, tidy)
-           ) %>% 
-  select(metric, model, coef) %>%
-  unnest(cols = c(coef)) %>%
-  filter(term == "year_scaled") %>%
-  select(-c(effect, group, term, statistic))
+Create dataframe of models to fit
+```{r}
+raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base"), 
+                         base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
+                         taxa = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)",
+                         realm = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|realm)",
+                         climate = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)") %>%
+  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
 
-metric_model_table <- bind_rows(SES_rand_slope, s_mod, jacc_base_mod)
+log_value_mods <- tibble(metric_name = c("S"), 
+                         base = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
+                         taxa = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)",
+                         realm = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|realm)",
+                         climate = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)") %>%
+  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
 
-#save coefficients and models
+mods <- bind_rows(raw_value_mods, log_value_mods) %>% 
+  select(metric_name, formula, model_id)
+```
+
+Fit models
+```{r}
+#Fit models with base settings
+mod_results <- pmap_dfr(mods, get_model, data = model_data)
+
+#get the models that had convergence warnings
+bad_models <- mod_results %>%
+  select(metric, model_id, warning) %>% 
+  filter(!is.na(warning)) %>% 
+  distinct() %>%
+  left_join(mods, by = c("metric" = "metric_name", "model_id"))
+
+#Turn of REML for these models 
+bad_mod_results <- pmap_dfr(bad_models %>% select(-warning), get_model, data = model_data, REML = FALSE)
+
+#One model still won't converge
+bad_models2 <- bad_mod_results %>%
+  select(metric, model_id, warning) %>% 
+  filter(!is.na(warning)) %>% 
+  distinct() %>%
+  left_join(mods, by = c("metric" = "metric_name", "model_id"))
+
+#Fit with a different algorithm
+bad_mod_results2 <- pmap_dfr(bad_models2 %>% select(-warning), get_model, data = model_data, control = lmerControl(optimizer = "bobyqa"))
+
+metric_model_table <- bind_rows(mod_results %>% filter(is.na(warning)), 
+          bad_mod_results %>% filter(is.na(warning)),
+          bad_mod_results2 %>% filter(is.na(warning)))
+
 usethis::use_data(metric_model_table)
 pins::pin(metric_model_table, board = "github")
+```
+
+```{r}
+# SES_rand_slope <- model_data %>%
+#   filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
+#   group_nest(metric) %>%
+#   mutate(#year_center = map(data, ~scale(.x$year)),
+#          model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
+#                                  data = .x)),
+#            # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
+#            #                       data = .x)),
+#            # test = map2(model, null_model, anova),
+#            # test_res = map(test, tidy),
+#            coef = map(model, tidy),
+#            ) %>% 
+#   select(metric, model, coef) %>%
+#   unnest(cols = c(coef)) %>%
+#   filter(term == "year_scaled") %>%
+#   select(-c(effect, group, term, statistic))
+# 
+# #check that they all used the same number of observations
+# map(SES_rand_slope$model, nobs)
+# 
+# #these two metrics are cranky and want their own year scaling to converge
+# 
+# #Jaccard Base
+# jacc_base_mod <-  model_data %>%
+#   filter(metric == ("Jaccard_base")) %>%
+#   mutate(year_scaled = scale(year)) %>%
+#   group_nest(metric) %>%
+#   mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
+#                                  data = .x)),
+#            coef = map(model, tidy),
+#            ) %>% 
+#   select(metric, model, coef) %>%
+#   unnest(cols = c(coef)) %>%
+#   filter(term == "year_scaled") %>%
+#   select(-c(effect, group, term, statistic))
+# 
+# # S model
+# s_mod <- model_data %>%
+#   filter(metric %in%  c("S")) %>%
+#   #it wants its own scaling to converge for some reason
+#   mutate(year_scaled = scale(year)) %>%
+#   group_nest(metric) %>%
+#   mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
+#                                  data = .x, REML = TRUE)),
+#            coef = map(model, tidy)
+#            ) %>% 
+#   select(metric, model, coef) %>%
+#   unnest(cols = c(coef)) %>%
+#   filter(term == "year_scaled") %>%
+#   select(-c(effect, group, term, statistic))
+# 
+# metric_model_table <- bind_rows(SES_rand_slope, s_mod, jacc_base_mod)
+# 
+# #save coefficients and models
+# usethis::use_data(metric_model_table)
+# pins::pin(metric_model_table, board = "github")
+# ```
+# Taxa model
+# ```{r}
+# taxa_rand_slope <- model_data %>%
+#   filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
+#   group_nest(metric) %>%
+#   mutate(#year_center = map(data, ~scale(.x$year)),
+#          model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa), 
+#                                  data = .x)),
+#            # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
+#            #                       data = .x)),
+#            # test = map2(model, null_model, anova),
+#            # test_res = map(test, tidy),
+#            coef = map(model, tidy),
+#            ) %>% 
+#   select(metric, model, coef) %>%
+#   unnest(cols = c(coef))
+# 
+# #Jaccard Base
+# jacc_taxa_base_mod <-  model_data %>%
+#   filter(metric == ("Jaccard_base")) %>%
+#   mutate(year_scaled = scale(year)) %>%
+#   group_nest(metric) %>%
+#   mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa), 
+#                                  data = .x)),
+#            coef = map(model, tidy),
+#            ) %>% 
+#   select(metric, model, coef) %>%
+#   unnest(cols = c(coef)) %>%
+#   filter(term == "year_scaled") %>%
+#   select(-c(effect, group, term, statistic))
+# 
+#  s_taxa_mod <- model_data %>%
+#   filter(metric %in%  c("S")) %>%
+#   #it wants its own scaling to converge for some reason
+#   mutate(year_scaled = scale(year)) %>%
+#   group_nest(metric) %>%
+#   mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa), 
+#                                  data = .x, REML = TRUE)),
+#            coef = map(model, tidy)
+#            ) %>% 
+#   select(metric, model, coef) %>%
+#   unnest(cols = c(coef)) %>%
+#   filter(term == "year_scaled") %>%
+#   select(-c(effect, group, term, statistic))
+# 
+# fric_taxa_mod <- model_data %>%
+#   filter(metric %in%  c("SES_FRic")) %>%
+#   #it wants its own scaling to converge for some reason
+#   mutate(year_scaled = scale(year)) %>%
+#   group_nest(metric) %>%
+#   mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa), 
+#                                  data = .x, REML = TRUE)),
+#            coef = map(model, tidy)
+#            ) %>% 
+#   select(metric, model, coef) %>%
+#   unnest(cols = c(coef)) %>%
+#   filter(term == "year_scaled") %>%
+#   select(-c(effect, group, term, statistic))
+# 
+# taxa_model_table <- bind_rows(taxa_rand_slope, jacc_taxa_base_mod, s_taxa_mod, fric_taxa_mod)
 ```
@@
 ```{r}
+library(multcomp)
+
 study_slope_data <- study_slopes %>%
@@
 
+taxa_slope <- study_slope_data %>%
+  mutate(taxa = as.factor(taxa)) %>%
+  group_nest(metric) %>%
+  mutate(#year_center = map(data, ~scale(.x$year)),
+         model = map(data, ~lmer(estimate ~ (1|study_id) + year_scaled + (year_scaled|study_id/rarefyID), 
+                                 data = .x)),
+           # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
+           #                       data = .x)),
+           # test = map2(model, null_model, anova),
+           # test_res = map(test, tidy),
+           coef = map(model, tidy),
+           ) %>% 
+  select(metric, model, coef) %>%
+  unnest(cols = c(coef)) %>%
+  filter(term == "year_scaled") %>%
+  select(-c(effect, group, term, statistic))
+
 #taxa model
@@
                                  data = .x)),
-           coef = map(model, tidy),
-           ) %>% 
-  select(metric, coef) %>%
-  unnest(cols = c(coef)) %>%
-  mutate(covariate = "taxa")
+           tukey_mod = map(model, ~glht(., mcp(taxa="Tukey"))),
+         tukey_coef = map(tukey_mod, tidy)) %>% 
+  select(metric, tukey_coef) %>%
+  unnest(cols = c(tukey_coef)) %>%
+  rename(covariate = term)
 
@@
                                  data = .x)),
-           coef = map(model, tidy),
-           ) %>% 
-  select(metric, coef) %>%
-  unnest(cols = c(coef)) %>%
-  mutate(covariate = "climate")
+           tukey_mod = map(model, ~glht(., mcp(climate="Tukey"))),
+         tukey_coef = map(tukey_mod, tidy)) %>% 
+  select(metric, tukey_coef) %>%
+  unnest(cols = c(tukey_coef)) %>%
+  rename(covariate = term)
 
@@
                                  data = .x)),
-           coef = map(model, tidy),
-           ) %>% 
-  select(metric, coef) %>%
-  unnest(cols = c(coef)) %>%
-  mutate(covariate = "realm")
+           tukey_mod = map(model, ~glht(., mcp(realm="Tukey"))),
+         tukey_coef = map(tukey_mod, tidy)) %>% 
+  select(metric, tukey_coef) %>%
+  unnest(cols = c(tukey_coef)) %>%
+  rename(covariate = term)
 

### 2021-08-03 analysis/05_model_metrics.Rmd
Modify function for extracting slopes for random effects levels to generalize for all random effects, not just study.

@@
   tidy(fit$result) %>%
-   mutate(metric = metric_name, model = list(fit), model_id = model_id, warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
+   mutate(metric = metric_name, model = list(fit$result), model_id = model_id, warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
 }
@@
 ```{r}
-get_study_slopes <- function(metric, model){
+get_study_slopes <- function(metric, model, model_id){
+
+  filter_name <- ifelse(model_id == "base", "study_id", model_id)
+
   year_scaled_coef <- tidy(model) %>%
@@
   study_ests <- broom.mixed::tidy(model, effects="ran_vals") %>%
-    filter(group == "study_id", term == "year_scaled") %>% 
+    filter(group == filter_name, term == "year_scaled") %>% 
     select(-c(effect, group, term)) %>%
-    rename(study_id = level, cond.std.error = std.error, cond.estimate = estimate) %>%
+    #rename({{ model_id }} := level, cond.std.error = std.error, cond.estimate = estimate) %>%
+    rename(cond.std.error = std.error, cond.estimate = estimate) %>%
     mutate(estimate = cond.estimate + year_scaled_est,
@@
            ),
-           metric = metric)
+      metric = metric,
+      model_id = model_id)
 }
 
-study_slopes <- pmap_dfr(select(metric_model_table,  metric, model), get_study_slopes)
+model_df <- metric_model_table %>%
+  select(metric, model, model_id) %>%
+  group_by(metric, model_id) %>% 
+  slice_head()
 
-slopes_wide <- study_slopes %>%
-  select(study_id, metric, estimate) %>%
+study_slopes <- pmap_dfr(model_df, get_study_slopes)
+
+study_slopes_wide <- study_slopes %>%
+  filter(model_id == "base") %>%
+  select(study_id = level, metric, estimate) %>%
   pivot_wider(names_from = "metric", values_from = "estimate")

### 2021-08-04 analysis/05_model_metrics.Rmd
Fix issue with assigning significance

@@
            sig = case_when(
-             lwr.ci < 0 & upr.ci > 0 ~ TRUE,
-             TRUE ~ FALSE
+             lwr.ci < 0 & upr.ci > 0 ~ FALSE,
+             TRUE ~ TRUE
            ),
@@
 
-study_slopes <- pmap_dfr(model_df, get_study_slopes)
+re_slopes <- pmap_dfr(model_df, get_study_slopes)
+
+study_slopes <- re_slopes %>% 
+  filter(model_id == "base") %>%
+  rename(study_id = level)
 
 study_slopes_wide <- study_slopes %>%
-  filter(model_id == "base") %>%
-  select(study_id = level, metric, estimate) %>%
+  select(study_id, metric, estimate) %>%
   pivot_wider(names_from = "metric", values_from = "estimate")
@@
 
-#And how many were significant for all but Jaccard?
+#And which studies were they?
 study_slopes %>% 
-  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "S")) %>% 
-  group_by(study_id) %>% filter(sum(sig) == 4) %>% pull(study_id) %>% 
-  n_distinct()
+  filter(sig == TRUE) %>% 
+  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
+  select(study_id, metric) %>%
+  distinct()
 
-#Breakdown of relationship between functional richness and species richness
-s_sig <- study_slopes %>% filter(metric == "S", sig == TRUE) %>% pull(study_id) %>% unique()
-s_sig_not_fric <- study_slopes %>% 
-  filter(study_id %in% s_sig, metric == "FRic", sig == FALSE) %>% 
-  pull(study_id) %>%
-  unique()
+#Which were significant for multiple metrics?
+study_slopes %>% 
+  filter(sig == TRUE) %>% 
+  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
+  group_by(study_id) %>%
+  filter(n() > 1) %>%
+  select(study_id, metric)
+```
 
-both_sig <- study_slopes %>% 
-  filter(study_id %in% s_sig, metric == "FRic", sig == TRUE) %>%
-  pull(study_id)
-
-both_pos
-fric_pos
-s_pos_fric_neg <- study_slopes %>%
-  filter(study_id %in% both_sig) %>% 
-  group_by(study_id) %>%
-  filter(sum(sign(estimate)) == 0)
+```{r}
+# #And how many were significant for all but Jaccard?
+# study_slopes %>% 
+#   filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "S")) %>% 
+#   group_by(study_id) %>% filter(sum(sig) == 4) %>% pull(study_id) %>% 
+#   n_distinct()
+# 
+# #Breakdown of relationship between functional richness and species richness
+# s_sig <- study_slopes %>% filter(metric == "S", sig == TRUE) %>% pull(study_id) %>% unique()
+# s_sig_not_fric <- study_slopes %>% 
+#   filter(study_id %in% s_sig, metric == "FRic", sig == FALSE) %>% 
+#   pull(study_id) %>%
+#   unique()
+# 
+# both_sig <- study_slopes %>% 
+#   filter(study_id %in% s_sig, metric == "FRic", sig == TRUE) %>%
+#   pull(study_id)
+# 
+# both_pos
+# fric_pos
+# s_pos_fric_neg <- study_slopes %>%
+#   filter(study_id %in% both_sig) %>% 
+#   group_by(study_id) %>%
+#   filter(sum(sign(estimate)) == 0)
 

### 2021-08-20 analysis/05_model_metrics.Rmd
Clean up

@@
 
-get_model(data = model_data, metric_name = "SES_FRic", "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)", model_id = "base")
+#get_model(data = model_data, metric_name = "SES_FRic", "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)", model_id = "base")
 ```
@@
 
+#get study level slopes for base, and levels for categorical random effect in non-base models
 re_slopes <- pmap_dfr(model_df, get_study_slopes)
 
+#get only study-level slopes for the base models
 study_slopes <- re_slopes %>% 
@@
 study_slope_data <- study_slopes %>%
-  select(study_id, metric, estimate) %>%
+  dplyr::select(study_id, metric, estimate) %>%
   left_join(meta %>% 
-              select(-rarefyid) %>% 
+              dplyr::select(-rarefyid) %>% 
               distinct() %>%
               mutate(study_id = as.character(study_id))) %>%
-  left_join(model_data %>% select(study_id, startyear, duration) %>% distinct())
-
-taxa_slope <- study_slope_data %>%
-  mutate(taxa = as.factor(taxa)) %>%
-  group_nest(metric) %>%
-  mutate(#year_center = map(data, ~scale(.x$year)),
-         model = map(data, ~lmer(estimate ~ (1|study_id) + year_scaled + (year_scaled|study_id/rarefyID), 
-                                 data = .x)),
-           # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
-           #                       data = .x)),
-           # test = map2(model, null_model, anova),
-           # test_res = map(test, tidy),
-           coef = map(model, tidy),
-           ) %>% 
-  select(metric, model, coef) %>%
-  unnest(cols = c(coef)) %>%
-  filter(term == "year_scaled") %>%
-  select(-c(effect, group, term, statistic))
-
-#taxa model
-taxa_slope <- study_slope_data %>%
-  mutate(taxa = as.factor(taxa)) %>%
-  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
-  group_nest(metric) %>%
-  mutate(model = map(data, ~glm(estimate ~ taxa, 
-                                 data = .x)),
-           tukey_mod = map(model, ~glht(., mcp(taxa="Tukey"))),
-         tukey_coef = map(tukey_mod, tidy)) %>% 
-  select(metric, tukey_coef) %>%
-  unnest(cols = c(tukey_coef)) %>%
-  rename(covariate = term)
-
-climate_slope <- study_slope_data %>%
-  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
-  group_nest(metric) %>%
-  mutate(#year_center = map(data, ~scale(.x$year)),
-         #model = map(data, ~lmer(estimate ~ taxa + (1|taxa), 
-         model = map(data, ~glm(estimate ~ climate, 
-                                 data = .x)),
-           tukey_mod = map(model, ~glht(., mcp(climate="Tukey"))),
-         tukey_coef = map(tukey_mod, tidy)) %>% 
-  select(metric, tukey_coef) %>%
-  unnest(cols = c(tukey_coef)) %>%
-  rename(covariate = term)
-
-realm_slope <- study_slope_data %>%
-  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
-  group_nest(metric) %>%
-  mutate(#year_center = map(data, ~scale(.x$year)),
-         #model = map(data, ~lmer(estimate ~ taxa + (1|taxa), 
-         model = map(data, ~glm(estimate ~ realm, 
-                                 data = .x)),
-           tukey_mod = map(model, ~glht(., mcp(realm="Tukey"))),
-         tukey_coef = map(tukey_mod, tidy)) %>% 
-  select(metric, tukey_coef) %>%
-  unnest(cols = c(tukey_coef)) %>%
-  rename(covariate = term)
+  left_join(model_data %>% dplyr::select(study_id, startyear, duration) %>% distinct())
 
 duration_slope <- study_slope_data %>%
-  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
   group_nest(metric) %>%
-  mutate(#year_center = map(data, ~scale(.x$year)),
-         #model = map(data, ~lmer(estimate ~ taxa + (1|taxa), 
-         model = map(data, ~glm(estimate ~ duration, 
+  mutate(model = map(data, ~glm(estimate ~ duration, 
                                  data = .x)),
@@
            ) %>% 
-  select(metric, coef) %>%
+  dplyr::select(metric, coef) %>%
   unnest(cols = c(coef)) %>%
@@
 startyear_slope <- study_slope_data %>%
-  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
   group_nest(metric) %>%
-  mutate(#year_center = map(data, ~scale(.x$year)),
-         #model = map(data, ~lmer(estimate ~ taxa + (1|taxa), 
-         model = map(data, ~glm(estimate ~ startyear, 
+  mutate(model = map(data, ~glm(estimate ~ startyear, 
                                  data = .x)),
@@
 #All model estimates in one dataframe
-slope_models <- bind_rows(taxa_slope, realm_slope, climate_slope, duration_slope, startyear_slope)
+slope_models <- bind_rows(duration_slope, startyear_slope)
 

### 2021-08-20 analysis/05_model_metrics.Rmd
Save out useful bits

@@
 re_slopes <- pmap_dfr(model_df, get_study_slopes)
+pins::pin(re_slopes, board = "github")
 
@@
 
-study_slopes_wide <- study_slopes %>%
-  select(study_id, metric, estimate) %>%
-  pivot_wider(names_from = "metric", values_from = "estimate")
+# study_slopes_wide <- study_slopes %>%
+#   select(study_id, metric, estimate) %>%
+#   pivot_wider(names_from = "metric", values_from = "estimate")
 ```
@@
 slope_models <- bind_rows(duration_slope, startyear_slope)
+pins::pin(slope_models, board = "github")
 

### 2021-08-23 analysis/visualization.Rmd
New color palette, fix overplotting for amphibians.

--- 2021-06-30 analysis/visualization.Rmd
@@
 library(stringr)
+library(cowplot)
 ```
@@
 pins::board_register_github(repo = "karinorman/biodivTS_data", branch = "master")
-country_data <- pins::pin_get("country-data-bind", board = "github") %>%
-  mutate(country = str_remove(country, "Commonwealth of The "),
-         country = str_remove(country, "Kingdom of "),
-         country = str_remove(country, "Republic of "),
-         country = str_remove(country, "Kingdom of "),
-         country = str_remove(country, "Federative Republic of "),
-         country = str_remove(country, "Federative "),) %>%
-  mutate(country = case_when(
-    country == "United States" ~ "USA",
-    country == "Russian Federation" ~ "Russia",
-    country == "United Great Britain and Northern Ireland" ~ "UK",
-    country == "Czechia" ~ "Czech Republic",
-    TRUE ~ country
-  )) %>% 
-  count(country, code)
-
-mapdata <- map_data("world") %>%
-  left_join(country_data, by = c("region" = "country"))
-
-mapdata %>%
-  ggplot(aes(long, lat)) +
-  geom_polygon(mapping = aes(group = group, fill = n)) +
-  ggthemes::theme_map() +
-  #scale_fill_continuous(color = "#32A251")
-  scale_fill_gradient(low = "white", high = "#39737c")
+# country_data <- pins::pin_get("country-data-bind", board = "github") %>%
+#   mutate(country = str_remove(country, "Commonwealth of The "),
+#          country = str_remove(country, "Kingdom of "),
+#          country = str_remove(country, "Republic of "),
+#          country = str_remove(country, "Kingdom of "),
+#          country = str_remove(country, "Federative Republic of "),
+#          country = str_remove(country, "Federative "),) %>%
+#   mutate(country = case_when(
+#     country == "United States" ~ "USA",
+#     country == "Russian Federation" ~ "Russia",
+#     country == "United Great Britain and Northern Ireland" ~ "UK",
+#     country == "Czechia" ~ "Czech Republic",
+#     TRUE ~ country
+#   )) %>% 
+#   count(country, code)
+# 
+# mapdata <- map_data("world") %>%
+#   left_join(country_data, by = c("region" = "country"))
+# 
+# mapdata %>%
+#   ggplot(aes(long, lat)) +
+#   geom_polygon(mapping = aes(group = group, fill = n)) +
+#   ggthemes::theme_map() +
+#   #scale_fill_continuous(color = "#32A251")
+#   scale_fill_gradient(low = "white", high = "#39737c")
 
 #ggsave("study_map.jpeg")
+```
+
+Define color palette
+```{r}
+my_pal <- c("#d9be3b", "#77976e", "#086788", "#d9be3b", "#99adbf", "#cc8214", "#bfac88")
 ```
@@
   coord_map(xlim=c(-180,180), ylim=c(-57, 90), projection = "mercator") +
-  geom_point(data = study_data, aes(rarefyID_x, rarefyID_y, color = taxa), alpha = 0.3, inherit.aes = FALSE) +
+  geom_point(data = study_data,
+             aes(rarefyID_x, rarefyID_y, color = taxa), alpha = 0.4, inherit.aes = FALSE) +
+  #plot the US amphibian point on top so we can see it
+  geom_point(data = study_data %>% 
+               #filter(taxa %in% c("Amphibians")), 
+               filter(taxa %in% c("Amphibians"), rarefyID_y > 0),
+             aes(rarefyID_x, rarefyID_y ), color = "#d9be3b", alpha = 0.8, inherit.aes = FALSE) +
   ggthemes::theme_map() +
-  theme(legend.title  = element_blank())
+  scale_color_manual(values = my_pal) +
+  theme(legend.title  = element_blank()) +
+  guides(color = guide_legend(override.aes= list(alpha = 0.8)))
 
@@
   facet_wrap(~taxa, scales = "free") +
-  theme(strip.background = element_blank(), strip.text = element_blank(),
-        legend.position = "none")
+  xlab("Timeseries Duration") +
+  theme(strip.background = element_blank(), 
+        strip.text = element_blank(),
+        legend.position = "none", 
+        axis.title = element_text(size=8)) +
+  scale_fill_manual(values = my_pal)
 

### 2021-08-23 analysis/05_model_metrics.Rmd
Move timeseries plots from the analysis script, update color palette

@@
 library(lmerTest)
-library(cowplot)
 ```
@@
 ```
-
-```{r}
-library(cowplot)
-library(RColorBrewer)
-
-plot_metric <- function(metric_name, ylabel, yval, xlabel, data, model_coef, color_list){
-
-  if(is.na(xlabel)) xlabel <- NULL
-
-  data <- data %>%
-    filter(metric == metric_name, !is.na(value), !is.infinite(value))
-
-  model_coef <- model_coef %>%
-    filter(metric == metric_name) %>%
-    select(term, estimate)
-
-  intercept <- model_coef %>% filter(term == "(Intercept)") %>% pull(estimate)
-  slope <- model_coef %>% filter(term == "year_scaled") %>% pull(estimate)
-
-  pred <- data %>%
-    select(year_scaled) %>%
-    distinct() %>%
-    mutate(pred = intercept + year_scaled * slope)
-
-  trend_plot <- data %>%
-    left_join(pred, by = "year_scaled") %>%
-    #left_join(metadata %>% select(study_id, climate)) %>%
-    ggplot(aes(x = year, y = !!sym(yval))) +
-    # geom_smooth(aes(color = climate, group = rarefyID), method = "glm", se = FALSE) +
-    # geom_point(aes(color = climate, group = rarefyID)) +
-    geom_point(aes(group = rarefyID) ,color = "grey", size = 0.25) +
-    geom_smooth(aes(color = climate, group = study_id), method = "glm", se = FALSE, size = 0.7) +
-    #geom_smooth(color = "black", method = "glm", se = FALSE) +
-    geom_line(mapping = aes(y = pred)) +
-    theme_classic() +
-    theme(axis.title = element_text(size=7), legend.position = "none") +
-    ylab(ylabel) +
-    xlab(xlabel) +
-    labs(color = "Climate") +
-    scale_x_continuous(breaks = scales::pretty_breaks(n = 6)) +
-    scale_colour_manual(values = color_list, drop = FALSE)
-
-    #scale_color_manual(values = c("#32a251", 
-               # "#ff7f0f", "#39737c", "#b85a0d", "#ffd94a"))
-
-  ggsave(paste0(here::here("figures"), "/",metric_name, "_trend.png"), trend_plot)
-  return(trend_plot)
-}
-
-model_coef <- metric_model_table %>%
-  select(metric, model) %>%
-   mutate(coef = map(model, tidy)) %>% 
-  select(-model) %>% 
-  unnest(cols = c(coef))
-
-pal <- brewer.pal(n = 5, name = "Dark2")
-group_colors <- c("Global" = pal[1], "Polar/Temperate" = pal[2], "Temperate/Tropical" = pal[3], "Tropical" = pal[4], "Temperate" = pal[5], "Overall Mean" = "#000000")
-
-plot_map <- tibble(metric_name = c("S", "Jaccard_base", "SES_FRic", "SES_FDiv", "SES_FEve"), 
-           ylabel = c("log(S)", "Jaccard", "Functional Richness SES", "Functional Divergence SES", "Functional Evenness SES"),
-           #ylabel = c("log(S)", "Jaccard", "FRic SES", "FDiv SES", "FEve SES"),
-           yval = c("logvalue", "value", "value", "value", "value"),
-           xlabel = c(NA, NA, NA, "year", "year"))
-m_plots <- pmap(plot_map, plot_metric, data = model_data, model_coef = model_coef, color_list = group_colors)
-
-plot_join <- plot_grid(plotlist=m_plots, labels = "AUTO", nrow = 3, align = 'vh',
-                       hjust = -1, axis = "l", scale = 0.9)
-
-leg_plot <- model_data %>%
-  bind_rows(data.frame(climate = c("Overall Mean", "Overall Mean"), year = c(2019, 2020), value = c(1, 1))) %>%
-  ggplot(aes(x = year, y = value)) +
-  geom_point(color = "grey", size = 0.50) +
-  geom_smooth(aes(color = climate, group = study_id), method = "glm", se = FALSE, size = 0.7) +
-  theme_classic() +
-  theme(axis.title = element_text(size=7)) +
-  scale_colour_manual(values = group_colors, drop = FALSE)
-
-# extract the legend from one of the plots
-legend <- get_legend(leg_plot)
-
-# add the legend to the row we made earlier. Give it one-third of 
-# the width of one plot (via rel_widths).
-#plot_join + draw_grob(legend,  2/3.3, -0.3, .3/3.3, 1, scale = 0.6)
-plot_join + draw_grob(legend,  2/3.3, -0.3,.3/3.3, 1, scale = 0.01)
-
-ggsave2(here::here("figures", "3met_long.jpeg"))
-```

### 2021-08-23 analysis/visualization.Rmd
Move timeseries plots from the analysis script, update color palette

@@
 library(cowplot)
+library(tidyverse)
 ```
@@
 ```{r}
-my_pal <- c("#d9be3b", "#77976e", "#086788", "#d9be3b", "#99adbf", "#cc8214", "#bfac88")
+my_pal <- c("#d9be3b", "#77976e", "#086788", "#cc8214", "#bfac88", "#99adbf")
 ```
@@
 ```
+
+Metric plots with trend lines
+```{r}
+metric_model_table <- pins::pin_get("metric-model-table", board = "github")
+model_data <- pins::pin_get("model-data", board = "github")
+
+plot_metric <- function(metric_name, ylabel, yval, xlabel, data, model_coef, color_list){
+
+  if(is.na(xlabel)) xlabel <- NULL
+
+  data <- data %>%
+    filter(metric == metric_name, !is.na(value), !is.infinite(value))
+
+  model_coef <- model_coef %>%
+    filter(metric == metric_name) %>%
+    select(term, estimate)
+
+  intercept <- model_coef %>% filter(term == "(Intercept)") %>% pull(estimate)
+  slope <- model_coef %>% filter(term == "year_scaled") %>% pull(estimate)
+
+  pred <- data %>%
+    select(year_scaled) %>%
+    distinct() %>%
+    mutate(pred = intercept + year_scaled * slope)
+
+  trend_plot <- data %>%
+    left_join(pred, by = "year_scaled") %>%
+    #left_join(metadata %>% select(study_id, climate)) %>%
+    ggplot(aes(x = year, y = !!sym(yval))) +
+    # geom_smooth(aes(color = climate, group = rarefyID), method = "glm", se = FALSE) +
+    # geom_point(aes(color = climate, group = rarefyID)) +
+    geom_point(aes(group = rarefyID) ,color = "grey", size = 0.25) +
+    geom_smooth(aes(color = climate, group = study_id), method = "glm", se = FALSE, size = 0.7) +
+    #geom_smooth(color = "black", method = "glm", se = FALSE) +
+    geom_line(mapping = aes(y = pred)) +
+    theme_classic() +
+    theme(axis.title = element_text(size=7), legend.position = "none") +
+    ylab(ylabel) +
+    xlab(xlabel) +
+    labs(color = "Climate") +
+    scale_x_continuous(breaks = scales::pretty_breaks(n = 6)) +
+    scale_colour_manual(values = color_list, drop = FALSE)
+
+  ggsave(paste0(here::here("figures"), "/",metric_name, "_trend.png"), trend_plot)
+  return(trend_plot)
+}
+
+model_coef <- metric_model_table %>%
+  filter(model_id == "base") %>%
+  select(metric, model) %>%
+   mutate(coef = map(model, broom.mixed::tidy)) %>% 
+  select(-model) %>% 
+  unnest(cols = c(coef))
+
+group_colors <- c("Global" = my_pal[1], "Polar/Temperate" = my_pal[5], "Temperate/Tropical" = my_pal[3], "Tropical" = my_pal[4], "Temperate" = my_pal[2], "Overall Mean" = "#000000")
+
+plot_map <- tibble(metric_name = c("S", "Jaccard_base", "SES_FRic", "SES_FDiv", "SES_FEve"), 
+           ylabel = c("log(S)", "Jaccard", "Functional Richness SES", "Functional Divergence SES", "Functional Evenness SES"),
+           #ylabel = c("log(S)", "Jaccard", "FRic SES", "FDiv SES", "FEve SES"),
+           yval = c("logvalue", "value", "value", "value", "value"),
+           xlabel = c(NA, NA, NA, "year", "year"))
+m_plots <- pmap(plot_map, plot_metric, data = model_data, model_coef = model_coef, color_list = group_colors)
+
+plot_join <- plot_grid(plotlist=m_plots, labels = "AUTO", nrow = 3, align = 'vh',
+                       hjust = -1, axis = "l", scale = 0.9)
+
+leg_plot <- model_data %>%
+  bind_rows(data.frame(climate = c("Overall Mean", "Overall Mean"), year = c(2019, 2020), value = c(1, 1))) %>%
+  ggplot(aes(x = year, y = value)) +
+  geom_point(color = "grey", size = 0.50) +
+  geom_smooth(aes(color = climate, group = study_id), method = "glm", se = FALSE, size = 0.7) +
+  theme_classic() +
+  theme(axis.title = element_text(size=7)) +
+  scale_colour_manual(values = group_colors, drop = FALSE)
+
+# extract the legend from one of the plots
+legend <- get_legend(leg_plot)
+
+# add the legend to the row we made earlier. Give it one-third of 
+# the width of one plot (via rel_widths).
+#plot_join + draw_grob(legend,  2/3.3, -0.3, .3/3.3, 1, scale = 0.6)
+plot_join + draw_grob(legend,  2/3.3, -0.3,.3/3.3, 1, scale = 0.01)
+
+ggsave2(here::here("figures", "3met_long.jpeg"))
+```

### 2021-09-01 paper/manuscript.Rmd
Clean up tables and update analysis.

@@
                caption = "Summary of the data in the final trait database.") %>%
-   kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
+  kable_classic() #%>% 
+  #kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
    #kableExtra::column_spec(2, width = "5cm") #%>%
@@
 Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package.
-Here we report functional richness (*FRic*), functional evenness (*FEve*), and functional diveregence (*FDiv*) which together describe three complementary characteristics of the functional space [@mason2005].
+Here we report functional richness (*FRic*), functional evenness (*FEve*), and functional diveregence (*FDiv*) which together describe three complementary characteristics of the functional space [@mason2005, @hillebrand2009].
 *FRic* assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
@@
 
-```{r metric-fi, fig.cap="Example distributions of species and abundance for a single trait axis for high and low values of our measured metrics.", echo=FALSE, strip.white=TRUE}
+```{r metric-fi, fig.cap="Example distributions of species and abundance for a single trait axis for high and low values of our measured metrics.
+Figure is modified from a similar illustration in Hillebrand & Matthiessen 2009.", echo=FALSE, strip.white=TRUE}
 knitr::include_graphics(here::here("figures/conceptual_fig.jpeg"))
@@
 
-*Overall Trends*
+We estimated general trends for each diversity metric using a Linear Mixed Effects model with a random slope and intercept for each timeseries nested within the study.
+To test for differences in trends by taxa, biome, and realm, each variable was modeled individually as random effects in models of the same structure.
+We obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model with no additional covariates.
 
-We estimated overall trends for each diversity metric using a Linear Mixed Effects model with a random slope and intercept for each timeseries nested within the study.
-Differences in trends by taxa, biome, and realm were each modeled individually as covariates in models of the same structure.
-
-*Study-level Trends*
-
-We obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model and assessed them in two ways.
-First, we looked at pairwise relationships between all metrics to assess what kinds of community shifts are commonly found together.
-We obtained correlations and bootstrapped p-values using *corr.test* and *corCI* functions in the *psych* R package (citation).
-Second, we assessed the relationship between all metrics simultaneously by performing a cluster analysis which grouped each study based on their similarities across all metrics.
-We used a hierarchical clustering algorithm through the function *hclust* from the R package *stats.* We included all standardized functional diversity metrics, species richness, and Jaccard similarity in the analysis.
-We report a qualitative interpretation of the groups from the cluster analysis.
-
-*Study-level Trends*
-
-We assessed the impact of study-level characteristics on slope estimates...
+We assessed the impact of timeseries duration and start year on study-level trends using General Linear Models with duration and start year as predictors.
 
@@
 
-<!-- Despite significant trends at the study level for the majority of studies,-->
-
 We found no significant overall trend in species richness or functional diversity metrics (Fig 2).
@@
 
-At the study level, the majority of studies (86%) experienced a significant temporal change in species richness and all functional diversity metrics.
+At the study level, only 7 of 54 studies experienced a significant trend for a metric other than Jaccard similarity.
+4 studies had a significant trend in species diversity, one positive and three negative.
+Two studies experienced a significant decrease in functional evenness, and one study experienced a significant decrease in both functional richness and functional divergence.
+
+```{r echo = FALSE}
+library(tidyverse)
+
+data.frame(sign = c("+", "-"), S = c("1", "3"), `Jaccard Similarity` = c("0", "37"), FRic = c("0", "1"), FEve = c("0", "2"), FDiv = c("0", "1")) %>%
+  column_to_rownames("sign") %>%
+  kable(caption = "Number of studies that experienced a signficiant trend in each calculated metric.")
+```
+
+Trends for multiple metrics were significantly related to the the sampling characteristics of indivdual studies.
+
+<!-- the majority of studies (86%) experienced a significant temporal change in species richness and all functional diversity metrics.
 92% of studies experienced a significant trend for species richness, 88% for functional richness, 98% for functional divergence, 96% for functional evenness, and 30% for Jaccard similarity.
@@
 There was only one significant pairwise correlation between the metrics, a positive relationship between Functional Divergence and Functional Evenness, indicating that communities experiencing an increase in Divergence frequently experience an increase in Evenness.
+-->
 
-Cluster analysis grouped all but a few studies into a cluster characterized by non-significant or very small significant slopes for all metrics.
+<!-- Cluster analysis grouped all but a few studies into a cluster characterized by non-significant or very small significant slopes for all metrics.
 We identified four other clusters containing six studies (list studies) that were characterized by outlier trends in at least one of the metrics and near-zero trends in the rest.
@@
 The third contained one study with strongly decreasing functional divergence and richness, and the fourth strongly decreasing species richness with increasing functional richness.
+-->
 
@@
 
-Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at overall trends in functional diversity and the range of changes happening in mammal, bird, and amphibian communities.
+Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at aggregate and local trends in functional diversity in mammal, bird, and amphibian communities.
 Surprisingly, we did not detect an overall trend in any of the calculated functional diversity metrics.
-While nearly all of the studies included in our data experienced a significant increase or decrease in at least one of the metrics, trends were observed in equal numbers resulting in an overall trend that was not significantly different from zero.
 As with previous species-based syntheses, we also found no overall trend in species diversity accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics may be consistent with similar well-documented species derived trends.
@@
 
+Communities demonstrated almost universal maintenance of functional composition.
+While the majority of the studies (\~70%) included in our data experienced significant species turnover, only three experienced a significant shift in any functional dimension.
+
 Individual studies in our dataset demonstrate the variety of species and functional shifts that can happen concurrently in a community.
+
 With only one significant pairwise relationship between study-level functional divergence slopes and functional evenness slopes, all other possible combinations of species and functional change were observed in at least one study.

### 2021-09-02 paper/manuscript.Rmd
Taking another stab at the discussion.

@@
 ---
-title: "Functional timeseries reveal no systemic shifts in vertebrate communities"
+title: "Evidence of widspread maintenence of functional composition in vertebrate communities"
 author: 
@@
 Assessments of broad-scale temporal change in functional diversity have previously been limited by a lack of functional trait data.
-The majority of work has therefore focused largely on system-specific studies with traits collected *in situ*.
+<!-- and computational power? --> The majority of work has therefore focused largely on system-specific studies with traits collected in situ.
 Ongoing efforts to assemble functional traits for a variety of taxa have made synthesis of existing community assemblage data and functional traits possible for the first time, providing initial insights into the ways functional diversity changes on a broad scale for specific taxa [e.g.
@@
 Here we perform the first multi-taxa, multi-system assessment of functional diversity change through time.
-We assess thousands of mammal, bird, and amphibian functional diversity timeseries to 1) determine how functional shifts compares to species change and 2) assess the types of functional shifts that are most common across communities.
-<!-- While our data represent the largest collection of functional diversity timeseries to date, they are not a representative sample of systems experiencing biodiversity change.
-We do not therefore seek to classify the global trend in functional diversity change, but rather assess the utility of a functional lens in understanding biodiversity change.-->
+We assess thousands of mammal, bird, and amphibian functional diversity timeseries to 1) establish a general trend of functional change 2) ???.
 
@@
 We found no significant overall trend in species richness or functional diversity metrics (Fig 2).
-There was also no temporal trend for studies broken down by taxa, biome or realm.
+There was also no temporal trend in those metrics for studies broken down by taxa, biome or realm.
 We did find a significant overall decrease in Jaccard similarity, indicating increasing turnover through time.
@@
 
-<!-- Our study represents the largest broad-scale multi-taxa assessment of functional change through time, giving a first look at overall trends in functional diversity and the range of changes happening in mammal, bird, and amphibian communities.
-Consistent with previous species-based syntheses, we found no overall trend in species diversity accompanied by increasing turnover through time [@dornelas2018].
-While nearly all of the studies included in our data experienced a significant increase or decrease in species richness, trends were observed in equal numbers resulting in an overall trend not significantly different from zero.
--->
-
-<!-- Surprisingly, we did not detect an overall trend in any of the functional diversity metrics.
-Similar to species richness, communities are experiencing a wide range of functional shifts that balance each other out.
-While our dataset likely under-samples areas with most direct human impact, anthropogenic change is hypothesized to impact even the most remote systems.
-Conservatively, our findings therefore indicate that in unperturbed mammal, bird, and amphibian communities there is no evidence of systemic functional richness loss or functional change.
-A lack of trend for data broken down by realm, biome, and taxonomic group gives further evidence that directional functional change is absent from all systems observed in our dataset.
--->
-
 Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at aggregate and local trends in functional diversity in mammal, bird, and amphibian communities.
@@
 While the majority of the studies (\~70%) included in our data experienced significant species turnover, only three experienced a significant shift in any functional dimension.
+This suggests certain characteristics of the functional space are conserved even in the face of significant change in species identity, namely the size of the functional space occupied by the community (FRic) and the distribution of species and individuals within that space (FEve and FDiv).
+On average, species additions have similar functional characteristics as lost species and therefore maintain the functional space structure.
+<!-- language to distinguish between "functional traits" and a species' role in the context of the functional space -->
 
-Individual studies in our dataset demonstrate the variety of species and functional shifts that can happen concurrently in a community.
+-   Paragraph about analog communities?
 
-With only one significant pairwise relationship between study-level functional divergence slopes and functional evenness slopes, all other possible combinations of species and functional change were observed in at least one study.
-Notably, there is no relationship between species richness change and change in functional richness, meaning that communities experiencing species losses do not necessarily experience accompanying functional richness, and in some cases may even experience an increase.
-<!-- need to put a disclaimer about what "increasing" means here - it's increased richness relative to the expectation for a community of that size --> These communities gain species that expand the boundaries of the trait space while losing interior species.
+-   Paragraph about implications for BEF work
+
+Our work challenges assumptions that functional loss is the default state of all or even many communities.
+While we do not directly assess the disturbance histories of included communities, many are from sites that experienced significant disturbance or manual treatment (see appendix for metadata of included studies).
+<!-- check this, get numbers? --> This is good news for ecosystem function!
+But we may also be measuring the wrong dimensions of the trait space or the wrong traits.
+Can't really say BEF needs to measure function directly because these communities aren't losing species or functional diversity - turnover is either within regular bounds or if not (if more turnover than normal) it's still maintaining community structure.
+And presuming that the original structure was good for function, maintaining it is also good for function!
+
+**What are we going to say about the disturbance/environmental change backgrounds of these communities??**
+
+While we did not directly assess the impact of potential disturbance or environmental change in the included communities
+
+While there are clear impacts on functional makeup in scenarios of disturbance (what other scenarios do we see??)
+
+-   some of these studies expereience a disturbance/treatment!
+
+-   even for those that didn't it is remarkable that these metrics were conserved so universally!
 
     The overall trend of increasing turnover is consistent with work indicating higher than expected background rates of turnover in contemporary communities [@gotelli2017; @magurran2018].
-However, many studies did not experience increasing turnover (65%), suggesting that it is not universal.
-Even for those studies without a turnover trend, significant functional shifts occurred.
-While some communities may be changing more rapidly than usual, that turnover is not necessarily an indication of specific kind of functional change.
 
@@
 
+<!-- Here's the prediction of the field, here is the data of the field, given that the data doesn't match the prediction it's on the field to resolve these issues. -->
+
 *Implications for Future Work*
 
-Here's the prediction of the field, here is the data of the field, given that the data doesn't match the prediction it's on the field to resolve these issues.
+Despite widespread concern about the loss of functional diversity, our analysis does not indicate significant change in functional composition is the rule for communities.
+Why
 
-Why do we find no significant trend in the global data despite extensive predictions to the contrary from both theory and local studies?
-First, there are a variety of data limitations to consider:
-
-This study represents best practices in biodiversity synthesis and functional diversity analysis, however methodological limitations may still be obscuring potential systematic shifts in functional composition.
-First, the Biotime database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
-While both of these challenges are pervasive in Ecology generally and could have potentially insidious implications for synthesis work, we can only use the data we have.
-We do however emphasize that our study included many timeseries that experienced a significant species loss (\~ 52%), indicating that this is not simply a synthesis of pristine systems.
+This study represents the best available biodiversity and trait data paired with best practices in biodiversity synthesis.
+Why then do we find no change despite extensive predictions to the contrary from both **theory** <!-- need more theory in the intro? --> and and many local studies?
+First, the Biotime database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017] (other sampling citation!), and the subset of data in this study exhibits a strong Northern Hemisphere bias.
+We may simply not have data from those areas experiencing the greatest functional shifts.
+We do however emphasize that even disturbed sites can maintain structure (citation), and therefore the inclusion of these sites would not necessarily change the result.
 
@@
 <!-- Further exploration of complementarity of traits... -->
+
+Notably, the metrics we calculated do not capture shifts in the location of the functional space.
+For example two communities could have very similar metric values but no overlap in their trait spaces.
+This is especially critical in the context of biodiversity change, as a species could be lost and replaced by a species with very different functional attributes, but the replacement would go undetected if the new species expanded the trait space by the same degree and had similar population attributes.
+This may be particularly relevant for communties that exist in a rapidly changing envrionment (both abiotic and biotic), as communities may maintain some characteristcs of their functional composition while shifting to track the new conditions.
+Metrics developed to assess shifts and overlaps in trait spaces are now being developed (citation), and further exploration of time series using those approaches may disentangle maintained characteristics from occupied trait space.
 

### 2021-09-02 paper/manuscript.Rmd
Clean up discussion, finish methods.

@@
 ```{r metric-fi, fig.cap="Example distributions of species and abundance for a single trait axis for high and low values of our measured metrics.
-Figure is modified from a similar illustration in Hillebrand & Matthiessen 2009.", echo=FALSE, strip.white=TRUE}
+Figure is modified from a similar illustration in Hillebrand \\& Matthiessen 2009.", echo=FALSE, strip.white=TRUE}
 knitr::include_graphics(here::here("figures/conceptual_fig.jpeg"))
@@
 
-Trends for multiple metrics were significantly related to the the sampling characteristics of indivdual studies.
+Study-level slopes for multiple metrics were significantly related to the duration and start year of studies.
+Slopes for species richness were significantly more negative with later start date and more positive shorter duration studies.
+Jaccard similarity and functional evenness both had significantly more negative slopes with more recent start year, whereas functional divergence was significantly more positive.
+Slopes for functional evenness were also significantly more positive for longer duration studies.
 
@@
 
-**What are we going to say about the disturbance/environmental change backgrounds of these communities??**
-
-While we did not directly assess the impact of potential disturbance or environmental change in the included communities
-
-While there are clear impacts on functional makeup in scenarios of disturbance (what other scenarios do we see??)
-
--   some of these studies expereience a disturbance/treatment!
-
--   even for those that didn't it is remarkable that these metrics were conserved so universally!
-
-    The overall trend of increasing turnover is consistent with work indicating higher than expected background rates of turnover in contemporary communities [@gotelli2017; @magurran2018].
-
-[@osullivan2021]
-
-<!-- Link turnover paragraph to background turnover rate, maintenence of species diversity (i.e.
-turnover is happening!
-what does that mean for speices div?
-maintained.
-for functional diversity?
-maintained.) -->
+*Implications for Future Work*
 
@@
 
-*Implications for Future Work*
+<!-- Despite widespread concern about the loss of functional diversity, our analysis does not indicate significant change in functional composition is the rule for communities. Why -->
 
-Despite widespread concern about the loss of functional diversity, our analysis does not indicate significant change in functional composition is the rule for communities.
-Why
+This study represents the best available data and biodiversity synthesis approaches.
+Why then do we find no change in functional diversity despite extensive predictions to the contrary from both **theory** <!-- need more theory in the intro? --> and and many local studies?
+First, the Biotime database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017] (other sampling citation!), and the subset of data in this study exhibits a strong Northern Hemisphere bias.
+We may simply not have data from those areas experiencing the greatest functional shifts [hughes2021].
+We do however emphasize that even disturbed communities maintain functional structure [@edwards2013; @matuoka2020], and the inclusion of more heavily impacted sites would not necessarily change the result.
 
-This study represents the best available biodiversity and trait data paired with best practices in biodiversity synthesis.
-Why then do we find no change despite extensive predictions to the contrary from both **theory** <!-- need more theory in the intro? --> and and many local studies?
-First, the Biotime database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017] (other sampling citation!), and the subset of data in this study exhibits a strong Northern Hemisphere bias.
-We may simply not have data from those areas experiencing the greatest functional shifts.
-We do however emphasize that even disturbed sites can maintain structure (citation), and therefore the inclusion of these sites would not necessarily change the result.
+\<!-
+- You'd imagine the pattern would still be visible even with North American birds...
 
-You'd imagine the pattern would still be visible even with North American birds...
+To further explore the issue we look for example at the North American Breeding Birdy Survey, included in this study and assessed in many other publications.
+--\>
 
@@
 Further investment in individual level trait data for long-term monitoring will be critical for assessing finer scale shifts that likely occur more commonly in communities and potentially preceded bigger changes.
+Likewise, available trait data may not capture the traits experiencing the greatest change.
 
-Third, while we use here the most common metrics for describing functional diversity there are a number of possible metrics for assessing a functional space, with new ones frequently proposed.
-There may be a significant shift in a characteristic of the functional space that was poorly captured by our chosen metrics.
-<!-- Further exploration of complementarity of traits... -->
-
-Notably, the metrics we calculated do not capture shifts in the location of the functional space.
+Third, while we use here the most common metrics for describing functional diversity they do not measure some potentially important aspects of the functional space.
+Most notably, the metrics we calculated do not capture shifts in the location of the functional space.
 For example two communities could have very similar metric values but no overlap in their trait spaces.
-This is especially critical in the context of biodiversity change, as a species could be lost and replaced by a species with very different functional attributes, but the replacement would go undetected if the new species expanded the trait space by the same degree and had similar population attributes.
-This may be particularly relevant for communties that exist in a rapidly changing envrionment (both abiotic and biotic), as communities may maintain some characteristcs of their functional composition while shifting to track the new conditions.
-Metrics developed to assess shifts and overlaps in trait spaces are now being developed (citation), and further exploration of time series using those approaches may disentangle maintained characteristics from occupied trait space.
+This is especially relevant in the context of biodiversity change as lost species could be replaced by a species with very different functional attributes, but the replacement would go undetected if the new species expanded the trait space by the same degree and had similar abundance.This may be particularly relevant for communities tracking shifting environmental conditions, as communities may maintain some characteristics of their functional composition while shifting to track the new conditions.
+Metrics developed to assess shifts and overlaps in trait spaces are now being developed (citation), and further exploration of time series using those approaches may disentangle trait space characteristics from the location of the occupied space.
 

### 2021-09-07 paper/manuscript.Rmd
add citation
Minor changes

@@
 And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @matuoka2020].
-It is therefore critical to establish whether or not functional loss is the general rule for communities, and at what rate functional loss is occurring.
+On paleoecological time scales, functional composition is often conserved alongside species diversity even in the face of drastic environmental change [@jackson2015; @barnosky2005].
+It is therefore critical to establish whether or not functional loss is the general rule for communities on time scales relevant for current environmental impacts.
 
@@
 Our final dataset included 2,443 timeseries from 53 studies in 21 countries and 15 biomes and 13 different traits (Fig 1).
+The earliest sample was in 1923 and the most recent was in 2016.
 Only four studies (consisting of 11 timeseries) came from Amphibian studies due to the limited availability of amphibian timeseries and low species richness values for assemblages (Table 1).
@@
 
-<!-- the majority of studies (86%) experienced a significant temporal change in species richness and all functional diversity metrics.
-92% of studies experienced a significant trend for species richness, 88% for functional richness, 98% for functional divergence, 96% for functional evenness, and 30% for Jaccard similarity.
-Only 4 (cite studies) out of 53 studies experienced no significant trend in species richness and were also characterized by significantly decreasing functional evenness.
-There was only one significant pairwise correlation between the metrics, a positive relationship between Functional Divergence and Functional Evenness, indicating that communities experiencing an increase in Divergence frequently experience an increase in Evenness.
--->
-
 <!-- Cluster analysis grouped all but a few studies into a cluster characterized by non-significant or very small significant slopes for all metrics.
@@
 While the majority of the studies (\~70%) included in our data experienced significant species turnover, only three experienced a significant shift in any functional dimension.
-This suggests certain characteristics of the functional space are conserved even in the face of significant change in species identity, namely the size of the functional space occupied by the community (FRic) and the distribution of species and individuals within that space (FEve and FDiv).
+This suggests certain characteristics of the functional space are conserved even in the face of significant change in species identity, specifically the size of the functional space occupied by the community (FRic) and the distribution of species and individuals within that space (FEve and FDiv).
 On average, species additions have similar functional characteristics as lost species and therefore maintain the functional space structure.
-<!-- language to distinguish between "functional traits" and a species' role in the context of the functional space -->
+<!-- While we did not explicitly assess how turnover rates compare to background turnover, --> <!-- language to distinguish between "functional traits" and a species' role in the context of the functional space -->
 

### 2021-09-07 paper/manuscript.Rmd
more citations

@@
 And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @matuoka2020].
-On paleoecological time scales, functional composition is often conserved alongside species diversity even in the face of drastic environmental change [@jackson2015; @barnosky2005].
+On paleoecological time scales, functional composition is often conserved alongside species diversity even in the face of drastic environmental change [@jackson2015; @barnosky2004].
 It is therefore critical to establish whether or not functional loss is the general rule for communities on time scales relevant for current environmental impacts.
@@
 Why then do we find no change in functional diversity despite extensive predictions to the contrary from both **theory** <!-- need more theory in the intro? --> and and many local studies?
-First, the Biotime database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017] (other sampling citation!), and the subset of data in this study exhibits a strong Northern Hemisphere bias.
-We may simply not have data from those areas experiencing the greatest functional shifts [hughes2021].
+
+First, the Biotime database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
+We may simply not have data from those areas experiencing the greatest functional shifts [@hughes2021].
 We do however emphasize that even disturbed communities maintain functional structure [@edwards2013; @matuoka2020], and the inclusion of more heavily impacted sites would not necessarily change the result.

### 2021-09-08 paper/manuscript.Rmd
Flesh out citations

@@
 And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @matuoka2020].
-On paleoecological time scales, functional composition is often conserved alongside species diversity even in the face of drastic environmental change [@jackson2015; @barnosky2004].
-It is therefore critical to establish whether or not functional loss is the general rule for communities on time scales relevant for current environmental impacts.
+On paleoecological time scales, functional composition is often conserved alongside species diversity even in the face of drastic environmental change [@barnosky2004 @jackson2015] .
+It is therefore critical to establish whether or not functional loss is the general rule for communities on time scales relevant for environmental impacts and management action.
 
@@
 fish @trindade-santos2020, birds @jarzyna2016; @barnagaud2017].
-<!-- so what do these studies say? does that go here? --> However, to date there has been no cross-taxa assessment of temporal functional change for a broad geographic and taxonomic extent.
+However, to date there has been no cross-taxa assessment of temporal functional change for a broad geographic and taxonomic extent.
 
 Here we perform the first multi-taxa, multi-system assessment of functional diversity change through time.
-We assess thousands of mammal, bird, and amphibian functional diversity timeseries to 1) establish a general trend of functional change 2) ???.
+We assess thousands of mammal, bird, and amphibian functional diversity timeseries to assess if systematic functional shifts are occurring and to classify the nature of those shifts.
 
@@
 This suggests certain characteristics of the functional space are conserved even in the face of significant change in species identity, specifically the size of the functional space occupied by the community (FRic) and the distribution of species and individuals within that space (FEve and FDiv).
-On average, species additions have similar functional characteristics as lost species and therefore maintain the functional space structure.
-<!-- While we did not explicitly assess how turnover rates compare to background turnover, --> <!-- language to distinguish between "functional traits" and a species' role in the context of the functional space -->
-
--   Paragraph about analog communities?
-
--   Paragraph about implications for BEF work
+On average, species additions have similar functional characteristics as lost species and therefore maintain the structure of the functional space.
+<!-- While we did not explicitly assess how turnover rates compare to background turnover, --> Our results give strong evidence that maintenance of functional structure is not just a phenomena of paleoecological time scales but also happens on annual time scales [@barnosky2004; @jackson2015].
 
 Our work challenges assumptions that functional loss is the default state of all or even many communities.
-While we do not directly assess the disturbance histories of included communities, many are from sites that experienced significant disturbance or manual treatment (see appendix for metadata of included studies).
-<!-- check this, get numbers? --> This is good news for ecosystem function!
+While we do not directly assess the disturbance histories of included communities, trends were consistent even for the longest running and most heavily impacted studies.
+The North American Breeding Survey for example is made up exclusively of roadside surveys and is therefore in relatively close proximity to human impact [@pardieck2020] .
+It is considered an authoritative dataset on the state of bird populations on the continent and underpins policy decisions about bird conservation [@rosenberg2019; @sauer2017].
+No more robust dataset exists to capture North American avian community change, yet we detected no general shifts in functional structure across the dataset.
+Further, none of the 5 included studies that experienced a manual manipulation (e.g. burning, grazing exclosure, etc) experienced any significant functional trends.
+
+<!-- Widespread functional stability has important implications for how we approach the relationship between biodiversity and ecosystem function.
+Just as links between declining species diversity and function may be less relevant than we thought (citations!) so too the realtionship between functional diversity and ecosystem function.
+his is good news for ecosystem function!
 But we may also be measuring the wrong dimensions of the trait space or the wrong traits.
@@
 And presuming that the original structure was good for function, maintaining it is also good for function!
+-->
 
@@
 This study represents the best available data and biodiversity synthesis approaches.
-Why then do we find no change in functional diversity despite extensive predictions to the contrary from both **theory** <!-- need more theory in the intro? --> and and many local studies?
+Why then do we find no change in functional diversity despite extensive predictions to the contrary from both **theory** <!-- need more theory in the intro? -->and many local studies?
 
@@
 We do however emphasize that even disturbed communities maintain functional structure [@edwards2013; @matuoka2020], and the inclusion of more heavily impacted sites would not necessarily change the result.
-
-\<!-
-- You'd imagine the pattern would still be visible even with North American birds...
-
-To further explore the issue we look for example at the North American Breeding Birdy Survey, included in this study and assessed in many other publications.
---\>
 

### 2021-09-08 paper/manuscript.Rmd
Clean up! Finish up hypervolume thoughts

@@
   column_to_rownames("sign") %>%
-  kable(caption = "Number of studies that experienced a signficiant trend in each calculated metric.")
+  kable(caption = "Number of studies that experienced a signficiant trend in each calculated metric out of 53 total studies.")
 ```
@@
 Slopes for functional evenness were also significantly more positive for longer duration studies.
-
-<!-- Cluster analysis grouped all but a few studies into a cluster characterized by non-significant or very small significant slopes for all metrics.
-We identified four other clusters containing six studies (list studies) that were characterized by outlier trends in at least one of the metrics and near-zero trends in the rest.
-The first contained two studies with the most strongly decreasing evenness.
-The second, two studies with the most strongly increasing functional divergence and increasing functional richness.
-The third contained one study with strongly decreasing functional divergence and richness, and the fourth strongly decreasing species richness with increasing functional richness.
--->
 
@@
 
-<!-- Widespread functional stability has important implications for how we approach the relationship between biodiversity and ecosystem function.
-Just as links between declining species diversity and function may be less relevant than we thought (citations!) so too the realtionship between functional diversity and ecosystem function.
-his is good news for ecosystem function!
-But we may also be measuring the wrong dimensions of the trait space or the wrong traits.
-Can't really say BEF needs to measure function directly because these communities aren't losing species or functional diversity - turnover is either within regular bounds or if not (if more turnover than normal) it's still maintaining community structure.
-And presuming that the original structure was good for function, maintaining it is also good for function!
--->
-
 *Implications for Future Work*
 
-<!-- Here's the prediction of the field, here is the data of the field, given that the data doesn't match the prediction it's on the field to resolve these issues. -->
-
-<!-- Despite widespread concern about the loss of functional diversity, our analysis does not indicate significant change in functional composition is the rule for communities. Why -->
-
 This study represents the best available data and biodiversity synthesis approaches.
-Why then do we find no change in functional diversity despite extensive predictions to the contrary from both **theory** <!-- need more theory in the intro? -->and many local studies?
+Why then do we find no change in functional diversity despite extensive predictions to the contrary from both theory and local studies?
 
@@
 For example two communities could have very similar metric values but no overlap in their trait spaces.
-This is especially relevant in the context of biodiversity change as lost species could be replaced by a species with very different functional attributes, but the replacement would go undetected if the new species expanded the trait space by the same degree and had similar abundance.This may be particularly relevant for communities tracking shifting environmental conditions, as communities may maintain some characteristics of their functional composition while shifting to track the new conditions.
-Metrics developed to assess shifts and overlaps in trait spaces are now being developed (citation), and further exploration of time series using those approaches may disentangle trait space characteristics from the location of the occupied space.
-
--   one opening sentence that links back to the null result
-
-Even with the right data and metrics, you may not see a result!
-Due to the wide variety of mechanisms at play we amy not expect to see a global trend even when the local trends are significant.
-
-Our results highlight a gap in both empirical and theoretical work linking shifts in community composition to functional change.
-While a large body of work assesses study-level insights into temporal functional responses (citations), a macroecological perspective illustrates the urgent need for future work exploring the factors important in determining functional change under different species change scenarios.
-Due to the many previously discussed barriers to directly assessing functional shifts, a framework for predicting functional change from community characteristics would be a substantial step forward in understanding biodiversity change.
-
-Much of the work linking species change (or loss) to ecological process either implicitly or explicitly assumes functional loss follows from species loss in natural communities.
-For example ....
-This assumption is clearly refuted by our results.
-
-Our results have important implications for how we approach work linking species loss to ecosystem processes, which often implicitly assume concurrent functional loss.
-There is no evidence of directional functional loss or even functional shifts accompany species loss in these natural communities.
-<!-- clarify that this is about relationship between concurrent patterns in a single community --> Building on a rich body of work linking species diversity changes to changes in ecosystem function, resilience, and stability (citations), it will be essential to tease apart effects due to species change from those due to functional shifts if we want to infer impacts in natural communities.
---\>
-
-We can not understand all functional loss in terms of species loss
-
--   lean in to four points in the abstraction
-
-    -   raises questions for future data collection and theoretical development
-
--   field that's full of idle speculation
-
--   find papers that make claims we're "testing"
-
-<!-- call for theory development, punt back to community ecology - we don't know what this means! -->
-
-Our results emphasize the importance of directly assessing functional change rather than assuming functional loss accompanies species loss.
-
--   getting functional data is costly, so we'd like to be able to say something about udnerlying functional composition without it
-
--   list of studies where they make functional inference without only species level data
-
-Also illustrates the lack of theoretical development surrounding the relationship between functional and species change, and when you would expect different kinds of relationships.
-
--   also theory around the implications of functional shifts for ecological processes
-
-**Notes**
-
-<!-- The absence of systematic loss of functional richness or functional divergence suggests that widespread biological homogenization is not occurring in sampled communities.
-Biological homogenization occurs when endemic or rare species are lost and replaced with generalist or non-native species, resulting in a loss of -->
-
-As the first broad scale analysis of temporal functional diversity change we make x broad conclusions.
-First,
-
-Above all our study shows that patterns of functional change can not be inferred from
-
--   how are these findings consistent with expectations for hypothesized shifts in functional change?
-
--   Biological homogenization is one of the hypothesized meaningful forces
-
--   other unmeasured taxa may be harder hit?
-    different patterns for fish assemblages
-
--   contextualize functional shifts in other temporal studies
-
-While species-level metrics are the only functional data available to perform an analysis of this scale, they are unable to detect functional shifts in communities with little to no turnover as they do not capture intraspecific shifts.
-
-Data limitations:
-
--   species-level traits
-
--   not a lot of changes in these systems already?
-
-Papers to look at:
-
-Magurran et al 2018
-
-Magurran et al 2019 (temporal beta diversity)
-
-McGill 2015
-
-Crossley 2020 - not sure what to do with it
-
-Liang et al 2019 - functional homogenization
-
-Finderup Nielsen et al 2019 - biotic homogenization
-
-Dornelas et al 2019
-
-MacLean et al 2018
-
-Hilldebrand 2018
-
-**Ideas to tie in:**
-
--   species richness loss linked to impacts on processes, in doing so often inferring functional loss, but the two do not go hand in hand in natural communities
-
--   no evidence of biological homogenization
-
--   how is this consistent with theory
-
--   land use history and disturbance showing different relationships between species and functional diversity (Cadotte 2011)
-
-#### Add to supplement
-
--   breakdown of number of studies/timeseries for each taxa/realm etc
-
-They are limitations of the field not our limitations
-
--   acknowledge those things without taking the blame
-
--   lean into it as a conclusion
-
--   we found a null result and its your fault
-
--   fixing x, y, z could resolve the null result, but there are good reasons it might not
-
--   talk about balance of local trends as the fourth reason
-
-There are reasons we haven't got trait data, hoped there would be an easy way out, but these results show that there isn't
-
-Is there a null result for some deep theoretical reason or
-
-we could've been lucky that we would see the pattern with crappy data, but it isn't
-
--   the thing we thought would be there isn't there, so what can we do to solve it
-
-clever way to do temporal coverage - maps by decade?
-
--   number of data points through time (bar chart colored by taxa)
-
-figure 1.
-illustrating functional diversity metrics
-
-figure 2.
-the processing workflow
+This is especially relevant in the context of biodiversity change as lost species could be replaced by a species with very different functional attributes, but the replacement would go undetected if the new species expanded the trait space by the same degree and had similar abundance.
+This scenario may be common in communities tracking changing environmental conditions.
+Approaches for assessing shifts and overlaps in functional space are still relatively new [@barros2016; @blonder2018; @mammola2019] a could shed critical insight into functional composition changes of this nature.

### 2021-09-08 paper/manuscript.Rmd
Update caption

@@
 
-```{r taxa-map, fig.cap="Map of timeseries locations with points color corresponding to taxa.", echo=FALSE}
+```{r taxa-map, fig.cap="A) Map of timeseries locations with points colored by taxa, and B) histograms of timeseries duration broken down by taxa.", echo=FALSE}
 knitr::include_graphics(here::here("figures/study_map_hist.jpeg"))

### 2021-09-08 paper/manuscript.Rmd
Citation

@@
 
-Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package.
-Here we report functional richness (*FRic*), functional evenness (*FEve*), and functional diveregence (*FDiv*) which together describe three complementary characteristics of the functional space [@mason2005, @hillebrand2009].
-*FRic* assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
-*FEve* describes how species are distributed across the trait space and how abundance is distributed across species.
-Higher values of *FEve* indicate more even spacing of species in the trait space and individuals across species.
-*FDiv* measures the degree to which species and their abundances maximize differences in the functional space.
-Higher values of *FDiv* therefore correspond to communities where many highly abundant species are on the edges of the trait space.
+Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package [@laliberté2010].
+Here we report functional richness (FRic), functional evenness (FEve), and functional diveregence (FDiv) which together describe three complementary characteristics of the functional space [@mason2005, @hillebrand2009].
+FRic assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
+FEve describes how species are distributed across the trait space and how abundance is distributed across species.
+Higher values of FEve indicate more even spacing of species in the trait space and individuals across species.
+FDiv measures the degree to which species and their abundances maximize differences in the functional space.
+Higher values of FDiv therefore correspond to communities where many highly abundant species are on the edges of the trait space.
 <!-- use these metrics because they're common, not because of their results -->
@@
 All continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
-The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= s^t$, where $s$ is the number of species and \$t\$ is the number of traits.
+The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= s^t$, where $s$ is the number of species and *t* is the number of traits.
 This restriction allows for a sufficient number of axes to capture the trait space while maintaining computational feasibility.

### 2021-09-10 paper/manuscript.Rmd
rejigger discussion structure, ramblings on the closing paragraph

@@
 
+This striking result could be a product of two possible processes which we discuss futher here.
+First, systemic maintenance of functional structure across communities, or second, a failure of existing data and methods to detect a true underlying change.
+
+*Potential Ecological Implications*
+
 Communities demonstrated almost universal maintenance of functional composition.
@@
 On average, species additions have similar functional characteristics as lost species and therefore maintain the structure of the functional space.
-<!-- While we did not explicitly assess how turnover rates compare to background turnover, --> Our results give strong evidence that maintenance of functional structure is not just a phenomena of paleoecological time scales but also happens on annual time scales [@barnosky2004; @jackson2015].
+<!-- While we did not explicitly assess how turnover rates compare to background turnover, --> Our results give strong evidence that maintenance of functional structure is not just a phenomena of paleoecological time scales but also annual time scales [@barnosky2004; @jackson2015].
 
@@
 
-*Implications for Future Work*
+*Potential Methodological Limitations*
 
@@
 The importance of intraspecific variation is well documented [@desroches2018; @violle2012], however individual-level traits are rarely collected along side monitoring data, especially for the longest running efforts.
-Further investment in individual level trait data for long-term monitoring will be critical for assessing finer scale shifts that likely occur more commonly in communities and potentially preceded bigger changes.
+Species-level traits may be obscuring more subtle shifts in the trait space happening within species.
 Likewise, available trait data may not capture the traits experiencing the greatest change.
@@
 For example two communities could have very similar metric values but no overlap in their trait spaces.
-This is especially relevant in the context of biodiversity change as lost species could be replaced by a species with very different functional attributes, but the replacement would go undetected if the new species expanded the trait space by the same degree and had similar abundance.
+This is especially relevant in the context of biodiversity change as a species loss could be replaced by a species with very different functional attributes, but the replacement would go undetected if the new species expanded the trait space by the same degree and had similar abundance.
 This scenario may be common in communities tracking changing environmental conditions.
 Approaches for assessing shifts and overlaps in functional space are still relatively new [@barros2016; @blonder2018; @mammola2019] a could shed critical insight into functional composition changes of this nature.
+
+*Future Work*
+
+Ecology is now in a pickle because both of these options are totally plausible.
+Where do we go from here?
+We need more data, but not data that continues to replicate existing biases (i.e. not more body masses from north american species).
+Targeted collection of even species-traits we may think are especially impacted by environemental change and have particular relevance for function.
+Geographic biases, esp long term monitoring.
+Simulation/modeling work to tease apart scenarios where we can reliable identify regulation in individual data sets, identify regulation vs dis regulation.

### 2021-09-14 paper/manuscript.Rmd
Discussion discussion discussion

@@
 ---
-title: "Evidence of widspread maintenence of functional composition in vertebrate communities"
+title: "Evidence of widespread maintenance of functional composition in vertebrate communities"
 author: 
@@
 Surprisingly, we did not detect an overall trend in any of the calculated functional diversity metrics.
-As with previous species-based syntheses, we also found no overall trend in species diversity accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics may be consistent with similar well-documented species derived trends.
+As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics may be consistent with similar well-documented species derived trends.
 We found no evidence of systemic functional richness loss or functional change.
@@
 
-This striking result could be a product of two possible processes which we discuss futher here.
-First, systemic maintenance of functional structure across communities, or second, a failure of existing data and methods to detect a true underlying change.
+This striking result could be a product of two possible processes, one ecological and one methodological.
+Null trends appear to give strong evidence of systemic maintenance of functional structure due to common ecological processes, however multiple limitations of current approaches in synthesis weaken the conclusions possible from our data due to potential statistical artefacts.
+We discuss both options further here.
 
-*Potential Ecological Implications*
+*Evidence of Ecological Processes*
 
@@
 
-Our work challenges assumptions that functional loss is the default state of all or even many communities.
+These results challenge assumptions that functional loss is the default state of all or even many communities.
 While we do not directly assess the disturbance histories of included communities, trends were consistent even for the longest running and most heavily impacted studies.
@@
 
-This study represents the best available data and biodiversity synthesis approaches.
-Why then do we find no change in functional diversity despite extensive predictions to the contrary from both theory and local studies?
+Here we approach the question of functional change using the best available data and biodiversity synthesis approaches.
+However, a number of gaps in best practices may be obscuring a true underlying trend.
 
 First, the Biotime database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
-We may simply not have data from those areas experiencing the greatest functional shifts [@hughes2021].
+We may simply not have data from those areas experiencing the greatest perturbation [@hughes2021].
 We do however emphasize that even disturbed communities maintain functional structure [@edwards2013; @matuoka2020], and the inclusion of more heavily impacted sites would not necessarily change the result.
@@
 Second, despite using the most comprehensive trait databases for these taxa, we were still limited to species-level means of the traits deemed important by database creators.
-The importance of intraspecific variation is well documented [@desroches2018; @violle2012], however individual-level traits are rarely collected along side monitoring data, especially for the longest running efforts.
+The importance of intraspecific variation is well documented [@desroches2018; @violle2012], however individual-level traits are rarely collected alongside monitoring data, especially for the longest running efforts.
 Species-level traits may be obscuring more subtle shifts in the trait space happening within species.
@@
 Most notably, the metrics we calculated do not capture shifts in the location of the functional space.
-For example two communities could have very similar metric values but no overlap in their trait spaces.
+For example, two communities could have very similar metric values but no overlap in their trait spaces.
 This is especially relevant in the context of biodiversity change as a species loss could be replaced by a species with very different functional attributes, but the replacement would go undetected if the new species expanded the trait space by the same degree and had similar abundance.
@@
 
+Our results illustrates a well known but increasingly pervasive problem in synthesis-based ecology of distinguishing true ecological patterns from methodological artifacts.
+We present here the conclusion best supported by available data with a number of caveats and suggestions for next steps forward.
+
+The most pressing need is for intentional and targeted data collection efforts.
+We join others in the call for increased monitoring in under sampled areas and continued efforts to centralize existing data sources [@gonzalez2016; @hughes2021; @vellend2017].
+Data that fill geographic, taxonomic and trait gaps should be prioritized over further collection of data that replicate existing biases.
+One relatively low-cost high-reward data investment is collation of additional species-level trait means.
+Intentional trait selection is critical for linking functional patterns to ecological process [@zhu2017] (MORE CITATION), however synthesis is constrained to the traits in a few taxa-specific databases.
+Prioritization of additional traits linked to environmental change drivers (response traits) or ecosystem function (effect traits) in particular would provide a powerful link to existing frameworks [@lavorel2002] .
+
+A more diverse set of traits would facillitate work linking macroecological analysis with theory of community change and ecosystem function.
+For example, identification of traits impacted by environmental change drivers (response traits) or which have the greatest impact on function (effect traits) [@lavorel2002].
+In addition to potentially better capturing change in communities, this framework would provide a much needed link between biogeographic patterns and theory.
+
+Our results give both a first look at the current state of functional change in communities and illustrate critical next steps for distinguishing true ecological patterns from artefacts.
+The clearest need is Understanding would be greatly improved by intentional and targeted data collection.
+We join others in emphasizing a call for increased monitoring in under sampled areas, particularly the global south, along with investment in collection of individual-level trait data.
+<!-- this is not a unique call, all of ecology would be benefited, we also recognize that this is a huge systemic problem that will require x, y, z investment --> While this clearly represents a long term investment in the future of ecology, there are also a number of things that can be done now.
+
+In the short term, broad-scale functional studies could benefit greatly from expansion of species-level trait values to allow for more intentional selection of relevant traits.
+A more diverse set of traits would allow ...
+especially those hypothesized to be particularly impacted by environmental change drivers (response traits) or which have the greatest impact on function (effect traits) [@lavorel2002].
+In addition to potentially better capturing change in communities, this framework would provide a much needed link between biogeographic patterns and theory.
+
+-   better geographic sampling
+
+-   better trait sampling - both intraspecific traits (in the field) and new trait axes (not necessarily in the field)
+
+-   we may need more theory to inform which traits are most relevant for environmental change and ecological function
+
 Ecology is now in a pickle because both of these options are totally plausible.
@@
 We need more data, but not data that continues to replicate existing biases (i.e. not more body masses from north american species).
-Targeted collection of even species-traits we may think are especially impacted by environemental change and have particular relevance for function.
+Targeted collection of even species-traits we may think are especially impacted by environmental change and have particular relevance for function.
 Geographic biases, esp long term monitoring.

### 2021-09-14 paper/manuscript.Rmd
clean up

@@
 One relatively low-cost high-reward data investment is collation of additional species-level trait means.
-Intentional trait selection is critical for linking functional patterns to ecological process [@zhu2017] (MORE CITATION), however synthesis is constrained to the traits in a few taxa-specific databases.
-Prioritization of additional traits linked to environmental change drivers (response traits) or ecosystem function (effect traits) in particular would provide a powerful link to existing frameworks [@lavorel2002] .
-
-A more diverse set of traits would facillitate work linking macroecological analysis with theory of community change and ecosystem function.
-For example, identification of traits impacted by environmental change drivers (response traits) or which have the greatest impact on function (effect traits) [@lavorel2002].
-In addition to potentially better capturing change in communities, this framework would provide a much needed link between biogeographic patterns and theory.
-
-Our results give both a first look at the current state of functional change in communities and illustrate critical next steps for distinguishing true ecological patterns from artefacts.
-The clearest need is Understanding would be greatly improved by intentional and targeted data collection.
-We join others in emphasizing a call for increased monitoring in under sampled areas, particularly the global south, along with investment in collection of individual-level trait data.
-<!-- this is not a unique call, all of ecology would be benefited, we also recognize that this is a huge systemic problem that will require x, y, z investment --> While this clearly represents a long term investment in the future of ecology, there are also a number of things that can be done now.
-
-In the short term, broad-scale functional studies could benefit greatly from expansion of species-level trait values to allow for more intentional selection of relevant traits.
-A more diverse set of traits would allow ...
-especially those hypothesized to be particularly impacted by environmental change drivers (response traits) or which have the greatest impact on function (effect traits) [@lavorel2002].
-In addition to potentially better capturing change in communities, this framework would provide a much needed link between biogeographic patterns and theory.
-
--   better geographic sampling
-
--   better trait sampling - both intraspecific traits (in the field) and new trait axes (not necessarily in the field)
-
--   we may need more theory to inform which traits are most relevant for environmental change and ecological function
-
-Ecology is now in a pickle because both of these options are totally plausible.
-Where do we go from here?
-We need more data, but not data that continues to replicate existing biases (i.e. not more body masses from north american species).
-Targeted collection of even species-traits we may think are especially impacted by environmental change and have particular relevance for function.
-Geographic biases, esp long term monitoring.
-Simulation/modeling work to tease apart scenarios where we can reliable identify regulation in individual data sets, identify regulation vs dis regulation.
+Intentional trait selection is critical for linking functional patterns to ecological process [@zhu2017], however synthesis is constrained to the traits in a few taxa-specific databases.
+Prioritization of additional traits linked to environmental change drivers (response traits) or ecosystem function (effect traits) in particular would provide a powerful link to existing frameworks for understanding functional change [@lavorel2002].

### 2021-09-16 paper/manuscript.Rmd
add citation

@@
 output: #word_document
-  rticles::elsevier_article:
+  rticles::pnas_article:
   includes:

### 2021-09-16 data-raw/unused_trait_sources/fishtraits.Rmd
Rename and rearrange data-raw folder

--- 2019-09-09 data-raw/fishtraits.Rmd
@@
 
+Oh wait, we actually have the whole database now, sigh. (sciencebase.gov/catalog/item/5a7c6e8ce4b00f54eb2318c0)
+```{r}
+#system.file("extdata", "fishtraits_db", package = "biodivTS")
+
+ft <- read_csv(here::here("inst", "extdata", "fishtraits_db", "FishTraits_14.3.csv"), na = c("NA", "-999", "-555", "-999.0", "-555.0", "-555.00", "-999.00")) %>%
+  rename_all(tolower) %>%
+  unite(scientificName, genus, species, sep = " ") %>%
+  select(scientificName, itistsn,
+                      diet_nonfeed = nonfeed, diet_benthic = benthic, diet_surwcol = surwcol, diet_algphyto = algphyto, diet_macvascu = macvascu, 
+                      diet_detritus = detritus, diet_inv = invlvfsh, diet_fshcrcrb = fshcrcrb, diet_blood = blood, diet_eggs = eggs, diet_other = other, 
+                      max_length = maxtl, matuage, longevity, fecundity, serial, season, starts_with("a_"), starts_with("b_"), starts_with("c1_"),
+                      euryhaline, mintemp, maxtemp, muck, claysilt, sand, gravel, cobble, boulder, bedrock, vegetat, debrdetr, lwd,
+                      pelagic, preflot, preflen, largeriv, smallriv, creek, sprgsubt, lacustrine, potanadr, lowland, upland, montane,
+                      slowcurr, modcurr, fastcurr) %>%
+  filter(scientificName != "Salvelinus aureolus") #this isn't a real entry, all NA's
+
+missing_lh <- ft %>% 
+  #select(starts_with("a_"), starts_with("b_")) %>%
+  filter_at(vars(starts_with("a_"), starts_with("b_"), starts_with("c1_")),  ~ !. %in% TRUE) %>%
+  pull(itistsn) %>%
+  unique()
+
+ft_clean <- ft %>%
+  gather(lifehistory, value, starts_with("a_"), starts_with("b_"), starts_with("c1_")) %>%
+  mutate(lifehistory = case_when(
+    itistsn %in% missing_hab ~ NA_character_,
+    TRUE ~ lifehistory
+    )
+  ) %>% 
+  filter(value == TRUE | is.na(lifehistory), itistsn != 0) %>%
+  select(-value) %>%
+  distinct()
+```
+
+ids
+```{r}
+fishtraits <- ft_clean %>%
+  mutate(itistsn = paste0("ITIS:", itistsn), 
+         id = get_ids(scientificName))
+
+#There are no instances where the database id doesn't match the id from taxadb
+fishtraits %>% filter(itistsn != id) %>% dim()
+
+#so we'll just keep that ID column
+fishtraits <- fishtraits %>%
+  mutate(id = itistsn) %>%
+  select(-itistsn)
+```
+
 ```{r}

### 2021-09-16 data-raw/unused_trait_sources/hfp_data.Rmd
Rename and rearrange data-raw folder

---
title: "Human Footprint"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Extract zip files
```{r}
# library(archive)
# hf_arhc <- archive_extract(here::here("data", "HumanFootprintv2.7z.zip"), here::here("data"))
```

Read in raster
```{r}
library(raster)
library(sf)
library(dplyr)
library(geosphere)
library(osmdata)

# met_data <- pins::pin_get("rarefied-metrics", board = "github")
# rarefyIDs <- met_data %>% pull(rarefyID) %>% unique()

hfp_1993 <- raster(here::here("data/Dryadv3/Maps/HFP1993_int.tif")) %>%
  reclassify(., c(100, 128, NA))

hfp_2009 <- raster(here::here("data/Dryadv3/Maps/HFP2009_int.tif")) %>%
  reclassify(., c(100, 128, NA))

#study-level points
# meta <- pins::pin_get("meta", board = "github")
# study_points <- st_as_sf(meta, coords = c("cent_lat", "cent_long"))

#get rarefyid centers
rare_cells <- pins::pin_get("rarefyID-cell-centre", board = "github")
points <- st_as_sf(rare_cells, coords = c("rarefyID_x", "rarefyID_y"), crs = "EPSG:4326") %>%
  st_transform(points, crs = st_crs(hfp_1993)) %>%
  #filter(rarefyID %in% rarefyIDs) %>% 
  dplyr::select(-cell_extent)

# Get 1993 data
ex_1993 <- extract(hfp_1993, points, buffer = 1000, fun = mean, na.rm = TRUE, sp = TRUE) %>%
  st_as_sf(crs = st_crs(hfp_1993))
ex_1993_bigbuff <- ex_1993 %>% filter(is.na(HFP1993_int)) %>% dplyr::select(-HFP1993_int) %>%
  extract(hfp_1993, ., buffer = 45000, fun = mean, na.rm = TRUE, sp = TRUE) %>%
  st_as_sf(crs = st_crs(hfp_1993))

# Get2009 data
ex_2009 <- extract(hfp_2009, points, buffer = 1000, fun = mean, na.rm = TRUE, sp = TRUE) %>%
  st_as_sf(crs = st_crs(hfp_1993))
ex_2009_bigbuff <- ex_2009 %>% filter(is.na(HFP2009_int)) %>% dplyr::select(-HFP2009_int) %>%
  extract(hfp_2009, ., buffer = 45000, fun = mean, na.rm = TRUE, sp = TRUE) %>%
  st_as_sf(crs = st_crs(hfp_1993))

hfp_data <- bind_rows(bind_rows(ex_1993, ex_1993_bigbuff) %>% mutate(hfpYear = 1993), 
          bind_rows(ex_2009, ex_2009_bigbuff) %>% mutate(hfpYear = 2009))

hfp_data <- left_join(bind_rows(ex_1993 %>% filter(!rarefyID %in% ex_1993_bigbuff$rarefyID), ex_1993_bigbuff) %>% 
                        rename(`1993` = HFP1993_int) %>%
                        st_drop_geometry(), 
          bind_rows(ex_2009 %>% filter(!rarefyID %in% ex_2009_bigbuff$rarefyID), ex_2009_bigbuff) %>% 
            rename(`2009` = HFP2009_int) %>%
            st_drop_geometry()) %>%
  pivot_longer(c(`1993`, `2009`), names_to = "year", values_to = "hfp")

pins::pin(hfp_data, board = "github")
```

Attempt at getting distance from shore
```{r}
#Get list of countries
# pins::board_register_github(repo = "karinorman/biodivTS_data", branch = "master")
# country_data <- pins::pin_get("country-data-bind", board = "github") %>%
#   mutate(country = str_remove(country, "Commonwealth of The "),
#          country = str_remove(country, "Kingdom of "),
#          country = str_remove(country, "Republic of "),
#          country = str_remove(country, "Kingdom of "),
#          country = str_remove(country, "Federative Republic of "),
#          country = str_remove(country, "Federative "),) %>%
#   mutate(country = case_when(
#     country == "United States" ~ "USA",
#     country == "Russian Federation" ~ "Russia",
#     country == "United Great Britain and Northern Ireland" ~ "UK",
#     country == "Czechia" ~ "Czech",
#     TRUE ~ country
#   )) 
# 
# map(unique(country_data$country), osm_box)
# 
# get_coasts <- function(country_name) {
#   osm_box <- getbb(place_name = country_name) %>%
#   opq() %>%
#   add_osm_feature("natural", "coastline") %>%
#   osmdata_sf()
# }
# osm_box <- getbb(place_name = "World") %>%
#   opq() %>%
#   add_osm_feature("natural", "coastline") %>%
#   osmdata_sf()
# 
# dist <- geosphere::dist2Line(p = st_as_sf(rare_cells, coords = c("rarefyID_x", "rarefyID_y"), crs = "EPSG:4326") %>% st_coordinates(), 
#                          line = st_coordinates(osm_box$osm_lines)[,1:2])
# 
# #combine initial data with distance to coastline
# df <- cbind( d1 %>% rename(y=lat,x=long),dist) %>%
#   mutate(miles=distance/1609)
# 
# st_as_sf(rare_cells, coords = c("rarefyID_x", "rarefyID_y"), crs = "EPSG:4326")
```

Plot
```{r}
# plot all points 
plot(hfp_1993) 
plot(points, pch = 19, add = TRUE)

plot(hfp_1993) 
plot(ex_1993_bigbuff %>% filter(is.na(HFP1993_int), rarefyID %in% model_data$rarefyID) , pch = 10, add = TRUE)
```

### 2021-09-16 analysis/other_Blowes_analysis/05_visualise_data.R
delete

@@ -1,167 +1,0 @@
-rm(list=ls())
-# some plots of teh data the model were fit to
-library(tidyverse)
-library(brms)
-
-#--------the data
-load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
-
-
-cell_count <- rarefied_medians %>%
-  group_by(biome_map, taxa_mod) %>%
-  dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-  ungroup()
-
-##	rejoin
-rarefied_medians <- left_join(cell_count, rarefied_medians, by=c('biome_map', 'taxa_mod'))
-
-##	filter to count data and biome/taxa combinations with >3 cells
-# these are the data that the model was fit to
-rarefied_medians <- rarefied_medians %>%
-  filter(BROAD_TYPE=='count' & n_cells > 3)
-
-# do any cells have multiple rarefyID's
-rarefied_medians %>%
-  group_by(cell) %>%
-  summarise(n_rarefyID = n_distinct(rarefyID),
-            n_studyID = n_distinct(STUDY_ID)) %>%
-  filter(n_rarefyID > 2)
-
-# create df with the unique rarefyIDs
-newdat <-  rarefied_medians %>%
-               distinct(biome_map, taxa_mod, STUDY_ID, rarefyID,
-                        SamplePool, SampleN, num_years,
-                        duration, startYear, endYear)
-
-# how many time series with duration > xx
-newdat %>% filter(duration > 20)
-newdat %>% filter(duration > 10)
-
-# cell-counts per taxa
-newdat %>%
-  mutate(total_cells = n_distinct(rarefyID)) %>%
-  group_by(REALM, biome_map, taxa_mod) %>%
-  summarise(n_cell = n_distinct(rarefyID),
-            percent_of_total = n_cell/unique(total_cells))# %>%
-  # write.csv(file = '~/Dropbox/BiogeoBioTIME/Biogeo Science submission/Biogeo Science Revision/figures/taxa_cell_count.csv')
-
-# want counts of SL and ML studies
-bt_grid_filtered <- pins::pin_get("bt-traitfiltered", board = "github")
-
-bt_grid_filtered %>%
-  filter(study_id %in% newdat$STUDY_ID) %>%
-  distinct(study_id, .keep_all = T) %>%
-  summarise(n.sl = sum(studymethod=='SL'),
-            n.ml = sum(studymethod=='ML'))
-
[…]

### 2021-09-16 analysis/other_Blowes_analysis/06_rarefysamplesturnoverbnh.R

@@ -1,132 +1,0 @@
-# Maria Dornelas 13.05.2013
-# rarefying for equal number of samples and calculating temporal turnover
-
-rarefysamplesturnoverbnh<-function(Year, SampleID, Species, Abundance, resamps) {
-#######################################################################
-  # takes as input a  Year, SampleID, Species, Abundance and number of resamples
-  # which should be in dataframe so that elements match
-  # calculates turnover:
-  # 1) between each year and the first year 
-  # 2) between pairs of adjacent years 
-  # 3) between each year and the last year of the time series
-  # for the rarefied pooled samples
-###########################################################################
-library(vegan)
-counter<-1
-simbaseline<-data.frame(array(NA,dim=c(length(unique(Year)),7)))
-names(simbaseline)<-c("Year", "Jaccard","Horn","Chao","Pearson", "Jtu", "Jne")
-
-simnext<-data.frame(array(NA,dim=c(length(unique(Year)),7)))
-names(simnext)<-c("Year", "Jaccard","Horn","Chao","Pearson", "Jtu", "Jne")
-
-simhind<-data.frame(array(NA,dim=c(length(unique(Year)),7)))
-names(simhind)<-c("Year", "Jaccard","Horn","Chao","Pearson", "Jtu", "Jne")
-
-counter2<-1
-
-# getting vector with number of samples per year
-nsamples<-c()
-for(y in unique(Year)){
-  nsamples<-c(nsamples, length(unique(SampleID[Year==y])))
-}
-t<-1
-minsample<-min(nsamples)
-for(repeats in 1:resamps){
-  raref<-data.frame(array(NA,dim=c(1,3)))
-  names(raref)<-c("Year","Species","Abundance")
-  for(y in unique(Year)){
-    #getting samples for this year
-    samps<-unique(SampleID[Year==y])
-    # re-sampling samples to equalize number of samples
-    sam<-as.character(sample(samps,minsample,replace=T))
-    # getting data that belongs to bootstraped samples
-    rarefyear<-data.frame(SampleID[which(SampleID %in% sam & Year == y)],
-                          Species[which(SampleID %in% sam & Year == y)],
-                          Abundance[which(SampleID %in% sam & Year == y)])
-    names(rarefyear)<-c("SampleID", "Species", "Abundance")
-    # calculating pooled abundances of eahc species to store
[…]

### 2021-09-16 analysis/other_Blowes_analysis/07_sim_repeat_sampling.R

@@ -1,168 +1,0 @@
-# source code to do turnover calculations
-rm(list=ls())
-
-library(tidyverse)
-source('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/06_rarefysamplesturnoverbnh.R')
-
-set.seed(403)
-
-# simulation parameters
-n_sim <- 1000
-# length of the time series for each simulation
-# time_series_length <- floor(runif(n_sim, min = 2, max = 50))
-# what is the empirical distribution of time series length in our analyses?
-load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
-cell_count <- rarefied_medians %>%
-  group_by(Biome, taxa_mod) %>%
-  dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-  ungroup() 
-
-##	rejoin
-rarefied_medians <- left_join(cell_count, rarefied_medians, by=c('Biome', 'taxa_mod'))
-
-# get parameters of poisson lognormal for duration
-duration_poilog <- poilog::poilogMLE(rarefied_medians %>% 
-                    group_by(rarefyID) %>%
-                    # distinct(num_years) %>%
-                    distinct(duration) %>% .$duration)
-
-##	filter to count data and biome/taxa combinations with >3 cells
-# these are the data that the model was fit to
-rarefied_medians <- rarefied_medians %>%
-  filter(BROAD_TYPE=='count' & n_cells > 3)
-# duration
-inset <- rarefied_medians %>% 
-  group_by(rarefyID) %>%
-  # distinct(num_years) %>%
-  distinct(duration) %>%
-  # summarise(mean(log(duration)),
-            # sd(log(duration)))
- ggplot() +
-  geom_density(aes(x = duration, fill = 'empirical')) +
-  # squint (or assume something) and it could be lognormal 
-  geom_density(data = data.frame(duration_sim = rlnorm(1000,2.3,0.65)),#poilog::rpoilog(n_distinct(rarefied_medians %>%
-                                                                             # distinct(duration) %>% 
-                                                                             # .$duration),
-                                                                # mu = duration_poilog$par[1],
-                                                                # sig = duration_poilog$par[2])),#sads::rsad(#S = 50000, 
-                                                           # frac = 1,
[…]

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/00_HLM_Jne.R

@@ -1,50 +1,0 @@
-##======================================================================
-##  Hierarchical models: fit to either all studies (S, Jaccard), or only those with
-##	count data (N, PIE, etc)
-##======================================================================
-rm(list=ls())
-
-##	
-#withr::with_libpaths(new = '/home/blowes/R/x86_64-pc-linux-gnu-library/3.3/', library(brms))
-library(brms)
-library(dplyr)
-library(tidyr)
-##======================================================================
-##	alternatively load averaged metrics
-load('/gpfs1/data/sChange/BioTime/data/rarefied_medians.Rdata')
-
-##======================================================================		
-##	count cells within realm/climate/taxa hierarchy
-cell_count <- rarefied_medians %>%
-        group_by(Biome, taxa_mod) %>%
-        dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-        ungroup()
-
-##	rejoin
-rarefied_medians <- left_join(cell_count, rarefied_medians, 
-	by=c('Biome','taxa_mod'))
-
-# need to set first year to zero dissimilarity for consecutive beta metrics
-#rarefied_medians <- rarefied_medians %>%
-#        mutate(Jne_next = ifelse(YEAR==startYear, 0, Jne_next))
-
-##	filter to ecoregions with >3 cells
-rarefied_medians <- rarefied_medians %>%
-        filter(n_cells > 3)
-
-##	set some weakly regularising priors...
-hier_prior <- c(set_prior(prior = 'normal(0,1)', class='b', coef='cYEAR'), 	# global slope
-	set_prior(prior = 'normal(0,2)', class='Intercept', coef=''), 		# global intercept
-	set_prior(prior = 'cauchy(0,2)', class='sd'),							# group-level intercepts and slopes
-	set_prior(prior = 'lkj(2)', class='cor'))
-
-Jne_norm_BTSRfyID <- brm(bf(Jne_base ~ cYEAR + (cYEAR|Biome/taxa_mod/STUDY_ID/rarefyID), 
-	family = brmsfamily('gaussian')),	
-	data= rarefied_medians %>% filter(BROAD_TYPE=='count'),
-	prior=hier_prior,
-	inits = '0',
-	iter = 2000,
-	cores = 4,
-	chains = 4)
-
-save(Jne_norm_BTSRfyID, file=Sys.getenv('OFILE'))

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/00_HLM_Jtu.R

@@ -1,50 +1,0 @@
-##======================================================================
-##  Hierarchical models: fit to either all studies (S, Jaccard), or only those with
-##	count data (N, PIE, etc)
-##======================================================================
-rm(list=ls())
-
-##	
-#withr::with_libpaths(new = '/home/blowes/R/x86_64-pc-linux-gnu-library/3.3/', library(brms))
-library(brms)
-library(dplyr)
-library(tidyr)
-##======================================================================
-##	alternatively load averaged metrics
-load('/gpfs1/data/sChange/BioTime/data/rarefied_medians.Rdata')
-
-##======================================================================		
-##	count cells within realm/climate/taxa hierarchy
-cell_count <- rarefied_medians %>%
-        group_by(Biome, taxa_mod) %>%
-        dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-        ungroup()
-
-##	rejoin
-rarefied_medians <- left_join(cell_count, rarefied_medians, 
-	by=c('Biome','taxa_mod'))
-
-# need to set first year to zero dissimilarity for consecutive beta metrics
-#rarefied_medians <- rarefied_medians %>%
-#        mutate(Jtu_next = ifelse(YEAR==startYear, 0, Jtu_next))
-
-##	filter to ecoregions with >3 cells
-rarefied_medians <- rarefied_medians %>%
-        filter(n_cells > 3 & BROAD_TYPE == 'count')
-
-##	set some weakly regularising priors...
-hier_prior <- c(set_prior(prior = 'normal(0,1)', class='b', coef='cYEAR'), 	# global slope
-	set_prior(prior = 'normal(0,2)', class='Intercept', coef=''), 		# global intercept
-	set_prior(prior = 'cauchy(0,2)', class='sd'),							# group-level intercepts and slopes
-	set_prior(prior = 'lkj(2)', class='cor'))
-
-Jtu_norm_BTSRfyID <- brm(bf(Jtu_base ~ cYEAR + (cYEAR|Biome/taxa_mod/STUDY_ID/rarefyID), 
-	family = brmsfamily('gaussian')),	
-	data= rarefied_medians,
-	prior=hier_prior,
-	inits = '0',
-	iter = 2000,
-	cores = 4,
-	chains = 4)
-
-save(Jtu_norm_BTSRfyID, file=Sys.getenv('OFILE'))

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/00_HLM_Jtu_betaFix.R

@@ -1,45 +1,0 @@
-##  Hierarchical models: fit to either all studies (S, Jaccard), or only those with
-##	count data (N, PIE, etc)
-##======================================================================
-rm(list=ls())
-
-##	
-library(brms)
-library(dplyr)
-##======================================================================
-##	alternatively load averaged metrics
-load('/gpfs1/data/sChange/BioTime/data/rarefied_medians.Rdata')
-##======================================================================
-##      count cells within Biome/taxa hierarchy, add STUDY_ID 
-cell_count <- rarefied_medians %>%
-        group_by(Biome, taxa_mod) %>%
-        dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-        ungroup()
-
-##      rejoin
-rarefied_medians <- left_join(cell_count, rarefied_medians, by=c('Biome','taxa_mod'))
-
-## filter to ecoregions with >3 cells
-rarefied_medians <- rarefied_medians %>%
-        filter(n_cells > 3) %>%
-	# add dirty fix for 0's and 1's
-	mutate(Jtu.fix = ifelse(Jtu_base==0, Jtu_base + 1e-6, Jtu_base),
-               Jtu.fix = ifelse(Jtu.fix==1, Jtu.fix - 1e-3, Jtu.fix))
-
-##	set some priors...
-hier_prior <- c(set_prior(prior = 'normal(0,0.05)', class='b', coef='cYEAR'), 	# global slope
-	set_prior(prior = 'normal(0,2)', class='Intercept', coef=''), 		# global intercept
-	set_prior(prior = 'cauchy(0,2.5)', class='sd'),							# group-level intercepts and slopes
-	set_prior(prior = 'lkj(2)', class='cor'))
-
-##	stan via brms
-Jtu_beta_BTSRfyID_countData <- brm(bf(Jtu.fix ~ cYEAR + (cYEAR|Biome/taxa_mod/STUDY_ID/rarefyID),
-	#zoi ~ REALM, coi ~ REALM, 
-        family = brmsfamily('Beta')),       
-        data= filter(rarefied_medians, BROAD_TYPE=='count'),
-        prior=hier_prior,
-        inits = '0',
-        #control = list(adapt_delta=0.99),
-        cores = 4)
-
-save(Jtu_beta_BTSRfyID_countData, file=Sys.getenv('OFILE'))

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/00_HLM_Jtu_z1i.R

@@ -1,42 +1,0 @@
-##  Hierarchical models: fit to either all studies (S, Jaccard), or only those with
-##	count data (N, PIE, etc)
-##======================================================================
-rm(list=ls())
-
-##	
-library(brms)
-library(dplyr)
-##======================================================================
-##	alternatively load averaged metrics
-load('/gpfs1/data/sChange/BioTime/data/rarefied_medians.Rdata')
-##======================================================================
-##      count cells within Biome/taxa hierarchy, add STUDY_ID 
-cell_count <- rarefied_medians %>%
-        group_by(Biome, taxa_mod) %>%
-        dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-        ungroup()
-
-##      rejoin
-rarefied_medians <- left_join(cell_count, rarefied_medians, by=c('Biome','taxa_mod'))
-
-## filter to ecoregions with >3 cells
-rarefied_medians <- rarefied_medians %>%
-        filter(n_cells > 3)
-
-##	set some priors...
-hier_prior <- c(set_prior(prior = 'normal(0,1)', class='b', coef='cYEAR'), 	# global slope
-	set_prior(prior = 'normal(0,2)', class='Intercept', coef=''), 		# global intercept
-	set_prior(prior = 'cauchy(0,2.5)', class='sd'),							# group-level intercepts and slopes
-	set_prior(prior = 'lkj(2)', class='cor'))
-
-##	stan via brms
-Jtu_z1i_realm_BTSRfyID_countData <- brm(bf(Jtu_base ~ cYEAR + (cYEAR|Biome/taxa_mod/STUDY_ID/rarefyID),
-	zoi ~ REALM, coi ~ REALM, 
-        family = brmsfamily('zero_one_inflated_beta')),       
-        data= filter(rarefied_medians, BROAD_TYPE=='count'),
-        prior=hier_prior,
-        inits = '0',
-        control = list(adapt_delta=0.99),
-        cores = 4)
-
-save(Jtu_z1i_realm_BTSRfyID_countData, file=Sys.getenv('OFILE'))

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/00_HLM_S.R

@@ -1,51 +1,0 @@
-##======================================================================
-##  Hierarchical models: fit to either all studies (S, Jaccard), or only those with
-##	count data (N, PIE, etc)
-##======================================================================
-rm(list=ls())
-
-##	
-#withr::with_libpaths(new = '/home/blowes/R/x86_64-pc-linux-gnu-library/3.3/', library(brms))
-library(brms)
-library(dplyr)
-library(tidyr)
-##======================================================================
-##	alternatively load averaged metrics
-load('/gpfs1/data/sChange/BioTime/data/rarefied_medians.Rdata')
-
-##======================================================================		
-# cell count for a hierarchy that includes study
-cell_count <- rarefied_medians %>%
-        group_by(Biome, taxa_mod) %>%
-        dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-        ungroup()
-
-##	rejoin
-rarefied_medians <- left_join(cell_count, rarefied_medians, by=c('Biome','taxa_mod'))
-
-##	filter to ecoregions with >3 cells
-rarefied_medians <- rarefied_medians %>%
-	ungroup() %>%
-	filter(BROAD_TYPE == 'count') %>%
-        filter(n_cells > 3)
-
-##	set some weakly regularising priors...
-hier_prior <- c(set_prior(prior = 'normal(0,0.2)', class='b', coef='cYEAR'), 	# global slope
-	set_prior(prior = 'normal(0,1)', class='Intercept', coef=''), 		# global intercept
-	#set_prior(prior = 'exponential(1)', class = 'sd'),
-	set_prior(prior = 'lkj(1)', class='cor')
-)
-
-S_pois_BTSRfyID_countData <- brm(bf(S ~ cYEAR + (cYEAR|Biome/taxa_mod/STUDY_ID/rarefyID), 
-	family = brmsfamily('poisson')),	
-	data= rarefied_medians,
-	prior=hier_prior,
-	iter = 2000,
-	warmup = 1000,
-	inits = '0',
-	#init_r = 0.001,
-	control = list(adapt_delta=0.95),
-	cores = 4,
-	chains = 4)
-
-save(S_pois_BTSRfyID_countData, file=Sys.getenv('OFILE'))

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/00a_nuts_diagnostics.R

@@ -1,92 +1,0 @@
-# code to check nuts (no u-turn sampler) diagnostics for all models
-# code to run remotely and save results to file
-rm(list=ls())
-
-library(brms)
-library(rstan)
-library(dplyr)
-library(tibble)
-
-# source nuts diagnostics file (from Michael Betancourt)
-source('/gpfs1/data/sChange/BioTime/hierarchical/final_results/stan_utility.R')
-
-# #------------load rlm_clm_txa models------------
-# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/results/Jne_norm_RCTSRfyID-4950830.Rdata')
-# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jne_norm_RCTSRfyID_slice1.Rdata')
-# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jne_norm_RCTSR_slice2.Rdata')
-# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jne_norm_RCTSRfyID_slice3.Rdata')
-# 
-# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/results/Jtu_norm_RCTSRfyID_count-4950828.Rdata')
-# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jtu_norm_RCTSRfyID_slice1-4951060.Rdata')
-# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jtu_norm_RCTSRfyID_slice2.Rdata')
-# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jtu_norm_RCTSRfyID_slice3.Rdata')
-#  
-# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/results/S_pois_RCTRfyID_count-4436001.Rdata')
-# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/S_pois_RCTSR_slice1.Rdata')
-# load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/S_pois_RCTSRfyID_slice2-4950841.Rdata')
[…]

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/01_getCoefs.R

@@ -1,348 +1,0 @@
-## get model coefs new nestedness and turnover model fit to count data
-##------data and libraries-----------------------
-rm(list=ls())
-
-library(brms)
-library(dplyr)
-#library(broom)
-library(tidyr)
-library(tibble)
-
-#----load data
-##----count data only, whole time period--------
-load('~/Desktop/revision_models/Jtu_norm_BTSRfyID_count.Rdata')
-load('~/Desktop/revision_models/Jne_norm_BTSRfyID_count-4950633.Rdata')
-load('~/Desktop/revision_models/S_pois_BTSRfyID_count-5050747.Rdata')
-
-
-##--species richness coefficients and random effects---------
-S_pois_new_ALL_global <- fixef(S_pois_BTSRfyID_countData, robust = TRUE, probs = c(0.05, 0.95)) %>%
-  as_tibble() %>%
-  mutate(
-    model = 'S_pois_new',
-    time_period = 'ALL')
-
-names(S_pois_new_ALL_global)[3:4] <- c('lower', 'upper')
-
-S_pois_new_ALL_global <- S_pois_new_ALL_global %>%
-  mutate(
-    change = ifelse(lower > 0, 'up',
-                    ifelse(upper < 0, 'down', 'neutral'))) 
-
-
-
-S_pois_new_ALL_groupCoefs <- coef(S_pois_BTSRfyID_countData, robust = TRUE, probs = c(0.05, 0.95)) 
-
-##  break into biome and rarefyID estimates
-S_pois_ALL_biome <- as_tibble(S_pois_new_ALL_groupCoefs[[1]]) %>%
-  mutate(Biome = rownames(S_pois_new_ALL_groupCoefs[[1]]))
-names(S_pois_ALL_biome)[c(3:4, 7:8)] <- c('lower_intercept', 'upper_intercept', 'lower_slope', 'upper_slope') 
-S_pois_newALL_biome <- S_pois_ALL_biome %>%
-  mutate(
-    model = 'S_pois_new',
-    time_period = 'ALL',
-    change = ifelse(lower_slope > 0, 'up',
[…]

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/02a_bt_coef_wrangle.R

@@ -1,162 +1,0 @@
-# coef wrangle for new models
-rm(list = ls())
-library(tidyverse)
-library(brms)
-
-load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/bt/model_fits_output/BT_model_coef_ranef.Rdata')
-
-
-##----- get raw data and process for metadata--------
-load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')  
-
-cell_count <- rarefied_medians %>%
-  group_by(Biome, taxa_mod) %>%
-  dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-  ungroup() 
-
-biome_cell_count <- rarefied_medians %>%
-  group_by(Biome) %>%
-  dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-  ungroup() 
-
-##	rejoin
-rarefied_medians <- left_join(cell_count, rarefied_medians, by=c('Biome', 'taxa_mod'))
-
-##	filter to count data and biome/taxa combinations with >3 cells
-# these are the data that the model was fit to
-rarefied_medians <- rarefied_medians %>%
-  filter(BROAD_TYPE=='count' & n_cells > 3) %>%
-  # simplify realm covariate for plotting
-  mutate(Realm2 = ifelse(REALM!='Marine', 'Terrestrial/Freshwater', 'Marine'))
-
-##-------wrangle global level estimates-----------
-bt_new_global <- BT_global_estimates %>% 
-  filter(model=='Jne_new_norm' | model=='Jtu_new_norm' | model=='S_pois_new') 
-
-bt_new_global$model_clean <- bt_new_global$model
-bt_new_global$model_clean <- factor(bt_new_global$model_clean, 
-                                          levels = c('S_pois_new', 'Jtu_new_norm', 'Jne_new_norm'),
-                                          labels = c('Species richness', 'Turnover component', 'Nestedness component'))
-
-##-------wrangle biome level coef and ranef estimates-----------
-bt_new_biome <- BT_biome_estimate %>% 
-  filter(model=='Jne_new_norm' | model=='Jtu_new_norm' | model=='S_pois_new') %>% 
-  inner_join(rarefied_medians %>% distinct(REALM, Biome),
-             by = 'Biome')
-
-
-bt_new_biome_ranef <- BT_biome_ranef %>% 
[…]

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/02b_studyLevel_coef_wrangle.R

@@ -1,170 +1,0 @@
-# rm(list=ls())
-## plot to study level estimates as a function of latitude
-library(tidyverse)
-library(brms)
-load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/bt/model_fits_output/BT_model_coef_ranef.Rdata')
-
-#--------rarefied_medians.Rdata for meta data not in model object
-load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
-
-rarefied_medians <- rarefied_medians %>%
-  group_by(Biome, taxa_mod) %>%
-  mutate(n_cells = n_distinct(rarefyID)) %>%
-  ungroup() %>%
-  filter(n_cells > 3)
-
-# want the centre_lat and studyID to join with study-level estimates
-BT_study_estimate <- BT_studyID_estimate %>% 
-  mutate(STUDY_ID2 = str_sub(level, start = -3),
-         # drop underscores
-         STUDY_ID = sub('_','', STUDY_ID2),
-         # remove study
-         bt = str_remove(level, STUDY_ID2),
-         # remove trailing underscore
-         bt = sub('_$', '', bt)) %>% 
-  # ok, now use existing hacky code to separate biome and taxa
-  separate(bt, c('b1', 'b2', 'b3', 'b4', 'b5', 'b6', 'b7', 't'), remove=F, sep='_') %>% 
-  mutate(taxa_mod = ifelse(!is.na(t), t,
-                           ifelse(!is.na(b7), b7,
-                                  ifelse(!is.na(b6), b6,
-                                         ifelse(!is.na(b5), b5,
-                                                ifelse(!is.na(b4), b4,
-                                                       ifelse(!is.na(b3), b3, b2)))))),
-         b7 = ifelse(b7==taxa_mod, '', b7),                                           
-         Biome = ifelse(is.na(b3), b1,
-                        ifelse(is.na(b4), paste(b1, b2, sep='_'),
-                               ifelse(is.na(b5), paste(b1, b2, b3, sep='_'),
-                                      ifelse(is.na(b6), paste(b1, b2, b3, b4, sep='_'),
-                                             ifelse(is.na(b7), paste(b1, b2, b3, b4, b5, sep='_'),
[…]

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/02c_bt_taxa_to_singleLocations.R

@@ -1,355 +1,0 @@
-# code to get geography for plotting study estimates: get one or more points for biome-taxa-study combinations
-source('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/bt/code/02b_studyLevel_coef_wrangle.R')
-world <- map_data('world')
-
-# function to find the convex hull of a polygon
-find_hull <- function(df) df[chull(df$rarefyID_x, df$rarefyID_y), ]
-# function to find centre of hull
-hull_centre <- function(rarefyID_x, rarefyID_y){
-  hc_x = geosphere::geomean(cbind(x = rarefyID_x, y = rarefyID_y))[1]
-  hc_y = geosphere::geomean(cbind(x = rarefyID_x, y = rarefyID_y))[2]
-  return(cbind.data.frame(hc_x, hc_y))
-}  
-
-##  get unique coords for each biome, taxa combination
-studies_ <- BT_study_estimate %>% 
-  distinct(STUDY_ID)
-
-coords <- rarefied_medians %>% 
-  distinct(REALM, Biome, taxa_mod, STUDY_ID, rarefyID, rarefyID_x, rarefyID_y) %>% 
-  # filter to the studies for which we have study-level slope estimates
-  filter(STUDY_ID %in% studies_$STUDY_ID)
-
-
-##--------marine realm wrangle------------------
-marine_hulls <- plyr::ddply(filter(coords, REALM=='Marine'),
-                     plyr::.(REALM, Biome, taxa_mod, STUDY_ID), find_hull) %>%
-  as_tibble()
-
-
-marine_nest <- marine_hulls %>%
-  group_by(REALM, Biome, taxa_mod, STUDY_ID) %>%
-  filter(n_distinct(rarefyID) > 1) %>%
-  nest(rarefyID_x, rarefyID_y) %>% 
-  mutate(hull_centre = purrr::map(data, ~ hull_centre(.x$rarefyID_x, .x$rarefyID_y)))
-  
-
-mar_hull_cent <- marine_nest %>% unnest(hull_centre)
-
-# we want a point for each unique combination of biome/taxa/study: ok
-mar_hull_cent %>% 
-  group_by(Biome, taxa_mod) %>% 
-  summarise(n.study = n_distinct(STUDY_ID), # 
-            points = n_distinct(hc_x, hc_y)) %>% 
-  filter(points!=n.study)
-
-# and we need points for the studies that only had one location 
[…]

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/02d_bt_biome_map_wrangle.R

@@ -1,240 +1,0 @@
-# read terrestrial shape file
-
-# code modified 080219: map results from new models (S, Jtu, Jne)
-
-# load new results and raw data
-load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/bt/model_fits_output/BT_model_coef_ranef.Rdata')
-load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
-
-
-# load biomes---------------------
-library(tidyverse)
-world <- map_data('world')
-
-
-library(sf)
-# library(tidyverse) 
- 
-mar_shp <- sf::st_read('~/Dropbox/BiogeoBioTIME/biomes/marineInUSe.shp', quiet = T) %>% select(-lat, -long)
-terr_shp <- sf::st_read('~/Dropbox/BiogeoBioTIME/biomes/terrInUSe.shp', quiet = T)  %>% select(-lat, -long)
-fresh_shp <- sf::st_read('~/Dropbox/BiogeoBioTIME/biomes/freshinUse.shp', quiet = T)  %>% select(-lat, -long)
-
-# create spatialdataframes (sp package) for use with geom_polygon
-marine_spdf <- as(mar_shp, 'Spatial')
-terr_spdf <- as(terr_shp, 'Spatial')
-fresh_spdf <- as(fresh_shp, 'Spatial')
-
-
-#-------process marine shp file: add slope ranef and flatten for ggplot---------
-mar_biomes <- rarefied_medians %>% 
-  filter(REALM=='Marine') %>%
-  distinct(Biome) %>%
-  mutate(Biome = gsub("_", " ", Biome))
-
-# filter spatialdataframe to these biomes only
-marine_spdf <- marine_spdf[which(marine_spdf$Biome %in% mar_biomes$Biome),]
-
-# put the slope ranef in after fixing names
-marine_biome_re <- bind_cols(BT_biome_ranef %>% 
-                               filter(model=='S_pois_new') %>% 
-                               mutate(S_biome_re = Slope_ranef) %>%
-                               select(S_biome_re, Biome),
-                             BT_biome_ranef %>% 
-                               filter(model=='Jtu_new_norm') %>% 
-                               mutate(Jtu_biome_re = Slope_ranef) %>%
-                               select(Jtu_biome_re),
-                             BT_biome_ranef %>% 
-                               filter(model=='Jne_new_norm') %>%
-                               mutate(Jne_biome_re = Slope_ranef) %>%
[…]

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/03a_bt_plot_global_estimates.R

@@ -1,133 +1,0 @@
-# update code for new models
-library(brms)
-library(tidyverse)
-
-setwd('~/Dropbox/1current/BioTime/local_code/hierarchical/6results/bt/figs/')
-
-# load model fits to extract posterior samples
-load('~/Desktop/revision_models/Jtu_norm_BTSRfyID_count.Rdata')
-load('~/Desktop/revision_models/S_pois_BTSRfyID_expPrior.Rdata')
-load('~/Desktop/revision_models/Jne_norm_BTSRfyID_count-4950633.Rdata')
-
-# want coefficient plot of the posterior distribution of the global slope
-jne_post <- posterior_samples(Jne_norm_BTSRfyID, pars = 'b_cYEAR', subset = 2000:3000)
-
-jne_summary <- jne_post %>%
-  summarise(
-    median = median(b_cYEAR),
-    lower50 = quantile(b_cYEAR, probs = 0.25),
-    upper50 = quantile(b_cYEAR, probs = 0.75),
-    lower90 = quantile(b_cYEAR, probs = 0.05),
-    upper90 = quantile(b_cYEAR, probs = 0.95))
-
-jtu_post <- posterior_samples(Jtu_norm_BTSRfyID, pars = 'b_cYEAR', subset = 2000:3000)
-
-jtu_summary <- jtu_post %>%
-  summarise(
-    median = median(b_cYEAR),
-    lower50 = quantile(b_cYEAR, probs = 0.25),
-    upper50 = quantile(b_cYEAR, probs = 0.75),
-    lower90 = quantile(b_cYEAR, probs = 0.05),
-    upper90 = quantile(b_cYEAR, probs = 0.95))
-
-s_post <- posterior_samples(S_pois_BTSRfyID_countData, pars = 'b_cYEAR', subset = 2000:3000)
-
-s_summary <- s_post %>%
-  summarise(
-    median = median(b_cYEAR),
-    lower50 = quantile(b_cYEAR, probs = 0.25),
-    upper50 = quantile(b_cYEAR, probs = 0.75),
-    lower90 = quantile(b_cYEAR, probs = 0.05),
-    upper90 = quantile(b_cYEAR, probs = 0.95),
-    lower95 = quantile(b_cYEAR, probs = 0.025),
-    upper95 = quantile(b_cYEAR, probs = 0.975))
-
-# tidy up (the model objects are big)
[…]

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/03b_bt_biome_map_plot.R

@@ -1,323 +1,0 @@
-# code to plot biome maps
-
-# run code to plot posterior distribution of global estimates for inset
-source('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/bt/code/03a_bt_plot_global_estimates.R')
-# run code to wrangle the data for mapping
-source('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/bt/code/02d_bt_biome_map_wrangle.R')
-
-#----------plot Jtu biome random effects and taxa estimates------------------
-# label for the legend
-Jtu_leg_label <- c(Jtu_re_quantile[1:4], 0, Jtu_re_quantile[5:6]) %>% signif(1) %>% as.numeric() %>% format(scientific = T)
-# alt legend label with the numbers representing biome estimates (not random effects)
-Jtu_estimate_label <- as.numeric(signif(c(Jtu_re_quantile[1:4], 0, Jtu_re_quantile[5:6]) + 
-                                          BT_global_estimates %>% filter(model=='Jtu_new_norm' & time_period=='ALL') %>% 
-                                          filter(term=='Slope') %>% .$Estimate, 2) )
-
-combined_label <- paste0(paste(Jtu_leg_label, Jtu_estimate_label, sep = ' ('),')') 
-
-mar_biome_plot_Jtu <-  ggplot() +
-  geom_polygon(data = filter(alldf, Realm2 == 'Marine'),
-               aes(x = long, y = lat, group = group, linetype = NA, fill = Jtu_re), 
-               alpha = 0.85) +#colour = 'black', 
-  geom_polygon(data=world, aes(long, lat, group = group), colour=NA, fill='#7f7f7f', size=0) +
-  geom_point(data = filter(coords2, REALM == 'Marine'),
-             aes(x = rarefyID_x, y = rarefyID_y, colour = Jtu_re), size = 0.5, stroke = 1.2) +
-  scale_fill_manual(name = '', values = c('1' = '#b2182b', '2' = '#d6604d', '3' = '#f4a582',
-                                          '4' = '#fddbc7', '5' = '#4393c3', '6' = '#2166ac'), 
-                    breaks = c('1', '2', '3', '4', '5', '6'),
-                    labels = c(combined_label[-1]), drop = FALSE) +
-  scale_colour_manual(guide = FALSE, values = c('1' = '#b2182b', '2' = '#d6604d', '3' = '#f4a582',
[…]

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/03c_bt_alpha_beta_plots.R

@@ -1,285 +1,0 @@
-# code to examine relationships between alpha and beta-diversity change at the study level (new model)
-
-#--Taxa-level plots------
-study_corr <- study_corr %>%
-  mutate(realm2 = ifelse(REALM=='Marine', REALM, 'Terrestrial and freshwater'),
-         realm_label = ifelse(REALM=='A.   Marine', REALM, 'B.   Terrestrial and freshwater'))
-
-# rename group of taxa labelled 'All' to 'Multiple'
-study_corr$taxa_mod2 <- factor(study_corr$taxa_mod, 
-                              levels = c("Fish", "Benthos", "Birds", "Invertebrates", "Plant", "Amphibians",
-                                         "All", "Marine invertebrates/plants", "Mammals"),
-                              labels = c("Fish", "Benthos", "Birds", "Invertebrates", "Plant", "Amphibians",
-                                         "Multiple taxa", "Marine invertebrates/plants", "Mammals"))
-
-# set shapes for plotting groups of taxa
-shapes_mod2 = c('Invertebrates' = 0, 'Fish' = 1, 'Benthos' = 2, 'Birds' = 15, 'Mammals' = 17,
-                'Plant' = 5, 'Multiple taxa' = 6, 'Marine invertebrates/plants' = 7, 'Amphibians' = 8)
-
-# shapes for realm
-shapes_realm <- c('Marine' = 15, 'Terrestrial' = 16, 'Freshwater' = 17)
-
-# colours for each taxa
-taxa_col = c('Multiple taxa' = '#e6f598',
-             'Amphibians' = '#fee08b',
-             'Benthos' = '#5e4fa2',
-             'Birds' = '#f46d43',
-             'Fish' = '#3288bd',
-             'Invertebrates' = '#abdda4',
-             'Mammals' = '#9e0142',
-             'Marine invertebrates/plants' = '#66c2a5',
-             'Plant' = '#fdae61')
-
-# colours for conceptual model relating changes in alpha & beta diversity
-# blue green (colour blind friendly - from colorbrewer2.org) l.blue, d.blue, l.green, d.green
-# quad_col = c('c1' = '#a6cee3', 'c2' = '#1f78b4', 'c3' = '#b2df8a', 'c4' = '#33a02c')
[…]

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/04_bt_sensitivity_plots.R

@@ -1,207 +1,0 @@
-##------------sensitivity plots for the BT models---------------------
-# update to plot results of new models (include studyID in hierarchy)
-rm(list=ls())
-
-library(brms)
-library(tidyverse)
-
-load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/bt/model_fits_output/BT_model_coef_ranef.Rdata')
-
-##----the data models were fit too
-load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
-
-##	create new covariate that is concatenation of realm_climate_taxa
-##	count cells within realm/climate/taxa hierarchy
-cell_count <- rarefied_medians %>%
-  group_by(Biome, taxa_mod) %>%
-  dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-  ungroup() 
-
-##	rejoin
-rarefied_medians <- left_join(cell_count, rarefied_medians, 
-                              by=c('Biome', 'taxa_mod'))
-
-##	filter to  count data with >3 cells
-rarefied_medians <- rarefied_medians %>%
-  filter(n_cells > 3 & BROAD_TYPE=='count')
-
-## get duration, num_years and other metadata for each rarefyID
-cell_meta <- rarefied_medians %>%
-  distinct(REALM, climate_mod, taxa_mod, STUDY_ID, rarefyID, 
-           rarefyID_x, rarefyID_y, num_years, duration, 
-           SamplePool, SampleN, startYear, endYear)
-
-initialS <- rarefied_medians %>% 
-  distinct(rarefyID, .keep_all=TRUE) %>%
-  filter(YEAR == startYear) %>%
-  mutate(initialS = S) %>%
-  select(rarefyID, initialS)
-
-cell_meta <- inner_join(initialS, 
-                        cell_meta,
-                        by = 'rarefyID')
-
-# clean up the rarefyID estimates
-bt_rarefyID_coef <- BTRfyID_rarefyID_coef %>%
-  filter(model=='Jtu_new_norm' | model=='Jne_new_norm' | model=='S_pois_new') %>% 
-  separate(level, c('b1', 'b2', 'b3', 'b4', 'b5', 'b6', 'b7', 'b8', 'b9', 't'), remove=F, sep='_') %>% 
-  mutate(cell = ifelse(!is.na(t), t,
-                       ifelse(!is.na(b9), b9,
-                              ifelse(!is.na(b8), b8,
-                                     ifelse(!is.na(b7), b7,
[…]

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/05_bt_beta_vs_norm.R

@@ -1,321 +1,0 @@
-## compare the coefficients of jtu_norm and jtu_beta
-rm(list=ls())
-library(tidyverse)
-library(ggridges)
-load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/bt/model_fits_output/BT_model_coef_ranef.Rdata')
-
-# get the z1i Jtu model too...
-load('~/Desktop/revision_models/Jtu_z1i_realm_BTSRfyID_count-4946085.Rdata')
-
-# posterior distributions of the coi parameter for supplement
-coi_post <- posterior_samples(Jtu_z1i_realm_BTSRfyID_countData,
-                              pars = c('^b_coi_', '^b_zoi_')) %>% 
-  as_tibble() 
-
-coi_post2 <- tibble(
-  Realm = rep(c('Freshwater', 'Marine', 'Terrestrial'), each = 4000),
-  coi = c(plogis(coi_post$b_coi_Intercept), 
-          plogis(coi_post$b_coi_Intercept + coi_post$b_coi_REALMMarine),
-          plogis(coi_post$b_coi_Intercept + coi_post$b_coi_REALMTerrestrial)))
-
-ggplot() +
-  geom_density_ridges(data = coi_post2,
-                      aes(x = coi, 
-                          y = Realm),
-                      linetype = 0,
-                      scale = 0.95, alpha = 0.6)
-
-jtu_z1i_fixef <- fixef(Jtu_z1i_realm_BTSRfyID_countData, robust = T, probs = c(0.05, 0.95)) %>% 
-  as_tibble() %>% 
-  mutate(parname = rownames(fixef(Jtu_z1i_realm_BTSRfyID_countData)))
-
-coi <- tibble(Realm = c('Freshwater', 'Marine', 'Terrestrial'),
-              # pr one = zoi * coi
-              pr_one = c((jtu_z1i_fixef %>% filter(parname=='coi_Intercept') %>% .$Estimate *
-                jtu_z1i_fixef %>% filter(parname=='zoi_Intercept') %>% .$Estimate) %>% plogis,
-                ((jtu_z1i_fixef %>% filter(parname=='coi_Intercept' |parname=='coi_REALMMarine') %>% .$Estimate %>% sum) *
-                   (jtu_z1i_fixef %>% filter(parname=='zoi_Intercept' |parname=='zoi_REALMMarine') %>% .$Estimate %>% sum)) %>% plogis,
[…]

### 2021-09-16 analysis/other_Blowes_analysis/bt/code/06_bt_rct_new_models_coefCompare.R

@@ -1,213 +1,0 @@
-# want to compare the slope estimates for studies and rarefyID's from
-# the bt versus the rct models
-
-rm(list=ls())
-
-library(tidyverse)
-
-load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/bt/model_fits_output/BT_model_coef_ranef.Rdata')
-load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/rct/model_fits_output/rlm_clm_txa_modelCoefs.Rdata')
-
-#--------rarefied_medians.Rdata for meta data not in model object
-load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
-
-rarefied_medians <- rarefied_medians %>%
-  group_by(Biome, taxa_mod) %>%
-  mutate(n_cells = n_distinct(rarefyID)) %>%
-  ungroup() %>%
-  filter(n_cells > 3)
-
-realm_climate_biome <- rarefied_medians %>%
-  distinct(REALM, climate_mod, Biome) %>%
-  ungroup()
-
-# want unique climate (latitudinal bands for each biome)
-# this is to compare like with like: some of the rct study estimates span multiple latitudinal bands
-realm_climate_biome <- realm_climate_biome %>%
-  # filter(Biome %in% biome_levels_model$Biome) %>%
-  mutate(climate_mod = ifelse(Biome=='Arctic', 'Polar', 
-                              ifelse(Biome=='Boreal_Forests_Taiga', 'Temperate',
-                                     ifelse(Biome=='Cold_Temperate_Northeast_Pacific', 'Temperate',
-                                            ifelse(Biome=='Continental_High_Antarctic', 'Polar',
-                                                   ifelse(Biome=='Hawaii', 'Tropical', 
-                                                          ifelse(Biome=='Northern_European_Seas', 'Temperate',
-                                                                 ifelse(Biome=='Subantarctic_Islands', 'Polar',
-                                                                        ifelse(Biome=='Tropical_and_Subtropical_Grasslands_Savannas_and_Shrublands', 'Temperate',
-                                                                               ifelse(Biome=='Tropical_and_Subtropical_Moist_Broadleaf_Forests', 'Tropical', 
[…]

### 2021-09-16 analysis/other_Blowes_analysis/rlt/code/00_HLM_Jne_rlt.R

@@ -1,56 +1,0 @@
-##======================================================================
-##  Hierarchical models: fit to either all studies (S, Jaccard), or only those with
-##	count data (N, PIE, etc)
-##======================================================================
-rm(list=ls())
-
-##	
-#withr::with_libpaths(new = '/home/blowes/R/x86_64-pc-linux-gnu-library/3.3/brms', library(brms))
-library(brms)
-library(dplyr)
-library(tidyr)
-##======================================================================
-##	alternatively load averaged metrics
-load('/gpfs1/data/sChange/BioTime/data/rarefied_medians.Rdata')
-
-##======================================================================		
-##	remove rarefyID's with only one year
-one_year_only <- rarefied_medians %>%
-        group_by(rarefyID) %>%
-        dplyr::summarise(duration = max(YEAR) - min(YEAR)) %>%
-        filter(duration <= 1)
-
-rarefied_medians <- rarefied_medians %>% filter(!(rarefyID %in% one_year_only$rarefyID))
-
-##	create new covariate that is concatenation of realm_climate_taxa
-rarefied_medians <- rarefied_medians %>%
-	unite(rlm_clm_txa, REALM, climate_mod, taxa_mod, remove=FALSE)
-
-##	count cells within realm/climate/taxa hierarchy
-cell_count <- rarefied_medians %>%
-	group_by(rlm_clm_txa) %>%
-	dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-	ungroup() 
-
-##	rejoin
-rarefied_medians <- left_join(cell_count, rarefied_medians, by='rlm_clm_txa')  # cell_count_alt[,c(1,5)]
-
-##	filter to ecoregions with >3 cells
-rarefied_medians <- rarefied_medians %>%
-	filter(n_cells > 3 & BROAD_TYPE=='count')
-	
-##	set some weakly regularising priors...
-hier_prior <- c(set_prior(prior = 'normal(0,1)', class='b', coef='cYEAR'), 	# global slope
-	set_prior(prior = 'normal(0,2)', class='Intercept', coef=''), 		# global intercept
-	set_prior(prior = 'cauchy(0,2.5)', class='sd'),							# group-level intercepts and slopes
-	set_prior(prior = 'lkj(2)', class='cor'))
-
[…]

### 2021-09-16 analysis/other_Blowes_analysis/rlt/code/00_HLM_Jtu_rlt.R

@@ -1,56 +1,0 @@
-##======================================================================
-##  Hierarchical models: fit to either all studies (S, Jaccard), or only those with
-##	count data (N, PIE, etc)
-##======================================================================
-rm(list=ls())
-
-##	
-#withr::with_libpaths(new = '/home/blowes/R/x86_64-pc-linux-gnu-library/3.3/brms', library(brms))
-library(brms)
-library(dplyr)
-library(tidyr)
-##======================================================================
-##	alternatively load averaged metrics
-load('/gpfs1/data/sChange/BioTime/data/rarefied_medians.Rdata')
-
-##======================================================================		
-##	remove rarefyID's with only one year
-one_year_only <- rarefied_medians %>%
-        group_by(rarefyID) %>%
-        dplyr::summarise(duration = max(YEAR) - min(YEAR)) %>%
-        filter(duration <= 1)
-
-rarefied_medians <- rarefied_medians %>% filter(!(rarefyID %in% one_year_only$rarefyID))
-
-##	create new covariate that is concatenation of realm_climate_taxa
-rarefied_medians <- rarefied_medians %>%
-	unite(rlm_clm_txa, REALM, climate_mod, taxa_mod, remove=FALSE)
-
-##	count cells within realm/climate/taxa hierarchy
-cell_count <- rarefied_medians %>%
-	group_by(rlm_clm_txa) %>%
-	dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-	ungroup() 
-
-##	rejoin
-rarefied_medians <- left_join(cell_count, rarefied_medians, by='rlm_clm_txa')  # cell_count_alt[,c(1,5)]
-
-##	filter to ecoregions with >3 cells
-rarefied_medians <- rarefied_medians %>%
-	filter(n_cells > 3 & BROAD_TYPE=='count')
-	
-##	set some weakly regularising priors...
-hier_prior <- c(set_prior(prior = 'normal(0,1)', class='b', coef='cYEAR'), 	# global slope
-	set_prior(prior = 'normal(0,2)', class='Intercept', coef=''), 		# global intercept
-	set_prior(prior = 'cauchy(0,2.5)', class='sd'),							# group-level intercepts and slopes
-	set_prior(prior = 'lkj(2)', class='cor'))
-
[…]

### 2021-09-16 analysis/other_Blowes_analysis/rlt/code/00_HLM_S_rlt.R

@@ -1,64 +1,0 @@
-##======================================================================
-##  Hierarchical models: fit to either all studies (S, Jaccard), or only those with
-##	count data (N, PIE, etc)
-##======================================================================
-rm(list=ls())
-
-##	
-#withr::with_libpaths(new = '/home/blowes/R/x86_64-pc-linux-gnu-library/3.3/brms', library(brms))
-library(brms)
-library(dplyr)
-library(tidyr)
-##======================================================================
-##	alternatively load averaged metrics
-load('/gpfs1/data/sChange/BioTime/data/rarefied_medians.Rdata')
-
-##======================================================================		
-##	remove rarefyID's with only one year
-one_year_only <- rarefied_medians %>%
-        group_by(rarefyID) %>%
-        dplyr::summarise(duration = max(YEAR) - min(YEAR)) %>%
-        filter(duration <= 1)
-
-rarefied_medians <- rarefied_medians %>% filter(!(rarefyID %in% one_year_only$rarefyID))
-
-##	create new covariate that is concatenation of realm_climate_taxa
-rarefied_medians <- rarefied_medians %>%
-	unite(rlm_clm_txa, REALM, climate_mod, taxa_mod, remove=FALSE)
-
-##	count cells within realm/climate/taxa hierarchy
-cell_count <- rarefied_medians %>%
-	group_by(rlm_clm_txa) %>%
-	dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-	ungroup() 
-
-#cell_count_alt <- rarefied_medians %>%
-#	group_by(REALM, climate_mod, taxa_mod) %>%
-#	dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-#	ungroup() %>%
-#	unite(rlm_clm_txa, REALM, climate_mod, taxa_mod, remove=FALSE)
-
-##	rejoin
-rarefied_medians <- left_join(cell_count, rarefied_medians, by='rlm_clm_txa')  # cell_count_alt[,c(1,5)]
-
-##	filter to ecoregions with >3 cells
-rarefied_medians <- rarefied_medians %>%
-	filter(n_cells > 3 & BROAD_TYPE=='count')
-	
-##	set some weakly regularising priors...
[…]

### 2021-09-16 analysis/other_Blowes_analysis/rlt/code/00a_nuts_diagnostics.R

@@ -1,56 +1,0 @@
-# code to check nuts (no u-turn sampler) diagnostics for all models
-# code to run remotely and save results to file
-rm(list=ls())
-
-library(brms)
-library(rstan)
-library(dplyr)
-library(tibble)
-
-# source nuts diagnostics file (from Michael Betancourt)
-source('/gpfs1/data/sChange/BioTime/hierarchical/final_results/stan_utility.R')
-
-#------------load rlm_clm_txa models------------
-load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/results/Jne_norm_RCTSRfyID-4950830.Rdata')
-load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jne_norm_RCTSRfyID_slice1.Rdata')
-load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jne_norm_RCTSR_slice2.Rdata')
-load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jne_norm_RCTSRfyID_slice3.Rdata')
-
-load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/results/Jtu_norm_RCTSRfyID_count-4950828.Rdata')
-load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jtu_norm_RCTSRfyID_slice1-4951060.Rdata')
-load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jtu_norm_RCTSRfyID_slice2.Rdata')
-load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/Jtu_norm_RCTSRfyID_slice3.Rdata')
-
-load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/results/S_pois_RCTRfyID_count-4436001.Rdata')
-load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/S_pois_RCTSR_slice1.Rdata')
-load('/gpfs1/data/sChange/BioTime/hierarchical/rlm_clm_txa/timeSlice/results/S_pois_RCTSRfyID_slice2-4950841.Rdata')
[…]

### 2021-09-16 analysis/other_Blowes_analysis/rlt/code/03_rlt_posterior_density_plots.R

@@ -1,531 +1,0 @@
-# plot posterior densities of the rct study level estimates (at the realm-climate level)
-rm(list=ls())
-# setwd('~/Desktop/revision_code/')
-# rm(list=ls())
-library(brms) 
-library(tidyverse)
-library(ggridges)
-#------posterior samples from biome and biome-taxa level----------
-load('~/Desktop/revision_models/S_pois_RCTSRfyID_count-4950831.Rdata')
-load('~/Desktop/revision_models/Jtu_norm_RCTSRfyID_count.Rdata')
-
-# first get posterior draws for overall trend
-rct_global_posterior <- tibble(
-  S_global = posterior_samples(S_pois_RCTSRfyID,
-                               pars = 'b_cYEAR',
-                               exact_match = TRUE,
-                               subset = floor(runif(n = 1000,
-                                                    min = 1, max = 2000))) %>% unlist() %>% as.numeric(),
-  Jtu_global = posterior_samples(Jtu_norm_RCTSRfyID,
-                                 pars = 'b_cYEAR',
-                                 exact_match = TRUE,
-                                 subset = floor(runif(n = 1000,
-                                                      min = 1, max = 2000))) %>% unlist() %>% as.numeric())
-
-rct_levels <- S_pois_RCTSRfyID$data %>% 
-  as_tibble() %>% 
-  distinct(rlm_clm_txa) %>% 
-  mutate(level = rlm_clm_txa) %>% 
-  nest(level)
-
-study_levels <- S_pois_RCTSRfyID$data %>% 
-  as_tibble() %>% 
-  distinct(`rlm_clm_txa:STUDY_ID`) %>% 
-  mutate(level = `rlm_clm_txa:STUDY_ID`) %>% 
-  nest(level)
-
-rct_sample_posterior <- rct_levels %>%
-  mutate(S_posteriorSamp = purrr::map(data, ~posterior_samples(S_pois_RCTSRfyID, 
-                                                               pars = paste('r_rlm_clm_txa[', as.character(.x$level), ',cYEAR]', sep=''),
-                                                               exact = TRUE,
-                                                               subset = floor(runif(n = 1000,
-                                                                                    min = 1, max = 2000))) %>% unlist() %>% as.numeric()),
-         Jtu_posterior = purrr::map(data, ~posterior_samples(Jtu_norm_RCTSRfyID,
-                                                             pars = paste('r_rlm_clm_txa[', as.character(.x$level), ',cYEAR]', sep =''),
-                                                             exact = TRUE,
[…]

### 2021-09-16 analysis/other_Blowes_analysis/rlt/code/04_rlt_coef_plots.R

@@ -1,318 +1,0 @@
-##------results of realm/climate/taxa hierarchical NEW MODELS-----------------------
-# code modified 210219
-##------data and libraries-----------------------
-rm(list=ls())
-
-library(brms)
-library(tidyverse)
-
-load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/rlt/model_fits_output/rlm_clm_txa_modelCoefs.Rdata')
-
-##---------add the number of time series used to estimate each slope-------------
-##----the data models were fit too
-load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
-
-##	create new covariate that is concatenation of realm_climate_taxa
-rarefied_medians <- rarefied_medians %>%
-  unite(rlm_clm_txa, REALM, climate_mod, taxa_mod, remove=FALSE)
-
-##	count cells within realm/climate/taxa hierarchy
-cell_count <- rarefied_medians %>%
-  group_by(rlm_clm_txa) %>%
-  dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-  ungroup() 
-
-##	rejoin
-rarefied_medians <- left_join(cell_count, rarefied_medians, by='rlm_clm_txa')
-
-##	filter to ecoregions with >3 cells
-rarefied_medians <- rarefied_medians %>%
-  filter(n_cells > 3 & BROAD_TYPE=='count')
-
-##  redo cell counts (this will be for the whole data set combined)
-cell_count3_ALL <- rarefied_medians %>%
-  group_by(rlm_clm_txa) %>%
-  dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-  ungroup() %>%
-  mutate(time_period='ALL')
-
-cell_count3_STUDY <- rarefied_medians %>%
-  group_by(rlm_clm_txa, STUDY_ID) %>%
-  dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-  ungroup() %>%
-  mutate(time_period='ALL')
-
-##  create the time sliced data
-# load('/gpfs1/data/sChange/BioTime/data/rarefied_medians.Rdata')
-# rarefied_medians <- rarefied_medians %>%
-#   unite(rlm_clm_txa, REALM, climate_mod, taxa_mod, remove=FALSE)
-ts1 <- filter(rarefied_medians, YEAR >= 1951 & YEAR <= 1970)
[…]

### 2021-09-16 analysis/other_Blowes_analysis/rlt/code/05_rlt_sensitivity_plots.R

@@ -1,251 +1,0 @@
-# code modified 120219: new models
-##------data and libraries-----------------------
-rm(list=ls())
-
-library(brms)
-library(tidyverse)
-
-load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/rct/model_fits_output/rlm_clm_txa_modelCoefs.Rdata')
-
-##---------add the number of time series used to estimate each slope-------------
-##----the data models were fit too
-load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
-
-##	create new covariate that is concatenation of realm_climate_taxa
-rarefied_medians <- rarefied_medians %>%
-  unite(rlm_clm_txa, REALM, climate_mod, taxa_mod, remove=FALSE)
-
-## get duration, num_years and other metadata for each rarefyID
-cell_meta <- rarefied_medians %>%
-  distinct(REALM, climate_mod, taxa_mod, rarefyID, rarefyID_x, rarefyID_y, num_years, duration, SamplePool, SampleN, startYear, endYear)
-
-initialS <- rarefied_medians %>% distinct(rarefyID, .keep_all=TRUE) %>%
-  filter(YEAR == startYear) %>%
-  mutate(initialS = S) %>%
-  select(rarefyID, initialS)
-
-cell_meta <- inner_join(cell_meta, initialS, by='rarefyID')
-
-
-## need to split level into rct and rarefyID
-rarefyID_estimate <- bind_rows(
-  # need to do this differently for the new model
-  rlm_clm_txa_rarefyID_estimate %>%
-  filter(model=='Jtu_new_norm' | model=='Jne_new_norm' | model=='S_pois_new') %>% 
-  separate(level, into = c('Realm', 'Climate', 'Taxa', 'STUDY_ID', 'b5' , 'b6'), sep = '_', remove = FALSE) %>%
-  unite(rarefyID, b5, b6))
-
-# check if we have metadata for all the cells in the model
-rarefyID_estimate %>% 
-  filter(!rarefyID %in% cell_meta$rarefyID)
-
-## combine with rarefyID estimates for plotting
-rarefyID_estimate <- right_join(cell_meta, rarefyID_estimate, by = 'rarefyID') %>%
-  mutate(change = ifelse(lower_slope > 0, 'up',
-                         ifelse(upper_slope < 0, 'down', 'neutral'))) %>%
-  unite(RealmClimate, c(Realm,Climate), remove = FALSE) 
-
[…]

### 2021-09-16 analysis/other_Blowes_analysis/rrt/00_continent_wrangle.R

@@ -1,141 +1,0 @@
-# code to create and check new regional grouping structure:
-# marine realm to combine biome_taxa
-# terrestrial and freshwater to combine continent and taxa
-library(tidyverse)
-library(sp)
-library(rworldmap)
-
-# Want to know how many studies in each group 
-load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
-
-
-# The single argument to this function, points, is a data.frame in which:
-#   - column 1 contains the longitude in degrees
-#   - column 2 contains the latitude in degrees
-coords2continent = function(points)
-{  
-  countriesSP <- getMap(resolution='low')
-  #countriesSP <- getMap(resolution='high') #you could use high res map from rworldxtra if you were concerned about detail
-  
-  # converting points to a SpatialPoints object
-  # setting CRS directly to that from rworldmap
-  pointsSP = SpatialPoints(points, proj4string=CRS(proj4string(countriesSP)))  
-  
-  
-  # use 'over' to get indices of the Polygons object containing each point 
-  indices = over(pointsSP, countriesSP)
-  
-  #indices$continent   # returns the continent (6 continent model)
-  indices$REGION   # returns the continent (7 continent model)
-  #indices$ADMIN  #returns country name
-  #indices$ISO3 # returns the ISO3 code 
-}
-
-continents <- rarefied_medians %>% 
-  distinct(REALM, Biome, STUDY_ID, rarefyID, rarefyID_x, rarefyID_y) %>% 
-  nest(rarefyID_x, rarefyID_y) %>% 
-  mutate(continent = purrr::map(data, ~coords2continent(data.frame(.x$rarefyID_x, .x$rarefyID_y))))
-
-continents <- continents %>% 
-  unnest(data, continent)
-
-world = map_data('world')
-# visual check
-ggplot() +
-  geom_polygon(data=world, aes(long, lat, group = group), colour=NA, fill='#f0f0f0', size=0) +
-  geom_point(data = filter(continents, REALM == 'Marine'),
-             aes(x = rarefyID_x, y = rarefyID_y, colour = Biome), size = 0.5, stroke = 1.2)+
-  scale_color_viridis_d(guide = F)
-
-# no problems for the marine biomes
[…]

### 2021-09-16 analysis/other_Blowes_analysis/rrt/01_continent_jne.R

@@ -1,54 +1,0 @@
-##======================================================================
-##  Hierarchical models: fit to either all studies (S, Jaccard), or only those with
-##	count data (N, PIE, etc)
-##======================================================================
-rm(list=ls())
-
-##	
-#withr::with_libpaths(new = '/home/blowes/R/x86_64-pc-linux-gnu-library/3.3/', library(brms))
-library(brms)
-library(dplyr)
-library(tidyr)
-##======================================================================
-##	alternatively load averaged metrics
-load('/gpfs1/data/sChange/BioTime/data/rarefied_medians_continents.Rdata')
-
-##======================================================================		
-##	create new covariate that is concatenation of region and taxa (with or without realm)
-rarefied_medians <- rarefied_medians %>%
-  unite(rlm_region_taxa, REALM, region, taxa_mod, remove=FALSE)
-
-# rarefied_medians <- rarefied_medians %>%
-#   unite(region_taxa, region, taxa_mod, remove=FALSE)
-
-##	count cells within realm/climate/taxa hierarchy
-cell_count <- rarefied_medians %>%
-  group_by(rlm_region_taxa) %>%
-  dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-  ungroup() 
-
-##	rejoin
-rarefied_medians <- left_join(cell_count, rarefied_medians, by='region_taxa')  
-
-##	filter to ecoregions with >3 cells
-rarefied_medians <- rarefied_medians %>%
-  filter(n_cells > 3 & BROAD_TYPE=='count')
-
-##	set some weakly regularising priors...
-hier_prior <- c(set_prior(prior = 'normal(0,1)', class='b', coef='cYEAR'), 	# global slope
-	set_prior(prior = 'normal(0,2)', class='Intercept', coef=''), 		# global intercept
-	set_prior(prior = 'cauchy(0,2)', class='sd'),							# group-level intercepts and slopes
-	set_prior(prior = 'lkj(2)', class='cor'))
-
-Jne_norm_CTSRfyID <- brm(bf(Jne_base ~ cYEAR + (cYEAR|rlm_region_taxa/STUDY_ID/rarefyID), 
-	family = brmsfamily('gaussian')),	
-	data= rarefied_medians,
-	prior=hier_prior,
-	inits = '0',
-	iter = 2000,
-	#warmup = 1500,
-	#control = list(adapt_delta=0.9),
-	cores = 4,
[…]

### 2021-09-16 analysis/other_Blowes_analysis/rrt/01_continent_jtu.R

@@ -1,54 +1,0 @@
-##======================================================================
-##  Hierarchical models: fit to either all studies (S, Jaccard), or only those with
-##	count data (N, PIE, etc)
-##======================================================================
-rm(list=ls())
-
-##	
-#withr::with_libpaths(new = '/home/blowes/R/x86_64-pc-linux-gnu-library/3.3/', library(brms))
-library(brms)
-library(dplyr)
-library(tidyr)
-##======================================================================
-##	alternatively load averaged metrics
-load('/gpfs1/data/sChange/BioTime/data/rarefied_medians_continents.Rdata')
-
-##======================================================================		
-##	create new covariate that is concatenation of region and taxa (with or without realm)
-rarefied_medians <- rarefied_medians %>%
-  unite(rlm_region_taxa, REALM, region, taxa_mod, remove=FALSE)
-
-# rarefied_medians <- rarefied_medians %>%
-#   unite(region_taxa, region, taxa_mod, remove=FALSE)
-
-##	count cells within realm/climate/taxa hierarchy
-cell_count <- rarefied_medians %>%
-  group_by(rlm_region_taxa) %>%
-  dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-  ungroup() 
-
-##	rejoin
-rarefied_medians <- left_join(cell_count, rarefied_medians, by='region_taxa')  
-
-##	filter to ecoregions with >3 cells
-rarefied_medians <- rarefied_medians %>%
-  filter(n_cells > 3 & BROAD_TYPE=='count')
-
-##	set some weakly regularising priors...
-hier_prior <- c(set_prior(prior = 'normal(0,1)', class='b', coef='cYEAR'), 	# global slope
-	set_prior(prior = 'normal(0,2)', class='Intercept', coef=''), 		# global intercept
-	set_prior(prior = 'cauchy(0,2)', class='sd'),							# group-level intercepts and slopes
-	set_prior(prior = 'lkj(2)', class='cor'))
-
-Jtu_norm_CTSRfyID <- brm(bf(Jtu_base ~ cYEAR + (cYEAR|rlm_region_taxa/STUDY_ID/rarefyID), 
-	family = brmsfamily('gaussian')),	
-	data= rarefied_medians,
-	prior=hier_prior,
-	inits = '0',
-	iter = 2000,
-	#warmup = 1500,
-	#control = list(adapt_delta=0.9),
[…]

### 2021-09-16 analysis/other_Blowes_analysis/rrt/01_continent_s.R

@@ -1,59 +1,0 @@
-##======================================================================
-##  Hierarchical models: fit to either all studies (S, Jaccard), or only those with
-##	count data (N, PIE, etc)
-##======================================================================
-rm(list=ls())
-
-##	
-#withr::with_libpaths(new = '/home/blowes/R/x86_64-pc-linux-gnu-library/3.3/', library(brms))
-library(brms)
-library(dplyr)
-library(tidyr)
-##======================================================================
-##	alternatively load averaged metrics
-load('/gpfs1/data/sChange/BioTime/data/rarefied_medians_continents.Rdata')
-
-##======================================================================		
-##	create new covariate that is concatenation of region and taxa (with or without realm)
-rarefied_medians <- rarefied_medians %>%
-  unite(rlm_region_taxa, REALM, region, taxa_mod, remove=FALSE)
-
-# rarefied_medians <- rarefied_medians %>%
-#   unite(region_taxa, region, taxa_mod, remove=FALSE)
-
-##	count cells within realm/climate/taxa hierarchy
-cell_count <- rarefied_medians %>%
-  group_by(rlm_region_taxa) %>%
-  dplyr::summarise(n_cells = n_distinct(rarefyID)) %>%
-  ungroup() 
-
-##	rejoin
-rarefied_medians <- left_join(cell_count, rarefied_medians, by='region_taxa')  
-
-##	filter to ecoregions with >3 cells
-rarefied_medians <- rarefied_medians %>%
-  filter(n_cells > 3 & BROAD_TYPE=='count')
-
-##	set some weakly regularising priors...
-hier_prior <- c(set_prior(prior = 'normal(0,0.2)', class='b', coef='cYEAR'), 	# global slope
-	set_prior(prior = 'normal(0,1)', class='Intercept', coef=''), 		# global intercept
-	#set_prior(prior = 'exponential(1)', class = 'sd'),
-	set_prior(prior = 'lkj(1)', class='cor')
-)
-
-S_pois_RTSRfyID_countData <- brm(bf(S ~ cYEAR + (cYEAR|rlm_region_taxa/STUDY_ID/rarefyID), 
-	family = brmsfamily('poisson')),	
-	data= rarefied_medians,
-	prior=hier_prior,
-	iter = 4000,
-	warmup = 1000,
-	#thin = 10,
-	inits = '0',
-	#init_r = 0.001,
[…]

### 2021-09-16 analysis/other_Blowes_analysis/rrt/02_get_coef.R

@@ -1,281 +1,0 @@
-## modify code for new region model that included realm: realm-region-taxa
-##------data and libraries-----------------------
-rm(list=ls())
-
-library(brms)
-library(dplyr)
-#library(broom)
-library(tidyr)
-library(tibble)
-
-#----load data
-##----count data only, whole time period--------
-load('~/Desktop/revision_models/rrt_models/S_pois_rlm_reg_txa-5211694.Rdata')
-##--species richness coefficients and random effects---------
-S_pois_new_ALL_global <- fixef(S_pois_RTSRfyID_countData, robust = TRUE, probs = c(0.05, 0.95)) %>%
-  as_tibble() %>%
-  mutate(
-    model = 'S_rlm_reg_txa',
-    time_period = 'ALL')
-
-names(S_pois_new_ALL_global)[3:4] <- c('lower', 'upper')
-
-S_pois_new_ALL_global <- S_pois_new_ALL_global %>%
-  mutate(
-    change = ifelse(lower > 0, 'up',
-                    ifelse(upper < 0, 'down', 'neutral'))) 
-
-
-
-S_pois_new_ALL_groupCoefs <- coef(S_pois_RTSRfyID_countData, robust = TRUE, probs = c(0.05, 0.95)) 
-
-##  break into rrt and rarefyID estimates
-S_pois_ALL_rrt <- as_tibble(S_pois_new_ALL_groupCoefs[[1]]) %>%
-  mutate(rrt = rownames(S_pois_new_ALL_groupCoefs[[1]]))
-names(S_pois_ALL_rrt)[c(3:4, 7:8)] <- c('lower_intercept', 'upper_intercept', 'lower_slope', 'upper_slope') 
-S_pois_ALL_rrt <- S_pois_ALL_rrt %>%
-  mutate(
-    model = 'S_rlm_reg_txa',
-    time_period = 'ALL',
-    change = ifelse(lower_slope > 0, 'up',
-                    ifelse(upper_slope < 0, 'down', 'neutral'))) 
-
-S_pois_new_ALL_studyID <- as_tibble(S_pois_new_ALL_groupCoefs[[2]]) %>%
-  mutate(level = rownames(S_pois_new_ALL_groupCoefs[[2]]))
[…]

### 2021-09-16 analysis/other_Blowes_analysis/rrt/03_coef_plots.R

@@ -1,301 +1,0 @@
-# code to plot coefficients of new model: this time with realm coded into groups: realm-region-taxa
-
-library(tidyverse)
-
-load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/rrt/model_coef_ranef/rlm_reg_txa_coef_ranef.Rdata')
-
-# first need to wrangle to levels for plots
-rrt_rrt_estimate <- rrt_rrt_estimate %>% 
-  # separate on the last (rightmost) underscore
-  separate(rrt, c('realm_region', 'taxa'), sep = '_(?!.*_)',
-           remove = F)
-
-rrt_rrt_estimate <- rrt_rrt_estimate %>% 
-  # separate realm and region
-  separate(realm_region, c('realm' , 'region'), sep = '(^[^_]*_)',
-           remove = F) %>% 
-  separate(realm_region, c('realm', 'nuu'), sep = '_', remove = F) %>% 
-  select(-nuu)
-
-# add realm metadata
-load('~/Dropbox/BiogeoBioTIME/rarefied_medians_continents.Rdata')
-
-rarefied_medians <- rarefied_medians %>% 
-  unite(rrt, REALM, region, taxa_mod, remove = F)
-
-rrt <- rarefied_medians %>% 
-  select(REALM, region, taxa_mod, rrt, Biome) %>% 
-  distinct(region, taxa_mod, rrt, .keep_all = T) %>% 
-  rename(taxa = taxa_mod)
-
-
-rrt_rrt_estimate <- rrt_rrt_estimate %>%
-  left_join(rrt,
-            by = c('rrt', 'region', 'taxa')) 
-
-# rename group of taxa labelled 'All' to 'Multiple'
-rrt_rrt_estimate$taxa2 <- factor(rrt_rrt_estimate$taxa, 
-                                      levels = c("Fish", "Benthos", "Birds", "Invertebrates", "Plant", "Amphibians",
-                                                 "All", "Marine invertebrates/plants", "Mammals"),
-                                      labels = c("Fish", "Benthos", "Birds", "Invertebrates", "Plant", "Amphibians",
-                                                 "Multiple taxa", "Marine invertebrates/plants", "Mammals"))
-
-# colours for each taxa
-taxa_col = c('Multiple taxa' = '#e6f598',
-             'Amphibians' = '#fee08b',
-             'Benthos' = '#5e4fa2',
-             'Birds' = '#f46d43',
-             'Fish' = '#3288bd',
-             'Invertebrates' = '#abdda4',
[…]

### 2021-09-16 analysis/other_Blowes_analysis/rrt/04_model_coef_compare.R

@@ -1,730 +1,0 @@
-# code to compare results of alternative models
-
-# high-level comparison - bt versus rrt: 
-# compare the biome/taxon level to the realm-region-taxon levels
-
-
-# low level comparisons: compare slope estimates for studies and cells that the different models share
-# biome-taxon versus realm-climate-taxon and bt versus realm-region-taxon
-
-rm(list=ls())
-
-library(tidyverse)
-
-# biome-taxa model
-load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/bt/model_fits_output/BT_model_coef_ranef.Rdata')
-# realm-climate-taxa model
-load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/rct/model_fits_output/rlm_clm_txa_modelCoefs.Rdata')
-# realm-region-taxa model
-load('~/Dropbox/1current/BioTime/BioGeo-BioDiv-Change/R/rrt/model_coef_ranef/rlm_reg_txa_coef_ranef.Rdata')
-
-#--------rarefied_medians.Rdata for meta data not in model object
-load('~/Dropbox/BiogeoBioTIME/rarefied_medians_continents.Rdata')
-rarefied_medians_region <- rarefied_medians
-load('~/Dropbox/BiogeoBioTIME/rarefied_medians.Rdata')
-
-rarefied_medians <- rarefied_medians %>%
-  group_by(Biome, taxa_mod) %>%
-  mutate(n_cells = n_distinct(rarefyID)) %>%
-  ungroup() %>%
-  filter(n_cells > 3)
-
-realm_climate_biome <- rarefied_medians %>%
-  distinct(REALM, climate_mod, Biome) %>%
-  ungroup()
-
-# want unique climate (latitudinal bands for each biome)
-# this is to compare like with like when comparing bt with rct:
-# both rct estimates and bimoe estimates span multiple latitudinal bands, 
-# here I constrain biomes and to later make comparisons with rct estimate
-# from the same latitude
-realm_climate_biome <- realm_climate_biome %>%
-  # filter(Biome %in% biome_levels_model$Biome) %>%
-  mutate(climate_mod = ifelse(Biome=='Arctic', 'Polar', 
-                              ifelse(Biome=='Boreal_Forests_Taiga', 'Temperate',
-                                     ifelse(Biome=='Cold_Temperate_Northeast_Pacific', 'Temperate',
-                                            ifelse(Biome=='Continental_High_Antarctic', 'Polar',
[…]

### 2021-09-16 paper/manuscript.Rmd
Minor reference and grammar changes

--- 2021-09-14 paper/manuscript.Rmd
@@
 
-#### *Data*
+#### Data
 
@@
                caption = "Summary of the data in the final trait database.") %>%
-  kable_classic() #%>% 
-  #kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
+  kable_classic() %>% 
+  kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
    #kableExtra::column_spec(2, width = "5cm") #%>%
@@
 
-#### *Diversity Metrics*
+#### Diversity Metrics
 
@@
 
-#### *Null Models*
+#### Null Models
 
@@
 
-#### *Analysis*
+#### Analysis
 
@@
 
-*Evidence of Ecological Processes*
+#### Evidence of Ecological Processes
 
@@
 
-*Potential Methodological Limitations*
+#### Potential Methodological Limitations
 
@@
 However, a number of gaps in best practices may be obscuring a true underlying trend.
-
 First, the Biotime database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
@@
 This scenario may be common in communities tracking changing environmental conditions.
-Approaches for assessing shifts and overlaps in functional space are still relatively new [@barros2016; @blonder2018; @mammola2019] a could shed critical insight into functional composition changes of this nature.
+Approaches for assessing shifts and overlaps in functional space are still relatively new [@barros2016; @blonder2018; @mammola2019] but could shed critical insight into functional composition changes of this nature.
 
-*Future Work*
+#### Future Work
 
-Our results illustrates a well known but increasingly pervasive problem in synthesis-based ecology of distinguishing true ecological patterns from methodological artifacts.
+Our results illustrate a well known but increasingly pervasive problem in synthesis-based ecology of distinguishing true ecological patterns from methodological artifacts.
 We present here the conclusion best supported by available data with a number of caveats and suggestions for next steps forward.
-
 The most pressing need is for intentional and targeted data collection efforts.
@@
 One relatively low-cost high-reward data investment is collation of additional species-level trait means.
-Intentional trait selection is critical for linking functional patterns to ecological process [@zhu2017], however synthesis is constrained to the traits in a few taxa-specific databases.
+Intentional trait selection is critical for linking functional patterns to ecological processes [@zhu2017], however synthesis is constrained to the traits in a few taxa-specific databases.
 Prioritization of additional traits linked to environmental change drivers (response traits) or ecosystem function (effect traits) in particular would provide a powerful link to existing frameworks for understanding functional change [@lavorel2002].
+
+### References

### 2021-09-16 analysis/01_filter_by_traits.Rmd
Clean up, annotate

@@
 ---
-title: "Check Coverage"
+title: "Merge biotime timeseries with trait data"
 author: "Kari Norman"
@@
 ```
+
+# Description
+This script merges trait data and biotime timeseries data, then filters out timeseries that don't have high enough trait coverage.
 
@@
 
-#Checking Coverage of ID's and trait data -> biotime
-
-check if all bird & mammal biotime data w/id's has an elton match. There are 18 species without a match.
+check if all bird & mammal biotime data w/id's has an elton match. There are 13 species without a match.
 ```{r}
@@
 
-and 25268 species without an ID, some of which are species codes
-```{r}
-biotime_data %>%
-  #filter(taxa == "Birds") %>%
-  select(id, sourceName, taxa) %>%
-  filter(is.na(id)) %>%
-  distinct() %>%
-  dim()
-```
-
-Are there any studies for which we have trait data for all species?
+Function for joining timeseries data with functional trait data. 
 ```{r}
@@
   biotime_data %>%
-    select(-scientificName) %>%
-    rename(biotimeName = sourceName) %>%
+    select(-scientificName) %>% #remove scientific name from biotime data
+    rename(biotimeName = sourceName) %>% #give original biotime name a better column name
   left_join(trait_data %>% rename(traitName = sourceName), by = "id",
-            na_matches = "never") %>%
+            na_matches = "never") %>% # join to trait data
    drop_na(scientificName)
@@
 
-Get data w/traits for all studies that have >85% trait coverage
+Get data w/traits for all studies that have >75% trait coverage
 
 `id` is `NA` if there was no `id` match
-`biotimeName` or `traitName` is `NA` if that ID is the result of an average of multliple species matches
+`biotimeName` or `traitName` is `NA` if that ID is the result of an average of multiple species matches
 `scientificName` is `NA` when there was no trait data for a given ID
@@
 
+#map across list of trait sources to merge timeseries with each one
 trait_merge <- map_dfr(traits, merge_data)
 
+#combine observations w/trait data back with observations with no trait data
 comp_biotime_traits <- trait_merge %>%
@@
 
+#get a percentage of trait coverage for each study
 coverage <- comp_biotime_traits %>%
@@
 
+#studies to include, since they meet the coverage requirement 
 inc_coverage <- coverage %>% filter(perc > percent)
@@
   group_by(rarefyid, year) %>%
-  filter(n_distinct(id) > 4) %>%
+  filter(n_distinct(id) > 4) %>% #more than four species for each observation
   ungroup()
@@
 
-Get biotime data filtered by trait coverage
+Get final columns 
 ```{r}
@@
 
-Split timeseries that have more than one class of species (split up birds and mammals)
+Some timeseries have observations from more than one class of species (e.g. birds and mammals), here we split them into two individual timeseries.
 ```{r}
@@
 
-#update that studies that needed to be split with new rarefyid breakdowns
+#update the studies that needed to be split with new rarefyid breakdowns
 bt_traitfiltered <- biotime_traits %>%
@@
 
+Final merged data set to save
 ```{r}

### 2021-09-16 analysis/02_rarefy_timeseries.Rmd
Rename file and move to rmd format

---
title: "Rarefy Timeseries"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script calculates diversity metrics for rarefied timeseries samples. It is a modified version of a script originally written by Shane Blowes and Sarah Supp found here: https://github.com/sChange-workshop/BioGeo-BioDiv-Change/blob/master/R/02_rarefy_griddedData_clusterVersion.R

```{r}
library(dplyr)
library(purrr)
library(tibble)
library(tidyr)
library(lazyeval)
library(vegan)
library(betapart)
library(doParallel)
library(foreach)
library(FD)
devtools::load_all()
```

Load data
```{r}
pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")

##	Get trait filtered data, rename colums to jive with the rest of the script
bt_grid_filtered <- pins::pin_get("bt-traitfiltered", board = "github") %>%
  rename_with(toupper) %>%
  rename(Species = SPECIES, StudyMethod = STUDYMETHOD, ObsEventID = OBSEVENTID,
         rarefyID = RAREFYID, cell = CELL, Abundance = ABUNDANCE, Biomass = BIOMASS)

trait_ref <- pins::pin_get("trait-ref", board = "github")
```

Function to rarefy data
```{r}
rarefy_diversity <- function(grid, type=c("count", "presence", "biomass"), resamples=100,
                             parallel, core_num, ...){

  #	CALCULATE RAREFIED METRICS for each study for all years
  #	restrict calculations to where there is abundance>0 AND
  #	following removal of NAs and 0's there are still more than 2 years

  # Check if the data is count abundance: if yes, calculate all rarefied metrics
  # Check if the data is presence or biomass: if yes, calculate only S and Jaccards
  # If is.na(ABUNDANCE_TYPE), then should calculate on the Biomass column
  # grid: input dataset
  # type: count, presence, or biomass
  # resamples: the number of bootstrap resampling events desired (default is 100)
  # trimsamples: TRUE means that years with < 1/2 the average number of samples should be removed to avoid excessive information loss. Default is FALSE.
  # calculate the number of sampling events per year, and find the minimum
  # resample the data to rarefy the diversity metrics
  # output a new dataframe

  if(type == "count" | type == "presence") { field = "Abundance"
  } else { field = "Biomass" }

  # Get the sample-size to rarefy to. How many sampling events per cell per year?
  # This is handled differently depending on the data type

  # define a filter for 'field' to be >0 and not NA
  zero_NA_filter <- interp(~y > x & !is.na(y),
                           .values = list(y = as.name(field), x = 0))

  #	define a function to calculate the sum(field) for use on the rarefied sample
  sum_field <- interp(~sum(as.numeric(var), na.rm=T),
                      var= as.name(field))

  nsamples <- ungroup(grid) %>%
    group_by(rarefyID, YEAR) %>%
    # remove 0's or NA's (this is to catch any places where abundance wasn't actually recorded
    # or the species is indicated as absent)
    filter_(.dots=zero_NA_filter) %>%
    # calculate how many observations per year per study
    dplyr::summarise(nsamples = n_distinct(ObsEventID))

  # Calculate the minimum number of samples per cell
  min_samp <- ungroup(nsamples) %>% group_by(rarefyID) %>%
    mutate(min_samp = min(nsamples), max_samp = max(nsamples)) %>%
    #check for studies that have consistent sampling and don't need to be rarefied
    mutate(rarefy = if_else(min_samp != max_samp, TRUE, FALSE)) %>%
    # retain only the rows with the minimum sample size for a given cell
    select(-max_samp, -nsamples, -YEAR) %>%
    distinct()

  #	Add the min_samp to the data and tidy a little
  grid <- inner_join(grid, min_samp)
  rm(min_samp)

  # Re-calculate metadata
  new_meta <- ungroup(grid) %>%
    # remove 0's or NA's (this is to catch any places where abundance wasn't actually recorded
    # or the species is indicated as absent)
    filter_(.dots=zero_NA_filter) %>%
    group_by(rarefyID) %>%
    summarise(
      # total number of species in rarefyID time-series
      SamplePool = n_distinct(Species),
      # total number of individuals
      SampleN = ifelse(type=='count', sum(as.numeric(Abundance)),
                       NA),
      # number of years sampled
      num_years = n_distinct(YEAR),
      # duration of time series, start and end points
      duration = max(YEAR) - min(YEAR) + 1,
      startYear = min(YEAR),
      endYear = max(YEAR),
      rarefied = unique(rarefy))

  #	Create dataframe where unique observations (i.e., the data of an
  #	ObsEventID's [individual species abundances])
  #	are nested within cells within years within studies
  bt_grid_nest <- ungroup(grid) %>%
    group_by(rarefyID, ObsEventID, cell, YEAR, min_samp) %>%
    # remove 0's or NA's (to catch any places where abundance wasn't actually recorded or the species is indicated as absent)
    filter_(.dots=zero_NA_filter) %>%
    # depending on type: nest(Species, Abundance) OR nest(Species, Biomass)
    nest(data=c("Species", field)) %>%
    # reduce to studies that have more than two time points for a given cell
    group_by(rarefyID) %>%
    #keeps all study_cells with 2 or more years of data
    filter(n_distinct(YEAR)>=2) %>%
    ungroup()

  ## loop to do rarefaction for each study
  for(i in 1:length(unique(bt_grid_nest$rarefyID))){
    #for(j in 1:3){

    #get ID for the study and filter data
    filter_id <- unique(bt_grid_nest$rarefyID)[i]

    #if(filter_id %in% filter(new_meta, rarefied == TRUE)$rarefyID){browser()}

    study <- bt_grid_nest %>%
      filter(rarefyID==filter_id)

    # get minimum sample size for rarefaction
    min_samp <- study %>% distinct(min_samp) %>% .$min_samp

    # check that there is only one cell represented (This shouldn't be a problem)
    if(length(unique(study$cell))>1) {
      stop(paste0("ERROR: ", filter_id, " contains more than one grid cell")) }

    # check there there is more than one year in the cell
    if(length(unique(study$YEAR))<2) {
      print(paste0("ERROR: ", filter_id, " does not have more than one year"))
      next }

    #check if we need to rarify
    if(isFALSE(unique(study$rarefy))){
      study_resamples <- 1
    }else{study_resamples <- resamples}

    ##	initialise df to store all biochange metrics
    rarefied_metrics <- data.frame()

    #set up parallelization
    # if(isTRUE(parallel) & study_resamples > 1){
    #   # set up parallel loop
    #   cl <- parallel::makeForkCluster(core_num) #not to overload your computer
    #   registerDoParallel(cl)
    # }else{
    #   registerDoSEQ()
    # }

    ##	rarefy rarefy_resamps times
    #rarefied_metrics <- foreach(j = 1:study_resamples, .combine=rbind) %dopar% {
    for(j in 1:study_resamples){
      print(paste('rarefaction', j, 'out of', study_resamples, 'for study_cell', i, '(', filter_id, ')',  'in', length(unique(bt_grid_nest$rarefyID))))

      rare_samp <- study %>%
        # rarefy to min_samp
        group_by(rarefyID, YEAR) %>%
        sample_n(size=min_samp) %>%
        # unpack and collate taxa from rarefied sample
        unnest(data) %>%
        # add unique counter for a resampling event
        mutate(rarefy_resamp = j) %>%
        # collate species within cells within years
        group_by(rarefyID, YEAR, cell, rarefy_resamp, Species) %>%
        dplyr::summarise_(
          Abundance=sum_field) %>%  #
        ungroup()

      # create community matrix of rarefied sample
      rare_comm  <- ungroup(rare_samp) %>%
        arrange(Species) %>% #order species so the trait matrix can in the same order
        spread(Species, Abundance, fill=0)

      #save out sample species matrix
      rare_comm_save <- rare_comm %>% mutate(type = type)
      path <- here::here("data", "rarefied_samples")
      dir.create(path)
      save(rare_comm_save, file=paste0(path, "/", type, "_cell", unique(study$rarefyID), "_sample", j, ".rda"))

      #get vector of years in the same order as the community matrix to tack back on to FD dataframe later
      years <- rare_comm$YEAR

      rare_comm <- rare_comm %>%
        column_to_rownames("YEAR") %>%
        dplyr::select(-rarefyID, -cell, -rarefy_resamp)

      # betapart requires presence/absence matrix for Jaccard calculations of turnover/nestedness
      rare_comm_binary <- with(rare_comm, ifelse(rare_comm > 0, 1, 0))

      # initialise matrices for calculating turnover
      simbaseline <- data.frame(array(NA, dim=c(length(unique(rare_samp$YEAR)), 11)))
      names(simbaseline)<-c('YEAR', 'cell', 'Jaccard_base','Horn_base','Chao_base','Pearson_base','Gains_base','Losses_base', 'Jbeta_base', 'Jtu_base', 'Jne_base')#, 'Jbeta_base_func', 'Jtu_base_func', 'Jne_base_func')

      simnext <- data.frame(array(NA, dim=c(length(unique(rare_samp$YEAR)), 11)))
      names(simnext)<-c('YEAR', 'cell', 'Jaccard_next','Horn_next','Chao_next','Pearson_next','Gains_next','Losses_next', 'Jbeta_next', 'Jtu_next', 'Jne_next')#, 'Jbeta_next_func', 'Jtu_next_func', 'Jne_next_func')

      simhind <- data.frame(array(NA, dim=c(length(unique(rare_samp$YEAR)), 11)))
      names(simhind)<-c('YEAR', 'cell', 'Jaccard_hind','Horn_hind','Chao_hind','Pearson_hind','Gains_hind','Losses_hind', 'Jbeta_hind', 'Jtu_hind', 'Jne_hind')#, 'Jbeta_hind_func', 'Jtu_hind_func', 'Jne_hind_func')

      counter2 <- 1

      if(type=="count"){

        # calculating between year similarities (NOT DISTANCE!) with Jaccard, Morisita-Horn, Chao and Pearson correlations
        Pearsoncor <- cor(t(log(rare_comm+1)), method='pearson')
        Jacsim <- as.matrix(1-vegdist(rare_comm, method='jaccard', binary=TRUE))
        Hornsim <- as.matrix(1-vegdist(rare_comm, method='horn'))
        Chaosim <- as.matrix(1-vegdist(rare_comm, method='chao'))
        Gainssim <- as.matrix(designdist(rare_comm, method = "B-J", terms = "binary",  abcd = FALSE, alphagamma = FALSE, "gains"))
        Lossessim <- as.matrix(designdist(rare_comm, method = "A-J", terms = "binary",  abcd = FALSE, alphagamma = FALSE, "losses"))
        # two steps for Jaccard components (so as calculation is done only once)
        J_components <- beta.pair(rare_comm_binary, index.family='jaccard')	# distance
        Jbeta <- as.matrix(J_components$beta.jac)
        Jtu <- as.matrix(J_components$beta.jtu)
        Jne <- as.matrix(J_components$beta.jne)
        n <- length(years)

        # calculate univariate metrics
        uni_metrics <- ungroup(rare_samp) %>%
          group_by(rarefyID, YEAR, cell, rarefy_resamp) %>%
          summarise(
            N = sum(as.numeric(Abundance)),
            S = n_distinct(Species),
            PIE = diversity(Abundance, index='simpson'),
            ENSPIE = diversity(Abundance, index='invsimpson'),
            pielou = diversity(Abundance)/log(S),
            Hill1 = renyi(Abundance,scales = 1,hill = T)[1]) %>%
          #Hill2 = renyi(Abundance,scales = 2,hill = T)[1]
          ungroup()

      }
      else {
        # ONLY the metrics we need for biomass and presence (S, and Jaccard's)
        # For presence data, first convert rare_comm to a binary species matrix (0,1)   #FIXME: Also do this for Biomass? Because of limiting calcs, the only thing this would affect is Pearson cor?
        if(type=="presence" | type=='biomass'){
          rare_comm[rare_comm >0 ] <- 1 }

        # calculating between year similarities (NOT DISTANCE!) with Jaccard, Morisita-Horn, Chao and Pearson correlations
        n <- length(unique(rare_samp$YEAR))
        Pearsoncor <- cor(t(log(rare_comm+1)), method='pearson')
        Jacsim <- as.matrix(1-vegdist(rare_comm, method='jaccard', binary=TRUE))
        Hornsim <- matrix(nrow=n, ncol=n, NA)
        Chaosim <- matrix(nrow=n, ncol=n, NA)
        Gainssim <- matrix(nrow=n, ncol=n, NA)
        Lossessim <- matrix(nrow=n, ncol=n, NA)
        # two steps for Jaccard components (so as calculation is done only once)
        J_components <- beta.pair(rare_comm, index.family='jaccard')	# distance
        Jbeta <- as.matrix(J_components$beta.jac)
        Jtu <- as.matrix(J_components$beta.jtu)
        Jne <- as.matrix(J_components$beta.jne)

        # calculate univariate metrics
        uni_metrics <- ungroup(rare_samp) %>%
          group_by(rarefyID, YEAR, cell, rarefy_resamp) %>%
          summarise(
            N = NA,
            S = n_distinct(Species),
            PIE = NA,
            ENSPIE = NA,
            pielou = NA,
            Hill1 = NA) %>%
          #Hill2 = NA
          ungroup()
      }

      #Collate metrics
      # How baseline is calculated.
      # NB: we do not compare first year with itself here (to be added before model fitting later)
      simbaseline[counter2:(counter2+n-2),] <- cbind(
        unique(rare_samp$YEAR)[2:n],
        unique(rare_samp$cell),
        Jacsim[2:n],
        Hornsim[2:n],
        Chaosim[2:n],
        Pearsoncor[2:n],
        Gainssim[2:n],
        Lossessim[2:n],
        Jbeta[2:n],
        Jtu[2:n],
        Jne[2:n]#,
        # Jbeta_func[2:n],
        # Jtu_func[2:n],
        # Jne_func[2:n]
      )

      # How consecutive is calculated.
      simnext[counter2:(counter2+n-2),] <- cbind(
        unique(rare_samp$YEAR)[2:n],
        unique(rare_samp$cell),
        Jacsim[row(Jacsim)-col(Jacsim)==1],
        Hornsim[row(Hornsim)-col(Hornsim)==1],
        Chaosim[row(Chaosim)-col(Chaosim)==1],
        Pearsoncor[row(Pearsoncor)-col(Pearsoncor)==1],
        Gainssim[row(Gainssim)-col(Gainssim)==1],
        Lossessim[row(Lossessim)-col(Lossessim)==1],
        Jbeta[row(Jbeta)-col(Jbeta)==1],
        Jtu[row(Jtu)-col(Jtu)==1],
        Jne[row(Jne)-col(Jne)==1]#,
        # Jbeta_func[row(Jbeta_func)-col(Jbeta_func)==1],
        # Jtu_func[row(Jtu_func)-col(Jtu_func)==1],
        # Jne_func[row(Jne_func)-col(Jne_func)==1]
      )

      # How hindcasting is calculated.
      simhind[counter2:(counter2+n-2),] <- cbind(
        unique(rare_samp$YEAR)[1:(n-1)],
        unique(rare_samp$cell),
        Jacsim[row(Jacsim)%in%1:(max(row(Jacsim))-1) & col(Jacsim)==max(col(Jacsim))],
        Hornsim[row(Hornsim)%in%1:(max(row(Hornsim))-1) & col(Hornsim)==max(col(Hornsim))],
        Chaosim[row(Chaosim)%in%1:(max(row(Chaosim))-1) & col(Chaosim)==max(col(Chaosim))],
        Pearsoncor[row(Pearsoncor)%in%1:(max(row(Pearsoncor))-1) & col(Pearsoncor)==max(col(Pearsoncor))],
        Gainssim[row(Gainssim)%in%1:(max(row(Gainssim))-1) & col(Gainssim)==max(col(Gainssim))],
        Lossessim[row(Lossessim)%in%1:(max(row(Lossessim))-1) & col(Lossessim)==max(col(Lossessim))],
        Jbeta[row(Jbeta)%in%1:(max(row(Jbeta))-1) & col(Jbeta)==max(col(Jbeta))],
        Jtu[row(Jtu)%in%1:(max(row(Jtu))-1) & col(Jtu)==max(col(Jtu))],
        Jne[row(Jne)%in%1:(max(row(Jne))-1) & col(Jne)==max(col(Jne))]#,
        # Jbeta_func[row(Jbeta_func)%in%1:(max(row(Jbeta_func))-1) & col(Jbeta_func)==max(col(Jbeta_func))],
        # Jtu_func[row(Jtu_func)%in%1:(max(row(Jtu_func))-1) & col(Jtu_func)==max(col(Jtu_func))],
        # Jne_func[row(Jne_func)%in%1:(max(row(Jne_func))-1) & col(Jne_func)==max(col(Jne_func))]
      )

      # combine univariate and turnover metrics
      biochange_metrics <- full_join(uni_metrics, simbaseline[-length(unique(rare_samp$YEAR)),], by=c('YEAR', 'cell')) %>%
        full_join(simnext[-length(unique(rare_samp$YEAR)),], by=c('YEAR', 'cell')) %>%
        full_join(simhind[-length(unique(rare_samp$YEAR)),], by=c('YEAR', 'cell'))

      # add to dataframe for all studies
      rarefied_metrics <- bind_rows(rarefied_metrics, biochange_metrics)

    }	# rarefyID loop (STUDY_CELL ID)

    rarefied_metrics <- inner_join(new_meta, rarefied_metrics) %>%
      mutate(type = type)

    #save out the files
    path <- here::here("data", "rarefied_metrics")
    dir.create(path)
    save(rarefied_metrics, file=paste0(path, "/rarefyID_", filter_id, type, ".rda"))

  }	# rarefaction loop
} # END function
```

Get rarefied diversity metrics
```{r}
## Separate true abundance (count) data from presence and biomass/cover data
bt_grid_abund <- bt_grid_filtered %>%
  filter(ABUNDANCE_TYPE %in% c("Count", "Density", "MeanCount"))

bt_grid_pres <- bt_grid_filtered %>%
  filter(ABUNDANCE_TYPE == "Presence/Absence")

## Get rarefied resample for each type of measurement (all data)
rarefy_abund <- rarefy_diversity(grid=bt_grid_abund, trait_data = trait_ref, type="count", trait_axes = "min", parallel = TRUE, resamples=200, core_num = 44)
rarefy_pres <- rarefy_diversity(grid=bt_grid_pres, trait_data = trait_ref, type="presence", trait_axes = "min", parallel = TRUE, resamples=200, core_num = 20)
```

Functions to calculate functional diversity metrics for the samples
```{r}
#helper function to load file
loadRData <- function(file_name){
  load(file_name)
  get(ls()[ls() != "file_name"])
}

#function to calculate functional diversity metrics for a single sample community
calc_FD_mets <- function(file_path, traits){
  #load community sample
  rare_comm_save <- loadRData(file_path)

  years <- rare_comm_save$YEAR

  #get species matrix
  species_mat <- rare_comm_save %>%
    column_to_rownames("YEAR") %>%
    dplyr::select(-c(rarefyID, cell, type, rarefy_resamp))

  # get trait matrix for the species in the sample
  traits <- get_traitMat(colnames(species_mat), trait_data = traits)

  # create FD function that returns an empty dataframe instead of erroring if FD can't be calculated
  get_FD_safe <- possibly(get_FD, otherwise = data_frame())
  betapart_safe <- possibly(functional.beta.pair, otherwise = data.frame())

  # calculate functional diversity for different data types
  if(unique(rare_comm_save$type) == "count"){
    FD_mets <- get_FD_safe(species_mat = species_mat, trait_mat = traits, year_list = years,
                           data_id = unique(rare_comm_save$rarefyID), 
                           samp_id = unique(rare_comm_save$rarefy_resamp),
                           w.abun = TRUE, m = "min", corr = "cailliez")
  } else{
    FD_mets <- get_FD_safe(species_mat = species_mat, trait_mat = traits, year_list = years,
                           data_id = unique(rare_comm_save$rarefyID), 
                           samp_id = unique(rare_comm_save$rarefy_resamp),
                           w.abun = FALSE, m = "min", corr = "cailliez")
  }

  #check if FD metrics could be calculated, otherwise return empty dataframe
  if (is.null(dim(FD_mets))){
    species_mat_binary <- with(species_mat, ifelse(species_mat > 0, 1, 0))
    J_func_components <- betapart_safe(x = species_mat_binary, traits = FD_mets$pca_traits, index.family='jaccard')	# distance
    #check if the betapart metrics could be calculated, otherwise return dataframe of just FD metrics
    if(is.null(dim(J_func_components))){
      Jbeta_func <- as.matrix(J_func_components$funct.beta.jac)
      Jtu_func <- as.matrix(J_func_components$funct.beta.jtu)
      Jne_func <- as.matrix(J_func_components$funct.beta.jne)

      n <- length(unique(rare_comm_save$YEAR))

      # initialise matrices for calculating turnover
      simbaseline <- data.frame(array(NA, dim=c(n, 5)))
      names(simbaseline)<-c('YEAR', 'cell', 'Jbeta_base_func', 'Jtu_base_func', 'Jne_base_func')

      simnext <- data.frame(array(NA, dim=c(n, 5)))
      names(simnext)<-c('YEAR', 'cell', 'Jbeta_next_func', 'Jtu_next_func', 'Jne_next_func')

      simhind <- data.frame(array(NA, dim=c(n, 5)))
      names(simhind)<-c('YEAR', 'cell', 'Jbeta_hind_func', 'Jtu_hind_func', 'Jne_hind_func')

      counter2 <- 1

      #Collate metrics
      # How baseline is calculated.
      # NB: we do not compare first year with itself here (to be added before model fitting later)
      simbaseline[counter2:(counter2+n-2),] <- cbind(
        unique(rare_comm_save$YEAR)[2:n],
        unique(rare_comm_save$cell),
        Jbeta_func[2:n],
        Jtu_func[2:n],
        Jne_func[2:n]
      )

      # How consecutive is calculated.
      simnext[counter2:(counter2+n-2),] <- cbind(
        unique(rare_comm_save$YEAR)[2:n],
        unique(rare_comm_save$cell),
        Jbeta_func[row(Jbeta_func)-col(Jbeta_func)==1],
        Jtu_func[row(Jtu_func)-col(Jtu_func)==1],
        Jne_func[row(Jne_func)-col(Jne_func)==1]
      )

      # How hindcasting is calculated.
      simhind[counter2:(counter2+n-2),] <- cbind(
        unique(rare_comm_save$YEAR)[1:(n-1)],
        unique(rare_comm_save$cell),
        Jbeta_func[row(Jbeta_func)%in%1:(max(row(Jbeta_func))-1) & col(Jbeta_func)==max(col(Jbeta_func))],
        Jtu_func[row(Jtu_func)%in%1:(max(row(Jtu_func))-1) & col(Jtu_func)==max(col(Jtu_func))],
        Jne_func[row(Jne_func)%in%1:(max(row(Jne_func))-1) & col(Jne_func)==max(col(Jne_func))]
      )

      # combine univariate and turnover metrics
      biochange_metrics <- full_join(simbaseline[-n,],
                                     FD_mets$fd_met %>% dplyr::select(-c(nbsp, sing.sp)),
                                     by=c('YEAR' = 'year')) %>%
        mutate(cell = unique(na.omit(cell))) %>%
        full_join(simnext[-n,], by=c('YEAR', 'cell')) %>%
        full_join(simhind[-n,], by=c('YEAR', 'cell')) %>%
        mutate(rarefy_resamp = unique(rare_comm_save$rarefy_resamp))

    } else{
      biochange_metrics <- FD_mets$fd_met %>%
        select(-c(nbsp, sing.sp)) %>%
        mutate(cell = unique(rare_comm_save$cell), rarefy_resamp = unique(rare_comm_save$rarefy_resamp)) %>%
        rename(YEAR = year)
    }
  }else{ biochange_metrics <- tibble() }

  #save out dataframe of rarefied metrics
  path <- here::here("data", "rarefied_metrics", "fd")
  dir.create(path)
  save(biochange_metrics, file=paste0(path, "/", unique(rare_comm_save$type),
                                      "_cell", unique(rare_comm_save$rarefyID),
                                      "_sample", unique(rare_comm_save$rarefy_resamp), ".rda"))

}

#get list of rarefaction sample files
path <- here::here("data", "rarefied_samples")
files <- dir(path, "*.rda") %>%
  paste0(path, "/", .)

#parallelized mapping across files to calculate functional diversity metrics
plan("multiprocess", workers = 30)
furrr::future_map(files, calc_FD_mets, traits = trait_ref)
```

### 2021-09-16 analysis/02_rarefy_timeseries.Rmd
more intuitive variable names

@@
 ##	Get trait filtered data, rename colums to jive with the rest of the script
-bt_grid_filtered <- pins::pin_get("bt-traitfiltered", board = "github") %>%
+bt_traitfiltered <- pins::pin_get("bt-traitfiltered", board = "github") %>%
   rename_with(toupper) %>%
@@
 ## Separate true abundance (count) data from presence and biomass/cover data
-bt_grid_abund <- bt_grid_filtered %>%
+bt_grid_abund <- bt_traitfiltered %>%
   filter(ABUNDANCE_TYPE %in% c("Count", "Density", "MeanCount"))
 
-bt_grid_pres <- bt_grid_filtered %>%
+bt_grid_pres <- bt_traitfiltered %>%
   filter(ABUNDANCE_TYPE == "Presence/Absence")

### 2021-09-16 analysis/03_rarefy_null_models.Rmd
more intuitive variable names

@@
   paste0(path, "/", .)
-
-# samples <- purrr::map_dfr(files, function(x) {load(x)
-#   return(rare_comm_save)
-# 	}) %>% distinct()
 ```
 
-Function to get all rarefied samples for an ID
-```{r}
-# get_rarefy_samps <- function(id, path = here::here("data", "rarefied_samples")){
-#   print(id)
-#   samp_data <- list.files(path = path, pattern = paste0( "*.", id, ".*")) %>% 
-#     paste0(path, "/", .) %>%
-#      purrr::map_dfr(., function(x) {load(x)
-#        return(rare_comm_save)
-#        }) 
-#   return(samp_data)
-# }
-```
 Get species pools for each timeseries
 ```{r}
-bt_grid_filtered <- pins::pin_get("bt-traitfiltered", board = "github")
+bt_traitfiltered <- pins::pin_get("bt-traitfiltered", board = "github")
 trait_ref <- pins::pin_get("trait-ref", board = "github")
 
-species_pools <- bt_grid_filtered %>%
+species_pools <- bt_traitfiltered %>%
   select(rarefyid, species) %>% 
@@
 
-rm(bt_grid_filtered)
+rm(bt_traitfiltered)
 ```

### 2021-09-17 analysis/03_rarefy_null_models.Rmd
clearer annotation

@@
 ```{r}
+#get list of files
 path <- here::here("data", "rarefied_samples")
@@
 
+#for each rarefyID, get complete list of species
 species_pools <- bt_traitfiltered %>%
@@
 ```{r}
+
+#for each file, sample n times; each file is data for one community
 get_plot_sample <- function(file, n){
   #print(filter_id)
-  load(file)
+  load(file) #loads file called rare_comm_save
+  #put in matrix form
   comm <- rare_comm_save %>%
@@
 
+  #get rarefyid and resample number 
   filter_rarefyid <- unique(comm$rarefyid) %>% unlist()
@@
 
-  # plot_data <- data %>% filter(rarefyID == filter_id) %>%
-  #   select(year, id, weight)
-
+  #only species occuring at that rarefyid should be in the species pool
   species_pool <- species_pools %>%
@@
 
+  #function to get sample species and abundances
   get_samp_species <- function(data){
@@
 
+  #loop over the number of samples
   site_samples <- data.frame()
@@
     iter_samples <- data.frame()
+    #for each year get a sample species list
     for(i in unique(comm$year)){
@@
 
+  #save out samples
   save_dir <- here::here("data", "null_samples")
@@
 ```{r}
+#get null samples for each rarefied sample
 future::plan(multiprocess, workers = 30)
 sample_occurrence <- future_map(files, get_plot_sample, n = 500, 
-                                .options = furrr_options(seed = TRUE)) #%>%
-  # bind_rows() %>%
-  # replace(is.na(.), 0)
+                                .options = furrr_options(seed = TRUE)) 
 ```
@@
 
+Calculated FD metrics for all the null samples
 ```{r}
@@
 
+  #get species matrix
   sample_species_mat <- plot_samples %>%
@@
 
-  #print(head(sample_species_mat))
+  #get trait matrix
   trait_mat <- get_traitMat(colnames(sample_species_mat), trait_data = traits)
@@
 
+  # check if data are counts or abundances, then calculated FD
   max_val <- max(unlist(map(sample_species_mat, max)))
@@
 
+#get list of null sample files
 null_files <-  dir(here::here("data", "null_samples"), "*.csv") %>%
@@
 ```{r}
+#get list of null fd sample files
 path <- here::here("data", "fd_null_output")
@@
 
+#function to read in null fd samples and get info to calculate stats
 process_nulls <- function(file_path, n){
@@
 
+#write out null table
 usethis::use_data(null_table)

### 2021-09-17 analysis/04_collate_rarefied_null.Rmd
Move to rmd

---
title: "Collate Rarefied Samples and Null Models"
output: html_document
---

```{r}
library(dplyr)
library(tibble)
library(stringr)
```

Read in files
```{r}
##	load species based metrics
path <- here::here("data", "rarefied_metrics")
filelist <-  dir(path, "*.rda") %>%
  paste0(path, "/", .)

species_metrics <- purrr::map_dfr(filelist, function(x) {load(x)
  return(rarefied_metrics)
	}) %>% distinct()

## load fd based metrics
path <- here::here("data", "rarefied_metrics", "fd")
filelist <-  dir(path, "*.rda") %>%
  paste0(path, "/", .)

#read in metric data and do some clean up before combining
fd_metrics <- purrr::map_dfr(filelist, function(x) {
    load(x)
    rarefyID <- str_match(x,"_cell\\s*(.*?)\\s*_sample")[,2]

    biochange_metrics <- biochange_metrics %>%
      mutate(rarefyID = rarefyID)
    return(biochange_metrics)
  }) %>% distinct()

## get null models
null_table <- pins::pin_get("null-table", board = "github")
```

Merge rarefied species and fd metrics and null models
```{r}
rarefied_metrics <- full_join(species_metrics, fd_metrics,
                              c("YEAR", "cell", "rarefy_resamp", "rarefyID")) %>%
  filter(S != 1) %>% # remove samples that had one species as they result in strange metric values
  select(-qual.FRic)

##	pull out new metadata
new_meta <- rarefied_metrics %>%
	distinct(rarefyID, SamplePool, SampleN, num_years, duration, startYear, endYear)

##	calculate the medians for all the metrics for studies that were rarefied
metric_cols <- ungroup(rarefied_metrics) %>%
  select(-c(SamplePool, SampleN, num_years, duration, startYear, endYear, CWM, type)) %>%
  pivot_longer(-c(rarefyID, rarefied, YEAR, cell, rarefy_resamp), names_to = "metric") %>%
  left_join(null_table %>% select(-c(richness, se, lowerCI, upperCI)),
            by = c("YEAR" = "year", "rarefyID" = "rarefyid", "rarefy_resamp" = "rarefy_resamp", "metric" = "metric"))

# get a dataframe of rarefyID's and rarefication samples that we don't have null_model for, and a count of the number of resamples that are missing
no_null <- metric_cols %>%
  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis"), is.na(mean)) %>%
  select(rarefyID, rarefy_resamp) %>%
  distinct()

missing_nulls <- null_table %>%
  right_join(no_null, by = c("rarefyid" = "rarefyID","rarefy_resamp")) %>%
  group_by(rarefyid) %>%
  summarise(n_missing_nulls = n_distinct(rarefy_resamp)) %>%
  mutate(commplete_null_samps = "FALSE")

# calculate SES and incorporate into metric columns
metric_cols <- metric_cols %>%
  mutate(SES = (value - mean)/sd) %>%
  #for years where species richness = richness of the species pool, FRic sd = 0, so use the original value with no SES correction
  mutate(SES = case_when(
    sd == 0 ~ 0,
    TRUE ~ SES)) %>%
  select(-c(mean, sd)) %>%
  #the next three lines go back to wide format for both SES and unadjusted values, then to long format incorporating SES values
  pivot_wider(names_from = "metric", values_from = c("value", "SES")) %>%
  rename_at(vars(starts_with("value")), ~str_replace(., "value_", "")) %>%
  pivot_longer(-c(rarefyID, rarefied, YEAR, cell, rarefy_resamp), names_to = "metric") %>%
  filter(!is.na(value))

rarefied_medians <- metric_cols %>%
  group_by(rarefyID, YEAR, cell, rarefied, metric) %>%
  summarise(value = median(value, na.rm = TRUE))

##	recombine with new metadata
rarefied_metrics <- inner_join(new_meta, rarefied_medians, by='rarefyID') %>%
  left_join(missing_nulls, by = c("rarefyID" = "rarefyid"))
  #inner_join(rarefied_ints)

##	save
usethis::use_data(rarefied_metrics)

pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
pins::pin(rarefied_metrics, board = "github")
```

### 2021-09-17 analysis/05_model_metrics.Rmd
Add some explanatory comments

@@
 library(tidyverse)
-#library(moderndive)
 library(lme4)
@@
 ```
-Add human footprint data
+
+Option to add human footprint data
 ```{r}
@@
 
-Convergence problems!! See here:
+Convergence problems! See here:
 https://joshua-nugent.github.io/allFit/
@@
 
-```{r}
-# SES_rand_slope <- model_data %>%
-#   filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_next")) %>%
-#   group_nest(metric) %>%
-#   mutate(#year_center = map(data, ~scale(.x$year)),
-#          model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
-#                                  data = .x)),
-#            # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
-#            #                       data = .x)),
-#            # test = map2(model, null_model, anova),
-#            # test_res = map(test, tidy),
-#            coef = map(model, tidy),
-#            ) %>% 
-#   select(metric, model, coef) %>%
-#   unnest(cols = c(coef)) %>%
-#   filter(term == "year_scaled") %>%
-#   select(-c(effect, group, term, statistic))
-# 
-# #check that they all used the same number of observations
-# map(SES_rand_slope$model, nobs)
-# 
-# #these two metrics are cranky and want their own year scaling to converge
-# 
-# #Jaccard Base
-# jacc_base_mod <-  model_data %>%
-#   filter(metric == ("Jaccard_base")) %>%
-#   mutate(year_scaled = scale(year)) %>%
-#   group_nest(metric) %>%
-#   mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
-#                                  data = .x)),
-#            coef = map(model, tidy),
-#            ) %>% 
-#   select(metric, model, coef) %>%
-#   unnest(cols = c(coef)) %>%
-#   filter(term == "year_scaled") %>%
-#   select(-c(effect, group, term, statistic))
-# 
-# # S model
-# s_mod <- model_data %>%
-#   filter(metric %in%  c("S")) %>%
-#   #it wants its own scaling to converge for some reason
-#   mutate(year_scaled = scale(year)) %>%
-#   group_nest(metric) %>%
-#   mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
-#                                  data = .x, REML = TRUE)),
-#            coef = map(model, tidy)
-#            ) %>% 
-#   select(metric, model, coef) %>%
-#   unnest(cols = c(coef)) %>%
-#   filter(term == "year_scaled") %>%
-#   select(-c(effect, group, term, statistic))
-# 
-# metric_model_table <- bind_rows(SES_rand_slope, s_mod, jacc_base_mod)
-# 
-# #save coefficients and models
-# usethis::use_data(metric_model_table)
-# pins::pin(metric_model_table, board = "github")
-# ```
-# Taxa model
-# ```{r}
-# taxa_rand_slope <- model_data %>%
-#   filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv")) %>%
-#   group_nest(metric) %>%
-#   mutate(#year_center = map(data, ~scale(.x$year)),
-#          model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa), 
-#                                  data = .x)),
-#            # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
-#            #                       data = .x)),
-#            # test = map2(model, null_model, anova),
-#            # test_res = map(test, tidy),
-#            coef = map(model, tidy),
-#            ) %>% 
-#   select(metric, model, coef) %>%
-#   unnest(cols = c(coef))
-# 
-# #Jaccard Base
-# jacc_taxa_base_mod <-  model_data %>%
-#   filter(metric == ("Jaccard_base")) %>%
-#   mutate(year_scaled = scale(year)) %>%
-#   group_nest(metric) %>%
-#   mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa), 
-#                                  data = .x)),
-#            coef = map(model, tidy),
-#            ) %>% 
-#   select(metric, model, coef) %>%
-#   unnest(cols = c(coef)) %>%
-#   filter(term == "year_scaled") %>%
-#   select(-c(effect, group, term, statistic))
-# 
-#  s_taxa_mod <- model_data %>%
-#   filter(metric %in%  c("S")) %>%
-#   #it wants its own scaling to converge for some reason
-#   mutate(year_scaled = scale(year)) %>%
-#   group_nest(metric) %>%
-#   mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa), 
-#                                  data = .x, REML = TRUE)),
-#            coef = map(model, tidy)
-#            ) %>% 
-#   select(metric, model, coef) %>%
-#   unnest(cols = c(coef)) %>%
-#   filter(term == "year_scaled") %>%
-#   select(-c(effect, group, term, statistic))
-# 
-# fric_taxa_mod <- model_data %>%
-#   filter(metric %in%  c("SES_FRic")) %>%
-#   #it wants its own scaling to converge for some reason
-#   mutate(year_scaled = scale(year)) %>%
-#   group_nest(metric) %>%
-#   mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa), 
-#                                  data = .x, REML = TRUE)),
-#            coef = map(model, tidy)
-#            ) %>% 
-#   select(metric, model, coef) %>%
-#   unnest(cols = c(coef)) %>%
-#   filter(term == "year_scaled") %>%
-#   select(-c(effect, group, term, statistic))
-# 
-# taxa_model_table <- bind_rows(taxa_rand_slope, jacc_taxa_base_mod, s_taxa_mod, fric_taxa_mod)
-```
-
 #### Process Model Output
@@
 
+  #get the statistics for the year variable
   year_scaled_coef <- tidy(model) %>%
@@
 
+  #get the year estimate
   year_scaled_est <- pull(year_scaled_coef, estimate)
 
+  #get year variation
   year_scaled_var <- year_scaled_coef %>% 
@@
 
+  #get study level conditional estimates and add to over all estimates
   study_ests <- broom.mixed::tidy(model, effects="ran_vals") %>%
@@
     select(-c(effect, group, term)) %>%
-    #rename({{ model_id }} := level, cond.std.error = std.error, cond.estimate = estimate) %>%
     rename(cond.std.error = std.error, cond.estimate = estimate) %>%
@@
   select(study_id, metric)
-```
-
-```{r}
-# #And how many were significant for all but Jaccard?
-# study_slopes %>% 
-#   filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "S")) %>% 
-#   group_by(study_id) %>% filter(sum(sig) == 4) %>% pull(study_id) %>% 
-#   n_distinct()
-# 
-# #Breakdown of relationship between functional richness and species richness
-# s_sig <- study_slopes %>% filter(metric == "S", sig == TRUE) %>% pull(study_id) %>% unique()
-# s_sig_not_fric <- study_slopes %>% 
-#   filter(study_id %in% s_sig, metric == "FRic", sig == FALSE) %>% 
-#   pull(study_id) %>%
-#   unique()
-# 
-# both_sig <- study_slopes %>% 
-#   filter(study_id %in% s_sig, metric == "FRic", sig == TRUE) %>%
-#   pull(study_id)
-# 
-# both_pos
-# fric_pos
-# s_pos_fric_neg <- study_slopes %>%
-#   filter(study_id %in% both_sig) %>% 
-#   group_by(study_id) %>%
-#   filter(sum(sign(estimate)) == 0)
-
 ```

### 2021-09-17 analysis/exporatory_scripts/bbs_timeseries.Rmd
Clean up folders

---
title: "Temporal Analysis"
author: "Kari Norman"
date: "9/19/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#devtools::install_github("karinorman/functional_diversity")

library(tidyverse)
library(dplyr)
library(FD)
library(furrr)
library(here)
library(foreach)
#library(functional.diversity)

piggyback::pb_download(file = c("data/trait.rda", "data/bbs.rda"), repo = "karinorman/functional_diversity", dest = here())

load(here("data", "trait.rda"))
load(here("data", "bbs.rda"))
```

Get FD and TD for each site for each year
```{r}
min_year <- 1969

bbs_wide <- bbs %>%
  select(site_id, scientific, year, abundance) %>%
  spread(scientific, abundance)

bbs_species_mat <- select(bbs_wide, -site_id, -year)
bbs_trait_mat <- get_trait_matrix(colnames(bbs_species_mat))

data_path <- paste0(here('data'), '/FD_bbs_yearly.RData')
  if (file.exists(data_path)){
    print("FD present")
    FD_file <- load(data_path)
  }else{
    print("No FD")
    FD <- as.data.frame(dbFD(bbs_trait_mat, bbs_species_mat, w.abun = TRUE))
    save(FD, file = data_path)
    pb_upload(data_path)
  }

FD <- cbind(bbs_wide[c('site_id', 'year')], FD) #dbFD() preserves row order, so we can just brute force labels back on 
#FD$site_id <- as.factor(FD$site_id)

#double check that labels are right by comparing richnesses calculated below
# richness <- bbs %>%
#   select(site_id, scientific, year) %>%
#   group_by(site_id, year) %>%
#   summarise(n = n())

```

Jarzyna & Jetz 2016 (doi: 10.1111/gcb.13571) calculates a few different metrics to measure temporal change.

Change relative to previous year: (previous year - current year)/current year x 100%
```{r}
yearly_delta <- FD %>%
  select(site_id, year, nbsp, FRic) %>%
  group_by(site_id) %>%
  arrange(year) %>%
  mutate(nbsp_delta = (lag(nbsp) - nbsp)/nbsp,
         fric_delta = (lag(FRic) - FRic)/FRic)
```

Simpson's dissimilarity
Still need to figure out how to deal with holes in the timeseries

Attempt using purrr
```{r}
library(fossil)

get_simpson <- function(site, start_year){
  sp1 <- filter(bbs, site_id == site & year == start_year) %>%
    select(species_id, abundance)

  sp2 <- filter(bbs, site_id == site & year == start_year + 1) %>%
    select(species_id, abundance)

  if(dim(sp2)[1] == 0){
    return(data.frame())
  } else{   
    #need a complete list of species in both sites with zeros for non-occurrences
    comb_sp <- full_join(sp1, sp2, by = "species_id") %>%
      replace_na(list(species_id = 0, abundance.x = 0, abundance.y = 0))

    simp <- simpson(comb_sp$abundance.x, comb_sp$abundance.y)
    return(data.frame(site_id = site, year = start_year + 1, simpson = simp))
  }
}

site_year_comb <- unique(bbs[c('site_id', 'year')]) #get a dataframe of all the site and year combinations that need to be mapped over

bbs_simpson <- map2_dfr(site_year_comb$site_id, site_year_comb$year, get_simpson)
```

For loop approach
```{r}
data_path <- paste0(here('data'), '/simpson_diversity.RData')
if (file.exists(data_path)){
  simpson_file <- load(data_path)
}else{

  site_list <- unique(bbs$site_id)

  #initialize for loops
  simpson_list <- list()
  rownum <- 1

  for (i in 1:length(site_list)) {
    site_data <- filter(bbs, site_id == site_list[i])
    site_years <- unique(site_data$year)

    for (j in 1:length(site_years)) {
      if (j != length(site_years)) { #this will skip the last year which doesn't have a next year to compare to
        sp1 <- filter(site_data, year == site_years[j]) %>%
          select(species_id, abundance)

        sp2 <- filter(site_data, year == site_years[j + 1]) %>%
          select(species_id, abundance)

        comb_sp <- full_join(sp1, sp2, by = "species_id") %>%
          replace_na(list(species_id = 0, abundance.x = 0, abundance.y = 0))

        simp <- simpson(comb_sp$abundance.x, comb_sp$abundance.y)
        simpson_list[[rownum]] <- data.frame(site_id = site_list[i], year = site_years[j+1], simpson = simp)

        rownum <- rownum + 1
      }
    }
  }

  bbs_simpson <- bind_rows(simpson_list)
  save(bbs_simpson, file = data_path)

  pb_upload(data_path)
}
```

Master dataframe of all the metrics
```{r}
metrics <- FD %>%
  select(-starts_with("CWM"), -qual.FRic, -sing.sp) %>%
  left_join(bbs_simpson, by = c("year", "site_id")) %>%
  left_join(yearly_delta, by = c("year", "site_id", "nbsp"))
```

Plotting!
```{r}
sites <- unique(metrics$site_id)
metrics %>%
  filter(site_id %in% sites[1:10]) %>%
  ggplot(aes(year, nbsp, color = as.factor(site_id))) + geom_line() #+
    #theme(legend.position = "none")
```

```{r}
metrics %>%
  group_by(year) %>%
  summarise(ave = mean(nbsp)) %>%
  mutate(ave_5 = RcppRoll::roll_mean(ave, n = 5, fill = NA)) %>%
  ggplot(aes(year, ave_5)) + geom_line()
```

### 2021-09-17 analysis/figures/conceptual_fig.Rmd
Clean up folders

--- 2021-07-26 analysis/conceptual_fig.Rmd
@@
         axis.ticks.x=element_blank(),
-        axis.title.y=element_text(angle = 0, vjust = 0.5),
+        axis.title.y=element_text(angle = 0, vjust = 0.5, face = "bold"),
         axis.text.y=element_blank(),
         axis.ticks.y=element_blank(),
-        plot.title = element_text(hjust = 0.5)) +
+        plot.title = element_text(hjust = 0.5, face = "bold")) +
   ggtitle("Low")
@@
         axis.ticks.y=element_blank(),
-        plot.title = element_text(hjust = 0.5)) +
+        plot.title = element_text(hjust = 0.5, face = "bold")) +
   ggtitle("High")
@@
         axis.ticks.x=element_blank(),
-        axis.title.y=element_text(angle = 0, vjust = 0.5),
+        axis.title.y=element_text(angle = 0, vjust = 0.5, face = "bold"),
         axis.text.y=element_blank(),
@@
                        hjust = -1, axis = "l", scale = 0.9, 
-                       label_x = "Trait Axis", label_y = "Abundance")
+                       #label_x = "Trait Axis", label_y = "Abundance",
+                       rel_widths = c(.5, 1, 1))
 
@@
 y.grob <- textGrob("Abundance",  
-            gp = gpar(fontsize = 14),
-            vjust = 15, rot=90)
+            gp = gpar(fontsize = 12),
+            vjust = 11, hjust = 0.25, rot=90)
 
-x.grob <- textGrob("Trait Axis", hjust = -1.25)
+x.grob <- textGrob("Trait Axis", gp = gpar(fontsize = 12), hjust = -0.5)
 

### 2021-09-20 paper/manuscript.Rmd
minor grammatical changes

@@
 Ecological communities are experiencing unprecedented change as a result of anthropogenic pressures such as climate change, land use change, and invasive species.
-Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011], and fundamental changes in some of the most well-studied systems (e.g.
-coral bleaching).
+Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011], and fundamental changes in some of the most well-studied systems [e.g.
+coral bleaching; @sully2019].
 At the local scale however, species diversity tells a different story.
@@
 And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @matuoka2020].
-On paleoecological time scales, functional composition is often conserved alongside species diversity even in the face of drastic environmental change [@barnosky2004 @jackson2015] .
+On paleoecological time scales, functional composition is often conserved alongside species diversity even in the face of drastic environmental change [@barnosky2004; @jackson2015] .
 It is therefore critical to establish whether or not functional loss is the general rule for communities on time scales relevant for environmental impacts and management action.
@@
 
-We obtained mammal, bird, and amphibian timeseries from the Biotime database, a global repository of high quality community assemblage timeseries.
+We obtained mammal, bird, and amphibian timeseries from the BioTIME database, a global repository of high quality assemblage timeseries.
 All studies included in the database follow consistent sampling protocols and represent full assemblages rather than populations of single species [@dornelas2018].
@@
 These databases include species-level means for traits that best define species' function in the community including body size, diet, and behavioral characteristics.
-For the full list of traits included in the analysis for each taxa see Table 1.
+For the full list of traits included in the analysis for each taxon see Table 1.
 <!-- limitations of species level trait data? -->
 
-In order to ensure taxonomic consistency across datasets, Biotime species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (citation), obtained through the taxadb R package [@norman2020].
+In order to ensure taxonomic consistency across datasets, BioTIME species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (citation), obtained through the taxadb R package [@norman2020].
 If more than one species in the assemblage data resolved to the same identifier, observations were considered the same species.
@@
 Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package [@laliberté2010].
-Here we report functional richness (FRic), functional evenness (FEve), and functional diveregence (FDiv) which together describe three complementary characteristics of the functional space [@mason2005, @hillebrand2009].
+Here we report functional richness (FRic), functional evenness (FEve), and functional divergence (FDiv) which together describe three complementary characteristics of the functional space [@mason2005, @hillebrand2009].
 FRic assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
@@
 
-We estimated general trends for each diversity metric using a Linear Mixed Effects model with a random slope and intercept for each timeseries nested within the study.
+We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each timeseries nested within the study.
 To test for differences in trends by taxa, biome, and realm, each variable was modeled individually as random effects in models of the same structure.
@@
 
-We assessed the impact of timeseries duration and start year on study-level trends using General Linear Models with duration and start year as predictors.
+We assessed the impact of timeseries duration and start year on study-level trends using general linear models with duration and start year as predictors.
 
@@
 However, a number of gaps in best practices may be obscuring a true underlying trend.
-First, the Biotime database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
+First, the BioTIME database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
 We may simply not have data from those areas experiencing the greatest perturbation [@hughes2021].

### 2021-09-21 paper/manuscript_word.Rmd
First commit, PNAS version

--- 2021-09-16 paper/manuscript.Rmd
@@
 ---
-title: "Evidence of widespread maintenance of functional composition in vertebrate communities"
-author: 
-  - name: "Kari E A Norman"
-    affiliation: ucb
-  - name: "Perry de Valpine"
-    affiliation: ucb
-  - name: "Carl Boettiger"
-    affiliation: ucb, ropensci
-address:
-  - code: ucb
-    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
-  - code: ropensci
-    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA" 
-abstract: |
-
-journal: "Ecology Letters"
-date: "`r Sys.Date()`"
+title: "Manuscript doc"
+output: word_document
 bibliography: refs.bib
-layout: 3p
-header-includes:
-   - \usepackage{lineno}
-   - \linenumbers
-output: #word_document
-  rticles::pnas_article:
-  includes:
-   in_header: preamble.tex
 editor_options: 
@@
 
-*Future Work*
+#### Future Work
 
@@
 We present here the conclusion best supported by available data with a number of caveats and suggestions for next steps forward.
-
 The most pressing need is for intentional and targeted data collection efforts.
@@
 Prioritization of additional traits linked to environmental change drivers (response traits) or ecosystem function (effect traits) in particular would provide a powerful link to existing frameworks for understanding functional change [@lavorel2002].
+
+### References

### 2021-09-22 paper/manuscript.Rmd
add ref
Add abstract, introduce response/effect framework earlier.

@@
 ---
-title: "Evidence of widespread maintenance of functional composition in vertebrate communities"
+title: "Evaluating the evidence of widespread maintenance of functional composition in vertebrate communities"
 author: 
@@
     address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA" 
-abstract: |
+abstract: Despite unprecedented environmental change due to anthropogenic pressure, recent work has found no overall trend in species diversity through time accompanied by increasing turnover.
+Functional diversity provides a potentially powerful alternative approach for understanding this change in community composition by linking shifts in species identity to mechanisms of ecosystem processes.
+Here we present the first multi-taxa, multi-system analysis of functional change through time, pairing thousands of vertebrate assemblage timeseries from the BioTIME database with existing functional trait databases.
+We found no overall trend in any calculated functional diversity metric after correcting for changes in species richness, despite similar species-based patterns of constant richness with increasing turnover.
+This pattern held at the study-level where only 3 of 54 studies experienced a significant trend in at least one functional diversity metric.
+Using best available data and biodiversity synthesis practices, our results give strong evidence that across a variety of taxa, climates, and biomes, functional characteristics are maintained even in the face of significant environmental change.
+We also discuss the potential for underlying functional shifts to be obscured by current approaches and data and call for targeted data collection efforts to combat existing biases in monitoring and trait data.
 
@@
 
-Functional diversity offers a potentially powerful alternative for detecting and describing community change by providing a mechanistic link between the organisms in a system and the processes they perform [@mcgill2006].
+Functional diversity offers a potentially powerful alternative for detecting and describing community change by providing a mechanistic link between species' response to environmental change (response traits) and the processes they perform (effect traits) [@lavorel2002; @mcgill2006; @suding2008].
 By describing the functional trait space rather than species, functional diversity metrics capture the disproportionate impact of losses or gains of functionally unique species.
@@
 This restriction allows for a sufficient number of axes to capture the trait space while maintaining computational feasibility.
-Metrics incorporated weighting based on species abundance where available.
+Metrics incorporated weighting based on species abundance where available, three studies had data on presence only.
 
@@
 Intentional trait selection is critical for linking functional patterns to ecological processes [@zhu2017], however synthesis is constrained to the traits in a few taxa-specific databases.
-Prioritization of additional traits linked to environmental change drivers (response traits) or ecosystem function (effect traits) in particular would provide a powerful link to existing frameworks for understanding functional change [@lavorel2002].
+Traits would be especially beneficial if they explicitly fit into existing frameworks for linking traits to processes, including the now widely accepted response and effect framework [@lavorel2002].
 

### 2021-09-22 analysis/05_model_metrics.Rmd
Add formula to model table output

--- 2021-08-23 analysis/05_model_metrics.Rmd
@@
   tidy(fit$result) %>%
-   mutate(metric = metric_name, model = list(fit$result), model_id = model_id, warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
+   mutate(metric = metric_name, model = list(fit$result), 
+          model_id = model_id, formula = formula, 
+          warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
 }

### 2021-09-22 analysis/exporatory_scripts/BAT_explore.Rmd

--- 2020-09-15 analysis/BAT_explore.Rmd
@@
 library(tidyverse)
+```
+
+Let's also do the functions from BAT package
+```{r}
+#function for getting PCA axes with at least 95% coverage
+axes95 <- function(mat){
+  axes <- pco(mat)
+  eigen <- axes$eig
+
+  threshold <- sum(eigen[eigen > 0]) * 0.95
+
+  sums <- cumsum(eigen)
+  num_axes <- Position(function(x) x > threshold, sums)
+
+  return(num_axes)
+}
+
+get_BAT_fd <- function(data, filter_plot){
+  require(BAT)
+  require(StatMatch)
+  require(labdsv)
+
+  data <- data %>% filter(plot_id == filter_plot)
+
+  print(max(data$record_id))
+
+  species_mat <- data %>%
+    select(year, id) %>% 
+    distinct() %>%
+    arrange(year, id) %>%
+    mutate(present = 1) %>%
+    spread(id, value = present, fill = 0) %>%
+    column_to_rownames("year")
+
+  trait_mat <- data %>%
+    select(-c(taxa, record_id, day, month, year, sample_desc, id_species, 
+              latitude, longitude, sum_abundance, sum_biomass, biotimeName, 
+              traitName, scientificName)
+    ) %>%
+    distinct() %>%
+    select_if(~ length(unique(.)) > 1) %>% #remove columns that have the same value for all columns
+    mutate_at(vars(-id, -ends_with("5cat")), list(~scales::rescale(.,to=c(0, 1)))) %>% #rescale variables
+    arrange(id) %>%
+    column_to_rownames("id")
+
+  #get PCA trait matrix
+  trait_dist <- gower.dist(trait_mat)
+  trait_pco <- pco(trait_dist, k = axes95(trait_dist))
+
+  traits <- trait_pco$points
+  rownames(traits) <- rownames(trait_mat)
+
+  #getFD metrics
+  kernelFD <- kernel.alpha(comm=as.matrix(species_mat),trait=traits,abund=FALSE,method="gaussian",return.hv=TRUE)
+  hv <- kernelFD[[2]]
+
+  disp <- kernel.dispersion(comm=hv,frac=0.1,func='dissimilarity') 
+  div <- kernel.dispersion(comm=hv,frac=0.6, func='divergence')
+  even <- kernel.evenness(comm=hv)
+
+  results <- data.frame(richness = as.numeric(kernelFD[[1]]), disp, div, even) %>% 
+    mutate(plot_id = filter_plot)
+
+  return(results)
+}
+
+get_BAT_fd(biotime_traits, 376300153)
 ```

### 2021-09-22 analysis/exporatory_scripts/func_distinctiveness.Rmd

---
title: "Functional Rarity"
output: html_document
---
```{r}
library(funrar)
library(biodivTS)
```

Calculate functional distinctiveness for each species based on the species pool (defined by all species observed over the course of the timeseries)

Get data
```{r}
pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")

##	Get the gridded data locally
bt_data <- pins::pin_get("bt-traitfiltered", board = "github") %>%
  rename_with(tolower)

trait_ref <- pins::pin_get("trait-ref", board = "github")

bt_abund <- bt_data %>%
  filter(abundance_type %in% c("Count", "Density", "MeanCount"))

bt_pres <- bt_data %>%
  filter(abundance_type == "Presence/Absence")
```

```{r}
loadRData <- function(file_name){
  load(file_name)
  get(ls()[ls() != "file_name"])
}

get_ts_distinct <- function(file_path, trait_data){

  rare_comm_save <- loadRData(file_path)

  species_mat <- rare_comm_save %>%
    column_to_rownames("YEAR") %>%
    dplyr::select(-c(rarefyID, cell, type, rarefy_resamp))

  rel_species <- make_relative(as.matrix(species_mat))

  trait_dist <- get_traitMat(colnames(species_mat), trait_data)

  trait_dist <- trait_dist %>%
    mutate_if(is.character, as.factor) %>%
    compute_dist_matrix(., metric = "gower")

  distinctive <- as.data.frame(distinctiveness(pres_matrix = rel_species, dist_matrix = trait_dist)) %>%
    rownames_to_column("year") %>%
    mutate(rarefyid = unique(rare_comm_save$rarefyID), 
           rarefy_resamp = unique(rare_comm_save$rarefy_resamp))

  return(distinctive)
}

path <- here::here("data", "rarefied_samples")
files <- dir(path, "*.rda") %>%
  paste0(path, "/", .)

plan("multiprocess", workers = 120)
yearly_distinct <- furrr::future_map_dfr(files, get_ts_distinct, trait_data = trait_ref) %>%
  pivot_longer(starts_with("ITIS:"), names_to = "species", values_to = "distinctiveness") %>%
  filter(!is.na(distinctiveness))
```

Get species-pool level classifications of distinctiveness/scarcity
```{r}
sp_data <- bt_data %>%
  select(species, rarefyid, abundance, abundance_type)

## Break data down into different kinds of observations to aggregate across years
# first presence/absence
presence_sp <- sp_data %>%
  filter(abundance_type == "Presence/Absence") %>%
  select(species, rarefyid) %>%
  distinct() %>%
  mutate(abundance = 1)

# and then things that act like counts
count_sp <- sp_data %>%
  filter(abundance_type != "Presence/Absence") %>%
  group_by(rarefyid, species, abundance_type) %>%
  mutate(abundance = sum(abundance))
```

### 2021-09-22 analysis/exporatory_scripts/invasive_data.Rmd

---
title: "GBIF invasive species"
author: "Kari Norman"
date: "3/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(rgbif)
library(geonames)
library(countrycode)
library(taxadb)
```

Get GBIF data
```{r}
#get all the datasets from the Invasive Species Specialist Group
datasets <- dataset_search(publishingOrg = "cdef28b1-db4e-4c58-aa71-3c5238c2d0b5", limit = 300)
datasets_list <- datasets$data

occ <- occ_download(dataset = "b351a324-77c4-41c9-a909-f30f77268bc4")
```

Figure out which country each study is located in 
```{r}
pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
options(geonamesUsername="karinorman")

obs_data <- pins::pin_get("bt-traitfiltered", board = "github") %>%
  select(study_id, species) %>%
  distinct() %>%
  mutate(scientificName = get_names(species, "itis"),
         gbifID = get_ids(scientificName, "gbif")) %>%
  mutate(gbifID = case_when(
    scientificName == "Oreortyx pictus" ~ "GBIF:2474320",
    scientificName == "Aegithalos caudatus" ~ "GBIF:2495000",
    scientificName == "Papio anubis" ~ "GBIF:5707341",
    scientificName == "Boana bischoffi" ~ "GBIF:160065393",
    scientificName == "Boana faber" ~ "GBIF:10863437",
    scientificName == "Ololygon rizibilis" ~ "GBIF:10896287",
    scientificName == "Ololygon perpusilla" ~ "GBIF:7341662",
    scientificName == "Boana pardalis" ~ "GBIF:10842532",
    scientificName == "Ololygon littoralis" ~ "GBIF:10733643",
    scientificName == "Brachycephalus pitanga" ~ "GBIF:10846405",
    TRUE ~ gbifID
  )) %>%
  separate(gbifID, c("gbif", "ID"), remove = F) %>%
  select(-gbif)

# "Boana caipora" and "Boana albomarginata" have no GBIF ID

meta <- pins::pin_get("meta", board = "github") %>%
  select(study_id, cent_lat, cent_long) %>%
  filter(study_id %in% obs_data$study_id)

# get_GNcountry <- possibly(GNcountryCode, otherwise = NA_character_)
# 
# get_country <- function(study_id, cent_lat, cent_long){
#   #con_data <- get_GNcountry(lat = cent_lat, lng = cent_long, lang = "en", radius = 50)
#   con_data <- get_GNcountry(lat = cent_lat, lng = cent_long, radius = 50)
#   name <- ifelse(length(con_data) == 1, con_data, con_data$countryName)
#   
#   return(tibble(study_id = study_id, country = name))
# }
# 
# country_data <- pmap_dfr(meta, get_country)
# missing_cd <- country_data %>% filter(is.na(country)) %>% pull(study_id)
# 
# country_data2 <- pmap_dfr(meta %>% filter(study_id %in% missing_cd), get_country)
# missing_cd2 <- country_data2 %>% filter(is.na(country)) %>% pull(study_id)
# 
# country_data_bind <- bind_rows(country_data %>% filter(!study_id %in% country_data2$study_id),country_data2) %>%
#   mutate(country = case_when(
#   study_id == 108 ~ "Namibia",
#   study_id %in% c(166, 169) ~ "United States",
#   study_id == 172 ~ "Morocco",
#   TRUE ~ country
# )) %>%
#   mutate(code = countryname(country, destination = "genc2c")) %>%
#   mutate(code = case_when(
#     code == "CA" ~ "US",
#     code == "PR" ~ "DO",
#     TRUE ~ code
#   ))

country_data_bind <- pins::pin_get("country-data-bind", board = "github")

obs_data <- obs_data %>%
  left_join(country_data_bind)
```

Read in Invasive Data, first using rgbif
```{r}
#rgbif invasive id's 
invasive_search <- occ_data(taxonKey = na.omit(obs_data$ID), establishmentMeans = "INVASIVE;INTRODUCED", limit = 3000)
invasive_data <- map_dfr(invasive_search, ~.x$data) %>%
  select(acceptedTaxonKey, countryCode, establishmentMeans)

gbif_invasive <- obs_data %>%
  mutate(ID = as.integer(ID)) %>%
  left_join(invasive_data, by = c("ID" = "acceptedTaxonKey", "code" = "countryCode")) %>%
  distinct() %>%
  filter(!is.na(establishmentMeans))
```

Read in invasive data from hand downloading from GBIF
```{r}
#invasive data from manual download  
# invasive_files <- list.dirs(here::here("data", "invasive")) %>%
#   select(acceptedTaxonKey, countryCode, establishmentMeans, occurrenceStatus) %>% 
#   distinct()
# 
# invasive_data <- map_dfr(invasive_files[2:21], function(x){
#   taxon <- read_tsv(paste0(x, "/", "taxon.txt"))
#   profile <- read_tsv(paste0(x, "/", "speciesprofile.txt"))
#   dist <- read_tsv(paste0(x, "/", "distribution.txt")) %>%
#     select(id, countryCode)
#   
#   data <- left_join(taxon, profile) %>% left_join(dist)
#   return(data)
# }) %>%
#   mutate(cleanScientificName = clean_names(scientificName, lowercase = FALSE),
#          itisID = get_ids(cleanScientificName, "itis"))

invasive_table <- pins::pin_get("invasive-data", board = "github") %>%
  mutate(cleanScientificName = clean_names(scientificName, lowercase = FALSE),
           itisID = get_ids(cleanScientificName, "itis"),
         gbifID = get_ids(cleanScientificName, "gbif"))

manual_invasive <-  obs_data %>%
  left_join(invasive_table %>% 
              select(itisID, countryCode, ITISisInvasive = isInvasive), 
            by = c("species" = "itisID", "code" = "countryCode")) %>%
  left_join(invasive_table %>% 
              select(gbifID, countryCode, GBIFisInvasive = isInvasive), 
            by = c("species" = "gbifID", "code" = "countryCode")) %>%
  distinct() %>% 
  filter(GBIFisInvasive == "invasive" | ITISisInvasive == "invasive")

```
Only six invasive species across three studies!
"Mus musculus"      "Rattus rattus"     "Rattus norvegicus" "Sus scrofa"        "Microtus arvalis"  "Apodemus agrarius"

Substituting years:
- orig: Canada, data: United States
- orig: Puerto Rico, data: Dominican Republic

### 2021-09-22 analysis/exporatory_scripts/problem_model.Rmd

--- 2021-04-09 analysis/problem_model.Rmd
@@
   filter(!is.na(value)) %>%
-  mutate(logvalue = log(value + 1),
-         year_scaled = scale(year)) # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
+  mutate(logvalue = log(value + 1)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
+  left_join(meta, c("rarefyID" = "rarefyid"))
 
+#find rarefyID's that have inf value for some metrics and exclude
 cont_inf <- metrics %>% filter(is.infinite(value), !metric %in% c("qual.FRic",  "SES_qual.FRic")) %>% pull(rarefyID)
@@
   filter(!rarefyID %in% cont_inf, !rarefyID %in% missing_null) %>%
-  mutate(study_id = substr(rarefyID, 1, 3))
+  group_by(metric) %>%
+  mutate(study_id =  str_extract(rarefyID, "[^_]+")) %>%
+  mutate(year_scaled = scale(year), 
+         scale_center = attributes(year_scaled)$`scaled:center`, 
+         scale = attributes(year_scaled)$`scaled:scale`) %>%
+  ungroup()
 ```
@@
 model_data %>%
-  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S_int")) %>%
+  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_base")) %>%
   ggplot() +
@@
 test_data <- model_data %>% 
-  filter(metric == "FEve") %>%
+  filter(metric == "FRic") %>%
   mutate(study_id = as.factor(study_id))
@@
 #lognormal
-ln_fit <- lmer(logvalue ~ year_center + (year_center|study_id/rarefyID), data = test_data,
+ln_fit <- lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), data = test_data,
                control = lmerControl(optCtrl = list(maxit = 1e9,
@@
 
+Get p-values for blups 
 ```{r}
-options(mc.cores = parallel::detectCores())
-stan_fit <- stan_lmer(value_scale ~ year_center + (year_center|study_id/rarefyID), data = test_data, iter = 500)
+cV <- ranef(ln_fit, effects = "ran_vals", drop = TRUE)
+ranvar <- attr(cV[[1]], "postVar")
 
-launch_shinystan(stan_fit,ppd = FALSE)
+ng <- dim(ranvar)[3]
+np <- dim(ranvar)[2]
+mm <- matrix(ranvar[cbind(rep(seq(np),ng),
+             rep(seq(np),ng),
+             rep(ng,each=np))],
+       byrow=TRUE,
+       nrow=ng)
+
+sumVar <- vcov(ln_fit)[1,1]+mm[,1]
+
+year_scaled_var <- tidy(ln_fit) %>%
+  filter(term == "year_scaled") %>% 
+  mutate(var = std.error*std.error) %>%
+  pull(var)
+
+study_ests <- broom.mixed::tidy(ln_fit, effects="ran_vals") %>%
+  filter(group == "study_id", term == "year_scaled") %>% 
+  rename(cond.std.error = std.error) %>%
+  mutate(cond.var = cond.std.error*cond.std.error, 
+         var = cond.var + year_scaled_var,
+         std.error = sqrt(var),
+         upr.ci = estimate + (2*std.error),
+         lwr.ci = estimate - (2*std.error),
+         sig = case_when(
+           lwr.ci < 0 & upr.ci > 0 ~ FALSE,
+           TRUE ~ TRUE
+         ))
 ```

### 2021-09-22 analysis/figures/visualization.Rmd

--- 2021-08-23 analysis/visualization.Rmd
@@
 
-ggsave(here::here("figures/study_map.jpeg"))
+ggsave(here::here("figures/study_map.jpeg"), bg = "white")
 ```
@@
 
-ggsave2(here::here("figures/study_map_hist.jpeg"))
+ggsave2(here::here("figures/study_map_hist.jpeg"), bg = "white")
 ```
@@
 
-group_colors <- c("Global" = my_pal[1], "Polar/Temperate" = my_pal[5], "Temperate/Tropical" = my_pal[3], "Tropical" = my_pal[4], "Temperate" = my_pal[2], "Overall Mean" = "#000000")
+group_colors <- c("Global" = my_pal[1], "Polar/Temperate" = my_pal[6], "Temperate/Tropical" = my_pal[3], "Tropical" = my_pal[4], "Temperate" = my_pal[2], "Overall Mean" = "#000000")
 
@@
 
-ggsave2(here::here("figures", "3met_long.jpeg"))
+ggsave2(here::here("figures", "3met_long.jpeg"), bg = "white")
 ```

### 2021-09-23 paper/manuscript.Rmd
Initial round of edits after Ben's comments

@@
 At the local scale however, species diversity tells a different story.
-Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of a significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017] .
-<!-- potential indication of biological homogenization due to shifting ranges and/or invasive species--> While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.
+Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of a significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017].
+While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.
 
@@
 By describing the functional trait space rather than species, functional diversity metrics capture the disproportionate impact of losses or gains of functionally unique species.
-Functional diversity metrics are therefore uniquely suited for assessing the implications of community shifts in the full range of species change scenarios.
+Functional diversity metrics are therefore uniquely suited for assessing community shifts underlying even constant species richness trends.
 
@@
 
+It is critical to establish whether or not functional loss is the general rule for communities.
 While functional loss is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], it is not necessarily inevitable even in scenarios of species loss [@daz2001].
@@
 And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @matuoka2020].
-On paleoecological time scales, functional composition is often conserved alongside species diversity even in the face of drastic environmental change [@barnosky2004; @jackson2015] .
-It is therefore critical to establish whether or not functional loss is the general rule for communities on time scales relevant for environmental impacts and management action.
+On paleoecological time scales functional composition shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with impact of species extinctions on functional diversity potentially mediated by organism size [@pimiento2017].
 
 Assessments of broad-scale temporal change in functional diversity have previously been limited by a lack of functional trait data.
-<!-- and computational power? --> The majority of work has therefore focused largely on system-specific studies with traits collected in situ.
+The majority of work has therefore focused largely on system-specific studies with traits collected in situ.
 Ongoing efforts to assemble functional traits for a variety of taxa have made synthesis of existing community assemblage data and functional traits possible for the first time, providing initial insights into the ways functional diversity changes on a broad scale for specific taxa [e.g.
@@
 Here we perform the first multi-taxa, multi-system assessment of functional diversity change through time.
-We assess thousands of mammal, bird, and amphibian functional diversity timeseries to assess if systematic functional shifts are occurring and to classify the nature of those shifts.
+We assess thousands of mammal, bird, and amphibian functional diversity timeseries to determine f there is a general trend of functional change.
+We distinguish between three possible scenarios of functional change: 1) significant loss of functional diversity with accompanying shifts in other functional metrics, 2) no functional diversity loss, but significant shifts in other functional metrics, 3) maintenance of functional diversity and composition.
+
+<!-- We expect to find a general trend of functional loss underlying similar patterns of constant species richness and increasing turnover, as communities exhibit the impacts of accelerating environmental change.
+-->
 
@@
 
-In order to ensure taxonomic consistency across datasets, BioTIME species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (citation), obtained through the taxadb R package [@norman2020].
+In order to ensure taxonomic consistency across datasets, BioTIME species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (retrieved 09-15-2020 from the on-line database, <https://doi.org/10.5066/F7KH0KBK>), obtained through the taxadb R package [@norman2020; @rcoreteam2021].
 If more than one species in the assemblage data resolved to the same identifier, observations were considered the same species.
@@
 Higher values of FDiv therefore correspond to communities where many highly abundant species are on the edges of the trait space.
-<!-- use these metrics because they're common, not because of their results -->
-
-```{r metric-fi, fig.cap="Example distributions of species and abundance for a single trait axis for high and low values of our measured metrics.
-Figure is modified from a similar illustration in Hillebrand \\& Matthiessen 2009.", echo=FALSE, strip.white=TRUE}
-knitr::include_graphics(here::here("figures/conceptual_fig.jpeg"))
-```
 
@@
 We obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model with no additional covariates.
-
 We assessed the impact of timeseries duration and start year on study-level trends using general linear models with duration and start year as predictors.
+All models were executed using the *lme4* package in R and p-values were calculated using the *lmerTest* package [@bates2015; @rcoreteam2021].
 
@@
 Non-significant overall trends indicate that an equal number of studies experienced increasing and decreasing trends in those metrics.
+<!-- check this -->
 
@@
 4 studies had a significant trend in species diversity, one positive and three negative.
-Two studies experienced a significant decrease in functional evenness, and one study experienced a significant decrease in both functional richness and functional divergence.
+Two studies experienced a significant decrease in functional evenness, and one study experienced a significant decrease in both functional richness and functional divergence (Table 2).
 
@@
 Slopes for functional evenness were also significantly more positive for longer duration studies.
+See supplement for estimates and p-values for all models.
 
@@
 While the majority of the studies (\~70%) included in our data experienced significant species turnover, only three experienced a significant shift in any functional dimension.
-This suggests certain characteristics of the functional space are conserved even in the face of significant change in species identity, specifically the size of the functional space occupied by the community (FRic) and the distribution of species and individuals within that space (FEve and FDiv).
+This suggests certain characteristics of the functional space are maintained even in the face of significant change in species identity, specifically the size of the functional space occupied by the community (FRic) and the distribution of species and individuals within that space (FEve and FDiv).
 On average, species additions have similar functional characteristics as lost species and therefore maintain the structure of the functional space.
-<!-- While we did not explicitly assess how turnover rates compare to background turnover, --> Our results give strong evidence that maintenance of functional structure is not just a phenomena of paleoecological time scales but also annual time scales [@barnosky2004; @jackson2015].
+<!-- Our results give strong evidence that maintenance of functional structure is not just a phenomena of paleoecological time scales but also annual time scales [@barnosky2004; @jackson2015]. -->
 
@@
 Further, none of the 5 included studies that experienced a manual manipulation (e.g. burning, grazing exclosure, etc) experienced any significant functional trends.
+
+where are the increases/decreases occurring?
+
+any differences in some of the taxonomic groups even if not significant?
+
+<!-- how does this reconcile with the large body of studies showing that extinctions primarily affect functionally rare species?
+It seems inconsistent with their findings --> Maintenance of functional richness also goes against predictions given by work linking extinction probability to functional uniqueness.
+A large body of work demonstrates that in x,y,z communities functionally unique species are more likely to go extinct.
+At the scale of a local community, we would expect these species would also be first to be locally extirpated as population sizes decrease and ranges shrink and become more sparse.
 

### 2021-09-27 paper/manuscript.Rmd
Revamp intro, word vomit of additional discussion paragraphs

@@
 And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @matuoka2020].
-On paleoecological time scales functional composition shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with impact of species extinctions on functional diversity potentially mediated by organism size [@pimiento2017].
+On paleoecological time scales functional composition shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with large impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].
 
 Assessments of broad-scale temporal change in functional diversity have previously been limited by a lack of functional trait data.
-The majority of work has therefore focused largely on system-specific studies with traits collected in situ.
+The majority of work has therefore focused largely on system-specific studies with traits collected *in situ*.
 Ongoing efforts to assemble functional traits for a variety of taxa have made synthesis of existing community assemblage data and functional traits possible for the first time, providing initial insights into the ways functional diversity changes on a broad scale for specific taxa [e.g.
@@
 Here we perform the first multi-taxa, multi-system assessment of functional diversity change through time.
-We assess thousands of mammal, bird, and amphibian functional diversity timeseries to determine f there is a general trend of functional change.
+We focus on mammal, bird, and amphibian species as a significant subset of the world's biodiversity heavily impacted by anthropogenic change (citation).
+While examining trends in plants, invertebrates, and other vertebrate species are of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
+In order to ensure comparability across taxa in trait type and data quality we therefore we focus on Anamalia here.
+
+We assess thousands of mammal, bird, and amphibian functional diversity timeseries to determine whether or not there is a general trend of functional change.
 We distinguish between three possible scenarios of functional change: 1) significant loss of functional diversity with accompanying shifts in other functional metrics, 2) no functional diversity loss, but significant shifts in other functional metrics, 3) maintenance of functional diversity and composition.
+Based on expectation due to human impacts, we expect to find a significant functional loss with further restructuring indicated by the additional metrics.
 
@@
 -->
+
+<!-- While trait data exists for other taxa in the BioTIME database, we focus on these groups because they have accepted species-level means, comparable traits collected across taxa, and near comprehensive species coverage.
+Plants in particular have a rich history of functional trait work, but are far more specoise and suffer from biases in trait coverage that could have meaningful impacts on interpretation of trends [@fitzjohn2014].
+-->
+
+<!--These taxa represent a significant portion of the world's biodiversity and we have no prior expectation that the inclusion of additional taxa would significantly alter the picture of functional change.
+W and as such should capture general trends -->
+
+<!--We have no prior expectation based on ecological theory that broad scale functional shifts would be significantly different between taxa, as communities of all different kinds of organisms are shaped by the same processes (citation).
+-->
+
+\
+<!--if the hypothesis of global biodiversity trends is true, there is no reason to believe that it would not be evident in animals alone, and only manifest when plants are included.
+Examining trends in functional traits of plants is certainly of equal interest, but raises additional challenges (species-level means, comparable traits cross-taxa, bias).  Therefore we focus on Anamalia here.-->
 
@@
 For the full list of traits included in the analysis for each taxon see Table 1.
-<!-- limitations of species level trait data? -->
 
@@
 We did find a significant overall decrease in Jaccard similarity, indicating increasing turnover through time.
-Non-significant overall trends indicate that an equal number of studies experienced increasing and decreasing trends in those metrics.
-<!-- check this -->
+Non-significant overall trends are the result of both non-significant trends a the study level, and a balance of positive and negative trends.
 
@@
 This striking result could be a product of two possible processes, one ecological and one methodological.
-Null trends appear to give strong evidence of systemic maintenance of functional structure due to common ecological processes, however multiple limitations of current approaches in synthesis weaken the conclusions possible from our data due to potential statistical artefacts.
+Null trends appear to give strong evidence of systemic maintenance of functional structure due to common ecological processes, however multiple limitations of current approaches in synthesis could potentially be obscuring a true underlying trend.
 We discuss both options further here.
@@
 
-where are the increases/decreases occurring?
+If this is true, it has significant implications for how we approach work on biodiversity and ecosystem function.
+Very few observed communities meet the criteria for biodviersity loss as defined in BEF work (loss of species or loss of functional diversity) and the communities that lost species maintained functional diversity.
+Need to measure the right traits, but also decide what kind of shifts are important for functional change.
+But this is also good news!
+Most communities appear to be just fine from a function standpoint following best practices.
+
+No significant differences between taxa.
+Indicates that we could've looked at any taonomic subset of the data and gotten the same result.
+But what does that mean for taxa we left out?
+We can only make educated guesses based on what we know of the ecology of taxa that were left out - taxa specific papers?
 
@@
 
+do we expect different results for different taxa?
+
 <!-- how does this reconcile with the large body of studies showing that extinctions primarily affect functionally rare species?
-It seems inconsistent with their findings --> Maintenance of functional richness also goes against predictions given by work linking extinction probability to functional uniqueness.
-A large body of work demonstrates that in x,y,z communities functionally unique species are more likely to go extinct.
-At the scale of a local community, we would expect these species would also be first to be locally extirpated as population sizes decrease and ranges shrink and become more sparse.
+It seems inconsistent with their findings --> Maintenance of functional composition contradicts likely scenarios of loss predicted due to rare species extinction.
+Rare species, defined by small populations and geographic restriction, are simultaneously more likely to be functionally distinct and at higher risk for extinction [@davies2004; @harnik2012; @loiseau2020].
+Assuming local extirpation is a precursor to global extinction, communities losing functionally rare species should exhibit strong functional shifts as lost species cannot be replaced by functionally similar species [@leitão2016] .
+Instead, these communities appear to only have turnover of species with functionally redundant species in the species pool.
+<!-- good news for ecosystem function! -->
 
@@
 Intentional trait selection is critical for linking functional patterns to ecological processes [@zhu2017], however synthesis is constrained to the traits in a few taxa-specific databases.
-Traits would be especially beneficial if they explicitly fit into existing frameworks for linking traits to processes, including the now widely accepted response and effect framework [@lavorel2002].
+Traits would be especially beneficial if they explicitly incorporate existing frameworks for linking traits to processes, including the now widely accepted response and effect framework [@lavorel2002].
+
+We also need theory that links ecological process to the kinds of changes actually happening in communities.
+Do the shifts we see matter?
+Are there shifts we didn't even measure that are critical for function and resilience?
 

### 2021-09-29 paper/manuscript.Rmd
Brain dump in discussion, add citations for studies.

@@
     wrap: sentence
+
+nocite: |
+  @bakker1990, @berezovikovn.n.2004, @enemar2004, @gido2019, @hogstad1993, @holmes1986, @jahncke2006, @krivenkov.g.1991, @melnikovy.i.2000, @monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2013, @monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2014, @monitori2014, @pardieck2020, @preston1960, @svensson2006, @svensson2010, @thorn2016changes, @thorn2016response, @usfs, @vermontcenterforecostudies2015, @vickery1984, @waide2017, @williamson1983, @woehlerericj., @zakharovv.d.1998, @meditss2011, @animald, @baltics, @piropno, @rossaferes1997, @scott, @silva, @ernest2009, @friggens2008, @jalilova.b.2014, @kartzinel2014, @kaufman2019, @krefting1974, @lightfoot, @malyshevy.s.2011, @meyer2008, @nedosekinv.y.2015, @prins1990, @rocha2017, @sheftelb.i., @stapp2014, @monitori2014, @bahamas, @popacet
 ---
@@
 And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @matuoka2020].
-On paleoecological time scales functional composition shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with large impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].
+On paleoecological time scales functional composition shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with significant impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].
 
@@
 Here we perform the first multi-taxa, multi-system assessment of functional diversity change through time.
-We focus on mammal, bird, and amphibian species as a significant subset of the world's biodiversity heavily impacted by anthropogenic change (citation).
+We focus on mammal, bird, and amphibian species as a significant subset of the world's biodiversity heavily impacted by anthropogenic change.
 While examining trends in plants, invertebrates, and other vertebrate species are of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
@@
 Based on expectation due to human impacts, we expect to find a significant functional loss with further restructuring indicated by the additional metrics.
-
-<!-- We expect to find a general trend of functional loss underlying similar patterns of constant species richness and increasing turnover, as communities exhibit the impacts of accelerating environmental change.
--->
-
-<!-- While trait data exists for other taxa in the BioTIME database, we focus on these groups because they have accepted species-level means, comparable traits collected across taxa, and near comprehensive species coverage.
-Plants in particular have a rich history of functional trait work, but are far more specoise and suffer from biases in trait coverage that could have meaningful impacts on interpretation of trends [@fitzjohn2014].
--->
-
-<!--These taxa represent a significant portion of the world's biodiversity and we have no prior expectation that the inclusion of additional taxa would significantly alter the picture of functional change.
-W and as such should capture general trends -->
-
-<!--We have no prior expectation based on ecological theory that broad scale functional shifts would be significantly different between taxa, as communities of all different kinds of organisms are shaped by the same processes (citation).
--->
-
-\
-<!--if the hypothesis of global biodiversity trends is true, there is no reason to believe that it would not be evident in animals alone, and only manifest when plants are included.
-Examining trends in functional traits of plants is certainly of equal interest, but raises additional challenges (species-level means, comparable traits cross-taxa, bias).  Therefore we focus on Anamalia here.-->
 
@@
   mutate_all(linebreak, align = "l") %>% 
-  knitr::kable(format = "latex", col.names = c("Taxa", "Number of \nTimeseries", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
-               caption = "Summary of the data in the final trait database.") %>%
-  kable_classic() %>% 
-  kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
+  kableExtra::kbl(format = "markdown", col.names = c("Taxa", "Number of \nTimeseries", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
+               caption = "Summary of the data in the final trait database.") #%>%
+  #kable_classic() #%>% 
+  #kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
    #kableExtra::column_spec(2, width = "5cm") #%>%
@@
 
+All code for reproducing analyses are archived at (zenodo link).
+Data products are archived at (zenodo link).
+Raw
+
 ### Results
@@
 
-If this is true, it has significant implications for how we approach work on biodiversity and ecosystem function.
-Very few observed communities meet the criteria for biodviersity loss as defined in BEF work (loss of species or loss of functional diversity) and the communities that lost species maintained functional diversity.
+Local maintenance of functional structure has important implications for biodiversity and ecosystem function (BEF).
+BEF work links ecosystem function to changes in species and functional diversity, however very few observed communities experienced changes along either of those characteristics.
+Do we then not need to worry about loss of function in most bird, mammal and amphibian communities?
+Always the option that we didn't measure traits related to function, though in this case I think we did.
+We did not however measure fucntional redundancy, which could be important for maintaining multifunctionality [@bagousse-pinguet2019] .
+Also may not be sensitive enough to detect significant decreases in abundance for the most dominant species, which may also be important
+
+-   Traits are not grounded in theory, might be measuring the wrong ones
+
+-
+
+Even if there were unobserved communities experiencing huge shifts,
+
+meet the criteria for biodiversity loss as defined in BEF work (loss of species or loss of functional diversity) and the communities that lost species maintained functional diversity.
 Need to measure the right traits, but also decide what kind of shifts are important for functional change.
@@
 
-<!-- how does this reconcile with the large body of studies showing that extinctions primarily affect functionally rare species?
-It seems inconsistent with their findings --> Maintenance of functional composition contradicts likely scenarios of loss predicted due to rare species extinction.
+<!-- how does this reconcile with the large body of studies showing that extinctions primarily affect functionally rare species? It seems inconsistent with their findings -->
+
+Maintenance of functional composition contradicts likely scenarios of loss predicted due to rare species extinction.
 Rare species, defined by small populations and geographic restriction, are simultaneously more likely to be functionally distinct and at higher risk for extinction [@davies2004; @harnik2012; @loiseau2020].

### 2021-09-29 paper/manuscript.Rmd
chippin' away at ye olde BEF paragraph

@@
 While examining trends in plants, invertebrates, and other vertebrate species are of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
-In order to ensure comparability across taxa in trait type and data quality we therefore we focus on Anamalia here.
+In order to ensure comparability across taxa in trait type and data quality we therefore we focus on these taxa.
 
@@
 While we do not directly assess the disturbance histories of included communities, trends were consistent even for the longest running and most heavily impacted studies.
-The North American Breeding Survey for example is made up exclusively of roadside surveys and is therefore in relatively close proximity to human impact [@pardieck2020] .
+The North American Breeding Survey for example is made up exclusively of roadside surveys and is therefore in relatively close proximity to human impact [@pardieck2020].
 It is considered an authoritative dataset on the state of bird populations on the continent and underpins policy decisions about bird conservation [@rosenberg2019; @sauer2017].
@@
 
-Local maintenance of functional structure has important implications for biodiversity and ecosystem function (BEF).
+What does local maintenance of functional structure mean for the function of these communities?
+The vast majority of experimental and observational work links function to changes in functional or species diversity.
+By those criteria, very few observed communities in our dataset are in a state of concern for declining functionality.
+However, we did not we did not directly measure some aspects of community structure important for ecosystem multifunctionality [e.g.
+dispersion, rarity, abundance of dominant species; @bagousse-pinguet2021].
+While our metrics indirectly incorporate dimensions of the functional space related to those characteristics, they are likely not sensitive enough to infer deterioration (or improvement) in those those aspects of structure.
+
+<!-- has important implications for biodiversity and ecosystem function (BEF).
 BEF work links ecosystem function to changes in species and functional diversity, however very few observed communities experienced changes along either of those characteristics.
 Do we then not need to worry about loss of function in most bird, mammal and amphibian communities?
-Always the option that we didn't measure traits related to function, though in this case I think we did.
+While -->
+
+<!-- Always the option that we didn't measure traits related to function, though in this case I think we did.
 We did not however measure fucntional redundancy, which could be important for maintaining multifunctionality [@bagousse-pinguet2019] .
-Also may not be sensitive enough to detect significant decreases in abundance for the most dominant species, which may also be important
+Also may not be sensitive enough to detect significant decreases in abundance for the most dominant species, which may also be important-->
 
--   Traits are not grounded in theory, might be measuring the wrong ones
-
--
-
-Even if there were unobserved communities experiencing huge shifts,
-
-meet the criteria for biodiversity loss as defined in BEF work (loss of species or loss of functional diversity) and the communities that lost species maintained functional diversity.
+<!-- meet the criteria for biodiversity loss as defined in BEF work (loss of species or loss of functional diversity) and the communities that lost species maintained functional diversity.
 Need to measure the right traits, but also decide what kind of shifts are important for functional change.
@@
 Most communities appear to be just fine from a function standpoint following best practices.
+-->
 

### 2021-09-30 paper/manuscript.Rmd
clean up

@@
 While our metrics indirectly incorporate dimensions of the functional space related to those characteristics, they are likely not sensitive enough to infer deterioration (or improvement) in those those aspects of structure.
+<!-- more citations please -->
 
-<!-- has important implications for biodiversity and ecosystem function (BEF).
-BEF work links ecosystem function to changes in species and functional diversity, however very few observed communities experienced changes along either of those characteristics.
-Do we then not need to worry about loss of function in most bird, mammal and amphibian communities?
-While -->
+<!-- any work looking at function in communities that are stable along these axes? -->
 
-<!-- Always the option that we didn't measure traits related to function, though in this case I think we did.
-We did not however measure fucntional redundancy, which could be important for maintaining multifunctionality [@bagousse-pinguet2019] .
-Also may not be sensitive enough to detect significant decreases in abundance for the most dominant species, which may also be important-->
-
-<!-- meet the criteria for biodiversity loss as defined in BEF work (loss of species or loss of functional diversity) and the communities that lost species maintained functional diversity.
-Need to measure the right traits, but also decide what kind of shifts are important for functional change.
-But this is also good news!
-Most communities appear to be just fine from a function standpoint following best practices.
--->
+While we did not directly measure changes in rare species, our results also contradict likely scenarios of loss predicted due to rare species extinction.
+Rare species, defined by small populations and geographic restriction, are simultaneously more likely to be functionally distinct and at higher risk for extinction [@davies2004; @harnik2012; @loiseau2020].
+Assuming local extirpation is a precursor to global extinction, communities losing functionally rare species should exhibit strong functional shifts as lost species cannot be replaced by functionally similar species [@leitão2016] .
+Instead, these communities appear to only have turnover of species with functionally redundant species in the species pool.
 
@@
 We can only make educated guesses based on what we know of the ecology of taxa that were left out - taxa specific papers?
-
-any differences in some of the taxonomic groups even if not significant?
-
-do we expect different results for different taxa?
-
-<!-- how does this reconcile with the large body of studies showing that extinctions primarily affect functionally rare species? It seems inconsistent with their findings -->
-
-Maintenance of functional composition contradicts likely scenarios of loss predicted due to rare species extinction.
-Rare species, defined by small populations and geographic restriction, are simultaneously more likely to be functionally distinct and at higher risk for extinction [@davies2004; @harnik2012; @loiseau2020].
-Assuming local extirpation is a precursor to global extinction, communities losing functionally rare species should exhibit strong functional shifts as lost species cannot be replaced by functionally similar species [@leitão2016] .
-Instead, these communities appear to only have turnover of species with functionally redundant species in the species pool.
-<!-- good news for ecosystem function! -->
 

### 2021-09-30 paper/manuscript.Rmd
Minor Perry edits

@@
 Functional diversity provides a potentially powerful alternative approach for understanding this change in community composition by linking shifts in species identity to mechanisms of ecosystem processes.
-Here we present the first multi-taxa, multi-system analysis of functional change through time, pairing thousands of vertebrate assemblage timeseries from the BioTIME database with existing functional trait databases.
+Here we present the first multi-taxa, multi-system analysis of functional change through time, pairing thousands of vertebrate assemblage time series from the BioTIME database with existing functional trait databases.
 We found no overall trend in any calculated functional diversity metric after correcting for changes in species richness, despite similar species-based patterns of constant richness with increasing turnover.
@@
    - \linenumbers
-output: #word_document
-  rticles::elsevier_article:
-  includes:
-   in_header: preamble.tex
+output: 
+  rticles::arxiv_article:
+    keep_tex: true
+  word_document: default
 editor_options: 
@@
 By describing the functional trait space rather than species, functional diversity metrics capture the disproportionate impact of losses or gains of functionally unique species.
-Functional diversity metrics are therefore uniquely suited for assessing community shifts underlying even constant species richness trends.
+Functional diversity metrics are therefore particularly well suited for assessing community shifts underlying even constant species richness trends.
 
@@
 
-We assess thousands of mammal, bird, and amphibian functional diversity timeseries to determine whether or not there is a general trend of functional change.
+We assess thousands of mammal, bird, and amphibian functional diversity time series to determine whether or not there is a general trend of functional change.
 We distinguish between three possible scenarios of functional change: 1) significant loss of functional diversity with accompanying shifts in other functional metrics, 2) no functional diversity loss, but significant shifts in other functional metrics, 3) maintenance of functional diversity and composition.
@@
 
-We obtained mammal, bird, and amphibian timeseries from the BioTIME database, a global repository of high quality assemblage timeseries.
+We obtained mammal, bird, and amphibian time series from the BioTIME database, a global repository of high quality assemblage time series.
 All studies included in the database follow consistent sampling protocols and represent full assemblages rather than populations of single species [@dornelas2018].
-Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual timeseries following a standardized spatial scale.
+Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual time series following a standardized spatial scale.
 Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018] for details on how sample extents were defined), with the area of each cell set to one standard deviation away from the mean of the single extent locations.
-All samples from a study within a single cell were considered to be a single timeseries, and species abundances were combined for all samples.
+All samples from a study within a single cell were considered to be a single time series, and species abundances were combined for all samples.
 
@@
   mutate_all(linebreak, align = "l") %>% 
-  kableExtra::kbl(format = "markdown", col.names = c("Taxa", "Number of \nTimeseries", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
+  kableExtra::kbl(format = "markdown", col.names = c("Taxa", "Number of \nTime Series", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
                caption = "Summary of the data in the final trait database.") #%>%
@@
 Many studies had a variable number of samples within years.
-To account for this inconsistency in sampling effort we used sample-based rarefaction by bootstrap resampling within years for each timeseries based on the smallest number of samples in a year for that timeseries.
+To account for this inconsistency in sampling effort we used sample-based rarefaction by bootstrap resampling within years for each time series based on the smallest number of samples in a year for that time series.
 
-Our final dataset included 2,443 timeseries from 53 studies in 21 countries and 15 biomes and 13 different traits (Fig 1).
+Our final dataset included 2,443 time series from 53 studies in 21 countries and 15 biomes and 13 different traits (Fig 1).
 The earliest sample was in 1923 and the most recent was in 2016.
-Only four studies (consisting of 11 timeseries) came from Amphibian studies due to the limited availability of amphibian timeseries and low species richness values for assemblages (Table 1).
+Only four studies (consisting of 11 time series) came from Amphibian studies due to the limited availability of amphibian time series and low species richness values for assemblages (Table 1).
 For a full breakdown of studies and their characteristics, see the supplement.
 
-```{r taxa-map, fig.cap="A) Map of timeseries locations with points colored by taxa, and B) histograms of timeseries duration broken down by taxa.", echo=FALSE}
+```{r taxa-map, fig.cap="A) Map of time series locations with points colored by taxa, and B) histograms of time series duration broken down by taxa.", echo=FALSE}
 knitr::include_graphics(here::here("figures/study_map_hist.jpeg"))
@@
 
-We calculated yearly metrics of functional and species diversity for each timeseries.
+We calculated yearly metrics of functional and species diversity for each time series.
 Species-based metrics include species richness (*S*) and Jaccard similarity (*J*) as a measure of turnover.
-Jaccard similarity was calculated relative to the first observed year for a timeseries.
+Jaccard similarity was calculated relative to the first observed year for a time series.
 A negative trend in *J* would therefore indicate increasing turnover.
@@
 All continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
-The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= s^t$, where $s$ is the number of species and *t* is the number of traits.
+The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
 This restriction allows for a sufficient number of axes to capture the trait space while maintaining computational feasibility.
@@
 
-We corrected for bias in functional diversity metrics due to changes in species richness by calculating the standardized effect size (SES) for each metric from null estimates [@swenson2012].
+To assess functional change independent of species richness we calculating the standardized effect size (SES) for each metric from null estimates [@swenson2012].
 Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
 Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species.
-Species pools included all species observed for a timeseries.
-This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for that timeseries.
+Species pools included all species observed for a time series.
+This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction for that time series.
 We used these values to calculate SES using the following formula: $SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
+We then calculated the median SES estimate for each metric from all the rarefaction samples for a time series.
 Unless otherwise specified, functional diversity results are reported for standardized values for the remainder of the paper.
@@
 
-We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each timeseries nested within the study.
-To test for differences in trends by taxa, biome, and realm, each variable was modeled individually as random effects in models of the same structure.
+We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study.
+To test for trends within taxa, biome, and realm, groups we also fit models with random slope and intercept effects for each one of these factors.
 We obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model with no additional covariates.
-We assessed the impact of timeseries duration and start year on study-level trends using general linear models with duration and start year as predictors.
+BLUP's provide an estimate for the conditional mean and variance of each level of a random effect.
+from which we calculated 95% confidence intervals to determine significance of study-level slopes.
+We assessed the impact of time series duration and start year on study-level trends using general linear models with duration and start year as predictors.
 All models were executed using the *lme4* package in R and p-values were calculated using the *lmerTest* package [@bates2015; @rcoreteam2021].
 
-```{r timeseries-plot, fig.cap="Plots of timeseries-level trends with line color corresponding to climatic region, with data points in grey and the overall metric mean in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES", echo=FALSE}
+```{r timeseries-plot, fig.cap="Plots of time series-level trends with line color corresponding to climatic region, with data points in grey and the overall metric mean in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES", echo=FALSE}
 knitr::include_graphics(here::here("figures/3met_long.jpeg"))
@@
 We found no significant overall trend in species richness or functional diversity metrics (Fig 2).
-There was also no temporal trend in those metrics for studies broken down by taxa, biome or realm.
+There was also no temporal trend in those metrics within taxa, biome or realm groups.
 We did find a significant overall decrease in Jaccard similarity, indicating increasing turnover through time.
-Non-significant overall trends are the result of both non-significant trends a the study level, and a balance of positive and negative trends.
+Non-significant overall trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies was plausibly 0.
 
@@
 While we do not directly assess the disturbance histories of included communities, trends were consistent even for the longest running and most heavily impacted studies.
-The North American Breeding Survey for example is made up exclusively of roadside surveys and is therefore in relatively close proximity to human impact [@pardieck2020].
-It is considered an authoritative dataset on the state of bird populations on the continent and underpins policy decisions about bird conservation [@rosenberg2019; @sauer2017].
+The North American Breeding Survey for example is considered an authoritative dataset on the state of bird populations on the continent and underpins policy decisions about bird conservation [@pardieck2020; @rosenberg2019; @sauer2017].
 No more robust dataset exists to capture North American avian community change, yet we detected no general shifts in functional structure across the dataset.
 Further, none of the 5 included studies that experienced a manual manipulation (e.g. burning, grazing exclosure, etc) experienced any significant functional trends.
+
+We observed no significant trends within taxa No significant differences between taxa.
+Indicates that we could've looked at any taonomic subset of the data and gotten the same result.
+But what does that mean for taxa we left out?
+We can only make educated guesses based on what we know of the ecology of taxa that were left out - taxa specific papers?
 
@@
 Instead, these communities appear to only have turnover of species with functionally redundant species in the species pool.
-
-No significant differences between taxa.
-Indicates that we could've looked at any taonomic subset of the data and gotten the same result.
-But what does that mean for taxa we left out?
-We can only make educated guesses based on what we know of the ecology of taxa that were left out - taxa specific papers?
 
@@
 
-Our results illustrate a well known but increasingly pervasive problem in synthesis-based ecology of distinguishing true ecological patterns from methodological artifacts.
+Our results illustrate a well known but increasingly pervasive problem in synthesis-based ecology of distinguishing true ecological patterns from methodological limitations.
 We present here the conclusion best supported by available data with a number of caveats and suggestions for next steps forward.

### 2021-10-01 paper/manuscript.Rmd
still going

@@
 We observed no significant trends within taxa No significant differences between taxa.
-Indicates that we could've looked at any taonomic subset of the data and gotten the same result.
+Indicates that we could've looked at any taxonomic subset of the data and gotten the same result.
 But what does that mean for taxa we left out?
@@
 What does local maintenance of functional structure mean for the function of these communities?
-The vast majority of experimental and observational work links function to changes in functional or species diversity.
-By those criteria, very few observed communities in our dataset are in a state of concern for declining functionality.
+The vast majority of experimental and observational work links declines in function to declines in functional or species diversity [@brose2016; @cadotte2011; @duffy2007] .
+By those criteria, very few observed communities in our dataset are in a state of concern for loss of functionality.
 However, we did not we did not directly measure some aspects of community structure important for ecosystem multifunctionality [e.g.
 dispersion, rarity, abundance of dominant species; @bagousse-pinguet2021].
-While our metrics indirectly incorporate dimensions of the functional space related to those characteristics, they are likely not sensitive enough to infer deterioration (or improvement) in those those aspects of structure.
+While our metrics indirectly incorporate dimensions of the functional space related to those community characteristics, they are likely not sensitive enough to infer deterioration (or improvement).
 <!-- more citations please -->
+
+-   experiments vs broad scale observational studies - do we look at FD in both?
 
@@
 Rare species, defined by small populations and geographic restriction, are simultaneously more likely to be functionally distinct and at higher risk for extinction [@davies2004; @harnik2012; @loiseau2020].
-Assuming local extirpation is a precursor to global extinction, communities losing functionally rare species should exhibit strong functional shifts as lost species cannot be replaced by functionally similar species [@leitão2016] .
-Instead, these communities appear to only have turnover of species with functionally redundant species in the species pool.
+Locally, communities losing functionally rare species should exhibit strong functional shifts as lost species cannot be replaced by functionally similar species [@leitão2016] .
+Instead, we observed patterns consistent with species replacement by functionally redundant species from the species pool.
 

### 2021-10-04 analysis/05_model_metrics.Rmd
Write out csv with study metadata
Include raw metric values

--- 2021-09-17 analysis/05_model_metrics.Rmd
@@
 pins::pin(model_data, board = "github")
+```
+
+Write out file with all necessary metadata for the supplement.
+```{r}
+meta %>% 
+  filter(study_id %in% model_data$study_id) %>% 
+  select(-c(link_id, date_study_added, sample_desc_name)) %>% 
+  write_csv(here::here("paper", "study_metadata.csv"))
 ```
@@
 ```{r}
-raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base"), 
+raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "FRic", "FEve", "FDiv", "Jaccard_base"), 
                          base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
@@
 ```
-
 #### Process Model Output

### 2021-10-05 paper/manuscript.Rmd
Include uncorrected metrics

@@
 We then calculated the median SES estimate for each metric from all the rarefaction samples for a time series.
-Unless otherwise specified, functional diversity results are reported for standardized values for the remainder of the paper.
-
 <!-- description of edge case where community = species pool -->
@@
 We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study.
-To test for trends within taxa, biome, and realm, groups we also fit models with random slope and intercept effects for each one of these factors.
+To test for trends within taxa, biome, and realm we also fit models with random slope and intercept effects for each one of these factors.
 We obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model with no additional covariates.
-BLUP's provide an estimate for the conditional mean and variance of each level of a random effect.
-from which we calculated 95% confidence intervals to determine significance of study-level slopes.
+BLUP's provide an estimate for the conditional mean and variance of each level of a random effect from which we calculated 95% confidence intervals to determine significance of study-level slopes.
 We assessed the impact of time series duration and start year on study-level trends using general linear models with duration and start year as predictors.
@@
 
-We found no significant overall trend in species richness or functional diversity metrics (Fig 2).
-There was also no temporal trend in those metrics within taxa, biome or realm groups.
+We found no significant overall trend in species richness or observed or null-corrected functional diversity metrics (Fig 2).
+There was also no temporal trend in those metrics within taxa, biome or realm factor levels.
 We did find a significant overall decrease in Jaccard similarity, indicating increasing turnover through time.
@@
 
-At the study level, only 7 of 54 studies experienced a significant trend for a metric other than Jaccard similarity.
-4 studies had a significant trend in species diversity, one positive and three negative.
-Two studies experienced a significant decrease in functional evenness, and one study experienced a significant decrease in both functional richness and functional divergence (Table 2).
+At the study level, 4 studies experienced a significance trend and only 10 of 54 studies for uncorrected metrics and 4 of 54 studies for corrected metrics experienced a significant trend for a metric other than Jaccard similarity (Table 2).
+Trends for all but one study were inconsistent between corrected and uncorrected metrics, with significance for one or the other.
 
@@
 
-data.frame(sign = c("+", "-"), S = c("1", "3"), `Jaccard Similarity` = c("0", "37"), FRic = c("0", "1"), FEve = c("0", "2"), FDiv = c("0", "1")) %>%
+data.frame(sign = c("+", "-"), S = c("1", "3"), `Jaccard Similarity` = c("0", "37"), FRic = c("2", "6"), FEve = c("1", "0"), FDiv = c("0", "1"), `SES FRic` = c("0", "1"), `SES FEve` = c("0", "2"), `SES FDiv` = c("0", "1")) %>%
   column_to_rownames("sign") %>%
@@
 We found no evidence of systemic functional richness loss or functional change.
-A lack of trend for data broken down by realm, biome, and taxonomic group gives further evidence that directional functional change is absent from all systems observed in our dataset.
+A lack of trend within realm, biome, or taxonomic group gives further evidence that directional functional change is absent from all systems observed in our dataset.
 
 This striking result could be a product of two possible processes, one ecological and one methodological.
-Null trends appear to give strong evidence of systemic maintenance of functional structure due to common ecological processes, however multiple limitations of current approaches in synthesis could potentially be obscuring a true underlying trend.
+Null trends appear to give strong evidence of systemic maintenance of functional structure due to common ecological processes, however multiple limitations of current approaches in synthesis could potentially be obscuring a true underlying global trend.
 We discuss both options further here.

### 2021-10-05 paper/manuscript.Rmd
more incorporating unstandardized metrics.

@@
 
-We found no significant overall trend in species richness or observed or null-corrected functional diversity metrics (Fig 2).
+We found no significant overall trend in species richness or functional diversity metrics (observed or standardized) (Fig 2).
 There was also no temporal trend in those metrics within taxa, biome or realm factor levels.
@@
 
-At the study level, 4 studies experienced a significance trend and only 10 of 54 studies for uncorrected metrics and 4 of 54 studies for corrected metrics experienced a significant trend for a metric other than Jaccard similarity (Table 2).
-Trends for all but one study were inconsistent between corrected and uncorrected metrics, with significance for one or the other.
+At the study level, 4 studies experienced a significance trend in species richness and only 10 of 54 studies for observed metrics and 3 of 54 studies for standardized metrics experienced a significant trend for a metric other than Jaccard similarity (Table 2).
+Most significant trends for observed functional metrics are in functional diversity and disappeared after standardization, indicating that richness increases were likely due to changes in the number of species.
+<!-- Trends for all but one study were inconsistent between corrected and uncorrected metrics, with significance for one or the other.
+--> Hypothesis testing for study-level trends is likely affected by multiple testing issues and some trends identified as significant are therefore potentially erroneous.
+Rather than interpreting changes in specific studies, we present these results as a general picture of the small number of studies experiencing a trend and highlight that even those studies are likely an overestimate of the number of significant trends.
 
@@
 Slopes for functional evenness were also significantly more positive for longer duration studies.
+Results were consistent between standardized and observed metrics with exception of functional evenness, which was negatively related to duration for observed data and positively related for standardized data.
 See supplement for estimates and p-values for all models.
@@
 Communities demonstrated almost universal maintenance of functional composition.
-While the majority of the studies (\~70%) included in our data experienced significant species turnover, only three experienced a significant shift in any functional dimension.
-This suggests certain characteristics of the functional space are maintained even in the face of significant change in species identity, specifically the size of the functional space occupied by the community (FRic) and the distribution of species and individuals within that space (FEve and FDiv).
+While the majority of the studies (\~70%) included in our data experienced significant species turnover, only three (for standardized metrics) experienced a significant shift in any functional dimension.
+<!-- General consistency between observed and standardized metrics underlines that changes in species richness were not a major factor determining functional composition changes.--> This suggests certain characteristics of the functional space are maintained even in the face of significant change in species identity, specifically the size of the functional space occupied by the community (FRic) and the distribution of species and individuals within that space (FEve and FDiv).
 On average, species additions have similar functional characteristics as lost species and therefore maintain the structure of the functional space.
-<!-- Our results give strong evidence that maintenance of functional structure is not just a phenomena of paleoecological time scales but also annual time scales [@barnosky2004; @jackson2015]. -->
+<!-- Our results give strong evidence that maintenance of functional structure is not just a phenomena of paleoecological time scales but also annual time scales [@barnosky2004; @jackson2015].
+-->
 
@@
 However, a number of gaps in best practices may be obscuring a true underlying trend.
-First, the BioTIME database is not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
+First, the BioTIME database, while the most comprehensive data source of time series available, is limited in temporal and geographic scope.
+Most time series span only a few years (Figure 1) and may not provide the statistical power necessary to detect trends.
+The database is also not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
 We may simply not have data from those areas experiencing the greatest perturbation [@hughes2021].

### 2021-10-06 analysis/05_model_metrics.Rmd
add interaction models

@@
     filter(metric == metric_name) %>%
-    mutate(year_scaled = scale(year)) %>%
+    mutate(year_scaled = scale(year),
+           duration_scaled = scale(duration, scale = FALSE),
+           startyear_scaled = scale(startyear, scale = FALSE)) %>%
     qlmer(formula = formula, data = ., ...)
@@
 ```
+
+Models with interactions
+```{r}
+int_raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "FRic", "FEve", "FDiv", "Jaccard_base"), 
+                         base_dur = "value ~ year_scaled + duration_scaled*year_scaled + (year_scaled|study_id/rarefyID)",
+                         taxa_dur = "value ~ year_scaled + duration_scaled*year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)",
+                         realm_dur = "value ~ year_scaled + duration_scaled*year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|realm)",
+                         climate_dur = "value ~ year_scaled + duration_scaled*year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)",
+                         base_srt = "value ~ year_scaled + startyear_scaled*year_scaled + (year_scaled|study_id/rarefyID)",
+                         taxa_srt = "value ~ year_scaled + startyear_scaled*year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)",
+                         realm_srt = "value ~ year_scaled + startyear_scaled*year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|realm)",
+                         climate_srt = "value ~ year_scaled + startyear_scaled*year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)") %>%
+  pivot_longer(cols = c(base_dur, taxa_dur, realm_dur, climate_dur, base_srt, taxa_srt, realm_srt, climate_srt), 
+               names_to = "model_id", values_to = "formula")
+
+# int_log_value_mods <- tibble(metric_name = c("S"), 
+#                          base = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
+#                          taxa = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)",
+#                          realm = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|realm)",
+#                          climate = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)") %>%
+#   pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
+
+# mods <- bind_rows(raw_value_mods, log_value_mods) %>% 
+#   select(metric_name, formula, model_id)
+
+#Fit models with base settings
+int_mod_results <- pmap_dfr(int_raw_value_mods, get_model, data = model_data)
+
+#get the models that had convergence warnings
+bad_models <- int_mod_results %>%
+  select(metric, model_id, warning) %>% 
+  filter(!is.na(warning)) %>% 
+  distinct() %>%
+  left_join(int_raw_value_mods, by = c("metric" = "metric_name", "model_id")) 
+```
+
 #### Process Model Output

### 2021-10-06 analysis/05_model_metrics.Rmd
covariates as fixed effects

--- 2021-10-04 analysis/05_model_metrics.Rmd
@@
 ```{r}
+# raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "FRic", "FEve", "FDiv", "Jaccard_base"), 
+#                          base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
+#                          taxa = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)",
+#                          realm = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|realm)",
+#                          climate = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)") %>%
+#   pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
+# 
+# log_value_mods <- tibble(metric_name = c("S"), 
+#                          base = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
+#                          taxa = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)",
+#                          realm = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|realm)",
+#                          climate = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)") %>%
+#   pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
+
 raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "FRic", "FEve", "FDiv", "Jaccard_base"), 
                          base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
-                         taxa = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)",
-                         realm = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|realm)",
-                         climate = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)") %>%
+                         taxa = "value ~ year_scaled + as.factor(taxa) + (year_scaled|study_id/rarefyID)",
+                         realm = "value ~ year_scaled + as.factor(realm) + (year_scaled|study_id/rarefyID)",
+                         climate = "value ~ year_scaled + as.factor(climate) + (year_scaled|study_id/rarefyID)") %>%
   pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
@@
                          base = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
-                         taxa = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)",
-                         realm = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|realm)",
-                         climate = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)") %>%
+                         taxa = "logvalue ~ year_scaled + as.factor(taxa) + (year_scaled|study_id/rarefyID)",
+                         realm = "logvalue ~ year_scaled + as.factor(realm) + (year_scaled|study_id/rarefyID)",
+                         climate = "logvalue ~ year_scaled + as.factor(climate) + (year_scaled|study_id/rarefyID)") %>%
   pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")

### 2021-10-07 paper/manuscript.Rmd
At table of model estimates.

@@
 knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE, out.width = "\\textwidth")
+options(knitr.kable.NA = '')
 
@@
   mutate_all(linebreak, align = "l") %>% 
-  kableExtra::kbl(format = "markdown", col.names = c("Taxa", "Number of \nTime Series", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
-               caption = "Summary of the data in the final trait database.") #%>%
-  #kable_classic() #%>% 
-  #kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
+  kableExtra::kable(format = "latex", col.names = c("Taxa", "Number of \nTime Series", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
+               caption = "Summary of the data in the final trait database.") %>%
+  kable_classic() %>% 
+  kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
    #kableExtra::column_spec(2, width = "5cm") #%>%
@@
 We did find a significant overall decrease in Jaccard similarity, indicating increasing turnover through time.
-Non-significant overall trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies was plausibly 0.
+Non-significant overall trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies was plausibly 0 (Table 2).
 
-At the study level, 4 studies experienced a significance trend in species richness and only 10 of 54 studies for observed metrics and 3 of 54 studies for standardized metrics experienced a significant trend for a metric other than Jaccard similarity (Table 2).
-Most significant trends for observed functional metrics are in functional diversity and disappeared after standardization, indicating that richness increases were likely due to changes in the number of species.
+```{r echo=FALSE}
+read_csv(here::here("figures/man_model_table.csv")) %>%
+  select(metric, everything()) %>%
+  mutate(metric = ifelse(metric == "Jaccard_base", "Jaccard", metric)) %>%
+  arrange(metric) %>%
+  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness (S) and standardized metrics.
+Additional model estimates can be found in the supplement.") %>%
+  kable_classic() %>% 
+  kable_styling(latex_options= c("scale_down", "hold_position"))
+```
+
+At the study level, 4 studies experienced a significance trend in species richness and only 10 of 54 studies for observed metrics and 3 of 54 studies for standardized metrics experienced a significant trend for a metric other than Jaccard similarity (Table 3).
+Most significant trends for observed functional metrics are in functional richness and disappeared after standardization, indicating that richness increases were likely due to changes in the number of species.
 <!-- Trends for all but one study were inconsistent between corrected and uncorrected metrics, with significance for one or the other.
@@
 On average, species additions have similar functional characteristics as lost species and therefore maintain the structure of the functional space.
-<!-- Our results give strong evidence that maintenance of functional structure is not just a phenomena of paleoecological time scales but also annual time scales [@barnosky2004; @jackson2015].
--->
+<!-- Our results give strong evidence that maintenance of functional structure is not just a phenomena of paleoecological time scales but also annual time scales [@barnosky2004; @jackson2015]. -->
 
@@
 By those criteria, very few observed communities in our dataset are in a state of concern for loss of functionality.
-However, we did not we did not directly measure some aspects of community structure important for ecosystem multifunctionality [e.g.
+However, we did not directly measure some aspects of community structure important for ecosystem multifunctionality [e.g.
 dispersion, rarity, abundance of dominant species; @bagousse-pinguet2021].
 While our metrics indirectly incorporate dimensions of the functional space related to those community characteristics, they are likely not sensitive enough to infer deterioration (or improvement).
-<!-- more citations please -->
+<!-- more citations please --> This highlights the need for further development of theory linking different aspects of functional change to specific implications for function.
 

### 2021-10-08 paper/manuscript.Rmd
BEF paragraph! At long last!

@@
 
-We observed no significant trends within taxa No significant differences between taxa.
-Indicates that we could've looked at any taxonomic subset of the data and gotten the same result.
-But what does that mean for taxa we left out?
-We can only make educated guesses based on what we know of the ecology of taxa that were left out - taxa specific papers?
-
-What does local maintenance of functional structure mean for the function of these communities?
-The vast majority of experimental and observational work links declines in function to declines in functional or species diversity [@brose2016; @cadotte2011; @duffy2007] .
-By those criteria, very few observed communities in our dataset are in a state of concern for loss of functionality.
-However, we did not directly measure some aspects of community structure important for ecosystem multifunctionality [e.g.
-dispersion, rarity, abundance of dominant species; @bagousse-pinguet2021].
-While our metrics indirectly incorporate dimensions of the functional space related to those community characteristics, they are likely not sensitive enough to infer deterioration (or improvement).
-<!-- more citations please --> This highlights the need for further development of theory linking different aspects of functional change to specific implications for function.
-
--   experiments vs broad scale observational studies - do we look at FD in both?
-
-<!-- any work looking at function in communities that are stable along these axes? -->
-
 While we did not directly measure changes in rare species, our results also contradict likely scenarios of loss predicted due to rare species extinction.
@@
 Instead, we observed patterns consistent with species replacement by functionally redundant species from the species pool.
+
+What does local maintenance of functional structure mean for ecosystem function?
+The vast majority of experimental and observational work links declines in function to declines in functional or species diversity [@brose2016; @cadotte2011; @duffy2007].
+By those criteria very few communities in our dataset are in a state of concern for loss of functionality.
+However, the dimensions of functional space most important for ecosystem function are still a topic of on going debate, and at least some known aspects important for multifunctionality were not measured here [e.g.
+dispersion, rarity, abundance of dominant species; @bagousse-pinguet2021].
+The fact that we observed so many communities maintaining structure across the most commonly used metrics for linking biodiversity to function calls into question how previous work translates to natural communities.
+Metrics need to be both closely linked to changes in ecosystem function and also experiencing shifts in natural communities to be meaningful.
+
+<!-- We observed no significant trends within taxa.
+Even Amphibians, considered highly sensitive to human impact, maintained functional structure through time.
+Indicates that we could've looked at any taxonomic subset of the data and gotten the same result.
+But what does that mean for taxa we left out?
+We can only make educated guesses based on what we know of the ecology of taxa that were left out - taxa specific papers?
+-->
 

### 2021-10-08 paper/manuscript.Rmd
Spruce up the estimates table.

@@
 At the local scale however, species diversity tells a different story.
-Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of a significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017].
+Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of a significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017; @li2020].
 While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.
@@
 
-In order to ensure taxonomic consistency across datasets, BioTIME species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (retrieved 09-15-2020 from the on-line database, <https://doi.org/10.5066/F7KH0KBK>), obtained through the taxadb R package [@norman2020; @rcoreteam2021].
+In order to ensure taxonomic consistency across datasets, BioTIME species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (retrieved 09-15-2020 from the on-line database, <https://doi.org/10.5066/F7KH0KBK>), obtained through the *taxadb* R package [@norman2020; @rcoreteam2021].
 If more than one species in the assemblage data resolved to the same identifier, observations were considered the same species.
@@
 We then calculated the median SES estimate for each metric from all the rarefaction samples for a time series.
-<!-- description of edge case where community = species pool -->
+SES estimates can be interpreted as how much of the functional characteristic (richness, evenness, divergence) was observed beyond what was expected by chance for a community of that species richness.
 
@@
   select(metric, everything()) %>%
-  mutate(metric = ifelse(metric == "Jaccard_base", "Jaccard", metric)) %>%
+  mutate(metric = ifelse(metric == "Jaccard_base", "Jaccard", metric),
+         group = case_when(
+           group == "rarefyID:study_id" ~ "timeseries:study",
+           group == "study_id" ~ "study",
+           TRUE ~ group),
+         term = case_when(
+           term == "(Intercept)" ~ "Intercept",
+           term == "year_scaled" ~ "Year",
+           term == "sd__(Intercept)" ~ "Intercept sd",
+           term == "cor__(Intercept).year_scaled" ~ "Intercept Year corr",
+           term == "sd__year_scaled" ~ "Year sd",
+           term == "sd__Observation" ~ "Observation sd",
+           TRUE ~ term
+         )) %>%
   arrange(metric) %>%
@@
 Additional model estimates can be found in the supplement.") %>%
+  collapse_rows(columns = 1, valign = "middle") %>%
   kable_classic() %>% 
@@
 
-We also need theory that links ecological process to the kinds of changes actually happening in communities.
+<!-- We also need theory that links ecological process to the kinds of changes actually happening in communities.
 Do the shifts we see matter?
 Are there shifts we didn't even measure that are critical for function and resilience?
+-->
 

### 2021-10-08 paper/manuscript.Rmd
Minor refrence mistakes.

@@
 Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011], and fundamental changes in some of the most well-studied systems [e.g.
-coral bleaching; @sully2019].
+coral bleaching, @sully2019].
 At the local scale however, species diversity tells a different story.
@@
 Ongoing efforts to assemble functional traits for a variety of taxa have made synthesis of existing community assemblage data and functional traits possible for the first time, providing initial insights into the ways functional diversity changes on a broad scale for specific taxa [e.g.
-fish @trindade-santos2020, birds @jarzyna2016; @barnagaud2017].
+fish, @trindade-santos2020; birds, @jarzyna2016; @barnagaud2017].
 However, to date there has been no cross-taxa assessment of temporal functional change for a broad geographic and taxonomic extent.
@@
 While examining trends in plants, invertebrates, and other vertebrate species are of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
-In order to ensure comparability across taxa in trait type and data quality we therefore we focus on these taxa.
+In order to ensure comparability across taxa in trait type and data quality we therefore focus on these taxa.
 
@@
 Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package [@laliberté2010].
-Here we report functional richness (FRic), functional evenness (FEve), and functional divergence (FDiv) which together describe three complementary characteristics of the functional space [@mason2005, @hillebrand2009].
+Here we report functional richness (FRic), functional evenness (FEve), and functional divergence (FDiv) which together describe three complementary characteristics of the functional space [@mason2005; @hillebrand2009].
 FRic assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
@@
 
-To assess functional change independent of species richness we calculating the standardized effect size (SES) for each metric from null estimates [@swenson2012].
+To assess functional change independent of species richness we calculated the standardized effect size (SES) for each metric from null estimates [@swenson2012].
 Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
@@
 
-At the study level, 4 studies experienced a significance trend in species richness and only 10 of 54 studies for observed metrics and 3 of 54 studies for standardized metrics experienced a significant trend for a metric other than Jaccard similarity (Table 3).
+At the study level, 4 studies experienced a significant trend in species richness and only 10 of 54 studies for observed metrics and 3 of 54 studies for standardized metrics experienced a significant trend for a metric other than Jaccard similarity (Table 3).
 Most significant trends for observed functional metrics are in functional richness and disappeared after standardization, indicating that richness increases were likely due to changes in the number of species.
@@
 By those criteria very few communities in our dataset are in a state of concern for loss of functionality.
-However, the dimensions of functional space most important for ecosystem function are still a topic of on going debate, and at least some known aspects important for multifunctionality were not measured here [e.g.
-dispersion, rarity, abundance of dominant species; @bagousse-pinguet2021].
+However, the dimensions of functional space most important for ecosystem function are still a topic of ongoing debate, and at least some known aspects important for multifunctionality were not measured here [e.g.
+dispersion, rarity, abundance of dominant species, @bagousse-pinguet2021].
 The fact that we observed so many communities maintaining structure across the most commonly used metrics for linking biodiversity to function calls into question how previous work translates to natural communities.

### 2021-10-11 paper/manuscript.Rmd
Minor typos, add citation, fix sig figs in model estimates table.

@@
 The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
-This restriction allows for a sufficient number of axes to capture the trait space while maintaining computational feasibility.
-Metrics incorporated weighting based on species abundance where available, three studies had data on presence only.
+This restriction allows for a sufficient number of axes to capture the trait space while maintaining computational feasibility [@blonder2018].
+Metrics incorporated weighting based on species abundance where available (three studies were presence only).
 
@@
            group == "study_id" ~ "study",
+           group == "Residual" ~ "residual",
            TRUE ~ group),
@@
          )) %>%
+  mutate(across(c(estimate, std.error, statistic, df, p.value), round, 2), 
+         p.value = ifelse(p.value == 0, "<0.001", p.value)) %>%
   arrange(metric) %>%

### 2021-10-11 paper/manuscript.Rmd
significance statement.

@@
 We assessed the impact of time series duration and start year on study-level trends using general linear models with duration and start year as predictors.
-All models were executed using the *lme4* package in R and p-values were calculated using the *lmerTest* package [@bates2015; @rcoreteam2021].
+All models were executed using the *lme4* package in R and p-values were calculated using the *lmerTest* package with a significance level of p \< 0.05 [@bates2015; @rcoreteam2021].
 

### 2021-10-12 paper/manuscript.Rmd
Add clarification of what kinds of traits we used and why.

@@
 Functional diversity provides a potentially powerful alternative approach for understanding this change in community composition by linking shifts in species identity to mechanisms of ecosystem processes.
-Here we present the first multi-taxa, multi-system analysis of functional change through time, pairing thousands of vertebrate assemblage time series from the BioTIME database with existing functional trait databases.
+Here we present the first multi-taxa, multi-system analysis of functional change through time, pairing thousands of vertebrate assemblage time series from the BioTIME database with existing functional trait representitive of a species' functional role.
 We found no overall trend in any calculated functional diversity metric after correcting for changes in species richness, despite similar species-based patterns of constant richness with increasing turnover.
 This pattern held at the study-level where only 3 of 54 studies experienced a significant trend in at least one functional diversity metric.
-Using best available data and biodiversity synthesis practices, our results give strong evidence that across a variety of taxa, climates, and biomes, functional characteristics are maintained even in the face of significant environmental change.
+Using best available data and biodiversity synthesis practices, our results give strong evidence that across a variety of taxa, climates, and biomes, functional characteristics are maintained even in the face of significant environmental and community change.
 We also discuss the potential for underlying functional shifts to be obscured by current approaches and data and call for targeted data collection efforts to combat existing biases in monitoring and trait data.
@@
 In order to ensure comparability across taxa in trait type and data quality we therefore focus on these taxa.
+Traits were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].
 
@@
 We gathered trait data from the Elton Trait Database [mammals and birds, @wilman2014] and Amphibio [amphibians, @oliveira2017].
-These databases include species-level means for traits that best define species' function in the community including body size, diet, and behavioral characteristics.
+These databases include species-level means for traits that best represent species' multifaceted function in the community including body size, diet, and behavioral characteristics.
 For the full list of traits included in the analysis for each taxon see Table 1.

### 2021-10-12 paper/manuscript.Rmd
Clarification of amphibians
Add info about trophic downgrading

@@
 On paleoecological time scales functional composition shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with significant impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].
+Some losses of functional diversity are indisputable on both paleocological and contemporary timescales such as continued trophic downgrading due to loss of large-bodied mammals, but implications of those losses for local diversity patterns are less clear [@estes2011; @smith2018] .
 
@@
 Only four studies (consisting of 11 time series) came from Amphibian studies due to the limited availability of amphibian time series and low species richness values for assemblages (Table 1).
+Amphibians are of particular concern due to impacts of habitat loss and pollution [@gibbons2000], so we include the data we have while acknowledging that general inference for Amphibians as a taxa is not possible with the timeseries available.
 For a full breakdown of studies and their characteristics, see the supplement.

### 2021-10-13 analysis/04_collate_rarefied_null.Rmd
Add description at the top.
typo
Add in CWM for continuous variables.

@@
 ---
+
+This script reads in all the rarefacation sample files, gets the median value for each rarefyID and year, and then merges with null model statistics.
 
@@
 
+#read in CWM metrics
+path <- here::here("data", "rarefied_metrics", "cwm")
+filelist <-  dir(path, "*.rda") %>%
+  paste0(path, "/", .)
+
+cwm_metrics <- purrr::map_dfr(filelist, function(x) {
+  load(x)
+  rarefyID <- str_match(x,"_cell\\s*(.*?)\\s*_sample")[,2]
+
+  if(dim(cwm_metrics)[1] == 0){
+    return(data.frame())
+  }else{
+
+    cwm_metrics <- cwm_metrics$CWM %>%
+      rownames_to_column("YEAR") %>%
+      mutate(YEAR = as.integer(YEAR)) %>%
+      left_join(cwm_metrics %>% select(-CWM)) %>%
+      mutate(rarefyID = rarefyID)
+
+    return(cwm_metrics)
+  }
+})
+
 ## get null models
@@
 ```{r}
-rarefied_metrics <- full_join(species_metrics, fd_metrics,
+rarefied_metrics <- full_join(species_metrics, fd_metrics %>% select(-CWM),
                               c("YEAR", "cell", "rarefy_resamp", "rarefyID")) %>%
+  full_join(cwm_metrics, c("YEAR", "cell", "rarefy_resamp", "rarefyID")) %>%
   filter(S != 1) %>% # remove samples that had one species as they result in strange metric values
@@
 ##	calculate the medians for all the metrics for studies that were rarefied
+## exclude categorical variables - they don't make sense as medians
 metric_cols <- ungroup(rarefied_metrics) %>%
-  select(-c(SamplePool, SampleN, num_years, duration, startYear, endYear, CWM, type)) %>%
+  select(-c(SamplePool, SampleN, num_years, duration, startYear, endYear, type,
+            diet_5cat, pelagicspecialist, diurnal, nocturnal, crepuscular, dir, lar)) %>%
   pivot_longer(-c(rarefyID, rarefied, YEAR, cell, rarefy_resamp), names_to = "metric") %>%

### 2021-10-14 analysis/05_model_metrics.Rmd
Save out table for manuscript
exclude some columns from estimates table

@@
 pins::pin(metric_model_table, board = "github")
+```
+
+Write out table of model estimates for manuscript
+```{r}
+metric_model_table %>% 
+  filter(model_id == "base", metric %in% c("S", "Jaccard_base", "SES_FRic", "SES_FEve", "SES_FDiv")) %>% 
+  select(-c(model, model_id, warning, formula, statistic, df)) %>% 
+  write_csv(here::here("figures", "man_model_table.csv"))
 ```

### 2021-10-15 paper/manuscript.Rmd
Minor changes
Response to Ben comments

@@
          )) %>%
-  mutate(across(c(estimate, std.error, statistic, df, p.value), round, 2), 
+  mutate(across(c(estimate, std.error, p.value), round, 2), 
          p.value = ifelse(p.value == 0, "<0.001", p.value)) %>%
@@
 Rare species, defined by small populations and geographic restriction, are simultaneously more likely to be functionally distinct and at higher risk for extinction [@davies2004; @harnik2012; @loiseau2020].
-Locally, communities losing functionally rare species should exhibit strong functional shifts as lost species cannot be replaced by functionally similar species [@leitão2016] .
-Instead, we observed patterns consistent with species replacement by functionally redundant species from the species pool.
+Locally, communities losing functionally rare species should exhibit strong functional shifts as lost species can eventually no longer be replaced by functionally similar species [@leitão2016].
+While we did not have a large enough time window in most time series to capture the complete extinction process, observed patterns were instead consistent with species replacement by functionally redundant species from the species pool.
 
@@
 By those criteria very few communities in our dataset are in a state of concern for loss of functionality.
-However, the dimensions of functional space most important for ecosystem function are still a topic of ongoing debate, and at least some known aspects important for multifunctionality were not measured here [e.g.
+However, shifts in metrics are only relevant if the underlying traits are those most critical for ecosystem function.
+We were limited in this analysis to the traits available rather than those with strong empirical links to function.
+Similarly, the dimensions of functional space most important for ecosystem function are still a topic of ongoing debate, and at least some known aspects important for multifunctionality were not measured here [e.g.
 dispersion, rarity, abundance of dominant species, @bagousse-pinguet2021].
-The fact that we observed so many communities maintaining structure across the most commonly used metrics for linking biodiversity to function calls into question how previous work translates to natural communities.
+Still, the fact that we observed so many communities maintaining structure across the most commonly used metrics for linking biodiversity and function calls into question how previous work translates to natural communities.
 Metrics need to be both closely linked to changes in ecosystem function and also experiencing shifts in natural communities to be meaningful.
@@
 The database is also not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
-We may simply not have data from those areas experiencing the greatest perturbation [@hughes2021].
-We do however emphasize that even disturbed communities maintain functional structure [@edwards2013; @matuoka2020], and the inclusion of more heavily impacted sites would not necessarily change the result.
+We may simply not have data from those areas experiencing the greatest perturbation [@hughes2021], particularly scenarios of conversion to urban, human-dominated landscapes.
+While evidence from other work shows even disturbed communities can maintain functional structure [@edwards2013; @matuoka2020], these results should not be interpreted as evidence of low functional impact in areas of heavy human disturbance.
 
@@
 
-Our results illustrate a well known but increasingly pervasive problem in synthesis-based ecology of distinguishing true ecological patterns from methodological limitations.
-We present here the conclusion best supported by available data with a number of caveats and suggestions for next steps forward.
-The most pressing need is for intentional and targeted data collection efforts.
+Here we make a significant first step in establishing a general trend for functional diversity through time across a variety of taxa and systems.
+We present the conclusion best supported by available data and acknowledge that it is still too early to confidently distinguish between true ecological pattern and methodological limitations.
+The most pressing next step is for intentional and targeted data collection efforts.
 We join others in the call for increased monitoring in under sampled areas and continued efforts to centralize existing data sources [@gonzalez2016; @hughes2021; @vellend2017].
@@
 Intentional trait selection is critical for linking functional patterns to ecological processes [@zhu2017], however synthesis is constrained to the traits in a few taxa-specific databases.
-Traits would be especially beneficial if they explicitly incorporate existing frameworks for linking traits to processes, including the now widely accepted response and effect framework [@lavorel2002].
-
-<!-- We also need theory that links ecological process to the kinds of changes actually happening in communities.
-Do the shifts we see matter?
-Are there shifts we didn't even measure that are critical for function and resilience?
--->
+Trait collection should explicitly consider existing frameworks for linking traits to processes [e.g.
+the response and effect framework @lavorel2002] to facillitate clear ecological interpretation of potential functional changes.
 

### 2021-10-18 analysis/05_model_metrics.Rmd
Add interaction models to get contrasts between levels and within level slopes.

--- 2021-10-06 analysis/05_model_metrics.Rmd
@@
   filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "Jaccard_next",
-                       "FRic", "FEve", "FDiv", "S"),
+                       "FRic", "FEve", "FDiv", "S",
+                       "bodymass_value"),
          !rarefyID %in% missing_null) %>%
@@
 
-  fit <- data %>%
+  fit_data <- data %>%
     filter(metric == metric_name) %>%
-    mutate(year_scaled = scale(year)) %>%
-    qlmer(formula = formula, data = ., ...)
+    mutate(year_scaled = scale(year),
+           duration_scaled = scale(duration, scale = FALSE),
+           startyear_scaled = scale(startyear, scale = FALSE),
+           taxa = as.factor(taxa),
+           realm = as.factor(realm),
+           climate = as.factor(climate))
 
+    fit <- qlmer(formula = formula, data = fit_data, ...)
+
+  if(!model_id %in% c("base", "base_dur", "base_srt")){
+    level_slopes <- get_fixed_slopes(metric = metric_name, model = fit$result, model_id = model_id, model_data = fit_data) %>%
+      mutate(effect = "group_slope", group = model_id) %>%
+      rename(std.error = SE, estimate = year_scaled.trend, term = model_id)
+
+    int_contrasts <- get_int_contrasts(metric = metric_name, model = fit$result, model_id = model_id, model_data = fit_data) %>%
+      mutate(group = model_id, effect = "contrast") %>%
+      rename(std.error = SE, statistic = t.ratio, term = contrast) %>%
+      select(-year_scaled)
+
+    df <- tidy(fit$result) %>% 
+      bind_rows(level_slopes, int_contrasts) %>%
+      mutate(metric = metric_name, model = list(fit$result), model_id = model_id, warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
+  }else{
   tidy(fit$result) %>%
@@
   }
+}
 
+get_int_contrasts <- function(metric, model, model_id, model_data){
+  formula <- paste("pairwise ~", model_id, "| year_scaled")
+  emm <- emmeans(model,  mode = "satterth", lmerTest.limit = 20105, specs = as.formula(formula), data = model_data)
+
+  return(as_tibble(emm$contrasts))
+}
+
+get_fixed_slopes <- function(metric, model, model_id, model_data){
+  formula <- paste("~", model_id)
+  fit_trends <- emtrends(model, specs = as.formula(formula), var="year_scaled",  mode = "satterth", lmerTest.limit = 20105, data = model_data) %>% as_tibble()
+
+  return(fit_trends)
+}
 #get_model(data = model_data, metric_name = "SES_FRic", "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)", model_id = "base")
@@
 ```{r}
-# raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "FRic", "FEve", "FDiv", "Jaccard_base"), 
-#                          base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
-#                          taxa = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)",
-#                          realm = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|realm)",
-#                          climate = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)") %>%
-#   pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
-# 
-# log_value_mods <- tibble(metric_name = c("S"), 
-#                          base = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
-#                          taxa = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)",
-#                          realm = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|realm)",
-#                          climate = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)") %>%
-#   pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
-
 raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "FRic", "FEve", "FDiv", "Jaccard_base"), 
                          base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
-                         taxa = "value ~ year_scaled + as.factor(taxa) + (year_scaled|study_id/rarefyID)",
-                         realm = "value ~ year_scaled + as.factor(realm) + (year_scaled|study_id/rarefyID)",
-                         climate = "value ~ year_scaled + as.factor(climate) + (year_scaled|study_id/rarefyID)") %>%
+                         taxa = "value ~ year_scaled * taxa + (year_scaled|study_id/rarefyID)",
+                         realm = "value ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
+                         climate = "value ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
   pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
@@
                          base = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
-                         taxa = "logvalue ~ year_scaled + as.factor(taxa) + (year_scaled|study_id/rarefyID)",
-                         realm = "logvalue ~ year_scaled + as.factor(realm) + (year_scaled|study_id/rarefyID)",
-                         climate = "logvalue ~ year_scaled + as.factor(climate) + (year_scaled|study_id/rarefyID)") %>%
+                         taxa = "logvalue ~ year_scaled * taxa + (year_scaled|study_id/rarefyID)",
+                         realm = "logvalue ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
+                         climate = "logvalue ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
   pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
@@
 
-#Turn of REML for these models 
+#Turn off REML for these models, and they fit! 
 bad_mod_results <- pmap_dfr(bad_models %>% select(-warning), get_model, data = model_data, REML = FALSE)
 
-#One model still won't converge
-bad_models2 <- bad_mod_results %>%
-  select(metric, model_id, warning) %>% 
-  filter(!is.na(warning)) %>% 
-  distinct() %>%
-  left_join(mods, by = c("metric" = "metric_name", "model_id"))
-
-#Fit with a different algorithm
-bad_mod_results2 <- pmap_dfr(bad_models2 %>% select(-warning), get_model, data = model_data, control = lmerControl(optimizer = "bobyqa"))
-
 metric_model_table <- bind_rows(mod_results %>% filter(is.na(warning)), 
-          bad_mod_results %>% filter(is.na(warning)),
-          bad_mod_results2 %>% filter(is.na(warning)))
+          bad_mod_results %>% filter(is.na(warning)))
 
@@
 ```
+
+Models with interactions
+```{r}
+int_raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "FRic", "FEve", "FDiv", "Jaccard_base"), 
+                         base_dur = "value ~ year_scaled + duration_scaled*year_scaled + (year_scaled|study_id/rarefyID)",
+                         base_srt = "value ~ year_scaled + startyear_scaled*year_scaled + (year_scaled|study_id/rarefyID)") %>%
+  pivot_longer(cols = c(base_dur, taxa_dur, realm_dur, climate_dur, base_srt, taxa_srt, realm_srt, climate_srt), 
+               names_to = "model_id", values_to = "formula")
+
+int_log_value_mods <- tibble(metric_name = c("S"),
+                         base_dur = "logvalue ~ year_scaled + duration_scaled*year_scaled + (year_scaled|study_id/rarefyID)",
+                         base_srt = "logvalue ~ year_scaled + startyear_scaled*year_scaled + (year_scaled|study_id/rarefyID)") %>%
+  pivot_longer(cols = c(base_dur, taxa_dur, realm_dur, climate_dur, base_srt, taxa_srt, realm_srt, climate_srt), 
+               names_to = "model_id", values_to = "formula")
+
+# mods <- bind_rows(raw_value_mods, log_value_mods) %>% 
+#   select(metric_name, formula, model_id)
+
+#Fit models with base settings
+int_mod_results <- pmap_dfr(bind_rows(int_raw_value_mods, int_log_value_mods), get_model, data = model_data)
+
+#get the models that had convergence warnings
+bad_models <- int_mod_results %>%
+  select(metric, model_id, warning) %>% 
+  filter(!is.na(warning)) %>% 
+  distinct() %>%
+  left_join(int_raw_value_mods, by = c("metric" = "metric_name", "model_id")) 
+```
+
 #### Process Model Output
@@
 ```
-
-Pairwise correlation analysis
-```{r}
-library(psych)
-#slope_corr <- corr.test(slopes_wide %>% ungroup() %>% select(starts_with("SES"), "S", "Jaccard_base"))
-
-slope_corr_bootstrap <- corCi(slopes_wide %>% ungroup() %>% select(starts_with("SES"), "S", "Jaccard_base"))
-```
-
-Clustering 
-```{r}
-library(clValid)
-scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)
-
-scale_slopes <- slopes_wide %>%  
-  drop_na() %>%
-  select(study_id, starts_with("SES"), S, Jaccard_base) %>%
-  column_to_rownames("study_id") %>%
-  mutate(across(.fns = scale2)) %>%
-  as.matrix()
-
-clmethods <- c("hierarchical","kmeans","pam")
-intern <- clValid(scale_slopes, nClust = 2:6, clMethods = clmethods, validation = "internal")
-stab <- clValid(scale_slopes, nClust = 2:6, clMethods = clmethods, validation = "stability")
-
-summary(intern)
-optimalScores(stab)
-
-#get the hierarchical clustering object
-clust <- intern@clusterObjs$hierarchical
-cat5 <- cutree(clust, k = 5)
-cat3 <- cutree(clust, k = 3)
-
-slopes_clust <- slopes_wide %>% 
-  ungroup() %>%
-  left_join(as.data.frame(cat3) %>%
-  rownames_to_column("study_id")) %>%
-  left_join(as.data.frame(cat5) %>%
-  rownames_to_column("study_id"))
-```
-
-```{r}
-slopes_clust %>% 
-  gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S", "Jaccard_base") %>% {
-  ggplot(., aes(x = .xvalue, y = .yvalue, color = as.factor(cat5))) +
-      geom_point() + 
-      #geom_smooth(method = 'lm') +
-      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
-      scale_color_brewer(type = 'qual')
-  }
-```
-
-####Figures
-
-Pairwise plots of study-level slopes
-```{r}
-#pairwise plots using GGally
-ggpairs(slopes_wide %>% 
-          select(-c("study_id", "FDiv", "FEve", "FRic")))
-
-#pairwise plots using GGPlot
-gatherpairs <- function(data, ..., 
-                        xkey = '.xkey', xvalue = '.xvalue',
-                        ykey = '.ykey', yvalue = '.yvalue',
-                        na.rm = FALSE, convert = FALSE, factor_key = FALSE) {
-  vars <- quos(...)
-  xkey <- enquo(xkey)
-  xvalue <- enquo(xvalue)
-  ykey <- enquo(ykey)
-  yvalue <- enquo(yvalue)
-
-  data %>% {
-    cbind(gather(., key = !!xkey, value = !!xvalue, !!!vars,
-                 na.rm = na.rm, convert = convert, factor_key = factor_key),
-          select(., !!!vars)) 
-  } %>% gather(., key = !!ykey, value = !!yvalue, !!!vars,
-               na.rm = na.rm, convert = convert, factor_key = factor_key)
-}
-
-slopes_wide %>% 
-  gatherpairs( "SES_FDiv", "SES_FEve", "SES_FRic", "S") %>% {
-  ggplot(., aes(x = .xvalue, y = .yvalue)) +
-      geom_point() + 
-      #geom_smooth(method = 'lm') +
-      facet_wrap(.xkey ~ .ykey, ncol = length(unique(.$.ykey)), scales = 'free', labeller = label_both) +
-      scale_color_brewer(type = 'qual')
-  }
-
-slopes_wide %>%
-  select(-c("FDiv", "FEve", "FRic")) %>%
-  pivot_longer(-one_of("study_id", "S"), names_to = "metric") %>%
-  ggplot(aes(x = S, y = value)) +
-  geom_point() +
-  facet_wrap(~metric) +
-  xlim(-1.3, 1.3) +
-  ylim(-0.6, 0.6) +
-    theme_classic() +
-    geom_hline(yintercept = 0) +
-    geom_vline(xintercept = 0) + 
-  ylab("Metric Slope") +
-  xlab("Richness Slope")
-
-slopes_wide %>%
-  select(-c("FDiv", "FEve", "FRic")) %>%
-  pivot_longer(-one_of("study_id", "SES_FRic"), names_to = "metric") %>%
-  ggplot(aes(x = SES_FRic, y = value)) +
-  geom_point() +
-  facet_wrap(~metric) +
-  xlim(-.25, .25) +
-  ylim(-0.6, 0.6) +
-    theme_classic() +
-    geom_hline(yintercept = 0) +
-    geom_vline(xintercept = 0) + 
-  ylab("Metric Slope") +
-  xlab("Functional Richness Slope")
-```

### 2021-10-18 analysis/exporatory_scripts/problem_model.Rmd
Add interaction models to get contrasts between levels and within level slopes.

@@
 
-#get metadata
-meta <- pins::pin_get("meta", board = "github") %>%
-  select(study_id, realm, climate, habitat, protected_area, biome_map)
-
-meta <- pins::pin_get("bt-traitfiltered", board = "github") %>%
-  select(study_id, rarefyid, taxa) %>% distinct() %>%
-  left_join(meta)
-
-#get metric data
-metrics <- pins::pin_get("rarefied-metrics", board = "github") %>%
-  # pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
-  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
-  filter(!is.na(value)) %>%
-  mutate(logvalue = log(value + 1)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
-  left_join(meta, c("rarefyID" = "rarefyid"))
-
-#find rarefyID's that have inf value for some metrics and exclude
-cont_inf <- metrics %>% filter(is.infinite(value), !metric %in% c("qual.FRic",  "SES_qual.FRic")) %>% pull(rarefyID)
-
-#get rarefyID's that don't have enough null samples
-missing_null <- bind_rows(metrics %>% 
-  filter(n_missing_nulls == 1, commplete_null_samps == FALSE),
-  metrics %>% filter(n_missing_nulls > 100)) %>%
-  pull(rarefyID) %>%
-  unique()
-
-model_data <- metrics %>% 
-  filter(!rarefyID %in% cont_inf, !rarefyID %in% missing_null) %>%
-  group_by(metric) %>%
-  mutate(study_id =  str_extract(rarefyID, "[^_]+")) %>%
-  mutate(year_scaled = scale(year), 
-         scale_center = attributes(year_scaled)$`scaled:center`, 
-         scale = attributes(year_scaled)$`scaled:scale`) %>%
-  ungroup()
+model_data <- pins::pin_get("model-data", board = "github")
 ```
@@
 ```{r}
-model_data %>%
-  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_base")) %>%
-  ggplot() +
-  geom_histogram(aes(x = logvalue),
-                 bins = 60) +
-  facet_wrap(~metric, scales = "free") +
-  labs(title = "Distribution of Log values")
-
-model_data %>%
-  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_base")) %>%
-  ggplot() +
-  geom_histogram(aes(x = value),
-                 bins = 60) +
-  facet_wrap(~metric, scales = "free") +
-  labs(title = "Distribution of raw values")
+# model_data %>%
+#   filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_base")) %>%
+#   ggplot() +
+#   geom_histogram(aes(x = logvalue),
+#                  bins = 60) +
+#   facet_wrap(~metric, scales = "free") +
+#   labs(title = "Distribution of Log values")
+# 
+# model_data %>%
+#   filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_base")) %>%
+#   ggplot() +
+#   geom_histogram(aes(x = value),
+#                  bins = 60) +
+#   facet_wrap(~metric, scales = "free") +
+#   labs(title = "Distribution of raw values")
 ```
@@
   filter(metric == "FRic") %>%
-  mutate(study_id = as.factor(study_id))
+  mutate(year_scaled = scale(year),
+           duration_scaled = scale(duration, scale = FALSE),
+           startyear_scaled = scale(startyear, scale = FALSE),
+         taxa = as.factor(taxa))
+```
 
+Different options for getting between group samples, fixed effects (test differences between intercept):
+```{r}
+library(emmeans)
+
+#fixed effects contrasts
+fit_fixed <- lmer(formula = "value ~ year_scaled*taxa + (year_scaled|study_id/rarefyID)", data = test_data)
+m_emm <-  emmeans(fit_fixed,  mode = "satterth", lmerTest.limit = 20105, pairwise ~ taxa | year_scaled)
+
+fit_cont <- as_tibble(m_emm$contrasts)
+fit_trends <- emtrends(fit_fixed, ~ taxa, var="year_scaled",  mode = "satterth", lmerTest.limit = 20105) %>% as_tibble()
+
+interactions::interact_plot(fit_fixed, pred = year_scaled, modx = taxa, plot.points = TRUE)
+```
+
+As a random effect:
+```{r}
+fit_rand <- lmer(formula = "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|taxa)", data = test_data)
+```
+
+```{r}
 #lognormal
-ln_fit <- lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), data = test_data,
-               control = lmerControl(optCtrl = list(maxit = 1e9,
-                                                    maxfun = 1e9,
-                                                    xtol_abs = 1e-11,
-                                                    ftol_abs = 1e-11,
-                                                    maxeval = 1e9)))
+ln_fit <- lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), data = test_data)
 

### 2021-10-19 analysis/05_model_metrics.Rmd
Get contrasts for slopes not for intercept differences.

@@
   if(!model_id %in% c("base", "base_dur", "base_srt")){
-    level_slopes <- get_fixed_slopes(metric = metric_name, model = fit$result, model_id = model_id, model_data = fit_data) %>%
+    slopes <- get_fixed_slopes(metric = metric_name, model = fit$result, model_id = model_id, model_data = fit_data)
+
+    group_slopes <- as_tibble(slopes$emtrends) %>%
       mutate(effect = "group_slope", group = model_id) %>%
@@
 
-    int_contrasts <- get_int_contrasts(metric = metric_name, model = fit$result, model_id = model_id, model_data = fit_data) %>%
-      mutate(group = model_id, effect = "contrast") %>%
-      rename(std.error = SE, statistic = t.ratio, term = contrast) %>%
-      select(-year_scaled)
+    slope_constrasts <- as_tibble(slopes$contrasts) %>%
+      mutate(effect = "contrast", group = model_id) %>%
+      rename(std.error = SE, term = contrast, statistic = t.ratio)
 
     df <- tidy(fit$result) %>% 
-      bind_rows(level_slopes, int_contrasts) %>%
+      bind_rows(group_slopes, slope_constrasts) %>%
       mutate(metric = metric_name, model = list(fit$result), model_id = model_id, warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
@@
 
-get_int_contrasts <- function(metric, model, model_id, model_data){
-  formula <- paste("pairwise ~", model_id, "| year_scaled")
-  emm <- emmeans(model,  mode = "satterth", lmerTest.limit = 20105, specs = as.formula(formula), data = model_data)
-
-  return(as_tibble(emm$contrasts))
-}
-
 get_fixed_slopes <- function(metric, model, model_id, model_data){
-  formula <- paste("~", model_id)
-  fit_trends <- emtrends(model, specs = as.formula(formula), var="year_scaled",  mode = "satterth", lmerTest.limit = 20105, data = model_data) %>% as_tibble()
+  formula <- paste("pairwise", "~", model_id)
+  fit_trends <- emtrends(model, specs = as.formula(formula), var="year_scaled",  mode = "satterth", lmerTest.limit = 20105, data = model_data)
 

### 2021-10-19 analysis/exporatory_scripts/problem_model.Rmd
Get contrasts for slopes not for intercept differences.

@@
 fit_fixed <- lmer(formula = "value ~ year_scaled*taxa + (year_scaled|study_id/rarefyID)", data = test_data)
+
+#intercept contrasts
 m_emm <-  emmeans(fit_fixed,  mode = "satterth", lmerTest.limit = 20105, pairwise ~ taxa | year_scaled)
+fit_cont <- as_tibble(m_emm$contrasts)
 
-fit_cont <- as_tibble(m_emm$contrasts)
-fit_trends <- emtrends(fit_fixed, ~ taxa, var="year_scaled",  mode = "satterth", lmerTest.limit = 20105) %>% as_tibble()
+#slope contrasts and within group slope estimates
+fit_trends <- emtrends(fit_fixed, pairwise ~ taxa, var="year_scaled",  mode = "satterth", lmerTest.limit = 20105) %>% as_tibble()
 

### 2021-10-19 analysis/05_model_metrics.Rmd
Add a column that checks p-value and confidence interval significance.

@@
 library(lmerTest)
+library(emmeans)
 ```
@@
 metric_model_table <- bind_rows(mod_results %>% filter(is.na(warning)), 
-          bad_mod_results %>% filter(is.na(warning)))
+          bad_mod_results %>% filter(is.na(warning))) %>%
+  mutate(significant = case_when(
+    p.value > 0.05 | lower.CL < 0 & upper.CL > 0 ~ FALSE,
+    !is.na(p.value) | !is.na(lower.CL) ~ TRUE,
+    TRUE ~ NA
+  ))
 
@@
 model_df <- metric_model_table %>%
+  filter(model_id == "base") %>%
   select(metric, model, model_id) %>%

### 2021-10-25 paper/manuscript.Rmd
Ben comments
Change modeling details for categorical study-level variables.

@@
 We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study.
-To test for trends within taxa, biome, and realm we also fit models with random slope and intercept effects for each one of these factors.
-We obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model with no additional covariates.
+We obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model.
 BLUP's provide an estimate for the conditional mean and variance of each level of a random effect from which we calculated 95% confidence intervals to determine significance of study-level slopes.
+To test for trends within and between different levels of taxa, biome, and realm we fit separate models with each of those covariates added as a predictor to the original model structure.
+We estimated within level slopes and calculated between level contrasts using the *emmeans* package [@lenth2021] .
 We assessed the impact of time series duration and start year on study-level trends using general linear models with duration and start year as predictors.
-All models were executed using the *lme4* package in R and p-values were calculated using the *lmerTest* package with a significance level of p \< 0.05 [@bates2015; @rcoreteam2021].
+All models were executed using the *lme4* package in R and p-values were calculated using the *lmerTest* package with a significance level of p \< 0.05 [@bates2015; @kuznetsova2017; @rcoreteam2021].
 
@@
 We found no significant overall trend in species richness or functional diversity metrics (observed or standardized) (Fig 2).
-There was also no temporal trend in those metrics within taxa, biome or realm factor levels.
 We did find a significant overall decrease in Jaccard similarity, indicating increasing turnover through time.
 Non-significant overall trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies was plausibly 0 (Table 2).
+Trends for different taxa, biomes, or realms were also non-significant with the exception of a significantly increasing slope for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing standardized functional richness slope for freshwater studies.
+Slopes for freshwater studies were also significantly smaller than slopes for terrestrial studies.
 

### 2021-10-25 analysis/02_rarefy_timeseries.Rmd

@@
 ```
+
+Get cwm's that were excluded
+Functions to calculate functional diversity metrics for the samples
+```{r}
+#helper function to load file
+loadRData <- function(file_name){
+  load(file_name)
+  get(ls()[ls() != "file_name"])
+}
+
+#function to calculate functional diversity metrics for a single sample community
+calc_cwm_mets <- function(file_path, traits){
+  #library(biodivTS)
+  #load community sample
+  rare_comm_save <- loadRData(file_path)
+
+  years <- rare_comm_save$YEAR
+
+  #get species matrix
+  species_mat <- rare_comm_save %>%
+    tibble::column_to_rownames("YEAR") %>%
+    dplyr::select(-c(rarefyID, cell, type, rarefy_resamp))
+
+  # get trait matrix for the species in the sample
+  traits <- biodivTS::get_traitMat(colnames(species_mat), trait_data = traits)
+
+  # create FD function that returns an empty dataframe instead of erroring if FD can't be calculated
+  get_FD_safe <- possibly(biodivTS::get_FD, otherwise = data_frame())
+
+  # calculate functional diversity for different data types
+  if(unique(rare_comm_save$type) == "count"){
+    FD_mets <- get_FD_safe(species_mat = species_mat, trait_mat = traits, year_list = years,
+                           data_id = unique(rare_comm_save$rarefyID), 
+                           samp_id = unique(rare_comm_save$rarefy_resamp),
+                           w.abun = TRUE, m = "min", corr = "cailliez",
+                           calc.FDiv = FALSE)
+  } else{
+    FD_mets <- get_FD_safe(species_mat = species_mat, trait_mat = traits, year_list = years,
+                           data_id = unique(rare_comm_save$rarefyID), 
+                           samp_id = unique(rare_comm_save$rarefy_resamp),
+                           w.abun = FALSE, m = "min", corr = "cailliez",
+                           calc.FDiv = FALSE)
+  }
+
+  #check if FD metrics could be calculated, otherwise return empty dataframe
+  if (is.null(dim(FD_mets))){
+
+    # combine univariate and turnover metrics
+    cwm_metrics <- FD_mets$fd_met %>%
+      select(year, starts_with("CWM")) %>%
+      mutate(cell = unique(rare_comm_save$cell), rarefy_resamp = unique(rare_comm_save$rarefy_resamp)) %>%
+      rename(YEAR = year)
+
+  }else{ cwm_metrics <- tibble() }
+
+  #save out dataframe of rarefied metrics
+  path <- here::here("data", "rarefied_metrics", "cwm")
+  dir.create(path)
+  save(cwm_metrics, file=paste0(path, "/", unique(rare_comm_save$type),
+                                      "_cell", unique(rare_comm_save$rarefyID),
+                                      "_sample", unique(rare_comm_save$rarefy_resamp), ".rda"))
+
+}
+
+#get list of rarefaction sample files
+path <- here::here("data", "rarefied_samples")
+files <- dir(path, "*.rda") %>%
+  paste0(path, "/", .)
+
+#parallelized mapping across files to calculate functional diversity metrics
+future::plan("multiprocess", workers = 30)
+furrr::future_map(files, calc_cwm_mets, traits = trait_ref)
+```

### 2021-10-25 analysis/05_model_metrics.Rmd

@@
 
+Write out table of model estimates for manuscript
+```{r}
+metric_model_table %>% 
+  filter(model_id == "base", metric %in% c("S", "Jaccard_base", "SES_FRic", "SES_FEve", "SES_FDiv")) %>% 
+  select(-c(model, model_id, warning, formula, statistic, df)) %>% 
+  write_csv(here::here("figures", "man_model_table.csv"))
+```
+
 Models with interactions

### 2021-10-25 paper/manuscript.Rmd
Minor typos.

@@
 Functional diversity provides a potentially powerful alternative approach for understanding this change in community composition by linking shifts in species identity to mechanisms of ecosystem processes.
-Here we present the first multi-taxa, multi-system analysis of functional change through time, pairing thousands of vertebrate assemblage time series from the BioTIME database with existing functional trait representitive of a species' functional role.
+Here we present the first multi-taxa, multi-system analysis of functional change through time, pairing thousands of vertebrate assemblage time series from the BioTIME database with existing functional traits representitive of a species' functional role.
 We found no overall trend in any calculated functional diversity metric after correcting for changes in species richness, despite similar species-based patterns of constant richness with increasing turnover.
@@
 To test for trends within and between different levels of taxa, biome, and realm we fit separate models with each of those covariates added as a predictor to the original model structure.
-We estimated within level slopes and calculated between level contrasts using the *emmeans* package [@lenth2021] .
+We estimated within level slopes and calculated between level contrasts using the *emmeans* package [@lenth2021].
 We assessed the impact of time series duration and start year on study-level trends using general linear models with duration and start year as predictors.
-All models were executed using the *lme4* package in R and p-values were calculated using the *lmerTest* package with a significance level of p \< 0.05 [@bates2015; @kuznetsova2017; @rcoreteam2021].
+All models were executed using the *lme4* package in R and p-values were calculated using the *lmerTest* package with a significance level of \$\\alpha \< 0.05\$ [@bates2015; @kuznetsova2017; @rcoreteam2021].
 
@@
 Trait collection should explicitly consider existing frameworks for linking traits to processes [e.g.
-the response and effect framework @lavorel2002] to facillitate clear ecological interpretation of potential functional changes.
+the response and effect framework @lavorel2002] to facilitate clear ecological interpretation of potential functional changes.
 

### 2021-10-25 paper/manuscript.Rmd
Minor changes to language about model interpretation.

@@
 Trends for different taxa, biomes, or realms were also non-significant with the exception of a significantly increasing slope for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing standardized functional richness slope for freshwater studies.
-Slopes for freshwater studies were also significantly smaller than slopes for terrestrial studies.
+However, with only two global studies and two freshwater studies these results are a characterization of the limited included data rather than overall trends.
 
@@
 We found no evidence of systemic functional richness loss or functional change.
-A lack of trend within realm, biome, or taxonomic group gives further evidence that directional functional change is absent from all systems observed in our dataset.
+A lack of trend for almost all realms, biomes, and taxonomic groups gives further evidence that directional functional change is absent from all systems observed in our dataset.
 

### 2021-12-16 paper/manuscript.Rmd
Fix the yaml so everything renders right

@@
 ---
-title: "Evaluating the evidence of widespread maintenance of functional composition in vertebrate communities"
-author: 
-  - name: "Kari E A Norman"
-    affiliation: ucb
+title: Evaluating the evidence of widespread maintenance of functional composition in vertebrate communities
+authors: 
+  - name: "Kari E. A. Norman"
+    affiliation: University of California, Berkeley
+    location: Berkeley, California
+    email: kari.norman@berkeley.edu
   - name: "Perry de Valpine"
-    affiliation: ucb
+    affiliation: University of California, Berkeley
+    location: Berkeley, California
+    email: pdevalpine@berkeley.edu
   - name: "Carl Boettiger"
-    affiliation: ucb, ropensci
-address:
-  - code: ucb
-    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
-  - code: ropensci
-    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA" 
-abstract: Despite unprecedented environmental change due to anthropogenic pressure, recent work has found no overall trend in species diversity through time accompanied by increasing turnover.
+    affiliation: University of California, Berkeley
+    location: Berkeley, California
+    email: cboettig@berkeley.edu
+abstract: |
+ Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing species turnover but no overall trend in species diversity through time.
 Functional diversity provides a potentially powerful alternative approach for understanding this change in community composition by linking shifts in species identity to mechanisms of ecosystem processes.
 Here we present the first multi-taxa, multi-system analysis of functional change through time, pairing thousands of vertebrate assemblage time series from the BioTIME database with existing functional traits representitive of a species' functional role.
-We found no overall trend in any calculated functional diversity metric after correcting for changes in species richness, despite similar species-based patterns of constant richness with increasing turnover.
-This pattern held at the study-level where only 3 of 54 studies experienced a significant trend in at least one functional diversity metric.
-Using best available data and biodiversity synthesis practices, our results give strong evidence that across a variety of taxa, climates, and biomes, functional characteristics are maintained even in the face of significant environmental and community change.
+We found no overall trend in any calculated functional diversity metric, despite similar species-based patterns of constant richness with increasing turnover.
+The lack of trend held even after correcting for changes in species richness and at the study-level, where only 3 of 54 studies experienced a significant trend in at least one functional diversity metric.
+Results give evidence that across a variety of taxa, climates, and biomes, these selected functional characteristics are maintained even in the face of significant environmental and community change.
 We also discuss the potential for underlying functional shifts to be obscured by current approaches and data and call for targeted data collection efforts to combat existing biases in monitoring and trait data.
@@
     keep_tex: true
-  word_document: default
+#  word_document: default
 editor_options: 
@@
 At the local scale however, species diversity tells a different story.
-Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of a significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017; @li2020].
+Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017; @li2020].
 While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.
@@
 We focus on mammal, bird, and amphibian species as a significant subset of the world's biodiversity heavily impacted by anthropogenic change.
-While examining trends in plants, invertebrates, and other vertebrate species are of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
-In order to ensure comparability across taxa in trait type and data quality we therefore focus on these taxa.
+While examining trends in plants, invertebrates, and other vertebrate species is of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
+In order to ensure comparability across taxa in trait type and data quality we therefore focus on mammals, birds, and amphibians.
 Traits were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].
 
-We assess thousands of mammal, bird, and amphibian functional diversity time series to determine whether or not there is a general trend of functional change.
+We assess thousands of mammal, bird, and amphibian functional diversity time series to determine whether or not there is a general trend of functional change, both in observed metrics and in metrics corrected for changes in species richness.
 We distinguish between three possible scenarios of functional change: 1) significant loss of functional diversity with accompanying shifts in other functional metrics, 2) no functional diversity loss, but significant shifts in other functional metrics, 3) maintenance of functional diversity and composition.
@@
 Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual time series following a standardized spatial scale.
-Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018] for details on how sample extents were defined), with the area of each cell set to one standard deviation away from the mean of the single extent locations.
+Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018 for details on how sample extents were defined], with the area of each cell set to one standard deviation away from the mean of the single extent locations.
 All samples from a study within a single cell were considered to be a single time series, and species abundances were combined for all samples.
@@
 We gathered trait data from the Elton Trait Database [mammals and birds, @wilman2014] and Amphibio [amphibians, @oliveira2017].
-These databases include species-level means for traits that best represent species' multifaceted function in the community including body size, diet, and behavioral characteristics.
+These databases include species-level means for traits that partially represent species' multifaceted function in the community including body size, diet, and behavioral characteristics.
 For the full list of traits included in the analysis for each taxon see Table 1.
@@
 Only four studies (consisting of 11 time series) came from Amphibian studies due to the limited availability of amphibian time series and low species richness values for assemblages (Table 1).
-Amphibians are of particular concern due to impacts of habitat loss and pollution [@gibbons2000], so we include the data we have while acknowledging that general inference for Amphibians as a taxa is not possible with the timeseries available.
+Amphibians are of particular concern due to impacts of habitat loss and pollution [@gibbons2000], so we include data while acknowledging that general inference for amphibians as a clade is not possible with the time series available.
 For a full breakdown of studies and their characteristics, see the supplement.
@@
 To test for trends within and between different levels of taxa, biome, and realm we fit separate models with each of those covariates added as a predictor to the original model structure.
-We estimated within level slopes and calculated between level contrasts using the *emmeans* package [@lenth2021].
+We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [@lenth2021].
 We assessed the impact of time series duration and start year on study-level trends using general linear models with duration and start year as predictors.
-All models were executed using the *lme4* package in R and p-values were calculated using the *lmerTest* package with a significance level of \$\\alpha \< 0.05\$ [@bates2015; @kuznetsova2017; @rcoreteam2021].
+All models were executed using the *lme4* package in R and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* package with a significance level of \$\\alpha = 0.05\$ [@bates2015; @kuznetsova2017; @rcoreteam2021].
 
@@
 ```
-
-All code for reproducing analyses are archived at (zenodo link).
-Data products are archived at (zenodo link).
-Raw
 
@@
 Trends for different taxa, biomes, or realms were also non-significant with the exception of a significantly increasing slope for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing standardized functional richness slope for freshwater studies.
-However, with only two global studies and two freshwater studies these results are a characterization of the limited included data rather than overall trends.
+However, with only two global studies and two freshwater studies these results are a characterization of the limited data rather than overall trends.
 
@@
 Locally, communities losing functionally rare species should exhibit strong functional shifts as lost species can eventually no longer be replaced by functionally similar species [@leitão2016].
-While we did not have a large enough time window in most time series to capture the complete extinction process, observed patterns were instead consistent with species replacement by functionally redundant species from the species pool.
+Observed patterns were instead consistent with species replacement by functionally redundant species from the species pool.
+Still, for many timeseries we likely did not have a large enough time window to capture community and species pool impoverishment due to extinction.
 
@@
 
+#### Policy Implications
+
+Our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
+These findings do not negate a substantial body of work linking functional degradation to direct human intervention in the form of land use change and intensification or habitat fragmentation [@flynn2009; @magioli2021; @tinoco2018], but rather illustrate the prevailing functional state for communities experiencing background levels of environmental change.
+Rather than assuming functional structure will be maintained in areas of concern, our work indicates that when measurements of functional diversity show significant shifts, it should be considered an indication of substantial community change and outside the normal expectation.
+
 #### Future Work
@@
 
+### Data Availability
+
+Code for the analyses in this chapter is archived on Zenodo at \<DOI:10.5281/zenodo.5514335\> [\@norman_karinormanbiodivts_2021].
+Data products are on Github at <https://github.com/karinorman/biodivTS_data>.
+Original data sources are open access and available at their respective providers.
+
+### Acknowledgments
+
+We thank the scientists who contributed to and maintain the biodiversity databases included in this study, including BioTIME, Amphibio, and Elton Traits.
+We thank Dr Shane Blowes and Dr Sarah Supp, whose code we adapted for initial data processing of the BioTIME database.
+Support for this work was provided by United States Department of Energy through the Computational Sciences Graduate Fellowship (DOE CSGF) under grant number DE-FG02-97ER25308 awarded to K.E.A.N..
+
 ### References

### 2021-12-16 analysis/04_collate_rarefied_null.Rmd

@@
                               c("YEAR", "cell", "rarefy_resamp", "rarefyID")) %>%
-  full_join(cwm_metrics, c("YEAR", "cell", "rarefy_resamp", "rarefyID")) %>%
+  full_join(cwm_metrics %>% rename_with(~paste0("CWM_", .), -c("YEAR", "cell", "rarefy_resamp", "rarefyID")), 
+            c("YEAR", "cell", "rarefy_resamp", "rarefyID")) %>%
   filter(S != 1) %>% # remove samples that had one species as they result in strange metric values
@@
   select(-c(SamplePool, SampleN, num_years, duration, startYear, endYear, type,
-            diet_5cat, pelagicspecialist, diurnal, nocturnal, crepuscular, dir, lar)) %>%
-  pivot_longer(-c(rarefyID, rarefied, YEAR, cell, rarefy_resamp), names_to = "metric") %>%
+            CWM_diet_5cat, CWM_pelagicspecialist, CWM_diurnal, CWM_nocturnal, 
+            CWM_crepuscular, CWM_dir, CWM_lar)) %>%
+  tidyr::pivot_longer(-c(rarefyID, rarefied, YEAR, cell, rarefy_resamp), names_to = "metric") %>%
   left_join(null_table %>% select(-c(richness, se, lowerCI, upperCI)),
@@
   #the next three lines go back to wide format for both SES and unadjusted values, then to long format incorporating SES values
-  pivot_wider(names_from = "metric", values_from = c("value", "SES")) %>%
+  tidyr::pivot_wider(names_from = "metric", values_from = c("value", "SES")) %>%
   rename_at(vars(starts_with("value")), ~str_replace(., "value_", "")) %>%
-  pivot_longer(-c(rarefyID, rarefied, YEAR, cell, rarefy_resamp), names_to = "metric") %>%
+  tidyr::pivot_longer(-c(rarefyID, rarefied, YEAR, cell, rarefy_resamp), names_to = "metric") %>%
   filter(!is.na(value))

### 2021-12-16 analysis/05_model_metrics.Rmd

@@
                        "FRic", "FEve", "FDiv", "S",
-                       "bodymass_value"),
+                       "CWM_bodymass_value"),
          !rarefyID %in% missing_null) %>%

### 2022-04-13 analysis/05_model_metrics.Rmd
Fit models for all CWM models, add scenarios where there aren't enough data to model fit function

--- 2021-10-19 analysis/05_model_metrics.Rmd
@@
   filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "Jaccard_next",
-                       "FRic", "FEve", "FDiv", "S",
-                       "bodymass_value"),
+                       "FRic", "FEve", "FDiv", "S","CWM_bodymass_value","CWM_diet_inv", 
+                       "CWM_diet_scav", "CWM_diet_vfish", "CWM_diet_vunk", "CWM_forstrat_ground", 
+                       "CWM_forstrat_wataroundsurf", "CWM_forstrat_watbelowsurf", "CWM_diet_fruit", 
+                       "CWM_diet_planto", "CWM_diet_seed", "CWM_diet_vend", "CWM_forstrat_understory", 
+                       "CWM_diet_vect", "CWM_diet_nect", "CWM_forstrat_aerial", "CWM_forstrat_canopy", 
+                       "CWM_forstrat_midhigh", "CWM_age_at_maturity_max_y", "CWM_age_at_maturity_min_y", 
+                       "CWM_body_size_mm", "CWM_litter_size_max_n", "CWM_litter_size_min_n", 
+                       "CWM_longevity_max_y", "CWM_offspring_size_max_mm", "CWM_offspring_size_min_mm"),
          !rarefyID %in% missing_null) %>%
@@
 
+  if(n_distinct(fit_data$rarefyID) < 2){
+    return(tibble())
+  }
+
+  if(!model_id %in% c("base", "base_dur", "base_srt")){
+    if(fit_data %>% pull(model_id) %>% n_distinct() >1) {
+
       fit <- qlmer(formula = formula, data = fit_data, ...)
 
-  if(!model_id %in% c("base", "base_dur", "base_srt")){
       slopes <- get_fixed_slopes(metric = metric_name, model = fit$result, model_id = model_id, model_data = fit_data)
@@
         bind_rows(group_slopes, slope_constrasts) %>%
-      mutate(metric = metric_name, model = list(fit$result), model_id = model_id, warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
+        mutate(metric = metric_name, model = list(fit$result), model_id = model_id,
+               n_ts = n_distinct(fit_data$rarefyID),
+               warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
     }else{
+      return(tibble())
+    }
+  }else{
+
+    fit <- qlmer(formula = formula, data = fit_data, ...)
+
     tidy(fit$result) %>%
-   mutate(metric = metric_name, model = list(fit$result), model_id = model_id, warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
+      mutate(metric = metric_name, model = list(fit$result), model_id = model_id, 
+             n_ts = n_distinct(fit_data$rarefyID),
+             warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
   }
@@
 ```{r}
-raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "FRic", "FEve", "FDiv", "Jaccard_base"), 
+metric_names <- unique(model_data$metric)
+raw_value_mods <- tibble(metrimod_resultsc_name = metric_names[metric_names != "S"], 
                          base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
@@
 
+Write out table of model estimates for manuscript
+```{r}
+metric_model_table %>% 
+  filter(model_id == "base", metric %in% c("S", "Jaccard_base", "SES_FRic", "SES_FEve", "SES_FDiv")) %>% 
+  select(-c(model, model_id, warning, statistic, df)) %>% 
+  write_csv(here::here("figures", "man_model_table.csv"))
+```
+
 Models with interactions
 ```{r}
-int_raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "FRic", "FEve", "FDiv", "Jaccard_base"), 
-                         base_dur = "value ~ year_scaled + duration_scaled*year_scaled + (year_scaled|study_id/rarefyID)",
-                         base_srt = "value ~ year_scaled + startyear_scaled*year_scaled + (year_scaled|study_id/rarefyID)") %>%
-  pivot_longer(cols = c(base_dur, taxa_dur, realm_dur, climate_dur, base_srt, taxa_srt, realm_srt, climate_srt), 
-               names_to = "model_id", values_to = "formula")
-
-int_log_value_mods <- tibble(metric_name = c("S"),
-                         base_dur = "logvalue ~ year_scaled + duration_scaled*year_scaled + (year_scaled|study_id/rarefyID)",
-                         base_srt = "logvalue ~ year_scaled + startyear_scaled*year_scaled + (year_scaled|study_id/rarefyID)") %>%
-  pivot_longer(cols = c(base_dur, taxa_dur, realm_dur, climate_dur, base_srt, taxa_srt, realm_srt, climate_srt), 
-               names_to = "model_id", values_to = "formula")
-
-# mods <- bind_rows(raw_value_mods, log_value_mods) %>% 
-#   select(metric_name, formula, model_id)
-
-#Fit models with base settings
-int_mod_results <- pmap_dfr(bind_rows(int_raw_value_mods, int_log_value_mods), get_model, data = model_data)
-
-#get the models that had convergence warnings
-bad_models <- int_mod_results %>%
-  select(metric, model_id, warning) %>% 
-  filter(!is.na(warning)) %>% 
-  distinct() %>%
-  left_join(int_raw_value_mods, by = c("metric" = "metric_name", "model_id")) 
+# int_raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "FRic", "FEve", "FDiv", "Jaccard_base"), 
+#                          base_dur = "value ~ year_scaled + duration_scaled*year_scaled + (year_scaled|study_id/rarefyID)",
+#                          base_srt = "value ~ year_scaled + startyear_scaled*year_scaled + (year_scaled|study_id/rarefyID)") %>%
+#   pivot_longer(cols = c(base_dur, taxa_dur, realm_dur, climate_dur, base_srt, taxa_srt, realm_srt, climate_srt), 
+#                names_to = "model_id", values_to = "formula")
+# 
+# int_log_value_mods <- tibble(metric_name = c("S"),
+#                          base_dur = "logvalue ~ year_scaled + duration_scaled*year_scaled + (year_scaled|study_id/rarefyID)",
+#                          base_srt = "logvalue ~ year_scaled + startyear_scaled*year_scaled + (year_scaled|study_id/rarefyID)") %>%
+#   pivot_longer(cols = c(base_dur, taxa_dur, realm_dur, climate_dur, base_srt, taxa_srt, realm_srt, climate_srt), 
+#                names_to = "model_id", values_to = "formula")
+# 
+# 
+# # mods <- bind_rows(raw_value_mods, log_value_mods) %>% 
+# #   select(metric_name, formula, model_id)
+# 
+# #Fit models with base settings
+# int_mod_results <- pmap_dfr(bind_rows(int_raw_value_mods, int_log_value_mods), get_model, data = model_data)
+# 
+# #get the models that had convergence warnings
+# bad_models <- int_mod_results %>%
+#   select(metric, model_id, warning) %>% 
+#   filter(!is.na(warning)) %>% 
+#   distinct() %>%
+#   left_join(int_raw_value_mods, by = c("metric" = "metric_name", "model_id")) 
 ```

### 2022-04-14 analysis/05_model_metrics.Rmd
Refit second round of stubborn models, get list of models that wouldn't fit

@@
 metric_names <- unique(model_data$metric)
-raw_value_mods <- tibble(metrimod_resultsc_name = metric_names[metric_names != "S"], 
+raw_value_mods <- tibble(metric_name = metric_names[metric_names != "S"], 
                          base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
@@
                          climate = "value ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
-  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
+  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula") %>%
+  #don't fit the CWM_litter_size_min_n and climate model, not enough data
+  filter(metric_name != "CWM_litter_size_min_n" | model_id != "climate")
 
@@
 
+# metric_model_table <- bind_rows(mod_results %>% filter(is.na(warning)), 
+#           bad_mod_results %>% filter(is.na(warning))) %>%
+#   mutate(significant = case_when(
+#     p.value > 0.05 | lower.CL < 0 & upper.CL > 0 ~ FALSE,
+#     !is.na(p.value) | !is.na(lower.CL) ~ TRUE,
+#     TRUE ~ NA
+#   ))
+
+#models that still won't converge
+bad_models2 <- bad_mod_results %>%
+  select(metric, model_id, warning) %>% 
+  filter(!is.na(warning)) %>% 
+  distinct() %>%
+  left_join(mods, by = c("metric" = "metric_name", "model_id"))
+
+#Fit with a different algorithm
+bad_mod_results2 <- pmap_dfr(bad_models2 %>% select(-warning), get_model, data = model_data, control = lmerControl(optimizer = "bobyqa"))
+
 metric_model_table <- bind_rows(mod_results %>% filter(is.na(warning)), 
-          bad_mod_results %>% filter(is.na(warning))) %>%
+          bad_mod_results %>% filter(is.na(warning)),
+          bad_mod_results2 %>% filter(is.na(warning))) %>%
   mutate(significant = case_when(
@@
     TRUE ~ NA
-  ))
+  )) %>%
+  select(-model)
 
@@
 pins::pin(metric_model_table, board = "github")
+```
+
+Get list of models that couldn't be fit due to model limitations 
+```{r}
+#models that were fit
+uniq_metric_model <- metric_model_table %>% select(metric, model_id) %>% distinct()
+#models we attempted to fit
+uniq_mods <- mods %>% select(metric, model_id) %>% distinct()
+
+setdiff(uniq_mods, uniq_metric_model)
 ```

### 2022-04-20 analysis/05_model_metrics.Rmd
Modify reading files in to be cleaner

@@
 #models we attempted to fit
-uniq_mods <- mods %>% select(metric, model_id) %>% distinct()
+uniq_mods <- mods %>% select(metric = metric_name, model_id) %>% distinct()
 
-setdiff(uniq_mods, uniq_metric_model)
+missing_mods <- setdiff(uniq_mods, uniq_metric_model)
 ```

### 2022-04-20 analysis/05_model_metrics.Rmd
Clarifications

@@
 
+#this is a list of models that weren't fit because there weren't enough timeseries OR they didn't converge
+#min_litter_size just won't converge
 missing_mods <- setdiff(uniq_mods, uniq_metric_model)
@@
   filter(model_id == "base", metric %in% c("S", "Jaccard_base", "SES_FRic", "SES_FEve", "SES_FDiv")) %>% 
-  select(-c(model, model_id, warning, statistic, df)) %>% 
+  select(-c(model_id, warning, statistic, df)) %>% 
   write_csv(here::here("figures", "man_model_table.csv"))

### 2022-04-21 paper/manuscript.Rmd
clean up fixes
Add cwm stuff

@@
 Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package [@laliberté2010].
-Here we report functional richness (FRic), functional evenness (FEve), and functional divergence (FDiv) which together describe three complementary characteristics of the functional space [@mason2005; @hillebrand2009].
-FRic assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
-FEve describes how species are distributed across the trait space and how abundance is distributed across species.
-Higher values of FEve indicate more even spacing of species in the trait space and individuals across species.
-FDiv measures the degree to which species and their abundances maximize differences in the functional space.
-Higher values of FDiv therefore correspond to communities where many highly abundant species are on the edges of the trait space.
+Here we report functional richness (*FRic*), functional evenness (*FEve*), and functional divergence (*FDiv*) which together describe three complementary characteristics of the functional space [@mason2005; @hillebrand2009].
+*FRic* assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
+*FEve* describes how species are distributed across the trait space and how abundance is distributed across species.
+Higher values of *FEve* indicate more even spacing of species in the trait space and individuals across species.
+*FDiv* measures the degree to which species and their abundances maximize differences in the functional space.
+Higher values of *FDiv* therefore correspond to communities where many highly abundant species are on the edges of the trait space.
+We also calculated the community-weighted mean (*CWM*) of all included traits to examine shifts in individual trait distributions.
 
@@
 
-To assess functional change independent of species richness we calculated the standardized effect size (SES) for each metric from null estimates [@swenson2012].
+To assess functional change independent of species richness we calculated the standardized effect size (SES) for each of the three summary functional diversity metrics (*FRic*, *FEve*, *FDiv*) from null estimates [@swenson2012].
 Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
@@
 We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study.
-We obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model.
+For clarity in assessing the large number of *CWM*'s, we report only these general trends for each trait *CWM*.
+All timeseries with data for a given trait were included in the model.
+Trends could not be estimated for the *CWM'*s of three traits due to limited data: maximum age at maturity, minimum age of maturity, and minimum litter size.
+We fit 24 individual *CWM* models.
+<!-- something here about models be segregated by taxa already, since most traits weren't for all taxa -->
+
+For all other species and functional diversity metrics (*S*, *J*, *FRic*, *FEve*, *FDiv)* we obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model.
 BLUP's provide an estimate for the conditional mean and variance of each level of a random effect from which we calculated 95% confidence intervals to determine significance of study-level slopes.
@@
 
-We found no significant overall trend in species richness or functional diversity metrics (observed or standardized) (Fig 2).
+We found no significant overall trend in species richness or summary functional diversity metrics (observed or standardized) (Fig 2).
 We did find a significant overall decrease in Jaccard similarity, indicating increasing turnover through time.
@@
 Trends for different taxa, biomes, or realms were also non-significant with the exception of a significantly increasing slope for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing standardized functional richness slope for freshwater studies.
-However, with only two global studies and two freshwater studies these results are a characterization of the limited data rather than overall trends.
+However, with only two global studies and two freshwater studies these results are likely a characterization of the limited data rather than overall trends.
+We found similar results for the *CWM* models, with a significant trend for only one trait, percentage of aerial foraging utilization, which showed a significant positive trend.
 
@@
   arrange(metric) %>%
-  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness (S) and standardized metrics.
-Additional model estimates can be found in the supplement.") %>%
+  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness (S) and standardized summary metrics.
+Additional model estimates, including CWM models, can be found in the supplement.") %>%
   collapse_rows(columns = 1, valign = "middle") %>%
@@
 A lack of trend for almost all realms, biomes, and taxonomic groups gives further evidence that directional functional change is absent from all systems observed in our dataset.
+Additionally, results from *CWM* models show that there were very few general shifts in functional trait values across communities.
 
@@
 While the majority of the studies (\~70%) included in our data experienced significant species turnover, only three (for standardized metrics) experienced a significant shift in any functional dimension.
-<!-- General consistency between observed and standardized metrics underlines that changes in species richness were not a major factor determining functional composition changes.--> This suggests certain characteristics of the functional space are maintained even in the face of significant change in species identity, specifically the size of the functional space occupied by the community (FRic) and the distribution of species and individuals within that space (FEve and FDiv).
+This suggests certain characteristics of the functional space are maintained even in the face of significant change in species identity, specifically the size of the functional space occupied by the community (*FRic*), the distribution of species and individuals within that space (*FEve* and *FDiv*), and the mean of individual trait distributions (*CWM*).
 On average, species additions have similar functional characteristics as lost species and therefore maintain the structure of the functional space.
-<!-- Our results give strong evidence that maintenance of functional structure is not just a phenomena of paleoecological time scales but also annual time scales [@barnosky2004; @jackson2015]. -->
 
@@
 
-While we did not directly measure changes in rare species, our results also contradict likely scenarios of loss predicted due to rare species extinction.
+Results are also seemingly inconsistent with predictions for trait shifts under global change.
+For example, mean body size is predicted to decrease as a result of climate change impacts and megafaunal loss [@sheridan2011], a phenomena which has already been well documented empirically and expirimentally in multiple taxa [@caruso2014; @forster2012; @huss2019; @tseng2018].
+We found no evidence of a trend in the community-weighted mean of body size.
+Still, species-level trait means can only capture distribution shifts due to complete species losses.
+Intraspecific changes may still be occurring in these communities, particularly if species are being impacted by environmental changes but have not yet been fully extirpated from a community.
+
+While we did not directly measure changes in rare species, our results further contradict likely scenarios of loss predicted due to rare species extinction.
 Rare species, defined by small populations and geographic restriction, are simultaneously more likely to be functionally distinct and at higher risk for extinction [@davies2004; @harnik2012; @loiseau2020].
@@
 Third, while we use here the most common metrics for describing functional diversity they do not measure some potentially important aspects of the functional space.
-Most notably, the metrics we calculated do not capture shifts in the location of the functional space.
+Most notably, the summary metrics we calculated do not capture shifts in the location of the functional space as a whole.
 For example, two communities could have very similar metric values but no overlap in their trait spaces.
@@
 This scenario may be common in communities tracking changing environmental conditions.
-Approaches for assessing shifts and overlaps in functional space are still relatively new [@barros2016; @blonder2018; @mammola2019] but could shed critical insight into functional composition changes of this nature.
+While trait *CWM's* capture axis shifts, approaches for assessing multidimensional shifts in functional space are still relatively new [@barros2016; @blonder2018; @mammola2019] but could shed critical insight into functional composition changes of this nature.
 
@@
 
-Code for the analyses in this chapter is archived on Zenodo at \<DOI:10.5281/zenodo.5514335\> [\@norman_karinormanbiodivts_2021].
+Code for the analyses in this chapter is archived on Zenodo at \<[[DOI:10.5281/zenodo.5514335\\\\](DOI:10.5281/zenodo.5514335\){.uri}]([DOI:10.5281/zenodo.5514335\\](DOI:10.5281/zenodo.5514335)%7B.uri%7D){.uri}\> [\@norman_karinormanbiodivts_2021].
 Data products are on Github at <https://github.com/karinorman/biodivTS_data>.

### 2022-04-21 analysis/04_collate_rarefied_null.Rmd

@@
 #read in CWM metrics
-path <- here::here("data", "rarefied_metrics", "cwm")
+path <- here::here("data", "rarefied_metrics", "cwm_noscale")
 filelist <-  dir(path, "*.rda") %>%
@@
 cwm_metrics <- purrr::map_dfr(filelist, function(x) {
+  #browser()
   load(x)
@@
 
-    cwm_metrics <- cwm_metrics$CWM %>%
-      rownames_to_column("YEAR") %>%
-      mutate(YEAR = as.integer(YEAR)) %>%
-      left_join(cwm_metrics %>% select(-CWM)) %>%
-      mutate(rarefyID = rarefyID)
+    cwm_metrics <- cwm_metrics %>%
+      unnest(CWM) %>%
+      mutate(YEAR = as.integer(YEAR), rarefyID = rarefyID)
 

### 2022-04-21 paper/manuscript.Rmd
Stupid linebreak things

@@
 
-Code for the analyses in this chapter is archived on Zenodo at \<[[DOI:10.5281/zenodo.5514335\\\\](DOI:10.5281/zenodo.5514335\){.uri}]([DOI:10.5281/zenodo.5514335\\](DOI:10.5281/zenodo.5514335)%7B.uri%7D){.uri}\> [\@norman_karinormanbiodivts_2021].
+Code for the analyses in this chapter is archived on Zenodo at \<[[DOI:10.5281/zenodo.5514335\\\\](DOI:10.5281/zenodo.5514335){.uri}](%5BDOI:10.5281/zenodo.5514335\%5D(DOI:10.5281/zenodo.5514335)%7B.uri%7D){.uri}\> [\@norman_karinormanbiodivts_2021].
 Data products are on Github at <https://github.com/karinorman/biodivTS_data>.

### 2022-04-25 paper/manuscript.Rmd
Perry edits

--- 2022-04-21 paper/manuscript.Rmd
@@
  Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing species turnover but no overall trend in species diversity through time.
-Functional diversity provides a potentially powerful alternative approach for understanding this change in community composition by linking shifts in species identity to mechanisms of ecosystem processes.
-Here we present the first multi-taxa, multi-system analysis of functional change through time, pairing thousands of vertebrate assemblage time series from the BioTIME database with existing functional traits representitive of a species' functional role.
-We found no overall trend in any calculated functional diversity metric, despite similar species-based patterns of constant richness with increasing turnover.
+Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to mechanisms of ecosystem processes.
+Here we present the first multi-taxa, multi-system analysis of functional change through time, pairing thousands of vertebrate assemblage time series from the BioTIME database with existing trait data representative of a species' functional role.
+We found no overall trend in any functional diversity metric, despite similar species-based patterns of constant richness with increasing turnover.
 The lack of trend held even after correcting for changes in species richness and at the study-level, where only 3 of 54 studies experienced a significant trend in at least one functional diversity metric.
@@
 
-Functional diversity offers a potentially powerful alternative for detecting and describing community change by providing a mechanistic link between species' response to environmental change (response traits) and the processes they perform (effect traits) [@lavorel2002; @mcgill2006; @suding2008].
-By describing the functional trait space rather than species, functional diversity metrics capture the disproportionate impact of losses or gains of functionally unique species.
+Functional diversity offers a potentially powerful alternative for detecting and describing community change by providing a mechanistic link between species' response to environmental change (*response traits*) and the processes they perform (*effect traits*) [@lavorel2002; @mcgill2006; @suding2008].
+By describing the functional trait space, functional diversity metrics capture the disproportionate impact of losses or gains of functionally unique species.
 Functional diversity metrics are therefore particularly well suited for assessing community shifts underlying even constant species richness trends.
@@
 Indeed, almost all hypothesized mechanisms underpinning the relationship between species diversity and ecosystem function are trait-dependent [@hillebrand2009].
-Determining functional trends therefore gives a more complete picture of potential trends in critical ecosystem processes.
+Determining functional trends therefore gives a more fundamental picture of potential trends in critical ecosystem processes.
 
-It is critical to establish whether or not functional loss is the general rule for communities.
+It is critical to establish whether or not functional loss is prevalent across communities.
 While functional loss is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], it is not necessarily inevitable even in scenarios of species loss [@daz2001].
@@
 For the full list of traits included in the analysis for each taxon see Table 1.
+Multiple traits (i.e.
+diet, foraging strata, activity seasonality, active diel period) were broken down into percentage or binary use for each level.
 
@@
 Higher values of *FDiv* therefore correspond to communities where many highly abundant species are on the edges of the trait space.
-We also calculated the community-weighted mean (*CWM*) of all included traits to examine shifts in individual trait distributions.
+We also calculated the community-weighted mean (*CWM*) of included traits to examine shifts in the distribution of each trait.
 
 All available trait data for each study were included in functional diversity calculations with the exception of traits that were the same value for all observed species in the study.
-All continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
+For variables with multiple levels each level was included as a separate trait axis.
+Continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
 The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
@@
 Species pools included all species observed for a time series.
-This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction for that time series.
+This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
 We used these values to calculate SES using the following formula: $SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
@@
 We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study.
-For clarity in assessing the large number of *CWM*'s, we report only these general trends for each trait *CWM*.
 All timeseries with data for a given trait were included in the model.
@@
 We fit 24 individual *CWM* models.
-<!-- something here about models be segregated by taxa already, since most traits weren't for all taxa -->
 
 For all other species and functional diversity metrics (*S*, *J*, *FRic*, *FEve*, *FDiv)* we obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model.
-BLUP's provide an estimate for the conditional mean and variance of each level of a random effect from which we calculated 95% confidence intervals to determine significance of study-level slopes.
-To test for trends within and between different levels of taxa, biome, and realm we fit separate models with each of those covariates added as a predictor to the original model structure.
+BLUP's provide estimates for the conditional mean and variance of a random effect from which we calculated 95% confidence intervals to determine significance of study-level slopes.
+
+To test for trends within and between different levels of taxa, biome, and realm we fit separate models with each of those factors added as a predictor to the original model structure.
 We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [@lenth2021].
@@
 Non-significant overall trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies was plausibly 0 (Table 2).
-Trends for different taxa, biomes, or realms were also non-significant with the exception of a significantly increasing slope for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing standardized functional richness slope for freshwater studies.
-However, with only two global studies and two freshwater studies these results are likely a characterization of the limited data rather than overall trends.
+Trends for different taxa, biomes, or realms were also non-significant with the exception of a significantly increasing trend for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing standardized functional richness slope for freshwater studies.
+However, with only two global studies and two freshwater studies these results are not general trends.
 We found similar results for the *CWM* models, with a significant trend for only one trait, percentage of aerial foraging utilization, which showed a significant positive trend.
@@
   mutate(metric = ifelse(metric == "Jaccard_base", "Jaccard", metric),
+         metric = ifelse(metric == "S", "Species Richness", metric),
          group = case_when(
@@
 Most significant trends for observed functional metrics are in functional richness and disappeared after standardization, indicating that richness increases were likely due to changes in the number of species.
-<!-- Trends for all but one study were inconsistent between corrected and uncorrected metrics, with significance for one or the other.
---> Hypothesis testing for study-level trends is likely affected by multiple testing issues and some trends identified as significant are therefore potentially erroneous.
-Rather than interpreting changes in specific studies, we present these results as a general picture of the small number of studies experiencing a trend and highlight that even those studies are likely an overestimate of the number of significant trends.
+However, the majority of the functional richness trends were negative, indicating that when change is occurring it is more often a loss.
+Hypothesis testing for study-level trends is likely affected by multiple testing issues and some trends identified as significant are therefore potentially spurious.
+Rather than interpreting changes in specific studies, we present these results as a general picture of how few studies experienced a significant trend and highlight that even that count is likely an overestimate.
 
@@
 Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at aggregate and local trends in functional diversity in mammal, bird, and amphibian communities.
-Surprisingly, we did not detect an overall trend in any of the calculated functional diversity metrics.
+Surprisingly, we did not detect an overall trend in any of the chosen functional diversity metrics.
 As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics may be consistent with similar well-documented species derived trends.
-We found no evidence of systemic functional richness loss or functional change.
+We found no evidence of systematic functional richness loss or functional change.
 A lack of trend for almost all realms, biomes, and taxonomic groups gives further evidence that directional functional change is absent from all systems observed in our dataset.
-Additionally, results from *CWM* models show that there were very few general shifts in functional trait values across communities.
+Additionally, results from *CWM* models show that there were very few general shifts in functional trait values.
 
 This striking result could be a product of two possible processes, one ecological and one methodological.
-Null trends appear to give strong evidence of systemic maintenance of functional structure due to common ecological processes, however multiple limitations of current approaches in synthesis could potentially be obscuring a true underlying global trend.
+Null trends appear to give strong evidence of systematic maintenance of functional structure due to common ecological processes, however multiple limitations of current approaches in synthesis could potentially be obscuring a true underlying global trend.
 We discuss both options further here.
@@
 These results challenge assumptions that functional loss is the default state of all or even many communities.
-While we do not directly assess the disturbance histories of included communities, trends were consistent even for the longest running and most heavily impacted studies.
+And while there is some evidence that functional loss may be more common than functional gains for communities experiencing functional change, there was no significant trend for even the longest running and most heavily impacted studies.
 The North American Breeding Survey for example is considered an authoritative dataset on the state of bird populations on the continent and underpins policy decisions about bird conservation [@pardieck2020; @rosenberg2019; @sauer2017].
@@
 Results are also seemingly inconsistent with predictions for trait shifts under global change.
-For example, mean body size is predicted to decrease as a result of climate change impacts and megafaunal loss [@sheridan2011], a phenomena which has already been well documented empirically and expirimentally in multiple taxa [@caruso2014; @forster2012; @huss2019; @tseng2018].
-We found no evidence of a trend in the community-weighted mean of body size.
-Still, species-level trait means can only capture distribution shifts due to complete species losses.
-Intraspecific changes may still be occurring in these communities, particularly if species are being impacted by environmental changes but have not yet been fully extirpated from a community.
+For example, mean body size is predicted to decrease as a result of climate change impacts and megafaunal loss [@sheridan2011], a phenomena which has already been well documented empirically and experimentally in multiple taxa [@caruso2014; @forster2012; @huss2019; @tseng2018].
+While the species-level trait means used here are not appropriate for assessing intraspecific body size shifts, we would expect to see shifts in community-wide means due to local losses of large-bodied species.
+Instead, we found no evidence of a trend in *CWM* body size.
 
@@
 
-Third, while we use here the most common metrics for describing functional diversity they do not measure some potentially important aspects of the functional space.
+Third, while we use here the most common metrics for describing functional diversity, they do not measure some potentially important aspects of the functional space.
 Most notably, the summary metrics we calculated do not capture shifts in the location of the functional space as a whole.
@@
 Our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
-These findings do not negate a substantial body of work linking functional degradation to direct human intervention in the form of land use change and intensification or habitat fragmentation [@flynn2009; @magioli2021; @tinoco2018], but rather illustrate the prevailing functional state for communities experiencing background levels of environmental change.
+These findings do not negate a substantial body of work linking functional degradation to direct human intervention in the form of land use change and intensification or habitat fragmentation [@flynn2009; @magioli2021; @tinoco2018], but rather illustrate trends for communities experiencing background levels of environmental change.
 Rather than assuming functional structure will be maintained in areas of concern, our work indicates that when measurements of functional diversity show significant shifts, it should be considered an indication of substantial community change and outside the normal expectation.
@@
 
-Code for the analyses in this chapter is archived on Zenodo at \<[[DOI:10.5281/zenodo.5514335\\\\](DOI:10.5281/zenodo.5514335\){.uri}]([DOI:10.5281/zenodo.5514335\\](DOI:10.5281/zenodo.5514335)%7B.uri%7D){.uri}\> [\@norman_karinormanbiodivts_2021].
+Code for the analyses in this chapter is archived on Zenodo at \<[[DOI:10.5281/zenodo.5514335\\\\](DOI:10.5281/zenodo.5514335){.uri}](%5BDOI:10.5281/zenodo.5514335%5D(DOI:10.5281/zenodo.5514335)%7B.uri%7D){.uri}\> [\@norman_karinormanbiodivts_2021].
 Data products are on Github at <https://github.com/karinorman/biodivTS_data>.

### 2022-04-25 analysis/05_model_metrics.Rmd
Get rid of extraneous stuff

@@
   write_csv(here::here("paper", "study_metadata.csv"))
-```
-
-Option to add human footprint data
-```{r}
-# hfp_data <- pins::pin_get("hfp-data", board = "github")
-# 
-# model_data_hfp <- model_data %>% 
-#   #dplyr::select(rarefyID, year) %>% 
-#   #distinct() %>%
-#   mutate(hfp_year = case_when(
-#     year < 2001 ~ 1993,
-#     TRUE ~ 2009
-#   ), 
-#   hfp_rarefyID = str_replace_all(rarefyID, pattern = "_bird", ""), 
-#   hfp_rarefyID = str_replace_all(hfp_rarefyID, pattern = "_mamm", "")) %>%
-#   left_join(hfp_data %>% dplyr::select(-STUDY_ID) %>% mutate(year = as.numeric(year)), by = c("hfp_year" = "year", "hfp_rarefyID" = "rarefyID")) %>%
-#   dplyr::select(-hfp_year)
-
 ```
@@
 
-Models with interactions
-```{r}
-# int_raw_value_mods <- tibble(metric_name = c("SES_FRic", "SES_FEve", "SES_FDiv", "FRic", "FEve", "FDiv", "Jaccard_base"), 
-#                          base_dur = "value ~ year_scaled + duration_scaled*year_scaled + (year_scaled|study_id/rarefyID)",
-#                          base_srt = "value ~ year_scaled + startyear_scaled*year_scaled + (year_scaled|study_id/rarefyID)") %>%
-#   pivot_longer(cols = c(base_dur, taxa_dur, realm_dur, climate_dur, base_srt, taxa_srt, realm_srt, climate_srt), 
-#                names_to = "model_id", values_to = "formula")
-# 
-# int_log_value_mods <- tibble(metric_name = c("S"),
-#                          base_dur = "logvalue ~ year_scaled + duration_scaled*year_scaled + (year_scaled|study_id/rarefyID)",
-#                          base_srt = "logvalue ~ year_scaled + startyear_scaled*year_scaled + (year_scaled|study_id/rarefyID)") %>%
-#   pivot_longer(cols = c(base_dur, taxa_dur, realm_dur, climate_dur, base_srt, taxa_srt, realm_srt, climate_srt), 
-#                names_to = "model_id", values_to = "formula")
-# 
-# 
-# # mods <- bind_rows(raw_value_mods, log_value_mods) %>% 
-# #   select(metric_name, formula, model_id)
-# 
-# #Fit models with base settings
-# int_mod_results <- pmap_dfr(bind_rows(int_raw_value_mods, int_log_value_mods), get_model, data = model_data)
-# 
-# #get the models that had convergence warnings
-# bad_models <- int_mod_results %>%
-#   select(metric, model_id, warning) %>% 
-#   filter(!is.na(warning)) %>% 
-#   distinct() %>%
-#   left_join(int_raw_value_mods, by = c("metric" = "metric_name", "model_id")) 
-```
-
 #### Process Model Output

### 2022-04-26 paper/manuscript.Rmd
Clean up table formatting.

@@
 For the full list of traits included in the analysis for each taxon see Table 1.
-Multiple traits (i.e.
-diet, foraging strata, activity seasonality, active diel period) were broken down into percentage or binary use for each level.
+Multiple traits (i.e. diet, foraging strata, activity seasonality, active diel period) were broken down into percentage or binary use for each level.
 
@@
   select(metric, everything()) %>%
+  filter(group != "Residual" | is.na(group)) %>%
   mutate(metric = ifelse(metric == "Jaccard_base", "Jaccard", metric),
          metric = ifelse(metric == "S", "Species Richness", metric),
+         effect = ifelse(effect == "ran_pars", "random", effect),
          group = case_when(
-           group == "rarefyID:study_id" ~ "timeseries:study",
+           group == "rarefyID:study_id" ~ "timeseries within study",
            group == "study_id" ~ "study",
@@
            term == "year_scaled" ~ "Year",
-           term == "sd__(Intercept)" ~ "Intercept sd",
-           term == "cor__(Intercept).year_scaled" ~ "Intercept Year corr",
-           term == "sd__year_scaled" ~ "Year sd",
-           term == "sd__Observation" ~ "Observation sd",
+           term == "sd__(Intercept)" ~ "SD Intercept",
+           term == "cor__(Intercept).year_scaled" ~ "Corr(Intercept, Year)",
+           term == "sd__year_scaled" ~ "SD Year",
+           term == "sd__Observation" ~ "SD Observation",
            TRUE ~ term
@@
          p.value = ifelse(p.value == 0, "<0.001", p.value)) %>%
-  arrange(metric) %>%
-  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness (S) and standardized summary metrics.
+  arrange(metric != "Species Richness", metric, !is.na(group), group, term == "Corr(Intercept, Year)", term) %>%
+  dplyr::rename(grouping = group) %>%
+  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics.
 Additional model estimates, including CWM models, can be found in the supplement.") %>%
-  collapse_rows(columns = 1, valign = "middle") %>%
-  kable_classic() %>% 
+  column_spec(3, width = "2cm") %>%
+  collapse_rows(columns = 1:3, valign = "middle") %>%
+  kable_paper() %>%
   kable_styling(latex_options= c("scale_down", "hold_position"))

### 2022-04-27 paper/manuscript.Rmd
forrrrmattting

@@
     wrap: sentence
+keywords:
+  - biodiversity change
+  - functional traits
+  - global change
+  - timeseries
 
@@
 
-### Introduction
+## Significance Statement {.unnumbered}
+
+The globe is in the midst of a biodiversity crisis due to anthropogenic impacts.
+It is critical to understand how ecological communities are changing in response xxx.
+Previous broad-scale assessments of biodiversity change through time have focused on metrics of species identity rather than the role species play in the community.
+Functional trait data offer a powerful opportunity to directly assess changes in community functional composition alongside species-based patterns.
+Here, we pair vertebrate community timeseries with functional trait data to take a first look at temporal trends in functional composition.
+We found no evidence of systematic functional loss across communities, with maintenance of all measured aspects
+
+# Introduction
 
@@
 
-### Material and Methods
+# Material and Methods
 
-#### Data
+## Data
 
@@
 
-#### Diversity Metrics
+## Diversity Metrics
 
@@
 
-#### Null Models
+## Null Models
 
@@
 
-#### Analysis
+## Analysis
 
@@
 
-### Results
+# Results
 
@@
 
-### Discussion
+# Discussion
 
@@
 
-#### Evidence of Ecological Processes
+## Evidence of Ecological Processes
 
@@
 
-<!-- We observed no significant trends within taxa.
-Even Amphibians, considered highly sensitive to human impact, maintained functional structure through time.
-Indicates that we could've looked at any taxonomic subset of the data and gotten the same result.
-But what does that mean for taxa we left out?
-We can only make educated guesses based on what we know of the ecology of taxa that were left out - taxa specific papers?
--->
-
-#### Potential Methodological Limitations
+## Potential Methodological Limitations
 
@@
 
-#### Policy Implications
+## Policy Implications
 
@@
 
-#### Future Work
+## Future Work
 
@@
 
-### Data Availability
+# Data Availability
 
@@
 
-### Acknowledgments
+# Acknowledgments
 
@@
 We thank Dr Shane Blowes and Dr Sarah Supp, whose code we adapted for initial data processing of the BioTIME database.
-Support for this work was provided by United States Department of Energy through the Computational Sciences Graduate Fellowship (DOE CSGF) under grant number DE-FG02-97ER25308 awarded to K.E.A.N..
+Support for this work was provided by U.S. Department of Energy through the Computational Sciences Graduate Fellowship (DE-FG02-97ER25308) awarded to K.E.A.N..
 
-### References
+# References

### 2022-04-27 paper/manuscript.Rmd
Rejigger significance statement

@@
 
-The globe is in the midst of a biodiversity crisis due to anthropogenic impacts.
-It is critical to understand how ecological communities are changing in response xxx.
-Previous broad-scale assessments of biodiversity change through time have focused on metrics of species identity rather than the role species play in the community.
-Functional trait data offer a powerful opportunity to directly assess changes in community functional composition alongside species-based patterns.
-Here, we pair vertebrate community timeseries with functional trait data to take a first look at temporal trends in functional composition.
-We found no evidence of systematic functional loss across communities, with maintenance of all measured aspects
+It is critical to understand how ecological communities are responding to the ongoing biodiversity crisis.
+Previous broad-scale assessments of community change through time have focused on metrics of species identity, which do not account for differences between species.
+Functional trait approaches offer a powerful alternative to species-based approaches by directly assessing changes in the species characteristics that define their ecological role.
+Here, we pair vertebrate community timeseries with functional trait data to take a first look at temporal trends in functional composition across biomes, realms, and taxa.
+We found no evidence of systematic functional loss across communities, with maintenance of all measured aspects of functional structure.
+We also identified key methodological holes that may be obscuring a potential underlying trend.
 

### 2022-04-27 paper/manuscript.Rmd
Cleanup, PNAS formatting

@@
   - name: "Kari E. A. Norman"
-    affiliation: University of California, Berkeley
-    location: Berkeley, California
+    affiliation: Université de Montréal
+    location: Montréal, Quebec
     email: kari.norman@berkeley.edu
@@
 bibliography: refs.bib
+csl: pnas.csl
 layout: 3p
@@
   - global change
-  - timeseries
+  - time series
 
@@
 Functional trait approaches offer a powerful alternative to species-based approaches by directly assessing changes in the species characteristics that define their ecological role.
-Here, we pair vertebrate community timeseries with functional trait data to take a first look at temporal trends in functional composition across biomes, realms, and taxa.
+Here, we pair vertebrate community time series with functional trait data to take a first look at temporal trends in functional composition across biomes, realms, and taxa.
 We found no evidence of systematic functional loss across communities, with maintenance of all measured aspects of functional structure.
@@
 
-# Material and Methods
-
-## Data
-
-We obtained mammal, bird, and amphibian time series from the BioTIME database, a global repository of high quality assemblage time series.
-All studies included in the database follow consistent sampling protocols and represent full assemblages rather than populations of single species [@dornelas2018].
-Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual time series following a standardized spatial scale.
-Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018 for details on how sample extents were defined], with the area of each cell set to one standard deviation away from the mean of the single extent locations.
-All samples from a study within a single cell were considered to be a single time series, and species abundances were combined for all samples.
-
-```{r echo = FALSE, results = "asis"}
-tibble(Taxa = c("Mammals", "Birds", "Amphibians"),
-       Timeseries_Count = c(48, 2380, 11),
-       Species_Count = c(184, 700, 184),
-       Trait_Source = c("Elton Traits", "Elton Traits", "Amphibio"),
-       Traits = c("body mass, diet, active diel period", "body mass, diet, nocturnality, \nforest foraging strata, pelagic specialist", "habitat, diet, active diel period, \nactivity seasonality, body mass, body length, \nmin maturation size, max maturation size, \nmin offspring size, max offspring size, \nreproductive output, breeding strategy")
-) %>%
-  mutate_all(linebreak, align = "l") %>% 
-  kableExtra::kable(format = "latex", col.names = c("Taxa", "Number of \nTime Series", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
-               caption = "Summary of the data in the final trait database.") %>%
-  kable_classic() %>% 
-  kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
-   #kableExtra::column_spec(2, width = "5cm") #%>%
-  # kableExtra::column_spec(4, width = "10cm")
-```
-
-We gathered trait data from the Elton Trait Database [mammals and birds, @wilman2014] and Amphibio [amphibians, @oliveira2017].
-These databases include species-level means for traits that partially represent species' multifaceted function in the community including body size, diet, and behavioral characteristics.
-For the full list of traits included in the analysis for each taxon see Table 1.
-Multiple traits (i.e. diet, foraging strata, activity seasonality, active diel period) were broken down into percentage or binary use for each level.
-
-In order to ensure taxonomic consistency across datasets, BioTIME species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (retrieved 09-15-2020 from the on-line database, <https://doi.org/10.5066/F7KH0KBK>), obtained through the *taxadb* R package [@norman2020; @rcoreteam2021].
-If more than one species in the assemblage data resolved to the same identifier, observations were considered the same species.
-For trait data, traits for all species of the same identifier were averaged.
-Only studies with at least 75% trait coverage were included and observations for species with no trait data were excluded.
-In order to have a sufficient number of species to calculate functional diversity metrics, years with fewer than 5 species observed were also excluded.
-
-Many studies had a variable number of samples within years.
-To account for this inconsistency in sampling effort we used sample-based rarefaction by bootstrap resampling within years for each time series based on the smallest number of samples in a year for that time series.
-
-Our final dataset included 2,443 time series from 53 studies in 21 countries and 15 biomes and 13 different traits (Fig 1).
-The earliest sample was in 1923 and the most recent was in 2016.
-Only four studies (consisting of 11 time series) came from Amphibian studies due to the limited availability of amphibian time series and low species richness values for assemblages (Table 1).
-Amphibians are of particular concern due to impacts of habitat loss and pollution [@gibbons2000], so we include data while acknowledging that general inference for amphibians as a clade is not possible with the time series available.
-For a full breakdown of studies and their characteristics, see the supplement.
-
-```{r taxa-map, fig.cap="A) Map of time series locations with points colored by taxa, and B) histograms of time series duration broken down by taxa.", echo=FALSE}
-knitr::include_graphics(here::here("figures/study_map_hist.jpeg"))
-```
-
-## Diversity Metrics
-
-We calculated yearly metrics of functional and species diversity for each time series.
-Species-based metrics include species richness (*S*) and Jaccard similarity (*J*) as a measure of turnover.
-Jaccard similarity was calculated relative to the first observed year for a time series.
-A negative trend in *J* would therefore indicate increasing turnover.
-
-Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package [@laliberté2010].
-Here we report functional richness (*FRic*), functional evenness (*FEve*), and functional divergence (*FDiv*) which together describe three complementary characteristics of the functional space [@mason2005; @hillebrand2009].
-*FRic* assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
-*FEve* describes how species are distributed across the trait space and how abundance is distributed across species.
-Higher values of *FEve* indicate more even spacing of species in the trait space and individuals across species.
-*FDiv* measures the degree to which species and their abundances maximize differences in the functional space.
-Higher values of *FDiv* therefore correspond to communities where many highly abundant species are on the edges of the trait space.
-We also calculated the community-weighted mean (*CWM*) of included traits to examine shifts in the distribution of each trait.
-
-All available trait data for each study were included in functional diversity calculations with the exception of traits that were the same value for all observed species in the study.
-For variables with multiple levels each level was included as a separate trait axis.
-Continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
-The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
-This restriction allows for a sufficient number of axes to capture the trait space while maintaining computational feasibility [@blonder2018].
-Metrics incorporated weighting based on species abundance where available (three studies were presence only).
-
-## Null Models
-
-To assess functional change independent of species richness we calculated the standardized effect size (SES) for each of the three summary functional diversity metrics (*FRic*, *FEve*, *FDiv*) from null estimates [@swenson2012].
-Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
-Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species.
-Species pools included all species observed for a time series.
-This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
-We used these values to calculate SES using the following formula: $SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
-We then calculated the median SES estimate for each metric from all the rarefaction samples for a time series.
-SES estimates can be interpreted as how much of the functional characteristic (richness, evenness, divergence) was observed beyond what was expected by chance for a community of that species richness.
-
-## Analysis
-
-We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study.
-All timeseries with data for a given trait were included in the model.
-Trends could not be estimated for the *CWM'*s of three traits due to limited data: maximum age at maturity, minimum age of maturity, and minimum litter size.
-We fit 24 individual *CWM* models.
-
-For all other species and functional diversity metrics (*S*, *J*, *FRic*, *FEve*, *FDiv)* we obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model.
-BLUP's provide estimates for the conditional mean and variance of a random effect from which we calculated 95% confidence intervals to determine significance of study-level slopes.
-
-To test for trends within and between different levels of taxa, biome, and realm we fit separate models with each of those factors added as a predictor to the original model structure.
-We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [@lenth2021].
-We assessed the impact of time series duration and start year on study-level trends using general linear models with duration and start year as predictors.
-All models were executed using the *lme4* package in R and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* package with a significance level of \$\\alpha = 0.05\$ [@bates2015; @kuznetsova2017; @rcoreteam2021].
-
-```{r timeseries-plot, fig.cap="Plots of time series-level trends with line color corresponding to climatic region, with data points in grey and the overall metric mean in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES", echo=FALSE}
-knitr::include_graphics(here::here("figures/3met_long.jpeg"))
-```
-
 # Results
 
-We found no significant overall trend in species richness or summary functional diversity metrics (observed or standardized) (Fig 2).
+We found no significant overall trend in species richness or summary functional diversity metrics (observed or standardized) (Fig \ref{fig:timeseriesPlot}).
 We did find a significant overall decrease in Jaccard similarity, indicating increasing turnover through time.
-Non-significant overall trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies was plausibly 0 (Table 2).
+Non-significant overall trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies was plausibly 0 (Table \ref{tab:resultsTab}).
 Trends for different taxa, biomes, or realms were also non-significant with the exception of a significantly increasing trend for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing standardized functional richness slope for freshwater studies.
@@
 
-```{r echo=FALSE}
+```{r resultsTab, echo=FALSE}
 read_csv(here::here("figures/man_model_table.csv")) %>%
@@
          group = case_when(
-           group == "rarefyID:study_id" ~ "timeseries within study",
+           group == "rarefyID:study_id" ~ "time series within study",
            group == "study_id" ~ "study",
@@
 
-At the study level, 4 studies experienced a significant trend in species richness and only 10 of 54 studies for observed metrics and 3 of 54 studies for standardized metrics experienced a significant trend for a metric other than Jaccard similarity (Table 3).
+At the study level, 4 studies experienced a significant trend in species richness and only 10 of 54 studies for observed metrics and 3 of 54 studies for standardized metrics experienced a significant trend for a metric other than Jaccard similarity (Table \ref{tab:trendTab}).
 Most significant trends for observed functional metrics are in functional richness and disappeared after standardization, indicating that richness increases were likely due to changes in the number of species.
@@
 
-```{r echo = FALSE}
+```{r trendTab, echo = FALSE}
 library(tidyverse)
@@
 Observed patterns were instead consistent with species replacement by functionally redundant species from the species pool.
-Still, for many timeseries we likely did not have a large enough time window to capture community and species pool impoverishment due to extinction.
+Still, for many time series we likely did not have a large enough time window to capture community and species pool impoverishment due to extinction.
 
@@
 First, the BioTIME database, while the most comprehensive data source of time series available, is limited in temporal and geographic scope.
-Most time series span only a few years (Figure 1) and may not provide the statistical power necessary to detect trends.
+Most time series span only a few years (Fig \ref{fig:taxaMap}) and may not provide the statistical power necessary to detect trends.
 The database is also not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
@@
 
+# Material and Methods
+
+## Data
+
+We obtained mammal, bird, and amphibian time series from the BioTIME database, a global repository of high quality assemblage time series.
+All studies included in the database follow consistent sampling protocols and represent full assemblages rather than populations of single species [@dornelas2018].
+Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual time series following a standardized spatial scale.
+Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018 for details on how sample extents were defined], with the area of each cell set to one standard deviation away from the mean of the single extent locations.
+All samples from a study within a single cell were considered to be a single time series, and species abundances were combined for all samples.
+
+```{r traitSources, echo = FALSE, results = "asis"}
+tibble(Taxa = c("Mammals", "Birds", "Amphibians"),
+       Timeseries_Count = c(48, 2380, 11),
+       Species_Count = c(184, 700, 184),
+       Trait_Source = c("Elton Traits", "Elton Traits", "Amphibio"),
+       Traits = c("body mass, diet, active diel period", "body mass, diet, nocturnality, \nforest foraging strata, pelagic specialist", "habitat, diet, active diel period, \nactivity seasonality, body mass, body length, \nmin maturation size, max maturation size, \nmin offspring size, max offspring size, \nreproductive output, breeding strategy")
+) %>%
+  mutate_all(linebreak, align = "l") %>% 
+  kableExtra::kable(format = "latex", col.names = c("Taxa", "Number of \nTime Series", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
+               caption = "Summary of the data in the final trait database.") %>%
+  kable_classic() %>% 
+  kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
+   #kableExtra::column_spec(2, width = "5cm") #%>%
+  # kableExtra::column_spec(4, width = "10cm")
+```
+
+We gathered trait data from the Elton Trait Database [mammals and birds, @wilman2014] and Amphibio [amphibians, @oliveira2017].
+These databases include species-level means for traits that partially represent species' multifaceted function in the community including body size, diet, and behavioral characteristics.
+For the full list of traits included in the analysis for each taxon see Table \ref{tab:traitSources}.
+Multiple traits (i.e. diet, foraging strata, activity seasonality, active diel period) were broken down into percentage or binary use for each level.
+
+In order to ensure taxonomic consistency across datasets, BioTIME species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (retrieved 09-15-2020 from the on-line database, <https://doi.org/10.5066/F7KH0KBK>), obtained through the `taxadb` R package [@norman2020; @rcoreteam2021].
+If more than one species in the assemblage data resolved to the same identifier, observations were considered the same species.
+For trait data, traits for all species of the same identifier were averaged.
+Only studies with at least 75% trait coverage were included and observations for species with no trait data were excluded.
+In order to have a sufficient number of species to calculate functional diversity metrics, years with fewer than 5 species observed were also excluded.
+
+Many studies had a variable number of samples within years.
+To account for this inconsistency in sampling effort we used sample-based rarefaction by bootstrap resampling within years for each time series based on the smallest number of samples in a year for that time series.
+
+Our final dataset included 2,443 time series from 53 studies in 21 countries and 15 biomes and 13 different traits (Fig \ref{fig:taxaMap}).
+The earliest sample was in 1923 and the most recent was in 2016.
+Only four studies (consisting of 11 time series) came from Amphibian studies due to the limited availability of amphibian time series and low species richness values for assemblages (Table \ref{tab:traitSources}).
+Amphibians are of particular concern due to impacts of habitat loss and pollution [@gibbons2000], so we include data while acknowledging that general inference for amphibians as a clade is not possible with the time series available.
+For a full breakdown of studies and their characteristics, see the supplement.
+
+```{r taxaMap, fig.cap="A) Map of time series locations with points colored by taxa, and B) histograms of time series duration broken down by taxa.", echo=FALSE}
+knitr::include_graphics(here::here("figures/study_map_hist.jpeg"))
+```
+
+## Diversity Metrics
+
+We calculated yearly metrics of functional and species diversity for each time series.
+Species-based metrics include species richness (*S*) and Jaccard similarity (*J*) as a measure of turnover.
+Jaccard similarity was calculated relative to the first observed year for a time series.
+A negative trend in *J* would therefore indicate increasing turnover.
+
+Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package [@laliberté2010].
+Here we report functional richness (*FRic*), functional evenness (*FEve*), and functional divergence (*FDiv*) which together describe three complementary characteristics of the functional space [@mason2005; @hillebrand2009].
+*FRic* assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
+*FEve* describes how species are distributed across the trait space and how abundance is distributed across species.
+Higher values of *FEve* indicate more even spacing of species in the trait space and individuals across species.
+*FDiv* measures the degree to which species and their abundances maximize differences in the functional space.
+Higher values of *FDiv* therefore correspond to communities where many highly abundant species are on the edges of the trait space.
+We also calculated the community-weighted mean (*CWM*) of included traits to examine shifts in the distribution of each trait.
+
+All available trait data for each study were included in functional diversity calculations with the exception of traits that were the same value for all observed species in the study.
+For variables with multiple levels each level was included as a separate trait axis.
+Continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
+The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
+This restriction allows for a sufficient number of axes to capture the trait space while maintaining computational feasibility [@blonder2018].
+Metrics incorporated weighting based on species abundance where available (three studies were presence only).
+
+## Null Models
+
+To assess functional change independent of species richness we calculated the standardized effect size (SES) for each of the three summary functional diversity metrics (*FRic*, *FEve*, *FDiv*) from null estimates [@swenson2012].
+Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
+Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species.
+Species pools included all species observed for a time series.
+This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
+We used these values to calculate SES using the following formula: $SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
+We then calculated the median SES estimate for each metric from all the rarefaction samples for a time series.
+SES estimates can be interpreted as how much of the functional characteristic (richness, evenness, divergence) was observed beyond what was expected by chance for a community of that species richness.
+
+## Analysis
+
+We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study.
+All time series with data for a given trait were included in the model.
+Trends could not be estimated for the *CWM'*s of three traits due to limited data: maximum age at maturity, minimum age of maturity, and minimum litter size.
+We fit 24 individual *CWM* models.
+
+For all other species and functional diversity metrics (*S*, *J*, *FRic*, *FEve*, *FDiv)* we obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model.
+BLUP's provide estimates for the conditional mean and variance of a random effect from which we calculated 95% confidence intervals to determine significance of study-level slopes.
+
+To test for trends within and between different levels of taxa, biome, and realm we fit separate models with each of those factors added as a predictor to the original model structure.
+We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [@lenth2021].
+We assessed the impact of time series duration and start year on study-level trends using general linear models with duration and start year as predictors.
+All models were executed using the *lme4* package in R and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* package with a significance level of \$\\alpha = 0.05\$ [@bates2015; @kuznetsova2017; @rcoreteam2021].
+
+```{r timeseriesPlot, fig.cap="Plots of time series-level trends with line color corresponding to climatic region, with data points in grey and the overall metric mean in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES", echo=FALSE}
+knitr::include_graphics(here::here("figures/3met_long.jpeg"))
+```
+
 # Data Availability

### 2022-04-27 analysis/05_model_metrics.Rmd

@@
                n_ts = n_distinct(fit_data$rarefyID),
+               formula = formula,
                warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
@@
              n_ts = n_distinct(fit_data$rarefyID),
+             formula = formula,
              warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
@@
 ```{r}
+#manuscript table
 metric_model_table %>% 
@@
   write_csv(here::here("figures", "man_model_table.csv"))
+
+#supplement of complete model output
+metric_model_table %>%
+  select(-c(model_id, n_ts, warning, lower.CL, upper.CL, significant)) %>%
+  write_csv(here::here("paper", "model_estimates_supp.csv"))
+
 ```

### 2022-04-27 analysis/05_model_metrics.Rmd
Make sure variable names propogate through

@@
 
-# metric_model_table <- bind_rows(mod_results %>% filter(is.na(warning)), 
-#           bad_mod_results %>% filter(is.na(warning))) %>%
-#   mutate(significant = case_when(
-#     p.value > 0.05 | lower.CL < 0 & upper.CL > 0 ~ FALSE,
-#     !is.na(p.value) | !is.na(lower.CL) ~ TRUE,
-#     TRUE ~ NA
-#   ))
-
 #models that still won't converge
@@
 
-metric_model_table <- bind_rows(mod_results %>% filter(is.na(warning)), 
+model_table <- bind_rows(mod_results %>% filter(is.na(warning)), 
           bad_mod_results %>% filter(is.na(warning)),
@@
     TRUE ~ NA
-  )) %>%
-  select(-model)
+  )) #%>%
+  #select(-model)
+
+metric_model_table <- model_table %>% select(-model)
 
@@
 
-model_df <- metric_model_table %>%
+model_df <- model_table %>%
   filter(model_id == "base") %>%
@@
 #### Analizing slopes
-Is slope a function of timeseries characteristics?
+Is slope a function of time series characteristics?
 ```{r}
@@
 #All model estimates in one dataframe
-slope_models <- bind_rows(duration_slope, startyear_slope)
-pins::pin(slope_models, board = "github")
+# slope_models <- bind_rows(duration_slope, startyear_slope)
+# pins::pin(slope_models, board = "github")
 

### 2022-04-27 analysis/02_rarefy_timeseries.Rmd
scrap the pins approach

@@
 ```{r}
-pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
+load(here::here("data/bt_traitfiltered.rda"))
 
-##	Get trait filtered data, rename colums to jive with the rest of the script
-bt_traitfiltered <- pins::pin_get("bt-traitfiltered", board = "github") %>%
+##	Get trait filtered data, rename columns to jive with the rest of the script
+bt_traitfiltered <- bt_traitfiltered
   rename_with(toupper) %>%
@@
 
-trait_ref <- pins::pin_get("trait-ref", board = "github")
+load(here::here("data/trait_ref.rda"))
 ```

### 2022-04-27 analysis/04_collate_rarefied_null.Rmd
scrap the pins approach

@@
 ## get null models
-null_table <- pins::pin_get("null-table", board = "github")
+load(here::here("data/null_table.rda"))
 ```
@@
 usethis::use_data(rarefied_metrics)
-
-pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
-pins::pin(rarefied_metrics, board = "github")
 ```

### 2022-04-27 paper/manuscript.Rmd
Fix zenodo urls

@@
 
-Code for the analyses in this chapter is archived on Zenodo at \<[[DOI:10.5281/zenodo.5514335\\\\](DOI:10.5281/zenodo.5514335){.uri}](%5BDOI:10.5281/zenodo.5514335%5D(DOI:10.5281/zenodo.5514335)%7B.uri%7D){.uri}\> [\@norman_karinormanbiodivts_2021].
-Data products are on Github at <https://github.com/karinorman/biodivTS_data>.
+Code for the analyses in this chapter is archived on Zenodo at [10.5281/zenodo.5514334](https://doi.org/10.5281/zenodo.5514334).
+Data products are also archived on Zenodo at [10.5281/zenodo.6499442](https://doi.org/10.5281/zenodo.6499442).
 Original data sources are open access and available at their respective providers.

### 2022-04-27 analysis/01_filter_by_traits.Rmd

@@
 ```{r}
-pins::board_register_github(name = "github", repo = "karinorman/biodivTS_data", branch = "master")
+#load data, objects named the same as the filename
 
-biotime_data <- pins::pin_get("biotime-data", board = "github")
-elton_mamm <- pins::pin_get("elton-mamm", board = "github")
-elton_bird <- pins::pin_get("elton-bird", board = "github")
-amphibio <- pins::pin_get("amphibio", board = "github")
+load(here::here("data/biotime_data.rda")) #load biotime_data
+load(here::here("data/elton_mamm.rda")) #load elton_mamm
+load(here::here("data/elton_bird.rda")) #load elton_bird
+load(here::here("data/amphibio.rda")) #load amphibio
 
@@
 usethis::use_data(bt_traitfiltered)
-pins::pin(bt_traitfiltered, board = "github")
 ```
@@
 usethis::use_data(trait_ref)
-pins::pin(trait_ref, board = "github")
 ```
@@
 ```{r}
-metadata <- pins::pin_get("metadata", board = "github")
+load(here::here("data/meta.rda")) #load meta
 
@@
   count(study_id, name = "n_plots") %>% 
-  left_join(metadata %>% select(study_id, realm, climate, habitat, biome_map, taxa, 
+  left_join(meta %>% select(study_id, realm, climate, habitat, biome_map, taxa, 
                                 organisms, start_year, end_year), by = "study_id") %>%
@@
 #save locally and in cache
-usethis::use_data(study_table)
-pins::pin(study_table, board = "github")
+#usethis::use_data(study_table)
 ```

### 2022-04-27 analysis/03_rarefy_null_models.Rmd
Add description of script contents.

@@
 
+This script calculates n null samples for each rarefied samples, calculates the metrics for those null samples, and outputs a table of the null model statistics.
+
 ```{r}
@@
 library(furrr)
-
-pins::board_register_github(repo = "karinorman/biodivTS_data", branch = "master")
 ```
@@
 ```{r}
-bt_traitfiltered <- pins::pin_get("bt-traitfiltered", board = "github")
-trait_ref <- pins::pin_get("trait-ref", board = "github")
+load(here::here("data/bt_traitfiltered.rda"))
+load(here::here("data/trait_ref.rda"))
 
@@
 usethis::use_data(null_table)
-pins::pin(null_table, board = "github")
 ```

### 2022-04-27 analysis/05_model_metrics.Rmd

@@
 ```{r}
-pins::board_register_github(repo = "karinorman/biodivTS_data", branch = "master")
+#get data
+load(here::here("data/meta.rda"))
+load(here::here("data/bt_traitfiltered.rda"))
+load(here::here("data/rarefied_metrics.rda"))
 
-#get metadata
-meta <- pins::pin_get("meta", board = "github") %>%
+meta <- meta %>%
   select(study_id, realm, climate, habitat, protected_area, biome_map)
 
-meta <- pins::pin_get("bt-traitfiltered", board = "github") %>%
+meta <- bt_traitfiltered %>% 
   select(study_id, rarefyid, taxa) %>% distinct() %>%
@@
 
+rm(bt_traitfiltered)
+
 #get metric data
-metrics <- pins::pin_get("rarefied-metrics", board = "github") %>%
-  # pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
+metrics <- rarefied_metrics %>%
   rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
@@
 
-pins::pin(model_data, board = "github")
+usethis::use_data(model_data)
 ```
@@
 usethis::use_data(metric_model_table)
-pins::pin(metric_model_table, board = "github")
 ```
@@
 re_slopes <- pmap_dfr(model_df, get_study_slopes)
-pins::pin(re_slopes, board = "github")
 
@@
 # slope_models <- bind_rows(duration_slope, startyear_slope)
-# pins::pin(slope_models, board = "github")
 

### 2022-04-27 data-raw/02_biotime_ids.Rmd

@@
 #file created by 01_biotime_grid.R
-pins::board_register_github(repo = "karinorman/biodivTS_data", branch = "master", name = "github")
-bt_grid <- pins::pin_get("bt-grid", board = "github")
+load(here::here("data/bt_grid_collate.rda"))
 
-data <- bt_grid %>%
+data <- bt_grid_collate %>%
   rename_with(tolower) %>%
@@
 
-rm(bt_grid)
+rm(bt_grid_collate)
 ```
@@
 usethis::use_data(biotime_data)
-
-#save to cache
-pins::pin(biotime_data, board = "github")
 ```

### 2022-04-27 paper/manuscript.Rmd
little things

@@
 The lack of trend held even after correcting for changes in species richness and at the study-level, where only 3 of 54 studies experienced a significant trend in at least one functional diversity metric.
-Results give evidence that across a variety of taxa, climates, and biomes, these selected functional characteristics are maintained even in the face of significant environmental and community change.
-We also discuss the potential for underlying functional shifts to be obscured by current approaches and data and call for targeted data collection efforts to combat existing biases in monitoring and trait data.
+Results give evidence that across a variety of taxa, climates, and biomes, functional characteristics are maintained even in the face of significant environmental and community change.
+We also discuss the potential for underlying functional shifts to be obscured by current approaches and data, calling for targeted data collection efforts to combat existing biases in monitoring and trait data.
 
@@
 We assessed the impact of time series duration and start year on study-level trends using general linear models with duration and start year as predictors.
-All models were executed using the *lme4* package in R and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* package with a significance level of \$\\alpha = 0.05\$ [@bates2015; @kuznetsova2017; @rcoreteam2021].
+All models were executed using the *lme4* package in R and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2021].
 

### 2022-04-28 paper/manuscript.Rmd
final stuff

@@
 
-It is critical to understand how ecological communities are responding to the ongoing biodiversity crisis.
-Previous broad-scale assessments of community change through time have focused on metrics of species identity, which do not account for differences between species.
-Functional trait approaches offer a powerful alternative to species-based approaches by directly assessing changes in the species characteristics that define their ecological role.
+How are ecological communities responding to anthropogenic change?
+Functional trait approaches allow us to assess this questions by directly measuring shifts in the characteristics that define a species' ecological role.
 Here, we pair vertebrate community time series with functional trait data to take a first look at temporal trends in functional composition across biomes, realms, and taxa.
 We found no evidence of systematic functional loss across communities, with maintenance of all measured aspects of functional structure.
-We also identified key methodological holes that may be obscuring a potential underlying trend.
+We also highlight the need for targeted data collection and methodological expansion to further assess functional trends.
 
@@
 We found similar results for the *CWM* models, with a significant trend for only one trait, percentage of aerial foraging utilization, which showed a significant positive trend.
+
+```{r timeseriesPlot, fig.cap="Plots of time series-level trends with line color corresponding to climatic region, with data points in grey and the overall metric mean in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES", echo=FALSE}
+knitr::include_graphics(here::here("figures/3met_long.jpeg"))
+```
 
@@
 
-```{r timeseriesPlot, fig.cap="Plots of time series-level trends with line color corresponding to climatic region, with data points in grey and the overall metric mean in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES", echo=FALSE}
-knitr::include_graphics(here::here("figures/3met_long.jpeg"))
-```
-
 # Data Availability
@@
 We thank Dr Shane Blowes and Dr Sarah Supp, whose code we adapted for initial data processing of the BioTIME database.
-Support for this work was provided by U.S. Department of Energy through the Computational Sciences Graduate Fellowship (DE-FG02-97ER25308) awarded to K.E.A.N..
+Support for this work was provided by U.S.
+Department of Energy through the Computational Sciences Graduate Fellowship (DE-FG02-97ER25308) awarded to KEAN.
+CB acknowledges support from NSF CAREER Award #1942280 & NIFA project CA-B-INS-0162-H.
 

### 2022-05-13 paper/manuscript.Rmd
Anonymize

@@
 title: Evaluating the evidence of widespread maintenance of functional composition in vertebrate communities
-authors: 
-  - name: "Kari E. A. Norman"
-    affiliation: Université de Montréal
-    location: Montréal, Quebec
-    email: kari.norman@berkeley.edu
-  - name: "Perry de Valpine"
-    affiliation: University of California, Berkeley
-    location: Berkeley, California
-    email: pdevalpine@berkeley.edu
-  - name: "Carl Boettiger"
-    affiliation: University of California, Berkeley
-    location: Berkeley, California
-    email: cboettig@berkeley.edu
+# authors: 
+#   - name: "Kari E. A. Norman"
+#     affiliation: Université de Montréal
+#     location: Montréal, Quebec
+#     email: kari.norman@berkeley.edu
+#   - name: "Perry de Valpine"
+#     affiliation: University of California, Berkeley
+#     location: Berkeley, California
+#     email: pdevalpine@berkeley.edu
+#   - name: "Carl Boettiger"
+#     affiliation: University of California, Berkeley
+#     location: Berkeley, California
+#     email: cboettig@berkeley.edu
 abstract: |
@@
 
-# Data Availability
+<!-- # Data Availability -->
 
-Code for the analyses in this chapter is archived on Zenodo at [10.5281/zenodo.5514334](https://doi.org/10.5281/zenodo.5514334).
-Data products are also archived on Zenodo at [10.5281/zenodo.6499442](https://doi.org/10.5281/zenodo.6499442).
-Original data sources are open access and available at their respective providers.
+<!-- Code for the analyses in this chapter is archived on Zenodo at [10.5281/zenodo.5514334](https://doi.org/10.5281/zenodo.5514334). -->
+<!-- Data products are also archived on Zenodo at [10.5281/zenodo.6499442](https://doi.org/10.5281/zenodo.6499442). -->
+<!-- Original data sources are open access and available at their respective providers. -->
 

### 2022-07-28 analysis/figures/visualization.Rmd
Options in comments for poster size text, fix some typos.

@@
   theme(legend.title  = element_blank()) +
+  #theme(legend.title  = element_blank(), legend.text = element_text(size = 12)) +
   guides(color = guide_legend(override.aes= list(alpha = 0.8)))
@@
   theme_classic() + 
+  #theme_classic(base_size = 12) + 
   facet_wrap(~taxa, scales = "free") +
-  xlab("Timeseries Duration") +
+  xlab("Time Series Duration") +
   theme(strip.background = element_blank(), 
@@
         legend.position = "none", 
-        axis.title = element_text(size=8)) +
+        axis.title = element_text(size=8),
+        ) +
   scale_fill_manual(values = my_pal)
@@
     theme_classic() + 
+    #theme_classic(base_size = 12) + 
+    #theme(legend.position = "none") +
     theme(axis.title = element_text(size=7), legend.position = "none") +
@@
 
-model_coef <- metric_model_table %>%
-  filter(model_id == "base") %>%
-  select(metric, model) %>%
-   mutate(coef = map(model, broom.mixed::tidy)) %>% 
-  select(-model) %>% 
-  unnest(cols = c(coef))
-
 group_colors <- c("Global" = my_pal[1], "Polar/Temperate" = my_pal[6], "Temperate/Tropical" = my_pal[3], "Tropical" = my_pal[4], "Temperate" = my_pal[2], "Overall Mean" = "#000000")
@@
            ylabel = c("log(S)", "Jaccard", "Functional Richness SES", "Functional Divergence SES", "Functional Evenness SES"),
-           #ylabel = c("log(S)", "Jaccard", "FRic SES", "FDiv SES", "FEve SES"),
+           #ylabel = c("log(S)", "Jaccard", "Functional\nRichness SES", "Functional\nDivergence SES", "Functional\nEvenness SES"),
            yval = c("logvalue", "value", "value", "value", "value"),
            xlabel = c(NA, NA, NA, "year", "year"))
-m_plots <- pmap(plot_map, plot_metric, data = model_data, model_coef = model_coef, color_list = group_colors)
+m_plots <- pmap(plot_map, plot_metric, data = model_data, 
+                model_coef = metric_model_table %>% filter(model_id == "base"), 
+                color_list = group_colors)
 
@@
   theme(axis.title = element_text(size=7)) +
-  scale_colour_manual(values = group_colors, drop = FALSE)
+  #theme(legend.title = element_text(size = 12), legend.text = element_text(size = 12)) +
+  scale_colour_manual(values = group_colors, drop = FALSE, name = "Climate")
 

### 2022-07-28 analysis/05_model_metrics.Rmd
Found some little bugs

@@
   #this study has many low species richness observations, so FD metrics weren't calculated, just drop it
-  filter(study_id != 348)
+  filter(study_id != 348) %>%
+  #somehow the taxa labels didn't get updated when the timeseries for multi-taxa studies were split up, do that now
+   mutate(new_taxa = str_split_fixed(rarefyID, "_", n = 3)[,3],
+            taxa = case_when(
+            new_taxa == "bird" ~ "Birds",
+               new_taxa == "mamm" ~ "Mammals", 
+               TRUE ~ as.character(taxa)
+           )) %>%
+  select(-new_taxa)
 

### 2022-07-28 data-raw/01_biotime_grid.R

@@ -106,7 +106,7 @@ res <- dg_closest_res_to_area(dgg, SL_extent_mean+SL_extent_sd)
 dgg <- dgsetres(dgg, res)
 
 ##	get the corresponding grid cells for all observations
-bt <- bt %>% mutate(cell = dgtransform(dgg, lat=lat_to_grid, lon=lon_to_grid))
+bt <- bt %>% mutate(cell = dgGEO_to_SEQNUM(dgg, in_lon_deg=lon_to_grid, in_lat_deg=lat_to_grid)$seqnum)
 
 ##	what just happened?
 check <- bt %>% group_by(StudyMethod, STUDY_ID) %>% summarise(n_cell = n_distinct(cell))

### 2022-07-28 data-raw/01_biotime_grid.R
save that guy for the mapping figure

@@ -217,6 +217,7 @@ rarefyID_cell_centre <- bind_rows(rarefyID_cell_centre, SL_coords, ML_coords)
 ##	not sure why but I have multiple entries for each rarefyID!
 rarefyID_cell_centre <- rarefyID_cell_centre %>% distinct(STUDY_ID, rarefyID, cell_extent, rarefyID_x, rarefyID_y)
 
+usethis::use_data(rarefyID_cell_centre)
 ##======================================================================
 ##	reduce to data required for rarefying, and rename a couple of columns
 ##	NB: we will rarefy to the smallest number of ObsEventID's within studies

### 2022-08-30 analysis/05_model_metrics.Rmd
minor changes to visualizations

@@
 ```{r}
-hist_numYears <- metrics %>%
+hist_numYears <- model_data %>%
   select(year, rarefyID) %>% 
@@
   theme(panel.grid.minor.x = element_blank())
+
+hist_duration <- model_data %>%
+  select(rarefyID, duration) %>% 
+  distinct() %>% 
+  ggplot() +
+  geom_histogram(aes(x = duration),
+                 binwidth = 2) +
+  scale_x_continuous(name = 'Number of years sampled',
+                     breaks = c(2,4,8,16,32,48,64,96)) +
+  #scale_y_continuous(breaks = c(0,2500,5000,7500,10000,12500, 15000, 17500, 20000,22500)) +
+  labs(tag = 'B',
+       y = 'Number of cells') +
+  theme_bw() +
+  theme(panel.grid.minor.x = element_blank())
 ```
@@
 ```{r}
-metrics %>%
-  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_hind")) %>%
-  ggplot() +
-  geom_histogram(aes(x = logvalue),
-                 bins = 60) +
-  facet_wrap(~metric, scales = "free") +
-  labs(title = "Distribution of Log values")
-
-metrics %>%
-  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "SES_FDis", "SES_FDiv", "SES_FEve", "SES_FRic", "Jaccard_hind")) %>%
-  ggplot() +
-  geom_histogram(aes(x = value),
-                 bins = 60) +
-  facet_wrap(~metric, scales = "free") +
-  labs(title = "Distribution of raw values")
+# metrics %>%
+#   filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_hind")) %>%
+#   ggplot() +
+#   geom_histogram(aes(x = logvalue),
+#                  bins = 60) +
+#   facet_wrap(~metric, scales = "free") +
+#   labs(title = "Distribution of Log values")
+# 
+# metrics %>%
+#   filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "SES_FDis", "SES_FDiv", "SES_FEve", "SES_FRic", "Jaccard_hind")) %>%
+#   ggplot() +
+#   geom_histogram(aes(x = value),
+#                  bins = 60) +
+#   facet_wrap(~metric, scales = "free") +
+#   labs(title = "Distribution of raw values")
 ```

### 2022-08-30 analysis/06_sensitivity_analysis.Rmd
first commit

--- 2022-04-27 analysis/05_model_metrics.Rmd
@@
 ---
-title: "Models"
+title: "Sensitivity Analysis"
 output: html_document
+date: "2022-08-30"
 ---
@@
 library(emmeans)
-```
-#### Read in data, get final dataset for models
-Get data and put it in long form 
-```{r}
-#get data
-load(here::here("data/meta.rda"))
-load(here::here("data/bt_traitfiltered.rda"))
-load(here::here("data/rarefied_metrics.rda"))
 
-meta <- meta %>%
-  select(study_id, realm, climate, habitat, protected_area, biome_map)
-
-meta <- bt_traitfiltered %>% 
-  select(study_id, rarefyid, taxa) %>% distinct() %>%
-  left_join(meta)
-
-rm(bt_traitfiltered)
-
-#get metric data
-metrics <- rarefied_metrics %>%
-  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
-  mutate(logvalue = log(value + 1)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
-  left_join(meta, c("rarefyID" = "rarefyid"))
-
-## Now the data that actually makes it into the model ##
-
-#get rarefyID's that don't have enough null samples
-missing_null <- bind_rows(metrics %>% 
-  filter(n_missing_nulls == 1, commplete_null_samps == FALSE),
-  metrics %>% filter(n_missing_nulls > 100)) %>%
-  pull(rarefyID) %>%
-  unique()
-
-model_data <- metrics %>% 
-  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "Jaccard_next",
-                       "FRic", "FEve", "FDiv", "S","CWM_bodymass_value","CWM_diet_inv", 
-                       "CWM_diet_scav", "CWM_diet_vfish", "CWM_diet_vunk", "CWM_forstrat_ground", 
-                       "CWM_forstrat_wataroundsurf", "CWM_forstrat_watbelowsurf", "CWM_diet_fruit", 
-                       "CWM_diet_planto", "CWM_diet_seed", "CWM_diet_vend", "CWM_forstrat_understory", 
-                       "CWM_diet_vect", "CWM_diet_nect", "CWM_forstrat_aerial", "CWM_forstrat_canopy", 
-                       "CWM_forstrat_midhigh", "CWM_age_at_maturity_max_y", "CWM_age_at_maturity_min_y", 
-                       "CWM_body_size_mm", "CWM_litter_size_max_n", "CWM_litter_size_min_n", 
-                       "CWM_longevity_max_y", "CWM_offspring_size_max_mm", "CWM_offspring_size_min_mm"),
-         !rarefyID %in% missing_null) %>%
-  group_by(metric) %>%
-  mutate(study_id =  str_extract(rarefyID, "[^_]+")) %>%
-  mutate(year_scaled = scale(year), 
-         scale_center = attributes(year_scaled)$`scaled:center`, 
-         scale = attributes(year_scaled)$`scaled:scale`) %>%
-  ungroup() %>%
-  #this study has many low species richness observations, so FD metrics weren't calculated, just drop it
-  filter(study_id != 348)
-
-usethis::use_data(model_data)
+load(here::here("data/model_data.rda"))
 ```
 
-Write out file with all necessary metadata for the supplement.
-```{r}
-meta %>% 
-  filter(study_id %in% model_data$study_id) %>% 
-  select(-c(link_id, date_study_added, sample_desc_name)) %>% 
-  write_csv(here::here("paper", "study_metadata.csv"))
-```
+Let's test the sensitivity of the general trends (and study level patterns?) to major data processing decisions.
 
-####Data exploration
-
-Visualize number of years for each timeseries
-```{r}
-hist_numYears <- metrics %>%
-  select(year, rarefyID) %>% 
-  distinct() %>% 
-  count(rarefyID, name = "num_years") %>% 
-  ggplot() +
-  geom_histogram(aes(x = num_years),
-                 binwidth = 2) +
-  scale_x_continuous(name = 'Number of years sampled',
-                     breaks = c(2,4,8,16,32,48,64,96)) +
-  #scale_y_continuous(breaks = c(0,2500,5000,7500,10000,12500, 15000, 17500, 20000,22500)) +
-  labs(tag = 'B',
-       y = 'Number of cells') +
-  theme_bw() +
-  theme(panel.grid.minor.x = element_blank())
-```
-
-Visualize distributions of response
-```{r}
-metrics %>%
-  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_hind")) %>%
-  ggplot() +
-  geom_histogram(aes(x = logvalue),
-                 bins = 60) +
-  facet_wrap(~metric, scales = "free") +
-  labs(title = "Distribution of Log values")
-
-metrics %>%
-  filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "SES_FDis", "SES_FDiv", "SES_FEve", "SES_FRic", "Jaccard_hind")) %>%
-  ggplot() +
-  geom_histogram(aes(x = value),
-                 bins = 60) +
-  facet_wrap(~metric, scales = "free") +
-  labs(title = "Distribution of raw values")
-```
-
-####Fit GLMs to overall patterns and data broken down by groups
-
-Convergence problems! See here:
-https://joshua-nugent.github.io/allFit/
-https://rstudio-pubs-static.s3.amazonaws.com/33653_57fc7b8e5d484c909b615d8633c01d51.html
-https://biologyforfun.wordpress.com/2018/04/09/help-i-have-convergence-warnings/
-
-Model checks:
-https://debruine.github.io/posts/normality/
-performance::check_model()
-
-Function approach to model
+Model fitting function
 ```{r}
@@
 }
-#get_model(data = model_data, metric_name = "SES_FRic", "value ~ year_scaled + (year_scaled|study_id/rarefyID) + (year_scaled|climate)", model_id = "base")
 ```
 
-Create dataframe of models to fit
+1. Min time series duration
 ```{r}
-metric_names <- unique(model_data$metric)
-raw_value_mods <- tibble(metric_name = metric_names[metric_names != "S"], 
-                         base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
-                         taxa = "value ~ year_scaled * taxa + (year_scaled|study_id/rarefyID)",
-                         realm = "value ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
-                         climate = "value ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
-  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula") %>%
-  #don't fit the CWM_litter_size_min_n and climate model, not enough data
-  filter(metric_name != "CWM_litter_size_min_n" | model_id != "climate")
+dur_dfs <- list(min3 = filter(model_data, duration > 2),
+                min4 = filter(model_data, duration > 3),
+                min5 = filter(model_data, duration > 4))
 
-log_value_mods <- tibble(metric_name = c("S"), 
-                         base = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
-                         taxa = "logvalue ~ year_scaled * taxa + (year_scaled|study_id/rarefyID)",
-                         realm = "logvalue ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
-                         climate = "logvalue ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
-  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
-
-mods <- bind_rows(raw_value_mods, log_value_mods) %>% 
-  select(metric_name, formula, model_id)
+duration_sens <- map_dfr(dur_dfs, ~pmap_dfr(mods, get_model, data = .x), .id = "dataID")
 ```
 
-Fit models
+2. Inclusion of Amphibians - nothing changes when you exclude them 
 ```{r}
-#Fit models with base settings
-mod_results <- pmap_dfr(mods, get_model, data = model_data)
+no_amph <- pmap_dfr(mods, get_model, data = model_data %>% filter(taxa != "Amphibians")) %>%
+  mutate(dataID = no_amph)
 
-#get the models that had convergence warnings
-bad_models <- mod_results %>%
-  select(metric, model_id, warning) %>% 
-  filter(!is.na(warning)) %>% 
-  distinct() %>%
-  left_join(mods, by = c("metric" = "metric_name", "model_id"))
+# this annoyingly doesn't work because of some dataframe column
+# sens_models <- bind_rows(bind_cols(select(duration_sens, -dataID), pull(duration_sens, dataID)) %>% select(-model, -formula), 
+#                          no_amph %>% select(-model, -formula))
 
-#Turn off REML for these models, and they fit! 
-bad_mod_results <- pmap_dfr(bad_models %>% select(-warning), get_model, data = model_data, REML = FALSE)
-
-#models that still won't converge
-bad_models2 <- bad_mod_results %>%
-  select(metric, model_id, warning) %>% 
-  filter(!is.na(warning)) %>% 
-  distinct() %>%
-  left_join(mods, by = c("metric" = "metric_name", "model_id"))
-
-#Fit with a different algorithm
-bad_mod_results2 <- pmap_dfr(bad_models2 %>% select(-warning), get_model, data = model_data, control = lmerControl(optimizer = "bobyqa"))
-
-model_table <- bind_rows(mod_results %>% filter(is.na(warning)), 
-          bad_mod_results %>% filter(is.na(warning)),
-          bad_mod_results2 %>% filter(is.na(warning))) %>%
-  mutate(significant = case_when(
-    p.value > 0.05 | lower.CL < 0 & upper.CL > 0 ~ FALSE,
-    !is.na(p.value) | !is.na(lower.CL) ~ TRUE,
-    TRUE ~ NA
-  )) #%>%
-  #select(-model)
-
-metric_model_table <- model_table %>% select(-model)
-
-usethis::use_data(metric_model_table)
+save(duration_sens, no_amph, file = here::here("data/sensitivity_models.rda"))
 ```
 
-Get list of models that couldn't be fit due to model limitations 
+3. Trait Coverage
 ```{r}
-#models that were fit
-uniq_metric_model <- metric_model_table %>% select(metric, model_id) %>% distinct()
-#models we attempted to fit
-uniq_mods <- mods %>% select(metric = metric_name, model_id) %>% distinct()
-
-#this is a list of models that weren't fit because there weren't enough timeseries OR they didn't converge
-#min_litter_size just won't converge
-missing_mods <- setdiff(uniq_mods, uniq_metric_model)
-```
-
-Write out table of model estimates for manuscript
-```{r}
-#manuscript table
-metric_model_table %>% 
-  filter(model_id == "base", metric %in% c("S", "Jaccard_base", "SES_FRic", "SES_FEve", "SES_FDiv")) %>% 
-  select(-c(model_id, warning, statistic, df)) %>% 
-  write_csv(here::here("figures", "man_model_table.csv"))
-
-#supplement of complete model output
-metric_model_table %>%
-  select(-c(model_id, n_ts, warning, lower.CL, upper.CL, significant)) %>%
-  write_csv(here::here("paper", "model_estimates_supp.csv"))
 
 ```
-
-#### Process Model Output
-
-Get study level slope estimates
-```{r}
-get_study_slopes <- function(metric, model, model_id){
-
-  filter_name <- ifelse(model_id == "base", "study_id", model_id)
-
-  #get the statistics for the year variable
-  year_scaled_coef <- tidy(model) %>%
-    filter(term == "year_scaled")
-
-  #get the year estimate
-  year_scaled_est <- pull(year_scaled_coef, estimate)
-
-  #get year variation
-  year_scaled_var <- year_scaled_coef %>% 
-    mutate(var = std.error*std.error) %>%
-    pull(var)
-
-  #get study level conditional estimates and add to over all estimates
-  study_ests <- broom.mixed::tidy(model, effects="ran_vals") %>%
-    filter(group == filter_name, term == "year_scaled") %>% 
-    select(-c(effect, group, term)) %>%
-    rename(cond.std.error = std.error, cond.estimate = estimate) %>%
-    mutate(estimate = cond.estimate + year_scaled_est,
-      cond.var = cond.std.error*cond.std.error, 
-           var = cond.var + year_scaled_var,
-           std.error = sqrt(var),
-           upr.ci = estimate + (2*std.error),
-           lwr.ci = estimate - (2*std.error),
-           sig = case_when(
-             lwr.ci < 0 & upr.ci > 0 ~ FALSE,
-             TRUE ~ TRUE
-           ),
-      metric = metric,
-      model_id = model_id)
-}
-
-model_df <- model_table %>%
-  filter(model_id == "base") %>%
-  select(metric, model, model_id) %>%
-  group_by(metric, model_id) %>% 
-  slice_head()
-
-#get study level slopes for base, and levels for categorical random effect in non-base models
-re_slopes <- pmap_dfr(model_df, get_study_slopes)
-
-#get only study-level slopes for the base models
-study_slopes <- re_slopes %>% 
-  filter(model_id == "base") %>%
-  rename(study_id = level)
-
-# study_slopes_wide <- study_slopes %>%
-#   select(study_id, metric, estimate) %>%
-#   pivot_wider(names_from = "metric", values_from = "estimate")
-```
-
-Summaries of study slopes
-```{r}
-#First - how many studies had a significant trend for each metric? 
-study_slopes %>% 
-  filter(sig == TRUE) %>% 
-  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
-  group_by(metric) %>% 
-  summarise(n_study_id = n_distinct(study_id)) %>%
-  mutate(percent = n_study_id/53)
-
-#And which studies were they?
-study_slopes %>% 
-  filter(sig == TRUE) %>% 
-  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
-  select(study_id, metric) %>%
-  distinct()
-
-#Which were significant for multiple metrics?
-study_slopes %>% 
-  filter(sig == TRUE) %>% 
-  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
-  group_by(study_id) %>%
-  filter(n() > 1) %>%
-  select(study_id, metric)
-```
-
-#### Analizing slopes
-Is slope a function of time series characteristics?
-```{r}
-library(multcomp)
-
-study_slope_data <- study_slopes %>%
-  dplyr::select(study_id, metric, estimate) %>%
-  left_join(meta %>% 
-              dplyr::select(-rarefyid) %>% 
-              distinct() %>%
-              mutate(study_id = as.character(study_id))) %>%
-  left_join(model_data %>% dplyr::select(study_id, startyear, duration) %>% distinct())
-
-duration_slope <- study_slope_data %>%
-  group_nest(metric) %>%
-  mutate(model = map(data, ~glm(estimate ~ duration, 
-                                 data = .x)),
-           coef = map(model, tidy),
-           ) %>% 
-  dplyr::select(metric, coef) %>%
-  unnest(cols = c(coef)) %>%
-  mutate(covariate = "duration")
-
-startyear_slope <- study_slope_data %>%
-  group_nest(metric) %>%
-  mutate(model = map(data, ~glm(estimate ~ startyear, 
-                                 data = .x)),
-           coef = map(model, tidy),
-           ) %>% 
-  dplyr::select(metric, coef) %>%
-  unnest(cols = c(coef)) %>%
-  mutate(covariate = "startyear")
-
-#All model estimates in one dataframe
-# slope_models <- bind_rows(duration_slope, startyear_slope)
-
-# Which predictor variables were significant?
-slope_models %>% filter(term != "(Intercept)", p.value < 0.05) %>% View()
-```

### 2022-08-30 analysis/06_sensitivity_analysis.Rmd
add models to the actual sensitivity script

@@
 
+Define models to fit
+```{r}
+metric_names <- unique(model_data$metric)
+raw_value_mods <- tibble(metric_name = metric_names[metric_names != "S"], 
+                         base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
+                         taxa = "value ~ year_scaled * taxa + (year_scaled|study_id/rarefyID)",
+                         realm = "value ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
+                         climate = "value ~ year_scaled * climate + (year_scaled|study_id/rarefyID)"
+                         ) %>%
+  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula") %>%
+  #don't fit the CWM_litter_size_min_n and climate model, not enough data
+  filter(metric_name != "CWM_litter_size_min_n" | model_id != "climate")
+
+log_value_mods <- tibble(metric_name = c("S"), 
+                         base = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
+                         taxa = "logvalue ~ year_scaled * taxa + (year_scaled|study_id/rarefyID)",
+                         realm = "logvalue ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
+                         climate = "logvalue ~ year_scaled * climate + (year_scaled|study_id/rarefyID)"
+                         ) %>%
+  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
+
+mods <- bind_rows(raw_value_mods, log_value_mods) %>% 
+  select(metric_name, formula, model_id)
+```
+
 1. Min time series duration
@@
 ```{r}
+load(here::here("data/study_table.rda"))
+
+model_data <- model_data %>%
+  left_join(study_table %>% select(study_id, coverage) %>% mutate(study_id = as.factor(study_id)))
+
+covg_dfs <- list(covg75 = filter(model_data),
+                covg85 = filter(model_data, coverage >= 0.85),
+                covg90 = filter(model_data, coverage >= 0.90))
+
+coverage_sens <- map_dfr(covg_dfs, ~pmap_dfr(mods %>% 
+                                               filter(metric_name %in% c("FRic", "FDiv", "FEve", "S", "SES_FRic",
+                                                                         "SES_FDiv", "SES_FEve"), 
+                                                      model_id == "base"), 
+                                             get_model, data = .x), .id = "dataID")
 

### 2022-09-19 analysis/01_filter_by_traits.Rmd
add models to the actual sensitivity script
This is what I really mean, it doesn't effect the final dataset though because all the studies I accidentally excluded were super low richness

@@
 #studies to include, since they meet the coverage requirement 
-inc_coverage <- coverage %>% filter(perc > percent)
+inc_coverage <- coverage %>% filter(perc >= percent)
 
@@
 #save locally and in cache
-#usethis::use_data(study_table)
+usethis::use_data(study_table)
 ```

### 2022-09-19 paper/manuscript_GEB.Rmd
first commit

--- 2022-05-13 paper/manuscript.Rmd
@@
 ---
-title: Evaluating the evidence of widespread maintenance of functional composition in vertebrate communities
+title: Evaluating the evidence of widespread maintenance of functional composition in bird and mammal communities
 # authors: 
@@
 We also discuss the potential for underlying functional shifts to be obscured by current approaches and data, calling for targeted data collection efforts to combat existing biases in monitoring and trait data.
+ Aim, Location, Time period, Major taxa studied, Methods, Results, Main conclusions
 
@@
 
-## Significance Statement {.unnumbered}
-
-How are ecological communities responding to anthropogenic change?
-Functional trait approaches allow us to assess this questions by directly measuring shifts in the characteristics that define a species' ecological role.
-Here, we pair vertebrate community time series with functional trait data to take a first look at temporal trends in functional composition across biomes, realms, and taxa.
-We found no evidence of systematic functional loss across communities, with maintenance of all measured aspects of functional structure.
-We also highlight the need for targeted data collection and methodological expansion to further assess functional trends.
-
 # Introduction
@@
 
+Debate over the meaing of general trends, their robustness, and relevance for conservation action is ongoing.
+
 Functional diversity offers a potentially powerful alternative for detecting and describing community change by providing a mechanistic link between species' response to environmental change (*response traits*) and the processes they perform (*effect traits*) [@lavorel2002; @mcgill2006; @suding2008].
@@
 Indeed, almost all hypothesized mechanisms underpinning the relationship between species diversity and ecosystem function are trait-dependent [@hillebrand2009].
-Determining functional trends therefore gives a more fundamental picture of potential trends in critical ecosystem processes.
+Determining functional trends therefore gives a more direct picture of potential trends in critical ecosystem processes.
 
 It is critical to establish whether or not functional loss is prevalent across communities.
-While functional loss is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], it is not necessarily inevitable even in scenarios of species loss [@daz2001].
+While functional loss is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], local-scale loss is not necessarily inevitable even in scenarios of species loss [@daz2001; @petchey2009].
 Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020].
@@
 On paleoecological time scales functional composition shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with significant impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].
-Some losses of functional diversity are indisputable on both paleocological and contemporary timescales such as continued trophic downgrading due to loss of large-bodied mammals, but implications of those losses for local diversity patterns are less clear [@estes2011; @smith2018] .
+Some losses of functional diversity are indisputable on both paleocological and contemporary timescales such as continued trophic downgrading <!-- or is it just body-size downgrading?
+we lost mega grazers too --> due to loss of large-bodied mammals, but implications of those losses for local diversity patterns are less clear [@estes2011; @smith2018] .
 
@@
 Here we perform the first multi-taxa, multi-system assessment of functional diversity change through time.
-We focus on mammal, bird, and amphibian species as a significant subset of the world's biodiversity heavily impacted by anthropogenic change.
+We focus on mammal and bird species as a significant subset of the world's biodiversity heavily impacted by anthropogenic change.
 While examining trends in plants, invertebrates, and other vertebrate species is of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
-In order to ensure comparability across taxa in trait type and data quality we therefore focus on mammals, birds, and amphibians.
+In order to ensure comparability across taxa in trait type and data quality we therefore focus on mammals and birds.
 Traits were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].
+An initial assessment of amphibian trends is included in the supplement, but excluded from general trend assessment here due to limited geographic coverage.
 
-We assess thousands of mammal, bird, and amphibian functional diversity time series to determine whether or not there is a general trend of functional change, both in observed metrics and in metrics corrected for changes in species richness.
+We assess thousands of mammal and bird functional diversity time series to determine whether or not there is a general trend of functional change, both in observed metrics and in metrics corrected for changes in species richness.
 We distinguish between three possible scenarios of functional change: 1) significant loss of functional diversity with accompanying shifts in other functional metrics, 2) no functional diversity loss, but significant shifts in other functional metrics, 3) maintenance of functional diversity and composition.
 Based on expectation due to human impacts, we expect to find a significant functional loss with further restructuring indicated by the additional metrics.
+
+# Material and Methods
+
+## Data
+
+We obtained mammal and bird time series from the BioTIME database, a global repository of high quality assemblage time series.
+All studies included in the database follow consistent sampling protocols and represent full assemblages rather than populations of single species [@dornelas2018].
+All time series include abundance of observed species.
+Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual time series following a standardized spatial scale.
+Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018 for details on how sample extents were defined], with the area of each cell set to one standard deviation away from the mean of the single extent locations.
+All samples from a study within a single cell were considered to be a single time series, and species abundances were combined for all samples.
+
+We used trait data from the Elton Trait Database, which consists of species-level means for traits that represent species' multifaceted function in the community [@wilman2014].
+Traits include: body mass, diet, active diel period, nocturnality, forest foraging strata, pelagic use.
+Multiple traits (i.e. diet, foraging strata, activity seasonality, active diel period) were broken down into percentage or binary use for each level.
+
+In order to ensure taxonomic consistency across datasets, BioTIME species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (retrieved 09-15-2020 from the on-line database, <https://doi.org/10.5066/F7KH0KBK>), obtained through the `taxadb` R package [@norman2020; @rcoreteam2021].
+If more than one species in the assemblage data resolved to the same identifier, observations were considered the same species.
+For trait data, traits for all species of the same identifier were averaged.
+Only studies for which at least 75% of species had trait data were included.
+In order to have a sufficient number of species to calculate functional diversity metrics, years with fewer than 5 species observed were also excluded.
+
+Sensitivity analyses were conducted for the trait coverage threshold and the duration of included time series.
+
+Many studies had a variable number of samples within years.
+To account for this inconsistency in sampling effort we used sample-based rarefaction by bootstrap resampling within years for each time series based on the smallest number of samples in a year for that time series.
+
+Our final dataset included 2,432 time series from 50 studies in 21 countries and 12 biomes and 6 different traits (Fig \ref{fig:taxaMap}).
+The earliest sample was in 1923 and the most recent was in 2014.
+For a full breakdown of studies and their characteristics, see the supplement.
+
+```{r taxaMap, fig.cap="A) Map of time series locations with points colored by taxa, and B) histograms of time series duration broken down by taxa.", echo=FALSE}
+knitr::include_graphics(here::here("figures/study_map_hist.jpeg"))
+```
+
+## Diversity Metrics
+
+We calculated yearly metrics of functional and species diversity for each time series.
+Species-based metrics include species richness (*S*) and Jaccard similarity (*J*) as a measure of turnover.
+Jaccard similarity was calculated relative to the first observed year for a time series.
+A negative trend in *J* would therefore indicate increasing turnover.
+We did not impose a correction for unobserved species as non-parametric estimators do not assign species identities to corrected richness values, and therefore could not be propogated to the functional diversity metrics.
+
+Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package [@laliberté2010].
+Here we report functional richness (*FRic*), functional evenness (*FEve*), and functional divergence (*FDiv*) which together describe three complementary characteristics of the functional space [@mason2005; @hillebrand2009].
+*FRic* assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
+*FEve* describes how species are distributed across the trait space and how abundance is distributed across species.
+Higher values of *FEve* indicate more even spacing of species in the trait space and individuals across species.
+*FDiv* measures the degree to which species and their abundances maximize differences in the functional space.
+Higher values of *FDiv* therefore correspond to communities where many highly abundant species are on the edges of the trait space.
+We also calculated the community-weighted mean (*CWM*) of included traits to examine shifts in the distribution of each trait.
+
+All available trait data for each study were included in functional diversity calculations with the exception of traits that were the same value for all observed species in the study.
+For variables with multiple levels each level was included as a separate trait axis.
+Continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
+The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
+This restriction allows for a sufficient number of axes to capture the trait space while maintaining computational feasibility [@blonder2018].
+Metrics incorporated weighting based on species abundance where available (three studies were presence only).
+
+## Null Models
+
+To assess functional change independent of species richness we calculated the standardized effect size (SES) for each of the three summary functional diversity metrics (*FRic*, *FEve*, *FDiv*) from null estimates [@swenson2012].
+Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
+Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species.
+Species pools included all species observed for a time series.
+This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
+We used these values to calculate SES using the following formula: $SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
+We then calculated the median SES estimate for each metric from all the rarefaction samples for a time series.
+SES estimates can be interpreted as how much of the functional characteristic (richness, evenness, divergence) was observed beyond what was expected by chance for a community of that species richness.
+
+## Analysis
+
+We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study.
+All time series with data for a given trait were included in the model.
+Trends could not be estimated for the *CWM'*s of three traits due to limited data: maximum age at maturity, minimum age of maturity, and minimum litter size.
+We fit 24 individual *CWM* models.
+
+For all other species and functional diversity metrics (*S*, *J*, *FRic*, *FEve*, *FDiv)* we obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model.
+BLUP's provide estimates for the conditional mean and variance of a random effect from which we calculated 95% confidence intervals to determine significance of study-level slopes.
+
+To test for trends within and between different levels of taxa, biome, and realm we fit separate models with each of those factors added as a predictor to the original model structure.
+We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [@lenth2021].
+We assessed the impact of time series duration and start year on study-level trends using general linear models with duration and start year as predictors.
+All models were executed using the *lme4* package in R and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2021].
 
@@
   column_to_rownames("sign") %>%
-  kable(caption = "Number of studies that experienced a signficiant trend in each calculated metric out of 53 total studies.")
+  kable(caption = "Number of studies that experienced a significant trend in each calculated metric out of 53 total studies.")
 ```
@@
 
-Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at aggregate and local trends in functional diversity in mammal, bird, and amphibian communities.
+Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at aggregate and local trends in functional diversity in mammal and bird communities.
 Surprisingly, we did not detect an overall trend in any of the chosen functional diversity metrics.
@@
 
-This striking result could be a product of two possible processes, one ecological and one methodological.
+<!-- This striking result could be a product of two possible processes, one ecological and one methodological.
 Null trends appear to give strong evidence of systematic maintenance of functional structure due to common ecological processes, however multiple limitations of current approaches in synthesis could potentially be obscuring a true underlying global trend.
-We discuss both options further here.
+We discuss both options further here.-->
+
+We discuss here the possible ecological processes underpinning maintenance of functional diversity, particularly in a global change context.
 
@@
 On average, species additions have similar functional characteristics as lost species and therefore maintain the structure of the functional space.
+
+While observed communities are clearly losing species (as demonstrated by the turnover in species identity), sufficient diversity is being maintained in the species pool to provide functionally comparable species as replacement.
+This phenomena could be the result of two possible process, potentially working in concert.
+First, functional redundancy in the species pool maintains resilience of the functional space by providing multiple potential species for a any given set of functional traits.
+Alternatively, while there may be low redundancy in the functional space, functional composition in the species pool may be replenished by changes in the composition of the species pool to to shifts in species ranges.
+<!-- what does global change say about trajectory of functional redundancy -->
+
+-   what is the expectation for change in functional redundancy (and species pool changes) due to global change?
+
+-   don't have the data here to directly assess temporal shifts in the species pool per say, so we can't distinguish between these process, buy you could do it x,y,z way
+
+-   paleoecological studies indicate that functional composition can be largely conserved for long periods of time (strong ecological process), and losses are only seen after catastrophic events (e.g.
+terminal pleistocene extinction).
+
+-   discuss possibility of a tipping point?
+    catastrophic loss following long period of stability, what might the early warning signs be of that?
+
+<!-- change this discussion to reconciling with other long term trends and where the differences may arise -->
 
@@
 Instead, we found no evidence of a trend in *CWM* body size.
+<!-- in the communities we sampled we trophic downgrading has probably already occurred! -->
+
+-   discussion of expectations for undersampled areas (Africa particularly that still has a lot of its megafauna)?
 
@@
 
-# Material and Methods
-
-## Data
-
-We obtained mammal, bird, and amphibian time series from the BioTIME database, a global repository of high quality assemblage time series.
-All studies included in the database follow consistent sampling protocols and represent full assemblages rather than populations of single species [@dornelas2018].
-Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual time series following a standardized spatial scale.
-Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018 for details on how sample extents were defined], with the area of each cell set to one standard deviation away from the mean of the single extent locations.
-All samples from a study within a single cell were considered to be a single time series, and species abundances were combined for all samples.
-
-```{r traitSources, echo = FALSE, results = "asis"}
-tibble(Taxa = c("Mammals", "Birds", "Amphibians"),
-       Timeseries_Count = c(48, 2380, 11),
-       Species_Count = c(184, 700, 184),
-       Trait_Source = c("Elton Traits", "Elton Traits", "Amphibio"),
-       Traits = c("body mass, diet, active diel period", "body mass, diet, nocturnality, \nforest foraging strata, pelagic specialist", "habitat, diet, active diel period, \nactivity seasonality, body mass, body length, \nmin maturation size, max maturation size, \nmin offspring size, max offspring size, \nreproductive output, breeding strategy")
-) %>%
-  mutate_all(linebreak, align = "l") %>% 
-  kableExtra::kable(format = "latex", col.names = c("Taxa", "Number of \nTime Series", "Number of \nSpecies", "Trait Source", "Traits"), escape = FALSE, 
-               caption = "Summary of the data in the final trait database.") %>%
-  kable_classic() %>% 
-  kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
-   #kableExtra::column_spec(2, width = "5cm") #%>%
-  # kableExtra::column_spec(4, width = "10cm")
-```
-
-We gathered trait data from the Elton Trait Database [mammals and birds, @wilman2014] and Amphibio [amphibians, @oliveira2017].
-These databases include species-level means for traits that partially represent species' multifaceted function in the community including body size, diet, and behavioral characteristics.
-For the full list of traits included in the analysis for each taxon see Table \ref{tab:traitSources}.
-Multiple traits (i.e. diet, foraging strata, activity seasonality, active diel period) were broken down into percentage or binary use for each level.
-
-In order to ensure taxonomic consistency across datasets, BioTIME species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (retrieved 09-15-2020 from the on-line database, <https://doi.org/10.5066/F7KH0KBK>), obtained through the `taxadb` R package [@norman2020; @rcoreteam2021].
-If more than one species in the assemblage data resolved to the same identifier, observations were considered the same species.
-For trait data, traits for all species of the same identifier were averaged.
-Only studies with at least 75% trait coverage were included and observations for species with no trait data were excluded.
-In order to have a sufficient number of species to calculate functional diversity metrics, years with fewer than 5 species observed were also excluded.
-
-Many studies had a variable number of samples within years.
-To account for this inconsistency in sampling effort we used sample-based rarefaction by bootstrap resampling within years for each time series based on the smallest number of samples in a year for that time series.
-
-Our final dataset included 2,443 time series from 53 studies in 21 countries and 15 biomes and 13 different traits (Fig \ref{fig:taxaMap}).
-The earliest sample was in 1923 and the most recent was in 2016.
-Only four studies (consisting of 11 time series) came from Amphibian studies due to the limited availability of amphibian time series and low species richness values for assemblages (Table \ref{tab:traitSources}).
-Amphibians are of particular concern due to impacts of habitat loss and pollution [@gibbons2000], so we include data while acknowledging that general inference for amphibians as a clade is not possible with the time series available.
-For a full breakdown of studies and their characteristics, see the supplement.
-
-```{r taxaMap, fig.cap="A) Map of time series locations with points colored by taxa, and B) histograms of time series duration broken down by taxa.", echo=FALSE}
-knitr::include_graphics(here::here("figures/study_map_hist.jpeg"))
-```
-
-## Diversity Metrics
-
-We calculated yearly metrics of functional and species diversity for each time series.
-Species-based metrics include species richness (*S*) and Jaccard similarity (*J*) as a measure of turnover.
-Jaccard similarity was calculated relative to the first observed year for a time series.
-A negative trend in *J* would therefore indicate increasing turnover.
-
-Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package [@laliberté2010].
-Here we report functional richness (*FRic*), functional evenness (*FEve*), and functional divergence (*FDiv*) which together describe three complementary characteristics of the functional space [@mason2005; @hillebrand2009].
-*FRic* assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
-*FEve* describes how species are distributed across the trait space and how abundance is distributed across species.
-Higher values of *FEve* indicate more even spacing of species in the trait space and individuals across species.
-*FDiv* measures the degree to which species and their abundances maximize differences in the functional space.
-Higher values of *FDiv* therefore correspond to communities where many highly abundant species are on the edges of the trait space.
-We also calculated the community-weighted mean (*CWM*) of included traits to examine shifts in the distribution of each trait.
-
-All available trait data for each study were included in functional diversity calculations with the exception of traits that were the same value for all observed species in the study.
-For variables with multiple levels each level was included as a separate trait axis.
-Continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
-The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
-This restriction allows for a sufficient number of axes to capture the trait space while maintaining computational feasibility [@blonder2018].
-Metrics incorporated weighting based on species abundance where available (three studies were presence only).
-
-## Null Models
-
-To assess functional change independent of species richness we calculated the standardized effect size (SES) for each of the three summary functional diversity metrics (*FRic*, *FEve*, *FDiv*) from null estimates [@swenson2012].
-Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
-Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species.
-Species pools included all species observed for a time series.
-This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
-We used these values to calculate SES using the following formula: $SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
-We then calculated the median SES estimate for each metric from all the rarefaction samples for a time series.
-SES estimates can be interpreted as how much of the functional characteristic (richness, evenness, divergence) was observed beyond what was expected by chance for a community of that species richness.
-
-## Analysis
-
-We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study.
-All time series with data for a given trait were included in the model.
-Trends could not be estimated for the *CWM'*s of three traits due to limited data: maximum age at maturity, minimum age of maturity, and minimum litter size.
-We fit 24 individual *CWM* models.
-
-For all other species and functional diversity metrics (*S*, *J*, *FRic*, *FEve*, *FDiv)* we obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model.
-BLUP's provide estimates for the conditional mean and variance of a random effect from which we calculated 95% confidence intervals to determine significance of study-level slopes.
-
-To test for trends within and between different levels of taxa, biome, and realm we fit separate models with each of those factors added as a predictor to the original model structure.
-We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [@lenth2021].
-We assessed the impact of time series duration and start year on study-level trends using general linear models with duration and start year as predictors.
-All models were executed using the *lme4* package in R and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2021].
-
 <!-- # Data Availability -->
@@
 <!-- Code for the analyses in this chapter is archived on Zenodo at [10.5281/zenodo.5514334](https://doi.org/10.5281/zenodo.5514334). -->
+
 <!-- Data products are also archived on Zenodo at [10.5281/zenodo.6499442](https://doi.org/10.5281/zenodo.6499442). -->
+
 <!-- Original data sources are open access and available at their respective providers. -->
 
-# Acknowledgments
+<!-- # Acknowledgments -->
 
-We thank the scientists who contributed to and maintain the biodiversity databases included in this study, including BioTIME, Amphibio, and Elton Traits.
-We thank Dr Shane Blowes and Dr Sarah Supp, whose code we adapted for initial data processing of the BioTIME database.
-Support for this work was provided by U.S.
+<!-- We thank the scientists who contributed to and maintain the biodiversity databases included in this study, including BioTIME, Amphibio, and Elton Traits. -->
+
+<!-- We thank Dr Shane Blowes and Dr Sarah Supp, whose code we adapted for initial data processing of the BioTIME database. -->
+
+<!-- Support for this work was provided by U.S.
 Department of Energy through the Computational Sciences Graduate Fellowship (DE-FG02-97ER25308) awarded to KEAN.
 CB acknowledges support from NSF CAREER Award #1942280 & NIFA project CA-B-INS-0162-H.
+-->
 

### 2022-09-19 analysis/05_model_metrics.Rmd

@@
 ```{r}
+#exclude amphibians
+model_data <- model_data %>% filter(taxa != "Amphibians")
+
 #Fit models with base settings
@@
 #this is a list of models that weren't fit because there weren't enough timeseries OR they didn't converge
-#min_litter_size just won't converge
+#most are amphibian traits or traits only found for one taxa (with taxa as a covariate), the nectar diet realm model wouldn't converge
 missing_mods <- setdiff(uniq_mods, uniq_metric_model)

### 2022-09-19 analysis/figures/visualization.Rmd
Update figures to exclude amphibians

@@
 ```{r}
-pins::board_register_github(repo = "karinorman/biodivTS_data", branch = "master")
-# country_data <- pins::pin_get("country-data-bind", board = "github") %>%
-#   mutate(country = str_remove(country, "Commonwealth of The "),
-#          country = str_remove(country, "Kingdom of "),
-#          country = str_remove(country, "Republic of "),
-#          country = str_remove(country, "Kingdom of "),
-#          country = str_remove(country, "Federative Republic of "),
-#          country = str_remove(country, "Federative "),) %>%
-#   mutate(country = case_when(
-#     country == "United States" ~ "USA",
-#     country == "Russian Federation" ~ "Russia",
-#     country == "United Great Britain and Northern Ireland" ~ "UK",
-#     country == "Czechia" ~ "Czech Republic",
-#     TRUE ~ country
-#   )) %>% 
-#   count(country, code)
-# 
-# mapdata <- map_data("world") %>%
-#   left_join(country_data, by = c("region" = "country"))
-# 
-# mapdata %>%
-#   ggplot(aes(long, lat)) +
-#   geom_polygon(mapping = aes(group = group, fill = n)) +
-#   ggthemes::theme_map() +
-#   #scale_fill_continuous(color = "#32A251")
-#   scale_fill_gradient(low = "white", high = "#39737c")
-
-#ggsave("study_map.jpeg")
+load(here::here("data/model_data.rda"))
 ```
@@
 #get rarefyid centers
-included_ids <- pins::pin_get("model-data", board = "github") %>%
+included_ids <- model_data %>%
   select(rarefyID, taxa, climate, realm, duration) %>%

### 2022-09-19 analysis/05_model_metrics.Rmd
get what we need from the slope summaries

@@
   filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
-  group_by(metric) %>% 
-  summarise(n_study_id = n_distinct(study_id)) %>%
-  mutate(percent = n_study_id/53)
+  mutate(positive = ifelse(estimate > 0, TRUE, FALSE)) %>%
+  group_by(metric, positive) %>% 
+  summarise(n_study_id = n_distinct(study_id))
 
@@
   filter(sig == TRUE) %>% 
-  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
+  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S", "FRic", "FEve", "FDiv")) %>% 
   select(study_id, metric) %>%
@@
   left_join(meta %>% 
-              dplyr::select(-rarefyid) %>% 
-              distinct() %>%
               mutate(study_id = as.character(study_id))) %>%

### 2022-09-19 analysis/figures/visualization.Rmd
Minor changes to adjust for no amphibians

@@
 load(here::here("data/model_data.rda"))
+load(here::here("data/rarefyID_cell_centre.rda"))
 ```
@@
 included_ids <- model_data %>%
+  filter(taxa != "Amphibians") %>%
   select(rarefyID, taxa, climate, realm, duration) %>%
@@
 
-study_data <- pins::pin_get("rarefyID-cell-centre", board = "github") %>%
+study_data <- rarefyID_cell_centre %>%
   inner_join(included_ids) #%>%
@@
   #plot the US amphibian point on top so we can see it
-  geom_point(data = study_data %>% 
-               #filter(taxa %in% c("Amphibians")), 
-               filter(taxa %in% c("Amphibians"), rarefyID_y > 0),
-             aes(rarefyID_x, rarefyID_y ), color = "#d9be3b", alpha = 0.8, inherit.aes = FALSE) +
+  # geom_point(data = study_data %>% 
+  #              #filter(taxa %in% c("Amphibians")), 
+  #              filter(taxa %in% c("Amphibians"), rarefyID_y > 0),
+  #            aes(rarefyID_x, rarefyID_y ), color = "#d9be3b", alpha = 0.8, inherit.aes = FALSE) +
   ggthemes::theme_map() +
-  scale_color_manual(values = my_pal) +
+  scale_color_manual(values = my_pal[2:3]) +
   theme(legend.title  = element_blank()) +
@@
         ) +
-  scale_fill_manual(values = my_pal)
+  scale_fill_manual(values = my_pal[2:3])
 
-plot_grid(study_map, duration_hist, labels = "AUTO", nrow = 2, align = 'vh', hjust = -1, axis = "l", scale = c(1,0.9), rel_heights = c(2, 1))
+plot_grid(study_map, duration_hist, labels = "AUTO", nrow = 2, align = 'vh', hjust = -1, axis = "l", scale = c(1,0.9), rel_heights = c(2, 1.25))
 
@@
 ```{r}
-metric_model_table <- pins::pin_get("metric-model-table", board = "github")
-model_data <- pins::pin_get("model-data", board = "github")
+load(here::here("data/metric_model_table.rda"))
 

### 2022-09-20 analysis/06_sensitivity_analysis.Rmd
Add sensitivity analysis stuff

@@
 ```{r}
-dur_dfs <- list(min3 = filter(model_data, duration > 2),
-                min4 = filter(model_data, duration > 3),
-                min5 = filter(model_data, duration > 4))
+dur_dfs <- list(min3 = filter(model_data, taxa != "Amphibians", duration > 2),
+                min4 = filter(model_data, taxa != "Amphibians", duration > 3),
+                min5 = filter(model_data, taxa != "Amphibians", duration > 4))
 
-duration_sens <- map_dfr(dur_dfs, ~pmap_dfr(mods, get_model, data = .x), .id = "dataID")
+duration_sens <- map_dfr(dur_dfs, ~pmap_dfr(filter(mods, model_id == "base"), get_model, data = .x), .id = "dataID")
 ```
@@
 ```{r}
-no_amph <- pmap_dfr(mods, get_model, data = model_data %>% filter(taxa != "Amphibians")) %>%
-  mutate(dataID = no_amph)
-
-# this annoyingly doesn't work because of some dataframe column
-# sens_models <- bind_rows(bind_cols(select(duration_sens, -dataID), pull(duration_sens, dataID)) %>% select(-model, -formula), 
-#                          no_amph %>% select(-model, -formula))
-
-save(duration_sens, no_amph, file = here::here("data/sensitivity_models.rda"))
+with_amph <- pmap_dfr(filter(mods, model_id == "base"), get_model, data = model_data) %>%
+  mutate(dataID = "with_amph")
 ```
@@
 
-covg_dfs <- list(covg75 = filter(model_data),
-                covg85 = filter(model_data, coverage >= 0.85),
-                covg90 = filter(model_data, coverage >= 0.90))
+covg_dfs <- list(covg75 = filter(model_data, taxa != "Amphibians"),
+                covg85 = filter(model_data, taxa != "Amphibians", coverage >= 0.85),
+                covg90 = filter(model_data, taxa != "Amphibians", coverage >= 0.90))
 
-coverage_sens <- map_dfr(covg_dfs, ~pmap_dfr(mods %>% 
-                                               filter(metric_name %in% c("FRic", "FDiv", "FEve", "S", "SES_FRic",
-                                                                         "SES_FDiv", "SES_FEve"), 
-                                                      model_id == "base"), 
+coverage_sens <- map_dfr(covg_dfs, ~pmap_dfr(filter(mods, model_id == "base"), 
                                              get_model, data = .x), .id = "dataID")
 
+sens_mods <- bind_rows(duration_sens, coverage_sens, with_amph) %>%
+  select(-model) %>%
+  filter(metric != "Jaccard_next") %>%
+  write_csv(here::here("paper", "sensitivity_models.csv"))
 ```

### 2022-11-03 paper/manuscript_GEB.Rmd
changes after removing amphibians
Line edits and substantial rewriting

@@
 ---
-title: Evaluating the evidence of widespread maintenance of functional composition in bird and mammal communities
+title: Evidence of widespread maintenance of functional composition in bird and mammal communities
 # authors: 
@@
 abstract: |
- Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing species turnover but no overall trend in species diversity through time.
+ Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing species turnover but no overall trend in species diversity through time at the local scale.
 Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to mechanisms of ecosystem processes.
 Here we present the first multi-taxa, multi-system analysis of functional change through time, pairing thousands of vertebrate assemblage time series from the BioTIME database with existing trait data representative of a species' functional role.
-We found no overall trend in any functional diversity metric, despite similar species-based patterns of constant richness with increasing turnover.
-The lack of trend held even after correcting for changes in species richness and at the study-level, where only 3 of 54 studies experienced a significant trend in at least one functional diversity metric.
-Results give evidence that across a variety of taxa, climates, and biomes, functional characteristics are maintained even in the face of significant environmental and community change.
+We found no overall trend in any functional diversity metric, despite data replicating species-based patterns of constant richness with increasing turnover.
+The lack of trend held even after correcting for changes in species richness and at the study-level, where the vast majority of studies experienced no significant trend in any functional diversity metric.
+Results give evidence that across a variety of taxa, climates, and biomes, functional characteristics are maintained even in the face of significant community change.
 We also discuss the potential for underlying functional shifts to be obscured by current approaches and data, calling for targeted data collection efforts to combat existing biases in monitoring and trait data.
- Aim, Location, Time period, Major taxa studied, Methods, Results, Main conclusions
 
@@
 
-Debate over the meaing of general trends, their robustness, and relevance for conservation action is ongoing.
+Debate over the meaning of general trends derived from limited data and their relevance for conservation is ongoing.
+Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, a lack of baselines, and for not taking into account disturbance histories of individual studies [@cardinale2014; @cardinale2018; @gonzalez2016].
+However, many of these criticisms are unavoidable, reflecting general limitations of ecological data [@primack2018].
+The pattern of increasing turnover accompanied by constant or increasing richness has also been reproduced in multiple datasets and taxa, giving strong evidence that it represents a true ecological phenomena [@barnagaud2017]; @dunic2017; @elahi2015].
 
-Functional diversity offers a potentially powerful alternative for detecting and describing community change by providing a mechanistic link between species' response to environmental change (*response traits*) and the processes they perform (*effect traits*) [@lavorel2002; @mcgill2006; @suding2008].
+Functional diversity offers a potentially powerful alternative to species-based approaches for detecting and describing community change by providing a mechanistic link between species' response to environmental change (*response traits*) and the processes they perform (*effect traits*) [@lavorel2002; @mcgill2006; @suding2008].
 By describing the functional trait space, functional diversity metrics capture the disproportionate impact of losses or gains of functionally unique species.
@@
 Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020].
-And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @matuoka2020].
+And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @larsen2018; @matuoka2020].
 On paleoecological time scales functional composition shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with significant impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].
-Some losses of functional diversity are indisputable on both paleocological and contemporary timescales such as continued trophic downgrading <!-- or is it just body-size downgrading?
-we lost mega grazers too --> due to loss of large-bodied mammals, but implications of those losses for local diversity patterns are less clear [@estes2011; @smith2018] .
+For some time periods, functional structure appears to be maintained for substantial portions of geological time [@hedberg2022].
+Some losses of functional diversity are indisputable on both paleocological and contemporary timescales such as continued trophic downgrading due to loss of large-bodied mammals, but implications of those losses for local diversity patterns are less clear [@estes2011; @smith2018] .
 
@@
 The majority of work has therefore focused largely on system-specific studies with traits collected *in situ*.
+This has challenged our ability to establish general rules for how functional composition may be changing through time, particularly in response to global change drivers.
 Ongoing efforts to assemble functional traits for a variety of taxa have made synthesis of existing community assemblage data and functional traits possible for the first time, providing initial insights into the ways functional diversity changes on a broad scale for specific taxa [e.g.
 fish, @trindade-santos2020; birds, @jarzyna2016; @barnagaud2017].
-However, to date there has been no cross-taxa assessment of temporal functional change for a broad geographic and taxonomic extent.
+However, to date there has been no cross-taxa assessment of temporal functional change for a broad geographic and taxonomic extent [@díaz2022].
 
-Here we perform the first multi-taxa, multi-system assessment of functional diversity change through time.
+Here we perform the first multi-taxa, multi-realm assessment of functional diversity change through time.
 We focus on mammal and bird species as a significant subset of the world's biodiversity heavily impacted by anthropogenic change.
@@
 Our final dataset included 2,432 time series from 50 studies in 21 countries and 12 biomes and 6 different traits (Fig \ref{fig:taxaMap}).
+Data came from both terrestrial and marine realms and five climates (Global, Polar/Temperate, Temperate, Temperate/Tropical, Tropical).
 The earliest sample was in 1923 and the most recent was in 2014.
@@
 A negative trend in *J* would therefore indicate increasing turnover.
-We did not impose a correction for unobserved species as non-parametric estimators do not assign species identities to corrected richness values, and therefore could not be propogated to the functional diversity metrics.
+We did not impose a correction for unobserved species as non-parametric estimators do not assign species identities to corrected richness values, and therefore could not be propagated to the functional diversity metrics.
 
@@
 This restriction allows for a sufficient number of axes to capture the trait space while maintaining computational feasibility [@blonder2018].
-Metrics incorporated weighting based on species abundance where available (three studies were presence only).
+Metrics incorporated weighting based on species abundance.
 
@@
 All time series with data for a given trait were included in the model.
-Trends could not be estimated for the *CWM'*s of three traits due to limited data: maximum age at maturity, minimum age of maturity, and minimum litter size.
-We fit 24 individual *CWM* models.
+We fit 18 individual *CWM* models.
 
@@
 We found no significant overall trend in species richness or summary functional diversity metrics (observed or standardized) (Fig \ref{fig:timeseriesPlot}).
-We did find a significant overall decrease in Jaccard similarity, indicating increasing turnover through time.
+We did find a significant overall decrease in Jaccard similarity, indicating accumulating changes in species composition.
 Non-significant overall trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies was plausibly 0 (Table \ref{tab:resultsTab}).
-Trends for different taxa, biomes, or realms were also non-significant with the exception of a significantly increasing trend for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing standardized functional richness slope for freshwater studies.
-However, with only two global studies and two freshwater studies these results are not general trends.
-We found similar results for the *CWM* models, with a significant trend for only one trait, percentage of aerial foraging utilization, which showed a significant positive trend.
+Trends for different taxa, biomes, or realms were also non-significant for richness and summary functional diversity metrics, with the exception of a significantly increasing trend for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly increasing functional richness slope for tropical studies.
+However, neither trend was exhibited for the standardized metric, indicating that differences were largely due to changes in species richness.
+Further, with only two global studies these results are not general trends.
+The general trends for *CWM* models were similarly not significant, with a significant positive trend for only percentage of aerial foraging utilization.
 
@@
   select(metric, everything()) %>%
+  select(-c(n_ts, formula, lower.CL, upper.CL, significant)) %>%
   filter(group != "Residual" | is.na(group)) %>%
@@
 
-At the study level, 4 studies experienced a significant trend in species richness and only 10 of 54 studies for observed metrics and 3 of 54 studies for standardized metrics experienced a significant trend for a metric other than Jaccard similarity (Table \ref{tab:trendTab}).
+At the study level, 4 studies experienced a significant trend in species richness and only 9 of 50 studies for observed metrics and 3 of 50 studies for standardized metrics experienced a significant trend for a metric other than Jaccard similarity (Table \ref{tab:trendTab}).
 Most significant trends for observed functional metrics are in functional richness and disappeared after standardization, indicating that richness increases were likely due to changes in the number of species.
@@
 
-data.frame(sign = c("+", "-"), S = c("1", "3"), `Jaccard Similarity` = c("0", "37"), FRic = c("2", "6"), FEve = c("1", "0"), FDiv = c("0", "1"), `SES FRic` = c("0", "1"), `SES FEve` = c("0", "2"), `SES FDiv` = c("0", "1")) %>%
+data.frame(sign = c("+", "-"), S = c("1", "3"), `Jaccard Similarity` = c("0", "31"), FRic = c("2", "5"), FEve = c("1", "0"), FDiv = c("0", "1"), `SES FRic` = c("0", "1"), `SES FEve` = c("0", "2"), `SES FDiv` = c("0", "1")) %>%
   column_to_rownames("sign") %>%
-  kable(caption = "Number of studies that experienced a significant trend in each calculated metric out of 53 total studies.")
+  kable(caption = "Number of studies that experienced a significant trend in each calculated metric out of 50 total studies.")
 ```
@@
 Study-level slopes for multiple metrics were significantly related to the duration and start year of studies.
-Slopes for species richness were significantly more negative with later start date and more positive shorter duration studies.
+Slopes for species richness were significantly more negative with later start date and more positive for longer duration studies.
 Jaccard similarity and functional evenness both had significantly more negative slopes with more recent start year, whereas functional divergence was significantly more positive.
 Slopes for functional evenness were also significantly more positive for longer duration studies.
-Results were consistent between standardized and observed metrics with exception of functional evenness, which was negatively related to duration for observed data and positively related for standardized data.
 See supplement for estimates and p-values for all models.
+
+We did find significant differences between taxa, realms, and climate for Jaccard similarity and some of the *CWM's*.
+For example while Jaccard similarity was decreasing in the general trend and there were significant within group slopes for Mammals, Birds, and Terrestrial communities, there was no significant slope for the marine realm, indicating that the general trend is mostly driven by turnover in terrestrial communities.
+For bird communities, we also found within group trends and between group differences for trends in foraging behavior.
+We found a significant increasing trend in utilization of the canopy in Tropical communities that was significantly different from the trends for Polar/Temperate and Temperate communities.
+There was a significant decrease in utilization of the understory for Terrestrial communities and significant increase in foraging below the water surface for global studies (but see the previous limitations of Global data).
+There was also a significant positive slope for *CWM* body mass for the single Temperate/Tropical study (with three time series) which were marine mammal communities.
+
+We found significant dietary shifts across communities.
+We found a significant increase in fruit consumption in Terrestrial communities and a significant decrease in nectar consumption in Tropical communities, a trend significantly different than that for Terrestrial communities.
+There was a significant increase in seed consumption in bird species, which was significantly different from the trend for Mammal communities.
+There was a significant increase in fish consumption for the two Global studies.
+Vertebrate consumption significantly decreased for Marine communities and for Global and Tropical communities.
+
+We assessed the sensitivity of general trend results to major data processing decision by rerunning models with increasingly conservative subsets of the data.
+After excluding time series with less than two, three, and four year durations, we found that the general trend for increase in CWM aerial foraging disappeared, but two other trends in CWM's appeared.
+After excluding time series two years long, a general trend appeared for increased CWM use of the forest canopy, and decreased CWM of seed consumption.
+These two trends remained as increasingly shorter time series were excluded suggesting the increase in aerial foraging was an erroneous finding while the other two trends were more robust.
+Percentage of species in a community with trait data did not appear to have an affect on general trend results, as general trends were unchanged with increasing cut offs for percentage of species with trait data.
+A complete list of models run in the sensitivity analysis and their results can be found in the supplement.
 
@@
 Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at aggregate and local trends in functional diversity in mammal and bird communities.
-Surprisingly, we did not detect an overall trend in any of the chosen functional diversity metrics.
+Surprisingly, we did not detect an overall trend in any of the summary functional diversity metrics.
 As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics may be consistent with similar well-documented species derived trends.
 We found no evidence of systematic functional richness loss or functional change.
-A lack of trend for almost all realms, biomes, and taxonomic groups gives further evidence that directional functional change is absent from all systems observed in our dataset.
-Additionally, results from *CWM* models show that there were very few general shifts in functional trait values.
+A lack of trend for almost all realms, biomes, and taxonomic groups gives further evidence that changes in functional structure are not occurring in the systems observed in our dataset.
 
-<!-- This striking result could be a product of two possible processes, one ecological and one methodological.
-Null trends appear to give strong evidence of systematic maintenance of functional structure due to common ecological processes, however multiple limitations of current approaches in synthesis could potentially be obscuring a true underlying global trend.
-We discuss both options further here.-->
+Very few of the studies in our data set experienced a significant trend in species richness for the observation period, our data therefore mostly represent communities undergoing significant change in species composition but not species counts.
+Results should therefore not be extrapolated to communities experiencing significant species losses or gains as in scenarios of recovery from disturbance or catastrophic human impact.
+While we cannot make generalizations about scenarios with richness losses or gains from out data, for the four studies that did experience significant species richness changes, only one exhibited any trend in a summary functional metric with a significant decrease in FRic but not SES FRic, indicating it was mostly functional loss due to species loss.
+We therefore have preliminary evidence that maintenance of functional structure can occur even in the face of species richness gains or losses, but additional data sources will be necessary to further explore how prevalent those scenarios are.
 
-We discuss here the possible ecological processes underpinning maintenance of functional diversity, particularly in a global change context.
-
-## Evidence of Ecological Processes
-
-Communities demonstrated almost universal maintenance of functional composition.
-While the majority of the studies (\~70%) included in our data experienced significant species turnover, only three (for standardized metrics) experienced a significant shift in any functional dimension.
+Communities demonstrated almost universal maintenance of functional structure.
+While the majority of the studies (\~70%) included in our data experienced significant species turnover, only three (for standardized metrics) experienced a significant shift in any summary functional metric.
 This suggests certain characteristics of the functional space are maintained even in the face of significant change in species identity, specifically the size of the functional space occupied by the community (*FRic*), the distribution of species and individuals within that space (*FEve* and *FDiv*), and the mean of individual trait distributions (*CWM*).
@@
 While observed communities are clearly losing species (as demonstrated by the turnover in species identity), sufficient diversity is being maintained in the species pool to provide functionally comparable species as replacement.
-This phenomena could be the result of two possible process, potentially working in concert.
-First, functional redundancy in the species pool maintains resilience of the functional space by providing multiple potential species for a any given set of functional traits.
-Alternatively, while there may be low redundancy in the functional space, functional composition in the species pool may be replenished by changes in the composition of the species pool to to shifts in species ranges.
-<!-- what does global change say about trajectory of functional redundancy -->
+This phenomena could be the result of two possible process potentially working in concert.
+First, functional redundancy in the species pool maintains resilience of the functional space by providing multiple potential species for any given set of functional traits.
+Alternatively, while there may be low redundancy in the functional space, functional composition in the species pool may be replenished by changes in the composition of the species pool due to shifts in species ranges.
+Addition of new species to the species pool could occur under multiple mechanisms of global change, including the introduction of non-native species, shifts in species ranges due to climate change, and spatial homogenization due to replacement of specialist species by generalist ones.
+Future work directly assessing changes in the species pool and attributing those shifts to specific mechanisms will be critical for understanding how functional structure is maintained and what may interrupt that maintenance.
 
--   what is the expectation for change in functional redundancy (and species pool changes) due to global change?
+Despite maintenance of summary functional metrics we did find multiple trends in *CWM*'s for studies coming from the same taxa, realms, or climates, many of which are consistent with previous findings for those contexts.
+For example, the decrease in insect consumption for tropical communities reflects well documented declines in tropical insectivorous birds [@sherry2021].
+Reduction in utilization of understory foraging for terrestrial communities could be the result of disturbance like human recreation or increased predation from introduced species.
+Other shifts in diet for both Birds and Mammals, including increasing fruit, seed, and vertebrate consumption and decreasing seed consumption in some climates point to important areas for further exploration.
+Generally, these findings indicate that even when some shifts in the functional space are occurring, often as a direct result of global change impacts, summary metrics describing functional structure can be maintained.
 
--   don't have the data here to directly assess temporal shifts in the species pool per say, so we can't distinguish between these process, buy you could do it x,y,z way
-
--   paleoecological studies indicate that functional composition can be largely conserved for long periods of time (strong ecological process), and losses are only seen after catastrophic events (e.g.
-terminal pleistocene extinction).
-
--   discuss possibility of a tipping point?
-    catastrophic loss following long period of stability, what might the early warning signs be of that?
-
-<!-- change this discussion to reconciling with other long term trends and where the differences may arise -->
-
-These results challenge assumptions that functional loss is the default state of all or even many communities.
-And while there is some evidence that functional loss may be more common than functional gains for communities experiencing functional change, there was no significant trend for even the longest running and most heavily impacted studies.
+These results challenge implicit assumptions that functional loss is the default state of all or even many communities [@carmona2021; @dirzo2014; @young2016].
+And while there is some evidence in our data that functional loss may be more common than functional gains for communities experiencing functional change, there was no significant trend for even the longest running and most heavily impacted studies.
 The North American Breeding Survey for example is considered an authoritative dataset on the state of bird populations on the continent and underpins policy decisions about bird conservation [@pardieck2020; @rosenberg2019; @sauer2017].
@@
 
-Results are also seemingly inconsistent with predictions for trait shifts under global change.
+Results are also in contrast to some predictions for trait shifts under global change.
 For example, mean body size is predicted to decrease as a result of climate change impacts and megafaunal loss [@sheridan2011], a phenomena which has already been well documented empirically and experimentally in multiple taxa [@caruso2014; @forster2012; @huss2019; @tseng2018].
 While the species-level trait means used here are not appropriate for assessing intraspecific body size shifts, we would expect to see shifts in community-wide means due to local losses of large-bodied species.
-Instead, we found no evidence of a trend in *CWM* body size.
-<!-- in the communities we sampled we trophic downgrading has probably already occurred! -->
+Instead, we found no evidence of a trend in *CWM* body size, with the exception of a single study of marine mammals in the Bahamas which showed a significant increase.
+Still, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing [@fritz2009].
+Trends could be significantly different for the same time periods in regions of sub-saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened [@ripple2015].
 
--   discussion of expectations for undersampled areas (Africa particularly that still has a lot of its megafauna)?
-
-While we did not directly measure changes in rare species, our results further contradict likely scenarios of loss predicted due to rare species extinction.
-Rare species, defined by small populations and geographic restriction, are simultaneously more likely to be functionally distinct and at higher risk for extinction [@davies2004; @harnik2012; @loiseau2020].
+While we did not directly measure changes in rare species, our results further contradict hypothesized patterns due to rare species extinction.
+Rare species, defined by small populations and geographic restriction, are simultaneously more likely to be functionally distinct and at higher risk for extinction [@davies2004; @fritz2009; @harnik2012; @loiseau2020].
 Locally, communities losing functionally rare species should exhibit strong functional shifts as lost species can eventually no longer be replaced by functionally similar species [@leitão2016].
@@
 dispersion, rarity, abundance of dominant species, @bagousse-pinguet2021].
-Still, the fact that we observed so many communities maintaining structure across the most commonly used metrics for linking biodiversity and function calls into question how previous work translates to natural communities.
-Metrics need to be both closely linked to changes in ecosystem function and also experiencing shifts in natural communities to be meaningful.
+Still, our results indicate that highs species turnover alone is not sufficient cause for concern for ecosystem function, since functional structure is largely conserved.
+However, that may not be true indefinitely if functionally redundant species are depleted from the species pool.
 
@@
 However, a number of gaps in best practices may be obscuring a true underlying trend.
+Here we discuss data limitations and the implications they may have for general trends.
 First, the BioTIME database, while the most comprehensive data source of time series available, is limited in temporal and geographic scope.
-Most time series span only a few years (Fig \ref{fig:taxaMap}) and may not provide the statistical power necessary to detect trends.
+Most time series span only a few years (Fig \ref{fig:taxaMap}) and may not provide the statistical power necessary to detect trends [@wauchope2019].
 The database is also not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
@@
 Likewise, available trait data may not capture the traits experiencing the greatest change.
+For example, we focus here mostly on ecological traits (i.e. foraging strategy, diet, etc) rather than life-history or reproductive traits.
 
-Third, while we use here the most common metrics for describing functional diversity, they do not measure some potentially important aspects of the functional space.
+<!-- Third, while we use here the most common metrics for describing functional diversity, they do not measure some potentially important aspects of the functional space.
 Most notably, the summary metrics we calculated do not capture shifts in the location of the functional space as a whole.
@@
 This scenario may be common in communities tracking changing environmental conditions.
-While trait *CWM's* capture axis shifts, approaches for assessing multidimensional shifts in functional space are still relatively new [@barros2016; @blonder2018; @mammola2019] but could shed critical insight into functional composition changes of this nature.
+While trait CWM's capture axis shifts, approaches for assessing multidimensional shifts in functional space are still relatively new [@barros2016; @blonder2018; @mammola2019] but could shed critical insight into functional composition changes of this nature.
+-->
 
@@
 Our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
-These findings do not negate a substantial body of work linking functional degradation to direct human intervention in the form of land use change and intensification or habitat fragmentation [@flynn2009; @magioli2021; @tinoco2018], but rather illustrate trends for communities experiencing background levels of environmental change.
+These findings compliment a substantial body of work linking functional degradation to species losses as a result of direct human intervention in the form of land use change and intensification or habitat fragmentation [@flynn2009; @magioli2021; @tinoco2018].
 Rather than assuming functional structure will be maintained in areas of concern, our work indicates that when measurements of functional diversity show significant shifts, it should be considered an indication of substantial community change and outside the normal expectation.
@@
 Here we make a significant first step in establishing a general trend for functional diversity through time across a variety of taxa and systems.
-We present the conclusion best supported by available data and acknowledge that it is still too early to confidently distinguish between true ecological pattern and methodological limitations.
-The most pressing next step is for intentional and targeted data collection efforts.
+We present the conclusion best supported by available data and acknowledge that overall trends could look very different for communities experiencing significant species losses or gains.
+Better understanding the functional make up of a species pool for a given community will be critical for evaluating a community's capacity for continued maintenance of functional structure, and be a first step in identifying which processes may be maintaining functional structure.
+Further, looking at species pool make up through time would give us some indicator of the trajectory for future maintenance.
+
+Future broad-scales analysis would also greatly benefit from intentional and targeted data collection efforts.
 We join others in the call for increased monitoring in under sampled areas and continued efforts to centralize existing data sources [@gonzalez2016; @hughes2021; @vellend2017].

### 2022-11-03 paper/manuscript_GEB.Rmd
new references too

@@
 title: Evidence of widespread maintenance of functional composition in bird and mammal communities
-# authors: 
-#   - name: "Kari E. A. Norman"
+authors:
+- name: "Kari E. A. Norman"
 #   affiliation: Université de Montréal
@@
 # - name: "Perry de Valpine"
-#     affiliation: University of California, Berkeley
+#    afiliation: University of California, Berkeley
 #   location: Berkeley, California
-#     email: pdevalpine@berkeley.edu
+#    eail: pdevalpine@berkeley.edu
 # - name: "Carl Boettiger"
@@
 However, many of these criticisms are unavoidable, reflecting general limitations of ecological data [@primack2018].
-The pattern of increasing turnover accompanied by constant or increasing richness has also been reproduced in multiple datasets and taxa, giving strong evidence that it represents a true ecological phenomena [@barnagaud2017]; @dunic2017; @elahi2015].
+The pattern of increasing turnover accompanied by constant or increasing richness has also been reproduced in multiple datasets and taxa, giving strong evidence that it represents a true ecological pattern [@barnagaud2017; @dunic2017; @elahi2015].
 
@@
 
-We used trait data from the Elton Trait Database, which consists of species-level means for traits that represent species' multifaceted function in the community [@wilman2014].
+We used trait data from the Elton Trait Database, which consists of species-level means for traits that represent species' multifaceted role in the community [@wilman2014].
 Traits include: body mass, diet, active diel period, nocturnality, forest foraging strata, pelagic use.
@@
 In order to have a sufficient number of species to calculate functional diversity metrics, years with fewer than 5 species observed were also excluded.
-
 Sensitivity analyses were conducted for the trait coverage threshold and the duration of included time series.
@@
 First, functional redundancy in the species pool maintains resilience of the functional space by providing multiple potential species for any given set of functional traits.
-Alternatively, while there may be low redundancy in the functional space, functional composition in the species pool may be replenished by changes in the composition of the species pool due to shifts in species ranges.
+Alternatively, while there may be low redundancy in the functional space, functional composition in the species pool may be replenished by changes in the composition of the species pool.
 Addition of new species to the species pool could occur under multiple mechanisms of global change, including the introduction of non-native species, shifts in species ranges due to climate change, and spatial homogenization due to replacement of specialist species by generalist ones.
@@
 Here we approach the question of functional change using the best available data and biodiversity synthesis approaches.
-However, a number of gaps in best practices may be obscuring a true underlying trend.
-Here we discuss data limitations and the implications they may have for general trends.
+However, some gaps in best practices may be obscuring a true underlying trend.
 First, the BioTIME database, while the most comprehensive data source of time series available, is limited in temporal and geographic scope.

### 2022-11-03 paper/sensitivity_supp.Rmd

---
title: "Sensitivity Analysis"
output:
  pdf_document: default
  html_document: default
---

We performed sensitivity analyses to assess whether or not the general trends for biodiversity metrics changed based on major data processing decisions. We looked at 1) the minimum duration of included time series, 2) the minimum cut off for percentage of species with traits from a time series, and 3) in inclusion of amphibian data. The table below summarizes all models run and the variables for which a significant trend was detected. A star indicates a trend that was not detected in the original analyses presented in the manuscript. The model_id corresponds to the model_id in the CSV of sensitivity model results.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(dplyr)
library(kableExtra)

tibble(sensitivity_test = c(rep("duration", 3), rep("trait coverage", 2), "amphibians"), 
       model_id = c("min3", "min4", "min5", "covg85", "covg90", "with_amph"),
       description = c("Exclude time series with fewer than 3 years",
                       "Exclude time series with fewer than 4 years",
                       "Exclude time series with fewer than 5 years",
                       "Exclude time series with less than 85% of species with trait data",
                       "Exclude time series with less than 90% of species with trait data",
                       "Include Amphibian data"),
       sig_metrics = c("CWM seed diet*, CWM canopy foraging*, CWM aerial foraging, Jaccard", 
       "CWM seed diet*, CWM canopy foraging*, FRic*, Jaccard", 
       "CWM seed diet*, CWM canopy foraging*, Jaccard", rep("Jaccard", 3))
) %>% kable(format = "latex") %>%
  column_spec(3:4, width = "3cm") %>%
  collapse_rows(columns = 1, valign = "middle") %>%
  kable_paper() %>%
  kable_styling(latex_options= c("hold_position"))
```

### 2022-12-07 analysis/05_model_metrics.Rmd
Model studies individually

@@
 ```{r}
-get_model <- function(metric_name, formula, model_id, data, ...){
+get_model <- function(metric_name, formula, model_id, data, study_id, ...){
   print(c(metric_name, model_id))
@@
 
+  if (missing(study_id)){
     fit_data <- data %>%
@@
              climate = as.factor(climate))
+  } else{
+    fit_data <- data %>%
+      filter(metric == metric_name) %>%
+      mutate(year_scaled = scale(year),
+             duration_scaled = scale(duration, scale = FALSE),
+             startyear_scaled = scale(startyear, scale = FALSE),
+             taxa = as.factor(taxa),
+             realm = as.factor(realm),
+             climate = as.factor(climate)) %>%
+      filter(study_id == study_id)
+  }
 
@@
 
+# get number of observations for each model
+# lmm_obs <- model_table %>% 
+#   filter(model_id == "base") %>% 
+#   select(metric, model) %>% 
+#   rowwise() %>% 
+#   mutate(nobs = nobs(model)) %>% 
+#   select(-model) %>% 
+#   distinct()
+
 metric_model_table <- model_table %>% select(-model)
@@
            std.error = sqrt(var),
-           upr.ci = estimate + (2*std.error),
-           lwr.ci = estimate - (2*std.error),
+           upr.ci = estimate + (1.96*std.error),
+           lwr.ci = estimate - (1.96*std.error),
            sig = case_when(
@@
 
-# study_slopes_wide <- study_slopes %>%
-#   select(study_id, metric, estimate) %>%
-#   pivot_wider(names_from = "metric", values_from = "estimate")
+usethis::use_data(study_slopes)
 ```
@@
 ```
+
+#### Looking at BBS specifically to compare slopes and significance
+```{r}
+bbs_models <- map_dfr(metric_names[metric_names != "S"], 
+                      ~get_model(data = model_data, 
+                                 metric_name = .x, "value ~ year_scaled + (year_scaled|rarefyID)", 
+                                 model_id = "base",
+                                 study_id = 195))
+
+bbs_models %>% 
+  filter(term == "year_scaled") %>% 
+  rowwise() %>% 
+  mutate(nobs = nobs(model)) %>% 
+  select(-model, -model_id, -formula, -n_ts, -warning, -group) %>% 
+  relocate(metric) %>% 
+  arrange(metric) %>% 
+  write_csv(here::here("data/bbs_slopes_indvmodels.csv"))
+```

### 2022-12-07 analysis/exporatory_scripts/bbs_explore.Rmd
Model studies individually

---
title: "BBS Models"
output: html_document
---
```{r}
library(tidyverse)
library(lme4)
library(broom.mixed)
library(lmerTest)
library(emmeans)
```

#### Read in data, get final dataset for models
```{r}
load(here::here("data/model_data.rda"))

model_data <- filter(model_data, taxa != "Amphibians")
```

####Fit GLMMs to overall patterns and data broken down by groups

Function to fit a model
```{r}
get_model <- function(metric_name, formula, model_id, data, study_id, ...){
  print(c(metric_name, model_id))
  qlmer <- quietly(lmer)

  #check if we need to filter for a single study
  if (missing(study_id)){
    fit_data <- data %>%
      filter(metric == metric_name) %>%
      mutate(year_scaled = scale(year),
             duration_scaled = scale(duration, scale = FALSE),
             startyear_scaled = scale(startyear, scale = FALSE),
             taxa = as.factor(taxa),
             realm = as.factor(realm),
             climate = as.factor(climate))
  } else{
    fit_data <- data %>%
      filter(metric == metric_name) %>%
      mutate(year_scaled = scale(year),
             duration_scaled = scale(duration, scale = FALSE),
             startyear_scaled = scale(startyear, scale = FALSE),
             taxa = as.factor(taxa),
             realm = as.factor(realm),
             climate = as.factor(climate)) %>%
      filter(study_id == study_id)
  }

  #fit the model
  fit <- qlmer(formula = formula, data = fit_data, ...)

  # get tidy version of the model output, capture the warnings
  tidy(fit$result) %>%
    mutate(metric = metric_name, model = list(fit$result), model_id = model_id, 
           n_ts = n_distinct(fit_data$rarefyID),
           nobs = nobs(fit$result),
           formula = formula,
           warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))

}

#get_model(data = model_data, metric_name = "SES_FRic", "value ~ year_scaled + (year_scaled|study_id/rarefyID)", model_id = "base")
```

Function to extract estimates for slopes of studies
```{r}
get_study_slopes <- function(metric, model, model_id){

  filter_name <- ifelse(model_id == "base", "study_id", model_id)

  #get the statistics for the year variable
  year_scaled_coef <- tidy(model) %>%
    filter(term == "year_scaled")

  #get the year estimate
  year_scaled_est <- pull(year_scaled_coef, estimate)

  #get year variation
  year_scaled_var <- year_scaled_coef %>% 
    mutate(var = std.error*std.error) %>%
    pull(var)

  #get study level conditional estimates and add to over all estimates
  study_ests <- broom.mixed::tidy(model, effects="ran_vals") %>%
    filter(group == filter_name, term == "year_scaled") %>% 
    select(-c(effect, group, term)) %>%
    rename(cond.std.error = std.error, cond.estimate = estimate) %>%
    mutate(estimate = cond.estimate + year_scaled_est,
      cond.var = cond.std.error*cond.std.error, 
           var = cond.var + year_scaled_var,
           std.error = sqrt(var),
           upr.ci = estimate + (1.96*std.error),
           lwr.ci = estimate - (1.96*std.error),
           sig = case_when(
             lwr.ci < 0 & upr.ci > 0 ~ FALSE,
             TRUE ~ TRUE
           ),
      metric = metric,
      model_id = model_id)
}
```

Fit models only for BBS (study 195)
```{r}
metric_names <- unique(model_data$metric)

# Fit all the models only to BBS data (filtering is done in the function)
bbs_models <- map_dfr(metric_names[metric_names != "S"], 
                      ~get_model(data = model_data, 
                                 metric_name = .x, "value ~ year_scaled + (year_scaled|rarefyID)", 
                                 model_id = "base",
                                 study_id = 195))

# Get table of estimates for the scaled year slope
bbs_models %>% 
  filter(term == "year_scaled") %>% 
  select(-model, -model_id, -formula, -n_ts, -warning, -group) %>% 
  relocate(metric) %>% 
  arrange(metric) %>% 
  View()
```

Fit full models and extract the BBS estimates
```{r}

# Fit full models with all the studies
full_models <- map_dfr(metric_names[metric_names != "S"], 
                      ~get_model(data = model_data, 
                                 metric_name = .x, "value ~ year_scaled + (year_scaled|study_id/rarefyID)", 
                                 model_id = "base"))

#get the metrics for which the models didn't converge
non_conv <- full_models %>%
  filter(!is.na(warning)) %>% 
  pull(metric) %>% 
  unique()

#Turn off REML for these models so they'll fit
refit_full_models <- map_dfr(non_conv, 
                      ~get_model(data = model_data, 
                                 metric_name = .x, "value ~ year_scaled + (year_scaled|study_id/rarefyID)", 
                                 model_id = "base", REML = FALSE))

# check that they all fit
unique(refit_full_models$warning)

full_models <- bind_rows(full_models %>% filter(is.na(warning)),
                         refit_full_models)

# Get a dataframe with one row for each model
model_df <- full_models %>%
  select(metric, model, model_id, nobs) %>%
  group_by(metric, model_id) %>% 
  slice_head() %>%
  ungroup()

# Get slopes for just BBS for each metric
bbs_slopes <- pmap_dfr(select(model_df, -nobs), get_study_slopes) %>%
  rename(study_id = level) %>%
  filter(study_id == 195) %>% 
  mutate(p.value = 2*(1-pnorm(abs(estimate), 0, std.error)) ) %>% 
  left_join(model_df %>% select(metric, nobs)) %>%
  relocate(metric) %>% 
  arrange(metric)
```

### 2022-12-07 paper/manuscript.Rmd

@@
 
-# Acknowledgments
+<!-- # Acknowledgments -->
 
-We thank the scientists who contributed to and maintain the biodiversity databases included in this study, including BioTIME, Amphibio, and Elton Traits.
-We thank Dr Shane Blowes and Dr Sarah Supp, whose code we adapted for initial data processing of the BioTIME database.
-Support for this work was provided by U.S.
+<!-- We thank the scientists who contributed to and maintain the biodiversity databases included in this study, including BioTIME, Amphibio, and Elton Traits. -->
+<!-- We thank Dr Shane Blowes and Dr Sarah Supp, whose code we adapted for initial data processing of the BioTIME database. -->
+<!-- Support for this work was provided by U.S.
 Department of Energy through the Computational Sciences Graduate Fellowship (DE-FG02-97ER25308) awarded to KEAN.
 CB acknowledges support from NSF CAREER Award #1942280 & NIFA project CA-B-INS-0162-H.
+-->
 

### 2022-12-07 analysis/05_model_metrics.Rmd
Add bootstrap approach for calculating study slopes

--- 2022-09-19 analysis/05_model_metrics.Rmd
@@
 
+Get study level slopes using bootstrap resampling
+```{r}
+extract_study_slopes <- function(model){
+
+  #get the statistics for the year variable
+  year_scaled_est <- tidy(model) %>%
+    filter(term == "year_scaled") %>%
+    pull(estimate)
+
+  estimate <- broom.mixed::tidy(model, effects="ran_vals") %>%
+    filter(group == "study_id", term == "year_scaled") %>% 
+    select(-c(effect, group, term)) %>%
+    rename(cond.estimate = estimate) %>%
+    mutate(estimate = cond.estimate + year_scaled_est) %>%
+    select(level, estimate) %>%
+    deframe()
+
+  return(estimate)
+}
+
+bootstrap_study_slopes <- function(metric, model, model_id, nsim){
+  print(c(metric, model_id))
+
+  fric_vars <- bootMer(model, FUN = extract_study_slopes, nsim = nsim, use.u = TRUE)
+
+  estimates <- enframe(fric_vars$t0, name = "study_id", value = "estimate")
+
+  as_tibble(fric_vars$t) %>% 
+    pivot_longer(everything(), names_to = "study_id", values_to = "values") %>%
+    group_by(study_id) %>%
+    summarize(n = n(), std.error = sd(values)/sqrt(n)) %>%
+    left_join(estimates) %>%
+    select(-n) %>%
+    mutate(metric = metric, model_id = model_id)
+}
+
+#future::plan("multisession", workers = 30)
+
+bootstrap_slopes <- pmap_dfr(model_df %>% filter(model_id == "base") %>% mutate(nsim = 100), bootstrap_study_slopes) %>%
+  # furrr::future_pmap_dfr(model_df %>% filter(model_id == "base") %>% mutate(nsim = 100), bootstrap_study_slopes, 
+  #                                          .options = furrr::furrr_options(seed = TRUE)) %>%
+  mutate(upr.ci = estimate + (2*std.error),
+         lwr.ci = estimate - (2*std.error),
+         sig = case_when(
+             lwr.ci < 0 & upr.ci > 0 ~ FALSE,
+             TRUE ~ TRUE
+           ))
+```
+
 Summaries of study slopes

### 2022-12-08 analysis/05_model_metrics.Rmd
Fix issues with fitting mixed model for single study (wasn't filter), add function for fitting glm instead for studies that only have one time series.

@@
 ```{r}
-get_model <- function(metric_name, formula, model_id, data, study_id, ...){
+get_model <- function(metric_name, formula, model_id, data, filter_study_id, ...){
   print(c(metric_name, model_id))
   qlmer <- quietly(lmer)
+  print(filter_study_id)
 
-  if (missing(study_id)){
+  if (missing(filter_study_id)){
     fit_data <- data %>%
@@
   } else{
+    print("Filter study")
+
     fit_data <- data %>%
@@
              climate = as.factor(climate)) %>%
-      filter(study_id == study_id)
+      filter(study_id == filter_study_id)
+
+    print(n_distinct(fit_data$study_id))
   }
 
+  print(dim(fit_data))
+
+  # check if there's only one time series
   if(n_distinct(fit_data$rarefyID) < 2){
@@
 
+  # # check if there are enough time points to fit
+  # year_count <- fit_data %>% count(rarefyID)
+  # if(any(year_count$n == 1)){
+  #   return(tibble())
+  # }
   if(!model_id %in% c("base", "base_dur", "base_srt")){
@@
 
-    tidy(fit$result) %>%
+    if (missing(filter_study_id)){
+      df_out <- tidy(fit$result) %>%
         mutate(metric = metric_name, model = list(fit$result), model_id = model_id, 
@@
                warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings))
+    }else
+      df_out <- tidy(fit$result) %>%
+        mutate(metric = metric_name, model = list(fit$result), model_id = model_id, 
+               n_ts = n_distinct(fit_data$rarefyID),
+               formula = formula,
+               warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings), 
+               study_id = filter_study_id)
   }
@@
 
-Create dataframe of models to fit
+Function to fit model for studies with only one time series (so a glm)
+```{r}
+get_glm_model <- function(metric_name, formula, model_id, data, filter_study_id, ...){
+  print(c(metric_name, model_id))
+  qglm <- quietly(glm)
+  print(filter_study_id)
+
+  fit_data <- data %>%
+    filter(metric == metric_name) %>%
+    mutate(year_scaled = scale(year),
+           duration_scaled = scale(duration, scale = FALSE),
+           startyear_scaled = scale(startyear, scale = FALSE),
+           taxa = as.factor(taxa),
+           realm = as.factor(realm),
+           climate = as.factor(climate)) %>%
+    filter(study_id == filter_study_id)
+
+  print(n_distinct(fit_data$study_id))
+
+  fit <- qglm(formula = formula, data = fit_data, ...)
+
+  df_out <- tidy(fit$result) %>%
+    mutate(metric = metric_name, model = list(fit$result), model_id = model_id, 
+           n_ts = n_distinct(fit_data$rarefyID),
+           formula = formula,
+           warning = ifelse(length(fit$warnings) == 0, NA, fit$warnings), 
+           study_id = filter_study_id)
+
+}
+```
+
+Create dataframe of mixed effects models to fit
 ```{r}
@@
 
+```
+
+#### Fit individual models
+```{r}
+#get list of studies that only have one time series
+single_ts <- model_data %>% 
+  select(study_id, rarefyID, metric) %>%
+  distinct() %>%
+  count(study_id, metric) %>%
+  filter(n == 1) %>%
+  select(-n) %>%
+  rename(metric_name = metric, filter_study_id = study_id)
+
+# get dataframe of mixed effects models to fit
+metric_df <- bind_rows(tibble(metric_name = metric_names[metric_names != "S"], 
+                         formula = "value ~ year_scaled + (year_scaled|rarefyID)"),
+          tibble(metric_name = c("S"), 
+                         formula = "logvalue ~ year_scaled + (year_scaled|rarefyID)")) %>%
+  mutate(model_id = "base")
+
+# exclude studies/metrics with only one rarefyID
+mixed_indv_model_df <- map_dfr(unique(model_data$study_id), ~metric_df %>% mutate(filter_study_id = .x)) %>%
+  anti_join(single_ts, by = c("metric_name", "filter_study_id"))
+
+# fit models for individual studies 
+mixed_indv_mod_results <- pmap_dfr(indv_model_df %>% filter(!metric_name %in% c("Jaccard_base", "Jaccard_next")), get_model, data = model_data)
+
+#get dataframe of glm's to fit
+glm_indv_model_df <- single_ts %>% 
+  mutate(model_id = "base",
+         formula = "value ~ year_scaled")
+
+#fit them
+glm_indv_mod_results <- pmap_dfr(glm_indv_model_df, get_glm_model, data = model_data)
+
+indv_mod_results <- bind_rows(mixed_indv_mod_results, glm_indv_mod_results)
 ```
@@
            ))
+
+slope_comp <- study_slopes %>% 
+  select(study_id, metric, sig_bulp = sig) %>%
+  left_join(bootstrap_slopes %>% 
+              select(study_id, metric, sig_boot = sig)) %>%
+  left_join(indv_mod_results %>%
+              filter(term == "year_scaled") %>%
+            mutate(sig_indv = if_else(p.value < 0.05, TRUE, FALSE)) %>%
+              select(study_id, metric, sig_indv))
 ```

### 2023-01-16 analysis/exporatory_scripts/transformations_outliers.Rmd
First commit

---
title: "Outlier assessment"
output: html_document
date: "2023-01-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(broom.mixed)
library(lmerTest)
library(emmeans)
```

#### Read in data, get final dataset for models
```{r}
load(here::here("data/model_data.rda"))

model_data <- filter(model_data, taxa != "Amphibians")
```

Let's look at models for FRic

### FRic model

```{r}
fric_data <- model_data %>% filter(metric == "FRic") %>%
  rename(FRic = value) %>%
   mutate(year_scaled = scale(year),
          taxa = as.factor(taxa),
             realm = as.factor(realm),
             climate = as.factor(climate),
          logFRic = log(FRic))
```

let's do the full model for all studies
```{r}
fric_full_fit <- lmer("FRic ~ year_scaled + (year_scaled|study_id/rarefyID)", data = fric_data, REML = FALSE)

logfric_full_fit <- lmer("logFRic ~ year_scaled + (year_scaled|study_id/rarefyID)", data = fric_data, control = lmerControl(optimizer = "bobyqa"))

# remove the outlier
logfric_full_fit_ex <- lmer("logFRic ~ year_scaled + (year_scaled|study_id/rarefyID)", data = fric_data %>% filter(study_id != 515), control = lmerControl(optimizer = "bobyqa"))
```

look at normality
```{r}
# hist(ranef(fric_full_fit)$study_id$`(Intercept)`)
# shapiro.test(ranef(fric_full_fit)$study_id$`(Intercept)`) # test of normality
hist(ranef(fric_full_fit)$study_id$`year_scaled`)
shapiro.test(ranef(fric_full_fit)$study_id$`year_scaled`)

hist(ranef(logfric_full_fit)$study_id$`(Intercept)`)
shapiro.test(ranef(logfric_full_fit)$study_id$`(Intercept)`)
hist(ranef(logfric_full_fit)$study_id$`year_scaled`)
shapiro.test(ranef(logfric_full_fit)$study_id$`year_scaled`)
```
```{r}
test_normality <- function(model_fit, metric_name){
  intercepts <- ranef(model_fit)$study_id$`(Intercept)`
  slopes <- ranef(model_fit)$study_id$`year_scaled`

  hist(intercepts, main = paste("intercepts for", metric_name))
  hist(slopes, main = paste("slopes for", metric_name))

  bind_rows(tidy(shapiro.test(intercepts)) %>% mutate(type = "intercept"),
            tidy(shapiro.test(slopes)) %>% mutate(type = "slope")) %>%
    mutate(metric = metric_name)
}

test_normality(logfric_full_fit, "log(FRic)")
```

# Look at normality for base models we've already fit

```{r}
# model_table comes from 05_model_metrics.Rmd

model_df <- model_table %>%
  filter(model_id == "base") %>%
  select(metric, model, model_id) %>%
  group_by(metric, model_id) %>% 
  slice_head()

mod_normality <- map2_dfr(model_df$model, model_df$metric, test_normality)
```

# Let's make almost everything log, not SES metrics because they're already normalized
```{r}
metric_names <- unique(model_data$metric)

log_mods <- tibble(metric_name = metric_names[!metric_names %in% c("SES_FRic", "SES_FDiv", "SES_FEve")], 
                         formula = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
                   model_id = "base")

log_mods_fit <- pmap_dfr(log_mods, get_model, data = model_data)

log_model_df <- log_mods_fit %>%
  filter(model_id == "base") %>%
  select(metric, model, model_id) %>%
  group_by(metric, model_id) %>% 
  slice_head()

logmod_normality <- map2_dfr(log_model_df$model, log_model_df$metric, test_normality)
```

# The logvalue in model_data is actually log(value + 1), which works for the CWM but not the summary metrics
```{r}
log_metrics <- c("FRic", "S", "FDiv", "FEve")
log_mods <- tibble(metric_name = log_metrics, 
                         formula = "log(value) ~ year_scaled + (year_scaled|study_id/rarefyID)",
                   model_id = "base")

logvalue_mods <- tibble(metric_name = metric_names[!metric_names %in% append(c("SES_FRic", "SES_FDiv", "SES_FEve"), log_metrics)], 
                         formula = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
                   model_id = "base")

trans_mods <- bind_rows(log_mods, logvalue_mods)

trans_mods_fit <- pmap_dfr(trans_mods, get_model, data = model_data) %>%
  mutate(REML = TRUE, bobyqa = FALSE)

# have to make sure they all converge #

#get the models that had convergence warnings
bad_models <- trans_mods_fit %>%
  select(metric, model_id, warning) %>% 
  filter(!is.na(warning)) %>% 
  distinct() %>%
  left_join(trans_mods, by = c("metric" = "metric_name", "model_id"))

#Turn off REML for these models, and they fit! 
bad_mod_results <- pmap_dfr(bad_models %>% select(-warning), get_model, data = model_data, REML = FALSE) %>%
  mutate(REML = FALSE, bobyqa = FALSE)

#models that still won't converge
bad_models2 <- bad_mod_results %>%
  select(metric, model_id, warning) %>% 
  filter(!is.na(warning)) %>% 
  distinct() %>%
  left_join(trans_mods, by = c("metric" = "metric_name", "model_id"))

#Fit with a different algorithm
bad_mod_results2 <- pmap_dfr(bad_models2 %>% select(-warning), get_model, data = model_data, control = lmerControl(optimizer = "bobyqa")) %>%
  mutate(REML = TRUE, bobyqa = TRUE)

transmodel_table <- bind_rows(trans_mods_fit %>% filter(is.na(warning)), 
          bad_mod_results %>% filter(is.na(warning)),
          bad_mod_results2 %>% filter(is.na(warning)))

# let's look at normality of the random effects
trans_model_df <- transmodel_table %>%
  select(metric, model, model_id, formula, REML, bobyqa) %>%
  group_by(metric, model_id) %>% 
  slice_head() %>%
  ungroup()

transmod_normality <- map2_dfr(trans_model_df$model, trans_model_df$metric, test_normality)
```

# Let's figure out if we have outlier issues, after visually inspecting the random effects

```{r}
s_mod <- trans_model_df %>% filter(metric == "S") %>% pull(model) %>% .[[1]]

intercepts <- ranef(s_mod)$study_id$`(Intercept)`
slopes <- ranef(s_mod)$study_id$`year_scaled`

which(slopes < -1) # It's row 22
ranef(s_mod)$study_id[22,] # It's study 377

s_fit <- lmer("log(value) ~ year_scaled + (year_scaled|study_id/rarefyID)", data = model_data %>% filter(metric == "S", study_id != 377), control = lmerControl(optimizer = "bobyqa"))
```

Let's make it a function
```{r}
check_outliers <- function(start_mod, metric_name, outlier_criterion, lmer_formula, REML = TRUE, bobyqa = FALSE){

  intercepts <- ranef(start_mod)$study_id$`(Intercept)`
  slopes <- ranef(start_mod)$study_id$`year_scaled`

  rows <- which(eval(rlang::parse_expr(outlier_criterion))) # It's row 22
  print(rows)
  if(length(rows) > 1) warning(paste("There is more than one study outside the outlier cutoff for", metric_name))

  study <- as.integer(rownames(ranef(start_mod)$study_id[rows,]))

  if (isTRUE(REML)){
    if (isTRUE(bobyqa)) {
        fit <- lmer(lmer_formula, 
              data = model_data %>% filter(metric == metric_name, study_id != study), 
              control = lmerControl(optimizer = "bobyqa"))
    } else {
      fit <- lmer(lmer_formula, 
              data = model_data %>% filter(metric == metric_name, study_id != study))
    }
  } else{
    if (isTRUE(bobyqa)) {
        fit <- lmer(lmer_formula, 
              data = model_data %>% filter(metric == metric_name, study_id != study), 
              REML = FALSE,
              control = lmerControl(optimizer = "bobyqa"))
    } else {
      fit <- lmer(lmer_formula, 
              data = model_data %>% filter(metric == metric_name, study_id != study),
              REML = FALSE)
    }
  }

  tibble(metric = metric_name,
         old_estimate = tidy(start_mod) %>% filter(term == "year_scaled") %>% pull(estimate), 
         old_p = tidy(start_mod) %>% filter(term == "year_scaled") %>% pull(p.value), 
         new_estimate = tidy(fit) %>% filter(term == "year_scaled") %>% pull(estimate), 
         new_p = tidy(fit) %>% filter(term == "year_scaled") %>% pull(p.value))

}

#check_outliers("S", "slopes < -1", "log(value) ~ year_scaled + (year_scaled|study_id/rarefyID)", REML = TRUE, bobyqa = TRUE)

# create dataframe for outlier criterion based on visualize of random effects
criterion <- tibble(metric_name = c("CWM_bodymass_value", "CWM_diet_inv", "CWM_diet_scav", "CWM_diet_vect", 
                                     "CWM_diet_vend", "CWM_diet_vunk", "CWM_forstrat_aerial", "CWM_forstrat_midhigh",
                                     "CWM_forstrat_understory", "Jaccard_base", "S"),
                    outlier_criterion = c("slopes > 0.4", "slopes < -0.5", "slopes > 0.6", "slopes > 0.4", 
                                          "slopes > 0.5", "slopes < -0.5", "slopes > 0.04", "slopes < -0.05", 
                                          "slopes < -0.4", "slopes > 0.02", "slopes < -1")
                    )

# get metric info for how models were fit
outlier_mods <- trans_model_df %>% 
  select(start_mod = model, metric_name = metric, lmer_formula = formula, REML, bobyqa) %>%
  right_join(criterion) %>%
  select(start_mod, metric_name, outlier_criterion, lmer_formula, REML, bobyqa)

outlier_comp <- pmap_dfr(outlier_mods, check_outliers)

```

### 2023-01-17 analysis/05_model_metrics.Rmd
Modify the response variable for better model fit, keep track of model fitting controls in the results table

@@
   qlmer <- quietly(lmer)
-  print(filter_study_id)
+  #print(filter_study_id)
 
@@
 metric_names <- unique(model_data$metric)
-raw_value_mods <- tibble(metric_name = metric_names[metric_names != "S"], 
+raw_metrics <- c("SES_FRic", "SES_FDiv", "SES_FEve")
+log_metrics <- c("FRic", "S", "FDiv", "FEve")
+logone_metrics <- metric_names[!metric_names %in% c(raw_metrics, log_metrics)]
+
+raw_value_mods <- tibble(metric_name = raw_metrics, 
                          base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
@@
 
-log_value_mods <- tibble(metric_name = c("S"), 
+log_mods <- tibble(metric_name = log_metrics, 
+                         base = "log(value) ~ year_scaled + (year_scaled|study_id/rarefyID)",
+                         taxa = "log(value) ~ year_scaled * taxa + (year_scaled|study_id/rarefyID)",
+                         realm = "log(value) ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
+                         climate = "log(value) ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
+  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
+
+logone_mods <- tibble(metric_name = logone_metrics, 
                          base = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
@@
 
-mods <- bind_rows(raw_value_mods, log_value_mods) %>% 
+mods <- bind_rows(raw_value_mods, log_mods, logone_mods) %>% 
   select(metric_name, formula, model_id)
@@
 #Fit models with base settings
-mod_results <- pmap_dfr(mods, get_model, data = model_data)
+mod_results <- pmap_dfr(mods, get_model, data = model_data) %>%
+  mutate(REML = TRUE, bobyqa = FALSE)
 
@@
 #Turn off REML for these models, and they fit! 
-bad_mod_results <- pmap_dfr(bad_models %>% select(-warning), get_model, data = model_data, REML = FALSE)
+bad_mod_results <- pmap_dfr(bad_models %>% select(-warning), get_model, data = model_data, REML = FALSE) %>%
+  mutate(REML = FALSE, bobyqa = FALSE)
 
@@
 #Fit with a different algorithm
-bad_mod_results2 <- pmap_dfr(bad_models2 %>% select(-warning), get_model, data = model_data, control = lmerControl(optimizer = "bobyqa"))
+bad_mod_results2 <- pmap_dfr(bad_models2 %>% select(-warning), get_model, data = model_data, control = lmerControl(optimizer = "bobyqa")) %>%
+  mutate(REML = TRUE, bobyqa = TRUE)
 
@@
 # fit models for individual studies 
-mixed_indv_mod_results <- pmap_dfr(indv_model_df %>% filter(!metric_name %in% c("Jaccard_base", "Jaccard_next")), get_model, data = model_data)
+mixed_indv_mod_results <- pmap_dfr(mixed_indv_model_df %>% filter(!metric_name %in% c("Jaccard_base", "Jaccard_next")), get_model, data = model_data)
 

### 2023-01-17 analysis/05_model_metrics.Rmd
Add outlier checks

@@
 
+#### Examine Outliers
+
+Let's identify any major outliers in slope random effects for the base models to see if they're impacting the overall trend.
+
+First two functions to visualize and exclude outliers
+
+```{r}
+test_normality <- function(model_fit, metric_name){
+  intercepts <- ranef(model_fit)$study_id$`(Intercept)`
+  slopes <- ranef(model_fit)$study_id$`year_scaled`
+
+  hist(intercepts, main = paste("intercepts for", metric_name))
+  hist(slopes, main = paste("slopes for", metric_name))
+
+  bind_rows(tidy(shapiro.test(intercepts)) %>% mutate(type = "intercept"),
+            tidy(shapiro.test(slopes)) %>% mutate(type = "slope")) %>%
+    mutate(metric = metric_name)
+}
+
+check_outliers <- function(start_mod, metric_name, outlier_criterion, lmer_formula, REML = TRUE, bobyqa = FALSE){
+
+  intercepts <- ranef(start_mod)$study_id$`(Intercept)`
+  slopes <- ranef(start_mod)$study_id$`year_scaled`
+
+  rows <- which(eval(rlang::parse_expr(outlier_criterion))) # It's row 22
+  print(rows)
+  if(length(rows) > 1) warning(paste("There is more than one study outside the outlier cutoff for", metric_name))
+
+  study <- as.integer(rownames(ranef(start_mod)$study_id[rows,]))
+
+  if (isTRUE(REML)){
+    if (isTRUE(bobyqa)) {
+        fit <- lmer(lmer_formula, 
+              data = model_data %>% filter(metric == metric_name, study_id != study), 
+              control = lmerControl(optimizer = "bobyqa"))
+    } else {
+      fit <- lmer(lmer_formula, 
+              data = model_data %>% filter(metric == metric_name, study_id != study))
+    }
+  } else{
+    if (isTRUE(bobyqa)) {
+        fit <- lmer(lmer_formula, 
+              data = model_data %>% filter(metric == metric_name, study_id != study), 
+              REML = FALSE,
+              control = lmerControl(optimizer = "bobyqa"))
+    } else {
+      fit <- lmer(lmer_formula, 
+              data = model_data %>% filter(metric == metric_name, study_id != study),
+              REML = FALSE)
+    }
+  }
+
+  tibble(metric = metric_name,
+         old_estimate = tidy(start_mod) %>% filter(term == "year_scaled") %>% pull(estimate), 
+         old_p = tidy(start_mod) %>% filter(term == "year_scaled") %>% pull(p.value), 
+         new_estimate = tidy(fit) %>% filter(term == "year_scaled") %>% pull(estimate), 
+         new_p = tidy(fit) %>% filter(term == "year_scaled") %>% pull(p.value))
+
+}
+```
+
+Visualize the random effects for the base models
+```{r}
+base_mods <- model_table %>% 
+  filter(model_id == "base") %>%
+  select(metric, model, formula, REML, bobyqa) %>%
+  group_by(metric) %>% 
+  slice_head() %>%
+  ungroup()
+
+mod_normality <- map2_dfr(base_mods$model, base_mods$metric, test_normality)
+```
+
+Based on the visualization, we identify some metrics that are suspect for outliers
+```{r}
+# create dataframe for outlier criterion based on visualize of random effects
+criterion <- tibble(metric_name = c("CWM_bodymass_value", "CWM_diet_inv", "CWM_diet_scav", "CWM_diet_vect", 
+                                     "CWM_diet_vend", "CWM_diet_vunk", "CWM_forstrat_aerial", "CWM_forstrat_midhigh",
+                                     "CWM_forstrat_understory", "Jaccard_base", "S"),
+                    outlier_criterion = c("slopes > 0.4", "slopes < -0.5", "slopes > 0.6", "slopes > 0.4", 
+                                          "slopes > 0.5", "slopes < -0.5", "slopes > 0.04", "slopes < -0.05", 
+                                          "slopes < -0.4", "slopes > 0.02", "slopes < -1")
+                    )
+
+# get info for how original models were fit
+outlier_mods <- trans_model_df %>% 
+  select(start_mod = model, metric_name = metric, lmer_formula = formula, REML, bobyqa) %>%
+  right_join(criterion) %>%
+  select(start_mod, metric_name, outlier_criterion, lmer_formula, REML, bobyqa)
+
+# refit without outliers and compare
+outlier_comp <- pmap_dfr(outlier_mods, check_outliers)
+```
+
 #### Fit individual models

### 2023-01-17 analysis/05_model_metrics.Rmd
Fit individual studies using the appropriate response variable, deal with convergence stuff

@@
 # get dataframe of mixed effects models to fit
-metric_df <- bind_rows(tibble(metric_name = metric_names[metric_names != "S"], 
+metric_df <- bind_rows(tibble(metric_name = raw_metrics, 
                          formula = "value ~ year_scaled + (year_scaled|rarefyID)"),
-          tibble(metric_name = c("S"), 
-                         formula = "logvalue ~ year_scaled + (year_scaled|rarefyID)")) %>%
+          tibble(metric_name = logone_metrics, 
+                         formula = "logvalue ~ year_scaled + (year_scaled|rarefyID)"),
+          tibble(metric_name = log_metrics, 
+                         formula = "log(value) ~ year_scaled + (year_scaled|rarefyID)")) %>%
   mutate(model_id = "base")
@@
 
+# they didn't all converge
+#get the models that had convergence warnings
+bad_indv_models <- mixed_indv_mod_results %>%
+  select(metric, model_id, study_id, warning) %>% 
+  filter(!is.na(warning)) %>% 
+  distinct() %>%
+  left_join(mixed_indv_model_df, 
+            by = c("metric" = "metric_name", "model_id", "study_id" = "filter_study_id")) %>%
+  rename( filter_study_id = study_id)
+
+#Turn off REML for these models, and they fit! 
+bad_indv_results <- pmap_dfr(bad_indv_models %>% select(-warning), get_model, data = model_data, REML = FALSE) %>%
+  mutate(REML = FALSE, bobyqa = FALSE)
+
+#models that still won't converge
+bad_indv_models2 <- bad_indv_results %>%
+  select(metric, model_id, warning, study_id) %>% 
+  filter(!is.na(warning)) %>% 
+  distinct() %>%
+  left_join(mixed_indv_model_df, 
+            by = c("metric" = "metric_name", "model_id", "study_id" = "filter_study_id")) %>%
+  rename( filter_study_id = study_id)
+
+#Fit with a different algorithm
+bad_indv_results2 <- pmap_dfr(bad_indv_models2 %>% select(-warning), get_model, data = model_data, control = lmerControl(optimizer = "bobyqa")) %>%
+  mutate(REML = TRUE, bobyqa = TRUE)
+
+##FIXME: some of these still don't converge##
+```
+
+Fit models for studies with only one time series
+```{r}
+
 #get dataframe of glm's to fit
@@
   mutate(model_id = "base",
-         formula = "value ~ year_scaled")
+         formula = case_when(
+           metric_name %in% raw_metrics ~ "value ~ year_scaled",
+           metric_name %in% log_metrics ~ "log(value) ~ year_scaled",
+           TRUE ~ "logvalue ~ year_scaled"
+           )
+         )
 
@@
 
-indv_mod_results <- bind_rows(mixed_indv_mod_results, glm_indv_mod_results)
+indv_mod_results <- bind_rows(mixed_indv_mod_results %>% filter(is.na(warning)), 
+          bad_indv_results %>% filter(is.na(warning)),
+          bad_indv_results2 %>% filter(is.na(warning)),
+          glm_indv_mod_results
+          )
 ```

### 2023-01-17 analysis/05_model_metrics.Rmd
fix copy paste issues in outlier section

@@
 
+unique_mod_df <- model_table %>%
+  filter(model_id == "base") %>%
+  select(metric, model, model_id, formula, REML, bobyqa) %>%
+  group_by(metric) %>% 
+  slice_head() %>%
+  ungroup()
+
 # get info for how original models were fit
-outlier_mods <- trans_model_df %>% 
+outlier_mods <- unique_mod_df %>% 
   select(start_mod = model, metric_name = metric, lmer_formula = formula, REML, bobyqa) %>%
@@
 ```{r}
-
 #get dataframe of glm's to fit

### 2023-01-19 analysis/05_model_metrics.Rmd
Figured out why bootstrap doesn't work for different optimization algorithms - pmap and quitely are obscuring the formula call. Had to do an ugly workaround.

@@
 ```{r}
-get_model <- function(metric_name, formula, model_id, data, filter_study_id, ...){
+get_model <- function(metric_name, formula, model_id, data, filter_study_id, bobyqa = FALSE, ...){
   print(c(metric_name, model_id))
@@
 
+      if (isFALSE(bobyqa)){
         fit <- qlmer(formula = formula, data = fit_data, ...)
+      } else{
+        fit <- qlmer(formula = formula, data = fit_data, control = lmerControl(optimizer = "bobyqa"), ...)
+        plain_fit <- lmer(formula = formula, data = fit_data, control = lmerControl(optimizer = "bobyqa"), ...)
+      }
 
@@
         bind_rows(group_slopes, slope_constrasts) %>%
-        mutate(metric = metric_name, model = list(fit$result), model_id = model_id,
+        mutate(metric = metric_name, 
+               model = ifelse(exists("plain_fit"), list(plain_fit), list(fit$result)), model_id = model_id,
                n_ts = n_distinct(fit_data$rarefyID),
@@
 
+    if (isFALSE(bobyqa)){
       fit <- qlmer(formula = formula, data = fit_data, ...)
+    } else {
+      fit <- qlmer(formula = formula, data = fit_data, control = lmerControl(optimizer = "bobyqa"), ...)
+      plain_fit <- lmer(formula = formula, data = fit_data, control = lmerControl(optimizer = "bobyqa"), ...)
+    }
 
@@
       df_out <- tidy(fit$result) %>%
-        mutate(metric = metric_name, model = list(fit$result), model_id = model_id, 
+        mutate(metric = metric_name, 
+               model = ifelse(exists("plain_fit"), list(plain_fit), list(fit$result)), model_id = model_id, 
                n_ts = n_distinct(fit_data$rarefyID),
@@
 #Fit with a different algorithm
-bad_mod_results2 <- pmap_dfr(bad_models2 %>% select(-warning), get_model, data = model_data, control = lmerControl(optimizer = "bobyqa")) %>%
+bad_mod_results2 <- pmap_dfr(bad_models2 %>% select(-warning), get_model, data = model_data, bobyqa = TRUE) %>%
   mutate(REML = TRUE, bobyqa = TRUE)
@@
 bootstrap_study_slopes <- function(metric, model, model_id, nsim){
+  browser()
   print(c(metric, model_id))
@@
 
-bootstrap_slopes <- pmap_dfr(model_df %>% filter(model_id == "base") %>% mutate(nsim = 100), bootstrap_study_slopes) %>%
+bootstrap_slopes <- pmap_dfr(model_df %>% filter(model_id == "base", metric == "CWM_bodymass_value") %>% mutate(nsim = 100), bootstrap_study_slopes) %>%
   # furrr::future_pmap_dfr(model_df %>% filter(model_id == "base") %>% mutate(nsim = 100), bootstrap_study_slopes, 

### 2023-02-09 analysis/05_model_metrics.Rmd
clean up

@@
   } else{
-    print("Filter study")
+    print(paste("Filter study", filter_study))
 
@@
 # get dataframe of mixed effects models to fit
-metric_df <- bind_rows(tibble(metric_name = raw_metrics, 
-                         formula = "value ~ year_scaled + (year_scaled|rarefyID)"),
-          tibble(metric_name = logone_metrics, 
-                         formula = "logvalue ~ year_scaled + (year_scaled|rarefyID)"),
-          tibble(metric_name = log_metrics, 
-                         formula = "log(value) ~ year_scaled + (year_scaled|rarefyID)")) %>%
-  mutate(model_id = "base")
+metric_df <- model_data %>%
+  select(filter_study_id = study_id, metric_name = metric) %>% 
+  distinct() %>% 
+  mutate(model_id = "base",
+         formula = case_when(
+           metric_name %in% raw_metrics ~ "value ~ year_scaled + (year_scaled|rarefyID)",
+           metric_name %in% log_metrics ~ "log(value) ~ year_scaled + (year_scaled|rarefyID)",
+           TRUE ~ "logvalue ~ year_scaled + (year_scaled|rarefyID)"
+           )
+         )
 
 # exclude studies/metrics with only one rarefyID
-mixed_indv_model_df <- map_dfr(unique(model_data$study_id), ~metric_df %>% mutate(filter_study_id = .x)) %>%
+mixed_indv_model_df <- metric_df %>%
   anti_join(single_ts, by = c("metric_name", "filter_study_id"))
@@
 
-##FIXME: some of these still don't converge##
+# There are 12 study/metric combos that can't be fit due to data limitations (too many rarefyID's )
+# tofit_mods <- bad_indv_results2 %>% filter(!is.na(warning)) %>% select(metric, study_id, formula) %>% distinct()
+# 
+# pmap_df(tofit_mods %>% select(metric_name = metric, filter_study_id = study_id), function(metric_name, filter_study_id) {model_data %>% 
+#          filter(metric_name == metric, study_id == filter_study_id) %>%
+#         mutate(n = n(), nrarefyID = n_distinct(rarefyID)) %>%
+#         select(metric, study_id, nrarefyID, n) %>%
+#         distinct()}) %>% View()
+
 ```
@@
 bootstrap_study_slopes <- function(metric, model, model_id, nsim){
-  browser()
+
   print(c(metric, model_id))
@@
 
-bootstrap_slopes <- pmap_dfr(model_df %>% filter(model_id == "base", metric == "CWM_bodymass_value") %>% mutate(nsim = 100), bootstrap_study_slopes) %>%
+bootstrap_slopes <- pmap_dfr(model_df %>% filter(model_id == "base") %>% mutate(nsim = 100), bootstrap_study_slopes) %>%
   # furrr::future_pmap_dfr(model_df %>% filter(model_id == "base") %>% mutate(nsim = 100), bootstrap_study_slopes, 
@@
   left_join(bootstrap_slopes %>% 
-              select(study_id, metric, sig_boot = sig)) %>%
+              select(study_id, metric, boot_estimate = estimate, boot_stde = std.error, sig_boot = sig)) %>%
   left_join(indv_mod_results %>%
@@
             mutate(sig_indv = if_else(p.value < 0.05, TRUE, FALSE)) %>%
-              select(study_id, metric, sig_indv))
+              select(study_id, metric, indv_estimate = estimate, indv_stde = std.error, sig_indv))
 ```

### 2023-02-14 analysis/05_model_metrics.Rmd

@@
 ```{r}
+#exclude amphibians
+model_data <- model_data %>% filter(taxa != "Amphibians")
+
 metric_names <- unique(model_data$metric)
-raw_metrics <- c("SES_FRic", "SES_FDiv", "SES_FEve")
-log_metrics <- c("FRic", "S", "FDiv", "FEve")
+raw_metrics <- c("SES_FRic", "SES_FDiv", "SES_FEve", 
+                 "CWM_diet_inv", "CWM_forstrat_ground", "CWM_forstrat_watbelowsurf", "FEve")
+log_metrics <- c("FRic", "S", "FDiv")
 logone_metrics <- metric_names[!metric_names %in% c(raw_metrics, log_metrics)]
@@
 ```{r}
-#exclude amphibians
-model_data <- model_data %>% filter(taxa != "Amphibians")
-
 #Fit models with base settings
@@
 
+Fit Jaccard Mixed models to see which will fit
+```{r}
+mixed_jacc_mod_results <- pmap_dfr(mixed_indv_model_df %>% filter(metric_name == "Jaccard_base"), possibly(get_model, otherwise = tibble()), data = model_data)
+
+non_conv_study <- mixed_jacc_mod_results %>% filter(!is.na(warning)) %>% pull(study_id) %>% unique()
+
+mixed_jacc_mod_results2 <- pmap_dfr(mixed_indv_model_df %>% filter(metric_name == "Jaccard_base", filter_study_id == non_conv_study), 
+                                    possibly(get_model, otherwise = tibble()), data = model_data, bobyqa = TRUE)
+
+mixed_jacc_mod_results <- bind_rows(mixed_jacc_mod_results %>% filter(study_id != non_conv_study),
+                                    mixed_jacc_mod_results2)
+```
+
 Fit models for studies with only one time series
@@
           bad_indv_results2 %>% filter(is.na(warning)),
-          glm_indv_mod_results
+          glm_indv_mod_results,
+          mixed_jacc_mod_results
           )
-```
 
-#### Process Model Output
-
-Get study level slope estimates
-```{r}
-get_study_slopes <- function(metric, model, model_id){
-
-  filter_name <- ifelse(model_id == "base", "study_id", model_id)
-
-  #get the statistics for the year variable
-  year_scaled_coef <- tidy(model) %>%
-    filter(term == "year_scaled")
-
-  #get the year estimate
-  year_scaled_est <- pull(year_scaled_coef, estimate)
-
-  #get year variation
-  year_scaled_var <- year_scaled_coef %>% 
-    mutate(var = std.error*std.error) %>%
-    pull(var)
-
-  #get study level conditional estimates and add to over all estimates
-  study_ests <- broom.mixed::tidy(model, effects="ran_vals") %>%
-    filter(group == filter_name, term == "year_scaled") %>% 
-    select(-c(effect, group, term)) %>%
-    rename(cond.std.error = std.error, cond.estimate = estimate) %>%
-    mutate(estimate = cond.estimate + year_scaled_est,
-      cond.var = cond.std.error*cond.std.error, 
-           var = cond.var + year_scaled_var,
-           std.error = sqrt(var),
-           upr.ci = estimate + (1.96*std.error),
-           lwr.ci = estimate - (1.96*std.error),
-           sig = case_when(
-             lwr.ci < 0 & upr.ci > 0 ~ FALSE,
-             TRUE ~ TRUE
-           ),
-      metric = metric,
-      model_id = model_id)
-}
-
-model_df <- model_table %>%
-  filter(model_id == "base") %>%
-  select(metric, model, model_id) %>%
-  group_by(metric, model_id) %>% 
-  slice_head()
-
-#get study level slopes for base, and levels for categorical random effect in non-base models
-re_slopes <- pmap_dfr(model_df, get_study_slopes)
-
-#get only study-level slopes for the base models
-study_slopes <- re_slopes %>% 
-  filter(model_id == "base") %>%
-  rename(study_id = level)
-
-usethis::use_data(study_slopes)
-```
-
-Get study level slopes using bootstrap resampling
-```{r}
-extract_study_slopes <- function(model){
-
-  #get the statistics for the year variable
-  year_scaled_est <- tidy(model) %>%
-    filter(term == "year_scaled") %>%
-    pull(estimate)
-
-  estimate <- broom.mixed::tidy(model, effects="ran_vals") %>%
-    filter(group == "study_id", term == "year_scaled") %>% 
-    select(-c(effect, group, term)) %>%
-    rename(cond.estimate = estimate) %>%
-    mutate(estimate = cond.estimate + year_scaled_est) %>%
-    select(level, estimate) %>%
-    deframe()
-
-  return(estimate)
-}
-
-bootstrap_study_slopes <- function(metric, model, model_id, nsim){
-
-  print(c(metric, model_id))
-
-  fric_vars <- bootMer(model, FUN = extract_study_slopes, nsim = nsim, use.u = TRUE)
-
-  estimates <- enframe(fric_vars$t0, name = "study_id", value = "estimate")
-
-  as_tibble(fric_vars$t) %>% 
-    pivot_longer(everything(), names_to = "study_id", values_to = "values") %>%
-    group_by(study_id) %>%
-    summarize(n = n(), std.error = sd(values)/sqrt(n)) %>%
-    left_join(estimates) %>%
-    select(-n) %>%
-    mutate(metric = metric, model_id = model_id)
-}
-
-#future::plan("multisession", workers = 30)
-
-bootstrap_slopes <- pmap_dfr(model_df %>% filter(model_id == "base") %>% mutate(nsim = 100), bootstrap_study_slopes) %>%
-  # furrr::future_pmap_dfr(model_df %>% filter(model_id == "base") %>% mutate(nsim = 100), bootstrap_study_slopes, 
-  #                                          .options = furrr::furrr_options(seed = TRUE)) %>%
-  mutate(upr.ci = estimate + (2*std.error),
-         lwr.ci = estimate - (2*std.error),
-         sig = case_when(
-             lwr.ci < 0 & upr.ci > 0 ~ FALSE,
-             TRUE ~ TRUE
-           ))
-
-slope_comp <- study_slopes %>% 
-  select(study_id, metric, sig_bulp = sig) %>%
-  left_join(bootstrap_slopes %>% 
-              select(study_id, metric, boot_estimate = estimate, boot_stde = std.error, sig_boot = sig)) %>%
-  left_join(indv_mod_results %>%
-              filter(term == "year_scaled") %>%
-            mutate(sig_indv = if_else(p.value < 0.05, TRUE, FALSE)) %>%
-              select(study_id, metric, indv_estimate = estimate, indv_stde = std.error, sig_indv))
+indv_mod_table <- indv_mod_results %>% select(-model)
+usethis::use_data(indv_mod_table)
 ```

### 2023-02-14 analysis/exporatory_scripts/compare_transformations.Rmd

---
title: "Outlier assessment"
output: html_document
date: "2023-01-13"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lme4)
library(broom.mixed)
library(lmerTest)
library(emmeans)
```

#### Read in data, get final dataset for models
```{r}
load(here::here("data/model_data.rda"))

model_data <- filter(model_data, taxa != "Amphibians")
```

Create dataframe of mixed effects models to fit
```{r}
# metric_names <- unique(model_data$metric)
# raw_metrics <- c("SES_FRic", "SES_FDiv", "SES_FEve")
# log_metrics <- c("FRic", "S", "FDiv", "FEve")
# logone_metrics <- metric_names[!metric_names %in% c(raw_metrics, log_metrics)]
# 
# raw_value_mods <- tibble(metric_name = raw_metrics, 
#                          base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
#                          taxa = "value ~ year_scaled * taxa + (year_scaled|study_id/rarefyID)",
#                          realm = "value ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
#                          climate = "value ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
#   pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula") %>%
#   #don't fit the CWM_litter_size_min_n and climate model, not enough data
#   filter(metric_name != "CWM_litter_size_min_n" | model_id != "climate")
# 
# log_mods <- tibble(metric_name = log_metrics, 
#                          base = "log(value) ~ year_scaled + (year_scaled|study_id/rarefyID)",
#                          taxa = "log(value) ~ year_scaled * taxa + (year_scaled|study_id/rarefyID)",
#                          realm = "log(value) ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
#                          climate = "log(value) ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
#   pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
# 
# logone_mods <- tibble(metric_name = logone_metrics, 
#                          base = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
#                          taxa = "logvalue ~ year_scaled * taxa + (year_scaled|study_id/rarefyID)",
#                          realm = "logvalue ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
#                          climate = "logvalue ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
#   pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
# 
# mods <- bind_rows(raw_value_mods, log_mods, logone_mods) %>% 
#   select(metric_name, formula, model_id)
```

Fit models
```{r}
# #exclude amphibians
# model_data <- model_data %>% filter(taxa != "Amphibians")
# 
# #Fit models with base settings
# mod_results <- pmap_dfr(mods, get_model, data = model_data) %>%
#   mutate(REML = TRUE, bobyqa = FALSE)
# 
# #get the models that had convergence warnings
# bad_models <- mod_results %>%
#   select(metric, model_id, warning) %>% 
#   filter(!is.na(warning)) %>% 
#   distinct() %>%
#   left_join(mods, by = c("metric" = "metric_name", "model_id")) 
# 
# #Turn off REML for these models, and they fit! 
# bad_mod_results <- pmap_dfr(bad_models %>% select(-warning), get_model, data = model_data, REML = FALSE) %>%
#   mutate(REML = FALSE, bobyqa = FALSE)
# 
# #models that still won't converge
# bad_models2 <- bad_mod_results %>%
#   select(metric, model_id, warning) %>% 
#   filter(!is.na(warning)) %>% 
#   distinct() %>%
#   left_join(mods, by = c("metric" = "metric_name", "model_id")) 
# 
# #Fit with a different algorithm
# bad_mod_results2 <- pmap_dfr(bad_models2 %>% select(-warning), get_model, data = model_data, bobyqa = TRUE) %>%
#   mutate(REML = TRUE, bobyqa = TRUE)
# 
# model_table <- bind_rows(mod_results %>% filter(is.na(warning)), 
#           bad_mod_results %>% filter(is.na(warning)),
#           bad_mod_results2 %>% filter(is.na(warning))) %>%
#   mutate(significant = case_when(
#     p.value > 0.05 | lower.CL < 0 & upper.CL > 0 ~ FALSE,
#     !is.na(p.value) | !is.na(lower.CL) ~ TRUE,
#     TRUE ~ NA
#   )) 
```

Need to also fit the non-log models
```{r}
nonlog_mods <- tibble(metric_name = metric_names[metric_names != "S"], 
                         base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
                         taxa = "value ~ year_scaled * taxa + (year_scaled|study_id/rarefyID)",
                         realm = "value ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
                         climate = "value ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula") %>%
  #don't fit the CWM_litter_size_min_n and climate model, not enough data
  filter(metric_name != "CWM_litter_size_min_n" | model_id != "climate") %>% 
  select(metric_name, formula, model_id)
```

Fit models
```{r}
#Fit models with base settings
nonlog_mod_results <- pmap_dfr(nonlog_mods, get_model, data = model_data)

#get the models that had convergence warnings
nonlog_bad_models <- nonlog_mod_results %>%
  select(metric, model_id, warning) %>% 
  filter(!is.na(warning)) %>% 
  distinct() %>%
  left_join(mods, by = c("metric" = "metric_name", "model_id"))

#Turn off REML for these models, and they fit! 
nonlog_bad_mod_results <- pmap_dfr(nonlog_bad_models %>% select(-warning), get_model, data = model_data, REML = FALSE)

#models that still won't converge
nonlog_bad_models2 <- nonlog_bad_mod_results %>%
  select(metric, model_id, warning) %>% 
  filter(!is.na(warning)) %>% 
  distinct() %>%
  left_join(mods, by = c("metric" = "metric_name", "model_id"))

#Fit with a different algorithm
nonlog_bad_mod_results2 <- pmap_dfr(nonlog_bad_models2 %>% select(-warning), get_model, data = model_data, control = lmerControl(optimizer = "bobyqa"))

nonlog_model_table <- bind_rows(nonlog_mod_results %>% filter(is.na(warning)), 
          nonlog_bad_mod_results %>% filter(is.na(warning)),
          nonlog_bad_mod_results2 %>% filter(is.na(warning))) %>%
  mutate(significant = case_when(
    p.value > 0.05 | lower.CL < 0 & upper.CL > 0 ~ FALSE,
    !is.na(p.value) | !is.na(lower.CL) ~ TRUE,
    TRUE ~ NA
  ))
```

Function to compare pre and post transformation normality
```{r}
library(DHARMa)

comp_normality <- function(metric, model, log_model){
  model_sim <- simulateResiduals(fittedModel = model, plot = F)
  log_model_sim <- simulateResiduals(fittedModel = log_model, plot = F)

  plot(model_sim, title = paste(metric, "untransformed"))
  plot(log_model_sim, title = paste(metric, "log transformed"))
}

model_comp_df <- nonlog_model_table %>%
  filter(model_id == "base") %>%
  select(metric, model) %>% 
  group_by(metric) %>%
  slice_head(n=1) %>%
  left_join(model_table %>% 
              filter(model_id == "base") %>%
              select(metric, log_model = model) %>% 
              group_by(metric) %>%
              slice_head(n=1))

pmap(model_comp_df, comp_normality)
```
Based on visual comparison, the following metrics are better *not* transformed: diet_inv, forstrat_ground, forstrat_watbelowsurf, FEve.

### 2023-02-14 analysis/exporatory_scripts/slope_est_comparison.Rmd

---
title: "Untitled"
output: html_document
date: "2023-02-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Get study level slope estimates using BLUP's
```{r}
get_study_slopes <- function(metric, model, model_id){

  filter_name <- ifelse(model_id == "base", "study_id", model_id)

  #get the statistics for the year variable
  year_scaled_coef <- tidy(model) %>%
    filter(term == "year_scaled")

  #get the year estimate
  year_scaled_est <- pull(year_scaled_coef, estimate)

  #get year variation
  year_scaled_var <- year_scaled_coef %>% 
    mutate(var = std.error*std.error) %>%
    pull(var)

  #get study level conditional estimates and add to over all estimates
  study_ests <- broom.mixed::tidy(model, effects="ran_vals") %>%
    filter(group == filter_name, term == "year_scaled") %>% 
    select(-c(effect, group, term)) %>%
    rename(cond.std.error = std.error, cond.estimate = estimate) %>%
    mutate(estimate = cond.estimate + year_scaled_est,
      cond.var = cond.std.error*cond.std.error, 
           var = cond.var + year_scaled_var,
           std.error = sqrt(var),
           upr.ci = estimate + (1.96*std.error),
           lwr.ci = estimate - (1.96*std.error),
           sig = case_when(
             lwr.ci < 0 & upr.ci > 0 ~ FALSE,
             TRUE ~ TRUE
           ),
      metric = metric,
      model_id = model_id)
}

model_df <- model_table %>%
  filter(model_id == "base") %>%
  select(metric, model, model_id) %>%
  group_by(metric, model_id) %>% 
  slice_head()

#get study level slopes for base, and levels for categorical random effect in non-base models
re_slopes <- pmap_dfr(model_df, get_study_slopes)

#get only study-level slopes for the base models
study_slopes <- re_slopes %>% 
  filter(model_id == "base") %>%
  rename(study_id = level)

usethis::use_data(study_slopes)
```

Get study level slopes using bootstrap resampling
```{r}
extract_study_slopes <- function(model){

  #get the statistics for the year variable
  year_scaled_est <- tidy(model) %>%
    filter(term == "year_scaled") %>%
    pull(estimate)

  estimate <- broom.mixed::tidy(model, effects="ran_vals") %>%
    filter(group == "study_id", term == "year_scaled") %>% 
    select(-c(effect, group, term)) %>%
    rename(cond.estimate = estimate) %>%
    mutate(estimate = cond.estimate + year_scaled_est) %>%
    select(level, estimate) %>%
    deframe()

  return(estimate)
}

bootstrap_study_slopes <- function(metric, model, model_id, nsim){

  print(c(metric, model_id))

  fric_vars <- bootMer(model, FUN = extract_study_slopes, nsim = nsim, use.u = TRUE)

  estimates <- enframe(fric_vars$t0, name = "study_id", value = "estimate")

  as_tibble(fric_vars$t) %>% 
    pivot_longer(everything(), names_to = "study_id", values_to = "values") %>%
    group_by(study_id) %>%
    summarize(n = n(), std.error = sd(values)/sqrt(n)) %>%
    left_join(estimates) %>%
    select(-n) %>%
    mutate(metric = metric, model_id = model_id)
}

#future::plan("multisession", workers = 30)

bootstrap_slopes <- pmap_dfr(model_df %>% filter(model_id == "base") %>% mutate(nsim = 100), bootstrap_study_slopes) %>%
  # furrr::future_pmap_dfr(model_df %>% filter(model_id == "base") %>% mutate(nsim = 100), bootstrap_study_slopes, 
  #                                          .options = furrr::furrr_options(seed = TRUE)) %>%
  mutate(upr.ci = estimate + (2*std.error),
         lwr.ci = estimate - (2*std.error),
         sig = case_when(
             lwr.ci < 0 & upr.ci > 0 ~ FALSE,
             TRUE ~ TRUE
           ))
```

```{r}
slope_comp <- study_slopes %>% 
  select(study_id, metric, blup_estimate = estimate, blup_stde = std.error, sig_bulp = sig) %>%
  left_join(bootstrap_slopes %>% 
              select(study_id, metric, boot_estimate = estimate, boot_stde = std.error, sig_boot = sig)) %>%
  left_join(indv_mod_results %>%
              filter(term == "year_scaled") %>%
            mutate(sig_indv = if_else(p.value < 0.05, TRUE, FALSE)) %>%
              select(study_id, metric, indv_estimate = estimate, indv_stde = std.error, sig_indv))
```

```{r}
load(here::here("data/indv_mod_table.rda"))

slopes_long <- slope_comp %>% 
    select(study_id, metric, blup_estimate, blup_stde, boot_estimate, boot_stde, indv_estimate, indv_stde) %>%
  pivot_longer(cols = c(blup_estimate, blup_stde, boot_estimate, boot_stde, indv_estimate, indv_stde),
  names_to = c("type", ".value"),
  names_pattern = "(.+)_(.+)"
  )

slopes_long %>% 
  #filter(metric %in% c("S", "Jaccard_base", "SES_FRic", "SES_FDiv", "SES_FEve")) %>%
  filter(metric == "Jaccard_base") %>%
  filter(type != "blup") %>%
  ggplot(aes(x = study_id, y = estimate)) +
  geom_point(aes(colour = type), size = 3.5) +
  geom_errorbar(aes(ymin = estimate-(2*stde), ymax = estimate + (2*stde))) + 
    facet_wrap(vars(metric))

```

### 2023-02-14 analysis/05_model_metrics.Rmd
Clean up, make sure everything is written out for the paper.

@@
 
-Visualize distributions of response
-```{r}
-# metrics %>%
-#   filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "Jaccard_hind")) %>%
-#   ggplot() +
-#   geom_histogram(aes(x = logvalue),
-#                  bins = 60) +
-#   facet_wrap(~metric, scales = "free") +
-#   labs(title = "Distribution of Log values")
-# 
-# metrics %>%
-#   filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S", "SES_FDis", "SES_FDiv", "SES_FEve", "SES_FRic", "Jaccard_hind")) %>%
-#   ggplot() +
-#   geom_histogram(aes(x = value),
-#                  bins = 60) +
-#   facet_wrap(~metric, scales = "free") +
-#   labs(title = "Distribution of raw values")
-```
-
 ####Fit GLMs to overall patterns and data broken down by groups
@@
   } else{
-    print(paste("Filter study", filter_study))
+    print(paste("Filter study", filter_study_id))
 
@@
     TRUE ~ NA
-  )) #%>%
-  #select(-model)
-
-# get number of observations for each model
-# lmm_obs <- model_table %>% 
-#   filter(model_id == "base") %>% 
-#   select(metric, model) %>% 
-#   rowwise() %>% 
-#   mutate(nobs = nobs(model)) %>% 
-#   select(-model) %>% 
-#   distinct()
+  ))
 
@@
 #this is a list of models that weren't fit because there weren't enough timeseries OR they didn't converge
-#most are amphibian traits or traits only found for one taxa (with taxa as a covariate), the nectar diet realm model wouldn't converge
+#most traits only found for one taxa with taxa as a covariate, the nectar diet realm model was data limited
 missing_mods <- setdiff(uniq_mods, uniq_metric_model)
-```
-
-Write out table of model estimates for manuscript
-```{r}
-#manuscript table
-metric_model_table %>% 
-  filter(model_id == "base", metric %in% c("S", "Jaccard_base", "SES_FRic", "SES_FEve", "SES_FDiv")) %>% 
-  select(-c(model_id, warning, statistic, df)) %>% 
-  write_csv(here::here("figures", "man_model_table.csv"))
-
-#supplement of complete model output
-metric_model_table %>%
-  select(-c(model_id, n_ts, warning, lower.CL, upper.CL, significant)) %>%
-  write_csv(here::here("paper", "model_estimates_supp.csv"))
-
 ```
@@
 check_outliers <- function(start_mod, metric_name, outlier_criterion, lmer_formula, REML = TRUE, bobyqa = FALSE){
+
+  print(metric_name)
 
@@
                                      "CWM_forstrat_understory", "Jaccard_base", "S"),
-                    outlier_criterion = c("slopes > 0.4", "slopes < -0.5", "slopes > 0.6", "slopes > 0.4", 
+                    outlier_criterion = c("slopes > 0.4", "slopes < -20", "slopes > 0.6", "slopes > 0.4", 
                                           "slopes > 0.5", "slopes < -0.5", "slopes > 0.04", "slopes < -0.05", 
@@
 # fit models for individual studies 
-mixed_indv_mod_results <- pmap_dfr(mixed_indv_model_df %>% filter(!metric_name %in% c("Jaccard_base", "Jaccard_next")), get_model, data = model_data)
+mixed_indv_mod_results <- pmap_dfr(mixed_indv_model_df %>% 
+                                     filter(!metric_name %in% c("Jaccard_base", "Jaccard_next")), 
+                                   get_model, data = model_data)
 
@@
 
+Write out table of model estimates for manuscript
+```{r}
+#manuscript table
+metric_model_table %>% 
+  filter(model_id == "base", metric %in% c("S", "Jaccard_base", "SES_FRic", "SES_FEve", "SES_FDiv")) %>% 
+  select(-c(model_id, warning, statistic, df)) %>% 
+  write_csv(here::here("figures", "man_model_table.csv"))
+
+#supplement of complete model output
+bind_rows(metric_model_table, indv_mod_table) %>%
+    select(-c(model_id, n_ts, warning, lower.CL, upper.CL, significant, REML, bobyqa)) %>%
+  write_csv(here::here("paper", "model_estimates_supp.csv"))
+
+```
+
 Summaries of study slopes
@@
 #First - how many studies had a significant trend for each metric? 
-study_slopes %>% 
-  filter(sig == TRUE) %>% 
+indv_mod_table %>% 
+  filter(term == "year_scaled", p.value < 0.05) %>% 
   filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
@@
 #And which studies were they?
-study_slopes %>% 
-  filter(sig == TRUE) %>% 
+indv_mod_table %>% 
+  filter(term == "year_scaled", p.value < 0.05) %>% 
   filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S", "FRic", "FEve", "FDiv")) %>% 
@@
 #Which were significant for multiple metrics?
-study_slopes %>% 
-  filter(sig == TRUE) %>% 
+indv_mod_table %>% 
+  filter(term == "year_scaled", p.value < 0.05) %>% 
   filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
@@
 ```
-
-#### Looking at BBS specifically to compare slopes and significance
-```{r}
-bbs_models <- map_dfr(metric_names[metric_names != "S"], 
-                      ~get_model(data = model_data, 
-                                 metric_name = .x, "value ~ year_scaled + (year_scaled|rarefyID)", 
-                                 model_id = "base",
-                                 study_id = 195))
-
-bbs_models %>% 
-  filter(term == "year_scaled") %>% 
-  rowwise() %>% 
-  mutate(nobs = nobs(model)) %>% 
-  select(-model, -model_id, -formula, -n_ts, -warning, -group) %>% 
-  relocate(metric) %>% 
-  arrange(metric) %>% 
-  write_csv(here::here("data/bbs_slopes_indvmodels.csv"))
-```

### 2023-02-14 analysis/05_model_metrics.Rmd
adapt for glm study model approach

@@
 
-study_slope_data <- study_slopes %>%
+study_slope_data <- indv_mod_table %>%
+  filter(term == "year_scaled") %>%
   dplyr::select(study_id, metric, estimate) %>%
@@
 #All model estimates in one dataframe
-# slope_models <- bind_rows(duration_slope, startyear_slope)
+slope_models <- bind_rows(duration_slope, startyear_slope)
 

### 2023-02-14 analysis/06_sensitivity_analysis.Rmd
update with log models amphibian exclusion

--- 2022-08-30 analysis/06_sensitivity_analysis.Rmd
@@
 load(here::here("data/model_data.rda"))
+#exclude amphibians
+model_data <- model_data %>% filter(taxa != "Amphibians")
 ```
@@
 
+Define models to fit
+```{r}
+metric_names <- unique(model_data$metric)
+raw_metrics <- c("SES_FRic", "SES_FDiv", "SES_FEve", 
+                 "CWM_diet_inv", "CWM_forstrat_ground", "CWM_forstrat_watbelowsurf", "FEve")
+log_metrics <- c("FRic", "S", "FDiv")
+logone_metrics <- metric_names[!metric_names %in% c(raw_metrics, log_metrics)]
+
+raw_value_mods <- tibble(metric_name = raw_metrics, 
+                         base = "value ~ year_scaled + (year_scaled|study_id/rarefyID)",
+                         taxa = "value ~ year_scaled * taxa + (year_scaled|study_id/rarefyID)",
+                         realm = "value ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
+                         climate = "value ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
+  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula") %>%
+  #don't fit the CWM_litter_size_min_n and climate model, not enough data
+  filter(metric_name != "CWM_litter_size_min_n" | model_id != "climate")
+
+log_mods <- tibble(metric_name = log_metrics, 
+                         base = "log(value) ~ year_scaled + (year_scaled|study_id/rarefyID)",
+                         taxa = "log(value) ~ year_scaled * taxa + (year_scaled|study_id/rarefyID)",
+                         realm = "log(value) ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
+                         climate = "log(value) ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
+  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
+
+logone_mods <- tibble(metric_name = logone_metrics, 
+                         base = "logvalue ~ year_scaled + (year_scaled|study_id/rarefyID)",
+                         taxa = "logvalue ~ year_scaled * taxa + (year_scaled|study_id/rarefyID)",
+                         realm = "logvalue ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
+                         climate = "logvalue ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
+  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
+
+mods <- bind_rows(raw_value_mods, log_mods, logone_mods) %>% 
+  select(metric_name, formula, model_id)
+```
+
 1. Min time series duration
@@
 
-duration_sens <- map_dfr(dur_dfs, ~pmap_dfr(mods, get_model, data = .x), .id = "dataID")
+duration_sens <- map_dfr(dur_dfs, ~pmap_dfr(filter(mods, model_id == "base"), get_model, data = .x), .id = "dataID")
 ```
 
-2. Inclusion of Amphibians - nothing changes when you exclude them 
+2. Trait Coverage
 ```{r}
-no_amph <- pmap_dfr(mods, get_model, data = model_data %>% filter(taxa != "Amphibians")) %>%
-  mutate(dataID = no_amph)
+load(here::here("data/study_table.rda"))
 
-# this annoyingly doesn't work because of some dataframe column
-# sens_models <- bind_rows(bind_cols(select(duration_sens, -dataID), pull(duration_sens, dataID)) %>% select(-model, -formula), 
-#                          no_amph %>% select(-model, -formula))
+model_data <- model_data %>%
+  left_join(study_table %>% select(study_id, coverage) %>% mutate(study_id = as.factor(study_id)))
 
-save(duration_sens, no_amph, file = here::here("data/sensitivity_models.rda"))
+covg_dfs <- list(covg75 = filter(model_data, taxa != "Amphibians"),
+                covg85 = filter(model_data, taxa != "Amphibians", coverage >= 0.85),
+                covg90 = filter(model_data, taxa != "Amphibians", coverage >= 0.90))
+
+coverage_sens <- map_dfr(covg_dfs, ~pmap_dfr(filter(mods, model_id == "base"), 
+                                             get_model, data = .x), .id = "dataID")
+
+sens_mods <- bind_rows(duration_sens, coverage_sens) %>%
+  select(-model) %>%
+  filter(metric != "Jaccard_next")
+
+usethis::use_data(sens_mods)
+
+write_csv(sens_mods, here::here("paper", "sensitivity_models.csv"))
 ```
-
-3. Trait Coverage
-```{r}
-
-```

### 2023-02-14 analysis/figures/visualization.Rmd
update figures with new model transformations

@@
 plot_map <- tibble(metric_name = c("S", "Jaccard_base", "SES_FRic", "SES_FDiv", "SES_FEve"), 
-           ylabel = c("log(S)", "Jaccard", "Functional Richness SES", "Functional Divergence SES", "Functional Evenness SES"),
+           ylabel = c("log(S)", "log(Jaccard + 1)", "Functional Richness SES", "Functional Divergence SES", "Functional Evenness SES"),
            #ylabel = c("log(S)", "Jaccard", "Functional\nRichness SES", "Functional\nDivergence SES", "Functional\nEvenness SES"),
-           yval = c("logvalue", "value", "value", "value", "value"),
+           yval = c("log(value)", "logvalue", "value", "value", "value"),
            xlabel = c(NA, NA, NA, "year", "year"))
-m_plots <- pmap(plot_map, plot_metric, data = model_data, 
+m_plots <- pmap(plot_map, plot_metric, data = model_data %>% mutate("log(value)" = log(value)), 
                 model_coef = metric_model_table %>% filter(model_id == "base"), 

### 2023-06-01 06_study_change.Rmd
First commit

---
title: "Study Level trends"
journal: "Ecology Letters"
date: "`r Sys.Date()`"
csl: pnas.csl
layout: 3p
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output: 
  rticles::arxiv_article:
    keep_tex: true
#  word_document: default
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
load(here::here("data/indv_mod_table.rda"))
load(here::here("data/model_data.rda"))
```

# Types of Biodiversity Change

```{r}
wide_table <- indv_mod_table %>% 
  filter(p.value < 0.05, term == "year_scaled", model_id == "base") %>% 
  dplyr::select(study_id, metric, estimate) %>% 
  pivot_wider(names_from = "metric", values_from = "estimate") %>% 
  dplyr::select(., -starts_with("CWM_")) %>% 
  dplyr::select(study_id, S, Jaccard_base, everything(), -Jaccard_next) %>%
  left_join(model_data %>% 
              dplyr::select(study_id, duration) %>%
              distinct() %>%
              group_by(study_id) %>% 
              mutate(duration = max(duration)) %>% distinct() %>% 
              mutate(study_id = as.character(study_id))) 
```
# Add decriptive columns to table 
```{r}
fun_change_table <- wide_table %>%
  mutate(rich_change = ifelse(!is.na(S), "Richness Change", "No Richness Change"),
         turnover = ifelse(!is.na(Jaccard_base), "Turnover", "No Turnover"),
         fun_change = case_when(
           if_all(c(FRic, SES_FRic, FDiv, SES_FDiv, FEve, SES_FEve), ~is.na(.)) ~ "No Functional Change",
           .default = "Functional Change"
         ))

labels = fun_change_table %>%
  dplyr::select(rich_change, turnover, fun_change) %>%
  distinct() %>%
  mutate(group = LETTERS[1:8])

fun_change_table <- fun_change_table %>%
  left_join(labels)

fun_change_table %>% 
  count(rich_change, turnover, fun_change, group) 
```

Let's split up the corrected and uncorrected functional metrics
```{r}
ses_wide_table <-  wide_table %>%
  dplyr::select(-FRic, -FDiv, -FEve) %>%
  mutate(rich_change = ifelse(!is.na(S), TRUE, FALSE),
         turnover = ifelse(!is.na(Jaccard_base), TRUE, FALSE),
         fun_change = case_when(
           if_all(c(SES_FRic, SES_FDiv, SES_FEve), ~is.na(.)) ~ FALSE,
           .default = TRUE
         ))

ses_wide_table %>% 
  count(rich_change, turnover, fun_change) 
```
```{r}
uncor_wide_table <-  wide_table %>%
  dplyr::select(-SES_FRic, -SES_FDiv, -SES_FEve) %>%
  mutate(rich_change = ifelse(!is.na(S), TRUE, FALSE),
         turnover = ifelse(!is.na(Jaccard_base), TRUE, FALSE),
         fun_change = case_when(
           if_all(c(FRic, FDiv, FEve), ~is.na(.)) ~ FALSE,
           .default = TRUE
         ))

uncor_wide_table %>% 
  count(rich_change, turnover, fun_change) 
```

# Look at distribution of durations
```{r}
duration_table <- fun_change_table %>%
  dplyr::select(study_id, rich_change, turnover, fun_change, group) %>%
  left_join(model_data %>% 
              dplyr::select(study_id, rarefyID, duration) %>% 
              distinct())

```

```{r}
plot_table <- duration_table %>%
  filter(group %in% c("A", "B", "C", "D"))

group_types <- as_labeller(
  c("A" = "No species change or turnover, functional change", "B" = "Turnover, no richness or functional change", "C" = "No change", "D" = "Turnover, species and functional Change")
)

my_pal <- c("#d9be3b", "#77976e", "#086788", "#cc8214", "#bfac88", "#99adbf")
group_colors <- c("A" = "#d9be3b", "B" = "#77976e", "C" = "#086788", "D" = "#cc8214")

ggplot(plot_table, aes(x = duration, fill = group)) +
  geom_histogram(data = dplyr::select(plot_table, -group), fill = "grey", alpha = .5) +
  geom_histogram(colour = "black") +
  facet_wrap(~ group, labeller = group_types) +
  guides(fill = FALSE) +  # to remove the legend
  scale_fill_manual(values = group_colors) +
  theme_classic()      
```
Let's make a pretty table now
```{r}
library(kableExtra)

fun_change_table %>% 
  dplyr::select(-group) %>%
  count(rich_change, turnover, fun_change) %>%
  pivot_wider(names_from = rich_change, values_from = n) %>%
  kable() %>%
  collapse_rows(columns = 1, valign = "middle")

```

### 2023-06-05 06_study_change.Rmd
Get the table working for knit at least!

@@
 title: "Study Level trends"
-journal: "Ecology Letters"
-date: "`r Sys.Date()`"
-csl: pnas.csl
-layout: 3p
+output: pdf_document
 header-includes:
-   - \usepackage{lineno}
-   - \linenumbers
-output: 
-  rticles::arxiv_article:
-    keep_tex: true
-#  word_document: default
-editor_options: 
-  markdown: 
-    wrap: sentence
+  - \usepackage{booktabs}
+date: "2023-06-05"
 ---
@@
 group_types <- as_labeller(
-  c("A" = "No species change or turnover, functional change", "B" = "Turnover, no richness or functional change", "C" = "No change", "D" = "Turnover, species and functional Change")
+  c("A" = "No richness change or turnover, functional change", "B" = "No turnover, richness and functional change", "C" = "No change", "D" = "Turnover, richness and functional Change")
 )
@@
 Let's make a pretty table now
-```{r}
+```{r, echo=FALSE}
 library(kableExtra)
@@
   pivot_wider(names_from = rich_change, values_from = n) %>%
-  kable() %>%
-  collapse_rows(columns = 1, valign = "middle")
-
+  select(fun_change, everything()) %>%
+  arrange(fun_change) %>%
+  mutate(`No Richness Change` = cell_spec(`No Richness Change`,
+                                           color = case_when(
+                                             `No Richness Change` == 12 ~ "#d9be3b",
+                                             `No Richness Change` == 11 ~ "#086788",
+                                             .default = "black"))) %>%
+  mutate(`Richness Change` = cell_spec(`Richness Change`,
+                                           color = case_when(
+                                             `Richness Change` == 6 ~ "#77976e",
+                                             `Richness Change` == 7 ~ "#cc8214",
+                                             .default = "black"))
+         ) %>%
+  kable(format = "latex", col.names = c("", "", "No Richness Change", "Richness Change"), 
+        booktabs = TRUE, align = "c", escape = FALSE) %>%
+  kable_styling("striped") %>%
+  row_spec(0, bold = TRUE) %>%
+  column_spec(1:2, bold = TRUE) %>%
+  collapse_rows(columns = 1, valign = "middle") %>%
+  kable_as_image("figures/change_table", file_format = "jpeg", latex_header_includes = c("\\\\usepackage{booktabs}"))
 ```

### 2023-06-05 06_study_change.Rmd
Give up on saving table as image

@@
 
-```{r}
+```{r, echo = FALSE, message = FALSE, fig.show = 'hold', fig.cap= "Break down of studies by the type of change they exhibited."}
 plot_table <- duration_table %>%
@@
   facet_wrap(~ group, labeller = group_types) +
+  ylab("count of time series") +
   guides(fill = FALSE) +  # to remove the legend
@@
   theme_classic()
-```
-Let's make a pretty table now
-```{r, echo=FALSE}
+
 library(kableExtra)
@@
   column_spec(1:2, bold = TRUE) %>%
-  collapse_rows(columns = 1, valign = "middle") %>%
-  kable_as_image("figures/change_table", file_format = "jpeg", latex_header_includes = c("\\\\usepackage{booktabs}"))
+  collapse_rows(columns = 1, valign = "middle") #%>%
+  #kable_as_image("figures/change_table", file_format = "jpeg", latex_header_includes = c("\\\\usepackage{booktabs}"))
 ```

### 2023-06-06 paper/manuscript_GEB.Rmd
Changing intro

@@
   - name: "Kari E. A. Norman"
-#   affiliation: Université de Montréal
-#   location: Montréal, Quebec
-#   email: kari.norman@berkeley.edu
-# - name: "Perry de Valpine"
-#    afiliation: University of California, Berkeley
-#   location: Berkeley, California
-#    eail: pdevalpine@berkeley.edu
-# - name: "Carl Boettiger"
-#   affiliation: University of California, Berkeley
-#   location: Berkeley, California
-#   email: cboettig@berkeley.edu
+    affiliation: Université de Montréal
+    location: Montréal, Quebec
+    email: kari.norman@berkeley.edu
+  - name: "Perry de Valpine"
+    affiliation: University of California, Berkeley
+    location: Berkeley, California
+    email: pdevalpine@berkeley.edu
+  - name: "Carl Boettiger"
+    affiliation: University of California, Berkeley
+    location: Berkeley, California
+    email: cboettig@berkeley.edu
 abstract: |
@@
 
-Debate over the meaning of general trends derived from limited data and their relevance for conservation is ongoing.
-Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, a lack of baselines, and for not taking into account disturbance histories of individual studies [@cardinale2014; @cardinale2018; @gonzalez2016].
-However, many of these criticisms are unavoidable, reflecting general limitations of ecological data [@primack2018].
-The pattern of increasing turnover accompanied by constant or increasing richness has also been reproduced in multiple datasets and taxa, giving strong evidence that it represents a true ecological pattern [@barnagaud2017; @dunic2017; @elahi2015].
+The meaning of general trends derived from limited data, and their relevance for conservation, is a topic of on going debate.
+Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and for not taking into account the ecological context of individual studies [@cardinale2014; @cardinale2018; @gonzalez2016].
+Many of these criticisms reflect limitations of ecological data on the whole, and proponents argue we must make the best of the imperfect data available [@primack2018].
+Still, there is general consensus on a few local-scale patterns that seem to be characteristic of the anthropocene, including the prevalence of species turnover across communities and patterns of both richness increase and decrease in different taxa and biomes [@dornelas2023].
+Understanding the implications of change in species assemblage for ecosystem stability and function is a critical next step in moving forward the biodiversity change conversation.
 
@@
 We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study.
+We fit 18 individual *CWM* models.
 All time series with data for a given trait were included in the model.
-We fit 18 individual *CWM* models.
-
-For all other species and functional diversity metrics (*S*, *J*, *FRic*, *FEve*, *FDiv)* we obtained study-level estimates of temporal change from the Best Linear Unbiased Predictors (BLUPs) for each overall trend model.
-BLUP's provide estimates for the conditional mean and variance of a random effect from which we calculated 95% confidence intervals to determine significance of study-level slopes.
+We estimated study level trends using individual linear models, for studies with more than one times series we fit a random slope and intercept for time series.
+Some study-level models could not be fit for five studies for at least one metric due to data limitations, but those studies were still included in the general models.
+They represented 12 of 1350 study-level models fit for each metric.
+For further details see the supplement.
+Where appropriate, response variables were $log$ or $log(x+1)$ transformed to better fit model assumptions of residual normality.
 
@@
 We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [@lenth2021].
-We assessed the impact of time series duration and start year on study-level trends using general linear models with duration and start year as predictors.
-All models were executed using the *lme4* package in R and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2021].
+We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
+All models in our analysis were fit using the *lme4* package in R and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2021].
 
@@
 However, neither trend was exhibited for the standardized metric, indicating that differences were largely due to changes in species richness.
-Further, with only two global studies these results are not general trends.
-The general trends for *CWM* models were similarly not significant, with a significant positive trend for only percentage of aerial foraging utilization.
+Further, with only two global studies these results are not true general trends.
+The general trends for *CWM* models were similarly not significant, with a significant positive trend for only percentage of fish in diet composition.
 
@@
   filter(group != "Residual" | is.na(group)) %>%
-  mutate(metric = ifelse(metric == "Jaccard_base", "Jaccard", metric),
-         metric = ifelse(metric == "S", "Species Richness", metric),
+  mutate(metric = ifelse(metric == "Jaccard_base", "log(Jaccard + 1)", metric),
+         metric = ifelse(metric == "S", "log(Species Richness)", metric),
          effect = ifelse(effect == "ran_pars", "random", effect),

### 2023-06-06 paper/manuscript_GEB.Rmd
Results rewrite study trends

@@
 Non-significant overall trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies was plausibly 0 (Table \ref{tab:resultsTab}).
-Trends for different taxa, biomes, or realms were also non-significant for richness and summary functional diversity metrics, with the exception of a significantly increasing trend for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly increasing functional richness slope for tropical studies.
-However, neither trend was exhibited for the standardized metric, indicating that differences were largely due to changes in species richness.
-Further, with only two global studies these results are not true general trends.
+Trends for different taxa, biomes, or realms were also non-significant for richness and summary functional diversity metrics, with the exception of a significantly increasing trend for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing functional richness slope for temperate/tropical studies and mammal studies.
+However, trends were not exhibited for the standardized metric indicating that differences were largely due to changes in species richness.
+Further, with only two global studies, the trend should not be considered truely general.
 The general trends for *CWM* models were similarly not significant, with a significant positive trend for only percentage of fish in diet composition.
@@
 
-At the study level, 4 studies experienced a significant trend in species richness and only 9 of 50 studies for observed metrics and 3 of 50 studies for standardized metrics experienced a significant trend for a metric other than Jaccard similarity (Table \ref{tab:trendTab}).
-Most significant trends for observed functional metrics are in functional richness and disappeared after standardization, indicating that richness increases were likely due to changes in the number of species.
-However, the majority of the functional richness trends were negative, indicating that when change is occurring it is more often a loss.
+We did find significant differences between taxa, realms, and climate for Jaccard similarity and some of the *CWM's*.
+For example while Jaccard similarity was decreasing in the general trend and there were significant within group slopes for Mammals, Birds, and Terrestrial communities, there was no significant slope for the marine realm, indicating that the general trend is mostly driven by turnover in terrestrial communities.
+For bird communities, we also found within group trends and between group differences for trends in foraging behavior.
+We found a significant increasing trend in utilization of the canopy in Tropical communities that was significantly different from the trends for Polar/Temperate and Temperate communities.
+There was a significant decrease in utilization of the understory for Terrestrial communities and significant increase in foraging below the water surface for global studies (but see the previous limitations of Global data).
+There was also a significant positive slope for *CWM* body mass for the single Temperate/Tropical study (with three time series) which were marine mammal communities.
+
+We found significant dietary shifts across communities, with a significant increase in fruit consumption in Terrestrial communities and a significant decrease in nectar consumption in Tropical communities, a trend significantly different than that for Terrestrial communities.
+There was a significant increase in seed consumption in bird species, which was significantly different from the trend for Mammal communities.
+There was a significant increase in fish consumption for the two Global studies.
+Vertebrate consumption significantly decreased for Marine studies and for studies of Global and Tropical communities.
+
+At the study level, 15 of 50 studies exhibited a significant trend in species richness and 11 exhibited significant turnover.
+For observed functional metrics, 25 of 50 studies exhibited a trend in a least one metric, and 15 of 50 studies exhibited a significant trend for at least one standardized metrics (Table \ref{tab:trendTab}).
+In general, there were more signifcant trends for uncorrected metrics, with some disappearing after correction, indicating that those trends were likely due to changes in the number of species.
 Hypothesis testing for study-level trends is likely affected by multiple testing issues and some trends identified as significant are therefore potentially spurious.
@@
 
-data.frame(sign = c("+", "-"), S = c("1", "3"), `Jaccard Similarity` = c("0", "31"), FRic = c("2", "5"), FEve = c("1", "0"), FDiv = c("0", "1"), `SES FRic` = c("0", "1"), `SES FEve` = c("0", "2"), `SES FDiv` = c("0", "1")) %>%
-  column_to_rownames("sign") %>%
-  kable(caption = "Number of studies that experienced a significant trend in each calculated metric out of 50 total studies.")
+load(here::here("data/indv_mod_table.rda"))
+
+indv_mod_table %>%
+  filter(term == "year_scaled", p.value < 0.05) %>%
+  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S", "FRic", "FEve", "FDiv")) %>%
+  mutate(positive = ifelse(estimate > 0, "+", "-")) %>%
+  group_by(metric, positive) %>%
+  summarise(n_study_id = n_distinct(study_id)) %>%
+  pivot_wider(names_from = "metric", values_from = "n_study_id", values_fill = 0) %>%
+  select(positive, S, Jaccard_base, FRic, FEve, FDiv, SES_FRic, SES_FEve, SES_FDiv) %>%
+  kable(format = "latex", 
+        col.names = c("", "S", "Jaccard Similarity", "FRic", "FEve", "FDiv", "SES FRic", "SES FEve", "SES FDiv"),
+        caption = "Number of studies that experienced a significant trend in each calculated metric out of 50 total studies.", 
+        booktabs = TRUE, align = "c", escape = FALSE) %>%
+  kable_styling("striped")
+
 ```
@@
 See supplement for estimates and p-values for all models.
-
-We did find significant differences between taxa, realms, and climate for Jaccard similarity and some of the *CWM's*.
-For example while Jaccard similarity was decreasing in the general trend and there were significant within group slopes for Mammals, Birds, and Terrestrial communities, there was no significant slope for the marine realm, indicating that the general trend is mostly driven by turnover in terrestrial communities.
-For bird communities, we also found within group trends and between group differences for trends in foraging behavior.
-We found a significant increasing trend in utilization of the canopy in Tropical communities that was significantly different from the trends for Polar/Temperate and Temperate communities.
-There was a significant decrease in utilization of the understory for Terrestrial communities and significant increase in foraging below the water surface for global studies (but see the previous limitations of Global data).
-There was also a significant positive slope for *CWM* body mass for the single Temperate/Tropical study (with three time series) which were marine mammal communities.
-
-We found significant dietary shifts across communities.
-We found a significant increase in fruit consumption in Terrestrial communities and a significant decrease in nectar consumption in Tropical communities, a trend significantly different than that for Terrestrial communities.
-There was a significant increase in seed consumption in bird species, which was significantly different from the trend for Mammal communities.
-There was a significant increase in fish consumption for the two Global studies.
-Vertebrate consumption significantly decreased for Marine communities and for Global and Tropical communities.
 

### 2023-06-06 analysis/figures/06_study_change.Rmd
Save histogram figure

--- 2023-06-05 06_study_change.Rmd
@@
 ```
+
 ```{r}
@@
 
-```{r, echo = FALSE, message = FALSE, fig.show = 'hold', fig.cap= "Break down of studies by the type of change they exhibited."}
+```{r}
 plot_table <- duration_table %>%
@@
 
+ggsave(here::here("figures/duration_hist.jpeg"), bg = "white")
+```
+
+```{r, echo = FALSE, message = FALSE, fig.show = 'hold', fig.cap= "Break down of studies by the type of change they exhibited."}
 library(kableExtra)
+
+knitr::include_graphics(here::here("figures/duration_hist.jpeg"))
 

### 2023-06-07 analysis/figures/06_study_change.Rmd
Minor figure changes, substantial rewrite

@@
   count(rich_change, turnover, fun_change, group)
+
+#usethis::use_data(fun_change_table)
 ```
@@
 group_types <- as_labeller(
-  c("A" = "No richness change or turnover, functional change", "B" = "No turnover, richness and functional change", "C" = "No change", "D" = "Turnover, richness and functional Change")
+  c("A" = "No richness change or turnover, functional change", "B" = "No turnover, richness and functional change", "C" = "No change", "D" = "Turnover, richness and functional change")
 )
@@
   ylab("count of time series") +
+  xlab("time series duration") +
   guides(fill = FALSE) +  # to remove the legend
@@
   kable_styling("striped") %>%
-  row_spec(0, bold = TRUE) %>%
-  column_spec(1:2, bold = TRUE) %>%
+  # row_spec(0, bold = TRUE) %>%
+  # column_spec(1:2, bold = TRUE) %>%
   collapse_rows(columns = 1, valign = "middle") #%>%
-  #kable_as_image("figures/change_table", file_format = "jpeg", latex_header_includes = c("\\\\usepackage{booktabs}"))
+  #save_kable(here::here("figures/change_table.jpeg"), latex_header_includes = c("\\\\usepackage{booktabs}"), keep_tex = TRUE)
+
 ```

### 2023-06-07 paper/manuscript_GEB.Rmd
Minor figure changes, substantial rewrite

@@
 As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics may be consistent with similar well-documented species derived trends.
-We found no evidence of systematic functional richness loss or functional change.
-A lack of trend for almost all realms, biomes, and taxonomic groups gives further evidence that changes in functional structure are not occurring in the systems observed in our dataset.
+We found no trend in functional change for almost all realms, biomes, and taxonomic groups, with evidence of functional richness loss for mammal studies.
+These results are consistent with... particularly as the trend was also significantly more negative the the trend for birds.
+
+Despite a lack of general trends in summary functional metrics, we did find multiple trends in *CWM*'s for studies coming from the same taxa, realms, or climates, many of which are consistent with previous findings for those contexts.
+For example, the decrease in insect consumption for tropical communities reflects well documented declines in tropical insectivorous birds [@sherry2021].
+Reduction in utilization of understory foraging for terrestrial communities could be the result of disturbance like human recreation or increased predation from introduced species.
+Other shifts in diet for both Birds and Mammals, including increasing fruit, seed, and vertebrate consumption and decreasing seed consumption in some climates point to important areas for further exploration.
+
+At the study level, biodiversity change fell into four dominant scenarios of change based on trends in richness, turnover, and functional metrics: no change, only functional change, richness and functional change with turnover, and richness and functional change without turnover (Table \ref{tab:changeScenarios}).
+Around a quarter of studies experienced no trend in richness, turnover, or functional metrics.
+These studies spanned the distribution of study durations excluding only the very longest running studies, with the longest no change time series lasting 23 years.
+
+Another quarter of the studies showed no trend in richness or turnover, but did show a significant shift in at least one summary functional metric.
+These communities are best aligned with scenarios of species replacement where only a few functional outliers replaced functionally indistinct species or vice versa (there were studies with positive and negative trends in all three functional metrics).
+Studies in the scenario of only functional change were heavily skewed toward shorter running time series, indicating that they may represent a limited snapshot of communities that on longer time scales would be exhibiting significant change in species based metrics, particularly as species replacement continues adding to turnover.
+Communities in this scenario, and the previous scenario of no functional change are consistent with hypotheses that species richness may be strongly regulated through time [@brown2001; @gotelli2017], and emphasize that maintenance of functional structure can be divorced from those processes.
+
+About 10 percent of studies, but the majority of time series in our dataset, exhibited richness and functional change but no turnover.
+These studies also spanned the the distribution of time series durations, excluding the longest running studies.
+With the exception of one study, trends in functional change for this group were only found for uncorrected metrics and disappeared after correction, indicating that functional change in this group is almost exclusively the result of species gains or losses rather than turnover of species with different functional traits.
+
+About 15 percent of studies, which were also exclusively the longest running studies, had a trend in turnover, richness, and functional change.
+The studies in this scenario fell into two groups: a positive trend in richness with a positive trend in functional richness (corrected and uncorrected), and a negative trend in species richness with a negative trend in uncorrected functional richness and no trend in corrected richness but a trend in other corrected functional metrics.
+This would indicate that when species are lost from a community, they are not more functionally unique than expected by chance, though they may produce shifts in other aspects of functional structure.
+When species are gained however, they disproportionately add to the functional richness of the community.
+
+Our results illuminate the relationship between species-based and functional metrics observed in real communities.
+Critically, particularly for the short observation time windows that characterize the majority of ecological data, functional shifts may be underlying seemingly static species-based community metrics.
+Species-based approaches may therefore be missing areas where significant change in the functional space could be impacting ecosystem function and resilience.
+In scenarios where there are significant species gains or losses over time, we found evidence that species losses may be relatively functionally indistinct, while species gains contribute disproportionately to the richness of a functional space.
+This finding is consistent with evidence that at least in brid communities common, functionally general species are being lost even as rare species are increasing [@inger2015; @rosenberg2019; @schipper2016].
+
+Our results are consistent with ecological expectations that community dissimilarity would be greater the larger the observation time window purely due to background processes, which is likely further accelerated by global change drivers [@dornelas2023].
+While we do not attempt to address how observed rates differ from background expectation here, we can make some assertions about the implication of turnover for functional change.
+Where turnover is occurring, it is almost always accompanied by a directional change in the functional space.
+Species are not therefore being replaced by exact functional counter parts.
+The fact that we observed both increases and decreases in corrected functional richness accompanying turnover is a reminder that replacements are not always an indication of functional loss.
+We also can't tease apart whether these are directional shifts due to changing environmental pressures, or simply shifts due to the available set of traits in the species pool.
 
@@
 We therefore have preliminary evidence that maintenance of functional structure can occur even in the face of species richness gains or losses, but additional data sources will be necessary to further explore how prevalent those scenarios are.
+
+-   there are a few clear types of change, and probability of falling in those groups does seem related to length of the time series
+
+```{r changeScenarios, echo = FALSE, message = FALSE, fig.show = 'hold', fig.cap= "Break down of studies by the type of change they exhibited."}
+
+load(here::here("data/fun_change_table.rda"))
+
+knitr::include_graphics(here::here("figures/duration_hist.jpeg"))
+
+fun_change_table %>% 
+  dplyr::select(-group) %>%
+  count(rich_change, turnover, fun_change) %>%
+  pivot_wider(names_from = rich_change, values_from = n) %>%
+  select(fun_change, everything()) %>%
+  arrange(fun_change) %>%
+  mutate(`No Richness Change` = cell_spec(`No Richness Change`,
+                                           color = case_when(
+                                             `No Richness Change` == 12 ~ "#d9be3b",
+                                             `No Richness Change` == 11 ~ "#086788",
+                                             .default = "black"))) %>%
+  mutate(`Richness Change` = cell_spec(`Richness Change`,
+                                           color = case_when(
+                                             `Richness Change` == 6 ~ "#77976e",
+                                             `Richness Change` == 7 ~ "#cc8214",
+                                             .default = "black"))
+         ) %>%
+  kable(format = "latex", col.names = c("", "", "No Richness Change", "Richness Change"), 
+        booktabs = TRUE, align = "c", escape = FALSE) %>%
+  kable_styling("striped") %>%
+  #row_spec(0, bold = TRUE) %>%
+  #column_spec(1:2, bold = TRUE) %>%
+  collapse_rows(columns = 1, valign = "middle")
+```
 
@@
 Future work directly assessing changes in the species pool and attributing those shifts to specific mechanisms will be critical for understanding how functional structure is maintained and what may interrupt that maintenance.
-
-Despite maintenance of summary functional metrics we did find multiple trends in *CWM*'s for studies coming from the same taxa, realms, or climates, many of which are consistent with previous findings for those contexts.
-For example, the decrease in insect consumption for tropical communities reflects well documented declines in tropical insectivorous birds [@sherry2021].
-Reduction in utilization of understory foraging for terrestrial communities could be the result of disturbance like human recreation or increased predation from introduced species.
-Other shifts in diet for both Birds and Mammals, including increasing fruit, seed, and vertebrate consumption and decreasing seed consumption in some climates point to important areas for further exploration.
-Generally, these findings indicate that even when some shifts in the functional space are occurring, often as a direct result of global change impacts, summary metrics describing functional structure can be maintained.
 
@@
 
-Our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
+While we found on over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
+In fact, study-level trends indicate quite the opposite, that functional shifts with unknown implications for ecosystem characteristics may be going undetected by common species-based approaches.
 These findings compliment a substantial body of work linking functional degradation to species losses as a result of direct human intervention in the form of land use change and intensification or habitat fragmentation [@flynn2009; @magioli2021; @tinoco2018].
@@
 ## Future Work
+
+While the trends we see here are statistically significant, we still do not know if they are ecologically significant.
+There is substantial work to be done to link degrees of change in functional structure to measurable outcomes in ecosystem processes.
 

### 2023-06-12 paper/manuscript_GEB.Rmd
reviiiisions

@@
 There is increasing evidence functional diversity is a better predictor of ecosystem function than species-based metrics [@cadotte2011; @gagic2015], and that different facets of functional diversity play essential roles in maintaining ecosystem stability [@craven2018; @morin2014].
-Indeed, almost all hypothesized mechanisms underpinning the relationship between species diversity and ecosystem function are trait-dependent [@hillebrand2009].
+Almost all hypothesized mechanisms underpinning the relationship between species diversity and ecosystem function are trait-dependent [@hillebrand2009].
 Determining functional trends therefore gives a more direct picture of potential trends in critical ecosystem processes.
@@
 For some time periods, functional structure appears to be maintained for substantial portions of geological time [@hedberg2022].
-Some losses of functional diversity are indisputable on both paleocological and contemporary timescales such as continued trophic downgrading due to loss of large-bodied mammals, but implications of those losses for local diversity patterns are less clear [@estes2011; @smith2018] .
+Still, some losses of functional diversity are indisputable on both paleocological and contemporary timescales such as continued trophic downgrading due to loss of large-bodied mammals, but implications of those losses for local diversity patterns are less clear [@estes2011; @smith2018] .
 
@@
 We assess thousands of mammal and bird functional diversity time series to determine whether or not there is a general trend of functional change, both in observed metrics and in metrics corrected for changes in species richness.
-We distinguish between three possible scenarios of functional change: 1) significant loss of functional diversity with accompanying shifts in other functional metrics, 2) no functional diversity loss, but significant shifts in other functional metrics, 3) maintenance of functional diversity and composition.
-Based on expectation due to human impacts, we expect to find a significant functional loss with further restructuring indicated by the additional metrics.
+We further assess the prevailing modes of change occurring at the study level to determine how functional change is related to species patterns, and which kinds of species and functional change are most common across communities.
+We distinguish between three possible scenarios of functional change: 1) significant functional change accompanied by changes in species-based metrics, 2) maintenance of functional structure despite significant species changes, 3) maintenance of functional and species metrics.
+Based on expectation due to human impacts, we expect to find significant functional changes, often accompanied by species change.
 
@@
 To test for trends within and between different levels of taxa, biome, and realm we fit separate models with each of those factors added as a predictor to the original model structure.
-We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [@lenth2021].
+We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.8.2, @lenth2022].
 We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
-All models in our analysis were fit using the *lme4* package in R and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2021].
+All models in our analysis were fit using the *lme4* (v1.1-30) package in R (v4.2.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].
 
@@
 For observed functional metrics, 25 of 50 studies exhibited a trend in a least one metric, and 15 of 50 studies exhibited a significant trend for at least one standardized metrics (Table \ref{tab:trendTab}).
-In general, there were more signifcant trends for uncorrected metrics, with some disappearing after correction, indicating that those trends were likely due to changes in the number of species.
+In general, there were more significant trends for uncorrected metrics, with some disappearing after correction, indicating that those trends were likely due to changes in the number of species.
 Hypothesis testing for study-level trends is likely affected by multiple testing issues and some trends identified as significant are therefore potentially spurious.
@@
 
-Study-level slopes for multiple metrics were significantly related to the duration and start year of studies.
-Slopes for species richness were significantly more negative with later start date and more positive for longer duration studies.
-Jaccard similarity and functional evenness both had significantly more negative slopes with more recent start year, whereas functional divergence was significantly more positive.
-Slopes for functional evenness were also significantly more positive for longer duration studies.
-See supplement for estimates and p-values for all models.
+Study-level slopes for summary functional metrics were significantly related start year of the time series for Jaccard similarity and functional evenness, both of which had significantly more negative slopes with more recent start year.
+No summary functional metrics were significantly related to the start year of the time series.
 
@@
 As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics may be consistent with similar well-documented species derived trends.
-We found no trend in functional change for almost all realms, biomes, and taxonomic groups, with evidence of functional richness loss for mammal studies.
-These results are consistent with... particularly as the trend was also significantly more negative the the trend for birds.
+We found no trend in functional change for almost all realms, biomes, and taxonomic groups, with evidence of functional richness loss in mammal studies only.
+These results are consistent with multiple studies linking anthropogenic drivers to loss of functional diversity in mammal communities [@brodie2021; @cox2022; @li2022]
 
@@
 These communities are best aligned with scenarios of species replacement where only a few functional outliers replaced functionally indistinct species or vice versa (there were studies with positive and negative trends in all three functional metrics).
-Studies in the scenario of only functional change were heavily skewed toward shorter running time series, indicating that they may represent a limited snapshot of communities that on longer time scales would be exhibiting significant change in species based metrics, particularly as species replacement continues adding to turnover.
+Studies in the scenario of only functional change were heavily skewed toward shorter running time series, indicating that they may represent a limited snapshot of communities that on longer time scales would be exhibiting significant change in species based metrics, particularly as species replacement continued adding to turnover.
 Communities in this scenario, and the previous scenario of no functional change are consistent with hypotheses that species richness may be strongly regulated through time [@brown2001; @gotelli2017], and emphasize that maintenance of functional structure can be divorced from those processes.
@@
 These studies also spanned the the distribution of time series durations, excluding the longest running studies.
-With the exception of one study, trends in functional change for this group were only found for uncorrected metrics and disappeared after correction, indicating that functional change in this group is almost exclusively the result of species gains or losses rather than turnover of species with different functional traits.
+With the exception of one study, trends in functional change for this group were only found for uncorrected metrics and disappeared after correcting for species richness, indicating that functional change in this group is almost exclusively the result of species gains or losses rather than turnover of species with different functional traits.
 
@@
 
-Our results illuminate the relationship between species-based and functional metrics observed in real communities.
+Our results illustrate the range of relationships between species-based and functional metrics observed in real communities.
 Critically, particularly for the short observation time windows that characterize the majority of ecological data, functional shifts may be underlying seemingly static species-based community metrics.
@@
 In scenarios where there are significant species gains or losses over time, we found evidence that species losses may be relatively functionally indistinct, while species gains contribute disproportionately to the richness of a functional space.
-This finding is consistent with evidence that at least in brid communities common, functionally general species are being lost even as rare species are increasing [@inger2015; @rosenberg2019; @schipper2016].
+This finding is consistent with evidence that at least in bird communities, common functionally general species are being lost even as rare species are increasing [@inger2015; @rosenberg2019; @schipper2016].
 
 Our results are consistent with ecological expectations that community dissimilarity would be greater the larger the observation time window purely due to background processes, which is likely further accelerated by global change drivers [@dornelas2023].
-While we do not attempt to address how observed rates differ from background expectation here, we can make some assertions about the implication of turnover for functional change.
+While we do not address how observed rates differ from background expectation here, we can make some assertions about the implication of turnover for functional change.
 Where turnover is occurring, it is almost always accompanied by a directional change in the functional space.
 Species are not therefore being replaced by exact functional counter parts.
-The fact that we observed both increases and decreases in corrected functional richness accompanying turnover is a reminder that replacements are not always an indication of functional loss.
-We also can't tease apart whether these are directional shifts due to changing environmental pressures, or simply shifts due to the available set of traits in the species pool.
+The fact that we observed both increases and decreases in corrected functional richness accompanying turnover is a reminder that replacements are not always an indication of functional loss, but can also lead to increased functional richness.
+While shifts in functional space may be the result of environmental pressure, we cannot distinguish between directional change and shifts due to the unavailability of an analog species in the species pool.
 
-Very few of the studies in our data set experienced a significant trend in species richness for the observation period, our data therefore mostly represent communities undergoing significant change in species composition but not species counts.
-Results should therefore not be extrapolated to communities experiencing significant species losses or gains as in scenarios of recovery from disturbance or catastrophic human impact.
-While we cannot make generalizations about scenarios with richness losses or gains from out data, for the four studies that did experience significant species richness changes, only one exhibited any trend in a summary functional metric with a significant decrease in FRic but not SES FRic, indicating it was mostly functional loss due to species loss.
-We therefore have preliminary evidence that maintenance of functional structure can occur even in the face of species richness gains or losses, but additional data sources will be necessary to further explore how prevalent those scenarios are.
-
--   there are a few clear types of change, and probability of falling in those groups does seem related to length of the time series
-
-```{r changeScenarios, echo = FALSE, message = FALSE, fig.show = 'hold', fig.cap= "Break down of studies by the type of change they exhibited."}
-
-load(here::here("data/fun_change_table.rda"))
+```{r changeScenarios, echo = FALSE, message = FALSE, fig.show = 'hold', fig.cap= "Comparision of the distribution of time series durations from each type of change to the overall distribution of time series durations, with no richness change or turnover and functional change in yellow, no turnover but richness and functional change in green, no change in blue, and turnover, richness and functional change studies in orange."}
 
@@
 
-fun_change_table %>% 
-  dplyr::select(-group) %>%
-  count(rich_change, turnover, fun_change) %>%
-  pivot_wider(names_from = rich_change, values_from = n) %>%
-  select(fun_change, everything()) %>%
-  arrange(fun_change) %>%
-  mutate(`No Richness Change` = cell_spec(`No Richness Change`,
-                                           color = case_when(
-                                             `No Richness Change` == 12 ~ "#d9be3b",
-                                             `No Richness Change` == 11 ~ "#086788",
-                                             .default = "black"))) %>%
-  mutate(`Richness Change` = cell_spec(`Richness Change`,
-                                           color = case_when(
-                                             `Richness Change` == 6 ~ "#77976e",
-                                             `Richness Change` == 7 ~ "#cc8214",
-                                             .default = "black"))
-         ) %>%
-  kable(format = "latex", col.names = c("", "", "No Richness Change", "Richness Change"), 
-        booktabs = TRUE, align = "c", escape = FALSE) %>%
-  kable_styling("striped") %>%
-  #row_spec(0, bold = TRUE) %>%
-  #column_spec(1:2, bold = TRUE) %>%
-  collapse_rows(columns = 1, valign = "middle")
+knitr::include_graphics(here::here("figures/functional_change_table.png"))
 ```
 
-Communities demonstrated almost universal maintenance of functional structure.
-While the majority of the studies (\~70%) included in our data experienced significant species turnover, only three (for standardized metrics) experienced a significant shift in any summary functional metric.
-This suggests certain characteristics of the functional space are maintained even in the face of significant change in species identity, specifically the size of the functional space occupied by the community (*FRic*), the distribution of species and individuals within that space (*FEve* and *FDiv*), and the mean of individual trait distributions (*CWM*).
-On average, species additions have similar functional characteristics as lost species and therefore maintain the structure of the functional space.
-
-While observed communities are clearly losing species (as demonstrated by the turnover in species identity), sufficient diversity is being maintained in the species pool to provide functionally comparable species as replacement.
-This phenomena could be the result of two possible process potentially working in concert.
-First, functional redundancy in the species pool maintains resilience of the functional space by providing multiple potential species for any given set of functional traits.
-Alternatively, while there may be low redundancy in the functional space, functional composition in the species pool may be replenished by changes in the composition of the species pool.
-Addition of new species to the species pool could occur under multiple mechanisms of global change, including the introduction of non-native species, shifts in species ranges due to climate change, and spatial homogenization due to replacement of specialist species by generalist ones.
-Future work directly assessing changes in the species pool and attributing those shifts to specific mechanisms will be critical for understanding how functional structure is maintained and what may interrupt that maintenance.
-
-These results challenge implicit assumptions that functional loss is the default state of all or even many communities [@carmona2021; @dirzo2014; @young2016].
-And while there is some evidence in our data that functional loss may be more common than functional gains for communities experiencing functional change, there was no significant trend for even the longest running and most heavily impacted studies.
-The North American Breeding Survey for example is considered an authoritative dataset on the state of bird populations on the continent and underpins policy decisions about bird conservation [@pardieck2020; @rosenberg2019; @sauer2017].
-No more robust dataset exists to capture North American avian community change, yet we detected no general shifts in functional structure across the dataset.
-Further, none of the 5 included studies that experienced a manual manipulation (e.g. burning, grazing exclosure, etc) experienced any significant functional trends.
-
-Results are also in contrast to some predictions for trait shifts under global change.
+Still, some results stand in contrast to predictions for trait shifts under global change.
 For example, mean body size is predicted to decrease as a result of climate change impacts and megafaunal loss [@sheridan2011], a phenomena which has already been well documented empirically and experimentally in multiple taxa [@caruso2014; @forster2012; @huss2019; @tseng2018].
@@
 
-While we did not directly measure changes in rare species, our results further contradict hypothesized patterns due to rare species extinction.
-Rare species, defined by small populations and geographic restriction, are simultaneously more likely to be functionally distinct and at higher risk for extinction [@davies2004; @fritz2009; @harnik2012; @loiseau2020].
-Locally, communities losing functionally rare species should exhibit strong functional shifts as lost species can eventually no longer be replaced by functionally similar species [@leitão2016].
-Observed patterns were instead consistent with species replacement by functionally redundant species from the species pool.
-Still, for many time series we likely did not have a large enough time window to capture community and species pool impoverishment due to extinction.
-
 What does local maintenance of functional structure mean for ecosystem function?
 The vast majority of experimental and observational work links declines in function to declines in functional or species diversity [@brose2016; @cadotte2011; @duffy2007].
-By those criteria very few communities in our dataset are in a state of concern for loss of functionality.
-However, shifts in metrics are only relevant if the underlying traits are those most critical for ecosystem function.
+By those criteria very few communities in our dataset are in a state of concern for loss of functionality (Table \ref{tab:trendTab}), though we did find evidence that mammal communities may be in greater danger than bird communities.
+Still, shifts in metrics are only relevant if the underlying traits are those most critical for ecosystem function.
 We were limited in this analysis to the traits available rather than those with strong empirical links to function.
@@
 dispersion, rarity, abundance of dominant species, @bagousse-pinguet2021].
-Still, our results indicate that highs species turnover alone is not sufficient cause for concern for ecosystem function, since functional structure is largely conserved.
-However, that may not be true indefinitely if functionally redundant species are depleted from the species pool.
+Critically, while trends for functional metrics may be statistically significant, the effect size necessary to be ecologically significant for key processes remains an outstanding question of the field.
 
@@
 Here we approach the question of functional change using the best available data and biodiversity synthesis approaches.
-However, some gaps in best practices may be obscuring a true underlying trend.
+While we found signatures of change at the study level, some gaps in best practices may be obscuring a true general trend at the aggregate level.
 First, the BioTIME database, while the most comprehensive data source of time series available, is limited in temporal and geographic scope.
@@
 
-<!-- Third, while we use here the most common metrics for describing functional diversity, they do not measure some potentially important aspects of the functional space.
+Third, while we use here the most common metrics for describing functional diversity, they do not measure some potentially important aspects of the functional space.
 Most notably, the summary metrics we calculated do not capture shifts in the location of the functional space as a whole.
@@
 While trait CWM's capture axis shifts, approaches for assessing multidimensional shifts in functional space are still relatively new [@barros2016; @blonder2018; @mammola2019] but could shed critical insight into functional composition changes of this nature.
--->
 
@@
 
-While we found on over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
-In fact, study-level trends indicate quite the opposite, that functional shifts with unknown implications for ecosystem characteristics may be going undetected by common species-based approaches.
-These findings compliment a substantial body of work linking functional degradation to species losses as a result of direct human intervention in the form of land use change and intensification or habitat fragmentation [@flynn2009; @magioli2021; @tinoco2018].
-Rather than assuming functional structure will be maintained in areas of concern, our work indicates that when measurements of functional diversity show significant shifts, it should be considered an indication of substantial community change and outside the normal expectation.
+While we found no over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
+In fact, study-level trends indicate quite the opposite, that functional shifts with unknown implications for ecosystem processes may be going undetected by common species-based approaches, particularly for short observation time windows.
+While the majority of studies in our dataset did not experience a significant functional loss, a substantial body of work links functional degradation to species losses as a result of direct human intervention in the form of land use change and intensification or habitat fragmentation [@flynn2009; @magioli2021; @tinoco2018].
 
 ## Future Work
+
+We present here four prevailing scenarios of change experienced in bird and mammal communities.
+Incorporation of additional data to fill geographic and taxanomic data gaps would be able to give us a clear picture which of these scenarios is most prevalent.
+However, data takes time to collect and new sources can be costly to incproporate.
+In the meantime then, we can make significant strides in linking the scenarios of change to common drivers.
+Do different drivers result in different signatures of change?
 
@@
 Here we make a significant first step in establishing a general trend for functional diversity through time across a variety of taxa and systems.
-We present the conclusion best supported by available data and acknowledge that overall trends could look very different for communities experiencing significant species losses or gains.
+We present the conclusion best supported by available data and acknowledge that overall trends would likely look different were underlying data a true global sample.
 Better understanding the functional make up of a species pool for a given community will be critical for evaluating a community's capacity for continued maintenance of functional structure, and be a first step in identifying which processes may be maintaining functional structure.

### 2023-06-15 analysis/05_model_metrics.Rmd
small update

@@
 library(emmeans)
+library(purrr)
 ```
@@
 # 
-# pmap_df(tofit_mods %>% select(metric_name = metric, filter_study_id = study_id), function(metric_name, filter_study_id) {model_data %>%
+# missing_indv_studies <- pmap_df(tofit_mods %>% select(metric_name = metric, filter_study_id = study_id), function(metric_name, filter_study_id) {model_data %>%
 #          filter(metric_name == metric, study_id == filter_study_id) %>%
@@
 #         select(metric, study_id, nrarefyID, n) %>%
-#         distinct()}) %>% View()
+#         distinct()})
 
@@
   select(-c(model_id, warning, statistic, df)) %>% 
-  write_csv(here::here("figures", "man_model_table.csv"))
+  readr::write_csv(here::here("figures", "man_model_table.csv"))
 
@@
     select(-c(model_id, n_ts, warning, lower.CL, upper.CL, significant, REML, bobyqa)) %>%
-  write_csv(here::here("paper", "model_estimates_supp.csv"))
+  readr::write_csv(here::here("paper", "model_estimates_supp.csv"))
 
@@
   filter(term == "year_scaled", p.value < 0.05) %>% 
-  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
+  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S", "FRic", "FEve", "FDiv")) %>% 
   mutate(positive = ifelse(estimate > 0, TRUE, FALSE)) %>%
   group_by(metric, positive) %>% 
-  summarise(n_study_id = n_distinct(study_id))
+  summarise(n_study_id = n_distinct(study_id)) %>%
+  pivot_wider(names_from = "metric", values_from = "n_study_id")
 

### 2023-06-15 paper/manuscript_GEB.Rmd
Rewrites! Title Change! Fix the table!

@@
 ---
-title: Evidence of widespread maintenance of functional composition in bird and mammal communities
+title: Prevailing scenarios of functional change in anthropocene bird and mammal communities
 authors:
@@
 abstract: |
- Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing species turnover but no overall trend in species diversity through time at the local scale.
+  Aim: Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing species turnover but no overall trend in species diversity through time at the local scale.
 Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to mechanisms of ecosystem processes.
-Here we present the first multi-taxa, multi-system analysis of functional change through time, pairing thousands of vertebrate assemblage time series from the BioTIME database with existing trait data representative of a species' functional role.
-We found no overall trend in any functional diversity metric, despite data replicating species-based patterns of constant richness with increasing turnover.
-The lack of trend held even after correcting for changes in species richness and at the study-level, where the vast majority of studies experienced no significant trend in any functional diversity metric.
-Results give evidence that across a variety of taxa, climates, and biomes, functional characteristics are maintained even in the face of significant community change.
-We also discuss the potential for underlying functional shifts to be obscured by current approaches and data, calling for targeted data collection efforts to combat existing biases in monitoring and trait data.
+Here we present the first multi-taxa, multi-system analysis of functional change through time.
+
+  Location: Global, with a North American focus
+
+  Time period: 1923-2014
+
+  Major taxa studied: Mammals, Birds
+
+  Methods: We paired thousands of bird and mammal assemblage time series from the BioTIME database with existing trait data representative of a species' functional role to reconstruct time series of functional diversity metrics.
+Using generalized linear mixed models we estimated general trends in those metrics and trends for individual studies.
+
+  Results: We found no overall trend in any functional diversity metric, despite data replicating species-based patterns of constant richness with increasing turnover.
+The lack of trend held even after correcting for changes in species richness.
+At the study-level, we identified four prevailing scenarios of species and functional change, which showed links to the duration of the observation window.
+
+  Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevelant than the other across many taxa, biomes, and realms.
+At the study-level, there were also a substantial number of time series exhibiting no species or functional change, however the majority of studies showed a shift in a species or functional metric.
+With no one prevailing scenario of change, it will be critical to link change scenarios to drivers of change, particularly to identify communities with capacity to resist drivers from those not experiencing substantial pressure from a driver.
 
@@
 The meaning of general trends derived from limited data, and their relevance for conservation, is a topic of on going debate.
-Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and for not taking into account the ecological context of individual studies [@cardinale2014; @cardinale2018; @gonzalez2016].
+Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and not taking into account the ecological context of individual studies [@cardinale2014; @cardinale2018; @gonzalez2016].
 Many of these criticisms reflect limitations of ecological data on the whole, and proponents argue we must make the best of the imperfect data available [@primack2018].
@@
 For some time periods, functional structure appears to be maintained for substantial portions of geological time [@hedberg2022].
-Still, some losses of functional diversity are indisputable on both paleocological and contemporary timescales such as continued trophic downgrading due to loss of large-bodied mammals, but implications of those losses for local diversity patterns are less clear [@estes2011; @smith2018] .
+Still, some losses of functional diversity are indisputable on both paleoecological and contemporary timescales such as continued trophic downgrading due to loss of large-bodied mammals, but implications of those losses for local diversity patterns are less clear [@estes2011; @smith2018] .
 
@@
 We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
+We also assessed the impact of potential study-level outliers on general trends, details on that process can be found in the supplement.
 All models in our analysis were fit using the *lme4* (v1.1-30) package in R (v4.2.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].
@@
 However, trends were not exhibited for the standardized metric indicating that differences were largely due to changes in species richness.
-Further, with only two global studies, the trend should not be considered truely general.
+Further, with only two global studies, the trend should not be considered truly general.
 The general trends for *CWM* models were similarly not significant, with a significant positive trend for only percentage of fish in diet composition.
@@
 read_csv(here::here("figures/man_model_table.csv")) %>%
-  select(metric, everything()) %>%
+  select(metric, everything(), -REML, -bobyqa) %>%
   select(-c(n_ts, formula, lower.CL, upper.CL, significant)) %>%
@@
 Hypothesis testing for study-level trends is likely affected by multiple testing issues and some trends identified as significant are therefore potentially spurious.
-Rather than interpreting changes in specific studies, we present these results as a general picture of how few studies experienced a significant trend and highlight that even that count is likely an overestimate.
+Rather than interpreting changes in specific studies, we present these results as a general picture of the kinds of trends experienced by communities.
 
@@
 Still, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing [@fritz2009].
-Trends could be significantly different for the same time periods in regions of sub-saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened [@ripple2015].
+Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened [@ripple2015].
 
@@
 dispersion, rarity, abundance of dominant species, @bagousse-pinguet2021].
-Critically, while trends for functional metrics may be statistically significant, the effect size necessary to be ecologically significant for key processes remains an outstanding question of the field.
 
@@
 In fact, study-level trends indicate quite the opposite, that functional shifts with unknown implications for ecosystem processes may be going undetected by common species-based approaches, particularly for short observation time windows.
-While the majority of studies in our dataset did not experience a significant functional loss, a substantial body of work links functional degradation to species losses as a result of direct human intervention in the form of land use change and intensification or habitat fragmentation [@flynn2009; @magioli2021; @tinoco2018].
+While the majority of studies in our dataset did not experience a significant functional loss, a substantial body of work links functional degradation to species losses as a result of direct human intervention in the form of land use change and intensification or habitat fragmentation, indicating that those studies are simply not representative of the kinds of impacts of greatest policy concern [@flynn2009; @magioli2021; @tinoco2018].
 
@@
 We present here four prevailing scenarios of change experienced in bird and mammal communities.
-Incorporation of additional data to fill geographic and taxanomic data gaps would be able to give us a clear picture which of these scenarios is most prevalent.
-However, data takes time to collect and new sources can be costly to incproporate.
-In the meantime then, we can make significant strides in linking the scenarios of change to common drivers.
-Do different drivers result in different signatures of change?
+While we offer an first discussion of which kinds of change are most common and why, assessing the true global prevalence will require continued efforts to fill data gaps, a well recognized challenge in ecology [@gonzalez2016; @hughes2021; @vellend2017].
+Still, a "scenarios of change" framework can provide structure for future work addressing functional shifts, particularly as we reconcile results from broad-scale syntheses with in-depth single system studies.
+It will be particularly critical to link forms of change to individual drivers to assess which drivers may impact species and functional diversity differently.
+Understanding those links will help identify where directly measuring functional structure instead of just species change is necessary for understand impacts on a given system.
+In addition, direct links to drivers will improve our ability to distinguish between communities experiencing no change due to a lack of perturbation from communities with high resilience in the face of disturbance.
 
-While the trends we see here are statistically significant, we still do not know if they are ecologically significant.
-There is substantial work to be done to link degrees of change in functional structure to measurable outcomes in ecosystem processes.
-
-Here we make a significant first step in establishing a general trend for functional diversity through time across a variety of taxa and systems.
-We present the conclusion best supported by available data and acknowledge that overall trends would likely look different were underlying data a true global sample.
-Better understanding the functional make up of a species pool for a given community will be critical for evaluating a community's capacity for continued maintenance of functional structure, and be a first step in identifying which processes may be maintaining functional structure.
-Further, looking at species pool make up through time would give us some indicator of the trajectory for future maintenance.
-
-Future broad-scales analysis would also greatly benefit from intentional and targeted data collection efforts.
-We join others in the call for increased monitoring in under sampled areas and continued efforts to centralize existing data sources [@gonzalez2016; @hughes2021; @vellend2017].
-Data that fill geographic, taxonomic and trait gaps should be prioritized over further collection of data that replicate existing biases.
-One relatively low-cost high-reward data investment is collation of additional species-level trait means.
-Intentional trait selection is critical for linking functional patterns to ecological processes [@zhu2017], however synthesis is constrained to the traits in a few taxa-specific databases.
-Trait collection should explicitly consider existing frameworks for linking traits to processes [e.g.
-the response and effect framework @lavorel2002] to facilitate clear ecological interpretation of potential functional changes.
+Here we identify trends that are statistically significant, however they may not necessarily be ecologically significant.
+While it is common to link changes in functional metrics to changes in ecosystem processes, those changes are less frequently discussed in terms of the size necessary for ecological impact.
+The degree of change in a process considered ecological meaningful is somewhat subjective and a function of the system and management context, making identifying the ecological meaning of broad-scale aggregate shifts even more challenging.
+It is further hampered by the use of species-level trait data, as the traits most closely linked to a process of interest may not be available [@zhu2017].
+Future trait collection that explicitly considers existing frameworks for linking traits to processes [e.g.
+the response and effect framework @lavorel2002] would facilitate improved ecological interpretation of potential functional changes.
 

### 2023-06-22 paper/GEB Submission/GEB_title_page.Rmd
Reorganization for GEB submission

--- 2023-06-15 paper/manuscript_GEB.Rmd
@@
 With no one prevailing scenario of change, it will be critical to link change scenarios to drivers of change, particularly to identify communities with capacity to resist drivers from those not experiencing substantial pressure from a driver.
-
 journal: "Ecology Letters"
@@
     wrap: sentence
-keywords:
-  - biodiversity change
-  - functional traits
-  - global change
-  - time series
-
-nocite: |
-  @bakker1990, @berezovikovn.n.2004, @enemar2004, @gido2019, @hogstad1993, @holmes1986, @jahncke2006, @krivenkov.g.1991, @melnikovy.i.2000, @monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2013, @monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2014, @monitori2014, @pardieck2020, @preston1960, @svensson2006, @svensson2010, @thorn2016changes, @thorn2016response, @usfs, @vermontcenterforecostudies2015, @vickery1984, @waide2017, @williamson1983, @woehlerericj., @zakharovv.d.1998, @meditss2011, @animald, @baltics, @piropno, @rossaferes1997, @scott, @silva, @ernest2009, @friggens2008, @jalilova.b.2014, @kartzinel2014, @kaufman2019, @krefting1974, @lightfoot, @malyshevy.s.2011, @meyer2008, @nedosekinv.y.2015, @prins1990, @rocha2017, @sheftelb.i., @stapp2014, @monitori2014, @bahamas, @popacet
 ---
 
-```{r message=FALSE, include = FALSE}
-library(kableExtra)
-library(magrittr)
-library(tidyverse)
-#library(printr)
-library(rticles)
-library(knitr)
+# Data Availability
 
-knitr::opts_chunk$set(cache=FALSE, message=FALSE, warning=FALSE, out.width = "\\textwidth")
-options(knitr.kable.NA = '')
+Code for the analyses in this chapter is archived on Zenodo at [10.5281/zenodo.5514334](https://doi.org/10.5281/zenodo.5514334).
 
-# no sci notation integers pleases
-options(scipen=999)
-```
+Data products are also archived on Zenodo at [10.5281/zenodo.6499442](https://doi.org/10.5281/zenodo.6499442).
 
-# Introduction
+Original data sources are open access and available at their respective providers.
 
-Ecological communities are experiencing unprecedented change as a result of anthropogenic pressures such as climate change, land use change, and invasive species.
-Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011], and fundamental changes in some of the most well-studied systems [e.g.
-coral bleaching, @sully2019].
-At the local scale however, species diversity tells a different story.
-Recent syntheses of local trends in biodiversity over time have found no net change in local species diversity despite ongoing turnover [@brown2001; @dornelas2014; @vellend2013; @vellend2017] and evidence of significant shifts in community composition underlying consistent species richness [@brose2016; @gotelli2017; @li2020].
-While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.
+# Funding
 
-The meaning of general trends derived from limited data, and their relevance for conservation, is a topic of on going debate.
-Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and not taking into account the ecological context of individual studies [@cardinale2014; @cardinale2018; @gonzalez2016].
-Many of these criticisms reflect limitations of ecological data on the whole, and proponents argue we must make the best of the imperfect data available [@primack2018].
-Still, there is general consensus on a few local-scale patterns that seem to be characteristic of the anthropocene, including the prevalence of species turnover across communities and patterns of both richness increase and decrease in different taxa and biomes [@dornelas2023].
-Understanding the implications of change in species assemblage for ecosystem stability and function is a critical next step in moving forward the biodiversity change conversation.
-
-Functional diversity offers a potentially powerful alternative to species-based approaches for detecting and describing community change by providing a mechanistic link between species' response to environmental change (*response traits*) and the processes they perform (*effect traits*) [@lavorel2002; @mcgill2006; @suding2008].
-By describing the functional trait space, functional diversity metrics capture the disproportionate impact of losses or gains of functionally unique species.
-Functional diversity metrics are therefore particularly well suited for assessing community shifts underlying even constant species richness trends.
-
-Beyond simply characterizing changes in community structure, trends in functional composition also have important implications for ecosystem stability, function, and resilience.
-There is increasing evidence functional diversity is a better predictor of ecosystem function than species-based metrics [@cadotte2011; @gagic2015], and that different facets of functional diversity play essential roles in maintaining ecosystem stability [@craven2018; @morin2014].
-Almost all hypothesized mechanisms underpinning the relationship between species diversity and ecosystem function are trait-dependent [@hillebrand2009].
-Determining functional trends therefore gives a more direct picture of potential trends in critical ecosystem processes.
-
-It is critical to establish whether or not functional loss is prevalent across communities.
-While functional loss is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], local-scale loss is not necessarily inevitable even in scenarios of species loss [@daz2001; @petchey2009].
-Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020].
-And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @larsen2018; @matuoka2020].
-On paleoecological time scales functional composition shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with significant impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].
-For some time periods, functional structure appears to be maintained for substantial portions of geological time [@hedberg2022].
-Still, some losses of functional diversity are indisputable on both paleoecological and contemporary timescales such as continued trophic downgrading due to loss of large-bodied mammals, but implications of those losses for local diversity patterns are less clear [@estes2011; @smith2018] .
-
-Assessments of broad-scale temporal change in functional diversity have previously been limited by a lack of functional trait data.
-The majority of work has therefore focused largely on system-specific studies with traits collected *in situ*.
-This has challenged our ability to establish general rules for how functional composition may be changing through time, particularly in response to global change drivers.
-Ongoing efforts to assemble functional traits for a variety of taxa have made synthesis of existing community assemblage data and functional traits possible for the first time, providing initial insights into the ways functional diversity changes on a broad scale for specific taxa [e.g.
-fish, @trindade-santos2020; birds, @jarzyna2016; @barnagaud2017].
-However, to date there has been no cross-taxa assessment of temporal functional change for a broad geographic and taxonomic extent [@díaz2022].
-
-Here we perform the first multi-taxa, multi-realm assessment of functional diversity change through time.
-We focus on mammal and bird species as a significant subset of the world's biodiversity heavily impacted by anthropogenic change.
-While examining trends in plants, invertebrates, and other vertebrate species is of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
-In order to ensure comparability across taxa in trait type and data quality we therefore focus on mammals and birds.
-Traits were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].
-An initial assessment of amphibian trends is included in the supplement, but excluded from general trend assessment here due to limited geographic coverage.
-
-We assess thousands of mammal and bird functional diversity time series to determine whether or not there is a general trend of functional change, both in observed metrics and in metrics corrected for changes in species richness.
-We further assess the prevailing modes of change occurring at the study level to determine how functional change is related to species patterns, and which kinds of species and functional change are most common across communities.
-We distinguish between three possible scenarios of functional change: 1) significant functional change accompanied by changes in species-based metrics, 2) maintenance of functional structure despite significant species changes, 3) maintenance of functional and species metrics.
-Based on expectation due to human impacts, we expect to find significant functional changes, often accompanied by species change.
-
-# Material and Methods
-
-## Data
-
-We obtained mammal and bird time series from the BioTIME database, a global repository of high quality assemblage time series.
-All studies included in the database follow consistent sampling protocols and represent full assemblages rather than populations of single species [@dornelas2018].
-All time series include abundance of observed species.
-Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual time series following a standardized spatial scale.
-Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018 for details on how sample extents were defined], with the area of each cell set to one standard deviation away from the mean of the single extent locations.
-All samples from a study within a single cell were considered to be a single time series, and species abundances were combined for all samples.
-
-We used trait data from the Elton Trait Database, which consists of species-level means for traits that represent species' multifaceted role in the community [@wilman2014].
-Traits include: body mass, diet, active diel period, nocturnality, forest foraging strata, pelagic use.
-Multiple traits (i.e. diet, foraging strata, activity seasonality, active diel period) were broken down into percentage or binary use for each level.
-
-In order to ensure taxonomic consistency across datasets, BioTIME species were paired with trait data based on their species identifier from the Integrated Taxonomic Information System database (retrieved 09-15-2020 from the on-line database, <https://doi.org/10.5066/F7KH0KBK>), obtained through the `taxadb` R package [@norman2020; @rcoreteam2021].
-If more than one species in the assemblage data resolved to the same identifier, observations were considered the same species.
-For trait data, traits for all species of the same identifier were averaged.
-Only studies for which at least 75% of species had trait data were included.
-In order to have a sufficient number of species to calculate functional diversity metrics, years with fewer than 5 species observed were also excluded.
-Sensitivity analyses were conducted for the trait coverage threshold and the duration of included time series.
-
-Many studies had a variable number of samples within years.
-To account for this inconsistency in sampling effort we used sample-based rarefaction by bootstrap resampling within years for each time series based on the smallest number of samples in a year for that time series.
-
-Our final dataset included 2,432 time series from 50 studies in 21 countries and 12 biomes and 6 different traits (Fig \ref{fig:taxaMap}).
-Data came from both terrestrial and marine realms and five climates (Global, Polar/Temperate, Temperate, Temperate/Tropical, Tropical).
-The earliest sample was in 1923 and the most recent was in 2014.
-For a full breakdown of studies and their characteristics, see the supplement.
-
-```{r taxaMap, fig.cap="A) Map of time series locations with points colored by taxa, and B) histograms of time series duration broken down by taxa.", echo=FALSE}
-knitr::include_graphics(here::here("figures/study_map_hist.jpeg"))
-```
-
-## Diversity Metrics
-
-We calculated yearly metrics of functional and species diversity for each time series.
-Species-based metrics include species richness (*S*) and Jaccard similarity (*J*) as a measure of turnover.
-Jaccard similarity was calculated relative to the first observed year for a time series.
-A negative trend in *J* would therefore indicate increasing turnover.
-We did not impose a correction for unobserved species as non-parametric estimators do not assign species identities to corrected richness values, and therefore could not be propagated to the functional diversity metrics.
-
-Functional diversity metrics were calculated using the *dbFD* function from the *FD* R package [@laliberté2010].
-Here we report functional richness (*FRic*), functional evenness (*FEve*), and functional divergence (*FDiv*) which together describe three complementary characteristics of the functional space [@mason2005; @hillebrand2009].
-*FRic* assesses the volume of the trait space occupied by species in the community, with higher values indicating communities with species of more extreme trait values.
-*FEve* describes how species are distributed across the trait space and how abundance is distributed across species.
-Higher values of *FEve* indicate more even spacing of species in the trait space and individuals across species.
-*FDiv* measures the degree to which species and their abundances maximize differences in the functional space.
-Higher values of *FDiv* therefore correspond to communities where many highly abundant species are on the edges of the trait space.
-We also calculated the community-weighted mean (*CWM*) of included traits to examine shifts in the distribution of each trait.
-
-All available trait data for each study were included in functional diversity calculations with the exception of traits that were the same value for all observed species in the study.
-For variables with multiple levels each level was included as a separate trait axis.
-Continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
-The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
-This restriction allows for a sufficient number of axes to capture the trait space while maintaining computational feasibility [@blonder2018].
-Metrics incorporated weighting based on species abundance.
-
-## Null Models
-
-To assess functional change independent of species richness we calculated the standardized effect size (SES) for each of the three summary functional diversity metrics (*FRic*, *FEve*, *FDiv*) from null estimates [@swenson2012].
-Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
-Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species.
-Species pools included all species observed for a time series.
-This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
-We used these values to calculate SES using the following formula: $SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
-We then calculated the median SES estimate for each metric from all the rarefaction samples for a time series.
-SES estimates can be interpreted as how much of the functional characteristic (richness, evenness, divergence) was observed beyond what was expected by chance for a community of that species richness.
-
-## Analysis
-
-We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study.
-We fit 18 individual *CWM* models.
-All time series with data for a given trait were included in the model.
-We estimated study level trends using individual linear models, for studies with more than one times series we fit a random slope and intercept for time series.
-Some study-level models could not be fit for five studies for at least one metric due to data limitations, but those studies were still included in the general models.
-They represented 12 of 1350 study-level models fit for each metric.
-For further details see the supplement.
-Where appropriate, response variables were $log$ or $log(x+1)$ transformed to better fit model assumptions of residual normality.
-
-To test for trends within and between different levels of taxa, biome, and realm we fit separate models with each of those factors added as a predictor to the original model structure.
-We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.8.2, @lenth2022].
-We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
-We also assessed the impact of potential study-level outliers on general trends, details on that process can be found in the supplement.
-All models in our analysis were fit using the *lme4* (v1.1-30) package in R (v4.2.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].
-
-# Results
-
-We found no significant overall trend in species richness or summary functional diversity metrics (observed or standardized) (Fig \ref{fig:timeseriesPlot}).
-We did find a significant overall decrease in Jaccard similarity, indicating accumulating changes in species composition.
-Non-significant overall trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies was plausibly 0 (Table \ref{tab:resultsTab}).
-Trends for different taxa, biomes, or realms were also non-significant for richness and summary functional diversity metrics, with the exception of a significantly increasing trend for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing functional richness slope for temperate/tropical studies and mammal studies.
-However, trends were not exhibited for the standardized metric indicating that differences were largely due to changes in species richness.
-Further, with only two global studies, the trend should not be considered truly general.
-The general trends for *CWM* models were similarly not significant, with a significant positive trend for only percentage of fish in diet composition.
-
-```{r timeseriesPlot, fig.cap="Plots of time series-level trends with line color corresponding to climatic region, with data points in grey and the overall metric mean in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES", echo=FALSE}
-knitr::include_graphics(here::here("figures/3met_long.jpeg"))
-```
-
-```{r resultsTab, echo=FALSE}
-read_csv(here::here("figures/man_model_table.csv")) %>%
-  select(metric, everything(), -REML, -bobyqa) %>%
-  select(-c(n_ts, formula, lower.CL, upper.CL, significant)) %>%
-  filter(group != "Residual" | is.na(group)) %>%
-  mutate(metric = ifelse(metric == "Jaccard_base", "log(Jaccard + 1)", metric),
-         metric = ifelse(metric == "S", "log(Species Richness)", metric),
-         effect = ifelse(effect == "ran_pars", "random", effect),
-         group = case_when(
-           group == "rarefyID:study_id" ~ "time series within study",
-           group == "study_id" ~ "study",
-           group == "Residual" ~ "residual",
-           TRUE ~ group),
-         term = case_when(
-           term == "(Intercept)" ~ "Intercept",
-           term == "year_scaled" ~ "Year",
-           term == "sd__(Intercept)" ~ "SD Intercept",
-           term == "cor__(Intercept).year_scaled" ~ "Corr(Intercept, Year)",
-           term == "sd__year_scaled" ~ "SD Year",
-           term == "sd__Observation" ~ "SD Observation",
-           TRUE ~ term
-         )) %>%
-  mutate(across(c(estimate, std.error, p.value), round, 2),
-         p.value = ifelse(p.value == 0, "<0.001", p.value)) %>%
-  arrange(metric != "Species Richness", metric, !is.na(group), group, term == "Corr(Intercept, Year)", term) %>%
-  dplyr::rename(grouping = group) %>%
-  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics.
-Additional model estimates, including CWM models, can be found in the supplement.") %>%
-  column_spec(3, width = "2cm") %>%
-  collapse_rows(columns = 1:3, valign = "middle") %>%
-  kable_paper() %>%
-  kable_styling(latex_options= c("scale_down", "hold_position"))
-```
-
-We did find significant differences between taxa, realms, and climate for Jaccard similarity and some of the *CWM's*.
-For example while Jaccard similarity was decreasing in the general trend and there were significant within group slopes for Mammals, Birds, and Terrestrial communities, there was no significant slope for the marine realm, indicating that the general trend is mostly driven by turnover in terrestrial communities.
-For bird communities, we also found within group trends and between group differences for trends in foraging behavior.
-We found a significant increasing trend in utilization of the canopy in Tropical communities that was significantly different from the trends for Polar/Temperate and Temperate communities.
-There was a significant decrease in utilization of the understory for Terrestrial communities and significant increase in foraging below the water surface for global studies (but see the previous limitations of Global data).
-There was also a significant positive slope for *CWM* body mass for the single Temperate/Tropical study (with three time series) which were marine mammal communities.
-
-We found significant dietary shifts across communities, with a significant increase in fruit consumption in Terrestrial communities and a significant decrease in nectar consumption in Tropical communities, a trend significantly different than that for Terrestrial communities.
-There was a significant increase in seed consumption in bird species, which was significantly different from the trend for Mammal communities.
-There was a significant increase in fish consumption for the two Global studies.
-Vertebrate consumption significantly decreased for Marine studies and for studies of Global and Tropical communities.
-
-At the study level, 15 of 50 studies exhibited a significant trend in species richness and 11 exhibited significant turnover.
-For observed functional metrics, 25 of 50 studies exhibited a trend in a least one metric, and 15 of 50 studies exhibited a significant trend for at least one standardized metrics (Table \ref{tab:trendTab}).
-In general, there were more significant trends for uncorrected metrics, with some disappearing after correction, indicating that those trends were likely due to changes in the number of species.
-Hypothesis testing for study-level trends is likely affected by multiple testing issues and some trends identified as significant are therefore potentially spurious.
-Rather than interpreting changes in specific studies, we present these results as a general picture of the kinds of trends experienced by communities.
-
-```{r trendTab, echo = FALSE}
-library(tidyverse)
-
-load(here::here("data/indv_mod_table.rda"))
-
-indv_mod_table %>%
-  filter(term == "year_scaled", p.value < 0.05) %>%
-  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S", "FRic", "FEve", "FDiv")) %>%
-  mutate(positive = ifelse(estimate > 0, "+", "-")) %>%
-  group_by(metric, positive) %>%
-  summarise(n_study_id = n_distinct(study_id)) %>%
-  pivot_wider(names_from = "metric", values_from = "n_study_id", values_fill = 0) %>%
-  select(positive, S, Jaccard_base, FRic, FEve, FDiv, SES_FRic, SES_FEve, SES_FDiv) %>%
-  kable(format = "latex", 
-        col.names = c("", "S", "Jaccard Similarity", "FRic", "FEve", "FDiv", "SES FRic", "SES FEve", "SES FDiv"),
-        caption = "Number of studies that experienced a significant trend in each calculated metric out of 50 total studies.", 
-        booktabs = TRUE, align = "c", escape = FALSE) %>%
-  kable_styling("striped")
-
-```
-
-Study-level slopes for summary functional metrics were significantly related start year of the time series for Jaccard similarity and functional evenness, both of which had significantly more negative slopes with more recent start year.
-No summary functional metrics were significantly related to the start year of the time series.
-
-We assessed the sensitivity of general trend results to major data processing decision by rerunning models with increasingly conservative subsets of the data.
-After excluding time series with less than two, three, and four year durations, we found that the general trend for increase in CWM aerial foraging disappeared, but two other trends in CWM's appeared.
-After excluding time series two years long, a general trend appeared for increased CWM use of the forest canopy, and decreased CWM of seed consumption.
-These two trends remained as increasingly shorter time series were excluded suggesting the increase in aerial foraging was an erroneous finding while the other two trends were more robust.
-Percentage of species in a community with trait data did not appear to have an affect on general trend results, as general trends were unchanged with increasing cut offs for percentage of species with trait data.
-A complete list of models run in the sensitivity analysis and their results can be found in the supplement.
-
-# Discussion
-
-Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at aggregate and local trends in functional diversity in mammal and bird communities.
-Surprisingly, we did not detect an overall trend in any of the summary functional diversity metrics.
-As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics may be consistent with similar well-documented species derived trends.
-We found no trend in functional change for almost all realms, biomes, and taxonomic groups, with evidence of functional richness loss in mammal studies only.
-These results are consistent with multiple studies linking anthropogenic drivers to loss of functional diversity in mammal communities [@brodie2021; @cox2022; @li2022]
-
-Despite a lack of general trends in summary functional metrics, we did find multiple trends in *CWM*'s for studies coming from the same taxa, realms, or climates, many of which are consistent with previous findings for those contexts.
-For example, the decrease in insect consumption for tropical communities reflects well documented declines in tropical insectivorous birds [@sherry2021].
-Reduction in utilization of understory foraging for terrestrial communities could be the result of disturbance like human recreation or increased predation from introduced species.
-Other shifts in diet for both Birds and Mammals, including increasing fruit, seed, and vertebrate consumption and decreasing seed consumption in some climates point to important areas for further exploration.
-
-At the study level, biodiversity change fell into four dominant scenarios of change based on trends in richness, turnover, and functional metrics: no change, only functional change, richness and functional change with turnover, and richness and functional change without turnover (Table \ref{tab:changeScenarios}).
-Around a quarter of studies experienced no trend in richness, turnover, or functional metrics.
-These studies spanned the distribution of study durations excluding only the very longest running studies, with the longest no change time series lasting 23 years.
-
-Another quarter of the studies showed no trend in richness or turnover, but did show a significant shift in at least one summary functional metric.
-These communities are best aligned with scenarios of species replacement where only a few functional outliers replaced functionally indistinct species or vice versa (there were studies with positive and negative trends in all three functional metrics).
-Studies in the scenario of only functional change were heavily skewed toward shorter running time series, indicating that they may represent a limited snapshot of communities that on longer time scales would be exhibiting significant change in species based metrics, particularly as species replacement continued adding to turnover.
-Communities in this scenario, and the previous scenario of no functional change are consistent with hypotheses that species richness may be strongly regulated through time [@brown2001; @gotelli2017], and emphasize that maintenance of functional structure can be divorced from those processes.
-
-About 10 percent of studies, but the majority of time series in our dataset, exhibited richness and functional change but no turnover.
-These studies also spanned the the distribution of time series durations, excluding the longest running studies.
-With the exception of one study, trends in functional change for this group were only found for uncorrected metrics and disappeared after correcting for species richness, indicating that functional change in this group is almost exclusively the result of species gains or losses rather than turnover of species with different functional traits.
-
-About 15 percent of studies, which were also exclusively the longest running studies, had a trend in turnover, richness, and functional change.
-The studies in this scenario fell into two groups: a positive trend in richness with a positive trend in functional richness (corrected and uncorrected), and a negative trend in species richness with a negative trend in uncorrected functional richness and no trend in corrected richness but a trend in other corrected functional metrics.
-This would indicate that when species are lost from a community, they are not more functionally unique than expected by chance, though they may produce shifts in other aspects of functional structure.
-When species are gained however, they disproportionately add to the functional richness of the community.
-
-Our results illustrate the range of relationships between species-based and functional metrics observed in real communities.
-Critically, particularly for the short observation time windows that characterize the majority of ecological data, functional shifts may be underlying seemingly static species-based community metrics.
-Species-based approaches may therefore be missing areas where significant change in the functional space could be impacting ecosystem function and resilience.
-In scenarios where there are significant species gains or losses over time, we found evidence that species losses may be relatively functionally indistinct, while species gains contribute disproportionately to the richness of a functional space.
-This finding is consistent with evidence that at least in bird communities, common functionally general species are being lost even as rare species are increasing [@inger2015; @rosenberg2019; @schipper2016].
-
-Our results are consistent with ecological expectations that community dissimilarity would be greater the larger the observation time window purely due to background processes, which is likely further accelerated by global change drivers [@dornelas2023].
-While we do not address how observed rates differ from background expectation here, we can make some assertions about the implication of turnover for functional change.
-Where turnover is occurring, it is almost always accompanied by a directional change in the functional space.
-Species are not therefore being replaced by exact functional counter parts.
-The fact that we observed both increases and decreases in corrected functional richness accompanying turnover is a reminder that replacements are not always an indication of functional loss, but can also lead to increased functional richness.
-While shifts in functional space may be the result of environmental pressure, we cannot distinguish between directional change and shifts due to the unavailability of an analog species in the species pool.
-
-```{r changeScenarios, echo = FALSE, message = FALSE, fig.show = 'hold', fig.cap= "Comparision of the distribution of time series durations from each type of change to the overall distribution of time series durations, with no richness change or turnover and functional change in yellow, no turnover but richness and functional change in green, no change in blue, and turnover, richness and functional change studies in orange."}
-
-knitr::include_graphics(here::here("figures/duration_hist.jpeg"))
-
-knitr::include_graphics(here::here("figures/functional_change_table.png"))
-```
-
-Still, some results stand in contrast to predictions for trait shifts under global change.
-For example, mean body size is predicted to decrease as a result of climate change impacts and megafaunal loss [@sheridan2011], a phenomena which has already been well documented empirically and experimentally in multiple taxa [@caruso2014; @forster2012; @huss2019; @tseng2018].
-While the species-level trait means used here are not appropriate for assessing intraspecific body size shifts, we would expect to see shifts in community-wide means due to local losses of large-bodied species.
-Instead, we found no evidence of a trend in *CWM* body size, with the exception of a single study of marine mammals in the Bahamas which showed a significant increase.
-Still, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing [@fritz2009].
-Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened [@ripple2015].
-
-What does local maintenance of functional structure mean for ecosystem function?
-The vast majority of experimental and observational work links declines in function to declines in functional or species diversity [@brose2016; @cadotte2011; @duffy2007].
-By those criteria very few communities in our dataset are in a state of concern for loss of functionality (Table \ref{tab:trendTab}), though we did find evidence that mammal communities may be in greater danger than bird communities.
-Still, shifts in metrics are only relevant if the underlying traits are those most critical for ecosystem function.
-We were limited in this analysis to the traits available rather than those with strong empirical links to function.
-Similarly, the dimensions of functional space most important for ecosystem function are still a topic of ongoing debate, and at least some known aspects important for multifunctionality were not measured here [e.g.
-dispersion, rarity, abundance of dominant species, @bagousse-pinguet2021].
-
-## Potential Methodological Limitations
-
-Here we approach the question of functional change using the best available data and biodiversity synthesis approaches.
-While we found signatures of change at the study level, some gaps in best practices may be obscuring a true general trend at the aggregate level.
-First, the BioTIME database, while the most comprehensive data source of time series available, is limited in temporal and geographic scope.
-Most time series span only a few years (Fig \ref{fig:taxaMap}) and may not provide the statistical power necessary to detect trends [@wauchope2019].
-The database is also not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
-We may simply not have data from those areas experiencing the greatest perturbation [@hughes2021], particularly scenarios of conversion to urban, human-dominated landscapes.
-While evidence from other work shows even disturbed communities can maintain functional structure [@edwards2013; @matuoka2020], these results should not be interpreted as evidence of low functional impact in areas of heavy human disturbance.
-
-Second, despite using the most comprehensive trait databases for these taxa, we were still limited to species-level means of the traits deemed important by database creators.
-The importance of intraspecific variation is well documented [@desroches2018; @violle2012], however individual-level traits are rarely collected alongside monitoring data, especially for the longest running efforts.
-Species-level traits may be obscuring more subtle shifts in the trait space happening within species.
-Likewise, available trait data may not capture the traits experiencing the greatest change.
-For example, we focus here mostly on ecological traits (i.e. foraging strategy, diet, etc) rather than life-history or reproductive traits.
-
-Third, while we use here the most common metrics for describing functional diversity, they do not measure some potentially important aspects of the functional space.
-Most notably, the summary metrics we calculated do not capture shifts in the location of the functional space as a whole.
-For example, two communities could have very similar metric values but no overlap in their trait spaces.
-This is especially relevant in the context of biodiversity change as a species loss could be replaced by a species with very different functional attributes, but the replacement would go undetected if the new species expanded the trait space by the same degree and had similar abundance.
-This scenario may be common in communities tracking changing environmental conditions.
-While trait CWM's capture axis shifts, approaches for assessing multidimensional shifts in functional space are still relatively new [@barros2016; @blonder2018; @mammola2019] but could shed critical insight into functional composition changes of this nature.
-
-## Policy Implications
-
-While we found no over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
-In fact, study-level trends indicate quite the opposite, that functional shifts with unknown implications for ecosystem processes may be going undetected by common species-based approaches, particularly for short observation time windows.
-While the majority of studies in our dataset did not experience a significant functional loss, a substantial body of work links functional degradation to species losses as a result of direct human intervention in the form of land use change and intensification or habitat fragmentation, indicating that those studies are simply not representative of the kinds of impacts of greatest policy concern [@flynn2009; @magioli2021; @tinoco2018].
-
-## Future Work
-
-We present here four prevailing scenarios of change experienced in bird and mammal communities.
-While we offer an first discussion of which kinds of change are most common and why, assessing the true global prevalence will require continued efforts to fill data gaps, a well recognized challenge in ecology [@gonzalez2016; @hughes2021; @vellend2017].
-Still, a "scenarios of change" framework can provide structure for future work addressing functional shifts, particularly as we reconcile results from broad-scale syntheses with in-depth single system studies.
-It will be particularly critical to link forms of change to individual drivers to assess which drivers may impact species and functional diversity differently.
-Understanding those links will help identify where directly measuring functional structure instead of just species change is necessary for understand impacts on a given system.
-In addition, direct links to drivers will improve our ability to distinguish between communities experiencing no change due to a lack of perturbation from communities with high resilience in the face of disturbance.
-
-Here we identify trends that are statistically significant, however they may not necessarily be ecologically significant.
-While it is common to link changes in functional metrics to changes in ecosystem processes, those changes are less frequently discussed in terms of the size necessary for ecological impact.
-The degree of change in a process considered ecological meaningful is somewhat subjective and a function of the system and management context, making identifying the ecological meaning of broad-scale aggregate shifts even more challenging.
-It is further hampered by the use of species-level trait data, as the traits most closely linked to a process of interest may not be available [@zhu2017].
-Future trait collection that explicitly considers existing frameworks for linking traits to processes [e.g.
-the response and effect framework @lavorel2002] would facilitate improved ecological interpretation of potential functional changes.
-
-<!-- # Data Availability -->
-
-<!-- Code for the analyses in this chapter is archived on Zenodo at [10.5281/zenodo.5514334](https://doi.org/10.5281/zenodo.5514334). -->
-
-<!-- Data products are also archived on Zenodo at [10.5281/zenodo.6499442](https://doi.org/10.5281/zenodo.6499442). -->
-
-<!-- Original data sources are open access and available at their respective providers. -->
-
-<!-- # Acknowledgments -->
-
-<!-- We thank the scientists who contributed to and maintain the biodiversity databases included in this study, including BioTIME, Amphibio, and Elton Traits. -->
-
-<!-- We thank Dr Shane Blowes and Dr Sarah Supp, whose code we adapted for initial data processing of the BioTIME database. -->
-
-<!-- Support for this work was provided by U.S.
+Support for this work was provided by U.S.
 Department of Energy through the Computational Sciences Graduate Fellowship (DE-FG02-97ER25308) awarded to KEAN.
 CB acknowledges support from NSF CAREER Award #1942280 & NIFA project CA-B-INS-0162-H.
--->
 
-# References
+# Biosketch
+
+Kari Norman’s research is generally focused on understanding the impacts of anthropogenic change on biodiversity.
+She brings together existing data sources to understand broadscale change and develops software tools to facilitate robust synthesis workflows.

### 2023-06-22 paper/GEB Submission/manuscript_GEB.Rmd
Reorganization for GEB submission

--- 2023-06-15 paper/manuscript_GEB.Rmd
@@
 title: Prevailing scenarios of functional change in anthropocene bird and mammal communities
-authors:
-  - name: "Kari E. A. Norman"
-    affiliation: Université de Montréal
-    location: Montréal, Quebec
-    email: kari.norman@berkeley.edu
-  - name: "Perry de Valpine"
-    affiliation: University of California, Berkeley
-    location: Berkeley, California
-    email: pdevalpine@berkeley.edu
-  - name: "Carl Boettiger"
-    affiliation: University of California, Berkeley
-    location: Berkeley, California
-    email: cboettig@berkeley.edu
 abstract: |
@@
 We fit 18 individual *CWM* models.
-All time series with data for a given trait were included in the model.
-We estimated study level trends using individual linear models, for studies with more than one times series we fit a random slope and intercept for time series.
+All time series with data for a given trait were included in the corresponding *CWM* model.
+We estimated study level trends using individual linear models.
+For studies with more than one times series we fit a random slope and intercept for time series.
 Some study-level models could not be fit for five studies for at least one metric due to data limitations, but those studies were still included in the general models.
@@
 We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
-We also assessed the impact of potential study-level outliers on general trends, details on that process can be found in the supplement.
+We also assessed the impact of potential study-level outliers on general trends.
+Details on that process can be found in the supplement.
 All models in our analysis were fit using the *lme4* (v1.1-30) package in R (v4.2.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].
@@
 
-Study-level slopes for summary functional metrics were significantly related start year of the time series for Jaccard similarity and functional evenness, both of which had significantly more negative slopes with more recent start year.
+Study-level slopes for summary functional metrics were significantly related to start year of the time series for Jaccard similarity and functional evenness, both of which had significantly more negative slopes with more recent start year.
 No summary functional metrics were significantly related to the start year of the time series.
@@
 We were limited in this analysis to the traits available rather than those with strong empirical links to function.
-Similarly, the dimensions of functional space most important for ecosystem function are still a topic of ongoing debate, and at least some known aspects important for multifunctionality were not measured here [e.g.
+Similarly, the dimensions of functional space most important for ecosystem function are still a topic of on going debate, and at least some known aspects important for multifunctionality were not measured here [e.g.
 dispersion, rarity, abundance of dominant species, @bagousse-pinguet2021].
@@
 Most time series span only a few years (Fig \ref{fig:taxaMap}) and may not provide the statistical power necessary to detect trends [@wauchope2019].
-The database is also not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study exhibits a strong Northern Hemisphere bias.
+The database is also not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study is predominantly from North America.
 We may simply not have data from those areas experiencing the greatest perturbation [@hughes2021], particularly scenarios of conversion to urban, human-dominated landscapes.
@@
 
-<!-- # Data Availability -->
+# Acknowledgments
 
-<!-- Code for the analyses in this chapter is archived on Zenodo at [10.5281/zenodo.5514334](https://doi.org/10.5281/zenodo.5514334). -->
+We thank the scientists who contributed to and maintain the biodiversity databases included in this study, including BioTIME, Amphibio, and Elton Traits.
 
-<!-- Data products are also archived on Zenodo at [10.5281/zenodo.6499442](https://doi.org/10.5281/zenodo.6499442). -->
-
-<!-- Original data sources are open access and available at their respective providers. -->
-
-<!-- # Acknowledgments -->
-
-<!-- We thank the scientists who contributed to and maintain the biodiversity databases included in this study, including BioTIME, Amphibio, and Elton Traits. -->
-
-<!-- We thank Dr Shane Blowes and Dr Sarah Supp, whose code we adapted for initial data processing of the BioTIME database. -->
-
-<!-- Support for this work was provided by U.S.
-Department of Energy through the Computational Sciences Graduate Fellowship (DE-FG02-97ER25308) awarded to KEAN.
-CB acknowledges support from NSF CAREER Award #1942280 & NIFA project CA-B-INS-0162-H.
--->
+We thank Dr Shane Blowes and Dr Sarah Supp, whose code we adapted for initial data processing of the BioTIME database.
 

### 2023-06-22 paper/GEB Submission/GEB_title_page.tex

--- 2023-06-22 paper/GEB Submission/GEB_title_page.Rmd
@@
----
-title: Prevailing scenarios of functional change in anthropocene bird and mammal communities
-authors:
-  - name: "Kari E. A. Norman"
-    affiliation: Université de Montréal
-    location: Montréal, Quebec
-    email: kari.norman@berkeley.edu
-  - name: "Perry de Valpine"
-    affiliation: University of California, Berkeley
-    location: Berkeley, California
-    email: pdevalpine@berkeley.edu
-  - name: "Carl Boettiger"
-    affiliation: University of California, Berkeley
-    location: Berkeley, California
-    email: cboettig@berkeley.edu
-abstract: |
-  Aim: Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing species turnover but no overall trend in species diversity through time at the local scale.
-Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to mechanisms of ecosystem processes.
-Here we present the first multi-taxa, multi-system analysis of functional change through time.
+\documentclass{article}
+
+\usepackage{arxiv}
+
+\usepackage[utf8]{inputenc} % allow utf-8 input
+\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
+\usepackage{hyperref}       % hyperlinks
+\usepackage{url}            % simple URL typesetting
+\usepackage{booktabs}       % professional-quality tables
+\usepackage{amsfonts}       % blackboard math symbols
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
+\usepackage{graphicx}
+
+\title{Prevailing scenarios of functional change in anthropocene bird
+and mammal communities}
+
+\author{
+    Kari E. A. Norman
+   \\
+    Université de Montréal \\
+  Montréal, Quebec \\
+  \texttt{\href{mailto:kari.norman@berkeley.edu}{\nolinkurl{kari.norman@berkeley.edu}}} \\
+   \And
+    Perry de Valpine
+   \\
+    University of California, Berkeley \\
+  Berkeley, California \\
+  \texttt{\href{mailto:pdevalpine@berkeley.edu}{\nolinkurl{pdevalpine@berkeley.edu}}} \\
+   \And
+    Carl Boettiger
+   \\
+    University of California, Berkeley \\
+  Berkeley, California \\
+  \texttt{\href{mailto:cboettig@berkeley.edu}{\nolinkurl{cboettig@berkeley.edu}}} \\
+  }
+
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+\usepackage{lineno}
+\linenumbers
+\begin{document}
+\maketitle
+
+\begin{abstract}
+Aim: Despite unprecedented environmental change due to anthropogenic
+pressure, recent work has found increasing species turnover but no
+overall trend in species diversity through time at the local scale.
+Functional diversity provides a potentially powerful alternative
+approach for understanding community composition by linking shifts in
+species identity to mechanisms of ecosystem processes. Here we present
+the first multi-taxa, multi-system analysis of functional change through
+time.
 
@@
 
-  Methods: We paired thousands of bird and mammal assemblage time series from the BioTIME database with existing trait data representative of a species' functional role to reconstruct time series of functional diversity metrics.
-Using generalized linear mixed models we estimated general trends in those metrics and trends for individual studies.
+Methods: We paired thousands of bird and mammal assemblage time series
+from the BioTIME database with existing trait data representative of a
+species' functional role to reconstruct time series of functional
+diversity metrics. Using generalized linear mixed models we estimated
+general trends in those metrics and trends for individual studies.
 
-  Results: We found no overall trend in any functional diversity metric, despite data replicating species-based patterns of constant richness with increasing turnover.
-The lack of trend held even after correcting for changes in species richness.
-At the study-level, we identified four prevailing scenarios of species and functional change, which showed links to the duration of the observation window.
+Results: We found no overall trend in any functional diversity metric,
+despite data replicating species-based patterns of constant richness
+with increasing turnover. The lack of trend held even after correcting
+for changes in species richness. At the study-level, we identified four
+prevailing scenarios of species and functional change, which showed
+links to the duration of the observation window.
 
-  Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevelant than the other across many taxa, biomes, and realms.
-At the study-level, there were also a substantial number of time series exhibiting no species or functional change, however the majority of studies showed a shift in a species or functional metric.
-With no one prevailing scenario of change, it will be critical to link change scenarios to drivers of change, particularly to identify communities with capacity to resist drivers from those not experiencing substantial pressure from a driver.
-journal: "Ecology Letters"
-date: "`r Sys.Date()`"
-bibliography: refs.bib
-csl: pnas.csl
-layout: 3p
-header-includes:
-   - \usepackage{lineno}
-   - \linenumbers
-output: 
-  rticles::arxiv_article:
-    keep_tex: true
-#  word_document: default
-editor_options: 
-  markdown: 
-    wrap: sentence
----
+Main Conclusions: General trends indicate that on the aggregate one type
+of functional shift is not more prevelant than the other across many
+taxa, biomes, and realms. At the study-level, there were also a
+substantial number of time series exhibiting no species or functional
+change, however the majority of studies showed a shift in a species or
+functional metric. With no one prevailing scenario of change, it will be
+critical to link change scenarios to drivers of change, particularly to
+identify communities with capacity to resist drivers from those not
+experiencing substantial pressure from a driver.
+\end{abstract}
 
-# Data Availability
+\hypertarget{data-availability}{%
+\section{Data Availability}\label{data-availability}}
 
-Code for the analyses in this chapter is archived on Zenodo at [10.5281/zenodo.5514334](https://doi.org/10.5281/zenodo.5514334).
+Code for the analyses in this chapter is archived on Zenodo at
+\href{https://doi.org/10.5281/zenodo.5514334}{10.5281/zenodo.5514334}.
 
-Data products are also archived on Zenodo at [10.5281/zenodo.6499442](https://doi.org/10.5281/zenodo.6499442).
+Data products are also archived on Zenodo at
+\href{https://doi.org/10.5281/zenodo.6499442}{10.5281/zenodo.6499442}.
 
-Original data sources are open access and available at their respective providers.
+Original data sources are open access and available at their respective
+providers.
 
-# Funding
+\hypertarget{funding}{%
+\section{Funding}\label{funding}}
 
-Support for this work was provided by U.S.
-Department of Energy through the Computational Sciences Graduate Fellowship (DE-FG02-97ER25308) awarded to KEAN.
-CB acknowledges support from NSF CAREER Award #1942280 & NIFA project CA-B-INS-0162-H.
+Support for this work was provided by U.S. Department of Energy through
+the Computational Sciences Graduate Fellowship (DE-FG02-97ER25308)
+awarded to KEAN. CB acknowledges support from NSF CAREER Award \#1942280
+\& NIFA project CA-B-INS-0162-H.
 
-# Biosketch
+\bibliographystyle{unsrt}
+\bibliography{refs.bib}
 
-Kari Norman’s research is generally focused on understanding the impacts of anthropogenic change on biodiversity.
-She brings together existing data sources to understand broadscale change and develops software tools to facilitate robust synthesis workflows.
+\end{document}

### 2023-06-22 paper/GEB Submission/manuscript_GEB.tex

\documentclass{article}

\usepackage{arxiv}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}

\title{Prevailing scenarios of functional change in anthropocene bird
and mammal communities}

\author{
  }

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\begin{document}
\maketitle

\begin{abstract}
Aim: Despite unprecedented environmental change due to anthropogenic
pressure, recent work has found increasing species turnover but no
overall trend in species diversity through time at the local scale.
Functional diversity provides a potentially powerful alternative
approach for understanding community composition by linking shifts in
species identity to mechanisms of ecosystem processes. Here we present
the first multi-taxa, multi-system analysis of functional change through
time.

Location: Global, with a North American focus

Time period: 1923-2014

Major taxa studied: Mammals, Birds

Methods: We paired thousands of bird and mammal assemblage time series
from the BioTIME database with existing trait data representative of a
species' functional role to reconstruct time series of functional
diversity metrics. Using generalized linear mixed models we estimated
general trends in those metrics and trends for individual studies.

Results: We found no overall trend in any functional diversity metric,
despite data replicating species-based patterns of constant richness
with increasing turnover. The lack of trend held even after correcting
for changes in species richness. At the study-level, we identified four
prevailing scenarios of species and functional change, which showed
links to the duration of the observation window.

Main Conclusions: General trends indicate that on the aggregate one type
of functional shift is not more prevelant than the other across many
taxa, biomes, and realms. At the study-level, there were also a
substantial number of time series exhibiting no species or functional
change, however the majority of studies showed a shift in a species or
functional metric. With no one prevailing scenario of change, it will be
critical to link change scenarios to drivers of change, particularly to
identify communities with capacity to resist drivers from those not
experiencing substantial pressure from a driver.
\end{abstract}

\keywords{
    biodiversity change
   \and
    functional traits
   \and
    global change
   \and
    time series
  }

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Ecological communities are experiencing unprecedented change as a result
of anthropogenic pressures such as climate change, land use change, and
invasive species. Impacts of these pressures are well documented at a
global scale by an accelerating global extinction rate (1), and
fundamental changes in some of the most well-studied systems (e.g.~coral
bleaching, 2). At the local scale however, species diversity tells a
different story. Recent syntheses of local trends in biodiversity over
time have found no net change in local species diversity despite ongoing
turnover (3--6) and evidence of significant shifts in community
composition underlying consistent species richness (7--9). While
communities are clearly changing, our most common species-based
approaches do not fully capture the nature of that change.

The meaning of general trends derived from limited data, and their
relevance for conservation, is a topic of on going debate. Global
analyses have been heavily criticized for geographic biases, lack of
data in the most heavily impacted areas, and not taking into account the
ecological context of individual studies (10--12). Many of these
criticisms reflect limitations of ecological data on the whole, and
proponents argue we must make the best of the imperfect data available
(13). Still, there is general consensus on a few local-scale patterns
that seem to be characteristic of the anthropocene, including the
prevalence of species turnover across communities and patterns of both
richness increase and decrease in different taxa and biomes (14).
Understanding the implications of change in species assemblage for
ecosystem stability and function is a critical next step in moving
forward the biodiversity change conversation.

Functional diversity offers a potentially powerful alternative to
species-based approaches for detecting and describing community change
by providing a mechanistic link between species' response to
environmental change (\emph{response traits}) and the processes they
perform (\emph{effect traits}) (15--17). By describing the functional
trait space, functional diversity metrics capture the disproportionate
impact of losses or gains of functionally unique species. Functional
diversity metrics are therefore particularly well suited for assessing
community shifts underlying even constant species richness trends.

Beyond simply characterizing changes in community structure, trends in
functional composition also have important implications for ecosystem
stability, function, and resilience. There is increasing evidence
functional diversity is a better predictor of ecosystem function than
species-based metrics (18, 19), and that different facets of functional
diversity play essential roles in maintaining ecosystem stability (20,
21). Almost all hypothesized mechanisms underpinning the relationship
between species diversity and ecosystem function are trait-dependent
(22). Determining functional trends therefore gives a more direct
picture of potential trends in critical ecosystem processes.

It is critical to establish whether or not functional loss is prevalent
across communities. While functional loss is frequently cited as one of
the most pressing concerns of the anthropocene (23--25), local-scale
loss is not necessarily inevitable even in scenarios of species loss
(26, 27). Forecasts of functional loss range from negligible (28) to
dire (29, 30). And while some observed trends show significant
functional loss (31) others document no loss even in some of the most
heavily impacted communities (32--34). On paleoecological time scales
functional composition shows mixed responses to environmental change and
extinction events (35, 36), with significant impacts of species
extinctions on functional diversity in some taxa and not others (37).
For some time periods, functional structure appears to be maintained for
substantial portions of geological time (38). Still, some losses of
functional diversity are indisputable on both paleoecological and
contemporary timescales such as continued trophic downgrading due to
loss of large-bodied mammals, but implications of those losses for local
diversity patterns are less clear (39, 40) .

Assessments of broad-scale temporal change in functional diversity have
previously been limited by a lack of functional trait data. The majority
of work has therefore focused largely on system-specific studies with
traits collected \emph{in situ}. This has challenged our ability to
establish general rules for how functional composition may be changing
through time, particularly in response to global change drivers. Ongoing
efforts to assemble functional traits for a variety of taxa have made
synthesis of existing community assemblage data and functional traits
possible for the first time, providing initial insights into the ways
functional diversity changes on a broad scale for specific taxa
(e.g.~fish, 41, birds, 42, 43). However, to date there has been no
cross-taxa assessment of temporal functional change for a broad
geographic and taxonomic extent (44).

Here we perform the first multi-taxa, multi-realm assessment of
functional diversity change through time. We focus on mammal and bird
species as a significant subset of the world's biodiversity heavily
impacted by anthropogenic change. While examining trends in plants,
invertebrates, and other vertebrate species is of equal interest, trait
data for those taxa raise additional challenges such as limited and
biased species coverage (45), a lack of accepted species-level means,
and differences in the types of traits collected. In order to ensure
comparability across taxa in trait type and data quality we therefore
focus on mammals and birds. Traits were intentionally selected to be
representative of a species' Eltonian niche, thereby summarizing the
functional role they play in the community (46). An initial assessment
of amphibian trends is included in the supplement, but excluded from
general trend assessment here due to limited geographic coverage.

We assess thousands of mammal and bird functional diversity time series
to determine whether or not there is a general trend of functional
change, both in observed metrics and in metrics corrected for changes in
species richness. We further assess the prevailing modes of change
occurring at the study level to determine how functional change is
related to species patterns, and which kinds of species and functional
change are most common across communities. We distinguish between three
possible scenarios of functional change: 1) significant functional
change accompanied by changes in species-based metrics, 2) maintenance
of functional structure despite significant species changes, 3)
maintenance of functional and species metrics. Based on expectation due
to human impacts, we expect to find significant functional changes,
often accompanied by species change.

\hypertarget{material-and-methods}{%
\section{Material and Methods}\label{material-and-methods}}

\hypertarget{data}{%
\subsection{Data}\label{data}}

We obtained mammal and bird time series from the BioTIME database, a
global repository of high quality assemblage time series. All studies
included in the database follow consistent sampling protocols and
represent full assemblages rather than populations of single species
(36). All time series include abundance of observed species. Following
best practices for the database (47), studies with multiple sample
locations were split into individual time series following a
standardized spatial scale. Scale was set by a global grid with cell
size determined based on the sample extent of studies with only a single
location (see 36 for details on how sample extents were defined), with
the area of each cell set to one standard deviation away from the mean
of the single extent locations. All samples from a study within a single
cell were considered to be a single time series, and species abundances
were combined for all samples.

We used trait data from the Elton Trait Database, which consists of
species-level means for traits that represent species' multifaceted role
in the community (46). Traits include: body mass, diet, active diel
period, nocturnality, forest foraging strata, pelagic use. Multiple
traits (i.e.~diet, foraging strata, activity seasonality, active diel
period) were broken down into percentage or binary use for each level.

In order to ensure taxonomic consistency across datasets, BioTIME
species were paired with trait data based on their species identifier
from the Integrated Taxonomic Information System database (retrieved
09-15-2020 from the on-line database,
\url{https://doi.org/10.5066/F7KH0KBK}), obtained through the
\texttt{taxadb} R package (48, 49). If more than one species in the
assemblage data resolved to the same identifier, observations were
considered the same species. For trait data, traits for all species of
the same identifier were averaged. Only studies for which at least 75\%
of species had trait data were included. In order to have a sufficient
number of species to calculate functional diversity metrics, years with
fewer than 5 species observed were also excluded. Sensitivity analyses
were conducted for the trait coverage threshold and the duration of
included time series.

Many studies had a variable number of samples within years. To account
for this inconsistency in sampling effort we used sample-based
rarefaction by bootstrap resampling within years for each time series
based on the smallest number of samples in a year for that time series.

Our final dataset included 2,432 time series from 50 studies in 21
countries and 12 biomes and 6 different traits (Fig \ref{fig:taxaMap}).
Data came from both terrestrial and marine realms and five climates
(Global, Polar/Temperate, Temperate, Temperate/Tropical, Tropical). The
earliest sample was in 1923 and the most recent was in 2014. For a full
breakdown of studies and their characteristics, see the supplement.

\begin{figure}
\includegraphics[width=\textwidth]{../../figures/study_map_hist} \caption{A) Map of time series locations with points colored by taxa, and B) histograms of time series duration broken down by taxa.}\label{fig:taxaMap}
\end{figure}

\hypertarget{diversity-metrics}{%
\subsection{Diversity Metrics}\label{diversity-metrics}}

We calculated yearly metrics of functional and species diversity for
each time series. Species-based metrics include species richness
(\emph{S}) and Jaccard similarity (\emph{J}) as a measure of turnover.
Jaccard similarity was calculated relative to the first observed year
for a time series. A negative trend in \emph{J} would therefore indicate
increasing turnover. We did not impose a correction for unobserved
species as non-parametric estimators do not assign species identities to
corrected richness values, and therefore could not be propagated to the
functional diversity metrics.

Functional diversity metrics were calculated using the \emph{dbFD}
function from the \emph{FD} R package (50). Here we report functional
richness (\emph{FRic}), functional evenness (\emph{FEve}), and
functional divergence (\emph{FDiv}) which together describe three
complementary characteristics of the functional space (22, 51).
\emph{FRic} assesses the volume of the trait space occupied by species
in the community, with higher values indicating communities with species
of more extreme trait values. \emph{FEve} describes how species are
distributed across the trait space and how abundance is distributed
across species. Higher values of \emph{FEve} indicate more even spacing
of species in the trait space and individuals across species.
\emph{FDiv} measures the degree to which species and their abundances
maximize differences in the functional space. Higher values of
\emph{FDiv} therefore correspond to communities where many highly
abundant species are on the edges of the trait space. We also calculated
the community-weighted mean (\emph{CWM}) of included traits to examine
shifts in the distribution of each trait.

All available trait data for each study were included in functional
diversity calculations with the exception of traits that were the same
value for all observed species in the study. For variables with multiple
levels each level was included as a separate trait axis. Continuous
traits were z-score scaled to give each trait equal weight in the trait
space (52, 53). The number of trait axes was limited to the maximum
number of traits that fulfills the criteria \(s >= 2^t\), where \(s\) is
the number of species and \emph{t} is the number of traits. This
restriction allows for a sufficient number of axes to capture the trait
space while maintaining computational feasibility (54). Metrics
incorporated weighting based on species abundance.

\hypertarget{null-models}{%
\subsection{Null Models}\label{null-models}}

To assess functional change independent of species richness we
calculated the standardized effect size (SES) for each of the three
summary functional diversity metrics (\emph{FRic}, \emph{FEve},
\emph{FDiv}) from null estimates (55). Null model corrections allow us
to assess the degree to which the observed functional diversity metric
deviates from the value expected by chance in a randomly assembled
community. Null estimates were calculated for each rarefied sample by
randomly sampling species from the species pool for each year and
randomly assigning observed abundances to species. Species pools
included all species observed for a time series. This process was
repeated 500 times to get an estimate and standard deviation of the null
expectation for the metric for each rarefaction sample for that time
series. We used these values to calculate SES using the following
formula: \(SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}\). We
then calculated the median SES estimate for each metric from all the
rarefaction samples for a time series. SES estimates can be interpreted
as how much of the functional characteristic (richness, evenness,
divergence) was observed beyond what was expected by chance for a
community of that species richness.

\hypertarget{analysis}{%
\subsection{Analysis}\label{analysis}}

We estimated general trends for each diversity metric using a linear
mixed effects model with a random slope and intercept for each study and
each time series nested within the study. We fit 18 individual
\emph{CWM} models. All time series with data for a given trait were
included in the corresponding \emph{CWM} model. We estimated study level
trends using individual linear models. For studies with more than one
times series we fit a random slope and intercept for time series. Some
study-level models could not be fit for five studies for at least one
metric due to data limitations, but those studies were still included in
the general models. They represented 12 of 1350 study-level models fit
for each metric. For further details see the supplement. Where
appropriate, response variables were \(log\) or \(log(x+1)\) transformed
to better fit model assumptions of residual normality.

To test for trends within and between different levels of taxa, biome,
and realm we fit separate models with each of those factors added as a
predictor to the original model structure. We estimated within-level
slopes and calculated between-level contrasts using the \emph{emmeans}
package (v1.8.2, 56). We assessed the impact of time series duration and
start year on study-level trends using linear models with duration and
start year as predictors. We also assessed the impact of potential
study-level outliers on general trends. Details on that process can be
found in the supplement. All models in our analysis were fit using the
\emph{lme4} (v1.1-30) package in R (v4.2.3) and p-values were calculated
by Satterthwaite's degrees of freedom method using the \emph{lmerTest}
(v3.1-3) package with a significance level of \(\alpha = 0.05\)
(57--59).

\hypertarget{results}{%
\section{Results}\label{results}}

We found no significant overall trend in species richness or summary
functional diversity metrics (observed or standardized) (Fig
\ref{fig:timeseriesPlot}). We did find a significant overall decrease in
Jaccard similarity, indicating accumulating changes in species
composition. Non-significant overall trends indicate that although some
studies experience increasing or decreasing trends, the average trend
across studies was plausibly 0 (Table \ref{tab:resultsTab}). Trends for
different taxa, biomes, or realms were also non-significant for richness
and summary functional diversity metrics, with the exception of a
significantly increasing trend for functional evenness of global studies
(characterized by having samples on multiple continents), and a
significantly decreasing functional richness slope for
temperate/tropical studies and mammal studies. However, trends were not
exhibited for the standardized metric indicating that differences were
largely due to changes in species richness. Further, with only two
global studies, the trend should not be considered truly general. The
general trends for \emph{CWM} models were similarly not significant,
with a significant positive trend for only percentage of fish in diet
composition.

\begin{figure}
\includegraphics[width=\textwidth]{../../figures/3met_long} \caption{Plots of time series-level trends with line color corresponding to climatic region, with data points in grey and the overall metric mean in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES}\label{fig:timeseriesPlot}
\end{figure}

\begin{table}[!h]

\caption{\label{tab:resultsTab}Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics. Additional model estimates, including CWM models, can be found in the supplement.}
\centering
\resizebox{\linewidth}{!}{
\begin{tabular}[t]{l|l|>{\raggedright\arraybackslash}p{2cm}|l|r|r|l}
\hline
metric & effect & grouping & term & estimate & std.error & p.value\\
\hline
 &  &  & Intercept & -0.25 & 0.11 & 0.03\\
\cline{4-7}
 & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & 0.00 & 0.04 & 0.96\\
\cline{2-7}
 &  &  & SD Intercept & 0.60 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.11 &  & \\
\cline{4-7}
 &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.16 &  & \\
\cline{3-7}
 &  &  & SD Intercept & 0.60 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.23 &  & \\
\cline{4-7}
\multirow{-8}{*}{\raggedright\arraybackslash SES\_FDiv} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & -0.01 &  & \\
\cline{1-7}
 &  &  & Intercept & 0.09 & 0.16 & 0.59\\
\cline{4-7}
 & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & 0.00 & 0.01 & 0.73\\
\cline{2-7}
 &  &  & SD Intercept & 1.04 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.02 &  & \\
\cline{4-7}
 &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & 1.00 &  & \\
\cline{3-7}
 &  &  & SD Intercept & 0.40 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.17 &  & \\
\cline{4-7}
\multirow{-8}{*}{\raggedright\arraybackslash SES\_FEve} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & -0.21 &  & \\
\cline{1-7}
 &  &  & Intercept & -0.26 & 0.07 & <0.001\\
\cline{4-7}
 & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & 0.04 & 0.04 & 0.35\\
\cline{2-7}
 &  &  & SD Intercept & 0.29 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.11 &  & \\
\cline{4-7}
 &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.27 &  & \\
\cline{3-7}
 &  &  & SD Intercept & 0.54 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.18 &  & \\
\cline{4-7}
\multirow{-8}{*}{\raggedright\arraybackslash SES\_FRic} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.06 &  & \\
\cline{1-7}
 &  &  & Intercept & 0.46 & 0.02 & <0.001\\
\cline{4-7}
 & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & -0.03 & 0.01 & <0.001\\
\cline{2-7}
 &  &  & SD Intercept & 0.09 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.02 &  & \\
\cline{4-7}
 &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.35 &  & \\
\cline{3-7}
 &  &  & SD Intercept & 0.08 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.02 &  & \\
\cline{4-7}
\multirow{-8}{*}{\raggedright\arraybackslash log(Jaccard + 1)} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.05 &  & \\
\cline{1-7}
 &  &  & Intercept & 2.40 & 0.10 & <0.001\\
\cline{4-7}
 & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & -0.06 & 0.05 & 0.17\\
\cline{2-7}
 &  &  & SD Intercept & 0.64 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.27 &  & \\
\cline{4-7}
 &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.70 &  & \\
\cline{3-7}
 &  &  & SD Intercept & 0.21 &  & \\
\cline{4-7}
 &  &  & SD Year & 0.09 &  & \\
\cline{4-7}
\multirow{-8}{*}{\raggedright\arraybackslash log(Species Richness)} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.34 &  & \\
\hline
\end{tabular}}
\end{table}

We did find significant differences between taxa, realms, and climate
for Jaccard similarity and some of the \emph{CWM's}. For example while
Jaccard similarity was decreasing in the general trend and there were
significant within group slopes for Mammals, Birds, and Terrestrial
communities, there was no significant slope for the marine realm,
indicating that the general trend is mostly driven by turnover in
terrestrial communities. For bird communities, we also found within
group trends and between group differences for trends in foraging
behavior. We found a significant increasing trend in utilization of the
canopy in Tropical communities that was significantly different from the
trends for Polar/Temperate and Temperate communities. There was a
significant decrease in utilization of the understory for Terrestrial
communities and significant increase in foraging below the water surface
for global studies (but see the previous limitations of Global data).
There was also a significant positive slope for \emph{CWM} body mass for
the single Temperate/Tropical study (with three time series) which were
marine mammal communities.

We found significant dietary shifts across communities, with a
significant increase in fruit consumption in Terrestrial communities and
a significant decrease in nectar consumption in Tropical communities, a
trend significantly different than that for Terrestrial communities.
There was a significant increase in seed consumption in bird species,
which was significantly different from the trend for Mammal communities.
There was a significant increase in fish consumption for the two Global
studies. Vertebrate consumption significantly decreased for Marine
studies and for studies of Global and Tropical communities.

At the study level, 15 of 50 studies exhibited a significant trend in
species richness and 11 exhibited significant turnover. For observed
functional metrics, 25 of 50 studies exhibited a trend in a least one
metric, and 15 of 50 studies exhibited a significant trend for at least
one standardized metrics (Table \ref{tab:trendTab}). In general, there
were more significant trends for uncorrected metrics, with some
disappearing after correction, indicating that those trends were likely
due to changes in the number of species. Hypothesis testing for
study-level trends is likely affected by multiple testing issues and
some trends identified as significant are therefore potentially
spurious. Rather than interpreting changes in specific studies, we
present these results as a general picture of the kinds of trends
experienced by communities.

\begin{table}

\caption{\label{tab:trendTab}Number of studies that experienced a significant trend in each calculated metric out of 50 total studies.}
\centering
\begin{tabular}[t]{ccccccccc}
\toprule
 & S & Jaccard Similarity & FRic & FEve & FDiv & SES FRic & SES FEve & SES FDiv\\
\midrule
+ & 7 & 0 & 7 & 5 & 3 & 4 & 4 & 3\\
- & 8 & 11 & 8 & 6 & 5 & 2 & 6 & 2\\
\bottomrule
\end{tabular}
\end{table}

Study-level slopes for summary functional metrics were significantly
related to start year of the time series for Jaccard similarity and
functional evenness, both of which had significantly more negative
slopes with more recent start year. No summary functional metrics were
significantly related to the start year of the time series.

We assessed the sensitivity of general trend results to major data
processing decision by rerunning models with increasingly conservative
subsets of the data. After excluding time series with less than two,
three, and four year durations, we found that the general trend for
increase in CWM aerial foraging disappeared, but two other trends in
CWM's appeared. After excluding time series two years long, a general
trend appeared for increased CWM use of the forest canopy, and decreased
CWM of seed consumption. These two trends remained as increasingly
shorter time series were excluded suggesting the increase in aerial
foraging was an erroneous finding while the other two trends were more
robust. Percentage of species in a community with trait data did not
appear to have an affect on general trend results, as general trends
were unchanged with increasing cut offs for percentage of species with
trait data. A complete list of models run in the sensitivity analysis
and their results can be found in the supplement.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Our study represents the largest broad-scale multi-taxa assessment of
functional change through time to date, giving a first look at aggregate
and local trends in functional diversity in mammal and bird communities.
Surprisingly, we did not detect an overall trend in any of the summary
functional diversity metrics. As with previous species-based syntheses,
we also found no overall trend in species richness accompanied by
increasing turnover through time (36), indicating that non-significant
trends in functional metrics may be consistent with similar
well-documented species derived trends. We found no trend in functional
change for almost all realms, biomes, and taxonomic groups, with
evidence of functional richness loss in mammal studies only. These
results are consistent with multiple studies linking anthropogenic
drivers to loss of functional diversity in mammal communities (60--62)

Despite a lack of general trends in summary functional metrics, we did
find multiple trends in \emph{CWM}'s for studies coming from the same
taxa, realms, or climates, many of which are consistent with previous
findings for those contexts. For example, the decrease in insect
consumption for tropical communities reflects well documented declines
in tropical insectivorous birds (63). Reduction in utilization of
understory foraging for terrestrial communities could be the result of
disturbance like human recreation or increased predation from introduced
species. Other shifts in diet for both Birds and Mammals, including
increasing fruit, seed, and vertebrate consumption and decreasing seed
consumption in some climates point to important areas for further
exploration.

At the study level, biodiversity change fell into four dominant
scenarios of change based on trends in richness, turnover, and
functional metrics: no change, only functional change, richness and
functional change with turnover, and richness and functional change
without turnover (Table \ref{tab:changeScenarios}). Around a quarter of
studies experienced no trend in richness, turnover, or functional
metrics. These studies spanned the distribution of study durations
excluding only the very longest running studies, with the longest no
change time series lasting 23 years.

Another quarter of the studies showed no trend in richness or turnover,
but did show a significant shift in at least one summary functional
metric. These communities are best aligned with scenarios of species
replacement where only a few functional outliers replaced functionally
indistinct species or vice versa (there were studies with positive and
negative trends in all three functional metrics). Studies in the
scenario of only functional change were heavily skewed toward shorter
running time series, indicating that they may represent a limited
snapshot of communities that on longer time scales would be exhibiting
significant change in species based metrics, particularly as species
replacement continued adding to turnover. Communities in this scenario,
and the previous scenario of no functional change are consistent with
hypotheses that species richness may be strongly regulated through time
(3, 8), and emphasize that maintenance of functional structure can be
divorced from those processes.

About 10 percent of studies, but the majority of time series in our
dataset, exhibited richness and functional change but no turnover. These
studies also spanned the the distribution of time series durations,
excluding the longest running studies. With the exception of one study,
trends in functional change for this group were only found for
uncorrected metrics and disappeared after correcting for species
richness, indicating that functional change in this group is almost
exclusively the result of species gains or losses rather than turnover
of species with different functional traits.

About 15 percent of studies, which were also exclusively the longest
running studies, had a trend in turnover, richness, and functional
change. The studies in this scenario fell into two groups: a positive
trend in richness with a positive trend in functional richness
(corrected and uncorrected), and a negative trend in species richness
with a negative trend in uncorrected functional richness and no trend in
corrected richness but a trend in other corrected functional metrics.
This would indicate that when species are lost from a community, they
are not more functionally unique than expected by chance, though they
may produce shifts in other aspects of functional structure. When
species are gained however, they disproportionately add to the
functional richness of the community.

Our results illustrate the range of relationships between species-based
and functional metrics observed in real communities. Critically,
particularly for the short observation time windows that characterize
the majority of ecological data, functional shifts may be underlying
seemingly static species-based community metrics. Species-based
approaches may therefore be missing areas where significant change in
the functional space could be impacting ecosystem function and
resilience. In scenarios where there are significant species gains or
losses over time, we found evidence that species losses may be
relatively functionally indistinct, while species gains contribute
disproportionately to the richness of a functional space. This finding
is consistent with evidence that at least in bird communities, common
functionally general species are being lost even as rare species are
increasing (64--66).

Our results are consistent with ecological expectations that community
dissimilarity would be greater the larger the observation time window
purely due to background processes, which is likely further accelerated
by global change drivers (14). While we do not address how observed
rates differ from background expectation here, we can make some
assertions about the implication of turnover for functional change.
Where turnover is occurring, it is almost always accompanied by a
directional change in the functional space. Species are not therefore
being replaced by exact functional counter parts. The fact that we
observed both increases and decreases in corrected functional richness
accompanying turnover is a reminder that replacements are not always an
indication of functional loss, but can also lead to increased functional
richness. While shifts in functional space may be the result of
environmental pressure, we cannot distinguish between directional change
and shifts due to the unavailability of an analog species in the species
pool.

\begin{figure}
\includegraphics[width=\textwidth]{../../figures/duration_hist} \includegraphics[width=\textwidth]{../../figures/functional_change_table} \caption{Comparision of the distribution of time series durations from each type of change to the overall distribution of time series durations, with no richness change or turnover and functional change in yellow, no turnover but richness and functional change in green, no change in blue, and turnover, richness and functional change studies in orange.}\label{fig:changeScenarios}
\end{figure}

Still, some results stand in contrast to predictions for trait shifts
under global change. For example, mean body size is predicted to
decrease as a result of climate change impacts and megafaunal loss (67),
a phenomena which has already been well documented empirically and
experimentally in multiple taxa (68--71). While the species-level trait
means used here are not appropriate for assessing intraspecific body
size shifts, we would expect to see shifts in community-wide means due
to local losses of large-bodied species. Instead, we found no evidence
of a trend in \emph{CWM} body size, with the exception of a single study
of marine mammals in the Bahamas which showed a significant increase.
Still, many of the studies in our dataset draw from areas that may have
experienced significant loss of large-bodied species before the
observation window, with contemporary loss rates slowing (72). Trends
could be significantly different for the same time periods in regions of
sub-Saharan Africa, for example, which has poor representation in our
dataset but where megafauna exist on the landscape and are increasingly
threatened (73).

What does local maintenance of functional structure mean for ecosystem
function? The vast majority of experimental and observational work links
declines in function to declines in functional or species diversity (7,
18, 74). By those criteria very few communities in our dataset are in a
state of concern for loss of functionality (Table \ref{tab:trendTab}),
though we did find evidence that mammal communities may be in greater
danger than bird communities. Still, shifts in metrics are only relevant
if the underlying traits are those most critical for ecosystem function.
We were limited in this analysis to the traits available rather than
those with strong empirical links to function. Similarly, the dimensions
of functional space most important for ecosystem function are still a
topic of on going debate, and at least some known aspects important for
multifunctionality were not measured here (e.g.~dispersion, rarity,
abundance of dominant species, 75).

\hypertarget{potential-methodological-limitations}{%
\subsection{Potential Methodological
Limitations}\label{potential-methodological-limitations}}

Here we approach the question of functional change using the best
available data and biodiversity synthesis approaches. While we found
signatures of change at the study level, some gaps in best practices may
be obscuring a true general trend at the aggregate level. First, the
BioTIME database, while the most comprehensive data source of time
series available, is limited in temporal and geographic scope. Most time
series span only a few years (Fig \ref{fig:taxaMap}) and may not provide
the statistical power necessary to detect trends (76). The database is
also not a representative sample of the world's biodiversity or areas of
greatest threat (6, 12), and the subset of data in this study is
predominantly from North America. We may simply not have data from those
areas experiencing the greatest perturbation (77), particularly
scenarios of conversion to urban, human-dominated landscapes. While
evidence from other work shows even disturbed communities can maintain
functional structure (32, 34), these results should not be interpreted
as evidence of low functional impact in areas of heavy human
disturbance.

Second, despite using the most comprehensive trait databases for these
taxa, we were still limited to species-level means of the traits deemed
important by database creators. The importance of intraspecific
variation is well documented (78, 79), however individual-level traits
are rarely collected alongside monitoring data, especially for the
longest running efforts. Species-level traits may be obscuring more
subtle shifts in the trait space happening within species. Likewise,
available trait data may not capture the traits experiencing the
greatest change. For example, we focus here mostly on ecological traits
(i.e.~foraging strategy, diet, etc) rather than life-history or
reproductive traits.

Third, while we use here the most common metrics for describing
functional diversity, they do not measure some potentially important
aspects of the functional space. Most notably, the summary metrics we
calculated do not capture shifts in the location of the functional space
as a whole. For example, two communities could have very similar metric
values but no overlap in their trait spaces. This is especially relevant
in the context of biodiversity change as a species loss could be
replaced by a species with very different functional attributes, but the
replacement would go undetected if the new species expanded the trait
space by the same degree and had similar abundance. This scenario may be
common in communities tracking changing environmental conditions. While
trait CWM's capture axis shifts, approaches for assessing
multidimensional shifts in functional space are still relatively new
(54, 80, 81) but could shed critical insight into functional composition
changes of this nature.

\hypertarget{policy-implications}{%
\subsection{Policy Implications}\label{policy-implications}}

While we found no over all trends in functional metrics, our results
should not be interpreted as an indication that the ongoing biodiversity
crisis is less severe than previously described, or that there is no
concern for functional change as a result of anthropogenic impact. In
fact, study-level trends indicate quite the opposite, that functional
shifts with unknown implications for ecosystem processes may be going
undetected by common species-based approaches, particularly for short
observation time windows. While the majority of studies in our dataset
did not experience a significant functional loss, a substantial body of
work links functional degradation to species losses as a result of
direct human intervention in the form of land use change and
intensification or habitat fragmentation, indicating that those studies
are simply not representative of the kinds of impacts of greatest policy
concern (31, 82, 83).

\hypertarget{future-work}{%
\subsection{Future Work}\label{future-work}}

We present here four prevailing scenarios of change experienced in bird
and mammal communities. While we offer an first discussion of which
kinds of change are most common and why, assessing the true global
prevalence will require continued efforts to fill data gaps, a well
recognized challenge in ecology (6, 12, 77). Still, a ``scenarios of
change'' framework can provide structure for future work addressing
functional shifts, particularly as we reconcile results from broad-scale
syntheses with in-depth single system studies. It will be particularly
critical to link forms of change to individual drivers to assess which
drivers may impact species and functional diversity differently.
Understanding those links will help identify where directly measuring
functional structure instead of just species change is necessary for
understand impacts on a given system. In addition, direct links to
drivers will improve our ability to distinguish between communities
experiencing no change due to a lack of perturbation from communities
with high resilience in the face of disturbance.

Here we identify trends that are statistically significant, however they
may not necessarily be ecologically significant. While it is common to
link changes in functional metrics to changes in ecosystem processes,
those changes are less frequently discussed in terms of the size
necessary for ecological impact. The degree of change in a process
considered ecological meaningful is somewhat subjective and a function
of the system and management context, making identifying the ecological
meaning of broad-scale aggregate shifts even more challenging. It is
further hampered by the use of species-level trait data, as the traits
most closely linked to a process of interest may not be available (84).
Future trait collection that explicitly considers existing frameworks
for linking traits to processes (e.g.~the response and effect framework
15) would facilitate improved ecological interpretation of potential
functional changes.

\hypertarget{acknowledgments}{%
\section{Acknowledgments}\label{acknowledgments}}

We thank the scientists who contributed to and maintain the biodiversity
databases included in this study, including BioTIME, Amphibio, and Elton
Traits.

We thank Dr Shane Blowes and Dr Sarah Supp, whose code we adapted for
initial data processing of the BioTIME database.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{0}{0}
\leavevmode\vadjust pre{\hypertarget{ref-barnosky2011}{}}%
\CSLLeftMargin{1. }%
\CSLRightInline{Barnosky AD, et al. (2011)
\href{https://doi.org/10.1038/nature09678}{Has the Earth{'}s sixth mass
extinction already arrived?} \emph{Nature} 471(7336):51--57.}

\leavevmode\vadjust pre{\hypertarget{ref-sully2019}{}}%
\CSLLeftMargin{2. }%
\CSLRightInline{Sully S, Burkepile DE, Donovan MK, Hodgson G, van Woesik
R (2019) \href{https://doi.org/10.1038/s41467-019-09238-2}{A global
analysis of coral bleaching over the past two decades}. \emph{Nature
Communications} 10(1):1264.}

\leavevmode\vadjust pre{\hypertarget{ref-brown2001}{}}%
\CSLLeftMargin{3. }%
\CSLRightInline{Brown JH, Ernest SKM, Parody JM, Haskell JP (2001)
\href{http://www.jstor.org/stable/4222853}{Regulation of diversity:
Maintenance of species richness in changing environments}.
\emph{Oecologia} 126(3):321--332.}

\leavevmode\vadjust pre{\hypertarget{ref-dornelas2014}{}}%
\CSLLeftMargin{4. }%
\CSLRightInline{Dornelas M, et al. (2014)
\href{https://doi.org/10.1126/science.1248484}{Assemblage Time Series
Reveal Biodiversity Change but Not Systematic Loss}. \emph{Science}
344(6181):296--299.}

\leavevmode\vadjust pre{\hypertarget{ref-vellend2013}{}}%
\CSLLeftMargin{5. }%
\CSLRightInline{Vellend M, et al. (2013)
\href{https://doi.org/10.1073/pnas.1312779110}{Global meta-analysis
reveals no net change in local-scale plant biodiversity over time}.
\emph{Proceedings of the National Academy of Sciences}
110(48):19456--19459.}

\leavevmode\vadjust pre{\hypertarget{ref-vellend2017}{}}%
\CSLLeftMargin{6. }%
\CSLRightInline{Vellend M, et al. (2017)
\href{https://doi.org/10.1002/ecy.1660}{Estimates of local biodiversity
change over time stand up to scrutiny}. \emph{Ecology} 98(2):583--590.}

\leavevmode\vadjust pre{\hypertarget{ref-brose2016}{}}%
\CSLLeftMargin{7. }%
\CSLRightInline{Brose U, Hillebrand H (2016)
\href{https://doi.org/10.1098/rstb.2015.0267}{Biodiversity and ecosystem
functioning in dynamic landscapes}. \emph{Philosophical Transactions of
the Royal Society B: Biological Sciences} 371(1694):20150267.}

\leavevmode\vadjust pre{\hypertarget{ref-gotelli2017}{}}%
\CSLLeftMargin{8. }%
\CSLRightInline{Gotelli NJ, et al. (2017) Community-level regulation of
temporal trends in biodiversity. \emph{Science Advances} 3(7).
doi:\href{https://doi.org/10.1126/sciadv.1700315}{10.1126/sciadv.1700315}.}

\leavevmode\vadjust pre{\hypertarget{ref-li2020}{}}%
\CSLLeftMargin{9. }%
\CSLRightInline{Li D, et al. (2020)
\href{https://doi.org/10.1098/rspb.2020.0777}{Changes in taxonomic and
phylogenetic diversity in the anthropocene}. \emph{Proceedings of the
Royal Society B: Biological Sciences} 287(1929):20200777.}

\leavevmode\vadjust pre{\hypertarget{ref-cardinale2014}{}}%
\CSLLeftMargin{10. }%
\CSLRightInline{Cardinale B (2014)
\href{https://doi.org/10.1126/science.344.6188.1098-a}{Overlooked local
biodiversity loss}. \emph{Science} 344(6188):1098--1098.}

\leavevmode\vadjust pre{\hypertarget{ref-cardinale2018}{}}%
\CSLLeftMargin{11. }%
\CSLRightInline{Cardinale BJ, Gonzalez A, Allington GRH, Loreau M (2018)
Is local biodiversity declining or not? A summary of the debate over
analysis of species richness time trends. \emph{Biological
Conservation}.
doi:\href{https://doi.org/10.1016/j.biocon.2017.12.021}{10.1016/j.biocon.2017.12.021}.}

\leavevmode\vadjust pre{\hypertarget{ref-gonzalez2016}{}}%
\CSLLeftMargin{12. }%
\CSLRightInline{Gonzalez A, et al. (2016)
\href{https://doi.org/10.1890/15-1759.1}{Estimating local biodiversity
change: a critique of papers claiming no net loss of local diversity}.
\emph{Ecology} 97(8):1949--1960.}

\leavevmode\vadjust pre{\hypertarget{ref-primack2018}{}}%
\CSLLeftMargin{13. }%
\CSLRightInline{Primack RB, et al. (2018)
\href{https://doi.org/10.1016/j.biocon.2017.12.023}{Biodiversity gains?
The debate on changes in local- vs global-scale species richness}.
\emph{Biological Conservation} 219:A1--A3.}

\leavevmode\vadjust pre{\hypertarget{ref-dornelas2023}{}}%
\CSLLeftMargin{14. }%
\CSLRightInline{Dornelas M, et al. (2023)
\href{https://doi.org/10.1098/rstb.2022.0199}{Looking back on
biodiversity change: Lessons for the road ahead}. \emph{Philosophical
Transactions of the Royal Society B: Biological Sciences}
378(1881):20220199.}

\leavevmode\vadjust pre{\hypertarget{ref-lavorel2002}{}}%
\CSLLeftMargin{15. }%
\CSLRightInline{Lavorel S, Garnier E (2002)
\href{https://doi.org/10.1046/j.1365-2435.2002.00664.x}{Predicting
changes in community composition and ecosystem functioning from plant
traits: revisiting the Holy Grail}. \emph{Functional Ecology}
16(5):545--556.}

\leavevmode\vadjust pre{\hypertarget{ref-mcgill2006}{}}%
\CSLLeftMargin{16. }%
\CSLRightInline{Mcgill B, Enquist B, Weiher E, Westoby M (2006)
\href{https://doi.org/10.1016/j.tree.2006.02.002}{Rebuilding community
ecology from functional traits}. \emph{Trends in Ecology \& Evolution}
21(4):178--185.}

\leavevmode\vadjust pre{\hypertarget{ref-suding2008}{}}%
\CSLLeftMargin{17. }%
\CSLRightInline{Suding KN, et al. (2008)
\href{https://doi.org/10.1111/j.1365-2486.2008.01557.x}{Scaling
environmental change through the community-level: a trait-based
response-and-effect framework for plants}. \emph{Global Change Biology}
14(5):1125--1140.}

\leavevmode\vadjust pre{\hypertarget{ref-cadotte2011}{}}%
\CSLLeftMargin{18. }%
\CSLRightInline{Cadotte MW, Carscadden K, Mirotchnick N (2011)
\href{https://doi.org/10.1111/j.1365-2664.2011.02048.x}{Beyond species:
functional diversity and the maintenance of ecological processes and
services}. \emph{Journal of Applied Ecology} 48(5):1079--1087.}

\leavevmode\vadjust pre{\hypertarget{ref-gagic2015}{}}%
\CSLLeftMargin{19. }%
\CSLRightInline{Gagic V, et al. (2015)
\href{https://doi.org/10.1098/rspb.2014.2620}{Functional identity and
diversity of animals predict ecosystem functioning better than
species-based indices}. \emph{Proceedings of the Royal Society B:
Biological Sciences} 282(1801):20142620.}

\leavevmode\vadjust pre{\hypertarget{ref-craven2018}{}}%
\CSLLeftMargin{20. }%
\CSLRightInline{Craven D, et al. (2018)
\href{https://doi.org/10.1038/s41559-018-0647-7}{Multiple facets of
biodiversity drive the diversity{\textendash}stability relationship}.
\emph{Nature Ecology \& Evolution}:1.}

\leavevmode\vadjust pre{\hypertarget{ref-morin2014}{}}%
\CSLLeftMargin{21. }%
\CSLRightInline{Morin X, Fahse L, Mazancourt C de, Scherer-Lorenzen M,
Bugmann H (2014) \href{https://doi.org/10.1111/ele.12357}{Temporal
stability in forest productivity increases with tree diversity due to
asynchrony in species dynamics}. \emph{Ecology Letters}
17(12):1526--1535.}

\leavevmode\vadjust pre{\hypertarget{ref-hillebrand2009}{}}%
\CSLLeftMargin{22. }%
\CSLRightInline{Hillebrand H, Matthiessen B (2009)
\href{https://doi.org/10.1111/j.1461-0248.2009.01388.x}{Biodiversity in
a complex world: consolidation and progress in functional biodiversity
research: Consolidation and progress in BDEF research}. \emph{Ecology
Letters} 12(12):1405--1419.}

\leavevmode\vadjust pre{\hypertarget{ref-cardinale2012}{}}%
\CSLLeftMargin{23. }%
\CSLRightInline{Cardinale BJ, et al. (2012)
\href{https://doi.org/10.1038/nature11148}{Biodiversity loss and its
impact on humanity}. \emph{Nature} 486(7401):59--67.}

\leavevmode\vadjust pre{\hypertarget{ref-dirzo2014}{}}%
\CSLLeftMargin{24. }%
\CSLRightInline{Dirzo R, et al. (2014)
\href{https://doi.org/10.1126/science.1251817}{Defaunation in the
Anthropocene}. \emph{Science} 345(6195):401--406.}

\leavevmode\vadjust pre{\hypertarget{ref-young2016}{}}%
\CSLLeftMargin{25. }%
\CSLRightInline{Young HS, McCauley DJ, Galetti M, Dirzo R (2016)
\href{https://doi.org/10.1146/annurev-ecolsys-112414-054142}{Patterns,
causes, and consequences of anthropocene defaunation}. \emph{Annual
Review of Ecology, Evolution, and Systematics} 47(1):333--358.}

\leavevmode\vadjust pre{\hypertarget{ref-daz2001}{}}%
\CSLLeftMargin{26. }%
\CSLRightInline{Dıáz S, Cabido M (2001)
\href{https://doi.org/10.1016/S0169-5347(01)02283-2}{Vive la différence:
Plant functional diversity matters to ecosystem processes}. \emph{Trends
in Ecology \& Evolution} 16(11):646--655.}

\leavevmode\vadjust pre{\hypertarget{ref-petchey2009}{}}%
\CSLLeftMargin{27. }%
\CSLRightInline{Petchey OL, Gaston KJ (2009)
\href{https://doi.org/10.1007/s12080-009-0041-9}{Effects on ecosystem
resilience of biodiversity, extinctions, and the structure of regional
species pools}. \emph{Theoretical Ecology} 2(3):177--187.}

\leavevmode\vadjust pre{\hypertarget{ref-gallagher2013}{}}%
\CSLLeftMargin{28. }%
\CSLRightInline{Gallagher RV, Hughes L, Leishman MR (2013)
\href{https://doi.org/10.1111/j.1600-0587.2012.07514.x}{Species loss and
gain in communities under future climate change: consequences for
functional diversity}. \emph{Ecography} 36(5):531--540.}

\leavevmode\vadjust pre{\hypertarget{ref-petchey2002}{}}%
\CSLLeftMargin{29. }%
\CSLRightInline{Petchey OL, Gaston KJ (2002)
\href{https://doi.org/10.1098/rspb.2002.2073}{Extinction and the loss of
functional diversity}. \emph{Proceedings of the Royal Society of London
Series B: Biological Sciences} 269(1501):1721--1727.}

\leavevmode\vadjust pre{\hypertarget{ref-pimiento2020}{}}%
\CSLLeftMargin{30. }%
\CSLRightInline{Pimiento C, et al. (2020)
\href{https://doi.org/10.1126/sciadv.aay7650}{Functional diversity of
marine megafauna in the Anthropocene}. \emph{Science Advances}
6(16):eaay7650.}

\leavevmode\vadjust pre{\hypertarget{ref-flynn2009}{}}%
\CSLLeftMargin{31. }%
\CSLRightInline{Flynn DFB, et al. (2009)
\href{https://doi.org/10.1111/j.1461-0248.2008.01255.x}{Loss of
functional diversity under land use intensification across multiple
taxa}. \emph{Ecology Letters} 12(1):22--33.}

\leavevmode\vadjust pre{\hypertarget{ref-edwards2013}{}}%
\CSLLeftMargin{32. }%
\CSLRightInline{Edwards FA, Edwards DP, Hamer KC, Davies RG (2013)
\href{https://doi.org/10.1111/ibi.12027}{Impacts of logging and
conversion of rainforest to oil palm on the functional diversity of
birds in Sundaland}. \emph{Ibis} 155(2):313--326.}

\leavevmode\vadjust pre{\hypertarget{ref-larsen2018}{}}%
\CSLLeftMargin{33. }%
\CSLRightInline{Larsen S, Chase JM, Durance I, Ormerod SJ (2018)
\href{https://doi.org/10.1002/ecy.2213}{Lifting the veil: richness
measurements fail to detect systematic biodiversity change over three
decades}. \emph{Ecology} 99(6):1316--1326.}

\leavevmode\vadjust pre{\hypertarget{ref-matuoka2020}{}}%
\CSLLeftMargin{34. }%
\CSLRightInline{Matuoka MA, Benchimol M, Almeida-Rocha JM de,
Morante-Filho JC (2020)
\href{https://doi.org/10.1016/j.ecolind.2020.106471}{Effects of
anthropogenic disturbances on bird functional diversity: A global
meta-analysis}. \emph{Ecological Indicators} 116:106471.}

\leavevmode\vadjust pre{\hypertarget{ref-jackson2015}{}}%
\CSLLeftMargin{35. }%
\CSLRightInline{Jackson ST, Blois JL (2015)
\href{https://doi.org/10.1073/pnas.1403664111}{Community ecology in a
changing environment: Perspectives from the Quaternary}.
\emph{Proceedings of the National Academy of Sciences}
112(16):4915--4921.}

\leavevmode\vadjust pre{\hypertarget{ref-dornelas2018}{}}%
\CSLLeftMargin{36. }%
\CSLRightInline{Dornelas M, et al. (2018)
\href{https://doi.org/10.1111/geb.12729}{BioTIME: A database of
biodiversity time series for the Anthropocene}. \emph{Global Ecology and
Biogeography} 27(7):760--786.}

\leavevmode\vadjust pre{\hypertarget{ref-pimiento2017}{}}%
\CSLLeftMargin{37. }%
\CSLRightInline{Pimiento C, et al. (2017)
\href{https://doi.org/10.1038/s41559-017-0223-6}{The Pliocene marine
megafauna extinction and its impact on functional diversity}.
\emph{Nature Ecology \& Evolution} 1(8):1100--1106.}

\leavevmode\vadjust pre{\hypertarget{ref-hedberg2022}{}}%
\CSLLeftMargin{38. }%
\CSLRightInline{Hedberg CP, Lyons SK, Smith FA (2022)
\href{https://doi.org/10.1111/geb.13428}{The hidden legacy of megafaunal
extinction: Loss of functional diversity and resilience over the Late
Quaternary at Hall{'}s Cave}. \emph{Global Ecology and Biogeography}
31(2):294--307.}

\leavevmode\vadjust pre{\hypertarget{ref-estes2011}{}}%
\CSLLeftMargin{39. }%
\CSLRightInline{Estes JA, et al. (2011)
\href{https://doi.org/10.1126/science.1205106}{Trophic Downgrading of
Planet Earth}. \emph{Science} 333(6040):301--306.}

\leavevmode\vadjust pre{\hypertarget{ref-smith2018}{}}%
\CSLLeftMargin{40. }%
\CSLRightInline{Smith FA, Elliott Smith RE, Lyons SK, Payne JL (2018)
\href{https://doi.org/10.1126/science.aao5987}{Body size downgrading of
mammals over the late quaternary}. \emph{Science} 360(6386):310--313.}

\leavevmode\vadjust pre{\hypertarget{ref-trindade-santos2020}{}}%
\CSLLeftMargin{41. }%
\CSLRightInline{Trindade-Santos I, Moyes F, Magurran AE (2020)
\href{https://doi.org/10.1098/rspb.2020.0889}{Global change in the
functional diversity of marine fisheries exploitation over the past 65
years}. \emph{Proceedings of the Royal Society B: Biological Sciences}
287(1933):20200889.}

\leavevmode\vadjust pre{\hypertarget{ref-jarzyna2016}{}}%
\CSLLeftMargin{42. }%
\CSLRightInline{Jarzyna MA, Jetz W (2016)
\href{https://doi.org/10.1111/gcb.13571}{A near half-century of temporal
change in different facets of avian diversity}. \emph{Global Change
Biology} 23(8):2999--3011.}

\leavevmode\vadjust pre{\hypertarget{ref-barnagaud2017}{}}%
\CSLLeftMargin{43. }%
\CSLRightInline{Barnagaud J-Y, Gaüzère P, Zuckerberg B, Princé K,
Svenning J-C (2017)
\href{https://doi.org/10.1007/s00442-017-3967-4}{Temporal changes in
bird functional diversity across the United States}. \emph{Oecologia}
185(4):737--748.}

\leavevmode\vadjust pre{\hypertarget{ref-duxedaz2022}{}}%
\CSLLeftMargin{44. }%
\CSLRightInline{Díaz S, Malhi Y (2022)
\href{https://doi.org/10.1146/annurev-environ-120120-054300}{Biodiversity:
Concepts, patterns, trends, and perspectives}. \emph{Annual Review of
Environment and Resources} 47(1):31--63.}

\leavevmode\vadjust pre{\hypertarget{ref-fitzjohn2014}{}}%
\CSLLeftMargin{45. }%
\CSLRightInline{FitzJohn RG, et al. (2014)
\href{https://doi.org/10.1111/1365-2745.12260}{How much of the world is
woody?} \emph{Journal of Ecology} 102(5):1266--1272.}

\leavevmode\vadjust pre{\hypertarget{ref-wilman2014}{}}%
\CSLLeftMargin{46. }%
\CSLRightInline{Wilman H, et al. (2014)
\href{https://doi.org/10.1890/13-1917.1}{EltonTraits 1.0: Species-level
foraging attributes of the world's birds and mammals: {\emph{Ecological
Archives}} E095-178}. \emph{Ecology} 95(7):2027--2027.}

\leavevmode\vadjust pre{\hypertarget{ref-blowes2019}{}}%
\CSLLeftMargin{47. }%
\CSLRightInline{Blowes SA, et al. (2019)
\href{https://doi.org/10.1126/science.aaw1620}{The geography of
biodiversity change in marine and terrestrial assemblages}.
\emph{Science} 366(6463):339--345.}

\leavevmode\vadjust pre{\hypertarget{ref-norman2020}{}}%
\CSLLeftMargin{48. }%
\CSLRightInline{Norman KEA, Chamberlain S, Boettiger C (2020)
\href{https://doi.org/10.1111/2041-210X.13440}{taxadb: A
high-performance local taxonomic database interface}. \emph{Methods in
Ecology and Evolution} 11(9):1153--1159.}

\leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2021}{}}%
\CSLLeftMargin{49. }%
\CSLRightInline{R Core Team (2021) \emph{R: A language and environment
for statistical computing} (R Foundation for Statistical Computing,
Vienna, Austria) Available at: \url{https://www.R-project.org/}.}

\leavevmode\vadjust pre{\hypertarget{ref-lalibertuxe92010}{}}%
\CSLLeftMargin{50. }%
\CSLRightInline{Laliberté E, Legendre P (2010)
\href{https://doi.org/10.1890/08-2244.1}{A distance-based framework for
measuring functional diversity from multiple traits}. \emph{Ecology}
91(1):299--305.}

\leavevmode\vadjust pre{\hypertarget{ref-mason2005}{}}%
\CSLLeftMargin{51. }%
\CSLRightInline{Mason NWH, Mouillot D, Lee WG, Wilson JB, Setälä H
(2005) \href{http://www.jstor.org/stable/3548774}{Functional richness,
functional evenness and functional divergence: The primary components of
functional diversity}. \emph{Oikos} 111(1):112--118.}

\leavevmode\vadjust pre{\hypertarget{ref-leps2006}{}}%
\CSLLeftMargin{52. }%
\CSLRightInline{Leps J, Bello F, Lavorel S, Berman S (2006) Quantifying
and interpreting functional diversity of natural communities: Practical
considerations matter. \emph{Preslia} 78:481--501.}

\leavevmode\vadjust pre{\hypertarget{ref-schleuter2010}{}}%
\CSLLeftMargin{53. }%
\CSLRightInline{Schleuter D, Daufresne M, Massol F, Argillier C (2010)
\href{https://doi.org/10.1890/08-2225.1}{A user's guide to functional
diversity indices}. \emph{Ecological Monographs} 80(3):469--484.}

\leavevmode\vadjust pre{\hypertarget{ref-blonder2018}{}}%
\CSLLeftMargin{54. }%
\CSLRightInline{Blonder B (2018)
\href{https://doi.org/10.1111/ecog.03187}{Hypervolume concepts in niche-
and trait-based ecology}. \emph{Ecography} 41(9):1441--1455.}

\leavevmode\vadjust pre{\hypertarget{ref-swenson2012}{}}%
\CSLLeftMargin{55. }%
\CSLRightInline{Swenson NG, et al. (2012)
\href{https://doi.org/10.1111/j.1466-8238.2011.00727.x}{The biogeography
and filtering of woody plant functional diversity in North and South
America}. \emph{Global Ecology and Biogeography} 21(8):798--808.}

\leavevmode\vadjust pre{\hypertarget{ref-lenth2022}{}}%
\CSLLeftMargin{56. }%
\CSLRightInline{Lenth RV (2022) \emph{Emmeans: Estimated marginal means,
aka least-squares means} Available at:
\url{https://github.com/rvlenth/emmeans}.}

\leavevmode\vadjust pre{\hypertarget{ref-bates2015}{}}%
\CSLLeftMargin{57. }%
\CSLRightInline{Bates D, Mächler M, Bolker B, Walker S (2015) Fitting
Linear Mixed-Effects Models Using {\textbf{lme4}}. \emph{Journal of
Statistical Software} 67(1).
doi:\href{https://doi.org/10.18637/jss.v067.i01}{10.18637/jss.v067.i01}.}

\leavevmode\vadjust pre{\hypertarget{ref-kuznetsova2017}{}}%
\CSLLeftMargin{58. }%
\CSLRightInline{Kuznetsova A, Brockhoff PB, Christensen RHB (2017)
\href{https://doi.org/10.18637/jss.v082.i13}{lmerTest package: Tests in
linear mixed effects models}. \emph{Journal of Statistical Software}
82(13):126.}

\leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2023}{}}%
\CSLLeftMargin{59. }%
\CSLRightInline{R Core Team (2023) \emph{R: A language and environment
for statistical computing} (R Foundation for Statistical Computing,
Vienna, Austria) Available at: \url{https://www.R-project.org/}.}

\leavevmode\vadjust pre{\hypertarget{ref-brodie2021}{}}%
\CSLLeftMargin{60. }%
\CSLRightInline{Brodie JF, Williams S, Garner B (2021) The decline of
mammal functional and evolutionary diversity worldwide.
\emph{Proceedings of the National Academy of Sciences} 118(3).
doi:\href{https://doi.org/10.1073/pnas.1921849118}{10.1073/pnas.1921849118}.}

\leavevmode\vadjust pre{\hypertarget{ref-cox2022}{}}%
\CSLLeftMargin{61. }%
\CSLRightInline{Cox DTC, Gardner AS, Gaston KJ (2022)
\href{https://doi.org/10.1126/sciadv.abn6008}{Global and regional
erosion of mammalian functional diversity across the diel cycle}.
\emph{Science Advances} 8(32):eabn6008.}

\leavevmode\vadjust pre{\hypertarget{ref-li2022}{}}%
\CSLLeftMargin{62. }%
\CSLRightInline{Li X, et al. (2022)
\href{https://doi.org/10.1111/cobi.13839}{Functional diversity loss and
change in nocturnal behavior of mammals under anthropogenic
disturbance}. \emph{Conservation Biology} 36(3):e13839.}

\leavevmode\vadjust pre{\hypertarget{ref-sherry2021}{}}%
\CSLLeftMargin{63. }%
\CSLRightInline{Sherry TW (2021) Sensitivity of tropical insectivorous
birds to the anthropocene: A review of multiple mechanisms and
conservation implications. \emph{Frontiers in Ecology and Evolution} 9.
Available at:
\url{https://www.frontiersin.org/articles/10.3389/fevo.2021.662873}.}

\leavevmode\vadjust pre{\hypertarget{ref-inger2015}{}}%
\CSLLeftMargin{64. }%
\CSLRightInline{Inger R, et al. (2015)
\href{https://doi.org/10.1111/ele.12387}{Common European birds are
declining rapidly while less abundant species' numbers are rising}.
\emph{Ecology Letters} 18(1):28--36.}

\leavevmode\vadjust pre{\hypertarget{ref-rosenberg2019}{}}%
\CSLLeftMargin{65. }%
\CSLRightInline{Rosenberg KV, et al. (2019) Decline of the North
American avifauna. \emph{Science}. Available at:
\url{https://www.science.org/doi/abs/10.1126/science.aaw1313}.}

\leavevmode\vadjust pre{\hypertarget{ref-schipper2016}{}}%
\CSLLeftMargin{66. }%
\CSLRightInline{Schipper AM, et al. (2016)
\href{https://doi.org/10.1111/gcb.13292}{Contrasting changes in the
abundance and diversity of North American bird assemblages from 1971 to
2010}. \emph{Global Change Biology} 22(12):3948--3959.}

\leavevmode\vadjust pre{\hypertarget{ref-sheridan2011}{}}%
\CSLLeftMargin{67. }%
\CSLRightInline{Sheridan JA, Bickford D (2011)
\href{https://doi.org/10.1038/nclimate1259}{Shrinking body size as an
ecological response to climate change}. \emph{Nature Climate Change}
1(8):401--406.}

\leavevmode\vadjust pre{\hypertarget{ref-caruso2014}{}}%
\CSLLeftMargin{68. }%
\CSLRightInline{Caruso NM, Sears MW, Adams DC, Lips KR (2014)
\href{https://doi.org/10.1111/gcb.12550}{Widespread rapid reductions in
body size of adult salamanders in response to climate change}.
\emph{Global Change Biology} 20(6):1751--1759.}

\leavevmode\vadjust pre{\hypertarget{ref-forster2012}{}}%
\CSLLeftMargin{69. }%
\CSLRightInline{Forster J, Hirst AG, Atkinson D (2012)
\href{https://doi.org/10.1073/pnas.1210460109}{Warming-induced
reductions in body size are greater in aquatic than terrestrial
species}. \emph{Proceedings of the National Academy of Sciences}
109(47):19310--19314.}

\leavevmode\vadjust pre{\hypertarget{ref-huss2019}{}}%
\CSLLeftMargin{70. }%
\CSLRightInline{Huss M, Lindmark M, Jacobson P, van Dorst RM, Gårdmark A
(2019) \href{https://doi.org/10.1111/gcb.14637}{Experimental evidence of
gradual size-dependent shifts in body size and growth of fish in
response to warming}. \emph{Global Change Biology} 25(7):2285--2295.}

\leavevmode\vadjust pre{\hypertarget{ref-tseng2018}{}}%
\CSLLeftMargin{71. }%
\CSLRightInline{Tseng M, et al. (2018)
\href{https://doi.org/10.1111/1365-2656.12789}{Decreases in beetle body
size linked to climate change and warming temperatures}. \emph{Journal
of Animal Ecology} 87(3):647--659.}

\leavevmode\vadjust pre{\hypertarget{ref-fritz2009}{}}%
\CSLLeftMargin{72. }%
\CSLRightInline{Fritz SA, Bininda-Emonds ORP, Purvis A (2009)
\href{https://doi.org/10.1111/j.1461-0248.2009.01307.x}{Geographical
variation in predictors of mammalian extinction risk: big is bad, but
only in the tropics}. \emph{Ecology Letters} 12(6):538--549.}

\leavevmode\vadjust pre{\hypertarget{ref-ripple2015}{}}%
\CSLLeftMargin{73. }%
\CSLRightInline{Ripple WJ, et al. (2015)
\href{https://doi.org/10.1126/sciadv.1400103}{Collapse of the world{'}s
largest herbivores}. \emph{Science Advances} 1(4):e1400103.}

\leavevmode\vadjust pre{\hypertarget{ref-duffy2007}{}}%
\CSLLeftMargin{74. }%
\CSLRightInline{Duffy JE, et al. (2007)
\href{https://doi.org/10.1111/j.1461-0248.2007.01037.x}{The functional
role of biodiversity in ecosystems: incorporating trophic complexity}.
\emph{Ecology Letters} 10(6):522--538.}

\leavevmode\vadjust pre{\hypertarget{ref-bagousse-pinguet2021}{}}%
\CSLLeftMargin{75. }%
\CSLRightInline{Bagousse-Pinguet YL, et al. (2021) Functional rarity and
evenness are key facets of biodiversity to boost multifunctionality.
\emph{Proceedings of the National Academy of Sciences} 118(7).
doi:\href{https://doi.org/10.1073/pnas.2019355118}{10.1073/pnas.2019355118}.}

\leavevmode\vadjust pre{\hypertarget{ref-wauchope2019}{}}%
\CSLLeftMargin{76. }%
\CSLRightInline{Wauchope HS, Amano T, Sutherland WJ, Johnston A (2019)
\href{https://doi.org/10.1111/2041-210X.13302}{When can we trust
population trends? A method for quantifying the effects of sampling
interval and duration}. \emph{Methods in Ecology and Evolution}
10(12):2067--2078.}

\leavevmode\vadjust pre{\hypertarget{ref-hughes2021}{}}%
\CSLLeftMargin{77. }%
\CSLRightInline{Hughes AC, et al. (2021)
\href{https://doi.org/10.1111/ecog.05926}{Sampling biases shape our view
of the natural world}. \emph{Ecography} 44(9):1259--1269.}

\leavevmode\vadjust pre{\hypertarget{ref-desroches2018}{}}%
\CSLLeftMargin{78. }%
\CSLRightInline{Des Roches S, et al. (2018)
\href{https://doi.org/10.1038/s41559-017-0402-5}{The ecological
importance of intraspecific variation}. \emph{Nature Ecology \&
Evolution} 2(1):57--64.}

\leavevmode\vadjust pre{\hypertarget{ref-violle2012}{}}%
\CSLLeftMargin{79. }%
\CSLRightInline{Violle C, et al. (2012)
\href{https://doi.org/10.1016/j.tree.2011.11.014}{The return of the
variance: Intraspecific variability in community ecology}. \emph{Trends
in Ecology \& Evolution} 27(4):244--252.}

\leavevmode\vadjust pre{\hypertarget{ref-barros2016}{}}%
\CSLLeftMargin{80. }%
\CSLRightInline{Barros C, Thuiller W, Georges D, Boulangeat I,
Münkemüller T (2016)
\href{https://doi.org/10.1111/ele.12617}{N-dimensional hypervolumes to
study stability of complex ecosystems}. \emph{Ecology Letters}
19(7):729--742.}

\leavevmode\vadjust pre{\hypertarget{ref-mammola2019}{}}%
\CSLLeftMargin{81. }%
\CSLRightInline{Mammola S (2019)
\href{https://doi.org/10.1111/jbi.13618}{Assessing similarity of
n-dimensional hypervolumes: Which metric to use?} \emph{Journal of
Biogeography} 46(9):2012--2023.}

\leavevmode\vadjust pre{\hypertarget{ref-magioli2021}{}}%
\CSLLeftMargin{82. }%
\CSLRightInline{Magioli M, et al. (2021)
\href{https://doi.org/10.1016/j.pecon.2021.02.006}{Land-use changes lead
to functional loss of terrestrial mammals in a Neotropical rainforest}.
\emph{Perspectives in Ecology and Conservation} 19(2):161--170.}

\leavevmode\vadjust pre{\hypertarget{ref-tinoco2018}{}}%
\CSLLeftMargin{83. }%
\CSLRightInline{Tinoco BA, Santillán VE, Graham CH (2018)
\href{https://doi.org/10.1002/ece3.3813}{Land use change has stronger
effects on functional diversity than taxonomic diversity in tropical
Andean hummingbirds}. \emph{Ecology and Evolution} 8(6):3478--3490.}

\leavevmode\vadjust pre{\hypertarget{ref-zhu2017}{}}%
\CSLLeftMargin{84. }%
\CSLRightInline{Zhu L, et al. (2017)
\href{https://doi.org/10.1038/s41598-017-03812-8}{Trait choice
profoundly affected the ecological conclusions drawn from functional
diversity measures}. \emph{Scientific Reports} 7(1):3643.}

\leavevmode\vadjust pre{\hypertarget{ref-pardieck2020}{}}%
\CSLLeftMargin{85. }%
\CSLRightInline{Pardieck KL, David ZJ, Lutmerding M, Aponte V, Hudson
M-AR (2020) North american breeding bird survey dataset 1966 - 2019,
version 2019.0.
doi:\href{https://doi.org/10.5066/P9J6QUF6}{10.5066/P9J6QUF6}.}

\leavevmode\vadjust pre{\hypertarget{ref-piropno}{}}%
\CSLLeftMargin{86. }%
\CSLRightInline{PIROP northwest atlantic 1965{\textendash}1992 - OBIS
SEAMAP Available at: \url{http://www.iobis.org/mapper/?dataset=2245}.}

\leavevmode\vadjust pre{\hypertarget{ref-silva}{}}%
\CSLLeftMargin{87. }%
\CSLRightInline{Silva FR da Brazil dataset 1.}

\leavevmode\vadjust pre{\hypertarget{ref-scott}{}}%
\CSLLeftMargin{88. }%
\CSLRightInline{Scott D, Metts B, Lance S The rainbow bay long-term
study. Available at: \url{http://srelherp.uga.edu/projects/rbay.htm}.}

\leavevmode\vadjust pre{\hypertarget{ref-rossaferes1997}{}}%
\CSLLeftMargin{89. }%
\CSLRightInline{Rossa-Feres D de C (1997) Community ecology of anura
amphibia at northwest region of sao paulo state, brazil: Microhabitat,
seasonality, diet and multidimensional niche. PhD thesis.}

\leavevmode\vadjust pre{\hypertarget{ref-bakker1990}{}}%
\CSLLeftMargin{90. }%
\CSLRightInline{Bakker C, Herman PMJ (1990) Phytoplankton in the
oosterschelde before, during and after the storm-surge barrier
(1982{\textendash}1990). Available at:
\url{http://www.iobis.org/mapper/?dataset=505}.}

\leavevmode\vadjust pre{\hypertarget{ref-berezovikovn.n.2004}{}}%
\CSLLeftMargin{91. }%
\CSLRightInline{Berezovikov, N.N. (2004) The birds of settlements in
markakol depression (southern altai). (249):3--15.}

\leavevmode\vadjust pre{\hypertarget{ref-enemar2004}{}}%
\CSLLeftMargin{92. }%
\CSLRightInline{Enemar A, Sjöstrand B, Andersson G, Von Proschwitz T
(2004) \href{https://doi.org/10.34080/os.v14.20236}{The 37-year dynamics
of a subalpine passerine bird community, with special emphasis on the
influence of environmental temperature and epirrita autumnata cycles}.
\emph{Ornis Svecica} 14(3):63--106.}

\leavevmode\vadjust pre{\hypertarget{ref-gido2019}{}}%
\CSLLeftMargin{93. }%
\CSLRightInline{Gido K (2019) CFP01 Fish population on selected
watersheds at Konza Prairie.
doi:\href{https://doi.org/10.6073/PASTA/BE5AD393AF83F9602AAE96423A280875}{10.6073/PASTA/BE5AD393AF83F9602AAE96423A280875}.}

\leavevmode\vadjust pre{\hypertarget{ref-hogstad1993}{}}%
\CSLLeftMargin{94. }%
\CSLRightInline{Hogstad O (1993)
\href{https://www.jstor.org/stable/23735355}{Structure and dynamics of a
passerine bird community in a spruce-dominated boreal forest. A 12-year
study}. \emph{Annales Zoologici Fennici} 30(1):43--54.}

\leavevmode\vadjust pre{\hypertarget{ref-holmes1986}{}}%
\CSLLeftMargin{95. }%
\CSLRightInline{Holmes RT, Sherry TW, Sturges FW (1986)
\href{https://doi.org/10.2307/2937074}{Bird Community Dynamics in a
Temperate Deciduous Forest: Long-Term Trends at Hubbard Brook}.
\emph{Ecological Monographs} 56(3):201--220.}

\leavevmode\vadjust pre{\hypertarget{ref-jahncke2006}{}}%
\CSLLeftMargin{96. }%
\CSLRightInline{Jahncke J, Rintoul C (2006) CalCOFI and NMFS seabird and
marine mammal observation data, 1987{\textendash}2006. Available at:
\url{http://www.iobis.org}.}

\leavevmode\vadjust pre{\hypertarget{ref-krivenkov.g.1991}{}}%
\CSLLeftMargin{97. }%
\CSLRightInline{Krivenko, V.G. (1991) Waterfowl and their protection.}

\leavevmode\vadjust pre{\hypertarget{ref-melnikovy.i.2000}{}}%
\CSLLeftMargin{98. }%
\CSLRightInline{Melnikov, Y.I., Melnikova, N., Pronkevich, V.V. (2000)
Migration of birds of prey in the mouth of the river irkut.
(108):3--17.}

\leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2013}{}}%
\CSLLeftMargin{99. }%
\CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
Ministry of Environment of Japan (2013) Monitoring site 1000 shorebird
survey. Available at:
\url{http://www.biodic.go.jp/moni1000/findings/data/index.html}.}

\leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2014}{}}%
\CSLLeftMargin{100. }%
\CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
Ministry of Environment of Japan (2014) Monitoring site 1000 village
survey - bird survey data (2005{\textendash}2012). Available at:
\url{http://www.biodic.go.jp/moni1000/findings/data/index.html)}.}

\leavevmode\vadjust pre{\hypertarget{ref-monitori2014}{}}%
\CSLLeftMargin{101. }%
\CSLRightInline{Monitoring site 1000 village survey - medium and large
mammal survey data (2006{\textendash}2012) (2014) Available at:
\url{http://www.biodic.go.jp/moni1000/findings/data/index.html}.}

\leavevmode\vadjust pre{\hypertarget{ref-preston1960}{}}%
\CSLLeftMargin{102. }%
\CSLRightInline{Preston FW (1960)
\href{https://doi.org/10.2307/1931793}{Time and space and the variation
of species}. \emph{Ecology} 41(4):612--627.}

\leavevmode\vadjust pre{\hypertarget{ref-svensson2006}{}}%
\CSLLeftMargin{103. }%
\CSLRightInline{Svensson S (2006) Species composition and population
fluctuations of alpine bird communities during 38 years in the
scandinavian mountain range. \emph{Ornis Svecica} 16:183--210.}

\leavevmode\vadjust pre{\hypertarget{ref-svensson2010}{}}%
\CSLLeftMargin{104. }%
\CSLRightInline{Svensson S, Thorner AM, Nyholm NEI (2010) Species
trends, turnover and composition of a woodland bird community in
southern sweden during a period of fifty-seven years. \emph{Ornis
Svecica} 20(1).
doi:\href{https://doi.org/10.34080/os.v20.22641}{10.34080/os.v20.22641}.}

\leavevmode\vadjust pre{\hypertarget{ref-thorn2016changes}{}}%
\CSLLeftMargin{105. }%
\CSLRightInline{Thorn S, et al. (2016)
\href{https://doi.org/10.1111/ele.12548}{Changes in the dominant
assembly mechanism drive species loss caused by declining resources}.
\emph{Ecology Letters} 19(2):163--170.}

\leavevmode\vadjust pre{\hypertarget{ref-usfs}{}}%
\CSLLeftMargin{106. }%
\CSLRightInline{USFS Landbird monitoring program (UMT-LBMP). Available
at: \url{http://www.avianknowledge.net/}.}

\leavevmode\vadjust pre{\hypertarget{ref-vermontcenterforecostudies2015}{}}%
\CSLLeftMargin{107. }%
\CSLRightInline{Vermont Center For Ecostudies, Lambert JD, Hart J (2015)
Mountain Birdwatch 1.0.
doi:\href{https://doi.org/10.5063/F1DN430G}{10.5063/F1DN430G}.}

\leavevmode\vadjust pre{\hypertarget{ref-vickery1984}{}}%
\CSLLeftMargin{108. }%
\CSLRightInline{Vickery WL, Nudds TD (1984)
\href{https://doi.org/10.2307/1939462}{Detection of Density-Dependent
Effects in Annual Duck Censuses}. \emph{Ecology} 65(1):96--104.}

\leavevmode\vadjust pre{\hypertarget{ref-waide2017}{}}%
\CSLLeftMargin{109. }%
\CSLRightInline{Waide R (2017) Bird abundance - point counts.
doi:\href{https://doi.org/10.6073/PASTA/91E6302E743BAC1E3E32781B869CE3D9}{10.6073/PASTA/91E6302E743BAC1E3E32781B869CE3D9}.}

\leavevmode\vadjust pre{\hypertarget{ref-williamson1983}{}}%
\CSLLeftMargin{110. }%
\CSLRightInline{Williamson M (1983)
\href{https://doi.org/10.2307/3544096}{The land-bird community of
skokholm: Ordination and turnover}. \emph{Oikos} 41(3):378--384.}

\leavevmode\vadjust pre{\hypertarget{ref-zakharovv.d.1998}{}}%
\CSLLeftMargin{111. }%
\CSLRightInline{Zakharov, V.D. (1998) Biodiversity of bird population of
terrestrial habitats in southern ural. Miass: IGZ. Available at:
\url{http://ashipunov.info/shipunov/school/books/zakharov1998_biorazn_nasel_ptits_mazemn_mestoob_juzhn_urala.pdf}.}

\leavevmode\vadjust pre{\hypertarget{ref-meditss2011}{}}%
\CSLLeftMargin{112. }%
\CSLRightInline{MEDITS seabird surveys 1999 / 2000 / 2002 (2011)
Available at:
\url{http://www.emodnet-biology.eu/component/imis/?module=dataset\&dasid=1979}.}

\leavevmode\vadjust pre{\hypertarget{ref-animald}{}}%
\CSLLeftMargin{113. }%
\CSLRightInline{Animal demography unit - coordinated waterbird counts
(CWAC) - AfrOBIS Available at: \url{http://www.iobis.org/}.}

\leavevmode\vadjust pre{\hypertarget{ref-baltics}{}}%
\CSLLeftMargin{114. }%
\CSLRightInline{Baltic seabirds transect surveys Available at:
\url{http://www.emodnet-biology.eu/component/imis/?module=dataset\&dasid=1971}.}

\leavevmode\vadjust pre{\hypertarget{ref-thorn2016response}{}}%
\CSLLeftMargin{115. }%
\CSLRightInline{Thorn S, et al. (2016)
\href{https://doi.org/10.1016/j.ecolind.2015.06.033}{Response of bird
assemblages to windstorm and salvage logging {\textemdash} Insights from
analyses of functional guild and indicator species}. \emph{Ecological
Indicators} 65:142--148.}

\leavevmode\vadjust pre{\hypertarget{ref-ernest2009}{}}%
\CSLLeftMargin{116. }%
\CSLRightInline{Ernest SKM, Valone TJ, Brown JH (2009)
\href{https://doi.org/10.1890/08-1222.1}{Long-term monitoring and
experimental manipulation of a Chihuahuan Desert ecosystem near Portal,
Arizona, USA: {\emph{Ecological Archives}} E090-118}. \emph{Ecology}
90(6):1708--1708.}

\leavevmode\vadjust pre{\hypertarget{ref-friggens2008}{}}%
\CSLLeftMargin{117. }%
\CSLRightInline{Friggens M (2008) Sevilleta LTER small mammal population
data. Available at: \url{http://sev.lternet.edu/data/sev-8}.}

\leavevmode\vadjust pre{\hypertarget{ref-jalilova.b.2014}{}}%
\CSLLeftMargin{118. }%
\CSLRightInline{Jalilov, A. B., Andreychev, A. V., Kuznetsov, V. A.
(2014) Monitoring and conservation of medium and large mammals in
chamzinsky district of the republic of mordovia.}

\leavevmode\vadjust pre{\hypertarget{ref-kartzinel2014}{}}%
\CSLLeftMargin{119. }%
\CSLRightInline{Kartzinel TR, et al. (2014)
\href{https://doi.org/10.1890/13-1023R.1}{Plant and small-mammal
responses to large-herbivore exclusion in an African savanna: five years
of the UHURU experiment: {\emph{Ecological Archives}} E095-064}.
\emph{Ecology} 95(3):787--787.}

\leavevmode\vadjust pre{\hypertarget{ref-kaufman2019}{}}%
\CSLLeftMargin{120. }%
\CSLRightInline{Kaufman D (2019) CSM01 Seasonal Summary of Numbers of
Small Mammals on 14 LTER Traplines in Prairie Habitats at Konza Prairie.
doi:\href{https://doi.org/10.6073/PASTA/9735A16A0018D85FF5EFB8B74FD100F4}{10.6073/PASTA/9735A16A0018D85FF5EFB8B74FD100F4}.}

\leavevmode\vadjust pre{\hypertarget{ref-krefting1974}{}}%
\CSLLeftMargin{121. }%
\CSLRightInline{Krefting LW, Ahlgren CE (1974)
\href{https://doi.org/10.2307/1935467}{Small Mammals and Vegetation
Changes After Fire in a Mixed Conifer-Hardwood Forest}. \emph{Ecology}
55(6):1391--1398.}

\leavevmode\vadjust pre{\hypertarget{ref-lightfoot}{}}%
\CSLLeftMargin{122. }%
\CSLRightInline{Lightfoot D, Schooley RL SMES rodent trapping data,
small mammal exclosure study. Available at:
\href{http://jornada.nmsu.edu/sites/jornada.nmsu.edu/files/data_files/JornadaStudy_086_smes_rodent_trapping_data_0.csv,\%20accessed\%202016.}{http://jornada.nmsu.edu/sites/jornada.nmsu.edu/files/data\_files/JornadaStudy\_086\_smes\_rodent\_trapping\_data\_0.csv,
accessed 2016.}}

\leavevmode\vadjust pre{\hypertarget{ref-malyshevy.s.2011}{}}%
\CSLLeftMargin{123. }%
\CSLRightInline{Malyshev, Y. S. (2011) On the diagnostic techniques of
ranks of the number dynamics cycles of small mammals. 1(6):92--106.}

\leavevmode\vadjust pre{\hypertarget{ref-meyer2008}{}}%
\CSLLeftMargin{124. }%
\CSLRightInline{Meyer CFJ, Kalko EKV (2008)
\href{https://doi.org/10.1111/j.1365-2699.2008.01916.x}{Assemblage-level
responses of phyllostomid bats to tropical forest fragmentation:
land-bridge islands as a model system}. \emph{Journal of Biogeography}
35(9):1711--1726.}

\leavevmode\vadjust pre{\hypertarget{ref-nedosekinv.y.2015}{}}%
\CSLLeftMargin{125. }%
\CSLRightInline{Nedosekin, V. Y. (2015) Long-term dynamics of the
population and the quantity of small mammals under conditions of the
reserve {"}galichya gora{"}.}

\leavevmode\vadjust pre{\hypertarget{ref-prins1990}{}}%
\CSLLeftMargin{126. }%
\CSLRightInline{Prins HHT, Douglas-Hamilton I (1990)
\href{https://doi.org/10.1007/BF00317566}{Stability in a multi-species
assemblage of large herbivores in East Africa}. \emph{Oecologia}
83(3):392--400.}

\leavevmode\vadjust pre{\hypertarget{ref-rocha2017}{}}%
\CSLLeftMargin{127. }%
\CSLRightInline{Rocha R, et al. (2017)
\href{https://doi.org/10.1007/s10980-016-0425-3}{Consequences of a
large-scale fragmentation experiment for Neotropical bats: disentangling
the relative importance of local and landscape-scale effects}.
\emph{Landscape Ecology} 32(1):31--45.}

\leavevmode\vadjust pre{\hypertarget{ref-stapp2014}{}}%
\CSLLeftMargin{128. }%
\CSLRightInline{Stapp P (2014) SGS-LTER long-term monitoring project:
Small mammals on trapping webs on the central plains experimental range,
nunn, colorado, USA 1994 -2006, ARS study number 118.
doi:\href{https://doi.org/10.6073/PASTA/2E311B4E40FEA38E573890F473807BA9}{10.6073/PASTA/2E311B4E40FEA38E573890F473807BA9}.}

\leavevmode\vadjust pre{\hypertarget{ref-bahamas}{}}%
\CSLLeftMargin{129. }%
\CSLRightInline{Bahamas marine mammal research organisation
opportunistic sightings - OBIS SEAMAP Available at:
\url{http://www.iobis.org}.}

\leavevmode\vadjust pre{\hypertarget{ref-popacet}{}}%
\CSLLeftMargin{130. }%
\CSLRightInline{POPA cetacean, seabird, and sea turtle sightings in the
azores area 1998{\textendash}2009 - OBIS SEAMAP Available at:
\url{http://www.iobis.org/mapper/?dataset=4257}.}

\end{CSLReferences}

\bibliographystyle{unsrt}
\bibliography{refs.bib}

\end{document}

### 2023-06-22 paper/supp1_sensitivity_outliers.Rmd
Add amphibian trends and missing individual studies

--- 2022-11-03 paper/sensitivity_supp.Rmd
@@
 ---
-title: "Sensitivity Analysis"
+title: "Supplement 1"
 output:
   pdf_document: default
-  html_document: default
 ---
+
+# Sensitivity Analysis
 
@@
        "CWM seed diet*, CWM canopy foraging*, Jaccard", rep("Jaccard", 3))
-) %>% kable(format = "latex") %>%
+) %>% kable(format = "latex", caption = "Summary of the results for the sensitivity analyses, including those for duration of time series, trait coverage, and amphibian inclusion.") %>%
   column_spec(3:4, width = "3cm") %>%
@@
 ```
+
+# Amphibian Trends
+
+There were 3 amphibian studies comprised of 11 time series in the bioTIME database with sufficient trait coverage to be included in the analysis.
+With such limited data, we excluded amphibians from the main results, but present preliminary trends here.
+Models fit with amphibian, mammal, and bird data combined showed qualitatively similar trend results to those reported in the manuscript that did not include amphibians.
+For models fit with only amphibian data, no general trends were detected for any metric.
+Table 2 gives model estimates for species-based and species corrected metrics.
+
+```{r resultsTab, echo=FALSE, message = FALSE, warning = FALSE}
+readr::read_csv(here::here("figures/amph_model_table.csv")) %>%
+  select(metric, everything(), -REML, -bobyqa) %>%
+  select(-c(n_ts, formula, model)) %>%
+  filter(group != "Residual" | is.na(group), metric %in% c("S", "Jaccard_base", "SES_FRic", "SES_FEve", "SES_FDiv")) %>%
+  mutate(metric = ifelse(metric == "Jaccard_base", "log(Jaccard + 1)", metric),
+         metric = ifelse(metric == "S", "log(Species Richness)", metric),
+         effect = ifelse(effect == "ran_pars", "random", effect),
+         group = case_when(
+           group == "rarefyID:study_id" ~ "time series within study",
+           group == "study_id" ~ "study",
+           group == "Residual" ~ "residual",
+           TRUE ~ group),
+         term = case_when(
+           term == "(Intercept)" ~ "Intercept",
+           term == "year_scaled" ~ "Year",
+           term == "sd__(Intercept)" ~ "SD Intercept",
+           term == "cor__(Intercept).year_scaled" ~ "Corr(Intercept, Year)",
+           term == "sd__year_scaled" ~ "SD Year",
+           term == "sd__Observation" ~ "SD Observation",
+           TRUE ~ term
+         )) %>%
+  mutate(across(c(estimate, std.error, p.value), round, 2),
+         p.value = ifelse(p.value == 0, "<0.001", p.value)) %>%
+  arrange(metric != "Species Richness", metric, !is.na(group), group, term == "Corr(Intercept, Year)", term) %>%
+  dplyr::rename(grouping = group) %>%
+  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics for Amphibian time series.") %>%
+  column_spec(3, width = "2cm") %>%
+  collapse_rows(columns = 1:3, valign = "middle") %>%
+  kable_paper() %>%
+  kable_styling(latex_options= c("scale_down", "hold_position"))
+```
+
+# Missing Individual Study Models
+
+We fit models separately for each study and metric to estimate study-level trends.
+11 models could not be fit because there were to many time series relative to the number of observations, and therefore not enough data to fit the appropriate mixed effects model structure.
+The missing models are in Table 3.
+
+```{r missingTab, echo=FALSE, message = FALSE, warning = FALSE}
+readr::read_csv(here::here("paper/missing_indv_studies.csv")) %>%
+  rename(`n time series` = nrarefyID, `n observations` = n) %>%
+  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics for Amphibian time series.") %>%
+  kable_paper() %>%
+  kable_styling(latex_options= c("scale_down", "hold_position"))
+```

### 2023-06-23 paper/GEB Submission/GEB_title_page.Rmd
Save out tables to create doc by hand.

@@
 ---
-title: Prevailing scenarios of functional change in anthropocene bird and mammal communities
+title: Prevailing scenarios of functional change in Anthropocene bird and mammal communities
 authors:
@@
 
-  Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevelant than the other across many taxa, biomes, and realms.
+  Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevalant than the other across many taxa, biomes, and realms.
 At the study-level, there were also a substantial number of time series exhibiting no species or functional change, however the majority of studies showed a shift in a species or functional metric.

### 2023-06-23 paper/GEB Submission/GEB_title_page.tex
Save out tables to create doc by hand.

@@
 
-\title{Prevailing scenarios of functional change in anthropocene bird
+\title{Prevailing scenarios of functional change in Anthropocene bird
 and mammal communities}
@@
 Main Conclusions: General trends indicate that on the aggregate one type
-of functional shift is not more prevelant than the other across many
+of functional shift is not more prevalant than the other across many
 taxa, biomes, and realms. At the study-level, there were also a
@@
 
+\hypertarget{biosketch}{%
+\section{Biosketch}\label{biosketch}}
+
+Kari Norman's research is generally focused on understanding the impacts
+of anthropogenic change on biodiversity. She brings together existing
+data sources to understand broadscale change and develops software tools
+to facilitate robust synthesis workflows.
+
 \bibliographystyle{unsrt}

### 2023-06-23 paper/GEB Submission/manuscript_GEB.Rmd
Save out tables to create doc by hand.

@@
 ---
-title: Prevailing scenarios of functional change in anthropocene bird and mammal communities
+title: Prevailing scenarios of functional change in Anthropocene bird and mammal communities
 abstract: |
@@
 We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
-We also assessed the impact of potential study-level outliers on general trends.
-Details on that process can be found in the supplement.
 All models in our analysis were fit using the *lme4* (v1.1-30) package in R (v4.2.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].

### 2023-06-23 paper/supp1_sensitivity_outliers.Rmd
Save out tables to create doc by hand.

--- 2022-11-03 paper/sensitivity_supp.Rmd
@@
 ---
-title: "Sensitivity Analysis"
+title: "Supplement 1"
 output:
-  pdf_document: default
-  html_document: default
+  #pdf_document: default
+  word_document: default
 ---
+
+# Sensitivity Analysis
 
@@
        "CWM seed diet*, CWM canopy foraging*, Jaccard", rep("Jaccard", 3))
-) %>% kable(format = "latex") %>%
+) %>% 
+  #kable(format = "latex", caption = "Summary of the results for the sensitivity analyses, including those for duration of time series, trait coverage, and amphibian inclusion.") %>%
+  kable(format = "latex") %>%
   column_spec(3:4, width = "3cm") %>%
@@
   kable_paper() %>%
-  kable_styling(latex_options= c("hold_position"))
+  kable_styling(latex_options= c("hold_position")) %>%
+  save_kable(here::here("figures/tables/sensitivity_table.jpeg"))
 ```
+
+# Amphibian Trends
+
+There were 3 amphibian studies comprised of 11 time series in the bioTIME database with sufficient trait coverage to be included in the analysis.
+With such limited data, we excluded amphibians from the main results, but present preliminary trends here.
+Models fit with amphibian, mammal, and bird data combined showed qualitatively similar trend results to those reported in the manuscript that did not include amphibians.
+For models fit with only amphibian data, no general trends were detected for any metric.
+Table 2 gives model estimates for species-based and species corrected metrics.
+
+```{r resultsTab, echo=FALSE, message = FALSE, warning = FALSE}
+readr::read_csv(here::here("figures/amph_model_table.csv")) %>%
+  select(metric, everything(), -REML, -bobyqa) %>%
+  select(-c(n_ts, formula, model)) %>%
+  filter(group != "Residual" | is.na(group), metric %in% c("S", "Jaccard_base", "SES_FRic", "SES_FEve", "SES_FDiv")) %>%
+  mutate(metric = ifelse(metric == "Jaccard_base", "log(Jaccard + 1)", metric),
+         metric = ifelse(metric == "S", "log(Species Richness)", metric),
+         effect = ifelse(effect == "ran_pars", "random", effect),
+         group = case_when(
+           group == "rarefyID:study_id" ~ "time series within study",
+           group == "study_id" ~ "study",
+           group == "Residual" ~ "residual",
+           TRUE ~ group),
+         term = case_when(
+           term == "(Intercept)" ~ "Intercept",
+           term == "year_scaled" ~ "Year",
+           term == "sd__(Intercept)" ~ "SD Intercept",
+           term == "cor__(Intercept).year_scaled" ~ "Corr(Intercept, Year)",
+           term == "sd__year_scaled" ~ "SD Year",
+           term == "sd__Observation" ~ "SD Observation",
+           TRUE ~ term
+         )) %>%
+  mutate(across(c(estimate, std.error, p.value), round, 2),
+         p.value = ifelse(p.value == 0, "<0.001", p.value)) %>%
+  arrange(metric != "Species Richness", metric, !is.na(group), group, term == "Corr(Intercept, Year)", term) %>%
+  dplyr::rename(grouping = group) %>%
+  #kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics for Amphibian time series.") %>%
+  kable(format = "latex") %>%
+  column_spec(3, width = "2cm") %>%
+  collapse_rows(columns = 1:3, valign = "middle") %>%
+  kable_paper() %>%
+  kable_styling(latex_options= c("scale_down", "hold_position")) %>%
+  save_kable(here::here("figures/tables/amph_table.jpeg"))
+```
+
+# Missing Individual Study Models
+
+We fit models separately for each study and metric to estimate study-level trends.
+11 models could not be fit because there were to many time series relative to the number of observations, and therefore not enough data to fit the appropriate mixed effects model structure.
+The missing models are in Table 3.
+
+```{r missingTab, echo=FALSE, message = FALSE, warning = FALSE}
+readr::read_csv(here::here("paper/missing_indv_studies.csv")) %>%
+  rename(`n time series` = nrarefyID, `n observations` = n) %>%
+  #kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics for Amphibian time series.") %>%
+  kable(format = "latex") %>%
+  kable_paper() %>%
+  kable_styling(latex_options= c("scale_down", "hold_position")) %>%
+  save_kable(here::here("figures/tables/missing_models_table.jpeg"))
+```

### 2023-06-23 paper/GEB Submission/manuscript_GEB.tex
Final submission tweaks.

@@
 species' functional role to reconstruct time series of functional
-diversity metrics. Using generalized linear mixed models we estimated
+diversity metrics. Using generalized linear mixed models, we estimated
 general trends in those metrics and trends for individual studies.
@@
 with increasing turnover. The lack of trend held even after correcting
-for changes in species richness. At the study-level, we identified four
+for changes in species richness. At the study-level, there were also a
+substantial number of time series exhibiting no species or functional
+change, however most studies showed a shift in a species or functional
+metric.
+
+Main Conclusions: General trends indicate that on the aggregate one type
+of functional shift is not more prevalant than the other across many
+taxa, biomes, and realms. At the study-level, we identified four
 prevailing scenarios of species and functional change, which showed
-links to the duration of the observation window.
-
-Main Conclusions: General trends indicate that on the aggregate one type
-of functional shift is not more prevelant than the other across many
-taxa, biomes, and realms. At the study-level, there were also a
-substantial number of time series exhibiting no species or functional
-change, however the majority of studies showed a shift in a species or
-functional metric. With no one prevailing scenario of change, it will be
-critical to link change scenarios to drivers of change, particularly to
-identify communities with capacity to resist drivers from those not
-experiencing substantial pressure from a driver.
+links to the duration of the observation window. With no one prevailing
+scenario of change, it will be critical to link change scenarios to
+drivers of change, particularly to identify communities with capacity to
+resist drivers from those not experiencing substantial pressure from a
+driver.
 \end{abstract}
@@
 analyses have been heavily criticized for geographic biases, lack of
-data in the most heavily impacted areas, and not taking into account the
+data in the most heavily impacted areas, and not considering the
 ecological context of individual studies (10--12). Many of these
@@
 biased species coverage (45), a lack of accepted species-level means,
-and differences in the types of traits collected. In order to ensure
+and differences in the types of traits collected. To ensure
 comparability across taxa in trait type and data quality we therefore
@@
 related to species patterns, and which kinds of species and functional
-change are most common across communities. We distinguish between three
-possible scenarios of functional change: 1) significant functional
-change accompanied by changes in species-based metrics, 2) maintenance
-of functional structure despite significant species changes, 3)
-maintenance of functional and species metrics. Based on expectation due
-to human impacts, we expect to find significant functional changes,
-often accompanied by species change.
+change are most common across communities. We expect to find significant
+functional changes happening alongside and independent of species
+richness changes as a result of ongoing anthropogenic impacts and global
+change drivers.
 
@@
 We obtained mammal and bird time series from the BioTIME database, a
-global repository of high quality assemblage time series. All studies
+global repository of high-quality assemblage time series. All studies
 included in the database follow consistent sampling protocols and
@@
 the number of species and \emph{t} is the number of traits. This
-restriction allows for a sufficient number of axes to capture the trait
-space while maintaining computational feasibility (54). Metrics
-incorporated weighting based on species abundance.
+restriction allows for enough axes to capture the trait space while
+maintaining computational feasibility (54). Metrics incorporated
+weighting based on species abundance.
 
@@
 start year on study-level trends using linear models with duration and
-start year as predictors. We also assessed the impact of potential
-study-level outliers on general trends. Details on that process can be
-found in the supplement. All models in our analysis were fit using the
+start year as predictors. All models in our analysis were fit using the
 \emph{lme4} (v1.1-30) package in R (v4.2.3) and p-values were calculated
@@
 We did find significant differences between taxa, realms, and climate
-for Jaccard similarity and some of the \emph{CWM's}. For example while
+for Jaccard similarity and some of the \emph{CWM's}. For example, while
 Jaccard similarity was decreasing in the general trend and there were
@@
 functional change with turnover, and richness and functional change
-without turnover (Table \ref{tab:changeScenarios}). Around a quarter of
-studies experienced no trend in richness, turnover, or functional
-metrics. These studies spanned the distribution of study durations
-excluding only the very longest running studies, with the longest no
-change time series lasting 23 years.
+without turnover (Table \ref{tab:changeScenarios}). In contrast to our
+expectation based on anthropogenic impacts, around a quarter of studies
+experienced no trend in richness, turnover, or functional metrics. These
+studies spanned the distribution of study durations excluding only the
+very longest running studies, with the longest no change time series
+lasting 23 years.
 
@@
 snapshot of communities that on longer time scales would be exhibiting
-significant change in species based metrics, particularly as species
+significant change in species-based metrics, particularly as species
 replacement continued adding to turnover. Communities in this scenario,
@@
 accompanying turnover is a reminder that replacements are not always an
-indication of functional loss, but can also lead to increased functional
+indication of functional loss but can also lead to increased functional
 richness. While shifts in functional space may be the result of
 environmental pressure, we cannot distinguish between directional change
-and shifts due to the unavailability of an analog species in the species
+and shifts due to the inavailability of an analog species in the species
 pool.
@@
 decrease as a result of climate change impacts and megafaunal loss (67),
-a phenomena which has already been well documented empirically and
+a phenomenon which has already been well documented empirically and
 experimentally in multiple taxa (68--71). While the species-level trait
@@
 We present here four prevailing scenarios of change experienced in bird
-and mammal communities. While we offer an first discussion of which
-kinds of change are most common and why, assessing the true global
-prevalence will require continued efforts to fill data gaps, a well
-recognized challenge in ecology (6, 12, 77). Still, a ``scenarios of
-change'' framework can provide structure for future work addressing
-functional shifts, particularly as we reconcile results from broad-scale
-syntheses with in-depth single system studies. It will be particularly
-critical to link forms of change to individual drivers to assess which
-drivers may impact species and functional diversity differently.
-Understanding those links will help identify where directly measuring
-functional structure instead of just species change is necessary for
-understand impacts on a given system. In addition, direct links to
-drivers will improve our ability to distinguish between communities
-experiencing no change due to a lack of perturbation from communities
-with high resilience in the face of disturbance.
+and mammal communities. While we offer a first discussion of which kinds
+of change are most common and why, assessing the true global prevalence
+will require continued efforts to fill data gaps, a well recognized
+challenge in ecology (6, 12, 77). Still, a ``scenarios of change''
+framework can provide structure for future work addressing functional
+shifts, particularly as we reconcile results from broad-scale syntheses
+with in-depth single system studies. It will be particularly critical to
+link forms of change to individual drivers to assess which drivers may
+impact species and functional diversity differently. Understanding those
+links will help identify where directly measuring functional structure
+instead of just species change is necessary for understand impacts on a
+given system. In addition, direct links to drivers will improve our
+ability to distinguish between communities experiencing no change due to
+a lack of perturbation from communities with high resilience in the face
+of disturbance.
 

### 2023-06-23 analysis/05_model_metrics.Rmd
Save out the missing models

@@
   mutate(REML = TRUE, bobyqa = TRUE)
+```
 
-# There are 12 study/metric combos that can't be fit due to data limitations (too many rarefyID's )
-# tofit_mods <- bad_indv_results2 %>% filter(!is.na(warning)) %>% select(metric, study_id, formula) %>% distinct()
-# 
-# missing_indv_studies <- pmap_df(tofit_mods %>% select(metric_name = metric, filter_study_id = study_id), function(metric_name, filter_study_id) {model_data %>%
-#          filter(metric_name == metric, study_id == filter_study_id) %>%
-#         mutate(n = n(), nrarefyID = n_distinct(rarefyID)) %>%
-#         select(metric, study_id, nrarefyID, n) %>%
-#         distinct()})
+There are 12 study/metric combos that can't be fit due to data limitations (too many rarefyID's )
+```{r}
+tofit_mods <- bad_indv_results2 %>% filter(!is.na(warning)) %>% select(metric, study_id, formula) %>% distinct()
 
+missing_indv_studies <- pmap_df(tofit_mods %>% select(metric_name = metric, filter_study_id = study_id), function(metric_name, filter_study_id) {model_data %>%
+         filter(metric_name == metric, study_id == filter_study_id) %>%
+        mutate(n = n(), nrarefyID = n_distinct(rarefyID)) %>%
+        select(metric, study_id, nrarefyID, n) %>%
+        distinct()})
+
+readr::write_csv(missing_indv_studies, here::here("paper/missing_indv_studies.csv"))
 ```

### 2023-06-23 paper/GEB Submission/manuscript_GEB.Rmd

@@
   Methods: We paired thousands of bird and mammal assemblage time series from the BioTIME database with existing trait data representative of a species' functional role to reconstruct time series of functional diversity metrics.
-Using generalized linear mixed models we estimated general trends in those metrics and trends for individual studies.
+Using generalized linear mixed models, we estimated general trends in those metrics and trends for individual studies.
 
@@
 The lack of trend held even after correcting for changes in species richness.
-At the study-level, we identified four prevailing scenarios of species and functional change, which showed links to the duration of the observation window.
+At the study-level, there were also a substantial number of time series exhibiting no species or functional change, however most studies showed a shift in a species or functional metric.
 
   Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevalant than the other across many taxa, biomes, and realms.
-At the study-level, there were also a substantial number of time series exhibiting no species or functional change, however the majority of studies showed a shift in a species or functional metric.
+At the study-level, we identified four prevailing scenarios of species and functional change, which showed links to the duration of the observation window.
 With no one prevailing scenario of change, it will be critical to link change scenarios to drivers of change, particularly to identify communities with capacity to resist drivers from those not experiencing substantial pressure from a driver.
@@
 The meaning of general trends derived from limited data, and their relevance for conservation, is a topic of on going debate.
-Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and not taking into account the ecological context of individual studies [@cardinale2014; @cardinale2018; @gonzalez2016].
+Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and not considering the ecological context of individual studies [@cardinale2014; @cardinale2018; @gonzalez2016].
 Many of these criticisms reflect limitations of ecological data on the whole, and proponents argue we must make the best of the imperfect data available [@primack2018].
@@
 While examining trends in plants, invertebrates, and other vertebrate species is of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
-In order to ensure comparability across taxa in trait type and data quality we therefore focus on mammals and birds.
+To ensure comparability across taxa in trait type and data quality we therefore focus on mammals and birds.
 Traits were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].
@@
 We further assess the prevailing modes of change occurring at the study level to determine how functional change is related to species patterns, and which kinds of species and functional change are most common across communities.
-We distinguish between three possible scenarios of functional change: 1) significant functional change accompanied by changes in species-based metrics, 2) maintenance of functional structure despite significant species changes, 3) maintenance of functional and species metrics.
-Based on expectation due to human impacts, we expect to find significant functional changes, often accompanied by species change.
+We expect to find significant functional changes happening alongside and independent of species richness changes as a result of ongoing anthropogenic impacts and global change drivers.
 
@@
 
-We obtained mammal and bird time series from the BioTIME database, a global repository of high quality assemblage time series.
+We obtained mammal and bird time series from the BioTIME database, a global repository of high-quality assemblage time series.
 All studies included in the database follow consistent sampling protocols and represent full assemblages rather than populations of single species [@dornelas2018].
@@
 The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
-This restriction allows for a sufficient number of axes to capture the trait space while maintaining computational feasibility [@blonder2018].
+This restriction allows for enough axes to capture the trait space while maintaining computational feasibility [@blonder2018].
 Metrics incorporated weighting based on species abundance.
@@
   kable_paper() %>%
-  kable_styling(latex_options= c("scale_down", "hold_position"))
+  kable_styling(latex_options= c("scale_down", "hold_position")) %>%
+  save_kable(here::here("figures/tables/results_table.jpeg"))
 ```
@@
 We did find significant differences between taxa, realms, and climate for Jaccard similarity and some of the *CWM's*.
-For example while Jaccard similarity was decreasing in the general trend and there were significant within group slopes for Mammals, Birds, and Terrestrial communities, there was no significant slope for the marine realm, indicating that the general trend is mostly driven by turnover in terrestrial communities.
+For example, while Jaccard similarity was decreasing in the general trend and there were significant within group slopes for Mammals, Birds, and Terrestrial communities, there was no significant slope for the marine realm, indicating that the general trend is mostly driven by turnover in terrestrial communities.
 For bird communities, we also found within group trends and between group differences for trends in foraging behavior.
@@
         booktabs = TRUE, align = "c", escape = FALSE) %>%
-  kable_styling("striped")
+  kable_styling("striped")%>%
+  save_kable(here::here("figures/tables/trend_table.jpeg"))
 
@@
 At the study level, biodiversity change fell into four dominant scenarios of change based on trends in richness, turnover, and functional metrics: no change, only functional change, richness and functional change with turnover, and richness and functional change without turnover (Table \ref{tab:changeScenarios}).
-Around a quarter of studies experienced no trend in richness, turnover, or functional metrics.
+In contrast to our expectation based on anthropogenic impacts, around a quarter of studies experienced no trend in richness, turnover, or functional metrics.
 These studies spanned the distribution of study durations excluding only the very longest running studies, with the longest no change time series lasting 23 years.
@@
 These communities are best aligned with scenarios of species replacement where only a few functional outliers replaced functionally indistinct species or vice versa (there were studies with positive and negative trends in all three functional metrics).
-Studies in the scenario of only functional change were heavily skewed toward shorter running time series, indicating that they may represent a limited snapshot of communities that on longer time scales would be exhibiting significant change in species based metrics, particularly as species replacement continued adding to turnover.
+Studies in the scenario of only functional change were heavily skewed toward shorter running time series, indicating that they may represent a limited snapshot of communities that on longer time scales would be exhibiting significant change in species-based metrics, particularly as species replacement continued adding to turnover.
 Communities in this scenario, and the previous scenario of no functional change are consistent with hypotheses that species richness may be strongly regulated through time [@brown2001; @gotelli2017], and emphasize that maintenance of functional structure can be divorced from those processes.
@@
 Species are not therefore being replaced by exact functional counter parts.
-The fact that we observed both increases and decreases in corrected functional richness accompanying turnover is a reminder that replacements are not always an indication of functional loss, but can also lead to increased functional richness.
-While shifts in functional space may be the result of environmental pressure, we cannot distinguish between directional change and shifts due to the unavailability of an analog species in the species pool.
+The fact that we observed both increases and decreases in corrected functional richness accompanying turnover is a reminder that replacements are not always an indication of functional loss but can also lead to increased functional richness.
+While shifts in functional space may be the result of environmental pressure, we cannot distinguish between directional change and shifts due to the inavailability of an analog species in the species pool.
 
-```{r changeScenarios, echo = FALSE, message = FALSE, fig.show = 'hold', fig.cap= "Comparision of the distribution of time series durations from each type of change to the overall distribution of time series durations, with no richness change or turnover and functional change in yellow, no turnover but richness and functional change in green, no change in blue, and turnover, richness and functional change studies in orange."}
+```{r changeScenarios, echo = FALSE, message = FALSE, fig.show = 'hold', fig.cap= "Comparison of the distribution of time series durations from each type of change to the overall distribution of time series durations, with no richness change or turnover and functional change in yellow, no turnover but richness and functional change in green, no change in blue, and turnover, richness and functional change studies in orange."}
 
@@
 Still, some results stand in contrast to predictions for trait shifts under global change.
-For example, mean body size is predicted to decrease as a result of climate change impacts and megafaunal loss [@sheridan2011], a phenomena which has already been well documented empirically and experimentally in multiple taxa [@caruso2014; @forster2012; @huss2019; @tseng2018].
+For example, mean body size is predicted to decrease as a result of climate change impacts and megafaunal loss [@sheridan2011], a phenomenon which has already been well documented empirically and experimentally in multiple taxa [@caruso2014; @forster2012; @huss2019; @tseng2018].
 While the species-level trait means used here are not appropriate for assessing intraspecific body size shifts, we would expect to see shifts in community-wide means due to local losses of large-bodied species.
@@
 We present here four prevailing scenarios of change experienced in bird and mammal communities.
-While we offer an first discussion of which kinds of change are most common and why, assessing the true global prevalence will require continued efforts to fill data gaps, a well recognized challenge in ecology [@gonzalez2016; @hughes2021; @vellend2017].
+While we offer a first discussion of which kinds of change are most common and why, assessing the true global prevalence will require continued efforts to fill data gaps, a well recognized challenge in ecology [@gonzalez2016; @hughes2021; @vellend2017].
 Still, a "scenarios of change" framework can provide structure for future work addressing functional shifts, particularly as we reconcile results from broad-scale syntheses with in-depth single system studies.

### 2024-01-19 paper/GEB Submission/manuscript_GEB.Rmd
Addressing reviewer comments.

@@
 options(scipen=999)
+
+colorize <- function(x, color) {
+  if (knitr::is_latex_output()) {
+    sprintf("\\textcolor{%s}{%s}", color, x)
+  } else if (knitr::is_latex_output()) {
+    sprintf("<span style='color: %s;'>%s</span>", color,
+      x)
+  } else x
+}
 ```
@@
 It is critical to establish whether or not functional loss is prevalent across communities.
-While functional loss is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], local-scale loss is not necessarily inevitable even in scenarios of species loss [@daz2001; @petchey2009].
+While functional loss is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], `r colorize("functional diversity may be maintained even when species are lost from a community", "blue")` [@daz2001; @petchey2009].
 Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020].
@@
 Continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
-The number of trait axes was limited to the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
+`r colorize("Before calculating diversity metrics, dbFD reduces the dimensionality of the trait space by performing PCoA.
+We limited the number of included PCoA axes to", "blue")` the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
 This restriction allows for enough axes to capture the trait space while maintaining computational feasibility [@blonder2018].
@@
 
-We estimated general trends for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study.
+We estimated general trends `r colorize("across bird and mammals communities", "blue")` for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study.
 We fit 18 individual *CWM* models.
@@
 Trends for different taxa, biomes, or realms were also non-significant for richness and summary functional diversity metrics, with the exception of a significantly increasing trend for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing functional richness slope for temperate/tropical studies and mammal studies.
-However, trends were not exhibited for the standardized metric indicating that differences were largely due to changes in species richness.
+However, trends were not exhibited for the standardized metric indicating that differences `r colorize("in functional diversity metrics", "blue")` were largely due to changes in species richness.
 Further, with only two global studies, the trend should not be considered truly general.
@@
 The studies in this scenario fell into two groups: a positive trend in richness with a positive trend in functional richness (corrected and uncorrected), and a negative trend in species richness with a negative trend in uncorrected functional richness and no trend in corrected richness but a trend in other corrected functional metrics.
-This would indicate that when species are lost from a community, they are not more functionally unique than expected by chance, though they may produce shifts in other aspects of functional structure.
-When species are gained however, they disproportionately add to the functional richness of the community.
+`r colorize("This would indicate that when species are gained in a community, they disproportionately add to the functional richness.
+When species are lost however, they are not more functionally unique than expected by chance though they may produce shifts in other aspects of functional space.", "blue")`
 

### 2024-01-19 paper/GEB Submission/manuscript_GEB.tex
Addressing reviewer comments.

@@
 across communities. While functional loss is frequently cited as one of
-the most pressing concerns of the anthropocene (23--25), local-scale
-loss is not necessarily inevitable even in scenarios of species loss
+the most pressing concerns of the anthropocene (23--25), functional
+diversity may be maintained even when species are lost from a community
 (26, 27). Forecasts of functional loss range from negligible (28) to
@@
 traits were z-score scaled to give each trait equal weight in the trait
-space (52, 53). The number of trait axes was limited to the maximum
-number of traits that fulfills the criteria \(s >= 2^t\), where \(s\) is
-the number of species and \emph{t} is the number of traits. This
-restriction allows for enough axes to capture the trait space while
-maintaining computational feasibility (54). Metrics incorporated
-weighting based on species abundance.
+space (52,
+53).\textcolor{blue}{Before calculating diversity metrics, dbFD reduces the dimensionality of the trait space by performing PCoA. We limited the number of included PCoA axes to}
+the maximum number of traits that fulfills the criteria \(s >= 2^t\),
+where \(s\) is the number of species and \emph{t} is the number of
+traits. This restriction allows for enough axes to capture the trait
+space while maintaining computational feasibility (54). Metrics
+incorporated weighting based on species abundance.
 
@@
 
-\begin{table}[!h]
-
-\caption{\label{tab:resultsTab}Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics.
-Additional model estimates, including CWM models, can be found in the supplement.}
-\centering
-\resizebox{\linewidth}{!}{
-\begin{tabular}[t]{l|l|>{\raggedright\arraybackslash}p{2cm}|l|r|r|l}
-\hline
-metric & effect & grouping & term & estimate & std.error & p.value\\
-\hline
- &  &  & Intercept & -0.25 & 0.11 & 0.03\\
-\cline{4-7}
- & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & 0.00 & 0.04 & 0.96\\
-\cline{2-7}
- &  &  & SD Intercept & 0.60 &  & \\
-\cline{4-7}
- &  &  & SD Year & 0.11 &  & \\
-\cline{4-7}
- &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.16 &  & \\
-\cline{3-7}
- &  &  & SD Intercept & 0.60 &  & \\
-\cline{4-7}
- &  &  & SD Year & 0.23 &  & \\
-\cline{4-7}
-\multirow{-8}{*}{\raggedright\arraybackslash SES\_FDiv} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & -0.01 &  & \\
-\cline{1-7}
- &  &  & Intercept & 0.09 & 0.16 & 0.59\\
-\cline{4-7}
- & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & 0.00 & 0.01 & 0.73\\
-\cline{2-7}
- &  &  & SD Intercept & 1.04 &  & \\
-\cline{4-7}
- &  &  & SD Year & 0.02 &  & \\
-\cline{4-7}
- &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & 1.00 &  & \\
-\cline{3-7}
- &  &  & SD Intercept & 0.40 &  & \\
-\cline{4-7}
- &  &  & SD Year & 0.17 &  & \\
-\cline{4-7}
-\multirow{-8}{*}{\raggedright\arraybackslash SES\_FEve} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & -0.21 &  & \\
-\cline{1-7}
- &  &  & Intercept & -0.26 & 0.07 & <0.001\\
-\cline{4-7}
- & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & 0.04 & 0.04 & 0.35\\
-\cline{2-7}
- &  &  & SD Intercept & 0.29 &  & \\
-\cline{4-7}
- &  &  & SD Year & 0.11 &  & \\
-\cline{4-7}
- &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.27 &  & \\
-\cline{3-7}
- &  &  & SD Intercept & 0.54 &  & \\
-\cline{4-7}
- &  &  & SD Year & 0.18 &  & \\
-\cline{4-7}
-\multirow{-8}{*}{\raggedright\arraybackslash SES\_FRic} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.06 &  & \\
-\cline{1-7}
- &  &  & Intercept & 0.46 & 0.02 & <0.001\\
-\cline{4-7}
- & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & -0.03 & 0.01 & <0.001\\
-\cline{2-7}
- &  &  & SD Intercept & 0.09 &  & \\
-\cline{4-7}
- &  &  & SD Year & 0.02 &  & \\
-\cline{4-7}
- &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.35 &  & \\
-\cline{3-7}
- &  &  & SD Intercept & 0.08 &  & \\
-\cline{4-7}
- &  &  & SD Year & 0.02 &  & \\
-\cline{4-7}
-\multirow{-8}{*}{\raggedright\arraybackslash log(Jaccard + 1)} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.05 &  & \\
-\cline{1-7}
- &  &  & Intercept & 2.40 & 0.10 & <0.001\\
-\cline{4-7}
- & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & -0.06 & 0.05 & 0.17\\
-\cline{2-7}
- &  &  & SD Intercept & 0.64 &  & \\
-\cline{4-7}
- &  &  & SD Year & 0.27 &  & \\
-\cline{4-7}
- &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.70 &  & \\
-\cline{3-7}
- &  &  & SD Intercept & 0.21 &  & \\
-\cline{4-7}
- &  &  & SD Year & 0.09 &  & \\
-\cline{4-7}
-\multirow{-8}{*}{\raggedright\arraybackslash log(Species Richness)} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.34 &  & \\
-\hline
-\end{tabular}}
-\end{table}
-
 We did find significant differences between taxa, realms, and climate
@@
 
-\begin{table}
-
-\caption{\label{tab:trendTab}Number of studies that experienced a significant trend in each calculated metric out of 50 total studies.}
-\centering
-\begin{tabular}[t]{ccccccccc}
-\toprule
- & S & Jaccard Similarity & FRic & FEve & FDiv & SES FRic & SES FEve & SES FDiv\\
-\midrule
-+ & 7 & 0 & 7 & 5 & 3 & 4 & 4 & 3\\
-- & 8 & 11 & 8 & 6 & 5 & 2 & 6 & 2\\
-\bottomrule
-\end{tabular}
-\end{table}
-
 Study-level slopes for summary functional metrics were significantly
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{../../figures/duration_hist} \includegraphics[width=\textwidth]{../../figures/functional_change_table} \caption{Comparision of the distribution of time series durations from each type of change to the overall distribution of time series durations, with no richness change or turnover and functional change in yellow, no turnover but richness and functional change in green, no change in blue, and turnover, richness and functional change studies in orange.}\label{fig:changeScenarios}
+\includegraphics[width=\textwidth]{../../figures/duration_hist} \includegraphics[width=\textwidth]{../../figures/functional_change_table} \caption{Comparison of the distribution of time series durations from each type of change to the overall distribution of time series durations, with no richness change or turnover and functional change in yellow, no turnover but richness and functional change in green, no change in blue, and turnover, richness and functional change studies in orange.}\label{fig:changeScenarios}
 \end{figure}

### 2024-03-11 analysis/exporatory_scripts/invasive_data.Rmd
Just... so many changes.

@@
 library(taxadb)
+```
+
+Try with the USDA master list of invasive species
+```{r}
+invasive <- read.csv(here::here("inst/extdata/USRIISv2csvFormat/USRIISv2_MasterList.csv"))
+load(here::here("bt_traitfiltered.rda"))
+
+# all invasive species have GBIF id's so let's match on that
+ids <- bt_traitfiltered %>%
+  select(species) %>% 
+  distinct() %>%
+  mutate(sciName = taxadb::get_names(species),
+         gbif = taxadb::get_ids(sciName, provider = "gbif"))
+
+included_invasives <- invasive %>%
+  filter(locality == "L48") %>% 
+  select(GBIF_taxonKey, establishmentMeans, degreeOfEstablishment) %>%
+  distinct() %>% 
+  mutate(gbif = paste0("GBIF:", GBIF_taxonKey)) %>%
+  inner_join(ids)
+```
+Get a count of invasives for each study %>%
+```{r}
+invasive_count <- bt_traitfiltered %>%
+  select(species, study_id, rarefyid, year) %>%
+  left_join(included_invasives %>% select(species) %>%
+              mutate(invasive = 1)) %>%
+  select(-species) %>%
+  distinct() %>%
+  group_by(study_id, rarefyid, year) %>%
+  summarize(invasive_count = sum(invasive, na.rm = TRUE))
+
 ```

### 2024-03-11 analysis/figures/06_study_change.Rmd
Just... so many changes.

@@
 ```
+
+Let's look at more specific combinations
+```{r}
+load(here::here("data/meta.rda"))
+
+change_groups <- wide_table %>% 
+  select(study_id, S, Jaccard_base, starts_with("SES")) %>%
+  mutate(across(!study_id, ~case_when(
+    . > 0 ~ "pos",
+    . < 0 ~ "neg",
+    .default = "0"
+  )))
+
+change_group_count <- change_groups %>%
+  select(-study_id) %>%
+  group_by(across(everything())) %>%
+  count()
+
+no_change <- change_groups %>% 
+  mutate(study_id = as.integer(study_id)) %>%
+  filter(across(!study_id, ~. == 0)) %>%
+  left_join(meta)
+
+redundancy <- change_groups %>% 
+  mutate(study_id = as.integer(study_id)) %>%
+  filter(S == "neg", SES_FRic == "0", SES_FDiv == "0", SES_FEve == "0") %>%
+  left_join(meta)
+
+redundancy_gain <- change_groups %>% 
+  mutate(study_id = as.integer(study_id)) %>%
+  filter(S == "pos", SES_FRic == "0", SES_FDiv == "0", SES_FEve == "0") %>%
+  left_join(meta)
+
+evenness <- change_groups %>% 
+  mutate(study_id = as.integer(study_id)) %>%
+  filter(across(!study_id & !SES_FEve, ~. == 0)) %>%
+  left_join(meta)
+
+increase <- change_groups %>% 
+  mutate(study_id = as.integer(study_id)) %>%
+  filter(Jaccard_base == "neg", S == "pos", SES_FRic == "pos", SES_FDiv == "0", SES_FEve == "0") %>%
+  left_join(meta)
+```

### 2024-03-11 paper/GEB Submission/manuscript_GEB.Rmd
Just... so many changes.

@@
 The meaning of general trends derived from limited data, and their relevance for conservation, is a topic of on going debate.
-Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and not considering the ecological context of individual studies [@cardinale2014; @cardinale2018; @gonzalez2016].
-Many of these criticisms reflect limitations of ecological data on the whole, and proponents argue we must make the best of the imperfect data available [@primack2018].
-Still, there is general consensus on a few local-scale patterns that seem to be characteristic of the anthropocene, including the prevalence of species turnover across communities and patterns of both richness increase and decrease in different taxa and biomes [@dornelas2023].
-Understanding the implications of change in species assemblage for ecosystem stability and function is a critical next step in moving forward the biodiversity change conversation.
+Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
+Many of these criticisms reflect limitations of ecological data on the whole, `r colorize("leading to a call for additional data not only to fill geographic and temporal gaps, but to flesh out key characteristics of communities", "blue")` [@dornelas2023; @primack2018].
 
-Functional diversity offers a potentially powerful alternative to species-based approaches for detecting and describing community change by providing a mechanistic link between species' response to environmental change (*response traits*) and the processes they perform (*effect traits*) [@lavorel2002; @mcgill2006; @suding2008].
+Functional diversity offers a potentially powerful addition to species-based approaches for detecting and describing community change by providing a mechanistic link between species' response to environmental change (*response traits*) and the processes they perform (*effect traits*) [@lavorel2002; @mcgill2006; @suding2008].
 By describing the functional trait space, functional diversity metrics capture the disproportionate impact of losses or gains of functionally unique species.
-Functional diversity metrics are therefore particularly well suited for assessing community shifts underlying even constant species richness trends.
+Functional diversity metrics `r colorize("may therefore illuminate joint responses from functionally similar species or communities undetectible by looking at species identity alone.", "blue")`
 
-Beyond simply characterizing changes in community structure, trends in functional composition also have important implications for ecosystem stability, function, and resilience.
-There is increasing evidence functional diversity is a better predictor of ecosystem function than species-based metrics [@cadotte2011; @gagic2015], and that different facets of functional diversity play essential roles in maintaining ecosystem stability [@craven2018; @morin2014].
-Almost all hypothesized mechanisms underpinning the relationship between species diversity and ecosystem function are trait-dependent [@hillebrand2009].
-Determining functional trends therefore gives a more direct picture of potential trends in critical ecosystem processes.
-
-It is critical to establish whether or not functional loss is prevalent across communities.
+`r colorize("The expectation for functional change across communities is not obvious from past work, and may or may not follow taxonomic trends", "blue")` [@dornelas2023].
 While functional loss is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], `r colorize("functional diversity may be maintained even when species are lost from a community", "blue")` [@daz2001; @petchey2009].
@@
 And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @larsen2018; @matuoka2020].
-On paleoecological time scales functional composition shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with significant impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].
+On paleoecological time scales, functional composition shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with significant impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].
 For some time periods, functional structure appears to be maintained for substantial portions of geological time [@hedberg2022].
-Still, some losses of functional diversity are indisputable on both paleoecological and contemporary timescales such as continued trophic downgrading due to loss of large-bodied mammals, but implications of those losses for local diversity patterns are less clear [@estes2011; @smith2018] .
+`r colorize("Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds [@barnagaud2017; @jarzyna2016] and ray-finned fishes, sharks, and rays", "blue")` [@trindade-santos2020].
 
-Assessments of broad-scale temporal change in functional diversity have previously been limited by a lack of functional trait data.
-The majority of work has therefore focused largely on system-specific studies with traits collected *in situ*.
-This has challenged our ability to establish general rules for how functional composition may be changing through time, particularly in response to global change drivers.
-Ongoing efforts to assemble functional traits for a variety of taxa have made synthesis of existing community assemblage data and functional traits possible for the first time, providing initial insights into the ways functional diversity changes on a broad scale for specific taxa [e.g.
-fish, @trindade-santos2020; birds, @jarzyna2016; @barnagaud2017].
-However, to date there has been no cross-taxa assessment of temporal functional change for a broad geographic and taxonomic extent [@díaz2022].
-
-Here we perform the first multi-taxa, multi-realm assessment of functional diversity change through time.
+Here we `r colorize("leverage ongoing efforts to assemble functional trait data and recent computational advances to", "blue")` perform the first multi-taxa, multi-realm assessment of functional diversity change through time.
 We focus on mammal and bird species as a significant subset of the world's biodiversity heavily impacted by anthropogenic change.
@@
 
-We assess thousands of mammal and bird functional diversity time series to determine whether or not there is a general trend of functional change, both in observed metrics and in metrics corrected for changes in species richness.
-We further assess the prevailing modes of change occurring at the study level to determine how functional change is related to species patterns, and which kinds of species and functional change are most common across communities.
-We expect to find significant functional changes happening alongside and independent of species richness changes as a result of ongoing anthropogenic impacts and global change drivers.
+We assess thousands of mammal and bird functional diversity time series to determine whether or not `r colorize("the addition of functional trait data gives a clarifying picture of biodiversity change across communities.
+Rather than testing specific hypotheses of change we present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination.
+We present results at three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm), and trends for individual studies.", "blue")`
 
@@
 For a full breakdown of studies and their characteristics, see the supplement.
+Our final dataset reflects many of the data biases that make global synthesis work challenging, including geographic bias, a bias away from areas currently under the greatest threat, and a bias towards shorter time series.
+We address these shortcomings and their potential impact on our results in the discussion.
 
@@
 We also calculated the community-weighted mean (*CWM*) of included traits to examine shifts in the distribution of each trait.
+`r colorize("Hereafter we refer to results for functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).", "blue")`
 
@@
 
-To assess functional change independent of species richness we calculated the standardized effect size (SES) for each of the three summary functional diversity metrics (*FRic*, *FEve*, *FDiv*) from null estimates [@swenson2012].
+To assess functional change independent of species richness we calculated the standardized effect size (SES) for each of the three functional diversity metrics (*FRic*, *FEve*, *FDiv*) from null estimates [@swenson2012].
 Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
 Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species.
-Species pools included all species observed for a time series.
+\` `r colorize("Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability", "blue")`.
 This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
@@
 
-We found no significant overall trend in species richness or summary functional diversity metrics (observed or standardized) (Fig \ref{fig:timeseriesPlot}).
+We found no significant overall trend in species richness or functional diversity metrics (observed or `r colorize("corrected", "blue")`) (Fig \ref{fig:timeseriesPlot}).
 We did find a significant overall decrease in Jaccard similarity, indicating accumulating changes in species composition.
 Non-significant overall trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies was plausibly 0 (Table \ref{tab:resultsTab}).
-Trends for different taxa, biomes, or realms were also non-significant for richness and summary functional diversity metrics, with the exception of a significantly increasing trend for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing functional richness slope for temperate/tropical studies and mammal studies.
-However, trends were not exhibited for the standardized metric indicating that differences `r colorize("in functional diversity metrics", "blue")` were largely due to changes in species richness.
+`r colorize("Within-group", "blue")` trends for different taxa, biomes, or realms were also non-significant for richness and functional diversity metrics, with the exception of a significantly increasing trend for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing functional richness slope for temperate/tropical studies and mammal studies.
+However, trends were not exhibited for the `r colorize("corrected", "blue")` metric indicating that differences `r colorize("in functional diversity metrics", "blue")` were largely due to changes in species richness.
 Further, with only two global studies, the trend should not be considered truly general.
@@
   dplyr::rename(grouping = group) %>%
-  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics.
+  kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness, Jaccard similarity, and corrected functional diversity metrics.
 Additional model estimates, including CWM models, can be found in the supplement.") %>%
@@
 At the study level, 15 of 50 studies exhibited a significant trend in species richness and 11 exhibited significant turnover.
-For observed functional metrics, 25 of 50 studies exhibited a trend in a least one metric, and 15 of 50 studies exhibited a significant trend for at least one standardized metrics (Table \ref{tab:trendTab}).
+For observed functional metrics, 25 of 50 studies exhibited a trend in a least one metric, and 15 of 50 studies exhibited a significant trend for at least one `r colorize("corrected", "blue")` metric (Table \ref{tab:trendTab}).
 In general, there were more significant trends for uncorrected metrics, with some disappearing after correction, indicating that those trends were likely due to changes in the number of species.
@@
 
-Study-level slopes for summary functional metrics were significantly related to start year of the time series for Jaccard similarity and functional evenness, both of which had significantly more negative slopes with more recent start year.
-No summary functional metrics were significantly related to the start year of the time series.
+Study-level slopes for functional diversity metrics were significantly related to start year of the time series for Jaccard similarity and functional evenness, both of which had significantly more negative slopes with more recent start year.
+No functional diversity metrics were significantly related to the start year of the time series.
 
@@
 Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at aggregate and local trends in functional diversity in mammal and bird communities.
-Surprisingly, we did not detect an overall trend in any of the summary functional diversity metrics.
+`r colorize("Our results show that the addition of functional traits illuminates a few consistent functional trends across communities, but largely complicates rather that clarifies the story of biodiversity change.
+While the characteristics of species clearly matter, instead of unifying the nature of communities' change, they more often distinguish them.
+In general, we found a few areas of consensus for models where communities were aggregated (general trends, trends by taxa, biome, and realm), and vast heterogeneity for study-level models.", "blue")`
+
+### Evidence of consensus
+
+`r colorize("The most stark area of consensus was the lack of general trend aggregating across communities.", "blue")` We did not detect an overall trend in any functional diversity metrics.
 As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics may be consistent with similar well-documented species derived trends.
-We found no trend in functional change for almost all realms, biomes, and taxonomic groups, with evidence of functional richness loss in mammal studies only.
-These results are consistent with multiple studies linking anthropogenic drivers to loss of functional diversity in mammal communities [@brodie2021; @cox2022; @li2022]
+We found no trend in functional change for almost all realms, biomes, and taxonomic groups.
+`r colorize("The one exception was a consensus trend indicating", "blue")` functional richness loss for mammal studies.
+These results are consistent with multiple studies linking anthropogenic drivers to loss of functional diversity in mammal communities [@brodie2021; @cox2022; @li2022].
 
-Despite a lack of general trends in summary functional metrics, we did find multiple trends in *CWM*'s for studies coming from the same taxa, realms, or climates, many of which are consistent with previous findings for those contexts.
+Despite a lack of general trends in functional diversity metrics, we did find evidence of compositional shifts in the trait space with multiple `significant` trends in *CWM*'s for studies coming from the same taxa, realms, or climates, `indicating areas of consensus change across communities`.
+Some of those trends are consistent with existing work in those contexts.
 For example, the decrease in insect consumption for tropical communities reflects well documented declines in tropical insectivorous birds [@sherry2021].
@@
 Other shifts in diet for both Birds and Mammals, including increasing fruit, seed, and vertebrate consumption and decreasing seed consumption in some climates point to important areas for further exploration.
+
+### Study heterogeneity
+
+`r colorize("At the study-level, our results run contrary to our hope for a clearer picture of biodiversity change through a functional lens.
+The lack of general trend in functional diversity metrics belies a huge range of positive and negative trends at the study-level that happen to average out to zero.
+In order to simplify discussion, we will talk about the implications for communities grouped by their concurrent change in functional diversity metrics, species richness and turnover split into positive, negative or no trend.
+While there are over 150 possible combinations of change direction (or no change) in the 5 metrics, we discuss here the six scenarios that occurred in more than one study: no change, species loss or gain only, loss of functional evenness, richness loss with species turnover, and increase in species and functional richness accompanied by significant turnover.
+We focus on these scenarios to combat potential spurious results due to multiple testing, as it is unlikely all observations of the same scenario are due to false positives.
+We include a break down of the number of studies in each scenario to illustrate relative number of studies in each group, but emphasize that the absolute number is likely impacted by multiple testing.", "blue")`
+
+The majority group of studies exhibited no trend in any species or functional diversity metric.
+Contrary to the expectation due to anthropogenic and global change stressors, these communities do not show significant changes over the course of the observation window.
+Studies in this group span the distribution of study durations excluding only the very longest running studies, with the longest no change time series lasting 23 years.
+They also included both bird and mammal studies and only four were located in protected areas, indicating that the lack of trend is not restricted to a specific ecological context or those communities most insulated from human impact.
+
+The lack of trend could be the result of multiple possible scenarios.
+First, these may be communities resisting perturbations or simply not experiencing significant perturbations.
+Given the studies in this group come from all possible taxa, realms, climates, and protection statuses this option is a strong possibility and offers some hope that there are areas of the globe where communities are fairing well for now.
+Alternatively, these may be communities that have experienced or continue to experience significant stress, but lost species or functional diversity outside the observation window.
+This could be true particularly for North American mammal communities where trophic downgrading and megafaunal losses occurred hundreds of years ago [citation].
+Third, these communities may be experiencing directional shifts undetectable by available data.
+For example, species-level trait data does not capture intraspecific shifts in the trait space, which can represent significant changes in overall functional composition and impact maintenance of ecological processes [citation].
+
+Three of the groups fit under the broad umbrella of changes in redundancy, including communities exhibiting gains or losses in species richness with no change in functional metrics or turnover and richness loss with significant turnover.
+These communities experience changes in species richness that do not impact the overall functional space, meaning species gains or losses represent a loss or increase in functional redundancy respectively.
+Where turnover is also occurring, species are being exchanged for functional analogs from the species pool.
+For communities exhibiting loss of redundancy, ecological processes are likely being maintained, but have a reduced capacity to respond to future stressors.
+These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future.
+Conversely, communities exhibiting a increase in redundancy (via an increase in species richness) are becoming better positioned to respond to future stressors, but are not actually expanding their functional capacity as may be assumed by looking at species gains alone.
+
+-   add redundancy literature context?
+
+The next group of studies exhibit an increase in species richness and functional richness with significant turnover.
+While these communities are losing some species that are functionally redundant or functional analogs of species additions, they are gaining even more species that expand the functional space.
+These communities have a hopeful trajectory, as they are improving their functional capacity and potentially their functional redundancy, likely leading to robust future ecological function.
+Notably, the studies that exhibited species richness increases (in this group with increasing functional richness, as well as communities without functional richness increases) were exclusively from terrestrial, temperate (sometimes temperate/polar) bird communities.
+We therefore have a strong indication that communities whose functional state has improved may be restricted to these specific contexts for x,y,z reasons.
+
+The final group exhibits decreasing functional evenness.
+The implications of this change current and future maintenance of functional structure is the least clear from our current understanding of the relationship between different aspects of functional diversity and ecological process.
+Maintaining evenness may be important for supporting multifunctionality in communities [@bagousse-pinguet2019]
+
+While the majority of studies fell into one of the groups discussed in detail above, 25% of the studies exhibited some different combination of change in metrics, underlining the vast heterogeneity of realized change scenarios.
+While we have an indication of the ecological characteristics most likely to lead to some of the change types (e.g.
+temperate bird communities are gaining species and functional richness), most types of change are exhibited across many different kinds of communities.
+Identifying key factors mediating the kind of biodiversity change a community experiences is critical for identifying areas of conservation concern and extrapolating our results to areas where functional trait data may not be available.
+While it is tempting to draw a one to one line between the degree of human impact and loss functional integrity (in the form of functional redundancy and richness), that relationship is not borne out by our rough metric of human impact or previous taxa or single community work.
 
@@
 While we do not address how observed rates differ from background expectation here, we can make some assertions about the implication of turnover for functional change.
-Where turnover is occurring, it is almost always accompanied by a directional change in the functional space.
+Where turnover is occurring, it is almost always accompanied by a `r colorize("trend in one of the functional diversity metrics, indicating", "blue")` directional change in the functional space.
 Species are not therefore being replaced by exact functional counter parts.
-The fact that we observed both increases and decreases in corrected functional richness accompanying turnover is a reminder that replacements are not always an indication of functional loss but can also lead to increased functional richness.
-While shifts in functional space may be the result of environmental pressure, we cannot distinguish between directional change and shifts due to the inavailability of an analog species in the species pool.
+The fact that we observed both increases and decreases in corrected functional richness accompanying turnover is a reminder that `r colorize("new species are not always replacing more functionally unique species, but are often more functionally unique themselves.", "blue")` <!-- While shifts in functional space may be the result of environmental pressure, we cannot distinguish between directional change and shifts due to the inavailability of an analog species in the species pool.
+Comment: Want to say that trends could be an indication of adaptation to changing environment OR the result of extirpations being replaced (or not) by non-analoge species.
+Can't really distinguish between those things.
+-->
 
@@
 dispersion, rarity, abundance of dominant species, @bagousse-pinguet2021].
-
-## Potential Methodological Limitations
-
-Here we approach the question of functional change using the best available data and biodiversity synthesis approaches.
-While we found signatures of change at the study level, some gaps in best practices may be obscuring a true general trend at the aggregate level.
-First, the BioTIME database, while the most comprehensive data source of time series available, is limited in temporal and geographic scope.
-Most time series span only a few years (Fig \ref{fig:taxaMap}) and may not provide the statistical power necessary to detect trends [@wauchope2019].
-The database is also not a representative sample of the world's biodiversity or areas of greatest threat [@gonzalez2016; @vellend2017], and the subset of data in this study is predominantly from North America.
-We may simply not have data from those areas experiencing the greatest perturbation [@hughes2021], particularly scenarios of conversion to urban, human-dominated landscapes.
-While evidence from other work shows even disturbed communities can maintain functional structure [@edwards2013; @matuoka2020], these results should not be interpreted as evidence of low functional impact in areas of heavy human disturbance.
-
-Second, despite using the most comprehensive trait databases for these taxa, we were still limited to species-level means of the traits deemed important by database creators.
-The importance of intraspecific variation is well documented [@desroches2018; @violle2012], however individual-level traits are rarely collected alongside monitoring data, especially for the longest running efforts.
-Species-level traits may be obscuring more subtle shifts in the trait space happening within species.
-Likewise, available trait data may not capture the traits experiencing the greatest change.
-For example, we focus here mostly on ecological traits (i.e. foraging strategy, diet, etc) rather than life-history or reproductive traits.
-
-Third, while we use here the most common metrics for describing functional diversity, they do not measure some potentially important aspects of the functional space.
-Most notably, the summary metrics we calculated do not capture shifts in the location of the functional space as a whole.
-For example, two communities could have very similar metric values but no overlap in their trait spaces.
-This is especially relevant in the context of biodiversity change as a species loss could be replaced by a species with very different functional attributes, but the replacement would go undetected if the new species expanded the trait space by the same degree and had similar abundance.
-This scenario may be common in communities tracking changing environmental conditions.
-While trait CWM's capture axis shifts, approaches for assessing multidimensional shifts in functional space are still relatively new [@barros2016; @blonder2018; @mammola2019] but could shed critical insight into functional composition changes of this nature.
 

### 2024-03-11 paper/GEB Submission/manuscript_GEB.tex
Just... so many changes.

@@
 analyses have been heavily criticized for geographic biases, lack of
-data in the most heavily impacted areas, and not considering the
-ecological context of individual studies (10--12). Many of these
-criticisms reflect limitations of ecological data on the whole, and
-proponents argue we must make the best of the imperfect data available
-(13). Still, there is general consensus on a few local-scale patterns
-that seem to be characteristic of the anthropocene, including the
-prevalence of species turnover across communities and patterns of both
-richness increase and decrease in different taxa and biomes (14).
-Understanding the implications of change in species assemblage for
-ecosystem stability and function is a critical next step in moving
-forward the biodiversity change conversation.
-
-Functional diversity offers a potentially powerful alternative to
+data in the most heavily impacted areas, and exclusion of individual
+studies' ecological context (10--12). Many of these criticisms reflect
+limitations of ecological data on the whole,
+\textcolor{blue}{leading to a call for additional data not only to fill geographic and temporal gaps, but to flesh out key characteristics of communities}
+(13, 14).
+
+Functional diversity offers a potentially powerful addition to
 species-based approaches for detecting and describing community change
@@
 impact of losses or gains of functionally unique species. Functional
-diversity metrics are therefore particularly well suited for assessing
-community shifts underlying even constant species richness trends.
-
-Beyond simply characterizing changes in community structure, trends in
-functional composition also have important implications for ecosystem
-stability, function, and resilience. There is increasing evidence
-functional diversity is a better predictor of ecosystem function than
-species-based metrics (18, 19), and that different facets of functional
-diversity play essential roles in maintaining ecosystem stability (20,
-21). Almost all hypothesized mechanisms underpinning the relationship
-between species diversity and ecosystem function are trait-dependent
-(22). Determining functional trends therefore gives a more direct
-picture of potential trends in critical ecosystem processes.
-
-It is critical to establish whether or not functional loss is prevalent
-across communities. While functional loss is frequently cited as one of
-the most pressing concerns of the anthropocene (23--25), functional
-diversity may be maintained even when species are lost from a community
-(26, 27). Forecasts of functional loss range from negligible (28) to
-dire (29, 30). And while some observed trends show significant
-functional loss (31) others document no loss even in some of the most
-heavily impacted communities (32--34). On paleoecological time scales
+diversity metrics
+\textcolor{blue}{may therefore illuminate joint responses from functionally similar species or communities undetectible by looking at species identity alone.}
+
+\textcolor{blue}{The expectation for functional change across communities is not obvious from past work, and may or may not follow taxonomic trends}
+(13). While functional loss is frequently cited as one of the most
+pressing concerns of the anthropocene (18--20),
+\textcolor{blue}{functional diversity may be maintained even when species are lost from a community}
+(21, 22). Forecasts of functional loss range from negligible (23) to
+dire (24, 25). And while some observed trends show significant
+functional loss (26) others document no loss even in some of the most
+heavily impacted communities (27--29). On paleoecological time scales,
 functional composition shows mixed responses to environmental change and
-extinction events (35, 36), with significant impacts of species
-extinctions on functional diversity in some taxa and not others (37).
+extinction events (30, 31), with significant impacts of species
+extinctions on functional diversity in some taxa and not others (32).
 For some time periods, functional structure appears to be maintained for
-substantial portions of geological time (38). Still, some losses of
-functional diversity are indisputable on both paleoecological and
-contemporary timescales such as continued trophic downgrading due to
-loss of large-bodied mammals, but implications of those losses for local
-diversity patterns are less clear (39, 40) .
-
-Assessments of broad-scale temporal change in functional diversity have
-previously been limited by a lack of functional trait data. The majority
-of work has therefore focused largely on system-specific studies with
-traits collected \emph{in situ}. This has challenged our ability to
-establish general rules for how functional composition may be changing
-through time, particularly in response to global change drivers. Ongoing
-efforts to assemble functional traits for a variety of taxa have made
-synthesis of existing community assemblage data and functional traits
-possible for the first time, providing initial insights into the ways
-functional diversity changes on a broad scale for specific taxa
-(e.g.~fish, 41, birds, 42, 43). However, to date there has been no
-cross-taxa assessment of temporal functional change for a broad
-geographic and taxonomic extent (44).
-
-Here we perform the first multi-taxa, multi-realm assessment of
-functional diversity change through time. We focus on mammal and bird
-species as a significant subset of the world's biodiversity heavily
-impacted by anthropogenic change. While examining trends in plants,
-invertebrates, and other vertebrate species is of equal interest, trait
-data for those taxa raise additional challenges such as limited and
-biased species coverage (45), a lack of accepted species-level means,
-and differences in the types of traits collected. To ensure
-comparability across taxa in trait type and data quality we therefore
-focus on mammals and birds. Traits were intentionally selected to be
-representative of a species' Eltonian niche, thereby summarizing the
-functional role they play in the community (46). An initial assessment
-of amphibian trends is included in the supplement, but excluded from
-general trend assessment here due to limited geographic coverage.
+substantial portions of geological time (33).
+\textcolor{blue}{Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds [@barnagaud2017; @jarzyna2016] and ray-finned fishes, sharks, and rays}
+(34).
+
+Here we
+\textcolor{blue}{leverage ongoing efforts to assemble functional trait data and computational advances to}
+perform the first multi-taxa, multi-realm assessment of functional
+diversity change through time. We focus on mammal and bird species as a
+significant subset of the world's biodiversity heavily impacted by
+anthropogenic change. While examining trends in plants, invertebrates,
+and other vertebrate species is of equal interest, trait data for those
+taxa raise additional challenges such as limited and biased species
+coverage (35), a lack of accepted species-level means, and differences
+in the types of traits collected. To ensure comparability across taxa in
+trait type and data quality we therefore focus on mammals and birds.
+Traits were intentionally selected to be representative of a species'
+Eltonian niche, thereby summarizing the functional role they play in the
+community (36). An initial assessment of amphibian trends is included in
+the supplement, but excluded from general trend assessment here due to
+limited geographic coverage.
 
 We assess thousands of mammal and bird functional diversity time series
-to determine whether or not there is a general trend of functional
-change, both in observed metrics and in metrics corrected for changes in
-species richness. We further assess the prevailing modes of change
-occurring at the study level to determine how functional change is
-related to species patterns, and which kinds of species and functional
-change are most common across communities. We expect to find significant
-functional changes happening alongside and independent of species
-richness changes as a result of ongoing anthropogenic impacts and global
-change drivers.
+to determine whether or not
+\textcolor{blue}{the addition of functional trait data gives a clarifying picture of biodiversity change across communities.
+Rather than testing specific hypotheses of change we present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination.
+We present results at three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm), and trends for individual studies.}
 
@@
 represent full assemblages rather than populations of single species
-(36). All time series include abundance of observed species. Following
-best practices for the database (47), studies with multiple sample
+(31). All time series include abundance of observed species. Following
+best practices for the database (37), studies with multiple sample
 locations were split into individual time series following a
@@
 size determined based on the sample extent of studies with only a single
-location (see 36 for details on how sample extents were defined), with
+location (see 31 for details on how sample extents were defined), with
 the area of each cell set to one standard deviation away from the mean
@@
 species-level means for traits that represent species' multifaceted role
-in the community (46). Traits include: body mass, diet, active diel
+in the community (36). Traits include: body mass, diet, active diel
 period, nocturnality, forest foraging strata, pelagic use. Multiple
@@
 \url{https://doi.org/10.5066/F7KH0KBK}), obtained through the
-\texttt{taxadb} R package (48, 49). If more than one species in the
+\texttt{taxadb} R package (38, 39). If more than one species in the
 assemblage data resolved to the same identifier, observations were
@@
 earliest sample was in 1923 and the most recent was in 2014. For a full
-breakdown of studies and their characteristics, see the supplement.
+breakdown of studies and their characteristics, see the supplement. Our
+final dataset reflects many of the data biases that make global
+synthesis work challenging, including geographic bias, a bias away from
+areas currently under the greatest threat, and a bias towards shorter
+time series. We address these shortcomings and their potential impact on
+our results in the discussion.
 
@@
 Functional diversity metrics were calculated using the \emph{dbFD}
-function from the \emph{FD} R package (50). Here we report functional
+function from the \emph{FD} R package (40). Here we report functional
 richness (\emph{FRic}), functional evenness (\emph{FEve}), and
 functional divergence (\emph{FDiv}) which together describe three
-complementary characteristics of the functional space (22, 51).
+complementary characteristics of the functional space (41, 42).
 \emph{FRic} assesses the volume of the trait space occupied by species
@@
 shifts in the distribution of each trait.
+\textcolor{blue}{Hereafter we refer to results for functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).}
 
@@
 traits were z-score scaled to give each trait equal weight in the trait
-space (52,
-53).\textcolor{blue}{Before calculating diversity metrics, dbFD reduces the dimensionality of the trait space by performing PCoA. We limited the number of included PCoA axes to}
+space (43, 44).
+\textcolor{blue}{Before calculating diversity metrics, dbFD reduces the dimensionality of the trait space by performing PCoA. We limited the number of included PCoA axes to}
 the maximum number of traits that fulfills the criteria \(s >= 2^t\),
@@
 traits. This restriction allows for enough axes to capture the trait
-space while maintaining computational feasibility (54). Metrics
+space while maintaining computational feasibility (45). Metrics
 incorporated weighting based on species abundance.
@@
 calculated the standardized effect size (SES) for each of the three
-summary functional diversity metrics (\emph{FRic}, \emph{FEve},
-\emph{FDiv}) from null estimates (55). Null model corrections allow us
-to assess the degree to which the observed functional diversity metric
-deviates from the value expected by chance in a randomly assembled
-community. Null estimates were calculated for each rarefied sample by
-randomly sampling species from the species pool for each year and
-randomly assigning observed abundances to species. Species pools
-included all species observed for a time series. This process was
-repeated 500 times to get an estimate and standard deviation of the null
-expectation for the metric for each rarefaction sample for that time
-series. We used these values to calculate SES using the following
-formula: \(SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}\). We
-then calculated the median SES estimate for each metric from all the
+functional diversity metrics (\emph{FRic}, \emph{FEve}, \emph{FDiv})
+from null estimates (46). Null model corrections allow us to assess the
+degree to which the observed functional diversity metric deviates from
+the value expected by chance in a randomly assembled community. Null
+estimates were calculated for each rarefied sample by randomly sampling
+species from the species pool for each year and randomly assigning
+observed abundances to species. `
+\textcolor{blue}{Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability}.
+This process was repeated 500 times to get an estimate and standard
+deviation of the null expectation for the metric for each rarefaction
+sample for that time series. We used these values to calculate SES using
+the following formula:
+\(SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}\). We then
+calculated the median SES estimate for each metric from all the
 rarefaction samples for a time series. SES estimates can be interpreted
@@
 
-We estimated general trends for each diversity metric using a linear
-mixed effects model with a random slope and intercept for each study and
-each time series nested within the study. We fit 18 individual
-\emph{CWM} models. All time series with data for a given trait were
-included in the corresponding \emph{CWM} model. We estimated study level
-trends using individual linear models. For studies with more than one
-times series we fit a random slope and intercept for time series. Some
-study-level models could not be fit for five studies for at least one
-metric due to data limitations, but those studies were still included in
-the general models. They represented 12 of 1350 study-level models fit
-for each metric. For further details see the supplement. Where
-appropriate, response variables were \(log\) or \(log(x+1)\) transformed
-to better fit model assumptions of residual normality.
+We estimated general trends
+\textcolor{blue}{across bird and mammals communities} for each diversity
+metric using a linear mixed effects model with a random slope and
+intercept for each study and each time series nested within the study.
+We fit 18 individual \emph{CWM} models. All time series with data for a
+given trait were included in the corresponding \emph{CWM} model. We
+estimated study level trends using individual linear models. For studies
+with more than one times series we fit a random slope and intercept for
+time series. Some study-level models could not be fit for five studies
+for at least one metric due to data limitations, but those studies were
+still included in the general models. They represented 12 of 1350
+study-level models fit for each metric. For further details see the
+supplement. Where appropriate, response variables were \(log\) or
+\(log(x+1)\) transformed to better fit model assumptions of residual
+normality.
 
@@
 slopes and calculated between-level contrasts using the \emph{emmeans}
-package (v1.8.2, 56). We assessed the impact of time series duration and
+package (v1.8.2, 47). We assessed the impact of time series duration and
 start year on study-level trends using linear models with duration and
@@
 (v3.1-3) package with a significance level of \(\alpha = 0.05\)
-(57--59).
+(48--50).
 
@@
 
-We found no significant overall trend in species richness or summary
-functional diversity metrics (observed or standardized) (Fig
+We found no significant overall trend in species richness or functional
+diversity metrics (observed or \textcolor{blue}{corrected}) (Fig
 \ref{fig:timeseriesPlot}). We did find a significant overall decrease in
@@
 studies experience increasing or decreasing trends, the average trend
-across studies was plausibly 0 (Table \ref{tab:resultsTab}). Trends for
-different taxa, biomes, or realms were also non-significant for richness
-and summary functional diversity metrics, with the exception of a
-significantly increasing trend for functional evenness of global studies
-(characterized by having samples on multiple continents), and a
-significantly decreasing functional richness slope for
-temperate/tropical studies and mammal studies. However, trends were not
-exhibited for the standardized metric indicating that differences were
-largely due to changes in species richness. Further, with only two
-global studies, the trend should not be considered truly general. The
-general trends for \emph{CWM} models were similarly not significant,
-with a significant positive trend for only percentage of fish in diet
-composition.
+across studies was plausibly 0 (Table \ref{tab:resultsTab}).
+\textcolor{blue}{Within-group} trends for different taxa, biomes, or
+realms were also non-significant for richness and functional diversity
+metrics, with the exception of a significantly increasing trend for
+functional evenness of global studies (characterized by having samples
+on multiple continents), and a significantly decreasing functional
+richness slope for temperate/tropical studies and mammal studies.
+However, trends were not exhibited for the \textcolor{blue}{corrected}
+metric indicating that differences
+\textcolor{blue}{in functional diversity metrics} were largely due to
+changes in species richness. Further, with only two global studies, the
+trend should not be considered truly general. The general trends for
+\emph{CWM} models were similarly not significant, with a significant
+positive trend for only percentage of fish in diet composition.
 
@@
 metric, and 15 of 50 studies exhibited a significant trend for at least
-one standardized metrics (Table \ref{tab:trendTab}). In general, there
-were more significant trends for uncorrected metrics, with some
-disappearing after correction, indicating that those trends were likely
-due to changes in the number of species. Hypothesis testing for
-study-level trends is likely affected by multiple testing issues and
+one \textcolor{blue}{corrected} metric (Table \ref{tab:trendTab}). In
+general, there were more significant trends for uncorrected metrics,
+with some disappearing after correction, indicating that those trends
+were likely due to changes in the number of species. Hypothesis testing
+for study-level trends is likely affected by multiple testing issues and
 some trends identified as significant are therefore potentially
@@
 
-Study-level slopes for summary functional metrics were significantly
+Study-level slopes for functional diversity metrics were significantly
 related to start year of the time series for Jaccard similarity and
 functional evenness, both of which had significantly more negative
-slopes with more recent start year. No summary functional metrics were
+slopes with more recent start year. No functional diversity metrics were
 significantly related to the start year of the time series.
@@
 and local trends in functional diversity in mammal and bird communities.
-Surprisingly, we did not detect an overall trend in any of the summary
-functional diversity metrics. As with previous species-based syntheses,
-we also found no overall trend in species richness accompanied by
-increasing turnover through time (36), indicating that non-significant
-trends in functional metrics may be consistent with similar
-well-documented species derived trends. We found no trend in functional
-change for almost all realms, biomes, and taxonomic groups, with
-evidence of functional richness loss in mammal studies only. These
-results are consistent with multiple studies linking anthropogenic
-drivers to loss of functional diversity in mammal communities (60--62)
-
-Despite a lack of general trends in summary functional metrics, we did
-find multiple trends in \emph{CWM}'s for studies coming from the same
-taxa, realms, or climates, many of which are consistent with previous
-findings for those contexts. For example, the decrease in insect
-consumption for tropical communities reflects well documented declines
-in tropical insectivorous birds (63). Reduction in utilization of
-understory foraging for terrestrial communities could be the result of
-disturbance like human recreation or increased predation from introduced
-species. Other shifts in diet for both Birds and Mammals, including
-increasing fruit, seed, and vertebrate consumption and decreasing seed
-consumption in some climates point to important areas for further
-exploration.
+\textcolor{blue}{Our results show that the addition of functional traits illuminates a few consistent functional trends across communities, but largely complicates rather that clarifies the story of biodiversity change.
+While the characteristics of species clearly matter, instead of unifying the nature of communities' change, they more often distinguish them.
+In general, we found a few areas of consensus for models where communities were aggregated (general trends, trends by taxa, biome, and realm), and vast heterogeneity for study-level models.}
+
+\hypertarget{evidence-of-consensus}{%
+\subsubsection{Evidence of consensus}\label{evidence-of-consensus}}
+
+We did not detect an overall trend in any functional diversity metrics.
+As with previous species-based syntheses, we also found no overall trend
+in species richness accompanied by increasing turnover through time
+(31), indicating that non-significant trends in functional metrics may
+be consistent with similar well-documented species derived trends. We
+found no trend in functional change for almost all realms, biomes, and
+taxonomic groups.
+\textcolor{blue}{The one exception was a consensus trend indicating}
+functional richness loss for mammal studies. These results are
+consistent with multiple studies linking anthropogenic drivers to loss
+of functional diversity in mammal communities (51--53).
+
+Despite a lack of general trends in functional diversity metrics, we did
+find evidence of compositional shifts in the trait space with multiple
+\texttt{significant} trends in \emph{CWM}'s for studies coming from the
+same taxa, realms, or climates,
+\texttt{indicating\ areas\ of\ consensus\ change\ across\ communities}.
+Some of those trends are consistent with existing work in those
+contexts. For example, the decrease in insect consumption for tropical
+communities reflects well documented declines in tropical insectivorous
+birds (54). Reduction in utilization of understory foraging for
+terrestrial communities could be the result of disturbance like human
+recreation or increased predation from introduced species. Other shifts
+in diet for both Birds and Mammals, including increasing fruit, seed,
+and vertebrate consumption and decreasing seed consumption in some
+climates point to important areas for further exploration.
+
+\hypertarget{study-heterogeneity}{%
+\subsubsection{Study heterogeneity}\label{study-heterogeneity}}
+
+The strongest picture the study-level results give is that trait level
+data further complicates the story of biodiversity change. The lack of
+general trend in functional diversity metrics belies a huge range of
+positive and negative trends at the study-level that happen to average
+out to zero. In order to simplify discussion, we will talk about the
+implications for communities grouped by their concurrent change in
+functional diversity metrics, species richness and turnover split into
+positive, negative or no trend. While there are over 150 possible
+combinations of change direction (or no change) in the 5 metrics, we
+discuss here the six scenarios that occurred in more than one study: no
+change, species loss or gain only, loss of functional evenness, richness
+loss with species turnover, and increase in species and functional
+richness accompanied by significant turnover. We focus on these
+scenarios to combat potential spurious results due to multiple testing,
+as it is unlikely all observations of the same scenario are spurious.
+The number of studies in each scenarios is broken down in table x to
+illustrate the relative number between groups rather than absolute
+number.
+
+The majority group of studies exhibited no trend in any species or
+functional diversity metric.
 
@@
 corrected richness but a trend in other corrected functional metrics.
-This would indicate that when species are lost from a community, they
-are not more functionally unique than expected by chance, though they
-may produce shifts in other aspects of functional structure. When
-species are gained however, they disproportionately add to the
-functional richness of the community.
+\textcolor{blue}{This would indicate that when species are gained in a community, they disproportionately add to the functional richness.
+When species are lost however, they are not more functionally unique than expected by chance though they may produce shifts in other aspects of functional space.}
 
@@
 functionally general species are being lost even as rare species are
-increasing (64--66).
+increasing (55--57).
 
@@
 purely due to background processes, which is likely further accelerated
-by global change drivers (14). While we do not address how observed
+by global change drivers (13). While we do not address how observed
 rates differ from background expectation here, we can make some
@@
 Where turnover is occurring, it is almost always accompanied by a
+\textcolor{blue}{trend in one of the functional diversity metrics, indicating}
 directional change in the functional space. Species are not therefore
@@
 observed both increases and decreases in corrected functional richness
-accompanying turnover is a reminder that replacements are not always an
-indication of functional loss but can also lead to increased functional
-richness. While shifts in functional space may be the result of
-environmental pressure, we cannot distinguish between directional change
-and shifts due to the inavailability of an analog species in the species
-pool.
+accompanying turnover is a reminder that
+\textcolor{blue}{new species are not always replacing more functionally unique species, but are often more functionally unique themselves.}
 
@@
 under global change. For example, mean body size is predicted to
-decrease as a result of climate change impacts and megafaunal loss (67),
+decrease as a result of climate change impacts and megafaunal loss (58),
 a phenomenon which has already been well documented empirically and
-experimentally in multiple taxa (68--71). While the species-level trait
+experimentally in multiple taxa (59--62). While the species-level trait
 means used here are not appropriate for assessing intraspecific body
@@
 experienced significant loss of large-bodied species before the
-observation window, with contemporary loss rates slowing (72). Trends
+observation window, with contemporary loss rates slowing (63). Trends
 could be significantly different for the same time periods in regions of
@@
 dataset but where megafauna exist on the landscape and are increasingly
-threatened (73).
+threatened (64).
 
@@
 declines in function to declines in functional or species diversity (7,
-18, 74). By those criteria very few communities in our dataset are in a
+65, 66). By those criteria very few communities in our dataset are in a
 state of concern for loss of functionality (Table \ref{tab:trendTab}),
@@
 multifunctionality were not measured here (e.g.~dispersion, rarity,
-abundance of dominant species, 75).
-
-\hypertarget{potential-methodological-limitations}{%
-\subsection{Potential Methodological
-Limitations}\label{potential-methodological-limitations}}
-
-Here we approach the question of functional change using the best
-available data and biodiversity synthesis approaches. While we found
-signatures of change at the study level, some gaps in best practices may
-be obscuring a true general trend at the aggregate level. First, the
-BioTIME database, while the most comprehensive data source of time
-series available, is limited in temporal and geographic scope. Most time
-series span only a few years (Fig \ref{fig:taxaMap}) and may not provide
-the statistical power necessary to detect trends (76). The database is
-also not a representative sample of the world's biodiversity or areas of
-greatest threat (6, 12), and the subset of data in this study is
-predominantly from North America. We may simply not have data from those
-areas experiencing the greatest perturbation (77), particularly
-scenarios of conversion to urban, human-dominated landscapes. While
-evidence from other work shows even disturbed communities can maintain
-functional structure (32, 34), these results should not be interpreted
-as evidence of low functional impact in areas of heavy human
-disturbance.
-
-Second, despite using the most comprehensive trait databases for these
-taxa, we were still limited to species-level means of the traits deemed
-important by database creators. The importance of intraspecific
-variation is well documented (78, 79), however individual-level traits
-are rarely collected alongside monitoring data, especially for the
-longest running efforts. Species-level traits may be obscuring more
-subtle shifts in the trait space happening within species. Likewise,
-available trait data may not capture the traits experiencing the
-greatest change. For example, we focus here mostly on ecological traits
-(i.e.~foraging strategy, diet, etc) rather than life-history or
-reproductive traits.
-
-Third, while we use here the most common metrics for describing
-functional diversity, they do not measure some potentially important
-aspects of the functional space. Most notably, the summary metrics we
-calculated do not capture shifts in the location of the functional space
-as a whole. For example, two communities could have very similar metric
-values but no overlap in their trait spaces. This is especially relevant
-in the context of biodiversity change as a species loss could be
-replaced by a species with very different functional attributes, but the
-replacement would go undetected if the new species expanded the trait
-space by the same degree and had similar abundance. This scenario may be
-common in communities tracking changing environmental conditions. While
-trait CWM's capture axis shifts, approaches for assessing
-multidimensional shifts in functional space are still relatively new
-(54, 80, 81) but could shed critical insight into functional composition
-changes of this nature.
+abundance of dominant species, 67).
 
@@
 are simply not representative of the kinds of impacts of greatest policy
-concern (31, 82, 83).
+concern (26, 68, 69).
 
@@
 will require continued efforts to fill data gaps, a well recognized
-challenge in ecology (6, 12, 77). Still, a ``scenarios of change''
+challenge in ecology (6, 12, 70). Still, a ``scenarios of change''
 framework can provide structure for future work addressing functional
@@
 further hampered by the use of species-level trait data, as the traits
-most closely linked to a process of interest may not be available (84).
+most closely linked to a process of interest may not be available (71).
 Future trait collection that explicitly considers existing frameworks
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-primack2018}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-dornelas2023}{}}%
 \CSLLeftMargin{13. }%
-\CSLRightInline{Primack RB, et al. (2018)
-\href{https://doi.org/10.1016/j.biocon.2017.12.023}{Biodiversity gains?
-The debate on changes in local- vs global-scale species richness}.
-\emph{Biological Conservation} 219:A1--A3.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-dornelas2023}{}}%
-\CSLLeftMargin{14. }%
 \CSLRightInline{Dornelas M, et al. (2023)
@@
 378(1881):20220199.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-primack2018}{}}%
+\CSLLeftMargin{14. }%
+\CSLRightInline{Primack RB, et al. (2018)
+\href{https://doi.org/10.1016/j.biocon.2017.12.023}{Biodiversity gains?
+The debate on changes in local- vs global-scale species richness}.
+\emph{Biological Conservation} 219:A1--A3.}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-cadotte2011}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-cardinale2012}{}}%
 \CSLLeftMargin{18. }%
-\CSLRightInline{Cadotte MW, Carscadden K, Mirotchnick N (2011)
-\href{https://doi.org/10.1111/j.1365-2664.2011.02048.x}{Beyond species:
-functional diversity and the maintenance of ecological processes and
-services}. \emph{Journal of Applied Ecology} 48(5):1079--1087.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gagic2015}{}}%
+\CSLRightInline{Cardinale BJ, et al. (2012)
+\href{https://doi.org/10.1038/nature11148}{Biodiversity loss and its
+impact on humanity}. \emph{Nature} 486(7401):59--67.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-dirzo2014}{}}%
 \CSLLeftMargin{19. }%
-\CSLRightInline{Gagic V, et al. (2015)
-\href{https://doi.org/10.1098/rspb.2014.2620}{Functional identity and
-diversity of animals predict ecosystem functioning better than
-species-based indices}. \emph{Proceedings of the Royal Society B:
-Biological Sciences} 282(1801):20142620.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-craven2018}{}}%
+\CSLRightInline{Dirzo R, et al. (2014)
+\href{https://doi.org/10.1126/science.1251817}{Defaunation in the
+Anthropocene}. \emph{Science} 345(6195):401--406.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-young2016}{}}%
 \CSLLeftMargin{20. }%
-\CSLRightInline{Craven D, et al. (2018)
-\href{https://doi.org/10.1038/s41559-018-0647-7}{Multiple facets of
-biodiversity drive the diversity{\textendash}stability relationship}.
-\emph{Nature Ecology \& Evolution}:1.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-morin2014}{}}%
+\CSLRightInline{Young HS, McCauley DJ, Galetti M, Dirzo R (2016)
+\href{https://doi.org/10.1146/annurev-ecolsys-112414-054142}{Patterns,
+causes, and consequences of anthropocene defaunation}. \emph{Annual
+Review of Ecology, Evolution, and Systematics} 47(1):333--358.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-daz2001}{}}%
 \CSLLeftMargin{21. }%
-\CSLRightInline{Morin X, Fahse L, Mazancourt C de, Scherer-Lorenzen M,
-Bugmann H (2014) \href{https://doi.org/10.1111/ele.12357}{Temporal
-stability in forest productivity increases with tree diversity due to
-asynchrony in species dynamics}. \emph{Ecology Letters}
-17(12):1526--1535.}
+\CSLRightInline{Dıáz S, Cabido M (2001)
+\href{https://doi.org/10.1016/S0169-5347(01)02283-2}{Vive la différence:
+Plant functional diversity matters to ecosystem processes}. \emph{Trends
+in Ecology \& Evolution} 16(11):646--655.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-petchey2009}{}}%
+\CSLLeftMargin{22. }%
+\CSLRightInline{Petchey OL, Gaston KJ (2009)
+\href{https://doi.org/10.1007/s12080-009-0041-9}{Effects on ecosystem
+resilience of biodiversity, extinctions, and the structure of regional
+species pools}. \emph{Theoretical Ecology} 2(3):177--187.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-gallagher2013}{}}%
+\CSLLeftMargin{23. }%
+\CSLRightInline{Gallagher RV, Hughes L, Leishman MR (2013)
+\href{https://doi.org/10.1111/j.1600-0587.2012.07514.x}{Species loss and
+gain in communities under future climate change: consequences for
+functional diversity}. \emph{Ecography} 36(5):531--540.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-petchey2002}{}}%
+\CSLLeftMargin{24. }%
+\CSLRightInline{Petchey OL, Gaston KJ (2002)
+\href{https://doi.org/10.1098/rspb.2002.2073}{Extinction and the loss of
+functional diversity}. \emph{Proceedings of the Royal Society of London
+Series B: Biological Sciences} 269(1501):1721--1727.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-pimiento2020}{}}%
+\CSLLeftMargin{25. }%
+\CSLRightInline{Pimiento C, et al. (2020)
+\href{https://doi.org/10.1126/sciadv.aay7650}{Functional diversity of
+marine megafauna in the Anthropocene}. \emph{Science Advances}
+6(16):eaay7650.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-flynn2009}{}}%
+\CSLLeftMargin{26. }%
+\CSLRightInline{Flynn DFB, et al. (2009)
+\href{https://doi.org/10.1111/j.1461-0248.2008.01255.x}{Loss of
+functional diversity under land use intensification across multiple
+taxa}. \emph{Ecology Letters} 12(1):22--33.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-edwards2013}{}}%
+\CSLLeftMargin{27. }%
+\CSLRightInline{Edwards FA, Edwards DP, Hamer KC, Davies RG (2013)
+\href{https://doi.org/10.1111/ibi.12027}{Impacts of logging and
+conversion of rainforest to oil palm on the functional diversity of
+birds in Sundaland}. \emph{Ibis} 155(2):313--326.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-larsen2018}{}}%
+\CSLLeftMargin{28. }%
+\CSLRightInline{Larsen S, Chase JM, Durance I, Ormerod SJ (2018)
+\href{https://doi.org/10.1002/ecy.2213}{Lifting the veil: richness
+measurements fail to detect systematic biodiversity change over three
+decades}. \emph{Ecology} 99(6):1316--1326.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-matuoka2020}{}}%
+\CSLLeftMargin{29. }%
+\CSLRightInline{Matuoka MA, Benchimol M, Almeida-Rocha JM de,
+Morante-Filho JC (2020)
+\href{https://doi.org/10.1016/j.ecolind.2020.106471}{Effects of
+anthropogenic disturbances on bird functional diversity: A global
+meta-analysis}. \emph{Ecological Indicators} 116:106471.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-jackson2015}{}}%
+\CSLLeftMargin{30. }%
+\CSLRightInline{Jackson ST, Blois JL (2015)
+\href{https://doi.org/10.1073/pnas.1403664111}{Community ecology in a
+changing environment: Perspectives from the Quaternary}.
+\emph{Proceedings of the National Academy of Sciences}
+112(16):4915--4921.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-dornelas2018}{}}%
+\CSLLeftMargin{31. }%
+\CSLRightInline{Dornelas M, et al. (2018)
+\href{https://doi.org/10.1111/geb.12729}{BioTIME: A database of
+biodiversity time series for the Anthropocene}. \emph{Global Ecology and
+Biogeography} 27(7):760--786.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-pimiento2017}{}}%
+\CSLLeftMargin{32. }%
+\CSLRightInline{Pimiento C, et al. (2017)
+\href{https://doi.org/10.1038/s41559-017-0223-6}{The Pliocene marine
+megafauna extinction and its impact on functional diversity}.
+\emph{Nature Ecology \& Evolution} 1(8):1100--1106.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hedberg2022}{}}%
+\CSLLeftMargin{33. }%
+\CSLRightInline{Hedberg CP, Lyons SK, Smith FA (2022)
+\href{https://doi.org/10.1111/geb.13428}{The hidden legacy of megafaunal
+extinction: Loss of functional diversity and resilience over the Late
+Quaternary at Hall{'}s Cave}. \emph{Global Ecology and Biogeography}
+31(2):294--307.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-trindade-santos2020}{}}%
+\CSLLeftMargin{34. }%
+\CSLRightInline{Trindade-Santos I, Moyes F, Magurran AE (2020)
+\href{https://doi.org/10.1098/rspb.2020.0889}{Global change in the
+functional diversity of marine fisheries exploitation over the past 65
+years}. \emph{Proceedings of the Royal Society B: Biological Sciences}
+287(1933):20200889.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-fitzjohn2014}{}}%
+\CSLLeftMargin{35. }%
+\CSLRightInline{FitzJohn RG, et al. (2014)
+\href{https://doi.org/10.1111/1365-2745.12260}{How much of the world is
+woody?} \emph{Journal of Ecology} 102(5):1266--1272.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-wilman2014}{}}%
+\CSLLeftMargin{36. }%
+\CSLRightInline{Wilman H, et al. (2014)
+\href{https://doi.org/10.1890/13-1917.1}{EltonTraits 1.0: Species-level
+foraging attributes of the world's birds and mammals: {\emph{Ecological
+Archives}} E095-178}. \emph{Ecology} 95(7):2027--2027.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-blowes2019}{}}%
+\CSLLeftMargin{37. }%
+\CSLRightInline{Blowes SA, et al. (2019)
+\href{https://doi.org/10.1126/science.aaw1620}{The geography of
+biodiversity change in marine and terrestrial assemblages}.
+\emph{Science} 366(6463):339--345.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-norman2020}{}}%
+\CSLLeftMargin{38. }%
+\CSLRightInline{Norman KEA, Chamberlain S, Boettiger C (2020)
+\href{https://doi.org/10.1111/2041-210X.13440}{taxadb: A
+high-performance local taxonomic database interface}. \emph{Methods in
+Ecology and Evolution} 11(9):1153--1159.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2021}{}}%
+\CSLLeftMargin{39. }%
+\CSLRightInline{R Core Team (2021) \emph{R: A language and environment
+for statistical computing} (R Foundation for Statistical Computing,
+Vienna, Austria) Available at: \url{https://www.R-project.org/}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-lalibertuxe92010}{}}%
+\CSLLeftMargin{40. }%
+\CSLRightInline{Laliberté E, Legendre P (2010)
+\href{https://doi.org/10.1890/08-2244.1}{A distance-based framework for
+measuring functional diversity from multiple traits}. \emph{Ecology}
+91(1):299--305.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-mason2005}{}}%
+\CSLLeftMargin{41. }%
+\CSLRightInline{Mason NWH, Mouillot D, Lee WG, Wilson JB, Setälä H
+(2005) \href{http://www.jstor.org/stable/3548774}{Functional richness,
+functional evenness and functional divergence: The primary components of
+functional diversity}. \emph{Oikos} 111(1):112--118.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-hillebrand2009}{}}%
-\CSLLeftMargin{22. }%
+\CSLLeftMargin{42. }%
 \CSLRightInline{Hillebrand H, Matthiessen B (2009)
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-cardinale2012}{}}%
-\CSLLeftMargin{23. }%
-\CSLRightInline{Cardinale BJ, et al. (2012)
-\href{https://doi.org/10.1038/nature11148}{Biodiversity loss and its
-impact on humanity}. \emph{Nature} 486(7401):59--67.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-dirzo2014}{}}%
-\CSLLeftMargin{24. }%
-\CSLRightInline{Dirzo R, et al. (2014)
-\href{https://doi.org/10.1126/science.1251817}{Defaunation in the
-Anthropocene}. \emph{Science} 345(6195):401--406.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-young2016}{}}%
-\CSLLeftMargin{25. }%
-\CSLRightInline{Young HS, McCauley DJ, Galetti M, Dirzo R (2016)
-\href{https://doi.org/10.1146/annurev-ecolsys-112414-054142}{Patterns,
-causes, and consequences of anthropocene defaunation}. \emph{Annual
-Review of Ecology, Evolution, and Systematics} 47(1):333--358.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-daz2001}{}}%
-\CSLLeftMargin{26. }%
-\CSLRightInline{Dıáz S, Cabido M (2001)
-\href{https://doi.org/10.1016/S0169-5347(01)02283-2}{Vive la différence:
-Plant functional diversity matters to ecosystem processes}. \emph{Trends
-in Ecology \& Evolution} 16(11):646--655.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-petchey2009}{}}%
-\CSLLeftMargin{27. }%
-\CSLRightInline{Petchey OL, Gaston KJ (2009)
-\href{https://doi.org/10.1007/s12080-009-0041-9}{Effects on ecosystem
-resilience of biodiversity, extinctions, and the structure of regional
-species pools}. \emph{Theoretical Ecology} 2(3):177--187.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gallagher2013}{}}%
-\CSLLeftMargin{28. }%
-\CSLRightInline{Gallagher RV, Hughes L, Leishman MR (2013)
-\href{https://doi.org/10.1111/j.1600-0587.2012.07514.x}{Species loss and
-gain in communities under future climate change: consequences for
-functional diversity}. \emph{Ecography} 36(5):531--540.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-petchey2002}{}}%
-\CSLLeftMargin{29. }%
-\CSLRightInline{Petchey OL, Gaston KJ (2002)
-\href{https://doi.org/10.1098/rspb.2002.2073}{Extinction and the loss of
-functional diversity}. \emph{Proceedings of the Royal Society of London
-Series B: Biological Sciences} 269(1501):1721--1727.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-pimiento2020}{}}%
-\CSLLeftMargin{30. }%
-\CSLRightInline{Pimiento C, et al. (2020)
-\href{https://doi.org/10.1126/sciadv.aay7650}{Functional diversity of
-marine megafauna in the Anthropocene}. \emph{Science Advances}
-6(16):eaay7650.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-flynn2009}{}}%
-\CSLLeftMargin{31. }%
-\CSLRightInline{Flynn DFB, et al. (2009)
-\href{https://doi.org/10.1111/j.1461-0248.2008.01255.x}{Loss of
-functional diversity under land use intensification across multiple
-taxa}. \emph{Ecology Letters} 12(1):22--33.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-edwards2013}{}}%
-\CSLLeftMargin{32. }%
-\CSLRightInline{Edwards FA, Edwards DP, Hamer KC, Davies RG (2013)
-\href{https://doi.org/10.1111/ibi.12027}{Impacts of logging and
-conversion of rainforest to oil palm on the functional diversity of
-birds in Sundaland}. \emph{Ibis} 155(2):313--326.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-larsen2018}{}}%
-\CSLLeftMargin{33. }%
-\CSLRightInline{Larsen S, Chase JM, Durance I, Ormerod SJ (2018)
-\href{https://doi.org/10.1002/ecy.2213}{Lifting the veil: richness
-measurements fail to detect systematic biodiversity change over three
-decades}. \emph{Ecology} 99(6):1316--1326.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-matuoka2020}{}}%
-\CSLLeftMargin{34. }%
-\CSLRightInline{Matuoka MA, Benchimol M, Almeida-Rocha JM de,
-Morante-Filho JC (2020)
-\href{https://doi.org/10.1016/j.ecolind.2020.106471}{Effects of
-anthropogenic disturbances on bird functional diversity: A global
-meta-analysis}. \emph{Ecological Indicators} 116:106471.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-jackson2015}{}}%
-\CSLLeftMargin{35. }%
-\CSLRightInline{Jackson ST, Blois JL (2015)
-\href{https://doi.org/10.1073/pnas.1403664111}{Community ecology in a
-changing environment: Perspectives from the Quaternary}.
-\emph{Proceedings of the National Academy of Sciences}
-112(16):4915--4921.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-dornelas2018}{}}%
-\CSLLeftMargin{36. }%
-\CSLRightInline{Dornelas M, et al. (2018)
-\href{https://doi.org/10.1111/geb.12729}{BioTIME: A database of
-biodiversity time series for the Anthropocene}. \emph{Global Ecology and
-Biogeography} 27(7):760--786.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-pimiento2017}{}}%
-\CSLLeftMargin{37. }%
-\CSLRightInline{Pimiento C, et al. (2017)
-\href{https://doi.org/10.1038/s41559-017-0223-6}{The Pliocene marine
-megafauna extinction and its impact on functional diversity}.
-\emph{Nature Ecology \& Evolution} 1(8):1100--1106.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hedberg2022}{}}%
-\CSLLeftMargin{38. }%
-\CSLRightInline{Hedberg CP, Lyons SK, Smith FA (2022)
-\href{https://doi.org/10.1111/geb.13428}{The hidden legacy of megafaunal
-extinction: Loss of functional diversity and resilience over the Late
-Quaternary at Hall{'}s Cave}. \emph{Global Ecology and Biogeography}
-31(2):294--307.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-estes2011}{}}%
-\CSLLeftMargin{39. }%
-\CSLRightInline{Estes JA, et al. (2011)
-\href{https://doi.org/10.1126/science.1205106}{Trophic Downgrading of
-Planet Earth}. \emph{Science} 333(6040):301--306.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-smith2018}{}}%
-\CSLLeftMargin{40. }%
-\CSLRightInline{Smith FA, Elliott Smith RE, Lyons SK, Payne JL (2018)
-\href{https://doi.org/10.1126/science.aao5987}{Body size downgrading of
-mammals over the late quaternary}. \emph{Science} 360(6386):310--313.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-trindade-santos2020}{}}%
-\CSLLeftMargin{41. }%
-\CSLRightInline{Trindade-Santos I, Moyes F, Magurran AE (2020)
-\href{https://doi.org/10.1098/rspb.2020.0889}{Global change in the
-functional diversity of marine fisheries exploitation over the past 65
-years}. \emph{Proceedings of the Royal Society B: Biological Sciences}
-287(1933):20200889.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-jarzyna2016}{}}%
-\CSLLeftMargin{42. }%
-\CSLRightInline{Jarzyna MA, Jetz W (2016)
-\href{https://doi.org/10.1111/gcb.13571}{A near half-century of temporal
-change in different facets of avian diversity}. \emph{Global Change
-Biology} 23(8):2999--3011.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-barnagaud2017}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-leps2006}{}}%
 \CSLLeftMargin{43. }%
-\CSLRightInline{Barnagaud J-Y, Gaüzère P, Zuckerberg B, Princé K,
-Svenning J-C (2017)
-\href{https://doi.org/10.1007/s00442-017-3967-4}{Temporal changes in
-bird functional diversity across the United States}. \emph{Oecologia}
-185(4):737--748.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-duxedaz2022}{}}%
-\CSLLeftMargin{44. }%
-\CSLRightInline{Díaz S, Malhi Y (2022)
-\href{https://doi.org/10.1146/annurev-environ-120120-054300}{Biodiversity:
-Concepts, patterns, trends, and perspectives}. \emph{Annual Review of
-Environment and Resources} 47(1):31--63.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-fitzjohn2014}{}}%
-\CSLLeftMargin{45. }%
-\CSLRightInline{FitzJohn RG, et al. (2014)
-\href{https://doi.org/10.1111/1365-2745.12260}{How much of the world is
-woody?} \emph{Journal of Ecology} 102(5):1266--1272.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-wilman2014}{}}%
-\CSLLeftMargin{46. }%
-\CSLRightInline{Wilman H, et al. (2014)
-\href{https://doi.org/10.1890/13-1917.1}{EltonTraits 1.0: Species-level
-foraging attributes of the world's birds and mammals: {\emph{Ecological
-Archives}} E095-178}. \emph{Ecology} 95(7):2027--2027.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-blowes2019}{}}%
-\CSLLeftMargin{47. }%
-\CSLRightInline{Blowes SA, et al. (2019)
-\href{https://doi.org/10.1126/science.aaw1620}{The geography of
-biodiversity change in marine and terrestrial assemblages}.
-\emph{Science} 366(6463):339--345.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-norman2020}{}}%
-\CSLLeftMargin{48. }%
-\CSLRightInline{Norman KEA, Chamberlain S, Boettiger C (2020)
-\href{https://doi.org/10.1111/2041-210X.13440}{taxadb: A
-high-performance local taxonomic database interface}. \emph{Methods in
-Ecology and Evolution} 11(9):1153--1159.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2021}{}}%
-\CSLLeftMargin{49. }%
-\CSLRightInline{R Core Team (2021) \emph{R: A language and environment
-for statistical computing} (R Foundation for Statistical Computing,
-Vienna, Austria) Available at: \url{https://www.R-project.org/}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-lalibertuxe92010}{}}%
-\CSLLeftMargin{50. }%
-\CSLRightInline{Laliberté E, Legendre P (2010)
-\href{https://doi.org/10.1890/08-2244.1}{A distance-based framework for
-measuring functional diversity from multiple traits}. \emph{Ecology}
-91(1):299--305.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-mason2005}{}}%
-\CSLLeftMargin{51. }%
-\CSLRightInline{Mason NWH, Mouillot D, Lee WG, Wilson JB, Setälä H
-(2005) \href{http://www.jstor.org/stable/3548774}{Functional richness,
-functional evenness and functional divergence: The primary components of
-functional diversity}. \emph{Oikos} 111(1):112--118.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-leps2006}{}}%
-\CSLLeftMargin{52. }%
 \CSLRightInline{Leps J, Bello F, Lavorel S, Berman S (2006) Quantifying
@@
 \leavevmode\vadjust pre{\hypertarget{ref-schleuter2010}{}}%
-\CSLLeftMargin{53. }%
+\CSLLeftMargin{44. }%
 \CSLRightInline{Schleuter D, Daufresne M, Massol F, Argillier C (2010)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-blonder2018}{}}%
-\CSLLeftMargin{54. }%
+\CSLLeftMargin{45. }%
 \CSLRightInline{Blonder B (2018)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-swenson2012}{}}%
-\CSLLeftMargin{55. }%
+\CSLLeftMargin{46. }%
 \CSLRightInline{Swenson NG, et al. (2012)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lenth2022}{}}%
-\CSLLeftMargin{56. }%
+\CSLLeftMargin{47. }%
 \CSLRightInline{Lenth RV (2022) \emph{Emmeans: Estimated marginal means,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bates2015}{}}%
-\CSLLeftMargin{57. }%
+\CSLLeftMargin{48. }%
 \CSLRightInline{Bates D, Mächler M, Bolker B, Walker S (2015) Fitting
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kuznetsova2017}{}}%
-\CSLLeftMargin{58. }%
+\CSLLeftMargin{49. }%
 \CSLRightInline{Kuznetsova A, Brockhoff PB, Christensen RHB (2017)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2023}{}}%
-\CSLLeftMargin{59. }%
+\CSLLeftMargin{50. }%
 \CSLRightInline{R Core Team (2023) \emph{R: A language and environment
@@
 \leavevmode\vadjust pre{\hypertarget{ref-brodie2021}{}}%
-\CSLLeftMargin{60. }%
+\CSLLeftMargin{51. }%
 \CSLRightInline{Brodie JF, Williams S, Garner B (2021) The decline of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-cox2022}{}}%
-\CSLLeftMargin{61. }%
+\CSLLeftMargin{52. }%
 \CSLRightInline{Cox DTC, Gardner AS, Gaston KJ (2022)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-li2022}{}}%
-\CSLLeftMargin{62. }%
+\CSLLeftMargin{53. }%
 \CSLRightInline{Li X, et al. (2022)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-sherry2021}{}}%
-\CSLLeftMargin{63. }%
+\CSLLeftMargin{54. }%
 \CSLRightInline{Sherry TW (2021) Sensitivity of tropical insectivorous
@@
 \leavevmode\vadjust pre{\hypertarget{ref-inger2015}{}}%
-\CSLLeftMargin{64. }%
+\CSLLeftMargin{55. }%
 \CSLRightInline{Inger R, et al. (2015)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rosenberg2019}{}}%
-\CSLLeftMargin{65. }%
+\CSLLeftMargin{56. }%
 \CSLRightInline{Rosenberg KV, et al. (2019) Decline of the North
@@
 \leavevmode\vadjust pre{\hypertarget{ref-schipper2016}{}}%
-\CSLLeftMargin{66. }%
+\CSLLeftMargin{57. }%
 \CSLRightInline{Schipper AM, et al. (2016)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-sheridan2011}{}}%
-\CSLLeftMargin{67. }%
+\CSLLeftMargin{58. }%
 \CSLRightInline{Sheridan JA, Bickford D (2011)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-caruso2014}{}}%
-\CSLLeftMargin{68. }%
+\CSLLeftMargin{59. }%
 \CSLRightInline{Caruso NM, Sears MW, Adams DC, Lips KR (2014)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-forster2012}{}}%
-\CSLLeftMargin{69. }%
+\CSLLeftMargin{60. }%
 \CSLRightInline{Forster J, Hirst AG, Atkinson D (2012)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-huss2019}{}}%
-\CSLLeftMargin{70. }%
+\CSLLeftMargin{61. }%
 \CSLRightInline{Huss M, Lindmark M, Jacobson P, van Dorst RM, Gårdmark A
@@
 \leavevmode\vadjust pre{\hypertarget{ref-tseng2018}{}}%
-\CSLLeftMargin{71. }%
+\CSLLeftMargin{62. }%
 \CSLRightInline{Tseng M, et al. (2018)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-fritz2009}{}}%
-\CSLLeftMargin{72. }%
+\CSLLeftMargin{63. }%
 \CSLRightInline{Fritz SA, Bininda-Emonds ORP, Purvis A (2009)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ripple2015}{}}%
-\CSLLeftMargin{73. }%
+\CSLLeftMargin{64. }%
 \CSLRightInline{Ripple WJ, et al. (2015)
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-cadotte2011}{}}%
+\CSLLeftMargin{65. }%
+\CSLRightInline{Cadotte MW, Carscadden K, Mirotchnick N (2011)
+\href{https://doi.org/10.1111/j.1365-2664.2011.02048.x}{Beyond species:
+functional diversity and the maintenance of ecological processes and
+services}. \emph{Journal of Applied Ecology} 48(5):1079--1087.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-duffy2007}{}}%
-\CSLLeftMargin{74. }%
+\CSLLeftMargin{66. }%
 \CSLRightInline{Duffy JE, et al. (2007)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bagousse-pinguet2021}{}}%
-\CSLLeftMargin{75. }%
+\CSLLeftMargin{67. }%
 \CSLRightInline{Bagousse-Pinguet YL, et al. (2021) Functional rarity and
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-wauchope2019}{}}%
-\CSLLeftMargin{76. }%
-\CSLRightInline{Wauchope HS, Amano T, Sutherland WJ, Johnston A (2019)
-\href{https://doi.org/10.1111/2041-210X.13302}{When can we trust
-population trends? A method for quantifying the effects of sampling
-interval and duration}. \emph{Methods in Ecology and Evolution}
-10(12):2067--2078.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hughes2021}{}}%
-\CSLLeftMargin{77. }%
-\CSLRightInline{Hughes AC, et al. (2021)
-\href{https://doi.org/10.1111/ecog.05926}{Sampling biases shape our view
-of the natural world}. \emph{Ecography} 44(9):1259--1269.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-desroches2018}{}}%
-\CSLLeftMargin{78. }%
-\CSLRightInline{Des Roches S, et al. (2018)
-\href{https://doi.org/10.1038/s41559-017-0402-5}{The ecological
-importance of intraspecific variation}. \emph{Nature Ecology \&
-Evolution} 2(1):57--64.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-violle2012}{}}%
-\CSLLeftMargin{79. }%
-\CSLRightInline{Violle C, et al. (2012)
-\href{https://doi.org/10.1016/j.tree.2011.11.014}{The return of the
-variance: Intraspecific variability in community ecology}. \emph{Trends
-in Ecology \& Evolution} 27(4):244--252.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-barros2016}{}}%
-\CSLLeftMargin{80. }%
-\CSLRightInline{Barros C, Thuiller W, Georges D, Boulangeat I,
-Münkemüller T (2016)
-\href{https://doi.org/10.1111/ele.12617}{N-dimensional hypervolumes to
-study stability of complex ecosystems}. \emph{Ecology Letters}
-19(7):729--742.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-mammola2019}{}}%
-\CSLLeftMargin{81. }%
-\CSLRightInline{Mammola S (2019)
-\href{https://doi.org/10.1111/jbi.13618}{Assessing similarity of
-n-dimensional hypervolumes: Which metric to use?} \emph{Journal of
-Biogeography} 46(9):2012--2023.}
-
 \leavevmode\vadjust pre{\hypertarget{ref-magioli2021}{}}%
-\CSLLeftMargin{82. }%
+\CSLLeftMargin{68. }%
 \CSLRightInline{Magioli M, et al. (2021)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-tinoco2018}{}}%
-\CSLLeftMargin{83. }%
+\CSLLeftMargin{69. }%
 \CSLRightInline{Tinoco BA, Santillán VE, Graham CH (2018)
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-hughes2021}{}}%
+\CSLLeftMargin{70. }%
+\CSLRightInline{Hughes AC, et al. (2021)
+\href{https://doi.org/10.1111/ecog.05926}{Sampling biases shape our view
+of the natural world}. \emph{Ecography} 44(9):1259--1269.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-zhu2017}{}}%
-\CSLLeftMargin{84. }%
+\CSLLeftMargin{71. }%
 \CSLRightInline{Zhu L, et al. (2017)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-pardieck2020}{}}%
-\CSLLeftMargin{85. }%
+\CSLLeftMargin{72. }%
 \CSLRightInline{Pardieck KL, David ZJ, Lutmerding M, Aponte V, Hudson
@@
 \leavevmode\vadjust pre{\hypertarget{ref-piropno}{}}%
-\CSLLeftMargin{86. }%
+\CSLLeftMargin{73. }%
 \CSLRightInline{PIROP northwest atlantic 1965{\textendash}1992 - OBIS
@@
 \leavevmode\vadjust pre{\hypertarget{ref-silva}{}}%
-\CSLLeftMargin{87. }%
+\CSLLeftMargin{74. }%
 \CSLRightInline{Silva FR da Brazil dataset 1.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-scott}{}}%
-\CSLLeftMargin{88. }%
+\CSLLeftMargin{75. }%
 \CSLRightInline{Scott D, Metts B, Lance S The rainbow bay long-term
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rossaferes1997}{}}%
-\CSLLeftMargin{89. }%
+\CSLLeftMargin{76. }%
 \CSLRightInline{Rossa-Feres D de C (1997) Community ecology of anura
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bakker1990}{}}%
-\CSLLeftMargin{90. }%
+\CSLLeftMargin{77. }%
 \CSLRightInline{Bakker C, Herman PMJ (1990) Phytoplankton in the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-berezovikovn.n.2004}{}}%
-\CSLLeftMargin{91. }%
+\CSLLeftMargin{78. }%
 \CSLRightInline{Berezovikov, N.N. (2004) The birds of settlements in
@@
 \leavevmode\vadjust pre{\hypertarget{ref-enemar2004}{}}%
-\CSLLeftMargin{92. }%
+\CSLLeftMargin{79. }%
 \CSLRightInline{Enemar A, Sjöstrand B, Andersson G, Von Proschwitz T
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gido2019}{}}%
-\CSLLeftMargin{93. }%
+\CSLLeftMargin{80. }%
 \CSLRightInline{Gido K (2019) CFP01 Fish population on selected
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hogstad1993}{}}%
-\CSLLeftMargin{94. }%
+\CSLLeftMargin{81. }%
 \CSLRightInline{Hogstad O (1993)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-holmes1986}{}}%
-\CSLLeftMargin{95. }%
+\CSLLeftMargin{82. }%
 \CSLRightInline{Holmes RT, Sherry TW, Sturges FW (1986)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-jahncke2006}{}}%
-\CSLLeftMargin{96. }%
+\CSLLeftMargin{83. }%
 \CSLRightInline{Jahncke J, Rintoul C (2006) CalCOFI and NMFS seabird and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-krivenkov.g.1991}{}}%
-\CSLLeftMargin{97. }%
+\CSLLeftMargin{84. }%
 \CSLRightInline{Krivenko, V.G. (1991) Waterfowl and their protection.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-melnikovy.i.2000}{}}%
-\CSLLeftMargin{98. }%
+\CSLLeftMargin{85. }%
 \CSLRightInline{Melnikov, Y.I., Melnikova, N., Pronkevich, V.V. (2000)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2013}{}}%
-\CSLLeftMargin{99. }%
+\CSLLeftMargin{86. }%
 \CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2014}{}}%
-\CSLLeftMargin{100. }%
+\CSLLeftMargin{87. }%
 \CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitori2014}{}}%
-\CSLLeftMargin{101. }%
+\CSLLeftMargin{88. }%
 \CSLRightInline{Monitoring site 1000 village survey - medium and large
@@
 \leavevmode\vadjust pre{\hypertarget{ref-preston1960}{}}%
-\CSLLeftMargin{102. }%
+\CSLLeftMargin{89. }%
 \CSLRightInline{Preston FW (1960)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-svensson2006}{}}%
-\CSLLeftMargin{103. }%
+\CSLLeftMargin{90. }%
 \CSLRightInline{Svensson S (2006) Species composition and population
@@
 \leavevmode\vadjust pre{\hypertarget{ref-svensson2010}{}}%
-\CSLLeftMargin{104. }%
+\CSLLeftMargin{91. }%
 \CSLRightInline{Svensson S, Thorner AM, Nyholm NEI (2010) Species
@@
 \leavevmode\vadjust pre{\hypertarget{ref-thorn2016changes}{}}%
-\CSLLeftMargin{105. }%
+\CSLLeftMargin{92. }%
 \CSLRightInline{Thorn S, et al. (2016)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-usfs}{}}%
-\CSLLeftMargin{106. }%
+\CSLLeftMargin{93. }%
 \CSLRightInline{USFS Landbird monitoring program (UMT-LBMP). Available
@@
 \leavevmode\vadjust pre{\hypertarget{ref-vermontcenterforecostudies2015}{}}%
-\CSLLeftMargin{107. }%
+\CSLLeftMargin{94. }%
 \CSLRightInline{Vermont Center For Ecostudies, Lambert JD, Hart J (2015)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-vickery1984}{}}%
-\CSLLeftMargin{108. }%
+\CSLLeftMargin{95. }%
 \CSLRightInline{Vickery WL, Nudds TD (1984)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-waide2017}{}}%
-\CSLLeftMargin{109. }%
+\CSLLeftMargin{96. }%
 \CSLRightInline{Waide R (2017) Bird abundance - point counts.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-williamson1983}{}}%
-\CSLLeftMargin{110. }%
+\CSLLeftMargin{97. }%
 \CSLRightInline{Williamson M (1983)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-zakharovv.d.1998}{}}%
-\CSLLeftMargin{111. }%
+\CSLLeftMargin{98. }%
 \CSLRightInline{Zakharov, V.D. (1998) Biodiversity of bird population of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-meditss2011}{}}%
-\CSLLeftMargin{112. }%
+\CSLLeftMargin{99. }%
 \CSLRightInline{MEDITS seabird surveys 1999 / 2000 / 2002 (2011)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-animald}{}}%
-\CSLLeftMargin{113. }%
+\CSLLeftMargin{100. }%
 \CSLRightInline{Animal demography unit - coordinated waterbird counts
@@
 \leavevmode\vadjust pre{\hypertarget{ref-baltics}{}}%
-\CSLLeftMargin{114. }%
+\CSLLeftMargin{101. }%
 \CSLRightInline{Baltic seabirds transect surveys Available at:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-thorn2016response}{}}%
-\CSLLeftMargin{115. }%
+\CSLLeftMargin{102. }%
 \CSLRightInline{Thorn S, et al. (2016)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ernest2009}{}}%
-\CSLLeftMargin{116. }%
+\CSLLeftMargin{103. }%
 \CSLRightInline{Ernest SKM, Valone TJ, Brown JH (2009)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-friggens2008}{}}%
-\CSLLeftMargin{117. }%
+\CSLLeftMargin{104. }%
 \CSLRightInline{Friggens M (2008) Sevilleta LTER small mammal population
@@
 \leavevmode\vadjust pre{\hypertarget{ref-jalilova.b.2014}{}}%
-\CSLLeftMargin{118. }%
+\CSLLeftMargin{105. }%
 \CSLRightInline{Jalilov, A. B., Andreychev, A. V., Kuznetsov, V. A.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kartzinel2014}{}}%
-\CSLLeftMargin{119. }%
+\CSLLeftMargin{106. }%
 \CSLRightInline{Kartzinel TR, et al. (2014)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kaufman2019}{}}%
-\CSLLeftMargin{120. }%
+\CSLLeftMargin{107. }%
 \CSLRightInline{Kaufman D (2019) CSM01 Seasonal Summary of Numbers of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-krefting1974}{}}%
-\CSLLeftMargin{121. }%
+\CSLLeftMargin{108. }%
 \CSLRightInline{Krefting LW, Ahlgren CE (1974)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lightfoot}{}}%
-\CSLLeftMargin{122. }%
+\CSLLeftMargin{109. }%
 \CSLRightInline{Lightfoot D, Schooley RL SMES rodent trapping data,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-malyshevy.s.2011}{}}%
-\CSLLeftMargin{123. }%
+\CSLLeftMargin{110. }%
 \CSLRightInline{Malyshev, Y. S. (2011) On the diagnostic techniques of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-meyer2008}{}}%
-\CSLLeftMargin{124. }%
+\CSLLeftMargin{111. }%
 \CSLRightInline{Meyer CFJ, Kalko EKV (2008)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nedosekinv.y.2015}{}}%
-\CSLLeftMargin{125. }%
+\CSLLeftMargin{112. }%
 \CSLRightInline{Nedosekin, V. Y. (2015) Long-term dynamics of the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-prins1990}{}}%
-\CSLLeftMargin{126. }%
+\CSLLeftMargin{113. }%
 \CSLRightInline{Prins HHT, Douglas-Hamilton I (1990)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rocha2017}{}}%
-\CSLLeftMargin{127. }%
+\CSLLeftMargin{114. }%
 \CSLRightInline{Rocha R, et al. (2017)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-stapp2014}{}}%
-\CSLLeftMargin{128. }%
+\CSLLeftMargin{115. }%
 \CSLRightInline{Stapp P (2014) SGS-LTER long-term monitoring project:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bahamas}{}}%
-\CSLLeftMargin{129. }%
+\CSLLeftMargin{116. }%
 \CSLRightInline{Bahamas marine mammal research organisation
@@
 \leavevmode\vadjust pre{\hypertarget{ref-popacet}{}}%
-\CSLLeftMargin{130. }%
+\CSLLeftMargin{117. }%
 \CSLRightInline{POPA cetacean, seabird, and sea turtle sightings in the

### 2024-03-12 analysis/05_model_metrics.Rmd
add protected area models

@@
         mutate(effect = "group_slope", group = model_id) %>%
-        rename(std.error = SE, estimate = year_scaled.trend, term = model_id)
+        rename(std.error = SE, estimate = year_scaled.trend, term = model_id) %>%
+        mutate(term = as.character(term))
 
@@
                          realm = "value ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
-                         climate = "value ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
-  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula") %>%
+                         climate = "value ~ year_scaled * climate + (year_scaled|study_id/rarefyID)",
+                         protected_area = "value ~ year_scaled * protected_area + (year_scaled|study_id/rarefyID)") %>%
+  pivot_longer(cols = c(base, taxa, realm, climate, protected_area), names_to = "model_id", values_to = "formula") %>%
   #don't fit the CWM_litter_size_min_n and climate model, not enough data
@@
                          realm = "log(value) ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
-                         climate = "log(value) ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
-  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
+                         climate = "log(value) ~ year_scaled * climate + (year_scaled|study_id/rarefyID)",
+                   protected_area = "value ~ year_scaled * protected_area + (year_scaled|study_id/rarefyID)") %>%
+  pivot_longer(cols = c(base, taxa, realm, climate, protected_area), names_to = "model_id", values_to = "formula")
 
@@
                          realm = "logvalue ~ year_scaled * realm + (year_scaled|study_id/rarefyID)",
-                         climate = "logvalue ~ year_scaled * climate + (year_scaled|study_id/rarefyID)") %>%
-  pivot_longer(cols = c(base, taxa, realm, climate), names_to = "model_id", values_to = "formula")
+                         climate = "logvalue ~ year_scaled * climate + (year_scaled|study_id/rarefyID)",
+                      protected_area = "value ~ year_scaled * protected_area + (year_scaled|study_id/rarefyID)") %>%
+  pivot_longer(cols = c(base, taxa, realm, climate, protected_area), names_to = "model_id", values_to = "formula")
 

### 2024-03-13 analysis/05_model_metrics.Rmd
Get the country labels for studies where available

@@
 load(here::here("data/rarefied_metrics.rda"))
+load(here::here("data/rarefyID_cell_centre.rda"))
 
-meta <- meta %>%
-  select(study_id, realm, climate, habitat, protected_area, biome_map)
+source(here::here("R/coords2continent.R"))
 
-meta <- bt_traitfiltered %>% 
+meta_clean <- meta %>%
+  select(study_id, realm, climate, general_treat, treatment, treat_comments, treat_date, habitat, protected_area, biome_map)
+
+meta_clean <- bt_traitfiltered %>% 
   select(study_id, rarefyid, taxa) %>% distinct() %>%
-  left_join(meta)
+  left_join(meta_clean) %>%
+  mutate(rarefyid_join = stringr::str_extract(rarefyid, "[^_]*_[^_]*")) %>%
+  left_join(rarefyID_cell_centre %>% select(-STUDY_ID), by = c("rarefyid_join" = "rarefyID")) %>%
+  mutate(country = coords2continent(tibble(rarefyID_x = .$rarefyID_x, rarefyID_y = .$rarefyID_y))) %>%
+  select(-c(rarefyid_join, cell_extent, rarefyID_x, rarefyID_y, rarefyid)) %>%
+  distinct()
+
+#remove NA entries if another country was assigned
+na_studies <- meta_clean %>% 
+  group_by(study_id) %>% 
+  filter(n() > 1, is.na(country))
+
+meta_clean <- setdiff(meta_clean, na_studies)
 
@@
 usethis::use_data(model_data)
+usethis::use_data(meta_clean)
 ```

### 2024-03-13 analysis/figures/06_study_change.Rmd
Get the country labels for studies where available

--- 2023-06-07 analysis/figures/06_study_change.Rmd
@@
 ```
+
+Let's look at more specific combinations
+```{r}
+load(here::here("data/meta_clean.rda"))
+
+change_groups <- wide_table %>% 
+  select(study_id, S, Jaccard_base, starts_with("SES")) %>%
+  mutate(across(!study_id, ~case_when(
+    . > 0 ~ "pos",
+    . < 0 ~ "neg",
+    .default = "0"
+  )))
+
+change_group_count <- change_groups %>%
+  select(-study_id) %>%
+  group_by(across(everything())) %>%
+  count()
+
+no_change <- change_groups %>% 
+  mutate(study_id = as.integer(study_id)) %>%
+  filter(across(!study_id, ~. == 0)) %>%
+  left_join(meta_clean)
+
+redundancy <- change_groups %>% 
+  mutate(study_id = as.integer(study_id)) %>%
+  filter(S == "neg", SES_FRic == "0", SES_FDiv == "0", SES_FEve == "0") %>%
+  left_join(meta_clean)
+
+redundancy_gain <- change_groups %>% 
+  mutate(study_id = as.integer(study_id)) %>%
+  filter(S == "pos", SES_FRic == "0", SES_FDiv == "0", SES_FEve == "0") %>%
+  left_join(meta_clean)
+
+evenness <- change_groups %>% 
+  mutate(study_id = as.integer(study_id)) %>%
+  filter(across(!study_id & !SES_FEve, ~. == 0)) %>%
+  left_join(meta_clean)
+
+increase <- change_groups %>% 
+  mutate(study_id = as.integer(study_id)) %>%
+  filter(Jaccard_base == "neg", S == "pos", SES_FRic == "pos", SES_FDiv == "0", SES_FEve == "0") %>%
+  left_join(meta_clean)
+```

### 2024-03-13 analysis/figures/visualization.Rmd
Get the country labels for studies where available

@@
 library(tidyverse)
+library(sp)
+library(rworldmap)
 ```
@@
   #mutate(taxa = as.factor(taxa), taxa = fct_relevel(taxa, "Birds"))
+
+# get country for each cell
+
+# The single argument to this function, points, is a data.frame in which:
+#   - column 1 contains the longitude in degrees
+#   - column 2 contains the latitude in degrees
+coords2continent = function(points){  
+  countriesSP <- getMap(resolution='low')
+  #countriesSP <- getMap(resolution='high') #you could use high res map from rworldxtra if you were concerned about detail
+
+  # converting points to a SpatialPoints object
+  # setting CRS directly to that from rworldmap
+  pointsSP = SpatialPoints(points, proj4string=CRS(proj4string(countriesSP)))
+
+  # use 'over' to get indices of the Polygons object containing each point 
+  indices = over(pointsSP, countriesSP)
+
+  #indices$continent   # returns the continent (6 continent model)
+  #indices$REGION   # returns the continent (7 continent model)
+  indices$ADMIN  #returns country name
+  #indices$ISO3 # returns the ISO3 code 
+}
+
+study_data <- study_data %>% 
+  mutate(country = coords2continent(study_data %>% select(rarefyID_x, rarefyID_y)))
 

### 2024-03-13 R/coords2continent.R

@@ -1,0 +1,20 @@
+# The single argument to this function, points, is a data.frame in which:
+#   - column 1 contains the longitude in degrees
+#   - column 2 contains the latitude in degrees
+coords2continent = function(points){
+  countriesSP <- getMap(resolution='low')
+  #countriesSP <- getMap(resolution='high') #you could use high res map from rworldxtra if you were concerned about detail
+
+  # converting points to a SpatialPoints object
+  # setting CRS directly to that from rworldmap
+  pointsSP = SpatialPoints(points, proj4string=CRS(proj4string(countriesSP)))
+
+
+  # use 'over' to get indices of the Polygons object containing each point
+  indices = over(pointsSP, countriesSP)
+
+  #indices$continent   # returns the continent (6 continent model)
+  #indices$REGION   # returns the continent (7 continent model)
+  indices$ADMIN  #returns country name
+  #indices$ISO3 # returns the ISO3 code
+}

### 2024-03-15 analysis/05_model_metrics.Rmd
Look at study locations.

@@
   mutate(country = coords2continent(tibble(rarefyID_x = .$rarefyID_x, rarefyID_y = .$rarefyID_y))) %>%
-  select(-c(rarefyid_join, cell_extent, rarefyID_x, rarefyID_y, rarefyid)) %>%
+  select(-c(rarefyid_join, cell_extent, rarefyID_x, rarefyID_y)) %>%
   distinct()
-
-#remove NA entries if another country was assigned
-na_studies <- meta_clean %>% 
-  group_by(study_id) %>% 
-  filter(n() > 1, is.na(country))
-
-meta_clean <- setdiff(meta_clean, na_studies)
 
@@
   mutate(logvalue = log(value + 1)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
-  left_join(meta, c("rarefyID" = "rarefyid"))
+  left_join(meta_clean, c("rarefyID" = "rarefyid"))
 
@@
   group_nest(metric) %>%
-  mutate(model = map(data, ~glm(estimate ~ duration, 
+  mutate(model = purrr::map(data, ~glm(estimate ~ duration, 
                                  data = .x)),
-           coef = map(model, tidy),
+           coef = purrr::map(model, tidy),
            ) %>% 
@@
   group_nest(metric) %>%
-  mutate(model = map(data, ~glm(estimate ~ startyear, 
+  mutate(model = purrr::map(data, ~glm(estimate ~ startyear, 
                                  data = .x)),
-           coef = map(model, tidy),
+           coef = purrr::map(model, tidy),
            ) %>% 

### 2024-03-15 paper/GEB Submission/manuscript_GEB.Rmd
Look at study locations.

@@
 ---
-title: Prevailing scenarios of functional change in Anthropocene bird and mammal communities
+title: Functional traits complicate the picture of temporal biodiversity change in bird and mammal communities
 abstract: |
@@
 Rather than testing specific hypotheses of change we present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination.
-We present results at three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm), and trends for individual studies.", "blue")`
+We present results at three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.", "blue")`
 
@@
 The earliest sample was in 1923 and the most recent was in 2014.
-For a full breakdown of studies and their characteristics, see the supplement.
+`r colorize("While it is not possible with this data to directly assess the level of human impact occurring for each study, we include binary protection status as a coarse indicator of impact level.
+However, protected areas were almost exclusively from temperate terrestrial studies (with one tropical study), so results are confounded by multiple other study characteristics.", "blue")` For a full breakdown of studies and their characteristics, see the supplement.
 Our final dataset reflects many of the data biases that make global synthesis work challenging, including geographic bias, a bias away from areas currently under the greatest threat, and a bias towards shorter time series.
@@
 
-To test for trends within and between different levels of taxa, biome, and realm we fit separate models with each of those factors added as a predictor to the original model structure.
+To test for trends within and between different levels of taxa, biome, realm, and protection status we fit separate models with each of those factors added as a predictor interacting with time to the original model structure.
 We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.8.2, @lenth2022].
@@
 Non-significant overall trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies was plausibly 0 (Table \ref{tab:resultsTab}).
-`r colorize("Within-group", "blue")` trends for different taxa, biomes, or realms were also non-significant for richness and functional diversity metrics, with the exception of a significantly increasing trend for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing functional richness slope for temperate/tropical studies and mammal studies.
+`r colorize("Within-group", "blue")` trends for different taxa, biomes, realms, or protection statuses were also non-significant for richness and functional diversity metrics, with the exception of a significantly increasing trend for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing functional richness slope for temperate/tropical studies and mammal studies.
 However, trends were not exhibited for the `r colorize("corrected", "blue")` metric indicating that differences `r colorize("in functional diversity metrics", "blue")` were largely due to changes in species richness.
@@
 For example, while Jaccard similarity was decreasing in the general trend and there were significant within group slopes for Mammals, Birds, and Terrestrial communities, there was no significant slope for the marine realm, indicating that the general trend is mostly driven by turnover in terrestrial communities.
+We also found a significant decrease in Jaccard similarity in unprotected areas only, with no trend for protected areas.
 For bird communities, we also found within group trends and between group differences for trends in foraging behavior.
@@
 Vertebrate consumption significantly decreased for Marine studies and for studies of Global and Tropical communities.
+`r colorize("There was also a significant increase in fish and fruit consumption for unprotected areas, but not protected.", "blue")`
 
@@
 
-Study-level slopes for functional diversity metrics were significantly related to start year of the time series for Jaccard similarity and functional evenness, both of which had significantly more negative slopes with more recent start year.
-No functional diversity metrics were significantly related to the start year of the time series.
+Study-level slopes were significantly related to start year of the time series for Jaccard similarity and functional evenness models, both of which had significantly more negative slopes with more recent start year.
+No functional diversity metrics were significantly related to the duration of the time series.
 
@@
 While the characteristics of species clearly matter, instead of unifying the nature of communities' change, they more often distinguish them.
-In general, we found a few areas of consensus for models where communities were aggregated (general trends, trends by taxa, biome, and realm), and vast heterogeneity for study-level models.", "blue")`
+In general, we found a few areas of consensus for models where communities were aggregated (general trends, trends by taxa, biome, realm, protection status), and vast heterogeneity for study-level models.", "blue")`
 
@@
 `r colorize("The most stark area of consensus was the lack of general trend aggregating across communities.", "blue")` We did not detect an overall trend in any functional diversity metrics.
-As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics may be consistent with similar well-documented species derived trends.
+As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends.
 We found no trend in functional change for almost all realms, biomes, and taxonomic groups.
@@
 
-Despite a lack of general trends in functional diversity metrics, we did find evidence of compositional shifts in the trait space with multiple `significant` trends in *CWM*'s for studies coming from the same taxa, realms, or climates, `indicating areas of consensus change across communities`.
+Despite a lack of general trends in functional diversity metrics, we did find evidence of compositional shifts in the trait space with multiple `significant` trends in *CWM*'s for studies coming from the same taxa, realms, climates, or protection statuses, `indicating areas of consensus change across communities`.
 Some of those trends are consistent with existing work in those contexts.
@@
 `r colorize("At the study-level, our results run contrary to our hope for a clearer picture of biodiversity change through a functional lens.
-The lack of general trend in functional diversity metrics belies a huge range of positive and negative trends at the study-level that happen to average out to zero.
-In order to simplify discussion, we will talk about the implications for communities grouped by their concurrent change in functional diversity metrics, species richness and turnover split into positive, negative or no trend.
+The lack of general trend in functional diversity metrics belies a huge range of positive and negative trends at the study-level.
+In order to simplify discussion, we will talk about the implications for ecosystem process and vulnerability for communities grouped by their concurrent change in functional diversity metrics, species richness and turnover split into positive, negative or no trend.
 While there are over 150 possible combinations of change direction (or no change) in the 5 metrics, we discuss here the six scenarios that occurred in more than one study: no change, species loss or gain only, loss of functional evenness, richness loss with species turnover, and increase in species and functional richness accompanied by significant turnover.
@@
 First, these may be communities resisting perturbations or simply not experiencing significant perturbations.
-Given the studies in this group come from all possible taxa, realms, climates, and protection statuses this option is a strong possibility and offers some hope that there are areas of the globe where communities are fairing well for now.
+Given the studies in this group come from all possible taxa, realms, climates, and protection statuses, evidence points to communities resisting perturbation, offering some hope that there are areas of the globe where communities are fairing well for now.
 Alternatively, these may be communities that have experienced or continue to experience significant stress, but lost species or functional diversity outside the observation window.
@@
 
-Three of the groups fit under the broad umbrella of changes in redundancy, including communities exhibiting gains or losses in species richness with no change in functional metrics or turnover and richness loss with significant turnover.
-These communities experience changes in species richness that do not impact the overall functional space, meaning species gains or losses represent a loss or increase in functional redundancy respectively.
-Where turnover is also occurring, species are being exchanged for functional analogs from the species pool.
-For communities exhibiting loss of redundancy, ecological processes are likely being maintained, but have a reduced capacity to respond to future stressors.
-These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future.
-Conversely, communities exhibiting a increase in redundancy (via an increase in species richness) are becoming better positioned to respond to future stressors, but are not actually expanding their functional capacity as may be assumed by looking at species gains alone.
-
--   add redundancy literature context?
+Three of the groups fit under the broad umbrella of changes in redundancy.
+By definition, if a community gains or loses species while functional metrics are unchanged, those species represent an increase or decrease in redundancy, so we include communities exhibiting gains or losses in species richness with no change in functional metrics or turnover and richness loss with significant turnover in this umbrella.
+For communities exhibiting loss of redundancy (declining species richness), ecological processes are likely being maintained, but capacity to respond to future stressors is reduced [@reich2012].
+These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future [@pimiento2020a].
+Conversely, communities exhibiting a increase in redundancy (via an increase in species richness) are becoming better positioned to respond to future stressors, but are not actually expanding their functional capacity as may be assumed by looking at species gains alone [@biggs2020].
 
@@
 These communities have a hopeful trajectory, as they are improving their functional capacity and potentially their functional redundancy, likely leading to robust future ecological function.
-Notably, the studies that exhibited species richness increases (in this group with increasing functional richness, as well as communities without functional richness increases) were exclusively from terrestrial, temperate (sometimes temperate/polar) bird communities.
-We therefore have a strong indication that communities whose functional state has improved may be restricted to these specific contexts for x,y,z reasons.
+Notably, the studies that exhibited species richness increases (in this group with increasing functional richness, as well as communities without functional richness increases) were exclusively from terrestrial, temperate (sometimes temperate/polar) bird communities and occurred in countries where there has been a significant investment in conservation over the last few decades (United States, Canada, Sweden).
+Our results are consistent with other functional work in these regions, showing increases in species richness and functional diversity for North American breeding birds over the last few decades after a period of decline[@jarzyna2016], and loss of common, functionally general species even as rare species are increasing in North America and Europe [@inger2015; @rosenberg2019; @schipper2016].
 
 The final group exhibits decreasing functional evenness.
-The implications of this change current and future maintenance of functional structure is the least clear from our current understanding of the relationship between different aspects of functional diversity and ecological process.
-Maintaining evenness may be important for supporting multifunctionality in communities [@bagousse-pinguet2019]
+The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability [@filstrup2019].
+If one or a few species become largely dominant in a community, function will be the result of only those species rather than spread out across a wealth of species responding independently to environmental change [@hillebrand2008; @hillebrand2009].
+Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities [@bagousse-pinguet2019].
+These communities are therefore of particular concern, especially considering evidence evenness is more sensitive to environmental change that richness [@filstrup2019].
 
 While the majority of studies fell into one of the groups discussed in detail above, 25% of the studies exhibited some different combination of change in metrics, underlining the vast heterogeneity of realized change scenarios.
-While we have an indication of the ecological characteristics most likely to lead to some of the change types (e.g.
-temperate bird communities are gaining species and functional richness), most types of change are exhibited across many different kinds of communities.
-Identifying key factors mediating the kind of biodiversity change a community experiences is critical for identifying areas of conservation concern and extrapolating our results to areas where functional trait data may not be available.
-While it is tempting to draw a one to one line between the degree of human impact and loss functional integrity (in the form of functional redundancy and richness), that relationship is not borne out by our rough metric of human impact or previous taxa or single community work.
+We have an indication of the ecological characteristics most likely to lead to some of the change types (e.g.
+temperate bird communities are gaining species and functional richness), but most types of change are exhibited across many different kinds of communities.
+Identifying key factors mediating the kind of biodiversity change a community experiences is critical for identifying areas of conservation concern and extrapolating results to areas where functional trait data may not be available.
+While it is tempting to draw a one to one line between the degree of human impact and loss of functional integrity (in the form of functional redundancy and richness), that relationship is not borne out by our rough metric of human impact or previous taxa-specific or single community work, which is largely mixed [@flynn2009; @luck2013; @magioli2021; @tinoco2018].
+<!-- huge differences between different taxa and the kind of disturbance/land use change -->
 
-At the study level, biodiversity change fell into four dominant scenarios of change based on trends in richness, turnover, and functional metrics: no change, only functional change, richness and functional change with turnover, and richness and functional change without turnover (Table \ref{tab:changeScenarios}).
-In contrast to our expectation based on anthropogenic impacts, around a quarter of studies experienced no trend in richness, turnover, or functional metrics.
-These studies spanned the distribution of study durations excluding only the very longest running studies, with the longest no change time series lasting 23 years.
+```{r changeScenarios, echo = FALSE, message = FALSE, fig.show = 'hold', fig.cap= "Comparison of the distribution of time series durations from each type of change to the overall distribution of time series durations, with no richness change or turnover and functional change in yellow, no turnover but richness and functional change in green, no change in blue, and turnover, richness and functional change studies in orange."}
+# 
+# knitr::include_graphics(here::here("figures/duration_hist.jpeg"))
+# 
+# knitr::include_graphics(here::here("figures/functional_change_table.png"))
+```
 
-Another quarter of the studies showed no trend in richness or turnover, but did show a significant shift in at least one summary functional metric.
-These communities are best aligned with scenarios of species replacement where only a few functional outliers replaced functionally indistinct species or vice versa (there were studies with positive and negative trends in all three functional metrics).
-Studies in the scenario of only functional change were heavily skewed toward shorter running time series, indicating that they may represent a limited snapshot of communities that on longer time scales would be exhibiting significant change in species-based metrics, particularly as species replacement continued adding to turnover.
-Communities in this scenario, and the previous scenario of no functional change are consistent with hypotheses that species richness may be strongly regulated through time [@brown2001; @gotelli2017], and emphasize that maintenance of functional structure can be divorced from those processes.
+### Areas of incongruence
 
-About 10 percent of studies, but the majority of time series in our dataset, exhibited richness and functional change but no turnover.
-These studies also spanned the the distribution of time series durations, excluding the longest running studies.
-With the exception of one study, trends in functional change for this group were only found for uncorrected metrics and disappeared after correcting for species richness, indicating that functional change in this group is almost exclusively the result of species gains or losses rather than turnover of species with different functional traits.
-
-About 15 percent of studies, which were also exclusively the longest running studies, had a trend in turnover, richness, and functional change.
-The studies in this scenario fell into two groups: a positive trend in richness with a positive trend in functional richness (corrected and uncorrected), and a negative trend in species richness with a negative trend in uncorrected functional richness and no trend in corrected richness but a trend in other corrected functional metrics.
-`r colorize("This would indicate that when species are gained in a community, they disproportionately add to the functional richness.
-When species are lost however, they are not more functionally unique than expected by chance though they may produce shifts in other aspects of functional space.", "blue")`
-
-Our results illustrate the range of relationships between species-based and functional metrics observed in real communities.
-Critically, particularly for the short observation time windows that characterize the majority of ecological data, functional shifts may be underlying seemingly static species-based community metrics.
-Species-based approaches may therefore be missing areas where significant change in the functional space could be impacting ecosystem function and resilience.
-In scenarios where there are significant species gains or losses over time, we found evidence that species losses may be relatively functionally indistinct, while species gains contribute disproportionately to the richness of a functional space.
-This finding is consistent with evidence that at least in bird communities, common functionally general species are being lost even as rare species are increasing [@inger2015; @rosenberg2019; @schipper2016].
-
-Our results are consistent with ecological expectations that community dissimilarity would be greater the larger the observation time window purely due to background processes, which is likely further accelerated by global change drivers [@dornelas2023].
-While we do not address how observed rates differ from background expectation here, we can make some assertions about the implication of turnover for functional change.
-Where turnover is occurring, it is almost always accompanied by a `r colorize("trend in one of the functional diversity metrics, indicating", "blue")` directional change in the functional space.
-Species are not therefore being replaced by exact functional counter parts.
-The fact that we observed both increases and decreases in corrected functional richness accompanying turnover is a reminder that `r colorize("new species are not always replacing more functionally unique species, but are often more functionally unique themselves.", "blue")` <!-- While shifts in functional space may be the result of environmental pressure, we cannot distinguish between directional change and shifts due to the inavailability of an analog species in the species pool.
-Comment: Want to say that trends could be an indication of adaptation to changing environment OR the result of extirpations being replaced (or not) by non-analoge species.
-Can't really distinguish between those things.
+`r colorize("The addition of functional trait illuminated a few other key areas where a synthesis approach is incongruous with system-specific studies or expectations based on anthropegenic impacts.", "blue")` For example, mean body size is predicted to decrease as a result of climate change impacts and megafaunal loss [@sheridan2011], a phenomenon which has already been well documented empirically and experimentally in multiple taxa [@caruso2014; @forster2012; @huss2019; @tseng2018].
+However, we found no evidence of a trend in *CWM* body size, with the exception of a single study of marine mammals in the Bahamas which showed a significant increase.
+`r colorize("While the lack of trend could be explained at least in part by data limitations such as shortened observation windows or missing intraspecific data, the fact that only one of the 41 studies for which study-level models were fit showed a change in mean body size, and in the opposite direction of the expectation, is suprising and warrants further reconciliation.", "blue")` <!-- Still, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing [@fritz2009].
+Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened [@ripple2015].
 -->
 
-```{r changeScenarios, echo = FALSE, message = FALSE, fig.show = 'hold', fig.cap= "Comparison of the distribution of time series durations from each type of change to the overall distribution of time series durations, with no richness change or turnover and functional change in yellow, no turnover but richness and functional change in green, no change in blue, and turnover, richness and functional change studies in orange."}
+Results for protected areas were also surprising given
 
-knitr::include_graphics(here::here("figures/duration_hist.jpeg"))
+-   changes in diet
 
-knitr::include_graphics(here::here("figures/functional_change_table.png"))
-```
-
-Still, some results stand in contrast to predictions for trait shifts under global change.
-For example, mean body size is predicted to decrease as a result of climate change impacts and megafaunal loss [@sheridan2011], a phenomenon which has already been well documented empirically and experimentally in multiple taxa [@caruso2014; @forster2012; @huss2019; @tseng2018].
-While the species-level trait means used here are not appropriate for assessing intraspecific body size shifts, we would expect to see shifts in community-wide means due to local losses of large-bodied species.
-Instead, we found no evidence of a trend in *CWM* body size, with the exception of a single study of marine mammals in the Bahamas which showed a significant increase.
-Still, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing [@fritz2009].
-Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened [@ripple2015].
-
-What does local maintenance of functional structure mean for ecosystem function?
-The vast majority of experimental and observational work links declines in function to declines in functional or species diversity [@brose2016; @cadotte2011; @duffy2007].
-By those criteria very few communities in our dataset are in a state of concern for loss of functionality (Table \ref{tab:trendTab}), though we did find evidence that mammal communities may be in greater danger than bird communities.
-Still, shifts in metrics are only relevant if the underlying traits are those most critical for ecosystem function.
-We were limited in this analysis to the traits available rather than those with strong empirical links to function.
-Similarly, the dimensions of functional space most important for ecosystem function are still a topic of on going debate, and at least some known aspects important for multifunctionality were not measured here [e.g.
-dispersion, rarity, abundance of dominant species, @bagousse-pinguet2021].
+-   turnover outside protected areas, not inside ( though might be driven by the fact that there's generally less turnover in terrestrial and temperate studies) [@blowes2019]
 
@@
 While we found no over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
-In fact, study-level trends indicate quite the opposite, that functional shifts with unknown implications for ecosystem processes may be going undetected by common species-based approaches, particularly for short observation time windows.
-While the majority of studies in our dataset did not experience a significant functional loss, a substantial body of work links functional degradation to species losses as a result of direct human intervention in the form of land use change and intensification or habitat fragmentation, indicating that those studies are simply not representative of the kinds of impacts of greatest policy concern [@flynn2009; @magioli2021; @tinoco2018].
+In fact, study-level trends indicate quite the opposite, that functional shifts with negative or yet unknown implications for ecosystem processes may be going undetected by common species-based approaches.
+For example, loss of evenness in communities with constant species richness may be an first sign of a community being impacted by environmental change, with negative implications for stability and function.
+
+One of the biggest threats to biodiversity is the wholesale conversion of natural areas to urban or human-dominated landscapes [@tilman2017].
+Typical long term monitoring such as those included in our study stops before this conversion occurs, leaving the resultant precipitous declines in biodiversity unrecorded [@cardinale2014].
+This is a known issue with the culture of long term monitoring, and our results should not be removed from that context.
+Rather, this studies captures communities that are likely experiencing a degree of human intervention, but are still largely nature-dominated.
 

### 2024-03-15 paper/GEB Submission/manuscript_GEB.tex
Look at study locations.

@@
 
-\title{Prevailing scenarios of functional change in Anthropocene bird
-and mammal communities}
+\title{Functional traits complicate the picture of temporal biodiversity
+change in bird and mammal communities}
 
@@
 Here we
-\textcolor{blue}{leverage ongoing efforts to assemble functional trait data and computational advances to}
+\textcolor{blue}{leverage ongoing efforts to assemble functional trait data and recent computational advances to}
 perform the first multi-taxa, multi-realm assessment of functional
@@
 Rather than testing specific hypotheses of change we present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination.
-We present results at three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm), and trends for individual studies.}
+We present results at three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.}
 
@@
 (Global, Polar/Temperate, Temperate, Temperate/Tropical, Tropical). The
-earliest sample was in 1923 and the most recent was in 2014. For a full
-breakdown of studies and their characteristics, see the supplement. Our
-final dataset reflects many of the data biases that make global
-synthesis work challenging, including geographic bias, a bias away from
-areas currently under the greatest threat, and a bias towards shorter
-time series. We address these shortcomings and their potential impact on
-our results in the discussion.
+earliest sample was in 1923 and the most recent was in 2014.
+\textcolor{blue}{While it is not possible with this data to directly assess the level of human impact occurring for each study, we include binary protection status as a coarse indicator of impact level.
+However, protected areas were almost exclusively from temperate terrestrial studies (with one tropical study), so results are confounded by multiple other study characteristics.}
+For a full breakdown of studies and their characteristics, see the
+supplement. Our final dataset reflects many of the data biases that make
+global synthesis work challenging, including geographic bias, a bias
+away from areas currently under the greatest threat, and a bias towards
+shorter time series. We address these shortcomings and their potential
+impact on our results in the discussion.
 
@@
 To test for trends within and between different levels of taxa, biome,
-and realm we fit separate models with each of those factors added as a
-predictor to the original model structure. We estimated within-level
-slopes and calculated between-level contrasts using the \emph{emmeans}
-package (v1.8.2, 47). We assessed the impact of time series duration and
-start year on study-level trends using linear models with duration and
-start year as predictors. All models in our analysis were fit using the
-\emph{lme4} (v1.1-30) package in R (v4.2.3) and p-values were calculated
-by Satterthwaite's degrees of freedom method using the \emph{lmerTest}
-(v3.1-3) package with a significance level of \(\alpha = 0.05\)
-(48--50).
+realm, and protection status we fit separate models with each of those
+factors added as a predictor interacting with time to the original model
+structure. We estimated within-level slopes and calculated between-level
+contrasts using the \emph{emmeans} package (v1.8.2, 47). We assessed the
+impact of time series duration and start year on study-level trends
+using linear models with duration and start year as predictors. All
+models in our analysis were fit using the \emph{lme4} (v1.1-30) package
+in R (v4.2.3) and p-values were calculated by Satterthwaite's degrees of
+freedom method using the \emph{lmerTest} (v3.1-3) package with a
+significance level of \(\alpha = 0.05\) (48--50).
 
@@
 across studies was plausibly 0 (Table \ref{tab:resultsTab}).
-\textcolor{blue}{Within-group} trends for different taxa, biomes, or
-realms were also non-significant for richness and functional diversity
-metrics, with the exception of a significantly increasing trend for
-functional evenness of global studies (characterized by having samples
-on multiple continents), and a significantly decreasing functional
-richness slope for temperate/tropical studies and mammal studies.
-However, trends were not exhibited for the \textcolor{blue}{corrected}
-metric indicating that differences
-\textcolor{blue}{in functional diversity metrics} were largely due to
-changes in species richness. Further, with only two global studies, the
-trend should not be considered truly general. The general trends for
-\emph{CWM} models were similarly not significant, with a significant
-positive trend for only percentage of fish in diet composition.
+\textcolor{blue}{Within-group} trends for different taxa, biomes,
+realms, or protection statuses were also non-significant for richness
+and functional diversity metrics, with the exception of a significantly
+increasing trend for functional evenness of global studies
+(characterized by having samples on multiple continents), and a
+significantly decreasing functional richness slope for
+temperate/tropical studies and mammal studies. However, trends were not
+exhibited for the \textcolor{blue}{corrected} metric indicating that
+differences \textcolor{blue}{in functional diversity metrics} were
+largely due to changes in species richness. Further, with only two
+global studies, the trend should not be considered truly general. The
+general trends for \emph{CWM} models were similarly not significant,
+with a significant positive trend for only percentage of fish in diet
+composition.
 
@@
 indicating that the general trend is mostly driven by turnover in
-terrestrial communities. For bird communities, we also found within
-group trends and between group differences for trends in foraging
-behavior. We found a significant increasing trend in utilization of the
-canopy in Tropical communities that was significantly different from the
-trends for Polar/Temperate and Temperate communities. There was a
-significant decrease in utilization of the understory for Terrestrial
-communities and significant increase in foraging below the water surface
-for global studies (but see the previous limitations of Global data).
-There was also a significant positive slope for \emph{CWM} body mass for
-the single Temperate/Tropical study (with three time series) which were
+terrestrial communities. We also found a significant decrease in Jaccard
+similarity in unprotected areas only, with no trend for protected areas.
+For bird communities, we also found within group trends and between
+group differences for trends in foraging behavior. We found a
+significant increasing trend in utilization of the canopy in Tropical
+communities that was significantly different from the trends for
+Polar/Temperate and Temperate communities. There was a significant
+decrease in utilization of the understory for Terrestrial communities
+and significant increase in foraging below the water surface for global
+studies (but see the previous limitations of Global data). There was
+also a significant positive slope for \emph{CWM} body mass for the
+single Temperate/Tropical study (with three time series) which were
 marine mammal communities.
@@
 studies and for studies of Global and Tropical communities.
+\textcolor{blue}{There was also a significant increase in fish and fruit consumption for unprotected areas, but not protected.}
 
@@
 While the characteristics of species clearly matter, instead of unifying the nature of communities' change, they more often distinguish them.
-In general, we found a few areas of consensus for models where communities were aggregated (general trends, trends by taxa, biome, and realm), and vast heterogeneity for study-level models.}
+In general, we found a few areas of consensus for models where communities were aggregated (general trends, trends by taxa, biome, realm, protection status), and vast heterogeneity for study-level models.}
 
@@
 
+\textcolor{blue}{The most stark area of consensus was the lack of general trend aggregating across communities.}
 We did not detect an overall trend in any functional diversity metrics.
@@
 in species richness accompanied by increasing turnover through time
-(31), indicating that non-significant trends in functional metrics may
-be consistent with similar well-documented species derived trends. We
-found no trend in functional change for almost all realms, biomes, and
+(31), indicating that non-significant trends in functional metrics are
+consistent with similar well-documented species derived trends. We found
+no trend in functional change for almost all realms, biomes, and
 taxonomic groups.
@@
 \texttt{significant} trends in \emph{CWM}'s for studies coming from the
-same taxa, realms, or climates,
+same taxa, realms, climates, or protection statuses,
 \texttt{indicating\ areas\ of\ consensus\ change\ across\ communities}.
@@
 
-The strongest picture the study-level results give is that trait level
-data further complicates the story of biodiversity change. The lack of
-general trend in functional diversity metrics belies a huge range of
-positive and negative trends at the study-level that happen to average
-out to zero. In order to simplify discussion, we will talk about the
-implications for communities grouped by their concurrent change in
-functional diversity metrics, species richness and turnover split into
-positive, negative or no trend. While there are over 150 possible
-combinations of change direction (or no change) in the 5 metrics, we
-discuss here the six scenarios that occurred in more than one study: no
-change, species loss or gain only, loss of functional evenness, richness
-loss with species turnover, and increase in species and functional
-richness accompanied by significant turnover. We focus on these
-scenarios to combat potential spurious results due to multiple testing,
-as it is unlikely all observations of the same scenario are spurious.
-The number of studies in each scenarios is broken down in table x to
-illustrate the relative number between groups rather than absolute
-number.
+\textcolor{blue}{At the study-level, our results run contrary to our hope for a clearer picture of biodiversity change through a functional lens.
+The lack of general trend in functional diversity metrics belies a huge range of positive and negative trends at the study-level.
+In order to simplify discussion, we will talk about the implications for ecosystem process and vulnerability for communities grouped by their concurrent change in functional diversity metrics, species richness and turnover split into positive, negative or no trend.
+While there are over 150 possible combinations of change direction (or no change) in the 5 metrics, we discuss here the six scenarios that occurred in more than one study: no change, species loss or gain only, loss of functional evenness, richness loss with species turnover, and increase in species and functional richness accompanied by significant turnover.
+We focus on these scenarios to combat potential spurious results due to multiple testing, as it is unlikely all observations of the same scenario are due to false positives.
+We include a break down of the number of studies in each scenario to illustrate relative number of studies in each group, but emphasize that the absolute number is likely impacted by multiple testing.}
 
 The majority group of studies exhibited no trend in any species or
-functional diversity metric.
-
-At the study level, biodiversity change fell into four dominant
-scenarios of change based on trends in richness, turnover, and
-functional metrics: no change, only functional change, richness and
-functional change with turnover, and richness and functional change
-without turnover (Table \ref{tab:changeScenarios}). In contrast to our
-expectation based on anthropogenic impacts, around a quarter of studies
-experienced no trend in richness, turnover, or functional metrics. These
-studies spanned the distribution of study durations excluding only the
-very longest running studies, with the longest no change time series
-lasting 23 years.
-
-Another quarter of the studies showed no trend in richness or turnover,
-but did show a significant shift in at least one summary functional
-metric. These communities are best aligned with scenarios of species
-replacement where only a few functional outliers replaced functionally
-indistinct species or vice versa (there were studies with positive and
-negative trends in all three functional metrics). Studies in the
-scenario of only functional change were heavily skewed toward shorter
-running time series, indicating that they may represent a limited
-snapshot of communities that on longer time scales would be exhibiting
-significant change in species-based metrics, particularly as species
-replacement continued adding to turnover. Communities in this scenario,
-and the previous scenario of no functional change are consistent with
-hypotheses that species richness may be strongly regulated through time
-(3, 8), and emphasize that maintenance of functional structure can be
-divorced from those processes.
-
-About 10 percent of studies, but the majority of time series in our
-dataset, exhibited richness and functional change but no turnover. These
-studies also spanned the the distribution of time series durations,
-excluding the longest running studies. With the exception of one study,
-trends in functional change for this group were only found for
-uncorrected metrics and disappeared after correcting for species
-richness, indicating that functional change in this group is almost
-exclusively the result of species gains or losses rather than turnover
-of species with different functional traits.
-
-About 15 percent of studies, which were also exclusively the longest
-running studies, had a trend in turnover, richness, and functional
-change. The studies in this scenario fell into two groups: a positive
-trend in richness with a positive trend in functional richness
-(corrected and uncorrected), and a negative trend in species richness
-with a negative trend in uncorrected functional richness and no trend in
-corrected richness but a trend in other corrected functional metrics.
-\textcolor{blue}{This would indicate that when species are gained in a community, they disproportionately add to the functional richness.
-When species are lost however, they are not more functionally unique than expected by chance though they may produce shifts in other aspects of functional space.}
-
-Our results illustrate the range of relationships between species-based
-and functional metrics observed in real communities. Critically,
-particularly for the short observation time windows that characterize
-the majority of ecological data, functional shifts may be underlying
-seemingly static species-based community metrics. Species-based
-approaches may therefore be missing areas where significant change in
-the functional space could be impacting ecosystem function and
-resilience. In scenarios where there are significant species gains or
-losses over time, we found evidence that species losses may be
-relatively functionally indistinct, while species gains contribute
-disproportionately to the richness of a functional space. This finding
-is consistent with evidence that at least in bird communities, common
-functionally general species are being lost even as rare species are
-increasing (55--57).
-
-Our results are consistent with ecological expectations that community
-dissimilarity would be greater the larger the observation time window
-purely due to background processes, which is likely further accelerated
-by global change drivers (13). While we do not address how observed
-rates differ from background expectation here, we can make some
-assertions about the implication of turnover for functional change.
-Where turnover is occurring, it is almost always accompanied by a
-\textcolor{blue}{trend in one of the functional diversity metrics, indicating}
-directional change in the functional space. Species are not therefore
-being replaced by exact functional counter parts. The fact that we
-observed both increases and decreases in corrected functional richness
-accompanying turnover is a reminder that
-\textcolor{blue}{new species are not always replacing more functionally unique species, but are often more functionally unique themselves.}
-
-\begin{figure}
-\includegraphics[width=\textwidth]{../../figures/duration_hist} \includegraphics[width=\textwidth]{../../figures/functional_change_table} \caption{Comparison of the distribution of time series durations from each type of change to the overall distribution of time series durations, with no richness change or turnover and functional change in yellow, no turnover but richness and functional change in green, no change in blue, and turnover, richness and functional change studies in orange.}\label{fig:changeScenarios}
-\end{figure}
-
-Still, some results stand in contrast to predictions for trait shifts
-under global change. For example, mean body size is predicted to
-decrease as a result of climate change impacts and megafaunal loss (58),
-a phenomenon which has already been well documented empirically and
-experimentally in multiple taxa (59--62). While the species-level trait
-means used here are not appropriate for assessing intraspecific body
-size shifts, we would expect to see shifts in community-wide means due
-to local losses of large-bodied species. Instead, we found no evidence
-of a trend in \emph{CWM} body size, with the exception of a single study
-of marine mammals in the Bahamas which showed a significant increase.
-Still, many of the studies in our dataset draw from areas that may have
-experienced significant loss of large-bodied species before the
-observation window, with contemporary loss rates slowing (63). Trends
-could be significantly different for the same time periods in regions of
-sub-Saharan Africa, for example, which has poor representation in our
-dataset but where megafauna exist on the landscape and are increasingly
-threatened (64).
-
-What does local maintenance of functional structure mean for ecosystem
-function? The vast majority of experimental and observational work links
-declines in function to declines in functional or species diversity (7,
-65, 66). By those criteria very few communities in our dataset are in a
-state of concern for loss of functionality (Table \ref{tab:trendTab}),
-though we did find evidence that mammal communities may be in greater
-danger than bird communities. Still, shifts in metrics are only relevant
-if the underlying traits are those most critical for ecosystem function.
-We were limited in this analysis to the traits available rather than
-those with strong empirical links to function. Similarly, the dimensions
-of functional space most important for ecosystem function are still a
-topic of on going debate, and at least some known aspects important for
-multifunctionality were not measured here (e.g.~dispersion, rarity,
-abundance of dominant species, 67).
+functional diversity metric. Contrary to the expectation due to
+anthropogenic and global change stressors, these communities do not show
+significant changes over the course of the observation window. Studies
+in this group span the distribution of study durations excluding only
+the very longest running studies, with the longest no change time series
+lasting 23 years. They also included both bird and mammal studies and
+only four were located in protected areas, indicating that the lack of
+trend is not restricted to a specific ecological context or those
+communities most insulated from human impact.
+
+The lack of trend could be the result of multiple possible scenarios.
+First, these may be communities resisting perturbations or simply not
+experiencing significant perturbations. Given the studies in this group
+come from all possible taxa, realms, climates, and protection statuses,
+evidence points to communities resisting perturbation, offering some
+hope that there are areas of the globe where communities are fairing
+well for now. Alternatively, these may be communities that have
+experienced or continue to experience significant stress, but lost
+species or functional diversity outside the observation window. This
+could be true particularly for North American mammal communities where
+trophic downgrading and megafaunal losses occurred hundreds of years ago
+{[}citation{]}. Third, these communities may be experiencing directional
+shifts undetectable by available data. For example, species-level trait
+data does not capture intraspecific shifts in the trait space, which can
+represent significant changes in overall functional composition and
+impact maintenance of ecological processes {[}citation{]}.
+
+Three of the groups fit under the broad umbrella of changes in
+redundancy. By definition, if a community gains or loses species while
+functional metrics are unchanged, those species represent an increase or
+decrease in redundancy, so we include communities exhibiting gains or
+losses in species richness with no change in functional metrics or
+turnover and richness loss with significant turnover in this umbrella.
+For communities exhibiting loss of redundancy (declining species
+richness), ecological processes are likely being maintained, but
+capacity to respond to future stressors is reduced (55). These
+communities are actually fairing better than expected looking at
+species-based metrics alone, but are also in a precarious position for
+maintaining ecological function into the future (56). Conversely,
+communities exhibiting a increase in redundancy (via an increase in
+species richness) are becoming better positioned to respond to future
+stressors, but are not actually expanding their functional capacity as
+may be assumed by looking at species gains alone (57).
+
+The next group of studies exhibit an increase in species richness and
+functional richness with significant turnover. While these communities
+are losing some species that are functionally redundant or functional
+analogs of species additions, they are gaining even more species that
+expand the functional space. These communities have a hopeful
+trajectory, as they are improving their functional capacity and
+potentially their functional redundancy, likely leading to robust future
+ecological function. Notably, the studies that exhibited species
+richness increases (in this group with increasing functional richness,
+as well as communities without functional richness increases) were
+exclusively from terrestrial, temperate (sometimes temperate/polar) bird
+communities and occurred in countries where there has been a significant
+investment in conservation over the last few decades (United States,
+Canada, Sweden). Our results are consistent with other functional work
+in these regions, showing increases in species richness and functional
+diversity for North American breeding birds over the last few decades
+after a period of decline(58), and loss of common, functionally general
+species even as rare species are increasing in North America and Europe
+(59--61).
+
+The final group exhibits decreasing functional evenness. The empirical
+link between changes in evenness and ecosystem process is the most
+poorly studied of the metrics, however theory indicates that increased
+evenness improves function and stability (62). If one or a few species
+become largely dominant in a community, function will be the result of
+only those species rather than spread out across a wealth of species
+responding independently to environmental change (42, 63). Further,
+there is some evidence that maintaining evenness is important for
+supporting multifunctionality in communities (64). These communities are
+therefore of particular concern, especially considering evidence
+evenness is more sensitive to environmental change that richness (62).
+
+While the majority of studies fell into one of the groups discussed in
+detail above, 25\% of the studies exhibited some different combination
+of change in metrics, underlining the vast heterogeneity of realized
+change scenarios. We have an indication of the ecological
+characteristics most likely to lead to some of the change types
+(e.g.~temperate bird communities are gaining species and functional
+richness), but most types of change are exhibited across many different
+kinds of communities. Identifying key factors mediating the kind of
+biodiversity change a community experiences is critical for identifying
+areas of conservation concern and extrapolating results to areas where
+functional trait data may not be available. While it is tempting to draw
+a one to one line between the degree of human impact and loss of
+functional integrity (in the form of functional redundancy and
+richness), that relationship is not borne out by our rough metric of
+human impact or previous taxa-specific or single community work, which
+is largely mixed (26, 65--67).
+
+\hypertarget{areas-of-incongruence}{%
+\subsubsection{Areas of incongruence}\label{areas-of-incongruence}}
+
+\textcolor{blue}{The addition of functional trait illuminated a few other key areas where a synthesis approach is incongruous with system-specific studies or expectations based on anthropegenic impacts.}
+For example, mean body size is predicted to decrease as a result of
+climate change impacts and megafaunal loss (68), a phenomenon which has
+already been well documented empirically and experimentally in multiple
+taxa (69--72). However, we found no evidence of a trend in \emph{CWM}
+body size, with the exception of a single study of marine mammals in the
+Bahamas which showed a significant increase.
+\textcolor{blue}{While the lack of trend could be explained at least in part by data limitations such as shortened observation windows or missing intraspecific data, the fact that only one of the 41 studies for which study-level models were fit showed a change in mean body size, and in the opposite direction of the expectation, is suprising and warrants further reconciliation.}
+
+Results for protected areas were also surprising given
+
+\begin{itemize}
+\item
+  changes in diet
+\item
+  turnover outside protected areas, not inside ( though might be driven
+  by the fact that there's generally less turnover in terrestrial and
+  temperate studies) (37)
+\end{itemize}
 
@@
 are simply not representative of the kinds of impacts of greatest policy
-concern (26, 68, 69).
+concern (67).
 
@@
 will require continued efforts to fill data gaps, a well recognized
-challenge in ecology (6, 12, 70). Still, a ``scenarios of change''
+challenge in ecology (6, 12, 73). Still, a ``scenarios of change''
 framework can provide structure for future work addressing functional
@@
 further hampered by the use of species-level trait data, as the traits
-most closely linked to a process of interest may not be available (71).
+most closely linked to a process of interest may not be available (74).
 Future trait collection that explicitly considers existing frameworks
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-reich2012}{}}%
+\CSLLeftMargin{55. }%
+\CSLRightInline{Reich PB, et al. (2012)
+\href{https://doi.org/10.1126/science.1217909}{Impacts of biodiversity
+loss escalate through time as redundancy fades}. \emph{Science (New
+York, NY)} 336(6081):589--592.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-pimiento2020a}{}}%
+\CSLLeftMargin{56. }%
+\CSLRightInline{Pimiento C, et al. (2020)
+\href{https://doi.org/10.1098/rspb.2020.1162}{Selective extinction
+against redundant species buffers functional diversity}.
+\emph{Proceedings of the Royal Society B: Biological Sciences}
+287(1931):20201162.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-biggs2020}{}}%
+\CSLLeftMargin{57. }%
+\CSLRightInline{Biggs CR, et al. (2020)
+\href{https://doi.org/10.1002/ecs2.3184}{Does functional redundancy
+affect ecological stability and resilience? A review and meta-analysis}.
+\emph{Ecosphere} 11(7):e03184.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-jarzyna2016}{}}%
+\CSLLeftMargin{58. }%
+\CSLRightInline{Jarzyna MA, Jetz W (2016)
+\href{https://doi.org/10.1111/gcb.13571}{A near half-century of temporal
+change in different facets of avian diversity}. \emph{Global Change
+Biology} 23(8):2999--3011.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-inger2015}{}}%
-\CSLLeftMargin{55. }%
+\CSLLeftMargin{59. }%
 \CSLRightInline{Inger R, et al. (2015)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rosenberg2019}{}}%
-\CSLLeftMargin{56. }%
+\CSLLeftMargin{60. }%
 \CSLRightInline{Rosenberg KV, et al. (2019) Decline of the North
@@
 \leavevmode\vadjust pre{\hypertarget{ref-schipper2016}{}}%
-\CSLLeftMargin{57. }%
+\CSLLeftMargin{61. }%
 \CSLRightInline{Schipper AM, et al. (2016)
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-filstrup2019}{}}%
+\CSLLeftMargin{62. }%
+\CSLRightInline{Filstrup CT, King KBS, McCullough IM (2019)
+\href{https://doi.org/10.1111/ele.13407}{Evenness effects mask richness
+effects on ecosystem functioning at macro-scales in lakes}.
+\emph{Ecology Letters} 22(12):2120--2129.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hillebrand2008}{}}%
+\CSLLeftMargin{63. }%
+\CSLRightInline{Hillebrand H, Bennett DM, Cadotte MW (2008)
+\href{https://doi.org/10.1890/07-1053.1}{Consequences of Dominance: A
+Review of Evenness Effects on Local and Regional Ecosystem Processes}.
+\emph{Ecology} 89(6):1510--1520.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-bagousse-pinguet2019}{}}%
+\CSLLeftMargin{64. }%
+\CSLRightInline{Bagousse-Pinguet YL, et al. (2019)
+\href{https://doi.org/10.1073/pnas.1815727116}{Phylogenetic, functional,
+and taxonomic richness have both positive and negative effects on
+ecosystem multifunctionality}. \emph{Proceedings of the National Academy
+of Sciences} 116(17):8419--8424.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-luck2013}{}}%
+\CSLLeftMargin{65. }%
+\CSLRightInline{Luck GW, Carter A, Smallbone L (2013)
+\href{https://doi.org/10.1371/journal.pone.0063671}{Changes in Bird
+Functional Diversity across Multiple Land Uses: Interpretations of
+Functional Redundancy Depend on Functional Group Identity}. \emph{PLOS
+ONE} 8(5):e63671.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-magioli2021}{}}%
+\CSLLeftMargin{66. }%
+\CSLRightInline{Magioli M, et al. (2021)
+\href{https://doi.org/10.1016/j.pecon.2021.02.006}{Land-use changes lead
+to functional loss of terrestrial mammals in a Neotropical rainforest}.
+\emph{Perspectives in Ecology and Conservation} 19(2):161--170.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-tinoco2018}{}}%
+\CSLLeftMargin{67. }%
+\CSLRightInline{Tinoco BA, Santillán VE, Graham CH (2018)
+\href{https://doi.org/10.1002/ece3.3813}{Land use change has stronger
+effects on functional diversity than taxonomic diversity in tropical
+Andean hummingbirds}. \emph{Ecology and Evolution} 8(6):3478--3490.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-sheridan2011}{}}%
-\CSLLeftMargin{58. }%
+\CSLLeftMargin{68. }%
 \CSLRightInline{Sheridan JA, Bickford D (2011)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-caruso2014}{}}%
-\CSLLeftMargin{59. }%
+\CSLLeftMargin{69. }%
 \CSLRightInline{Caruso NM, Sears MW, Adams DC, Lips KR (2014)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-forster2012}{}}%
-\CSLLeftMargin{60. }%
+\CSLLeftMargin{70. }%
 \CSLRightInline{Forster J, Hirst AG, Atkinson D (2012)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-huss2019}{}}%
-\CSLLeftMargin{61. }%
+\CSLLeftMargin{71. }%
 \CSLRightInline{Huss M, Lindmark M, Jacobson P, van Dorst RM, Gårdmark A
@@
 \leavevmode\vadjust pre{\hypertarget{ref-tseng2018}{}}%
-\CSLLeftMargin{62. }%
+\CSLLeftMargin{72. }%
 \CSLRightInline{Tseng M, et al. (2018)
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-fritz2009}{}}%
-\CSLLeftMargin{63. }%
-\CSLRightInline{Fritz SA, Bininda-Emonds ORP, Purvis A (2009)
-\href{https://doi.org/10.1111/j.1461-0248.2009.01307.x}{Geographical
-variation in predictors of mammalian extinction risk: big is bad, but
-only in the tropics}. \emph{Ecology Letters} 12(6):538--549.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ripple2015}{}}%
-\CSLLeftMargin{64. }%
-\CSLRightInline{Ripple WJ, et al. (2015)
-\href{https://doi.org/10.1126/sciadv.1400103}{Collapse of the world{'}s
-largest herbivores}. \emph{Science Advances} 1(4):e1400103.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-cadotte2011}{}}%
-\CSLLeftMargin{65. }%
-\CSLRightInline{Cadotte MW, Carscadden K, Mirotchnick N (2011)
-\href{https://doi.org/10.1111/j.1365-2664.2011.02048.x}{Beyond species:
-functional diversity and the maintenance of ecological processes and
-services}. \emph{Journal of Applied Ecology} 48(5):1079--1087.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-duffy2007}{}}%
-\CSLLeftMargin{66. }%
-\CSLRightInline{Duffy JE, et al. (2007)
-\href{https://doi.org/10.1111/j.1461-0248.2007.01037.x}{The functional
-role of biodiversity in ecosystems: incorporating trophic complexity}.
-\emph{Ecology Letters} 10(6):522--538.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-bagousse-pinguet2021}{}}%
-\CSLLeftMargin{67. }%
-\CSLRightInline{Bagousse-Pinguet YL, et al. (2021) Functional rarity and
-evenness are key facets of biodiversity to boost multifunctionality.
-\emph{Proceedings of the National Academy of Sciences} 118(7).
-doi:\href{https://doi.org/10.1073/pnas.2019355118}{10.1073/pnas.2019355118}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-magioli2021}{}}%
-\CSLLeftMargin{68. }%
-\CSLRightInline{Magioli M, et al. (2021)
-\href{https://doi.org/10.1016/j.pecon.2021.02.006}{Land-use changes lead
-to functional loss of terrestrial mammals in a Neotropical rainforest}.
-\emph{Perspectives in Ecology and Conservation} 19(2):161--170.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tinoco2018}{}}%
-\CSLLeftMargin{69. }%
-\CSLRightInline{Tinoco BA, Santillán VE, Graham CH (2018)
-\href{https://doi.org/10.1002/ece3.3813}{Land use change has stronger
-effects on functional diversity than taxonomic diversity in tropical
-Andean hummingbirds}. \emph{Ecology and Evolution} 8(6):3478--3490.}
-
 \leavevmode\vadjust pre{\hypertarget{ref-hughes2021}{}}%
-\CSLLeftMargin{70. }%
+\CSLLeftMargin{73. }%
 \CSLRightInline{Hughes AC, et al. (2021)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-zhu2017}{}}%
-\CSLLeftMargin{71. }%
+\CSLLeftMargin{74. }%
 \CSLRightInline{Zhu L, et al. (2017)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-pardieck2020}{}}%
-\CSLLeftMargin{72. }%
+\CSLLeftMargin{75. }%
 \CSLRightInline{Pardieck KL, David ZJ, Lutmerding M, Aponte V, Hudson
@@
 \leavevmode\vadjust pre{\hypertarget{ref-piropno}{}}%
-\CSLLeftMargin{73. }%
+\CSLLeftMargin{76. }%
 \CSLRightInline{PIROP northwest atlantic 1965{\textendash}1992 - OBIS
@@
 \leavevmode\vadjust pre{\hypertarget{ref-silva}{}}%
-\CSLLeftMargin{74. }%
+\CSLLeftMargin{77. }%
 \CSLRightInline{Silva FR da Brazil dataset 1.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-scott}{}}%
-\CSLLeftMargin{75. }%
+\CSLLeftMargin{78. }%
 \CSLRightInline{Scott D, Metts B, Lance S The rainbow bay long-term
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rossaferes1997}{}}%
-\CSLLeftMargin{76. }%
+\CSLLeftMargin{79. }%
 \CSLRightInline{Rossa-Feres D de C (1997) Community ecology of anura
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bakker1990}{}}%
-\CSLLeftMargin{77. }%
+\CSLLeftMargin{80. }%
 \CSLRightInline{Bakker C, Herman PMJ (1990) Phytoplankton in the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-berezovikovn.n.2004}{}}%
-\CSLLeftMargin{78. }%
+\CSLLeftMargin{81. }%
 \CSLRightInline{Berezovikov, N.N. (2004) The birds of settlements in
@@
 \leavevmode\vadjust pre{\hypertarget{ref-enemar2004}{}}%
-\CSLLeftMargin{79. }%
+\CSLLeftMargin{82. }%
 \CSLRightInline{Enemar A, Sjöstrand B, Andersson G, Von Proschwitz T
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gido2019}{}}%
-\CSLLeftMargin{80. }%
+\CSLLeftMargin{83. }%
 \CSLRightInline{Gido K (2019) CFP01 Fish population on selected
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hogstad1993}{}}%
-\CSLLeftMargin{81. }%
+\CSLLeftMargin{84. }%
 \CSLRightInline{Hogstad O (1993)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-holmes1986}{}}%
-\CSLLeftMargin{82. }%
+\CSLLeftMargin{85. }%
 \CSLRightInline{Holmes RT, Sherry TW, Sturges FW (1986)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-jahncke2006}{}}%
-\CSLLeftMargin{83. }%
+\CSLLeftMargin{86. }%
 \CSLRightInline{Jahncke J, Rintoul C (2006) CalCOFI and NMFS seabird and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-krivenkov.g.1991}{}}%
-\CSLLeftMargin{84. }%
+\CSLLeftMargin{87. }%
 \CSLRightInline{Krivenko, V.G. (1991) Waterfowl and their protection.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-melnikovy.i.2000}{}}%
-\CSLLeftMargin{85. }%
+\CSLLeftMargin{88. }%
 \CSLRightInline{Melnikov, Y.I., Melnikova, N., Pronkevich, V.V. (2000)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2013}{}}%
-\CSLLeftMargin{86. }%
+\CSLLeftMargin{89. }%
 \CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2014}{}}%
-\CSLLeftMargin{87. }%
+\CSLLeftMargin{90. }%
 \CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitori2014}{}}%
-\CSLLeftMargin{88. }%
+\CSLLeftMargin{91. }%
 \CSLRightInline{Monitoring site 1000 village survey - medium and large
@@
 \leavevmode\vadjust pre{\hypertarget{ref-preston1960}{}}%
-\CSLLeftMargin{89. }%
+\CSLLeftMargin{92. }%
 \CSLRightInline{Preston FW (1960)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-svensson2006}{}}%
-\CSLLeftMargin{90. }%
+\CSLLeftMargin{93. }%
 \CSLRightInline{Svensson S (2006) Species composition and population
@@
 \leavevmode\vadjust pre{\hypertarget{ref-svensson2010}{}}%
-\CSLLeftMargin{91. }%
+\CSLLeftMargin{94. }%
 \CSLRightInline{Svensson S, Thorner AM, Nyholm NEI (2010) Species
@@
 \leavevmode\vadjust pre{\hypertarget{ref-thorn2016changes}{}}%
-\CSLLeftMargin{92. }%
+\CSLLeftMargin{95. }%
 \CSLRightInline{Thorn S, et al. (2016)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-usfs}{}}%
-\CSLLeftMargin{93. }%
+\CSLLeftMargin{96. }%
 \CSLRightInline{USFS Landbird monitoring program (UMT-LBMP). Available
@@
 \leavevmode\vadjust pre{\hypertarget{ref-vermontcenterforecostudies2015}{}}%
-\CSLLeftMargin{94. }%
+\CSLLeftMargin{97. }%
 \CSLRightInline{Vermont Center For Ecostudies, Lambert JD, Hart J (2015)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-vickery1984}{}}%
-\CSLLeftMargin{95. }%
+\CSLLeftMargin{98. }%
 \CSLRightInline{Vickery WL, Nudds TD (1984)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-waide2017}{}}%
-\CSLLeftMargin{96. }%
+\CSLLeftMargin{99. }%
 \CSLRightInline{Waide R (2017) Bird abundance - point counts.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-williamson1983}{}}%
-\CSLLeftMargin{97. }%
+\CSLLeftMargin{100. }%
 \CSLRightInline{Williamson M (1983)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-zakharovv.d.1998}{}}%
-\CSLLeftMargin{98. }%
+\CSLLeftMargin{101. }%
 \CSLRightInline{Zakharov, V.D. (1998) Biodiversity of bird population of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-meditss2011}{}}%
-\CSLLeftMargin{99. }%
+\CSLLeftMargin{102. }%
 \CSLRightInline{MEDITS seabird surveys 1999 / 2000 / 2002 (2011)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-animald}{}}%
-\CSLLeftMargin{100. }%
+\CSLLeftMargin{103. }%
 \CSLRightInline{Animal demography unit - coordinated waterbird counts
@@
 \leavevmode\vadjust pre{\hypertarget{ref-baltics}{}}%
-\CSLLeftMargin{101. }%
+\CSLLeftMargin{104. }%
 \CSLRightInline{Baltic seabirds transect surveys Available at:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-thorn2016response}{}}%
-\CSLLeftMargin{102. }%
+\CSLLeftMargin{105. }%
 \CSLRightInline{Thorn S, et al. (2016)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ernest2009}{}}%
-\CSLLeftMargin{103. }%
+\CSLLeftMargin{106. }%
 \CSLRightInline{Ernest SKM, Valone TJ, Brown JH (2009)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-friggens2008}{}}%
-\CSLLeftMargin{104. }%
+\CSLLeftMargin{107. }%
 \CSLRightInline{Friggens M (2008) Sevilleta LTER small mammal population
@@
 \leavevmode\vadjust pre{\hypertarget{ref-jalilova.b.2014}{}}%
-\CSLLeftMargin{105. }%
+\CSLLeftMargin{108. }%
 \CSLRightInline{Jalilov, A. B., Andreychev, A. V., Kuznetsov, V. A.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kartzinel2014}{}}%
-\CSLLeftMargin{106. }%
+\CSLLeftMargin{109. }%
 \CSLRightInline{Kartzinel TR, et al. (2014)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kaufman2019}{}}%
-\CSLLeftMargin{107. }%
+\CSLLeftMargin{110. }%
 \CSLRightInline{Kaufman D (2019) CSM01 Seasonal Summary of Numbers of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-krefting1974}{}}%
-\CSLLeftMargin{108. }%
+\CSLLeftMargin{111. }%
 \CSLRightInline{Krefting LW, Ahlgren CE (1974)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lightfoot}{}}%
-\CSLLeftMargin{109. }%
+\CSLLeftMargin{112. }%
 \CSLRightInline{Lightfoot D, Schooley RL SMES rodent trapping data,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-malyshevy.s.2011}{}}%
-\CSLLeftMargin{110. }%
+\CSLLeftMargin{113. }%
 \CSLRightInline{Malyshev, Y. S. (2011) On the diagnostic techniques of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-meyer2008}{}}%
-\CSLLeftMargin{111. }%
+\CSLLeftMargin{114. }%
 \CSLRightInline{Meyer CFJ, Kalko EKV (2008)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nedosekinv.y.2015}{}}%
-\CSLLeftMargin{112. }%
+\CSLLeftMargin{115. }%
 \CSLRightInline{Nedosekin, V. Y. (2015) Long-term dynamics of the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-prins1990}{}}%
-\CSLLeftMargin{113. }%
+\CSLLeftMargin{116. }%
 \CSLRightInline{Prins HHT, Douglas-Hamilton I (1990)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rocha2017}{}}%
-\CSLLeftMargin{114. }%
+\CSLLeftMargin{117. }%
 \CSLRightInline{Rocha R, et al. (2017)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-stapp2014}{}}%
-\CSLLeftMargin{115. }%
+\CSLLeftMargin{118. }%
 \CSLRightInline{Stapp P (2014) SGS-LTER long-term monitoring project:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bahamas}{}}%
-\CSLLeftMargin{116. }%
+\CSLLeftMargin{119. }%
 \CSLRightInline{Bahamas marine mammal research organisation
@@
 \leavevmode\vadjust pre{\hypertarget{ref-popacet}{}}%
-\CSLLeftMargin{117. }%
+\CSLLeftMargin{120. }%
 \CSLRightInline{POPA cetacean, seabird, and sea turtle sightings in the

### 2024-03-18 R/coords2continent.R
get package names

@@ -2,16 +2,16 @@
 #   - column 1 contains the longitude in degrees
 #   - column 2 contains the latitude in degrees
 coords2continent = function(points){
-  countriesSP <- getMap(resolution='low')
+  countriesSP <- rworldmap::getMap(resolution='low')
   #countriesSP <- getMap(resolution='high') #you could use high res map from rworldxtra if you were concerned about detail
 
   # converting points to a SpatialPoints object
   # setting CRS directly to that from rworldmap
-  pointsSP = SpatialPoints(points, proj4string=CRS(proj4string(countriesSP)))
+  pointsSP = sp::SpatialPoints(points, proj4string = sp::CRS(sp::proj4string(countriesSP)))
 
 
   # use 'over' to get indices of the Polygons object containing each point
-  indices = over(pointsSP, countriesSP)
+  indices = sp::over(pointsSP, countriesSP)
 
   #indices$continent   # returns the continent (6 continent model)
   #indices$REGION   # returns the continent (7 continent model)

### 2024-03-18 analysis/05_model_metrics.Rmd
tweaks, update output

@@
 
-There are 12 study/metric combos that can't be fit due to data limitations (too many rarefyID's )
+There are 13 study/metric combos that can't be fit due to data limitations (too many rarefyID's )
 ```{r}
@@
 bind_rows(metric_model_table, indv_mod_table) %>%
-    select(-c(model_id, n_ts, warning, lower.CL, upper.CL, significant, REML, bobyqa)) %>%
+    select(-c(n_ts, warning, lower.CL, upper.CL, significant, REML, bobyqa)) %>%
   readr::write_csv(here::here("paper", "model_estimates_supp.csv"))

### 2024-03-19 analysis/05_model_metrics.Rmd
Little munging bug

@@
 ```{r}
-mixed_jacc_mod_results <- pmap_dfr(mixed_indv_model_df %>% filter(metric_name == "Jaccard_base"), possibly(get_model, otherwise = tibble()), data = model_data)
+mixed_jacc_mod_results <- pmap_dfr(mixed_indv_model_df %>% filter(metric_name == "Jaccard_base"), 
+                                   possibly(get_model, otherwise = tibble()), data = model_data)
 
-non_conv_study <- mixed_jacc_mod_results %>% filter(!is.na(warning)) %>% pull(study_id) %>% unique()
+non_conv_study <- mixed_jacc_mod_results %>% 
+  filter(!is.na(warning)) %>% pull(study_id) %>% unique()
 
-mixed_jacc_mod_results2 <- pmap_dfr(mixed_indv_model_df %>% filter(metric_name == "Jaccard_base", filter_study_id == non_conv_study), 
+mixed_jacc_mod_results2 <- pmap_dfr(mixed_indv_model_df %>% filter(metric_name == "Jaccard_base", filter_study_id %in% non_conv_study), 
                                     possibly(get_model, otherwise = tibble()), data = model_data, bobyqa = TRUE)
 
-mixed_jacc_mod_results <- bind_rows(mixed_jacc_mod_results %>% filter(study_id != non_conv_study),
+mixed_jacc_mod_results_fin <- bind_rows(mixed_jacc_mod_results %>% filter(!study_id %in% non_conv_study),
                                     mixed_jacc_mod_results2)
@@
           glm_indv_mod_results,
-          mixed_jacc_mod_results
+          mixed_jacc_mod_results_fin
           )

### 2024-03-25 analysis/figures/visualization.Rmd
Move function out to .R script

--- 2023-02-14 analysis/figures/visualization.Rmd
@@
 #library(forcats)
+source(here::here("R/coords2continent.R"))
 
@@
   #mutate(taxa = as.factor(taxa), taxa = fct_relevel(taxa, "Birds"))
+
+study_data <- study_data %>% 
+  mutate(country = coords2continent(study_data %>% select(rarefyID_x, rarefyID_y)))
 

### 2024-03-25 paper/GEB Submission/manuscript_GEB.Rmd
Mooore edits

@@
 abstract: |
-  Aim: Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing species turnover but no overall trend in species diversity through time at the local scale.
+  Aim: Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing dissimilarity due to turnover but no overall trend in species diversity through time at the local scale.
 Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to mechanisms of ecosystem processes.
@@
 
-  Results: We found no overall trend in any functional diversity metric, despite data replicating species-based patterns of constant richness with increasing turnover.
-The lack of trend held even after correcting for changes in species richness.
-At the study-level, there were also a substantial number of time series exhibiting no species or functional change, however most studies showed a shift in a species or functional metric.
+  Results: We found no overall trend in any functional diversity metric, which held even after correcting for changes in species richness.
+At the study-level, there was substantial heterogeneity in the direction and type of functional change exhibited.
+None of the included study characteristics such as taxa, realm, biome, or protection status were significantly related to the type of change exhibited by a community.
 
   Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevalant than the other across many taxa, biomes, and realms.
-At the study-level, we identified four prevailing scenarios of species and functional change, which showed links to the duration of the observation window.
-With no one prevailing scenario of change, it will be critical to link change scenarios to drivers of change, particularly to identify communities with capacity to resist drivers from those not experiencing substantial pressure from a driver.
+At the study-level, the majority group of studies showed no species or functional trends, but other types of change were manifest for multiple studies, including studies experiencing changes in functional redundancy, increased species and functional richness, or loss of functional richness.
+With no one prevailing scenario of change, it will be critical to link change scenarios to ecological context.
 
@@
 
-The meaning of general trends derived from limited data, and their relevance for conservation, is a topic of on going debate.
-Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
+`r colorize("Using general trends derived from limited data as a diagnostic for the state of biodiversity or directing conservation is a topic of ongoing debate.", "blue")` Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
 Many of these criticisms reflect limitations of ecological data on the whole, `r colorize("leading to a call for additional data not only to fill geographic and temporal gaps, but to flesh out key characteristics of communities", "blue")` [@dornelas2023; @primack2018].
@@
 `r colorize("The expectation for functional change across communities is not obvious from past work, and may or may not follow taxonomic trends", "blue")` [@dornelas2023].
-While functional loss is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], `r colorize("functional diversity may be maintained even when species are lost from a community", "blue")` [@daz2001; @petchey2009].
+While `r colorize("loss of functional diversity", "blue")` is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], `r colorize("functional diversity may be maintained even when species are lost from a community", "blue")` [@daz2001; @petchey2009].
 Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020].
 And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @larsen2018; @matuoka2020].
-On paleoecological time scales, functional composition shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with significant impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].
-For some time periods, functional structure appears to be maintained for substantial portions of geological time [@hedberg2022].
+On paleoecological time scales, `r colorize("the functional space", "blue")` shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with significant impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].
+For some time periods, functional diversity appears to be maintained for substantial portions of geological time [@hedberg2022].
 `r colorize("Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds [@barnagaud2017; @jarzyna2016] and ray-finned fishes, sharks, and rays", "blue")` [@trindade-santos2020].
+`r colorize("Additionally, these taxa have been the subject of substantial conservation investment and policy action, so are not necessarily representative of trends for other taxa or the same taxa in different locations.", "blue")`
 
 Here we `r colorize("leverage ongoing efforts to assemble functional trait data and recent computational advances to", "blue")` perform the first multi-taxa, multi-realm assessment of functional diversity change through time.
-We focus on mammal and bird species as a significant subset of the world's biodiversity heavily impacted by anthropogenic change.
-While examining trends in plants, invertebrates, and other vertebrate species is of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
+We focus on mammal and bird species as `r colorize("a subset of the world's biodiversity of particular conservation concern that is heavily impacted by anthropogenic change.", "blue")`While examining trends in plants, invertebrates, and other vertebrate species is of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
 To ensure comparability across taxa in trait type and data quality we therefore focus on mammals and birds.
-Traits were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].
+`r colorize("We include body mass, foraging and dietary traits", "blue")` that were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].
 An initial assessment of amphibian trends is included in the supplement, but excluded from general trend assessment here due to limited geographic coverage.
@@
 
-We obtained mammal and bird time series from the BioTIME database, a global repository of high-quality assemblage time series.
-All studies included in the database follow consistent sampling protocols and represent full assemblages rather than populations of single species [@dornelas2018].
-All time series include abundance of observed species.
+We obtained mammal and bird time series from the BioTIME database, a global repository of high-quality assemblage time series `r colorize("collected from the literature and ongoing monitoring efforts", "blue")`.
+`r colorize("Data is structured such that a study encompass all data collected following consistent sampling protocols and may contain multiple time series from separate locations.
+", "blue")` Time series represent full assemblages rather than populations of single species [@dornelas2018], and all time series include abundance of observed species.
 Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual time series following a standardized spatial scale.
@@
 Our final dataset included 2,432 time series from 50 studies in 21 countries and 12 biomes and 6 different traits (Fig \ref{fig:taxaMap}).
-Data came from both terrestrial and marine realms and five climates (Global, Polar/Temperate, Temperate, Temperate/Tropical, Tropical).
+Data came from both terrestrial and marine realms and five biomes (Global, Polar/Temperate, Temperate, Temperate/Tropical, Tropical).
 The earliest sample was in 1923 and the most recent was in 2014.
@@
 Jaccard similarity was calculated relative to the first observed year for a time series.
-A negative trend in *J* would therefore indicate increasing turnover.
+A negative trend in *J* would therefore indicate `r colorize("decreasing similiarity", "blue")`.
 We did not impose a correction for unobserved species as non-parametric estimators do not assign species identities to corrected richness values, and therefore could not be propagated to the functional diversity metrics.
@@
 SES estimates can be interpreted as how much of the functional characteristic (richness, evenness, divergence) was observed beyond what was expected by chance for a community of that species richness.
+`r colorize("This approach will be less accurate for shorter time series, as we likley will not have captured all available species in the true species pool, but it is impossible to know whether the mean estimate from the null model is an over or under estimate without knowing the functional characteristics of the missing species.", "blue")`
 
@@
 
-We estimated general trends `r colorize("across bird and mammals communities", "blue")` for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study.
-We fit 18 individual *CWM* models.
+We estimated general trends `r colorize("across bird and mammals communities", "blue")` for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study, `r colorize("methods which deal well with the inherent imbalances in our data.", "blue")` We fit 18 individual *CWM* models, `r colorize("one for each trait included in the analsysis", "blue")`.
 All time series with data for a given trait were included in the corresponding *CWM* model.
@@
 We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.8.2, @lenth2022].
-We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
+`r colorize("For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore refrained from interpreting results for levels where there were fewer than three studies.", "blue")` We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
 All models in our analysis were fit using the *lme4* (v1.1-30) package in R (v4.2.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].
@@
 
-We found no significant overall trend in species richness or functional diversity metrics (observed or `r colorize("corrected", "blue")`) (Fig \ref{fig:timeseriesPlot}).
+We found no significant overall `r colorize("temporal", "blue")` trend in species richness or functional diversity metrics `r colorize("including functional richness, evenness, or divergence", "blue")` (observed or `r colorize("corrected", "blue")`) (Fig \ref{fig:timeseriesPlot}).
 We did find a significant overall decrease in Jaccard similarity, indicating accumulating changes in species composition.
-Non-significant overall trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies was plausibly 0 (Table \ref{tab:resultsTab}).
-`r colorize("Within-group", "blue")` trends for different taxa, biomes, realms, or protection statuses were also non-significant for richness and functional diversity metrics, with the exception of a significantly increasing trend for functional evenness of global studies (characterized by having samples on multiple continents), and a significantly decreasing functional richness slope for temperate/tropical studies and mammal studies.
+Non-significant overall `r colorize("temporal", "blue")`trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies `r colorize("was not significantly different from zero", "blue")`\` (Table \ref{tab:resultsTab}).
+`r colorize("Within-group", "blue")` trends for different taxa, biomes, realms, or protection statuses were also non-significant for richness and functional diversity metrics, with the exception of a significantly increasing trend for functional evenness of marine studies, and a significantly decreasing functional richness slope for mammal studies.
 However, trends were not exhibited for the `r colorize("corrected", "blue")` metric indicating that differences `r colorize("in functional diversity metrics", "blue")` were largely due to changes in species richness.
-Further, with only two global studies, the trend should not be considered truly general.
 The general trends for *CWM* models were similarly not significant, with a significant positive trend for only percentage of fish in diet composition.
 
-```{r timeseriesPlot, fig.cap="Plots of time series-level trends with line color corresponding to climatic region, with data points in grey and the overall metric mean in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES", echo=FALSE}
+```{r timeseriesPlot, fig.cap="Plots of time series-level trends with line color corresponding to climatic region, with data points in grey and the overall mean slope for a metric in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES", echo=FALSE}
 knitr::include_graphics(here::here("figures/3met_long.jpeg"))
@@
 
-We did find significant differences between taxa, realms, and climate for Jaccard similarity and some of the *CWM's*.
-For example, while Jaccard similarity was decreasing in the general trend and there were significant within group slopes for Mammals, Birds, and Terrestrial communities, there was no significant slope for the marine realm, indicating that the general trend is mostly driven by turnover in terrestrial communities.
+We did find significant differences between taxa, realms, and biomes for Jaccard similarity and some of the *CWM's*.
+For example, while Jaccard similarity was decreasing in the general trend and there were significant within group slopes for Mammals, Birds, Terrestrial, Temperate, and Tropical communities, there was no significant slope for the marine realm, indicating that the general trend is mostly driven by turnover in terrestrial communities.
 We also found a significant decrease in Jaccard similarity in unprotected areas only, with no trend for protected areas.
-For bird communities, we also found within group trends and between group differences for trends in foraging behavior.
-We found a significant increasing trend in utilization of the canopy in Tropical communities that was significantly different from the trends for Polar/Temperate and Temperate communities.
-There was a significant decrease in utilization of the understory for Terrestrial communities and significant increase in foraging below the water surface for global studies (but see the previous limitations of Global data).
-There was also a significant positive slope for *CWM* body mass for the single Temperate/Tropical study (with three time series) which were marine mammal communities.
+We found significant dietary shifts across communities, with a significant general trend of increasing fish consumption, which was also reflected in increasing fish consumption trends for Bird studies and Temperate studies.
+We found a significantly increasing trend in nectar consumption for Temperate studies but a significantly decreasing tend for Tropical studies.
+We found significantly decreasing trend for vertebrate consumption in Marine studies and Tropical studies.
 
-We found significant dietary shifts across communities, with a significant increase in fruit consumption in Terrestrial communities and a significant decrease in nectar consumption in Tropical communities, a trend significantly different than that for Terrestrial communities.
-There was a significant increase in seed consumption in bird species, which was significantly different from the trend for Mammal communities.
-There was a significant increase in fish consumption for the two Global studies.
-Vertebrate consumption significantly decreased for Marine studies and for studies of Global and Tropical communities.
-`r colorize("There was also a significant increase in fish and fruit consumption for unprotected areas, but not protected.", "blue")`
-
-At the study level, 15 of 50 studies exhibited a significant trend in species richness and 11 exhibited significant turnover.
-For observed functional metrics, 25 of 50 studies exhibited a trend in a least one metric, and 15 of 50 studies exhibited a significant trend for at least one `r colorize("corrected", "blue")` metric (Table \ref{tab:trendTab}).
+At the study level, 16 of 50 studies exhibited a significant trend in species richness and 11 exhibited significant turnover.
+For uncorrected functional diversity metrics, 25 of 50 studies exhibited a trend in a least one metric, and 15 of 50 studies exhibited a significant trend for at least one `r colorize("corrected", "blue")` metric (Table \ref{tab:trendTab}).
 In general, there were more significant trends for uncorrected metrics, with some disappearing after correction, indicating that those trends were likely due to changes in the number of species.
@@
 We assessed the sensitivity of general trend results to major data processing decision by rerunning models with increasingly conservative subsets of the data.
-After excluding time series with less than two, three, and four year durations, we found that the general trend for increase in CWM aerial foraging disappeared, but two other trends in CWM's appeared.
-After excluding time series two years long, a general trend appeared for increased CWM use of the forest canopy, and decreased CWM of seed consumption.
-These two trends remained as increasingly shorter time series were excluded suggesting the increase in aerial foraging was an erroneous finding while the other two trends were more robust.
-Percentage of species in a community with trait data did not appear to have an affect on general trend results, as general trends were unchanged with increasing cut offs for percentage of species with trait data.
+Trends for Jaccard similarity and fish consumption were not sensitive to either time series duration or trait coverage.
+After excluding time series with less than three years we found an increasing trend for body mass that remained after excluding time series of less than four and five years.
+We also found a decreasing trend in functional richness for models with a minimum of three years and a minimum of five years.
+A decreasing functional richness trend was found for models with a min of 85% trait coverage, but not the other two coverage levels.
 A complete list of models run in the sensitivity analysis and their results can be found in the supplement.
@@
 
-`r colorize("The most stark area of consensus was the lack of general trend aggregating across communities.", "blue")` We did not detect an overall trend in any functional diversity metrics.
-As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing turnover through time [@dornelas2018], indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends.
-We found no trend in functional change for almost all realms, biomes, and taxonomic groups.
-`r colorize("The one exception was a consensus trend indicating", "blue")` functional richness loss for mammal studies.
-These results are consistent with multiple studies linking anthropogenic drivers to loss of functional diversity in mammal communities [@brodie2021; @cox2022; @li2022].
+`r colorize("The most stark area of consensus was the lack of general trend aggregating across communities.", "blue")` We did not detect an overall trend in any functional diversity metric, `r colorize("corrected or uncorrected", "blue")`.
+As with previous species-based syntheses, we also found no overall trend in species richness accompanied by `r colorize("increasing dissimilarity", "blue")` through time [@dornelas2018], indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends.
+We found significant turnover for all biomes, realm, and taxa except for marine studies, which stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial [@blowes2019].
+However, previous global estimates are dominated by fish communities which we exclude here and may be driving the overall trend while disguising relative stasis in marine bird and mammal communities.
 
-Despite a lack of general trends in functional diversity metrics, we did find evidence of compositional shifts in the trait space with multiple `significant` trends in *CWM*'s for studies coming from the same taxa, realms, climates, or protection statuses, `indicating areas of consensus change across communities`.
-Some of those trends are consistent with existing work in those contexts.
-For example, the decrease in insect consumption for tropical communities reflects well documented declines in tropical insectivorous birds [@sherry2021].
-Reduction in utilization of understory foraging for terrestrial communities could be the result of disturbance like human recreation or increased predation from introduced species.
-Other shifts in diet for both Birds and Mammals, including increasing fruit, seed, and vertebrate consumption and decreasing seed consumption in some climates point to important areas for further exploration.
+We found evidence of functional richness loss in mammal studies, with a loss trend significantly different from zero and significantly different from the trend for birds, indicating a potential loss of functional capacity in these communities.
+This result is consistent with previous work linking anthropogenic drivers such as habitat loss, poaching, and human modification and development to loss of functional diversity in mammal communities across the globe [@brodie2021; @cox2022; @li2022].
+However we did not find trends outside the random expectation for species loss, in contrast to expectations for multiple continents based on IUCN risk categories [@brodie2021].
+In general, functional capacity of mammals shows greater declines that bird communities.
+
+We also found evidence of increasing evenness for Marine studies.
+The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability [@filstrup2019].
+If one or a few species become largely dominant in a community, function will be the result of only those species rather than spread out across a wealth of species responding independently to environmental change [@hillebrand2008; @hillebrand2009].
+Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities [@bagousse-pinguet2019].
+The majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery, which is also indicated in recent species-based work [@spatz2023].
+
+Compositional shifts shown by significant trends in *CWM'*s were restricted to dietary traits, with no evidence of a trend in foraging traits or body mass.
+Increases in fish consumption for bird and temperate studies my be the product of increased availability of easy fish from commercial fishing operations as fish are processed or stocking of inland waterways for sport fishing.
+However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch.
+Decreases in nectar consumption for tropical studies is likely due declines in bat species, as 4 of the 5 tropical studies are for mammal and bats were the only nectar consuming mammals sampled in our data.
+Interpretation of the increases for temperate studies is less clear, as temperate studies included many bird communities with a variety of nectar consumers.
+Similarly, decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.
 
@@
 The lack of general trend in functional diversity metrics belies a huge range of positive and negative trends at the study-level.
-In order to simplify discussion, we will talk about the implications for ecosystem process and vulnerability for communities grouped by their concurrent change in functional diversity metrics, species richness and turnover split into positive, negative or no trend.
-While there are over 150 possible combinations of change direction (or no change) in the 5 metrics, we discuss here the six scenarios that occurred in more than one study: no change, species loss or gain only, loss of functional evenness, richness loss with species turnover, and increase in species and functional richness accompanied by significant turnover.
+In order to simplify discussion, we will talk about the implications for ecosystem process and vulnerability for communities grouped by their concurrent change in corrected functional diversity metrics, species richness and turnover split into positive, negative or no trend.
+While there are over 150 possible combinations of change direction (or no change) in the 5 metrics, we discuss here the six scenarios that occurred in more than one study: no change, species loss or gain only, loss of functional evenness, species richness loss with species turnover, and increase in species and functional richness accompanied by significant turnover.
 We focus on these scenarios to combat potential spurious results due to multiple testing, as it is unlikely all observations of the same scenario are due to false positives.
@@
 First, these may be communities resisting perturbations or simply not experiencing significant perturbations.
-Given the studies in this group come from all possible taxa, realms, climates, and protection statuses, evidence points to communities resisting perturbation, offering some hope that there are areas of the globe where communities are fairing well for now.
+Given the studies in this group come from all possible taxa, realms, biomes, and protection statuses, evidence points to communities resisting perturbation, offering some hope that there are areas of the globe where communities are fairing well for now.
 Alternatively, these may be communities that have experienced or continue to experience significant stress, but lost species or functional diversity outside the observation window.
@@
 Third, these communities may be experiencing directional shifts undetectable by available data.
-For example, species-level trait data does not capture intraspecific shifts in the trait space, which can represent significant changes in overall functional composition and impact maintenance of ecological processes [citation].
+For example, species-level trait data does not capture intraspecific shifts in the trait space, which can represent significant changes in the functional `r colorize("space as a whole", "blue")` and impact maintenance of ecological processes.
 
 Three of the groups fit under the broad umbrella of changes in redundancy.
-By definition, if a community gains or loses species while functional metrics are unchanged, those species represent an increase or decrease in redundancy, so we include communities exhibiting gains or losses in species richness with no change in functional metrics or turnover and richness loss with significant turnover in this umbrella.
+By definition, if a community gains or loses species while functional metrics are unchanged, those species represent an increase or decrease in redundancy, so we include communities exhibiting gains or losses in species richness with no change in functional metrics or turnover and species richness loss with significant turnover in this umbrella.
 For communities exhibiting loss of redundancy (declining species richness), ecological processes are likely being maintained, but capacity to respond to future stressors is reduced [@reich2012].
@@
 The final group exhibits decreasing functional evenness.
-The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability [@filstrup2019].
-If one or a few species become largely dominant in a community, function will be the result of only those species rather than spread out across a wealth of species responding independently to environmental change [@hillebrand2008; @hillebrand2009].
-Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities [@bagousse-pinguet2019].
-These communities are therefore of particular concern, especially considering evidence evenness is more sensitive to environmental change that richness [@filstrup2019].
+These communities are of particular concern not just because of the role evenness plays in maintaining functionality, but because of increasing evidence evenness is more sensitive to environmental change than species richness [@filstrup2019].
+Formal tests of evenness changes as an early warning signal of more catastrophic shifts in functional diversity will be critical next step for understanding the full implication of evenness loss for ecosystem processes.
 
@@
 Identifying key factors mediating the kind of biodiversity change a community experiences is critical for identifying areas of conservation concern and extrapolating results to areas where functional trait data may not be available.
-While it is tempting to draw a one to one line between the degree of human impact and loss of functional integrity (in the form of functional redundancy and richness), that relationship is not borne out by our rough metric of human impact or previous taxa-specific or single community work, which is largely mixed [@flynn2009; @luck2013; @magioli2021; @tinoco2018].
+While it is tempting to draw a one to one line between the degree of human impact and loss of functional integrity (in the form of functional redundancy and richness), that relationship is not borne out by our rough metric of human impact (protection status) or previous taxa-specific or single community work, which is largely mixed [@flynn2009; @luck2013; @magioli2021; @tinoco2018].
 <!-- huge differences between different taxa and the kind of disturbance/land use change -->
@@
 
-`r colorize("The addition of functional trait illuminated a few other key areas where a synthesis approach is incongruous with system-specific studies or expectations based on anthropegenic impacts.", "blue")` For example, mean body size is predicted to decrease as a result of climate change impacts and megafaunal loss [@sheridan2011], a phenomenon which has already been well documented empirically and experimentally in multiple taxa [@caruso2014; @forster2012; @huss2019; @tseng2018].
-However, we found no evidence of a trend in *CWM* body size, with the exception of a single study of marine mammals in the Bahamas which showed a significant increase.
-`r colorize("While the lack of trend could be explained at least in part by data limitations such as shortened observation windows or missing intraspecific data, the fact that only one of the 41 studies for which study-level models were fit showed a change in mean body size, and in the opposite direction of the expectation, is suprising and warrants further reconciliation.", "blue")` <!-- Still, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing [@fritz2009].
+`r colorize("The addition of functional trait illuminated a few key areas where a synthesis approach is incongruous with system-specific studies or expectations based on anthropegenic impacts.", "blue")` For example, we found no evidence of general trends in *CWM* body mass or for studies aggregated by their characteristics despite predicitions that mean body mass is decreasing as a result of climate change impacts and megafaunal loss [@sheridan2011].
+This phenomenon has also been well documented empirically and experimentally in multiple taxa [@caruso2014; @forster2012; @huss2019; @tseng2018].
+Still, climate change impacts are predicted to be most extreme for ectotherms and impact intraspecific traits first before species are lost, neither of which are represented by our data.
+Further, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing [@fritz2009].
 Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened [@ripple2015].
--->
 
-Results for protected areas were also surprising given
-
--   changes in diet
-
--   turnover outside protected areas, not inside ( though might be driven by the fact that there's generally less turnover in terrestrial and temperate studies) [@blowes2019]
+Results for protected areas were also surprising given the hope that protection insulates communities from functional degradation.
+The only difference we found between protected and unprotected areas was significant turnover occurring outside protected areas and no significant turnover inside protected areas.
+While this is an indicator things are generally more static within protect areas, protection has no bearing on whether not there is change in the functional space and the nature of that change.
+Looking at individual studies, there is a mix of both losses and gains in almost all functional diversity and composition metrics across communities found in protected areas.<!-- turnover outside protected areas, not inside ( though might be driven by the fact that there's generally less turnover in terrestrial and temperate studies) [@blowes2019] -->
 
@@
 
-We present here four prevailing scenarios of change experienced in bird and mammal communities.
-While we offer a first discussion of which kinds of change are most common and why, assessing the true global prevalence will require continued efforts to fill data gaps, a well recognized challenge in ecology [@gonzalez2016; @hughes2021; @vellend2017].
-Still, a "scenarios of change" framework can provide structure for future work addressing functional shifts, particularly as we reconcile results from broad-scale syntheses with in-depth single system studies.
-It will be particularly critical to link forms of change to individual drivers to assess which drivers may impact species and functional diversity differently.
-Understanding those links will help identify where directly measuring functional structure instead of just species change is necessary for understand impacts on a given system.
-In addition, direct links to drivers will improve our ability to distinguish between communities experiencing no change due to a lack of perturbation from communities with high resilience in the face of disturbance.
-
-Here we identify trends that are statistically significant, however they may not necessarily be ecologically significant.
-While it is common to link changes in functional metrics to changes in ecosystem processes, those changes are less frequently discussed in terms of the size necessary for ecological impact.
-The degree of change in a process considered ecological meaningful is somewhat subjective and a function of the system and management context, making identifying the ecological meaning of broad-scale aggregate shifts even more challenging.
-It is further hampered by the use of species-level trait data, as the traits most closely linked to a process of interest may not be available [@zhu2017].
-Future trait collection that explicitly considers existing frameworks for linking traits to processes [e.g.
-the response and effect framework @lavorel2002] would facilitate improved ecological interpretation of potential functional changes.
+The addition of functional trait data to the biodiversity change conversation is yet another illustration of how critical context is for understanding ecological patterns [@catford2022; @spake2023].
+Here we assessed how key community characteristics such as biome, realm, taxa, and protection status may partition variation in the functional trends exhibited and found that none of those characteristics seem to make the picture of change clearer.
+Further work linking specific ecological contexts to types of functional change will be critical for identifying new areas of concern, especially when those communities may not otherwise be showing significant changes in species richness.
+The identification of contexts where functional change is most likely occurring will be a significant asset in directing the future collection of trait data, a main barrier for taking a functional approach to biodiversity change.
 

### 2024-03-25 paper/GEB Submission/manuscript_GEB.tex
Mooore edits

@@
 Aim: Despite unprecedented environmental change due to anthropogenic
-pressure, recent work has found increasing species turnover but no
-overall trend in species diversity through time at the local scale.
-Functional diversity provides a potentially powerful alternative
+pressure, recent work has found increasing dissimilarity due to turnover
+but no overall trend in species diversity through time at the local
+scale. Functional diversity provides a potentially powerful alternative
 approach for understanding community composition by linking shifts in
@@
 Results: We found no overall trend in any functional diversity metric,
-despite data replicating species-based patterns of constant richness
-with increasing turnover. The lack of trend held even after correcting
-for changes in species richness. At the study-level, there were also a
-substantial number of time series exhibiting no species or functional
-change, however most studies showed a shift in a species or functional
-metric.
+which held even after correcting for changes in species richness. At the
+study-level, there was substantial heterogeneity in the direction and
+type of functional change exhibited. None of the included study
+characteristics such as taxa, realm, biome, or protection status were
+significantly related to the type of change exhibited by a community.
 
@@
 of functional shift is not more prevalant than the other across many
-taxa, biomes, and realms. At the study-level, we identified four
-prevailing scenarios of species and functional change, which showed
-links to the duration of the observation window. With no one prevailing
-scenario of change, it will be critical to link change scenarios to
-drivers of change, particularly to identify communities with capacity to
-resist drivers from those not experiencing substantial pressure from a
-driver.
+taxa, biomes, and realms. At the study-level, the majority group of
+studies showed no species or functional trends, but other types of
+change were manifest for multiple studies, including studies
+experiencing changes in functional redundancy, increased species and
+functional richness, or loss of functional richness. With no one
+prevailing scenario of change, it will be critical to link change
+scenarios to ecological context.
 \end{abstract}
@@
 
-The meaning of general trends derived from limited data, and their
-relevance for conservation, is a topic of on going debate. Global
-analyses have been heavily criticized for geographic biases, lack of
-data in the most heavily impacted areas, and exclusion of individual
+\textcolor{blue}{Using general trends derived from limited data as a diagnostic for the state of biodiversity or directing conservation is a topic of ongoing debate.}
+Global analyses have been heavily criticized for geographic biases, lack
+of data in the most heavily impacted areas, and exclusion of individual
 studies' ecological context (10--12). Many of these criticisms reflect
@@
 \textcolor{blue}{The expectation for functional change across communities is not obvious from past work, and may or may not follow taxonomic trends}
-(13). While functional loss is frequently cited as one of the most
-pressing concerns of the anthropocene (18--20),
+(13). While \textcolor{blue}{loss of functional diversity} is frequently
+cited as one of the most pressing concerns of the anthropocene (18--20),
 \textcolor{blue}{functional diversity may be maintained even when species are lost from a community}
@@
 heavily impacted communities (27--29). On paleoecological time scales,
-functional composition shows mixed responses to environmental change and
-extinction events (30, 31), with significant impacts of species
-extinctions on functional diversity in some taxa and not others (32).
-For some time periods, functional structure appears to be maintained for
-substantial portions of geological time (33).
+\textcolor{blue}{the functional space} shows mixed responses to
+environmental change and extinction events (30, 31), with significant
+impacts of species extinctions on functional diversity in some taxa and
+not others (32). For some time periods, functional diversity appears to
+be maintained for substantial portions of geological time (33).
 \textcolor{blue}{Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds [@barnagaud2017; @jarzyna2016] and ray-finned fishes, sharks, and rays}
 (34).
+\textcolor{blue}{Additionally, these taxa have been the subject of substantial conservation investment and policy action, so are not necessarily representative of trends for other taxa or the same taxa in different locations.}
 
@@
 perform the first multi-taxa, multi-realm assessment of functional
-diversity change through time. We focus on mammal and bird species as a
-significant subset of the world's biodiversity heavily impacted by
-anthropogenic change. While examining trends in plants, invertebrates,
-and other vertebrate species is of equal interest, trait data for those
-taxa raise additional challenges such as limited and biased species
-coverage (35), a lack of accepted species-level means, and differences
-in the types of traits collected. To ensure comparability across taxa in
-trait type and data quality we therefore focus on mammals and birds.
-Traits were intentionally selected to be representative of a species'
-Eltonian niche, thereby summarizing the functional role they play in the
+diversity change through time. We focus on mammal and bird species as
+\textcolor{blue}{a subset of the world's biodiversity of particular conservation concern that is heavily impacted by anthropogenic change.}While
+examining trends in plants, invertebrates, and other vertebrate species
+is of equal interest, trait data for those taxa raise additional
+challenges such as limited and biased species coverage (35), a lack of
+accepted species-level means, and differences in the types of traits
+collected. To ensure comparability across taxa in trait type and data
+quality we therefore focus on mammals and birds.
+\textcolor{blue}{We include body mass, foraging and dietary traits} that
+were intentionally selected to be representative of a species' Eltonian
+niche, thereby summarizing the functional role they play in the
 community (36). An initial assessment of amphibian trends is included in
@@
 We obtained mammal and bird time series from the BioTIME database, a
-global repository of high-quality assemblage time series. All studies
-included in the database follow consistent sampling protocols and
-represent full assemblages rather than populations of single species
-(31). All time series include abundance of observed species. Following
-best practices for the database (37), studies with multiple sample
-locations were split into individual time series following a
+global repository of high-quality assemblage time series
+\textcolor{blue}{collected from the literature and ongoing monitoring efforts}.
+\textcolor{blue}{Data is structured such that a study encompass all data collected following consistent sampling protocols and may contain multiple time series from separate locations. }
+Time series represent full assemblages rather than populations of single
+species (31), and all time series include abundance of observed species.
+Following best practices for the database (37), studies with multiple
+sample locations were split into individual time series following a
 standardized spatial scale. Scale was set by a global grid with cell
@@
 countries and 12 biomes and 6 different traits (Fig \ref{fig:taxaMap}).
-Data came from both terrestrial and marine realms and five climates
+Data came from both terrestrial and marine realms and five biomes
 (Global, Polar/Temperate, Temperate, Temperate/Tropical, Tropical). The
@@
 for a time series. A negative trend in \emph{J} would therefore indicate
-increasing turnover. We did not impose a correction for unobserved
-species as non-parametric estimators do not assign species identities to
-corrected richness values, and therefore could not be propagated to the
-functional diversity metrics.
+\textcolor{blue}{decreasing similiarity}. We did not impose a correction
+for unobserved species as non-parametric estimators do not assign
+species identities to corrected richness values, and therefore could not
+be propagated to the functional diversity metrics.
 
@@
 community of that species richness.
+\textcolor{blue}{This approach will be less accurate for shorter time series, as we likley will not have captured all available species in the true species pool, but it is impossible to know whether the mean estimate from the null model is an over or under estimate without knowing the functional characteristics of the missing species.}
 
@@
 metric using a linear mixed effects model with a random slope and
-intercept for each study and each time series nested within the study.
-We fit 18 individual \emph{CWM} models. All time series with data for a
-given trait were included in the corresponding \emph{CWM} model. We
-estimated study level trends using individual linear models. For studies
-with more than one times series we fit a random slope and intercept for
-time series. Some study-level models could not be fit for five studies
-for at least one metric due to data limitations, but those studies were
-still included in the general models. They represented 12 of 1350
-study-level models fit for each metric. For further details see the
-supplement. Where appropriate, response variables were \(log\) or
-\(log(x+1)\) transformed to better fit model assumptions of residual
-normality.
+intercept for each study and each time series nested within the study,
+\textcolor{blue}{methods which deal well with the inherent imbalances in our data.}
+We fit 18 individual \emph{CWM} models,
+\textcolor{blue}{one for each trait included in the analsysis}. All time
+series with data for a given trait were included in the corresponding
+\emph{CWM} model. We estimated study level trends using individual
+linear models. For studies with more than one times series we fit a
+random slope and intercept for time series. Some study-level models
+could not be fit for five studies for at least one metric due to data
+limitations, but those studies were still included in the general
+models. They represented 12 of 1350 study-level models fit for each
+metric. For further details see the supplement. Where appropriate,
+response variables were \(log\) or \(log(x+1)\) transformed to better
+fit model assumptions of residual normality.
 
@@
 structure. We estimated within-level slopes and calculated between-level
-contrasts using the \emph{emmeans} package (v1.8.2, 47). We assessed the
-impact of time series duration and start year on study-level trends
-using linear models with duration and start year as predictors. All
-models in our analysis were fit using the \emph{lme4} (v1.1-30) package
-in R (v4.2.3) and p-values were calculated by Satterthwaite's degrees of
-freedom method using the \emph{lmerTest} (v3.1-3) package with a
-significance level of \(\alpha = 0.05\) (48--50).
+contrasts using the \emph{emmeans} package (v1.8.2, 47).
+\textcolor{blue}{For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore refrained from interpreting results for levels where there were fewer than three studies.}
+We assessed the impact of time series duration and start year on
+study-level trends using linear models with duration and start year as
+predictors. All models in our analysis were fit using the \emph{lme4}
+(v1.1-30) package in R (v4.2.3) and p-values were calculated by
+Satterthwaite's degrees of freedom method using the \emph{lmerTest}
+(v3.1-3) package with a significance level of \(\alpha = 0.05\)
+(48--50).
 
@@
 
-We found no significant overall trend in species richness or functional
-diversity metrics (observed or \textcolor{blue}{corrected}) (Fig
+We found no significant overall \textcolor{blue}{temporal} trend in
+species richness or functional diversity metrics
+\textcolor{blue}{including functional richness, evenness, or divergence}
+(observed or \textcolor{blue}{corrected}) (Fig
 \ref{fig:timeseriesPlot}). We did find a significant overall decrease in
 Jaccard similarity, indicating accumulating changes in species
-composition. Non-significant overall trends indicate that although some
-studies experience increasing or decreasing trends, the average trend
-across studies was plausibly 0 (Table \ref{tab:resultsTab}).
-\textcolor{blue}{Within-group} trends for different taxa, biomes,
-realms, or protection statuses were also non-significant for richness
-and functional diversity metrics, with the exception of a significantly
-increasing trend for functional evenness of global studies
-(characterized by having samples on multiple continents), and a
-significantly decreasing functional richness slope for
-temperate/tropical studies and mammal studies. However, trends were not
-exhibited for the \textcolor{blue}{corrected} metric indicating that
-differences \textcolor{blue}{in functional diversity metrics} were
-largely due to changes in species richness. Further, with only two
-global studies, the trend should not be considered truly general. The
-general trends for \emph{CWM} models were similarly not significant,
-with a significant positive trend for only percentage of fish in diet
-composition.
+composition. Non-significant overall \textcolor{blue}{temporal}trends
+indicate that although some studies experience increasing or decreasing
+trends, the average trend across studies
+\textcolor{blue}{was not significantly different from zero}` (Table
+\ref{tab:resultsTab}). \textcolor{blue}{Within-group} trends for
+different taxa, biomes, realms, or protection statuses were also
+non-significant for richness and functional diversity metrics, with the
+exception of a significantly increasing trend for functional evenness of
+marine studies, and a significantly decreasing functional richness slope
+for mammal studies. However, trends were not exhibited for the
+\textcolor{blue}{corrected} metric indicating that differences
+\textcolor{blue}{in functional diversity metrics} were largely due to
+changes in species richness. The general trends for \emph{CWM} models
+were similarly not significant, with a significant positive trend for
+only percentage of fish in diet composition.
 
 \begin{figure}
-\includegraphics[width=\textwidth]{../../figures/3met_long} \caption{Plots of time series-level trends with line color corresponding to climatic region, with data points in grey and the overall metric mean in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES}\label{fig:timeseriesPlot}
+\includegraphics[width=\textwidth]{../../figures/3met_long} \caption{Plots of time series-level trends with line color corresponding to climatic region, with data points in grey and the overall mean slope for a metric in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES}\label{fig:timeseriesPlot}
 \end{figure}
 
-We did find significant differences between taxa, realms, and climate
-for Jaccard similarity and some of the \emph{CWM's}. For example, while
+We did find significant differences between taxa, realms, and biomes for
+Jaccard similarity and some of the \emph{CWM's}. For example, while
 Jaccard similarity was decreasing in the general trend and there were
-significant within group slopes for Mammals, Birds, and Terrestrial
-communities, there was no significant slope for the marine realm,
-indicating that the general trend is mostly driven by turnover in
-terrestrial communities. We also found a significant decrease in Jaccard
-similarity in unprotected areas only, with no trend for protected areas.
-For bird communities, we also found within group trends and between
-group differences for trends in foraging behavior. We found a
-significant increasing trend in utilization of the canopy in Tropical
-communities that was significantly different from the trends for
-Polar/Temperate and Temperate communities. There was a significant
-decrease in utilization of the understory for Terrestrial communities
-and significant increase in foraging below the water surface for global
-studies (but see the previous limitations of Global data). There was
-also a significant positive slope for \emph{CWM} body mass for the
-single Temperate/Tropical study (with three time series) which were
-marine mammal communities.
-
-We found significant dietary shifts across communities, with a
-significant increase in fruit consumption in Terrestrial communities and
-a significant decrease in nectar consumption in Tropical communities, a
-trend significantly different than that for Terrestrial communities.
-There was a significant increase in seed consumption in bird species,
-which was significantly different from the trend for Mammal communities.
-There was a significant increase in fish consumption for the two Global
-studies. Vertebrate consumption significantly decreased for Marine
-studies and for studies of Global and Tropical communities.
-\textcolor{blue}{There was also a significant increase in fish and fruit consumption for unprotected areas, but not protected.}
-
-At the study level, 15 of 50 studies exhibited a significant trend in
-species richness and 11 exhibited significant turnover. For observed
-functional metrics, 25 of 50 studies exhibited a trend in a least one
-metric, and 15 of 50 studies exhibited a significant trend for at least
-one \textcolor{blue}{corrected} metric (Table \ref{tab:trendTab}). In
-general, there were more significant trends for uncorrected metrics,
-with some disappearing after correction, indicating that those trends
-were likely due to changes in the number of species. Hypothesis testing
-for study-level trends is likely affected by multiple testing issues and
-some trends identified as significant are therefore potentially
-spurious. Rather than interpreting changes in specific studies, we
-present these results as a general picture of the kinds of trends
-experienced by communities.
-
-Study-level slopes for functional diversity metrics were significantly
-related to start year of the time series for Jaccard similarity and
-functional evenness, both of which had significantly more negative
-slopes with more recent start year. No functional diversity metrics were
-significantly related to the start year of the time series.
+significant within group slopes for Mammals, Birds, Terrestrial,
+Temperate, and Tropical communities, there was no significant slope for
+the marine realm, indicating that the general trend is mostly driven by
+turnover in terrestrial communities. We also found a significant
+decrease in Jaccard similarity in unprotected areas only, with no trend
+for protected areas. We found significant dietary shifts across
+communities, with a significant general trend of increasing fish
+consumption, which was also reflected in increasing fish consumption
+trends for Bird studies and Temperate studies. We found a significantly
+increasing trend in nectar consumption for Temperate studies but a
+significantly decreasing tend for Tropical studies. We found
+significantly decreasing trend for vertebrate consumption in Marine
+studies and Tropical studies.
+
+At the study level, 16 of 50 studies exhibited a significant trend in
+species richness and 11 exhibited significant turnover. For uncorrected
+functional diversity metrics, 25 of 50 studies exhibited a trend in a
+least one metric, and 15 of 50 studies exhibited a significant trend for
+at least one \textcolor{blue}{corrected} metric (Table
+\ref{tab:trendTab}). In general, there were more significant trends for
+uncorrected metrics, with some disappearing after correction, indicating
+that those trends were likely due to changes in the number of species.
+Hypothesis testing for study-level trends is likely affected by multiple
+testing issues and some trends identified as significant are therefore
+potentially spurious. Rather than interpreting changes in specific
+studies, we present these results as a general picture of the kinds of
+trends experienced by communities.
+
+Study-level slopes were significantly related to start year of the time
+series for Jaccard similarity and functional evenness models, both of
+which had significantly more negative slopes with more recent start
+year. No functional diversity metrics were significantly related to the
+duration of the time series.
 
@@
 processing decision by rerunning models with increasingly conservative
-subsets of the data. After excluding time series with less than two,
-three, and four year durations, we found that the general trend for
-increase in CWM aerial foraging disappeared, but two other trends in
-CWM's appeared. After excluding time series two years long, a general
-trend appeared for increased CWM use of the forest canopy, and decreased
-CWM of seed consumption. These two trends remained as increasingly
-shorter time series were excluded suggesting the increase in aerial
-foraging was an erroneous finding while the other two trends were more
-robust. Percentage of species in a community with trait data did not
-appear to have an affect on general trend results, as general trends
-were unchanged with increasing cut offs for percentage of species with
-trait data. A complete list of models run in the sensitivity analysis
-and their results can be found in the supplement.
+subsets of the data. Trends for Jaccard similarity and fish consumption
+were not sensitive to either time series duration or trait coverage.
+After excluding time series with less than three years we found an
+increasing trend for body mass that remained after excluding time series
+of less than four and five years. We also found a decreasing trend in
+functional richness for models with a minimum of three years and a
+minimum of five years. A decreasing functional richness trend was found
+for models with a min of 85\% trait coverage, but not the other two
+coverage levels. A complete list of models run in the sensitivity
+analysis and their results can be found in the supplement.
 
@@
 \textcolor{blue}{The most stark area of consensus was the lack of general trend aggregating across communities.}
-We did not detect an overall trend in any functional diversity metrics.
-As with previous species-based syntheses, we also found no overall trend
-in species richness accompanied by increasing turnover through time
-(31), indicating that non-significant trends in functional metrics are
-consistent with similar well-documented species derived trends. We found
-no trend in functional change for almost all realms, biomes, and
-taxonomic groups.
-\textcolor{blue}{The one exception was a consensus trend indicating}
-functional richness loss for mammal studies. These results are
-consistent with multiple studies linking anthropogenic drivers to loss
-of functional diversity in mammal communities (51--53).
-
-Despite a lack of general trends in functional diversity metrics, we did
-find evidence of compositional shifts in the trait space with multiple
-\texttt{significant} trends in \emph{CWM}'s for studies coming from the
-same taxa, realms, climates, or protection statuses,
-\texttt{indicating\ areas\ of\ consensus\ change\ across\ communities}.
-Some of those trends are consistent with existing work in those
-contexts. For example, the decrease in insect consumption for tropical
-communities reflects well documented declines in tropical insectivorous
-birds (54). Reduction in utilization of understory foraging for
-terrestrial communities could be the result of disturbance like human
-recreation or increased predation from introduced species. Other shifts
-in diet for both Birds and Mammals, including increasing fruit, seed,
-and vertebrate consumption and decreasing seed consumption in some
-climates point to important areas for further exploration.
+We did not detect an overall trend in any functional diversity metric,
+\textcolor{blue}{corrected or uncorrected}. As with previous
+species-based syntheses, we also found no overall trend in species
+richness accompanied by \textcolor{blue}{increasing dissimilarity}
+through time (31), indicating that non-significant trends in functional
+metrics are consistent with similar well-documented species derived
+trends. We found significant turnover for all biomes, realm, and taxa
+except for marine studies, which stands in contrast to other global
+estimates of biodiversity change that found higher turnover in marine
+systems than terrestrial (37). However, previous global estimates are
+dominated by fish communities which we exclude here and may be driving
+the overall trend while disguising relative stasis in marine bird and
+mammal communities.
+
+We found evidence of functional richness loss in mammal studies, with a
+loss trend significantly different from zero and significantly different
+from the trend for birds, indicating a potential loss of functional
+capacity in these communities. This result is consistent with previous
+work linking anthropogenic drivers such as habitat loss, poaching, and
+human modification and development to loss of functional diversity in
+mammal communities across the globe (51--53). However we did not find
+trends outside the random expectation for species loss, in contrast to
+expectations for multiple continents based on IUCN risk categories (51).
+In general, functional capacity of mammals shows greater declines that
+bird communities.
+
+We also found evidence of increasing evenness for Marine studies. The
+empirical link between changes in evenness and ecosystem process is the
+most poorly studied of the metrics, however theory indicates that
+increased evenness improves function and stability (54). If one or a few
+species become largely dominant in a community, function will be the
+result of only those species rather than spread out across a wealth of
+species responding independently to environmental change (42, 55).
+Further, there is some evidence that maintaining evenness is important
+for supporting multifunctionality in communities (56). The majority of
+marine studies are of seabird communities (7 of 9), so the general trend
+may be an early indicator of pay off from investments in sea bird
+recovery, which is also indicated in recent species-based work (57).
+
+Compositional shifts shown by significant trends in \emph{CWM'}s were
+restricted to dietary traits, with no evidence of a trend in foraging
+traits or body mass. Increases in fish consumption for bird and
+temperate studies my be the product of increased availability of easy
+fish from commercial fishing operations as fish are processed or
+stocking of inland waterways for sport fishing. However, fisheries also
+have negative impacts on waterbirds through competition for fish or bird
+mortality due to by catch. Decreases in nectar consumption for tropical
+studies is likely due declines in bat species, as 4 of the 5 tropical
+studies are for mammal and bats were the only nectar consuming mammals
+sampled in our data. Interpretation of the increases for temperate
+studies is less clear, as temperate studies included many bird
+communities with a variety of nectar consumers. Similarly, decreasing
+trends for vertebrate consumption in marine and tropical studies are not
+obviously consistent with known changes in those contexts and warrant
+further examination.
 
@@
 experiencing significant perturbations. Given the studies in this group
-come from all possible taxa, realms, climates, and protection statuses,
+come from all possible taxa, realms, biomes, and protection statuses,
 evidence points to communities resisting perturbation, offering some
@@
 data does not capture intraspecific shifts in the trait space, which can
-represent significant changes in overall functional composition and
-impact maintenance of ecological processes {[}citation{]}.
+represent significant changes in the functional
+\textcolor{blue}{space as a whole} and impact maintenance of ecological
+processes.
 
@@
 losses in species richness with no change in functional metrics or
-turnover and richness loss with significant turnover in this umbrella.
-For communities exhibiting loss of redundancy (declining species
-richness), ecological processes are likely being maintained, but
-capacity to respond to future stressors is reduced (55). These
+turnover and species richness loss with significant turnover in this
+umbrella. For communities exhibiting loss of redundancy (declining
+species richness), ecological processes are likely being maintained, but
+capacity to respond to future stressors is reduced (58). These
 communities are actually fairing better than expected looking at
 species-based metrics alone, but are also in a precarious position for
-maintaining ecological function into the future (56). Conversely,
+maintaining ecological function into the future (59). Conversely,
 communities exhibiting a increase in redundancy (via an increase in
@@
 stressors, but are not actually expanding their functional capacity as
-may be assumed by looking at species gains alone (57).
+may be assumed by looking at species gains alone (60).
 
@@
 diversity for North American breeding birds over the last few decades
-after a period of decline(58), and loss of common, functionally general
+after a period of decline(61), and loss of common, functionally general
 species even as rare species are increasing in North America and Europe
-(59--61).
-
-The final group exhibits decreasing functional evenness. The empirical
-link between changes in evenness and ecosystem process is the most
-poorly studied of the metrics, however theory indicates that increased
-evenness improves function and stability (62). If one or a few species
-become largely dominant in a community, function will be the result of
-only those species rather than spread out across a wealth of species
-responding independently to environmental change (42, 63). Further,
-there is some evidence that maintaining evenness is important for
-supporting multifunctionality in communities (64). These communities are
-therefore of particular concern, especially considering evidence
-evenness is more sensitive to environmental change that richness (62).
+(62--64).
+
+The final group exhibits decreasing functional evenness. These
+communities are of particular concern not just because of the role
+evenness plays in maintaining functionality, but because of increasing
+evidence evenness is more sensitive to environmental change than species
+richness (54). Formal tests of evenness changes as an early warning
+signal of more catastrophic shifts in functional diversity will be
+critical next step for understanding the full implication of evenness
+loss for ecosystem processes.
 
@@
 richness), that relationship is not borne out by our rough metric of
-human impact or previous taxa-specific or single community work, which
-is largely mixed (26, 65--67).
+human impact (protection status) or previous taxa-specific or single
+community work, which is largely mixed (26, 65--67).
 
@@
 
-\textcolor{blue}{The addition of functional trait illuminated a few other key areas where a synthesis approach is incongruous with system-specific studies or expectations based on anthropegenic impacts.}
-For example, mean body size is predicted to decrease as a result of
-climate change impacts and megafaunal loss (68), a phenomenon which has
-already been well documented empirically and experimentally in multiple
-taxa (69--72). However, we found no evidence of a trend in \emph{CWM}
-body size, with the exception of a single study of marine mammals in the
-Bahamas which showed a significant increase.
-\textcolor{blue}{While the lack of trend could be explained at least in part by data limitations such as shortened observation windows or missing intraspecific data, the fact that only one of the 41 studies for which study-level models were fit showed a change in mean body size, and in the opposite direction of the expectation, is suprising and warrants further reconciliation.}
-
-Results for protected areas were also surprising given
-
-\begin{itemize}
-\item
-  changes in diet
-\item
-  turnover outside protected areas, not inside ( though might be driven
-  by the fact that there's generally less turnover in terrestrial and
-  temperate studies) (37)
-\end{itemize}
+\textcolor{blue}{The addition of functional trait illuminated a few key areas where a synthesis approach is incongruous with system-specific studies or expectations based on anthropegenic impacts.}
+For example, we found no evidence of general trends in \emph{CWM} body
+mass or for studies aggregated by their characteristics despite
+predicitions that mean body mass is decreasing as a result of climate
+change impacts and megafaunal loss (68). This phenomenon has also been
+well documented empirically and experimentally in multiple taxa
+(69--72). Still, climate change impacts are predicted to be most extreme
+for ectotherms and impact intraspecific traits first before species are
+lost, neither of which are represented by our data. Further, many of the
+studies in our dataset draw from areas that may have experienced
+significant loss of large-bodied species before the observation window,
+with contemporary loss rates slowing (73). Trends could be significantly
+different for the same time periods in regions of sub-Saharan Africa,
+for example, which has poor representation in our dataset but where
+megafauna exist on the landscape and are increasingly threatened (74).
+
+Results for protected areas were also surprising given the hope that
+protection insulates communities from functional degradation. The only
+difference we found between protected and unprotected areas was
+significant turnover occurring outside protected areas and no
+significant turnover inside protected areas. While this is an indicator
+things are generally more static within protect areas, protection has no
+bearing on whether not there is change in the functional space and the
+nature of that change. Looking at individual studies, there is a mix of
+both losses and gains in almost all functional diversity and composition
+metrics across communities found in protected areas.
 
@@
 fact, study-level trends indicate quite the opposite, that functional
-shifts with unknown implications for ecosystem processes may be going
-undetected by common species-based approaches, particularly for short
-observation time windows. While the majority of studies in our dataset
-did not experience a significant functional loss, a substantial body of
-work links functional degradation to species losses as a result of
-direct human intervention in the form of land use change and
-intensification or habitat fragmentation, indicating that those studies
-are simply not representative of the kinds of impacts of greatest policy
-concern (67).
+shifts with negative or yet unknown implications for ecosystem processes
+may be going undetected by common species-based approaches. For example,
+loss of evenness in communities with constant species richness may be an
+first sign of a community being impacted by environmental change, with
+negative implications for stability and function.
+
+One of the biggest threats to biodiversity is the wholesale conversion
+of natural areas to urban or human-dominated landscapes (75). Typical
+long term monitoring such as those included in our study stops before
+this conversion occurs, leaving the resultant precipitous declines in
+biodiversity unrecorded (10). This is a known issue with the culture of
+long term monitoring, and our results should not be removed from that
+context. Rather, this studies captures communities that are likely
+experiencing a degree of human intervention, but are still largely
+nature-dominated.
 
@@
 
-We present here four prevailing scenarios of change experienced in bird
-and mammal communities. While we offer a first discussion of which kinds
-of change are most common and why, assessing the true global prevalence
-will require continued efforts to fill data gaps, a well recognized
-challenge in ecology (6, 12, 73). Still, a ``scenarios of change''
-framework can provide structure for future work addressing functional
-shifts, particularly as we reconcile results from broad-scale syntheses
-with in-depth single system studies. It will be particularly critical to
-link forms of change to individual drivers to assess which drivers may
-impact species and functional diversity differently. Understanding those
-links will help identify where directly measuring functional structure
-instead of just species change is necessary for understand impacts on a
-given system. In addition, direct links to drivers will improve our
-ability to distinguish between communities experiencing no change due to
-a lack of perturbation from communities with high resilience in the face
-of disturbance.
-
-Here we identify trends that are statistically significant, however they
-may not necessarily be ecologically significant. While it is common to
-link changes in functional metrics to changes in ecosystem processes,
-those changes are less frequently discussed in terms of the size
-necessary for ecological impact. The degree of change in a process
-considered ecological meaningful is somewhat subjective and a function
-of the system and management context, making identifying the ecological
-meaning of broad-scale aggregate shifts even more challenging. It is
-further hampered by the use of species-level trait data, as the traits
-most closely linked to a process of interest may not be available (74).
-Future trait collection that explicitly considers existing frameworks
-for linking traits to processes (e.g.~the response and effect framework
-15) would facilitate improved ecological interpretation of potential
-functional changes.
+The addition of functional trait data to the biodiversity change
+conversation is yet another illustration of how critical context is for
+understanding ecological patterns (76, 77). Here we assessed how key
+community characteristics such as biome, realm, taxa, and protection
+status may partition variation in the functional trends exhibited and
+found that none of those characteristics seem to make the picture of
+change clearer. Further work linking specific ecological contexts to
+types of functional change will be critical for identifying new areas of
+concern, especially when those communities may not otherwise be showing
+significant changes in species richness. The identification of contexts
+where functional change is most likely occurring will be a significant
+asset in directing the future collection of trait data, a main barrier
+for taking a functional approach to biodiversity change.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sherry2021}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-filstrup2019}{}}%
 \CSLLeftMargin{54. }%
-\CSLRightInline{Sherry TW (2021) Sensitivity of tropical insectivorous
-birds to the anthropocene: A review of multiple mechanisms and
-conservation implications. \emph{Frontiers in Ecology and Evolution} 9.
-Available at:
-\url{https://www.frontiersin.org/articles/10.3389/fevo.2021.662873}.}
+\CSLRightInline{Filstrup CT, King KBS, McCullough IM (2019)
+\href{https://doi.org/10.1111/ele.13407}{Evenness effects mask richness
+effects on ecosystem functioning at macro-scales in lakes}.
+\emph{Ecology Letters} 22(12):2120--2129.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hillebrand2008}{}}%
+\CSLLeftMargin{55. }%
+\CSLRightInline{Hillebrand H, Bennett DM, Cadotte MW (2008)
+\href{https://doi.org/10.1890/07-1053.1}{Consequences of Dominance: A
+Review of Evenness Effects on Local and Regional Ecosystem Processes}.
+\emph{Ecology} 89(6):1510--1520.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-bagousse-pinguet2019}{}}%
+\CSLLeftMargin{56. }%
+\CSLRightInline{Bagousse-Pinguet YL, et al. (2019)
+\href{https://doi.org/10.1073/pnas.1815727116}{Phylogenetic, functional,
+and taxonomic richness have both positive and negative effects on
+ecosystem multifunctionality}. \emph{Proceedings of the National Academy
+of Sciences} 116(17):8419--8424.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-spatz2023}{}}%
+\CSLLeftMargin{57. }%
+\CSLRightInline{Spatz DR, et al. (2023)
+\href{https://doi.org/10.1073/pnas.2214574120}{Tracking the global
+application of conservation translocation and social attraction to
+reverse seabird declines}. \emph{Proceedings of the National Academy of
+Sciences} 120(16):e2214574120.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-reich2012}{}}%
-\CSLLeftMargin{55. }%
+\CSLLeftMargin{58. }%
 \CSLRightInline{Reich PB, et al. (2012)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-pimiento2020a}{}}%
-\CSLLeftMargin{56. }%
+\CSLLeftMargin{59. }%
 \CSLRightInline{Pimiento C, et al. (2020)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-biggs2020}{}}%
-\CSLLeftMargin{57. }%
+\CSLLeftMargin{60. }%
 \CSLRightInline{Biggs CR, et al. (2020)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-jarzyna2016}{}}%
-\CSLLeftMargin{58. }%
+\CSLLeftMargin{61. }%
 \CSLRightInline{Jarzyna MA, Jetz W (2016)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-inger2015}{}}%
-\CSLLeftMargin{59. }%
+\CSLLeftMargin{62. }%
 \CSLRightInline{Inger R, et al. (2015)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rosenberg2019}{}}%
-\CSLLeftMargin{60. }%
+\CSLLeftMargin{63. }%
 \CSLRightInline{Rosenberg KV, et al. (2019) Decline of the North
@@
 \leavevmode\vadjust pre{\hypertarget{ref-schipper2016}{}}%
-\CSLLeftMargin{61. }%
+\CSLLeftMargin{64. }%
 \CSLRightInline{Schipper AM, et al. (2016)
@@
 2010}. \emph{Global Change Biology} 22(12):3948--3959.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-filstrup2019}{}}%
-\CSLLeftMargin{62. }%
-\CSLRightInline{Filstrup CT, King KBS, McCullough IM (2019)
-\href{https://doi.org/10.1111/ele.13407}{Evenness effects mask richness
-effects on ecosystem functioning at macro-scales in lakes}.
-\emph{Ecology Letters} 22(12):2120--2129.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hillebrand2008}{}}%
-\CSLLeftMargin{63. }%
-\CSLRightInline{Hillebrand H, Bennett DM, Cadotte MW (2008)
-\href{https://doi.org/10.1890/07-1053.1}{Consequences of Dominance: A
-Review of Evenness Effects on Local and Regional Ecosystem Processes}.
-\emph{Ecology} 89(6):1510--1520.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-bagousse-pinguet2019}{}}%
-\CSLLeftMargin{64. }%
-\CSLRightInline{Bagousse-Pinguet YL, et al. (2019)
-\href{https://doi.org/10.1073/pnas.1815727116}{Phylogenetic, functional,
-and taxonomic richness have both positive and negative effects on
-ecosystem multifunctionality}. \emph{Proceedings of the National Academy
-of Sciences} 116(17):8419--8424.}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-hughes2021}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-fritz2009}{}}%
 \CSLLeftMargin{73. }%
-\CSLRightInline{Hughes AC, et al. (2021)
-\href{https://doi.org/10.1111/ecog.05926}{Sampling biases shape our view
-of the natural world}. \emph{Ecography} 44(9):1259--1269.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zhu2017}{}}%
+\CSLRightInline{Fritz SA, Bininda-Emonds ORP, Purvis A (2009)
+\href{https://doi.org/10.1111/j.1461-0248.2009.01307.x}{Geographical
+variation in predictors of mammalian extinction risk: big is bad, but
+only in the tropics}. \emph{Ecology Letters} 12(6):538--549.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-ripple2015}{}}%
 \CSLLeftMargin{74. }%
-\CSLRightInline{Zhu L, et al. (2017)
-\href{https://doi.org/10.1038/s41598-017-03812-8}{Trait choice
-profoundly affected the ecological conclusions drawn from functional
-diversity measures}. \emph{Scientific Reports} 7(1):3643.}
+\CSLRightInline{Ripple WJ, et al. (2015)
+\href{https://doi.org/10.1126/sciadv.1400103}{Collapse of the world{'}s
+largest herbivores}. \emph{Science Advances} 1(4):e1400103.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-tilman2017}{}}%
+\CSLLeftMargin{75. }%
+\CSLRightInline{Tilman D, et al. (2017)
+\href{https://doi.org/10.1038/nature22900}{Future threats to
+biodiversity and pathways to their prevention}. \emph{Nature}
+546(7656):73--81.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-catford2022}{}}%
+\CSLLeftMargin{76. }%
+\CSLRightInline{Catford JA, Wilson JRU, Pyšek P, Hulme PE, Duncan RP
+(2022) \href{https://doi.org/10.1016/j.tree.2021.09.007}{Addressing
+context dependence in ecology}. \emph{Trends in Ecology \& Evolution}
+37(2):158--170.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-spake2023}{}}%
+\CSLLeftMargin{77. }%
+\CSLRightInline{Spake R, et al. (2023)
+\href{https://doi.org/10.1111/brv.12939}{Understanding {`}it depends{'}
+in ecology: a guide to hypothesising, visualising and interpreting
+statistical interactions}. \emph{Biological Reviews} 98(4):983--1002.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-pardieck2020}{}}%
-\CSLLeftMargin{75. }%
+\CSLLeftMargin{78. }%
 \CSLRightInline{Pardieck KL, David ZJ, Lutmerding M, Aponte V, Hudson
@@
 \leavevmode\vadjust pre{\hypertarget{ref-piropno}{}}%
-\CSLLeftMargin{76. }%
+\CSLLeftMargin{79. }%
 \CSLRightInline{PIROP northwest atlantic 1965{\textendash}1992 - OBIS
@@
 \leavevmode\vadjust pre{\hypertarget{ref-silva}{}}%
-\CSLLeftMargin{77. }%
+\CSLLeftMargin{80. }%
 \CSLRightInline{Silva FR da Brazil dataset 1.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-scott}{}}%
-\CSLLeftMargin{78. }%
+\CSLLeftMargin{81. }%
 \CSLRightInline{Scott D, Metts B, Lance S The rainbow bay long-term
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rossaferes1997}{}}%
-\CSLLeftMargin{79. }%
+\CSLLeftMargin{82. }%
 \CSLRightInline{Rossa-Feres D de C (1997) Community ecology of anura
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bakker1990}{}}%
-\CSLLeftMargin{80. }%
+\CSLLeftMargin{83. }%
 \CSLRightInline{Bakker C, Herman PMJ (1990) Phytoplankton in the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-berezovikovn.n.2004}{}}%
-\CSLLeftMargin{81. }%
+\CSLLeftMargin{84. }%
 \CSLRightInline{Berezovikov, N.N. (2004) The birds of settlements in
@@
 \leavevmode\vadjust pre{\hypertarget{ref-enemar2004}{}}%
-\CSLLeftMargin{82. }%
+\CSLLeftMargin{85. }%
 \CSLRightInline{Enemar A, Sjöstrand B, Andersson G, Von Proschwitz T
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gido2019}{}}%
-\CSLLeftMargin{83. }%
+\CSLLeftMargin{86. }%
 \CSLRightInline{Gido K (2019) CFP01 Fish population on selected
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hogstad1993}{}}%
-\CSLLeftMargin{84. }%
+\CSLLeftMargin{87. }%
 \CSLRightInline{Hogstad O (1993)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-holmes1986}{}}%
-\CSLLeftMargin{85. }%
+\CSLLeftMargin{88. }%
 \CSLRightInline{Holmes RT, Sherry TW, Sturges FW (1986)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-jahncke2006}{}}%
-\CSLLeftMargin{86. }%
+\CSLLeftMargin{89. }%
 \CSLRightInline{Jahncke J, Rintoul C (2006) CalCOFI and NMFS seabird and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-krivenkov.g.1991}{}}%
-\CSLLeftMargin{87. }%
+\CSLLeftMargin{90. }%
 \CSLRightInline{Krivenko, V.G. (1991) Waterfowl and their protection.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-melnikovy.i.2000}{}}%
-\CSLLeftMargin{88. }%
+\CSLLeftMargin{91. }%
 \CSLRightInline{Melnikov, Y.I., Melnikova, N., Pronkevich, V.V. (2000)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2013}{}}%
-\CSLLeftMargin{89. }%
+\CSLLeftMargin{92. }%
 \CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2014}{}}%
-\CSLLeftMargin{90. }%
+\CSLLeftMargin{93. }%
 \CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitori2014}{}}%
-\CSLLeftMargin{91. }%
+\CSLLeftMargin{94. }%
 \CSLRightInline{Monitoring site 1000 village survey - medium and large
@@
 \leavevmode\vadjust pre{\hypertarget{ref-preston1960}{}}%
-\CSLLeftMargin{92. }%
+\CSLLeftMargin{95. }%
 \CSLRightInline{Preston FW (1960)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-svensson2006}{}}%
-\CSLLeftMargin{93. }%
+\CSLLeftMargin{96. }%
 \CSLRightInline{Svensson S (2006) Species composition and population
@@
 \leavevmode\vadjust pre{\hypertarget{ref-svensson2010}{}}%
-\CSLLeftMargin{94. }%
+\CSLLeftMargin{97. }%
 \CSLRightInline{Svensson S, Thorner AM, Nyholm NEI (2010) Species
@@
 \leavevmode\vadjust pre{\hypertarget{ref-thorn2016changes}{}}%
-\CSLLeftMargin{95. }%
+\CSLLeftMargin{98. }%
 \CSLRightInline{Thorn S, et al. (2016)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-usfs}{}}%
-\CSLLeftMargin{96. }%
+\CSLLeftMargin{99. }%
 \CSLRightInline{USFS Landbird monitoring program (UMT-LBMP). Available
@@
 \leavevmode\vadjust pre{\hypertarget{ref-vermontcenterforecostudies2015}{}}%
-\CSLLeftMargin{97. }%
+\CSLLeftMargin{100. }%
 \CSLRightInline{Vermont Center For Ecostudies, Lambert JD, Hart J (2015)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-vickery1984}{}}%
-\CSLLeftMargin{98. }%
+\CSLLeftMargin{101. }%
 \CSLRightInline{Vickery WL, Nudds TD (1984)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-waide2017}{}}%
-\CSLLeftMargin{99. }%
+\CSLLeftMargin{102. }%
 \CSLRightInline{Waide R (2017) Bird abundance - point counts.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-williamson1983}{}}%
-\CSLLeftMargin{100. }%
+\CSLLeftMargin{103. }%
 \CSLRightInline{Williamson M (1983)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-zakharovv.d.1998}{}}%
-\CSLLeftMargin{101. }%
+\CSLLeftMargin{104. }%
 \CSLRightInline{Zakharov, V.D. (1998) Biodiversity of bird population of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-meditss2011}{}}%
-\CSLLeftMargin{102. }%
+\CSLLeftMargin{105. }%
 \CSLRightInline{MEDITS seabird surveys 1999 / 2000 / 2002 (2011)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-animald}{}}%
-\CSLLeftMargin{103. }%
+\CSLLeftMargin{106. }%
 \CSLRightInline{Animal demography unit - coordinated waterbird counts
@@
 \leavevmode\vadjust pre{\hypertarget{ref-baltics}{}}%
-\CSLLeftMargin{104. }%
+\CSLLeftMargin{107. }%
 \CSLRightInline{Baltic seabirds transect surveys Available at:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-thorn2016response}{}}%
-\CSLLeftMargin{105. }%
+\CSLLeftMargin{108. }%
 \CSLRightInline{Thorn S, et al. (2016)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ernest2009}{}}%
-\CSLLeftMargin{106. }%
+\CSLLeftMargin{109. }%
 \CSLRightInline{Ernest SKM, Valone TJ, Brown JH (2009)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-friggens2008}{}}%
-\CSLLeftMargin{107. }%
+\CSLLeftMargin{110. }%
 \CSLRightInline{Friggens M (2008) Sevilleta LTER small mammal population
@@
 \leavevmode\vadjust pre{\hypertarget{ref-jalilova.b.2014}{}}%
-\CSLLeftMargin{108. }%
+\CSLLeftMargin{111. }%
 \CSLRightInline{Jalilov, A. B., Andreychev, A. V., Kuznetsov, V. A.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kartzinel2014}{}}%
-\CSLLeftMargin{109. }%
+\CSLLeftMargin{112. }%
 \CSLRightInline{Kartzinel TR, et al. (2014)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kaufman2019}{}}%
-\CSLLeftMargin{110. }%
+\CSLLeftMargin{113. }%
 \CSLRightInline{Kaufman D (2019) CSM01 Seasonal Summary of Numbers of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-krefting1974}{}}%
-\CSLLeftMargin{111. }%
+\CSLLeftMargin{114. }%
 \CSLRightInline{Krefting LW, Ahlgren CE (1974)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lightfoot}{}}%
-\CSLLeftMargin{112. }%
+\CSLLeftMargin{115. }%
 \CSLRightInline{Lightfoot D, Schooley RL SMES rodent trapping data,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-malyshevy.s.2011}{}}%
-\CSLLeftMargin{113. }%
+\CSLLeftMargin{116. }%
 \CSLRightInline{Malyshev, Y. S. (2011) On the diagnostic techniques of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-meyer2008}{}}%
-\CSLLeftMargin{114. }%
+\CSLLeftMargin{117. }%
 \CSLRightInline{Meyer CFJ, Kalko EKV (2008)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nedosekinv.y.2015}{}}%
-\CSLLeftMargin{115. }%
+\CSLLeftMargin{118. }%
 \CSLRightInline{Nedosekin, V. Y. (2015) Long-term dynamics of the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-prins1990}{}}%
-\CSLLeftMargin{116. }%
+\CSLLeftMargin{119. }%
 \CSLRightInline{Prins HHT, Douglas-Hamilton I (1990)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rocha2017}{}}%
-\CSLLeftMargin{117. }%
+\CSLLeftMargin{120. }%
 \CSLRightInline{Rocha R, et al. (2017)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-stapp2014}{}}%
-\CSLLeftMargin{118. }%
+\CSLLeftMargin{121. }%
 \CSLRightInline{Stapp P (2014) SGS-LTER long-term monitoring project:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bahamas}{}}%
-\CSLLeftMargin{119. }%
+\CSLLeftMargin{122. }%
 \CSLRightInline{Bahamas marine mammal research organisation
@@
 \leavevmode\vadjust pre{\hypertarget{ref-popacet}{}}%
-\CSLLeftMargin{120. }%
+\CSLLeftMargin{123. }%
 \CSLRightInline{POPA cetacean, seabird, and sea turtle sightings in the

### 2024-03-26 paper/GEB Submission/manuscript_GEB.Rmd
Add table, change text color

@@
   Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevalant than the other across many taxa, biomes, and realms.
-At the study-level, the majority group of studies showed no species or functional trends, but other types of change were manifest for multiple studies, including studies experiencing changes in functional redundancy, increased species and functional richness, or loss of functional richness.
+At the study-level, the majority group of studies showed no species or functional trends, but other types of change were found for multiple studies, including studies experiencing changes in functional redundancy, increased species and functional richness, or loss of functional richness.
 With no one prevailing scenario of change, it will be critical to link change scenarios to ecological context.
@@
 
-`r colorize("Using general trends derived from limited data as a diagnostic for the state of biodiversity or directing conservation is a topic of ongoing debate.", "blue")` Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
+`r colorize("Using general trends derived from limited data as a diagnostic for the state of biodiversity or directing conservation action is a topic of ongoing debate.", "blue")` Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
 Many of these criticisms reflect limitations of ecological data on the whole, `r colorize("leading to a call for additional data not only to fill geographic and temporal gaps, but to flesh out key characteristics of communities", "blue")` [@dornelas2023; @primack2018].
@@
 For some time periods, functional diversity appears to be maintained for substantial portions of geological time [@hedberg2022].
-`r colorize("Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds [@barnagaud2017; @jarzyna2016] and ray-finned fishes, sharks, and rays", "blue")` [@trindade-santos2020].
-`r colorize("Additionally, these taxa have been the subject of substantial conservation investment and policy action, so are not necessarily representative of trends for other taxa or the same taxa in different locations.", "blue")`
+`r colorize("Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds", "blue")` [@barnagaud2017; @jarzyna2016] `r colorize("and ray-finned fishes, sharks, and rays", "blue")` [@trindade-santos2020].
 
@@
 To ensure comparability across taxa in trait type and data quality we therefore focus on mammals and birds.
-`r colorize("We include body mass, foraging and dietary traits", "blue")` that were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].
+`r colorize("We include body mass, dietary, foraging and other behavioral traits", "blue")` that were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].
 An initial assessment of amphibian trends is included in the supplement, but excluded from general trend assessment here due to limited geographic coverage.
@@
 Rather than testing specific hypotheses of change we present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination.
-We present results at three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.", "blue")`
+We present results for three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.", "blue")`
 
@@
 We obtained mammal and bird time series from the BioTIME database, a global repository of high-quality assemblage time series `r colorize("collected from the literature and ongoing monitoring efforts", "blue")`.
-`r colorize("Data is structured such that a study encompass all data collected following consistent sampling protocols and may contain multiple time series from separate locations.
-", "blue")` Time series represent full assemblages rather than populations of single species [@dornelas2018], and all time series include abundance of observed species.
+`r colorize("Data is structured such that a study encompass all data collected following consistent sampling protocols and may contain multiple time series from separate locations.", "blue")` Time series represent full assemblages rather than populations of single species [@dornelas2018], and all time series include abundance of observed species.
 Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual time series following a standardized spatial scale.
 Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018 for details on how sample extents were defined], with the area of each cell set to one standard deviation away from the mean of the single extent locations.
+`r colorize("The resulting cell size for our data was approximately 95 ", "blue")`km^2^.
 All samples from a study within a single cell were considered to be a single time series, and species abundances were combined for all samples.
@@
 We used trait data from the Elton Trait Database, which consists of species-level means for traits that represent species' multifaceted role in the community [@wilman2014].
-Traits include: body mass, diet, active diel period, nocturnality, forest foraging strata, pelagic use.
-Multiple traits (i.e. diet, foraging strata, activity seasonality, active diel period) were broken down into percentage or binary use for each level.
+Traits include: body mass, diet, nocturnality, forest foraging strata, pelagic use.
+With the exception of body mass, traits were broken down into percentage or binary use for each level of the trait type (Table \ref{tab:traitTable}).
+
+```{r traitTable, echo=FALSE}
+tibble(Trait = c(rep("Diet", 10), rep("Foraging Strata", 7), rep("Pelagic Specialist", 2), rep("Nocturnal", 2), "Body Mass"),
+       Category = c("Invertebrate", "Mammals and Birds", "Reptiles", "Fish", "Unknown Vertebrates", "Scavenging", "Fruit", "Nectar", "Seeds", "Other Plant", 
+                    "Below water surface", "water surface", "ground", "understory", "> 2m, below canopy", "canopy", "aerial",
+                    "yes", "no",
+                    "yes", "no",
+                    "-"),
+       `Data Type` = c(rep("percentage consumed", 10), rep("percentage of use", 7), rep("binary", 4), "continous, in grams")) %>%
+  kable(format = "latex", caption = "Description of the traits included in the analysis broken down by categories at data type.") %>%
+  collapse_rows(columns = 1:3) %>%
+  kable_paper()
+```
 
@@
 Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species.
-\` `r colorize("Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability", "blue")`.
+`r colorize("Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability", "blue")`.
 This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
@@
 Some study-level models could not be fit for five studies for at least one metric due to data limitations, but those studies were still included in the general models.
-They represented 12 of 1350 study-level models fit for each metric.
+They represented 13 of 1350 study-level models fit for each metric.
 For further details see the supplement.
@@
 We did find a significant overall decrease in Jaccard similarity, indicating accumulating changes in species composition.
-Non-significant overall `r colorize("temporal", "blue")`trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies `r colorize("was not significantly different from zero", "blue")`\` (Table \ref{tab:resultsTab}).
+Non-significant overall `r colorize("temporal", "blue")` trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies `r colorize("was not significantly different from zero", "blue")`\` (Table \ref{tab:resultsTab}).
 `r colorize("Within-group", "blue")` trends for different taxa, biomes, realms, or protection statuses were also non-significant for richness and functional diversity metrics, with the exception of a significantly increasing trend for functional evenness of marine studies, and a significantly decreasing functional richness slope for mammal studies.
@@
   kable_paper() %>%
-  kable_styling(latex_options= c("scale_down", "hold_position")) %>%
-  save_kable(here::here("figures/tables/results_table.jpeg"))
+  kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
+  #save_kable(here::here("figures/tables/results_table.jpeg"))
 ```
@@
         booktabs = TRUE, align = "c", escape = FALSE) %>%
-  kable_styling("striped")%>%
-  save_kable(here::here("figures/tables/trend_table.jpeg"))
+  kable_styling("striped") #%>%
+  #save_kable(here::here("figures/tables/trend_table.jpeg"))
 
@@
 
-`r colorize("The most stark area of consensus was the lack of general trend aggregating across communities.", "blue")` We did not detect an overall trend in any functional diversity metric, `r colorize("corrected or uncorrected", "blue")`.
-As with previous species-based syntheses, we also found no overall trend in species richness accompanied by `r colorize("increasing dissimilarity", "blue")` through time [@dornelas2018], indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends.
-We found significant turnover for all biomes, realm, and taxa except for marine studies, which stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial [@blowes2019].
-However, previous global estimates are dominated by fish communities which we exclude here and may be driving the overall trend while disguising relative stasis in marine bird and mammal communities.
+`r colorize("The most stark area of consensus was the lack of general trend aggregating across communities.
+We did not detect an overall trend in any functional diversity metric, corrected or uncorrected.
+As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing dissimilarity through time", "blue")` [@dornelas2018]`r colorize(", indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends.
+We found significant turnover for all biomes, realm, and taxa except for marine studies, which stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial", "blue")` [@blowes2019].
+`r colorize("However, previous global estimates are dominated by fish communities which we exclude here and may be driving the overall trend while disguising relative stasis in marine bird and mammal communities.", "blue")`
 
-We found evidence of functional richness loss in mammal studies, with a loss trend significantly different from zero and significantly different from the trend for birds, indicating a potential loss of functional capacity in these communities.
-This result is consistent with previous work linking anthropogenic drivers such as habitat loss, poaching, and human modification and development to loss of functional diversity in mammal communities across the globe [@brodie2021; @cox2022; @li2022].
-However we did not find trends outside the random expectation for species loss, in contrast to expectations for multiple continents based on IUCN risk categories [@brodie2021].
-In general, functional capacity of mammals shows greater declines that bird communities.
+`r colorize("We found evidence of functional richness loss in mammal studies, with a loss trend significantly different from zero and significantly different from the trend for birds, indicating a potential loss of functional capacity in these communities.
+This result is consistent with previous work linking anthropogenic drivers such as habitat loss, poaching, and human modification and development to loss of functional diversity in mammal communities across the globe", "blue")` [@brodie2021; @cox2022; @li2022].
+`r colorize("However we did not find trends outside the random expectation for species loss, in contrast to expectations for multiple continents based on IUCN risk categories", "blue")` [@brodie2021].
+`r colorize("In general, functional capacity of mammals shows greater declines that bird communities.", "blue")`
 
-We also found evidence of increasing evenness for Marine studies.
-The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability [@filstrup2019].
-If one or a few species become largely dominant in a community, function will be the result of only those species rather than spread out across a wealth of species responding independently to environmental change [@hillebrand2008; @hillebrand2009].
-Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities [@bagousse-pinguet2019].
-The majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery, which is also indicated in recent species-based work [@spatz2023].
+`r colorize("We also found evidence of increasing evenness for Marine studies.
+The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability ", "blue")` [@filstrup2019], `r colorize("as the dominance of one or a few species makes function sensitive to the ability of those species to respond to environmental change ", "blue")`[@hillebrand2008; @hillebrand2009].
+`r colorize("Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities ", "blue")`[@bagousse-pinguet2019].
+`r colorize("The majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery, which is also indicated in recent species-based work ", "blue")`[@spatz2023].
 
-Compositional shifts shown by significant trends in *CWM'*s were restricted to dietary traits, with no evidence of a trend in foraging traits or body mass.
+`r colorize("Compositional shifts shown by significant trends in CWM's were restricted to dietary traits, with no evidence of a trend in foraging traits or body mass.
 Increases in fish consumption for bird and temperate studies my be the product of increased availability of easy fish from commercial fishing operations as fish are processed or stocking of inland waterways for sport fishing.
@@
 Interpretation of the increases for temperate studies is less clear, as temperate studies included many bird communities with a variety of nectar consumers.
-Similarly, decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.
+Similarly, decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.", "blue")`
 
@@
 While there are over 150 possible combinations of change direction (or no change) in the 5 metrics, we discuss here the six scenarios that occurred in more than one study: no change, species loss or gain only, loss of functional evenness, species richness loss with species turnover, and increase in species and functional richness accompanied by significant turnover.
-We focus on these scenarios to combat potential spurious results due to multiple testing, as it is unlikely all observations of the same scenario are due to false positives.
-We include a break down of the number of studies in each scenario to illustrate relative number of studies in each group, but emphasize that the absolute number is likely impacted by multiple testing.", "blue")`
+We focus on these scenarios to combat potential spurious results due to multiple testing, as it is unlikely all observations of the same scenario are due to false positives.", "blue")`
 
-The majority group of studies exhibited no trend in any species or functional diversity metric.
+`r colorize("The majority group of studies (17 studies) exhibited no trend in any species or functional diversity metric.
 Contrary to the expectation due to anthropogenic and global change stressors, these communities do not show significant changes over the course of the observation window.
 Studies in this group span the distribution of study durations excluding only the very longest running studies, with the longest no change time series lasting 23 years.
-They also included both bird and mammal studies and only four were located in protected areas, indicating that the lack of trend is not restricted to a specific ecological context or those communities most insulated from human impact.
+They also included both bird and mammal studies and only four were located in protected areas, indicating that the lack of trend is not restricted to a specific ecological context or those communities most insulated from human impact.", "blue")`
 
-The lack of trend could be the result of multiple possible scenarios.
+`r colorize("The lack of trend could be the result of multiple possible scenarios.
 First, these may be communities resisting perturbations or simply not experiencing significant perturbations.
@@
 Alternatively, these may be communities that have experienced or continue to experience significant stress, but lost species or functional diversity outside the observation window.
-This could be true particularly for North American mammal communities where trophic downgrading and megafaunal losses occurred hundreds of years ago [citation].
-Third, these communities may be experiencing directional shifts undetectable by available data.
-For example, species-level trait data does not capture intraspecific shifts in the trait space, which can represent significant changes in the functional `r colorize("space as a whole", "blue")` and impact maintenance of ecological processes.
+This could be true particularly for North American mammal communities where trophic downgrading and megafaunal losses occurred hundreds of years ago ", "blue")`[@estes2011].
+`r colorize("Third, these communities may be experiencing directional shifts undetectable by available data.
+For example, species-level trait data does not capture intraspecific shifts in the trait space, which can represent significant changes in the functional space as a whole and impact maintenance of ecological processes.", "blue")`
 
-Three of the groups fit under the broad umbrella of changes in redundancy.
-By definition, if a community gains or loses species while functional metrics are unchanged, those species represent an increase or decrease in redundancy, so we include communities exhibiting gains or losses in species richness with no change in functional metrics or turnover and species richness loss with significant turnover in this umbrella.
-For communities exhibiting loss of redundancy (declining species richness), ecological processes are likely being maintained, but capacity to respond to future stressors is reduced [@reich2012].
-These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future [@pimiento2020a].
-Conversely, communities exhibiting a increase in redundancy (via an increase in species richness) are becoming better positioned to respond to future stressors, but are not actually expanding their functional capacity as may be assumed by looking at species gains alone [@biggs2020].
+`r colorize("Three of the groups fit under the broad umbrella of changes in redundancy.
+By definition, if a community gains or loses species while functional metrics are unchanged, those species represent an increase or decrease in redundancy, so we include communities exhibiting gains (2 studies) or losses (5 studies) in species richness with no change in functional metrics or turnover and species richness loss with significant turnover (2 studies) in this umbrella.
+For communities exhibiting loss of redundancy (declining species richness), ecological processes are likely being maintained, but capacity to respond to future stressors is reduced ", "blue")`[@reich2012].
+`r colorize("These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future ", "blue")`[@pimiento2020a].
+`r colorize("Conversely, communities exhibiting a increase in redundancy (via an increase in species richness) are becoming better positioned to respond to future stressors, but are not actually expanding their functional capacity as may be assumed by looking at species gains alone", "blue")` [@biggs2020].
 
-The next group of studies exhibit an increase in species richness and functional richness with significant turnover.
+`r colorize("The next group of studies (3 studies) exhibit an increase in species richness and functional richness with significant turnover.
 While these communities are losing some species that are functionally redundant or functional analogs of species additions, they are gaining even more species that expand the functional space.
@@
 Notably, the studies that exhibited species richness increases (in this group with increasing functional richness, as well as communities without functional richness increases) were exclusively from terrestrial, temperate (sometimes temperate/polar) bird communities and occurred in countries where there has been a significant investment in conservation over the last few decades (United States, Canada, Sweden).
-Our results are consistent with other functional work in these regions, showing increases in species richness and functional diversity for North American breeding birds over the last few decades after a period of decline[@jarzyna2016], and loss of common, functionally general species even as rare species are increasing in North America and Europe [@inger2015; @rosenberg2019; @schipper2016].
+Our results are consistent with other functional work in these regions, showing increases in species richness and functional diversity for North American breeding birds over the last few decades after a period of decline", "blue")` [@jarzyna2016] `r colorize(", and loss of common, functionally general species even as rare species are increasing in North America and Europe ", "blue")` [@inger2015; @rosenberg2019; @schipper2016].
 
-The final group exhibits decreasing functional evenness.
-These communities are of particular concern not just because of the role evenness plays in maintaining functionality, but because of increasing evidence evenness is more sensitive to environmental change than species richness [@filstrup2019].
-Formal tests of evenness changes as an early warning signal of more catastrophic shifts in functional diversity will be critical next step for understanding the full implication of evenness loss for ecosystem processes.
+`r colorize("The final group (3 studies) exhibits decreasing functional evenness.
+These communities are of particular concern not just because of the role evenness plays in maintaining functionality, but because of increasing evidence evenness is more sensitive to environmental change than species richness", "blue")` [@filstrup2019].
+`r colorize("Formal tests of evenness changes as an early warning signal of more catastrophic shifts in functional diversity will be critical next step for understanding the full implication of evenness loss for ecosystem processes.", "blue")`
 
-While the majority of studies fell into one of the groups discussed in detail above, 25% of the studies exhibited some different combination of change in metrics, underlining the vast heterogeneity of realized change scenarios.
-We have an indication of the ecological characteristics most likely to lead to some of the change types (e.g.
+`r colorize("While the majority of studies fell into one of the groups discussed in detail above, 25", "blue")`% `r colorize("of the studies exhibited some different combination of change in metrics, underlining the vast heterogeneity of realized change scenarios.", "blue")` `r colorize("We have an indication of the ecological characteristics most likely to lead to some of the change types (e.g.
 temperate bird communities are gaining species and functional richness), but most types of change are exhibited across many different kinds of communities.
-Identifying key factors mediating the kind of biodiversity change a community experiences is critical for identifying areas of conservation concern and extrapolating results to areas where functional trait data may not be available.
-While it is tempting to draw a one to one line between the degree of human impact and loss of functional integrity (in the form of functional redundancy and richness), that relationship is not borne out by our rough metric of human impact (protection status) or previous taxa-specific or single community work, which is largely mixed [@flynn2009; @luck2013; @magioli2021; @tinoco2018].
+", "blue")` `r colorize("Identifying key factors mediating the kind of biodiversity change a community experiences is critical for identifying areas of conservation concern and extrapolating results to areas where functional trait data may not be available.
+While it is tempting to draw a one to one line between the degree of human impact and loss of functional integrity (in the form of functional redundancy and richness), that relationship is not borne out by our rough metric of human impact (protection status) or previous taxa-specific or single community work, which is largely mixed", "blue")` [@flynn2009; @luck2013; @magioli2021; @tinoco2018].
+
 <!-- huge differences between different taxa and the kind of disturbance/land use change -->
@@
 
-`r colorize("The addition of functional trait illuminated a few key areas where a synthesis approach is incongruous with system-specific studies or expectations based on anthropegenic impacts.", "blue")` For example, we found no evidence of general trends in *CWM* body mass or for studies aggregated by their characteristics despite predicitions that mean body mass is decreasing as a result of climate change impacts and megafaunal loss [@sheridan2011].
-This phenomenon has also been well documented empirically and experimentally in multiple taxa [@caruso2014; @forster2012; @huss2019; @tseng2018].
-Still, climate change impacts are predicted to be most extreme for ectotherms and impact intraspecific traits first before species are lost, neither of which are represented by our data.
-Further, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing [@fritz2009].
-Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened [@ripple2015].
+`r colorize("The addition of functional trait data illuminates a few key areas where a synthesis approach is incongruous with system-specific studies or expectations based on anthropegenic impacts.
+For example, we found no evidence of general trends in the *CWM* for body mass or for studies aggregated by their characteristics despite predictions that mean body mass is decreasing as a result of climate change impacts and megafaunal loss ", "blue")`[@sheridan2011] `r colorize(", and downgrading documented empirically and experimentally in multiple taxa ", "blue")`[@caruso2014; @forster2012; @huss2019; @tseng2018].
+`r colorize("Still, climate change impacts are predicted to be most extreme for ectotherms and impact intraspecific traits first before species are lost, neither of which are represented by our data.
+Further, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].
+`r colorize("Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened", "blue")` [@ripple2015].
 
-Results for protected areas were also surprising given the hope that protection insulates communities from functional degradation.
+`r colorize("Results for protected areas were also surprising given the assumption that protection insulates communities from functional degradation.
 The only difference we found between protected and unprotected areas was significant turnover occurring outside protected areas and no significant turnover inside protected areas.
-While this is an indicator things are generally more static within protect areas, protection has no bearing on whether not there is change in the functional space and the nature of that change.
-Looking at individual studies, there is a mix of both losses and gains in almost all functional diversity and composition metrics across communities found in protected areas.<!-- turnover outside protected areas, not inside ( though might be driven by the fact that there's generally less turnover in terrestrial and temperate studies) [@blowes2019] -->
+Rather than the result of protection itself, this is likely due to the fact that the majority of studies were form terrestrial, temperate studies where turnover is known to be lower ", "blue")`[@blowes2019].
+`r colorize("While things may generally be more static within protect areas, the functional dimensions of protected communities fared no better or worse on average than their non-protected counter parts.
+Looking at individual studies, there is a mix of both losses and gains in almost all functional diversity and composition metrics across communities found in protected areas.", "blue")`
 
@@
 
-While we found no over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
+`r colorize("While we found no over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
 In fact, study-level trends indicate quite the opposite, that functional shifts with negative or yet unknown implications for ecosystem processes may be going undetected by common species-based approaches.
-For example, loss of evenness in communities with constant species richness may be an first sign of a community being impacted by environmental change, with negative implications for stability and function.
+For example, loss of evenness in communities with constant species richness may be an first sign of a community being impacted by environmental change, with negative implications for stability and function.", "blue")`
 
-One of the biggest threats to biodiversity is the wholesale conversion of natural areas to urban or human-dominated landscapes [@tilman2017].
-Typical long term monitoring such as those included in our study stops before this conversion occurs, leaving the resultant precipitous declines in biodiversity unrecorded [@cardinale2014].
-This is a known issue with the culture of long term monitoring, and our results should not be removed from that context.
-Rather, this studies captures communities that are likely experiencing a degree of human intervention, but are still largely nature-dominated.
+`r colorize("One of the biggest threats to biodiversity is the wholesale conversion of natural areas to urban or human-dominated landscapes ", "blue")`[@tilman2017].
+`r colorize("Typical long term monitoring such as those included in our study stops before this conversion occurs, leaving the resultant precipitous declines in biodiversity unrecorded", "blue")` [@cardinale2014].
+`r colorize("This is a known issue with the culture of long term monitoring, and our results should not be removed from that context.
+Rather, this studies captures communities that are likely experiencing a degree of human intervention, but are still largely nature-dominated.", "blue")`
 
@@
 The addition of functional trait data to the biodiversity change conversation is yet another illustration of how critical context is for understanding ecological patterns [@catford2022; @spake2023].
-Here we assessed how key community characteristics such as biome, realm, taxa, and protection status may partition variation in the functional trends exhibited and found that none of those characteristics seem to make the picture of change clearer.
+`r colorize("Here we assessed how key community characteristics such as biome, realm, taxa, and protection status may partition variation in the functional trends exhibited and found that, on the whole, these broad classifications did not seem to make the picture of change clearer.
 Further work linking specific ecological contexts to types of functional change will be critical for identifying new areas of concern, especially when those communities may not otherwise be showing significant changes in species richness.
-The identification of contexts where functional change is most likely occurring will be a significant asset in directing the future collection of trait data, a main barrier for taking a functional approach to biodiversity change.
+The identification of contexts where functional change is most likely occurring will be a significant asset in directing the future collection of trait data, the main barrier for taking a functional approach to biodiversity change.", "blue")`
 

### 2024-03-26 paper/GEB Submission/manuscript_GEB.tex
Add table, change text color

@@
 
-\textcolor{blue}{Using general trends derived from limited data as a diagnostic for the state of biodiversity or directing conservation is a topic of ongoing debate.}
+\textcolor{blue}{Using general trends derived from limited data as a diagnostic for the state of biodiversity or directing conservation action is a topic of ongoing debate.}
 Global analyses have been heavily criticized for geographic biases, lack
@@
 be maintained for substantial portions of geological time (33).
-\textcolor{blue}{Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds [@barnagaud2017; @jarzyna2016] and ray-finned fishes, sharks, and rays}
-(34).
-\textcolor{blue}{Additionally, these taxa have been the subject of substantial conservation investment and policy action, so are not necessarily representative of trends for other taxa or the same taxa in different locations.}
+\textcolor{blue}{Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds}
+(34, 35) \textcolor{blue}{and ray-finned fishes, sharks, and rays} (36).
 
@@
 is of equal interest, trait data for those taxa raise additional
-challenges such as limited and biased species coverage (35), a lack of
+challenges such as limited and biased species coverage (37), a lack of
 accepted species-level means, and differences in the types of traits
@@
 quality we therefore focus on mammals and birds.
-\textcolor{blue}{We include body mass, foraging and dietary traits} that
-were intentionally selected to be representative of a species' Eltonian
-niche, thereby summarizing the functional role they play in the
-community (36). An initial assessment of amphibian trends is included in
+\textcolor{blue}{We include body mass, dietary, foraging and other behavioral traits}
+that were intentionally selected to be representative of a species'
+Eltonian niche, thereby summarizing the functional role they play in the
+community (38). An initial assessment of amphibian trends is included in
 the supplement, but excluded from general trend assessment here due to
@@
 Rather than testing specific hypotheses of change we present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination.
-We present results at three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.}
+We present results for three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.}
 
@@
 \textcolor{blue}{collected from the literature and ongoing monitoring efforts}.
-\textcolor{blue}{Data is structured such that a study encompass all data collected following consistent sampling protocols and may contain multiple time series from separate locations. }
+\textcolor{blue}{Data is structured such that a study encompass all data collected following consistent sampling protocols and may contain multiple time series from separate locations.}
 Time series represent full assemblages rather than populations of single
 species (31), and all time series include abundance of observed species.
-Following best practices for the database (37), studies with multiple
+Following best practices for the database (39), studies with multiple
 sample locations were split into individual time series following a
@@
 the area of each cell set to one standard deviation away from the mean
-of the single extent locations. All samples from a study within a single
-cell were considered to be a single time series, and species abundances
-were combined for all samples.
+of the single extent locations.
+\textcolor{blue}{The resulting cell size for our data was approximately 95 }km\textsuperscript{2}.
+All samples from a study within a single cell were considered to be a
+single time series, and species abundances were combined for all
+samples.
 
@@
 species-level means for traits that represent species' multifaceted role
-in the community (36). Traits include: body mass, diet, active diel
-period, nocturnality, forest foraging strata, pelagic use. Multiple
-traits (i.e.~diet, foraging strata, activity seasonality, active diel
-period) were broken down into percentage or binary use for each level.
+in the community (38). Traits include: body mass, diet, nocturnality,
+forest foraging strata, pelagic use. With the exception of body mass,
+traits were broken down into percentage or binary use for each level of
+the trait type (Table \ref{tab:traitTable}).
+
+\begin{table}
+
+\caption{\label{tab:traitTable}Description of the traits included in the analysis broken down by categories at data type.}
+\centering
+\begin{tabular}[t]{l|l|l}
+\hline
+Trait & Category & Data Type\\
+\hline
+ & Invertebrate & \\
+\cline{2-2}
+ & Mammals and Birds & \\
+\cline{2-2}
+ & Reptiles & \\
+\cline{2-2}
+ & Fish & \\
+\cline{2-2}
+ & Unknown Vertebrates & \\
+\cline{2-2}
+ & Scavenging & \\
+\cline{2-2}
+ & Fruit & \\
+\cline{2-2}
+ & Nectar & \\
+\cline{2-2}
+ & Seeds & \\
+\cline{2-2}
+\multirow{-10}{*}{\raggedright\arraybackslash Diet} & Other Plant & \multirow{-10}{*}{\raggedright\arraybackslash percentage consumed}\\
+\cline{1-3}
+ & Below water surface & \\
+\cline{2-2}
+ & water surface & \\
+\cline{2-2}
+ & ground & \\
+\cline{2-2}
+ & understory & \\
+\cline{2-2}
+ & > 2m, below canopy & \\
+\cline{2-2}
+ & canopy & \\
+\cline{2-2}
+\multirow{-7}{*}{\raggedright\arraybackslash Foraging Strata} & aerial & \multirow{-7}{*}{\raggedright\arraybackslash percentage of use}\\
+\cline{1-3}
+ & yes & \\
+\cline{2-2}
+\multirow{-2}{*}{\raggedright\arraybackslash Pelagic Specialist} & no & \\
+\cline{1-2}
+ & yes & \\
+\cline{2-2}
+\multirow{-2}{*}{\raggedright\arraybackslash Nocturnal} & no & \multirow{-4}{*}{\raggedright\arraybackslash binary}\\
+\cline{1-3}
+Body Mass & - & continous, in grams\\
+\hline
+\end{tabular}
+\end{table}
 
@@
 \url{https://doi.org/10.5066/F7KH0KBK}), obtained through the
-\texttt{taxadb} R package (38, 39). If more than one species in the
+\texttt{taxadb} R package (40, 41). If more than one species in the
 assemblage data resolved to the same identifier, observations were
@@
 Functional diversity metrics were calculated using the \emph{dbFD}
-function from the \emph{FD} R package (40). Here we report functional
+function from the \emph{FD} R package (42). Here we report functional
 richness (\emph{FRic}), functional evenness (\emph{FEve}), and
 functional divergence (\emph{FDiv}) which together describe three
-complementary characteristics of the functional space (41, 42).
+complementary characteristics of the functional space (43, 44).
 \emph{FRic} assesses the volume of the trait space occupied by species
@@
 traits were z-score scaled to give each trait equal weight in the trait
-space (43, 44).
+space (45, 46).
 \textcolor{blue}{Before calculating diversity metrics, dbFD reduces the dimensionality of the trait space by performing PCoA. We limited the number of included PCoA axes to}
@@
 traits. This restriction allows for enough axes to capture the trait
-space while maintaining computational feasibility (45). Metrics
+space while maintaining computational feasibility (47). Metrics
 incorporated weighting based on species abundance.
@@
 functional diversity metrics (\emph{FRic}, \emph{FEve}, \emph{FDiv})
-from null estimates (46). Null model corrections allow us to assess the
+from null estimates (48). Null model corrections allow us to assess the
 degree to which the observed functional diversity metric deviates from
@@
 species from the species pool for each year and randomly assigning
-observed abundances to species. `
+observed abundances to species.
 \textcolor{blue}{Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability}.
@@
 limitations, but those studies were still included in the general
-models. They represented 12 of 1350 study-level models fit for each
+models. They represented 13 of 1350 study-level models fit for each
 metric. For further details see the supplement. Where appropriate,
@@
 structure. We estimated within-level slopes and calculated between-level
-contrasts using the \emph{emmeans} package (v1.8.2, 47).
+contrasts using the \emph{emmeans} package (v1.8.2, 49).
 \textcolor{blue}{For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore refrained from interpreting results for levels where there were fewer than three studies.}
@@
 (v3.1-3) package with a significance level of \(\alpha = 0.05\)
-(48--50).
+(50--52).
 
@@
 Jaccard similarity, indicating accumulating changes in species
-composition. Non-significant overall \textcolor{blue}{temporal}trends
+composition. Non-significant overall \textcolor{blue}{temporal} trends
 indicate that although some studies experience increasing or decreasing
@@
 \end{figure}
+
+\begin{table}[!h]
+
+\caption{\label{tab:resultsTab}Model estimates and statistics for general trend models for species richness, Jaccard similarity, and corrected functional diversity metrics.
+Additional model estimates, including CWM models, can be found in the supplement.}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{l|l|>{\raggedright\arraybackslash}p{2cm}|l|r|r|l}
+\hline
+metric & effect & grouping & term & estimate & std.error & p.value\\
+\hline
+ &  &  & Intercept & -0.25 & 0.11 & 0.03\\
+\cline{4-7}
+ & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & 0.00 & 0.04 & 0.96\\
+\cline{2-7}
+ &  &  & SD Intercept & 0.60 &  & \\
+\cline{4-7}
+ &  &  & SD Year & 0.11 &  & \\
+\cline{4-7}
+ &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.16 &  & \\
+\cline{3-7}
+ &  &  & SD Intercept & 0.60 &  & \\
+\cline{4-7}
+ &  &  & SD Year & 0.23 &  & \\
+\cline{4-7}
+\multirow{-8}{*}{\raggedright\arraybackslash SES\_FDiv} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & -0.01 &  & \\
+\cline{1-7}
+ &  &  & Intercept & 0.10 & 0.17 & 0.56\\
+\cline{4-7}
+ & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & -0.01 & 0.02 & 0.62\\
+\cline{2-7}
+ &  &  & SD Intercept & 1.08 &  & \\
+\cline{4-7}
+ &  &  & SD Year & 0.04 &  & \\
+\cline{4-7}
+ &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.53 &  & \\
+\cline{3-7}
+ &  &  & SD Intercept & 0.40 &  & \\
+\cline{4-7}
+ &  &  & SD Year & 0.17 &  & \\
+\cline{4-7}
+\multirow{-8}{*}{\raggedright\arraybackslash SES\_FEve} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & -0.21 &  & \\
+\cline{1-7}
+ &  &  & Intercept & -0.26 & 0.07 & <0.001\\
+\cline{4-7}
+ & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & 0.04 & 0.04 & 0.35\\
+\cline{2-7}
+ &  &  & SD Intercept & 0.29 &  & \\
+\cline{4-7}
+ &  &  & SD Year & 0.11 &  & \\
+\cline{4-7}
+ &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.27 &  & \\
+\cline{3-7}
+ &  &  & SD Intercept & 0.54 &  & \\
+\cline{4-7}
+ &  &  & SD Year & 0.18 &  & \\
+\cline{4-7}
+\multirow{-8}{*}{\raggedright\arraybackslash SES\_FRic} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.06 &  & \\
+\cline{1-7}
+ &  &  & Intercept & 0.46 & 0.02 & <0.001\\
+\cline{4-7}
+ & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & -0.03 & 0.00 & <0.001\\
+\cline{2-7}
+ &  &  & SD Intercept & 0.09 &  & \\
+\cline{4-7}
+ &  &  & SD Year & 0.02 &  & \\
+\cline{4-7}
+ &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.36 &  & \\
+\cline{3-7}
+ &  &  & SD Intercept & 0.08 &  & \\
+\cline{4-7}
+ &  &  & SD Year & 0.02 &  & \\
+\cline{4-7}
+\multirow{-8}{*}{\raggedright\arraybackslash log(Jaccard + 1)} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.05 &  & \\
+\cline{1-7}
+ &  &  & Intercept & 2.40 & 0.10 & <0.001\\
+\cline{4-7}
+ & \multirow{-2}{*}{\raggedright\arraybackslash fixed} & \multirow{-2}{2cm}{\raggedright\arraybackslash } & Year & -0.06 & 0.05 & 0.17\\
+\cline{2-7}
+ &  &  & SD Intercept & 0.64 &  & \\
+\cline{4-7}
+ &  &  & SD Year & 0.27 &  & \\
+\cline{4-7}
+ &  & \multirow{-3}{2cm}{\raggedright\arraybackslash study} & Corr(Intercept, Year) & -0.70 &  & \\
+\cline{3-7}
+ &  &  & SD Intercept & 0.21 &  & \\
+\cline{4-7}
+ &  &  & SD Year & 0.09 &  & \\
+\cline{4-7}
+\multirow{-8}{*}{\raggedright\arraybackslash log(Species Richness)} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.34 &  & \\
+\hline
+\end{tabular}}
+\end{table}
 
@@
 
+\begin{table}
+
+\caption{\label{tab:trendTab}Number of studies that experienced a significant trend in each calculated metric out of 50 total studies.}
+\centering
+\begin{tabular}[t]{ccccccccc}
+\toprule
+ & S & Jaccard Similarity & FRic & FEve & FDiv & SES FRic & SES FEve & SES FDiv\\
+\midrule
++ & 7 & 0 & 7 & 5 & 3 & 4 & 4 & 3\\
+- & 9 & 11 & 8 & 6 & 5 & 2 & 6 & 2\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 Study-level slopes were significantly related to start year of the time
@@
 
-\textcolor{blue}{The most stark area of consensus was the lack of general trend aggregating across communities.}
-We did not detect an overall trend in any functional diversity metric,
-\textcolor{blue}{corrected or uncorrected}. As with previous
-species-based syntheses, we also found no overall trend in species
-richness accompanied by \textcolor{blue}{increasing dissimilarity}
-through time (31), indicating that non-significant trends in functional
-metrics are consistent with similar well-documented species derived
-trends. We found significant turnover for all biomes, realm, and taxa
-except for marine studies, which stands in contrast to other global
-estimates of biodiversity change that found higher turnover in marine
-systems than terrestrial (37). However, previous global estimates are
-dominated by fish communities which we exclude here and may be driving
-the overall trend while disguising relative stasis in marine bird and
-mammal communities.
-
-We found evidence of functional richness loss in mammal studies, with a
-loss trend significantly different from zero and significantly different
-from the trend for birds, indicating a potential loss of functional
-capacity in these communities. This result is consistent with previous
-work linking anthropogenic drivers such as habitat loss, poaching, and
-human modification and development to loss of functional diversity in
-mammal communities across the globe (51--53). However we did not find
-trends outside the random expectation for species loss, in contrast to
-expectations for multiple continents based on IUCN risk categories (51).
-In general, functional capacity of mammals shows greater declines that
-bird communities.
-
-We also found evidence of increasing evenness for Marine studies. The
-empirical link between changes in evenness and ecosystem process is the
-most poorly studied of the metrics, however theory indicates that
-increased evenness improves function and stability (54). If one or a few
-species become largely dominant in a community, function will be the
-result of only those species rather than spread out across a wealth of
-species responding independently to environmental change (42, 55).
-Further, there is some evidence that maintaining evenness is important
-for supporting multifunctionality in communities (56). The majority of
-marine studies are of seabird communities (7 of 9), so the general trend
-may be an early indicator of pay off from investments in sea bird
-recovery, which is also indicated in recent species-based work (57).
-
-Compositional shifts shown by significant trends in \emph{CWM'}s were
-restricted to dietary traits, with no evidence of a trend in foraging
-traits or body mass. Increases in fish consumption for bird and
-temperate studies my be the product of increased availability of easy
-fish from commercial fishing operations as fish are processed or
-stocking of inland waterways for sport fishing. However, fisheries also
-have negative impacts on waterbirds through competition for fish or bird
-mortality due to by catch. Decreases in nectar consumption for tropical
-studies is likely due declines in bat species, as 4 of the 5 tropical
-studies are for mammal and bats were the only nectar consuming mammals
-sampled in our data. Interpretation of the increases for temperate
-studies is less clear, as temperate studies included many bird
-communities with a variety of nectar consumers. Similarly, decreasing
-trends for vertebrate consumption in marine and tropical studies are not
-obviously consistent with known changes in those contexts and warrant
-further examination.
+\textcolor{blue}{The most stark area of consensus was the lack of general trend aggregating across communities.
+We did not detect an overall trend in any functional diversity metric, corrected or uncorrected.
+As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing dissimilarity through time}
+(31)\textcolor{blue}{, indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends.
+We found significant turnover for all biomes, realm, and taxa except for marine studies, which stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial}
+(39).
+\textcolor{blue}{However, previous global estimates are dominated by fish communities which we exclude here and may be driving the overall trend while disguising relative stasis in marine bird and mammal communities.}
+
+\textcolor{blue}{We found evidence of functional richness loss in mammal studies, with a loss trend significantly different from zero and significantly different from the trend for birds, indicating a potential loss of functional capacity in these communities.
+This result is consistent with previous work linking anthropogenic drivers such as habitat loss, poaching, and human modification and development to loss of functional diversity in mammal communities across the globe}
+(53--55).
+\textcolor{blue}{However we did not find trends outside the random expectation for species loss, in contrast to expectations for multiple continents based on IUCN risk categories}
+(53).
+\textcolor{blue}{In general, functional capacity of mammals shows greater declines that bird communities.}
+
+\textcolor{blue}{We also found evidence of increasing evenness for Marine studies.
+The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability }
+(56),
+\textcolor{blue}{as the dominance of one or a few species makes function sensitive to the ability of those species to respond to environmental change }(44,
+57).
+\textcolor{blue}{Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities }(58).
+\textcolor{blue}{The majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery, which is also indicated in recent species-based work }(59).
+
+\textcolor{blue}{Compositional shifts shown by significant trends in CWM's were restricted to dietary traits, with no evidence of a trend in foraging traits or body mass.
+Increases in fish consumption for bird and temperate studies my be the product of increased availability of easy fish from commercial fishing operations as fish are processed or stocking of inland waterways for sport fishing.
+However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch.
+Decreases in nectar consumption for tropical studies is likely due declines in bat species, as 4 of the 5 tropical studies are for mammal and bats were the only nectar consuming mammals sampled in our data.
+Interpretation of the increases for temperate studies is less clear, as temperate studies included many bird communities with a variety of nectar consumers.
+Similarly, decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.}
 
@@
 While there are over 150 possible combinations of change direction (or no change) in the 5 metrics, we discuss here the six scenarios that occurred in more than one study: no change, species loss or gain only, loss of functional evenness, species richness loss with species turnover, and increase in species and functional richness accompanied by significant turnover.
-We focus on these scenarios to combat potential spurious results due to multiple testing, as it is unlikely all observations of the same scenario are due to false positives.
-We include a break down of the number of studies in each scenario to illustrate relative number of studies in each group, but emphasize that the absolute number is likely impacted by multiple testing.}
-
-The majority group of studies exhibited no trend in any species or
-functional diversity metric. Contrary to the expectation due to
-anthropogenic and global change stressors, these communities do not show
-significant changes over the course of the observation window. Studies
-in this group span the distribution of study durations excluding only
-the very longest running studies, with the longest no change time series
-lasting 23 years. They also included both bird and mammal studies and
-only four were located in protected areas, indicating that the lack of
-trend is not restricted to a specific ecological context or those
-communities most insulated from human impact.
-
-The lack of trend could be the result of multiple possible scenarios.
-First, these may be communities resisting perturbations or simply not
-experiencing significant perturbations. Given the studies in this group
-come from all possible taxa, realms, biomes, and protection statuses,
-evidence points to communities resisting perturbation, offering some
-hope that there are areas of the globe where communities are fairing
-well for now. Alternatively, these may be communities that have
-experienced or continue to experience significant stress, but lost
-species or functional diversity outside the observation window. This
-could be true particularly for North American mammal communities where
-trophic downgrading and megafaunal losses occurred hundreds of years ago
-{[}citation{]}. Third, these communities may be experiencing directional
-shifts undetectable by available data. For example, species-level trait
-data does not capture intraspecific shifts in the trait space, which can
-represent significant changes in the functional
-\textcolor{blue}{space as a whole} and impact maintenance of ecological
-processes.
-
-Three of the groups fit under the broad umbrella of changes in
-redundancy. By definition, if a community gains or loses species while
-functional metrics are unchanged, those species represent an increase or
-decrease in redundancy, so we include communities exhibiting gains or
-losses in species richness with no change in functional metrics or
-turnover and species richness loss with significant turnover in this
-umbrella. For communities exhibiting loss of redundancy (declining
-species richness), ecological processes are likely being maintained, but
-capacity to respond to future stressors is reduced (58). These
-communities are actually fairing better than expected looking at
-species-based metrics alone, but are also in a precarious position for
-maintaining ecological function into the future (59). Conversely,
-communities exhibiting a increase in redundancy (via an increase in
-species richness) are becoming better positioned to respond to future
-stressors, but are not actually expanding their functional capacity as
-may be assumed by looking at species gains alone (60).
-
-The next group of studies exhibit an increase in species richness and
-functional richness with significant turnover. While these communities
-are losing some species that are functionally redundant or functional
-analogs of species additions, they are gaining even more species that
-expand the functional space. These communities have a hopeful
-trajectory, as they are improving their functional capacity and
-potentially their functional redundancy, likely leading to robust future
-ecological function. Notably, the studies that exhibited species
-richness increases (in this group with increasing functional richness,
-as well as communities without functional richness increases) were
-exclusively from terrestrial, temperate (sometimes temperate/polar) bird
-communities and occurred in countries where there has been a significant
-investment in conservation over the last few decades (United States,
-Canada, Sweden). Our results are consistent with other functional work
-in these regions, showing increases in species richness and functional
-diversity for North American breeding birds over the last few decades
-after a period of decline(61), and loss of common, functionally general
-species even as rare species are increasing in North America and Europe
-(62--64).
-
-The final group exhibits decreasing functional evenness. These
-communities are of particular concern not just because of the role
-evenness plays in maintaining functionality, but because of increasing
-evidence evenness is more sensitive to environmental change than species
-richness (54). Formal tests of evenness changes as an early warning
-signal of more catastrophic shifts in functional diversity will be
-critical next step for understanding the full implication of evenness
-loss for ecosystem processes.
-
-While the majority of studies fell into one of the groups discussed in
-detail above, 25\% of the studies exhibited some different combination
-of change in metrics, underlining the vast heterogeneity of realized
-change scenarios. We have an indication of the ecological
-characteristics most likely to lead to some of the change types
-(e.g.~temperate bird communities are gaining species and functional
-richness), but most types of change are exhibited across many different
-kinds of communities. Identifying key factors mediating the kind of
-biodiversity change a community experiences is critical for identifying
-areas of conservation concern and extrapolating results to areas where
-functional trait data may not be available. While it is tempting to draw
-a one to one line between the degree of human impact and loss of
-functional integrity (in the form of functional redundancy and
-richness), that relationship is not borne out by our rough metric of
-human impact (protection status) or previous taxa-specific or single
-community work, which is largely mixed (26, 65--67).
+We focus on these scenarios to combat potential spurious results due to multiple testing, as it is unlikely all observations of the same scenario are due to false positives.}
+
+\textcolor{blue}{The majority group of studies (17 studies) exhibited no trend in any species or functional diversity metric.
+Contrary to the expectation due to anthropogenic and global change stressors, these communities do not show significant changes over the course of the observation window.
+Studies in this group span the distribution of study durations excluding only the very longest running studies, with the longest no change time series lasting 23 years.
+They also included both bird and mammal studies and only four were located in protected areas, indicating that the lack of trend is not restricted to a specific ecological context or those communities most insulated from human impact.}
+
+\textcolor{blue}{The lack of trend could be the result of multiple possible scenarios.
+First, these may be communities resisting perturbations or simply not experiencing significant perturbations.
+Given the studies in this group come from all possible taxa, realms, biomes, and protection statuses, evidence points to communities resisting perturbation, offering some hope that there are areas of the globe where communities are fairing well for now.
+Alternatively, these may be communities that have experienced or continue to experience significant stress, but lost species or functional diversity outside the observation window.
+This could be true particularly for North American mammal communities where trophic downgrading and megafaunal losses occurred hundreds of years ago }(60).
+\textcolor{blue}{Third, these communities may be experiencing directional shifts undetectable by available data.
+For example, species-level trait data does not capture intraspecific shifts in the trait space, which can represent significant changes in the functional space as a whole and impact maintenance of ecological processes.}
+
+\textcolor{blue}{Three of the groups fit under the broad umbrella of changes in redundancy.
+By definition, if a community gains or loses species while functional metrics are unchanged, those species represent an increase or decrease in redundancy, so we include communities exhibiting gains (2 studies) or losses (5 studies) in species richness with no change in functional metrics or turnover and species richness loss with significant turnover (2 studies) in this umbrella.
+For communities exhibiting loss of redundancy (declining species richness), ecological processes are likely being maintained, but capacity to respond to future stressors is reduced }(61).
+\textcolor{blue}{These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future }(62).
+\textcolor{blue}{Conversely, communities exhibiting a increase in redundancy (via an increase in species richness) are becoming better positioned to respond to future stressors, but are not actually expanding their functional capacity as may be assumed by looking at species gains alone}
+(63).
+
+\textcolor{blue}{The next group of studies (3 studies) exhibit an increase in species richness and functional richness with significant turnover.
+While these communities are losing some species that are functionally redundant or functional analogs of species additions, they are gaining even more species that expand the functional space.
+These communities have a hopeful trajectory, as they are improving their functional capacity and potentially their functional redundancy, likely leading to robust future ecological function.
+Notably, the studies that exhibited species richness increases (in this group with increasing functional richness, as well as communities without functional richness increases) were exclusively from terrestrial, temperate (sometimes temperate/polar) bird communities and occurred in countries where there has been a significant investment in conservation over the last few decades (United States, Canada, Sweden).
+Our results are consistent with other functional work in these regions, showing increases in species richness and functional diversity for North American breeding birds over the last few decades after a period of decline}
+(35)
+\textcolor{blue}{, and loss of common, functionally general species even as rare species are increasing in North America and Europe }
+(64--66).
+
+\textcolor{blue}{The final group (3 studies) exhibits decreasing functional evenness.
+These communities are of particular concern not just because of the role evenness plays in maintaining functionality, but because of increasing evidence evenness is more sensitive to environmental change than species richness}
+(56).
+\textcolor{blue}{Formal tests of evenness changes as an early warning signal of more catastrophic shifts in functional diversity will be critical next step for understanding the full implication of evenness loss for ecosystem processes.}
+
+\textcolor{blue}{While the majority of studies fell into one of the groups discussed in detail above, 25}\%
+\textcolor{blue}{of the studies exhibited some different combination of change in metrics, underlining the vast heterogeneity of realized change scenarios.}
+\textcolor{blue}{We have an indication of the ecological characteristics most likely to lead to some of the change types (e.g.
+temperate bird communities are gaining species and functional richness), but most types of change are exhibited across many different kinds of communities.
+}
+\textcolor{blue}{Identifying key factors mediating the kind of biodiversity change a community experiences is critical for identifying areas of conservation concern and extrapolating results to areas where functional trait data may not be available.
+While it is tempting to draw a one to one line between the degree of human impact and loss of functional integrity (in the form of functional redundancy and richness), that relationship is not borne out by our rough metric of human impact (protection status) or previous taxa-specific or single community work, which is largely mixed}
+(26, 67--69).
 
@@
 
-\textcolor{blue}{The addition of functional trait illuminated a few key areas where a synthesis approach is incongruous with system-specific studies or expectations based on anthropegenic impacts.}
-For example, we found no evidence of general trends in \emph{CWM} body
-mass or for studies aggregated by their characteristics despite
-predicitions that mean body mass is decreasing as a result of climate
-change impacts and megafaunal loss (68). This phenomenon has also been
-well documented empirically and experimentally in multiple taxa
-(69--72). Still, climate change impacts are predicted to be most extreme
-for ectotherms and impact intraspecific traits first before species are
-lost, neither of which are represented by our data. Further, many of the
-studies in our dataset draw from areas that may have experienced
-significant loss of large-bodied species before the observation window,
-with contemporary loss rates slowing (73). Trends could be significantly
-different for the same time periods in regions of sub-Saharan Africa,
-for example, which has poor representation in our dataset but where
-megafauna exist on the landscape and are increasingly threatened (74).
-
-Results for protected areas were also surprising given the hope that
-protection insulates communities from functional degradation. The only
-difference we found between protected and unprotected areas was
-significant turnover occurring outside protected areas and no
-significant turnover inside protected areas. While this is an indicator
-things are generally more static within protect areas, protection has no
-bearing on whether not there is change in the functional space and the
-nature of that change. Looking at individual studies, there is a mix of
-both losses and gains in almost all functional diversity and composition
-metrics across communities found in protected areas.
+\textcolor{blue}{The addition of functional trait data illuminates a few key areas where a synthesis approach is incongruous with system-specific studies or expectations based on anthropegenic impacts.
+For example, we found no evidence of general trends in the *CWM* for body mass or for studies aggregated by their characteristics despite predictions that mean body mass is decreasing as a result of climate change impacts and megafaunal loss }(70)
+\textcolor{blue}{, and downgrading documented empirically and experimentally in multiple taxa }(71--74).
+\textcolor{blue}{Still, climate change impacts are predicted to be most extreme for ectotherms and impact intraspecific traits first before species are lost, neither of which are represented by our data.
+Further, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing}
+(75).
+\textcolor{blue}{Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened}
+(76).
+
+\textcolor{blue}{Results for protected areas were also surprising given the assumption that protection insulates communities from functional degradation.
+The only difference we found between protected and unprotected areas was significant turnover occurring outside protected areas and no significant turnover inside protected areas.
+Rather than the result of protection itself, this is likely due to the fact that the majority of studies were form terrestrial, temperate studies where turnover is known to be lower }(39).
+\textcolor{blue}{While things may generally be more static within protect areas, the functional dimensions of protected communities fared no better or worse on average than their non-protected counter parts.
+Looking at individual studies, there is a mix of both losses and gains in almost all functional diversity and composition metrics across communities found in protected areas.}
 
@@
 
-While we found no over all trends in functional metrics, our results
-should not be interpreted as an indication that the ongoing biodiversity
-crisis is less severe than previously described, or that there is no
-concern for functional change as a result of anthropogenic impact. In
-fact, study-level trends indicate quite the opposite, that functional
-shifts with negative or yet unknown implications for ecosystem processes
-may be going undetected by common species-based approaches. For example,
-loss of evenness in communities with constant species richness may be an
-first sign of a community being impacted by environmental change, with
-negative implications for stability and function.
-
-One of the biggest threats to biodiversity is the wholesale conversion
-of natural areas to urban or human-dominated landscapes (75). Typical
-long term monitoring such as those included in our study stops before
-this conversion occurs, leaving the resultant precipitous declines in
-biodiversity unrecorded (10). This is a known issue with the culture of
-long term monitoring, and our results should not be removed from that
-context. Rather, this studies captures communities that are likely
-experiencing a degree of human intervention, but are still largely
-nature-dominated.
+\textcolor{blue}{While we found no over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
+In fact, study-level trends indicate quite the opposite, that functional shifts with negative or yet unknown implications for ecosystem processes may be going undetected by common species-based approaches.
+For example, loss of evenness in communities with constant species richness may be an first sign of a community being impacted by environmental change, with negative implications for stability and function.}
+
+\textcolor{blue}{One of the biggest threats to biodiversity is the wholesale conversion of natural areas to urban or human-dominated landscapes }(77).
+\textcolor{blue}{Typical long term monitoring such as those included in our study stops before this conversion occurs, leaving the resultant precipitous declines in biodiversity unrecorded}
+(10).
+\textcolor{blue}{This is a known issue with the culture of long term monitoring, and our results should not be removed from that context.
+Rather, this studies captures communities that are likely experiencing a degree of human intervention, but are still largely nature-dominated.}
 
@@
 conversation is yet another illustration of how critical context is for
-understanding ecological patterns (76, 77). Here we assessed how key
-community characteristics such as biome, realm, taxa, and protection
-status may partition variation in the functional trends exhibited and
-found that none of those characteristics seem to make the picture of
-change clearer. Further work linking specific ecological contexts to
-types of functional change will be critical for identifying new areas of
-concern, especially when those communities may not otherwise be showing
-significant changes in species richness. The identification of contexts
-where functional change is most likely occurring will be a significant
-asset in directing the future collection of trait data, a main barrier
-for taking a functional approach to biodiversity change.
+understanding ecological patterns (78, 79).
+\textcolor{blue}{Here we assessed how key community characteristics such as biome, realm, taxa, and protection status may partition variation in the functional trends exhibited and found that, on the whole, these broad classifications did not seem to make the picture of change clearer.
+Further work linking specific ecological contexts to types of functional change will be critical for identifying new areas of concern, especially when those communities may not otherwise be showing significant changes in species richness.
+The identification of contexts where functional change is most likely occurring will be a significant asset in directing the future collection of trait data, the main barrier for taking a functional approach to biodiversity change.}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-barnagaud2017}{}}%
+\CSLLeftMargin{34. }%
+\CSLRightInline{Barnagaud J-Y, Gaüzère P, Zuckerberg B, Princé K,
+Svenning J-C (2017)
+\href{https://doi.org/10.1007/s00442-017-3967-4}{Temporal changes in
+bird functional diversity across the United States}. \emph{Oecologia}
+185(4):737--748.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-jarzyna2016}{}}%
+\CSLLeftMargin{35. }%
+\CSLRightInline{Jarzyna MA, Jetz W (2016)
+\href{https://doi.org/10.1111/gcb.13571}{A near half-century of temporal
+change in different facets of avian diversity}. \emph{Global Change
+Biology} 23(8):2999--3011.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-trindade-santos2020}{}}%
-\CSLLeftMargin{34. }%
+\CSLLeftMargin{36. }%
 \CSLRightInline{Trindade-Santos I, Moyes F, Magurran AE (2020)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-fitzjohn2014}{}}%
-\CSLLeftMargin{35. }%
+\CSLLeftMargin{37. }%
 \CSLRightInline{FitzJohn RG, et al. (2014)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wilman2014}{}}%
-\CSLLeftMargin{36. }%
+\CSLLeftMargin{38. }%
 \CSLRightInline{Wilman H, et al. (2014)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-blowes2019}{}}%
-\CSLLeftMargin{37. }%
+\CSLLeftMargin{39. }%
 \CSLRightInline{Blowes SA, et al. (2019)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-norman2020}{}}%
-\CSLLeftMargin{38. }%
+\CSLLeftMargin{40. }%
 \CSLRightInline{Norman KEA, Chamberlain S, Boettiger C (2020)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2021}{}}%
-\CSLLeftMargin{39. }%
+\CSLLeftMargin{41. }%
 \CSLRightInline{R Core Team (2021) \emph{R: A language and environment
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lalibertuxe92010}{}}%
-\CSLLeftMargin{40. }%
+\CSLLeftMargin{42. }%
 \CSLRightInline{Laliberté E, Legendre P (2010)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-mason2005}{}}%
-\CSLLeftMargin{41. }%
+\CSLLeftMargin{43. }%
 \CSLRightInline{Mason NWH, Mouillot D, Lee WG, Wilson JB, Setälä H
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hillebrand2009}{}}%
-\CSLLeftMargin{42. }%
+\CSLLeftMargin{44. }%
 \CSLRightInline{Hillebrand H, Matthiessen B (2009)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-leps2006}{}}%
-\CSLLeftMargin{43. }%
+\CSLLeftMargin{45. }%
 \CSLRightInline{Leps J, Bello F, Lavorel S, Berman S (2006) Quantifying
@@
 \leavevmode\vadjust pre{\hypertarget{ref-schleuter2010}{}}%
-\CSLLeftMargin{44. }%
+\CSLLeftMargin{46. }%
 \CSLRightInline{Schleuter D, Daufresne M, Massol F, Argillier C (2010)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-blonder2018}{}}%
-\CSLLeftMargin{45. }%
+\CSLLeftMargin{47. }%
 \CSLRightInline{Blonder B (2018)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-swenson2012}{}}%
-\CSLLeftMargin{46. }%
+\CSLLeftMargin{48. }%
 \CSLRightInline{Swenson NG, et al. (2012)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lenth2022}{}}%
-\CSLLeftMargin{47. }%
+\CSLLeftMargin{49. }%
 \CSLRightInline{Lenth RV (2022) \emph{Emmeans: Estimated marginal means,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bates2015}{}}%
-\CSLLeftMargin{48. }%
+\CSLLeftMargin{50. }%
 \CSLRightInline{Bates D, Mächler M, Bolker B, Walker S (2015) Fitting
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kuznetsova2017}{}}%
-\CSLLeftMargin{49. }%
+\CSLLeftMargin{51. }%
 \CSLRightInline{Kuznetsova A, Brockhoff PB, Christensen RHB (2017)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2023}{}}%
-\CSLLeftMargin{50. }%
+\CSLLeftMargin{52. }%
 \CSLRightInline{R Core Team (2023) \emph{R: A language and environment
@@
 \leavevmode\vadjust pre{\hypertarget{ref-brodie2021}{}}%
-\CSLLeftMargin{51. }%
+\CSLLeftMargin{53. }%
 \CSLRightInline{Brodie JF, Williams S, Garner B (2021) The decline of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-cox2022}{}}%
-\CSLLeftMargin{52. }%
+\CSLLeftMargin{54. }%
 \CSLRightInline{Cox DTC, Gardner AS, Gaston KJ (2022)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-li2022}{}}%
-\CSLLeftMargin{53. }%
+\CSLLeftMargin{55. }%
 \CSLRightInline{Li X, et al. (2022)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-filstrup2019}{}}%
-\CSLLeftMargin{54. }%
+\CSLLeftMargin{56. }%
 \CSLRightInline{Filstrup CT, King KBS, McCullough IM (2019)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hillebrand2008}{}}%
-\CSLLeftMargin{55. }%
+\CSLLeftMargin{57. }%
 \CSLRightInline{Hillebrand H, Bennett DM, Cadotte MW (2008)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bagousse-pinguet2019}{}}%
-\CSLLeftMargin{56. }%
+\CSLLeftMargin{58. }%
 \CSLRightInline{Bagousse-Pinguet YL, et al. (2019)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-spatz2023}{}}%
-\CSLLeftMargin{57. }%
+\CSLLeftMargin{59. }%
 \CSLRightInline{Spatz DR, et al. (2023)
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-estes2011}{}}%
+\CSLLeftMargin{60. }%
+\CSLRightInline{Estes JA, et al. (2011)
+\href{https://doi.org/10.1126/science.1205106}{Trophic Downgrading of
+Planet Earth}. \emph{Science} 333(6040):301--306.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-reich2012}{}}%
-\CSLLeftMargin{58. }%
+\CSLLeftMargin{61. }%
 \CSLRightInline{Reich PB, et al. (2012)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-pimiento2020a}{}}%
-\CSLLeftMargin{59. }%
+\CSLLeftMargin{62. }%
 \CSLRightInline{Pimiento C, et al. (2020)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-biggs2020}{}}%
-\CSLLeftMargin{60. }%
+\CSLLeftMargin{63. }%
 \CSLRightInline{Biggs CR, et al. (2020)
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-jarzyna2016}{}}%
-\CSLLeftMargin{61. }%
-\CSLRightInline{Jarzyna MA, Jetz W (2016)
-\href{https://doi.org/10.1111/gcb.13571}{A near half-century of temporal
-change in different facets of avian diversity}. \emph{Global Change
-Biology} 23(8):2999--3011.}
-
 \leavevmode\vadjust pre{\hypertarget{ref-inger2015}{}}%
-\CSLLeftMargin{62. }%
+\CSLLeftMargin{64. }%
 \CSLRightInline{Inger R, et al. (2015)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rosenberg2019}{}}%
-\CSLLeftMargin{63. }%
+\CSLLeftMargin{65. }%
 \CSLRightInline{Rosenberg KV, et al. (2019) Decline of the North
@@
 \leavevmode\vadjust pre{\hypertarget{ref-schipper2016}{}}%
-\CSLLeftMargin{64. }%
+\CSLLeftMargin{66. }%
 \CSLRightInline{Schipper AM, et al. (2016)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-luck2013}{}}%
-\CSLLeftMargin{65. }%
+\CSLLeftMargin{67. }%
 \CSLRightInline{Luck GW, Carter A, Smallbone L (2013)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-magioli2021}{}}%
-\CSLLeftMargin{66. }%
+\CSLLeftMargin{68. }%
 \CSLRightInline{Magioli M, et al. (2021)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-tinoco2018}{}}%
-\CSLLeftMargin{67. }%
+\CSLLeftMargin{69. }%
 \CSLRightInline{Tinoco BA, Santillán VE, Graham CH (2018)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-sheridan2011}{}}%
-\CSLLeftMargin{68. }%
+\CSLLeftMargin{70. }%
 \CSLRightInline{Sheridan JA, Bickford D (2011)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-caruso2014}{}}%
-\CSLLeftMargin{69. }%
+\CSLLeftMargin{71. }%
 \CSLRightInline{Caruso NM, Sears MW, Adams DC, Lips KR (2014)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-forster2012}{}}%
-\CSLLeftMargin{70. }%
+\CSLLeftMargin{72. }%
 \CSLRightInline{Forster J, Hirst AG, Atkinson D (2012)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-huss2019}{}}%
-\CSLLeftMargin{71. }%
+\CSLLeftMargin{73. }%
 \CSLRightInline{Huss M, Lindmark M, Jacobson P, van Dorst RM, Gårdmark A
@@
 \leavevmode\vadjust pre{\hypertarget{ref-tseng2018}{}}%
-\CSLLeftMargin{72. }%
+\CSLLeftMargin{74. }%
 \CSLRightInline{Tseng M, et al. (2018)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-fritz2009}{}}%
-\CSLLeftMargin{73. }%
+\CSLLeftMargin{75. }%
 \CSLRightInline{Fritz SA, Bininda-Emonds ORP, Purvis A (2009)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ripple2015}{}}%
-\CSLLeftMargin{74. }%
+\CSLLeftMargin{76. }%
 \CSLRightInline{Ripple WJ, et al. (2015)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-tilman2017}{}}%
-\CSLLeftMargin{75. }%
+\CSLLeftMargin{77. }%
 \CSLRightInline{Tilman D, et al. (2017)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-catford2022}{}}%
-\CSLLeftMargin{76. }%
+\CSLLeftMargin{78. }%
 \CSLRightInline{Catford JA, Wilson JRU, Pyšek P, Hulme PE, Duncan RP
@@
 \leavevmode\vadjust pre{\hypertarget{ref-spake2023}{}}%
-\CSLLeftMargin{77. }%
+\CSLLeftMargin{79. }%
 \CSLRightInline{Spake R, et al. (2023)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-pardieck2020}{}}%
-\CSLLeftMargin{78. }%
+\CSLLeftMargin{80. }%
 \CSLRightInline{Pardieck KL, David ZJ, Lutmerding M, Aponte V, Hudson
@@
 \leavevmode\vadjust pre{\hypertarget{ref-piropno}{}}%
-\CSLLeftMargin{79. }%
+\CSLLeftMargin{81. }%
 \CSLRightInline{PIROP northwest atlantic 1965{\textendash}1992 - OBIS
@@
 \leavevmode\vadjust pre{\hypertarget{ref-silva}{}}%
-\CSLLeftMargin{80. }%
+\CSLLeftMargin{82. }%
 \CSLRightInline{Silva FR da Brazil dataset 1.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-scott}{}}%
-\CSLLeftMargin{81. }%
+\CSLLeftMargin{83. }%
 \CSLRightInline{Scott D, Metts B, Lance S The rainbow bay long-term
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rossaferes1997}{}}%
-\CSLLeftMargin{82. }%
+\CSLLeftMargin{84. }%
 \CSLRightInline{Rossa-Feres D de C (1997) Community ecology of anura
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bakker1990}{}}%
-\CSLLeftMargin{83. }%
+\CSLLeftMargin{85. }%
 \CSLRightInline{Bakker C, Herman PMJ (1990) Phytoplankton in the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-berezovikovn.n.2004}{}}%
-\CSLLeftMargin{84. }%
+\CSLLeftMargin{86. }%
 \CSLRightInline{Berezovikov, N.N. (2004) The birds of settlements in
@@
 \leavevmode\vadjust pre{\hypertarget{ref-enemar2004}{}}%
-\CSLLeftMargin{85. }%
+\CSLLeftMargin{87. }%
 \CSLRightInline{Enemar A, Sjöstrand B, Andersson G, Von Proschwitz T
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gido2019}{}}%
-\CSLLeftMargin{86. }%
+\CSLLeftMargin{88. }%
 \CSLRightInline{Gido K (2019) CFP01 Fish population on selected
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hogstad1993}{}}%
-\CSLLeftMargin{87. }%
+\CSLLeftMargin{89. }%
 \CSLRightInline{Hogstad O (1993)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-holmes1986}{}}%
-\CSLLeftMargin{88. }%
+\CSLLeftMargin{90. }%
 \CSLRightInline{Holmes RT, Sherry TW, Sturges FW (1986)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-jahncke2006}{}}%
-\CSLLeftMargin{89. }%
+\CSLLeftMargin{91. }%
 \CSLRightInline{Jahncke J, Rintoul C (2006) CalCOFI and NMFS seabird and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-krivenkov.g.1991}{}}%
-\CSLLeftMargin{90. }%
+\CSLLeftMargin{92. }%
 \CSLRightInline{Krivenko, V.G. (1991) Waterfowl and their protection.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-melnikovy.i.2000}{}}%
-\CSLLeftMargin{91. }%
+\CSLLeftMargin{93. }%
 \CSLRightInline{Melnikov, Y.I., Melnikova, N., Pronkevich, V.V. (2000)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2013}{}}%
-\CSLLeftMargin{92. }%
+\CSLLeftMargin{94. }%
 \CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2014}{}}%
-\CSLLeftMargin{93. }%
+\CSLLeftMargin{95. }%
 \CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitori2014}{}}%
-\CSLLeftMargin{94. }%
+\CSLLeftMargin{96. }%
 \CSLRightInline{Monitoring site 1000 village survey - medium and large
@@
 \leavevmode\vadjust pre{\hypertarget{ref-preston1960}{}}%
-\CSLLeftMargin{95. }%
+\CSLLeftMargin{97. }%
 \CSLRightInline{Preston FW (1960)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-svensson2006}{}}%
-\CSLLeftMargin{96. }%
+\CSLLeftMargin{98. }%
 \CSLRightInline{Svensson S (2006) Species composition and population
@@
 \leavevmode\vadjust pre{\hypertarget{ref-svensson2010}{}}%
-\CSLLeftMargin{97. }%
+\CSLLeftMargin{99. }%
 \CSLRightInline{Svensson S, Thorner AM, Nyholm NEI (2010) Species
@@
 \leavevmode\vadjust pre{\hypertarget{ref-thorn2016changes}{}}%
-\CSLLeftMargin{98. }%
+\CSLLeftMargin{100. }%
 \CSLRightInline{Thorn S, et al. (2016)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-usfs}{}}%
-\CSLLeftMargin{99. }%
+\CSLLeftMargin{101. }%
 \CSLRightInline{USFS Landbird monitoring program (UMT-LBMP). Available
@@
 \leavevmode\vadjust pre{\hypertarget{ref-vermontcenterforecostudies2015}{}}%
-\CSLLeftMargin{100. }%
+\CSLLeftMargin{102. }%
 \CSLRightInline{Vermont Center For Ecostudies, Lambert JD, Hart J (2015)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-vickery1984}{}}%
-\CSLLeftMargin{101. }%
+\CSLLeftMargin{103. }%
 \CSLRightInline{Vickery WL, Nudds TD (1984)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-waide2017}{}}%
-\CSLLeftMargin{102. }%
+\CSLLeftMargin{104. }%
 \CSLRightInline{Waide R (2017) Bird abundance - point counts.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-williamson1983}{}}%
-\CSLLeftMargin{103. }%
+\CSLLeftMargin{105. }%
 \CSLRightInline{Williamson M (1983)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-zakharovv.d.1998}{}}%
-\CSLLeftMargin{104. }%
+\CSLLeftMargin{106. }%
 \CSLRightInline{Zakharov, V.D. (1998) Biodiversity of bird population of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-meditss2011}{}}%
-\CSLLeftMargin{105. }%
+\CSLLeftMargin{107. }%
 \CSLRightInline{MEDITS seabird surveys 1999 / 2000 / 2002 (2011)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-animald}{}}%
-\CSLLeftMargin{106. }%
+\CSLLeftMargin{108. }%
 \CSLRightInline{Animal demography unit - coordinated waterbird counts
@@
 \leavevmode\vadjust pre{\hypertarget{ref-baltics}{}}%
-\CSLLeftMargin{107. }%
+\CSLLeftMargin{109. }%
 \CSLRightInline{Baltic seabirds transect surveys Available at:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-thorn2016response}{}}%
-\CSLLeftMargin{108. }%
+\CSLLeftMargin{110. }%
 \CSLRightInline{Thorn S, et al. (2016)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ernest2009}{}}%
-\CSLLeftMargin{109. }%
+\CSLLeftMargin{111. }%
 \CSLRightInline{Ernest SKM, Valone TJ, Brown JH (2009)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-friggens2008}{}}%
-\CSLLeftMargin{110. }%
+\CSLLeftMargin{112. }%
 \CSLRightInline{Friggens M (2008) Sevilleta LTER small mammal population
@@
 \leavevmode\vadjust pre{\hypertarget{ref-jalilova.b.2014}{}}%
-\CSLLeftMargin{111. }%
+\CSLLeftMargin{113. }%
 \CSLRightInline{Jalilov, A. B., Andreychev, A. V., Kuznetsov, V. A.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kartzinel2014}{}}%
-\CSLLeftMargin{112. }%
+\CSLLeftMargin{114. }%
 \CSLRightInline{Kartzinel TR, et al. (2014)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kaufman2019}{}}%
-\CSLLeftMargin{113. }%
+\CSLLeftMargin{115. }%
 \CSLRightInline{Kaufman D (2019) CSM01 Seasonal Summary of Numbers of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-krefting1974}{}}%
-\CSLLeftMargin{114. }%
+\CSLLeftMargin{116. }%
 \CSLRightInline{Krefting LW, Ahlgren CE (1974)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lightfoot}{}}%
-\CSLLeftMargin{115. }%
+\CSLLeftMargin{117. }%
 \CSLRightInline{Lightfoot D, Schooley RL SMES rodent trapping data,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-malyshevy.s.2011}{}}%
-\CSLLeftMargin{116. }%
+\CSLLeftMargin{118. }%
 \CSLRightInline{Malyshev, Y. S. (2011) On the diagnostic techniques of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-meyer2008}{}}%
-\CSLLeftMargin{117. }%
+\CSLLeftMargin{119. }%
 \CSLRightInline{Meyer CFJ, Kalko EKV (2008)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nedosekinv.y.2015}{}}%
-\CSLLeftMargin{118. }%
+\CSLLeftMargin{120. }%
 \CSLRightInline{Nedosekin, V. Y. (2015) Long-term dynamics of the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-prins1990}{}}%
-\CSLLeftMargin{119. }%
+\CSLLeftMargin{121. }%
 \CSLRightInline{Prins HHT, Douglas-Hamilton I (1990)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rocha2017}{}}%
-\CSLLeftMargin{120. }%
+\CSLLeftMargin{122. }%
 \CSLRightInline{Rocha R, et al. (2017)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-stapp2014}{}}%
-\CSLLeftMargin{121. }%
+\CSLLeftMargin{123. }%
 \CSLRightInline{Stapp P (2014) SGS-LTER long-term monitoring project:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bahamas}{}}%
-\CSLLeftMargin{122. }%
+\CSLLeftMargin{124. }%
 \CSLRightInline{Bahamas marine mammal research organisation
@@
 \leavevmode\vadjust pre{\hypertarget{ref-popacet}{}}%
-\CSLLeftMargin{123. }%
+\CSLLeftMargin{125. }%
 \CSLRightInline{POPA cetacean, seabird, and sea turtle sightings in the

### 2024-04-01 paper/GEB Submission/manuscript_GEB.Rmd
Reformat discussion

@@
 
-`r colorize("The expectation for functional change across communities is not obvious from past work, and may or may not follow taxonomic trends", "blue")` [@dornelas2023].
+`r colorize("The expectation for functional diversity change across communities is not obvious from past work, and may or may not follow taxonomic trends", "blue")` [@dornelas2023].
 While `r colorize("loss of functional diversity", "blue")` is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], `r colorize("functional diversity may be maintained even when species are lost from a community", "blue")` [@daz2001; @petchey2009].
@@
 
-Here we `r colorize("leverage ongoing efforts to assemble functional trait data and recent computational advances to", "blue")` perform the first multi-taxa, multi-realm assessment of functional diversity change through time.
+We have stronger expectations for changes in the prevalence of some individual traits.
+For example, animal body size is expected to decrease as a result of climate change, a phenomena that has been documented in multiple taxa empirically and experimentally [@caruso2014; @forster2012; @huss2019; @sheridan2011; @tseng2018].
+There is some evidence that this holds true for birds [@vanbuskirk2010], but the picture is likely more complicated for mammals, where urbanization may actually lead to larger body sizes due to novel food sources [@hantak2021], even as megafaunal loss leads to on average smaller community body size.
+For dietary traits, recent work documenting insect declines (citation) points to potentially significant negative impacts on insectivorous animals [@canaday1996; @sekercioglu2002; @tallamy2021].
+Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species [@cooke2019].
+Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems [@bain2020], and loss of neotropical understory foragers even in protect areas [@pollock2022].
+
+Here we `r colorize("leverage ongoing efforts to assemble functional trait data and recent computational advances to", "blue")` perform the first multi-taxa, multi-realm assessment of functional diversity and composition change through time.
 We focus on mammal and bird species as `r colorize("a subset of the world's biodiversity of particular conservation concern that is heavily impacted by anthropogenic change.", "blue")`While examining trends in plants, invertebrates, and other vertebrate species is of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
@@
 We assess thousands of mammal and bird functional diversity time series to determine whether or not `r colorize("the addition of functional trait data gives a clarifying picture of biodiversity change across communities.
-Rather than testing specific hypotheses of change we present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination.
-We present results for three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.", "blue")`
+We present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination.
+We present results for three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.", "blue")` We further evaluate evidence for key predictions of individual trait prevalence from the literature, including changes in mean body size, dietary traits, and foraging strata.\`
 
@@
 ```{r traitTable, echo=FALSE}
-tibble(Trait = c(rep("Diet", 10), rep("Foraging Strata", 7), rep("Pelagic Specialist", 2), rep("Nocturnal", 2), "Body Mass"),
+tibble(Trait = c(rep("Diet", 10), rep("Foraging Strata", 7), rep("Pelagic Specialist", 2), rep("Nocturnal", 2), rep("Crepuscular", 2), rep("Diurnal", 2), "Body Mass"),
        Category = c("Invertebrate", "Mammals and Birds", "Reptiles", "Fish", "Unknown Vertebrates", "Scavenging", "Fruit", "Nectar", "Seeds", "Other Plant", 
                     "Below water surface", "water surface", "ground", "understory", "> 2m, below canopy", "canopy", "aerial",
-                    "yes", "no",
-                    "yes", "no",
+                    rep(c("yes", "no"), 4),
                     "-"),
-       `Data Type` = c(rep("percentage consumed", 10), rep("percentage of use", 7), rep("binary", 4), "continous, in grams")) %>%
+       Taxa = c(rep("Bird and Mammal", 10), rep("Bird", 7), rep("Bird", 2), rep("Bird and Mammal", 2), rep("Mammal", 4), "Bird and Mammal"),
+       `Data Type` = c(rep("percentage consumed", 10), rep("percentage of use", 7), rep("binary", 8), "continous, in grams")) %>%
   kable(format = "latex", caption = "Description of the traits included in the analysis broken down by categories at data type.") %>%
-  collapse_rows(columns = 1:3) %>%
+  collapse_rows(columns = 1:4) %>%
+  column_spec(3:4, width = "3cm") %>%
   kable_paper()
@@
 Higher values of *FDiv* therefore correspond to communities where many highly abundant species are on the edges of the trait space.
-We also calculated the community-weighted mean (*CWM*) of included traits to examine shifts in the distribution of each trait.
-`r colorize("Hereafter we refer to results for functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).", "blue")`
+
+We also calculated the community-weighted mean (*CWM*) of all continuous traits to examine `r colorize("turnover", "blue")` in the distribution of each trait.
+`r colorize("Wholesale shifts in the trait space due to changes of trait means could occur even while the shape of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained.
+CWM's are therefore a way to assess whether or not that turnover is occurring and what the nature of the shift may be.
+Hereafter we refer to results for functional metrics in two groups: functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).", "blue")`
 
@@
 
-### Evidence of consensus
+## Evidence of consensus
 
@@
 `r colorize("However we did not find trends outside the random expectation for species loss, in contrast to expectations for multiple continents based on IUCN risk categories", "blue")` [@brodie2021].
-`r colorize("In general, functional capacity of mammals shows greater declines that bird communities.", "blue")`
+`r colorize("In general, functional capacity of mammals showed greater declines that bird communities.", "blue")`
 
-`r colorize("We also found evidence of increasing evenness for Marine studies.
+## Areas of incongruence
+
+`r colorize("The addition of functional trait data illuminated multiple areas where a synthesis approach is incongruous with system-specific studies or simply unexpected.
+For example, results for protected areas were surprising given the assumption that protection insulates communities from functional degradation.
+The only difference we found between protected and unprotected areas was significant turnover occurring outside protected areas and no significant turnover inside protected areas.
+Rather than the result of protection itself, this is likely due to the fact that the majority of studies were from terrestrial, temperate studies where turnover is known to be lower ", "blue")`[@blowes2019].
+`r colorize("While things may generally be more static within protect areas, the functional dimensions of protected communities fared no better or worse on average than their non-protected counter parts.
+Looking at individual studies, there is a mix of both losses and gains in almost all functional diversity and composition metrics across communities found in protected areas.", "blue")`
+
+`r colorize("We also found evidence of increasing evenness for marine studies, which was not predicted by previous work in marine systems.
 The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability ", "blue")` [@filstrup2019], `r colorize("as the dominance of one or a few species makes function sensitive to the ability of those species to respond to environmental change ", "blue")`[@hillebrand2008; @hillebrand2009].
 `r colorize("Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities ", "blue")`[@bagousse-pinguet2019].
-`r colorize("The majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery, which is also indicated in recent species-based work ", "blue")`[@spatz2023].
+`r colorize("An increase in evenness is therefore a positive indicator for these communities, despite the fact that marine bird and mammal communities are some of the most threatened communities in the world.
+Still, this result may be a function of data limitations as the majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery, which is also indicated in recent species-based work ", "blue")`[@spatz2023].
 
-`r colorize("Compositional shifts shown by significant trends in CWM's were restricted to dietary traits, with no evidence of a trend in foraging traits or body mass.
-Increases in fish consumption for bird and temperate studies my be the product of increased availability of easy fish from commercial fishing operations as fish are processed or stocking of inland waterways for sport fishing.
-However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch.
-Decreases in nectar consumption for tropical studies is likely due declines in bat species, as 4 of the 5 tropical studies are for mammal and bats were the only nectar consuming mammals sampled in our data.
+Our results were also inconsistent with all predictions for changes in trait prevalence, including foraging strata, body size, and dietary traits.
+`r colorize("We found no changes in the prevalence of different foraging strategies, despite documented losses of understory birds in the neotropics and some evidence that agricultural incursion particularly threatens arboreal species.
+These shifts may therefore be the result of specific contexts, and not generalizable to bird communities across the globe.
+We found no evidence of general trends in the *CWM* for body mass for birds or mammals, indicating that either body size is not changing significantly due to climate change, opposing pressures such as urbanization are overshadowing climate change impacts, or more likely current shifts are happening at an intraspecific level and and have not yet propogated to the species losses that would be captured by our data.
+Further, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].
+`r colorize("Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened", "blue")` [@ripple2015].
+
+Rather than declines in insectivory and increases in invertivory predicted by previous work, we instead found changes in the degree of consumption of fish, nectar, and vertebrates.
+`r colorize("Increases in fish consumption for bird and temperate studies may be the product of increased availability of easy fish from commercial fishing operations or the stocking of inland waterways for sport fishing.
+However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch, making these results suprising.
+Decreases in nectar consumption for tropical studies is likely due to declines in bat species, as 4 of the 5 tropical studies are for mammal and bats were the only nectar consuming mammals sampled in our data.
 Interpretation of the increases for temperate studies is less clear, as temperate studies included many bird communities with a variety of nectar consumers.
@@
 
-### Study heterogeneity
+## Study heterogeneity
 
@@
 
-<!-- huge differences between different taxa and the kind of disturbance/land use change -->
-
-```{r changeScenarios, echo = FALSE, message = FALSE, fig.show = 'hold', fig.cap= "Comparison of the distribution of time series durations from each type of change to the overall distribution of time series durations, with no richness change or turnover and functional change in yellow, no turnover but richness and functional change in green, no change in blue, and turnover, richness and functional change studies in orange."}
-# 
-# knitr::include_graphics(here::here("figures/duration_hist.jpeg"))
-# 
-# knitr::include_graphics(here::here("figures/functional_change_table.png"))
-```
-
-### Areas of incongruence
-
-`r colorize("The addition of functional trait data illuminates a few key areas where a synthesis approach is incongruous with system-specific studies or expectations based on anthropegenic impacts.
-For example, we found no evidence of general trends in the *CWM* for body mass or for studies aggregated by their characteristics despite predictions that mean body mass is decreasing as a result of climate change impacts and megafaunal loss ", "blue")`[@sheridan2011] `r colorize(", and downgrading documented empirically and experimentally in multiple taxa ", "blue")`[@caruso2014; @forster2012; @huss2019; @tseng2018].
-`r colorize("Still, climate change impacts are predicted to be most extreme for ectotherms and impact intraspecific traits first before species are lost, neither of which are represented by our data.
-Further, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].
-`r colorize("Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened", "blue")` [@ripple2015].
-
-`r colorize("Results for protected areas were also surprising given the assumption that protection insulates communities from functional degradation.
-The only difference we found between protected and unprotected areas was significant turnover occurring outside protected areas and no significant turnover inside protected areas.
-Rather than the result of protection itself, this is likely due to the fact that the majority of studies were form terrestrial, temperate studies where turnover is known to be lower ", "blue")`[@blowes2019].
-`r colorize("While things may generally be more static within protect areas, the functional dimensions of protected communities fared no better or worse on average than their non-protected counter parts.
-Looking at individual studies, there is a mix of both losses and gains in almost all functional diversity and composition metrics across communities found in protected areas.", "blue")`
-
 ## Policy Implications

### 2024-04-01 paper/GEB Submission/manuscript_GEB.tex
Reformat discussion

@@
 studies showed no species or functional trends, but other types of
-change were manifest for multiple studies, including studies
-experiencing changes in functional redundancy, increased species and
-functional richness, or loss of functional richness. With no one
-prevailing scenario of change, it will be critical to link change
-scenarios to ecological context.
+change were found for multiple studies, including studies experiencing
+changes in functional redundancy, increased species and functional
+richness, or loss of functional richness. With no one prevailing
+scenario of change, it will be critical to link change scenarios to
+ecological context.
 \end{abstract}
@@
 
-\textcolor{blue}{The expectation for functional change across communities is not obvious from past work, and may or may not follow taxonomic trends}
+\textcolor{blue}{The expectation for functional diversity change across communities is not obvious from past work, and may or may not follow taxonomic trends}
 (13). While \textcolor{blue}{loss of functional diversity} is frequently
@@
 
+We have stronger expectations for changes in the prevalence of some
+individual traits. For example, animal body size is expected to decrease
+as a result of climate change, a phenomena that has been documented in
+multiple taxa empirically and experimentally (37--41). There is some
+evidence that this holds true for birds (42), but the picture is likely
+more complicated for mammals, where urbanization may actually lead to
+larger body sizes due to novel food sources (43), even as megafaunal
+loss leads to on average smaller community body size. For dietary
+traits, recent work documenting insect declines (citation) points to
+potentially significant negative impacts on insectivorous animals
+(44--46). Predicted extinctions based on species-level vulnerability
+point to further dietary shifts, favoring increases in invertivorous
+species (47). Some systems also show significant shifts in the
+prevalence of different kinds of foragers in birds, for example loss of
+arboreal foragers in agricultural systems (48), and loss of neotropical
+understory foragers even in protect areas (49).
+
 Here we
@@
 perform the first multi-taxa, multi-realm assessment of functional
-diversity change through time. We focus on mammal and bird species as
+diversity and composition change through time. We focus on mammal and
+bird species as
 \textcolor{blue}{a subset of the world's biodiversity of particular conservation concern that is heavily impacted by anthropogenic change.}While
@@
 is of equal interest, trait data for those taxa raise additional
-challenges such as limited and biased species coverage (37), a lack of
+challenges such as limited and biased species coverage (50), a lack of
 accepted species-level means, and differences in the types of traits
@@
 Eltonian niche, thereby summarizing the functional role they play in the
-community (38). An initial assessment of amphibian trends is included in
+community (51). An initial assessment of amphibian trends is included in
 the supplement, but excluded from general trend assessment here due to
@@
 \textcolor{blue}{the addition of functional trait data gives a clarifying picture of biodiversity change across communities.
-Rather than testing specific hypotheses of change we present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination.
+We present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination.
 We present results for three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.}
+We further evaluate evidence for key predictions of individual trait
+prevalence from the literature, including changes in mean body size,
+dietary traits, and foraging strata.`
 
@@
 species (31), and all time series include abundance of observed species.
-Following best practices for the database (39), studies with multiple
+Following best practices for the database (52), studies with multiple
 sample locations were split into individual time series following a
@@
 species-level means for traits that represent species' multifaceted role
-in the community (38). Traits include: body mass, diet, nocturnality,
+in the community (51). Traits include: body mass, diet, nocturnality,
 forest foraging strata, pelagic use. With the exception of body mass,
@@
 \centering
-\begin{tabular}[t]{l|l|l}
+\begin{tabular}[t]{l|l|>{\raggedright\arraybackslash}p{3cm}|>{\raggedright\arraybackslash}p{3cm}}
 \hline
-Trait & Category & Data Type\\
+Trait & Category & Taxa & Data Type\\
 \hline
- & Invertebrate & \\
-\cline{2-2}
- & Mammals and Birds & \\
-\cline{2-2}
- & Reptiles & \\
-\cline{2-2}
- & Fish & \\
-\cline{2-2}
- & Unknown Vertebrates & \\
-\cline{2-2}
- & Scavenging & \\
-\cline{2-2}
- & Fruit & \\
-\cline{2-2}
- & Nectar & \\
-\cline{2-2}
- & Seeds & \\
-\cline{2-2}
-\multirow{-10}{*}{\raggedright\arraybackslash Diet} & Other Plant & \multirow{-10}{*}{\raggedright\arraybackslash percentage consumed}\\
+ & Invertebrate &  & \\
+\cline{2-2}
+ & Mammals and Birds &  & \\
+\cline{2-2}
+ & Reptiles &  & \\
+\cline{2-2}
+ & Fish &  & \\
+\cline{2-2}
+ & Unknown Vertebrates &  & \\
+\cline{2-2}
+ & Scavenging &  & \\
+\cline{2-2}
+ & Fruit &  & \\
+\cline{2-2}
+ & Nectar &  & \\
+\cline{2-2}
+ & Seeds &  & \\
+\cline{2-2}
+\multirow{-10}{*}{\raggedright\arraybackslash Diet} & Other Plant & \multirow{-10}{*}{\raggedright\arraybackslash Bird and Mammal} & \multirow{-10}{*}{\raggedright\arraybackslash percentage consumed}\\
+\cline{1-4}
+ & Below water surface &  & \\
+\cline{2-2}
+ & water surface &  & \\
+\cline{2-2}
+ & ground &  & \\
+\cline{2-2}
+ & understory &  & \\
+\cline{2-2}
+ & > 2m, below canopy &  & \\
+\cline{2-2}
+ & canopy &  & \\
+\cline{2-2}
+\multirow{-7}{*}{\raggedright\arraybackslash Foraging Strata} & aerial &  & \multirow{-7}{*}{\raggedright\arraybackslash percentage of use}\\
+\cline{1-2}
+\cline{4-4}
+ & yes &  & \\
+\cline{2-2}
+\multirow{-2}{*}{\raggedright\arraybackslash Pelagic Specialist} & no & \multirow{-9}{*}{\raggedright\arraybackslash Bird} & \\
 \cline{1-3}
- & Below water surface & \\
-\cline{2-2}
- & water surface & \\
-\cline{2-2}
- & ground & \\
-\cline{2-2}
- & understory & \\
-\cline{2-2}
- & > 2m, below canopy & \\
-\cline{2-2}
- & canopy & \\
-\cline{2-2}
-\multirow{-7}{*}{\raggedright\arraybackslash Foraging Strata} & aerial & \multirow{-7}{*}{\raggedright\arraybackslash percentage of use}\\
+ & yes &  & \\
+\cline{2-2}
+\multirow{-2}{*}{\raggedright\arraybackslash Nocturnal} & no & \multirow{-2}{*}{\raggedright\arraybackslash Bird and Mammal} & \\
 \cline{1-3}
- & yes & \\
-\cline{2-2}
-\multirow{-2}{*}{\raggedright\arraybackslash Pelagic Specialist} & no & \\
+ & yes &  & \\
+\cline{2-2}
+\multirow{-2}{*}{\raggedright\arraybackslash Crepuscular} & no &  & \\
 \cline{1-2}
- & yes & \\
-\cline{2-2}
-\multirow{-2}{*}{\raggedright\arraybackslash Nocturnal} & no & \multirow{-4}{*}{\raggedright\arraybackslash binary}\\
-\cline{1-3}
-Body Mass & - & continous, in grams\\
+ & yes &  & \\
+\cline{2-2}
+\multirow{-2}{*}{\raggedright\arraybackslash Diurnal} & no & \multirow{-4}{*}{\raggedright\arraybackslash Mammal} & \multirow{-8}{*}{\raggedright\arraybackslash binary}\\
+\cline{1-4}
+Body Mass & - & Bird and Mammal & continous, in grams\\
 \hline
@@
 \url{https://doi.org/10.5066/F7KH0KBK}), obtained through the
-\texttt{taxadb} R package (40, 41). If more than one species in the
+\texttt{taxadb} R package (53, 54). If more than one species in the
 assemblage data resolved to the same identifier, observations were
@@
 Functional diversity metrics were calculated using the \emph{dbFD}
-function from the \emph{FD} R package (42). Here we report functional
+function from the \emph{FD} R package (55). Here we report functional
 richness (\emph{FRic}), functional evenness (\emph{FEve}), and
 functional divergence (\emph{FDiv}) which together describe three
-complementary characteristics of the functional space (43, 44).
+complementary characteristics of the functional space (56, 57).
 \emph{FRic} assesses the volume of the trait space occupied by species
@@
 \emph{FDiv} therefore correspond to communities where many highly
-abundant species are on the edges of the trait space. We also calculated
-the community-weighted mean (\emph{CWM}) of included traits to examine
-shifts in the distribution of each trait.
-\textcolor{blue}{Hereafter we refer to results for functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).}
+abundant species are on the edges of the trait space.
+
+We also calculated the community-weighted mean (\emph{CWM}) of all
+continuous traits to examine \textcolor{blue}{turnover} in the
+distribution of each trait.
+\textcolor{blue}{Wholesale shifts in the trait space due to changes of trait means could occur even while the shape of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained.
+CWM's are therefore a way to assess whether or not that turnover is occurring and what the nature of the shift may be.
+Hereafter we refer to results for functional metrics in two groups: functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).}
 
@@
 traits were z-score scaled to give each trait equal weight in the trait
-space (45, 46).
+space (58, 59).
 \textcolor{blue}{Before calculating diversity metrics, dbFD reduces the dimensionality of the trait space by performing PCoA. We limited the number of included PCoA axes to}
@@
 traits. This restriction allows for enough axes to capture the trait
-space while maintaining computational feasibility (47). Metrics
+space while maintaining computational feasibility (60). Metrics
 incorporated weighting based on species abundance.
@@
 functional diversity metrics (\emph{FRic}, \emph{FEve}, \emph{FDiv})
-from null estimates (48). Null model corrections allow us to assess the
+from null estimates (61). Null model corrections allow us to assess the
 degree to which the observed functional diversity metric deviates from
@@
 structure. We estimated within-level slopes and calculated between-level
-contrasts using the \emph{emmeans} package (v1.8.2, 49).
+contrasts using the \emph{emmeans} package (v1.8.2, 62).
 \textcolor{blue}{For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore refrained from interpreting results for levels where there were fewer than three studies.}
@@
 (v3.1-3) package with a significance level of \(\alpha = 0.05\)
-(50--52).
+(63--65).
 
@@
 \hypertarget{evidence-of-consensus}{%
-\subsubsection{Evidence of consensus}\label{evidence-of-consensus}}
+\subsection{Evidence of consensus}\label{evidence-of-consensus}}
 
@@
 We found significant turnover for all biomes, realm, and taxa except for marine studies, which stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial}
-(39).
+(52).
 \textcolor{blue}{However, previous global estimates are dominated by fish communities which we exclude here and may be driving the overall trend while disguising relative stasis in marine bird and mammal communities.}
@@
 This result is consistent with previous work linking anthropogenic drivers such as habitat loss, poaching, and human modification and development to loss of functional diversity in mammal communities across the globe}
-(53--55).
+(66--68).
 \textcolor{blue}{However we did not find trends outside the random expectation for species loss, in contrast to expectations for multiple continents based on IUCN risk categories}
-(53).
-\textcolor{blue}{In general, functional capacity of mammals shows greater declines that bird communities.}
-
-\textcolor{blue}{We also found evidence of increasing evenness for Marine studies.
+(66).
+\textcolor{blue}{In general, functional capacity of mammals showed greater declines that bird communities.}
+
+\hypertarget{areas-of-incongruence}{%
+\subsection{Areas of incongruence}\label{areas-of-incongruence}}
+
+\textcolor{blue}{The addition of functional trait data illuminates a few key areas where a synthesis approach is incongruous with system-specific studies or simply unexpected.
+For example, results for protected areas were surprising given the assumption that protection insulates communities from functional degradation.
+The only difference we found between protected and unprotected areas was significant turnover occurring outside protected areas and no significant turnover inside protected areas.
+Rather than the result of protection itself, this is likely due to the fact that the majority of studies were from terrestrial, temperate studies where turnover is known to be lower }(52).
+\textcolor{blue}{While things may generally be more static within protect areas, the functional dimensions of protected communities fared no better or worse on average than their non-protected counter parts.
+Looking at individual studies, there is a mix of both losses and gains in almost all functional diversity and composition metrics across communities found in protected areas.}
+
+\textcolor{blue}{We also found evidence of increasing evenness for marine studies, which was not predicted by previous work in marine systems.
 The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability }
-(56),
-\textcolor{blue}{as the dominance of one or a few species makes function sensitive to the ability of those species to respond to environmental change }(44,
-57).
-\textcolor{blue}{Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities }(58).
-\textcolor{blue}{The majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery, which is also indicated in recent species-based work }(59).
-
-\textcolor{blue}{Compositional shifts shown by significant trends in CWM's were restricted to dietary traits, with no evidence of a trend in foraging traits or body mass.
-Increases in fish consumption for bird and temperate studies my be the product of increased availability of easy fish from commercial fishing operations as fish are processed or stocking of inland waterways for sport fishing.
-However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch.
-Decreases in nectar consumption for tropical studies is likely due declines in bat species, as 4 of the 5 tropical studies are for mammal and bats were the only nectar consuming mammals sampled in our data.
+(69),
+\textcolor{blue}{as the dominance of one or a few species makes function sensitive to the ability of those species to respond to environmental change }(57,
+70).
+\textcolor{blue}{Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities }(71).
+\textcolor{blue}{An increase in evenness is therefore a positive indicator for these communities, despite the fact that marine bird and mammal communities are some of the most threatened communities in the world.
+Still, this result may be a function of data limitations as the majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery, which is also indicated in recent species-based work }(72).
+
+Our results were also inconsistent with all predictions for changes in
+trait prevalence, including foraging strata, body size, and dietary
+traits.
+\textcolor{blue}{We found no changes in the prevalence of different foraging strategies, despite documented losses of understory birds in the neotropics and some evidence that agricultural incursion particularly threatens arboreal species.
+These shifts may therefore be the result of specific contexts, and not generalizable to bird communities across the globe.
+We found no evidence of general trends in the *CWM* for body mass for birds or mammals, indicating that either body size is not changing significantly due to climate change, opposing pressures such as urbanization are overshadowing climate change impacts, or more likely current shifts are happening at an intraspecific level and and have not yet propogated to the species losses that would be captured by our data.
+Further, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing}
+(73).
+\textcolor{blue}{Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened}
+(74).
+
+Rather than declines in insectivory and increases in invertivory
+predicted by previous work, we instead found changes in the degree of
+consumption of fish, nectar, and vertebrates.
+\textcolor{blue}{Increases in fish consumption for bird and temperate studies may be the product of increased availability of easy fish from commercial fishing operations as fish are processed or stocking of inland waterways for sport fishing.
+However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch, which makes these results suprising.
+Decreases in nectar consumption for tropical studies is likely due to declines in bat species, as 4 of the 5 tropical studies are for mammal and bats were the only nectar consuming mammals sampled in our data.
 Interpretation of the increases for temperate studies is less clear, as temperate studies included many bird communities with a variety of nectar consumers.
@@
 \hypertarget{study-heterogeneity}{%
-\subsubsection{Study heterogeneity}\label{study-heterogeneity}}
+\subsection{Study heterogeneity}\label{study-heterogeneity}}
 
@@
 Alternatively, these may be communities that have experienced or continue to experience significant stress, but lost species or functional diversity outside the observation window.
-This could be true particularly for North American mammal communities where trophic downgrading and megafaunal losses occurred hundreds of years ago }(60).
+This could be true particularly for North American mammal communities where trophic downgrading and megafaunal losses occurred hundreds of years ago }(75).
 \textcolor{blue}{Third, these communities may be experiencing directional shifts undetectable by available data.
@@
 By definition, if a community gains or loses species while functional metrics are unchanged, those species represent an increase or decrease in redundancy, so we include communities exhibiting gains (2 studies) or losses (5 studies) in species richness with no change in functional metrics or turnover and species richness loss with significant turnover (2 studies) in this umbrella.
-For communities exhibiting loss of redundancy (declining species richness), ecological processes are likely being maintained, but capacity to respond to future stressors is reduced }(61).
-\textcolor{blue}{These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future }(62).
+For communities exhibiting loss of redundancy (declining species richness), ecological processes are likely being maintained, but capacity to respond to future stressors is reduced }(76).
+\textcolor{blue}{These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future }(77).
 \textcolor{blue}{Conversely, communities exhibiting a increase in redundancy (via an increase in species richness) are becoming better positioned to respond to future stressors, but are not actually expanding their functional capacity as may be assumed by looking at species gains alone}
-(63).
+(78).
 
@@
 \textcolor{blue}{, and loss of common, functionally general species even as rare species are increasing in North America and Europe }
-(64--66).
+(79--81).
 
@@
 These communities are of particular concern not just because of the role evenness plays in maintaining functionality, but because of increasing evidence evenness is more sensitive to environmental change than species richness}
-(56).
+(69).
 \textcolor{blue}{Formal tests of evenness changes as an early warning signal of more catastrophic shifts in functional diversity will be critical next step for understanding the full implication of evenness loss for ecosystem processes.}
@@
 While it is tempting to draw a one to one line between the degree of human impact and loss of functional integrity (in the form of functional redundancy and richness), that relationship is not borne out by our rough metric of human impact (protection status) or previous taxa-specific or single community work, which is largely mixed}
-(26, 67--69).
-
-\hypertarget{areas-of-incongruence}{%
-\subsubsection{Areas of incongruence}\label{areas-of-incongruence}}
-
-\textcolor{blue}{The addition of functional trait data illuminates a few key areas where a synthesis approach is incongruous with system-specific studies or expectations based on anthropegenic impacts.
-For example, we found no evidence of general trends in the *CWM* for body mass or for studies aggregated by their characteristics despite predictions that mean body mass is decreasing as a result of climate change impacts and megafaunal loss }(70)
-\textcolor{blue}{, and downgrading documented empirically and experimentally in multiple taxa }(71--74).
-\textcolor{blue}{Still, climate change impacts are predicted to be most extreme for ectotherms and impact intraspecific traits first before species are lost, neither of which are represented by our data.
-Further, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing}
-(75).
-\textcolor{blue}{Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened}
-(76).
-
-\textcolor{blue}{Results for protected areas were also surprising given the assumption that protection insulates communities from functional degradation.
-The only difference we found between protected and unprotected areas was significant turnover occurring outside protected areas and no significant turnover inside protected areas.
-Rather than the result of protection itself, this is likely due to the fact that the majority of studies were form terrestrial, temperate studies where turnover is known to be lower }(39).
-\textcolor{blue}{While things may generally be more static within protect areas, the functional dimensions of protected communities fared no better or worse on average than their non-protected counter parts.
-Looking at individual studies, there is a mix of both losses and gains in almost all functional diversity and composition metrics across communities found in protected areas.}
+(26, 82--84).
 
@@
 
-\textcolor{blue}{One of the biggest threats to biodiversity is the wholesale conversion of natural areas to urban or human-dominated landscapes }(77).
+\textcolor{blue}{One of the biggest threats to biodiversity is the wholesale conversion of natural areas to urban or human-dominated landscapes }(85).
 \textcolor{blue}{Typical long term monitoring such as those included in our study stops before this conversion occurs, leaving the resultant precipitous declines in biodiversity unrecorded}
@@
 conversation is yet another illustration of how critical context is for
-understanding ecological patterns (78, 79).
+understanding ecological patterns (86, 87).
 \textcolor{blue}{Here we assessed how key community characteristics such as biome, realm, taxa, and protection status may partition variation in the functional trends exhibited and found that, on the whole, these broad classifications did not seem to make the picture of change clearer.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-caruso2014}{}}%
+\CSLLeftMargin{37. }%
+\CSLRightInline{Caruso NM, Sears MW, Adams DC, Lips KR (2014)
+\href{https://doi.org/10.1111/gcb.12550}{Widespread rapid reductions in
+body size of adult salamanders in response to climate change}.
+\emph{Global Change Biology} 20(6):1751--1759.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-forster2012}{}}%
+\CSLLeftMargin{38. }%
+\CSLRightInline{Forster J, Hirst AG, Atkinson D (2012)
+\href{https://doi.org/10.1073/pnas.1210460109}{Warming-induced
+reductions in body size are greater in aquatic than terrestrial
+species}. \emph{Proceedings of the National Academy of Sciences}
+109(47):19310--19314.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-huss2019}{}}%
+\CSLLeftMargin{39. }%
+\CSLRightInline{Huss M, Lindmark M, Jacobson P, van Dorst RM, Gårdmark A
+(2019) \href{https://doi.org/10.1111/gcb.14637}{Experimental evidence of
+gradual size-dependent shifts in body size and growth of fish in
+response to warming}. \emph{Global Change Biology} 25(7):2285--2295.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-sheridan2011}{}}%
+\CSLLeftMargin{40. }%
+\CSLRightInline{Sheridan JA, Bickford D (2011)
+\href{https://doi.org/10.1038/nclimate1259}{Shrinking body size as an
+ecological response to climate change}. \emph{Nature Climate Change}
+1(8):401--406.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-tseng2018}{}}%
+\CSLLeftMargin{41. }%
+\CSLRightInline{Tseng M, et al. (2018)
+\href{https://doi.org/10.1111/1365-2656.12789}{Decreases in beetle body
+size linked to climate change and warming temperatures}. \emph{Journal
+of Animal Ecology} 87(3):647--659.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-vanbuskirk2010}{}}%
+\CSLLeftMargin{42. }%
+\CSLRightInline{Van Buskirk J, Mulvihill RS, Leberman RC (2010)
+\href{https://www.jstor.org/stable/25700339}{Declining body sizes in
+north american birds associated with climate change}. \emph{Oikos}
+119(6):1047--1055.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hantak2021}{}}%
+\CSLLeftMargin{43. }%
+\CSLRightInline{Hantak MM, McLean BS, Li D, Guralnick RP (2021)
+\href{https://doi.org/10.1038/s42003-021-02505-3}{Mammalian body size is
+determined by interactions between climate, urbanization, and ecological
+traits}. \emph{Communications Biology} 4(1):1--10.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-canaday1996}{}}%
+\CSLLeftMargin{44. }%
+\CSLRightInline{Canaday C (1996)
+\href{https://doi.org/10.1016/0006-3207(95)00115-8}{Loss of
+insectivorous birds along a gradient of human impact in amazonia}.
+\emph{Biological Conservation} 77(1):63--77.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-sekercioglu2002}{}}%
+\CSLLeftMargin{45. }%
+\CSLRightInline{Şekercioḡlu ÇH, et al. (2002)
+\href{https://doi.org/10.1073/pnas.012616199}{Disappearance of
+insectivorous birds from tropical forest fragments}. \emph{Proceedings
+of the National Academy of Sciences} 99(1):263--267.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-tallamy2021}{}}%
+\CSLLeftMargin{46. }%
+\CSLRightInline{Tallamy DW, Shriver WG (2021)
+\href{https://doi.org/10.1093/ornithapp/duaa059}{Are declines in insects
+and insectivorous birds related?} \emph{Ornithological Applications}
+123(1):duaa059.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-cooke2019}{}}%
+\CSLLeftMargin{47. }%
+\CSLRightInline{Cooke RSC, Eigenbrod F, Bates AE (2019)
+\href{https://doi.org/10.1038/s41467-019-10284-z}{Projected losses of
+global mammal and bird ecological strategies}. \emph{Nature
+Communications} 10(1):1--8.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-bain2020}{}}%
+\CSLLeftMargin{48. }%
+\CSLRightInline{Bain GC, et al. (2020)
+\href{https://doi.org/10.1098/rsos.200076}{Changing bird communities of
+an agricultural landscape: Declines in arboreal foragers, increases in
+large species}. \emph{Royal Society Open Science} 7(3):200076.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-pollock2022}{}}%
+\CSLLeftMargin{49. }%
+\CSLRightInline{Pollock HS, et al. (2022)
+\href{https://doi.org/10.1073/pnas.2108731119}{Long-term monitoring
+reveals widespread and severe declines of understory birds in a
+protected neotropical forest}. \emph{Proceedings of the National Academy
+of Sciences} 119(16):e2108731119.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-fitzjohn2014}{}}%
-\CSLLeftMargin{37. }%
+\CSLLeftMargin{50. }%
 \CSLRightInline{FitzJohn RG, et al. (2014)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wilman2014}{}}%
-\CSLLeftMargin{38. }%
+\CSLLeftMargin{51. }%
 \CSLRightInline{Wilman H, et al. (2014)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-blowes2019}{}}%
-\CSLLeftMargin{39. }%
+\CSLLeftMargin{52. }%
 \CSLRightInline{Blowes SA, et al. (2019)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-norman2020}{}}%
-\CSLLeftMargin{40. }%
+\CSLLeftMargin{53. }%
 \CSLRightInline{Norman KEA, Chamberlain S, Boettiger C (2020)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2021}{}}%
-\CSLLeftMargin{41. }%
+\CSLLeftMargin{54. }%
 \CSLRightInline{R Core Team (2021) \emph{R: A language and environment
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lalibertuxe92010}{}}%
-\CSLLeftMargin{42. }%
+\CSLLeftMargin{55. }%
 \CSLRightInline{Laliberté E, Legendre P (2010)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-mason2005}{}}%
-\CSLLeftMargin{43. }%
+\CSLLeftMargin{56. }%
 \CSLRightInline{Mason NWH, Mouillot D, Lee WG, Wilson JB, Setälä H
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hillebrand2009}{}}%
-\CSLLeftMargin{44. }%
+\CSLLeftMargin{57. }%
 \CSLRightInline{Hillebrand H, Matthiessen B (2009)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-leps2006}{}}%
-\CSLLeftMargin{45. }%
+\CSLLeftMargin{58. }%
 \CSLRightInline{Leps J, Bello F, Lavorel S, Berman S (2006) Quantifying
@@
 \leavevmode\vadjust pre{\hypertarget{ref-schleuter2010}{}}%
-\CSLLeftMargin{46. }%
+\CSLLeftMargin{59. }%
 \CSLRightInline{Schleuter D, Daufresne M, Massol F, Argillier C (2010)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-blonder2018}{}}%
-\CSLLeftMargin{47. }%
+\CSLLeftMargin{60. }%
 \CSLRightInline{Blonder B (2018)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-swenson2012}{}}%
-\CSLLeftMargin{48. }%
+\CSLLeftMargin{61. }%
 \CSLRightInline{Swenson NG, et al. (2012)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lenth2022}{}}%
-\CSLLeftMargin{49. }%
+\CSLLeftMargin{62. }%
 \CSLRightInline{Lenth RV (2022) \emph{Emmeans: Estimated marginal means,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bates2015}{}}%
-\CSLLeftMargin{50. }%
+\CSLLeftMargin{63. }%
 \CSLRightInline{Bates D, Mächler M, Bolker B, Walker S (2015) Fitting
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kuznetsova2017}{}}%
-\CSLLeftMargin{51. }%
+\CSLLeftMargin{64. }%
 \CSLRightInline{Kuznetsova A, Brockhoff PB, Christensen RHB (2017)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rcoreteam2023}{}}%
-\CSLLeftMargin{52. }%
+\CSLLeftMargin{65. }%
 \CSLRightInline{R Core Team (2023) \emph{R: A language and environment
@@
 \leavevmode\vadjust pre{\hypertarget{ref-brodie2021}{}}%
-\CSLLeftMargin{53. }%
+\CSLLeftMargin{66. }%
 \CSLRightInline{Brodie JF, Williams S, Garner B (2021) The decline of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-cox2022}{}}%
-\CSLLeftMargin{54. }%
+\CSLLeftMargin{67. }%
 \CSLRightInline{Cox DTC, Gardner AS, Gaston KJ (2022)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-li2022}{}}%
-\CSLLeftMargin{55. }%
+\CSLLeftMargin{68. }%
 \CSLRightInline{Li X, et al. (2022)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-filstrup2019}{}}%
-\CSLLeftMargin{56. }%
+\CSLLeftMargin{69. }%
 \CSLRightInline{Filstrup CT, King KBS, McCullough IM (2019)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hillebrand2008}{}}%
-\CSLLeftMargin{57. }%
+\CSLLeftMargin{70. }%
 \CSLRightInline{Hillebrand H, Bennett DM, Cadotte MW (2008)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bagousse-pinguet2019}{}}%
-\CSLLeftMargin{58. }%
+\CSLLeftMargin{71. }%
 \CSLRightInline{Bagousse-Pinguet YL, et al. (2019)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-spatz2023}{}}%
-\CSLLeftMargin{59. }%
+\CSLLeftMargin{72. }%
 \CSLRightInline{Spatz DR, et al. (2023)
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-fritz2009}{}}%
+\CSLLeftMargin{73. }%
+\CSLRightInline{Fritz SA, Bininda-Emonds ORP, Purvis A (2009)
+\href{https://doi.org/10.1111/j.1461-0248.2009.01307.x}{Geographical
+variation in predictors of mammalian extinction risk: big is bad, but
+only in the tropics}. \emph{Ecology Letters} 12(6):538--549.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-ripple2015}{}}%
+\CSLLeftMargin{74. }%
+\CSLRightInline{Ripple WJ, et al. (2015)
+\href{https://doi.org/10.1126/sciadv.1400103}{Collapse of the world{'}s
+largest herbivores}. \emph{Science Advances} 1(4):e1400103.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-estes2011}{}}%
-\CSLLeftMargin{60. }%
+\CSLLeftMargin{75. }%
 \CSLRightInline{Estes JA, et al. (2011)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-reich2012}{}}%
-\CSLLeftMargin{61. }%
+\CSLLeftMargin{76. }%
 \CSLRightInline{Reich PB, et al. (2012)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-pimiento2020a}{}}%
-\CSLLeftMargin{62. }%
+\CSLLeftMargin{77. }%
 \CSLRightInline{Pimiento C, et al. (2020)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-biggs2020}{}}%
-\CSLLeftMargin{63. }%
+\CSLLeftMargin{78. }%
 \CSLRightInline{Biggs CR, et al. (2020)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-inger2015}{}}%
-\CSLLeftMargin{64. }%
+\CSLLeftMargin{79. }%
 \CSLRightInline{Inger R, et al. (2015)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rosenberg2019}{}}%
-\CSLLeftMargin{65. }%
+\CSLLeftMargin{80. }%
 \CSLRightInline{Rosenberg KV, et al. (2019) Decline of the North
@@
 \leavevmode\vadjust pre{\hypertarget{ref-schipper2016}{}}%
-\CSLLeftMargin{66. }%
+\CSLLeftMargin{81. }%
 \CSLRightInline{Schipper AM, et al. (2016)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-luck2013}{}}%
-\CSLLeftMargin{67. }%
+\CSLLeftMargin{82. }%
 \CSLRightInline{Luck GW, Carter A, Smallbone L (2013)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-magioli2021}{}}%
-\CSLLeftMargin{68. }%
+\CSLLeftMargin{83. }%
 \CSLRightInline{Magioli M, et al. (2021)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-tinoco2018}{}}%
-\CSLLeftMargin{69. }%
+\CSLLeftMargin{84. }%
 \CSLRightInline{Tinoco BA, Santillán VE, Graham CH (2018)
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sheridan2011}{}}%
-\CSLLeftMargin{70. }%
-\CSLRightInline{Sheridan JA, Bickford D (2011)
-\href{https://doi.org/10.1038/nclimate1259}{Shrinking body size as an
-ecological response to climate change}. \emph{Nature Climate Change}
-1(8):401--406.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-caruso2014}{}}%
-\CSLLeftMargin{71. }%
-\CSLRightInline{Caruso NM, Sears MW, Adams DC, Lips KR (2014)
-\href{https://doi.org/10.1111/gcb.12550}{Widespread rapid reductions in
-body size of adult salamanders in response to climate change}.
-\emph{Global Change Biology} 20(6):1751--1759.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-forster2012}{}}%
-\CSLLeftMargin{72. }%
-\CSLRightInline{Forster J, Hirst AG, Atkinson D (2012)
-\href{https://doi.org/10.1073/pnas.1210460109}{Warming-induced
-reductions in body size are greater in aquatic than terrestrial
-species}. \emph{Proceedings of the National Academy of Sciences}
-109(47):19310--19314.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-huss2019}{}}%
-\CSLLeftMargin{73. }%
-\CSLRightInline{Huss M, Lindmark M, Jacobson P, van Dorst RM, Gårdmark A
-(2019) \href{https://doi.org/10.1111/gcb.14637}{Experimental evidence of
-gradual size-dependent shifts in body size and growth of fish in
-response to warming}. \emph{Global Change Biology} 25(7):2285--2295.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tseng2018}{}}%
-\CSLLeftMargin{74. }%
-\CSLRightInline{Tseng M, et al. (2018)
-\href{https://doi.org/10.1111/1365-2656.12789}{Decreases in beetle body
-size linked to climate change and warming temperatures}. \emph{Journal
-of Animal Ecology} 87(3):647--659.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-fritz2009}{}}%
-\CSLLeftMargin{75. }%
-\CSLRightInline{Fritz SA, Bininda-Emonds ORP, Purvis A (2009)
-\href{https://doi.org/10.1111/j.1461-0248.2009.01307.x}{Geographical
-variation in predictors of mammalian extinction risk: big is bad, but
-only in the tropics}. \emph{Ecology Letters} 12(6):538--549.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-ripple2015}{}}%
-\CSLLeftMargin{76. }%
-\CSLRightInline{Ripple WJ, et al. (2015)
-\href{https://doi.org/10.1126/sciadv.1400103}{Collapse of the world{'}s
-largest herbivores}. \emph{Science Advances} 1(4):e1400103.}
-
 \leavevmode\vadjust pre{\hypertarget{ref-tilman2017}{}}%
-\CSLLeftMargin{77. }%
+\CSLLeftMargin{85. }%
 \CSLRightInline{Tilman D, et al. (2017)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-catford2022}{}}%
-\CSLLeftMargin{78. }%
+\CSLLeftMargin{86. }%
 \CSLRightInline{Catford JA, Wilson JRU, Pyšek P, Hulme PE, Duncan RP
@@
 \leavevmode\vadjust pre{\hypertarget{ref-spake2023}{}}%
-\CSLLeftMargin{79. }%
+\CSLLeftMargin{87. }%
 \CSLRightInline{Spake R, et al. (2023)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-pardieck2020}{}}%
-\CSLLeftMargin{80. }%
+\CSLLeftMargin{88. }%
 \CSLRightInline{Pardieck KL, David ZJ, Lutmerding M, Aponte V, Hudson
@@
 \leavevmode\vadjust pre{\hypertarget{ref-piropno}{}}%
-\CSLLeftMargin{81. }%
+\CSLLeftMargin{89. }%
 \CSLRightInline{PIROP northwest atlantic 1965{\textendash}1992 - OBIS
@@
 \leavevmode\vadjust pre{\hypertarget{ref-silva}{}}%
-\CSLLeftMargin{82. }%
+\CSLLeftMargin{90. }%
 \CSLRightInline{Silva FR da Brazil dataset 1.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-scott}{}}%
-\CSLLeftMargin{83. }%
+\CSLLeftMargin{91. }%
 \CSLRightInline{Scott D, Metts B, Lance S The rainbow bay long-term
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rossaferes1997}{}}%
-\CSLLeftMargin{84. }%
+\CSLLeftMargin{92. }%
 \CSLRightInline{Rossa-Feres D de C (1997) Community ecology of anura
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bakker1990}{}}%
-\CSLLeftMargin{85. }%
+\CSLLeftMargin{93. }%
 \CSLRightInline{Bakker C, Herman PMJ (1990) Phytoplankton in the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-berezovikovn.n.2004}{}}%
-\CSLLeftMargin{86. }%
+\CSLLeftMargin{94. }%
 \CSLRightInline{Berezovikov, N.N. (2004) The birds of settlements in
@@
 \leavevmode\vadjust pre{\hypertarget{ref-enemar2004}{}}%
-\CSLLeftMargin{87. }%
+\CSLLeftMargin{95. }%
 \CSLRightInline{Enemar A, Sjöstrand B, Andersson G, Von Proschwitz T
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gido2019}{}}%
-\CSLLeftMargin{88. }%
+\CSLLeftMargin{96. }%
 \CSLRightInline{Gido K (2019) CFP01 Fish population on selected
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hogstad1993}{}}%
-\CSLLeftMargin{89. }%
+\CSLLeftMargin{97. }%
 \CSLRightInline{Hogstad O (1993)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-holmes1986}{}}%
-\CSLLeftMargin{90. }%
+\CSLLeftMargin{98. }%
 \CSLRightInline{Holmes RT, Sherry TW, Sturges FW (1986)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-jahncke2006}{}}%
-\CSLLeftMargin{91. }%
+\CSLLeftMargin{99. }%
 \CSLRightInline{Jahncke J, Rintoul C (2006) CalCOFI and NMFS seabird and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-krivenkov.g.1991}{}}%
-\CSLLeftMargin{92. }%
+\CSLLeftMargin{100. }%
 \CSLRightInline{Krivenko, V.G. (1991) Waterfowl and their protection.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-melnikovy.i.2000}{}}%
-\CSLLeftMargin{93. }%
+\CSLLeftMargin{101. }%
 \CSLRightInline{Melnikov, Y.I., Melnikova, N., Pronkevich, V.V. (2000)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2013}{}}%
-\CSLLeftMargin{94. }%
+\CSLLeftMargin{102. }%
 \CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2014}{}}%
-\CSLLeftMargin{95. }%
+\CSLLeftMargin{103. }%
 \CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitori2014}{}}%
-\CSLLeftMargin{96. }%
+\CSLLeftMargin{104. }%
 \CSLRightInline{Monitoring site 1000 village survey - medium and large
@@
 \leavevmode\vadjust pre{\hypertarget{ref-preston1960}{}}%
-\CSLLeftMargin{97. }%
+\CSLLeftMargin{105. }%
 \CSLRightInline{Preston FW (1960)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-svensson2006}{}}%
-\CSLLeftMargin{98. }%
+\CSLLeftMargin{106. }%
 \CSLRightInline{Svensson S (2006) Species composition and population
@@
 \leavevmode\vadjust pre{\hypertarget{ref-svensson2010}{}}%
-\CSLLeftMargin{99. }%
+\CSLLeftMargin{107. }%
 \CSLRightInline{Svensson S, Thorner AM, Nyholm NEI (2010) Species
@@
 \leavevmode\vadjust pre{\hypertarget{ref-thorn2016changes}{}}%
-\CSLLeftMargin{100. }%
+\CSLLeftMargin{108. }%
 \CSLRightInline{Thorn S, et al. (2016)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-usfs}{}}%
-\CSLLeftMargin{101. }%
+\CSLLeftMargin{109. }%
 \CSLRightInline{USFS Landbird monitoring program (UMT-LBMP). Available
@@
 \leavevmode\vadjust pre{\hypertarget{ref-vermontcenterforecostudies2015}{}}%
-\CSLLeftMargin{102. }%
+\CSLLeftMargin{110. }%
 \CSLRightInline{Vermont Center For Ecostudies, Lambert JD, Hart J (2015)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-vickery1984}{}}%
-\CSLLeftMargin{103. }%
+\CSLLeftMargin{111. }%
 \CSLRightInline{Vickery WL, Nudds TD (1984)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-waide2017}{}}%
-\CSLLeftMargin{104. }%
+\CSLLeftMargin{112. }%
 \CSLRightInline{Waide R (2017) Bird abundance - point counts.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-williamson1983}{}}%
-\CSLLeftMargin{105. }%
+\CSLLeftMargin{113. }%
 \CSLRightInline{Williamson M (1983)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-zakharovv.d.1998}{}}%
-\CSLLeftMargin{106. }%
+\CSLLeftMargin{114. }%
 \CSLRightInline{Zakharov, V.D. (1998) Biodiversity of bird population of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-meditss2011}{}}%
-\CSLLeftMargin{107. }%
+\CSLLeftMargin{115. }%
 \CSLRightInline{MEDITS seabird surveys 1999 / 2000 / 2002 (2011)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-animald}{}}%
-\CSLLeftMargin{108. }%
+\CSLLeftMargin{116. }%
 \CSLRightInline{Animal demography unit - coordinated waterbird counts
@@
 \leavevmode\vadjust pre{\hypertarget{ref-baltics}{}}%
-\CSLLeftMargin{109. }%
+\CSLLeftMargin{117. }%
 \CSLRightInline{Baltic seabirds transect surveys Available at:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-thorn2016response}{}}%
-\CSLLeftMargin{110. }%
+\CSLLeftMargin{118. }%
 \CSLRightInline{Thorn S, et al. (2016)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ernest2009}{}}%
-\CSLLeftMargin{111. }%
+\CSLLeftMargin{119. }%
 \CSLRightInline{Ernest SKM, Valone TJ, Brown JH (2009)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-friggens2008}{}}%
-\CSLLeftMargin{112. }%
+\CSLLeftMargin{120. }%
 \CSLRightInline{Friggens M (2008) Sevilleta LTER small mammal population
@@
 \leavevmode\vadjust pre{\hypertarget{ref-jalilova.b.2014}{}}%
-\CSLLeftMargin{113. }%
+\CSLLeftMargin{121. }%
 \CSLRightInline{Jalilov, A. B., Andreychev, A. V., Kuznetsov, V. A.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kartzinel2014}{}}%
-\CSLLeftMargin{114. }%
+\CSLLeftMargin{122. }%
 \CSLRightInline{Kartzinel TR, et al. (2014)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kaufman2019}{}}%
-\CSLLeftMargin{115. }%
+\CSLLeftMargin{123. }%
 \CSLRightInline{Kaufman D (2019) CSM01 Seasonal Summary of Numbers of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-krefting1974}{}}%
-\CSLLeftMargin{116. }%
+\CSLLeftMargin{124. }%
 \CSLRightInline{Krefting LW, Ahlgren CE (1974)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lightfoot}{}}%
-\CSLLeftMargin{117. }%
+\CSLLeftMargin{125. }%
 \CSLRightInline{Lightfoot D, Schooley RL SMES rodent trapping data,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-malyshevy.s.2011}{}}%
-\CSLLeftMargin{118. }%
+\CSLLeftMargin{126. }%
 \CSLRightInline{Malyshev, Y. S. (2011) On the diagnostic techniques of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-meyer2008}{}}%
-\CSLLeftMargin{119. }%
+\CSLLeftMargin{127. }%
 \CSLRightInline{Meyer CFJ, Kalko EKV (2008)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nedosekinv.y.2015}{}}%
-\CSLLeftMargin{120. }%
+\CSLLeftMargin{128. }%
 \CSLRightInline{Nedosekin, V. Y. (2015) Long-term dynamics of the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-prins1990}{}}%
-\CSLLeftMargin{121. }%
+\CSLLeftMargin{129. }%
 \CSLRightInline{Prins HHT, Douglas-Hamilton I (1990)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rocha2017}{}}%
-\CSLLeftMargin{122. }%
+\CSLLeftMargin{130. }%
 \CSLRightInline{Rocha R, et al. (2017)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-stapp2014}{}}%
-\CSLLeftMargin{123. }%
+\CSLLeftMargin{131. }%
 \CSLRightInline{Stapp P (2014) SGS-LTER long-term monitoring project:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bahamas}{}}%
-\CSLLeftMargin{124. }%
+\CSLLeftMargin{132. }%
 \CSLRightInline{Bahamas marine mammal research organisation
@@
 \leavevmode\vadjust pre{\hypertarget{ref-popacet}{}}%
-\CSLLeftMargin{125. }%
+\CSLLeftMargin{133. }%
 \CSLRightInline{POPA cetacean, seabird, and sea turtle sightings in the

### 2024-04-16 paper/GEB Submission/manuscript_GEB.Rmd
more edits, hooray

@@
   Aim: Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing dissimilarity due to turnover but no overall trend in species diversity through time at the local scale.
-Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to mechanisms of ecosystem processes.
+Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to the characteristics that confer ecosystem processes.
 Here we present the first multi-taxa, multi-system analysis of functional change through time.
@@
 
-  Methods: We paired thousands of bird and mammal assemblage time series from the BioTIME database with existing trait data representative of a species' functional role to reconstruct time series of functional diversity metrics.
+  Methods: We paired thousands of bird and mammal assemblage time series from the BioTIME database with existing trait data representative of a species' functional role to reconstruct time series of functional diversity and composition metrics.
 Using generalized linear mixed models, we estimated general trends in those metrics and trends for individual studies.
 
-  Results: We found no overall trend in any functional diversity metric, which held even after correcting for changes in species richness.
+  Results: We found evidence of decreasing functional richeness for mammal communities and increasing functional evenness in marine communities, but otherwise no overall trends in functional diversity metrics, which held even after correcting for changes in species richness.
+With the exception of those two trends, study characteristics such as taxa, realm, biome, or protection status did not distinguish between types of change exhibited by communities.
 At the study-level, there was substantial heterogeneity in the direction and type of functional change exhibited.
-None of the included study characteristics such as taxa, realm, biome, or protection status were significantly related to the type of change exhibited by a community.
+We further did not find evidence to support multiple predictions for inidividual traits, including decreasing body size, dietary shifts, or changes in foraging strata.
 
   Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevalant than the other across many taxa, biomes, and realms.
+Loss of functional richness in mammal communities indicates that they are on average faring worse than bird communities.
 At the study-level, the majority group of studies showed no species or functional trends, but other types of change were found for multiple studies, including studies experiencing changes in functional redundancy, increased species and functional richness, or loss of functional richness.
@@
 
-We have stronger expectations for changes in the prevalence of some individual traits.
-For example, animal body size is expected to decrease as a result of climate change, a phenomena that has been documented in multiple taxa empirically and experimentally [@caruso2014; @forster2012; @huss2019; @sheridan2011; @tseng2018].
-There is some evidence that this holds true for birds [@vanbuskirk2010], but the picture is likely more complicated for mammals, where urbanization may actually lead to larger body sizes due to novel food sources [@hantak2021], even as megafaunal loss leads to on average smaller community body size.
-For dietary traits, recent work documenting insect declines (citation) points to potentially significant negative impacts on insectivorous animals [@canaday1996; @sekercioglu2002; @tallamy2021].
-Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species [@cooke2019].
-Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems [@bain2020], and loss of neotropical understory foragers even in protect areas [@pollock2022].
+`r colorize("We have stronger expectations for changes in the prevalence of some individual traits.
+For example, animal body size is expected to decrease as a result of climate change, a phenomena that has been documented in multiple taxa empirically and experimentally", "blue")` [@caruso2014; @forster2012; @huss2019; @sheridan2011; @tseng2018].
+`r colorize("There is some evidence that this holds true for birds ", "blue")`[@vanbuskirk2010]`r colorize(", but the picture is likely more complicated for mammals, where urbanization may actually lead to larger body sizes due to novel food sources", "blue")` [@hantak2021]`r colorize(", even as megafaunal loss leads to on average smaller community body size.
+For dietary traits, recent work documenting insect declines (citation) points to potentially significant negative impacts on insectivorous animals", "blue")` [@canaday1996; @sekercioglu2002; @tallamy2021].
+`r colorize("Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species ", "blue")`[@cooke2019].
+`r colorize("Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems", "blue")` [@bain2020]`r colorize(", and loss of neotropical understory foragers even in protect areas", "blue")` [@pollock2022].
 
@@
 We assess thousands of mammal and bird functional diversity time series to determine whether or not `r colorize("the addition of functional trait data gives a clarifying picture of biodiversity change across communities.
-We present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination.
-We present results for three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.", "blue")` We further evaluate evidence for key predictions of individual trait prevalence from the literature, including changes in mean body size, dietary traits, and foraging strata.\`
+We present a few areas of change consensus across communities, and even more scenarios where functional trends are unexplained and warrant further theoretical and experimental examination.
+We assess functional diversity trends at three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.
+We further evaluate evidence of key predictions for individual traits using trait-level compositional trends, including changes in mean body size, dietary traits, and foraging strata.", "blue")`
 
@@
 `r colorize("Wholesale shifts in the trait space due to changes of trait means could occur even while the shape of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained.
-CWM's are therefore a way to assess whether or not that turnover is occurring and what the nature of the shift may be.
+CWM's are therefore a way to assess whether or not turnover is occurring and what the nature of the shift may be.
 Hereafter we refer to results for functional metrics in two groups: functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).", "blue")`
@@
 We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.8.2, @lenth2022].
-`r colorize("For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore refrained from interpreting results for levels where there were fewer than three studies.", "blue")` We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
+`r colorize("For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.", "blue")` We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
 All models in our analysis were fit using the *lme4* (v1.1-30) package in R (v4.2.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].
@@
 We did find a significant overall decrease in Jaccard similarity, indicating accumulating changes in species composition.
-Non-significant overall `r colorize("temporal", "blue")` trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies `r colorize("was not significantly different from zero", "blue")`\` (Table \ref{tab:resultsTab}).
+Non-significant overall `r colorize("temporal", "blue")` trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies `r colorize("was not significantly different from zero", "blue")` (Table \ref{tab:resultsTab}).
 `r colorize("Within-group", "blue")` trends for different taxa, biomes, realms, or protection statuses were also non-significant for richness and functional diversity metrics, with the exception of a significantly increasing trend for functional evenness of marine studies, and a significantly decreasing functional richness slope for mammal studies.
@@
 `r colorize("However we did not find trends outside the random expectation for species loss, in contrast to expectations for multiple continents based on IUCN risk categories", "blue")` [@brodie2021].
-`r colorize("In general, functional capacity of mammals showed greater declines that bird communities.", "blue")`
+`r colorize("In general, functional capacity of mammals showed greater declines than bird communities.", "blue")`
 
@@
 `r colorize("An increase in evenness is therefore a positive indicator for these communities, despite the fact that marine bird and mammal communities are some of the most threatened communities in the world.
-Still, this result may be a function of data limitations as the majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery, which is also indicated in recent species-based work ", "blue")`[@spatz2023].
+Still, this result may be a function of data limitations as the majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery also found in recent species-based work ", "blue")`[@spatz2023].
 
-Our results were also inconsistent with all predictions for changes in trait prevalence, including foraging strata, body size, and dietary traits.
-`r colorize("We found no changes in the prevalence of different foraging strategies, despite documented losses of understory birds in the neotropics and some evidence that agricultural incursion particularly threatens arboreal species.
-These shifts may therefore be the result of specific contexts, and not generalizable to bird communities across the globe.
-We found no evidence of general trends in the *CWM* for body mass for birds or mammals, indicating that either body size is not changing significantly due to climate change, opposing pressures such as urbanization are overshadowing climate change impacts, or more likely current shifts are happening at an intraspecific level and and have not yet propogated to the species losses that would be captured by our data.
+`r colorize("Our results were also inconsistent with all predictions for changes in trait prevalence, including foraging strata, body size, and dietary traits.
+We found no changes in the prevalence of different foraging strategies, despite documented losses of understory birds in the neotropics and some evidence that agricultural incursion particularly threatens arboreal species.
+Those shifts may therefore be the result of specific contexts and not generalizable to bird communities across the globe.
+We found no evidence of general trends in the CWM for body mass for birds or mammals, indicating that either body size is not changing significantly due to climate change, opposing pressures such as urbanization are overshadowing climate change impacts, or more likely current shifts are happening at an intraspecific level and and have not yet propogated to the species losses that would be captured by our data.
 Further, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].
@@
 
-Rather than declines in insectivory and increases in invertivory predicted by previous work, we instead found changes in the degree of consumption of fish, nectar, and vertebrates.
-`r colorize("Increases in fish consumption for bird and temperate studies may be the product of increased availability of easy fish from commercial fishing operations or the stocking of inland waterways for sport fishing.
+`r colorize("Rather than declines in insectivory and increases in invertivory predicted by previous work, we instead found changes in the degree of consumption of fish, nectar, and vertebrates.
+Increases in fish consumption for bird and temperate studies may be the product of increased availability of easy fish from commercial fishing operations or the stocking of inland waterways for sport fishing.
 However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch, making these results suprising.
-Decreases in nectar consumption for tropical studies is likely due to declines in bat species, as 4 of the 5 tropical studies are for mammal and bats were the only nectar consuming mammals sampled in our data.
+Decreases in nectar consumption for tropical studies is likely due to declines in bat species, as 4 of the 5 tropical studies are for mammals and bats were the only nectar consuming mammals sampled in our data.
 Interpretation of the increases for temperate studies is less clear, as temperate studies included many bird communities with a variety of nectar consumers.
@@
 `r colorize("At the study-level, our results run contrary to our hope for a clearer picture of biodiversity change through a functional lens.
-The lack of general trend in functional diversity metrics belies a huge range of positive and negative trends at the study-level.
-In order to simplify discussion, we will talk about the implications for ecosystem process and vulnerability for communities grouped by their concurrent change in corrected functional diversity metrics, species richness and turnover split into positive, negative or no trend.
+The lack of general trend in functional diversity metrics belies a huge range of positive and negative trends at the study-level in both species and functional metrics.
+For example, it appears at the aggregate level that species turnover does not translate to functional turnover based on a lack of trends in CWM's.
+However at the study-level, all studies experiencing species turnover are also experiencing significant turnover for multiple individual traits, there is simply no consistency across communities in the kind of turnover occurring to be manifest in general trends.", "blue")`
+
+`r colorize("In order to simplify discussion, we will talk about the implications for ecosystem process and vulnerability for communities grouped by their concurrent change in corrected functional diversity metrics, species richness and turnover split into positive, negative or no trend.
 While there are over 150 possible combinations of change direction (or no change) in the 5 metrics, we discuss here the six scenarios that occurred in more than one study: no change, species loss or gain only, loss of functional evenness, species richness loss with species turnover, and increase in species and functional richness accompanied by significant turnover.
@@
 `r colorize("Three of the groups fit under the broad umbrella of changes in redundancy.
-By definition, if a community gains or loses species while functional metrics are unchanged, those species represent an increase or decrease in redundancy, so we include communities exhibiting gains (2 studies) or losses (5 studies) in species richness with no change in functional metrics or turnover and species richness loss with significant turnover (2 studies) in this umbrella.
+By definition, if a community gains or loses species while functional diversity metrics are unchanged, those species represent an increase or decrease in redundancy, so we include communities exhibiting gains (2 studies) or losses (5 studies) in species richness with no change in functional metrics or turnover and species richness loss with significant turnover (2 studies) in this umbrella.
 For communities exhibiting loss of redundancy (declining species richness), ecological processes are likely being maintained, but capacity to respond to future stressors is reduced ", "blue")`[@reich2012].
@@
 These communities have a hopeful trajectory, as they are improving their functional capacity and potentially their functional redundancy, likely leading to robust future ecological function.
-Notably, the studies that exhibited species richness increases (in this group with increasing functional richness, as well as communities without functional richness increases) were exclusively from terrestrial, temperate (sometimes temperate/polar) bird communities and occurred in countries where there has been a significant investment in conservation over the last few decades (United States, Canada, Sweden).
+Notably, the studies that exhibited species richness increases were exclusively from terrestrial, temperate (sometimes temperate/polar) bird communities and occurred in countries where there has been a significant investment in conservation over the last few decades (United States, Canada, Sweden).
 Our results are consistent with other functional work in these regions, showing increases in species richness and functional diversity for North American breeding birds over the last few decades after a period of decline", "blue")` [@jarzyna2016] `r colorize(", and loss of common, functionally general species even as rare species are increasing in North America and Europe ", "blue")` [@inger2015; @rosenberg2019; @schipper2016].
@@
 While it is tempting to draw a one to one line between the degree of human impact and loss of functional integrity (in the form of functional redundancy and richness), that relationship is not borne out by our rough metric of human impact (protection status) or previous taxa-specific or single community work, which is largely mixed", "blue")` [@flynn2009; @luck2013; @magioli2021; @tinoco2018].
+
+Study-level results also illustrate the relationship between species turnover and functional turnover.
+For all studies exhibiting increasing dissimilarity through time (11 studies), there were multipe significant trend in functional composition (CWM's) and diversity metrics.
+However, there was little consensus in the types of shift, as indicated by lack of general trends for most metrics.
 
@@
 
-The addition of functional trait data to the biodiversity change conversation is yet another illustration of how critical context is for understanding ecological patterns [@catford2022; @spake2023].
-`r colorize("Here we assessed how key community characteristics such as biome, realm, taxa, and protection status may partition variation in the functional trends exhibited and found that, on the whole, these broad classifications did not seem to make the picture of change clearer.
+`r colorize("The addition of functional trait data to the biodiversity change conversation is yet another illustration of how critical context is for understanding ecological patterns [@catford2022; @spake2023].
+Here we assessed how key community characteristics such as biome, realm, taxa, and protection status may partition variation in the functional trends exhibited and found that, on the whole, these broad classifications did not seem to make the picture of change clearer.
 Further work linking specific ecological contexts to types of functional change will be critical for identifying new areas of concern, especially when those communities may not otherwise be showing significant changes in species richness.

### 2024-04-16 paper/GEB Submission/manuscript_GEB.tex
more edits, hooray

@@
 approach for understanding community composition by linking shifts in
-species identity to mechanisms of ecosystem processes. Here we present
-the first multi-taxa, multi-system analysis of functional change through
-time.
+species identity to the characteristics that confer ecosystem processes.
+Here we present the first multi-taxa, multi-system analysis of
+functional change through time.
 
@@
 species' functional role to reconstruct time series of functional
-diversity metrics. Using generalized linear mixed models, we estimated
-general trends in those metrics and trends for individual studies.
-
-Results: We found no overall trend in any functional diversity metric,
-which held even after correcting for changes in species richness. At the
-study-level, there was substantial heterogeneity in the direction and
-type of functional change exhibited. None of the included study
-characteristics such as taxa, realm, biome, or protection status were
-significantly related to the type of change exhibited by a community.
+diversity and composition metrics. Using generalized linear mixed
+models, we estimated general trends in those metrics and trends for
+individual studies.
+
+Results: We found evidence of decreasing functional richeness for mammal
+communities and increasing functional evenness in marine communities,
+but otherwise no overall trends in functional diversity metrics, which
+held even after correcting for changes in species richness. With the
+exception of those two trends, study characteristics such as taxa,
+realm, biome, or protection status did not distinguish between types of
+change exhibited by communities. At the study-level, there was
+substantial heterogeneity in the direction and type of functional change
+exhibited. We further did not find evidence to support multiple
+predictions for inidividual traits, including decreasing body size,
+dietary shifts, or changes in foraging strata.
 
@@
 of functional shift is not more prevalant than the other across many
-taxa, biomes, and realms. At the study-level, the majority group of
-studies showed no species or functional trends, but other types of
-change were found for multiple studies, including studies experiencing
-changes in functional redundancy, increased species and functional
-richness, or loss of functional richness. With no one prevailing
-scenario of change, it will be critical to link change scenarios to
-ecological context.
+taxa, biomes, and realms. Loss of functional richness in mammal
+communities indicates that they are on average faring worse than bird
+communities. At the study-level, the majority group of studies showed no
+species or functional trends, but other types of change were found for
+multiple studies, including studies experiencing changes in functional
+redundancy, increased species and functional richness, or loss of
+functional richness. With no one prevailing scenario of change, it will
+be critical to link change scenarios to ecological context.
 \end{abstract}
@@
 
-We have stronger expectations for changes in the prevalence of some
-individual traits. For example, animal body size is expected to decrease
-as a result of climate change, a phenomena that has been documented in
-multiple taxa empirically and experimentally (37--41). There is some
-evidence that this holds true for birds (42), but the picture is likely
-more complicated for mammals, where urbanization may actually lead to
-larger body sizes due to novel food sources (43), even as megafaunal
-loss leads to on average smaller community body size. For dietary
-traits, recent work documenting insect declines (citation) points to
-potentially significant negative impacts on insectivorous animals
-(44--46). Predicted extinctions based on species-level vulnerability
-point to further dietary shifts, favoring increases in invertivorous
-species (47). Some systems also show significant shifts in the
-prevalence of different kinds of foragers in birds, for example loss of
-arboreal foragers in agricultural systems (48), and loss of neotropical
-understory foragers even in protect areas (49).
+\textcolor{blue}{We have stronger expectations for changes in the prevalence of some individual traits.
+For example, animal body size is expected to decrease as a result of climate change, a phenomena that has been documented in multiple taxa empirically and experimentally}
+(37--41).
+\textcolor{blue}{There is some evidence that this holds true for birds }(42)\textcolor{blue}{, but the picture is likely more complicated for mammals, where urbanization may actually lead to larger body sizes due to novel food sources}
+(43)\textcolor{blue}{, even as megafaunal loss leads to on average smaller community body size.
+For dietary traits, recent work documenting insect declines (citation) points to potentially significant negative impacts on insectivorous animals}
+(44--46).
+\textcolor{blue}{Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species }(47).
+\textcolor{blue}{Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems}
+(48)\textcolor{blue}{, and loss of neotropical understory foragers even in protect areas}
+(49).
 
@@
 \textcolor{blue}{the addition of functional trait data gives a clarifying picture of biodiversity change across communities.
-We present a few areas of change consensus across communities, and even more scenarios where the relationship between functional and species change are unexplained and warrant further theoretical and experimental examination.
-We present results for three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.}
-We further evaluate evidence for key predictions of individual trait
-prevalence from the literature, including changes in mean body size,
-dietary traits, and foraging strata.`
+We present a few areas of change consensus across communities, and even more scenarios where functional trends are unexplained and warrant further theoretical and experimental examination.
+We assess functional diversity trends at three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.
+We further evaluate evidence of key predictions for individual traits using trait-level compositional trends, including changes in mean body size, dietary traits, and foraging strata.}
 
@@
 \textcolor{blue}{Wholesale shifts in the trait space due to changes of trait means could occur even while the shape of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained.
-CWM's are therefore a way to assess whether or not that turnover is occurring and what the nature of the shift may be.
+CWM's are therefore a way to assess whether or not turnover is occurring and what the nature of the shift may be.
 Hereafter we refer to results for functional metrics in two groups: functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).}
@@
 trends, the average trend across studies
-\textcolor{blue}{was not significantly different from zero}` (Table
+\textcolor{blue}{was not significantly different from zero} (Table
 \ref{tab:resultsTab}). \textcolor{blue}{Within-group} trends for
@@
 (66).
-\textcolor{blue}{In general, functional capacity of mammals showed greater declines that bird communities.}
+\textcolor{blue}{In general, functional capacity of mammals showed greater declines than bird communities.}
 
@@
 
-\textcolor{blue}{The addition of functional trait data illuminates a few key areas where a synthesis approach is incongruous with system-specific studies or simply unexpected.
+\textcolor{blue}{The addition of functional trait data illuminated multiple areas where a synthesis approach is incongruous with system-specific studies or simply unexpected.
 For example, results for protected areas were surprising given the assumption that protection insulates communities from functional degradation.
@@
 \textcolor{blue}{An increase in evenness is therefore a positive indicator for these communities, despite the fact that marine bird and mammal communities are some of the most threatened communities in the world.
-Still, this result may be a function of data limitations as the majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery, which is also indicated in recent species-based work }(72).
-
-Our results were also inconsistent with all predictions for changes in
-trait prevalence, including foraging strata, body size, and dietary
-traits.
-\textcolor{blue}{We found no changes in the prevalence of different foraging strategies, despite documented losses of understory birds in the neotropics and some evidence that agricultural incursion particularly threatens arboreal species.
-These shifts may therefore be the result of specific contexts, and not generalizable to bird communities across the globe.
-We found no evidence of general trends in the *CWM* for body mass for birds or mammals, indicating that either body size is not changing significantly due to climate change, opposing pressures such as urbanization are overshadowing climate change impacts, or more likely current shifts are happening at an intraspecific level and and have not yet propogated to the species losses that would be captured by our data.
+Still, this result may be a function of data limitations as the majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery also found in recent species-based work }(72).
+
+\textcolor{blue}{Our results were also inconsistent with all predictions for changes in trait prevalence, including foraging strata, body size, and dietary traits.
+We found no changes in the prevalence of different foraging strategies, despite documented losses of understory birds in the neotropics and some evidence that agricultural incursion particularly threatens arboreal species.
+Those shifts may therefore be the result of specific contexts and not generalizable to bird communities across the globe.
+We found no evidence of general trends in the CWM for body mass for birds or mammals, indicating that either body size is not changing significantly due to climate change, opposing pressures such as urbanization are overshadowing climate change impacts, or more likely current shifts are happening at an intraspecific level and and have not yet propogated to the species losses that would be captured by our data.
 Further, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing}
@@
 
-Rather than declines in insectivory and increases in invertivory
-predicted by previous work, we instead found changes in the degree of
-consumption of fish, nectar, and vertebrates.
-\textcolor{blue}{Increases in fish consumption for bird and temperate studies may be the product of increased availability of easy fish from commercial fishing operations as fish are processed or stocking of inland waterways for sport fishing.
-However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch, which makes these results suprising.
-Decreases in nectar consumption for tropical studies is likely due to declines in bat species, as 4 of the 5 tropical studies are for mammal and bats were the only nectar consuming mammals sampled in our data.
+\textcolor{blue}{Rather than declines in insectivory and increases in invertivory predicted by previous work, we instead found changes in the degree of consumption of fish, nectar, and vertebrates.
+Increases in fish consumption for bird and temperate studies may be the product of increased availability of easy fish from commercial fishing operations or the stocking of inland waterways for sport fishing.
+However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch, making these results suprising.
+Decreases in nectar consumption for tropical studies is likely due to declines in bat species, as 4 of the 5 tropical studies are for mammals and bats were the only nectar consuming mammals sampled in our data.
 Interpretation of the increases for temperate studies is less clear, as temperate studies included many bird communities with a variety of nectar consumers.
@@
 \textcolor{blue}{At the study-level, our results run contrary to our hope for a clearer picture of biodiversity change through a functional lens.
-The lack of general trend in functional diversity metrics belies a huge range of positive and negative trends at the study-level.
-In order to simplify discussion, we will talk about the implications for ecosystem process and vulnerability for communities grouped by their concurrent change in corrected functional diversity metrics, species richness and turnover split into positive, negative or no trend.
+The lack of general trend in functional diversity metrics belies a huge range of positive and negative trends at the study-level in both species and functional metrics.
+For example, it appears at the aggregate level that species turnover does not translate to functional turnover based on a lack of trends in CWM's.
+However at the study-level, all studies experiencing species turnover are also experiencing significant turnover for multiple individual traits, there is simply no consistency across communities in the kind of turnover occurring to be manifest in general trends.}
+
+\textcolor{blue}{In order to simplify discussion, we will talk about the implications for ecosystem process and vulnerability for communities grouped by their concurrent change in corrected functional diversity metrics, species richness and turnover split into positive, negative or no trend.
 While there are over 150 possible combinations of change direction (or no change) in the 5 metrics, we discuss here the six scenarios that occurred in more than one study: no change, species loss or gain only, loss of functional evenness, species richness loss with species turnover, and increase in species and functional richness accompanied by significant turnover.
@@
 \textcolor{blue}{Three of the groups fit under the broad umbrella of changes in redundancy.
-By definition, if a community gains or loses species while functional metrics are unchanged, those species represent an increase or decrease in redundancy, so we include communities exhibiting gains (2 studies) or losses (5 studies) in species richness with no change in functional metrics or turnover and species richness loss with significant turnover (2 studies) in this umbrella.
+By definition, if a community gains or loses species while functional diversity metrics are unchanged, those species represent an increase or decrease in redundancy, so we include communities exhibiting gains (2 studies) or losses (5 studies) in species richness with no change in functional metrics or turnover and species richness loss with significant turnover (2 studies) in this umbrella.
 For communities exhibiting loss of redundancy (declining species richness), ecological processes are likely being maintained, but capacity to respond to future stressors is reduced }(76).
@@
 These communities have a hopeful trajectory, as they are improving their functional capacity and potentially their functional redundancy, likely leading to robust future ecological function.
-Notably, the studies that exhibited species richness increases (in this group with increasing functional richness, as well as communities without functional richness increases) were exclusively from terrestrial, temperate (sometimes temperate/polar) bird communities and occurred in countries where there has been a significant investment in conservation over the last few decades (United States, Canada, Sweden).
+Notably, the studies that exhibited species richness increases were exclusively from terrestrial, temperate (sometimes temperate/polar) bird communities and occurred in countries where there has been a significant investment in conservation over the last few decades (United States, Canada, Sweden).
 Our results are consistent with other functional work in these regions, showing increases in species richness and functional diversity for North American breeding birds over the last few decades after a period of decline}
@@
 
+Study-level results also illustrate the relationship between species
+turnover and functional turnover. For all studies exhibiting increasing
+dissimilarity through time (11 studies), there were multipe significant
+trend in functional composition (CWM's) and diversity metrics. However,
+there was little consensus in the types of shift, as indicated by lack
+of general trends for most metrics.
+
 \hypertarget{policy-implications}{%
@@
 
-The addition of functional trait data to the biodiversity change
-conversation is yet another illustration of how critical context is for
-understanding ecological patterns (86, 87).
-\textcolor{blue}{Here we assessed how key community characteristics such as biome, realm, taxa, and protection status may partition variation in the functional trends exhibited and found that, on the whole, these broad classifications did not seem to make the picture of change clearer.
+\textcolor{blue}{The addition of functional trait data to the biodiversity change conversation is yet another illustration of how critical context is for understanding ecological patterns [@catford2022; @spake2023].
+Here we assessed how key community characteristics such as biome, realm, taxa, and protection status may partition variation in the functional trends exhibited and found that, on the whole, these broad classifications did not seem to make the picture of change clearer.
 Further work linking specific ecological contexts to types of functional change will be critical for identifying new areas of concern, especially when those communities may not otherwise be showing significant changes in species richness.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-catford2022}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-pardieck2020}{}}%
 \CSLLeftMargin{86. }%
-\CSLRightInline{Catford JA, Wilson JRU, Pyšek P, Hulme PE, Duncan RP
-(2022) \href{https://doi.org/10.1016/j.tree.2021.09.007}{Addressing
-context dependence in ecology}. \emph{Trends in Ecology \& Evolution}
-37(2):158--170.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-spake2023}{}}%
-\CSLLeftMargin{87. }%
-\CSLRightInline{Spake R, et al. (2023)
-\href{https://doi.org/10.1111/brv.12939}{Understanding {`}it depends{'}
-in ecology: a guide to hypothesising, visualising and interpreting
-statistical interactions}. \emph{Biological Reviews} 98(4):983--1002.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-pardieck2020}{}}%
-\CSLLeftMargin{88. }%
 \CSLRightInline{Pardieck KL, David ZJ, Lutmerding M, Aponte V, Hudson
@@
 \leavevmode\vadjust pre{\hypertarget{ref-piropno}{}}%
-\CSLLeftMargin{89. }%
+\CSLLeftMargin{87. }%
 \CSLRightInline{PIROP northwest atlantic 1965{\textendash}1992 - OBIS
@@
 \leavevmode\vadjust pre{\hypertarget{ref-silva}{}}%
-\CSLLeftMargin{90. }%
+\CSLLeftMargin{88. }%
 \CSLRightInline{Silva FR da Brazil dataset 1.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-scott}{}}%
-\CSLLeftMargin{91. }%
+\CSLLeftMargin{89. }%
 \CSLRightInline{Scott D, Metts B, Lance S The rainbow bay long-term
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rossaferes1997}{}}%
-\CSLLeftMargin{92. }%
+\CSLLeftMargin{90. }%
 \CSLRightInline{Rossa-Feres D de C (1997) Community ecology of anura
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bakker1990}{}}%
-\CSLLeftMargin{93. }%
+\CSLLeftMargin{91. }%
 \CSLRightInline{Bakker C, Herman PMJ (1990) Phytoplankton in the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-berezovikovn.n.2004}{}}%
-\CSLLeftMargin{94. }%
+\CSLLeftMargin{92. }%
 \CSLRightInline{Berezovikov, N.N. (2004) The birds of settlements in
@@
 \leavevmode\vadjust pre{\hypertarget{ref-enemar2004}{}}%
-\CSLLeftMargin{95. }%
+\CSLLeftMargin{93. }%
 \CSLRightInline{Enemar A, Sjöstrand B, Andersson G, Von Proschwitz T
@@
 \leavevmode\vadjust pre{\hypertarget{ref-gido2019}{}}%
-\CSLLeftMargin{96. }%
+\CSLLeftMargin{94. }%
 \CSLRightInline{Gido K (2019) CFP01 Fish population on selected
@@
 \leavevmode\vadjust pre{\hypertarget{ref-hogstad1993}{}}%
-\CSLLeftMargin{97. }%
+\CSLLeftMargin{95. }%
 \CSLRightInline{Hogstad O (1993)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-holmes1986}{}}%
-\CSLLeftMargin{98. }%
+\CSLLeftMargin{96. }%
 \CSLRightInline{Holmes RT, Sherry TW, Sturges FW (1986)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-jahncke2006}{}}%
-\CSLLeftMargin{99. }%
+\CSLLeftMargin{97. }%
 \CSLRightInline{Jahncke J, Rintoul C (2006) CalCOFI and NMFS seabird and
@@
 \leavevmode\vadjust pre{\hypertarget{ref-krivenkov.g.1991}{}}%
-\CSLLeftMargin{100. }%
+\CSLLeftMargin{98. }%
 \CSLRightInline{Krivenko, V.G. (1991) Waterfowl and their protection.}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-melnikovy.i.2000}{}}%
-\CSLLeftMargin{101. }%
+\CSLLeftMargin{99. }%
 \CSLRightInline{Melnikov, Y.I., Melnikova, N., Pronkevich, V.V. (2000)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2013}{}}%
-\CSLLeftMargin{102. }%
+\CSLLeftMargin{100. }%
 \CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitoringsite1000projectbiodiversitycenterministryofenvironmentofjapan2014}{}}%
-\CSLLeftMargin{103. }%
+\CSLLeftMargin{101. }%
 \CSLRightInline{Monitoring Site 1000 Project, Biodiversity Center,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-monitori2014}{}}%
-\CSLLeftMargin{104. }%
+\CSLLeftMargin{102. }%
 \CSLRightInline{Monitoring site 1000 village survey - medium and large
@@
 \leavevmode\vadjust pre{\hypertarget{ref-preston1960}{}}%
-\CSLLeftMargin{105. }%
+\CSLLeftMargin{103. }%
 \CSLRightInline{Preston FW (1960)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-svensson2006}{}}%
-\CSLLeftMargin{106. }%
+\CSLLeftMargin{104. }%
 \CSLRightInline{Svensson S (2006) Species composition and population
@@
 \leavevmode\vadjust pre{\hypertarget{ref-svensson2010}{}}%
-\CSLLeftMargin{107. }%
+\CSLLeftMargin{105. }%
 \CSLRightInline{Svensson S, Thorner AM, Nyholm NEI (2010) Species
@@
 \leavevmode\vadjust pre{\hypertarget{ref-thorn2016changes}{}}%
-\CSLLeftMargin{108. }%
+\CSLLeftMargin{106. }%
 \CSLRightInline{Thorn S, et al. (2016)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-usfs}{}}%
-\CSLLeftMargin{109. }%
+\CSLLeftMargin{107. }%
 \CSLRightInline{USFS Landbird monitoring program (UMT-LBMP). Available
@@
 \leavevmode\vadjust pre{\hypertarget{ref-vermontcenterforecostudies2015}{}}%
-\CSLLeftMargin{110. }%
+\CSLLeftMargin{108. }%
 \CSLRightInline{Vermont Center For Ecostudies, Lambert JD, Hart J (2015)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-vickery1984}{}}%
-\CSLLeftMargin{111. }%
+\CSLLeftMargin{109. }%
 \CSLRightInline{Vickery WL, Nudds TD (1984)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-waide2017}{}}%
-\CSLLeftMargin{112. }%
+\CSLLeftMargin{110. }%
 \CSLRightInline{Waide R (2017) Bird abundance - point counts.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-williamson1983}{}}%
-\CSLLeftMargin{113. }%
+\CSLLeftMargin{111. }%
 \CSLRightInline{Williamson M (1983)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-zakharovv.d.1998}{}}%
-\CSLLeftMargin{114. }%
+\CSLLeftMargin{112. }%
 \CSLRightInline{Zakharov, V.D. (1998) Biodiversity of bird population of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-meditss2011}{}}%
-\CSLLeftMargin{115. }%
+\CSLLeftMargin{113. }%
 \CSLRightInline{MEDITS seabird surveys 1999 / 2000 / 2002 (2011)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-animald}{}}%
-\CSLLeftMargin{116. }%
+\CSLLeftMargin{114. }%
 \CSLRightInline{Animal demography unit - coordinated waterbird counts
@@
 \leavevmode\vadjust pre{\hypertarget{ref-baltics}{}}%
-\CSLLeftMargin{117. }%
+\CSLLeftMargin{115. }%
 \CSLRightInline{Baltic seabirds transect surveys Available at:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-thorn2016response}{}}%
-\CSLLeftMargin{118. }%
+\CSLLeftMargin{116. }%
 \CSLRightInline{Thorn S, et al. (2016)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-ernest2009}{}}%
-\CSLLeftMargin{119. }%
+\CSLLeftMargin{117. }%
 \CSLRightInline{Ernest SKM, Valone TJ, Brown JH (2009)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-friggens2008}{}}%
-\CSLLeftMargin{120. }%
+\CSLLeftMargin{118. }%
 \CSLRightInline{Friggens M (2008) Sevilleta LTER small mammal population
@@
 \leavevmode\vadjust pre{\hypertarget{ref-jalilova.b.2014}{}}%
-\CSLLeftMargin{121. }%
+\CSLLeftMargin{119. }%
 \CSLRightInline{Jalilov, A. B., Andreychev, A. V., Kuznetsov, V. A.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kartzinel2014}{}}%
-\CSLLeftMargin{122. }%
+\CSLLeftMargin{120. }%
 \CSLRightInline{Kartzinel TR, et al. (2014)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-kaufman2019}{}}%
-\CSLLeftMargin{123. }%
+\CSLLeftMargin{121. }%
 \CSLRightInline{Kaufman D (2019) CSM01 Seasonal Summary of Numbers of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-krefting1974}{}}%
-\CSLLeftMargin{124. }%
+\CSLLeftMargin{122. }%
 \CSLRightInline{Krefting LW, Ahlgren CE (1974)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-lightfoot}{}}%
-\CSLLeftMargin{125. }%
+\CSLLeftMargin{123. }%
 \CSLRightInline{Lightfoot D, Schooley RL SMES rodent trapping data,
@@
 \leavevmode\vadjust pre{\hypertarget{ref-malyshevy.s.2011}{}}%
-\CSLLeftMargin{126. }%
+\CSLLeftMargin{124. }%
 \CSLRightInline{Malyshev, Y. S. (2011) On the diagnostic techniques of
@@
 \leavevmode\vadjust pre{\hypertarget{ref-meyer2008}{}}%
-\CSLLeftMargin{127. }%
+\CSLLeftMargin{125. }%
 \CSLRightInline{Meyer CFJ, Kalko EKV (2008)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-nedosekinv.y.2015}{}}%
-\CSLLeftMargin{128. }%
+\CSLLeftMargin{126. }%
 \CSLRightInline{Nedosekin, V. Y. (2015) Long-term dynamics of the
@@
 \leavevmode\vadjust pre{\hypertarget{ref-prins1990}{}}%
-\CSLLeftMargin{129. }%
+\CSLLeftMargin{127. }%
 \CSLRightInline{Prins HHT, Douglas-Hamilton I (1990)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rocha2017}{}}%
-\CSLLeftMargin{130. }%
+\CSLLeftMargin{128. }%
 \CSLRightInline{Rocha R, et al. (2017)
@@
 \leavevmode\vadjust pre{\hypertarget{ref-stapp2014}{}}%
-\CSLLeftMargin{131. }%
+\CSLLeftMargin{129. }%
 \CSLRightInline{Stapp P (2014) SGS-LTER long-term monitoring project:
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bahamas}{}}%
-\CSLLeftMargin{132. }%
+\CSLLeftMargin{130. }%
 \CSLRightInline{Bahamas marine mammal research organisation
@@
 \leavevmode\vadjust pre{\hypertarget{ref-popacet}{}}%
-\CSLLeftMargin{133. }%
+\CSLLeftMargin{131. }%
 \CSLRightInline{POPA cetacean, seabird, and sea turtle sightings in the

### 2024-04-22 analysis/05_model_metrics.Rmd
Add adjusted p-values

@@
 ```
+Perform p-value correction
+```{r}
+# Let's get a dataframe of p-values for the models we actually used
+
+# base models for all metrics
+
+base <- metric_model_table %>%
+  filter(model_id == "base", term == "year_scaled")
+
+# within group and between group trends for all metrics
+
+contrast <- metric_model_table %>%
+  filter(model_id != "base", effect == "contrast")
+
+group_slope <-  metric_model_table %>%
+  filter(model_id != "base", effect == "group_slope") %>%
+  mutate(tval = estimate/std.error,
+         p.value = 2*pt(-abs(tval), df))
+
+# study-level trends for species and functional diversity metrics
+
+study_trend <- indv_mod_table %>%
+  filter(term == "year_scaled")
+
+pval_df <- bind_rows(base, contrast, group_slope, study_trend) %>%
+  filter(metric != "Jaccard_next")
+
+adj.p <- p.adjust(pval_df$p.value, method = "BH")
+
+pval_df <- bind_cols(pval_df, p.adjust = adj.p)
+
+# # Which of the general model trends are likely spurious
+# pval_df %>% 
+#   filter(p.value < 0.05, p.adjust > 0.05) %>% 
+#   filter(is.na(study_id)) %>% 
+#   View()
+# 
+# # Which of the study-level trends are likely spurious
+# pval_df %>% 
+#   filter(p.value < 0.05, p.adjust > 0.05) %>% 
+#   filter(!is.na(study_id)) %>% 
+#   View()
+
+# add adjusted p value to model table
+
+model_output <- bind_rows(metric_model_table, indv_mod_table) %>%
+  left_join(pval_df) %>%
+  select(-significant, -tval)
+
+usethis::use_data(model_output)
+```
 
@@
 #supplement of complete model output
-bind_rows(metric_model_table, indv_mod_table) %>%
-    select(-c(n_ts, warning, lower.CL, upper.CL, significant, REML, bobyqa)) %>%
+model_output %>%
+    select(-c(n_ts, warning, lower.CL, upper.CL, REML, bobyqa)) %>%
   readr::write_csv(here::here("paper", "model_estimates_supp.csv"))
@@
 #First - how many studies had a significant trend for each metric? 
-indv_mod_table %>% 
+model_output %>% 
+  filter(!is.na(study_id)) %>%
   filter(term == "year_scaled", p.value < 0.05) %>% 
@@
 #And which studies were they?
-indv_mod_table %>% 
+model_output %>% 
+  filter(!is.na(study_id)) %>%
   filter(term == "year_scaled", p.value < 0.05) %>% 
@@
 #Which were significant for multiple metrics?
-indv_mod_table %>% 
+model_output %>% 
+  filter(!is.na(study_id)) %>%
   filter(term == "year_scaled", p.value < 0.05) %>% 

### 2024-04-22 analysis/figures/06_study_change.Rmd
Look at study locations.
Add adjusted p-values

@@
 ```{r}
-load(here::here("data/indv_mod_table.rda"))
+load(here::here("data/model_output.rda"))
 load(here::here("data/model_data.rda"))
@@
 ```{r}
-wide_table <- indv_mod_table %>% 
-  filter(p.value < 0.05, term == "year_scaled", model_id == "base") %>% 
+wide_table <- model_output %>%
+  filter(!is.na(study_id)) %>%
+  filter(p.adjust < 0.05, term == "year_scaled", model_id == "base") %>% 
   dplyr::select(study_id, metric, estimate) %>% 
@@
   dplyr::select(., -starts_with("CWM_")) %>% 
-  dplyr::select(study_id, S, Jaccard_base, everything(), -Jaccard_next) %>%
+  dplyr::select(study_id, S, Jaccard_base, everything()) %>%
   left_join(model_data %>% 
@@
 
+study_meta <- meta_clean %>% 
+  select(-rarefyid) %>%
+  distinct()
+
+#remove NA entries if another country was assigned
+na_studies <- study_meta %>% 
+  group_by(study_id) %>% 
+  filter(n() > 1, is.na(country))
+
+study_meta <- setdiff(study_meta, na_studies)
+
+# change_groups <- wide_table %>% 
+#   select(study_id, S, Jaccard_base, starts_with("SES")) %>%
+#   mutate(across(!study_id, ~case_when(
+#     . > 0 ~ "pos",
+#     . < 0 ~ "neg",
+#     .default = "0"
+#   )))
+
 change_groups <- wide_table %>% 
-  select(study_id, S, Jaccard_base, starts_with("SES")) %>%
+  select(study_id, S, Jaccard_base, FRic, FEve, FDiv) %>%
   mutate(across(!study_id, ~case_when(

### 2024-04-22 paper/GEB Submission/manuscript_GEB.Rmd
Add adjusted p-values

@@
   Aim: Despite unprecedented environmental change due to anthropogenic pressure, recent work has found increasing dissimilarity due to turnover but no overall trend in species diversity through time at the local scale.
-Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to the characteristics that confer ecosystem processes.
+Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to the characteristics that underpin ecosystem processes.
 Here we present the first multi-taxa, multi-system analysis of functional change through time.
@@
 
-  Methods: We paired thousands of bird and mammal assemblage time series from the BioTIME database with existing trait data representative of a species' functional role to reconstruct time series of functional diversity and composition metrics.
+  Methods: We paired thousands of bird and mammal assemblage time series from the BioTIME database with existing trait data representative of species' functional roles to reconstruct time series of functional diversity and composition metrics.
 Using generalized linear mixed models, we estimated general trends in those metrics and trends for individual studies.
 
-  Results: We found evidence of decreasing functional richeness for mammal communities and increasing functional evenness in marine communities, but otherwise no overall trends in functional diversity metrics, which held even after correcting for changes in species richness.
-With the exception of those two trends, study characteristics such as taxa, realm, biome, or protection status did not distinguish between types of change exhibited by communities.
+  Results: We found no overall trends in functional diversity metrics.
+Study characteristics such as taxa, realm, biome, or protection status did not distinguish between types of change exhibited by communities.
 At the study-level, there was substantial heterogeneity in the direction and type of functional change exhibited.
-We further did not find evidence to support multiple predictions for inidividual traits, including decreasing body size, dietary shifts, or changes in foraging strata.
+We further did not find evidence to support multiple predictions for specific traits, including decreasing body size, dietary shifts, or changes in foraging strata.
 
   Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevalant than the other across many taxa, biomes, and realms.
-Loss of functional richness in mammal communities indicates that they are on average faring worse than bird communities.
-At the study-level, the majority group of studies showed no species or functional trends, but other types of change were found for multiple studies, including studies experiencing changes in functional redundancy, increased species and functional richness, or loss of functional richness.
-With no one prevailing scenario of change, it will be critical to link change scenarios to ecological context.
+At the study-level, the majority of studies showed no species or functional trends, but other types of change were found for multiple studies, including changes in functional redundancy, increased species and functional richness, or loss of functional richness.
+With no prevailing scenario of change, it will be critical to link types of change to ecological context.
 
@@
 
-Ecological communities are experiencing unprecedented change as a result of anthropogenic pressures such as climate change, land use change, and invasive species.
-Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011], and fundamental changes in some of the most well-studied systems [e.g.
+Ecological communities are experiencing unprecedented changes as a result of anthropogenic pressures such as climate change, land use change, and invasive species.
+Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011] and fundamental changes in some of the most well-studied systems [e.g.
 coral bleaching, @sully2019].
@@
 
-`r colorize("The expectation for functional diversity change across communities is not obvious from past work, and may or may not follow taxonomic trends", "blue")` [@dornelas2023].
+`r colorize("The expectation for functional diversity change across communities is not obvious from past work, and may or may not follow species richness trends", "blue")` [@dornelas2023].
 While `r colorize("loss of functional diversity", "blue")` is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], `r colorize("functional diversity may be maintained even when species are lost from a community", "blue")` [@daz2001; @petchey2009].
@@
 For example, animal body size is expected to decrease as a result of climate change, a phenomena that has been documented in multiple taxa empirically and experimentally", "blue")` [@caruso2014; @forster2012; @huss2019; @sheridan2011; @tseng2018].
-`r colorize("There is some evidence that this holds true for birds ", "blue")`[@vanbuskirk2010]`r colorize(", but the picture is likely more complicated for mammals, where urbanization may actually lead to larger body sizes due to novel food sources", "blue")` [@hantak2021]`r colorize(", even as megafaunal loss leads to on average smaller community body size.
-For dietary traits, recent work documenting insect declines (citation) points to potentially significant negative impacts on insectivorous animals", "blue")` [@canaday1996; @sekercioglu2002; @tallamy2021].
+`r colorize("There is some evidence that this holds true for birds ", "blue")`[@vanbuskirk2010]`r colorize(", but the picture is likely more complicated for mammals, where urbanization may actually lead to larger body sizes due to novel food sources", "blue")` [@hantak2021]`` r colorize(", even as megafaunal loss leads to on average smaller community body size.
+For dietary traits, recent work documenting insect declines", "blue")` ``[@wagner2021]`` `r colorize("points to potentially significant negative impacts on insectivorous animals", "blue") `` [@canaday1996; @sekercioglu2002; @tallamy2021].
 `r colorize("Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species ", "blue")`[@cooke2019].
@@
 Non-significant overall `r colorize("temporal", "blue")` trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies `r colorize("was not significantly different from zero", "blue")` (Table \ref{tab:resultsTab}).
-`r colorize("Within-group", "blue")` trends for different taxa, biomes, realms, or protection statuses were also non-significant for richness and functional diversity metrics, with the exception of a significantly increasing trend for functional evenness of marine studies, and a significantly decreasing functional richness slope for mammal studies.
-However, trends were not exhibited for the `r colorize("corrected", "blue")` metric indicating that differences `r colorize("in functional diversity metrics", "blue")` were largely due to changes in species richness.
+`r colorize("Within-group", "blue")` trends for different taxa, biomes, realms, or protection statuses were also non-significant for richness and functional diversity metrics.
 The general trends for *CWM* models were similarly not significant, with a significant positive trend for only percentage of fish in diet composition.
@@
 We did find significant differences between taxa, realms, and biomes for Jaccard similarity and some of the *CWM's*.
-For example, while Jaccard similarity was decreasing in the general trend and there were significant within group slopes for Mammals, Birds, Terrestrial, Temperate, and Tropical communities, there was no significant slope for the marine realm, indicating that the general trend is mostly driven by turnover in terrestrial communities.
+For example, while Jaccard similarity was decreasing in the general trend and there were significant within group slopes for birds, terrestrial, and temperate studies, there was no significant slope for marine, mammal, or tropical studies.
 We also found a significant decrease in Jaccard similarity in unprotected areas only, with no trend for protected areas.
-We found significant dietary shifts across communities, with a significant general trend of increasing fish consumption, which was also reflected in increasing fish consumption trends for Bird studies and Temperate studies.
-We found a significantly increasing trend in nectar consumption for Temperate studies but a significantly decreasing tend for Tropical studies.
+We found significant dietary shifts across communities, with a significant general trend of increasing fish consumption, which was also reflected in increasing fish consumption trends for bird studies and temperate studies.
 We found significantly decreasing trend for vertebrate consumption in Marine studies and Tropical studies.
+The trend for marine vertebrate consumption was significantly more negative than for terrestrial studies and the temperate trend was significantly more negative than tropical studies.
 
-At the study level, 16 of 50 studies exhibited a significant trend in species richness and 11 exhibited significant turnover.
-For uncorrected functional diversity metrics, 25 of 50 studies exhibited a trend in a least one metric, and 15 of 50 studies exhibited a significant trend for at least one `r colorize("corrected", "blue")` metric (Table \ref{tab:trendTab}).
+At the study level, 12 of 50 studies exhibited a significant trend in species richness and 11 exhibited significant turnover.
+For uncorrected functional diversity metrics, 11 of 50 studies exhibited a trend in a least one metric, and 7 of 50 studies exhibited a significant trend for at least one `r colorize("corrected", "blue")` metric (Table \ref{tab:trendTab}).
 In general, there were more significant trends for uncorrected metrics, with some disappearing after correction, indicating that those trends were likely due to changes in the number of species.
-Hypothesis testing for study-level trends is likely affected by multiple testing issues and some trends identified as significant are therefore potentially spurious.
-Rather than interpreting changes in specific studies, we present these results as a general picture of the kinds of trends experienced by communities.
 
@@
 
-load(here::here("data/indv_mod_table.rda"))
+load(here::here("data/model_output.rda"))
 
-indv_mod_table %>%
-  filter(term == "year_scaled", p.value < 0.05) %>%
+model_output %>%
+  filter(term == "year_scaled", p.adjust < 0.05) %>%
   filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S", "FRic", "FEve", "FDiv")) %>%
@@
 As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing dissimilarity through time", "blue")` [@dornelas2018]`r colorize(", indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends.
-We found significant turnover for all biomes, realm, and taxa except for marine studies, which stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial", "blue")` [@blowes2019].
+We found significant turnover for most biomes, realm, and taxa with the exception of marine, tropical and mammal studies.
+The lack of turnover in marine studies stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial", "blue")` [@blowes2019].
 `r colorize("However, previous global estimates are dominated by fish communities which we exclude here and may be driving the overall trend while disguising relative stasis in marine bird and mammal communities.", "blue")`
 
-`r colorize("We found evidence of functional richness loss in mammal studies, with a loss trend significantly different from zero and significantly different from the trend for birds, indicating a potential loss of functional capacity in these communities.
+```{=html}
+<!--`r colorize("We found evidence of functional richness loss in mammal studies, with a loss trend significantly different from zero and significantly different from the trend for birds, indicating a potential loss of functional capacity in these communities.
 This result is consistent with previous work linking anthropogenic drivers such as habitat loss, poaching, and human modification and development to loss of functional diversity in mammal communities across the globe", "blue")` [@brodie2021; @cox2022; @li2022].
 `r colorize("However we did not find trends outside the random expectation for species loss, in contrast to expectations for multiple continents based on IUCN risk categories", "blue")` [@brodie2021].
-`r colorize("In general, functional capacity of mammals showed greater declines than bird communities.", "blue")`
-
+`r colorize("In general, functional capacity of mammals showed greater declines than bird communities.", "blue")`-->
+```
 ## Areas of incongruence
@@
 
-`r colorize("We also found evidence of increasing evenness for marine studies, which was not predicted by previous work in marine systems.
+```{=html}
+<!--`r colorize("We also found evidence of increasing evenness for marine studies, which was not predicted by previous work in marine systems.
 The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability ", "blue")` [@filstrup2019], `r colorize("as the dominance of one or a few species makes function sensitive to the ability of those species to respond to environmental change ", "blue")`[@hillebrand2008; @hillebrand2009].
@@
 `r colorize("An increase in evenness is therefore a positive indicator for these communities, despite the fact that marine bird and mammal communities are some of the most threatened communities in the world.
-Still, this result may be a function of data limitations as the majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery also found in recent species-based work ", "blue")`[@spatz2023].
-
+Still, this result may be a function of data limitations as the majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery also found in recent species-based work ", "blue")`[@spatz2023].-->
+```
 `r colorize("Our results were also inconsistent with all predictions for changes in trait prevalence, including foraging strata, body size, and dietary traits.
@@
 
-`r colorize("Rather than declines in insectivory and increases in invertivory predicted by previous work, we instead found changes in the degree of consumption of fish, nectar, and vertebrates.
+`r colorize("Rather than declines in insectivory and increases in invertivory predicted by previous work, we instead found changes in the degree of consumption of fish and vertebrates.
 Increases in fish consumption for bird and temperate studies may be the product of increased availability of easy fish from commercial fishing operations or the stocking of inland waterways for sport fishing.
 However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch, making these results suprising.
-Decreases in nectar consumption for tropical studies is likely due to declines in bat species, as 4 of the 5 tropical studies are for mammals and bats were the only nectar consuming mammals sampled in our data.
-Interpretation of the increases for temperate studies is less clear, as temperate studies included many bird communities with a variety of nectar consumers.
-Similarly, decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.", "blue")`
+Decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.", "blue")`
 

### 2024-04-22 paper/GEB Submission/manuscript_GEB.tex
Add adjusted p-values

@@
 contrasts using the \emph{emmeans} package (v1.8.2, 62).
-\textcolor{blue}{For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore refrained from interpreting results for levels where there were fewer than three studies.}
+\textcolor{blue}{For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.}
 We assessed the impact of time series duration and start year on
@@
 different taxa, biomes, realms, or protection statuses were also
-non-significant for richness and functional diversity metrics, with the
-exception of a significantly increasing trend for functional evenness of
-marine studies, and a significantly decreasing functional richness slope
-for mammal studies. However, trends were not exhibited for the
-\textcolor{blue}{corrected} metric indicating that differences
-\textcolor{blue}{in functional diversity metrics} were largely due to
-changes in species richness. The general trends for \emph{CWM} models
-were similarly not significant, with a significant positive trend for
-only percentage of fish in diet composition.
+non-significant for richness and functional diversity metrics. The
+general trends for \emph{CWM} models were similarly not significant,
+with a significant positive trend for only percentage of fish in diet
+composition.
 
@@
 Jaccard similarity was decreasing in the general trend and there were
-significant within group slopes for Mammals, Birds, Terrestrial,
-Temperate, and Tropical communities, there was no significant slope for
-the marine realm, indicating that the general trend is mostly driven by
-turnover in terrestrial communities. We also found a significant
-decrease in Jaccard similarity in unprotected areas only, with no trend
-for protected areas. We found significant dietary shifts across
-communities, with a significant general trend of increasing fish
-consumption, which was also reflected in increasing fish consumption
-trends for Bird studies and Temperate studies. We found a significantly
-increasing trend in nectar consumption for Temperate studies but a
-significantly decreasing tend for Tropical studies. We found
-significantly decreasing trend for vertebrate consumption in Marine
-studies and Tropical studies.
-
-At the study level, 16 of 50 studies exhibited a significant trend in
+significant within group slopes for birds, terrestrial, and temperate
+studies, there was no significant slope for marine, mammal, or tropical
+studies. We also found a significant decrease in Jaccard similarity in
+unprotected areas only, with no trend for protected areas. We found
+significant dietary shifts across communities, with a significant
+general trend of increasing fish consumption, which was also reflected
+in increasing fish consumption trends for bird studies and temperate
+studies. We found significantly decreasing trend for vertebrate
+consumption in Marine studies and Tropical studies. The trend for marine
+vertebrate consumption was significantly more negative than for
+terrestrial studies and the temperate trend was significantly more
+negative than tropical studies.
+
+At the study level, 12 of 50 studies exhibited a significant trend in
 species richness and 11 exhibited significant turnover. For uncorrected
-functional diversity metrics, 25 of 50 studies exhibited a trend in a
-least one metric, and 15 of 50 studies exhibited a significant trend for
+functional diversity metrics, 11 of 50 studies exhibited a trend in a
+least one metric, and 7 of 50 studies exhibited a significant trend for
 at least one \textcolor{blue}{corrected} metric (Table
@@
 that those trends were likely due to changes in the number of species.
-Hypothesis testing for study-level trends is likely affected by multiple
-testing issues and some trends identified as significant are therefore
-potentially spurious. Rather than interpreting changes in specific
-studies, we present these results as a general picture of the kinds of
-trends experienced by communities.
 
@@
 \midrule
-+ & 7 & 0 & 7 & 5 & 3 & 4 & 4 & 3\\
-- & 9 & 11 & 8 & 6 & 5 & 2 & 6 & 2\\
++ & 2 & 0 & 5 & 3 & 1 & 4 & 0 & 1\\
+- & 9 & 12 & 2 & 0 & 3 & 1 & 1 & 2\\
 \bottomrule

### 2024-04-24 paper/GEB Submission/manuscript_GEB.tex
rebase?

@@
 \begin{table}
-
+\centering
 \caption{\label{tab:traitTable}Description of the traits included in the analysis broken down by categories at data type.}
@@
 \cline{2-2}
-\multirow{-10}{*}{\raggedright\arraybackslash Diet} & Other Plant & \multirow{-10}{*}{\raggedright\arraybackslash Bird and Mammal} & \multirow{-10}{*}{\raggedright\arraybackslash percentage consumed}\\
+\multirow{-10}{*}[4.5\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash Diet} & Other Plant & \multirow{-10}{*}{\raggedright\arraybackslash Bird and Mammal} & \multirow{-10}{*}{\raggedright\arraybackslash percentage consumed}\\
 \cline{1-4}
@@
 \cline{2-2}
-\multirow{-7}{*}{\raggedright\arraybackslash Foraging Strata} & aerial &  & \multirow{-7}{*}{\raggedright\arraybackslash percentage of use}\\
+\multirow{-7}{*}[3\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash Foraging Strata} & aerial &  & \multirow{-7}{*}{\raggedright\arraybackslash percentage of use}\\
 \cline{1-2}
@@
 \cline{2-2}
-\multirow{-2}{*}{\raggedright\arraybackslash Pelagic Specialist} & no & \multirow{-9}{*}{\raggedright\arraybackslash Bird} & \\
+\multirow{-2}{*}[0.5\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash Pelagic Specialist} & no & \multirow{-9}{*}[0.5\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash Bird} & \\
 \cline{1-3}
@@
 \cline{2-2}
-\multirow{-2}{*}{\raggedright\arraybackslash Nocturnal} & no & \multirow{-2}{*}{\raggedright\arraybackslash Bird and Mammal} & \\
+\multirow{-2}{*}[0.5\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash Nocturnal} & no & \multirow{-2}{*}{\raggedright\arraybackslash Bird and Mammal} & \\
 \cline{1-3}
@@
 \cline{2-2}
-\multirow{-2}{*}{\raggedright\arraybackslash Crepuscular} & no &  & \\
+\multirow{-2}{*}[0.5\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash Crepuscular} & no &  & \\
 \cline{1-2}
@@
 \cline{2-2}
-\multirow{-2}{*}{\raggedright\arraybackslash Diurnal} & no & \multirow{-4}{*}{\raggedright\arraybackslash Mammal} & \multirow{-8}{*}{\raggedright\arraybackslash binary}\\
+\multirow{-2}{*}[0.5\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash Diurnal} & no & \multirow{-4}{*}{\raggedright\arraybackslash Mammal} & \multirow{-8}{*}{\raggedright\arraybackslash binary}\\
 \cline{1-4}
@@
 \begin{table}[!h]
-
+\centering\centering
 \caption{\label{tab:resultsTab}Model estimates and statistics for general trend models for species richness, Jaccard similarity, and corrected functional diversity metrics.
@@
 \centering
-\resizebox{\linewidth}{!}{
+\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
 \begin{tabular}[t]{l|l|>{\raggedright\arraybackslash}p{2cm}|l|r|r|l}
@@
 \cline{4-7}
-\multirow{-8}{*}{\raggedright\arraybackslash SES\_FDiv} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & -0.01 &  & \\
+\multirow{-8}{*}[1\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash SES\_FDiv} & \multirow{-6}{*}[0.5\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & -0.01 &  & \\
 \cline{1-7}
@@
 \cline{4-7}
-\multirow{-8}{*}{\raggedright\arraybackslash SES\_FEve} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & -0.21 &  & \\
+\multirow{-8}{*}[1\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash SES\_FEve} & \multirow{-6}{*}[0.5\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & -0.21 &  & \\
 \cline{1-7}
@@
 \cline{4-7}
-\multirow{-8}{*}{\raggedright\arraybackslash SES\_FRic} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.06 &  & \\
+\multirow{-8}{*}[1\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash SES\_FRic} & \multirow{-6}{*}[0.5\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.06 &  & \\
 \cline{1-7}
@@
 \cline{4-7}
-\multirow{-8}{*}{\raggedright\arraybackslash log(Jaccard + 1)} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.05 &  & \\
+\multirow{-8}{*}[1\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash log(Jaccard + 1)} & \multirow{-6}{*}[0.5\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.05 &  & \\
 \cline{1-7}
@@
 \cline{4-7}
-\multirow{-8}{*}{\raggedright\arraybackslash log(Species Richness)} & \multirow{-6}{*}{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.34 &  & \\
+\multirow{-8}{*}[1\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash log(Species Richness)} & \multirow{-6}{*}[0.5\dimexpr\aboverulesep+\belowrulesep+\cmidrulewidth]{\raggedright\arraybackslash random} & \multirow{-3}{2cm}{\raggedright\arraybackslash time series within study} & Corr(Intercept, Year) & 0.34 &  & \\
 \hline
@@
 \begin{table}
-
+\centering
 \caption{\label{tab:trendTab}Number of studies that experienced a significant trend in each calculated metric out of 50 total studies.}

### 2024-04-24 analysis/05_model_metrics.Rmd
fix p.values not updated

@@
 model_output <- bind_rows(metric_model_table, indv_mod_table) %>%
+  select(-p.value) %>% 
   left_join(pval_df) %>%
@@
   filter(!is.na(study_id)) %>%
-  filter(term == "year_scaled", p.value < 0.05) %>% 
+  filter(term == "year_scaled", p.adjust < 0.05) %>% 
   filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S", "FRic", "FEve", "FDiv")) %>% 
@@
   filter(!is.na(study_id)) %>%
-  filter(term == "year_scaled", p.value < 0.05) %>% 
+  filter(term == "year_scaled", p.adjust < 0.05) %>% 
   filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 

### 2024-04-25 analysis/07_study_change.Rmd
Rename, move

--- 2024-04-22 analysis/figures/06_study_change.Rmd
@@
 
+all_metrics <- c("S", "Jaccard_base", "FRic", "FEve", "FDiv", "SES_FRic", "SES_FEve", "SES_FDiv")
+
 change_groups <- wide_table %>% 
-  select(study_id, S, Jaccard_base, FRic, FEve, FDiv) %>%
+  select(study_id,all_of(all_metrics)) %>%
   mutate(across(!study_id, ~case_when(
@@
 change_group_count <- change_groups %>%
-  select(-study_id) %>%
+  select(-study_id, -starts_with("SES_")) %>%
   group_by(across(everything())) %>%
@@
 
+change_group_count_ses <- change_groups %>%
+  select(-c(study_id, FRic, FEve, FDiv)) %>%
+  group_by(across(everything())) %>%
+  count()
+
+change_group_count_complete <- change_groups %>%
+    select(-c(study_id)) %>%
+    group_by(across(everything())) %>%
+    count() %>%
+  mutate(change_description = case_when(
+    S == "pos" ~ "increase in richness, not beyond random expectation", 
+    S == "neg" ~ "loss of redundancy", 
+    S == "0" & Jaccard_base == "neg" ~ "changes in functional space due to turnover",
+    n == 6 ~ "no change", 
+    .default = NA
+  ))
+
+change_group_ses_desc <- change_group_count_ses %>%
+   mutate(change_description = case_when(
+    S == "pos" ~ "increase in species richness, sometimes leading to increase in functional richness", 
+    S == "neg" ~ "loss of redundancy", 
+    S == "0" ~ "functional change independent of species richness change",
+    n == 7 ~ "no change", 
+    .default = NA
+  ))
+```
+
+```{r}
 no_change <- change_groups %>% 

### 2024-04-25 analysis/08_amphibian_trends.Rmd
Rename, move

--- 2023-02-14 analysis/06_sensitivity_analysis.Rmd
@@
 ---
-title: "Sensitivity Analysis"
+title: "Amphibians"
 output: html_document
-date: "2022-08-30"
+date: "2023-06-22"
 ---
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
 
@@
 load(here::here("data/model_data.rda"))
-#exclude amphibians
-model_data <- model_data %>% filter(taxa != "Amphibians")
+amph_data <- model_data %>%
+  filter(taxa == "Amphibians")
 ```
-
-Let's test the sensitivity of the general trends (and study level patterns?) to major data processing decisions.
 
@@
 }
-```
+````
 
-Define models to fit
 ```{r}
@@
 
-1. Min time series duration
 ```{r}
-dur_dfs <- list(min3 = filter(model_data, duration > 2),
-                min4 = filter(model_data, duration > 3),
-                min5 = filter(model_data, duration > 4))
+#Fit models with base settings
+mod_results <- pmap_dfr(mods %>% filter(model_id == "base"), get_model, data = amph_data) %>%
+  mutate(REML = TRUE, bobyqa = FALSE)
 
-duration_sens <- map_dfr(dur_dfs, ~pmap_dfr(filter(mods, model_id == "base"), get_model, data = .x), .id = "dataID")
+#manuscript table
+mod_results %>% 
+  filter(model_id == "base") %>% 
+  select(-c(model_id, warning, statistic, df)) %>% 
+  readr::write_csv(here::here("figures", "amph_model_table.csv"))
 ```
-
-2. Trait Coverage
-```{r}
-load(here::here("data/study_table.rda"))
-
-model_data <- model_data %>%
-  left_join(study_table %>% select(study_id, coverage) %>% mutate(study_id = as.factor(study_id)))
-
-covg_dfs <- list(covg75 = filter(model_data, taxa != "Amphibians"),
-                covg85 = filter(model_data, taxa != "Amphibians", coverage >= 0.85),
-                covg90 = filter(model_data, taxa != "Amphibians", coverage >= 0.90))
-
-coverage_sens <- map_dfr(covg_dfs, ~pmap_dfr(filter(mods, model_id == "base"), 
-                                             get_model, data = .x), .id = "dataID")
-
-sens_mods <- bind_rows(duration_sens, coverage_sens) %>%
-  select(-model) %>%
-  filter(metric != "Jaccard_next")
-
-usethis::use_data(sens_mods)
-
-write_csv(sens_mods, here::here("paper", "sensitivity_models.csv"))
-```

### 2024-04-26 analysis/07_study_change.Rmd
Clean up how we think about change groups.

@@
   ))
+
+usethis::use_data(change_group_ses_desc)
 ```
@@
 ```{r}
-no_change <- change_groups %>% 
+change <-  change_groups %>% 
   mutate(study_id = as.integer(study_id)) %>%
-  filter(across(!study_id, ~. == 0)) %>%
-  left_join(meta_clean)
+  filter(if_any(all_of(c("S", "Jaccard_base", "SES_FRic", "SES_FEve", "SES_FDiv")), ~. != 0))
+
+no_change <- study_meta %>% filter(!study_id %in% change$study_id, study_id %in% model_output$study_id)
 
@@
   filter(S == "neg", SES_FRic == "0", SES_FDiv == "0", SES_FEve == "0") %>%
-  left_join(meta_clean)
+  left_join(study_meta)
 
@@
   filter(S == "pos", SES_FRic == "0", SES_FDiv == "0", SES_FEve == "0") %>%
-  left_join(meta_clean)
+  left_join(study_meta)
 
@@
   filter(across(!study_id & !SES_FEve, ~. == 0)) %>%
-  left_join(meta_clean)
+  left_join(study_meta)
 
@@
   filter(Jaccard_base == "neg", S == "pos", SES_FRic == "pos", SES_FDiv == "0", SES_FEve == "0") %>%
-  left_join(meta_clean)
+  left_join(study_meta)
 ```

### 2024-04-26 paper/GEB Submission/manuscript_GEB.Rmd
Adjust for results changes with multiple testing

@@
 For example, animal body size is expected to decrease as a result of climate change, a phenomena that has been documented in multiple taxa empirically and experimentally", "blue")` [@caruso2014; @forster2012; @huss2019; @sheridan2011; @tseng2018].
-`r colorize("There is some evidence that this holds true for birds ", "blue")`[@vanbuskirk2010]`r colorize(", but the picture is likely more complicated for mammals, where urbanization may actually lead to larger body sizes due to novel food sources", "blue")` [@hantak2021]`` r colorize(", even as megafaunal loss leads to on average smaller community body size.
-For dietary traits, recent work documenting insect declines", "blue")` ``[@wagner2021]`` `r colorize("points to potentially significant negative impacts on insectivorous animals", "blue") `` [@canaday1996; @sekercioglu2002; @tallamy2021].
-`r colorize("Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species ", "blue")`[@cooke2019].
-`r colorize("Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems", "blue")` [@bain2020]`r colorize(", and loss of neotropical understory foragers even in protect areas", "blue")` [@pollock2022].
+`r colorize("There is some evidence that this holds true for birds ", "blue")`[@vanbuskirk2010]`r colorize(", but the picture is likely more complicated for mammals, where urbanization may actually lead to larger body sizes due to novel food sources", "blue")` [@hantak2021] `r colorize(", even as megafaunal loss leads to on average smaller community body size.
+For dietary traits, recent work documenting insect declines", "blue")` [@wagner2021] ``` r colorize("points to potentially significant negative impacts on insectivorous animals", "blue") `` [@canaday1996; @sekercioglu2002; @tallamy2021].
+```r colorize("Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species ", "blue")`[@cooke2019].`r colorize("Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems", "blue")`[@bain2020]`r colorize(", and loss of neotropical understory foragers even in protected areas", "blue")\` [@pollock2022].
 
@@
 
-We assess thousands of mammal and bird functional diversity time series to determine whether or not `r colorize("the addition of functional trait data gives a clarifying picture of biodiversity change across communities.
-We present a few areas of change consensus across communities, and even more scenarios where functional trends are unexplained and warrant further theoretical and experimental examination.
-We assess functional diversity trends at three different levels: general trends across communities, trends for communities with similar characteristics (taxa, biome, realm, protection status), and trends for individual studies.
-We further evaluate evidence of key predictions for individual traits using trait-level compositional trends, including changes in mean body size, dietary traits, and foraging strata.", "blue")`
+We synthesize thousands of mammal and bird functional diversity time series to assess mean trends across communities.
+We evaluate 1) whether or not there is a prevailing type of functional change across communities and if community characteristics explain variation in those trends, 2) hypotheses of change for individual trait axes, and 3) the implications of study-level trends for the ecology of those communities.
+To address these goals, we present results for general trends that aggregate across all communities, categorical trends that aggregate across communities with similar characteristics, and study-level trends.
 
@@
 
-To test for trends within and between different levels of taxa, biome, realm, and protection status we fit separate models with each of those factors added as a predictor interacting with time to the original model structure.
+To test for categorical trends within and between different levels of taxa, biome, realm, and protection status we fit separate models with each of those factors added as a predictor interacting with time to the original model structure.
 We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.8.2, @lenth2022].
 `r colorize("For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.", "blue")` We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
-All models in our analysis were fit using the *lme4* (v1.1-30) package in R (v4.2.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].
+All models in our analysis were fit using the *lme4* (v1.1-30) package in R (v4.2.1) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].
+
+To correct for false positives due to multiple testing, we assess significance of trends based on adjusted p-values.
+Adjusted p-values were calculated using the Benjamini-Hochberg procedure, which controls the false discovery rate.
+The procedure was implemented by the function `p.adjust` in the R package *stats* (v4.2.1).
 
@@
 
-We found no significant overall `r colorize("temporal", "blue")` trend in species richness or functional diversity metrics `r colorize("including functional richness, evenness, or divergence", "blue")` (observed or `r colorize("corrected", "blue")`) (Fig \ref{fig:timeseriesPlot}).
+We found no significant general `r colorize("temporal", "blue")` trend in species richness or functional diversity metrics `r colorize("including functional richness, evenness, or divergence", "blue")` (observed or `r colorize("corrected", "blue")`) (Fig \ref{fig:timeseriesPlot}).
 We did find a significant overall decrease in Jaccard similarity, indicating accumulating changes in species composition.
 Non-significant overall `r colorize("temporal", "blue")` trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies `r colorize("was not significantly different from zero", "blue")` (Table \ref{tab:resultsTab}).
-`r colorize("Within-group", "blue")` trends for different taxa, biomes, realms, or protection statuses were also non-significant for richness and functional diversity metrics.
-The general trends for *CWM* models were similarly not significant, with a significant positive trend for only percentage of fish in diet composition.
+`r colorize("Categorical", "blue")` trends for different taxa, biomes, realms, or protection statuses were also non-significant for species richness and functional diversity metrics.
 
@@
 
-We did find significant differences between taxa, realms, and biomes for Jaccard similarity and some of the *CWM's*.
-For example, while Jaccard similarity was decreasing in the general trend and there were significant within group slopes for birds, terrestrial, and temperate studies, there was no significant slope for marine, mammal, or tropical studies.
+We did find significant categorical trends for some taxa, realms, and biomes for Jaccard similarity and two dietary *CWM's*.
+While Jaccard similarity was decreasing in the general trend and there were significant within group slopes for birds, terrestrial, and temperate studies, there was no significant slope for marine, mammal, or tropical studies.
 We also found a significant decrease in Jaccard similarity in unprotected areas only, with no trend for protected areas.
@@
 
-At the study level, 12 of 50 studies exhibited a significant trend in species richness and 11 exhibited significant turnover.
+At the study level, 11 of 50 studies exhibited a significant trend in species richness and 11 exhibited significant turnover.
 For uncorrected functional diversity metrics, 11 of 50 studies exhibited a trend in a least one metric, and 7 of 50 studies exhibited a significant trend for at least one `r colorize("corrected", "blue")` metric (Table \ref{tab:trendTab}).
@@
 
-Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at aggregate and local trends in functional diversity in mammal and bird communities.
-`r colorize("Our results show that the addition of functional traits illuminates a few consistent functional trends across communities, but largely complicates rather that clarifies the story of biodiversity change.
-While the characteristics of species clearly matter, instead of unifying the nature of communities' change, they more often distinguish them.
-In general, we found a few areas of consensus for models where communities were aggregated (general trends, trends by taxa, biome, realm, protection status), and vast heterogeneity for study-level models.", "blue")`
-
-## Evidence of consensus
-
-`r colorize("The most stark area of consensus was the lack of general trend aggregating across communities.
-We did not detect an overall trend in any functional diversity metric, corrected or uncorrected.
+Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at general, categorical, and study-level local trends in functional diversity in mammal and bird communities.
+`We did not detect an overall trend in any functional diversity metric, corrected or uncorrected.
 As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing dissimilarity through time", "blue")` [@dornelas2018]`r colorize(", indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends.
@@
 The lack of turnover in marine studies stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial", "blue")` [@blowes2019].
-`r colorize("However, previous global estimates are dominated by fish communities which we exclude here and may be driving the overall trend while disguising relative stasis in marine bird and mammal communities.", "blue")`
+`r colorize("However, previous global estimates are dominated by fish communities which we exclude here and are likely driving the overall trend while disguising relative stasis in marine bird and mammal communities.", "blue")`
 
-```{=html}
-<!--`r colorize("We found evidence of functional richness loss in mammal studies, with a loss trend significantly different from zero and significantly different from the trend for birds, indicating a potential loss of functional capacity in these communities.
-This result is consistent with previous work linking anthropogenic drivers such as habitat loss, poaching, and human modification and development to loss of functional diversity in mammal communities across the globe", "blue")` [@brodie2021; @cox2022; @li2022].
-`r colorize("However we did not find trends outside the random expectation for species loss, in contrast to expectations for multiple continents based on IUCN risk categories", "blue")` [@brodie2021].
-`r colorize("In general, functional capacity of mammals showed greater declines than bird communities.", "blue")`-->
-```
-## Areas of incongruence
+-   head nod to mammal and tropical communities
 
-`r colorize("The addition of functional trait data illuminated multiple areas where a synthesis approach is incongruous with system-specific studies or simply unexpected.
+`r colorize("Many of our results were incongruous with system-specific studies or simply unexpected.
 For example, results for protected areas were surprising given the assumption that protection insulates communities from functional degradation.
@@
 
-```{=html}
-<!--`r colorize("We also found evidence of increasing evenness for marine studies, which was not predicted by previous work in marine systems.
-The empirical link between changes in evenness and ecosystem process is the most poorly studied of the metrics, however theory indicates that increased evenness improves function and stability ", "blue")` [@filstrup2019], `r colorize("as the dominance of one or a few species makes function sensitive to the ability of those species to respond to environmental change ", "blue")`[@hillebrand2008; @hillebrand2009].
-`r colorize("Further, there is some evidence that maintaining evenness is important for supporting multifunctionality in communities ", "blue")`[@bagousse-pinguet2019].
-`r colorize("An increase in evenness is therefore a positive indicator for these communities, despite the fact that marine bird and mammal communities are some of the most threatened communities in the world.
-Still, this result may be a function of data limitations as the majority of marine studies are of seabird communities (7 of 9), so the general trend may be an early indicator of pay off from investments in sea bird recovery also found in recent species-based work ", "blue")`[@spatz2023].-->
-```
 `r colorize("Our results were also inconsistent with all predictions for changes in trait prevalence, including foraging strata, body size, and dietary traits.
@@
 
-`r colorize("At the study-level, our results run contrary to our hope for a clearer picture of biodiversity change through a functional lens.
-The lack of general trend in functional diversity metrics belies a huge range of positive and negative trends at the study-level in both species and functional metrics.
-For example, it appears at the aggregate level that species turnover does not translate to functional turnover based on a lack of trends in CWM's.
-However at the study-level, all studies experiencing species turnover are also experiencing significant turnover for multiple individual traits, there is simply no consistency across communities in the kind of turnover occurring to be manifest in general trends.", "blue")`
-
-`r colorize("In order to simplify discussion, we will talk about the implications for ecosystem process and vulnerability for communities grouped by their concurrent change in corrected functional diversity metrics, species richness and turnover split into positive, negative or no trend.
-While there are over 150 possible combinations of change direction (or no change) in the 5 metrics, we discuss here the six scenarios that occurred in more than one study: no change, species loss or gain only, loss of functional evenness, species richness loss with species turnover, and increase in species and functional richness accompanied by significant turnover.
-We focus on these scenarios to combat potential spurious results due to multiple testing, as it is unlikely all observations of the same scenario are due to false positives.", "blue")`
-
-`r colorize("The majority group of studies (17 studies) exhibited no trend in any species or functional diversity metric.
+Trends for individual studies fell into three broad groups of change: no change in species or functional diversity metrics, loss of redundancy, and functional change independent of species richness.
+`r colorize("The majority of studies (25 studies) exhibited no trend in any species or functional diversity metric.
 Contrary to the expectation due to anthropogenic and global change stressors, these communities do not show significant changes over the course of the observation window.
-Studies in this group span the distribution of study durations excluding only the very longest running studies, with the longest no change time series lasting 23 years.
-They also included both bird and mammal studies and only four were located in protected areas, indicating that the lack of trend is not restricted to a specific ecological context or those communities most insulated from human impact.", "blue")`
+Studies in this group span the distribution of study durations excluding only the very longest running studies, with the longest no change time series lasting 33 years.
+They also included both bird and mammal studies and only seven were located in protected areas, indicating that the lack of trend is not restricted to a specific ecological context or those communities most insulated from human impact.", "blue")`
 
@@
 For example, species-level trait data does not capture intraspecific shifts in the trait space, which can represent significant changes in the functional space as a whole and impact maintenance of ecological processes.", "blue")`
+
+The next biggest group of studies (14 studies) exhibited some functional
 

### 2024-04-30 analysis/07_study_change.Rmd
Make sure labels are right

@@
     S == "0" & Jaccard_base == "neg" ~ "changes in functional space due to turnover",
-    n == 6 ~ "no change", 
+    n == 7 ~ "no change", 
     .default = NA
@@
    mutate(change_description = case_when(
-    S == "pos" ~ "increase in species richness, sometimes leading to increase in functional richness", 
+    S == "pos" & n != "7" ~ "increase in species richness", 
     S == "neg" ~ "loss of redundancy", 
-    S == "0" ~ "functional change independent of species richness change",
-    n == 7 ~ "no change", 
+    S == "0" ~ "functional change",
+    n == "7" ~ "no change", 
     .default = NA

### 2024-04-30 paper/GEB Submission/manuscript_GEB.Rmd
Changes to discussion

@@
 ---
-title: Functional traits complicate the picture of temporal biodiversity change in bird and mammal communities
+title: No general trend for multiple dimensions of functional diversity in bird and mammal communities
 abstract: |
@@
 Study characteristics such as taxa, realm, biome, or protection status did not distinguish between types of change exhibited by communities.
-At the study-level, there was substantial heterogeneity in the direction and type of functional change exhibited.
-We further did not find evidence to support multiple predictions for specific traits, including decreasing body size, dietary shifts, or changes in foraging strata.
+We found evidence of an increasing trend in fish consumption across all communities, but no evidence to support multiple predictions for specific traits, including decreasing body size, dietary shifts, or changes in foraging strata.
 
   Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevalant than the other across many taxa, biomes, and realms.
-At the study-level, the majority of studies showed no species or functional trends, but other types of change were found for multiple studies, including changes in functional redundancy, increased species and functional richness, or loss of functional richness.
-With no prevailing scenario of change, it will be critical to link types of change to ecological context.
+At the study-level, the majority of studies showed no species or functional trends, with the remainder of the studies falling into broad categories of functional shifts independent of species richness, functional redundancy loss, and increasing species richness.
 
@@
 
-`r colorize("Using general trends derived from limited data as a diagnostic for the state of biodiversity or directing conservation action is a topic of ongoing debate.", "blue")` Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
+`r colorize("Using general trends derived from limited data as a diagnostic for the state of biodiversity is a topic of ongoing debate.", "blue")` Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
 Many of these criticisms reflect limitations of ecological data on the whole, `r colorize("leading to a call for additional data not only to fill geographic and temporal gaps, but to flesh out key characteristics of communities", "blue")` [@dornelas2023; @primack2018].
@@
 `r colorize("There is some evidence that this holds true for birds ", "blue")`[@vanbuskirk2010]`r colorize(", but the picture is likely more complicated for mammals, where urbanization may actually lead to larger body sizes due to novel food sources", "blue")` [@hantak2021] `r colorize(", even as megafaunal loss leads to on average smaller community body size.
-For dietary traits, recent work documenting insect declines", "blue")` [@wagner2021] ``` r colorize("points to potentially significant negative impacts on insectivorous animals", "blue") `` [@canaday1996; @sekercioglu2002; @tallamy2021].
-```r colorize("Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species ", "blue")`[@cooke2019].`r colorize("Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems", "blue")`[@bain2020]`r colorize(", and loss of neotropical understory foragers even in protected areas", "blue")\` [@pollock2022].
+For dietary traits, recent work documenting insect declines", "blue")` [@wagner2021] `r colorize("points to potentially significant negative impacts on insectivorous animals", "blue")` [@canaday1996; @sekercioglu2002; @tallamy2021].
+`r colorize("Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species ", "blue")`[@cooke2019].`r colorize("Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems", "blue")`[@bain2020]`r colorize(", and loss of neotropical understory foragers even in protected areas", "blue")` [@pollock2022].
 
@@
 We obtained mammal and bird time series from the BioTIME database, a global repository of high-quality assemblage time series `r colorize("collected from the literature and ongoing monitoring efforts", "blue")`.
-`r colorize("Data is structured such that a study encompass all data collected following consistent sampling protocols and may contain multiple time series from separate locations.", "blue")` Time series represent full assemblages rather than populations of single species [@dornelas2018], and all time series include abundance of observed species.
+`r colorize("Data is structured such that a study comprises all data collected following consistent sampling protocols and may contain multiple time series from separate locations.", "blue")` Time series represent full assemblages rather than populations of single species [@dornelas2018], and all time series include abundance of observed species.
 Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual time series following a standardized spatial scale.
@@
 `r colorize("The resulting cell size for our data was approximately 95 ", "blue")`km^2^.
-All samples from a study within a single cell were considered to be a single time series, and species abundances were combined for all samples.
+All samples from a study within a single cell were considered to be a single time series location, and species abundances were combined for all samples.
 
@@
 Traits include: body mass, diet, nocturnality, forest foraging strata, pelagic use.
-With the exception of body mass, traits were broken down into percentage or binary use for each level of the trait type (Table \ref{tab:traitTable}).
+For categorical traits, different category levels are represented as binary or percentage use (Table \ref{tab:traitTable}).
 
@@
 The earliest sample was in 1923 and the most recent was in 2014.
-`r colorize("While it is not possible with this data to directly assess the level of human impact occurring for each study, we include binary protection status as a coarse indicator of impact level.
+`r colorize("While it is not possible with available data to directly assess the level of human impact occurring for each study, we include binary protection status as a coarse indicator of impact level.
 However, protected areas were almost exclusively from temperate terrestrial studies (with one tropical study), so results are confounded by multiple other study characteristics.", "blue")` For a full breakdown of studies and their characteristics, see the supplement.
@@
 
-We also calculated the community-weighted mean (*CWM*) of all continuous traits to examine `r colorize("turnover", "blue")` in the distribution of each trait.
+We also calculated the community-weighted mean (*CWM*) of all continuous traits to examine `r colorize("changes", "blue")` in the distribution of each trait.
 `r colorize("Wholesale shifts in the trait space due to changes of trait means could occur even while the shape of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained.
@@
 
-We estimated general trends `r colorize("across bird and mammals communities", "blue")` for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each time series nested within the study, `r colorize("methods which deal well with the inherent imbalances in our data.", "blue")` We fit 18 individual *CWM* models, `r colorize("one for each trait included in the analsysis", "blue")`.
+We estimated general trends `r colorize("across bird and mammals communities", "blue")` for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each location nested within the study, `r colorize("methods which deal well with the inherent imbalances in our data.", "blue")` We fit 18 individual *CWM* models, `r colorize("one for each trait included in the analsysis", "blue")`.
 All time series with data for a given trait were included in the corresponding *CWM* model.
 We estimated study level trends using individual linear models.
-For studies with more than one times series we fit a random slope and intercept for time series.
+For studies with more than one location we fit a random slope and intercept for time series.
 Some study-level models could not be fit for five studies for at least one metric due to data limitations, but those studies were still included in the general models.
@@
 
-We did find significant categorical trends for some taxa, realms, and biomes for Jaccard similarity and two dietary *CWM's*.
-While Jaccard similarity was decreasing in the general trend and there were significant within group slopes for birds, terrestrial, and temperate studies, there was no significant slope for marine, mammal, or tropical studies.
+We did find significant categorical trends for Jaccard similarity and two dietary *CWM's*.
+While Jaccard similarity was decreasing in the general trend and there were significant categorical slopes for birds, terrestrial, and temperate studies, there was no significant slope for marine, mammal, or tropical studies.
 We also found a significant decrease in Jaccard similarity in unprotected areas only, with no trend for protected areas.
@@
 Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at general, categorical, and study-level local trends in functional diversity in mammal and bird communities.
-`We did not detect an overall trend in any functional diversity metric, corrected or uncorrected.
-As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing dissimilarity through time", "blue")` [@dornelas2018]`r colorize(", indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends.
-We found significant turnover for most biomes, realm, and taxa with the exception of marine, tropical and mammal studies.
+`r colorize("We did not detect an overall trend in any functional diversity metric, corrected or uncorrected.
+As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing dissimilarity through time", "blue")` [@dornelas2018]`r colorize(", indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends.", "blue")`
+
+`r colorize("Categorical trends in turnover highlight the types of communities experiencing change in community composition.
+For example, we found significant turnover for most biomes, realms, and taxa with the exception of marine, tropical, and mammal studies.
 The lack of turnover in marine studies stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial", "blue")` [@blowes2019].
-`r colorize("However, previous global estimates are dominated by fish communities which we exclude here and are likely driving the overall trend while disguising relative stasis in marine bird and mammal communities.", "blue")`
+`r colorize("However, previous global estimates are dominated by fish communities which we exclude here and are likely driving the overall trend while disguising relative stasis in marine bird and mammal communities.
+We also found significant turnover for unprotected areas and no significant turnover in protected areas, indicating that in general unprotected communities are experiencing bigger shifts in community composition through time.", "blue")`
 
--   head nod to mammal and tropical communities
+`r colorize("Surprisingly, we found no evidence for some of the strongest predictions for anthropogenic impact.
+For example, we found no trend for mean body mass in bird or mammal communities.
+This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization are overshadowing climate change impacts, or more likely current shifts are happening at an intraspecific level not captured by our data.
+Additionally, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].
+`r colorize("Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened", "blue")` [@ripple2015].
 
-`r colorize("Many of our results were incongruous with system-specific studies or simply unexpected.
-For example, results for protected areas were surprising given the assumption that protection insulates communities from functional degradation.
-The only difference we found between protected and unprotected areas was significant turnover occurring outside protected areas and no significant turnover inside protected areas.
-Rather than the result of protection itself, this is likely due to the fact that the majority of studies were from terrestrial, temperate studies where turnover is known to be lower ", "blue")`[@blowes2019].
-`r colorize("While things may generally be more static within protect areas, the functional dimensions of protected communities fared no better or worse on average than their non-protected counter parts.
-Looking at individual studies, there is a mix of both losses and gains in almost all functional diversity and composition metrics across communities found in protected areas.", "blue")`
-
-`r colorize("Our results were also inconsistent with all predictions for changes in trait prevalence, including foraging strata, body size, and dietary traits.
+`r colorize("We similarly did not find evidence of changes in trait distributions predicted by previous global change work.
 We found no changes in the prevalence of different foraging strategies, despite documented losses of understory birds in the neotropics and some evidence that agricultural incursion particularly threatens arboreal species.
 Those shifts may therefore be the result of specific contexts and not generalizable to bird communities across the globe.
-We found no evidence of general trends in the CWM for body mass for birds or mammals, indicating that either body size is not changing significantly due to climate change, opposing pressures such as urbanization are overshadowing climate change impacts, or more likely current shifts are happening at an intraspecific level and and have not yet propogated to the species losses that would be captured by our data.
-Further, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].
-`r colorize("Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened", "blue")` [@ripple2015].
-
-`r colorize("Rather than declines in insectivory and increases in invertivory predicted by previous work, we instead found changes in the degree of consumption of fish and vertebrates.
+Rather than declines in insectivory and increases in invertivory predicted by changes in resource availability, we instead found changes in the degree of consumption of fish and vertebrates.
 Increases in fish consumption for bird and temperate studies may be the product of increased availability of easy fish from commercial fishing operations or the stocking of inland waterways for sport fishing.
@@
 
-## Study heterogeneity
+## Study-level functional change
 
-Trends for individual studies fell into three broad groups of change: no change in species or functional diversity metrics, loss of redundancy, and functional change independent of species richness.
-`r colorize("The majority of studies (25 studies) exhibited no trend in any species or functional diversity metric.
+```{r studyChange, echo = FALSE}
+
+load(here::here("data/change_group_ses_desc.rda"))
+
+change_group_ses_desc[change_group_ses_desc == "pos"] <- "+"
+change_group_ses_desc[change_group_ses_desc == "neg"] <- "-"
+
+change_group_ses_desc %>%
+  select(type = change_description, everything()) %>%
+  mutate(n = as.character(n),
+           n = ifelse(n == "7", "32", n)) %>%
+  mutate(type = ifelse(n == "32", "no change", type)) %>%
+  rename(`Study Count` = `n`) %>%
+     kable(format = "latex",
+        caption = "Table giving the count of number of studies for different kinds of functional change, broken down into no change functional change without species richness change, loss of redundancy, and increases in species richness.", align = "c",
+        latex_options= c("scale_down", "hold_position")) %>% 
+  collapse_rows(columns = c(1,2)) %>%
+  column_spec(1, width = "4cm")
+
+```
+
+`r colorize("While trends aggregating across studies did not indicate one prevailing kind of functional change, post hoc examination of trends for individual studies revealed four distinct change groups.
+These groups are categorized based on the concurrent trends in species and functional metrics and can be broadly described as: no change in species or functional diversity metrics, functional change independent of species richness, loss of redundancy, and increase in species richness", "blue")` (Table \ref{tab:studyChange}).
+`r colorize("The majority of studies (32 studies) exhibited no trend in any species or functional diversity metric.
 Contrary to the expectation due to anthropogenic and global change stressors, these communities do not show significant changes over the course of the observation window.
@@
 `r colorize("Third, these communities may be experiencing directional shifts undetectable by available data.
-For example, species-level trait data does not capture intraspecific shifts in the trait space, which can represent significant changes in the functional space as a whole and impact maintenance of ecological processes.", "blue")`
+For example, species-level trait data do not capture intraspecific shifts in the trait space, which can have significant implications for ecological processes.", "blue")`
 
-The next biggest group of studies (14 studies) exhibited some functional
+`r colorize("The next largest group of studies (7 studies) were those with no species richness trend that showed a trend in at least one functional diversity metric.
+Based on a species approach alone, these communities apear to either be experiencing no change or some change in species identity, with unknown implications for conservation concern or ecological process.
+However, with the addition of a functional perspective we found multiple kinds of underlying functional change including increases and decreases in corrected functional richness and divergence and loss of corrected functional evenness.
+We therefore found evidence of both increases and decreases in the functional capacity of communities independent of species-based metrics, reinforcing the importance of moving beyond soley species-based approaches for biodiversity assessment", "blue")` [@hillebrand2017; @santini2017].
 
-`r colorize("Three of the groups fit under the broad umbrella of changes in redundancy.
-By definition, if a community gains or loses species while functional diversity metrics are unchanged, those species represent an increase or decrease in redundancy, so we include communities exhibiting gains (2 studies) or losses (5 studies) in species richness with no change in functional metrics or turnover and species richness loss with significant turnover (2 studies) in this umbrella.
-For communities exhibiting loss of redundancy (declining species richness), ecological processes are likely being maintained, but capacity to respond to future stressors is reduced ", "blue")`[@reich2012].
-`r colorize("These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future ", "blue")`[@pimiento2020a].
-`r colorize("Conversely, communities exhibiting a increase in redundancy (via an increase in species richness) are becoming better positioned to respond to future stressors, but are not actually expanding their functional capacity as may be assumed by looking at species gains alone", "blue")` [@biggs2020].
+`r colorize("9 studies showed loss of functional redundancy.
+These studies are characterized by a loss of species richness with no shift in functional diversity metrics, indicating the lost species were functionally similar enough to another species to not impact the overall functional structure.
+Redundancy is critical for maintaining ecological processes in the face of disturbance as it ensures  functional roles are maintained", "blue")` [@biggs2020]`r colorize(", and a loss of functional redundancy likeley reduces a community's capacity to respond to future stressors ", "blue")`[@reich2012]`r colorize(".
+These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future", "blue")` [@pimiento2020a].
 
-`r colorize("The next group of studies (3 studies) exhibit an increase in species richness and functional richness with significant turnover.
-While these communities are losing some species that are functionally redundant or functional analogs of species additions, they are gaining even more species that expand the functional space.
+`r colorize("The final two studies showed an increase in species richness and uncorrected functional richness, with only one showing an increase in functional richness beyond what was expected by chance (corrected functional richness).
 These communities have a hopeful trajectory, as they are improving their functional capacity and potentially their functional redundancy, likely leading to robust future ecological function.
-Notably, the studies that exhibited species richness increases were exclusively from terrestrial, temperate (sometimes temperate/polar) bird communities and occurred in countries where there has been a significant investment in conservation over the last few decades (United States, Canada, Sweden).
-Our results are consistent with other functional work in these regions, showing increases in species richness and functional diversity for North American breeding birds over the last few decades after a period of decline", "blue")` [@jarzyna2016] `r colorize(", and loss of common, functionally general species even as rare species are increasing in North America and Europe ", "blue")` [@inger2015; @rosenberg2019; @schipper2016].
+Notably, both of these studies were from temperate, terrestrial bird communities in Scandinavian countries, which have some of the highest investments in conservation in the world.
+Still, results are consistent with other broad-scale studies of bird communities, which found loss of common, functionally general species alongside increases in rare species ", "blue")` [@inger2015; @rosenberg2019; @schipper2016].
 
-`r colorize("The final group (3 studies) exhibits decreasing functional evenness.
-These communities are of particular concern not just because of the role evenness plays in maintaining functionality, but because of increasing evidence evenness is more sensitive to environmental change than species richness", "blue")` [@filstrup2019].
-`r colorize("Formal tests of evenness changes as an early warning signal of more catastrophic shifts in functional diversity will be critical next step for understanding the full implication of evenness loss for ecosystem processes.", "blue")`
-
-`r colorize("While the majority of studies fell into one of the groups discussed in detail above, 25", "blue")`% `r colorize("of the studies exhibited some different combination of change in metrics, underlining the vast heterogeneity of realized change scenarios.", "blue")` `r colorize("We have an indication of the ecological characteristics most likely to lead to some of the change types (e.g.
-temperate bird communities are gaining species and functional richness), but most types of change are exhibited across many different kinds of communities.
-", "blue")` `r colorize("Identifying key factors mediating the kind of biodiversity change a community experiences is critical for identifying areas of conservation concern and extrapolating results to areas where functional trait data may not be available.
-While it is tempting to draw a one to one line between the degree of human impact and loss of functional integrity (in the form of functional redundancy and richness), that relationship is not borne out by our rough metric of human impact (protection status) or previous taxa-specific or single community work, which is largely mixed", "blue")` [@flynn2009; @luck2013; @magioli2021; @tinoco2018].
-
-Study-level results also illustrate the relationship between species turnover and functional turnover.
-For all studies exhibiting increasing dissimilarity through time (11 studies), there were multipe significant trend in functional composition (CWM's) and diversity metrics.
-However, there was little consensus in the types of shift, as indicated by lack of general trends for most metrics.
-
-## Policy Implications
+## Conservation Implications
 
@@
 
-## Future Work
-
-`r colorize("The addition of functional trait data to the biodiversity change conversation is yet another illustration of how critical context is for understanding ecological patterns [@catford2022; @spake2023].
-Here we assessed how key community characteristics such as biome, realm, taxa, and protection status may partition variation in the functional trends exhibited and found that, on the whole, these broad classifications did not seem to make the picture of change clearer.
-Further work linking specific ecological contexts to types of functional change will be critical for identifying new areas of concern, especially when those communities may not otherwise be showing significant changes in species richness.
-The identification of contexts where functional change is most likely occurring will be a significant asset in directing the future collection of trait data, the main barrier for taking a functional approach to biodiversity change.", "blue")`
-
 # Acknowledgments

### 2024-05-13 analysis/05_model_metrics.Rmd
Move sensitivy analysis into main analysis script so p-values can be adjusted.

@@
   write_csv(here::here("paper", "study_metadata.csv"))
-```
-
-####Data exploration
-
-Visualize number of years for each timeseries
-```{r}
-hist_numYears <- model_data %>%
-  select(year, rarefyID) %>% 
-  distinct() %>% 
-  count(rarefyID, name = "num_years") %>% 
-  ggplot() +
-  geom_histogram(aes(x = num_years),
-                 binwidth = 2) +
-  scale_x_continuous(name = 'Number of years sampled',
-                     breaks = c(2,4,8,16,32,48,64,96)) +
-  #scale_y_continuous(breaks = c(0,2500,5000,7500,10000,12500, 15000, 17500, 20000,22500)) +
-  labs(tag = 'B',
-       y = 'Number of cells') +
-  theme_bw() +
-  theme(panel.grid.minor.x = element_blank())
-
-hist_duration <- model_data %>%
-  select(rarefyID, duration) %>% 
-  distinct() %>% 
-  ggplot() +
-  geom_histogram(aes(x = duration),
-                 binwidth = 2) +
-  scale_x_continuous(name = 'Number of years sampled',
-                     breaks = c(2,4,8,16,32,48,64,96)) +
-  #scale_y_continuous(breaks = c(0,2500,5000,7500,10000,12500, 15000, 17500, 20000,22500)) +
-  labs(tag = 'B',
-       y = 'Number of cells') +
-  theme_bw() +
-  theme(panel.grid.minor.x = element_blank())
 ```
@@
 ```
-Perform p-value correction
+
+#### Sensitivity analysis for duration and trait coverage
+
+1. Min time series duration
+```{r}
+dur_dfs <- list(min3 = filter(model_data, duration > 2),
+                min4 = filter(model_data, duration > 3),
+                min5 = filter(model_data, duration > 4))
+
+duration_sens <- map_dfr(dur_dfs, ~pmap_dfr(filter(mods, model_id == "base"), get_model, data = .x), .id = "dataID")
+
+# first group of bad models, try with REML off
+bad_dur_models <- duration_sens %>%
+  select(metric, model_id, warning, dataID) %>% 
+  filter(!is.na(warning)) %>% 
+  distinct()
+
+min3_bad <- bad_dur_models %>% filter(dataID == "min3") %>% pull(metric)
+min4_bad <- bad_dur_models %>% filter(dataID == "min4") %>% pull(metric)
+min5_bad <- bad_dur_models %>% filter(dataID == "min5") %>% pull(metric)
+
+refit_dur <- bind_rows(pmap_dfr(filter(mods, model_id == "base", metric_name %in% min3_bad), get_model, data = dur_dfs$min3, REML = FALSE) %>% mutate(dataID = "min3"),
+                       pmap_dfr(filter(mods, model_id == "base", metric_name %in% min4_bad), get_model, data = dur_dfs$min4, REML = FALSE) %>% mutate(dataID = "min4"),
+                       pmap_dfr(filter(mods, model_id == "base", metric_name %in% min5_bad), get_model, data = dur_dfs$min5, REML = FALSE) %>% mutate(dataID = "min5")
+)
+
+# two left that didn't converge
+bad_dur_models2 <- refit_dur %>%
+  select(metric, model_id, warning, dataID) %>% 
+  filter(!is.na(warning)) %>% 
+  distinct()
+
+refit_dur_2 <- bind_rows(pmap_dfr(filter(mods, model_id == "base", metric_name == "CWM_diet_planto"), get_model, data = dur_dfs$min3, bobyqa = TRUE) %>% mutate(dataID = "min3"),
+                       pmap_dfr(filter(mods, model_id == "base", metric_name == "CWM_forstrat_midhigh"), get_model, data = dur_dfs$min5, bobyqa = TRUE) %>% mutate(dataID = "min5"))
+
+duration_sens_mods <- bind_rows(duration_sens %>% filter(is.na(warning)),
+                                refit_dur %>% filter(is.na(warning)),
+                                refit_dur_2 %>% filter(is.na(warning)))
+```
+
+2. Trait Coverage
+```{r}
+load(here::here("data/study_table.rda"))
+
+model_data <- model_data %>%
+  left_join(study_table %>% select(study_id, coverage) %>% mutate(study_id = as.factor(study_id)))
+
+covg_dfs <- list(covg75 = filter(model_data, taxa != "Amphibians"),
+                covg85 = filter(model_data, taxa != "Amphibians", coverage >= 0.85),
+                covg90 = filter(model_data, taxa != "Amphibians", coverage >= 0.90))
+
+coverage_sens <- map_dfr(covg_dfs, ~pmap_dfr(filter(mods, model_id == "base"), 
+                                             get_model, data = .x), .id = "dataID")
+
+# first group of bad models, try with REML off
+bad_cov_models <- coverage_sens %>%
+  select(metric, model_id, warning, dataID) %>% 
+  filter(!is.na(warning)) %>% 
+  distinct()
+
+covg75_bad <- bad_cov_models %>% filter(dataID == "covg75") %>% pull(metric)
+covg85_bad <- bad_cov_models %>% filter(dataID == "covg85") %>% pull(metric)
+covg90_bad <- bad_cov_models %>% filter(dataID == "covg90") %>% pull(metric)
+
+refit_cov <- bind_rows(pmap_dfr(filter(mods, model_id == "base", metric_name %in% covg75_bad), 
+                                get_model, data = covg_dfs$covg75, REML = FALSE) %>% mutate(dataID = "covg75"),
+                       pmap_dfr(filter(mods, model_id == "base", metric_name %in% covg85_bad), 
+                                get_model, data = covg_dfs$covg85, REML = FALSE) %>% mutate(dataID = "covg85"),
+                       pmap_dfr(filter(mods, model_id == "base", metric_name %in% covg90_bad), 
+                                get_model, data = covg_dfs$covg90, REML = FALSE) %>% mutate(dataID = "covg90")
+)
+
+coverage_sens_mods <- bind_rows(coverage_sens %>% filter(is.na(warning)),
+                                refit_cov %>% filter(is.na(warning)))
+
+sens_mods <- bind_rows(duration_sens_mods, coverage_sens_mods) %>%
+  select(-model, -warning) %>%
+  filter(metric != "Jaccard_next")
+```
+
+#### Perform p-value correction
 ```{r}
@@
 
-pval_df <- bind_rows(base, contrast, group_slope, study_trend) %>%
+# sensitivity models
+sens_trends <- sens_mods %>%
+  filter(term == "year_scaled")
+
+pval_df <- bind_rows(base, contrast, group_slope, study_trend, sens_trends) %>%
   filter(metric != "Jaccard_next")
@@
 
+sensitivity_output <- sens_mods %>%
+  select(-p.value) %>%
+  left_join(pval_df %>% select(p.value, p.adjust, effect, term, model_id, dataID, metric), na_matches = "never") %>%
+  rename(sens_level = dataID)
+
 usethis::use_data(model_output)
+usethis::use_data(sensitivity_output)
+
+write_csv(sensitivity_output, here::here("paper", "sensitivity_models.csv"))
 ```
@@
 
-Summaries of study slopes
-```{r}
-#First - how many studies had a significant trend for each metric? 
-model_output %>% 
-  filter(!is.na(study_id)) %>%
-  filter(term == "year_scaled", p.adjust < 0.05) %>% 
-  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S", "FRic", "FEve", "FDiv")) %>% 
-  mutate(positive = ifelse(estimate > 0, TRUE, FALSE)) %>%
-  group_by(metric, positive) %>% 
-  summarise(n_study_id = n_distinct(study_id)) %>%
-  pivot_wider(names_from = "metric", values_from = "n_study_id")
-
-#And which studies were they?
-model_output %>% 
-  filter(!is.na(study_id)) %>%
-  filter(term == "year_scaled", p.adjust < 0.05) %>% 
-  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S", "FRic", "FEve", "FDiv")) %>% 
-  select(study_id, metric) %>%
-  distinct()
-
-#Which were significant for multiple metrics?
-model_output %>% 
-  filter(!is.na(study_id)) %>%
-  filter(term == "year_scaled", p.adjust < 0.05) %>% 
-  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
-  group_by(study_id) %>%
-  filter(n() > 1) %>%
-  select(study_id, metric)
-```
-
 #### Analizing slopes
@@
 library(multcomp)
+
+study_time_data <- model_data %>% 
+  dplyr::select(study_id, startyear, duration) %>% 
+  distinct() %>%
+  group_by(study_id) %>%
+  filter(startyear == min(startyear))
 

### 2024-05-14 analysis/05_model_metrics.Rmd
Fix the duration models

@@
   group_by(study_id) %>%
-  filter(startyear == min(startyear))
+  mutate(startyear = min(startyear), duration = max(duration)) %>%
+  distinct()
 
-study_slope_data <- indv_mod_table %>%
-  filter(term == "year_scaled") %>%
+study_slope_data <- model_output %>%
+  filter(!is.na(study_id), term == "year_scaled") %>%
   dplyr::select(study_id, metric, estimate) %>%
-  left_join(meta %>% 
-              mutate(study_id = as.character(study_id))) %>%
-  left_join(model_data %>% dplyr::select(study_id, startyear, duration) %>% distinct())
+  left_join(study_time_data)
 

### 2024-05-16 analysis/05_model_metrics.Rmd
Add min 10 years to sensitivity analysis

@@
                 min4 = filter(model_data, duration > 3),
-                min5 = filter(model_data, duration > 4))
+                min5 = filter(model_data, duration > 4),
+                min10 = filter(model_data, duration > 9))
 
@@
 min5_bad <- bad_dur_models %>% filter(dataID == "min5") %>% pull(metric)
+min10_bad <- bad_dur_models %>% filter(dataID == "min10") %>% pull(metric)
 
@@
                        pmap_dfr(filter(mods, model_id == "base", metric_name %in% min4_bad), get_model, data = dur_dfs$min4, REML = FALSE) %>% mutate(dataID = "min4"),
-                       pmap_dfr(filter(mods, model_id == "base", metric_name %in% min5_bad), get_model, data = dur_dfs$min5, REML = FALSE) %>% mutate(dataID = "min5")
+                       pmap_dfr(filter(mods, model_id == "base", metric_name %in% min5_bad), get_model, data = dur_dfs$min5, REML = FALSE) %>% mutate(dataID = "min5"),
+                       pmap_dfr(filter(mods, model_id == "base", metric_name %in% min10_bad), get_model, data = dur_dfs$min10, REML = FALSE) %>% mutate(dataID = "min10")
 )

### 2024-05-16 paper/GEB Submission/manuscript_GEB.Rmd
Ben edits, reformat table

@@
 
-  Results: We found no overall trends in functional diversity metrics.
+  Results: We found no overall temporal trends in functional diversity metrics.
 Study characteristics such as taxa, realm, biome, or protection status did not distinguish between types of change exhibited by communities.
@@
 `r colorize("Using general trends derived from limited data as a diagnostic for the state of biodiversity is a topic of ongoing debate.", "blue")` Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
-Many of these criticisms reflect limitations of ecological data on the whole, `r colorize("leading to a call for additional data not only to fill geographic and temporal gaps, but to flesh out key characteristics of communities", "blue")` [@dornelas2023; @primack2018].
+Many of these criticisms reflect limitations of ecological data on the whole, `r colorize("leading to a call for additional data not only to fill geographic and temporal gaps, but to fill in key characteristics of communities", "blue")` [@dornelas2023; @primack2018].
 
@@
 Here we `r colorize("leverage ongoing efforts to assemble functional trait data and recent computational advances to", "blue")` perform the first multi-taxa, multi-realm assessment of functional diversity and composition change through time.
-We focus on mammal and bird species as `r colorize("a subset of the world's biodiversity of particular conservation concern that is heavily impacted by anthropogenic change.", "blue")`While examining trends in plants, invertebrates, and other vertebrate species is of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
+We focus on mammal and bird species as `r colorize("subsets of the world's biodiversity of particular conservation concern that is heavily impacted by anthropogenic change.", "blue")`While examining trends in plants, invertebrates, and other vertebrate species is of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
 To ensure comparability across taxa in trait type and data quality we therefore focus on mammals and birds.
 `r colorize("We include body mass, dietary, foraging and other behavioral traits", "blue")` that were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].
-An initial assessment of amphibian trends is included in the supplement, but excluded from general trend assessment here due to limited geographic coverage.
 
@@
 We obtained mammal and bird time series from the BioTIME database, a global repository of high-quality assemblage time series `r colorize("collected from the literature and ongoing monitoring efforts", "blue")`.
-`r colorize("Data is structured such that a study comprises all data collected following consistent sampling protocols and may contain multiple time series from separate locations.", "blue")` Time series represent full assemblages rather than populations of single species [@dornelas2018], and all time series include abundance of observed species.
+`r colorize("Data are structured such that a study comprises all data collected following consistent sampling protocols and may contain multiple time series from separate locations.", "blue")` Time series represent full assemblages rather than populations of single species [@dornelas2018], and all time series include abundance of observed species.
 Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual time series following a standardized spatial scale.
@@
 In order to have a sufficient number of species to calculate functional diversity metrics, years with fewer than 5 species observed were also excluded.
-Sensitivity analyses were conducted for the trait coverage threshold and the duration of included time series.
+Sensitivity analyses were conducted for the trait coverage threshold and duration of included time series by rerunning analyses with increasingly conservative cutoffs including 85% and 90 % coverage trait coverage and durations longer than 3, 4, 5, and 10 years.
 
@@
 
-To correct for false positives due to multiple testing, we assess significance of trends based on adjusted p-values.
+To correct for false positives due to multiple testing, we assess significance of the trends for the fixed effect of time based on adjusted p-values.
 Adjusted p-values were calculated using the Benjamini-Hochberg procedure, which controls the false discovery rate.
@@
 In general, there were more significant trends for uncorrected metrics, with some disappearing after correction, indicating that those trends were likely due to changes in the number of species.
+Study-level slopes were not significantly related to start year or duration of time series for any metric.
 
@@
   kable(format = "latex", 
-        col.names = c("", "S", "Jaccard Similarity", "FRic", "FEve", "FDiv", "SES FRic", "SES FEve", "SES FDiv"),
-        caption = "Number of studies that experienced a significant trend in each calculated metric out of 50 total studies.", 
+        #col.names = c("", "S", "Jaccard Similarity", "FRic", "FEve", "FDiv", "SES FRic", "SES FEve", "SES FDiv"),
+        col.names = c("", "richness", "Jaccard Similarity", "richness", "evenness", "divergence", "SES richness", "SES evenness", "SES divergence"),
+        caption = "Number of studies that experienced a significant trend in each calculated metric out of 50 total studies.
+Both uncorrected and corrected (SES) versions of functional metrics are included.",
         booktabs = TRUE, align = "c", escape = FALSE) %>%
+  kableExtra::add_header_above(c("", "species" = 2, "functional" = 6)) %>%
   kable_styling("striped") #%>%
@@
 
-Study-level slopes were significantly related to start year of the time series for Jaccard similarity and functional evenness models, both of which had significantly more negative slopes with more recent start year.
-No functional diversity metrics were significantly related to the duration of the time series.
-
 We assessed the sensitivity of general trend results to major data processing decision by rerunning models with increasingly conservative subsets of the data.
@@
 After excluding time series with less than three years we found an increasing trend for body mass that remained after excluding time series of less than four and five years.
-We also found a decreasing trend in functional richness for models with a minimum of three years and a minimum of five years.
-A decreasing functional richness trend was found for models with a min of 85% trait coverage, but not the other two coverage levels.
+The body size trend did however disappear after excluding time series of less then 10 years.
 A complete list of models run in the sensitivity analysis and their results can be found in the supplement.
@@
 For example, we found no trend for mean body mass in bird or mammal communities.
-This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization are overshadowing climate change impacts, or more likely current shifts are happening at an intraspecific level not captured by our data.
-Additionally, many of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].
+In fact, sensitivity results indicated that after excluding the shortest time series there is a robust increasing trend in body size.
+This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization or carnivore restablishment are overshadowing climate change impacts, or current shifts are happening at an intraspecific level not captured by our data.
+Additionally, the majority of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].
 `r colorize("Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened", "blue")` [@ripple2015].
@@
 Rather than declines in insectivory and increases in invertivory predicted by changes in resource availability, we instead found changes in the degree of consumption of fish and vertebrates.
-Increases in fish consumption for bird and temperate studies may be the product of increased availability of easy fish from commercial fishing operations or the stocking of inland waterways for sport fishing.
-However, fisheries also have negative impacts on waterbirds through competition for fish or bird mortality due to by catch, making these results suprising.
-Decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.", "blue")`
+Neither increasing trends for fish consumption nor decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.", "blue")`
 
@@
      kable(format = "latex",
-        caption = "Table giving the count of number of studies for different kinds of functional change, broken down into no change functional change without species richness change, loss of redundancy, and increases in species richness.", align = "c",
+        caption = "Table giving the count of number of studies for different kinds of functional change, broken down into no change functional change without species richness change, loss of redundancy, and increases in species richness.
+Trends direction is indicate by a positive or negative sign, with a 0 indicating no trend.", align = "c",
         latex_options= c("scale_down", "hold_position")) %>% 
@@
 First, these may be communities resisting perturbations or simply not experiencing significant perturbations.
-Given the studies in this group come from all possible taxa, realms, biomes, and protection statuses, evidence points to communities resisting perturbation, offering some hope that there are areas of the globe where communities are fairing well for now.
+Given the studies in this group come from all possible taxa, realms, biomes, and protection statuses, evidence points to communities resisting perturbation.
 Alternatively, these may be communities that have experienced or continue to experience significant stress, but lost species or functional diversity outside the observation window.
@@
 `r colorize("The final two studies showed an increase in species richness and uncorrected functional richness, with only one showing an increase in functional richness beyond what was expected by chance (corrected functional richness).
-These communities have a hopeful trajectory, as they are improving their functional capacity and potentially their functional redundancy, likely leading to robust future ecological function.
-Notably, both of these studies were from temperate, terrestrial bird communities in Scandinavian countries, which have some of the highest investments in conservation in the world.
+While this can be an indication of increasing functional capacity or reestabilishment of previously lost species, it can also be the result of less desirable process like the introduction of invasive species.
+Notably, both of these studies were from temperate, terrestrial bird communities in Scandinavian countries with a unique environment and conservation context.
 Still, results are consistent with other broad-scale studies of bird communities, which found loss of common, functionally general species alongside increases in rare species ", "blue")` [@inger2015; @rosenberg2019; @schipper2016].

### 2024-05-23 paper/GEB Submission/manuscript_GEB.Rmd
Additional intro paragraph, other minor edits from Ben.

@@
 Study characteristics such as taxa, realm, biome, or protection status did not distinguish between types of change exhibited by communities.
-We found evidence of an increasing trend in fish consumption across all communities, but no evidence to support multiple predictions for specific traits, including decreasing body size, dietary shifts, or changes in foraging strata.
+We found evidence of an increasing temporal trend in fish consumption across all communities with , but no evidence to support multiple predictions for specific traits, including decreasing body size, dietary shifts, or changes in foraging strata.
 
-  Main Conclusions: General trends indicate that on the aggregate one type of functional shift is not more prevalant than the other across many taxa, biomes, and realms.
+  Main Conclusions: General temporal trends indicate that on the aggregate one type of functional shift is not more prevalant than the other across many taxa, biomes, and realms.
 At the study-level, the majority of studies showed no species or functional trends, with the remainder of the studies falling into broad categories of functional shifts independent of species richness, functional redundancy loss, and increasing species richness.
@@
 
+Broad-scale trends in functional diversity are the aggregation of local-scale responses to disturbance and persistent pressures like environmental change or the introduction of invasive species.
+On acute timescales, impacts on communities can range from reduction in abundance across the trait space to complete gains or losses in some aspect of the trait space [@boersma2016; @larson2021].
+As communities recover or adapt to persistent pressures, changes in the functional space are determined by community assembly processes and limited by the species available in the regional species pool [@green2022].
+Lost species may be replaced by functionally similar species when available, or novel species may colonize niches newly created by changing environmental conditions.
+Critically, similar shifts in the functional space may be the result of both positive and negative conservation scenarios.
+For example, expansion of the functional trait space may be evidence of recovery from disturbance or the introduction of a functionally unique invasive species.
+
 `r colorize("The expectation for functional diversity change across communities is not obvious from past work, and may or may not follow species richness trends", "blue")` [@dornelas2023].
@@
 
-Many studies had a variable number of samples within years.
-To account for this inconsistency in sampling effort we used sample-based rarefaction by bootstrap resampling within years for each time series based on the smallest number of samples in a year for that time series.
+`r colorize("41 of the 2432 time series included in our dataset had a variable number of samples within years.", "blue")` To account for this inconsistency in sampling effort we used sample-based rarefaction by bootstrap resampling within years for each time series based on the smallest number of samples in a year for that time series.
+`r colorize("We made 200 rarefied samples for over sampled years and calculated diversity metrics for each sample, then took the median for each metric as the value for that year and time series.
+28 of the affect studies required under sampling by only one sample, and only three time series required under sampling by more than ten samples.", "blue")`
 
-Our final dataset included 2,432 time series from 50 studies in 21 countries and 12 biomes and 6 different traits (Fig \ref{fig:taxaMap}).
+Our final dataset included 2,432 time series from 50 studies in 21 countries and 12 biomes using 6 different traits (Fig \ref{fig:taxaMap}).
 Data came from both terrestrial and marine realms and five biomes (Global, Polar/Temperate, Temperate, Temperate/Tropical, Tropical).
@@
 SES estimates can be interpreted as how much of the functional characteristic (richness, evenness, divergence) was observed beyond what was expected by chance for a community of that species richness.
-`r colorize("This approach will be less accurate for shorter time series, as we likley will not have captured all available species in the true species pool, but it is impossible to know whether the mean estimate from the null model is an over or under estimate without knowing the functional characteristics of the missing species.", "blue")`
+`r colorize("This approach will be less accurate for shorter time series, as we likely will not have captured all available species in the true species pool, but it is impossible to know whether the mean estimate from the null model is an over or under estimate without knowing the functional characteristics of the missing species.", "blue")`
 
@@
 Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at general, categorical, and study-level local trends in functional diversity in mammal and bird communities.
+Our work complements efforts on a similar scale assessing temporal change in species and phyllogenetic diversity across taxa [@daru2021; @dornelas2014; @li2020; @vellend2013l; @vellend2017] or functional diversity change for a single taxa [@barnagaud2017; @jarzyna2016; @trindade-santos2020].
 `r colorize("We did not detect an overall trend in any functional diversity metric, corrected or uncorrected.
@@
 For example, we found no trend for mean body mass in bird or mammal communities.
-In fact, sensitivity results indicated that after excluding the shortest time series there is a robust increasing trend in body size.
+In fact, sensitivity results indicated that after excluding the shortest time series there is evidence of an increasing trend in body size.
 This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization or carnivore restablishment are overshadowing climate change impacts, or current shifts are happening at an intraspecific level not captured by our data.

### 2024-05-23 analysis/figures/visualization.Rmd
Attempt at the hexbin

@@
 
+Hexbin map
+```{r}
+coast <- rnaturalearth::ne_coastline(scale = "small", returnclass = "sf")
+
+basemap <- ggplot(data = coast) + geom_sf(color = "grey") + 
+  ggthemes::theme_map() #+
+  #coord_sf(crs = 3857)  #+
+  #ggthemes::theme_map()
+
+#xlim=c(-180,180), ylim=c(-57, 90), 
+bird_hexmap <- basemap +
+  geom_hex(data = study_data %>% select(rarefyID, rarefyID_x, rarefyID_y, taxa) %>% 
+             distinct() %>%
+             filter(taxa == "Birds"),
+             aes(x = rarefyID_x, y = rarefyID_y), color = "grey",# bins = 50, color = my_pal[2]) 
+                 bins = 35) +
+  scale_fill_gradient2()
+             scale_fill_gradient(low = "#bbe9ff", high = "#086788")
+
+mammal_hexmap <- basemap +
+  geom_hex(data = study_data %>% select(rarefyID, rarefyID_x, rarefyID_y, taxa) %>% 
+             distinct() %>%
+             filter(taxa == "Mammals"),
+             aes(x = rarefyID_x, y = rarefyID_y), color = "grey",# bins = 50, color = my_pal[2]) 
+                 bins = 35) +
+  scale_fill_gradient2()
+```
+
 Metric plots with trend lines

### 2024-05-29 paper/GEB Submission/manuscript_GEB.Rmd
more changes

@@
 
-  Main Conclusions: General temporal trends indicate that on the aggregate one type of functional shift is not more prevalant than the other across many taxa, biomes, and realms.
-At the study-level, the majority of studies showed no species or functional trends, with the remainder of the studies falling into broad categories of functional shifts independent of species richness, functional redundancy loss, and increasing species richness.
+  Main Conclusions: General temporal trends indicate that on the aggregate one type of functional change is not more prevalant than the other across many taxa, biomes, and realms.
+At the study-level, the majority of studies showed no species or functional trends, with the remainder of the studies falling into broad categories of functional changes independent of species richness, functional redundancy loss, and increasing species richness.
 
@@
 While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.
-
-`r colorize("Using general trends derived from limited data as a diagnostic for the state of biodiversity is a topic of ongoing debate.", "blue")` Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
+`r colorize("Still, using general trends derived from limited data as a diagnostic for the state of biodiversity is a topic of ongoing debate.", "blue")` Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
 Many of these criticisms reflect limitations of ecological data on the whole, `r colorize("leading to a call for additional data not only to fill geographic and temporal gaps, but to fill in key characteristics of communities", "blue")` [@dornelas2023; @primack2018].
 
-Functional diversity offers a potentially powerful addition to species-based approaches for detecting and describing community change by providing a mechanistic link between species' response to environmental change (*response traits*) and the processes they perform (*effect traits*) [@lavorel2002; @mcgill2006; @suding2008].
-By describing the functional trait space, functional diversity metrics capture the disproportionate impact of losses or gains of functionally unique species.
-Functional diversity metrics `r colorize("may therefore illuminate joint responses from functionally similar species or communities undetectible by looking at species identity alone.", "blue")`
+Functional diversity offers a potentially powerful addition to species-based approaches for detecting and describing community change by capturing changes in the traits that underpin fundamental ecological processes [@lavorel2002; @mcgill2006; @suding2008].
+Functional diversity metrics capture complimentary aspects of the trait space including the volume of the multidimensional space (functional diversity), how species and abundance are distributed across the space (functional evenness and divergence), and where the space occurs across the range of possible trait distributions (community weighted trait means).
+Considered together, they describe the impacts of changes in species identity on the building blocks of stability, function, and resilience.
 
-Broad-scale trends in functional diversity are the aggregation of local-scale responses to disturbance and persistent pressures like environmental change or the introduction of invasive species.
-On acute timescales, impacts on communities can range from reduction in abundance across the trait space to complete gains or losses in some aspect of the trait space [@boersma2016; @larson2021].
-As communities recover or adapt to persistent pressures, changes in the functional space are determined by community assembly processes and limited by the species available in the regional species pool [@green2022].
-Lost species may be replaced by functionally similar species when available, or novel species may colonize niches newly created by changing environmental conditions.
-Critically, similar shifts in the functional space may be the result of both positive and negative conservation scenarios.
-For example, expansion of the functional trait space may be evidence of recovery from disturbance or the introduction of a functionally unique invasive species.
+Broad-scale trends in functional diversity are the aggregation of local-scale responses to acute pulse disturbance, persistent press disturbances, and periods of relative stasis [@inamine2022].
+Pulse disturbances have a sudden one time impact on the community, changing abundance distributions (functional evenness or divergence) or reducing the volume of the functional space [@boersma2016; @hillebrand2020; @larson2021].
+While press disturbances can also lead to loss of functional volume as species cannot adapt rapidly to new conditions, it can also lead to wholesale shifts in the location of the functional space as species fill in new environmental space.
+Rather than identifying outcomes of one of these specific processes, observational time series of functional diversity therefore capture the cumulative affects of multiple potentially competing processes.
 
-`r colorize("The expectation for functional diversity change across communities is not obvious from past work, and may or may not follow species richness trends", "blue")` [@dornelas2023].
-While `r colorize("loss of functional diversity", "blue")` is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], `r colorize("functional diversity may be maintained even when species are lost from a community", "blue")` [@daz2001; @petchey2009].
-Forecasts of functional loss range from negligible [@gallagher2013] to dire [@petchey2002; @pimiento2020].
-And while some observed trends show significant functional loss [@flynn2009] others document no loss even in some of the most heavily impacted communities [@edwards2013; @larsen2018; @matuoka2020].
-On paleoecological time scales, `r colorize("the functional space", "blue")` shows mixed responses to environmental change and extinction events [@jackson2015; @dornelas2018], with significant impacts of species extinctions on functional diversity in some taxa and not others [@pimiento2017].
-For some time periods, functional diversity appears to be maintained for substantial portions of geological time [@hedberg2022].
-`r colorize("Contemporary, broad-scale examinations of functional change are limited to only a few taxa-focused studies, but show for example functional richness increases for both North American birds", "blue")` [@barnagaud2017; @jarzyna2016] `r colorize("and ray-finned fishes, sharks, and rays", "blue")` [@trindade-santos2020].
-
+`r colorize("The expectation for functional diversity change across communities is not obvious from past work, and may or may not follow species richness trends", "blue")` [@dornelas2023; @larson2021].
+`r colorize("While loss of functional diversityis frequently cited as one of the most pressing concerns of the anthropocene ", "blue")`[@cardinale2012; @dirzo2014; @young2016], observational trends show significant functional loss in some studies [@flynn2009] and no loss even in the most heavily impacted communities for others [@edwards2013; @larsen2018; @matuoka2020].
+`r colorize("Broad-scale studies are limited, but show for example functional richness increases for North American birds", "blue")` [@barnagaud2017; @jarzyna2016].
 `r colorize("We have stronger expectations for changes in the prevalence of some individual traits.
 For example, animal body size is expected to decrease as a result of climate change, a phenomena that has been documented in multiple taxa empirically and experimentally", "blue")` [@caruso2014; @forster2012; @huss2019; @sheridan2011; @tseng2018].
-`r colorize("There is some evidence that this holds true for birds ", "blue")`[@vanbuskirk2010]`r colorize(", but the picture is likely more complicated for mammals, where urbanization may actually lead to larger body sizes due to novel food sources", "blue")` [@hantak2021] `r colorize(", even as megafaunal loss leads to on average smaller community body size.
-For dietary traits, recent work documenting insect declines", "blue")` [@wagner2021] `r colorize("points to potentially significant negative impacts on insectivorous animals", "blue")` [@canaday1996; @sekercioglu2002; @tallamy2021].
+`r colorize("For dietary traits, recent work documenting insect declines", "blue")` [@wagner2021] `r colorize("points to potentially significant negative impacts on insectivorous animals", "blue")` [@canaday1996; @sekercioglu2002; @tallamy2021].
 `r colorize("Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species ", "blue")`[@cooke2019].`r colorize("Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems", "blue")`[@bain2020]`r colorize(", and loss of neotropical understory foragers even in protected areas", "blue")` [@pollock2022].
@@
 Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at general, categorical, and study-level local trends in functional diversity in mammal and bird communities.
-Our work complements efforts on a similar scale assessing temporal change in species and phyllogenetic diversity across taxa [@daru2021; @dornelas2014; @li2020; @vellend2013l; @vellend2017] or functional diversity change for a single taxa [@barnagaud2017; @jarzyna2016; @trindade-santos2020].
+Our work complements efforts on a similar scale assessing temporal change in species and phyllogenetic diversity across taxa [@daru2021; @dornelas2014; @li2020; @vellend2013; @vellend2017] or functional diversity change for a single taxa [@barnagaud2017; @jarzyna2016; @trindade-santos2020].
 `r colorize("We did not detect an overall trend in any functional diversity metric, corrected or uncorrected.
@@
 In fact, sensitivity results indicated that after excluding the shortest time series there is evidence of an increasing trend in body size.
-This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization or carnivore restablishment are overshadowing climate change impacts, or current shifts are happening at an intraspecific level not captured by our data.
+This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization or carnivore restablishment are overshadowing climate change impacts", "blue") `[@hantak2021]`r colorize(", or current shifts are happening at an intraspecific level not captured by our data.
 Additionally, the majority of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].
@@
 Rather than declines in insectivory and increases in invertivory predicted by changes in resource availability, we instead found changes in the degree of consumption of fish and vertebrates.
-Neither increasing trends for fish consumption nor decreasing trends for vertebrate consumption in marine and tropical studies are not obviously consistent with known changes in those contexts and warrant further examination.", "blue")`
+Neither increasing trends for fish consumption nor decreasing trends for vertebrate consumption in marine and tropical studies are obviously consistent with known changes in those contexts and warrant further examination.", "blue")`
 
@@
 `r colorize("While we found no over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
-In fact, study-level trends indicate quite the opposite, that functional shifts with negative or yet unknown implications for ecosystem processes may be going undetected by common species-based approaches.
+In fact, study-level trends indicate quite the opposite, that functional changes with negative or yet unknown implications for ecosystem processes may be going undetected by common species-based approaches.
 For example, loss of evenness in communities with constant species richness may be an first sign of a community being impacted by environmental change, with negative implications for stability and function.", "blue")`

### 2024-05-29 paper/GEB Submission/manuscript_GEB.Rmd
line breaks

@@
 Pulse disturbances have a sudden one time impact on the community, changing abundance distributions (functional evenness or divergence) or reducing the volume of the functional space [@boersma2016; @hillebrand2020; @larson2021].
-While press disturbances can also lead to loss of functional volume as species cannot adapt rapidly to new conditions, it can also lead to wholesale shifts in the location of the functional space as species fill in new environmental space.
+While press disturbances can also lead to loss of functional volume as species fail to adapt to new conditions, it can also lead to wholesale shifts in the location of the functional space as species fill in new environmental space.
 Rather than identifying outcomes of one of these specific processes, observational time series of functional diversity therefore capture the cumulative affects of multiple potentially competing processes.
@@
 `r colorize("The expectation for functional diversity change across communities is not obvious from past work, and may or may not follow species richness trends", "blue")` [@dornelas2023; @larson2021].
-`r colorize("While loss of functional diversityis frequently cited as one of the most pressing concerns of the anthropocene ", "blue")`[@cardinale2012; @dirzo2014; @young2016], observational trends show significant functional loss in some studies [@flynn2009] and no loss even in the most heavily impacted communities for others [@edwards2013; @larsen2018; @matuoka2020].
+`r colorize("While loss of functional diversity is frequently cited as one of the most pressing concerns of the anthropocene ", "blue")`[@cardinale2012; @dirzo2014; @young2016], observational trends show significant functional loss in some studies [@flynn2009] and no loss even in the most heavily impacted communities for others [@edwards2013; @larsen2018; @matuoka2020].
 `r colorize("Broad-scale studies are limited, but show for example functional richness increases for North American birds", "blue")` [@barnagaud2017; @jarzyna2016].
@@
 `r colorize("For dietary traits, recent work documenting insect declines", "blue")` [@wagner2021] `r colorize("points to potentially significant negative impacts on insectivorous animals", "blue")` [@canaday1996; @sekercioglu2002; @tallamy2021].
-`r colorize("Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species ", "blue")`[@cooke2019].`r colorize("Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems", "blue")`[@bain2020]`r colorize(", and loss of neotropical understory foragers even in protected areas", "blue")` [@pollock2022].
+`r colorize("Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species ", "blue")`[@cooke2019].
+`r colorize("Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems", "blue")`[@bain2020]`r colorize(", and loss of neotropical understory foragers even in protected areas", "blue")` [@pollock2022].
 
@@
 In fact, sensitivity results indicated that after excluding the shortest time series there is evidence of an increasing trend in body size.
-This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization or carnivore restablishment are overshadowing climate change impacts", "blue") `[@hantak2021]`r colorize(", or current shifts are happening at an intraspecific level not captured by our data.
+This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization or carnivore restablishment are overshadowing climate change impacts", "blue")`[@hantak2021]`r colorize(", or current shifts are happening at an intraspecific level not captured by our data.
 Additionally, the majority of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].

### 2024-05-30 analysis/06_study_change.Rmd
Minor changes after going over reviewer comments again.

--- 2024-04-30 analysis/07_study_change.Rmd
@@
 ```{r}
-change <-  change_groups %>% 
-  mutate(study_id = as.integer(study_id)) %>%
-  filter(if_any(all_of(c("S", "Jaccard_base", "SES_FRic", "SES_FEve", "SES_FDiv")), ~. != 0))
-
-no_change <- study_meta %>% filter(!study_id %in% change$study_id, study_id %in% model_output$study_id)
-
-redundancy <- change_groups %>% 
-  mutate(study_id = as.integer(study_id)) %>%
-  filter(S == "neg", SES_FRic == "0", SES_FDiv == "0", SES_FEve == "0") %>%
-  left_join(study_meta)
-
-redundancy_gain <- change_groups %>% 
-  mutate(study_id = as.integer(study_id)) %>%
-  filter(S == "pos", SES_FRic == "0", SES_FDiv == "0", SES_FEve == "0") %>%
-  left_join(study_meta)
-
-evenness <- change_groups %>% 
-  mutate(study_id = as.integer(study_id)) %>%
-  filter(across(!study_id & !SES_FEve, ~. == 0)) %>%
-  left_join(study_meta)
-
-increase <- change_groups %>% 
-  mutate(study_id = as.integer(study_id)) %>%
-  filter(Jaccard_base == "neg", S == "pos", SES_FRic == "pos", SES_FDiv == "0", SES_FEve == "0") %>%
-  left_join(study_meta)
+# change <-  change_groups %>% 
+#   mutate(study_id = as.integer(study_id)) %>%
+#   filter(if_any(all_of(c("S", "Jaccard_base", "SES_FRic", "SES_FEve", "SES_FDiv")), ~. != 0))
+# 
+# no_change <- study_meta %>% filter(!study_id %in% change$study_id, study_id %in% model_output$study_id)
+# 
+# redundancy <- change_groups %>% 
+#   mutate(study_id = as.integer(study_id)) %>%
+#   filter(S == "neg", SES_FRic == "0", SES_FDiv == "0", SES_FEve == "0") %>%
+#   left_join(study_meta)
+# 
+# redundancy_gain <- change_groups %>% 
+#   mutate(study_id = as.integer(study_id)) %>%
+#   filter(S == "pos", SES_FRic == "0", SES_FDiv == "0", SES_FEve == "0") %>%
+#   left_join(study_meta)
+# 
+#   
+# evenness <- change_groups %>% 
+#   mutate(study_id = as.integer(study_id)) %>%
+#   filter(across(!study_id & !SES_FEve, ~. == 0)) %>%
+#   left_join(study_meta)
+# 
+# increase <- change_groups %>% 
+#   mutate(study_id = as.integer(study_id)) %>%
+#   filter(Jaccard_base == "neg", S == "pos", SES_FRic == "pos", SES_FDiv == "0", SES_FEve == "0") %>%
+#   left_join(study_meta)
 ```

### 2024-05-30 paper/GEB Submission/manuscript_GEB.Rmd
Minor changes after going over reviewer comments again.

@@
 Functional diversity metrics capture complimentary aspects of the trait space including the volume of the multidimensional space (functional diversity), how species and abundance are distributed across the space (functional evenness and divergence), and where the space occurs across the range of possible trait distributions (community weighted trait means).
-Considered together, they describe the impacts of changes in species identity on the building blocks of stability, function, and resilience.
+Considered together, they describe the aspects of the trait space that mediate ecological process.
 
@@
 `r colorize("While loss of functional diversity is frequently cited as one of the most pressing concerns of the anthropocene ", "blue")`[@cardinale2012; @dirzo2014; @young2016], observational trends show significant functional loss in some studies [@flynn2009] and no loss even in the most heavily impacted communities for others [@edwards2013; @larsen2018; @matuoka2020].
-`r colorize("Broad-scale studies are limited, but show for example functional richness increases for North American birds", "blue")` [@barnagaud2017; @jarzyna2016].
+`r colorize("Broad-scale observational studies are limited, but show for example functional richness increases for North American birds", "blue")` [@barnagaud2017; @jarzyna2016].
 `r colorize("We have stronger expectations for changes in the prevalence of some individual traits.
@@
 We therefore found evidence of both increases and decreases in the functional capacity of communities independent of species-based metrics, reinforcing the importance of moving beyond soley species-based approaches for biodiversity assessment", "blue")` [@hillebrand2017; @santini2017].
+All but one of the studies in this group experienced significant turnover, which was also reflected in trends for multiple *CWM*'s, indicating that the overall functional space was shifting even when some characteristics of the functional space were maintained.
 
 `r colorize("9 studies showed loss of functional redundancy.
-These studies are characterized by a loss of species richness with no shift in functional diversity metrics, indicating the lost species were functionally similar enough to another species to not impact the overall functional structure.
+These studies are characterized by a loss of species richness with no shift in functional diversity metrics, indicating the lost species were functionally similar enough to another species to not impact the overall functional diversity.
 Redundancy is critical for maintaining ecological processes in the face of disturbance as it ensures  functional roles are maintained", "blue")` [@biggs2020]`r colorize(", and a loss of functional redundancy likeley reduces a community's capacity to respond to future stressors ", "blue")`[@reich2012]`r colorize(".

### 2024-05-30 paper/GEB Submission/manuscript_GEB.Rmd
Update references.

@@
 While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.
-`r colorize("Still, using general trends derived from limited data as a diagnostic for the state of biodiversity is a topic of ongoing debate.", "blue")` Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
+`r colorize("Still, using general trends derived from limited data as a diagnostic for the state of biodiversity is a topic of ongoing debate", "blue")` [@boënnec2024; @johnson2024].
+Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
 Many of these criticisms reflect limitations of ecological data on the whole, `r colorize("leading to a call for additional data not only to fill geographic and temporal gaps, but to fill in key characteristics of communities", "blue")` [@dornelas2023; @primack2018].

### 2024-05-30 paper/GEB Submission/manuscript_GEB.Rmd
Update package versions.

@@
 To test for categorical trends within and between different levels of taxa, biome, realm, and protection status we fit separate models with each of those factors added as a predictor interacting with time to the original model structure.
-We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.8.2, @lenth2022].
+We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
 `r colorize("For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.", "blue")` We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
-All models in our analysis were fit using the *lme4* (v1.1-30) package in R (v4.2.1) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].
+All models in our analysis were fit using the *lme4* (v1.1-35.3) package in R (v4.3.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].
 
@@
 Adjusted p-values were calculated using the Benjamini-Hochberg procedure, which controls the false discovery rate.
-The procedure was implemented by the function `p.adjust` in the R package *stats* (v4.2.1).
+The procedure was implemented by the function `p.adjust` in the R package *stats* (v4.3.3).
 

### 2024-05-30 analysis/02_rarefy_timeseries.Rmd
Accidentally erased at some point

@@
 ##	Get trait filtered data, rename columns to jive with the rest of the script
-bt_traitfiltered <- bt_traitfiltered
+bt_traitfiltered <- bt_traitfiltered %>%
   rename_with(toupper) %>%

### 2024-06-25 paper/GEB Submission/manuscript_GEB.Rmd
Final perry edits

@@
 Study characteristics such as taxa, realm, biome, or protection status did not distinguish between types of change exhibited by communities.
-We found evidence of an increasing temporal trend in fish consumption across all communities with , but no evidence to support multiple predictions for specific traits, including decreasing body size, dietary shifts, or changes in foraging strata.
+We found evidence of an increasing temporal trend in fish consumption across all communities but no evidence to support multiple predictions for specific traits, including decreasing body size, dietary shifts, or changes in foraging strata.
 
@@
 Pulse disturbances have a sudden one time impact on the community, changing abundance distributions (functional evenness or divergence) or reducing the volume of the functional space [@boersma2016; @hillebrand2020; @larson2021].
-While press disturbances can also lead to loss of functional volume as species fail to adapt to new conditions, it can also lead to wholesale shifts in the location of the functional space as species fill in new environmental space.
+While press disturbances can also lead to loss of functional volume as species fail to adapt to new conditions, they can also lead to wholesale shifts in the location of the functional space as species fill in new environmental space.
 Rather than identifying outcomes of one of these specific processes, observational time series of functional diversity therefore capture the cumulative affects of multiple potentially competing processes.
@@
 `r colorize("We have stronger expectations for changes in the prevalence of some individual traits.
-For example, animal body size is expected to decrease as a result of climate change, a phenomena that has been documented in multiple taxa empirically and experimentally", "blue")` [@caruso2014; @forster2012; @huss2019; @sheridan2011; @tseng2018].
-`r colorize("For dietary traits, recent work documenting insect declines", "blue")` [@wagner2021] `r colorize("points to potentially significant negative impacts on insectivorous animals", "blue")` [@canaday1996; @sekercioglu2002; @tallamy2021].
+For example, animal body size is expected to decrease as a result of climate change, a phenomenon that has been documented in multiple taxa empirically and experimentally", "blue")` [@caruso2014; @forster2012; @huss2019; @sheridan2011; @tseng2018].
+`r colorize("For dietary traits, recent work documenting insect declines", "blue")` [@wagner2021] `r colorize("points to potentially profound negative impacts on insectivorous animals", "blue")` [@canaday1996; @sekercioglu2002; @tallamy2021].
 `r colorize("Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species ", "blue")`[@cooke2019].
@@
 We evaluate 1) whether or not there is a prevailing type of functional change across communities and if community characteristics explain variation in those trends, 2) hypotheses of change for individual trait axes, and 3) the implications of study-level trends for the ecology of those communities.
-To address these goals, we present results for general trends that aggregate across all communities, categorical trends that aggregate across communities with similar characteristics, and study-level trends.
+To address these goals, we present results for general trends that aggregate across all communities, category-specifictrends that aggregate across communities with similar characteristics, and study-level trends.
 
@@
 We used trait data from the Elton Trait Database, which consists of species-level means for traits that represent species' multifaceted role in the community [@wilman2014].
-Traits include: body mass, diet, nocturnality, forest foraging strata, pelagic use.
+Traits include: body mass, diet, nocturnality, forest foraging strata, and pelagic use.
 For categorical traits, different category levels are represented as binary or percentage use (Table \ref{tab:traitTable}).
@@
 `r colorize("41 of the 2432 time series included in our dataset had a variable number of samples within years.", "blue")` To account for this inconsistency in sampling effort we used sample-based rarefaction by bootstrap resampling within years for each time series based on the smallest number of samples in a year for that time series.
-`r colorize("We made 200 rarefied samples for over sampled years and calculated diversity metrics for each sample, then took the median for each metric as the value for that year and time series.
-28 of the affect studies required under sampling by only one sample, and only three time series required under sampling by more than ten samples.", "blue")`
+`r colorize("We made 200 rarefied samples for over-sampled years and calculated diversity metrics for each sample, then took the median for each metric as the value for that year and time series.
+28 of the rarefied studies required under-sampling by only one sample, and only three time series required under-sampling by more than ten samples.", "blue")`
 
@@
 SES estimates can be interpreted as how much of the functional characteristic (richness, evenness, divergence) was observed beyond what was expected by chance for a community of that species richness.
+Hereafter we refer to the *SES* as the corrected metric in comparison to the observed metric.
 `r colorize("This approach will be less accurate for shorter time series, as we likely will not have captured all available species in the true species pool, but it is impossible to know whether the mean estimate from the null model is an over or under estimate without knowing the functional characteristics of the missing species.", "blue")`
@@
 
-We estimated general trends `r colorize("across bird and mammals communities", "blue")` for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each location nested within the study, `r colorize("methods which deal well with the inherent imbalances in our data.", "blue")` We fit 18 individual *CWM* models, `r colorize("one for each trait included in the analsysis", "blue")`.
+We estimated general trends `r colorize("across bird and mammals communities", "blue")` for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each location nested within the study, `r colorize("methods which deal well with the inherent imbalances in our data.", "blue")` We fit 18 individual *CWM* models, `r colorize("one for each trait included in the analysis", "blue")`.
 All time series with data for a given trait were included in the corresponding *CWM* model.
 We estimated study level trends using individual linear models.
-For studies with more than one location we fit a random slope and intercept for time series.
-Some study-level models could not be fit for five studies for at least one metric due to data limitations, but those studies were still included in the general models.
+For studies with more than one location we fit a random slope and intercept for location.
+Study-level models could not be fit for five studies for at least one metric due to data limitations, but those studies were still included in the general models.
 They represented 13 of 1350 study-level models fit for each metric.
@@
 
-To test for categorical trends within and between different levels of taxa, biome, realm, and protection status we fit separate models with each of those factors added as a predictor interacting with time to the original model structure.
+To test for category-specifictrends within and between different levels of taxa, biome, realm, and protection status we fit separate models with each of those factors added as a predictor interacting with time to the original model structure.
 We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
@@
 
-To correct for false positives due to multiple testing, we assess significance of the trends for the fixed effect of time based on adjusted p-values.
+To correct for false positives due to multiple testing, we assess significance of temporal trends based on adjusted p-values.
 Adjusted p-values were calculated using the Benjamini-Hochberg procedure, which controls the false discovery rate.
@@
 Non-significant overall `r colorize("temporal", "blue")` trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies `r colorize("was not significantly different from zero", "blue")` (Table \ref{tab:resultsTab}).
-`r colorize("Categorical", "blue")` trends for different taxa, biomes, realms, or protection statuses were also non-significant for species richness and functional diversity metrics.
+Trends by category for different taxa, biomes, realms, or protection statuses were also non-significant for species richness and functional diversity metrics.
 
-```{r timeseriesPlot, fig.cap="Plots of time series-level trends with line color corresponding to climatic region, with data points in grey and the overall mean slope for a metric in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES", echo=FALSE}
+```{r timeseriesPlot, fig.cap="Plots of study-level trends with line color corresponding to climatic region, with data points in grey and the overall mean slope for a metric in black for A) log species richness, B) Jaccard similarity, C) Functional Richness SES, D) Functional Divergence SES, and E) Functional Evenness SES", echo=FALSE}
 knitr::include_graphics(here::here("figures/3met_long.jpeg"))
@@
 
-We did find significant categorical trends for Jaccard similarity and two dietary *CWM's*.
-While Jaccard similarity was decreasing in the general trend and there were significant categorical slopes for birds, terrestrial, and temperate studies, there was no significant slope for marine, mammal, or tropical studies.
+We did find significant category-specific trends for Jaccard similarity and two dietary *CWM's*.
+While Jaccard similarity was decreasing in the general trend and there were significant category-specific trends for birds, terrestrial, and temperate studies, there was no significant trend for marine, mammal, or tropical studies.
 We also found a significant decrease in Jaccard similarity in unprotected areas only, with no trend for protected areas.
 We found significant dietary shifts across communities, with a significant general trend of increasing fish consumption, which was also reflected in increasing fish consumption trends for bird studies and temperate studies.
-We found significantly decreasing trend for vertebrate consumption in Marine studies and Tropical studies.
+We found significantly decreasing trend for vertebrate consumption in marine studies and tropical studies.
 The trend for marine vertebrate consumption was significantly more negative than for terrestrial studies and the temperate trend was significantly more negative than tropical studies.
@@
 At the study level, 11 of 50 studies exhibited a significant trend in species richness and 11 exhibited significant turnover.
-For uncorrected functional diversity metrics, 11 of 50 studies exhibited a trend in a least one metric, and 7 of 50 studies exhibited a significant trend for at least one `r colorize("corrected", "blue")` metric (Table \ref{tab:trendTab}).
-In general, there were more significant trends for uncorrected metrics, with some disappearing after correction, indicating that those trends were likely due to changes in the number of species.
+For observed functional diversity metrics, 11 of 50 studies exhibited a trend in at least one metric, and 7 of 50 studies exhibited a significant trend for at least one `r colorize("corrected", "blue")` metric (Table \ref{tab:trendTab}).
+In general, there were more significant trends for observed metrics, with some disappearing after correction, indicating that those trends were likely due to changes in the number of species.
 Study-level slopes were not significantly related to start year or duration of time series for any metric.
@@
         caption = "Number of studies that experienced a significant trend in each calculated metric out of 50 total studies.
-Both uncorrected and corrected (SES) versions of functional metrics are included.",
+Both observed and corrected (SES) versions of functional metrics are included.",
         booktabs = TRUE, align = "c", escape = FALSE) %>%
@@
 Our study represents the largest broad-scale multi-taxa assessment of functional change through time to date, giving a first look at general, categorical, and study-level local trends in functional diversity in mammal and bird communities.
-Our work complements efforts on a similar scale assessing temporal change in species and phyllogenetic diversity across taxa [@daru2021; @dornelas2014; @li2020; @vellend2013; @vellend2017] or functional diversity change for a single taxa [@barnagaud2017; @jarzyna2016; @trindade-santos2020].
-`r colorize("We did not detect an overall trend in any functional diversity metric, corrected or uncorrected.
+Our work complements efforts on a similar scale assessing temporal change in species and phylogenetic diversity across taxa [@daru2021; @dornelas2014; @li2020; @vellend2013; @vellend2017] or functional diversity change for a single taxa [@barnagaud2017; @jarzyna2016; @trindade-santos2020].
+`r colorize("We did not detect an overall trend in any functional diversity metric, corrected or observed.
 As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing dissimilarity through time", "blue")` [@dornelas2018]`r colorize(", indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends.", "blue")`
 
-`r colorize("Categorical trends in turnover highlight the types of communities experiencing change in community composition.
+`r colorize("category-specifictrends in turnover highlight the types of communities experiencing change in community composition.
 For example, we found significant turnover for most biomes, realms, and taxa with the exception of marine, tropical, and mammal studies.
@@
 Contrary to the expectation due to anthropogenic and global change stressors, these communities do not show significant changes over the course of the observation window.
-Studies in this group span the distribution of study durations excluding only the very longest running studies, with the longest no change time series lasting 33 years.
+Studies in this group span the distribution of study durations, excluding only the very longest running studies, with the longest no-change time series lasting 33 years.
 They also included both bird and mammal studies and only seven were located in protected areas, indicating that the lack of trend is not restricted to a specific ecological context or those communities most insulated from human impact.", "blue")`
@@
 However, with the addition of a functional perspective we found multiple kinds of underlying functional change including increases and decreases in corrected functional richness and divergence and loss of corrected functional evenness.
-We therefore found evidence of both increases and decreases in the functional capacity of communities independent of species-based metrics, reinforcing the importance of moving beyond soley species-based approaches for biodiversity assessment", "blue")` [@hillebrand2017; @santini2017].
+We therefore found evidence of both increases and decreases in the functional capacity of communities independent of species-based metrics, reinforcing the importance of moving beyond solely species-based approaches for biodiversity assessment", "blue")` [@hillebrand2017; @santini2017].
 All but one of the studies in this group experienced significant turnover, which was also reflected in trends for multiple *CWM*'s, indicating that the overall functional space was shifting even when some characteristics of the functional space were maintained.
 
-`r colorize("9 studies showed loss of functional redundancy.
+`r colorize("Nine studies showed loss of functional redundancy.
 These studies are characterized by a loss of species richness with no shift in functional diversity metrics, indicating the lost species were functionally similar enough to another species to not impact the overall functional diversity.
-Redundancy is critical for maintaining ecological processes in the face of disturbance as it ensures  functional roles are maintained", "blue")` [@biggs2020]`r colorize(", and a loss of functional redundancy likeley reduces a community's capacity to respond to future stressors ", "blue")`[@reich2012]`r colorize(".
+Redundancy is critical for maintaining ecological processes in the face of disturbance as it ensures  functional roles are maintained", "blue")` [@biggs2020]`r colorize(", and a loss of functional redundancy likely reduces a community's capacity to respond to future stressors ", "blue")`[@reich2012]`r colorize(".
 These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future", "blue")` [@pimiento2020a].
 
-`r colorize("The final two studies showed an increase in species richness and uncorrected functional richness, with only one showing an increase in functional richness beyond what was expected by chance (corrected functional richness).
-While this can be an indication of increasing functional capacity or reestabilishment of previously lost species, it can also be the result of less desirable process like the introduction of invasive species.
+`r colorize("The final two studies showed an increase in species richness and observed functional richness, with only one showing an increase in functional richness beyond what was expected by chance (corrected functional richness).
+While this can be an indication of increasing functional capacity or reestabilishment of previously lost species, it can also be the result of less desirable processes like the introduction of invasive species.
 Notably, both of these studies were from temperate, terrestrial bird communities in Scandinavian countries with a unique environment and conservation context.
@@
 `r colorize("One of the biggest threats to biodiversity is the wholesale conversion of natural areas to urban or human-dominated landscapes ", "blue")`[@tilman2017].
-`r colorize("Typical long term monitoring such as those included in our study stops before this conversion occurs, leaving the resultant precipitous declines in biodiversity unrecorded", "blue")` [@cardinale2014].
-`r colorize("This is a known issue with the culture of long term monitoring, and our results should not be removed from that context.
-Rather, this studies captures communities that are likely experiencing a degree of human intervention, but are still largely nature-dominated.", "blue")`
+`r colorize("Typical long-term monitoring data such as those included in our study stops before this conversion occurs, leaving the resultant precipitous declines in biodiversity unrecorded", "blue")` [@cardinale2014].
+`r colorize("This is a known issue with the culture of long-term monitoring, and our results should not be removed from that context.
+Rather, this study captures communities that are likely experiencing a degree of human intervention, but are still largely nature-dominated.", "blue")`
 

### 2024-06-25 analysis/figures/visualization.Rmd
Fixing overplotting once and for all.

@@
   coord_map(xlim=c(-180,180), ylim=c(-57, 90), projection = "mercator") +
-  geom_point(data = study_data,
+  geom_point(data = study_data %>% filter(taxa == "Birds"),
+             aes(rarefyID_x, rarefyID_y, color = taxa), alpha = 0.4, inherit.aes = FALSE) +
+  geom_point(data = study_data %>% filter(taxa == "Mammals"),
              aes(rarefyID_x, rarefyID_y, color = taxa), alpha = 0.4, inherit.aes = FALSE) +

### 2024-09-23 paper/GEB Submission/manuscript_GEB.Rmd
pre-submission minor changes

@@
 ```{r traitTable, echo=FALSE}
+
 tibble(Trait = c(rep("Diet", 10), rep("Foraging Strata", 7), rep("Pelagic Specialist", 2), rep("Nocturnal", 2), rep("Crepuscular", 2), rep("Diurnal", 2), "Body Mass"),
@@
   column_spec(3:4, width = "3cm") %>%
-  kable_paper()
+  kable_paper() #%>%
+  #save_kable(here::here("figures/tables/trait_table.jpeg"))
 ```
@@
 
-We estimated general trends `r colorize("across bird and mammals communities", "blue")` for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each location nested within the study, `r colorize("methods which deal well with the inherent imbalances in our data.", "blue")` We fit 18 individual *CWM* models, `r colorize("one for each trait included in the analysis", "blue")`.
+We estimated general trends `r colorize("across bird and mammal communities", "blue")` for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each location nested within the study, `r colorize("methods which deal well with the inherent imbalances in our data.", "blue")` We fit 18 individual *CWM* models, `r colorize("one for each trait included in the analysis", "blue")`.
 All time series with data for a given trait were included in the corresponding *CWM* model.
@@
         caption = "Table giving the count of number of studies for different kinds of functional change, broken down into no change functional change without species richness change, loss of redundancy, and increases in species richness.
-Trends direction is indicate by a positive or negative sign, with a 0 indicating no trend.", align = "c",
+Trends direction is indicated by a positive or negative sign, with a 0 indicating no trend.", align = "c",
         latex_options= c("scale_down", "hold_position")) %>% 
   collapse_rows(columns = c(1,2)) %>%
-  column_spec(1, width = "4cm")
+  column_spec(1, width = "4cm") #%>%
+  #save_kable(here::here("figures/tables/study_change_table.jpeg"))
 

### 2024-09-23 paper/supp1_sensitivity_outliers.Rmd
pre-submission minor changes

--- 2022-11-03 paper/sensitivity_supp.Rmd
@@
 ---
-title: "Sensitivity Analysis"
+title: "Supplement 1"
 output:
   pdf_document: default
-  html_document: default
+  #word_document: default
 ---
+
+# Sensitivity Analysis
 
@@
 
-tibble(sensitivity_test = c(rep("duration", 3), rep("trait coverage", 2), "amphibians"), 
-       model_id = c("min3", "min4", "min5", "covg85", "covg90", "with_amph"),
+tibble(sensitivity_test = c(rep("duration", 4), rep("trait coverage", 3)), 
+       model_id = c("min3", "min4", "min5", "min10", "covg75", "covg85", "covg90"),
        description = c("Exclude time series with fewer than 3 years",
@@
                        "Exclude time series with fewer than 5 years",
+                       "Exclude time series with fewer than 10 years",
+                       "Exclude time series with less than 75% of species with trait data",
                        "Exclude time series with less than 85% of species with trait data",
-                       "Exclude time series with less than 90% of species with trait data",
-                       "Include Amphibian data"),
-       sig_metrics = c("CWM seed diet*, CWM canopy foraging*, CWM aerial foraging, Jaccard", 
-       "CWM seed diet*, CWM canopy foraging*, FRic*, Jaccard", 
-       "CWM seed diet*, CWM canopy foraging*, Jaccard", rep("Jaccard", 3))
-) %>% kable(format = "latex") %>%
+                       "Exclude time series with less than 90% of species with trait data"),
+       sig_metrics = c("CWM body mass*, CWM fish diet, Jaccard", 
+       "CWM body mass*, CWM fish diet, FRic*, Jaccard", 
+       "CWM body mass*, CWM fish diet, FRic*, Jaccard",
+       "CWM body mass*, CWM fish diet, Jaccard", rep("Jaccard, CWM fish diet", 3))
+) %>% 
+  #kable(format = "latex", caption = "Summary of the results for the sensitivity analyses, including those for duration of time series, trait coverage, and amphibian inclusion.") %>%
+  kable(format = "latex") %>%
   column_spec(3:4, width = "3cm") %>%
@@
   kable_paper() %>%
-  kable_styling(latex_options= c("hold_position"))
+  kable_styling(latex_options= c("hold_position")) #%>%
+  #save_kable(here::here("figures/tables/sensitivity_table.jpeg"))
 ```
+
+# Amphibian Trends
+
+There were 3 amphibian studies comprised of 11 time series in the bioTIME database with sufficient trait coverage to be included in the analysis.
+With such limited data, we excluded amphibians from the main results, but present preliminary trends here.
+Models fit with amphibian, mammal, and bird data combined showed qualitatively similar trend results to those reported in the manuscript that did not include amphibians.
+For models fit with only amphibian data, no general trends were detected for any metric.
+Table 2 gives model estimates for species-based and species corrected metrics.
+
+```{r resultsTab, echo=FALSE, message = FALSE, warning = FALSE}
+readr::read_csv(here::here("figures/amph_model_table.csv")) %>%
+  select(metric, everything(), -REML, -bobyqa) %>%
+  select(-c(n_ts, formula, model)) %>%
+  filter(group != "Residual" | is.na(group), metric %in% c("S", "Jaccard_base", "SES_FRic", "SES_FEve", "SES_FDiv")) %>%
+  mutate(metric = ifelse(metric == "Jaccard_base", "log(Jaccard + 1)", metric),
+         metric = ifelse(metric == "S", "log(Species Richness)", metric),
+         effect = ifelse(effect == "ran_pars", "random", effect),
+         group = case_when(
+           group == "rarefyID:study_id" ~ "time series within study",
+           group == "study_id" ~ "study",
+           group == "Residual" ~ "residual",
+           TRUE ~ group),
+         term = case_when(
+           term == "(Intercept)" ~ "Intercept",
+           term == "year_scaled" ~ "Year",
+           term == "sd__(Intercept)" ~ "SD Intercept",
+           term == "cor__(Intercept).year_scaled" ~ "Corr(Intercept, Year)",
+           term == "sd__year_scaled" ~ "SD Year",
+           term == "sd__Observation" ~ "SD Observation",
+           TRUE ~ term
+         )) %>%
+  mutate(across(c(estimate, std.error, p.value), round, 2),
+         p.value = ifelse(p.value == 0, "<0.001", p.value)) %>%
+  arrange(metric != "Species Richness", metric, !is.na(group), group, term == "Corr(Intercept, Year)", term) %>%
+  dplyr::rename(grouping = group) %>%
+  #kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics for Amphibian time series.") %>%
+  kable(format = "latex") %>%
+  column_spec(3, width = "2cm") %>%
+  collapse_rows(columns = 1:3, valign = "middle") %>%
+  kable_paper() %>%
+  kable_styling(latex_options= c("scale_down", "hold_position")) #%>%
+  #save_kable(here::here("figures/tables/amph_table.jpeg"))
+```
+
+# Missing Individual Study Models
+
+We fit models separately for each study and metric to estimate study-level trends.
+11 models could not be fit because there were to many time series relative to the number of observations, and therefore not enough data to fit the appropriate mixed effects model structure.
+The missing models are in Table 3.
+
+```{r missingTab, echo=FALSE, message = FALSE, warning = FALSE}
+readr::read_csv(here::here("paper/missing_indv_studies.csv")) %>%
+  rename(`n time series` = nrarefyID, `n observations` = n) %>%
+  #kable(format = "latex", caption = "Model estimates and statistics for general trend models for species richness, Jaccard similarity, and standardized functional diversity metrics for Amphibian time series.") %>%
+  kable(format = "latex") %>%
+  kable_paper() %>%
+  kable_styling(latex_options= c("scale_down", "hold_position"))# %>%
+  #save_kable(here::here("figures/tables/missing_models_table.jpeg"))
+```

### 2024-09-23 paper/GEB Submission/manuscript_GEB.Rmd
Get rid of text coloring from previous revision

@@
 While communities are clearly changing, our most common species-based approaches do not fully capture the nature of that change.
-`r colorize("Still, using general trends derived from limited data as a diagnostic for the state of biodiversity is a topic of ongoing debate", "blue")` [@boënnec2024; @johnson2024].
+Still, using general trends derived from limited data as a diagnostic for the state of biodiversity is a topic of ongoing debate [@boënnec2024; @johnson2024].
 Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
-Many of these criticisms reflect limitations of ecological data on the whole, `r colorize("leading to a call for additional data not only to fill geographic and temporal gaps, but to fill in key characteristics of communities", "blue")` [@dornelas2023; @primack2018].
+Many of these criticisms reflect limitations of ecological data on the whole, leading to a call for additional data not only to fill geographic and temporal gaps, but to fill in key characteristics of communities [@dornelas2023; @primack2018].
 
@@
 
-`r colorize("The expectation for functional diversity change across communities is not obvious from past work, and may or may not follow species richness trends", "blue")` [@dornelas2023; @larson2021].
-`r colorize("While loss of functional diversity is frequently cited as one of the most pressing concerns of the anthropocene ", "blue")`[@cardinale2012; @dirzo2014; @young2016], observational trends show significant functional loss in some studies [@flynn2009] and no loss even in the most heavily impacted communities for others [@edwards2013; @larsen2018; @matuoka2020].
-`r colorize("Broad-scale observational studies are limited, but show for example functional richness increases for North American birds", "blue")` [@barnagaud2017; @jarzyna2016].
-`r colorize("We have stronger expectations for changes in the prevalence of some individual traits.
-For example, animal body size is expected to decrease as a result of climate change, a phenomenon that has been documented in multiple taxa empirically and experimentally", "blue")` [@caruso2014; @forster2012; @huss2019; @sheridan2011; @tseng2018].
-`r colorize("For dietary traits, recent work documenting insect declines", "blue")` [@wagner2021] `r colorize("points to potentially profound negative impacts on insectivorous animals", "blue")` [@canaday1996; @sekercioglu2002; @tallamy2021].
-`r colorize("Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species ", "blue")`[@cooke2019].
-`r colorize("Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems", "blue")`[@bain2020]`r colorize(", and loss of neotropical understory foragers even in protected areas", "blue")` [@pollock2022].
+The expectation for functional diversity change across communities is not obvious from past work, and may or may not follow species richness trends [@dornelas2023; @larson2021].
+While loss of functional diversity is frequently cited as one of the most pressing concerns of the anthropocene [@cardinale2012; @dirzo2014; @young2016], observational trends show significant functional loss in some studies [@flynn2009] and no loss even in the most heavily impacted communities for others [@edwards2013; @larsen2018; @matuoka2020].
+Broad-scale observational studies are limited, but show for example functional richness increases for North American birds [@barnagaud2017; @jarzyna2016].
+We have stronger expectations for changes in the prevalence of some individual traits.
+For example, animal body size is expected to decrease as a result of climate change, a phenomenon that has been documented in multiple taxa empirically and experimentally [@caruso2014; @forster2012; @huss2019; @sheridan2011; @tseng2018].
+For dietary traits, recent work documenting insect declines [@wagner2021] points to potentially profound negative impacts on insectivorous animals [@canaday1996; @sekercioglu2002; @tallamy2021].
+Predicted extinctions based on species-level vulnerability point to further dietary shifts, favoring increases in invertivorous species [@cooke2019].
+Some systems also show significant shifts in the prevalence of different kinds of foragers in birds, for example loss of arboreal foragers in agricultural systems[@bain2020], and loss of neotropical understory foragers even in protected areas [@pollock2022].
 
-Here we `r colorize("leverage ongoing efforts to assemble functional trait data and recent computational advances to", "blue")` perform the first multi-taxa, multi-realm assessment of functional diversity and composition change through time.
-We focus on mammal and bird species as `r colorize("subsets of the world's biodiversity of particular conservation concern that is heavily impacted by anthropogenic change.", "blue")`While examining trends in plants, invertebrates, and other vertebrate species is of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
+Here we leverage ongoing efforts to assemble functional trait data and recent computational advances to perform the first multi-taxa, multi-realm assessment of functional diversity and composition change through time.
+We focus on mammal and bird species as subsets of the world's biodiversity of particular conservation concern that is heavily impacted by anthropogenic change.
+While examining trends in plants, invertebrates, and other vertebrate species is of equal interest, trait data for those taxa raise additional challenges such as limited and biased species coverage [@fitzjohn2014], a lack of accepted species-level means, and differences in the types of traits collected.
 To ensure comparability across taxa in trait type and data quality we therefore focus on mammals and birds.
-`r colorize("We include body mass, dietary, foraging and other behavioral traits", "blue")` that were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].
+We include body mass, dietary, foraging and other behavioral traits that were intentionally selected to be representative of a species' Eltonian niche, thereby summarizing the functional role they play in the community [@wilman2014].
 
@@
 
-We obtained mammal and bird time series from the BioTIME database, a global repository of high-quality assemblage time series `r colorize("collected from the literature and ongoing monitoring efforts", "blue")`.
-`r colorize("Data are structured such that a study comprises all data collected following consistent sampling protocols and may contain multiple time series from separate locations.", "blue")` Time series represent full assemblages rather than populations of single species [@dornelas2018], and all time series include abundance of observed species.
+We obtained mammal and bird time series from the BioTIME database, a global repository of high-quality assemblage time series collected from the literature and ongoing monitoring efforts.
+Data are structured such that a study comprises all data collected following consistent sampling protocols and may contain multiple time series from separate locations.
+Time series represent full assemblages rather than populations of single species [@dornelas2018], and all time series include abundance of observed species.
 Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual time series following a standardized spatial scale.
 Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018 for details on how sample extents were defined], with the area of each cell set to one standard deviation away from the mean of the single extent locations.
-`r colorize("The resulting cell size for our data was approximately 95 ", "blue")`km^2^.
+The resulting cell size for our data was approximately 95 km^2^.
 All samples from a study within a single cell were considered to be a single time series location, and species abundances were combined for all samples.
@@
 
-`r colorize("41 of the 2432 time series included in our dataset had a variable number of samples within years.", "blue")` To account for this inconsistency in sampling effort we used sample-based rarefaction by bootstrap resampling within years for each time series based on the smallest number of samples in a year for that time series.
-`r colorize("We made 200 rarefied samples for over-sampled years and calculated diversity metrics for each sample, then took the median for each metric as the value for that year and time series.
-28 of the rarefied studies required under-sampling by only one sample, and only three time series required under-sampling by more than ten samples.", "blue")`
+41 of the 2432 time series included in our dataset had a variable number of samples within years.
+To account for this inconsistency in sampling effort we used sample-based rarefaction by bootstrap resampling within years for each time series based on the smallest number of samples in a year for that time series.
+We made 200 rarefied samples for over-sampled years and calculated diversity metrics for each sample, then took the median for each metric as the value for that year and time series.
+28 of the rarefied studies required under-sampling by only one sample, and only three time series required under-sampling by more than ten samples.
 
@@
 The earliest sample was in 1923 and the most recent was in 2014.
-`r colorize("While it is not possible with available data to directly assess the level of human impact occurring for each study, we include binary protection status as a coarse indicator of impact level.
-However, protected areas were almost exclusively from temperate terrestrial studies (with one tropical study), so results are confounded by multiple other study characteristics.", "blue")` For a full breakdown of studies and their characteristics, see the supplement.
+While it is not possible with available data to directly assess the level of human impact occurring for each study, we include binary protection status as a coarse indicator of impact level.
+However, protected areas were almost exclusively from temperate terrestrial studies (with one tropical study), so results are confounded by multiple other study characteristics.
+For a full breakdown of studies and their characteristics, see the supplement.
 Our final dataset reflects many of the data biases that make global synthesis work challenging, including geographic bias, a bias away from areas currently under the greatest threat, and a bias towards shorter time series.
@@
 Jaccard similarity was calculated relative to the first observed year for a time series.
-A negative trend in *J* would therefore indicate `r colorize("decreasing similiarity", "blue")`.
+A negative trend in *J* would therefore indicate decreasing similiarity.
 We did not impose a correction for unobserved species as non-parametric estimators do not assign species identities to corrected richness values, and therefore could not be propagated to the functional diversity metrics.
@@
 
-We also calculated the community-weighted mean (*CWM*) of all continuous traits to examine `r colorize("changes", "blue")` in the distribution of each trait.
-`r colorize("Wholesale shifts in the trait space due to changes of trait means could occur even while the shape of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained.
+We also calculated the community-weighted mean (*CWM*) of all continuous traits to examine changes in the distribution of each trait.
+Wholesale shifts in the trait space due to changes of trait means could occur even while the shape of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained.
 CWM's are therefore a way to assess whether or not turnover is occurring and what the nature of the shift may be.
-Hereafter we refer to results for functional metrics in two groups: functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).", "blue")`
+Hereafter we refer to results for functional metrics in two groups: functional diversity metrics (FRic, FEve, FDiv) and composition metrics (trait CWM's).
 
@@
 Continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
-`r colorize("Before calculating diversity metrics, dbFD reduces the dimensionality of the trait space by performing PCoA.
-We limited the number of included PCoA axes to", "blue")` the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
+Before calculating diversity metrics, dbFD reduces the dimensionality of the trait space by performing PCoA.
+We limited the number of included PCoA axes to the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
 This restriction allows for enough axes to capture the trait space while maintaining computational feasibility [@blonder2018].
@@
 Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species.
-`r colorize("Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability", "blue")`.
+Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability.
 This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
@@
 Hereafter we refer to the *SES* as the corrected metric in comparison to the observed metric.
-`r colorize("This approach will be less accurate for shorter time series, as we likely will not have captured all available species in the true species pool, but it is impossible to know whether the mean estimate from the null model is an over or under estimate without knowing the functional characteristics of the missing species.", "blue")`
+This approach will be less accurate for shorter time series, as we likely will not have captured all available species in the true species pool, but it is impossible to know whether the mean estimate from the null model is an over or under estimate without knowing the functional characteristics of the missing species.
 
@@
 
-We estimated general trends `r colorize("across bird and mammal communities", "blue")` for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each location nested within the study, `r colorize("methods which deal well with the inherent imbalances in our data.", "blue")` We fit 18 individual *CWM* models, `r colorize("one for each trait included in the analysis", "blue")`.
+We estimated general trends across bird and mammal communities for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each location nested within the study, methods which deal well with the inherent imbalances in our data.
+We fit 18 individual *CWM* models, one for each trait included in the analysis.
 All time series with data for a given trait were included in the corresponding *CWM* model.
@@
 We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
-`r colorize("For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.", "blue")` We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
+For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.
+We assessed the impact of time series duration and start year on study-level trends using linear models with duration and start year as predictors.
 All models in our analysis were fit using the *lme4* (v1.1-35.3) package in R (v4.3.3) and p-values were calculated by Satterthwaite's degrees of freedom method using the *lmerTest* (v3.1-3) package with a significance level of $\alpha = 0.05$ [@bates2015; @kuznetsova2017; @rcoreteam2023].
@@
 
-We found no significant general `r colorize("temporal", "blue")` trend in species richness or functional diversity metrics `r colorize("including functional richness, evenness, or divergence", "blue")` (observed or `r colorize("corrected", "blue")`) (Fig \ref{fig:timeseriesPlot}).
+We found no significant general temporal trend in species richness or functional diversity metrics including functional richness, evenness, or divergence (observed or corrected) (Fig \ref{fig:timeseriesPlot}).
 We did find a significant overall decrease in Jaccard similarity, indicating accumulating changes in species composition.
-Non-significant overall `r colorize("temporal", "blue")` trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies `r colorize("was not significantly different from zero", "blue")` (Table \ref{tab:resultsTab}).
+Non-significant overall temporal trends indicate that although some studies experience increasing or decreasing trends, the average trend across studies was not significantly different from zero (Table \ref{tab:resultsTab}).
 Trends by category for different taxa, biomes, realms, or protection statuses were also non-significant for species richness and functional diversity metrics.
@@
 At the study level, 11 of 50 studies exhibited a significant trend in species richness and 11 exhibited significant turnover.
-For observed functional diversity metrics, 11 of 50 studies exhibited a trend in at least one metric, and 7 of 50 studies exhibited a significant trend for at least one `r colorize("corrected", "blue")` metric (Table \ref{tab:trendTab}).
+For observed functional diversity metrics, 11 of 50 studies exhibited a trend in at least one metric, and 7 of 50 studies exhibited a significant trend for at least one corrected metric (Table \ref{tab:trendTab}).
 In general, there were more significant trends for observed metrics, with some disappearing after correction, indicating that those trends were likely due to changes in the number of species.
@@
 Our work complements efforts on a similar scale assessing temporal change in species and phylogenetic diversity across taxa [@daru2021; @dornelas2014; @li2020; @vellend2013; @vellend2017] or functional diversity change for a single taxa [@barnagaud2017; @jarzyna2016; @trindade-santos2020].
-`r colorize("We did not detect an overall trend in any functional diversity metric, corrected or observed.
-As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing dissimilarity through time", "blue")` [@dornelas2018]`r colorize(", indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends.", "blue")`
+We did not detect an overall trend in any functional diversity metric, corrected or observed.
+As with previous species-based syntheses, we also found no overall trend in species richness accompanied by increasing dissimilarity through time [@dornelas2018], indicating that non-significant trends in functional metrics are consistent with similar well-documented species derived trends.
 
-`r colorize("category-specifictrends in turnover highlight the types of communities experiencing change in community composition.
+category-specifictrends in turnover highlight the types of communities experiencing change in community composition.
 For example, we found significant turnover for most biomes, realms, and taxa with the exception of marine, tropical, and mammal studies.
-The lack of turnover in marine studies stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial", "blue")` [@blowes2019].
-`r colorize("However, previous global estimates are dominated by fish communities which we exclude here and are likely driving the overall trend while disguising relative stasis in marine bird and mammal communities.
-We also found significant turnover for unprotected areas and no significant turnover in protected areas, indicating that in general unprotected communities are experiencing bigger shifts in community composition through time.", "blue")`
+The lack of turnover in marine studies stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial [@blowes2019].
+However, previous global estimates are dominated by fish communities which we exclude here and are likely driving the overall trend while disguising relative stasis in marine bird and mammal communities.
+We also found significant turnover for unprotected areas and no significant turnover in protected areas, indicating that in general unprotected communities are experiencing bigger shifts in community composition through time.
 
-`r colorize("Surprisingly, we found no evidence for some of the strongest predictions for anthropogenic impact.
+Surprisingly, we found no evidence for some of the strongest predictions for anthropogenic impact.
 For example, we found no trend for mean body mass in bird or mammal communities.
 In fact, sensitivity results indicated that after excluding the shortest time series there is evidence of an increasing trend in body size.
-This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization or carnivore restablishment are overshadowing climate change impacts", "blue")`[@hantak2021]`r colorize(", or current shifts are happening at an intraspecific level not captured by our data.
-Additionally, the majority of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing", "blue")` [@fritz2009].
-`r colorize("Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened", "blue")` [@ripple2015].
+This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization or carnivore restablishment are overshadowing climate change impacts[@hantak2021], or current shifts are happening at an intraspecific level not captured by our data.
+Additionally, the majority of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing [@fritz2009].
+Trends could be significantly different for the same time periods in regions of sub-Saharan Africa, for example, which has poor representation in our dataset but where megafauna exist on the landscape and are increasingly threatened [@ripple2015].
 
-`r colorize("We similarly did not find evidence of changes in trait distributions predicted by previous global change work.
+We similarly did not find evidence of changes in trait distributions predicted by previous global change work.
 We found no changes in the prevalence of different foraging strategies, despite documented losses of understory birds in the neotropics and some evidence that agricultural incursion particularly threatens arboreal species.
@@
 Rather than declines in insectivory and increases in invertivory predicted by changes in resource availability, we instead found changes in the degree of consumption of fish and vertebrates.
-Neither increasing trends for fish consumption nor decreasing trends for vertebrate consumption in marine and tropical studies are obviously consistent with known changes in those contexts and warrant further examination.", "blue")`
+Neither increasing trends for fish consumption nor decreasing trends for vertebrate consumption in marine and tropical studies are obviously consistent with known changes in those contexts and warrant further examination.
 
@@
 
-`r colorize("While trends aggregating across studies did not indicate one prevailing kind of functional change, post hoc examination of trends for individual studies revealed four distinct change groups.
-These groups are categorized based on the concurrent trends in species and functional metrics and can be broadly described as: no change in species or functional diversity metrics, functional change independent of species richness, loss of redundancy, and increase in species richness", "blue")` (Table \ref{tab:studyChange}).
-`r colorize("The majority of studies (32 studies) exhibited no trend in any species or functional diversity metric.
+While trends aggregating across studies did not indicate one prevailing kind of functional change, post hoc examination of trends for individual studies revealed four distinct change groups.
+These groups are categorized based on the concurrent trends in species and functional metrics and can be broadly described as: no change in species or functional diversity metrics, functional change independent of species richness, loss of redundancy, and increase in species richness (Table \ref{tab:studyChange}).
+The majority of studies (32 studies) exhibited no trend in any species or functional diversity metric.
 Contrary to the expectation due to anthropogenic and global change stressors, these communities do not show significant changes over the course of the observation window.
 Studies in this group span the distribution of study durations, excluding only the very longest running studies, with the longest no-change time series lasting 33 years.
-They also included both bird and mammal studies and only seven were located in protected areas, indicating that the lack of trend is not restricted to a specific ecological context or those communities most insulated from human impact.", "blue")`
+They also included both bird and mammal studies and only seven were located in protected areas, indicating that the lack of trend is not restricted to a specific ecological context or those communities most insulated from human impact.
 
-`r colorize("The lack of trend could be the result of multiple possible scenarios.
+The lack of trend could be the result of multiple possible scenarios.
 First, these may be communities resisting perturbations or simply not experiencing significant perturbations.
@@
 Alternatively, these may be communities that have experienced or continue to experience significant stress, but lost species or functional diversity outside the observation window.
-This could be true particularly for North American mammal communities where trophic downgrading and megafaunal losses occurred hundreds of years ago ", "blue")`[@estes2011].
-`r colorize("Third, these communities may be experiencing directional shifts undetectable by available data.
-For example, species-level trait data do not capture intraspecific shifts in the trait space, which can have significant implications for ecological processes.", "blue")`
+This could be true particularly for North American mammal communities where trophic downgrading and megafaunal losses occurred hundreds of years ago [@estes2011].
+Third, these communities may be experiencing directional shifts undetectable by available data.
+For example, species-level trait data do not capture intraspecific shifts in the trait space, which can have significant implications for ecological processes.
 
-`r colorize("The next largest group of studies (7 studies) were those with no species richness trend that showed a trend in at least one functional diversity metric.
+The next largest group of studies (7 studies) were those with no species richness trend that showed a trend in at least one functional diversity metric.
 Based on a species approach alone, these communities apear to either be experiencing no change or some change in species identity, with unknown implications for conservation concern or ecological process.
 However, with the addition of a functional perspective we found multiple kinds of underlying functional change including increases and decreases in corrected functional richness and divergence and loss of corrected functional evenness.
-We therefore found evidence of both increases and decreases in the functional capacity of communities independent of species-based metrics, reinforcing the importance of moving beyond solely species-based approaches for biodiversity assessment", "blue")` [@hillebrand2017; @santini2017].
+We therefore found evidence of both increases and decreases in the functional capacity of communities independent of species-based metrics, reinforcing the importance of moving beyond solely species-based approaches for biodiversity assessment [@hillebrand2017; @santini2017].
 All but one of the studies in this group experienced significant turnover, which was also reflected in trends for multiple *CWM*'s, indicating that the overall functional space was shifting even when some characteristics of the functional space were maintained.
 
-`r colorize("Nine studies showed loss of functional redundancy.
+Nine studies showed loss of functional redundancy.
 These studies are characterized by a loss of species richness with no shift in functional diversity metrics, indicating the lost species were functionally similar enough to another species to not impact the overall functional diversity.
-Redundancy is critical for maintaining ecological processes in the face of disturbance as it ensures  functional roles are maintained", "blue")` [@biggs2020]`r colorize(", and a loss of functional redundancy likely reduces a community's capacity to respond to future stressors ", "blue")`[@reich2012]`r colorize(".
-These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future", "blue")` [@pimiento2020a].
+Redundancy is critical for maintaining ecological processes in the face of disturbance as it ensures  functional roles are maintained [@biggs2020], and a loss of functional redundancy likely reduces a community's capacity to respond to future stressors [@reich2012].
+These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future [@pimiento2020a].
 
-`r colorize("The final two studies showed an increase in species richness and observed functional richness, with only one showing an increase in functional richness beyond what was expected by chance (corrected functional richness).
-While this can be an indication of increasing functional capacity or reestabilishment of previously lost species, it can also be the result of less desirable processes like the introduction of invasive species.
+The final two studies showed an increase in species richness and observed functional richness, with only one showing an increase in functional richness beyond what was expected by chance (corrected functional richness).
+While this can be an indication of increasing functional capacity or reestablishment of previously lost species, it can also be the result of less desirable processes like the introduction of invasive species.
 Notably, both of these studies were from temperate, terrestrial bird communities in Scandinavian countries with a unique environment and conservation context.
-Still, results are consistent with other broad-scale studies of bird communities, which found loss of common, functionally general species alongside increases in rare species ", "blue")` [@inger2015; @rosenberg2019; @schipper2016].
+Still, results are consistent with other broad-scale studies of bird communities, which found loss of common, functionally general species alongside increases in rare species  [@inger2015; @rosenberg2019; @schipper2016].
 
@@
 
-`r colorize("While we found no over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
+While we found no over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
 In fact, study-level trends indicate quite the opposite, that functional changes with negative or yet unknown implications for ecosystem processes may be going undetected by common species-based approaches.
-For example, loss of evenness in communities with constant species richness may be an first sign of a community being impacted by environmental change, with negative implications for stability and function.", "blue")`
+For example, loss of evenness in communities with constant species richness may be an first sign of a community being impacted by environmental change, with negative implications for stability and function.
 
-`r colorize("One of the biggest threats to biodiversity is the wholesale conversion of natural areas to urban or human-dominated landscapes ", "blue")`[@tilman2017].
-`r colorize("Typical long-term monitoring data such as those included in our study stops before this conversion occurs, leaving the resultant precipitous declines in biodiversity unrecorded", "blue")` [@cardinale2014].
-`r colorize("This is a known issue with the culture of long-term monitoring, and our results should not be removed from that context.
-Rather, this study captures communities that are likely experiencing a degree of human intervention, but are still largely nature-dominated.", "blue")`
+One of the biggest threats to biodiversity is the wholesale conversion of natural areas to urban or human-dominated landscapes [@tilman2017].
+Typical long-term monitoring data such as those included in our study stops before this conversion occurs, leaving the resultant precipitous declines in biodiversity unrecorded [@cardinale2014].
+This is a known issue with the culture of long-term monitoring, and our results should not be removed from that context.
+Rather, this study captures communities that are likely experiencing a degree of human intervention, but are still largely nature-dominated.
 

### 2024-09-23 paper/GEB Submission/manuscript_GEB.Rmd
Minor spelling changes

@@
 We evaluate 1) whether or not there is a prevailing type of functional change across communities and if community characteristics explain variation in those trends, 2) hypotheses of change for individual trait axes, and 3) the implications of study-level trends for the ecology of those communities.
-To address these goals, we present results for general trends that aggregate across all communities, category-specifictrends that aggregate across communities with similar characteristics, and study-level trends.
+To address these goals, we present results for general trends that aggregate across all communities, category-specific trends that aggregate across communities with similar characteristics, and study-level trends.
 
@@
 Jaccard similarity was calculated relative to the first observed year for a time series.
-A negative trend in *J* would therefore indicate decreasing similiarity.
+A negative trend in *J* would therefore indicate decreasing similarity.
 We did not impose a correction for unobserved species as non-parametric estimators do not assign species identities to corrected richness values, and therefore could not be propagated to the functional diversity metrics.
@@
 
-To test for category-specifictrends within and between different levels of taxa, biome, realm, and protection status we fit separate models with each of those factors added as a predictor interacting with time to the original model structure.
+To test for category-specific trends within and between different levels of taxa, biome, realm, and protection status we fit separate models with each of those factors added as a predictor interacting with time to the original model structure.
 We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
@@
 
-category-specifictrends in turnover highlight the types of communities experiencing change in community composition.
+category-specific trends in turnover highlight the types of communities experiencing change in community composition.
 For example, we found significant turnover for most biomes, realms, and taxa with the exception of marine, tropical, and mammal studies.
@@
 The next largest group of studies (7 studies) were those with no species richness trend that showed a trend in at least one functional diversity metric.
-Based on a species approach alone, these communities apear to either be experiencing no change or some change in species identity, with unknown implications for conservation concern or ecological process.
+Based on a species approach alone, these communities appear to either be experiencing no change or some change in species identity, with unknown implications for conservation concern or ecological process.
 However, with the addition of a functional perspective we found multiple kinds of underlying functional change including increases and decreases in corrected functional richness and divergence and loss of corrected functional evenness.

### 2024-09-24 paper/GEB Submission/manuscript_GEB.Rmd
Line edits

@@
 Study characteristics such as taxa, realm, biome, or protection status did not distinguish between types of change exhibited by communities.
-We found evidence of an increasing temporal trend in fish consumption across all communities but no evidence to support multiple predictions for specific traits, including decreasing body size, dietary shifts, or changes in foraging strata.
+We found evidence of a temporal increase in fish consumption across all communities but no evidence to support multiple predictions for specific traits, including decreasing body size, dietary shifts, or changes in foraging strata.
 
@@
 Still, using general trends derived from limited data as a diagnostic for the state of biodiversity is a topic of ongoing debate [@boënnec2024; @johnson2024].
-Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
+Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, `r colorize("short time windows biased toward the present day", "blue")` and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
 Many of these criticisms reflect limitations of ecological data on the whole, leading to a call for additional data not only to fill geographic and temporal gaps, but to fill in key characteristics of communities [@dornelas2023; @primack2018].
 
-Functional diversity offers a potentially powerful addition to species-based approaches for detecting and describing community change by capturing changes in the traits that underpin fundamental ecological processes [@lavorel2002; @mcgill2006; @suding2008].
-Functional diversity metrics capture complimentary aspects of the trait space including the volume of the multidimensional space (functional diversity), how species and abundance are distributed across the space (functional evenness and divergence), and where the space occurs across the range of possible trait distributions (community weighted trait means).
+Functional diversity offers a powerful addition to species-based approaches for detecting and describing community change by capturing changes in the traits that underpin fundamental ecological processes [@lavorel2002; @mcgill2006; @suding2008].
+Functional diversity metrics capture complimentary aspects of the trait space including the volume of the multidimensional space (functional diversity), how species and abundance are distributed across the space (functional evenness and divergence), and `r colorize("how the space maps to individual trait distributions", "blue")` (community weighted trait means).
 Considered together, they describe the aspects of the trait space that mediate ecological process.
 
-Broad-scale trends in functional diversity are the aggregation of local-scale responses to acute pulse disturbance, persistent press disturbances, and periods of relative stasis [@inamine2022].
+`r colorize("Trends in functional diversity across broad geographic extents", "blue")` are the aggregation of local-scale responses to acute pulse disturbance, persistent press disturbances, and periods of relative stasis [@inamine2022].
 Pulse disturbances have a sudden one time impact on the community, changing abundance distributions (functional evenness or divergence) or reducing the volume of the functional space [@boersma2016; @hillebrand2020; @larson2021].
@@
 
-Our final dataset included 2,432 time series from 50 studies in 21 countries and 12 biomes using 6 different traits (Fig \ref{fig:taxaMap}).
+Our final dataset included 2,432 time series from 50 studies in 21 countries and 12 biomes using `r colorize("7", "blue")` different traits (Fig \ref{fig:taxaMap}).
 Data came from both terrestrial and marine realms and five biomes (Global, Polar/Temperate, Temperate, Temperate/Tropical, Tropical).
@@
 
-We also calculated the community-weighted mean (*CWM*) of all continuous traits to examine changes in the distribution of each trait.
-Wholesale shifts in the trait space due to changes of trait means could occur even while the shape of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained.
+We also calculated the community-weighted mean (*CWM*) of all continuous traits `r colorize("(diet, foraging strata, body mass)", "blue")` to examine changes in the distribution of each trait.
+Wholesale shifts in the `r colorize("observed", "blue")` trait space due to changes of trait means could occur even while the shape `r colorize("and dispersion", "blue")` of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained `r colorize("through time", "blue")`.
 CWM's are therefore a way to assess whether or not turnover is occurring and what the nature of the shift may be.
@@
 We estimated general trends across bird and mammal communities for each diversity metric using a linear mixed effects model with a random slope and intercept for each study and each location nested within the study, methods which deal well with the inherent imbalances in our data.
-We fit 18 individual *CWM* models, one for each trait included in the analysis.
-All time series with data for a given trait were included in the corresponding *CWM* model.
+`r colorize("We fit individual models for each of the 18 CWM's calculated for continuous traits or trait categories (see diet, foraging strata, and body mass traits in Table 1).", "blue")` All time series with data for a given trait were included in the corresponding *CWM* model.
 We estimated study level trends using individual linear models.
@@
 Where appropriate, response variables were $log$ or $log(x+1)$ transformed to better fit model assumptions of residual normality.
+`r colorize("The log transformation was applied to all metrics except the SES corrected functional diversity metrics, FEve, and CWM's for`invertebrate`consumption, ground foraging strata, and below water surface foraging strata.", "blue")`
 
-To test for category-specific trends within and between different levels of taxa, biome, realm, and protection status we fit separate models with each of those factors added as a predictor interacting with time to the original model structure.
-We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
+'r colorize("We tested for category-specific trends in species and functional diversity within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")` We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
 For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.
@@
 
-category-specific trends in turnover highlight the types of communities experiencing change in community composition.
+Category-specific trends in `r colorize("species", "blue")` turnover highlight the types of communities experiencing change in community composition.
 For example, we found significant turnover for most biomes, realms, and taxa with the exception of marine, tropical, and mammal studies.
@@
 Alternatively, these may be communities that have experienced or continue to experience significant stress, but lost species or functional diversity outside the observation window.
-This could be true particularly for North American mammal communities where trophic downgrading and megafaunal losses occurred hundreds of years ago [@estes2011].
+This could be true particularly for North American mammal communities where trophic downgrading and megafaunal losses occurred `r colorize("thousands", "blue")` of years ago [@estes2011].
 Third, these communities may be experiencing directional shifts undetectable by available data.

### 2024-09-26 paper/GEB Submission/manuscript_GEB.Rmd
Finish line edits

@@
 Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, `r colorize("short time windows biased toward the present day", "blue")` and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
-Many of these criticisms reflect limitations of ecological data on the whole, leading to a call for additional data not only to fill geographic and temporal gaps, but to fill in key characteristics of communities [@dornelas2023; @primack2018].
+Many of these criticisms reflect limitations of ecological data on the whole [@chapman2024], leading to a call for additional data not only to fill geographic and temporal gaps, but to fill in key characteristics of communities [@dornelas2023; @primack2018].
 
@@
 Where appropriate, response variables were $log$ or $log(x+1)$ transformed to better fit model assumptions of residual normality.
-`r colorize("The log transformation was applied to all metrics except the SES corrected functional diversity metrics, FEve, and CWM's for`invertebrate`consumption, ground foraging strata, and below water surface foraging strata.", "blue")`
+`r colorize("The log transformation was applied to all metrics except the SES corrected functional diversity metrics, FEve, and CWM's for invertebrate consumption, ground foraging strata, and below water surface foraging strata.", "blue")`
 
-'r colorize("We tested for category-specific trends in species and functional diversity within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")` We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
+'r colorize("We tested for category-specific trends in all diversity metrics within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")\` We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
 For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.
@@
 We did find significant category-specific trends for Jaccard similarity and two dietary *CWM's*.
-While Jaccard similarity was decreasing in the general trend and there were significant category-specific trends for birds, terrestrial, and temperate studies, there was no significant trend for marine, mammal, or tropical studies.
-We also found a significant decrease in Jaccard similarity in unprotected areas only, with no trend for protected areas.
-We found significant dietary shifts across communities, with a significant general trend of increasing fish consumption, which was also reflected in increasing fish consumption trends for bird studies and temperate studies.
-We found significantly decreasing trend for vertebrate consumption in marine studies and tropical studies.
+While Jaccard similarity was decreasing in the general trend and there were significant category-specific `r colorize("turnover", "blue")` trends for birds, terrestrial, and temperate studies, there was no significant trend `r colorize("in Jaccard similarity", "blue")` for marine, mammal, or tropical studies.
+`r colorize("Jaccard similarity decreased", "blue")` in unprotected areas only, with no trend for protected areas.
+We found significant dietary shifts across communities, with a significant increasing general trend in fish consumption, which was also reflected in increasing fish consumption trends for bird studies and temperate studies.
+`r colorize("Vertebrate consumption significantly declined", "blue")` in marine studies and tropical studies.
 The trend for marine vertebrate consumption was significantly more negative than for terrestrial studies and the temperate trend was significantly more negative than tropical studies.
@@
 For observed functional diversity metrics, 11 of 50 studies exhibited a trend in at least one metric, and 7 of 50 studies exhibited a significant trend for at least one corrected metric (Table \ref{tab:trendTab}).
-In general, there were more significant trends for observed metrics, with some disappearing after correction, indicating that those trends were likely due to changes in the number of species.
+In general, there were more significant trends for observed functional diversity metrics, with some disappearing after correction, indicating that those trends were likely due to changes in the number of species.
 Study-level slopes were not significantly related to start year or duration of time series for any metric.
@@
 
-We assessed the sensitivity of general trend results to major data processing decision by rerunning models with increasingly conservative subsets of the data.
+We assessed the sensitivity of general trend results `r colorize("for all metrics", "blue")` to major data processing decision by rerunning models with increasingly conservative subsets of the data.
 Trends for Jaccard similarity and fish consumption were not sensitive to either time series duration or trait coverage.
@@
 The lack of turnover in marine studies stands in contrast to other global estimates of biodiversity change that found higher turnover in marine systems than terrestrial [@blowes2019].
-However, previous global estimates are dominated by fish communities which we exclude here and are likely driving the overall trend while disguising relative stasis in marine bird and mammal communities.
+However, previous global estimates are dominated by fish communities which we exclude here and are likely driving the overall `r colorize("turnover", "blue")` trend while disguising relative stasis in marine bird and mammal communities.
 We also found significant turnover for unprotected areas and no significant turnover in protected areas, indicating that in general unprotected communities are experiencing bigger shifts in community composition through time.
@@
 Surprisingly, we found no evidence for some of the strongest predictions for anthropogenic impact.
-For example, we found no trend for mean body mass in bird or mammal communities.
+For example, we found no trend in mean body mass in bird or mammal communities.
 In fact, sensitivity results indicated that after excluding the shortest time series there is evidence of an increasing trend in body size.
-This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization or carnivore restablishment are overshadowing climate change impacts[@hantak2021], or current shifts are happening at an intraspecific level not captured by our data.
+This result could be an indication that either body size is not changing significantly due to climate change, opposing pressures such as urbanization or carnivore reestablishment are overshadowing climate change impacts[@hantak2021], or current shifts are happening at an intraspecific level not captured by our data.
 Additionally, the majority of the studies in our dataset draw from areas that may have experienced significant loss of large-bodied species before the observation window, with contemporary loss rates slowing [@fritz2009].

### 2024-10-10 paper/GEB Submission/manuscript_GEB.Rmd
Edits edits edits

@@
   Methods: We paired thousands of bird and mammal assemblage time series from the BioTIME database with existing trait data representative of species' functional roles to reconstruct time series of functional diversity and composition metrics.
+Our dataset included 2,432 time series of yearly observations from 50 studies with a maximum spatial extent of 95 km2.
 Using generalized linear mixed models, we estimated general trends in those metrics and trends for individual studies.
@@
 Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual time series following a standardized spatial scale.
-Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @dornelas2018 for details on how sample extents were defined], with the area of each cell set to one standard deviation away from the mean of the single extent locations.
+Scale was set by a global grid with cell size determined based on the sample extent of studies with only a single location [see @blowes2019 for details on how sample extents were defined], with the area of each cell set to one standard deviation away from the mean of the single extent locations.
 The resulting cell size for our data was approximately 95 km^2^.
 All samples from a study within a single cell were considered to be a single time series location, and species abundances were combined for all samples.
+`r colorize("This approach therefore sets an upper bound on the definition of "local" for studies with multiple locations.
+Our dataset included 29 single location studies (average spatial extent of ~4 km2), and 24 multi location studies aggregated by cell.", "blue")`
 
@@
 Continuous traits were z-score scaled to give each trait equal weight in the trait space [@leps2006; @schleuter2010].
-Before calculating diversity metrics, dbFD reduces the dimensionality of the trait space by performing PCoA.
-We limited the number of included PCoA axes to the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
+`r colorize("In preparation for calculating diversity metrics dbFD calculates a trait distance matrix using Gower's distance and reduces the dimensionality of that trait space by performing PCoA.", "blue")` We limited the number of included PCoA axes to the maximum number of traits that fulfills the criteria $s >= 2^t$, where $s$ is the number of species and *t* is the number of traits.
 This restriction allows for enough axes to capture the trait space while maintaining computational feasibility [@blonder2018].
@@
 Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability.
-This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
+`r colorize("This approach assumes that the likelihood of a species being present is equally likely through time and that the species abundance distribution is conserved.", "blue")` This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
 We used these values to calculate SES using the following formula: $SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
@@
 We therefore found evidence of both increases and decreases in the functional capacity of communities independent of species-based metrics, reinforcing the importance of moving beyond solely species-based approaches for biodiversity assessment [@hillebrand2017; @santini2017].
-All but one of the studies in this group experienced significant turnover, which was also reflected in trends for multiple *CWM*'s, indicating that the overall functional space was shifting even when some characteristics of the functional space were maintained.
 
@@
 
+Study-level trends in *CWM*'s offer insight into how the functional space might be shifting through time, and how that functional turnover may or may not relate to species turnover.
+Every study that experienced a trend in a species or functional diversity metric (Table \ref{tab:studyChange}; functional change, loss of redundancy, and increase in species richness study types), with the exception of 3 "loss of redundancy" studies, also exhibited a trend in at least one *CWM*.
+This linking between *CWM* and other metrics indicates that changes in the shape of the functional space, and additions or losses of species, are almost always accompanied by a functional shift.
+However, shifts in the functional space were not restricted to change studies, as we also found significant *CWM* trends in 7 of the 32 "no change" studies.
+Functional turnover was also qualitatively divorced from species turnover, with just as many studies exhibiting functional turnover (at least one significant *CWM* trend) with species turnover as not.
+These studies give evidence that even minimal changes in species composition undetected by trends in species-based metrics can translate to meaningful changes to one or a few trait axes.
+
 ## Conservation Implications
@@
 In fact, study-level trends indicate quite the opposite, that functional changes with negative or yet unknown implications for ecosystem processes may be going undetected by common species-based approaches.
-For example, loss of evenness in communities with constant species richness may be an first sign of a community being impacted by environmental change, with negative implications for stability and function.
+For example, loss of evenness in communities with constant species richness may be a first sign of a community being impacted by environmental change, with negative implications for stability and function.
 

### 2024-10-10 paper/GEB Submission/manuscript_GEB.Rmd
Additional turnover paragraph, change to abundance description.

@@
 We obtained mammal and bird time series from the BioTIME database, a global repository of high-quality assemblage time series collected from the literature and ongoing monitoring efforts.
-Data are structured such that a study comprises all data collected following consistent sampling protocols and may contain multiple time series from separate locations.
-Time series represent full assemblages rather than populations of single species [@dornelas2018], and all time series include abundance of observed species.
+Data are structured such that a study comprises all data collected following consistent sampling protocols and may contain multiple site locations.
+Samples at each site represent full assemblages rather than populations of single species [@dornelas2018], `r colorize("and include both abundance and presence/absence data", "blue")`.
 Following best practices for the database [@blowes2019], studies with multiple sample locations were split into individual time series following a standardized spatial scale.
@@
 The resulting cell size for our data was approximately 95 km^2^.
-All samples from a study within a single cell were considered to be a single time series location, and species abundances were combined for all samples.
-`r colorize("This approach therefore sets an upper bound on the definition of "local" for studies with multiple locations.
+All samples from a study within a single cell were considered to be a single time series location and species abundances, if available, were combined for all samples.
+`r colorize("This approach therefore sets an upper bound on the definition of local for studies with multiple locations.
 Our dataset included 29 single location studies (average spatial extent of ~4 km2), and 24 multi location studies aggregated by cell.", "blue")`
@@
 This restriction allows for enough axes to capture the trait space while maintaining computational feasibility [@blonder2018].
-Metrics incorporated weighting based on species abundance.
+Metrics incorporated weighting based on species abundance when available.
 
@@
 Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
-Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species.
+Null estimates were calculated for each rarefied sample by randomly sampling species from the species pool for each year and randomly assigning observed abundances to species for abundance-based studies.
 Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability.
@@
 
-Study-level trends in *CWM*'s offer insight into how the functional space might be shifting through time, and how that functional turnover may or may not relate to species turnover.
-Every study that experienced a trend in a species or functional diversity metric (Table \ref{tab:studyChange}; functional change, loss of redundancy, and increase in species richness study types), with the exception of 3 "loss of redundancy" studies, also exhibited a trend in at least one *CWM*.
-This linking between *CWM* and other metrics indicates that changes in the shape of the functional space, and additions or losses of species, are almost always accompanied by a functional shift.
-However, shifts in the functional space were not restricted to change studies, as we also found significant *CWM* trends in 7 of the 32 "no change" studies.
-Functional turnover was also qualitatively divorced from species turnover, with just as many studies exhibiting functional turnover (at least one significant *CWM* trend) with species turnover as not.
-These studies give evidence that even minimal changes in species composition undetected by trends in species-based metrics can translate to meaningful changes to one or a few trait axes.
+`r colorize("Study-level trends in CWM's offer insight into how the functional space might be shifting through time, and how that functional turnover may or may not relate to species turnover.
+Every study that experienced a trend in a species or functional diversity metric (Table", "blue")` \ref{tab:studyChange}`r colorize("; functional change, loss of redundancy, and increase in species richness study types), with the exception of 3 \"loss of redundancy\" studies, also exhibited a trend in at least one CWM.This linking between CWM and other metrics indicates that changes in the shape of the functional space, and additions or losses of species, are almost always accompanied by a functional shift.
+However, shifts in the functional space were not restricted to change studies, as we also found significant CWM trends in 7 of the 32 \"no change\" studies.
+Functional turnover was also qualitatively divorced from species turnover, with just as many studies exhibiting functional turnover (at least one significant CWM trend) with species turnover as not.
+These post hoc observations give evidence that even minimal changes in species composition undetected by trends in species-based metrics can translate to significant changes to one or a few trait axes.", "blue")`
 

### 2024-10-14 analysis/06_study_change.Rmd
Look at CWM changes for each study

@@
 
+Look at CWM's
+```{r}
+cwm_counts <- model_output %>% 
+  filter(!is.na(study_id), model_id == "base", p.adjust < 0.05, 
+         !metric %in% c("S", "FRic", "FEve", "FDiv", "SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base")) %>% 
+  select(metric, study_id) %>% 
+  count(metric)
+
+# get study-level significant CWM counts with their other change characteristics
+cwm_counts %>% 
+  full_join(change_groups) %>% 
+  select(-c("FRic", "FEve", "FDiv")) %>% 
+  View()
+```
+
 ```{r}

### 2024-10-14 paper/GEB Submission/manuscript_GEB.Rmd
Change hypotheses'wording

@@
 
-Functional diversity offers a powerful addition to species-based approaches for detecting and describing community change by capturing changes in the traits that underpin fundamental ecological processes [@lavorel2002; @mcgill2006; @suding2008].
+Functional diversity offers a powerful addition to species-based approaches for detecting and describing community change by capturing changes in the traits that underpin fundamental ecological processes [@lavorel2002; @mcgill2006; @streit2023; @suding2008].
 Functional diversity metrics capture complimentary aspects of the trait space including the volume of the multidimensional space (functional diversity), how species and abundance are distributed across the space (functional evenness and divergence), and `r colorize("how the space maps to individual trait distributions", "blue")` (community weighted trait means).
@@
 We synthesize thousands of mammal and bird functional diversity time series to assess mean trends across communities.
-We evaluate 1) whether or not there is a prevailing type of functional change across communities and if community characteristics explain variation in those trends, 2) hypotheses of change for individual trait axes, and 3) the implications of study-level trends for the ecology of those communities.
+We evaluate 1) whether or not there is consistent change `r colorize("in the functional trait space shape (functional diversity) or location (community weighted trait means)", "blue")` across communities and if community characteristics explain variation in that change, 2) hypotheses of change for individual trait axes `r colorize("including: decreasing body size, decreasing insectivory, increasing invertivory, decreasing arboreal and understory foraging,", "blue")` and 3) the implications of study-level trends for the ecology of those communities.
 To address these goals, we present results for general trends that aggregate across all communities, category-specific trends that aggregate across communities with similar characteristics, and study-level trends.
@@
 28 of the rarefied studies required under-sampling by only one sample, and only three time series required under-sampling by more than ten samples.
+`r colorize("For time series with and without rarefaction, samples were pooled within each year and abundances were summed (where available).", "blue")`
 
@@
 `r colorize("Study-level trends in CWM's offer insight into how the functional space might be shifting through time, and how that functional turnover may or may not relate to species turnover.
-Every study that experienced a trend in a species or functional diversity metric (Table", "blue")` \ref{tab:studyChange}`r colorize("; functional change, loss of redundancy, and increase in species richness study types), with the exception of 3 \"loss of redundancy\" studies, also exhibited a trend in at least one CWM.This linking between CWM and other metrics indicates that changes in the shape of the functional space, and additions or losses of species, are almost always accompanied by a functional shift.
+Every study that experienced a trend in a species or functional diversity metric (Table", "blue")` \ref{tab:studyChange}`r colorize("; functional change, loss of redundancy, and increase in species richness study types), with the exception of 3 \"loss of redundancy\" studies, also exhibited a trend in at least one CWM.
+This link between CWM's and other metrics indicates that changes in the shape of the functional space, and additions or losses of species, are almost always accompanied by a functional shift.
 However, shifts in the functional space were not restricted to change studies, as we also found significant CWM trends in 7 of the 32 \"no change\" studies.

### 2024-10-28 paper/GEB Submission/manuscript_GEB.Rmd
Minor abastract change

@@
 Study characteristics such as taxa, realm, biome, or protection status did not distinguish between types of change exhibited by communities.
-We found evidence of a temporal increase in fish consumption across all communities but no evidence to support multiple predictions for specific traits, including decreasing body size, dietary shifts, or changes in foraging strata.
+We found evidence of a temporal increase in fish consumption across all communities but no evidence to support multiple predictions for specific traits, including decreasing body size, dietary shifts, or changes in bird foraging strata.
 
@@
 output: 
-  rticles::arxiv_article:
-    keep_tex: true
-#  word_document: default
+  # rticles::arxiv_article:
+  #   keep_tex: true
+  word_document: default
 editor_options: 
@@
 Species pools were unique for each time series and included all species observed over the course of sampling, therefore accounting for geographic restrictions in species availability.
-`r colorize("This approach assumes that the likelihood of a species being present is equally likely through time and that the species abundance distribution is conserved.", "blue")` This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
+`r colorize("This approach assumes the species pool is constant through time and that the species abundance distribution is conserved.", "blue")` This process was repeated 500 times to get an estimate and standard deviation of the null expectation for the metric for each rarefaction sample for that time series.
 We used these values to calculate SES using the following formula: $SES = [F_{obs} - mean_{(F_{null})}]/SD_{(F_{null})}$.
@@
 
-'r colorize("We tested for category-specific trends in all diversity metrics within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")\` We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
+'r colorize("We tested for category-specific trends in all diversity metrics within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")` We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
 For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.
@@
 
-# Acknowledgments
-
-We thank the scientists who contributed to and maintain the biodiversity databases included in this study, including BioTIME, Amphibio, and Elton Traits.
-
-We thank Dr Shane Blowes and Dr Sarah Supp, whose code we adapted for initial data processing of the BioTIME database.
-
 # References

### 2024-10-28 paper/GEB Submission/manuscript_GEB.Rmd
tiny typos, change citation style

@@
 bibliography: refs.bib
-csl: pnas.csl
+csl: apa-6th-edition.csl
 layout: 3p
@@
 
-Ecological communities are experiencing unprecedented changes as a result of anthropogenic pressures such as climate change, land use change, and invasive species.
+Ecological communities are experiencing unprecedented changes due to anthropogenic pressures such as climate change, land use change, and invasive species.
 Impacts of these pressures are well documented at a global scale by an accelerating global extinction rate [@barnosky2011] and fundamental changes in some of the most well-studied systems [e.g.
@@
 Global analyses have been heavily criticized for geographic biases, lack of data in the most heavily impacted areas, `r colorize("short time windows biased toward the present day", "blue")` and exclusion of individual studies' ecological context [@cardinale2014; @cardinale2018; @gonzalez2016].
-Many of these criticisms reflect limitations of ecological data on the whole [@chapman2024], leading to a call for additional data not only to fill geographic and temporal gaps, but to fill in key characteristics of communities [@dornelas2023; @primack2018].
+Many of these criticisms reflect limitations of ecological data overall [@chapman2024], leading to a call for additional data not only to fill geographic and temporal gaps, but to fill in key characteristics of communities [@dornelas2023; @primack2018].
 
@@
 `r colorize("Trends in functional diversity across broad geographic extents", "blue")` are the aggregation of local-scale responses to acute pulse disturbance, persistent press disturbances, and periods of relative stasis [@inamine2022].
-Pulse disturbances have a sudden one time impact on the community, changing abundance distributions (functional evenness or divergence) or reducing the volume of the functional space [@boersma2016; @hillebrand2020; @larson2021].
+Pulse disturbances have a sudden one-time impact on the community, changing abundance distributions (functional evenness or divergence) or reducing the volume of the functional space [@boersma2016; @hillebrand2020; @larson2021].
 While press disturbances can also lead to loss of functional volume as species fail to adapt to new conditions, they can also lead to wholesale shifts in the location of the functional space as species fill in new environmental space.
-Rather than identifying outcomes of one of these specific processes, observational time series of functional diversity therefore capture the cumulative affects of multiple potentially competing processes.
+Rather than identifying outcomes of one of these specific processes, observational time series of functional diversity therefore capture the cumulative effects of multiple potentially competing processes.
 
@@
 Redundancy is critical for maintaining ecological processes in the face of disturbance as it ensures functional roles are maintained [@biggs2020], and a loss of functional redundancy likely reduces a community's capacity to respond to future stressors [@reich2012].
-These communities are actually fairing better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future [@pimiento2020a].
+These communities are actually faring better than expected looking at species-based metrics alone, but are also in a precarious position for maintaining ecological function into the future [@pimiento2020a].
 

### 2024-11-12 analysis/figures/visualization.Rmd
Make effect size plot

@@
 ```
+
+Effect size plots
+
+```{r}
+load(here::here("data/model_output.rda"))
+
+durations <- model_data %>%
+  select(study_id, duration) %>%
+  group_by(study_id) %>%
+  mutate(duration = max(duration)) %>% 
+  distinct()
+
+slopes <- model_output %>% 
+  filter(term == "year_scaled", model_id == "base",
+         !metric %in% c("FDiv", "FRic", "FEve", "Jaccard_next")) %>% 
+  select(metric, estimate, p.adjust, study_id) %>%
+  group_by(metric) %>%
+  #rowwise() %>%
+  mutate(across(estimate, ~scale(.)[,1])) %>%
+  ungroup() %>%
+    mutate(ind = case_when(
+    estimate > 0 & p.adjust < 0.05 ~ "blue", 
+    estimate < 0 & p.adjust < 0.05 ~ "red", 
+    .default = "grey"
+  )) %>%
+  left_join(durations) %>%
+  mutate(duration = as.factor(duration))
+
+colgraid <- colorRampPalette(c("white", "#77976e"))
+graid_pal <- colgraid(26)
+
+plot_slopes <- function(metric_name, plot_label, keep_x_axis, color){
+
+  if (metric_name == "empty") return(NULL)
+
+  colgraid <- colorRampPalette(c("white", color))
+  graid_pal <- colgraid(26)
+
+  plot_data <- slopes %>%
+    filter(metric == metric_name)
+
+  gen_est <- plot_data %>% filter(is.na(study_id))
+
+  metric_plot <- plot_data %>% 
+    filter(!is.na(study_id)) %>% 
+    ggplot() +
+    geom_dotplot(aes(x = estimate, fill = duration), 
+                 #stackdir = "center",
+                 stackgroups = TRUE,
+                 method = "histodot", 
+                 binwidth = 0.15, dotsize = 1.15
+    ) +
+    scale_fill_manual(values = graid_pal, guide = "none", drop = FALSE) +
+    geom_vline(aes(xintercept = gen_est$estimate), linetype = "longdash") +
+    theme_minimal() +
+    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
+          axis.title.y = element_blank(), 
+          axis.text.y=element_blank(), 
+          axis.title=element_text(size=14,face="bold")) +
+    #labs(title = plot_label) +
+    xlim(-3.15, 3.15) +
+    #ylim(0, 2.5) +
+    annotate('text', x = -2.5, y = 0.8, label = plot_label, size = unit(7, "pt"))
+
+  if(isFALSE(keep_x_axis)) {
+    metric_plot <- metric_plot + theme(axis.text.x = element_blank(), axis.title.x = element_blank())
+  } else{
+    metric_plot <- metric_plot + xlim(-3.15, 3.15) +
+      geom_hline(aes(yintercept = 0))
+  }
+
+  if(gen_est$p.adjust < 0.05){
+    metric_plot <- metric_plot + annotate('text', x = (gen_est$estimate+.2), y = 0.8, label = "*", size = unit(16, "pt"))
+  }
+
+  return(metric_plot)
+}
+
+muted <- khroma::color("muted")(5)
+
+slope_plot_df <- data.frame(metric_name = c("S", "Jaccard_base", 
+                     "SES_FRic", "SES_FEve", "SES_FDiv", 
+                     "CWM_diet_inv", "CWM_diet_scav", "CWM_diet_vfish", "CWM_diet_vunk", "CWM_diet_planto",
+                     "CWM_diet_seed", "CWM_diet_vend", "CWM_diet_nect", "CWM_diet_fruit", "CWM_diet_vect", 
+                     "CWM_forstrat_ground", "CWM_forstrat_watbelowsurf", "CWM_forstrat_wataroundsurf", 
+                     "CWM_forstrat_understory", "CWM_forstrat_aerial", 
+                     "CWM_forstrat_canopy", "CWM_forstrat_midhigh", "CWM_bodymass_value", rep("empty", 2)),
+                     plot_label = c("Species\nRichness", "Jaccard\nSimilarity", 
+                     "FRic SES", "FEve SES", "FDiv SES", 
+                     "Invertebrates", "Scavenge", "Fish", "Other\nVertebrates", "Other Plants",
+                     "Seeds", "Mammals\n& Birds", "Nectar", "Fruit", "Reptiles &\nAmphibians", 
+                     "Ground", "Below Water\nSurface", "Around Water\nSurface", 
+                     "Understory", "Aerial", "Canopy", "Midhigh", 
+                     "Body Mass", rep("empty", 2)), 
+                     keep_x_axis = c(FALSE, FALSE, 
+                     FALSE, FALSE, TRUE, 
+                     FALSE, FALSE, FALSE, FALSE, TRUE,
+                     FALSE, FALSE, FALSE, FALSE, TRUE, 
+                     FALSE, FALSE, FALSE, 
+                     FALSE, TRUE, 
+                     FALSE, FALSE, TRUE, rep(FALSE, 2)),
+                     color = c(rep(muted[1], 2), rep(muted[2], 3), rep(muted[3], 10), rep(muted[4], 7), rep(muted[5], 3))) %>%
+  mutate(across(keep_x_axis, as.logical))
+
+slope_plots <- pmap(slope_plot_df, plot_slopes)
+
+slope_plot_grid <- plot_grid(plotlist=slope_plots, nrow = 5, align = 'vh',
+                       hjust = -1, axis = "l", byrow = FALSE)
+
+duration_legend_plot <- slopes %>%
+  ggplot(aes(y = metric, x = estimate, color = duration)) +
+  geom_point() +
+  scale_color_gradient(low = "white", high = "#77976e")
+
+duration_legend <- get_legend(duration_legend_plot)
+
+test_plot <- slope_plot_grid + draw_grob(duration_legend,  .8, -.2, .3/3.3, 1, scale = 5)
+
+cowplot::save_plot(here::here("figures/slope_plot_grid.png"), slope_plot_grid, ncol = 7, nrow = 5, bg = "white", base_asp = 1.18)
+# plot_grid(plotlist=slope_plots, labels = "AUTO", nrow = 3, align = 'vh',
+#                        hjust = -1, axis = "l", scale = 0.9)
+```

### 2024-11-12 analysis/figures/visualization.Rmd
Export legends for each of the colors, save out with better aspect ratio

@@
   left_join(durations) %>%
-  mutate(duration = as.factor(duration))
+  mutate(duration_fac = as.factor(duration))
 
@@
     ggplot() +
-    geom_dotplot(aes(x = estimate, fill = duration), 
-                 #stackdir = "center",
+    geom_dotplot(aes(x = estimate, fill = duration_fac), 
                  stackgroups = TRUE,
@@
           axis.text.y=element_blank(), 
-          axis.title=element_text(size=14,face="bold")) +
-    #labs(title = plot_label) +
+          axis.title=element_text(size=14)) ++
     xlim(-3.15, 3.15) +
-    #ylim(0, 2.5) +
+    xlab("temporal trend") +
     annotate('text', x = -2.5, y = 0.8, label = plot_label, size = unit(7, "pt"))
@@
 
+cowplot::save_plot(here::here("figures/slope_plot_grid.png"), slope_plot_grid, ncol = 7, nrow = 5, bg = "white", base_asp = 1.18)
+
+get_legends <- function(color){
   duration_legend_plot <- slopes %>%
@@
     geom_point() +
-  scale_color_gradient(low = "white", high = "#77976e")
+    scale_color_gradient(low = "white", high = color) +
+    theme(legend.title = element_blank(), legend.text = element_blank(), 
+          legend.background = element_rect(fill='transparent'))
 
   duration_legend <- get_legend(duration_legend_plot)
+  leg <- grid.arrange(duration_legend)
+  ggsave(here::here(paste0("figures/slope_legends/", color, "_legend.png")), leg, width = .6, height = 1.5,
+         bg = "transparent")
+}
 
-test_plot <- slope_plot_grid + draw_grob(duration_legend,  .8, -.2, .3/3.3, 1, scale = 5)
+map(muted, get_legends)
 
-cowplot::save_plot(here::here("figures/slope_plot_grid.png"), slope_plot_grid, ncol = 7, nrow = 5, bg = "white", base_asp = 1.18)
-# plot_grid(plotlist=slope_plots, labels = "AUTO", nrow = 3, align = 'vh',
-#                        hjust = -1, axis = "l", scale = 0.9)
 ```

### 2024-11-12 paper/GEB Submission/manuscript_GEB.Rmd
Language edits from editor

@@
 Functional diversity provides a potentially powerful alternative approach for understanding community composition by linking shifts in species identity to the characteristics that underpin ecosystem processes.
-Here we present the first multi-taxa, multi-system analysis of functional change through time.
+Here we present the first multi-taxa, multi-system analysis of functional diversity and composition through time.
 
@@
 
-  Main Conclusions: General temporal trends indicate that on the aggregate one type of functional change is not more prevalant than the other across many taxa, biomes, and realms.
-At the study-level, the majority of studies showed no species or functional trends, with the remainder of the studies falling into broad categories of functional changes independent of species richness, functional redundancy loss, and increasing species richness.
+  Main Conclusions: General temporal trends indicate that on the aggregate studies do not exhibit consistent changes in any aspect of functional diversity across many taxa, biomes, and realms.
+At the study-level, the majority of studies showed no temporal trends in species or functional diversity, with the remainder of the studies falling into broad categories of functional diversity change independent of species richness, loss of functional redundancy (species richness declines while functional richness is maintained), and increasing species richness sometimes accompanied by increasing functional richness.
 
@@
 
-To assess functional change independent of species richness we calculated the standardized effect size (SES) for each of the three functional diversity metrics (*FRic*, *FEve*, *FDiv*) from null estimates [@swenson2012].
+To assess changes in functional diversity independent of species richness we calculated the standardized effect size (SES) for each of the three functional diversity metrics (*FRic*, *FEve*, *FDiv*) from null estimates [@swenson2012].
 Null model corrections allow us to assess the degree to which the observed functional diversity metric deviates from the value expected by chance in a randomly assembled community.
@@
 
-'r colorize("We tested for category-specific trends in all diversity metrics within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")` We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
+'r colorize("We tested for category-specific trends in all diversity metrics within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")\` We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
 For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.
@@
 
-While trends aggregating across studies did not indicate one prevailing kind of functional change, post hoc examination of trends for individual studies revealed four distinct change groups.
-These groups are categorized based on the concurrent trends in species and functional metrics and can be broadly described as: no change in species or functional diversity metrics, functional change independent of species richness, loss of redundancy, and increase in species richness (Table \ref{tab:studyChange}).
+While trends aggregating across studies did not indicate one prevailing kind of change in functional diversity or functional composition (with the exception of changes in fish consumption), post hoc examination of trends for individual studies revealed four distinct change groups.
+These groups are categorized based on the concurrent trends in species and functional metrics and can be broadly described as: no change in species or functional diversity metrics, change in functional diversity independent of species richness, loss of redundancy (species richness declines while functional richness is maintained), and increase in species richness sometimes accompanied by increasing functional richness (Table \ref{tab:studyChange}).
 The majority of studies (32 studies) exhibited no trend in any species or functional diversity metric.
@@
 Based on a species approach alone, these communities appear to either be experiencing no change or some change in species identity, with unknown implications for conservation concern or ecological process.
-However, with the addition of a functional perspective we found multiple kinds of underlying functional change including increases and decreases in corrected functional richness and divergence and loss of corrected functional evenness.
+However, with the addition of a functional perspective we found multiple kinds of underlying change in functional diversity including increases and decreases in corrected functional richness and divergence and loss of corrected functional evenness.
 We therefore found evidence of both increases and decreases in the functional capacity of communities independent of species-based metrics, reinforcing the importance of moving beyond solely species-based approaches for biodiversity assessment [@hillebrand2017; @santini2017].
@@
 `r colorize("Study-level trends in CWM's offer insight into how the functional space might be shifting through time, and how that functional turnover may or may not relate to species turnover.
-Every study that experienced a trend in a species or functional diversity metric (Table", "blue")` \ref{tab:studyChange}`r colorize("; functional change, loss of redundancy, and increase in species richness study types), with the exception of 3 \"loss of redundancy\" studies, also exhibited a trend in at least one CWM.
+Every study that experienced a trend in a species or functional diversity metric (Table", "blue")` \ref{tab:studyChange}`r colorize("; functional diversity change, loss of redundancy, and increase in species richness study types), with the exception of 3 \"loss of redundancy\" studies, also exhibited a trend in at least one CWM.
 This link between CWM's and other metrics indicates that changes in the shape of the functional space, and additions or losses of species, are almost always accompanied by a functional shift.
@@
 
-While we found no over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for functional change as a result of anthropogenic impact.
-In fact, study-level trends indicate quite the opposite, that functional changes with negative or yet unknown implications for ecosystem processes may be going undetected by common species-based approaches.
+While we found no over all trends in functional metrics, our results should not be interpreted as an indication that the ongoing biodiversity crisis is less severe than previously described, or that there is no concern for changes in functional diversity as a result of anthropogenic impact.
+In fact, study-level trends indicate quite the opposite, that changes in functional diversity and composition with negative or yet unknown implications for ecosystem processes may be going undetected by common species-based approaches.
 For example, loss of evenness in communities with constant species richness may be a first sign of a community being impacted by environmental change, with negative implications for stability and function.

### 2024-11-13 analysis/05_model_metrics.Rmd
Attempts and mantel test

@@
 ```
+Check if Jaccard distance is correlated with time with Mantel test
+```{r}
+library(vegan)
+
+jaccard_dist <- model_data %>% 
+    select(rarefyID, value, year, metric) %>% 
+    pivot_wider(names_from = metric, values_from = value) %>% select(rarefyID, year, Jaccard_base) %>%
+  unite(obsid, rarefyID, year, remove = FALSE) %>%
+  rename(baselineID = rarefyID) %>%
+  mutate(#baseline = is.na(Jaccard_base), 
+         jacc_dist = 1-Jaccard_base) %>%
+  group_by(baselineID) %>%
+  mutate(time_dist = year - min(year)) %>%
+  ungroup() %>%
+  filter(!is.na(jacc_dist)) %>%
+  select(-Jaccard_base, -year)
+
+get_dist_mat <- function(df, dist_name){
+  dist_row <- df %>% 
+  select(obsid, baselineID, dist_name) %>% 
+  pivot_wider(names_from = obsid, values_from = dist_name) %>% 
+  mutate(baselineID = NA)
+
+rownames(dist_row) <- "baselineID"
+
+dist_mat <- dist_row %>%
+  t() %>% as.data.frame()
+
+dist_mat[,colnames(dist_row)[-1]] <- NA
+
+return(as.dist(dist_mat))
+}
+
+# do it one by one for each rarefyID
+get_mantel_stat <- function(rarefyID){
+  dist_data <- jaccard_dist %>%
+  filter(baselineID == rarefyID)
+
+  jacc_dist_mat <- get_dist_mat(dist_data, "jacc_dist")
+time_dist_mat <- get_dist_mat(dist_data, "time_dist")
+
+mantel_test <- mantel(jacc_dist_mat, time_dist_mat, method = "spearman", permutations = 9999, na.rm = TRUE)
+
+data.frame(rarefyID = rarefyID, p.value = mantel_test$signif)
+}
+
+mantel_stats <- purrr::map_dfr(unique(jaccard_dist$baselineID), get_mantel_stat)
+```
+
+```{r}
+# create master matrices with all timeseries
+master_jacc_dist <- purrr::map_dfr(unique(jaccard_dist$baselineID), function(x){
+  #browser()
+  dist_row <- jaccard_dist %>% 
+    filter(baselineID == x) %>%
+  select(obsid, baselineID, "jacc_dist") %>% 
+  pivot_wider(names_from = obsid, values_from = "jacc_dist")
+
+baseline_name <- unique(dist_row$baselineID)
+
+dist_mat <- dist_row %>%
+  mutate(baselineID = NA) %>%
+  rename(!!baseline_name := baselineID) %>%
+  t() %>% as.data.frame() %>%
+  rename(!!baseline_name := V1)
+
+dist_mat[,colnames(dist_row)[-1]] <- NA
+
+return(dist_mat)
+}) %>%
+  as.dist()
+
+master_time_dist <- purrr::map_dfr(unique(jaccard_dist$baselineID), function(x){
+  #browser()
+  dist_row <- jaccard_dist %>% 
+    filter(baselineID == x) %>%
+  select(obsid, baselineID, "time_dist") %>% 
+  pivot_wider(names_from = obsid, values_from = "time_dist")
+
+baseline_name <- unique(dist_row$baselineID)
+
+dist_mat <- dist_row %>%
+  mutate(baselineID = NA) %>%
+  rename(!!baseline_name := baselineID) %>%
+  t() %>% as.data.frame() %>%
+  rename(!!baseline_name := V1)
+
+dist_mat[,colnames(dist_row)[-1]] <- NA
+
+return(dist_mat)
+}) %>%
+  as.dist()
+
+mantel_test <- mantel(master_jacc_dist, master_time_dist, method = "spearman", permutations = 9999, na.rm = TRUE)
+```

### 2024-11-14 paper/GEB Submission/manuscript_GEB.Rmd
Small changes, fix table label

@@
 output: 
-  # rticles::arxiv_article:
-  #   keep_tex: true
-  word_document: default
+   rticles::arxiv_article:
+     keep_tex: true
+  #word_document: default
 editor_options: 
@@
 
-We also calculated the community-weighted mean (*CWM*) of all continuous traits `r colorize("(diet, foraging strata, body mass)", "blue")` to examine changes in the distribution of each trait.
+To assess compositional changes, we also calculated the community-weighted mean (*CWM*) of all continuous traits `r colorize("(diet, foraging strata, body mass)", "blue")` to examine changes in the distribution of each trait.
 Wholesale shifts in the `r colorize("observed", "blue")` trait space due to changes of trait means could occur even while the shape `r colorize("and dispersion", "blue")` of the multidimensional trait space, as defined by functional richness, evenness, and divergence, is maintained `r colorize("through time", "blue")`.
@@
 
-'r colorize("We tested for category-specific trends in all diversity metrics within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")\` We estimated within-level slopes and calculated between-level contrasts using the *emmeans* package [v1.10.1, @lenth2022].
+'r colorize("We tested for category-specific trends in all diversity metrics within and between different levels of taxa, biome, realm, and protection status by fitting separate models with each of those factors added as a predictor interacting with time to the original model structure.", "blue")\` Fore each linear mixed effects model we estimated within-level slopes and calculated between-level contrasts for each categorical variable using the *emmeans* package [v1.10.1, @lenth2022].
 For some levels of the categorical variables we did not have a sufficient number of studies to estimate a general trend, we therefore only interpret results for levels with more than three studies.
@@
            n = ifelse(n == "7", "32", n)) %>%
-  mutate(type = ifelse(n == "32", "no change", type)) %>%
+  mutate(type = ifelse(n == "32", "no change", type),
+         type = ifelse(type == "functional change", "functional diversity change", type)) %>%
   rename(`Study Count` = `n`) %>%

### 2024-11-15 analysis/figures/visualization.Rmd
Plot of group slopes

@@
              aes(rarefyID_x, rarefyID_y, color = taxa), alpha = 0.4, inherit.aes = FALSE) +
-  #plot the US amphibian point on top so we can see it
+  #plot the US amphibian point on bottom so we can see it
   # geom_point(data = study_data %>% 
@@
 
+metric_names <- unique(model_output$metric)
+raw_metrics <- c("SES_FRic", "SES_FDiv", "SES_FEve", 
+                 "CWM_diet_inv", "CWM_forstrat_ground", "CWM_forstrat_watbelowsurf", "FEve")
+log_metrics <- c("FRic", "S", "FDiv")
+logone_metrics <- metric_names[!metric_names %in% c(raw_metrics, log_metrics)]
+
 durations <- model_data %>%
@@
   select(metric, estimate, p.adjust, study_id) %>%
+  mutate(estimate_backtrans = case_when(
+           metric %in% log_metrics ~ exp(estimate),
+           metric %in% logone_metrics ~ 1 + exp(estimate),
+           .default = estimate
+         )) %>%
   group_by(metric) %>%
-  #rowwise() %>%
-  mutate(across(estimate, ~scale(.)[,1])) %>%
+  mutate(estimate_scale = scale(estimate_backtrans)[,1]) %>%
   ungroup() %>%
@@
     ggplot() +
-    geom_dotplot(aes(x = estimate, fill = duration_fac), 
+    geom_dotplot(aes(x = estimate_scale, fill = duration_fac), 
                  stackgroups = TRUE,
@@
     scale_fill_manual(values = graid_pal, guide = "none", drop = FALSE) +
-    geom_vline(aes(xintercept = gen_est$estimate), linetype = "longdash") +
+    geom_vline(aes(xintercept = gen_est$estimate_scale), linetype = "longdash") +
     theme_minimal() +
@@
           axis.text.y=element_blank(), 
-          axis.title=element_text(size=14)) ++
+          axis.title=element_text(size=14)) +
     xlim(-3.15, 3.15) +
-    xlab("temporal trend") +
+    xlab("temporal slope estimate") +
     annotate('text', x = -2.5, y = 0.8, label = plot_label, size = unit(7, "pt"))
@@
   if(gen_est$p.adjust < 0.05){
-    metric_plot <- metric_plot + annotate('text', x = (gen_est$estimate+.2), y = 0.8, label = "*", size = unit(16, "pt"))
+    metric_plot <- metric_plot + annotate('text', x = (gen_est$estimate_scale + .2), y = 0.8, label = "*", size = unit(16, "pt"))
   }
@@
 
-slope_plot_grid <- plot_grid(plotlist=slope_plots, nrow = 5, align = 'vh',
+slope_plot_grid <- cowplot::plot_grid(plotlist=slope_plots, nrow = 5, align = 'vh',
                        hjust = -1, axis = "l", byrow = FALSE)
@@
 ```
+Plots of significant contrasts
+```{r}
+contrast_data <- model_output %>%
+  filter(metric %in% c("Jaccard_base", "CWM_diet_vfish", "CWM_diet_vunk"), 
+         model_id != "base", 
+         effect %in% c("fixed", "contrast", "group_slope")) %>%
+  select(effect, term, estimate, p.adjust, metric, model_id)
+
+# test to plot taxa models
+test_plot_data <- contrast_data %>%
+  filter(model_id == "taxa", metric == "Jaccard_base")
+
+bird_params <- list(slope = test_plot_data %>% filter(term == "Birds") %>% pull(estimate),
+                    intercept = test_plot_data %>% filter(term == "(Intercept)") %>% pull(estimate),
+                    sig = test_plot_data %>% filter(term == "Birds") %>% pull(p.adjust) < 0.05)
+mammal_params <- list(slope = test_plot_data %>% filter(term == "Mammals") %>% pull(estimate),
+                      intercept = (test_plot_data %>% filter(term == "taxaMammals") %>% pull(estimate) + 
+                                     bird_params$intercept))
+ggplot() + theme_minimal() +
+  xlim(0, 1) +
+  ylim(0, 3) +
+  geom_abline(slope = bird_params$slope, intercept = bird_params$intercept, color = "red") +
+  geom_abline(slope = mammal_params$slope, intercept = mammal_params$intercept,)
+
+model_data %>%
+  filter(taxa != "Amphibians", metric == "Jaccard_base") %>%
+  ggplot(aes(x = year_scaled, y = value, color = taxa)) +
+  geom_point() +
+  geom_abline(slope = -0.033233485, intercept = 0.42832, color = "red") +
+  geom_abline(slope = -0.026364796, intercept = 0.42832 + 0.095965269)
+```
+Let's try dot plots of each categorical slope
+```{r}
+library(patchwork)
+
+group_slopes <- model_output %>% 
+    filter(effect == "group_slope", !metric %in% c("Jaccard_next", "FRic", "FEve", "FDiv"), 
+           !term %in% c("Global", "Polar/Temperate")) %>% 
+  select(model_id, term, estimate, p.adjust, metric) %>%
+  mutate(sig = ifelse(p.adjust < 0.05, "yes", "no")) %>%
+  left_join(data.frame(metric = c("S", "Jaccard_base", 
+                     "SES_FRic", "SES_FEve", "SES_FDiv", 
+                     "CWM_diet_inv", "CWM_diet_scav", "CWM_diet_vfish", "CWM_diet_vunk", "CWM_diet_planto",
+                     "CWM_diet_seed", "CWM_diet_vend", "CWM_diet_nect", "CWM_diet_fruit", "CWM_diet_vect", 
+                     "CWM_forstrat_ground", "CWM_forstrat_watbelowsurf", "CWM_forstrat_wataroundsurf", 
+                     "CWM_forstrat_understory", "CWM_forstrat_aerial", 
+                     "CWM_forstrat_canopy", "CWM_forstrat_midhigh", "CWM_bodymass_value"),
+                     plot_label = c("Species Richness", "Jaccard Similarity", 
+                     "FRic SES", "FEve SES", "FDiv SES", 
+                     "CWM Invertebrate Diet", "CWM Scavenge Diet", "CWM Fish Diet", "CWM Other Vertebrates Diet", "CWM Other Plants Diet",
+                     "CWM Seeds Diet", "CWM Mammals & Birds Diet", "CWM Nectar Diet", "CWM Fruit Diet", "CWM Reptiles & Amphibians Diet", 
+                     "CWM Ground Foraging", "CWM Below Water Surface Foraging", "CWM Around Water Surface Foraging", 
+                     "CWM Understory Foraging", "CWM Aerial Foraging", "CWM Canopy Foraging", "CWM Midhigh Foraging", 
+                     "CWM Body Mass"))) %>%
+  mutate(estimate_backtrans = case_when(
+           metric %in% log_metrics ~ exp(estimate),
+           metric %in% logone_metrics ~ 1 + exp(estimate),
+           .default = estimate
+         )) %>%
+  group_by(metric) %>%
+  mutate(numeric_metric = as.factor(cur_group_id()),
+         estimate_scale = scale(estimate_backtrans)[,1]) %>%
+  ungroup()
+
+sig_contrasts <- model_output %>%
+  filter(effect == "contrast", p.adjust < 0.05)
+
+get_estimate <- function(metric_name, term_name){
+  model_output %>% 
+    filter(metric == metric_name, term == term_name) %>%
+    pull(estimate_scale)
+}
+
+climate_plot <- group_slopes %>%
+  filter(model_id == "climate") %>%
+  ggplot(aes(y = numeric_metric, x = estimate_scale, color = term)) +
+  geom_point(aes(shape = sig), size = 3) +
+  #ylim(-7,7) +
+  #scale_x_discrete(expand=c(0.001, 4)) +
+  theme_bw() +
+  theme(axis.title.y = element_blank(), legend.title = element_blank(),
+        legend.position = "bottom", 
+        panel.grid.major.x = element_blank(),
+        panel.grid.minor.x = element_blank(),
+        panel.border = element_blank(),
+        axis.line = element_line(colour = "black")
+        ) +
+  scale_shape_manual(values = c(1, 19), guide = "none") +
+  #ggpubr::geom_bracket(data = group_slopes %>% filter(model_id == "climate"), xmin = unlist(get_estimate("CWM_diet_vunk", "Marine")), xmax = unlist(get_estimate("CWM_diet_vunk", "Terrestrial")), y.position = 11, label = "") +
+  scale_y_discrete(labels = group_slopes %>% select(numeric_metric, plot_label) %>% distinct() %>% tibble::deframe(), drop = FALSE) +
+  scale_color_manual(values = group_colors) +
+  geom_vline(xintercept = 0, linetype = "dashed") +
+  guides(color=guide_legend(nrow=2, byrow=TRUE)) +
+  xlab("temporal slope estimate")
+
+climate_plot
+
+taxa_plot <- group_slopes %>%
+  filter(model_id == "taxa") %>%
+  ggplot(aes(y = numeric_metric, x = estimate_scale, color = term)) +
+  geom_point(aes(shape = sig), size = 3) +
+  #ylim(-7,7) +
+  #scale_x_discrete(expand=c(0.001, 4)) +
+  theme_bw() +
+  theme(axis.title.y = element_blank(), legend.title = element_blank(),
+        axis.line.y = element_blank(), axis.text.y = element_blank(), 
+        axis.ticks.y = element_blank(),
+        legend.position = "bottom",
+        panel.grid.major.x = element_blank(),
+        panel.grid.minor.x = element_blank(),
+        panel.border = element_blank(),
+        axis.line = element_line(colour = "black")
+        ) +
+  scale_shape_manual(values = c(1, 19), guide = "none") +
+  scale_color_manual(values = my_pal) +
+  geom_vline(xintercept = 0, linetype = "dashed")+
+  guides(color=guide_legend(nrow=2, ncol = 2, byrow=TRUE)) +
+  scale_y_discrete(labels = group_slopes %>% select(numeric_metric, plot_label) %>% distinct() %>% tibble::deframe(), drop = FALSE) +
+  xlab("temporal slope estimate")
+
+taxa_plot
+
+pa_plot <- group_slopes %>%
+  filter(model_id == "protected_area") %>%
+  mutate(term = ifelse(term == TRUE, "Protected", "Unprotected")) %>%
+  ggplot(aes(y = numeric_metric, x = estimate_scale, color = term)) +
+  geom_point(aes(shape = sig), size = 3) +
+  #ylim(-7,7) +
+  #scale_x_discrete(expand=c(0.001, 4)) +
+  theme_bw() +
+  theme(axis.title.y = element_blank(), legend.title = element_blank(),
+        legend.position = "bottom",
+        axis.line.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(),
+        panel.border = element_blank(),
+        panel.grid.major.x = element_blank(),
+        panel.grid.minor.x = element_blank(),
+        axis.line = element_line(colour = "black")
+        ) +
+  scale_shape_manual(values = c(1, 19), guide = "none")+
+  scale_color_manual(values = my_pal) +
+  geom_vline(xintercept = 0, linetype = "dashed") +
+  guides(color=guide_legend(nrow=2, ncol = 2, byrow=TRUE)) +
+  scale_y_discrete(labels = group_slopes %>% select(numeric_metric, plot_label) %>% distinct() %>% tibble::deframe(), drop = FALSE) +
+  xlab("temporal slope estimate")
+
+pa_plot
+
+realm_plot <- group_slopes %>%
+  filter(model_id == "realm") %>%
+  ggplot(aes(y = numeric_metric, x = estimate_scale, color = term)) +
+  geom_point(aes(shape = sig), size = 3) +
+  #ylim(-7,7) +
+  #scale_x_discrete(expand=c(0.001, 4)) +
+  theme_bw() +
+  theme(axis.title.y = element_blank(), legend.title = element_blank(),
+        axis.line.y = element_blank(), axis.text.y = element_blank(), 
+        axis.ticks.y = element_blank(),
+        legend.position = "bottom",
+        panel.grid.major.x = element_blank(),
+        panel.grid.minor.x = element_blank(),
+        panel.border = element_blank(),
+        axis.line = element_line(colour = "black")
+        ) +
+  scale_shape_manual(values = c(1, 19), guide = "none") +
+  scale_color_manual(values = my_pal) +
+  geom_vline(xintercept = 0, linetype = "dashed")+
+  guides(color=guide_legend(nrow=2, ncol = 2, byrow=TRUE)) +
+  scale_y_discrete(labels = group_slopes %>% select(numeric_metric, plot_label) %>% distinct() %>% tibble::deframe(), drop = FALSE) +
+  xlab("temporal slope estimate")
+
+realm_plot
+
+# contrasts_plot_grid <- cowplot::plot_grid(plotlist = list(climate_plot, taxa_plot, pa_plot, realm_plot), nrow = 1, rel_widths = c(1.7, 1, 1, 1), rel_heights = c(1, .6, .6, .6))
+# cowplot::save_plot(here::here("figures/contrasts_plot_grid.png"), contrasts_plot_grid, ncol = 4, nrow = 1, bg = "white", base_asp = 1, base_height = 5)
+
+contrasts_plot_grid <- climate_plot + taxa_plot + pa_plot + realm_plot + plot_layout(ncol = 4, axis_titles = "collect")
+ggsave(here::here("figures/contrasts_plot_grid.png"), contrasts_plot_grid, width = 15, heigh = 6)
+```

### 2024-11-15 analysis/05_model_metrics.Rmd
Fix protected area models

@@
                          climate = "log(value) ~ year_scaled * climate + (year_scaled|study_id/rarefyID)",
-                   protected_area = "value ~ year_scaled * protected_area + (year_scaled|study_id/rarefyID)") %>%
+                   protected_area = "log(value) ~ year_scaled * protected_area + (year_scaled|study_id/rarefyID)") %>%
   pivot_longer(cols = c(base, taxa, realm, climate, protected_area), names_to = "model_id", values_to = "formula")
@@
                          climate = "logvalue ~ year_scaled * climate + (year_scaled|study_id/rarefyID)",
-                      protected_area = "value ~ year_scaled * protected_area + (year_scaled|study_id/rarefyID)") %>%
+                      protected_area = "logvalue ~ year_scaled * protected_area + (year_scaled|study_id/rarefyID)") %>%
   pivot_longer(cols = c(base, taxa, realm, climate, protected_area), names_to = "model_id", values_to = "formula")

### 2024-11-18 analysis/figures/visualization.Rmd
Fiddling with dot plots

@@
                  stackgroups = TRUE,
-                 method = "histodot", 
-                 binwidth = 0.15, dotsize = 1.15
+                 method = "histodot"#, 
+                 #binwidth = 0.15, dotsize = 1.15
     ) +
@@
           axis.title=element_text(size=14)) +
-    xlim(-3.15, 3.15) +
+    xlim(-4.2, 7) +
     xlab("temporal slope estimate") +
@@
   } else{
-    metric_plot <- metric_plot + xlim(-3.15, 3.15) +
+    metric_plot <- metric_plot + xlim(-4.2, 7) +
       geom_hline(aes(yintercept = 0))

### 2024-11-18 analysis/figures/visualization.Rmd
Figured out the pairwise results annotation!

@@
 get_estimate <- function(metric_name, term_name){
-  model_output %>% 
+  group_slopes %>% 
     filter(metric == metric_name, term == term_name) %>%
@@
   geom_point(aes(shape = sig), size = 3) +
+  geom_segment(aes(x = (get_estimate("CWM_diet_vunk", "Tropical") + 0.1), y = 11, xend = (get_estimate("CWM_diet_vunk", "Temperate") - .095)), color = "black") +
   #ylim(-7,7) +
@@
   geom_point(aes(shape = sig), size = 3) +
+   geom_segment(aes(x = (get_estimate("CWM_diet_vunk", "Marine") + 0.075), y = 11, xend = (get_estimate("CWM_diet_vunk", "Terrestrial")) - 0.075), color = "black") +
   #ylim(-7,7) +

### 2024-11-18 analysis/figures/visualization.Rmd
Don't need this any more

@@
 ```
-Plots of significant contrasts
-```{r}
-contrast_data <- model_output %>%
-  filter(metric %in% c("Jaccard_base", "CWM_diet_vfish", "CWM_diet_vunk"), 
-         model_id != "base", 
-         effect %in% c("fixed", "contrast", "group_slope")) %>%
-  select(effect, term, estimate, p.adjust, metric, model_id)
 
-# test to plot taxa models
-test_plot_data <- contrast_data %>%
-  filter(model_id == "taxa", metric == "Jaccard_base")
-
-bird_params <- list(slope = test_plot_data %>% filter(term == "Birds") %>% pull(estimate),
-                    intercept = test_plot_data %>% filter(term == "(Intercept)") %>% pull(estimate),
-                    sig = test_plot_data %>% filter(term == "Birds") %>% pull(p.adjust) < 0.05)
-mammal_params <- list(slope = test_plot_data %>% filter(term == "Mammals") %>% pull(estimate),
-                      intercept = (test_plot_data %>% filter(term == "taxaMammals") %>% pull(estimate) + 
-                                     bird_params$intercept))
-ggplot() + theme_minimal() +
-  xlim(0, 1) +
-  ylim(0, 3) +
-  geom_abline(slope = bird_params$slope, intercept = bird_params$intercept, color = "red") +
-  geom_abline(slope = mammal_params$slope, intercept = mammal_params$intercept,)
-
-model_data %>%
-  filter(taxa != "Amphibians", metric == "Jaccard_base") %>%
-  ggplot(aes(x = year_scaled, y = value, color = taxa)) +
-  geom_point() +
-  geom_abline(slope = -0.033233485, intercept = 0.42832, color = "red") +
-  geom_abline(slope = -0.026364796, intercept = 0.42832 + 0.095965269)
-```
 Let's try dot plots of each categorical slope

### 2024-11-19 analysis/figures/visualization.Rmd
Add foraging bird models back in, fiddling with saving out

@@
   left_join(durations) %>%
-  mutate(duration_fac = as.factor(duration))
+  mutate(duration_fac = as.factor(duration)) %>%
+  filter(estimate_scale < 3.5, estimate_scale > -3.5)
 
@@
                  stackgroups = TRUE,
-                 method = "histodot"#, 
-                 #binwidth = 0.15, dotsize = 1.15
+                 method = "histodot",
+                 binwidth = 0.1, dotsize = 1.4
     ) +
@@
     geom_vline(aes(xintercept = gen_est$estimate_scale), linetype = "longdash") +
+    #coord_fixed(ratio=5)+#, ylim = c(0,5)) +
+    #scale_y_continuous(limits=c(0, 1), expand = c(0, 0), breaks = seq(0, 1,1/10), labels=seq(0,10)) +
     theme_minimal() +
@@
           axis.title=element_text(size=14)) +
-    xlim(-4.2, 7) +
+    xlim(-3.5, 3.5) +
     xlab("temporal slope estimate") +
-    annotate('text', x = -2.5, y = 0.8, label = plot_label, size = unit(7, "pt"))
+    annotate('text', x = -2.5, y = 0.8, label = plot_label, size = unit(6, "pt"))
 
@@
   } else{
-    metric_plot <- metric_plot + xlim(-4.2, 7) +
+    metric_plot <- metric_plot + xlim(-3.5, 3.5) +
       geom_hline(aes(yintercept = 0))
@@
   if(gen_est$p.adjust < 0.05){
-    metric_plot <- metric_plot + annotate('text', x = (gen_est$estimate_scale + .2), y = 0.8, label = "*", size = unit(16, "pt"))
+    metric_plot <- metric_plot + annotate('text', x = (gen_est$estimate_scale + .4), y = 0.8, label = "*", size = unit(16, "pt"))
   }
@@
 }
+
+keep <- pmap(slope_plot_df %>% filter(metric_name == "CWM_diet_planto"), plot_slopes)
+keep
 
@@
 
-cowplot::save_plot(here::here("figures/slope_plot_grid.png"), slope_plot_grid, ncol = 7, nrow = 5, bg = "white", base_asp = 1.18)
+cowplot::save_plot(here::here("figures/slope_plot_grid.png"), slope_plot_grid, ncol = 7, nrow = 5, bg = "white", base_asp = .9)
 
@@
            !term %in% c("Global", "Polar/Temperate")) %>% 
+  #get the general model estimates for foraging CWM's, since they contain only bird communities
+  bind_rows(model_output %>% filter(model_id == "base", term == "year_scaled", is.na(study_id),
+                        metric %in% c("CWM_forstrat_ground", "CWM_forstrat_watbelowsurf", "CWM_forstrat_wataroundsurf", 
+                                                                                 "CWM_forstrat_understory", "CWM_forstrat_aerial", 
+                                                                                 "CWM_forstrat_canopy", "CWM_forstrat_midhigh")) %>%
+              mutate(model_id = "taxa", term = "Birds")) %>%
   select(model_id, term, estimate, p.adjust, metric) %>%

### 2024-11-21 analysis/exporatory_scripts/mantel_test.Rmd
First commit

---
title: "Mantel Test"
output: html_document
---

```{r}
library(dplyr)
```

get rarefied samples
```{r}
#helper function to load file
loadRData <- function(file_name){
  load(file_name)
  get(ls()[ls() != "file_name"])
}

load(here::here("data/model_data.rda"))
long_duration <- model_data %>% select(rarefyID, duration) %>%
  filter(duration > 2) %>%
  pull(rarefyID)

#get list of rarefaction sample files
path <- here::here("data", "rarefied_samples")
files <- dir(path, "*.rda") %>%
  paste0(path, "/", .)

file_df <- data.frame(file_path = files) %>%
  mutate(rarefyID = str_match(files,"_cell\\s*(.*?)\\s*_sample")[,2]) %>%
  filter(rarefyID %in% long_duration)

get_dist <- function(files){

    rare_comm <- loadRData(files[1])

    # meta <- rare_comm %>%
    #   select(rarefyID, cell, rarefy_resamp, type) %>%
    #   distinct()

    rare_comm <- rare_comm %>%
      column_to_rownames("YEAR") %>%
      dplyr::select(-rarefyID, -cell, -rarefy_resamp, -type)

    mat <- vegdist(rare_comm, method='jaccard', binary=TRUE)

    return(mat)
  }

rarefyid_distmat <- unique(file_df$rarefyID) %>%
  purrr::set_names() %>% 
  purrr::map(function(id){

  #browser()

  files <- file_df %>% filter(rarefyID == id) %>%
    pull(file_path)

  if (length(files) > 1) {

  mat_list <- purrr::map(files, get_dist)
  dist_mat <- Reduce("+", mat_list) / length(mat_list)

  } else {

    dist_mat <- get_dist(files)

  }
  return(dist_mat)
})

dim_check <- purrr::map(rarefyid_distmat, ~ dim(.x)[1] != 2)
```

```{r}
# remove time series with only two time points, won't be able to permute
sub_rarefyid_distmat <- rarefyid_distmat[unlist(dim_check)]

get_mantel <- function(jacc_dist_mat){
  years <- as.integer(names(jacc_dist_mat))
  names(years) <- names(jacc_dist_mat)
  year_dist <- dist(years)

  mantel_out <- mantel(jacc_dist_mat, year_dist, method = "spearman", permutations = 9999)

  return(data.frame(p.value = mantel_out$signif, perm_num = mantel_out$permutations))
}

 mantel_list <- #names(sub_rarefyid_distmat) %>%
#   purrr::set_names() %>% 
  purrr::map(sub_rarefyid_distmat, get_mantel)

 names(mantel_list) <- names(sub_rarefyid_distmat)

 mantel_df <- bind_rows(mantel_list, .id = "rarefyID") %>%
   mutate(p.adjust =  p.adjust(mantel_df$p.value, method = "BH"))

 usethis::use_data(mantel_df)
```

### 2024-12-03 analysis/05_model_metrics.Rmd
Add jaccard hind cast model and figures. Move mantel test stuff.

--- 2024-05-16 analysis/05_model_metrics.Rmd
@@
 model_data <- metrics %>% 
-  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "Jaccard_next",
+  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "Jaccard_hind",
                        "FRic", "FEve", "FDiv", "S","CWM_bodymass_value","CWM_diet_inv", 
@@
 mixed_indv_mod_results <- pmap_dfr(mixed_indv_model_df %>% 
-                                     filter(!metric_name %in% c("Jaccard_base", "Jaccard_next")), 
+                                     filter(!metric_name %in% c("Jaccard_base", "Jaccard_hind")), 
                                    get_model, data = model_data)
@@
 base <- metric_model_table %>%
-  filter(model_id == "base", term == "year_scaled")
+  filter(model_id == "base", term == "year_scaled",
+         metric != "Jaccard_hind")
 

### 2024-12-03 analysis/figures/visualization.Rmd
Add jaccard hind cast model and figures. Move mantel test stuff.

@@
 
-plot_metric <- function(metric_name, ylabel, yval, xlabel, data, model_coef, color_list){
+plot_metric <- function(metric_name, ylabel, yval, xlabel, data, model_coef, color_list, legend){
 
@@
     #theme(legend.position = "none") +
-    theme(axis.title = element_text(size=7), legend.position = "none") +
     ylab(ylabel) +
@@
     scale_colour_manual(values = color_list, drop = FALSE)
+
+  if (legend == TRUE){
+    trend_plot <- trend_plot + theme(axis.title = element_text(size=7)) 
+  } else{
+    trend_plot <- trend_plot + theme(axis.title = element_text(size=7), legend.position = "none")
+  }
 
@@
 ggsave2(here::here("figures", "3met_long.jpeg"), bg = "white")
+```
+
+Get hindcast Jaccard plot for supplement
+```{r}
+jaccard_hind_plot <- plot_metric("Jaccard_hind", ylabel = "log(Jaccard + 1)", yval = "logvalue", xlabel = "year", data = model_data %>% mutate("log(value)" = log(value)), 
+            model_coef = metric_model_table %>% filter(model_id == "base"), 
+            color_list = group_colors,
+            legend = TRUE)
+
+jaccard_hind_plot + draw_grob(legend,  2/3.3, -0.3,.3/3.3, 1, scale = 0.01)
+
+ggsave(here::here("figures", "jaccard_hind_plot.jpeg"), bg = "white", width = 6, height = 4)
 ```

### 2024-12-10 analysis/figures/visualization.Rmd
Get rid of the scaling, annotate with new x axes, make it look pretty

@@
   filter(term == "year_scaled", model_id == "base",
-         !metric %in% c("FDiv", "FRic", "FEve", "Jaccard_next")) %>% 
+         !metric %in% c("FDiv", "FRic", "FEve", "Jaccard_next", "Jaccard_hind")) %>% 
   select(metric, estimate, p.adjust, study_id) %>%
@@
 
+  # plotting parameters
+  lim_buffer <- plot_data %>% filter(!is.na(estimate)) %>% pull(estimate) %>% sd()/4
+  lim <- max(abs(plot_data$estimate))
+
   gen_est <- plot_data %>% filter(is.na(study_id))
@@
     ggplot() +
-    geom_dotplot(aes(x = estimate_scale, fill = duration_fac),
+    geom_dotplot(aes(x = estimate, fill = duration_fac),
                  stackgroups = TRUE,
-                 method = "histodot",
-                 binwidth = 0.1, dotsize = 1.4
+                 method = "histodot"#,
+                 #binwidth = 0.1, dotsize = 1.4
     ) +
     scale_fill_manual(values = graid_pal, guide = "none", drop = FALSE) +
-    geom_vline(aes(xintercept = gen_est$estimate_scale), linetype = "longdash") +
+    geom_vline(aes(xintercept = gen_est$estimate), linetype = "longdash") +
     #coord_fixed(ratio=5)+#, ylim = c(0,5)) +
@@
           axis.title.y = element_blank(), 
-          axis.text.y=element_blank(), 
-          axis.title=element_text(size=14)) +
-    xlim(-3.5, 3.5) +
+          axis.text.y=element_blank()#, 
+          #axis.title=element_text(size=14)
+          ) +
+    xlim(-lim - lim_buffer, 
+         lim + lim_buffer) +
     xlab("temporal slope estimate") +
-    annotate('text', x = -2.5, y = 0.8, label = plot_label, size = unit(6, "pt"))
+    geom_hline(aes(yintercept = 0)) +
+    annotate('text', x = (-lim + (2*lim_buffer)), y = 0.8, label = plot_label, size = unit(4, "pt"))
 
   if(isFALSE(keep_x_axis)) {
-    metric_plot <- metric_plot + theme(axis.text.x = element_blank(), axis.title.x = element_blank())
+    #metric_plot <- metric_plot + theme(axis.text.x = element_blank(), axis.title.x = element_blank())
   } else{
-    metric_plot <- metric_plot + xlim(-3.5, 3.5) +
-      geom_hline(aes(yintercept = 0))
+    #metric_plot <- metric_plot + xlim(-3.5, 3.5) +
+      #geom_hline(aes(yintercept = 0))
   }
@@
   if(gen_est$p.adjust < 0.05){
-    metric_plot <- metric_plot + annotate('text', x = (gen_est$estimate_scale + .4), y = 0.8, label = "*", size = unit(16, "pt"))
+    metric_plot <- metric_plot + annotate('text', x = (gen_est$estimate + (lim_buffer*2)), y = 0.8, label = "*", size = unit(16, "pt"))
   }
@@
                      "FRic SES", "FEve SES", "FDiv SES", 
-                     "Invertebrates", "Scavenge", "Fish", "Other\nVertebrates", "Other Plants",
+                     "Invertebrates", "Scavenge", "Fish", "Other\nVertebrates", "Other\nPlants",
                      "Seeds", "Mammals\n& Birds", "Nectar", "Fruit", "Reptiles &\nAmphibians", 
-                     "Ground", "Below Water\nSurface", "Around Water\nSurface", 
+                     "Ground", "Below Water\nSurface", "Around\nWater\nSurface", 
                      "Understory", "Aerial", "Canopy", "Midhigh", 
@@
 
-cowplot::save_plot(here::here("figures/slope_plot_grid.png"), slope_plot_grid, ncol = 7, nrow = 5, bg = "white", base_asp = .9)
+cowplot::save_plot(here::here("figures/slope_plot_grid_test.png"), slope_plot_grid, ncol = 7, nrow = 5, bg = "white", base_asp = .7)
 

### 2024-12-13 figures/export_tables.Rmd
export tables for publication formatting using flextable

---
title: "Editable tables"
output: html_document
date: "2024-12-13"
---

```{r}
library(kableExtra)
library(magrittr)
library(tidyverse)
#library(printr)
library(rticles)
library(knitr)
library(flextable)

```

```{r}
traitTable <- tibble(Trait = c(rep("Diet", 10), rep("Foraging Strata", 7), rep("Pelagic Specialist", 2), rep("Nocturnal", 2), rep("Crepuscular", 2), rep("Diurnal", 2), "Body Mass"),
       Category = c("Invertebrate", "Mammals and Birds", "Reptiles", "Fish", "Unknown Vertebrates", "Scavenging", "Fruit", "Nectar", "Seeds", "Other Plant",
                    "Below water surface", "water surface", "ground", "understory", "> 2m, below canopy", "canopy", "aerial",
                    rep(c("yes", "no"), 4),
                    "-"),
       Taxa = c(rep("Bird and Mammal", 10), rep("Bird", 7), rep("Bird", 2), rep("Bird and Mammal", 2), rep("Mammal", 4), "Bird and Mammal"),
       `Data Type` = c(rep("percentage consumed", 10), rep("percentage of use", 7), rep("binary", 8), "continous, in grams")) %>%
  flextable()
  # kable(format = "latex", caption = "Description of the traits included in the analysis broken down by categories at data type.") %>%
  # collapse_rows(columns = 1:4) %>%
  # column_spec(3:4, width = "3cm") %>%
  # kable_paper() #%>%
  # #save_kable(here::here("figures/tables/trait_table.jpeg"))

traitTable <- traitTable %>% merge_v() %>% border_inner_h() %>% border_inner_v()

```

```{r}
load(here::here("data/model_output.rda"))

change_table <- model_output %>%
  filter(term == "year_scaled", p.adjust < 0.05) %>%
  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S", "FRic", "FEve", "FDiv")) %>%
  mutate(positive = ifelse(estimate > 0, "+", "-")) %>%
  group_by(metric, positive) %>%
  summarise(n_study_id = n_distinct(study_id)) %>%
  pivot_wider(names_from = "metric", values_from = "n_study_id", values_fill = 0) %>%
  select(positive, S, Jaccard_base, FRic, FEve, FDiv, SES_FRic, SES_FEve, SES_FDiv) %>%
  flextable() %>% 
  set_header_labels(positive = "", S = "S", Jaccard_base = "Jaccard Similarity", FRic = "FRic", 
                    FEve = "FEve", FDiv = "FDiv", SES_FRic = "SES FRic", SES_FEve = "SES FEve", SES_FDiv = "SES FDiv") %>%
  add_header_row(
  values = c("", "species", "functional"),
  colwidths = c(1, 2, 6))
  # kable(format = "latex", 
  #       #col.names = c("", "S", "Jaccard Similarity", "FRic", "FEve", "FDiv", "SES FRic", "SES FEve", "SES FDiv"),
  #       col.names = c("", "richness", "Jaccard Similarity", "richness", "evenness", "divergence", "SES richness", "SES evenness", "SES divergence"),
  #       caption = "Number of studies that experienced a significant trend in each calculated metric out of 50 total studies. Both observed and corrected (SES) versions of functional metrics are included.", 
  #       booktabs = TRUE, align = "c", escape = FALSE) %>%
  # kableExtra::add_header_above(c("", "species" = 2, "functional" = 6)) %>%
  # kable_styling("striped") #%>%
  # #save_kable(here::here("figures/tables/trend_table.jpeg"))

change_table

```

```{r}
load(here::here("data/change_group_ses_desc.rda"))

change_group_ses_desc[change_group_ses_desc == "pos"] <- "+"
change_group_ses_desc[change_group_ses_desc == "neg"] <- "-"

change_groups_table <- change_group_ses_desc %>%
  select(type = change_description, everything()) %>%
  mutate(n = as.character(n),
           n = ifelse(n == "7", "32", n)) %>%
  mutate(type = ifelse(n == "32", "no change", type),
         type = ifelse(type == "functional change", "functional diversity change", type)) %>%
  rename(`Study Count` = `n`) %>%
  flextable() %>%
  merge_v(j = "type")  %>%
  border_inner_h() %>%
  border_inner_v() %>%
  set_header_labels(type = "type", S = "S", Jaccard_base = "Jaccard Similarity", FRic = "FRic", 
                    FEve = "FEve", FDiv = "FDiv", SES_FRic = "SES FRic", SES_FEve = "SES FEve", SES_FDiv = "SES FDiv")

  #    kable(format = "latex",
  #       caption = "Table giving the count of number of studies for different kinds of functional change, broken down into no change functional change without species richness change, loss of redundancy, and increases in species richness. Trends direction is indicated by a positive or negative sign, with a 0 indicating no trend.", align = "c",
  #       latex_options= c("scale_down", "hold_position")) %>% 
  # collapse_rows(columns = c(1,2)) %>%
  # column_spec(1, width = "4cm") #%>%
  #save_kable(here::here("figures/tables/study_change_table.jpeg"))

change_groups_table

```

```{r}
save_as_docx(
  "Table 1" = traitTable, "Table 2" = change_table, "Table 3" = change_groups_table,
  path = here::here("figures/table_doc.docx"))
```

### 2024-12-19 analysis/figures/visualization.Rmd
Save high quality figures.

@@
 
-ggsave(here::here("figures/study_map.jpeg"), bg = "white")
+ggsave(here::here("figures/study_map.jpeg"), bg = "white", dpi = 800)
 ```
@@
 
-ggsave2(here::here("figures/study_map_hist.jpeg"), bg = "white")
+ggsave2(here::here("figures/study_map_hist.jpeg"), bg = "white", dpi = 1000, width = 180, units = "mm")
 ```
@@
                 model_coef = metric_model_table %>% filter(model_id == "base"), 
-                color_list = group_colors)
+                color_list = group_colors, 
+                legend = FALSE)
 
@@
 
-ggsave2(here::here("figures", "3met_long.jpeg"), bg = "white")
+ggsave2(here::here("figures", "3met_long.jpeg"), bg = "white", width = 180, unit = "mm", dpi = 1000)
 ```
@@
 
-keep <- pmap(slope_plot_df %>% filter(metric_name == "CWM_diet_planto"), plot_slopes)
-keep
-
 muted <- khroma::color("muted")(5)
@@
 
-cowplot::save_plot(here::here("figures/slope_plot_grid_test.png"), slope_plot_grid, ncol = 7, nrow = 5, bg = "white", base_asp = .7)
+cowplot::save_plot(here::here("figures/slope_plot_grid_test.png"), slope_plot_grid, ncol = 7, nrow = 5, bg = "white", base_asp = .7, dpi = 1000)
 
@@
   duration_legend <- get_legend(duration_legend_plot)
-  leg <- grid.arrange(duration_legend)
+  leg <- gridExtra::grid.arrange(duration_legend)
   ggsave(here::here(paste0("figures/slope_legends/", color, "_legend.png")), leg, width = .6, height = 1.5,
-         bg = "transparent")
+         bg = "transparent", dpi = 800)
 }
@@
 contrasts_plot_grid <- climate_plot + taxa_plot + pa_plot + realm_plot + plot_layout(ncol = 4, axis_titles = "collect")
-ggsave(here::here("figures/contrasts_plot_grid.png"), contrasts_plot_grid, width = 15, heigh = 6)
+ggsave(here::here("figures/contrasts_plot_grid.jpg"), contrasts_plot_grid, width = 15, heigh = 6, dpi = 1000)
 ```
